### 2021-10-04 Multilevel Missing Data/Multilevel Missing Data.Rmd
Add markdown in JSS style for tutorial paper

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    # use this syntax to add text on several lines
    address: |
      | First line
      | Second line
    email: \email{name@company.com}
    url: http://rstudio.com
  - name: Gerko Vink
    affiliation: 'Utrecht University'
    # To add another line, use \AND at the end of the previous one as above
  - name: Thomas Debray
    # address: |
    #   | Department of Statistics and Mathematics,
    #   | Faculty of Biosciences,
    #   | Universitat Autònoma de Barcelona
    affiliation: |
      | University Medical Centre Utrecht
    # use a different affiliation in adress field (differently formated here)
    affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  # For running headers, if needed
  short:     "\\pkg{mice}: Multilevel"
abstract: >
  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

# Introduction

This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.

## Code formatting

In general, don't use Markdown, but use the more precise LaTeX commands instead:

* \proglang{Java}
* \pkg{plyr}

One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).

If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.

# \proglang{R} code {short-title="R code" #r-code}

Can be inserted in regular R markdown blocks.

```{r}
x <- 1:10
x
```

## Features specific to \pkg{rticles} {short-title="Features specific to rticles"}

* Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
* Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format.

### 2021-10-04 Multilevel Missing Data/Multilevel-Missing-Data.tex
Add markdown in JSS style for tutorial paper

\documentclass[
]{jss}

\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Hanne Oberman\\Utrecht University \And Gerko Vink\\Utrecht
University \And Thomas Debray\\University Medical Centre Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{\pkg{mice}: Multilevel}

\Abstract{
Tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
Including methods for ignorable and non-ignorable missingness.
}

\Keywords{missing data, multilevel, clustering, \proglang{R}}
\Plainkeywords{missing data, multilevel, clustering, R}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Hanne Oberman\\
    Utrecht University\\
    First line\\
Second line\\
  E-mail: \email{name@company.com}\\
  URL: \url{http://rstudio.com}\\~\\
      }

% Pandoc syntax highlighting

% Pandoc citation processing

\usepackage{amsmath}

\begin{document}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

This template demonstrates some of the basic LaTeX that you need to know
to create a JSS article.

\hypertarget{code-formatting}{%
\subsection{Code formatting}\label{code-formatting}}

In general, don't use Markdown, but use the more precise LaTeX commands
instead:

\begin{itemize}
\item
  \proglang{Java}
\item
  \pkg{plyr}
\end{itemize}

One exception is inline code, which can be written inside a pair of
backticks (i.e., using the Markdown syntax).

If you want to use LaTeX commands in headers, you need to provide a
\texttt{short-title} attribute. You can also provide a custom identifier
if necessary. See the header of Section \ref{r-code} for example.

\section[R code]{\proglang{R} code}\label{r-code}

Can be inserted in regular R markdown blocks.

\begin{CodeChunk}
\begin{CodeInput}
R> x <- 1:10
R> x
\end{CodeInput}
\begin{CodeOutput}
 [1]  1  2  3  4  5  6  7  8  9 10
\end{CodeOutput}
\end{CodeChunk}

\subsection[Features specific to rticles]{Features specific to
\pkg{rticles}}\label{features-specific-to}

\begin{itemize}
\tightlist
\item
  Adding short titles to section headers is a feature specific to
  \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is
  currently not supported by Pandoc and we will update this template if
  \href{https://github.com/jgm/pandoc/issues/4409}{it is officially
  supported in the future}.
\item
  Using the \texttt{\textbackslash{}AND} syntax in the \texttt{author}
  field to add authors on a new line. This is a specific to the
  \texttt{rticles::jss\_article} format.
\end{itemize}

\end{document}

### 2021-10-06 Multilevel Missing Data/Multilevel-Missing-Data.tex
Fill in basic info in Jstat template

@@
 Tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
-Including methods for ignorable and non-ignorable missingness.
+Including methods for ignorable and non-ignorable missingness. AKA
+missing on so many levels.
 }
@@
 Second line\\
-  E-mail: \email{name@company.com}\\
-  URL: \url{http://rstudio.com}\\~\\
+  E-mail: \email{h.i.oberman@uu.nl}\\
+  URL: \url{https://hanneoberman.github.io/}\\~\\
       }
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-This template demonstrates some of the basic LaTeX that you need to know
-to create a JSS article.
-
-\hypertarget{code-formatting}{%
-\subsection{Code formatting}\label{code-formatting}}
-
-In general, don't use Markdown, but use the more precise LaTeX commands
-instead:
+\hypertarget{outline-door-thomas}{%
+\section{Outline door Thomas}\label{outline-door-thomas}}
 
@@
 \item
-  \proglang{Java}
+  Waarom zijn standaard (ad hoc) methoden niet geschikt?
 \item
-  \pkg{plyr}
+  Wat is heterogeniteit? (i.e.~variability binnen studies
+  vs.~variability tussen studies). Random effects voor intercept term,
+  predictor effects, en/of variance residual error
+\item
+  JOMO aanpak (misschien buiten beschouwing laten?)
+\item
+  MICE aanpak: one-stage imputation versus two-stage imputation methods
+\item
+  Imputatie van studie-level versus patient-level covariates
+\item
+  Omgaan met MNAR
 \end{itemize}
 
-One exception is inline code, which can be written inside a pair of
-backticks (i.e., using the Markdown syntax).
-
-If you want to use LaTeX commands in headers, you need to provide a
-\texttt{short-title} attribute. You can also provide a custom identifier
-if necessary. See the header of Section \ref{r-code} for example.
-
-\section[R code]{\proglang{R} code}\label{r-code}
-
-Can be inserted in regular R markdown blocks.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> x <- 1:10
-R> x
-\end{CodeInput}
-\begin{CodeOutput}
- [1]  1  2  3  4  5  6  7  8  9 10
-\end{CodeOutput}
-\end{CodeChunk}
-
-\subsection[Features specific to rticles]{Features specific to
-\pkg{rticles}}\label{features-specific-to}
+\hypertarget{references}{%
+\section{References}\label{references}}
 
 \begin{itemize}
-\tightlist
 \item
-  Adding short titles to section headers is a feature specific to
-  \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is
-  currently not supported by Pandoc and we will update this template if
-  \href{https://github.com/jgm/pandoc/issues/4409}{it is officially
-  supported in the future}.
+  \href{Chapter:\%20Dealing\%20with\%20Missing\%20Data\%20in\%20an\%20IPD\%20Meta-analysis}{https://onlinelibrary.wiley.com/doi/abs/10.1002/9781119333784.ch18}
 \item
-  Using the \texttt{\textbackslash{}AND} syntax in the \texttt{author}
-  field to add authors on a new line. This is a specific to the
-  \texttt{rticles::jss\_article} format.
+  \href{Vignette:\%20gerkovink.com/miceVignettes/Multi_level}{https://www.gerkovink.com/miceVignettes/Multi\_level/Multi\_level\_data.html}
 \end{itemize}

### 2021-10-06 Multilevel Missing Data/Multilevel Missing Data.Rmd

@@
       | Second line
-    email: \email{name@company.com}
-    url: http://rstudio.com
+    email: \email{h.i.oberman@uu.nl}
+    url: https://hanneoberman.github.io/
   - name: Gerko Vink
@@
 abstract: >
-  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness.
+  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. AKA missing on so many levels.
 keywords:
@@
 
+# Outline door Thomas
+
+-	Waarom zijn standaard (ad hoc) methoden niet geschikt?
+
+-	Wat is heterogeniteit? (i.e. variability binnen studies vs. variability tussen studies). Random effects voor intercept term, predictor effects, en/of variance residual error
+
+-	JOMO aanpak (misschien buiten beschouwing laten?)
+
+-	MICE aanpak: one-stage imputation versus two-stage imputation methods
+
+-	Imputatie van studie-level versus patient-level covariates
+
+-	Omgaan met MNAR
+
+# References
+
+- [https://onlinelibrary.wiley.com/doi/abs/10.1002/9781119333784.ch18](Chapter: Dealing with Missing Data in an IPD Meta-analysis)
+
+- [https://www.gerkovink.com/miceVignettes/Multi_level/Multi_level_data.html](Vignette: gerkovink.com/miceVignettes/Multi_level)
+
 # Introduction
-
-This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.
-
-## Code formatting
-
-In general, don't use Markdown, but use the more precise LaTeX commands instead:
-
-* \proglang{Java}
-* \pkg{plyr}
-
-One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).
-
-If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.
-
-# \proglang{R} code {short-title="R code" #r-code}
-
-Can be inserted in regular R markdown blocks.
-
-```{r}
-x <- 1:10
-x
-```
-
-## Features specific to \pkg{rticles} {short-title="Features specific to rticles"}
-
-* Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter).
-This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
-* Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format.

### 2021-10-11 Multilevel Missing Data/Multilevel Missing Data.Rmd
Translate outline

@@
 
--	Waarom zijn standaard (ad hoc) methoden niet geschikt?
+-	Why are standard (ad hoc) methods not well suited?
 
--	Wat is heterogeniteit? (i.e. variability binnen studies vs. variability tussen studies). Random effects voor intercept term, predictor effects, en/of variance residual error
+-	What is heterogeniety? (i.e. variability within studies vs. variability between studies). Random effects for intercept term, predictor effects, and/or variance residual error
 
--	JOMO aanpak (misschien buiten beschouwing laten?)
+- Systematic vs sporadic missingness
 
--	MICE aanpak: one-stage imputation versus two-stage imputation methods
+-	MICE approach: one-stage imputation versus two-stage imputation methods
 
--	Imputatie van studie-level versus patient-level covariates
+-	Imputation of study level versus patient-level covariates
 
--	Omgaan met MNAR
+-	Handling MNAR
+
+- Case study: metamisc::impact()
+
+# Think about adding
+
+- JOMO in mice
 

### 2021-10-11 Multilevel Missing Data/Multilevel-Missing-Data.tex

@@
 \item
-  Waarom zijn standaard (ad hoc) methoden niet geschikt?
+  Why are standard (ad hoc) methods not well suited?
 \item
-  Wat is heterogeniteit? (i.e.~variability binnen studies
-  vs.~variability tussen studies). Random effects voor intercept term,
-  predictor effects, en/of variance residual error
+  What is heterogeniety? (i.e.~variability within studies
+  vs.~variability between studies). Random effects for intercept term,
+  predictor effects, and/or variance residual error
 \item
-  JOMO aanpak (misschien buiten beschouwing laten?)
+  Systematic vs sporadic missingness
 \item
-  MICE aanpak: one-stage imputation versus two-stage imputation methods
+  MICE approach: one-stage imputation versus two-stage imputation
+  methods
 \item
-  Imputatie van studie-level versus patient-level covariates
+  Imputation of study level versus patient-level covariates
 \item
-  Omgaan met MNAR
+  Handling MNAR
+\item
+  Case study: metamisc::impact()
+\end{itemize}
+
+\hypertarget{think-about-adding}{%
+\section{Think about adding}\label{think-about-adding}}
+
+\begin{itemize}
+\tightlist
+\item
+  JOMO in mice
 \end{itemize}

### 2021-10-18 Multilevel Missing Data/Multilevel Missing Data.Rmd
Add sub-sections and references to paper outline

@@
     affiliation: Utrecht University
-    # use this syntax to add text on several lines
     address: |
-      | First line
-      | Second line
+      | Padualaan 14
+      | 3584 CH Utrecht
     email: \email{h.i.oberman@uu.nl}
     url: https://hanneoberman.github.io/
+  - name: Johanna Munoz Avila
+    affiliation: University Medical Center Utrecht
+  - name: Valentijn
+    affiliation: University Medical Center Utrecht \AND
   - name: Gerko Vink
-    affiliation: 'Utrecht University'
-    # To add another line, use \AND at the end of the previous one as above
+    affiliation: Utrecht University
   - name: Thomas Debray
-    # address: |
-    #   | Department of Statistics and Mathematics,
-    #   | Faculty of Biosciences,
-    #   | Universitat Autònoma de Barcelona
-    affiliation: |
-      | University Medical Centre Utrecht
-    # use a different affiliation in adress field (differently formated here)
-    affiliation2: Universitat Autònoma de Barcelona
+    affiliation: University Medical Center Utrecht 
+      # | Julius Center for Health Sciences and Primary Care, 
+      # | University Medical Center Utrecht, Utrecht, The Netherlands
+    #affiliation2:
 title:
@@
 output: rticles::jss_article
+bibliography: ../References/multilevelmice.bib
 ---
@@
 options(prompt = 'R> ', continue = '+ ')
+library(tidyverse)
+
 ```
 
-# Outline door Thomas
+# Introduction
 
--	Why are standard (ad hoc) methods not well suited?
+## Multilevel data
 
--	What is heterogeniety? (i.e. variability within studies vs. variability between studies). Random effects for intercept term, predictor effects, and/or variance residual error
+- What is clustering/multilevel data?
+In this paper, we discuss grouped observations, not longitudinal data (within-patient clustering).
+What do we mean by clustering?
+In the medical field: Clustering by studies (IPDMA), hospitals in registries, multi-center studies etc.
+In other fields: e.g.
+official stats clustering at country-level, or social sciences clustering at school-level (related to the sampling design).
 
-- Systematic vs sporadic missingness
+- What is heterogeniety (i.e.
+variability within studies vs.
+variability between studies)?
+ADD: Figure to show difference between patient-level datapoints vs cluster-level datapoints.
+And different data frame formats.
 
--	MICE approach: one-stage imputation versus two-stage imputation methods
+- What methods are required to analyze multilevel data?	Random effects for intercept term, predictor effects, and/or variance residual error. Add references, e.g. @hox17.
 
--	Imputation of study level versus patient-level covariates
+## Missing data
 
--	Handling MNAR
+- Why/where does missingness occur? I.e., not only patient-level but also cluster-level.
 
-- Case study: metamisc::impact()
+- How can we categorize this?
+Systematic vs sporadic missingness, see @resc13.
+ADD: visualization of systematic vs sporadic missingness.
+Within systematic we have always missing (same value per cluster) and non-measured variables (may differ per patient).
+TODO: adjust md pattern to match text.
+
+```{r patterns, echo=FALSE}
+# monotone vs non-monotone
+patterns <- rbind(
+  tibble(pat = rep("Monotone", 5), rownr = 5:1, X = rep(1, 5), Y = c(1,1,1,0, 0), Z = c(1, rep(0,4))) %>% pivot_longer(cols = c(X, Y, Z)),
+  tibble(pat = rep("Non-monotone", 5), rownr = 5:1, X = c(0, rep(1,4)), Y = c(1,1,1,0, 0), Z = c(1, rep(0,3), 1)) %>% pivot_longer(cols = c(X, Y, Z))) %>%
+  mutate(across(everything(), as.factor))
+ggplot(patterns, aes(x = name, y = rownr, fill = value, type = pat)) +
+  geom_tile(color = "black") +
+  scale_fill_brewer(palette = "Set1") +
+  facet_wrap(~pat) +
+  theme_void() +
+  theme(legend.position = "none") 
+```
+
+-	Why are standard (ad hoc) missing data methods not well suited?
+
+- What other methods are available?
+General overview of approaches, see @audi18 @grun18.
+E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.
+
+-	Additional problem addressed in this paper: handling MNAR.
+
+## Aim of this paper
+
+- Provide practical guidelines with code snippets for imputation of incomplete multilevel data.
+
+- Case study options: metamisc::impact (IPD), GREAT (IPD).
+
+# Workflows
+
+## Modeling choices
+
+- ADD: Section with more in-depth info on different data types and suitable methods. Add some equations, but not too much -> grow in complexity.
+
+- Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeneiality) --> explain predictor matrix.
+Just mention that we should think about random intercept/slope, but just refer to other papers.
+This paper is just a software tutorial.
+--> Typically, at least random intercepts, but often random slopes as well.
+Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there.
+Sometimes there's little reason to assume some variable is affected by heterogeniety.
+We won't focus on congeneiality, that's too complex for 2-3 papers.
+
+## Conditional models
+
+- How to define the imputation model(s) in mice?
+
+- What do the different implementations look like?
+
+## Pooling
+
+- Pooling 'regular' parameters vs more 'exotic' parameters (SE of residual errors, or autocorreletion)
+
+- ADD: export mids objects to other packages like lme4 or coxme(?)
 
@@
 
-- JOMO in mice
+- JOMO in mice --> on the side for now
+
+- Additional levels of clustering
 
 # References
-
-- [https://onlinelibrary.wiley.com/doi/abs/10.1002/9781119333784.ch18](Chapter: Dealing with Missing Data in an IPD Meta-analysis)
-
-- [https://www.gerkovink.com/miceVignettes/Multi_level/Multi_level_data.html](Vignette: gerkovink.com/miceVignettes/Multi_level)
-
-# Introduction

### 2021-10-18 Multilevel Missing Data/Multilevel-Missing-Data.tex
Add sub-sections and references to paper outline

@@
 \author{
-Hanne Oberman\\Utrecht University \And Gerko Vink\\Utrecht
-University \And Thomas Debray\\University Medical Centre Utrecht
+Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
+Medical Center Utrecht \And Valentijn\\University Medical Center Utrecht
+\AND Gerko Vink\\Utrecht University \And Thomas Debray\\University
+Medical Center Utrecht
 }
@@
 
-\Plainauthor{Hanne Oberman, Gerko Vink, Thomas Debray}
+\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn, Gerko
+Vink, Thomas Debray}
 \Plaintitle{Imputation of Incomplete Multilevel Data with mice}
@@
     Utrecht University\\
-    First line\\
-Second line\\
+    Padualaan 14\\
+3584 CH Utrecht\\
   E-mail: \email{h.i.oberman@uu.nl}\\
@@
 
-\hypertarget{outline-door-thomas}{%
-\section{Outline door Thomas}\label{outline-door-thomas}}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+\hypertarget{multilevel-data}{%
+\subsection{Multilevel data}\label{multilevel-data}}
 
@@
 \item
-  Why are standard (ad hoc) methods not well suited?
+  What is clustering/multilevel data? In this paper, we discuss grouped
+  observations, not longitudinal data (within-patient clustering). What
+  do we mean by clustering? In the medical field: Clustering by studies
+  (IPDMA), hospitals in registries, multi-center studies etc. In other
+  fields: e.g.~official stats clustering at country-level, or social
+  sciences clustering at school-level (related to the sampling design).
 \item
-  What is heterogeniety? (i.e.~variability within studies
-  vs.~variability between studies). Random effects for intercept term,
-  predictor effects, and/or variance residual error
+  What is heterogeniety (i.e.~variability within studies vs.~variability
+  between studies)? ADD: Figure to show difference between patient-level
+  datapoints vs cluster-level datapoints. And different data frame
+  formats.
 \item
-  Systematic vs sporadic missingness
+  What methods are required to analyze multilevel data? Random effects
+  for intercept term, predictor effects, and/or variance residual error.
+  Add references, e.g. \citet{hox17}.
+\end{itemize}
+
+\hypertarget{missing-data}{%
+\subsection{Missing data}\label{missing-data}}
+
+\begin{itemize}
 \item
-  MICE approach: one-stage imputation versus two-stage imputation
-  methods
+  Why/where does missingness occur? I.e., not only patient-level but
+  also cluster-level.
 \item
-  Imputation of study level versus patient-level covariates
+  How can we categorize this? Systematic vs sporadic missingness, see
+  \citet{resc13}. ADD: visualization of systematic vs sporadic
+  missingness. Within systematic we have always missing (same value per
+  cluster) and non-measured variables (may differ per patient). TODO:
+  adjust md pattern to match text.
+\end{itemize}
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Multilevel-Missing-Data_files/figure-latex/patterns-1} \end{center}
+
+\end{CodeChunk}
+
+\begin{itemize}
 \item
-  Handling MNAR
+  Why are standard (ad hoc) missing data methods not well suited?
 \item
-  Case study: metamisc::impact()
+  What other methods are available? General overview of approaches, see
+  \citet{audi18} \citet{grun18}. E.g., imputation of study level versus
+  patient-level covariates, and one-stage imputation versus two-stage
+  imputation methods.
+\item
+  Additional problem addressed in this paper: handling MNAR.
+\end{itemize}
+
+\hypertarget{aim-of-this-paper}{%
+\subsection{Aim of this paper}\label{aim-of-this-paper}}
+
+\begin{itemize}
+\item
+  Provide practical guidelines with code snippets for imputation of
+  incomplete multilevel data.
+\item
+  Case study options: metamisc::impact (IPD), GREAT (IPD).
+\end{itemize}
+
+\hypertarget{workflows}{%
+\section{Workflows}\label{workflows}}
+
+\hypertarget{modeling-choices}{%
+\subsection{Modeling choices}\label{modeling-choices}}
+
+\begin{itemize}
+\item
+  ADD: Section with more in-depth info on different data types and
+  suitable methods. Add some equations, but not too much -\textgreater{}
+  grow in complexity.
+\item
+  Ideally we impute with random everything and heteroscedastic errors:
+  most generic method (no worry about congeneiality) --\textgreater{}
+  explain predictor matrix. Just mention that we should think about
+  random intercept/slope, but just refer to other papers. This paper is
+  just a software tutorial. --\textgreater{} Typically, at least random
+  intercepts, but often random slopes as well. Refer to other papers for
+  background, we'll focus just on the software implementation of the
+  situations mentioned there. Sometimes there's little reason to assume
+  some variable is affected by heterogeniety. We won't focus on
+  congeneiality, that's too complex for 2-3 papers.
+\end{itemize}
+
+\hypertarget{conditional-models}{%
+\subsection{Conditional models}\label{conditional-models}}
+
+\begin{itemize}
+\item
+  How to define the imputation model(s) in mice?
+\item
+  What do the different implementations look like?
+\end{itemize}
+
+\hypertarget{pooling}{%
+\subsection{Pooling}\label{pooling}}
+
+\begin{itemize}
+\item
+  Pooling `regular' parameters vs more `exotic' parameters (SE of
+  residual errors, or autocorreletion)
+\item
+  ADD: export mids objects to other packages like lme4 or coxme(?)
 \end{itemize}
@@
 \begin{itemize}
-\tightlist
 \item
-  JOMO in mice
+  JOMO in mice --\textgreater{} on the side for now
+\item
+  Additional levels of clustering
 \end{itemize}
 
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begin{itemize}
-\item
-  \href{Chapter:\%20Dealing\%20with\%20Missing\%20Data\%20in\%20an\%20IPD\%20Meta-analysis}{https://onlinelibrary.wiley.com/doi/abs/10.1002/9781119333784.ch18}
-\item
-  \href{Vignette:\%20gerkovink.com/miceVignettes/Multi_level}{https://www.gerkovink.com/miceVignettes/Multi\_level/Multi\_level\_data.html}
-\end{itemize}
+\renewcommand\refname{References}
+\bibliography{../References/multilevelmice.bib}
 

### 2021-11-01 Multilevel Missing Data/Multilevel Missing Data.Rmd
Outline paper incl modeling steps

@@
   - name: Johanna Munoz Avila
+    affiliation: University Medical Center Utrecht \AND
+  - name: Valentijn de Jong
     affiliation: University Medical Center Utrecht 
-  - name: Valentijn
-    affiliation: University Medical Center Utrecht \AND
   - name: Gerko Vink
-    affiliation: Utrecht University
+    affiliation: Utrecht University \AND
   - name: Thomas Debray
@@
 abstract: >
-  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. AKA missing on so many levels.
+  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. 
 keywords:
   # at least one keyword must be supplied
-  formatted: [missing data, multilevel, clustering, "\\proglang{R}"]
-  plain:     [missing data, multilevel, clustering, R]
+  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
+  plain:     [missing data, multilevel, clustering, mice, R]
 preamble: >
@@
 
-- What is clustering/multilevel data?
-In this paper, we discuss grouped observations, not longitudinal data (within-patient clustering).
-What do we mean by clustering?
+- What is clustering/multilevel data? In this paper, we discuss grouped observations, not longitudinal data (within-patient clustering).  -> ADD: timeseries also in Discussion section.
+
+- What do we mean by clustering?
 In the medical field: Clustering by studies (IPDMA), hospitals in registries, multi-center studies etc.
@@
 
-- What is heterogeniety (i.e.
-variability within studies vs.
-variability between studies)?
-ADD: Figure to show difference between patient-level datapoints vs cluster-level datapoints.
-And different data frame formats.
+- What is heterogeniety? I.e. variability within studies vs. variability between studies
 
-- What methods are required to analyze multilevel data?	Random effects for intercept term, predictor effects, and/or variance residual error. Add references, e.g. @hox17.
+- What does multilevel data look like?
+ADD: figure to show difference between patient-level datapoints vs cluster-level datapoints.
+Maybe also add different data frame formats (or just explain in text that there's long and wide formats).
+
+- What methods are required to analyze multilevel data?	Add references, e.g. @hox17. At least explain difference random effects for intercept term, predictor effects, and/or variance residual error.
 
@@
 
-- Why/where does missingness occur? I.e., not only patient-level but also cluster-level.
+- Why/where does missingness occur in multilevel data? I.e., not only patient-level but also cluster-level.
 
@@
 TODO: adjust md pattern to match text.
+-> syst may vary or same for all patients (observations/participants).
 
@@
 patterns <- rbind(
-  tibble(pat = rep("Monotone", 5), rownr = 5:1, X = rep(1, 5), Y = c(1,1,1,0, 0), Z = c(1, rep(0,4))) %>% pivot_longer(cols = c(X, Y, Z)),
-  tibble(pat = rep("Non-monotone", 5), rownr = 5:1, X = c(0, rep(1,4)), Y = c(1,1,1,0, 0), Z = c(1, rep(0,3), 1)) %>% pivot_longer(cols = c(X, Y, Z))) %>%
+  tibble(pat = rep("Systematic", 5), rownr = 5:1, X = rep(1, 5), Y = c(1,1,1,0, 0), Z = c(1, rep(0,4))) %>% pivot_longer(cols = c(X, Y, Z)),
+  tibble(pat = rep("Sporadic", 5), rownr = 5:1, X = c(0, rep(1,4)), Y = c(1,1,1,0, 0), Z = c(1, rep(0,3), 1)) %>% pivot_longer(cols = c(X, Y, Z))) %>%
   mutate(across(everything(), as.factor))
@@
 
+-	ADD: missingness mech here
+
 - Why are standard (ad hoc) missing data methods not well suited?
 
-- What other methods are available?
+- What types of multilevel methods are available?
 General overview of approaches, see @audi18 @grun18.
 E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.
-
--	Additional problem addressed in this paper: handling MNAR.
 
@@
 
-- Case study options: metamisc::impact (IPD), GREAT (IPD).
+- We focus on the workflow for conditional modeling (not JOMO) in `mice`. Refer to other packages: `mitml`, `miceadds`.
+
+- Case study options: `metamisc::impact` (real data on traumatic brain injuries, IPD), `mice::popularity` (simulated data with MNAR/MAR mixture, schools). -> Check Gelman's data/NSRI data.
 
@@
 
+## Case study
+
+- Introduce case study and set scope of this tutorial: We're providing an overview of implementations.
+It's up-to the reader to decide which strategy suits their data.
+So we won't go into detail for the different methods (and equations).
+This paper is just a software tutorial.
+We'll keep it practical.
+-> ADD: some kind of help function that suggests a suitable pred matrix to the user, given a certain analysis model.
+
 ## Modeling choices
 
-- ADD: Section with more in-depth info on different data types and suitable methods. Add some equations, but not too much -> grow in complexity.
+- Which models will we discuss? We'll build the model to grow in complexity. The final model is the most complex but also the most versatile.
 
-- Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeneiality) --> explain predictor matrix.
-Just mention that we should think about random intercept/slope, but just refer to other papers.
-This paper is just a software tutorial.
---> Typically, at least random intercepts, but often random slopes as well.
-Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there.
-Sometimes there's little reason to assume some variable is affected by heterogeniety.
-We won't focus on congeneiality, that's too complex for 2-3 papers.
+- Note on model complexity: Typically, we should at least use random intercepts, but often random slopes as well.
+Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeniality, but don't mention the term) ->  Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there.
+Sometimes there's little reason to assume some variable is affected by heterogeneity.
+-> Refer to Meng, Vincent, and a paper by Grund on congeniality and random slopes.
+
+- Step 0: study as a predictor, AKA multilevel imputation for dummies. Doesn't work for syst missing.
 
@@
 
-- How to define the imputation model(s) in mice?
+- How to define the imputation model(s) in `mice`?
 
@@
 
+- Step 1: Intercept
+
+- Step 2: Slope
+
+- Step 3: Residuals
+
+- Heckman model for MNAR
+
 ## Pooling
+
+- Analysis of scientific interest.
+
+- Pooling using `mitml`.
 
@@
 
-- ADD: export mids objects to other packages like lme4 or coxme(?)
+- ADD: export mids objects to other packages like `lme4` or `coxme`(?)
 
-# Think about adding
+# Discussion
 
-- JOMO in mice --> on the side for now
+- JOMO in `mice` --> on the side for now
 
@@
 
+- Timeseries: and polynomial relationship in the clustering.
+
 # References

### 2021-11-01 Multilevel Missing Data/Multilevel-Missing-Data.tex
Outline paper incl modeling steps

@@
 Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
-Medical Center Utrecht \And Valentijn\\University Medical Center Utrecht
-\AND Gerko Vink\\Utrecht University \And Thomas Debray\\University
-Medical Center Utrecht
+Medical Center Utrecht \AND Valentijn de Jong\\University Medical
+Center Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
+Debray\\University Medical Center Utrecht
 }
@@
 
-\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn, Gerko
-Vink, Thomas Debray}
+\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
+Jong, Gerko Vink, Thomas Debray}
 \Plaintitle{Imputation of Incomplete Multilevel Data with mice}
@@
 Tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
-Including methods for ignorable and non-ignorable missingness. AKA
-missing on so many levels.
+Including methods for ignorable and non-ignorable missingness.
 }
 
-\Keywords{missing data, multilevel, clustering, \proglang{R}}
-\Plainkeywords{missing data, multilevel, clustering, R}
+\Keywords{missing
+data, multilevel, clustering, \pkg{mice}, \proglang{R}}
+\Plainkeywords{missing data, multilevel, clustering, mice, R}
 
@@
   What is clustering/multilevel data? In this paper, we discuss grouped
-  observations, not longitudinal data (within-patient clustering). What
-  do we mean by clustering? In the medical field: Clustering by studies
-  (IPDMA), hospitals in registries, multi-center studies etc. In other
-  fields: e.g.~official stats clustering at country-level, or social
-  sciences clustering at school-level (related to the sampling design).
+  observations, not longitudinal data (within-patient clustering).
+  -\textgreater{} ADD: timeseries also in Discussion section.
 \item
-  What is heterogeniety (i.e.~variability within studies vs.~variability
-  between studies)? ADD: Figure to show difference between patient-level
-  datapoints vs cluster-level datapoints. And different data frame
-  formats.
+  What do we mean by clustering? In the medical field: Clustering by
+  studies (IPDMA), hospitals in registries, multi-center studies etc. In
+  other fields: e.g.~official stats clustering at country-level, or
+  social sciences clustering at school-level (related to the sampling
+  design).
 \item
-  What methods are required to analyze multilevel data? Random effects
-  for intercept term, predictor effects, and/or variance residual error.
-  Add references, e.g. \citet{hox17}.
+  What is heterogeniety? I.e. variability within studies vs.~variability
+  between studies
+\item
+  What does multilevel data look like? ADD: figure to show difference
+  between patient-level datapoints vs cluster-level datapoints. Maybe
+  also add different data frame formats (or just explain in text that
+  there's long and wide formats).
+\item
+  What methods are required to analyze multilevel data? Add references,
+  e.g. \citet{hox17}. At least explain difference random effects for
+  intercept term, predictor effects, and/or variance residual error.
 \end{itemize}
@@
 \item
-  Why/where does missingness occur? I.e., not only patient-level but
-  also cluster-level.
+  Why/where does missingness occur in multilevel data? I.e., not only
+  patient-level but also cluster-level.
 \item
@@
   cluster) and non-measured variables (may differ per patient). TODO:
-  adjust md pattern to match text.
+  adjust md pattern to match text. -\textgreater{} syst may vary or same
+  for all patients (observations/participants).
 \end{itemize}
@@
 \item
+  ADD: missingness mech here
+\item
   Why are standard (ad hoc) missing data methods not well suited?
 \item
-  What other methods are available? General overview of approaches, see
-  \citet{audi18} \citet{grun18}. E.g., imputation of study level versus
-  patient-level covariates, and one-stage imputation versus two-stage
-  imputation methods.
-\item
-  Additional problem addressed in this paper: handling MNAR.
+  What types of multilevel methods are available? General overview of
+  approaches, see \citet{audi18} \citet{grun18}. E.g., imputation of
+  study level versus patient-level covariates, and one-stage imputation
+  versus two-stage imputation methods.
 \end{itemize}
@@
 \item
-  Case study options: metamisc::impact (IPD), GREAT (IPD).
+  We focus on the workflow for conditional modeling (not JOMO) in
+  \texttt{mice}. Refer to other packages: \texttt{mitml}.
+\item
+  Case study options: \texttt{metamisc::impact} (real data on traumatic
+  brain injuries, IPD), \texttt{mice::popularity} (simulated data with
+  MNAR/MAR mixture, schools). -\textgreater{} Check Gelman's data/NSRI
+  data.
 \end{itemize}
@@
 \section{Workflows}\label{workflows}}
+
+\hypertarget{case-study}{%
+\subsection{Case study}\label{case-study}}
+
+\begin{itemize}
+\tightlist
+\item
+  Introduce case study and set scope of this tutorial: We're providing
+  an overview of implementations. It's up-to the reader to decide which
+  strategy suits their data. So we won't go into detail for the
+  different methods (and equations). This paper is just a software
+  tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
+  help function that suggests a suitable pred matrix to the user, given
+  a certain analysis model.
+\end{itemize}
 
@@
 \item
-  ADD: Section with more in-depth info on different data types and
-  suitable methods. Add some equations, but not too much -\textgreater{}
-  grow in complexity.
+  Which models will we discuss? We'll build the model to grow in
+  complexity. The final model is the most complex but also the most
+  versatile.
 \item
-  Ideally we impute with random everything and heteroscedastic errors:
-  most generic method (no worry about congeneiality) --\textgreater{}
-  explain predictor matrix. Just mention that we should think about
-  random intercept/slope, but just refer to other papers. This paper is
-  just a software tutorial. --\textgreater{} Typically, at least random
-  intercepts, but often random slopes as well. Refer to other papers for
-  background, we'll focus just on the software implementation of the
-  situations mentioned there. Sometimes there's little reason to assume
-  some variable is affected by heterogeniety. We won't focus on
-  congeneiality, that's too complex for 2-3 papers.
+  Note on model complexity: Typically, we should at least use random
+  intercepts, but often random slopes as well. Ideally we impute with
+  random everything and heteroscedastic errors: most generic method (no
+  worry about congeniality, but don't mention the term) -\textgreater{}
+  Refer to other papers for background, we'll focus just on the software
+  implementation of the situations mentioned there. Sometimes there's
+  little reason to assume some variable is affected by heterogeneity.
+  -\textgreater{} Refer to Meng, Vincent, and a paper by Grund on
+  congeniality and random slopes.
+\item
+  Step 0: study as a predictor, AKA multilevel imputation for dummies.
+  Doesn't work for syst missing.
 \end{itemize}
@@
 \item
-  How to define the imputation model(s) in mice?
+  How to define the imputation model(s) in \texttt{mice}?
 \item
   What do the different implementations look like?
+\item
+  Step 1: Intercept
+\item
+  Step 2: Slope
+\item
+  Step 3: Residuals
+\item
+  Heckman model for MNAR
 \end{itemize}
@@
 \item
+  Analysis of scientific interest.
+\item
+  Pooling using \texttt{mitml}.
+\item
   Pooling `regular' parameters vs more `exotic' parameters (SE of
@@
 \item
-  ADD: export mids objects to other packages like lme4 or coxme(?)
+  ADD: export mids objects to other packages like \texttt{lme4} or
+  \texttt{coxme}(?)
 \end{itemize}
 
-\hypertarget{think-about-adding}{%
-\section{Think about adding}\label{think-about-adding}}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
   Additional levels of clustering
+\item
+  Timeseries: and polynomial relationship in the clustering.
 \end{itemize}

### 2021-11-01 Multilevel Missing Data/Multilevel Missing Data.Rmd
Add impact data

@@
 library(tidyverse)
-
+library(mice)
+library(metamisc)
 ```
@@
 
-# Workflows
-
-## Case study
-
 - Introduce case study and set scope of this tutorial: We're providing an overview of implementations.
@@
 -> ADD: some kind of help function that suggests a suitable pred matrix to the user, given a certain analysis model.
+
+# Workflows
+
+## Case study
+
+- We'll use the IMPACT data (`metamisc::impact`) and simulated MNAR data (based on `mice::popularity`).
+
+- IMPACT: A data frame with 11022 observations on the following 11 variables:
+`name` Name of the study,
+`type` Type of study, RCT: randomized controlled trial,OBS: observational cohort,
+`age` Age of the patient,
+`motor_score` Glasgow Coma Scale motor score,
+`pupil` Pupillary reactivity,
+`ct` Marshall Computerized Tomography classification,
+`hypox` Hypoxia (0=no, 1=yes),
+`hypots` Hypotension (0=no, 1=yes),
+`tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+`edh` Epidural hematoma (0=no, 1=yes),
+`mort` 6-month mortality (0=alive, 1=dead).
+
+```{r impact}
+# load data
+data("impact")
+# descriptive statistics
+psych::describe(impact)[,c(2:5,8:9)]
+```
 

### 2021-11-01 Multilevel Missing Data/Multilevel-Missing-Data.tex
Add impact data

@@
   We focus on the workflow for conditional modeling (not JOMO) in
-  \texttt{mice}. Refer to other packages: \texttt{mitml}.
+  \texttt{mice}. Refer to other packages: \texttt{mitml},
+  \texttt{miceadds}.
 \item
@@
   data.
+\item
+  Introduce case study and set scope of this tutorial: We're providing
+  an overview of implementations. It's up-to the reader to decide which
+  strategy suits their data. So we won't go into detail for the
+  different methods (and equations). This paper is just a software
+  tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
+  help function that suggests a suitable pred matrix to the user, given
+  a certain analysis model.
 \end{itemize}
@@
 \begin{itemize}
-\tightlist
 \item
-  Introduce case study and set scope of this tutorial: We're providing
-  an overview of implementations. It's up-to the reader to decide which
-  strategy suits their data. So we won't go into detail for the
-  different methods (and equations). This paper is just a software
-  tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
-  help function that suggests a suitable pred matrix to the user, given
-  a certain analysis model.
+  We'll use the IMPACT data (\texttt{metamisc::impact}) and simulated
+  MNAR data (based on \texttt{mice::popularity}).
+\item
+  IMPACT: A data frame with 11022 observations on the following 11
+  variables: \texttt{name} Name of the study, \texttt{type} Type of
+  study, RCT: randomized controlled trial,OBS: observational cohort,
+  \texttt{age} Age of the patient, \texttt{motor\_score} Glasgow Coma
+  Scale motor score, \texttt{pupil} Pupillary reactivity, \texttt{ct}
+  Marshall Computerized Tomography classification, \texttt{hypox}
+  Hypoxia (0=no, 1=yes), \texttt{hypots} Hypotension (0=no, 1=yes),
+  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+  \texttt{edh} Epidural hematoma (0=no, 1=yes), \texttt{mort} 6-month
+  mortality (0=alive, 1=dead).
 \end{itemize}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # load data
+R> data("impact")
+R> # descriptive statistics
+R> psych::describe(impact)[,c(2:5,8:9)]
+\end{CodeInput}
+\begin{CodeOutput}
+                 n  mean    sd median min max
+name*        11022  8.32  4.37      8   1  15
+type*        11022  1.73  0.44      2   1   2
+age          11022 34.93 15.89     31  14  93
+motor_score* 11022  2.59  1.18      3   1   4
+pupil*       11022  1.46  0.71      1   1   3
+ct*          11022  1.98  0.89      2   1   3
+hypox        11022  0.22  0.41      0   0   1
+hypots       11022  0.17  0.38      0   0   1
+tsah         11022  0.45  0.50      0   0   1
+edh          11022  0.13  0.34      0   0   1
+mort         11022  0.26  0.44      0   0   1
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 \item
-  JOMO in mice --\textgreater{} on the side for now
+  JOMO in \texttt{mice} --\textgreater{} on the side for now
 \item

### 2021-11-01 Manuscript/Manuscript.Rmd
Add popularity data and missing data pattern plot

--- 2021-11-01 Multilevel Missing Data/Multilevel Missing Data.Rmd
@@
 options(prompt = 'R> ', continue = '+ ')
+
+# packages
 library(tidyverse)
+library(mice)
+library(miceadds)
+library(metamisc)
+library(pan)
 
+# functions
+miceadds::source.all(path = "../R")
+
+# plot parameters
+plot_col <- mice:::mdc(1:2) %>% setNames(c("observed", "missing"))
 ```
@@
 patterns <- rbind(
-  tibble(pat = rep("Systematic", 5), rownr = 5:1, X = rep(1, 5), Y = c(1,1,1,0, 0), Z = c(1, rep(0,4))) %>% pivot_longer(cols = c(X, Y, Z)),
-  tibble(pat = rep("Sporadic", 5), rownr = 5:1, X = c(0, rep(1,4)), Y = c(1,1,1,0, 0), Z = c(1, rep(0,3), 1)) %>% pivot_longer(cols = c(X, Y, Z))) %>%
-  mutate(across(everything(), as.factor))
-ggplot(patterns, aes(x = name, y = rownr, fill = value, type = pat)) +
+  tibble(
+    pat = rep("systematic", 5),
+    rownr = 5:1,
+    X1 = rep("observed", 5),
+    X2 = c(rep("observed", 3), rep("missing", 2)),
+    X3 = c(rep("missing", 3),  rep("observed", 2))
+  ) %>% pivot_longer(cols = c(X1, X2, X3)),
+  tibble(
+    pat = rep("sporadic", 5),
+    rownr = 5:1,
+    X1 = c("observed", "missing", rep("observed", 3)),
+    X2 = c(rep("observed", 2), rep("missing", 3)),
+    X3 = c(rep("missing", 3),  rep("observed", 2))
+  ) %>% pivot_longer(cols = c(X1, X2, X3))
+) %>%
+  mutate(across(everything(), as.factor),
+         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE))
+# plot
+ggplot(patterns, aes(
+  x = name,
+  y = rownr,
+  fill = value,
+  type = pat
+)) +
   geom_tile(color = "black") +
-  scale_fill_brewer(palette = "Set1") +
-  facet_wrap(~pat) +
-  theme_void() +
-  theme(legend.position = "none") 
+  scale_y_discrete(labels = c(2, 2, 1, 1, 1)) +
+  scale_fill_manual(values = plot_col, name = "Datapoint") +
+  facet_wrap( ~ pat, ncol = 2) +
+  theme_minimal() +
+  theme(legend.position = "bottom") +
+  labs(x = "Variable", y = "Cluster")
 ```
 
--	ADD: missingness mech here
+-	What kinds of missingness are there? ADD: missingness mechanisms here.
 
@@
 E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.
+
+- Additional difficulty that is addressed in this tutorial: MNAR data.
 
@@
 -> ADD: some kind of help function that suggests a suitable pred matrix to the user, given a certain analysis model.
+
+# Workflows
+
+## Case study
+
+- We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).
+
+- `impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
+    * `name` Name of the study,
+    * `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
+    * `age` Age of the patient,
+    * `motor_score` Glasgow Coma Scale motor score,
+    * `pupil` Pupillary reactivity,
+    * `ct` Marshall Computerized Tomography classification,
+    * `hypox` Hypoxia (0=no, 1=yes),
+    * `hypots` Hypotension (0=no, 1=yes),
+    * `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+    * `edh` Epidural hematoma (0=no, 1=yes),
+    * `mort` 6-month mortality (0=alive, 1=dead).
+
+```{r impact}
+# load data
+data("impact")
+# # descriptive statistics
+# by(impact, impact$name, summary) 
+# psych::describe(impact)[,c(2:5,8:9)]
+# missingness
+md_pat(impact)
+```
+-> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf), the data is already imputed (Steyerberg et al, 2008).
+
+- `popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables: 
+    * `pupil`	Pupil number within class,
+    * `class`	Class number,
+    * `extrav`	Pupil extraversion,
+    * `sex`	Pupil gender,
+    * `texp`	Teacher experience (years),
+    * `popular`	Pupil popularity,
+    * `popteach`	Teacher popularity.
+
+```{r pop}
+# load data
+pop <- readRDS("../Data/popNCR.RDS")
+# missingness
+md_pat(pop)
+```
 

### 2021-11-01 Manuscript/Manuscript.tex
Clean up repository, updat readme
Add popularity data and missing data pattern plot

--- 2021-11-01 Multilevel Missing Data/Multilevel-Missing-Data.tex
@@
 
-\begin{center}\includegraphics{Multilevel-Missing-Data_files/figure-latex/patterns-1} \end{center}
+\begin{center}\includegraphics{Manuscript_files/figure-latex/patterns-1} \end{center}
 
@@
 \item
-  ADD: missingness mech here
+  What kinds of missingness are there? ADD: missingness mechanisms here.
 \item
@@
   versus two-stage imputation methods.
+\item
+  Additional difficulty that is addressed in this tutorial: MNAR data.
 \end{itemize}
@@
 \begin{itemize}
+\item
+  We'll use the IMPACT data (\texttt{metamisc::impact}) and a MAR/MNAR
+  version of the \texttt{mice::popmis} data (i.e., a variation on the
+  Hox (2010) popularity data, where the missingness in the variables is
+  either missing at random (MAR) or missing not at random (MNAR)).
+\item
+  \texttt{impact} is traumatic brain injury data with patients clustered
+  in studies, \(n_{\text{participants}} = 11022\) and
+  \(n_{\text{clusters}} = 15\), on the following 11 variables:
+
+  \begin{itemize}
   \tightlist
   \item
-  Introduce case study and set scope of this tutorial: We're providing
-  an overview of implementations. It's up-to the reader to decide which
-  strategy suits their data. So we won't go into detail for the
-  different methods (and equations). This paper is just a software
-  tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
-  help function that suggests a suitable pred matrix to the user, given
-  a certain analysis model.
+    \texttt{name} Name of the study,
+  \item
+    \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
+    observational cohort),
+  \item
+    \texttt{age} Age of the patient,
+  \item
+    \texttt{motor\_score} Glasgow Coma Scale motor score,
+  \item
+    \texttt{pupil} Pupillary reactivity,
+  \item
+    \texttt{ct} Marshall Computerized Tomography classification,
+  \item
+    \texttt{hypox} Hypoxia (0=no, 1=yes),
+  \item
+    \texttt{hypots} Hypotension (0=no, 1=yes),
+  \item
+    \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+  \item
+    \texttt{edh} Epidural hematoma (0=no, 1=yes),
+  \item
+    \texttt{mort} 6-month mortality (0=alive, 1=dead).
   \end{itemize}
+\end{itemize}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # load data
+R> data("impact")
+R> # # descriptive statistics
+R> # by(impact, impact$name, summary) 
+R> # psych::describe(impact)[,c(2:5,8:9)]
+R> # missingness
+R> md_pat(impact)
+\end{CodeInput}
+\begin{CodeOutput}
+ /\     /\
+{  `---'  }
+{  O   O  }
+==>  V <==  No need for mice. This data set is completely observed.
+ \  \|/  /
+  `-----'
+\end{CodeOutput}
+\begin{CodeOutput}
+      name type age motor_score pupil ct hypox hypots tsah edh mort  
+11022    1    1   1           1     1  1     1      1    1   1    1 0
+         0    0   0           0     0  0     0      0    0   0    0 0
+\end{CodeOutput}
+\end{CodeChunk}
+
+-\textgreater{} Why are there no missings? According to the
+\href{https://cran.r-project.org/web/packages/metamisc/metamisc.pdf}{vignette},
+the data is already imputed (Steyerberg et al, 2008).
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{popNCR} is a simulated dataset with pupils clustered in
+  classes, \(n_{\text{participants}} = 2000\),
+  \(n_{\text{clusters}} = 100\), on 7 variables:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \texttt{pupil} Pupil number within class,
+  \item
+    \texttt{class} Class number,
+  \item
+    \texttt{extrav} Pupil extraversion,
+  \item
+    \texttt{sex} Pupil gender,
+  \item
+    \texttt{texp} Teacher experience (years),
+  \item
+    \texttt{popular} Pupil popularity,
+  \item
+    \texttt{popteach} Teacher popularity.
+  \end{itemize}
+\end{itemize}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # load data
+R> pop <- readRDS("../Data/popNCR.RDS")
+R> # missingness
+R> md_pat(pop)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-1} \end{center}
+
+\end{CodeChunk}
 

### 2021-11-01 R/plot_md_pat.R

@@ -1,0 +1,93 @@
+# missing data pattern
+md_pat <- function(dat) {
+  # get md pattern and store additional info
+  pat <- mice::md.pattern(dat, plot = FALSE)
+  vrb <- colnames(pat)[-ncol(pat)]
+  vrb_labs <- abbreviate(vrb)
+  vrb_caption <- purrr::map2_chr(vrb_labs, vrb, ~{paste(.x, " = ", .y)}) %>% paste(collapse = ", ")
+  colnames(pat) <- c(vrb, "NA_per_pat")
+  pat_freq <- as.numeric(rownames(pat))[-nrow(pat)]
+  NA_per_pat <- as.numeric(pat[, ncol(pat)])[-nrow(pat)]
+  NA_per_vrb <- as.numeric(pat[nrow(pat),])[-ncol(pat)]
+  NA_total <- pat[nrow(pat), ncol(pat)]
+  # make the pattern tidy
+  if (NA_total == 0) {
+    pat_long <-
+      data.frame(
+        NA_per_vrb = 0,
+        NA_per_pat = 0,
+        pat_freq = pat_freq,
+        pat_nr = 1,
+        obs = 1
+      ) %>% cbind(vrb = vrb)
+  } 
+  if (NA_total != 0) {
+    pat_long <- pat[-nrow(pat),] %>%
+    cbind(., pat_freq, pat_nr = 1:nrow(.)) %>%
+    as.data.frame() %>%
+    tidyr::pivot_longer(cols = all_of(vrb),
+                        names_to = "vrb",
+                        values_to = "obs") %>%
+    cbind(NA_per_vrb, .)
+  }
+  # plot the md pattern
+  p <- pat_long %>%
+    ggplot2::ggplot() +
+    ggplot2::geom_tile(ggplot2::aes(
+      x = vrb,
+      y = pat_nr,
+      fill = as.factor(obs),
+      group = NA_per_pat
+    ),
+    color = "black") +
+    # set axes
+    ggplot2::scale_x_discrete(
+      limits = vrb,
+      position = "bottom",
+      labels = as.character(NA_per_vrb),
+      expand = c(0, 0)
+    ) +
+    ggplot2::scale_y_reverse(
+      breaks = 1:max(pat_long$pat_nr),
+      labels = as.character(pat_freq),
+      expand = c(0, 0),
+      sec.axis = ggplot2::dup_axis(labels = as.character(NA_per_pat),
[…]

### 2021-11-04 Manuscript/Manuscript.Rmd
Add some more references to manuscript

@@
 
-- What is heterogeniety? I.e. variability within studies vs. variability between studies
+- What is heterogeneity? I.e. variability within studies vs. variability between studies
 
@@
 
-- What methods are required to analyze multilevel data?	Add references, e.g. @hox17. At least explain difference random effects for intercept term, predictor effects, and/or variance residual error.
+- What methods are required to analyze multilevel data?	Add references, e.g.
+@hox17 and @jong21.
+At least explain difference random effects for intercept term, predictor effects, and/or variance residual error.
 
@@
 
--	What kinds of missingness are there? ADD: missingness mechanisms here.
+-	What kinds of missingness are there? ADD: missingness mechanisms here. See e.g. @yuce08 and @hox15.
 
@@
 - What types of multilevel methods are available?
-General overview of approaches, see @audi18 @grun18.
+General overview of approaches, see @audi18 and @grun18.
 E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.
@@
 
-- We focus on the workflow for conditional modeling (not JOMO) in `mice`. Refer to other packages: `mitml`, `miceadds`.
+- We focus on the workflow for conditional modeling (not JOMO) in `mice`. Refer to other packages: `mitml`, `miceadds`, `mdmb`.
 
-- Case study options: `metamisc::impact` (real data on traumatic brain injuries, IPD), `mice::popularity` (simulated data with MNAR/MAR mixture, schools). -> Check Gelman's data/NSRI data.
+- Case study options: `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s), `mice::popularity` (simulated data on school kids, with MNAR/MAR mixture).
+TODO: Check example data Gelman.
 
@@
 We'll keep it practical.
--> ADD: some kind of help function that suggests a suitable pred matrix to the user, given a certain analysis model.
+-> ADD: some kind of help function that suggests a suitable predictor matrix to the user, given a certain analysis model.
 
@@
 
-## Case study
+We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).
 
-- We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).
+## Case study I: IMPACT
 
-- `impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
+`impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
     * `name` Name of the study,
@@
 
-- `popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables: 
+## Case Study II: Popularity
+
+`popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables: 
     * `pupil`	Pupil number within class,
@@
 
-- Pooling 'regular' parameters vs more 'exotic' parameters (SE of residual errors, or autocorreletion)
+- Pooling 'regular' parameters vs more 'exotic' parameters (SE of residual errors, or autocorrelation)
 
-- ADD: export mids objects to other packages like `lme4` or `coxme`(?)
+- ADD: export `mids` objects to other packages like `lme4` or `coxme`?
 

### 2021-11-04 Manuscript/Manuscript.tex
Add some more references to manuscript

--- 2021-11-01 Multilevel Missing Data/Multilevel-Missing-Data.tex
@@
 \item
-  What is heterogeniety? I.e. variability within studies vs.~variability
+  What is heterogeneity? I.e. variability within studies vs.~variability
   between studies
@@
   What methods are required to analyze multilevel data? Add references,
-  e.g. \citet{hox17}. At least explain difference random effects for
-  intercept term, predictor effects, and/or variance residual error.
+  e.g. \citet{hox17} and \citet{jong21}. At least explain difference
+  random effects for intercept term, predictor effects, and/or variance
+  residual error.
 \end{itemize}
@@
 \item
-  ADD: missingness mech here
+  What kinds of missingness are there? ADD: missingness mechanisms here.
+  See e.g. \citet{yuce08} and \citet{hox15}.
 \item
@@
   What types of multilevel methods are available? General overview of
-  approaches, see \citet{audi18} \citet{grun18}. E.g., imputation of
+  approaches, see \citet{audi18} and \citet{grun18}. E.g., imputation of
   study level versus patient-level covariates, and one-stage imputation
   versus two-stage imputation methods.
+\item
+  Additional difficulty that is addressed in this tutorial: MNAR data.
 \end{itemize}
@@
   We focus on the workflow for conditional modeling (not JOMO) in
-  \texttt{mice}. Refer to other packages: \texttt{mitml}.
+  \texttt{mice}. Refer to other packages: \texttt{mitml},
+  \texttt{miceadds}, \texttt{mdmb}.
 \item
-  Case study options: \texttt{metamisc::impact} (real data on traumatic
-  brain injuries, IPD), \texttt{mice::popularity} (simulated data with
-  MNAR/MAR mixture, schools). -\textgreater{} Check Gelman's data/NSRI
-  data.
-\end{itemize}
-
-\hypertarget{workflows}{%
-\section{Workflows}\label{workflows}}
-
-\hypertarget{case-study}{%
-\subsection{Case study}\label{case-study}}
-
-\begin{itemize}
-\tightlist
+  Case study options: \texttt{metamisc::impact} (real IPD on traumatic
+  brain injuries, without \texttt{NA}s), \texttt{mice::popularity}
+  (simulated data on school kids, with MNAR/MAR mixture). TODO: Check
+  example data Gelman.
 \item
@@
   tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
-  help function that suggests a suitable pred matrix to the user, given
-  a certain analysis model.
+  help function that suggests a suitable predictor matrix to the user,
+  given a certain analysis model.
 \end{itemize}
+
+\hypertarget{workflows}{%
+\section{Workflows}\label{workflows}}
+
+We'll use the IMPACT data (\texttt{metamisc::impact}) and a MAR/MNAR
+version of the \texttt{mice::popmis} data (i.e., a variation on the Hox
+(2010) popularity data, where the missingness in the variables is either
+missing at random (MAR) or missing not at random (MNAR)).
+
+\hypertarget{case-study-i-impact}{%
+\subsection{Case study I: IMPACT}\label{case-study-i-impact}}
+
+\texttt{impact} is traumatic brain injury data with patients clustered
+in studies, \(n_{\text{participants}} = 11022\) and
+\(n_{\text{clusters}} = 15\), on the following 11 variables: *
+\texttt{name} Name of the study, * \texttt{type} Type of study (RCT:
+randomized controlled trial, OBS: observational cohort), * \texttt{age}
+Age of the patient, * \texttt{motor\_score} Glasgow Coma Scale motor
+score, * \texttt{pupil} Pupillary reactivity, * \texttt{ct} Marshall
+Computerized Tomography classification, * \texttt{hypox} Hypoxia (0=no,
+1=yes), * \texttt{hypots} Hypotension (0=no, 1=yes), * \texttt{tsah}
+Traumatic subarachnoid hemorrhage (0=no, 1=yes), * \texttt{edh} Epidural
+hematoma (0=no, 1=yes), * \texttt{mort} 6-month mortality (0=alive,
+1=dead).
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # load data
+R> data("impact")
+R> # # descriptive statistics
+R> # by(impact, impact$name, summary) 
+R> # psych::describe(impact)[,c(2:5,8:9)]
+R> # missingness
+R> md_pat(impact)
+\end{CodeInput}
+\begin{CodeOutput}
+ /\     /\
+{  `---'  }
+{  O   O  }
+==>  V <==  No need for mice. This data set is completely observed.
+ \  \|/  /
+  `-----'
+\end{CodeOutput}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}
+
+\end{CodeChunk}
+
+-\textgreater{} Why are there no missings? According to the
+\href{https://cran.r-project.org/web/packages/metamisc/metamisc.pdf}{vignette},
+the data is already imputed (Steyerberg et al, 2008).
+
+\hypertarget{case-study-ii-popularity}{%
+\subsection{Case Study II: Popularity}\label{case-study-ii-popularity}}
+
+\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
+\(n_{\text{participants}} = 2000\), \(n_{\text{clusters}} = 100\), on 7
+variables: * \texttt{pupil} Pupil number within class, * \texttt{class}
+Class number, * \texttt{extrav} Pupil extraversion, * \texttt{sex} Pupil
+gender, * \texttt{texp} Teacher experience (years), * \texttt{popular}
+Pupil popularity, * \texttt{popteach} Teacher popularity.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # load data
+R> pop <- readRDS("../Data/popNCR.RDS")
+R> # missingness
+R> md_pat(pop)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-1} \end{center}
+
+\end{CodeChunk}
 
@@
   Pooling `regular' parameters vs more `exotic' parameters (SE of
-  residual errors, or autocorreletion)
+  residual errors, or autocorrelation)
 \item
-  ADD: export mids objects to other packages like \texttt{lme4} or
-  \texttt{coxme}(?)
+  ADD: export \texttt{mids} objects to other packages like \texttt{lme4}
+  or \texttt{coxme}?
 \end{itemize}

### 2021-11-04 R/plot_md_pat.R

@@ -3,7 +3,7 @@ md_pat <- function(dat) {
   # get md pattern and store additional info
   pat <- mice::md.pattern(dat, plot = FALSE)
   vrb <- colnames(pat)[-ncol(pat)]
-  vrb_labs <- abbreviate(vrb)
+  vrb_labs <- abbreviate(vrb, minlength = 3)
   vrb_caption <- purrr::map2_chr(vrb_labs, vrb, ~{paste(.x, " = ", .y)}) %>% paste(collapse = ", ")
   colnames(pat) <- c(vrb, "NA_per_pat")
   pat_freq <- as.numeric(rownames(pat))[-nrow(pat)]
@@ -63,7 +63,7 @@ md_pat <- function(dat) {
       subtitle = paste0("Total number of missing entries: ",
                         NA_total,
                         "\n\n"),
-      caption = paste("\n Note.", vrb_caption)
+      caption = stringr::str_wrap(paste("\n\nNote.", vrb_caption))
     ) +
     ggplot2::geom_text(
       ggplot2::aes(
@@ -84,6 +84,7 @@ md_pat <- function(dat) {
         r = 10,
         unit = "pt"
       ),
+      plot.caption.position = "plot",
       axis.title.y.right = ggplot2::element_text(margin = ggplot2::margin(l = 10)),
       panel.grid = ggplot2::element_blank()
     ) +

### 2021-11-04 Manuscript/Manuscript.tex
Add data exploration popularity dataset

@@
 
-\texttt{impact} is traumatic brain injury data with patients clustered
-in studies, \(n_{\text{participants}} = 11022\) and
-\(n_{\text{clusters}} = 15\), on the following 11 variables: *
-\texttt{name} Name of the study, * \texttt{type} Type of study (RCT:
-randomized controlled trial, OBS: observational cohort), * \texttt{age}
-Age of the patient, * \texttt{motor\_score} Glasgow Coma Scale motor
-score, * \texttt{pupil} Pupillary reactivity, * \texttt{ct} Marshall
-Computerized Tomography classification, * \texttt{hypox} Hypoxia (0=no,
-1=yes), * \texttt{hypots} Hypotension (0=no, 1=yes), * \texttt{tsah}
-Traumatic subarachnoid hemorrhage (0=no, 1=yes), * \texttt{edh} Epidural
-hematoma (0=no, 1=yes), * \texttt{mort} 6-month mortality (0=alive,
-1=dead).
+\begin{itemize}
+\item
+  What does the data look like? \texttt{impact} is traumatic brain
+  injury data with patients clustered in studies,
+  \(n_{\text{participants}} = 11022\) and \(n_{\text{clusters}} = 15\),
+  on the following 11 variables:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \texttt{name} Name of the study,
+  \item
+    \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
+    observational cohort),
+  \item
+    \texttt{age} Age of the patient,
+  \item
+    \texttt{motor\_score} Glasgow Coma Scale motor score,
+  \item
+    \texttt{pupil} Pupillary reactivity,
+  \item
+    \texttt{ct} Marshall Computerized Tomography classification,
+  \item
+    \texttt{hypox} Hypoxia (0=no, 1=yes),
+  \item
+    \texttt{hypots} Hypotension (0=no, 1=yes),
+  \item
+    \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+  \item
+    \texttt{edh} Epidural hematoma (0=no, 1=yes),
+  \item
+    \texttt{mort} 6-month mortality (0=alive, 1=dead).
+  \end{itemize}
+\end{itemize}
 
 \begin{CodeChunk}
-\begin{CodeInput}
-R> # load data
-R> data("impact")
-R> # # descriptive statistics
-R> # by(impact, impact$name, summary) 
-R> # psych::describe(impact)[,c(2:5,8:9)]
-R> # missingness
-R> md_pat(impact)
-\end{CodeInput}
 \begin{CodeOutput}
@@
 
-\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
-\(n_{\text{participants}} = 2000\), \(n_{\text{clusters}} = 100\), on 7
-variables: * \texttt{pupil} Pupil number within class, * \texttt{class}
-Class number, * \texttt{extrav} Pupil extraversion, * \texttt{sex} Pupil
-gender, * \texttt{texp} Teacher experience (years), * \texttt{popular}
-Pupil popularity, * \texttt{popteach} Teacher popularity.
+\begin{itemize}
+\item
+  What does the data look like? \texttt{popNCR} is a simulated dataset
+  with pupils clustered in classes, \(n_{\text{participants}} = 2000\),
+  \(n_{\text{clusters}} = 100\), on 7 variables:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \texttt{pupil} Pupil number within class,
+  \item
+    \texttt{class} Class number,
+  \item
+    \texttt{extrav} Pupil extraversion,
+  \item
+    \texttt{sex} Pupil gender,
+  \item
+    \texttt{texp} Teacher experience (years),
+  \item
+    \texttt{popular} Pupil popularity,
+  \item
+    \texttt{popteach} Teacher popularity.
+  \end{itemize}
+\item
+  What are the ICCs? For \texttt{popular} the ICC is 0.33. For
+  \texttt{popteach} it is 0.31.
+\item
+  What does the missingness look like? Induced MAR/MNAR missingness.
+  Missing data pattern:
+\end{itemize}
 
 \begin{CodeChunk}
-\begin{CodeInput}
-R> # load data
-R> pop <- readRDS("../Data/popNCR.RDS")
-R> # missingness
-R> md_pat(pop)
-\end{CodeInput}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-1} \end{center}
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_pat-1} \end{center}
+
+\end{CodeChunk}
+
+\begin{itemize}
+\item
+  Does the missing data of popular depend on popteach? One could for
+  example check this by making a histogram of popteach separately for
+  the pupils with known popularity and missing popularity. We do see
+  that the histogram for the missing popular (TRUE) is further to the
+  right than the histogram for observed popular (FALSE). This would
+  indicate a right-tailed MAR missingness. In fact this is exactly what
+  happens, because we created the missingness in these data ourselves.
+  But we can make it observable by examining the relations between the
+  missingness in popular and the observed data in popteach.
+\item
+  Does the missingness in teacher popularity depend on pupil popularity?
+  Yes: there is a dependency. The relation seems to be right-tailed.
+\end{itemize}
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_dist-1} \end{center}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_dist-2} \end{center}
 
@@
 \item
-  Step 0: study as a predictor, AKA multilevel imputation for dummies.
-  Doesn't work for syst missing.
-\end{itemize}
-
-\hypertarget{conditional-models}{%
-\subsection{Conditional models}\label{conditional-models}}
-
-\begin{itemize}
+  Step 0: As predictor
 \item
-  How to define the imputation model(s) in \texttt{mice}?
+  Step 1: Random intercepts
 \item
-  What do the different implementations look like?
-\item
-  Step 1: Intercept
-\item
-  Step 2: Slope
+  Step 2: Random slopes
 \item
@@
   Heckman model for MNAR
+\item
+  What do the different implementations look like? How to define the
+  imputation model(s) in \texttt{mice}?
+\end{itemize}
+
+\hypertarget{step-0}{%
+\subsection{Step 0}\label{step-0}}
+
+\begin{itemize}
+\item
+  AKA multilevel imputation for dummies.
+\item
+  Doesn't work for systematic missingness.
+\end{itemize}
+
+\hypertarget{step-1-3}{%
+\subsection{Step 1-3}\label{step-1-3}}
+
+\begin{itemize}
+\tightlist
+\item
+  TODO: fill in.
 \end{itemize}

### 2021-11-04 R/plot_md.R

@@ -1,5 +1,5 @@
 # missing data pattern
-md_pat <- function(dat) {
+plot_md_pat <- function(dat) {
   # get md pattern and store additional info
   pat <- mice::md.pattern(dat, plot = FALSE)
   vrb <- colnames(pat)[-ncol(pat)]
@@ -92,3 +92,78 @@ md_pat <- function(dat) {
   # output
   return(p)
 }
+
+#' Title Create a histogram/bar plot of a variable conditional on missingness in another variable
+#'
+#' @param dat An incomplete dataset of class dataframe
+#' @param x A variable to plot
+#' @param z A second variable to plot
+#'
+#' @return A ggplot object with two facets (variable z observed vs. missing)
+#' @export
+plot_NA_cond <- function(dat, x, z, bins = NULL) {
+  # define graphing elements to add to plot
+  if (is.numeric(dat[[x]])) {
+    # for continuous variables
+    geom <- ggplot2::geom_histogram(fill = "white", binwidth = bins)
+  } else {
+    # for categorical variables
+    geom <- ggplot2::geom_bar(fill = "white")
+  }
+  # create facet labels
+  facet_labs <- c(paste(z, "observed"), paste(z, "missing")) %>%
+    setNames(c("observed", "missing"))
+  # preprocess the data
+  d <- dat[!is.na(dat[[x]]),] %>%
+    dplyr::mutate(conditional = factor(
+      is.na(.data[[z]]),
+      levels = c(FALSE, TRUE),
+      labels = c("observed", "missing")
+    )) 
+  
+  # plot
+  p <- d %>% ggplot2::ggplot(ggplot2::aes(x = .data[[x]],
+                                          color = conditional,
+                                          fill = "white")) +
+    geom +
+    # split by conditional variable
+    ggplot2::facet_wrap(
+      ~ conditional,
+      ncol = 1,
+      scales = "free_y",
+      labeller = ggplot2::labeller(conditional = facet_labs)
+    ) +
+    # style
+    ggplot2::scale_color_manual(
+        values = c(
+          "observed" = mice:::mdc(1),
+          "missing" = mice:::mdc(2),
+          "imputed" = mice:::mdc(2)
+        )
+      ) +
+      ggplot2::scale_fill_manual(
+        values = c(
+          "observed" = mice:::mdc(1),
+          "missing" = mice:::mdc(2),
+          "imputed" = mice:::mdc(2)
+        )
+      ) +
[…]

### 2021-11-05 Manuscript/Manuscript.Rmd

@@
 
+- What does the data look like?
 `impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
-    * `name` Name of the study,
-    * `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
-    * `age` Age of the patient,
-    * `motor_score` Glasgow Coma Scale motor score,
-    * `pupil` Pupillary reactivity,
-    * `ct` Marshall Computerized Tomography classification,
-    * `hypox` Hypoxia (0=no, 1=yes),
-    * `hypots` Hypotension (0=no, 1=yes),
-    * `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
-    * `edh` Epidural hematoma (0=no, 1=yes),
-    * `mort` 6-month mortality (0=alive, 1=dead).
 
-```{r impact}
+  - `name` Name of the study,
+  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
+  - `age` Age of the patient,
+  - `motor_score` Glasgow Coma Scale motor score,
+  - `pupil` Pupillary reactivity,
+  - `ct` Marshall Computerized Tomography classification,
+  - `hypox` Hypoxia (0=no, 1=yes),
+  - `hypots` Hypotension (0=no, 1=yes),
+  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+  - `edh` Epidural hematoma (0=no, 1=yes),
+  - `mort` 6-month mortality (0=alive, 1=dead).
+
+```{r impact, echo=FALSE}
 # load data
@@
 # missingness
-md_pat(impact)
+plot_md_pat(impact)
 ```
@@
 
-`popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables: 
-    * `pupil`	Pupil number within class,
-    * `class`	Class number,
-    * `extrav`	Pupil extraversion,
-    * `sex`	Pupil gender,
-    * `texp`	Teacher experience (years),
-    * `popular`	Pupil popularity,
-    * `popteach`	Teacher popularity.
+- What does the data look like? `popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables:
 
-```{r pop}
+  - `pupil`	Pupil number within class,
+  - `class`	Class number,
+  - `extrav`	Pupil extraversion,
+  - `sex`	Pupil gender,
+  - `texp`	Teacher experience (years),
+  - `popular`	Pupil popularity,
+  - `popteach`	Teacher popularity.
+
+```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
 # load data
 pop <- readRDS("../Data/popNCR.RDS")
+# ICCs
+icc_pop <- multilevel::ICC1(aov(popular ~ as.factor(class), data = pop))
+icc_teach <- multilevel::ICC1(aov(popteach ~ as.factor(class), data = pop))
+```
+- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`.
+
+- What does the missingness look like? Induced MAR/MNAR missingness. Missing data pattern:
+
+```{r pop_pat, echo=FALSE, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
 # missingness
-md_pat(pop)
+plot_md_pat(pop)
+```
+
+- Does the missing data of popular depend on popteach?
+One could for example check this by making a histogram of popteach separately for the pupils with known popularity and missing popularity.
+We do see that the histogram for the missing popular (TRUE) is further to the right than the histogram for observed popular (FALSE).
+This would indicate a right-tailed MAR missingness.
+In fact this is exactly what happens, because we created the missingness in these data ourselves.
+But we can make it observable by examining the relations between the missingness in popular and the observed data in popteach.
+
+- Does the missingness in teacher popularity depend on pupil popularity? Yes: there is a dependency. The relation seems to be right-tailed.
+
+```{r pop_dist, echo=FALSE, message=FALSE, warning=FALSE}
+# distributions
+plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
+plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
 ```
@@
 
-- Step 0: study as a predictor, AKA multilevel imputation for dummies. Doesn't work for syst missing.
+- Step 0: As predictor
 
-## Conditional models
+- Step 1: Random intercepts
 
-- How to define the imputation model(s) in `mice`?
-
-- What do the different implementations look like?
-
-- Step 1: Intercept
-
-- Step 2: Slope
+- Step 2: Random slopes
 
@@
 - Heckman model for MNAR
+
+- What do the different implementations look like? How to define the imputation model(s) in `mice`?
+
+## Step 0
+
+- AKA multilevel imputation for dummies.
+
+- Doesn't work for systematic missingness.
+
+## Step 1-3
+
+- TODO: fill in.
 

### 2021-11-22 Manuscript/Manuscript.Rmd
Update md pattern, add new case study

@@
     rownr = 5:1,
+    # obs = rep("observed", 5),
+    cluster = rep("observed", 5),
     X1 = rep("observed", 5),
@@
     X3 = c(rep("missing", 3),  rep("observed", 2))
-  ) %>% pivot_longer(cols = c(X1, X2, X3)),
+  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3)),
   tibble(
@@
     rownr = 5:1,
+    # obs = rep("observed", 5),
+    cluster = rep("observed", 5),
     X1 = c("observed", "missing", rep("observed", 3)),
@@
     X3 = c(rep("missing", 3),  rep("observed", 2))
-  ) %>% pivot_longer(cols = c(X1, X2, X3))
+  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3))
 ) %>%
   mutate(across(everything(), as.factor),
-         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE))
+         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE),
+         text = rep(c(rep(c("1","", "", ""), 3), rep(c("2","", "", ""), 2)),2))
 # plot
@@
   y = rownr,
-  fill = value,
-  type = pat
+  color = value,
+  type = pat,
+  width = 0.8, 
+  height = 0.8,
+  label = text
 )) +
-  geom_tile(color = "black") +
-  scale_y_discrete(labels = c(2, 2, 1, 1, 1)) +
-  scale_fill_manual(values = plot_col, name = "Datapoint") +
-  facet_wrap( ~ pat, ncol = 2) +
+  geom_tile(fill = "white", size = 2) +
+  geom_text() +
+  scale_x_discrete(position = "top") +
+  scale_y_discrete(labels = 5:1) +
+  scale_color_manual(values = plot_col, name = "Datapoint") +
+  facet_wrap( ~ pat, ncol = 2, strip.position = "bottom") +
   theme_minimal() +
-  theme(legend.position = "bottom") +
-  labs(x = "Variable", y = "Cluster")
+  theme(
+    legend.position = "none", 
+    panel.grid.major = element_blank()) +
+  labs(x = "", y = "Observation (e.g., patient)")
+# TODO: add patient nr and cluster
+# TODO: add grid.lines = "none"
+
 ```
@@
 We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).
+-> ask whether we can use the Heckman repo data or simulate data ourselves
+
+Heckman options:
+
+- leiden85
+
+- GJRM::hiv (https://rdrr.io/github/egeminiani/GJRM/man/hiv.html)
+
+- simulating
+
+- IMPACT
 
@@
 plot_md_pat(impact)
+# TODO: rotate variable labels
 ```
 -> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf), the data is already imputed (Steyerberg et al, 2008).
+
+- MAR miss varying by cluster.
+Obs data patt differ per cluster.
+E.g., in cluster 1 miss depends on age but not in cluster two.
+Split the dataframe and run `ampute()` on each cluster.
+-> TODO: also make MNAR missingness to heckman model.
+Maybe based on `ct` variable?
+Inclusion-selection variable.
+-> otherwise: use `leiden85` data on blood pressure with MNAR.
+Then run cox regression like the boshuizen article but with living situation as clusters.
+-> TODO: get analyses from https://www.gerkovink.com/mimp/Contents/Exercises/Day%203%20-%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html.
+
+- ADD: `multilevel_ampute()` wrapper function in `mice`.
 
@@
 ```
-- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`.
+- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would be wise to use multilevel modeling.
 
@@
 plot_md_pat(pop)
+# TODO: add pattern by cluster, text mining example square size
+# TODO: try to make it a shepley plot 
 ```
 
-- Does the missing data of popular depend on popteach?
-One could for example check this by making a histogram of popteach separately for the pupils with known popularity and missing popularity.
-We do see that the histogram for the missing popular (TRUE) is further to the right than the histogram for observed popular (FALSE).
-This would indicate a right-tailed MAR missingness.
-In fact this is exactly what happens, because we created the missingness in these data ourselves.
-But we can make it observable by examining the relations between the missingness in popular and the observed data in popteach.
-
-- Does the missingness in teacher popularity depend on pupil popularity? Yes: there is a dependency. The relation seems to be right-tailed.
+- Does the missing data of `popular` depend on `popteach`?
+One could for example check this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity.
+And the other way around: Does the missingness in teacher popularity depend on pupil popularity?
 
@@
 plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
+# TODO: check if up-side-down plot works
+# or maybe overlapping densities
+# TODO: check smoothing in geom_density function and make it the inverse of the sample size
+# TODO: add functions to mice
 ```
+
+-  We do see that the histogram for the missing `popular` is further to the right than the histogram for observed `popular`.
+This would indicate a right-tailed MAR missingness.
+In fact this is exactly what happens, because we created the missingness in these data ourselves.
+But we made it observable by examining the relations between the missingness in popular and the observed data in `popteach`.
+There is also a dependency between the missingness in teacher popularity and pupil popularity.
+The relation seems to be right-tailed.
+
+## Amputation
+
+-
 
@@
 Sometimes there's little reason to assume some variable is affected by heterogeneity.
--> Refer to Meng, Vincent, and a paper by Grund on congeniality and random slopes.
+-> Refer to @meng94, an Audigier paper, and a paper by Grund on congeniality and random slopes.
 
-- Step 0: As predictor
+- Step 0: As predictor + CCA to scare off users
 
@@
 
-## Step 1-3
+## Step 1-3 + MNAR
 

### 2021-11-22 Manuscript/Manuscript.tex
Update md pattern, add new case study

@@
 missing at random (MAR) or missing not at random (MNAR)).
+-\textgreater{} ask whether we can use the Heckman repo data or simulate
+data ourselves
+
+Heckman options:
+
+\begin{itemize}
+\item
+  leiden85
+\item
+  GJRM::hiv
+\item
+  simulating
+\item
+  IMPACT
+\end{itemize}
 
@@
 
+\begin{itemize}
+\item
+  MAR miss varying by cluster. Obs data patt differ per cluster. E.g.,
+  in cluster 1 miss depends on age but not in cluster two. Split the
+  dataframe and run \texttt{ampute()} on each cluster. -\textgreater{}
+  TODO: also make MNAR missingness to heckman model. Maybe based on
+  \texttt{ct} variable? Inclusion-selection variable. -\textgreater{}
+  otherwise: use \texttt{leiden85} data on blood pressure with MNAR.
+  Then run cox regression like the boshuizen article but with living
+  situation as clusters. -\textgreater{} TODO: get analyses from
+  \url{https://www.gerkovink.com/mimp/Contents/Exercises/Day\%203\%20-\%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html}.
+\item
+  ADD: \texttt{multilevel\_ampute()} wrapper function in \texttt{mice}.
+\end{itemize}
+
 \hypertarget{case-study-ii-popularity}{%
@@
   What are the ICCs? For \texttt{popular} the ICC is 0.33. For
-  \texttt{popteach} it is 0.31.
+  \texttt{popteach} it is 0.31. It would be wise to use multilevel
+  modeling.
 \item
@@
 \begin{itemize}
+\tightlist
 \item
-  Does the missing data of popular depend on popteach? One could for
-  example check this by making a histogram of popteach separately for
-  the pupils with known popularity and missing popularity. We do see
-  that the histogram for the missing popular (TRUE) is further to the
-  right than the histogram for observed popular (FALSE). This would
-  indicate a right-tailed MAR missingness. In fact this is exactly what
-  happens, because we created the missingness in these data ourselves.
-  But we can make it observable by examining the relations between the
-  missingness in popular and the observed data in popteach.
-\item
-  Does the missingness in teacher popularity depend on pupil popularity?
-  Yes: there is a dependency. The relation seems to be right-tailed.
+  Does the missing data of \texttt{popular} depend on \texttt{popteach}?
+  One could for example check this by making a histogram of
+  \texttt{popteach} separately for the pupils with known popularity and
+  missing popularity. And the other way around: Does the missingness in
+  teacher popularity depend on pupil popularity?
 \end{itemize}
@@
 \end{CodeChunk}
+
+\begin{itemize}
+\tightlist
+\item
+  We do see that the histogram for the missing \texttt{popular} is
+  further to the right than the histogram for observed \texttt{popular}.
+  This would indicate a right-tailed MAR missingness. In fact this is
+  exactly what happens, because we created the missingness in these data
+  ourselves. But we made it observable by examining the relations
+  between the missingness in popular and the observed data in
+  \texttt{popteach}. There is also a dependency between the missingness
+  in teacher popularity and pupil popularity. The relation seems to be
+  right-tailed.
+\end{itemize}
+
+\hypertarget{amputation}{%
+\subsection{Amputation}\label{amputation}}
+
+\begin{itemize}
+\tightlist
+\item
+\end{itemize}
 
@@
   little reason to assume some variable is affected by heterogeneity.
-  -\textgreater{} Refer to Meng, Vincent, and a paper by Grund on
-  congeniality and random slopes.
+  -\textgreater{} Refer to \citet{meng94}, an Audigier paper, and a
+  paper by Grund on congeniality and random slopes.
 \item
-  Step 0: As predictor
+  Step 0: As predictor + CCA to scare off users
 \item
@@
 
-\hypertarget{step-1-3}{%
-\subsection{Step 1-3}\label{step-1-3}}
+\hypertarget{step-1-3-mnar}{%
+\subsection{Step 1-3 + MNAR}\label{step-1-3-mnar}}
 

### 2021-11-26 Manuscript/Manuscript.Rmd
Add steps from mice vignette up-to multilevel models

@@
 # load data
-pop <- readRDS("../Data/popNCR.RDS")
+pop <- readRDS("Data/popNCR.RDS")
 # ICCs
@@
 - Does the missing data of `popular` depend on `popteach`?
-One could for example check this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity.
-And the other way around: Does the missingness in teacher popularity depend on pupil popularity?
+Does the missingness in teacher popularity depend on pupil popularity?
+-> Check this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity, and the other way around.
 
@@
 # distributions
-plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
-plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
+# plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
+# plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
 # TODO: check if up-side-down plot works
@@
 # TODO: add functions to mice
+
+miss_col <- mice:::mdc(1:2) %>% setNames(c("FALSE", "TRUE"))
+
+# x = "popular"
+# z = "popteach"
+plot_conditional <- function(data, x, z, ...){
+  ggplot(data, aes(x = get(x), color = !is.na(get(z)))) +
+  geom_density() +
+  scale_color_manual(values = miss_col, name = paste("Missing", z)) +
+  theme_classic() +
+  theme(legend.position = "bottom") +
+  labs(x = x)
+}
+plot_conditional(pop, x = "popular", z = "popteach")
+plot_conditional(pop, x = "popteach", z = "popular")
 ```
 
--  We do see that the histogram for the missing `popular` is further to the right than the histogram for observed `popular`.
+-  We do see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
 This would indicate a right-tailed MAR missingness.
@@
 There is also a dependency between the missingness in teacher popularity and pupil popularity.
-The relation seems to be right-tailed.
+The relation seems to be right-tailed as well.
+
+- We can impute the missingness the 'standard' way, ignoring the multilevel structure of the data. This is surely invalid, given the high ICCs, but we'll do it anyways.
+
+- We'll use Bayesian linear regression to impute the continuous variables, and logistic regression to impute the binary variable `sex`.
+We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.
+
+```{r pop-ignore}
+# dry run to get imputation parameters
+ini <- mice(pop, maxit = 0)
+
+# extract imputation methods and adjust
+(meth <- ini$meth)
+meth[c("extrav", "texp", "popular", "popteach")] <- "norm"
+
+# extract predictor matrix and adjust
+(pred <- ini$pred)
+pred[, c("class", "pupil")] <- 0
+
+# impute the data, ignoring the cluster structure
+imp_ignore <- mice(pop, m = 10, meth = meth, pred = pred, print = FALSE)
+
+# check convergence of the imputation model
+plot(imp_ignore)
+
+# compare descriptives before and after imputation
+psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+psych::describe(mice::complete(imp_ignore))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
+
+# compare ICCs before and after imputation
+data.frame(vars = names(pop[c(6, 7, 5)]), 
+           observed = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+                        multilevel::ICC1(aov(popteach ~ class, pop)), 
+                        multilevel::ICC1(aov(texp ~ class, pop))), 
+           norm     = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignore))), 
+                        multilevel::ICC1(aov(popteach ~ class, complete(imp_ignore))), 
+                        multilevel::ICC1(aov(texp ~ class, complete(imp_ignore)))))
+
+# further inspection of the imputations
+densityplot(imp_ignore)
+```
+
+- As the ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `class`.
+This tells us that the multilevel structure of the data should be taken into account.
+If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.
+
+- We can also observe that the teacher experience increases slightly after imputation.
+This is due to the MNAR missingness in `texp`.
+Higher values for `texp` have a larger probability to be missing.
+This may not a problem, however, if at least one pupil in each class has teacher experience recorded, we can deductively impute the correct (i.e.
+true) value for every pupil in the class.
+
+- We'll now use `class` as a predictor to impute all other variables.
+
+```{r pop-predictor}
+# adjust the predictor matrix
+pred <- ini$pred 
+pred[, "pupil"] <- 0
+
+# impute the data, cluster as predictor
+imp_predictor <- mice(pop, m = 10, meth = meth, pred = pred, print = FALSE)
+
+# check convergence of the imputation model
+plot(imp_predictor)
+
+# compare descriptives before and after imputation
+psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
+
+# compare ICCs before and after imputation
+data.frame(vars = names(pop[c(6, 7, 5)]), 
+           observed = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+                        multilevel::ICC1(aov(popteach ~ class, pop)), 
+                        multilevel::ICC1(aov(texp ~ class, pop))), 
+           norm     = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
+                        multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
+                        multilevel::ICC1(aov(texp ~ class, complete(imp_predictor)))))
+
+# further inspection of the imputations
+densityplot(imp_predictor)
+```
+
+- Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MNAR.
 

### 2021-11-26 R/plot_md.R

@@ -167,3 +167,4 @@ plot_NA_cond <- function(dat, x, z, bins = NULL) {
   return(p)
 }
 
+# plot conditional densities

### 2021-11-29 Manuscript/Manuscript.Rmd
Feedback on figures (md pattern, conditinal distributions, distributions after imputation)

--- 2021-11-22 Manuscript/Manuscript.Rmd
@@
   labs(x = "", y = "Observation (e.g., patient)")
-# TODO: add patient nr and cluster
-# TODO: add grid.lines = "none"
+# TODO: make it one plot with sporadic and systematic
+# TODO: participant/response/*sample units*/records instead of obs on the y axis
 
@@
 # TODO: rotate variable labels
+
+# TODO: check md pattern in clusters, should differ, otherwise create new missingness
+# MAYBE: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters instead
+# TODO: e-mail Joop for DGM or look at adrew and jennifer book for data
+# sim.multi from psych
 ```
@@
 # TODO: add functions to mice
+
+miss_col <- mice:::mdc(1:2) %>% setNames(c("FALSE", "TRUE"))
+
+# x = "popular"
+# z = "popteach"
+plot_conditional <- function(data, x, z, cluster, ...){
+  ggplot(data, aes(x = get(x), color = !is.na(get(z)))) + # group = get(cluster)
+  geom_density(alpha = 0.01) +
+  scale_color_manual(values = miss_col, name = z) +
+  theme_classic() +
+  theme(legend.position = "bottom") +
+  labs(x = x)
+}
+plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
+plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
+# TODO: pooteach obs/mis
+# TODO: add title
+# TODO: add facets for some clusters
+# or add propensity score distribution `is.na(popular) ~ .`
+# TODO: make it average of cluster dens, not marginal
 ```
 
--  We do see that the histogram for the missing `popular` is further to the right than the histogram for observed `popular`.
+-  We do see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
 This would indicate a right-tailed MAR missingness.
@@
 There is also a dependency between the missingness in teacher popularity and pupil popularity.
-The relation seems to be right-tailed.
+The relation seems to be right-tailed as well.
+
+- We can impute the missingness the 'standard' way, ignoring the multilevel structure of the data. This is surely invalid, given the high ICCs, but we'll do it anyways.
+
+- We'll use predictive mean matching to impute the continuous variables (some appear to be somewhat ordinal), and logistic regression to impute the binary variable `sex`.
+We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.
+
+```{r pop-ignored}
+# dry run to get imputation parameters
+ini <- mice(pop, maxit = 0)
+
+# extract predictor matrix and adjust
+pred <- ini$pred
+pred[, c("class", "pupil")] <- 0
+pred
+
+# impute the data, ignoring the cluster structure
+imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+
+# check convergence of the imputation model
+plot(imp_ignored)
+
+# compare descriptives before and after imputation
+psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
+# TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
+# TODO: pool mean median and sd
+# feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
+# TODO: add FMI for each of the estimates? at least for the mean
+
+# further inspection of the imputations
+densityplot(imp_ignored)
+
+# compare ICCs before and after imputation
+ICCs <- data.frame(
+  vars = c("popular", "popteach", "texp"), 
+  incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+               multilevel::ICC1(aov(popteach ~ class, pop)),
+               multilevel::ICC1(aov(texp ~ class, pop))), 
+  ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
+              multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
+              multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
+  )
+ICCs
+
+```
+
+- As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `class`.
+This tells us that the multilevel structure of the data should be taken into account.
+If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.
+
+- We can also observe that the teacher experience increases slightly after imputation.
+This is due to the MNAR missingness in `texp`.
+Higher values for `texp` have a larger probability to be missing.
+This may not a problem, however, if at least one pupil in each class has teacher experience recorded, we can deductively impute the correct (i.e.
+true) value for every pupil in the class.
+
+- We'll now use `class` as a predictor to impute all other variables.
+
+```{r pop-predictor}
+# adjust the predictor matrix
+pred <- ini$pred 
+pred[, "pupil"] <- 0
+pred
+
+# impute the data, cluster as predictor
+imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+
+# check logged events
+head(imp_predictor$loggedEvents)
+## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
+
+# check convergence of the imputation model
+plot(imp_predictor)
+
+# compare descriptives before and after imputation
+psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
+
+# further inspection of the imputations
+densityplot(imp_predictor)
+
+# compare ICCs before and after imputation
+ICCs <- ICCs %>% cbind(
+           predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
+                        multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
+                        multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
+           )
+ICCs
+```
+
+- Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MNAR.
+
+- The ICCs are way more in line with the ICCs in the incomplete data. But this is a quick and dirty way of imputing multilevel data. We *should* be using a multilevel model.
 

### 2021-11-29 Manuscript/Manuscript.tex
Add steps from mice vignette up-to multilevel models
Feedback on figures (md pattern, conditinal distributions, distributions after imputation)

@@
 \item
-  GJRM::hiv
+  GJRM::hiv (\url{https://rdrr.io/github/egeminiani/GJRM/man/hiv.html})
 \item
@@
   Does the missing data of \texttt{popular} depend on \texttt{popteach}?
-  One could for example check this by making a histogram of
-  \texttt{popteach} separately for the pupils with known popularity and
-  missing popularity. And the other way around: Does the missingness in
-  teacher popularity depend on pupil popularity?
+  Does the missingness in teacher popularity depend on pupil popularity?
+  -\textgreater{} Check this by making a histogram of \texttt{popteach}
+  separately for the pupils with known popularity and missing
+  popularity, and the other way around.
 \end{itemize}
@@
 \begin{itemize}
-\tightlist
 \item
-  We do see that the histogram for the missing \texttt{popular} is
-  further to the right than the histogram for observed \texttt{popular}.
-  This would indicate a right-tailed MAR missingness. In fact this is
-  exactly what happens, because we created the missingness in these data
-  ourselves. But we made it observable by examining the relations
-  between the missingness in popular and the observed data in
-  \texttt{popteach}. There is also a dependency between the missingness
-  in teacher popularity and pupil popularity. The relation seems to be
-  right-tailed.
+  We do see that the distribution for the missing \texttt{popular} is
+  further to the right than the distribution for observed
+  \texttt{popular}. This would indicate a right-tailed MAR missingness.
+  In fact this is exactly what happens, because we created the
+  missingness in these data ourselves. But we made it observable by
+  examining the relations between the missingness in popular and the
+  observed data in \texttt{popteach}. There is also a dependency between
+  the missingness in teacher popularity and pupil popularity. The
+  relation seems to be right-tailed as well.
+\item
+  We can impute the missingness the `standard' way, ignoring the
+  multilevel structure of the data. This is surely invalid, given the
+  high ICCs, but we'll do it anyways.
+\item
+  We'll use predictive mean matching to impute the continuous variables
+  (some appear to be somewhat ordinal), and logistic regression to
+  impute the binary variable \texttt{sex}. We do not use the observation
+  identifier \texttt{pupil} or cluster identifier \texttt{class} as
+  predictors to impute other variables.
+\end{itemize}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # dry run to get imputation parameters
+R> ini <- mice(pop, maxit = 0)
+R> 
+R> # extract predictor matrix and adjust
+R> pred <- ini$pred
+R> pred[, c("class", "pupil")] <- 0
+R> pred 
+\end{CodeInput}
+\begin{CodeOutput}
+         pupil class extrav sex texp popular popteach
+pupil        0     0      1   1    1       1        1
+class        0     0      1   1    1       1        1
+extrav       0     0      0   1    1       1        1
+sex          0     0      1   0    1       1        1
+texp         0     0      1   1    0       1        1
+popular      0     0      1   1    1       0        1
+popteach     0     0      1   1    1       1        0
+\end{CodeOutput}
+\begin{CodeInput}
+R> # impute the data, ignoring the cluster structure
+R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+R> 
+R> # check convergence of the imputation model
+R> plot(imp_ignored)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-1} \end{center}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-2} \end{center}
+
+\begin{CodeInput}
+R> # compare descriptives before and after imputation
+R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+\end{CodeInput}
+\begin{CodeOutput}
+            n  mean median min   max    sd
+pupil    2000 10.65   11.0   1  26.0  5.97
+class*   2000 50.37   51.0   1 100.0 29.08
+extrav   1484  5.31    5.0   1  10.0  1.29
+sex*     1504  1.56    2.0   1   2.0  0.50
+texp     1024 11.80   12.0   2  25.0  6.26
+popular  1490  4.83    4.8   0   9.1  1.34
+popteach 1472  4.83    5.0   1  10.0  1.36
+\end{CodeOutput}
+\begin{CodeInput}
+R> psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
+\end{CodeInput}
+\begin{CodeOutput}
+            n  mean median min   max    sd
+pupil    2000 10.65     11   1  26.0  5.97
+class*   2000 50.37     51   1 100.0 29.08
+extrav   2000  5.24      5   1  10.0  1.26
+sex*     2000  1.51      2   1   2.0  0.50
+texp     2000 12.63     13   2  25.0  6.27
+popular  2000  5.01      5   0   9.1  1.37
+popteach 2000  5.02      5   1  10.0  1.39
+\end{CodeOutput}
+\begin{CodeInput}
+R> # further inspection of the imputations
+R> densityplot(imp_ignored)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-3} \end{center}
+
+\begin{CodeInput}
+R> # compare ICCs before and after imputation
+R> ICCs <- data.frame(
++   vars = c("popular", "popteach", "texp"), 
++   incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
++                multilevel::ICC1(aov(popteach ~ class, pop)),
++                multilevel::ICC1(aov(texp ~ class, pop))), 
++   ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
++               multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
++               multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
++   )
+R> ICCs
+\end{CodeInput}
+\begin{CodeOutput}
+      vars incomplete   ignored
+1  popular  0.3280070 0.2743092
+2 popteach  0.3138658 0.2479930
+3     texp  1.0000000 0.4510366
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{itemize}
+\item
+  As the original ICCs show, 100\% of the variance in \texttt{texp} can
+  be attributed to the clustering variable \texttt{class}. This tells us
+  that the multilevel structure of the data should be taken into
+  account. If we don't, we'll end up with incorrect imputations, biasing
+  the effect of the clusters towards zero.
+\item
+  We can also observe that the teacher experience increases slightly
+  after imputation. This is due to the MNAR missingness in
+  \texttt{texp}. Higher values for \texttt{texp} have a larger
+  probability to be missing. This may not a problem, however, if at
+  least one pupil in each class has teacher experience recorded, we can
+  deductively impute the correct (i.e.~true) value for every pupil in
+  the class.
+\item
+  We'll now use \texttt{class} as a predictor to impute all other
+  variables.
+\end{itemize}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # adjust the predictor matrix
+R> pred <- ini$pred 
+R> pred[, "pupil"] <- 0
+R> pred
+\end{CodeInput}
+\begin{CodeOutput}
+         pupil class extrav sex texp popular popteach
+pupil        0     1      1   1    1       1        1
+class        0     0      1   1    1       1        1
+extrav       0     1      0   1    1       1        1
+sex          0     1      1   0    1       1        1
+texp         0     1      1   1    0       1        1
+popular      0     1      1   1    1       0        1
+popteach     0     1      1   1    1       1        0
+\end{CodeOutput}
+\begin{CodeInput}
+R> # impute the data, cluster as predictor
+R> imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+\end{CodeInput}
+\begin{CodeOutput}
+Warning: Number of logged events: 190
+\end{CodeOutput}
+\begin{CodeInput}
+R> # check logged events
+R> head(imp_predictor$loggedEvents)
+\end{CodeInput}
+\begin{CodeOutput}
+  it im      dep meth  out
+1  1  1  popular  pmm texp
+2  1  1 popteach  pmm texp
+3  1  2  popular  pmm texp
+4  1  2 popteach  pmm texp
+5  1  3  popular  pmm texp
+6  1  3 popteach  pmm texp
+\end{CodeOutput}
+\begin{CodeInput}
+R> ## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
+R> 
+R> # check convergence of the imputation model
+R> plot(imp_predictor)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-1} \end{center}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-2} \end{center}
+
+\begin{CodeInput}
+R> # compare descriptives before and after imputation
+R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+\end{CodeInput}
+\begin{CodeOutput}
+            n  mean median min   max    sd
+pupil    2000 10.65   11.0   1  26.0  5.97
+class*   2000 50.37   51.0   1 100.0 29.08
+extrav   1484  5.31    5.0   1  10.0  1.29
+sex*     1504  1.56    2.0   1   2.0  0.50
+texp     1024 11.80   12.0   2  25.0  6.26
+popular  1490  4.83    4.8   0   9.1  1.34
+popteach 1472  4.83    5.0   1  10.0  1.36
+\end{CodeOutput}
+\begin{CodeInput}
+R> psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
+\end{CodeInput}
+\begin{CodeOutput}
+            n  mean median min   max    sd
+pupil    2000 10.65   11.0   1  26.0  5.97
+class*   2000 50.37   51.0   1 100.0 29.08
+extrav   2000  5.20    5.0   1  10.0  1.28
+sex*     2000  1.52    2.0   1   2.0  0.50
+texp     2000 14.26   15.0   2  25.0  6.55
+popular  2000  5.05    5.1   0   9.1  1.38
+popteach 2000  5.03    5.0   1  10.0  1.39
+\end{CodeOutput}
+\begin{CodeInput}
+R> # further inspection of the imputations
+R> densityplot(imp_predictor)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-3} \end{center}
+
+\begin{CodeInput}
+R> # compare ICCs before and after imputation
+R> ICCs <- ICCs %>% cbind(
++            predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
++                         multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
++                         multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
++            )
+R> ICCs
+\end{CodeInput}
+\begin{CodeOutput}
+      vars incomplete   ignored predictor
+1  popular  0.3280070 0.2743092 0.3694899
+2 popteach  0.3138658 0.2479930 0.3335612
+3     texp  1.0000000 0.4510366 1.0000000
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{itemize}
+\item
+  Now, we can clearly see that the imputed values of \texttt{texp} are
+  higher than the observed values, which is in line with right-tailed
+  MNAR.
+\item
+  The ICCs are way more in line with the ICCs in the incomplete data.
+  But this is a quick and dirty way of imputing multilevel data. We
+  \emph{should} be using a multilevel model.
 \end{itemize}

### 2021-11-29 Manuscript/Manuscript.Rmd
Make full sentences from bullets in first part introduction. Edit and add explanation missingness figure.

@@
 miceadds::source.all(path = "../R")
+icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
 
@@
 
-- What is clustering/multilevel data? In this paper, we discuss grouped observations, not longitudinal data (within-patient clustering).  -> ADD: timeseries also in Discussion section.
+We talk of multilevel data when there is some kind of hierarchy or clustering in a dataset.
+In the typical case, individuals are nested within groups, but there are many different types of multilevel data.
+In the medical field clustering occurs at e.g., the hospitals/center level in registry data, or at the study-level in meta-analyses (IPDMA).
+In the social sciences and official statistics we can find clustering e.g.
+at the country-level, or as imposed by the sampling design.
+In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.
+For reasons of brevity, we only discuss clustering between units, not longitudinal data (within-unit clustering).
 
-- What do we mean by clustering?
-In the medical field: Clustering by studies (IPDMA), hospitals in registries, multi-center studies etc.
-In other fields: e.g.
-official stats clustering at country-level, or social sciences clustering at school-level (related to the sampling design).
-
-- What is heterogeneity? I.e. variability within studies vs. variability between studies
-
-- What does multilevel data look like?
-ADD: figure to show difference between patient-level datapoints vs cluster-level datapoints.
-Maybe also add different data frame formats (or just explain in text that there's long and wide formats).
-
-- What methods are required to analyze multilevel data?	Add references, e.g.
+Multilevel data requires special care when doing any sort of analysis.
+The cluster to which a unit belongs may influence the unit-level observations, and since clusters are made up of units, clusters depend on units as well [@hox17].
+[Explain ICC here?
+The percentage of variance attributed to the cluster-level is expressed by the intra-class coefficient (ICC).
+The ICC can also be interpreted as the expected correlation between two randomly sampled units in same cluster.] These relations can and should be taken into account when developing analysis models for multilevel data.
+At least, such models include a separate intercept term for each cluster.
+But there may also be random predictor effects and/or random error terms (residual error variances), see e.g.
 @hox17 and @jong21.
-At least explain difference random effects for intercept term, predictor effects, and/or variance residual error.
+Heterogeneity refers to variability within clusters vs.
+variability between clusters.
+There are many names for models that take clustering into account.
+Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
 
@@
 Systematic vs sporadic missingness, see @resc13.
-ADD: visualization of systematic vs sporadic missingness.
-Within systematic we have always missing (same value per cluster) and non-measured variables (may differ per patient).
-TODO: adjust md pattern to match text.
--> syst may vary or same for all patients (observations/participants).
+Within systematic we have two flavors: unobserved constants (same value per cluster) and non-measured random variables (which may differ per unit within clusters).
+In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 2 clusters, and 3 variables of interest.
+Variable `X1` is completely observed.
+Variable `X2` is systematically missing, `X3` is sporadically missing.
+The unobserved value for units 4 and 5 on variable `X2` may be the same or different, defining which type of systematic missingness is happening.
 
 ```{r patterns, echo=FALSE}
-# monotone vs non-monotone
-patterns <- rbind(
-  tibble(
-    pat = rep("systematic", 5),
+# missingness indicator plot (monotone vs non-monotone)
+miss_ind <- tibble(
     rownr = 5:1,
-    # obs = rep("observed", 5),
+    unit = rep("observed", 5),
     cluster = rep("observed", 5),
@@
     X2 = c(rep("observed", 3), rep("missing", 2)),
-    X3 = c(rep("missing", 3),  rep("observed", 2))
-  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3)),
-  tibble(
-    pat = rep("sporadic", 5),
-    rownr = 5:1,
-    # obs = rep("observed", 5),
-    cluster = rep("observed", 5),
-    X1 = c("observed", "missing", rep("observed", 3)),
-    X2 = c(rep("observed", 2), rep("missing", 3)),
-    X3 = c(rep("missing", 3),  rep("observed", 2))
-  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3))
-) %>%
-  mutate(across(everything(), as.factor),
-         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE),
-         text = rep(c(rep(c("1","", "", ""), 3), rep(c("2","", "", ""), 2)),2))
+    X3 = c("missing", rep("observed", 3), "missing")
+  ) %>% pivot_longer(cols = c(unit, cluster, X1, X2, X3))
+
+# add text 
+miss_ind$text <- ""
+miss_ind[miss_ind$name == "unit", "text"] <- as.character(1:5)
+miss_ind[miss_ind$name == "cluster", "text"] <- as.character(c(1,1,1,2,2))
+
 # plot
-ggplot(patterns, aes(
+miss_ind %>% 
+  mutate(name = factor(name, levels = unique(name))) %>% 
+ggplot(aes(
   x = name,
@@
   color = value,
-  type = pat,
   width = 0.8, 
@@
   geom_tile(fill = "white", size = 2) +
-  geom_text() +
+  geom_text(show.legend = FALSE) +
   scale_x_discrete(position = "top") +
   scale_y_discrete(labels = 5:1) +
-  scale_color_manual(values = plot_col, name = "Datapoint") +
-  facet_wrap( ~ pat, ncol = 2, strip.position = "bottom") +
+  scale_color_manual(values = plot_col, name = "") +
   theme_minimal() +
   theme(
-    legend.position = "none", 
+    legend.position = "bottom", 
     panel.grid.major = element_blank()) +
-  labs(x = "", y = "Observation (e.g., patient)")
+  labs(x = "", y = "")
 # TODO: make it one plot with sporadic and systematic
@@
 ```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
-# load data
-pop <- readRDS("../Data/popNCR.RDS")
+# load data and merge classes into schools for this tutorial -> but this cannot work because how texp and popteach are defined...
+pop <- readRDS("../Data/popNCR.RDS") %>% mutate(school = round(0.5 + as.numeric(class)/10))
+
 # ICCs
@@
 
-miss_col <- mice:::mdc(1:2) %>% setNames(c("FALSE", "TRUE"))
+plot_col <- mice:::mdc(c("obs", "mis")) %>% setNames(c("observed", "missing"))
 
@@
 plot_conditional <- function(data, x, z, cluster, ...){
-  ggplot(data, aes(x = get(x), color = !is.na(get(z)))) + # group = get(cluster)
-  geom_density(alpha = 0.01) +
-  scale_color_manual(values = miss_col, name = z) +
+  ggplot(data, aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
+  geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
+  geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
+  geom_density() +
+  scale_color_manual(values = plot_col, name = z) +
   theme_classic() +
@@
 }
+
 plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
@@
 
-- Timeseries: and polynomial relationship in the clustering.
+- More complex data types: timeseries and polynomial relationship in the clustering.
 

### 2021-11-29 Manuscript/Manuscript.tex
Make full sentences from bullets in first part introduction. Edit and add explanation missingness figure.

@@
 
-\begin{itemize}
-\item
-  What is clustering/multilevel data? In this paper, we discuss grouped
-  observations, not longitudinal data (within-patient clustering).
-  -\textgreater{} ADD: timeseries also in Discussion section.
-\item
-  What do we mean by clustering? In the medical field: Clustering by
-  studies (IPDMA), hospitals in registries, multi-center studies etc. In
-  other fields: e.g.~official stats clustering at country-level, or
-  social sciences clustering at school-level (related to the sampling
-  design).
-\item
-  What is heterogeneity? I.e. variability within studies vs.~variability
-  between studies
-\item
-  What does multilevel data look like? ADD: figure to show difference
-  between patient-level datapoints vs cluster-level datapoints. Maybe
-  also add different data frame formats (or just explain in text that
-  there's long and wide formats).
-\item
-  What methods are required to analyze multilevel data? Add references,
-  e.g. \citet{hox17} and \citet{jong21}. At least explain difference
-  random effects for intercept term, predictor effects, and/or variance
-  residual error.
-\end{itemize}
+We talk of multilevel data when there is some kind of hierarchy or
+clustering in a dataset. In the typical case, individuals are nested
+within groups, but there are many different types of multilevel data. In
+the medical field clustering occurs at e.g., the hospitals/center level
+in registry data, or at the study-level in meta-analyses (IPDMA). In the
+social sciences and official statistics we can find clustering e.g.~at
+the country-level, or as imposed by the sampling design. In this paper,
+we will refer to the grouping variable as `cluster', and the grouped
+variable as `(sample) unit'. For reasons of brevity, we only discuss
+clustering between units, not longitudinal data (within-unit
+clustering).
+
+Multilevel data requires special care when doing any sort of analysis.
+The cluster to which a unit belongs may influence the unit-level
+observations, and since clusters are made up of units, clusters depend
+on units as well \citep{hox17}. {[}Explain ICC here? The percentage of
+variance attributed to the cluster-level is expressed by the intra-class
+coefficient (ICC). The ICC can also be interpreted as the expected
+correlation between two randomly sampled units in same cluster.{]} These
+relations can and should be taken into account when developing analysis
+models for multilevel data. At least, such models include a separate
+intercept term for each cluster. But there may also be random predictor
+effects and/or random error terms (residual error variances), see e.g.
+\citet{hox17} and \citet{jong21}. Heterogeneity refers to variability
+within clusters vs.~variability between clusters. There are many names
+for models that take clustering into account. Some popular examples are
+`multilevel models', `hierarchical models', `mixed effect models' and
+`random effect models'.
 
@@
   How can we categorize this? Systematic vs sporadic missingness, see
-  \citet{resc13}. ADD: visualization of systematic vs sporadic
-  missingness. Within systematic we have always missing (same value per
-  cluster) and non-measured variables (may differ per patient). TODO:
-  adjust md pattern to match text. -\textgreater{} syst may vary or same
-  for all patients (observations/participants).
+  \citet{resc13}. Within systematic we have two flavors: unobserved
+  constants (same value per cluster) and non-measured random variables
+  (which may differ per unit within clusters). In figure 1, we show a
+  dataset with units in the rows and variables in the columns, there are
+  5 units nested within 2 clusters, and 3 variables of interest.
+  Variable \texttt{X1} is completely observed. Variable \texttt{X2} is
+  systematically missing, \texttt{X3} is sporadically missing. The
+  unobserved value for units 4 and 5 on variable \texttt{X2} may be the
+  same or different, defining which type of systematic missingness is
+  happening.
 \end{itemize}
@@
 \begin{CodeOutput}
-         pupil class extrav sex texp popular popteach
-pupil        0     0      1   1    1       1        1
-class        0     0      1   1    1       1        1
-extrav       0     0      0   1    1       1        1
-sex          0     0      1   0    1       1        1
-texp         0     0      1   1    0       1        1
-popular      0     0      1   1    1       0        1
-popteach     0     0      1   1    1       1        0
+         pupil class extrav sex texp popular popteach school
+pupil        0     0      1   1    1       1        1      1
+class        0     0      1   1    1       1        1      1
+extrav       0     0      0   1    1       1        1      1
+sex          0     0      1   0    1       1        1      1
+texp         0     0      1   1    0       1        1      1
+popular      0     0      1   1    1       0        1      1
+popteach     0     0      1   1    1       1        0      1
+school       0     0      1   1    1       1        1      0
 \end{CodeOutput}
@@
 popteach 1472  4.83    5.0   1  10.0  1.36
+school   2000  5.54    6.0   1  10.0  2.89
 \end{CodeOutput}
@@
 class*   2000 50.37     51   1 100.0 29.08
-extrav   2000  5.24      5   1  10.0  1.26
-sex*     2000  1.51      2   1   2.0  0.50
-texp     2000 12.63     13   2  25.0  6.27
-popular  2000  5.01      5   0   9.1  1.37
+extrav   2000  5.25      5   1  10.0  1.27
+sex*     2000  1.53      2   1   2.0  0.50
+texp     2000 12.48     12   2  25.0  6.29
+popular  2000  4.99      5   0   9.1  1.35
 popteach 2000  5.02      5   1  10.0  1.39
+school   2000  5.54      6   1  10.0  2.89
 \end{CodeOutput}
 \begin{CodeInput}
+R> # TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
+R> # TODO: pool mean median and sd
+R> # feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
+R> # TODO: add FMI for each of the estimates? at least for the mean
+R> 
 R> # further inspection of the imputations
@@
       vars incomplete   ignored
-1  popular  0.3280070 0.2743092
-2 popteach  0.3138658 0.2479930
-3     texp  1.0000000 0.4510366
+1  popular  0.3280070 0.2716802
+2 popteach  0.3138658 0.2528468
+3     texp  1.0000000 0.4395296
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-         pupil class extrav sex texp popular popteach
-pupil        0     1      1   1    1       1        1
-class        0     0      1   1    1       1        1
-extrav       0     1      0   1    1       1        1
-sex          0     1      1   0    1       1        1
-texp         0     1      1   1    0       1        1
-popular      0     1      1   1    1       0        1
-popteach     0     1      1   1    1       1        0
+         pupil class extrav sex texp popular popteach school
+pupil        0     1      1   1    1       1        1      1
+class        0     0      1   1    1       1        1      1
+extrav       0     1      0   1    1       1        1      1
+sex          0     1      1   0    1       1        1      1
+texp         0     1      1   1    0       1        1      1
+popular      0     1      1   1    1       0        1      1
+popteach     0     1      1   1    1       1        0      1
+school       0     1      1   1    1       1        1      0
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-Warning: Number of logged events: 190
+Warning: Number of logged events: 325
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-  it im      dep meth  out
-1  1  1  popular  pmm texp
-2  1  1 popteach  pmm texp
-3  1  2  popular  pmm texp
-4  1  2 popteach  pmm texp
-5  1  3  popular  pmm texp
-6  1  3 popteach  pmm texp
+  it im      dep   meth
+1  1  1   extrav    pmm
+2  1  1      sex logreg
+3  1  1     texp    pmm
+4  1  1  popular    pmm
+5  1  1 popteach    pmm
+6  1  1 popteach    pmm
+                                                                                                                                                                                                                                                       out
+1                                                                                                                                                                                                                                                   school
+2                                                                                                                                                                                                                                                   school
+3                                                                                                                                                                                                                                                   school
+4                                                                                                                                                                                                                                             texp, school
+5                                                                                                                                                                                                                                                     texp
+6 mice detected that your data are (nearly) multi-collinear.\nIt applied a ridge penalty to continue calculations, but the results can be unstable.\nDoes your dataset contain duplicates, linear transformation, or factors with unique respondent names?
 \end{CodeOutput}
@@
             n  mean median min   max    sd
-pupil    2000 10.65   11.0   1  26.0  5.97
-class*   2000 50.37   51.0   1 100.0 29.08
-extrav   2000  5.20    5.0   1  10.0  1.28
-sex*     2000  1.52    2.0   1   2.0  0.50
-texp     2000 14.26   15.0   2  25.0  6.55
-popular  2000  5.05    5.1   0   9.1  1.38
-popteach 2000  5.03    5.0   1  10.0  1.39
+pupil    2000 10.65     11   1  26.0  5.97
+class*   2000 50.37     51   1 100.0 29.08
+extrav   2000  6.20      6   1  10.0  2.00
+sex*     2000  1.52      2   1   2.0  0.50
+texp     2000 14.26     15   2  25.0  6.55
+popular  2000  5.03      5   0   9.1  1.35
+popteach 2000  5.03      5   1  10.0  1.38
+school   2000  5.54      6   1  10.0  2.89
 \end{CodeOutput}
@@
       vars incomplete   ignored predictor
-1  popular  0.3280070 0.2743092 0.3694899
-2 popteach  0.3138658 0.2479930 0.3335612
-3     texp  1.0000000 0.4510366 1.0000000
+1  popular  0.3280070 0.2716802 0.3550567
+2 popteach  0.3138658 0.2528468 0.3306035
+3     texp  1.0000000 0.4395296 1.0000000
 \end{CodeOutput}
@@
 \item
-  Timeseries: and polynomial relationship in the clustering.
+  More complex data types: timeseries and polynomial relationship in the
+  clustering.
 \end{itemize}

### 2021-12-02 Manuscript/Manuscript.Rmd
First draft introduction section

@@
 In the typical case, individuals are nested within groups, but there are many different types of multilevel data.
-In the medical field clustering occurs at e.g., the hospitals/center level in registry data, or at the study-level in meta-analyses (IPDMA).
+In the medical field, clustering occurs at e.g., the hospitals/center level in registry data, or at the study-level in meta-analyses (IPDMA).
 In the social sciences and official statistics we can find clustering e.g.
 at the country-level, or as imposed by the sampling design.
-In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.
-For reasons of brevity, we only discuss clustering between units, not longitudinal data (within-unit clustering).
+In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.^[Add that we'll only discuss two levels?] For reasons of brevity, we only discuss clustering between units, not within units (such as in timeseries or longitudinal data).
 
-Multilevel data requires special care when doing any sort of analysis.
+Analyzing multilevel data requires special care, compared to 'regular', single level data.
 The cluster to which a unit belongs may influence the unit-level observations, and since clusters are made up of units, clusters depend on units as well [@hox17].
-[Explain ICC here?
+These relations can and should be taken into account when developing analysis models for multilevel data.^[Explain ICC here?
 The percentage of variance attributed to the cluster-level is expressed by the intra-class coefficient (ICC).
-The ICC can also be interpreted as the expected correlation between two randomly sampled units in same cluster.] These relations can and should be taken into account when developing analysis models for multilevel data.
-At least, such models include a separate intercept term for each cluster.
-But there may also be random predictor effects and/or random error terms (residual error variances), see e.g.
-@hox17 and @jong21.
-Heterogeneity refers to variability within clusters vs.
-variability between clusters.
-There are many names for models that take clustering into account.
+The ICC can also be interpreted as the expected correlation between two randomly sampled units in same cluster.
+So if the ICC is high, a lot of variability in a variable is due to the clustering, which should be modeled accordingly.] Multilevel models typically include separate intercepts for each cluster, which relieves one restriction imposed by single-level models: equal group means across clusters.
+Additionally, there may be random predictor effects and/or random error terms (residual error variances), see e.g.
+@hox17 and @jong21.^[Add that heterogeneity refers to variability within clusters.] There are many names for models that take clustering into account.
 Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
@@
 
-- Why/where does missingness occur in multilevel data? I.e., not only patient-level but also cluster-level.
-
-- How can we categorize this?
-Systematic vs sporadic missingness, see @resc13.
-Within systematic we have two flavors: unobserved constants (same value per cluster) and non-measured random variables (which may differ per unit within clusters).
+Multilevel data is not spared the ubiquitous problem of missing information.
+Just as in single level data, missingness may occur at the unit level.
+But with multiple levels of data comes the potential for missingness at multiple levels.
+Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13.
 In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 2 clusters, and 3 variables of interest.
 Variable `X1` is completely observed.
-Variable `X2` is systematically missing, `X3` is sporadically missing.
-The unobserved value for units 4 and 5 on variable `X2` may be the same or different, defining which type of systematic missingness is happening.
+Variable `X2` is systematically missing, `X3` is sporadically missing.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (same value per cluster) and non-measured random variables (which may differ per unit within clusters).
+In Figure 1, the former implies that the unobserved values for units 4 and 5 on variable `X2` would be equal.
+With the latter, the values may differ.
+Depending on the missing data pattern, there are more or less optimal way of accommodating the missingness.
+
+Ignoring missing data in research endeavors is almost never a good idea.
+Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power.
+Instead, the missingness should be accommodated *before* or *within* the analysis of scientific interest.
+Especially the former is very generic and popular.
+Imputing (i.e., filling in) the missing values splits the missing data problem from the scientific problem.
+The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which solves the missingness one variable at a time, iteratively.
+\pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
+In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.^[Discuss missingness mechanisms before this point, add references @yuce08 and @hox15.]
 
@@
 
--	What kinds of missingness are there? ADD: missingness mechanisms here. See e.g. @yuce08 and @hox15.
+<!-- -	What kinds of missingness are there? ADD: missingness mechanisms here. See e.g. @yuce08 and @hox15. -->
 
-- Why are standard (ad hoc) missing data methods not well suited?
+<!-- - Why are standard (ad hoc) missing data methods not well suited?  -->
 
-- What types of multilevel methods are available?
+<!-- - What types of multilevel methods are available?
 General overview of approaches, see @audi18 and @grun18.
 E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.
+-->
 
-- Additional difficulty that is addressed in this tutorial: MNAR data.
+<!-- - Additional difficulty that is addressed in this tutorial: MNAR data. -->
 
@@
 
-- Provide practical guidelines with code snippets for imputation of incomplete multilevel data.
+This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}.
+We provide practical guidelines and code snippets for different missing data situations.
+For reasons of brevity, we focus on imputation by chained equations, although JOMO is available in \pkg{mice} as well.
+Other useful packages for incomplete multilevel data include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed.
+We're providing an overview of implementations.
+It's up-to the reader to decide which multilevel strategy suits their data.
+So we won't go into detail for the different methods (and equations).
+Refer to @meng94, an Audigier paper, and a paper by Grund on congeniality and random slopes.
+This paper is just a software tutorial.
+We'll keep it practical.]
 
-- We focus on the workflow for conditional modeling (not JOMO) in `mice`. Refer to other packages: `mitml`, `miceadds`, `mdmb`.
+We structure this tutorial around three case studies:
 
-- Case study options: `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s), `mice::popularity` (simulated data on school kids, with MNAR/MAR mixture).
-TODO: Check example data Gelman.
+- `mice::popmis` (simulated data on school kids, with MNAR/MAR mixture);
 
-- Introduce case study and set scope of this tutorial: We're providing an overview of implementations.
-It's up-to the reader to decide which strategy suits their data.
-So we won't go into detail for the different methods (and equations).
-This paper is just a software tutorial.
-We'll keep it practical.
--> ADD: some kind of help function that suggests a suitable predictor matrix to the user, given a certain analysis model.
+- `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s);
+
+- `GJRM::hiv` (simulated patient data on HIV, without `NA`s)
+
+For each case study we focus on a different aspect to illustrate how to impute incomplete multilevel data.
 
@@
 
-We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).
--> ask whether we can use the Heckman repo data or simulate data ourselves
+<!-- We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).
+-> ask whether we can use the Heckman repo data or simulate data ourselves -->
 
-Heckman options:
+<!-- Heckman options: -->
 
-- leiden85
+<!-- - leiden85 -->
 
-- GJRM::hiv (https://rdrr.io/github/egeminiani/GJRM/man/hiv.html)
+<!-- - GJRM::hiv (https://rdrr.io/github/egeminiani/GJRM/man/hiv.html) -->
 
-- simulating
+<!-- - simulating -->
 
-- IMPACT
+<!-- - IMPACT -->
 
-## Case study I: IMPACT
+We introduce three case studies to illustrate the workflow.
+In the `mice::popmis` data, we show the advantages of including the multilevel structure of the data into the imputation model.
+In the `metamisc::impact` data we'll show how to induce missingness and solve it in real-world data.
+In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model.
 
-- What does the data look like?
-`impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
+For each case study we'll look at least at: 1) the incomplete data; 2) the imputation model; 3) the imputed data; and 4) how to obtain pooled estimates for the analysis of scientific interest.
 
-  - `name` Name of the study,
-  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
-  - `age` Age of the patient,
-  - `motor_score` Glasgow Coma Scale motor score,
-  - `pupil` Pupillary reactivity,
-  - `ct` Marshall Computerized Tomography classification,
-  - `hypox` Hypoxia (0=no, 1=yes),
-  - `hypots` Hypotension (0=no, 1=yes),
-  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
-  - `edh` Epidural hematoma (0=no, 1=yes),
-  - `mort` 6-month mortality (0=alive, 1=dead).
+## Case Study I: Popularity
 
-```{r impact, echo=FALSE}
-# load data
-data("impact")
-# # descriptive statistics
-# by(impact, impact$name, summary) 
-# psych::describe(impact)[,c(2:5,8:9)]
-# missingness
-plot_md_pat(impact)
-# TODO: rotate variable labels
-
-# TODO: check md pattern in clusters, should differ, otherwise create new missingness
-# MAYBE: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters instead
-# TODO: e-mail Joop for DGM or look at adrew and jennifer book for data
-# sim.multi from psych
-```
--> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf), the data is already imputed (Steyerberg et al, 2008).
-
-- MAR miss varying by cluster.
-Obs data patt differ per cluster.
-E.g., in cluster 1 miss depends on age but not in cluster two.
-Split the dataframe and run `ampute()` on each cluster.
--> TODO: also make MNAR missingness to heckman model.
-Maybe based on `ct` variable?
-Inclusion-selection variable.
--> otherwise: use `leiden85` data on blood pressure with MNAR.
-Then run cox regression like the boshuizen article but with living situation as clusters.
--> TODO: get analyses from https://www.gerkovink.com/mimp/Contents/Exercises/Day%203%20-%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html.
-
-- ADD: `multilevel_ampute()` wrapper function in `mice`.
-
-## Case Study II: Popularity
-
-- What does the data look like? `popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables:
+`popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables:
 
@@
 
+### The incomplete data
+
 ```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
@@
 ```
-- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would be wise to use multilevel modeling.
 
@@
 ```
+
+- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would be wise to use multilevel modeling.
 
@@
 # TODO: make it average of cluster dens, not marginal
+
+# Feedback: 
+# you only see the extreme ones in the positive sense, especially with 100 clusters
+# alternatively add a quartile line around the density with geom_ribbon
+
 ```
@@
 
-- We can impute the missingness the 'standard' way, ignoring the multilevel structure of the data. This is surely invalid, given the high ICCs, but we'll do it anyways.
+### The imputation model
+
+- We could ignore the multilevel structure of the data. This is likely invalid, given the high ICCs, but we'll do it anyways.
 
@@
 
-## Amputation
+## Case study II: IMPACT
 
--
+`impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
 
-## Modeling choices
+  - `name` Name of the study,
+  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
+  - `age` Age of the patient,
+  - `motor_score` Glasgow Coma Scale motor score,
+  - `pupil` Pupillary reactivity,
+  - `ct` Marshall Computerized Tomography classification,
+  - `hypox` Hypoxia (0=no, 1=yes),
+  - `hypots` Hypotension (0=no, 1=yes),
+  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+  - `edh` Epidural hematoma (0=no, 1=yes),
+  - `mort` 6-month mortality (0=alive, 1=dead).
 
-- Which models will we discuss? We'll build the model to grow in complexity. The final model is the most complex but also the most versatile.
+The data is already imputed (Steyerberg et al, 2008), so we'll induce missingness ourselves.
+For example, MAR missingness varying by cluster.^[Observed data pattern should differ per cluster.
+So, in cluster 1, the missingness would depend on age, but not in cluster two.
+Split the dataframe and run `ampute()` on each cluster.]
 
-- Note on model complexity: Typically, we should at least use random intercepts, but often random slopes as well.
+```{r impact, echo=FALSE}
+# load data
+data("impact")
+# # descriptive statistics
+# by(impact, impact$name, summary) 
+# psych::describe(impact)[,c(2:5,8:9)]
+# missingness
+plot_md_pat(impact)
+# TODO: rotate variable labels
+
+# TODO: check md pattern in clusters, should differ, otherwise create new missingness
+# MAYBE: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters instead
+# TODO: e-mail Joop for DGM or look at adrew and jennifer book for data
+# sim.multi from psych
+```
+<!-- -> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf),  -->
+
+<!-- TODO: also make MNAR missingness to heckman model.
+Maybe based on `ct` variable?
+Inclusion-selection variable.
+-> otherwise: use `leiden85` data on blood pressure with MNAR.
+Then run cox regression like the boshuizen article but with living situation as clusters.
+-> TODO: get analyses from https://www.gerkovink.com/mimp/Contents/Exercises/Day%203%20-%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html.
+-->
+
+<!-- ## Amputation -->
+
+<!-- -  -->
+
+<!-- ## Modeling choices -->
+
+<!-- - Which models will we discuss? We'll build the model to grow in complexity. The final model is the most complex but also the most versatile.  -->
+
+<!-- - Note on model complexity: Typically, we should at least use random intercepts, but often random slopes as well.
 Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeniality, but don't mention the term) ->  Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there.
 Sometimes there's little reason to assume some variable is affected by heterogeneity.
--> Refer to @meng94, an Audigier paper, and a paper by Grund on congeniality and random slopes.
+->  -->
 
-- Step 0: As predictor + CCA to scare off users
+<!-- - Step 0: As predictor + CCA to scare off users -->
 
-- Step 1: Random intercepts
+<!-- - Step 1: Random intercepts -->
 
-- Step 2: Random slopes
+<!-- - Step 2: Random slopes -->
 
-- Step 3: Residuals
+<!-- - Step 3: Residuals -->
 
-- Heckman model for MNAR
+<!-- - Heckman model for MNAR -->
 
-- What do the different implementations look like? How to define the imputation model(s) in `mice`?
+<!-- - What do the different implementations look like? How to define the imputation model(s) in \pkg{mice}? -->
 
-## Step 0
+<!-- ## Step 0 -->
 
-- AKA multilevel imputation for dummies.
+<!-- - AKA multilevel imputation for dummies.  -->
 
-- Doesn't work for systematic missingness.
+<!-- - Doesn't work for systematic missingness.  -->
 
-## Step 1-3 + MNAR
+<!-- ## Step 1-3 + MNAR -->
 
-- TODO: fill in.
+<!-- - TODO: fill in. -->
 
@@
 
-- ADD: export `mids` objects to other packages like `lme4` or `coxme`?
-
 # Discussion
 
-- JOMO in `mice` --> on the side for now
+- JOMO in \pkg{mice} -> on the side for now
 
@@
 
+# Think about
+
+- Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model.
+
+- Adding a `multilevel_ampute()` wrapper function in mice.
+
+- Exporting `mids` objects to other packages like `lme4` or `coxme`?
+
 # References

### 2021-12-02 Manuscript/Manuscript.tex
First draft introduction section

@@
 within groups, but there are many different types of multilevel data. In
-the medical field clustering occurs at e.g., the hospitals/center level
+the medical field, clustering occurs at e.g., the hospitals/center level
 in registry data, or at the study-level in meta-analyses (IPDMA). In the
@@
 we will refer to the grouping variable as `cluster', and the grouped
-variable as `(sample) unit'. For reasons of brevity, we only discuss
-clustering between units, not longitudinal data (within-unit
-clustering).
+variable as `(sample) unit'.\footnote{Add that we'll only discuss two
+  levels?} For reasons of brevity, we only discuss clustering between
+units, not within units (such as in timeseries or longitudinal data).
 
-Multilevel data requires special care when doing any sort of analysis.
-The cluster to which a unit belongs may influence the unit-level
-observations, and since clusters are made up of units, clusters depend
-on units as well \citep{hox17}. {[}Explain ICC here? The percentage of
-variance attributed to the cluster-level is expressed by the intra-class
+Analyzing multilevel data requires special care, compared to `regular',
+single level data. The cluster to which a unit belongs may influence the
+unit-level observations, and since clusters are made up of units,
+clusters depend on units as well \citep{hox17}. These relations can and
+should be taken into account when developing analysis models for
+multilevel data.\footnote{Explain ICC here? The percentage of variance
+  attributed to the cluster-level is expressed by the intra-class
   coefficient (ICC). The ICC can also be interpreted as the expected
-correlation between two randomly sampled units in same cluster.{]} These
-relations can and should be taken into account when developing analysis
-models for multilevel data. At least, such models include a separate
-intercept term for each cluster. But there may also be random predictor
-effects and/or random error terms (residual error variances), see e.g.
-\citet{hox17} and \citet{jong21}. Heterogeneity refers to variability
-within clusters vs.~variability between clusters. There are many names
-for models that take clustering into account. Some popular examples are
-`multilevel models', `hierarchical models', `mixed effect models' and
-`random effect models'.
+  correlation between two randomly sampled units in same cluster. So if
+  the ICC is high, a lot of variability in a variable is due to the
+  clustering, which should be modeled accordingly.} Multilevel models
+typically include separate intercepts for each cluster, which relieves
+one restriction imposed by single-level models: equal group means across
+clusters. Additionally, there may be random predictor effects and/or
+random error terms (residual error variances), see e.g. \citet{hox17}
+and \citet{jong21}.\footnote{Add that heterogeneity refers to
+  variability within clusters.} There are many names for models that
+take clustering into account. Some popular examples are `multilevel
+models', `hierarchical models', `mixed effect models' and `random effect
+models'.
 
@@
 
-\begin{itemize}
-\item
-  Why/where does missingness occur in multilevel data? I.e., not only
-  patient-level but also cluster-level.
-\item
-  How can we categorize this? Systematic vs sporadic missingness, see
-  \citet{resc13}. Within systematic we have two flavors: unobserved
+Multilevel data is not spared the ubiquitous problem of missing
+information. Just as in single level data, missingness may occur at the
+unit level. But with multiple levels of data comes the potential for
+missingness at multiple levels. Missingness in multilevel data can be
+categorized into two general patterns: systematical missingness and
+sporadic missingness, see \citet{resc13}. In figure 1, we show a dataset
+with units in the rows and variables in the columns, there are 5 units
+nested within 2 clusters, and 3 variables of interest. Variable
+\texttt{X1} is completely observed. Variable \texttt{X2} is
+systematically missing, \texttt{X3} is sporadically missing.\footnote{Explain
+  why.} Systematic missingness can be further subdivided into unobserved
 constants (same value per cluster) and non-measured random variables
-  (which may differ per unit within clusters). In figure 1, we show a
-  dataset with units in the rows and variables in the columns, there are
-  5 units nested within 2 clusters, and 3 variables of interest.
-  Variable \texttt{X1} is completely observed. Variable \texttt{X2} is
-  systematically missing, \texttt{X3} is sporadically missing. The
-  unobserved value for units 4 and 5 on variable \texttt{X2} may be the
-  same or different, defining which type of systematic missingness is
-  happening.
-\end{itemize}
+(which may differ per unit within clusters). In Figure 1, the former
+implies that the unobserved values for units 4 and 5 on variable
+\texttt{X2} would be equal. With the latter, the values may differ.
+Depending on the missing data pattern, there are more or less optimal
+way of accommodating the missingness.
+
+Ignoring missing data in research endeavors is almost never a good idea.
+Complete case analysis (i.e., excluding all units with one or more
+missing entries) can introduce bias in statistical inference and lowers
+statistical power. Instead, the missingness should be accommodated
+\emph{before} or \emph{within} the analysis of scientific interest.
+Especially the former is very generic and popular. Imputing (i.e.,
+filling in) the missing values splits the missing data problem from the
+scientific problem. The \proglang{R} package \pkg{mice} has become the
+de-facto standard for imputation by chained equations, which solves the
+missingness one variable at a time, iteratively. \pkg{mice} is known to
+yield valid inferences under many different missing data circumstances
+\citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
+the context of multilevel data, under varying missing data
+mechanisms.\footnote{Discuss missingness mechanisms before this point,
+  add references \citet{yuce08} and \citet{hox15}.}
 
@@
 
-\begin{itemize}
-\item
-  What kinds of missingness are there? ADD: missingness mechanisms here.
-  See e.g. \citet{yuce08} and \citet{hox15}.
-\item
-  Why are standard (ad hoc) missing data methods not well suited?
-\item
-  What types of multilevel methods are available? General overview of
-  approaches, see \citet{audi18} and \citet{grun18}. E.g., imputation of
-  study level versus patient-level covariates, and one-stage imputation
-  versus two-stage imputation methods.
-\item
-  Additional difficulty that is addressed in this tutorial: MNAR data.
-\end{itemize}
-
 \hypertarget{aim-of-this-paper}{%
@@
 
+This papers serves as a tutorial for imputing incomplete multilevel data
+with \pkg{mice}. We provide practical guidelines and code snippets for
+different missing data situations. For reasons of brevity, we focus on
+imputation by chained equations, although JOMO is available in
+\pkg{mice} as well. Other useful packages for incomplete multilevel data
+include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.\footnote{Rephrase:
+  Some level of knowledge on multilevel models is assumed. We're
+  providing an overview of implementations. It's up-to the reader to
+  decide which multilevel strategy suits their data. So we won't go into
+  detail for the different methods (and equations). Refer to
+  \citet{meng94}, an Audigier paper, and a paper by Grund on
+  congeniality and random slopes. This paper is just a software
+  tutorial. We'll keep it practical.}
+
+We structure this tutorial around three case studies:
+
 \begin{itemize}
 \item
-  Provide practical guidelines with code snippets for imputation of
-  incomplete multilevel data.
+  \texttt{mice::popmis} (simulated data on school kids, with MNAR/MAR
+  mixture);
 \item
-  We focus on the workflow for conditional modeling (not JOMO) in
-  \texttt{mice}. Refer to other packages: \texttt{mitml},
-  \texttt{miceadds}, \texttt{mdmb}.
+  \texttt{metamisc::impact} (real IPD on traumatic brain injuries,
+  without \texttt{NA}s);
 \item
-  Case study options: \texttt{metamisc::impact} (real IPD on traumatic
-  brain injuries, without \texttt{NA}s), \texttt{mice::popularity}
-  (simulated data on school kids, with MNAR/MAR mixture). TODO: Check
-  example data Gelman.
-\item
-  Introduce case study and set scope of this tutorial: We're providing
-  an overview of implementations. It's up-to the reader to decide which
-  strategy suits their data. So we won't go into detail for the
-  different methods (and equations). This paper is just a software
-  tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
-  help function that suggests a suitable predictor matrix to the user,
-  given a certain analysis model.
+  \texttt{GJRM::hiv} (simulated patient data on HIV, without
+  \texttt{NA}s)
 \end{itemize}
+
+For each case study we focus on a different aspect to illustrate how to
+impute incomplete multilevel data.
 
@@
 
-We'll use the IMPACT data (\texttt{metamisc::impact}) and a MAR/MNAR
-version of the \texttt{mice::popmis} data (i.e., a variation on the Hox
-(2010) popularity data, where the missingness in the variables is either
-missing at random (MAR) or missing not at random (MNAR)).
--\textgreater{} ask whether we can use the Heckman repo data or simulate
-data ourselves
+We introduce three case studies to illustrate the workflow. In the
+\texttt{mice::popmis} data, we show the advantages of including the
+multilevel structure of the data into the imputation model. In the
+\texttt{metamisc::impact} data we'll show how to induce missingness and
+solve it in real-world data. In the \texttt{GJRM::hiv} we provide novel
+methodology\footnote{not really, the methods exist already, but how to
+  show that this is something new and exciting?} for imputing MNAR
+missingness according to the Heckman model.
 
-Heckman options:
+For each case study we'll look at least at: 1) the incomplete data; 2)
+the imputation model; 3) the imputed data; and 4) how to obtain pooled
+estimates for the analysis of scientific interest.
 
-\begin{itemize}
-\item
-  leiden85
-\item
-  GJRM::hiv (\url{https://rdrr.io/github/egeminiani/GJRM/man/hiv.html})
-\item
-  simulating
-\item
-  IMPACT
-\end{itemize}
+\hypertarget{case-study-i-popularity}{%
+\subsection{Case Study I: Popularity}\label{case-study-i-popularity}}
 
-\hypertarget{case-study-i-impact}{%
-\subsection{Case study I: IMPACT}\label{case-study-i-impact}}
-
-\begin{itemize}
-\item
-  What does the data look like? \texttt{impact} is traumatic brain
-  injury data with patients clustered in studies,
-  \(n_{\text{participants}} = 11022\) and \(n_{\text{clusters}} = 15\),
-  on the following 11 variables:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \texttt{name} Name of the study,
-  \item
-    \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
-    observational cohort),
-  \item
-    \texttt{age} Age of the patient,
-  \item
-    \texttt{motor\_score} Glasgow Coma Scale motor score,
-  \item
-    \texttt{pupil} Pupillary reactivity,
-  \item
-    \texttt{ct} Marshall Computerized Tomography classification,
-  \item
-    \texttt{hypox} Hypoxia (0=no, 1=yes),
-  \item
-    \texttt{hypots} Hypotension (0=no, 1=yes),
-  \item
-    \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
-  \item
-    \texttt{edh} Epidural hematoma (0=no, 1=yes),
-  \item
-    \texttt{mort} 6-month mortality (0=alive, 1=dead).
-  \end{itemize}
-\end{itemize}
-
-\begin{CodeChunk}
-\begin{CodeOutput}
- /\     /\
-{  `---'  }
-{  O   O  }
-==>  V <==  No need for mice. This data set is completely observed.
- \  \|/  /
-  `-----'
-\end{CodeOutput}
-
-\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}
-
-\end{CodeChunk}
-
--\textgreater{} Why are there no missings? According to the
-\href{https://cran.r-project.org/web/packages/metamisc/metamisc.pdf}{vignette},
-the data is already imputed (Steyerberg et al, 2008).
-
-\begin{itemize}
-\item
-  MAR miss varying by cluster. Obs data patt differ per cluster. E.g.,
-  in cluster 1 miss depends on age but not in cluster two. Split the
-  dataframe and run \texttt{ampute()} on each cluster. -\textgreater{}
-  TODO: also make MNAR missingness to heckman model. Maybe based on
-  \texttt{ct} variable? Inclusion-selection variable. -\textgreater{}
-  otherwise: use \texttt{leiden85} data on blood pressure with MNAR.
-  Then run cox regression like the boshuizen article but with living
-  situation as clusters. -\textgreater{} TODO: get analyses from
-  \url{https://www.gerkovink.com/mimp/Contents/Exercises/Day\%203\%20-\%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html}.
-\item
-  ADD: \texttt{multilevel\_ampute()} wrapper function in \texttt{mice}.
-\end{itemize}
-
-\hypertarget{case-study-ii-popularity}{%
-\subsection{Case Study II: Popularity}\label{case-study-ii-popularity}}
-
-\begin{itemize}
-\item
-  What does the data look like? \texttt{popNCR} is a simulated dataset
-  with pupils clustered in classes, \(n_{\text{participants}} = 2000\),
-  \(n_{\text{clusters}} = 100\), on 7 variables:
+\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
+\(n_{\text{participants}} = 2000\), \(n_{\text{clusters}} = 100\), on 7
+variables:
 
@@
 \end{itemize}
-\item
-  What are the ICCs? For \texttt{popular} the ICC is 0.33. For
-  \texttt{popteach} it is 0.31. It would be wise to use multilevel
-  modeling.
+
+\hypertarget{the-incomplete-data}{%
+\subsubsection{The incomplete data}\label{the-incomplete-data}}
+
+\begin{itemize}
+\tightlist
 \item
@@
 \begin{itemize}
-\tightlist
+\item
+  What are the ICCs? For \texttt{popular} the ICC is 0.33. For
+  \texttt{popteach} it is 0.31. It would be wise to use multilevel
+  modeling.
 \item
@@
 \begin{itemize}
+\tightlist
 \item
@@
   relation seems to be right-tailed as well.
+\end{itemize}
+
+\hypertarget{the-imputation-model}{%
+\subsubsection{The imputation model}\label{the-imputation-model}}
+
+\begin{itemize}
 \item
-  We can impute the missingness the `standard' way, ignoring the
-  multilevel structure of the data. This is surely invalid, given the
-  high ICCs, but we'll do it anyways.
+  We could ignore the multilevel structure of the data. This is likely
+  invalid, given the high ICCs, but we'll do it anyways.
 \item
@@
 class*   2000 50.37     51   1 100.0 29.08
-extrav   2000  5.25      5   1  10.0  1.27
-sex*     2000  1.53      2   1   2.0  0.50
-texp     2000 12.48     12   2  25.0  6.29
-popular  2000  4.99      5   0   9.1  1.35
-popteach 2000  5.02      5   1  10.0  1.39
+extrav   2000  5.23      5   1  10.0  1.25
+sex*     2000  1.51      2   1   2.0  0.50
+texp     2000 12.55     12   2  25.0  6.27
+popular  2000  4.99      5   0   9.1  1.32
+popteach 2000  5.01      5   1  10.0  1.38
 school   2000  5.54      6   1  10.0  2.89
@@
       vars incomplete   ignored
-1  popular  0.3280070 0.2716802
-2 popteach  0.3138658 0.2528468
-3     texp  1.0000000 0.4395296
+1  popular  0.3280070 0.2754195
+2 popteach  0.3138658 0.2467415
+3     texp  1.0000000 0.4443760
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-Warning: Number of logged events: 325
+Warning: Number of logged events: 312
 \end{CodeOutput}
@@
 4  1  1  popular    pmm
-5  1  1 popteach    pmm
+5  1  1  popular    pmm
 6  1  1 popteach    pmm
@@
 3                                                                                                                                                                                                                                                   school
-4                                                                                                                                                                                                                                             texp, school
-5                                                                                                                                                                                                                                                     texp
-6 mice detected that your data are (nearly) multi-collinear.\nIt applied a ridge penalty to continue calculations, but the results can be unstable.\nDoes your dataset contain duplicates, linear transformation, or factors with unique respondent names?
+4                                                                                                                                                                                                                                                     texp
+5 mice detected that your data are (nearly) multi-collinear.\nIt applied a ridge penalty to continue calculations, but the results can be unstable.\nDoes your dataset contain duplicates, linear transformation, or factors with unique respondent names?
+6                                                                                                                                                                                                                                             texp, school
 \end{CodeOutput}
@@
             n  mean median min   max    sd
-pupil    2000 10.65     11   1  26.0  5.97
-class*   2000 50.37     51   1 100.0 29.08
-extrav   2000  6.20      6   1  10.0  2.00
-sex*     2000  1.52      2   1   2.0  0.50
-texp     2000 14.26     15   2  25.0  6.55
-popular  2000  5.03      5   0   9.1  1.35
-popteach 2000  5.03      5   1  10.0  1.38
-school   2000  5.54      6   1  10.0  2.89
+pupil    2000 10.65   11.0   1  26.0  5.97
+class*   2000 50.37   51.0   1 100.0 29.08
+extrav   2000  5.22    5.0   1  10.0  1.29
+sex*     2000  1.53    2.0   1   2.0  0.50
+texp     2000 14.26   15.0   2  25.0  6.55
+popular  2000  5.59    5.4   0   9.1  1.81
+popteach 2000  5.03    5.0   1  10.0  1.40
+school   2000  5.54    6.0   1  10.0  2.89
 \end{CodeOutput}
@@
       vars incomplete   ignored predictor
-1  popular  0.3280070 0.2716802 0.3550567
-2 popteach  0.3138658 0.2528468 0.3306035
-3     texp  1.0000000 0.4395296 1.0000000
+1  popular  0.3280070 0.2754195 0.2496278
+2 popteach  0.3138658 0.2467415 0.3483987
+3     texp  1.0000000 0.4443760 1.0000000
 \end{CodeOutput}
@@
 
-\hypertarget{amputation}{%
-\subsection{Amputation}\label{amputation}}
+\hypertarget{case-study-ii-impact}{%
+\subsection{Case study II: IMPACT}\label{case-study-ii-impact}}
+
+\texttt{impact} is traumatic brain injury data with patients clustered
+in studies, \(n_{\text{participants}} = 11022\) and
+\(n_{\text{clusters}} = 15\), on the following 11 variables:
 
@@
 \item
+  \texttt{name} Name of the study,
+\item
+  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
+  observational cohort),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{motor\_score} Glasgow Coma Scale motor score,
+\item
+  \texttt{pupil} Pupillary reactivity,
+\item
+  \texttt{ct} Marshall Computerized Tomography classification,
+\item
+  \texttt{hypox} Hypoxia (0=no, 1=yes),
+\item
+  \texttt{hypots} Hypotension (0=no, 1=yes),
+\item
+  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+\item
+  \texttt{edh} Epidural hematoma (0=no, 1=yes),
+\item
+  \texttt{mort} 6-month mortality (0=alive, 1=dead).
 \end{itemize}
 
-\hypertarget{modeling-choices}{%
-\subsection{Modeling choices}\label{modeling-choices}}
+The data is already imputed (Steyerberg et al, 2008), so we'll induce
+missingness ourselves. For example, MAR missingness varying by
+cluster.\footnote{Observed data pattern should differ per cluster. So,
+  in cluster 1, the missingness would depend on age, but not in cluster
+  two. Split the dataframe and run \texttt{ampute()} on each cluster.}
 
-\begin{itemize}
-\item
-  Which models will we discuss? We'll build the model to grow in
-  complexity. The final model is the most complex but also the most
-  versatile.
-\item
-  Note on model complexity: Typically, we should at least use random
-  intercepts, but often random slopes as well. Ideally we impute with
-  random everything and heteroscedastic errors: most generic method (no
-  worry about congeniality, but don't mention the term) -\textgreater{}
-  Refer to other papers for background, we'll focus just on the software
-  implementation of the situations mentioned there. Sometimes there's
-  little reason to assume some variable is affected by heterogeneity.
-  -\textgreater{} Refer to \citet{meng94}, an Audigier paper, and a
-  paper by Grund on congeniality and random slopes.
-\item
-  Step 0: As predictor + CCA to scare off users
-\item
-  Step 1: Random intercepts
-\item
-  Step 2: Random slopes
-\item
-  Step 3: Residuals
-\item
-  Heckman model for MNAR
-\item
-  What do the different implementations look like? How to define the
-  imputation model(s) in \texttt{mice}?
-\end{itemize}
+\begin{CodeChunk}
+\begin{CodeOutput}
+ /\     /\
+{  `---'  }
+{  O   O  }
+==>  V <==  No need for mice. This data set is completely observed.
+ \  \|/  /
+  `-----'
+\end{CodeOutput}
 
-\hypertarget{step-0}{%
-\subsection{Step 0}\label{step-0}}
+\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}
 
-\begin{itemize}
-\item
-  AKA multilevel imputation for dummies.
-\item
-  Doesn't work for systematic missingness.
-\end{itemize}
-
-\hypertarget{step-1-3-mnar}{%
-\subsection{Step 1-3 + MNAR}\label{step-1-3-mnar}}
-
-\begin{itemize}
-\tightlist
-\item
-  TODO: fill in.
-\end{itemize}
+\end{CodeChunk}
 
@@
   residual errors, or autocorrelation)
-\item
-  ADD: export \texttt{mids} objects to other packages like \texttt{lme4}
-  or \texttt{coxme}?
 \end{itemize}
@@
 \item
-  JOMO in \texttt{mice} --\textgreater{} on the side for now
+  JOMO in \pkg{mice} -\textgreater{} on the side for now
 \item
@@
 
+\hypertarget{think-about}{%
+\section{Think about}\label{think-about}}
+
+\begin{itemize}
+\item
+  Adding some kind of help function to mice that suggests a suitable
+  predictor matrix to the user, given a certain analysis model.
+\item
+  Adding a \texttt{multilevel\_ampute()} wrapper function in mice.
+\item
+  Exporting \texttt{mids} objects to other packages like \texttt{lme4}
+  or \texttt{coxme}?
+\end{itemize}
+
 \renewcommand\refname{References}

### 2021-12-02 Manuscript/Manuscript.Rmd
Full text version of introduction and first part case study 1

@@
 
-Multilevel data is not spared the ubiquitous problem of missing information.
-Just as in single level data, missingness may occur at the unit level.
-But with multiple levels of data comes the potential for missingness at multiple levels.
-Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13.
-In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 2 clusters, and 3 variables of interest.
-Variable `X1` is completely observed.
-Variable `X2` is systematically missing, `X3` is sporadically missing.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (same value per cluster) and non-measured random variables (which may differ per unit within clusters).
-In Figure 1, the former implies that the unobserved values for units 4 and 5 on variable `X2` would be equal.
-With the latter, the values may differ.
-Depending on the missing data pattern, there are more or less optimal way of accommodating the missingness.
-
-Ignoring missing data in research endeavors is almost never a good idea.
-Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power.
-Instead, the missingness should be accommodated *before* or *within* the analysis of scientific interest.
-Especially the former is very generic and popular.
-Imputing (i.e., filling in) the missing values splits the missing data problem from the scientific problem.
-The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which solves the missingness one variable at a time, iteratively.
-\pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
-In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.^[Discuss missingness mechanisms before this point, add references @yuce08 and @hox15.]
-
-```{r patterns, echo=FALSE}
+```{r patterns, fig.height=4, fig.width=3.5, fig.cap = "Missingness in multilevel data", echo=FALSE}
 # missingness indicator plot (monotone vs non-monotone)
@@
 ```
+
+Multilevel data is not spared the ubiquitous problem of missing information.
+Just as in single level data, missingness may occur at the unit level.
+But with multiple levels of data comes the potential for missingness at multiple levels.
+Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13.
+In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 2 clusters, and 3 variables of interest.
+Variable `X1` is completely observed.
+Variable `X2` is systematically missing, `X3` is sporadically missing.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (same value per cluster) and non-measured random variables (which may differ per unit within clusters).
+In Figure 1, the former implies that the unobserved values for units 4 and 5 on variable `X2` would be equal.
+With the latter, the values may differ.
+Depending on the missing data pattern, there are more or less optimal way of accommodating the missingness.
+
+Ignoring missing data in research endeavors is almost never a good idea.
+Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power.
+Instead, the missingness should be accommodated *before* or *within* the analysis of scientific interest.
+Especially the former is very generic and popular.
+Imputing (i.e., filling in) the missing values splits the missing data problem from the scientific problem.
+The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which solves the missingness one variable at a time, iteratively.
+\pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
+In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.^[Discuss missingness mechanisms before this point, add references @yuce08 and @hox15.]
 
@@
 
+```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
+# load data and merge classes into schools for this tutorial -> but this cannot work because how texp and popteach are defined...
+pop <- readRDS("../Data/popNCR.RDS") %>% mutate(school = round(0.5 + as.numeric(class)/10))
+
+# ICCs
+icc_pop <- multilevel::ICC1(aov(popular ~ as.factor(class), data = pop))
+icc_teach <- multilevel::ICC1(aov(popteach ~ as.factor(class), data = pop))
+```
+
 `popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables:
@@
 
-### The incomplete data
+### Incomplete data
 
-```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
-# load data and merge classes into schools for this tutorial -> but this cannot work because how texp and popteach are defined...
-pop <- readRDS("../Data/popNCR.RDS") %>% mutate(school = round(0.5 + as.numeric(class)/10))
-
-# ICCs
-icc_pop <- multilevel::ICC1(aov(popular ~ as.factor(class), data = pop))
-icc_teach <- multilevel::ICC1(aov(popteach ~ as.factor(class), data = pop))
-```
-
-- What does the missingness look like? Induced MAR/MNAR missingness. Missing data pattern:
-
-```{r pop_pat, echo=FALSE, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
+```{r pop_pat, echo=FALSE, fig.height=6, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
 # missingness
@@
 
-- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would be wise to use multilevel modeling.
+The popularity data is created such that there are strong relations between the incomplete variables and the clustering variable `class`.
+We can express this using the intra-class correlation (ICC).
+For `popular` the ICC is `r round(icc_pop, 2)`.
+For `popteach` it is `r round(icc_teach, 2)`.
+It would thus be wise to use multilevel modeling.
 
-- Does the missing data of `popular` depend on `popteach`?
-Does the missingness in teacher popularity depend on pupil popularity?
--> Check this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity, and the other way around.
+The missingness in this dataset is induced conform MAR and MNAR mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows the systematic nature of the missingness.
 
-```{r pop_dist, echo=FALSE, message=FALSE, warning=FALSE}
+To develop the best imputation model, we need to know whether the missingness in one variable depends on the observed values of other variables.
+Visual inspection usually suffices.
+We'll highlight only two variables to illustrate, but ideally one would inspect all relations.
+The questions we'll ask are: 'Does the missing data of `popular` depend on `popteach`?' and 'Does the missingness in teacher popularity depend on pupil popularity?' We'll evaluate this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity, and the other way around.
+
+In Figure \ref{fig:pop_dist} we see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
+This would indicate a right-tailed MAR missingness.
+In fact, this is exactly what happens, because the missingness in these data was created manually.
+Now, we've made it observable by examining the relations between the missingness in popular and the observed data in `popteach`.
+There is also a dependency between the missingness in teacher popularity and pupil popularity.
+The relation seems to be right-tailed as well.
+
+```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
 # distributions
@@
 
--  We do see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
-This would indicate a right-tailed MAR missingness.
-In fact this is exactly what happens, because we created the missingness in these data ourselves.
-But we made it observable by examining the relations between the missingness in popular and the observed data in `popteach`.
-There is also a dependency between the missingness in teacher popularity and pupil popularity.
-The relation seems to be right-tailed as well.
+### Imputation model
 
-### The imputation model
+The first imputation model that we'll use is likely to be invalid.
+In this model, we ignore the multilevel structure of the data, despite the high ICCs.
+This is purely to illustrate the effects of ignoring the clustering in our imputation effort.
 
-- We could ignore the multilevel structure of the data. This is likely invalid, given the high ICCs, but we'll do it anyways.
-
-- We'll use predictive mean matching to impute the continuous variables (some appear to be somewhat ordinal), and logistic regression to impute the binary variable `sex`.
+We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`.
 We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.
 
-```{r pop-ignored}
+```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
 # dry run to get imputation parameters
@@
 pred[, c("class", "pupil")] <- 0
-pred
 
@@
 imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+```
 
-# check convergence of the imputation model
-plot(imp_ignored)
+### Imputed data
 
-# compare descriptives before and after imputation
-psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
-psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
+```{r pop_ignored_eval, echo=FALSE, message=FALSE, warning=FALSE}
+# # check convergence of the imputation model
+# plot(imp_ignored)
+
+# # compare descriptives before and after imputation
+# psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+# psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
 # TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
@@
 ICCs
-
 ```
 
-- As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `class`.
+As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `class`.
 This tells us that the multilevel structure of the data should be taken into account.
@@
 
-- We can also observe that the teacher experience increases slightly after imputation.
+We can also observe that the teacher experience increases slightly after imputation.
 This is due to the MNAR missingness in `texp`.
@@
 
-- We'll now use `class` as a predictor to impute all other variables.
+### Imputation model
 
-```{r pop-predictor}
+We'll now use `class` as a predictor to impute all other variables.
+This is still not recommended practice, since it only works under certain circumstances and results may be biased.
+But at least, it includes some multilevel aspect.
+Colloquially, this is 'multilevel imputation for dummies'.
+
+```{r pop_predictor, echo=FALSE, message=FALSE, warning=FALSE}
 # adjust the predictor matrix
@@
 pred[, "pupil"] <- 0
-pred
 
@@
 imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+```
 
-# check logged events
-head(imp_predictor$loggedEvents)
+### Imputed data
+
+```{r pop_predictor_eval, echo=FALSE, message=FALSE, warning=FALSE}
+# # check logged events
+# head(imp_predictor$loggedEvents)
 ## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
 
-# check convergence of the imputation model
-plot(imp_predictor)
+# # check convergence of the imputation model
+# plot(imp_predictor)
 
-# compare descriptives before and after imputation
-psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
-psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
+# # compare descriptives before and after imputation
+# psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+# psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
 
@@
 
-- Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MNAR.
+Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MNAR.
 
-- The ICCs are way more in line with the ICCs in the incomplete data. But this is a quick and dirty way of imputing multilevel data. We *should* be using a multilevel model.
+The ICCs are way more in line with the ICCs in the incomplete data. But this is a quick and dirty way of imputing multilevel data. We *should* be using a multilevel model.
+
+### Imputation model
+
+To include...
 
@@
 # psych::describe(impact)[,c(2:5,8:9)]
-# missingness
-plot_md_pat(impact)
+# # missingness
+# plot_md_pat(impact)
 # TODO: rotate variable labels

### 2021-12-02 Manuscript/Manuscript.tex

@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{Manuscript_files/figure-latex/patterns-1}
+
+}
+
+\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
+\end{figure}
+\end{CodeChunk}
+
 Multilevel data is not spared the ubiquitous problem of missing
@@
   add references \citet{yuce08} and \citet{hox15}.}
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Manuscript_files/figure-latex/patterns-1} \end{center}
-
-\end{CodeChunk}
 
@@
 
-\hypertarget{the-incomplete-data}{%
-\subsubsection{The incomplete data}\label{the-incomplete-data}}
-
-\begin{itemize}
-\tightlist
-\item
-  What does the missingness look like? Induced MAR/MNAR missingness.
-  Missing data pattern:
-\end{itemize}
+\hypertarget{incomplete-data}{%
+\subsubsection{Incomplete data}\label{incomplete-data}}
 
 \begin{CodeChunk}
+\begin{figure}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_pat-1} \end{center}
+{\centering \includegraphics{Manuscript_files/figure-latex/pop_pat-1}
 
+}
+
+\caption[Missing data pattern in the popularity data]{Missing data pattern in the popularity data}\label{fig:pop_pat}
+\end{figure}
 \end{CodeChunk}
 
-\begin{itemize}
-\item
-  What are the ICCs? For \texttt{popular} the ICC is 0.33. For
-  \texttt{popteach} it is 0.31. It would be wise to use multilevel
-  modeling.
-\item
-  Does the missing data of \texttt{popular} depend on \texttt{popteach}?
-  Does the missingness in teacher popularity depend on pupil popularity?
-  -\textgreater{} Check this by making a histogram of \texttt{popteach}
-  separately for the pupils with known popularity and missing
-  popularity, and the other way around.
-\end{itemize}
+The popularity data is created such that there are strong relations
+between the incomplete variables and the clustering variable
+\texttt{class}. We can express this using the intra-class correlation
+(ICC). For \texttt{popular} the ICC is 0.33. For \texttt{popteach} it is
+0.31. It would thus be wise to use multilevel modeling.
+
+The missingness in this dataset is induced conform MAR and MNAR
+mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows
+the systematic nature of the missingness.
+
+To develop the best imputation model, we need to know whether the
+missingness in one variable depends on the observed values of other
+variables. Visual inspection usually suffices. We'll highlight only two
+variables to illustrate, but ideally one would inspect all relations.
+The questions we'll ask are: `Does the missing data of \texttt{popular}
+depend on \texttt{popteach}?' and `Does the missingness in teacher
+popularity depend on pupil popularity?' We'll evaluate this by making a
+histogram of \texttt{popteach} separately for the pupils with known
+popularity and missing popularity, and the other way around.
+
+In Figure \ref{fig:pop_dist} we see that the distribution for the
+missing \texttt{popular} is further to the right than the distribution
+for observed \texttt{popular}. This would indicate a right-tailed MAR
+missingness. In fact, this is exactly what happens, because the
+missingness in these data was created manually. Now, we've made it
+observable by examining the relations between the missingness in popular
+and the observed data in \texttt{popteach}. There is also a dependency
+between the missingness in teacher popularity and pupil popularity. The
+relation seems to be right-tailed as well.
 
 \begin{CodeChunk}
+\begin{figure}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_dist-1} \end{center}
+{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-1}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_dist-2} \end{center}
+}
 
+\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-1}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-2}
+
+}
+
+\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-2}
+\end{figure}
 \end{CodeChunk}
 
-\begin{itemize}
-\tightlist
-\item
-  We do see that the distribution for the missing \texttt{popular} is
-  further to the right than the distribution for observed
-  \texttt{popular}. This would indicate a right-tailed MAR missingness.
-  In fact this is exactly what happens, because we created the
-  missingness in these data ourselves. But we made it observable by
-  examining the relations between the missingness in popular and the
-  observed data in \texttt{popteach}. There is also a dependency between
-  the missingness in teacher popularity and pupil popularity. The
-  relation seems to be right-tailed as well.
-\end{itemize}
+\hypertarget{imputation-model}{%
+\subsubsection{Imputation model}\label{imputation-model}}
 
-\hypertarget{the-imputation-model}{%
-\subsubsection{The imputation model}\label{the-imputation-model}}
+The first imputation model that we'll use is likely to be invalid. In
+this model, we ignore the multilevel structure of the data, despite the
+high ICCs. This is purely to illustrate the effects of ignoring the
+clustering in our imputation effort.
 
-\begin{itemize}
-\item
-  We could ignore the multilevel structure of the data. This is likely
-  invalid, given the high ICCs, but we'll do it anyways.
-\item
 We'll use predictive mean matching to impute the continuous variables
-  (some appear to be somewhat ordinal), and logistic regression to
-  impute the binary variable \texttt{sex}. We do not use the observation
-  identifier \texttt{pupil} or cluster identifier \texttt{class} as
-  predictors to impute other variables.
-\end{itemize}
+and logistic regression to impute the binary variable \texttt{sex}. We
+do not use the observation identifier \texttt{pupil} or cluster
+identifier \texttt{class} as predictors to impute other variables.
 
@@
 R> pred[, c("class", "pupil")] <- 0
-R> pred 
-\end{CodeInput}
-\begin{CodeOutput}
-         pupil class extrav sex texp popular popteach school
-pupil        0     0      1   1    1       1        1      1
-class        0     0      1   1    1       1        1      1
-extrav       0     0      0   1    1       1        1      1
-sex          0     0      1   0    1       1        1      1
-texp         0     0      1   1    0       1        1      1
-popular      0     0      1   1    1       0        1      1
-popteach     0     0      1   1    1       1        0      1
-school       0     0      1   1    1       1        1      0
-\end{CodeOutput}
-\begin{CodeInput}
+R> 
 R> # impute the data, ignoring the cluster structure
 R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
-R> 
-R> # check convergence of the imputation model
-R> plot(imp_ignored)
 \end{CodeInput}
+\end{CodeChunk}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-1} \end{center}
+\hypertarget{imputed-data}{%
+\subsubsection{Imputed data}\label{imputed-data}}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-2} \end{center}
+\begin{CodeChunk}
 
-\begin{CodeInput}
-R> # compare descriptives before and after imputation
-R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
-\end{CodeInput}
-\begin{CodeOutput}
-            n  mean median min   max    sd
-pupil    2000 10.65   11.0   1  26.0  5.97
-class*   2000 50.37   51.0   1 100.0 29.08
-extrav   1484  5.31    5.0   1  10.0  1.29
-sex*     1504  1.56    2.0   1   2.0  0.50
-texp     1024 11.80   12.0   2  25.0  6.26
-popular  1490  4.83    4.8   0   9.1  1.34
-popteach 1472  4.83    5.0   1  10.0  1.36
-school   2000  5.54    6.0   1  10.0  2.89
-\end{CodeOutput}
-\begin{CodeInput}
-R> psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
-\end{CodeInput}
-\begin{CodeOutput}
-            n  mean median min   max    sd
-pupil    2000 10.65     11   1  26.0  5.97
-class*   2000 50.37     51   1 100.0 29.08
-extrav   2000  5.23      5   1  10.0  1.25
-sex*     2000  1.51      2   1   2.0  0.50
-texp     2000 12.55     12   2  25.0  6.27
-popular  2000  4.99      5   0   9.1  1.32
-popteach 2000  5.01      5   1  10.0  1.38
-school   2000  5.54      6   1  10.0  2.89
-\end{CodeOutput}
-\begin{CodeInput}
-R> # TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
-R> # TODO: pool mean median and sd
-R> # feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
-R> # TODO: add FMI for each of the estimates? at least for the mean
-R> 
-R> # further inspection of the imputations
-R> densityplot(imp_ignored)
-\end{CodeInput}
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_ignored_eval-1} \end{center}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-3} \end{center}
-
-\begin{CodeInput}
-R> # compare ICCs before and after imputation
-R> ICCs <- data.frame(
-+   vars = c("popular", "popteach", "texp"), 
-+   incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
-+                multilevel::ICC1(aov(popteach ~ class, pop)),
-+                multilevel::ICC1(aov(texp ~ class, pop))), 
-+   ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
-+               multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
-+               multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
-+   )
-R> ICCs
-\end{CodeInput}
 \begin{CodeOutput}
       vars incomplete   ignored
-1  popular  0.3280070 0.2754195
-2 popteach  0.3138658 0.2467415
-3     texp  1.0000000 0.4443760
+1  popular  0.3280070 0.2958571
+2 popteach  0.3138658 0.2561198
+3     texp  1.0000000 0.4696024
 \end{CodeOutput}
@@
 
-\begin{itemize}
-\item
-  As the original ICCs show, 100\% of the variance in \texttt{texp} can
-  be attributed to the clustering variable \texttt{class}. This tells us
-  that the multilevel structure of the data should be taken into
-  account. If we don't, we'll end up with incorrect imputations, biasing
-  the effect of the clusters towards zero.
-\item
-  We can also observe that the teacher experience increases slightly
-  after imputation. This is due to the MNAR missingness in
-  \texttt{texp}. Higher values for \texttt{texp} have a larger
-  probability to be missing. This may not a problem, however, if at
-  least one pupil in each class has teacher experience recorded, we can
-  deductively impute the correct (i.e.~true) value for every pupil in
-  the class.
-\item
+As the original ICCs show, 100\% of the variance in \texttt{texp} can be
+attributed to the clustering variable \texttt{class}. This tells us that
+the multilevel structure of the data should be taken into account. If we
+don't, we'll end up with incorrect imputations, biasing the effect of
+the clusters towards zero.
+
+We can also observe that the teacher experience increases slightly after
+imputation. This is due to the MNAR missingness in \texttt{texp}. Higher
+values for \texttt{texp} have a larger probability to be missing. This
+may not a problem, however, if at least one pupil in each class has
+teacher experience recorded, we can deductively impute the correct
+(i.e.~true) value for every pupil in the class.
+
+\hypertarget{imputation-model-1}{%
+\subsubsection{Imputation model}\label{imputation-model-1}}
+
 We'll now use \texttt{class} as a predictor to impute all other
-  variables.
-\end{itemize}
+variables. This is still not recommended practice, since it only works
+under certain circumstances and results may be biased. But at least, it
+includes some multilevel aspect. Colloquially, this is `multilevel
+imputation for dummies'.
+
+\hypertarget{imputed-data-1}{%
+\subsubsection{Imputed data}\label{imputed-data-1}}
 
 \begin{CodeChunk}
-\begin{CodeInput}
-R> # adjust the predictor matrix
-R> pred <- ini$pred 
-R> pred[, "pupil"] <- 0
-R> pred
-\end{CodeInput}
-\begin{CodeOutput}
-         pupil class extrav sex texp popular popteach school
-pupil        0     1      1   1    1       1        1      1
-class        0     0      1   1    1       1        1      1
-extrav       0     1      0   1    1       1        1      1
-sex          0     1      1   0    1       1        1      1
-texp         0     1      1   1    0       1        1      1
-popular      0     1      1   1    1       0        1      1
-popteach     0     1      1   1    1       1        0      1
-school       0     1      1   1    1       1        1      0
-\end{CodeOutput}
-\begin{CodeInput}
-R> # impute the data, cluster as predictor
-R> imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-Warning: Number of logged events: 312
-\end{CodeOutput}
-\begin{CodeInput}
-R> # check logged events
-R> head(imp_predictor$loggedEvents)
-\end{CodeInput}
-\begin{CodeOutput}
-  it im      dep   meth
-1  1  1   extrav    pmm
-2  1  1      sex logreg
-3  1  1     texp    pmm
-4  1  1  popular    pmm
-5  1  1  popular    pmm
-6  1  1 popteach    pmm
-                                                                                                                                                                                                                                                       out
-1                                                                                                                                                                                                                                                   school
-2                                                                                                                                                                                                                                                   school
-3                                                                                                                                                                                                                                                   school
-4                                                                                                                                                                                                                                                     texp
-5 mice detected that your data are (nearly) multi-collinear.\nIt applied a ridge penalty to continue calculations, but the results can be unstable.\nDoes your dataset contain duplicates, linear transformation, or factors with unique respondent names?
-6                                                                                                                                                                                                                                             texp, school
-\end{CodeOutput}
-\begin{CodeInput}
-R> ## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
-R> 
-R> # check convergence of the imputation model
-R> plot(imp_predictor)
-\end{CodeInput}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-1} \end{center}
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_predictor_eval-1} \end{center}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-2} \end{center}
-
-\begin{CodeInput}
-R> # compare descriptives before and after imputation
-R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
-\end{CodeInput}
-\begin{CodeOutput}
-            n  mean median min   max    sd
-pupil    2000 10.65   11.0   1  26.0  5.97
-class*   2000 50.37   51.0   1 100.0 29.08
-extrav   1484  5.31    5.0   1  10.0  1.29
-sex*     1504  1.56    2.0   1   2.0  0.50
-texp     1024 11.80   12.0   2  25.0  6.26
-popular  1490  4.83    4.8   0   9.1  1.34
-popteach 1472  4.83    5.0   1  10.0  1.36
-school   2000  5.54    6.0   1  10.0  2.89
-\end{CodeOutput}
-\begin{CodeInput}
-R> psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
-\end{CodeInput}
-\begin{CodeOutput}
-            n  mean median min   max    sd
-pupil    2000 10.65   11.0   1  26.0  5.97
-class*   2000 50.37   51.0   1 100.0 29.08
-extrav   2000  5.22    5.0   1  10.0  1.29
-sex*     2000  1.53    2.0   1   2.0  0.50
-texp     2000 14.26   15.0   2  25.0  6.55
-popular  2000  5.59    5.4   0   9.1  1.81
-popteach 2000  5.03    5.0   1  10.0  1.40
-school   2000  5.54    6.0   1  10.0  2.89
-\end{CodeOutput}
-\begin{CodeInput}
-R> # further inspection of the imputations
-R> densityplot(imp_predictor)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-3} \end{center}
-
-\begin{CodeInput}
-R> # compare ICCs before and after imputation
-R> ICCs <- ICCs %>% cbind(
-+            predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
-+                         multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
-+                         multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
-+            )
-R> ICCs
-\end{CodeInput}
 \begin{CodeOutput}
       vars incomplete   ignored predictor
-1  popular  0.3280070 0.2754195 0.2496278
-2 popteach  0.3138658 0.2467415 0.3483987
-3     texp  1.0000000 0.4443760 1.0000000
+1  popular  0.3280070 0.2958571 0.3581435
+2 popteach  0.3138658 0.2561198 0.3446899
+3     texp  1.0000000 0.4696024 1.0000000
 \end{CodeOutput}
@@
 
-\begin{itemize}
-\item
 Now, we can clearly see that the imputed values of \texttt{texp} are
@@
 MNAR.
-\item
-  The ICCs are way more in line with the ICCs in the incomplete data.
-  But this is a quick and dirty way of imputing multilevel data. We
+
+The ICCs are way more in line with the ICCs in the incomplete data. But
+this is a quick and dirty way of imputing multilevel data. We
 \emph{should} be using a multilevel model.
-\end{itemize}
+
+\hypertarget{imputation-model-2}{%
+\subsubsection{Imputation model}\label{imputation-model-2}}
+
+To include\ldots{}
 
@@
   two. Split the dataframe and run \texttt{ampute()} on each cluster.}
-
-\begin{CodeChunk}
-\begin{CodeOutput}
- /\     /\
-{  `---'  }
-{  O   O  }
-==>  V <==  No need for mice. This data set is completely observed.
- \  \|/  /
-  `-----'
-\end{CodeOutput}
-
-\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}
-
-\end{CodeChunk}
 

### 2021-12-06 Manuscript/Manuscript.Rmd
Use correct icc function, add meeting comments

@@
 # TODO: participant/response/*sample units*/records instead of obs on the y axis
+# TODO: cluster 3
+# TODO: add border/line
+# TODO: remove unit
+# TODO: add accolade
 
@@
 # ICCs
-icc_pop <- multilevel::ICC1(aov(popular ~ as.factor(class), data = pop))
-icc_teach <- multilevel::ICC1(aov(popteach ~ as.factor(class), data = pop))
+icc_pop <- icc(popular ~ as.factor(class), data = pop)
+icc_teach <- icc(popteach ~ as.factor(class), data = pop)
 ```
@@
   ggplot(data, aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
-  geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
-  geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
+  geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
+  geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
   geom_density() +
@@
 # alternatively add a quartile line around the density with geom_ribbon
+
+# marginal
 
@@
   vars = c("popular", "popteach", "texp"), 
-  incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
-               multilevel::ICC1(aov(popteach ~ class, pop)),
-               multilevel::ICC1(aov(texp ~ class, pop))), 
-  ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
-              multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
-              multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
+  incomplete = c(icc(popular ~ class, pop), 
+               icc(popteach ~ class, pop),
+               icc(texp ~ class, pop)), 
+  ignored = c(icc(popular ~ class, complete(imp_ignored)), 
+              icc(popteach ~ class, complete(imp_ignored)), 
+              icc(texp ~ class, complete(imp_ignored)))
   )
@@
 
-```{r pop_predictor, echo=FALSE, message=FALSE, warning=FALSE}
+```{r pop_predictor, message=FALSE, warning=FALSE}
 # adjust the predictor matrix
@@
 ICCs <- ICCs %>% cbind(
-           predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
-                        multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
-                        multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
+           predictor = c(icc(popular ~ class, complete(imp_predictor)), 
+                        icc(popteach ~ class, complete(imp_predictor)), 
+                        icc(texp ~ class, complete(imp_predictor)))
            )
@@
 
+- Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt.
+
 # References

### 2021-12-06 Manuscript/Manuscript.tex
Use correct icc function, add meeting comments

@@
       vars incomplete   ignored
-1  popular  0.3280070 0.2958571
-2 popteach  0.3138658 0.2561198
-3     texp  1.0000000 0.4696024
+1  popular  0.3280070 0.2756762
+2 popteach  0.3138658 0.2456775
+3     texp  1.0000000 0.4129275
 \end{CodeOutput}
@@
 
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # adjust the predictor matrix
+R> pred <- ini$pred 
+R> pred[, "pupil"] <- 0
+R> 
+R> # impute the data, cluster as predictor
+R> imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
 \hypertarget{imputed-data-1}{%
@@
       vars incomplete   ignored predictor
-1  popular  0.3280070 0.2958571 0.3581435
-2 popteach  0.3138658 0.2561198 0.3446899
-3     texp  1.0000000 0.4696024 1.0000000
+1  popular  0.3280070 0.2756762 0.3587923
+2 popteach  0.3138658 0.2456775 0.1998760
+3     texp  1.0000000 0.4129275 1.0000000
 \end{CodeOutput}
@@
   or \texttt{coxme}?
+\item
+  Adding a ICC=0 dataset to show that even if there is no clustering it
+  doesn't hurt.
 \end{itemize}

### 2021-12-13 R/impact_functions.r
add missingness in impact data

@@ -1,0 +1,81 @@
+library(foreign)
+library(dplyr)
+
+fit_missingness_models <- function(fname) {
+  data <- read.spss(fname, to.data.frame = TRUE)
+  
+  # Names of incomplete variables
+  vnames_incomplete <- c("d_pupil", "ct_class", "hypoxia", "hypotens", "tsah", "edh") #i_pupil 
+  vnames_complete <- c("age", "d_motor")
+  
+  data$name <- ""
+  data$name[data$trial == "74"] <- "TINT" #1118
+  data$name[data$trial == "75"] <- "TIUS" #1041
+  data$name[data$trial == "76"] <- "SLIN" #409
+  data$name[data$trial == "77"] <- "SAP" #919
+  data$name[data$trial == "79"] <- "PEG" #1510
+  data$name[data$trial == "80"] <- "HIT I" # 350
+  data$name[data$trial == "81"] <- "UK4" #791
+  data$name[data$trial == "83"] <- "TCDB" #603
+  data$name[data$trial == "84"] <- "SKB" #126
+  data$name[data$trial == "85"] <- "EBIC" #822
+  data$name[data$trial == "86"] <- "HIT II" #819
+  data$name[data$trial == "88"] <- "NABIS" #385
+  data$name[data$trial == "89"] <- "CSTAT" # 517
+  data$name[data$trial == "90"] <- "PHARMOS" #856
+  data$name[data$trial == "91"] <- "APOE" # 756
+  
+  
+  results <- data.frame(trial = character(),
+                        variable_missing = character(),
+                        percent_missing = numeric(),
+                        "beta_Intercept" = numeric(),
+                        beta_d_pupil =  numeric(),
+                        beta_ct_class = numeric(),
+                        beta_hypoxia = numeric(),
+                        beta_hypotens = numeric(),
+                        beta_tsah = numeric(),
+                        beta_edh = numeric(),
+                        beta_age = numeric(),
+                        beta_d_motor = numeric())
+  
+
+
+
+  # Iterate over the studies
+  studyids <- unique(data$trial)
+  
+  # 
+  for (sid in studyids) {
+    dat <- subset(data, trial == sid)[,c("name", vnames_complete, vnames_incomplete)]
+    
+    nmissing <- colSums(apply(dat, 2, is.na))
+    pct_missing <- 100*nmissing/nrow(dat)
+    
+    for (vname in vnames_incomplete) {
+      dat$observed <- ifelse(is.na(dat[,vname]),0, 1)
[…]

### 2021-12-13 R/impact_functions.r
Add categorization on key variables; include outcome variable

@@ -5,8 +5,8 @@ fit_missingness_models <- function(fname) {
   data <- read.spss(fname, to.data.frame = TRUE)
   
   # Names of incomplete variables
-  vnames_incomplete <- c("d_pupil", "ct_class", "hypoxia", "hypotens", "tsah", "edh") #i_pupil 
-  vnames_complete <- c("age", "d_motor")
+  vnames_incomplete <- c("pupil", "ct", "hypoxia", "hypotens", "tsah", "edh")  
+  vnames_complete <- c("age", "motor_score", "d_mort")
   
   data$name <- ""
   data$name[data$trial == "74"] <- "TINT" #1118
@@ -25,19 +25,32 @@ fit_missingness_models <- function(fname) {
   data$name[data$trial == "90"] <- "PHARMOS" #856
   data$name[data$trial == "91"] <- "APOE" # 756
   
+  # Change coding of variables to match the coding in metamisc
+  data$motor_score <- factor(data$d_motor, levels = c(1, 2, 3, 4, 5, 6),
+                             labels = c("1/2", "1/2", "3", "4", "5/6", "5/6"))
+  data$pupil <- factor(data$d_pupil, levels = c(1, 2, 3),
+                       labels = c("Both", "One", "None"))
+  data$ct <- factor(data$ct_class, levels = c(1, 2, 3, 4, 5),
+                    labels = c("I/II", "I/II", "III", "IV/V", "IV/V"))
+  
   
   results <- data.frame(trial = character(),
                         variable_missing = character(),
                         percent_missing = numeric(),
                         "beta_Intercept" = numeric(),
-                        beta_d_pupil =  numeric(),
-                        beta_ct_class = numeric(),
+                        beta_pupilOne =  numeric(),
+                        beta_pupilNone =  numeric(),
+                        beta_ctIII = numeric(),
+                        beta_ctIV.V = numeric(),
                         beta_hypoxia = numeric(),
                         beta_hypotens = numeric(),
                         beta_tsah = numeric(),
                         beta_edh = numeric(),
                         beta_age = numeric(),
-                        beta_d_motor = numeric())
+                        beta_motor_score3 = numeric(),
+                        beta_motor_score4 = numeric(),
+                        "beta_motor_score5.6" = numeric(),
+                        beta_d_mort = numeric())
   
 
 
@@ -65,13 +78,19 @@ fit_missingness_models <- function(fname) {
       
       cnames <- names(coef(fit))
[…]

### 2021-12-13 Manuscript/Manuscript.Rmd
Clean up manuscript file

@@
     affiliation: University Medical Center Utrecht 
-      # | Julius Center for Health Sciences and Primary Care, 
-      # | University Medical Center Utrecht, Utrecht, The Netherlands
-    #affiliation2:
 title:
   formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
-  # If you use tex in the formatted title, also supply version without
   plain:     "Imputation of Incomplete Multilevel Data with mice"
-  # For running headers, if needed
-  short:     "\\pkg{mice}: Multilevel"
+  short:     "Multilevel \\pkg{mice}"
 abstract: >
@@
   \usepackage{amsmath}
+header-includes: 
+ - \usepackage{graphicx} 
+ - \usepackage{mathtools}
 output: rticles::jss_article
 bibliography: ../References/multilevelmice.bib
+editor_options: 
+  chunk_output_type: inline
 ---
@@
 options(prompt = 'R> ', continue = '+ ')
+
+# seed
+set.seed(123)
 
@@
 
+$$
+\begin{matrix} 
+                          &X1     & X2          & X3            \\
+   \hline
+   \mathrm{cluster 1}\{   &x_{11} &x_{12}       &\mathtt{NA}    \\
+                          &x_{21} &x_{22}       &x_{32}         \\
+   \mathrm{cluster 2}\{   &x_{31} &\mathtt{NA}  &x_{33}         \\
+                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
+   \mathrm{cluster 3}\{   &x_{51} &x_{52}       &x_{53}         \\
+\end{matrix} 
+$$
+
 Multilevel data is not spared the ubiquitous problem of missing information.
@@
 
-<!-- -	What kinds of missingness are there? ADD: missingness mechanisms here. See e.g. @yuce08 and @hox15. -->
-
-<!-- - Why are standard (ad hoc) missing data methods not well suited?  -->
-
 <!-- - What types of multilevel methods are available?
@@
 -->
-
-<!-- - Additional difficulty that is addressed in this tutorial: MNAR data. -->
 
@@
 
-# Workflows
-
-<!-- We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).
--> ask whether we can use the Heckman repo data or simulate data ourselves -->
-
-<!-- Heckman options: -->
-
-<!-- - leiden85 -->
-
-<!-- - GJRM::hiv (https://rdrr.io/github/egeminiani/GJRM/man/hiv.html) -->
-
-<!-- - simulating -->
-
-<!-- - IMPACT -->
-
 We introduce three case studies to illustrate the workflow.
@@
 
-## Case Study I: Popularity
+# Case Study I: Popularity
 
@@
 # load data and merge classes into schools for this tutorial -> but this cannot work because how texp and popteach are defined...
-pop <- readRDS("../Data/popNCR.RDS") %>% mutate(school = round(0.5 + as.numeric(class)/10))
+pop <- readRDS("../Data/popNCR2.RDS") %>% mutate(class = as.factor(class))
 
 # ICCs
-icc_pop <- icc(popular ~ as.factor(class), data = pop)
-icc_teach <- icc(popteach ~ as.factor(class), data = pop)
+icc_pop <- icc(popular ~ class, data = pop)
+icc_teach <- icc(popteach ~ class, data = pop)
 ```
 
-`popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables:
+`popNCR` is a simulated dataset with pupils clustered in classes, where the number of cases $n = 2000$, and the number of clusters $N = 100$, on 7 variables:
 
@@
 
-```{r pop_pat, echo=FALSE, fig.height=6, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
+```{r pop_pat, echo=FALSE, fig.height=4, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
 # missingness
@@
 ```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-# distributions
-# plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
-# plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
-# TODO: check if up-side-down plot works
-# or maybe overlapping densities
-# TODO: check smoothing in geom_density function and make it the inverse of the sample size
-# TODO: add functions to mice
-
+# plotting colors
 plot_col <- mice:::mdc(c("obs", "mis")) %>% setNames(c("observed", "missing"))
 
-# x = "popular"
-# z = "popteach"
+# distributions
 plot_conditional <- function(data, x, z, cluster, ...){
@@
 
-plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
+# plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
 plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
-# TODO: pooteach obs/mis
+# TODO: check if up-side-down plot works
+# or maybe overlapping densities
+# TODO: check smoothing in geom_density function and make it the inverse of the sample size
+# TODO: add functions to mice
+# TODO: legend: popteach obs/mis
 # TODO: add title
@@
 # alternatively add a quartile line around the density with geom_ribbon
-
-# marginal
-
 ```
@@
 # impute the data, ignoring the cluster structure
-imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
 ```
@@
 # impute the data, cluster as predictor
-imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)
+
 ```
@@
 # compare ICCs before and after imputation
-ICCs <- ICCs %>% cbind(
+ICCs <- ICCs %>% mutate(
            predictor = c(icc(popular ~ class, complete(imp_predictor)), 
@@
 
+```{r pop_norm}
+pred <- ini$pred
+pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
+#-2 for the cluster variable, 2 for random effects
+meth <- ini$meth
+meth <- c("", "", "", "", "", "2l.norm", "")
+imp_norm <-
+  mice(
+    pop %>% mutate(class = as.integer(class)),
+    pred = pred,
+    meth = meth,
+    maxit = 1,
+    print = FALSE
+  )
+```
+
+```{r pop_norm_eval}
+# plot(imp_norm)
+densityplot(imp_norm, ~popular)
+ICCs <- ICCs %>% mutate(
+           norm = c(icc(popular ~ as.factor(class), complete(imp_norm)), 
+                    icc(popteach ~ as.factor(class), complete(imp_norm)), 
+                    icc(texp ~ as.factor(class), complete(imp_norm)))
+           )
+ICCs
+```
+
+```{r pop_pan}
+pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
+meth <- c("", "", "", "", "", "2l.pan", "")
+imp_pan <-
+  mice(
+    pop %>% mutate(class = as.integer(class)),
+    pred = pred,
+    meth = meth,
+    maxit = 1,
+    print = FALSE
+  )
+
+```
+
+```{r pop_pan_eval}
+# plot(imp_pan)
+densityplot(imp_pan, ~popular)
+ICCs <- ICCs %>% mutate(
+           pan = c(icc(popular ~ as.factor(class), complete(imp_pan)), 
+                    icc(popteach ~ as.factor(class), complete(imp_pan)), 
+                    icc(texp ~ as.factor(class), complete(imp_pan)))
+           )
+ICCs
+```
+
 ## Case study II: IMPACT
 
-`impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
+`impact` is traumatic brain injury data with patients, $n = 11022$, clustered in studies, $N = 15$. With the following 11 variables:
 
@@
 ```
-<!-- -> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf),  -->
-
-<!-- TODO: also make MNAR missingness to heckman model.
-Maybe based on `ct` variable?
-Inclusion-selection variable.
--> otherwise: use `leiden85` data on blood pressure with MNAR.
-Then run cox regression like the boshuizen article but with living situation as clusters.
--> TODO: get analyses from https://www.gerkovink.com/mimp/Contents/Exercises/Day%203%20-%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html.
--->
-
-<!-- ## Amputation -->
-
-<!-- -  -->
-
-<!-- ## Modeling choices -->
-
-<!-- - Which models will we discuss? We'll build the model to grow in complexity. The final model is the most complex but also the most versatile.  -->
-
-<!-- - Note on model complexity: Typically, we should at least use random intercepts, but often random slopes as well.
-Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeniality, but don't mention the term) ->  Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there.
-Sometimes there's little reason to assume some variable is affected by heterogeneity.
-->  -->
-
-<!-- - Step 0: As predictor + CCA to scare off users -->
-
-<!-- - Step 1: Random intercepts -->
-
-<!-- - Step 2: Random slopes -->
-
-<!-- - Step 3: Residuals -->
-
-<!-- - Heckman model for MNAR -->
-
-<!-- - What do the different implementations look like? How to define the imputation model(s) in \pkg{mice}? -->
-
-<!-- ## Step 0 -->
-
-<!-- - AKA multilevel imputation for dummies.  -->
-
-<!-- - Doesn't work for systematic missingness.  -->
-
-<!-- ## Step 1-3 + MNAR -->
-
-<!-- - TODO: fill in. -->
 

### 2021-12-13 Manuscript/Manuscript.tex
Clean up manuscript file

--- 2021-12-02 Manuscript/Manuscript.tex
@@
 \Plaintitle{Imputation of Incomplete Multilevel Data with mice}
-\Shorttitle{\pkg{mice}: Multilevel}
+\Shorttitle{Multilevel \pkg{mice}}
 
@@
 % Pandoc citation processing
+
+\usepackage{graphicx}
+\usepackage{mathtools}
 
@@
 \end{CodeChunk}
+
+\[
+\begin{matrix} 
+                          &X1     & X2          & X3            \\
+   \hline
+   \mathrm{cluster 1}\{   &x_{11} &x_{12}       &\mathtt{NA}    \\
+                          &x_{21} &x_{22}       &x_{32}         \\
+   \mathrm{cluster 2}\{   &x_{31} &\mathtt{NA}  &x_{33}         \\
+                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
+   \mathrm{cluster 3}\{   &x_{51} &x_{52}       &x_{53}         \\
+\end{matrix} 
+\]
 
@@
 
-\hypertarget{workflows}{%
-\section{Workflows}\label{workflows}}
-
 We introduce three case studies to illustrate the workflow. In the
@@
 \hypertarget{case-study-i-popularity}{%
-\subsection{Case Study I: Popularity}\label{case-study-i-popularity}}
+\section{Case Study I: Popularity}\label{case-study-i-popularity}}
 
 \texttt{popNCR} is a simulated dataset with pupils clustered in classes,
-\(n_{\text{participants}} = 2000\), \(n_{\text{clusters}} = 100\), on 7
-variables:
+where the number of cases \(n = 2000\), and the number of clusters
+\(N = 100\), on 7 variables:
 
@@
 (ICC). For \texttt{popular} the ICC is 0.33. For \texttt{popteach} it is
-0.31. It would thus be wise to use multilevel modeling.
+0.34. It would thus be wise to use multilevel modeling.
 
@@
 
-\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-1}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-2}
-
-}
-
-\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-2}
+\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist}
 \end{figure}
@@
 R> # impute the data, ignoring the cluster structure
-R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
+R> imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
 \end{CodeInput}
@@
       vars incomplete   ignored
-1  popular  0.3280070 0.2958571
-2 popteach  0.3138658 0.2561198
-3     texp  1.0000000 0.4696024
+1  popular  0.3280070 0.3338626
+2 popteach  0.3414766 0.3414766
+3     texp  1.0000000 1.0000000
 \end{CodeOutput}
@@
 
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # adjust the predictor matrix
+R> pred <- ini$pred 
+R> pred[, "pupil"] <- 0
+R> 
+R> # impute the data, cluster as predictor
+R> imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
 \hypertarget{imputed-data-1}{%
@@
       vars incomplete   ignored predictor
-1  popular  0.3280070 0.2958571 0.3581435
-2 popteach  0.3138658 0.2561198 0.3446899
-3     texp  1.0000000 0.4696024 1.0000000
+1  popular  0.3280070 0.3338626 0.3726455
+2 popteach  0.3414766 0.3414766 0.3414766
+3     texp  1.0000000 1.0000000 1.0000000
 \end{CodeOutput}
@@
 
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred <- ini$pred
+R> pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
+R> #-2 for the cluster variable, 2 for random effects
+R> meth <- ini$meth
+R> meth <- c("", "", "", "", "", "2l.norm", "")
+R> imp_norm <-
++   mice(
++     pop %>% mutate(class = as.integer(class)),
++     pred = pred,
++     meth = meth,
++     maxit = 1,
++     print = FALSE
++   )
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # plot(imp_norm)
+R> densityplot(imp_norm, ~popular)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_norm_eval-1} \end{center}
+
+\begin{CodeInput}
+R> ICCs <- ICCs %>% mutate(
++            norm = c(icc(popular ~ as.factor(class), complete(imp_norm)), 
++                     icc(popteach ~ as.factor(class), complete(imp_norm)), 
++                     icc(texp ~ as.factor(class), complete(imp_norm)))
++            )
+R> ICCs
+\end{CodeInput}
+\begin{CodeOutput}
+      vars incomplete   ignored predictor      norm
+1  popular  0.3280070 0.3338626 0.3726455 0.3667167
+2 popteach  0.3414766 0.3414766 0.3414766 0.3414766
+3     texp  1.0000000 1.0000000 1.0000000 1.0000000
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
+R> meth <- c("", "", "", "", "", "2l.pan", "")
+R> imp_pan <-
++   mice(
++     pop %>% mutate(class = as.integer(class)),
++     pred = pred,
++     meth = meth,
++     maxit = 1,
++     print = FALSE
++   )
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # plot(imp_pan)
+R> densityplot(imp_pan, ~popular)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_pan_eval-1} \end{center}
+
+\begin{CodeInput}
+R> ICCs <- ICCs %>% mutate(
++            pan = c(icc(popular ~ as.factor(class), complete(imp_pan)), 
++                     icc(popteach ~ as.factor(class), complete(imp_pan)), 
++                     icc(texp ~ as.factor(class), complete(imp_pan)))
++            )
+R> ICCs
+\end{CodeInput}
+\begin{CodeOutput}
+      vars incomplete   ignored predictor      norm       pan
+1  popular  0.3280070 0.3338626 0.3726455 0.3667167 0.3719078
+2 popteach  0.3414766 0.3414766 0.3414766 0.3414766 0.3414766
+3     texp  1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
+\end{CodeOutput}
+\end{CodeChunk}
+
 \hypertarget{case-study-ii-impact}{%
@@
 
-\texttt{impact} is traumatic brain injury data with patients clustered
-in studies, \(n_{\text{participants}} = 11022\) and
-\(n_{\text{clusters}} = 15\), on the following 11 variables:
+\texttt{impact} is traumatic brain injury data with patients,
+\(n = 11022\), clustered in studies, \(N = 15\). With the following 11
+variables:
 

### 2021-12-13 Manuscript/Manuscript.Rmd
Add plotting functions for imputed data

@@
 miceadds::source.all(path = "../R")
+`%nin%` <- Negate(`%in%`)
 icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
@@
 at the country-level, or as imposed by the sampling design.
-In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.^[Add that we'll only discuss two levels?] For reasons of brevity, we only discuss clustering between units, not within units (such as in timeseries or longitudinal data).
+In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.^[Alternatives to unit: participant/response/record.] For reasons of brevity, we only discuss clustering between units, not within units (such as in timeseries or longitudinal data).^[Also, add that we'll only discuss two levels, no more?]
 
@@
 ```{r patterns, fig.height=4, fig.width=3.5, fig.cap = "Missingness in multilevel data", echo=FALSE}
-# missingness indicator plot (monotone vs non-monotone)
-miss_ind <- tibble(
-    rownr = 5:1,
-    unit = rep("observed", 5),
-    cluster = rep("observed", 5),
-    X1 = rep("observed", 5),
-    X2 = c(rep("observed", 3), rep("missing", 2)),
-    X3 = c("missing", rep("observed", 3), "missing")
-  ) %>% pivot_longer(cols = c(unit, cluster, X1, X2, X3))
-
-# add text 
-miss_ind$text <- ""
-miss_ind[miss_ind$name == "unit", "text"] <- as.character(1:5)
-miss_ind[miss_ind$name == "cluster", "text"] <- as.character(c(1,1,1,2,2))
-
-# plot
-miss_ind %>% 
-  mutate(name = factor(name, levels = unique(name))) %>% 
-ggplot(aes(
-  x = name,
-  y = rownr,
-  color = value,
-  width = 0.8, 
-  height = 0.8,
-  label = text
-)) +
-  geom_tile(fill = "white", size = 2) +
-  geom_text(show.legend = FALSE) +
-  scale_x_discrete(position = "top") +
-  scale_y_discrete(labels = 5:1) +
-  scale_color_manual(values = plot_col, name = "") +
-  theme_minimal() +
-  theme(
-    legend.position = "bottom", 
-    panel.grid.major = element_blank()) +
-  labs(x = "", y = "")
-# TODO: make it one plot with sporadic and systematic
-# TODO: participant/response/*sample units*/records instead of obs on the y axis
-# TODO: cluster 3
+plot_md_set()
 # TODO: add border/line
@@
 
-<!-- - What types of multilevel methods are available?
-General overview of approaches, see @audi18 and @grun18.
-E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.
--->
-
 ## Aim of this paper
@@
 So we won't go into detail for the different methods (and equations).
-Refer to @meng94, an Audigier paper, and a paper by Grund on congeniality and random slopes.
+Refer to @meng94, @audi18, and @grun18.
 This paper is just a software tutorial.
-We'll keep it practical.]
+We'll keep it practical.] We introduce three case studies to illustrate the workflow.
 
@@
 For each case study we focus on a different aspect to illustrate how to impute incomplete multilevel data.
-
-We introduce three case studies to illustrate the workflow.
 In the `mice::popmis` data, we show the advantages of including the multilevel structure of the data into the imputation model.
@@
 
-`popNCR` is a simulated dataset with pupils clustered in classes, where the number of cases $n = 2000$, and the number of clusters $N = 100$, on 7 variables:
+`popNCR` is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:
 
@@
 ```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-# plotting colors
-plot_col <- mice:::mdc(c("obs", "mis")) %>% setNames(c("observed", "missing"))
-
-# distributions
-plot_conditional <- function(data, x, z, cluster, ...){
-  ggplot(data, aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
-  geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
-  geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
-  geom_density() +
-  scale_color_manual(values = plot_col, name = z) +
-  theme_classic() +
-  theme(legend.position = "bottom") +
-  labs(x = x)
-}
-
-# plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
 plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
 # TODO: check if up-side-down plot works
-# or maybe overlapping densities
 # TODO: check smoothing in geom_density function and make it the inverse of the sample size
 # TODO: add functions to mice
-# TODO: legend: popteach obs/mis
-# TODO: add title
+# TODO: add title and informative legend
 # TODO: add facets for some clusters
@@
 # TODO: make it average of cluster dens, not marginal
-
-# Feedback: 
-# you only see the extreme ones in the positive sense, especially with 100 clusters
 # alternatively add a quartile line around the density with geom_ribbon
@@
 # further inspection of the imputations
-densityplot(imp_ignored)
+plot_box(imp_ignored, x = "popular", strip = TRUE)
 
@@
 # further inspection of the imputations
-densityplot(imp_predictor)
+plot_box(imp_predictor, x = "popular", strip = TRUE)
 
@@
 meth <- c("", "", "", "", "", "2l.norm", "")
-imp_norm <-
+imp_norm_2l <-
   mice(
@@
 # plot(imp_norm)
-densityplot(imp_norm, ~popular)
+plot_box(imp_norm_2l, x = "popular", strip = TRUE)
+
 ICCs <- ICCs %>% mutate(
-           norm = c(icc(popular ~ as.factor(class), complete(imp_norm)), 
-                    icc(popteach ~ as.factor(class), complete(imp_norm)), 
-                    icc(texp ~ as.factor(class), complete(imp_norm)))
+           norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
+                    icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
+                    icc(texp ~ as.factor(class), complete(imp_norm_2l)))
            )
@@
 meth <- c("", "", "", "", "", "2l.pan", "")
-imp_pan <-
+imp_pan_2l <-
   mice(
@@
 # plot(imp_pan)
-densityplot(imp_pan, ~popular)
+plot_box(imp_pan_2l, x = "popular", strip = TRUE)
 ICCs <- ICCs %>% mutate(
-           pan = c(icc(popular ~ as.factor(class), complete(imp_pan)), 
-                    icc(popteach ~ as.factor(class), complete(imp_pan)), 
-                    icc(texp ~ as.factor(class), complete(imp_pan)))
+           pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
+                    icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
+                    icc(texp ~ as.factor(class), complete(imp_pan_2l)))
            )
@@
 
-## Case study II: IMPACT
+# Case study II: IMPACT
 
@@
 
-## Pooling
+# Case study III: HIV
 
-- Analysis of scientific interest.
-
-- Pooling using `mitml`.
-
-- Pooling 'regular' parameters vs more 'exotic' parameters (SE of residual errors, or autocorrelation)
+Toy example from [Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
 

### 2021-12-13 Manuscript/Manuscript.tex
Add plotting functions for imputed data

@@
 we will refer to the grouping variable as `cluster', and the grouped
-variable as `(sample) unit'.\footnote{Add that we'll only discuss two
-  levels?} For reasons of brevity, we only discuss clustering between
-units, not within units (such as in timeseries or longitudinal data).
+variable as `(sample) unit'.\footnote{Alternatives to unit:
+  participant/response/record.} For reasons of brevity, we only discuss
+clustering between units, not within units (such as in timeseries or
+longitudinal data).\footnote{Also, add that we'll only discuss two
+  levels, no more?}
 
@@
   detail for the different methods (and equations). Refer to
-  \citet{meng94}, an Audigier paper, and a paper by Grund on
-  congeniality and random slopes. This paper is just a software
-  tutorial. We'll keep it practical.}
+  \citet{meng94}, \citet{audi18}, and \citet{grun18}. This paper is just
+  a software tutorial. We'll keep it practical.} We introduce three case
+studies to illustrate the workflow.
 
@@
 For each case study we focus on a different aspect to illustrate how to
-impute incomplete multilevel data.
-
-We introduce three case studies to illustrate the workflow. In the
-\texttt{mice::popmis} data, we show the advantages of including the
-multilevel structure of the data into the imputation model. In the
-\texttt{metamisc::impact} data we'll show how to induce missingness and
-solve it in real-world data. In the \texttt{GJRM::hiv} we provide novel
-methodology\footnote{not really, the methods exist already, but how to
-  show that this is something new and exciting?} for imputing MNAR
-missingness according to the Heckman model.
+impute incomplete multilevel data. In the \texttt{mice::popmis} data, we
+show the advantages of including the multilevel structure of the data
+into the imputation model. In the \texttt{metamisc::impact} data we'll
+show how to induce missingness and solve it in real-world data. In the
+\texttt{GJRM::hiv} we provide novel methodology\footnote{not really, the
+  methods exist already, but how to show that this is something new and
+  exciting?} for imputing MNAR missingness according to the Heckman
+model.
 
@@
 \texttt{popNCR} is a simulated dataset with pupils clustered in classes,
-where the number of cases \(n = 2000\), and the number of clusters
+where the number of units \(n = 2000\), and the number of clusters
 \(N = 100\), on 7 variables:
@@
       vars incomplete   ignored predictor
-1  popular  0.3280070 0.3338626 0.3726455
+1  popular  0.3280070 0.3338626 0.3871020
 2 popteach  0.3414766 0.3414766 0.3414766
@@
 R> meth <- c("", "", "", "", "", "2l.norm", "")
-R> imp_norm <-
+R> imp_norm_2l <-
 +   mice(
@@
 R> # plot(imp_norm)
-R> densityplot(imp_norm, ~popular)
+R> plot_box(imp_norm_2l, x = "popular", strip = TRUE)
 \end{CodeInput}
@@
 R> ICCs <- ICCs %>% mutate(
-+            norm = c(icc(popular ~ as.factor(class), complete(imp_norm)), 
-+                     icc(popteach ~ as.factor(class), complete(imp_norm)), 
-+                     icc(texp ~ as.factor(class), complete(imp_norm)))
++            norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
++                     icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
++                     icc(texp ~ as.factor(class), complete(imp_norm_2l)))
 +            )
@@
       vars incomplete   ignored predictor      norm
-1  popular  0.3280070 0.3338626 0.3726455 0.3667167
+1  popular  0.3280070 0.3338626 0.3871020 0.3612830
 2 popteach  0.3414766 0.3414766 0.3414766 0.3414766
@@
 R> meth <- c("", "", "", "", "", "2l.pan", "")
-R> imp_pan <-
+R> imp_pan_2l <-
 +   mice(
@@
 R> # plot(imp_pan)
-R> densityplot(imp_pan, ~popular)
+R> plot_box(imp_pan_2l, x = "popular", strip = TRUE)
 \end{CodeInput}
@@
 R> ICCs <- ICCs %>% mutate(
-+            pan = c(icc(popular ~ as.factor(class), complete(imp_pan)), 
-+                     icc(popteach ~ as.factor(class), complete(imp_pan)), 
-+                     icc(texp ~ as.factor(class), complete(imp_pan)))
++            pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
++                     icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
++                     icc(texp ~ as.factor(class), complete(imp_pan_2l)))
 +            )
@@
       vars incomplete   ignored predictor      norm       pan
-1  popular  0.3280070 0.3338626 0.3726455 0.3667167 0.3719078
+1  popular  0.3280070 0.3338626 0.3871020 0.3612830 0.3745108
 2 popteach  0.3414766 0.3414766 0.3414766 0.3414766 0.3414766
@@
 \hypertarget{case-study-ii-impact}{%
-\subsection{Case study II: IMPACT}\label{case-study-ii-impact}}
+\section{Case study II: IMPACT}\label{case-study-ii-impact}}
 
@@
 
-\hypertarget{pooling}{%
-\subsection{Pooling}\label{pooling}}
+\hypertarget{case-study-iii-hiv}{%
+\section{Case study III: HIV}\label{case-study-iii-hiv}}
 
-\begin{itemize}
-\item
-  Analysis of scientific interest.
-\item
-  Pooling using \texttt{mitml}.
-\item
-  Pooling `regular' parameters vs more `exotic' parameters (SE of
-  residual errors, or autocorrelation)
-\end{itemize}
+Toy example from
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{Heckman
+Github repo}.
 

### 2021-12-13 R/impact_functions.r

@@ -89,12 +89,22 @@ fit_missingness_models <- function(fname) {
       if (pct_missing[vname] > 0 &  pct_missing[vname] < 100) {
         results[nrow(results),cnames] <- coef(fit)
       }
-      
-      
-      
     }
   }
   
   return(results)
+
+}
+
+ampute_impact <- function(dat, p_obs){
+  # pre-process coefficients
+  p_obs[is.na(p_obs)] <- 0
+  betas <- split(p_obs, p_obs$trial)
+  betas <- purrr::map(betas, ~{.[, 4:17] %>% as.data.frame(row.names = unique(p_obs$variable_missing))})
+
+  # create model matrix
+  mm <- purrr::map(split(dat, dat$name), ~{model.matrix(name ~ pupil + ct + hypox + hypots + tsah + edh + age + motor_score + mort, .x) %>% as.data.frame()})
   
+  # calculate missingness probability
+  a <- purrr::map2(betas, mm, ~{as.matrix(.y) %*% t(.x[])})
 }

### 2021-12-13 R/plot_md.R

@@ -1,3 +1,44 @@
+# plotting functions for incomplete data
+# missingness indicator plot (monotone vs non-monotone)
+plot_md_set <- function(...){
+  miss_ind <- tibble(
+  rownr = 5:1,
+  unit = rep("observed", 5),
+  cluster = rep("observed", 5),
+  X1 = rep("observed", 5),
+  X2 = c(rep("observed", 3), rep("missing", 2)),
+  X3 = c("missing", rep("observed", 3), "missing")
+) %>% pivot_longer(cols = c(unit, cluster, X1, X2, X3))
+
+# add text 
+miss_ind$text <- ""
+miss_ind[miss_ind$name == "unit", "text"] <- as.character(1:5)
+miss_ind[miss_ind$name == "cluster", "text"] <- as.character(c(1,1,1,2,2))
+
+# plot
+miss_ind %>% 
+  mutate(name = factor(name, levels = unique(name))) %>% 
+  ggplot(aes(
+    x = name,
+    y = rownr,
+    color = value,
+    width = 0.8, 
+    height = 0.8,
+    label = text
+  )) +
+  geom_tile(fill = "white", size = 2) +
+  geom_text(show.legend = FALSE) +
+  scale_x_discrete(position = "top") +
+  scale_y_discrete(labels = 5:1) +
+  scale_color_manual(values = plot_col, name = "") +
+  theme_minimal() +
+  theme(
+    legend.position = "bottom", 
+    panel.grid.major = element_blank()) +
+  labs(x = "", y = "")
+}
+
+
 # missing data pattern
 plot_md_pat <- function(dat) {
   # get md pattern and store additional info
@@ -93,6 +134,19 @@ plot_md_pat <- function(dat) {
   return(p)
 }
 
+# plot conditional distributions
+plot_conditional <- function(data, x, z, cluster, ...){
+  ggplot(data, aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
+    geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
+    geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
+    geom_density() +
+    scale_color_manual(values = plot_col, name = z) +
+    theme_classic() +
+    theme(legend.position = "bottom") +
[…]

### 2021-12-13 Manuscript/Manuscript.Rmd
Add note about univariate missingness in popularity data

@@
 ```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
-# load data and merge classes into schools for this tutorial -> but this cannot work because how texp and popteach are defined...
+# load data 
 pop <- readRDS("../Data/popNCR2.RDS") %>% mutate(class = as.factor(class))
@@
 
-The missingness in this dataset is induced conform MAR and MNAR mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows the systematic nature of the missingness.
+The missingness in this dataset is induced conform MAR and MNAR mechanisms.
+The missing data pattern, Figure \ref{fig:pop_pat}, shows that just one variable is incomplete **[the next part is not yet updated to reflect this]**.
 

### 2021-12-15 Manuscript/Manuscript.tex
Edit missing data matrix

@@
 \Abstract{
-Tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
-Including methods for ignorable and non-ignorable missingness.
+Multilevel data is not spared the ubiquitous problem of missing
+information. This is a tutorial paper on imputing incomplete multilevel
+data with \pkg{mice}. Including methods for ignorable and non-ignorable
+missingness. Footnotes in the current version show work in
+progress/under construction. The last section is not part of the
+manuscript, but purely for reminders.
 }
@@
 
-We talk of multilevel data when there is some kind of hierarchy or
-clustering in a dataset. In the typical case, individuals are nested
-within groups, but there are many different types of multilevel data. In
-the medical field, clustering occurs at e.g., the hospitals/center level
-in registry data, or at the study-level in meta-analyses (IPDMA). In the
-social sciences and official statistics we can find clustering e.g.~at
-the country-level, or as imposed by the sampling design. In this paper,
-we will refer to the grouping variable as `cluster', and the grouped
-variable as `(sample) unit'.\footnote{Alternatives to unit:
-  participant/response/record.} For reasons of brevity, we only discuss
-clustering between units, not within units (such as in timeseries or
-longitudinal data).\footnote{Also, add that we'll only discuss two
-  levels, no more?}
+Research into any field with a hierarchical or clustered nature of
+observations may yield multilevel data. In the typical case, individuals
+are nested within groups, but there are many different types of
+multilevel data. In the medical field, clustering occurs at e.g., the
+hospital or center level in registry data, or at the study-level in
+meta-analyses (IPDMA\footnote{remove or write in full?}). In the social
+sciences and official statistics we can find clustering e.g.~at the
+country-level, or as imposed by a multi-stage sampling design. For the
+sake of legibility, we will refer to the grouping variable as `cluster',
+and the grouped variable as `(sample) unit' throughout this
+paper.\footnote{Alternatives to unit: participant/response/record.} And,
+for reasons of brevity, we only discuss clustering between units, not
+within units (e.g., timeseries or longitudinal data).\footnote{Also, add
+  that we'll only discuss two levels, not more?}
 
 Analyzing multilevel data requires special care, compared to `regular',
-single level data. The cluster to which a unit belongs may influence the
-unit-level observations, and since clusters are made up of units,
-clusters depend on units as well \citep{hox17}. These relations can and
-should be taken into account when developing analysis models for
-multilevel data.\footnote{Explain ICC here? The percentage of variance
-  attributed to the cluster-level is expressed by the intra-class
-  coefficient (ICC). The ICC can also be interpreted as the expected
-  correlation between two randomly sampled units in same cluster. So if
-  the ICC is high, a lot of variability in a variable is due to the
-  clustering, which should be modeled accordingly.} Multilevel models
-typically include separate intercepts for each cluster, which relieves
-one restriction imposed by single-level models: equal group means across
-clusters. Additionally, there may be random predictor effects and/or
-random error terms (residual error variances), see e.g. \citet{hox17}
-and \citet{jong21}.\footnote{Add that heterogeneity refers to
-  variability within clusters.} There are many names for models that
-take clustering into account. Some popular examples are `multilevel
-models', `hierarchical models', `mixed effect models' and `random effect
-models'.
+single level data. For instance, there may be cross-level interactions
+between unit-level variables and cluster-level variables. The cluster to
+which a unit belongs may influence the unit-level observations, and vice
+versa for the the units that make up the cluster \citep{hox17}. These
+relations can and should be taken into account when developing analysis
+models for multilevel data.\footnote{Explain ICC here? The percentage of
+  variance attributed to the cluster-level is expressed by the
+  intra-class coefficient (ICC). The ICC can also be interpreted as the
+  expected correlation between two randomly sampled units in same
+  cluster. So if the ICC is high, a lot of variability in a variable is
+  due to the clustering, which should be modeled accordingly.}
+Multilevel models typically include separate intercepts for each
+cluster, which relieves one restriction imposed by single-level models:
+equal group means across clusters. Additionally, there may be random
+predictor effects and/or random error terms (residual error variances),
+see e.g. \citet{hox17} and \citet{jong21}.\footnote{Add that
+  heterogeneity refers to variability within clusters.} There are many
+names for models that take clustering into account. Some popular
+examples are `multilevel models', `hierarchical models', `mixed effect
+models' and `random effect models'.
 
@@
 
+Multilevel data is not spared the ubiquitous problem of missing
+information. Just as in single level data, missingness may occur at the
+unit level. But with multiple levels of data comes the potential for
+missingness at multiple levels. Missingness in multilevel data can be
+categorized into two general patterns: systematical missingness and
+sporadic missingness, see \citet{resc13}. In figure 1, we show a dataset
+with units in the rows and variables in the columns, there are 5 units
+nested within 3 clusters, and 3 variables of interest. Variable
+\texttt{X1} is completely observed. Variable \texttt{X2} is
+systematically missing, \texttt{X3} is sporadically missing.\footnote{Explain
+  why.} Systematic missingness can be further subdivided into unobserved
+constants (i.e., the same value within clusters) and non-measured random
+variables (which may differ per unit within clusters). In Figure 1, the
+former implies that the unobserved values for units 3 and 4 on variable
+\texttt{X2} would be equal. With the latter, the values may differ.
+Depending on the missing data pattern\footnote{add missing data
+  mechanisms here?}, there are more or less optimal way of accommodating
+the missingness.
+
 \[
@@
 \]
-
-Multilevel data is not spared the ubiquitous problem of missing
-information. Just as in single level data, missingness may occur at the
-unit level. But with multiple levels of data comes the potential for
-missingness at multiple levels. Missingness in multilevel data can be
-categorized into two general patterns: systematical missingness and
-sporadic missingness, see \citet{resc13}. In figure 1, we show a dataset
-with units in the rows and variables in the columns, there are 5 units
-nested within 2 clusters, and 3 variables of interest. Variable
-\texttt{X1} is completely observed. Variable \texttt{X2} is
-systematically missing, \texttt{X3} is sporadically missing.\footnote{Explain
-  why.} Systematic missingness can be further subdivided into unobserved
-constants (same value per cluster) and non-measured random variables
-(which may differ per unit within clusters). In Figure 1, the former
-implies that the unobserved values for units 4 and 5 on variable
-\texttt{X2} would be equal. With the latter, the values may differ.
-Depending on the missing data pattern, there are more or less optimal
-way of accommodating the missingness.
 
@@
 different missing data situations. For reasons of brevity, we focus on
-imputation by chained equations, although JOMO is available in
-\pkg{mice} as well. Other useful packages for incomplete multilevel data
-include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.\footnote{Rephrase:
+imputation by chained equations\footnote{add that JOMO is available in
+  \pkg{mice} as well?}. Other useful packages for incomplete multilevel
+data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.\footnote{Rephrase:
   Some level of knowledge on multilevel models is assumed. We're
@@
   \citet{meng94}, \citet{audi18}, and \citet{grun18}. This paper is just
-  a software tutorial. We'll keep it practical.} We introduce three case
-studies to illustrate the workflow.
+  a software tutorial. We'll keep it practical.}
 
@@
   exciting?} for imputing MNAR missingness according to the Heckman
-model.
-
-For each case study we'll look at least at: 1) the incomplete data; 2)
-the imputation model; 3) the imputed data; and 4) how to obtain pooled
-estimates for the analysis of scientific interest.
+model. For all case studies we discuss the nature of the incomplete
+data, the imputation model(s), and evaluation of the imputed data.
 
@@
 mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows
-the systematic nature of the missingness.
+that just one variable is incomplete \textbf{{[}the next part is not yet
+updated to reflect this{]}}.
 

### 2021-12-15 R/impact_functions.r

@@ -96,15 +96,23 @@ fit_missingness_models <- function(fname) {
 
 }
 
-ampute_impact <- function(dat, p_obs){
+ampute_impact <- function(impact, p_obs){
+  # just for function development 
+  # load("../Data/p_observed_impact.rda")
+  # p_obs <- results
+  
   # pre-process coefficients
   p_obs[is.na(p_obs)] <- 0
   betas <- split(p_obs, p_obs$trial)
   betas <- purrr::map(betas, ~{.[, 4:17] %>% as.data.frame(row.names = unique(p_obs$variable_missing))})
 
-  # create model matrix
-  mm <- purrr::map(split(dat, dat$name), ~{model.matrix(name ~ pupil + ct + hypox + hypots + tsah + edh + age + motor_score + mort, .x) %>% as.data.frame()})
+  # create model matrices
+  mm <- purrr::map(split(impact, impact$name), ~{model.matrix(name ~ pupil + ct + hypox + hypots + tsah + edh + age + motor_score + mort, .x) %>% as.data.frame()})
   
-  # calculate missingness probability
-  a <- purrr::map2(betas, mm, ~{as.matrix(.y) %*% t(.x[])})
+  # calculate linear predictor for missingness probability and convert to prob_obs
+  pr <- purrr::map2(betas, mm, ~{as.matrix(.y) %*% t(.x[]) %>% plogis() %>% as.data.frame()})
+  
+  # select lowest values to make incomplete
+  
+  
 }

### 2021-12-15 Manuscript/Manuscript.Rmd

@@
 abstract: >
-  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. 
+  Multilevel data is not spared the ubiquitous problem of missing information.
+This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
+Including methods for ignorable and non-ignorable missingness.
+Footnotes in the current version show work in progress/under construction.
+The last section is not part of the manuscript, but purely for reminders.
 keywords:
@@
 
-We talk of multilevel data when there is some kind of hierarchy or clustering in a dataset.
+Research into any field with a hierarchical or clustered nature of observations may yield multilevel data.
 In the typical case, individuals are nested within groups, but there are many different types of multilevel data.
-In the medical field, clustering occurs at e.g., the hospitals/center level in registry data, or at the study-level in meta-analyses (IPDMA).
+In the medical field, clustering occurs at e.g., the hospital or center level in registry data, or at the study-level in meta-analyses (IPDMA^[remove or write in full?]).
 In the social sciences and official statistics we can find clustering e.g.
-at the country-level, or as imposed by the sampling design.
-In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.^[Alternatives to unit: participant/response/record.] For reasons of brevity, we only discuss clustering between units, not within units (such as in timeseries or longitudinal data).^[Also, add that we'll only discuss two levels, no more?]
+at the country-level, or as imposed by a multi-stage sampling design.
+For the sake of legibility, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit' throughout this paper.^[Alternatives to unit: participant/response/record.] And, for reasons of brevity, we only discuss clustering between units, not within units (e.g., timeseries or longitudinal data).^[Also, add that we'll only discuss two levels, not more?]
 
 Analyzing multilevel data requires special care, compared to 'regular', single level data.
-The cluster to which a unit belongs may influence the unit-level observations, and since clusters are made up of units, clusters depend on units as well [@hox17].
+For instance, there may be cross-level interactions between unit-level variables and cluster-level variables.
+The cluster to which a unit belongs may influence the unit-level observations, and vice versa for the the units that make up the cluster [@hox17].
 These relations can and should be taken into account when developing analysis models for multilevel data.^[Explain ICC here?
@@
 
-```{r patterns, fig.height=4, fig.width=3.5, fig.cap = "Missingness in multilevel data", echo=FALSE}
+```{r patterns, fig.height=4, fig.width=3.2, fig.cap = "Missingness in multilevel data", echo=FALSE}
 plot_md_set()
@@
 ```
+
+Multilevel data is not spared the ubiquitous problem of missing information.
+Just as in single level data, missingness may occur at the unit level.
+But with multiple levels of data comes the potential for missingness at multiple levels.
+Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13.
+In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 3 clusters, and 3 variables of interest.
+Variable `X1` is completely observed.
+Variable `X2` is systematically missing, `X3` is sporadically missing.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters).
+In Figure 1, the former implies that the unobserved values for units 3 and 4 on variable `X2` would be equal.
+With the latter, the values may differ.
+Depending on the missing data pattern^[add missing data mechanisms here?], there are more or less optimal way of accommodating the missingness.
 
@@
 $$
-
-Multilevel data is not spared the ubiquitous problem of missing information.
-Just as in single level data, missingness may occur at the unit level.
-But with multiple levels of data comes the potential for missingness at multiple levels.
-Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13.
-In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 2 clusters, and 3 variables of interest.
-Variable `X1` is completely observed.
-Variable `X2` is systematically missing, `X3` is sporadically missing.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (same value per cluster) and non-measured random variables (which may differ per unit within clusters).
-In Figure 1, the former implies that the unobserved values for units 4 and 5 on variable `X2` would be equal.
-With the latter, the values may differ.
-Depending on the missing data pattern, there are more or less optimal way of accommodating the missingness.
 
@@
 We provide practical guidelines and code snippets for different missing data situations.
-For reasons of brevity, we focus on imputation by chained equations, although JOMO is available in \pkg{mice} as well.
-Other useful packages for incomplete multilevel data include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed.
+For reasons of brevity, we focus on imputation by chained equations^[add that JOMO is available in \pkg{mice} as well?].
+Other useful packages for incomplete multilevel data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed.
 We're providing an overview of implementations.
@@
 This paper is just a software tutorial.
-We'll keep it practical.] We introduce three case studies to illustrate the workflow.
+We'll keep it practical.]
 
@@
 In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model.
-
-For each case study we'll look at least at: 1) the incomplete data; 2) the imputation model; 3) the imputed data; and 4) how to obtain pooled estimates for the analysis of scientific interest.
+For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data.
 
@@
 plot_box(imp_ignored, x = "popular", strip = TRUE)
+# TODO: think about multimodality
 

### 2021-12-15 Manuscript/Manuscript.Rmd

@@
 library(metamisc)
+library(GJRM)
 library(pan)
@@
 
-Toy example from [Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
+Toy example from [Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R). We will use the following variables:
+
+  - `region` Cluster variable,
+  - `hiv` HIV diagnosis (0=no, 1=yes),
+  - `age` Age of the patient,
+  - `marital` Marital status,
+  - `condom` Condom use during last intercourse,
+  - `smoke` Smoker (levels).
+
+```{r hiv, echo=FALSE}
+# data
+data("hiv")
+hiv <- hiv %>% 
+  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
+  mutate(region = as.integer(region))
+# md pattern
+plot_md_pat(hiv)
+```
 

### 2021-12-16 Manuscript/Manuscript.Rmd
Some suggestions to the intro

@@
  - \usepackage{mathtools}
+ - \usepackage{ulem}
 output: rticles::jss_article
@@
 
-Research into any field with a hierarchical or clustered nature of observations may yield multilevel data.
+~~Research into any field with a hierarchical or clustered nature of observations may yield multilevel data.
 In the typical case, individuals are nested within groups, but there are many different types of multilevel data.
-In the medical field, clustering occurs at e.g., the hospital or center level in registry data, or at the study-level in meta-analyses (IPDMA^[remove or write in full?]).
+In the medical field, clustering occurs at e.g., the hospital or center level in registry data, or at the study-level in meta-analyses (IPDMA).
 In the social sciences and official statistics we can find clustering e.g.
 at the country-level, or as imposed by a multi-stage sampling design.
-For the sake of legibility, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit' throughout this paper.^[Alternatives to unit: participant/response/record.] And, for reasons of brevity, we only discuss clustering between units, not within units (e.g., timeseries or longitudinal data).^[Also, add that we'll only discuss two levels, not more?]
+For the sake of legibility, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit' throughout this paper.
+And, for reasons of brevity, we only discuss clustering between units, not within units (e.g., timeseries or longitudinal data).
+Also, add that we'll only discuss two levels, not more?~~
 
-Analyzing multilevel data requires special care, compared to 'regular', single level data.
-For instance, there may be cross-level interactions between unit-level variables and cluster-level variables.
-The cluster to which a unit belongs may influence the unit-level observations, and vice versa for the the units that make up the cluster [@hox17].
-These relations can and should be taken into account when developing analysis models for multilevel data.^[Explain ICC here?
-The percentage of variance attributed to the cluster-level is expressed by the intra-class coefficient (ICC).
-The ICC can also be interpreted as the expected correlation between two randomly sampled units in same cluster.
-So if the ICC is high, a lot of variability in a variable is due to the clustering, which should be modeled accordingly.] Multilevel models typically include separate intercepts for each cluster, which relieves one restriction imposed by single-level models: equal group means across clusters.
+In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.
+Ignoring such structures may be harmful to the inferences and can yield biased results.
+On the other hand, analyzing such multilevel data requires specialized techniques that take the clustered structure into account.
+Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
+The cluster to which a unit belongs may then influence the unit-level observations - and vice versa - for each of the units that make up the cluster [@hox17].
+These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
+
+The variability due to clustering is often measured by means of the intra-class coefficient (ICC).
+The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.
+Multilevel models typically accommodate for this variability by including separate intercepts for each cluster.
+Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters.
 Additionally, there may be random predictor effects and/or random error terms (residual error variances), see e.g.
@@
 
-Multilevel data is not spared the ubiquitous problem of missing information.
-Just as in single level data, missingness may occur at the unit level.
-But with multiple levels of data comes the potential for missingness at multiple levels.
-Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13.
-In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 3 clusters, and 3 variables of interest.
-Variable `X1` is completely observed.
-Variable `X2` is systematically missing, `X3` is sporadically missing.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters).
-In Figure 1, the former implies that the unobserved values for units 3 and 4 on variable `X2` would be equal.
-With the latter, the values may differ.
-Depending on the missing data pattern^[add missing data mechanisms here?], there are more or less optimal way of accommodating the missingness.
+The process of analyzing multilevel data is further complicated when not all observations are observed.
+Just as with single level data, missingness may occur at the unit level.
+But with multiple levels of data comes the potential for clustered missingness.
+Missingness in multilevel data can therefore be categorized into two general patterns: systematic missingness and sporadic missingness, see @resc13.
+We have visualized the difference between these types of missingness in Figure 1.
+The figure shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n=5$ units and $p=3$ columns distributed over 3 clusters.
+Column `X1` is completely observed.
+Column `X2` is systematically missing and column `X3` is sporadically missing.^[Explain why.
+GV: niet nodig]
+
+Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters).
+In Figure 1, the former would imply that the unobserved values for units 3 and 4 on column `X2` are identical.
+With the latter, the values would differ.
+The optimal strategy for dealing with the missingness may therefore depend on the observed missing data pattern.
+^[add missing data mechanisms here?
+GV: No, patterns are different from mechanisms]
 
@@
    \hline
-   \mathrm{cluster 1}\{   &x_{11} &x_{12}       &\mathtt{NA}    \\
-                          &x_{21} &x_{22}       &x_{32}         \\
+   \mathrm{cluster 1}\{ \ &x_{11} &x_{12}       &\mathtt{NA}    \\
+                             \ &x_{21}  &x_{22}       &x_{32}         \\
    \mathrm{cluster 2}\{  &x_{31} &\mathtt{NA}  &x_{33}         \\
@@
 $$
+[GV Ik vind de latex tabel niet tof. ]
 
-Ignoring missing data in research endeavors is almost never a good idea.
+Ignoring the missingness in analyses can be extremely harmful to inferences.
 Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power.
 Instead, the missingness should be accommodated *before* or *within* the analysis of scientific interest.
-Especially the former is very generic and popular.
-Imputing (i.e., filling in) the missing values splits the missing data problem from the scientific problem.
-The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which solves the missingness one variable at a time, iteratively.
+Especially the former is very generic and popular ans is widely known as imputation.
+Imputing (i.e., filling in) the missing values seperates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed.
+The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
 \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].

### 2021-12-16 Manuscript/Manuscript.tex
Some suggestions to the intro

@@
 \usepackage{mathtools}
+\usepackage{ulem}
 
@@
 
-Research into any field with a hierarchical or clustered nature of
+\sout{Research into any field with a hierarchical or clustered nature of
 observations may yield multilevel data. In the typical case, individuals
@@
 hospital or center level in registry data, or at the study-level in
-meta-analyses (IPDMA\footnote{remove or write in full?}). In the social
-sciences and official statistics we can find clustering e.g.~at the
-country-level, or as imposed by a multi-stage sampling design. For the
-sake of legibility, we will refer to the grouping variable as `cluster',
-and the grouped variable as `(sample) unit' throughout this
-paper.\footnote{Alternatives to unit: participant/response/record.} And,
-for reasons of brevity, we only discuss clustering between units, not
-within units (e.g., timeseries or longitudinal data).\footnote{Also, add
-  that we'll only discuss two levels, not more?}
+meta-analyses (IPDMA). In the social sciences and official statistics we
+can find clustering e.g.~at the country-level, or as imposed by a
+multi-stage sampling design. For the sake of legibility, we will refer
+to the grouping variable as `cluster', and the grouped variable as
+`(sample) unit' throughout this paper. And, for reasons of brevity, we
+only discuss clustering between units, not within units (e.g.,
+timeseries or longitudinal data). Also, add that we'll only discuss two
+levels, not more?}
 
-Analyzing multilevel data requires special care, compared to `regular',
-single level data. For instance, there may be cross-level interactions
-between unit-level variables and cluster-level variables. The cluster to
-which a unit belongs may influence the unit-level observations, and vice
-versa for the the units that make up the cluster \citep{hox17}. These
-relations can and should be taken into account when developing analysis
-models for multilevel data.\footnote{Explain ICC here? The percentage of
-  variance attributed to the cluster-level is expressed by the
-  intra-class coefficient (ICC). The ICC can also be interpreted as the
-  expected correlation between two randomly sampled units in same
-  cluster. So if the ICC is high, a lot of variability in a variable is
-  due to the clustering, which should be modeled accordingly.}
-Multilevel models typically include separate intercepts for each
-cluster, which relieves one restriction imposed by single-level models:
-equal group means across clusters. Additionally, there may be random
-predictor effects and/or random error terms (residual error variances),
-see e.g. \citet{hox17} and \citet{jong21}.\footnote{Add that
-  heterogeneity refers to variability within clusters.} There are many
-names for models that take clustering into account. Some popular
-examples are `multilevel models', `hierarchical models', `mixed effect
-models' and `random effect models'.
+In many contemporary data analysis efforts, some form of hierarchical or
+clustered structure is recorded. Ignoring such structures may be harmful
+to the inferences and can yield biased results. On the other hand,
+analyzing such multilevel data requires specialized techniques that take
+the clustered structure into account. Imagine a case where cross-level
+interactions between unit-level variables and cluster-level variables
+are present. The cluster to which a unit belongs may then influence the
+unit-level observations - and vice versa - for each of the units that
+make up the cluster \citep{hox17}. These relations can and should be
+taken into account when developing analysis models for multilevel data
+for the simple reason that groups of observations share some common
+variance.
+
+The variability due to clustering is often measured by means of the
+intra-class coefficient (ICC). The ICC can be seen as the percentage of
+variance that can be attributed to the cluster-level, where a high ICC
+would indicate that a lot of variability is due to the cluster
+structure. Multilevel models typically accommodate for this variability
+by including separate intercepts for each cluster. Such fixed effects
+relieve the restriction imposed by single-level models: equal group
+means across clusters. Additionally, there may be random predictor
+effects and/or random error terms (residual error variances), see e.g.
+\citet{hox17} and \citet{jong21}.\footnote{Add that heterogeneity refers
+  to variability within clusters.} There are many names for models that
+take clustering into account. Some popular examples are `multilevel
+models', `hierarchical models', `mixed effect models' and `random effect
+models'.
 
@@
 
-Multilevel data is not spared the ubiquitous problem of missing
-information. Just as in single level data, missingness may occur at the
-unit level. But with multiple levels of data comes the potential for
-missingness at multiple levels. Missingness in multilevel data can be
-categorized into two general patterns: systematical missingness and
-sporadic missingness, see \citet{resc13}. In figure 1, we show a dataset
-with units in the rows and variables in the columns, there are 5 units
-nested within 3 clusters, and 3 variables of interest. Variable
-\texttt{X1} is completely observed. Variable \texttt{X2} is
-systematically missing, \texttt{X3} is sporadically missing.\footnote{Explain
-  why.} Systematic missingness can be further subdivided into unobserved
+The process of analyzing multilevel data is further complicated when not
+all observations are observed. Just as with single level data,
+missingness may occur at the unit level. But with multiple levels of
+data comes the potential for clustered missingness. Missingness in
+multilevel data can therefore be categorized into two general patterns:
+systematic missingness and sporadic missingness, see \citet{resc13}. We
+have visualized the difference between these types of missingness in
+Figure 1. The figure shows an \(n \times p\) set
+\(\mathbf{X} = X_1, \dots, X_p\), with \(n=5\) units and \(p=3\) columns
+distributed over 3 clusters. Column \texttt{X1} is completely observed.
+Column \texttt{X2} is systematically missing and column \texttt{X3} is
+sporadically missing.\footnote{Explain why. GV: niet nodig}
+
+Systematic missingness can be further subdivided into unobserved
 constants (i.e., the same value within clusters) and non-measured random
 variables (which may differ per unit within clusters). In Figure 1, the
-former implies that the unobserved values for units 3 and 4 on variable
-\texttt{X2} would be equal. With the latter, the values may differ.
-Depending on the missing data pattern\footnote{add missing data
-  mechanisms here?}, there are more or less optimal way of accommodating
-the missingness.
+former would imply that the unobserved values for units 3 and 4 on
+column \texttt{X2} are identical. With the latter, the values would
+differ. The optimal strategy for dealing with the missingness may
+therefore depend on the observed missing data pattern. \footnote{add
+  missing data mechanisms here? GV: No, patterns are different from
+  mechanisms}
 
@@
 \end{matrix} 
-\]
+\] {[}GV Ik vind de latex tabel niet tof. {]}
 
-Ignoring missing data in research endeavors is almost never a good idea.
-Complete case analysis (i.e., excluding all units with one or more
-missing entries) can introduce bias in statistical inference and lowers
-statistical power. Instead, the missingness should be accommodated
-\emph{before} or \emph{within} the analysis of scientific interest.
-Especially the former is very generic and popular. Imputing (i.e.,
-filling in) the missing values splits the missing data problem from the
-scientific problem. The \proglang{R} package \pkg{mice} has become the
-de-facto standard for imputation by chained equations, which solves the
-missingness one variable at a time, iteratively. \pkg{mice} is known to
-yield valid inferences under many different missing data circumstances
+Ignoring the missingness in analyses can be extremely harmful to
+inferences. Complete case analysis (i.e., excluding all units with one
+or more missing entries) can introduce bias in statistical inference and
+lowers statistical power. Instead, the missingness should be
+accommodated \emph{before} or \emph{within} the analysis of scientific
+interest. Especially the former is very generic and popular ans is
+widely known as imputation. Imputing (i.e., filling in) the missing
+values seperates the missing data problem from the scientific problem:
+missing data are replaced by plausible values whereafter the completed
+data is analysed as if it were completely observed. The \proglang{R}
+package \pkg{mice} has become the de-facto standard for imputation by
+chained equations, which iteratively solves the missingness on a
+variable-by-variable basis. \pkg{mice} is known to yield valid
+inferences under many different missing data circumstances
 \citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
@@
 \href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{Heckman
-Github repo}.
+Github repo}. We will use the following variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{region} Cluster variable,
+\item
+  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{marital} Marital status,
+\item
+  \texttt{condom} Condom use during last intercourse,
+\item
+  \texttt{smoke} Smoker (levels).
+\end{itemize}
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/hiv-1} \end{center}
+
+\end{CodeChunk}
 

### 2021-12-16 Manuscript/Manuscript.Rmd
attempt to add accolade to md plot

@@
 Instead, the missingness should be accommodated *before* or *within* the analysis of scientific interest.
-Especially the former is very generic and popular ans is widely known as imputation.
-Imputing (i.e., filling in) the missing values seperates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed.
+Especially the former is very generic and popular and is widely known as imputation.
+Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed.
 The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.

### 2021-12-16 Manuscript/Manuscript.tex
attempt to add accolade to md plot

@@
 accommodated \emph{before} or \emph{within} the analysis of scientific
-interest. Especially the former is very generic and popular ans is
+interest. Especially the former is very generic and popular and is
 widely known as imputation. Imputing (i.e., filling in) the missing
-values seperates the missing data problem from the scientific problem:
+values separates the missing data problem from the scientific problem:
 missing data are replaced by plausible values whereafter the completed

### 2021-12-16 R/plot_md.R

@@ -23,11 +23,10 @@ miss_ind %>%
     y = rownr,
     color = value,
     width = 0.8, 
-    height = 0.8,
-    label = text
-  )) +
+    height = 0.8
+    )) +
   geom_tile(fill = "white", size = 2) +
-  geom_text(show.legend = FALSE) +
+  geom_label(data = data.frame(name = "cluster", rownr = 1, value = "observed", label = "{"), aes(label = label)) +
   scale_x_discrete(position = "top") +
   scale_y_discrete(labels = 5:1) +
   scale_color_manual(values = plot_col, name = "") +
@@ -38,7 +37,7 @@ miss_ind %>%
   labs(x = "", y = "")
 }
 
-
+plot_md_set()
 # missing data pattern
 plot_md_pat <- function(dat) {
   # get md pattern and store additional info

### 2021-12-16 R/plot_md.R
Add clusters to missing data figure

@@ -3,41 +3,39 @@
 plot_md_set <- function(...){
   miss_ind <- tibble(
   rownr = 5:1,
-  # unit = rep("observed", 5),
-  cluster = rep("observed", 5),
   X1 = rep("observed", 5),
   X2 = c(rep("observed", 2), rep("missing", 2), "observed"),
   X3 = c("missing", rep("observed", 2), "missing", "observed")
-) %>% pivot_longer(cols = c(cluster, X1, X2, X3))
-
-# add text 
-miss_ind$text <- ""
-# miss_ind[miss_ind$name == "unit", "text"] <- as.character(1:5)
-miss_ind[miss_ind$name == "cluster", "text"] <- as.character(c(1,1,2,2,3))
+) %>% pivot_longer(cols = c(X1, X2, X3))
 
 # plot
 miss_ind %>% 
   mutate(name = factor(name, levels = unique(name))) %>% 
   ggplot(aes(
     x = name,
-    y = rownr,
+    y = as.factor(rownr),
     color = value,
     width = 0.8, 
     height = 0.8
     )) +
-  geom_tile(fill = "white", size = 2) +
-  geom_label(data = data.frame(name = "cluster", rownr = 1, value = "observed", label = "{"), aes(label = label)) +
-  scale_x_discrete(position = "top") +
-  scale_y_discrete(labels = 5:1) +
-  scale_color_manual(values = plot_col, name = "") +
+  geom_tile(fill = "white", size = 1.2) +
+  scale_x_discrete(position = "top", labels = c("", "", "X1", "X2", "X3")) +
+  scale_y_discrete() +
+  scale_color_manual(values = plot_col, name = "Legend:") +
   theme_minimal() +
+  labs(x = "", y = "") +
+  annotate(geom = "text", x = "X-1", y = "4", label = "Cluster 1 ", vjust = -1, hjust = 0, size = 3) +
+  annotate(geom = "text", x = "X-1", y = "2", label = "Cluster 2 ", vjust = -1, hjust = 0, size = 3) +
+  annotate(geom = "text", x = "X-1", y = "0", label = "Cluster N ", vjust = -1, hjust = 0, size = 3) +
+  annotate(geom = "text", x = "X0", y = "4", label = "{", vjust = 0, hjust = 0, size = 16) +
+  annotate(geom = "text", x = "X0", y = "2", label = "{", vjust = 0, hjust = 0, size = 16) +
[…]

### 2021-12-16 Manuscript/Manuscript.Rmd

@@
 
-```{r patterns, fig.height=4, fig.width=3.2, fig.cap = "Missingness in multilevel data", echo=FALSE}
+```{r patterns, fig.height=3, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
 plot_md_set()
-# TODO: add border/line
-# TODO: remove unit
-# TODO: add accolade
-
 ```
@@
 Column `X2` is systematically missing and column `X3` is sporadically missing.^[Explain why.
-GV: niet nodig]
+GV: not necessary]
 
@@
 GV: No, patterns are different from mechanisms]
-
-$$
-\begin{matrix} 
-                          &X1     & X2          & X3            \\
-   \hline
-   \mathrm{cluster 1}\{ \ &x_{11} &x_{12}       &\mathtt{NA}    \\
-                             \ &x_{21}  &x_{22}       &x_{32}         \\
-   \mathrm{cluster 2}\{  &x_{31} &\mathtt{NA}  &x_{33}         \\
-                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
-   \mathrm{cluster 3}\{  &x_{51} &x_{52}       &x_{53}         \\
-\end{matrix} 
-$$
-[GV Ik vind de latex tabel niet tof. ]
 

### 2021-12-16 R/load_popularity.R
Generate more cases for the popularity dataset

@@ -1,0 +1,21 @@
+# generate incomplete data based on the popularity dataset
+
+# load complete data
+pop_complete <-
+  read.spss("Data/popular2.sav", to.data.frame = TRUE)[, 2:7]
+
+icc(popular ~as.factor(class), pop_complete)
+
+pop_synth <- pop_complete %>% 
+  split(.$class) %>% 
+  .[1:3] %>% 
+  purrr::map_dfr(., ~{
+  mice::mice(
+    .,
+    m = 19,
+    maxit = 1,
+    method = "pmm",
+    where = cbind(FALSE, matrix(TRUE, nrow(.), ncol(.)-1))
+  ) %>%
+  mice::complete("long", include = TRUE)})
+

### 2021-12-20 Manuscript/Manuscript.Rmd
Add parameters estimates popularity data

@@
 # functions
-miceadds::source.all(path = "../R")
 `%nin%` <- Negate(`%in%`)
 icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
+miceadds::source.all(path = "../R")
 
@@
 And, for reasons of brevity, we only discuss clustering between units, not within units (e.g., timeseries or longitudinal data).
-Also, add that we'll only discuss two levels, not more?~~
+Also, add that we'll only discuss two levels, not more?~~ Add IPDMA back in: patient level data from multiple studies
 
@@
 plot_md_set()
+# TODO: make cluster part f the data (blue)
+# TODO: dots between 2 and N, for each variable
 ```
@@
 ICCs
+
+#TODO: think about interactive clusters for the strippplots
+#TODO: switch off outliers in boxplot?
 ```
@@
 
+Here we can set `maxit` to 1...
+
 # Discussion

### 2021-12-20 R/load_popularity.R

@@ -1,21 +1,27 @@
 # generate incomplete data based on the popularity dataset
 
 # load complete data
+library(lme4)
 pop_complete <-
-  read.spss("Data/popular2.sav", to.data.frame = TRUE)[, 2:7]
+  foreign::read.spss("../Data/popular2.sav", to.data.frame = TRUE)[, 2:7]
 
-icc(popular ~as.factor(class), pop_complete)
+mod <- lmer(popular ~ sex + extrav + texp + extrav * texp + (1 | class) + (1 | texp) +  (1 | extrav), pop_complete) %>% broom.mixed::tidy()
 
-pop_synth <- pop_complete %>% 
-  split(.$class) %>% 
-  .[1:3] %>% 
-  purrr::map_dfr(., ~{
-  mice::mice(
-    .,
-    m = 19,
-    maxit = 1,
-    method = "pmm",
-    where = cbind(FALSE, matrix(TRUE, nrow(.), ncol(.)-1))
-  ) %>%
-  mice::complete("long", include = TRUE)})
+# # synthesize some more observations
+# icc(popular ~as.factor(class), pop_complete)
+# 
+# pop_synth <- pop_complete %>%
+#   split(.$class) %>%
+#   .[1:3] %>%
+#   purrr::map_dfr(., ~{
+#   mice::mice(
+#     .,
+#     m = 19,
+#     maxit = 1,
+#     method = "pmm",
+#     where = cbind(FALSE, matrix(TRUE, nrow(.), ncol(.)-1))
+#   ) %>%
+#   mice::complete("long", include = TRUE)})
+
+# maybe just use this https://debruine.github.io/tutorials/sim-lmer.html
 

### 2021-12-21 Manuscript/Manuscript.Rmd
Valentijn's comments and suggestions.

--- 2021-12-16 Manuscript/Manuscript.Rmd
@@
     affiliation: University Medical Center Utrecht \AND
-  - name: Valentijn de Jong
-    affiliation: University Medical Center Utrecht 
+  - name: Valentijn M.T. de Jong
+    affiliation: University Medical Center Utrecht, EMA
   - name: Gerko Vink
     affiliation: Utrecht University \AND
-  - name: Thomas Debray
+  - name: Thomas P.A. Debray
     affiliation: University Medical Center Utrecht 
+      # | Julius Center for Health Sciences and Primary Care, 
+      # | University Medical Center Utrecht, Utrecht, The Netherlands
+    #affiliation2:
 title:
   formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
+  # If you use tex in the formatted title, also supply version without
   plain:     "Imputation of Incomplete Multilevel Data with mice"
-  short:     "Multilevel \\pkg{mice}"
+  # For running headers, if needed
+  short:     "\\pkg{mice}: Multilevel"
 abstract: >
-  Multilevel data is not spared the ubiquitous problem of missing information.
-This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
-Including methods for ignorable and non-ignorable missingness.
-Footnotes in the current version show work in progress/under construction.
-The last section is not part of the manuscript, but purely for reminders.
+  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. 
 keywords:
@@
   \usepackage{amsmath}
-header-includes: 
- - \usepackage{graphicx} 
- - \usepackage{mathtools}
- - \usepackage{ulem}
 output: rticles::jss_article
 bibliography: ../References/multilevelmice.bib
-editor_options: 
-  chunk_output_type: inline
 ---
@@
 options(prompt = 'R> ', continue = '+ ')
-
-# seed
-set.seed(123)
 
@@
 library(metamisc)
-library(GJRM)
 library(pan)
@@
 miceadds::source.all(path = "../R")
-`%nin%` <- Negate(`%in%`)
 icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
@@
 
-~~Research into any field with a hierarchical or clustered nature of observations may yield multilevel data.
+We talk of multilevel data when there is some kind of hierarchy or clustering in a data set.
 In the typical case, individuals are nested within groups, but there are many different types of multilevel data.
-In the medical field, clustering occurs at e.g., the hospital or center level in registry data, or at the study-level in meta-analyses (IPDMA).
+In the medical field, clustering occurs at e.g., the hospitals/center level in registry data, or at the study-level in individual participant data meta-analyses (IPDMA).
 In the social sciences and official statistics we can find clustering e.g.
-at the country-level, or as imposed by a multi-stage sampling design.
-For the sake of legibility, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit' throughout this paper.
-And, for reasons of brevity, we only discuss clustering between units, not within units (e.g., timeseries or longitudinal data).
-Also, add that we'll only discuss two levels, not more?~~
+at the country-level, or as imposed by the sampling design.
+In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.^[Add that we'll only discuss two levels?] For reasons of brevity, we only discuss clustering between units, not within units (such as in timeseries or longitudinal data).
 
-In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.
-Ignoring such structures may be harmful to the inferences and can yield biased results.
-On the other hand, analyzing such multilevel data requires specialized techniques that take the clustered structure into account.
-Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
-The cluster to which a unit belongs may then influence the unit-level observations - and vice versa - for each of the units that make up the cluster [@hox17].
-These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
-
-The variability due to clustering is often measured by means of the intra-class coefficient (ICC).
-The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.
-Multilevel models typically accommodate for this variability by including separate intercepts for each cluster.
-Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters.
+Analyzing multilevel data requires special care, compared to 'regular', single level data.
+The cluster to which a unit belongs may influence the unit-level observations, and since clusters are made up of units, clusters depend on units as well [@hox17].
+These relations can and should be taken into account when developing analysis models for multilevel data.^[Explain ICC here?
+The percentage of variance attributed to the cluster-level is expressed by the intra-class coefficient (ICC).
+The ICC can also be interpreted as the expected correlation between two randomly sampled units in same cluster.
+So if the ICC is high, a lot of variability in a variable is due to the clustering, which should be modeled accordingly.] Multilevel models typically include separate intercepts for each cluster, which relieves one restriction imposed by single-level models: equal group means across clusters.
 Additionally, there may be random predictor effects and/or random error terms (residual error variances), see e.g.
-@hox17 and @jong21.^[Add that heterogeneity refers to variability within clusters.] There are many names for models that take clustering into account.
+@hox17 and @jong21.^[Add that heterogeneity refers to variability within clusters.
+VDJ: heterogeneity can refer to variability of sample units, i.e.
+variance.
+It can also refer to variability across data sets: variability in measurement instruments, definitions, but also to variability of treatment/predictor effects across data sets.
+So we need to make clear which one we are referring to.] There are many names for models that take clustering into account.
 Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
@@
 
-```{r patterns, fig.height=3, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
-plot_md_set()
+```{r patterns, fig.height=4, fig.width=3.5, fig.cap = "Missingness in multilevel data", echo=FALSE}
+# missingness indicator plot (monotone vs non-monotone)
+miss_ind <- tibble(
+    rownr = 5:1,
+    unit = rep("observed", 5),
+    cluster = rep("observed", 5),
+    X1 = rep("observed", 5),
+    X2 = c(rep("observed", 3), rep("missing", 2)),
+    X3 = c("missing", rep("observed", 3), "missing")
+  ) %>% pivot_longer(cols = c(unit, cluster, X1, X2, X3))
+
+# add text 
+miss_ind$text <- ""
+miss_ind[miss_ind$name == "unit", "text"] <- as.character(1:5)
+miss_ind[miss_ind$name == "cluster", "text"] <- as.character(c(1,1,1,2,2))
+
+# plot
+miss_ind %>% 
+  mutate(name = factor(name, levels = unique(name))) %>% 
+ggplot(aes(
+  x = name,
+  y = rownr,
+  color = value,
+  width = 0.8, 
+  height = 0.8,
+  label = text
+)) +
+  geom_tile(fill = "white", size = 2) +
+  geom_text(show.legend = FALSE) +
+  scale_x_discrete(position = "top") +
+  scale_y_discrete(labels = 5:1) +
+  scale_color_manual(values = plot_col, name = "") +
+  theme_minimal() +
+  theme(
+    legend.position = "bottom", 
+    panel.grid.major = element_blank()) +
+  labs(x = "", y = "")
+# TODO: make it one plot with sporadic and systematic
+# TODO: participant/response/*sample units*/records instead of obs on the y axis
+
 ```
 
-The process of analyzing multilevel data is further complicated when not all observations are observed.
-Just as with single level data, missingness may occur at the unit level.
-But with multiple levels of data comes the potential for clustered missingness.
-Missingness in multilevel data can therefore be categorized into two general patterns: systematic missingness and sporadic missingness, see @resc13.
-We have visualized the difference between these types of missingness in Figure 1.
-The figure shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n=5$ units and $p=3$ columns distributed over 3 clusters.
-Column `X1` is completely observed.
-Column `X2` is systematically missing and column `X3` is sporadically missing.^[Explain why.
-GV: not necessary]
+Multilevel data is not spared the ubiquitous problem of missing information.
+Just as in single level data, missingness may occur at the unit level.
+But with multiple levels of data comes the potential for missingness at multiple levels.
+Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13.
+In figure 1, we show a data set with units in the rows and variables in the columns, there are 5 units nested within 2 clusters, and 3 variables of interest.
+Variable `X1` is completely observed.
+Variable `X2` is systematically missing in cluster 2, `X3` is sporadically missing in both clusters.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (same value per cluster) and non-measured random variables (which may differ per unit within clusters).
+In Figure 1, the former implies that the unobserved values for units 4 and 5 on variable `X2` would be equal.
+With the latter, the values may differ.
+The optimal way of accomodating the missingness may depend on the missing data pattern.
 
-Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters).
-In Figure 1, the former would imply that the unobserved values for units 3 and 4 on column `X2` are identical.
-With the latter, the values would differ.
-The optimal strategy for dealing with the missingness may therefore depend on the observed missing data pattern.
-^[add missing data mechanisms here?
-GV: No, patterns are different from mechanisms]
-
-Ignoring the missingness in analyses can be extremely harmful to inferences.
+Ignoring missing data in research endeavors is almost never a good idea.
 Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power.
 Instead, the missingness should be accommodated *before* or *within* the analysis of scientific interest.
-Especially the former is very generic and popular and is widely known as imputation.
-Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed.
-The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
+^[Can you give an example of each?
+With within do you mean integrating missing data out, or automatic imputation in a Bayesian MCMC analysis?.
+One might also think that with before you mean contacting the sample units to obtain the true values, which would then conflict with the remainder of the paragraph.]Especially the former is very generic and popular.
+Imputing (i.e., filling in) the missing values splits the missing data problem from the scientific problem.
+The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which solves the missingness one variable at a time, iteratively.
 \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
 In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.^[Discuss missingness mechanisms before this point, add references @yuce08 and @hox15.]
+
+<!-- -	What kinds of missingness are there? ADD: missingness mechanisms here. See e.g. @yuce08 and @hox15. -->
+
+<!-- - Why are standard (ad hoc) missing data methods not well suited?  -->
+
+<!-- - What types of multilevel methods are available?
+General overview of approaches, see @audi18 and @grun18.
+E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.
+-->
+
+<!-- - Additional difficulty that is addressed in this tutorial: MNAR data. -->
 
@@
 We provide practical guidelines and code snippets for different missing data situations.
-For reasons of brevity, we focus on imputation by chained equations^[add that JOMO is available in \pkg{mice} as well?].
-Other useful packages for incomplete multilevel data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed.
+For reasons of brevity, we focus on imputation by chained equations, although JOMO is available in \pkg{mice} as well.
+Other useful packages for incomplete multilevel data include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed.
 We're providing an overview of implementations.
@@
 So we won't go into detail for the different methods (and equations).
-Refer to @meng94, @audi18, and @grun18.
+Refer to @meng94, an Audigier paper, and a paper by Grund on congeniality and random slopes.
 This paper is just a software tutorial.
@@
 For each case study we focus on a different aspect to illustrate how to impute incomplete multilevel data.
+
+# Workflows
+
+<!-- We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).
+-> ask whether we can use the Heckman repo data or simulate data ourselves -->
+
+<!-- Heckman options: -->
+
+<!-- - leiden85 -->
+
+<!-- - GJRM::hiv (https://rdrr.io/github/egeminiani/GJRM/man/hiv.html) -->
+
+<!-- - simulating -->
+
+<!-- - IMPACT -->
+
+We introduce three case studies to illustrate the workflow.
 In the `mice::popmis` data, we show the advantages of including the multilevel structure of the data into the imputation model.
@@
 In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model.
-For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data.
 
-# Case Study I: Popularity
+For each case study we'll look at least at: 1) the incomplete data; 2) the imputation model; 3) the imputed data; and 4) how to obtain pooled estimates for the analysis of scientific interest.
+
+## Case Study I: Popularity
 
 ```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
-# load data 
-pop <- readRDS("../Data/popNCR2.RDS") %>% mutate(class = as.factor(class))
+# load data and merge classes into schools for this tutorial -> but this cannot work because how texp and popteach are defined...
+pop <- readRDS("../Data/popNCR.RDS") %>% mutate(school = round(0.5 + as.numeric(class)/10))
 
 # ICCs
-icc_pop <- icc(popular ~ class, data = pop)
-icc_teach <- icc(popteach ~ class, data = pop)
+icc_pop <- multilevel::ICC1(aov(popular ~ as.factor(class), data = pop))
+icc_teach <- multilevel::ICC1(aov(popteach ~ as.factor(class), data = pop))
 ```
 
-`popNCR` is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:
+`popNCR` is a simulated data set with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables:
 
@@
 
-```{r pop_pat, echo=FALSE, fig.height=4, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
+```{r pop_pat, echo=FALSE, fig.height=6, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
 # missingness
@@
 
-The missingness in this dataset is induced conform MAR and MNAR mechanisms.
-The missing data pattern, Figure \ref{fig:pop_pat}, shows that just one variable is incomplete **[the next part is not yet updated to reflect this]**.
+The missingness in this data set is induced conform MAR and MNAR mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows the systematic nature of the missingness.
 
@@
 In Figure \ref{fig:pop_dist} we see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
-This would indicate a right-tailed MAR missingness.
-In fact, this is exactly what happens, because the missingness in these data was created manually.
+^[Isn't it the other way around?.] This would indicate a right-tailed MAR missingness.
+In fact, we know that this is exactly what happened, because the missingness in these data was created manually.
 Now, we've made it observable by examining the relations between the missingness in popular and the observed data in `popteach`.
@@
 ```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
+# distributions
+# plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
+# plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
 # TODO: check if up-side-down plot works
+# or maybe overlapping densities
 # TODO: check smoothing in geom_density function and make it the inverse of the sample size
 # TODO: add functions to mice
-# TODO: add title and informative legend
+
+plot_col <- mice:::mdc(c("obs", "mis")) %>% setNames(c("observed", "missing"))
+
+# x = "popular"
+# z = "popteach"
+plot_conditional <- function(data, x, z, cluster, ...){
+  ggplot(data, aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
+  geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
+  geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
+  geom_density() +
+  scale_color_manual(values = plot_col, name = z) +
+  theme_classic() +
+  theme(legend.position = "bottom") +
+  labs(x = x)
+}
+
+plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
+plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
+# TODO: pooteach obs/mis
+# TODO: add title
 # TODO: add facets for some clusters
@@
 # TODO: make it average of cluster dens, not marginal
+
+# Feedback: 
+# you only see the extreme ones in the positive sense, especially with 100 clusters
 # alternatively add a quartile line around the density with geom_ribbon
+
 ```
@@
 
-The first imputation model that we'll use is likely to be invalid.
-In this model, we ignore the multilevel structure of the data, despite the high ICCs.
-This is purely to illustrate the effects of ignoring the clustering in our imputation effort.
+In the first imputation model that we'll use, we ignore the multilevel structure of the data despite the high ICCs.
+This model is likely to be invalid.
+We apply it purely to illustrate the effects of ignoring the clustering in our imputation effort.
 
 We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`.
-We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.
+We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.^[Explain why.]
 
@@
 # impute the data, ignoring the cluster structure
-imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
+imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
 ```
@@
 # further inspection of the imputations
-plot_box(imp_ignored, x = "popular", strip = TRUE)
-# TODO: think about multimodality
+densityplot(imp_ignored)
 
@@
   vars = c("popular", "popteach", "texp"), 
-  incomplete = c(icc(popular ~ class, pop), 
-               icc(popteach ~ class, pop),
-               icc(texp ~ class, pop)), 
-  ignored = c(icc(popular ~ class, complete(imp_ignored)), 
-              icc(popteach ~ class, complete(imp_ignored)), 
-              icc(texp ~ class, complete(imp_ignored)))
+  incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+               multilevel::ICC1(aov(popteach ~ class, pop)),
+               multilevel::ICC1(aov(texp ~ class, pop))), 
+  ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
+              multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
+              multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
   )
@@
 This is due to the MNAR missingness in `texp`.
-Higher values for `texp` have a larger probability to be missing.
-This may not a problem, however, if at least one pupil in each class has teacher experience recorded, we can deductively impute the correct (i.e.
+^[Mention somewhere around here that texp an unobserved constant, which you mentioned in the introduction?] Higher values for `texp` have a larger probability to be missing.
+This may not be a problem, however, if at least one pupil in each class has teacher experience recorded, we can deductively impute the correct (i.e.
 true) value for every pupil in the class.
+^[But is it reallly MNAR if we can restore it by simply using the observed data?]
 
@@
 We'll now use `class` as a predictor to impute all other variables.
-This is still not recommended practice, since it only works under certain circumstances and results may be biased.
-But at least, it includes some multilevel aspect.
+This is not recommended practice, since it only works under certain circumstances ^[Explain why.
+Because it does not borrow information across classes?
+or because of this specific example where texp is given by class?] and results may be biased.
+^[Why would it be biased?
+Typically fixed effects models (i.e.
+estimated per cluster) are unbiased, are you saying this is different for imputation models?] But at least, it includes some multilevel aspect.
 Colloquially, this is 'multilevel imputation for dummies'.
 
-```{r pop_predictor, message=FALSE, warning=FALSE}
+```{r pop_predictor, echo=FALSE, message=FALSE, warning=FALSE}
 # adjust the predictor matrix
@@
 # impute the data, cluster as predictor
-imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)
-
+imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
 ```
@@
 # further inspection of the imputations
-plot_box(imp_predictor, x = "popular", strip = TRUE)
+densityplot(imp_predictor)
 
 # compare ICCs before and after imputation
-ICCs <- ICCs %>% mutate(
-           predictor = c(icc(popular ~ class, complete(imp_predictor)), 
-                        icc(popteach ~ class, complete(imp_predictor)), 
-                        icc(texp ~ class, complete(imp_predictor)))
+ICCs <- ICCs %>% cbind(
+           predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
+                        multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
+                        multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
            )
@@
 
-```{r pop_norm}
-pred <- ini$pred
-pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
-#-2 for the cluster variable, 2 for random effects
-meth <- ini$meth
-meth <- c("", "", "", "", "", "2l.norm", "")
-imp_norm_2l <-
-  mice(
-    pop %>% mutate(class = as.integer(class)),
-    pred = pred,
-    meth = meth,
-    maxit = 1,
-    print = FALSE
-  )
-```
+## Case study II: IMPACT
 
-```{r pop_norm_eval}
-# plot(imp_norm)
-plot_box(imp_norm_2l, x = "popular", strip = TRUE)
-
-ICCs <- ICCs %>% mutate(
-           norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
-                    icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
-                    icc(texp ~ as.factor(class), complete(imp_norm_2l)))
-           )
-ICCs
-```
-
-```{r pop_pan}
-pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
-meth <- c("", "", "", "", "", "2l.pan", "")
-imp_pan_2l <-
-  mice(
-    pop %>% mutate(class = as.integer(class)),
-    pred = pred,
-    meth = meth,
-    maxit = 1,
-    print = FALSE
-  )
-
-```
-
-```{r pop_pan_eval}
-# plot(imp_pan)
-plot_box(imp_pan_2l, x = "popular", strip = TRUE)
-ICCs <- ICCs %>% mutate(
-           pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
-                    icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
-                    icc(texp ~ as.factor(class), complete(imp_pan_2l)))
-           )
-ICCs
-```
-
-# Case study II: IMPACT
-
-`impact` is traumatic brain injury data with patients, $n = 11022$, clustered in studies, $N = 15$. With the following 11 variables:
+`impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
 
@@
 ```
+<!-- -> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf),  -->
 
-# Case study III: HIV
+<!-- TODO: also make MNAR missingness to heckman model.
+Maybe based on `ct` variable?
+Inclusion-selection variable.
+-> otherwise: use `leiden85` data on blood pressure with MNAR.
+Then run cox regression like the boshuizen article but with living situation as clusters.
+-> TODO: get analyses from https://www.gerkovink.com/mimp/Contents/Exercises/Day%203%20-%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html.
+-->
 
-Toy example from [Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R). We will use the following variables:
+<!-- ## Amputation -->
 
-  - `region` Cluster variable,
-  - `hiv` HIV diagnosis (0=no, 1=yes),
-  - `age` Age of the patient,
-  - `marital` Marital status,
-  - `condom` Condom use during last intercourse,
-  - `smoke` Smoker (levels).
+<!-- -  -->
 
-```{r hiv, echo=FALSE}
-# data
-data("hiv")
-hiv <- hiv %>% 
-  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
-  mutate(region = as.integer(region))
-# md pattern
-plot_md_pat(hiv)
-```
+<!-- ## Modeling choices -->
+
+<!-- - Which models will we discuss? We'll build the model to grow in complexity. The final model is the most complex but also the most versatile.  -->
+
+<!-- - Note on model complexity: Typically, we should at least use random intercepts, but often random slopes as well.
+Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeniality, but don't mention the term) ->  Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there.
+Sometimes there's little reason to assume some variable is affected by heterogeneity.
+->  -->
+
+<!-- - Step 0: As predictor + CCA to scare off users -->
+
+<!-- - Step 1: Random intercepts -->
+
+<!-- - Step 2: Random slopes -->
+
+<!-- - Step 3: Residuals -->
+
+<!-- - Heckman model for MNAR -->
+
+<!-- - What do the different implementations look like? How to define the imputation model(s) in \pkg{mice}? -->
+
+<!-- ## Step 0 -->
+
+<!-- - AKA multilevel imputation for dummies.  -->
+
+<!-- - Doesn't work for systematic missingness.  -->
+
+<!-- ## Step 1-3 + MNAR -->
+
+<!-- - TODO: fill in. -->
+
+## Pooling
+
+- Analysis of scientific interest.
+
+- Pooling using `mitml`.
+
+- Pooling 'regular' parameters vs more 'exotic' parameters (SE of residual errors, or autocorrelation)
 
@@
 
-- Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt.
-
 # References

### 2021-12-21 Manuscript/Manuscript.tex
Add parameters estimates popularity data
Valentijn's comments and suggestions.

--- 2021-12-15 Manuscript/Manuscript.tex
@@
 Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
-Medical Center Utrecht \AND Valentijn de Jong\\University Medical
-Center Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
-Debray\\University Medical Center Utrecht
+Medical Center Utrecht \AND Valentijn M.T. de Jong\\University
+Medical Center Utrecht, EMA \And Gerko Vink\\Utrecht University
+\AND Thomas P.A. Debray\\University Medical Center Utrecht
 }
@@
 
-\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
-Jong, Gerko Vink, Thomas Debray}
+\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn M.T. de
+Jong, Gerko Vink, Thomas P.A. Debray}
 \Plaintitle{Imputation of Incomplete Multilevel Data with mice}
-\Shorttitle{Multilevel \pkg{mice}}
+\Shorttitle{\pkg{mice}: Multilevel}
 
 \Abstract{
-Multilevel data is not spared the ubiquitous problem of missing
-information. This is a tutorial paper on imputing incomplete multilevel
-data with \pkg{mice}. Including methods for ignorable and non-ignorable
-missingness. Footnotes in the current version show work in
-progress/under construction. The last section is not part of the
-manuscript, but purely for reminders.
+Tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
+Including methods for ignorable and non-ignorable missingness.
 }
@@
 
-\usepackage{graphicx}
-\usepackage{mathtools}
-
 \usepackage{amsmath}
@@
 
-Research into any field with a hierarchical or clustered nature of
-observations may yield multilevel data. In the typical case, individuals
-are nested within groups, but there are many different types of
-multilevel data. In the medical field, clustering occurs at e.g., the
-hospital or center level in registry data, or at the study-level in
-meta-analyses (IPDMA\footnote{remove or write in full?}). In the social
-sciences and official statistics we can find clustering e.g.~at the
-country-level, or as imposed by a multi-stage sampling design. For the
-sake of legibility, we will refer to the grouping variable as `cluster',
-and the grouped variable as `(sample) unit' throughout this
-paper.\footnote{Alternatives to unit: participant/response/record.} And,
-for reasons of brevity, we only discuss clustering between units, not
-within units (e.g., timeseries or longitudinal data).\footnote{Also, add
-  that we'll only discuss two levels, not more?}
+We talk of multilevel data when there is some kind of hierarchy or
+clustering in a data set. In the typical case, individuals are nested
+within groups, but there are many different types of multilevel data. In
+the medical field, clustering occurs at e.g., the hospitals/center level
+in registry data, or at the study-level in individual participant data
+meta-analyses (IPDMA). In the social sciences and official statistics we
+can find clustering e.g.~at the country-level, or as imposed by the
+sampling design. In this paper, we will refer to the grouping variable
+as `cluster', and the grouped variable as `(sample) unit'.\footnote{Add
+  that we'll only discuss two levels?} For reasons of brevity, we only
+discuss clustering between units, not within units (such as in
+timeseries or longitudinal data).
 
 Analyzing multilevel data requires special care, compared to `regular',
-single level data. For instance, there may be cross-level interactions
-between unit-level variables and cluster-level variables. The cluster to
-which a unit belongs may influence the unit-level observations, and vice
-versa for the the units that make up the cluster \citep{hox17}. These
-relations can and should be taken into account when developing analysis
-models for multilevel data.\footnote{Explain ICC here? The percentage of
-  variance attributed to the cluster-level is expressed by the
-  intra-class coefficient (ICC). The ICC can also be interpreted as the
-  expected correlation between two randomly sampled units in same
-  cluster. So if the ICC is high, a lot of variability in a variable is
-  due to the clustering, which should be modeled accordingly.}
-Multilevel models typically include separate intercepts for each
-cluster, which relieves one restriction imposed by single-level models:
-equal group means across clusters. Additionally, there may be random
-predictor effects and/or random error terms (residual error variances),
-see e.g. \citet{hox17} and \citet{jong21}.\footnote{Add that
-  heterogeneity refers to variability within clusters.} There are many
-names for models that take clustering into account. Some popular
-examples are `multilevel models', `hierarchical models', `mixed effect
-models' and `random effect models'.
+single level data. The cluster to which a unit belongs may influence the
+unit-level observations, and since clusters are made up of units,
+clusters depend on units as well \citep{hox17}. These relations can and
+should be taken into account when developing analysis models for
+multilevel data.\footnote{Explain ICC here? The percentage of variance
+  attributed to the cluster-level is expressed by the intra-class
+  coefficient (ICC). The ICC can also be interpreted as the expected
+  correlation between two randomly sampled units in same cluster. So if
+  the ICC is high, a lot of variability in a variable is due to the
+  clustering, which should be modeled accordingly.} Multilevel models
+typically include separate intercepts for each cluster, which relieves
+one restriction imposed by single-level models: equal group means across
+clusters. Additionally, there may be random predictor effects and/or
+random error terms (residual error variances), see e.g. \citet{hox17}
+and \citet{jong21}.\footnote{Add that heterogeneity refers to
+  variability within clusters. VDJ: heterogeneity can refer to
+  variability of sample units, i.e.~variance. It can also refer to
+  variability across data sets: variability in measurement instruments,
+  definitions, but also to variability of treatment/predictor effects
+  across data sets} There are many names for models that take clustering
+into account. Some popular examples are `multilevel models',
+`hierarchical models', `mixed effect models' and `random effect models'.
 
@@
 categorized into two general patterns: systematical missingness and
-sporadic missingness, see \citet{resc13}. In figure 1, we show a dataset
-with units in the rows and variables in the columns, there are 5 units
-nested within 3 clusters, and 3 variables of interest. Variable
+sporadic missingness, see \citet{resc13}. In figure 1, we show a data
+set with units in the rows and variables in the columns, there are 5
+units nested within 2 clusters, and 3 variables of interest. Variable
 \texttt{X1} is completely observed. Variable \texttt{X2} is
-systematically missing, \texttt{X3} is sporadically missing.\footnote{Explain
-  why.} Systematic missingness can be further subdivided into unobserved
-constants (i.e., the same value within clusters) and non-measured random
-variables (which may differ per unit within clusters). In Figure 1, the
-former implies that the unobserved values for units 3 and 4 on variable
-\texttt{X2} would be equal. With the latter, the values may differ.
-Depending on the missing data pattern\footnote{add missing data
-  mechanisms here?}, there are more or less optimal way of accommodating
-the missingness.
-
-\[
-\begin{matrix} 
-                          &X1     & X2          & X3            \\
-   \hline
-   \mathrm{cluster 1}\{   &x_{11} &x_{12}       &\mathtt{NA}    \\
-                          &x_{21} &x_{22}       &x_{32}         \\
-   \mathrm{cluster 2}\{   &x_{31} &\mathtt{NA}  &x_{33}         \\
-                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
-   \mathrm{cluster 3}\{   &x_{51} &x_{52}       &x_{53}         \\
-\end{matrix} 
-\]
+systematically missing in cluster 2, \texttt{X3} is sporadically missing
+in both clusters.\footnote{Explain why.} Systematic missingness can be
+further subdivided into unobserved constants (same value per cluster)
+and non-measured random variables (which may differ per unit within
+clusters). In Figure 1, the former implies that the unobserved values
+for units 4 and 5 on variable \texttt{X2} would be equal. With the
+latter, the values may differ. The optimal way of accomodating the
+missingness may depend on the missing data pattern.
 
@@
 \emph{before} or \emph{within} the analysis of scientific interest.
-Especially the former is very generic and popular. Imputing (i.e.,
-filling in) the missing values splits the missing data problem from the
-scientific problem. The \proglang{R} package \pkg{mice} has become the
-de-facto standard for imputation by chained equations, which solves the
+\footnote{Can you give an example of each? With within do you mean
+  integrating missing data out, or automatic imputation in a Bayesian
+  MCMC analysis?. One might also think that with before you mean
+  contacting the sample units to obtain the true values, which would
+  then conflict with the remainder of the paragraph.}Especially the
+former is very generic and popular. Imputing (i.e., filling in) the
+missing values splits the missing data problem from the scientific
+problem. The \proglang{R} package \pkg{mice} has become the de-facto
+standard for imputation by chained equations, which solves the
 missingness one variable at a time, iteratively. \pkg{mice} is known to
@@
 different missing data situations. For reasons of brevity, we focus on
-imputation by chained equations\footnote{add that JOMO is available in
-  \pkg{mice} as well?}. Other useful packages for incomplete multilevel
-data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.\footnote{Rephrase:
+imputation by chained equations, although JOMO is available in
+\pkg{mice} as well. Other useful packages for incomplete multilevel data
+include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.\footnote{Rephrase:
   Some level of knowledge on multilevel models is assumed. We're
@@
   detail for the different methods (and equations). Refer to
-  \citet{meng94}, \citet{audi18}, and \citet{grun18}. This paper is just
-  a software tutorial. We'll keep it practical.}
+  \citet{meng94}, an Audigier paper, and a paper by Grund on
+  congeniality and random slopes. This paper is just a software
+  tutorial. We'll keep it practical.}
 
@@
 For each case study we focus on a different aspect to illustrate how to
-impute incomplete multilevel data. In the \texttt{mice::popmis} data, we
-show the advantages of including the multilevel structure of the data
-into the imputation model. In the \texttt{metamisc::impact} data we'll
-show how to induce missingness and solve it in real-world data. In the
-\texttt{GJRM::hiv} we provide novel methodology\footnote{not really, the
-  methods exist already, but how to show that this is something new and
-  exciting?} for imputing MNAR missingness according to the Heckman
-model. For all case studies we discuss the nature of the incomplete
-data, the imputation model(s), and evaluation of the imputed data.
+impute incomplete multilevel data.
+
+\hypertarget{workflows}{%
+\section{Workflows}\label{workflows}}
+
+We introduce three case studies to illustrate the workflow. In the
+\texttt{mice::popmis} data, we show the advantages of including the
+multilevel structure of the data into the imputation model. In the
+\texttt{metamisc::impact} data we'll show how to induce missingness and
+solve it in real-world data. In the \texttt{GJRM::hiv} we provide novel
+methodology\footnote{not really, the methods exist already, but how to
+  show that this is something new and exciting?} for imputing MNAR
+missingness according to the Heckman model.
+
+For each case study we'll look at least at: 1) the incomplete data; 2)
+the imputation model; 3) the imputed data; and 4) how to obtain pooled
+estimates for the analysis of scientific interest.
 
 \hypertarget{case-study-i-popularity}{%
-\section{Case Study I: Popularity}\label{case-study-i-popularity}}
+\subsection{Case Study I: Popularity}\label{case-study-i-popularity}}
 
-\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
-where the number of units \(n = 2000\), and the number of clusters
-\(N = 100\), on 7 variables:
+\texttt{popNCR} is a simulated data set with pupils clustered in
+classes, \(n_{\text{participants}} = 2000\),
+\(n_{\text{clusters}} = 100\), on 7 variables:
 
@@
 (ICC). For \texttt{popular} the ICC is 0.33. For \texttt{popteach} it is
-0.34. It would thus be wise to use multilevel modeling.
+0.31. It would thus be wise to use multilevel modeling.
 
-The missingness in this dataset is induced conform MAR and MNAR
+The missingness in this data set is induced conform MAR and MNAR
 mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows
-that just one variable is incomplete \textbf{{[}the next part is not yet
-updated to reflect this{]}}.
+the systematic nature of the missingness.
 
@@
 missing \texttt{popular} is further to the right than the distribution
-for observed \texttt{popular}. This would indicate a right-tailed MAR
-missingness. In fact, this is exactly what happens, because the
-missingness in these data was created manually. Now, we've made it
-observable by examining the relations between the missingness in popular
-and the observed data in \texttt{popteach}. There is also a dependency
-between the missingness in teacher popularity and pupil popularity. The
-relation seems to be right-tailed as well.
+for observed \texttt{popular}. \footnote{Isn't it the other way around?.}
+This would indicate a right-tailed MAR missingness. In fact, we know
+that this is exactly what happened, because the missingness in these
+data was created manually. Now, we've made it observable by examining
+the relations between the missingness in popular and the observed data
+in \texttt{popteach}. There is also a dependency between the missingness
+in teacher popularity and pupil popularity. The relation seems to be
+right-tailed as well.
 
@@
 
-\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist}
+\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-1}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-2}
+
+}
+
+\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-2}
 \end{figure}
@@
 
-The first imputation model that we'll use is likely to be invalid. In
-this model, we ignore the multilevel structure of the data, despite the
-high ICCs. This is purely to illustrate the effects of ignoring the
+In the first imputation model that we'll use, we ignore the multilevel
+structure of the data despite the high ICCs. This model is likely to be
+invalid. We apply it purely to illustrate the effects of ignoring the
 clustering in our imputation effort.
@@
 do not use the observation identifier \texttt{pupil} or cluster
-identifier \texttt{class} as predictors to impute other variables.
+identifier \texttt{class} as predictors to impute other
+variables.\footnote{Explain why.}
 
@@
 R> # impute the data, ignoring the cluster structure
-R> imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
+R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
 \end{CodeInput}
@@
       vars incomplete   ignored
-1  popular  0.3280070 0.3338626
-2 popteach  0.3414766 0.3414766
-3     texp  1.0000000 1.0000000
+1  popular  0.3280070 0.2867315
+2 popteach  0.3138658 0.2541582
+3     texp  1.0000000 0.4329458
 \end{CodeOutput}
@@
 We can also observe that the teacher experience increases slightly after
-imputation. This is due to the MNAR missingness in \texttt{texp}. Higher
-values for \texttt{texp} have a larger probability to be missing. This
-may not a problem, however, if at least one pupil in each class has
-teacher experience recorded, we can deductively impute the correct
-(i.e.~true) value for every pupil in the class.
+imputation. This is due to the MNAR missingness in \texttt{texp}.
+\footnote{Mention somewhere around here that texp an unobserved
+  constant, which you mentioned in the introduction?} Higher values for
+\texttt{texp} have a larger probability to be missing. This may not be a
+problem, however, if at least one pupil in each class has teacher
+experience recorded, we can deductively impute the correct (i.e.~true)
+value for every pupil in the class. \footnote{But is it reallly MNAR if
+  we can restore it by simply using the observed data?}
 
@@
 We'll now use \texttt{class} as a predictor to impute all other
-variables. This is still not recommended practice, since it only works
-under certain circumstances and results may be biased. But at least, it
-includes some multilevel aspect. Colloquially, this is `multilevel
-imputation for dummies'.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # adjust the predictor matrix
-R> pred <- ini$pred 
-R> pred[, "pupil"] <- 0
-R> 
-R> # impute the data, cluster as predictor
-R> imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
+variables. This is not recommended practice, since it only works under
+certain circumstances \footnote{Explain why. Because it does not borrow
+  information across classes? or because of this specific example where
+  texp is given by class?} and results may be biased. \footnote{Why
+  would it be biased? Typically fixed effects models (i.e.~estimated per
+  cluster) are unbiased, are you saying this is different for imputation
+  models?} But at least, it includes some multilevel aspect.
+Colloquially, this is `multilevel imputation for dummies'.
 
@@
       vars incomplete   ignored predictor
-1  popular  0.3280070 0.3338626 0.3871020
-2 popteach  0.3414766 0.3414766 0.3414766
-3     texp  1.0000000 1.0000000 1.0000000
+1  popular  0.3280070 0.2867315 0.3825608
+2 popteach  0.3138658 0.2541582 0.2541437
+3     texp  1.0000000 0.4329458 1.0000000
 \end{CodeOutput}
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> pred <- ini$pred
-R> pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
-R> #-2 for the cluster variable, 2 for random effects
-R> meth <- ini$meth
-R> meth <- c("", "", "", "", "", "2l.norm", "")
-R> imp_norm_2l <-
-+   mice(
-+     pop %>% mutate(class = as.integer(class)),
-+     pred = pred,
-+     meth = meth,
-+     maxit = 1,
-+     print = FALSE
-+   )
-\end{CodeInput}
-\end{CodeChunk}
+\hypertarget{case-study-ii-impact}{%
+\subsection{Case study II: IMPACT}\label{case-study-ii-impact}}
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # plot(imp_norm)
-R> plot_box(imp_norm_2l, x = "popular", strip = TRUE)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_norm_eval-1} \end{center}
-
-\begin{CodeInput}
-R> ICCs <- ICCs %>% mutate(
-+            norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
-+                     icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
-+                     icc(texp ~ as.factor(class), complete(imp_norm_2l)))
-+            )
-R> ICCs
-\end{CodeInput}
-\begin{CodeOutput}
-      vars incomplete   ignored predictor      norm
-1  popular  0.3280070 0.3338626 0.3871020 0.3612830
-2 popteach  0.3414766 0.3414766 0.3414766 0.3414766
-3     texp  1.0000000 1.0000000 1.0000000 1.0000000
-\end{CodeOutput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
-R> meth <- c("", "", "", "", "", "2l.pan", "")
-R> imp_pan_2l <-
-+   mice(
-+     pop %>% mutate(class = as.integer(class)),
-+     pred = pred,
-+     meth = meth,
-+     maxit = 1,
-+     print = FALSE
-+   )
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # plot(imp_pan)
-R> plot_box(imp_pan_2l, x = "popular", strip = TRUE)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_pan_eval-1} \end{center}
-
-\begin{CodeInput}
-R> ICCs <- ICCs %>% mutate(
-+            pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
-+                     icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
-+                     icc(texp ~ as.factor(class), complete(imp_pan_2l)))
-+            )
-R> ICCs
-\end{CodeInput}
-\begin{CodeOutput}
-      vars incomplete   ignored predictor      norm       pan
-1  popular  0.3280070 0.3338626 0.3871020 0.3612830 0.3745108
-2 popteach  0.3414766 0.3414766 0.3414766 0.3414766 0.3414766
-3     texp  1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
-\end{CodeOutput}
-\end{CodeChunk}
-
-\hypertarget{case-study-ii-impact}{%
-\section{Case study II: IMPACT}\label{case-study-ii-impact}}
-
-\texttt{impact} is traumatic brain injury data with patients,
-\(n = 11022\), clustered in studies, \(N = 15\). With the following 11
-variables:
+\texttt{impact} is traumatic brain injury data with patients clustered
+in studies, \(n_{\text{participants}} = 11022\) and
+\(n_{\text{clusters}} = 15\), on the following 11 variables:
 
@@
 
-\hypertarget{case-study-iii-hiv}{%
-\section{Case study III: HIV}\label{case-study-iii-hiv}}
+\hypertarget{pooling}{%
+\subsection{Pooling}\label{pooling}}
 
-Toy example from
-\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{Heckman
-Github repo}.
+\begin{itemize}
+\item
+  Analysis of scientific interest.
+\item
+  Pooling using \texttt{mitml}.
+\item
+  Pooling `regular' parameters vs more `exotic' parameters (SE of
+  residual errors, or autocorrelation)
+\end{itemize}
 
@@
   or \texttt{coxme}?
-\item
-  Adding a ICC=0 dataset to show that even if there is no clustering it
-  doesn't hurt.
 \end{itemize}

### 2022-01-14 Manuscript/Manuscript.Rmd
Integrate comments last meeting

--- 2021-12-20 Manuscript/Manuscript.Rmd
@@
 
-~~Research into any field with a hierarchical or clustered nature of observations may yield multilevel data.
-In the typical case, individuals are nested within groups, but there are many different types of multilevel data.
-In the medical field, clustering occurs at e.g., the hospital or center level in registry data, or at the study-level in meta-analyses (IPDMA).
-In the social sciences and official statistics we can find clustering e.g.
-at the country-level, or as imposed by a multi-stage sampling design.
-For the sake of legibility, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit' throughout this paper.
-And, for reasons of brevity, we only discuss clustering between units, not within units (e.g., timeseries or longitudinal data).
-Also, add that we'll only discuss two levels, not more?~~ Add IPDMA back in: patient level data from multiple studies
-
-In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.
-Ignoring such structures may be harmful to the inferences and can yield biased results.
+In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.^[Add IPDMA back in: patient level data from multiple studies.] Ignoring such structures may be harmful to the inferences and can yield biased results.
 On the other hand, analyzing such multilevel data requires specialized techniques that take the clustered structure into account.
@@
 In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.^[Discuss missingness mechanisms before this point, add references @yuce08 and @hox15.]
+
+<!-- "We find it useful to distinguish the missingness pattern, which describes which values are missing and observed in the data matrix and the missingness mechanism (or mechanisms), which concerns the relationship between missingness and the values of variables in the data matrix" (Little and Rubin, p.
+8).
+-->
 

### 2022-01-14 Manuscript/Manuscript.tex

--- 2021-12-16 Manuscript/Manuscript.tex
@@
 \usepackage[utf8]{inputenc}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
-% Pandoc syntax highlighting
-
-% Pandoc citation processing
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
-\sout{Research into any field with a hierarchical or clustered nature of
-observations may yield multilevel data. In the typical case, individuals
-are nested within groups, but there are many different types of
-multilevel data. In the medical field, clustering occurs at e.g., the
-hospital or center level in registry data, or at the study-level in
-meta-analyses (IPDMA). In the social sciences and official statistics we
-can find clustering e.g.~at the country-level, or as imposed by a
-multi-stage sampling design. For the sake of legibility, we will refer
-to the grouping variable as `cluster', and the grouped variable as
-`(sample) unit' throughout this paper. And, for reasons of brevity, we
-only discuss clustering between units, not within units (e.g.,
-timeseries or longitudinal data). Also, add that we'll only discuss two
-levels, not more?}
-
 In many contemporary data analysis efforts, some form of hierarchical or
-clustered structure is recorded. Ignoring such structures may be harmful
-to the inferences and can yield biased results. On the other hand,
-analyzing such multilevel data requires specialized techniques that take
-the clustered structure into account. Imagine a case where cross-level
-interactions between unit-level variables and cluster-level variables
-are present. The cluster to which a unit belongs may then influence the
-unit-level observations - and vice versa - for each of the units that
-make up the cluster \citep{hox17}. These relations can and should be
-taken into account when developing analysis models for multilevel data
-for the simple reason that groups of observations share some common
-variance.
+clustered structure is recorded.\footnote{Add IPDMA back in: patient
+  level data from multiple studies.} Ignoring such structures may be
+harmful to the inferences and can yield biased results. On the other
+hand, analyzing such multilevel data requires specialized techniques
+that take the clustered structure into account. Imagine a case where
+cross-level interactions between unit-level variables and cluster-level
+variables are present. The cluster to which a unit belongs may then
+influence the unit-level observations - and vice versa - for each of the
+units that make up the cluster \citep{hox17}. These relations can and
+should be taken into account when developing analysis models for
+multilevel data for the simple reason that groups of observations share
+some common variance.
 
@@
 Column \texttt{X2} is systematically missing and column \texttt{X3} is
-sporadically missing.\footnote{Explain why. GV: niet nodig}
+sporadically missing.\footnote{Explain why. GV: not necessary}
 
@@
   mechanisms}
-
-\[
-\begin{matrix} 
-                          &X1     & X2          & X3            \\
-   \hline
-   \mathrm{cluster 1}\{ \ &x_{11} &x_{12}       &\mathtt{NA}    \\
-                             \ &x_{21}  &x_{22}       &x_{32}         \\
-   \mathrm{cluster 2}\{  &x_{31} &\mathtt{NA}  &x_{33}         \\
-                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
-   \mathrm{cluster 3}\{  &x_{51} &x_{52}       &x_{53}         \\
-\end{matrix} 
-\] {[}GV Ik vind de latex tabel niet tof. {]}
 
@@
 
+Here we can set \texttt{maxit} to 1\ldots{}
+
 \hypertarget{discussion}{%

### 2022-01-14 Manuscript/Manuscript.Rmd
add cluster to missing data pattern

@@
 Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
-The cluster to which a unit belongs may then influence the unit-level observations - and vice versa - for each of the units that make up the cluster [@hox17].
+The cluster to which a unit belongs may then influence the unit-level observations--and vice versa--for each of the units that make up the cluster [@hox17].
 These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
@@
 Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power.
-Instead, the missingness should be accommodated *before* or *within* the analysis of scientific interest.
+Instead, the missingness should be accommodated \emph{before} or _within_ the analysis of scientific interest.
 Especially the former is very generic and popular and is widely known as imputation.

### 2022-01-14 Manuscript/Manuscript.tex
add cluster to missing data pattern

@@
 variables are present. The cluster to which a unit belongs may then
-influence the unit-level observations - and vice versa - for each of the
+influence the unit-level observations--and vice versa--for each of the
 units that make up the cluster \citep{hox17}. These relations can and
@@
 lowers statistical power. Instead, the missingness should be
-accommodated \emph{before} or \emph{within} the analysis of scientific
+accommodated \emph{before} or \textbf{within} the analysis of scientific
 interest. Especially the former is very generic and popular and is

### 2022-01-14 R/plot_incomplete.R

@@ -2,11 +2,13 @@
 # missingness indicator plot (monotone vs non-monotone)
 plot_md_set <- function(...){
   miss_ind <- tibble(
-  rownr = 5:1,
-  X1 = rep("observed", 5),
-  X2 = c(rep("observed", 2), rep("missing", 2), "observed"),
-  X3 = c("missing", rep("observed", 2), "missing", "observed")
-) %>% pivot_longer(cols = c(X1, X2, X3))
+    rownr = 5:1,
+    cluster = rep("observed", 5),
+    X1 = rep("observed", 5),
+    X2 = c(rep("observed", 2), rep("missing", 2), "observed"),
+    X3 = c("missing", rep("observed", 2), "missing", "observed")
+  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3)) %>% 
+    cbind(text = c(1, "", "", "", 1, "", "", "", 2, "", "", "", 2, "", "", "", "N", "", "", ""))
 
 # plot
 miss_ind %>% 
@@ -19,17 +21,18 @@ miss_ind %>%
     height = 0.8
     )) +
   geom_tile(fill = "white", size = 1.2) +
-  scale_x_discrete(position = "top", labels = c("", "", "X1", "X2", "X3")) +
+  geom_text(aes(label = text), color = "black") +
+  scale_x_discrete(position = "top", labels = c("cluster", "X1", "X2", "X3")) +
   scale_y_discrete() +
   scale_color_manual(values = plot_col, name = "Legend:") +
   theme_minimal() +
   labs(x = "", y = "") +
-  annotate(geom = "text", x = "X-1", y = "4", label = "Cluster 1 ", vjust = -1, hjust = 0, size = 3) +
-  annotate(geom = "text", x = "X-1", y = "2", label = "Cluster 2 ", vjust = -1, hjust = 0, size = 3) +
-  annotate(geom = "text", x = "X-1", y = "0", label = "Cluster N ", vjust = -1, hjust = 0, size = 3) +
-  annotate(geom = "text", x = "X0", y = "4", label = "{", vjust = 0, hjust = 0, size = 16) +
-  annotate(geom = "text", x = "X0", y = "2", label = "{", vjust = 0, hjust = 0, size = 16) +
-  annotate(geom = "text", x = "X0", y = "0", label = "{...", vjust = 0, hjust = 0, size = 16) +
[…]

### 2022-01-19 Manuscript/Manuscript.Rmd
use subset of popularity data as case study

@@
 icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
+# icc for imputed data as function in mice
 miceadds::source.all(path = "../R")
@@
 
-- `mice::popmis` (simulated data on school kids, with MNAR/MAR mixture);
+- `mice::popmis` (simulated data on school kids, n = 199 units, N = 10 clusters, with MNAR/MAR mixture);
 
-- `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s);
+- `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s);^[add cross-level interaction?]
 
@@
 ```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
+# randomly select a few classes
+clusters <- sample(1:100, 10)
+
 # load data 
-pop <- readRDS("../Data/popNCR2.RDS") %>% mutate(class = as.factor(class))
+pop <- readRDS("../Data/popNCR2.RDS") %>% 
+  filter(class %in% clusters) %>% 
+  mutate(class = as.factor(class))
 
@@
 
-`popNCR` is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:
+`popNCR2` is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:
 
@@
 
-```{r pop_pat, echo=FALSE, fig.height=4, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
+```{r pop_pat, echo=FALSE, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
 # missingness
@@
   vars = c("popular", "popteach", "texp"), 
-  incomplete = c(icc(popular ~ class, pop), 
+  CCA = c(icc(popular ~ class, pop), 
                icc(popteach ~ class, pop),

### 2022-01-19 Manuscript/Manuscript.tex
use subset of popularity data as case study

@@
 lowers statistical power. Instead, the missingness should be
-accommodated \emph{before} or \textbf{within} the analysis of scientific
+accommodated \emph{before} or \emph{within} the analysis of scientific
 interest. Especially the former is very generic and popular and is
@@
 \item
-  \texttt{mice::popmis} (simulated data on school kids, with MNAR/MAR
-  mixture);
+  \texttt{mice::popmis} (simulated data on school kids, n = 199 units, N
+  = 10 clusters, with MNAR/MAR mixture);
 \item
   \texttt{metamisc::impact} (real IPD on traumatic brain injuries,
-  without \texttt{NA}s);
+  without \texttt{NA}s);\footnote{add cross-level interaction?}
 \item
@@
 
-\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
-where the number of units \(n = 2000\), and the number of clusters
-\(N = 100\), on 7 variables:
+\texttt{popNCR2} is a simulated dataset with pupils clustered in
+classes, where the number of units \(n = 2000\), and the number of
+clusters \(N = 100\), on 7 variables:
 
@@
 \texttt{class}. We can express this using the intra-class correlation
-(ICC). For \texttt{popular} the ICC is 0.33. For \texttt{popteach} it is
-0.34. It would thus be wise to use multilevel modeling.
+(ICC). For \texttt{popular} the ICC is 0.4. For \texttt{popteach} it is
+0.36. It would thus be wise to use multilevel modeling.
 
@@
 \begin{CodeOutput}
-      vars incomplete   ignored
-1  popular  0.3280070 0.3338626
-2 popteach  0.3414766 0.3414766
+      vars       CCA   ignored
+1  popular 0.3989463 0.4235643
+2 popteach 0.3605902 0.3605902
 3     texp 1.0000000 1.0000000
@@
 \begin{CodeOutput}
-      vars incomplete   ignored predictor
-1  popular  0.3280070 0.3338626 0.3871020
-2 popteach  0.3414766 0.3414766 0.3414766
+      vars       CCA   ignored predictor
+1  popular 0.3989463 0.4235643 0.3904372
+2 popteach 0.3605902 0.3605902 0.3605902
 3     texp 1.0000000 1.0000000 1.0000000
@@
 \begin{CodeOutput}
-      vars incomplete   ignored predictor      norm
-1  popular  0.3280070 0.3338626 0.3871020 0.3612830
-2 popteach  0.3414766 0.3414766 0.3414766 0.3414766
+      vars       CCA   ignored predictor      norm
+1  popular 0.3989463 0.4235643 0.3904372 0.4165006
+2 popteach 0.3605902 0.3605902 0.3605902 0.3605902
 3     texp 1.0000000 1.0000000 1.0000000 1.0000000
@@
 \begin{CodeOutput}
-      vars incomplete   ignored predictor      norm       pan
-1  popular  0.3280070 0.3338626 0.3871020 0.3612830 0.3745108
-2 popteach  0.3414766 0.3414766 0.3414766 0.3414766 0.3414766
+      vars       CCA   ignored predictor      norm       pan
+1  popular 0.3989463 0.4235643 0.3904372 0.4165006 0.3827595
+2 popteach 0.3605902 0.3605902 0.3605902 0.3605902 0.3605902
 3     texp 1.0000000 1.0000000 1.0000000 1.0000000 1.0000000

### 2022-01-19 R/load_popularity.R

@@ -7,6 +7,10 @@ pop_complete <-
 
 mod <- lmer(popular ~ sex + extrav + texp + extrav * texp + (1 | class) + (1 | texp) +  (1 | extrav), pop_complete) %>% broom.mixed::tidy()
 
+# parameters
+N = 5
+n = 25
+
 # # synthesize some more observations
 # icc(popular ~as.factor(class), pop_complete)
 # 

### 2022-01-19 R/plot_imputed.R

@@ -89,7 +89,8 @@ plot_box <- function(imp, x, strip = FALSE) {
         color = datapoint
       ),
       width = 0.5,
-      alpha = 0.5) +
+      alpha = 0.5,
+      outlier.shape = NA) +
       scale_y_discrete(limits = rev) +
       ggplot2::ylab("Imputation (0 = observed data)")
   # } else {

### 2022-01-19 R/plot_incomplete.R

@@ -2,13 +2,21 @@
 # missingness indicator plot (monotone vs non-monotone)
 plot_md_set <- function(...){
   miss_ind <- tibble(
-    rownr = 5:1,
-    cluster = rep("observed", 5),
-    X1 = rep("observed", 5),
-    X2 = c(rep("observed", 2), rep("missing", 2), "observed"),
-    X3 = c("missing", rep("observed", 2), "missing", "observed")
+    rownr = 6:1,
+    cluster = c(rep("observed", 4), "blank", "observed"),
+    X1 = c(rep("observed", 4), "blank", "observed"),
+    X2 = c(rep("observed", 2), rep("missing", 2), "blank", "observed"),
+    X3 = c("missing", rep("observed", 2), "missing", "blank", "observed")
   ) %>% pivot_longer(cols = c(cluster, X1, X2, X3)) %>% 
-    cbind(text = c(1, "", "", "", 1, "", "", "", 2, "", "", "", 2, "", "", "", "N", "", "", ""))
+    cbind(text = c(
+      1, "", "", "", 
+      1, "", "", "", 
+      2, "", "", "", 
+      2, "", "", "", 
+      "...", "...", "...", "...", 
+      #"\u22EE", "\u22EE", "\u22EE", "\u22EE", 
+      "N", "", "", "")
+      )
 
 # plot
 miss_ind %>% 
@@ -24,7 +32,7 @@ miss_ind %>%
   geom_text(aes(label = text), color = "black") +
   scale_x_discrete(position = "top", labels = c("cluster", "X1", "X2", "X3")) +
   scale_y_discrete() +
-  scale_color_manual(values = plot_col, name = "Legend:") +
+  scale_color_manual(values = c(plot_col, blank = "white"), name = "Legend:") +
   theme_minimal() +
   labs(x = "", y = "") +
   # annotate(geom = "text", x = "X-1", y = "4", label = "Cluster 1 ", vjust = -1, hjust = 0, size = 3) +
@@ -35,7 +43,8 @@ miss_ind %>%
   # annotate(geom = "text", x = "X0", y = "0", label = "{...", vjust = 0, hjust = 0, size = 16) +
   theme(
     axis.text.y = element_blank(),
-    # legend.position = "bottom", 
+    # text=element_text(family="Arial Unicode MS"),
+    legend.position = "none", 
     panel.grid.major = element_blank())
 }
 
@@ -134,12 +143,27 @@ plot_md_pat <- function(dat) {
   return(p)
 }
 
-# plot conditional distributions
[…]

### 2022-01-20 Manuscript/Manuscript.Rmd
Add Heckman model function and notes

@@
   - `condom` Condom use during last intercourse,
-  - `smoke` Smoker (levels).
+  - `smoke` Smoker (levels; inclusion restriction variable).
+
+There is a specific MNAR mechanism.
+
+- main model = analysis model = predicting the prevalence of hiv
+- people who have hiv are more reluctant to taking the test, not to reveal their status
+- solution: take variables from analysis model like condom use, marital status, age, etc. that relate to hiv prevalence
+- exclusion restriction variable = instrumental variable (from economics) = e.g.
+how kind the nurse was = smoke in this dataset = should be correlated to the missingness indicator, but *not* the variable itself
+- include these variables in the selection equation
+- take the missingness indicator of the variable of interest for the selection equation as DV
+- heckman models estimate BOTH selection and mean models simultaneously, because linked through the error term
+- if the errors are not correlated, the models are not linked at all -> not MNAR, but just MAR
+- if the rho is significantly different from zero, the errors are correlated and the models are related so we have MNAR
+- now, we add clustering
+- copulas = estimate 2 equations separately and use another equation to link them
+- for each of the clusters, est the copula separately, so we have all coeffs for each cluster, incl the rho and sigma (the errors)
+- put together all of the estimates through a meta-analysis model to get population distribution of the estimates
+- use the population parameter estimates to draw marginal parameters from the distribution to get a cluster/study parameter
+- with systematical miss we can est the values of the imputed values based on the marginal distribution
+- optional to add shrinkage: the marginal in account as well
+- the hiv data was simulated by the copula function/model developer
+- we could add a systematically missing variable in the data, because these are all sporadically
 

### 2022-01-21 Manuscript/Manuscript.Rmd
check assumptions Heckman model

@@
 
-```{r pop_norm_eval}
+```{r pop_norm_eval, echo=FALSE}
 # plot(imp_norm)
@@
 
-```{r pop_pan_eval}
+```{r pop_pan_eval, echo=FALSE}
 # plot(imp_pan)
@@
 
-Here we can set `maxit` to 1...
+From the missing data pattern we see that we can set `maxit` to 1, since there is only one variable with missingness.
+
+```{r include=FALSE}
+# inclusion restriction variable 
+mod_main <- glm(hiv ~ ., family = "binomial", data = hiv) %>% 
+  broom::tidy(conf.int = TRUE) %>% 
+  .[7, c(2,6,7)] %>% 
+  round(., 3)
+mod_select <- glm(is.na(hiv) ~ ., family = "binomial", data = hiv) %>% 
+  broom::tidy(conf.int = TRUE) %>% 
+  .[7, c(2,6,7)] %>% 
+  round(., 3)
+```
+
+The inclusion restriction variable should be a predictor of the missingness indicator for the variable of interest, but *not* of the actual value of the variable of interest.
+In this case, the data were simulated to adhere to this requirement.
+Namely, $\beta_{smoke}$ = `r mod_main$estimate`, 95% CI [`r paste0(mod_main$conf.low, ", ", mod_main$conf.high)`] for the analysis model (`formula = hiv ~ .`), and $\beta_{smoke}$ = `r mod_select$estimate`, 95% CI [`r paste0(mod_select$conf.low, ", ", mod_select$conf.high)`] for the selection model (`formula = is.na(hiv) ~ .`).
+This means the assumptions for the Heckman model are met.
 

### 2022-01-21 Manuscript/Manuscript.tex
check assumptions Heckman model

@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> # plot(imp_norm)
-R> plot_box(imp_norm_2l, x = "popular", strip = TRUE)
-\end{CodeInput}
 
@@
 
-\begin{CodeInput}
-R> ICCs <- ICCs %>% mutate(
-+            norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
-+                     icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
-+                     icc(texp ~ as.factor(class), complete(imp_norm_2l)))
-+            )
-R> ICCs
-\end{CodeInput}
 \begin{CodeOutput}
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> # plot(imp_pan)
-R> plot_box(imp_pan_2l, x = "popular", strip = TRUE)
-\end{CodeInput}
 
@@
 
-\begin{CodeInput}
-R> ICCs <- ICCs %>% mutate(
-+            pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
-+                     icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
-+                     icc(texp ~ as.factor(class), complete(imp_pan_2l)))
-+            )
-R> ICCs
-\end{CodeInput}
 \begin{CodeOutput}
@@
 \item
-  \texttt{smoke} Smoker (levels).
+  \texttt{smoke} Smoker (levels; inclusion restriction variable).
+\end{itemize}
+
+There is a specific MNAR mechanism.
+
+\begin{itemize}
+\tightlist
+\item
+  main model = analysis model = predicting the prevalence of hiv
+\item
+  people who have hiv are more reluctant to taking the test, not to
+  reveal their status
+\item
+  solution: take variables from analysis model like condom use, marital
+  status, age, etc. that relate to hiv prevalence
+\item
+  exclusion restriction variable = instrumental variable (from
+  economics) = e.g.~how kind the nurse was = smoke in this dataset =
+  should be correlated to the missingness indicator, but \emph{not} the
+  variable itself
+\item
+  include these variables in the selection equation
+\item
+  take the missingness indicator of the variable of interest for the
+  selection equation as DV
+\item
+  heckman models estimate BOTH selection and mean models simultaneously,
+  because linked through the error term
+\item
+  if the errors are not correlated, the models are not linked at all
+  -\textgreater{} not MNAR, but just MAR
+\item
+  if the rho is significantly different from zero, the errors are
+  correlated and the models are related so we have MNAR
+\item
+  now, we add clustering
+\item
+  copulas = estimate 2 equations separately and use another equation to
+  link them
+\item
+  for each of the clusters, est the copula separately, so we have all
+  coeffs for each cluster, incl the rho and sigma (the errors)
+\item
+  put together all of the estimates through a meta-analysis model to get
+  population distribution of the estimates
+\item
+  use the population parameter estimates to draw marginal parameters
+  from the distribution to get a cluster/study parameter
+\item
+  with systematical miss we can est the values of the imputed values
+  based on the marginal distribution
+\item
+  optional to add shrinkage: the marginal in account as well
+\item
+  the hiv data was simulated by the copula function/model developer
+\item
+  we could add a systematically missing variable in the data, because
+  these are all sporadically
 \end{itemize}
@@
 
-Here we can set \texttt{maxit} to 1\ldots{}
+From the missing data pattern we see that we can set \texttt{maxit} to
+1, since there is only one variable with missingness.
+
+The inclusion restriction variable should be a predictor of the
+missingness indicator for the variable of interest, but \emph{not} of
+the actual value of the variable of interest. In this case, the data
+were simulated to adhere to this requirement. Namely, \(\beta_{smoke}\)
+= -0.064, 95\% CI {[}-0.256, 0.126{]} for the analysis model
+(\texttt{formula\ =\ hiv\ \textasciitilde{}\ .}), and \(\beta_{smoke}\)
+= -0.265, 95\% CI {[}-0.422, -0.11{]} for the selection model
+(\texttt{formula\ =\ is.na(hiv)\ \textasciitilde{}\ .}). This means the
+assumptions for the Heckman model are met.
 

### 2022-01-21 R/plot_imputed.R

@@ -62,7 +62,7 @@ plot_strip <- function(imp, x) {
 }
 
 # boxplot
-plot_box <- function(imp, x, strip = FALSE) {
+plot_box <- function(imp, x, cluster = NULL, strip = FALSE) {
   # plot
   if (strip){
     p <- imp %>% plot_imps(x) +
@@ -109,7 +109,9 @@ plot_box <- function(imp, x, strip = FALSE) {
   #       ) %>% setNames(0:imp$m))
   #     )
   # }
-
+  if(!is.null(cluster)){
+    p <- p + facet_wrap(get(cluster))
+    }
     # output
   return(p)
 }

### 2022-01-21 R/plot_incomplete.R

@@ -159,11 +159,10 @@ plot_md_pat <- function(dat) {
 plot_conditional <- function(data, x, z, cluster, N = 4, ...){
   data %>% 
     filter(as.numeric(get(cluster)) < N+1) %>% 
+    mutate(cluster = factor(get(cluster), labels = paste("Cluster", 1:N))) %>% 
     ggplot(aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
-    # geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.1, fill = plot_col[2], color = NA) +
-    # geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.1, fill = plot_col[1], color = NA) +
     geom_density() +
-    facet_wrap(~get(cluster), labeller = label_both) +
+    facet_wrap(~cluster) +
     scale_color_manual(values = plot_col, name = z) +
     theme_classic() +
     theme(legend.position = "bottom") +

### 2022-01-26 Manuscript/Manuscript.Rmd
Ampute impact data

@@
 library(miceadds)
-library(metamisc)
-library(GJRM)
-library(pan)
+# library(metamisc)
+# library(GJRM)
+# library(pan)
 
@@
 # alternatively add a quartile line around the density with geom_ribbon
+
+## comments 24/1
+# pop obs and pop miss in legend
+# add to caption
+# add fill
+# overlap coeff??
+# make densities proportional
+# 
 ```
@@
 
-```{r impact, echo=FALSE}
+```{r impact, echo=FALSE, fig.width=5}
 # load data
-data("impact")
+data("impact", package = "metamisc")
+impact_NA <- readRDS("../Data/impact_NA.RDS")
 # # descriptive statistics
@@
 # # missingness
-# plot_md_pat(impact)
+plot_md_pat(impact_NA)
+
+# md pattern in first cluster
+plot_md_pat(impact_NA[impact_NA$name == "TINT", ])
+plot_md_pat(impact_NA[impact_NA$name == "APOE", ])
 # TODO: rotate variable labels
@@
 # data
-data("hiv")
+data("hiv", package = "GJRM")
 hiv <- hiv %>% 
@@
 
-The inclusion restriction variable should be a predictor of the missingness indicator for the variable of interest, but *not* of the actual value of the variable of interest.
+The inclusion restriction variable should be a predictor of the the actual value of the variable of interest, but *not* of missingness indicator for the variable of interest.
 In this case, the data were simulated to adhere to this requirement.

### 2022-01-26 Manuscript/Manuscript.tex
Ampute impact data

@@
 
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-2} \end{center}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-3} \end{center}
+
+\end{CodeChunk}
+
 \hypertarget{case-study-iii-hiv}{%
@@
 
-The inclusion restriction variable should be a predictor of the
-missingness indicator for the variable of interest, but \emph{not} of
-the actual value of the variable of interest. In this case, the data
-were simulated to adhere to this requirement. Namely, \(\beta_{smoke}\)
-= -0.064, 95\% CI {[}-0.256, 0.126{]} for the analysis model
+The inclusion restriction variable should be a predictor of the the
+actual value of the variable of interest, but \emph{not} of missingness
+indicator for the variable of interest. In this case, the data were
+simulated to adhere to this requirement. Namely, \(\beta_{smoke}\) =
+-0.064, 95\% CI {[}-0.256, 0.126{]} for the analysis model
 (\texttt{formula\ =\ hiv\ \textasciitilde{}\ .}), and \(\beta_{smoke}\)

### 2022-01-26 R/plot_incomplete.R

@@ -53,8 +53,8 @@ plot_md_pat <- function(dat) {
   # get md pattern and store additional info
   pat <- mice::md.pattern(dat, plot = FALSE)
   vrb <- colnames(pat)[-ncol(pat)]
-  vrb_labs <- abbreviate(vrb, minlength = 3)
-  vrb_caption <- purrr::map2_chr(vrb_labs, vrb, ~{paste(.x, " = ", .y)}) %>% paste(collapse = ", ")
+  vrb_labs <- abbreviate(vrb, minlength = 5)
+  vrb_caption <- paste(vrb_labs[vrb_labs!=vrb], vrb[vrb_labs!=vrb], sep = " = ") %>% paste(collapse = ", ")
   colnames(pat) <- c(vrb, "NA_per_pat")
   pat_freq <- as.numeric(rownames(pat))[-nrow(pat)]
   NA_per_pat <- as.numeric(pat[, ncol(pat)])[-nrow(pat)]

### 2022-01-28 Manuscript/Manuscript.Rmd
Add references about cluster-as-predictor method

@@
 
-- `mice::popmis` (simulated data on school kids, n = 199 units, N = 10 clusters, with MNAR/MAR mixture);
+- `mice::popmis` (simulated data on school kids, n = 199 units, N = 10 clusters, with MAR missingness in the outcome);
+
+- `GJRM::hiv` (simulated patient data on HIV, with MNAR missingness in the outcome)
 
 - `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s);^[add cross-level interaction?]
-
-- `GJRM::hiv` (simulated patient data on HIV, without `NA`s)
 
@@
 In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model.
-For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data.
+For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data:
 
-# Case Study I: Popularity
+A. Choose an analysis model (so the imputation model will be compatible with the analyses);
+B. Evaluate the incomplete data;
+C. Develop the imputation model(s);
+D. Impute the missingness;
+E. Evaluate the imputations.
+
+# How (not) to impute (Case Study I: Popularity)
 
@@
 ```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
+# plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
 # TODO: check if up-side-down plot works
@@
 
-### Imputation model
+### Complete case analysis (not recommended)
+
+Complete case analysis ignores the observations with missingness altogether, which may even introduce bias in MCAR situations.
+
+### Imputation ignoring the cluster variable (not recommended)
 
@@
 ```
-
-### Imputed data
 
@@
 
-### Imputation model
+### Imputation with the cluster variable as predictor (not recommended)
 
 We'll now use `class` as a predictor to impute all other variables.
-This is still not recommended practice, since it only works under certain circumstances and results may be biased.
+This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
 But at least, it includes some multilevel aspect.
+This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende2016].
 Colloquially, this is 'multilevel imputation for dummies'.
+
+<!-- "Beyond the potential bias, there are other reasons why a multilevel imputation model should be preferred if possible.
+First, the model is guaranteed to produce unbiased results even in extreme cases with very high missingness rates and/or small intraclass correlations.
+Second, with the multilevel model, it is possible to include additional variables that are constant on the cluster level, for example, information on the teachers if the cluster levels are chosen to be students within classes or background information on the schools if the clustering is modeled as students within schools.
+This approach is not possible with fixed effects imputation.
+While this limitation should not be problematic in terms of bias (the combined school effects are still modeled through the cluster specific intercepts), using this additional information will make the random effects imputation model more efficient." -->
 
@@
 
-Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MNAR.
+Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MAR.
 
@@
 
-### Imputation model
+### Imputation with random effects
 
-To include...
+With `2l.norm` we impute the outcome with a multilevel model  assuming random slopes for each variable in the imputation model and homogeneous within-cluster variance.
 
@@
 
+### Imputation with random effects and heterogeneity
+
+This method assumes random slopes for each variable in the imputation model. In contrast to `2l.norm` this method allows a cluster-specific residual error variance.
+
+<!-- "By default, 2l.pan includes the intercept as both a fixed and a random effect.
+...
+There are 5 different codes you can use in the predictor matrix for variables imputed with 2l.pan.
+The person identifier is coded as -2 (this is different from 2l.norm).
+To include predictor variables with fixed or random effects, these variables are coded with 1 or 2, respectively.
+If coded as 2, the corresponding fixed effect is automatically included.
+In addition, 2l.pan offers the codes 3 and 4, which have similar meanings as 1 and 2 but will include an additional fixed effect for the person mean of that variable.
+This is useful if you're trying to model within- and between-person effects of time-varying predictor variables." -->
+
 ```{r pop_pan}
@@
 
-# Case study II: IMPACT
+# How to handle non-random selection (Case study II: HIV)
+
+Data are simulated and included in the `GJRM` package. We will use the following variables:
+
+  - `region` Cluster variable,
+  - `hiv` HIV diagnosis (0=no, 1=yes),
+  - `age` Age of the patient,
+  - `marital` Marital status,
+  - `condom` Condom use during last intercourse,
+  - `smoke` Smoker (levels; inclusion restriction variable).
+
+The imputation of these date is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
+
+<!-- - main model = analysis model = predicting the prevalence of hiv -->
+<!-- - people who have hiv are more reluctant to taking the test, not to reveal their status -->
+<!-- - solution: take variables from analysis model like condom use, marital status, age, etc. that relate to hiv prevalence -->
+<!-- - exclusion restriction variable = instrumental variable (from economics) = e.g.
+how kind the nurse was = smoke in this dataset = should be correlated to the missingness indicator, but *not* the variable itself -->
+<!-- - include these variables in the selection equation -->
+<!-- - take the missingness indicator of the variable of interest for the selection equation as DV -->
+<!-- - heckman models estimate BOTH selection and mean models simultaneously, because linked through the error term -->
+<!-- - if the errors are not correlated, the models are not linked at all -> not MNAR, but just MAR -->
+<!-- - if the rho is significantly different from zero, the errors are correlated and the models are related so we have MNAR -->
+<!-- - now, we add clustering -->
+<!-- - copulas = estimate 2 equations separately and use another equation to link them -->
+<!-- - for each of the clusters, est the copula separately, so we have all coeffs for each cluster, incl the rho and sigma (the errors) -->
+<!-- - put together all of the estimates through a meta-analysis model to get population distribution of the estimates -->
+<!-- - use the population parameter estimates to draw marginal parameters from the distribution to get a cluster/study parameter -->
+<!-- - with systematical miss we can est the values of the imputed values based on the marginal distribution -->
+<!-- - optional to add shrinkage: the marginal in account as well -->
+<!-- - the hiv data was simulated by the copula function/model developer -->
+<!-- - we could add a systematically missing variable in the data, because these are all sporadically -->
+
+```{r hiv, echo=FALSE}
+# data
+data("hiv", package = "GJRM")
+hiv <- hiv %>% 
+  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
+  mutate(region = as.integer(region))
+# md pattern
+plot_md_pat(hiv)
+```
+
+From the missing data pattern we see that we can set `maxit` to 1, since there is only one variable with missingness.
+
+```{r include=FALSE}
+# inclusion restriction variable 
+mod_main <- glm(hiv ~ ., family = "binomial", data = hiv) %>% 
+  broom::tidy(conf.int = TRUE) %>% 
+  .[7, c(2,6,7)] %>% 
+  round(., 3)
+mod_select <- glm(is.na(hiv) ~ ., family = "binomial", data = hiv) %>% 
+  broom::tidy(conf.int = TRUE) %>% 
+  .[7, c(2,6,7)] %>% 
+  round(., 3)
+```
+
+The inclusion restriction variable should be a predictor of the the actual value of the variable of interest, but *not* of missingness indicator for the variable of interest.
+In this case, the data were simulated to adhere to this requirement.
+Namely, $\beta_{smoke}$ = `r mod_main$estimate`, 95% CI [`r paste0(mod_main$conf.low, ", ", mod_main$conf.high)`] for the analysis model (`formula = hiv ~ .`), and $\beta_{smoke}$ = `r mod_select$estimate`, 95% CI [`r paste0(mod_select$conf.low, ", ", mod_select$conf.high)`] for the selection model (`formula = is.na(hiv) ~ .`).
+This means the assumptions for the Heckman-type selection model are met.
+
+# How to handle multivariate missingness (Case study III: IMPACT)
 
@@
 
-# Case study III: HIV
-
-Toy example from [Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R). We will use the following variables:
-
-  - `region` Cluster variable,
-  - `hiv` HIV diagnosis (0=no, 1=yes),
-  - `age` Age of the patient,
-  - `marital` Marital status,
-  - `condom` Condom use during last intercourse,
-  - `smoke` Smoker (levels; inclusion restriction variable).
-
-There is a specific MNAR mechanism.
-
-- main model = analysis model = predicting the prevalence of hiv
-- people who have hiv are more reluctant to taking the test, not to reveal their status
-- solution: take variables from analysis model like condom use, marital status, age, etc. that relate to hiv prevalence
-- exclusion restriction variable = instrumental variable (from economics) = e.g.
-how kind the nurse was = smoke in this dataset = should be correlated to the missingness indicator, but *not* the variable itself
-- include these variables in the selection equation
-- take the missingness indicator of the variable of interest for the selection equation as DV
-- heckman models estimate BOTH selection and mean models simultaneously, because linked through the error term
-- if the errors are not correlated, the models are not linked at all -> not MNAR, but just MAR
-- if the rho is significantly different from zero, the errors are correlated and the models are related so we have MNAR
-- now, we add clustering
-- copulas = estimate 2 equations separately and use another equation to link them
-- for each of the clusters, est the copula separately, so we have all coeffs for each cluster, incl the rho and sigma (the errors)
-- put together all of the estimates through a meta-analysis model to get population distribution of the estimates
-- use the population parameter estimates to draw marginal parameters from the distribution to get a cluster/study parameter
-- with systematical miss we can est the values of the imputed values based on the marginal distribution
-- optional to add shrinkage: the marginal in account as well
-- the hiv data was simulated by the copula function/model developer
-- we could add a systematically missing variable in the data, because these are all sporadically
-
-```{r hiv, echo=FALSE}
-# data
-data("hiv", package = "GJRM")
-hiv <- hiv %>% 
-  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
-  mutate(region = as.integer(region))
-# md pattern
-plot_md_pat(hiv)
-```
-
-From the missing data pattern we see that we can set `maxit` to 1, since there is only one variable with missingness.
-
-```{r include=FALSE}
-# inclusion restriction variable 
-mod_main <- glm(hiv ~ ., family = "binomial", data = hiv) %>% 
-  broom::tidy(conf.int = TRUE) %>% 
-  .[7, c(2,6,7)] %>% 
-  round(., 3)
-mod_select <- glm(is.na(hiv) ~ ., family = "binomial", data = hiv) %>% 
-  broom::tidy(conf.int = TRUE) %>% 
-  .[7, c(2,6,7)] %>% 
-  round(., 3)
-```
-
-The inclusion restriction variable should be a predictor of the the actual value of the variable of interest, but *not* of missingness indicator for the variable of interest.
-In this case, the data were simulated to adhere to this requirement.
-Namely, $\beta_{smoke}$ = `r mod_main$estimate`, 95% CI [`r paste0(mod_main$conf.low, ", ", mod_main$conf.high)`] for the analysis model (`formula = hiv ~ .`), and $\beta_{smoke}$ = `r mod_select$estimate`, 95% CI [`r paste0(mod_select$conf.low, ", ", mod_select$conf.high)`] for the selection model (`formula = is.na(hiv) ~ .`).
-This means the assumptions for the Heckman model are met.
-
 # Discussion

### 2022-01-28 Manuscript/Manuscript.tex
Add references about cluster-as-predictor method

--- 2022-01-19 Manuscript/Manuscript.tex
@@
   \texttt{mice::popmis} (simulated data on school kids, n = 199 units, N
-  = 10 clusters, with MNAR/MAR mixture);
+  = 10 clusters, with MAR missingness in the outcome);
+\item
+  \texttt{GJRM::hiv} (simulated patient data on HIV, with MNAR
+  missingness in the outcome)
 \item
@@
   without \texttt{NA}s);\footnote{add cross-level interaction?}
-\item
-  \texttt{GJRM::hiv} (simulated patient data on HIV, without
-  \texttt{NA}s)
 \end{itemize}
@@
 model. For all case studies we discuss the nature of the incomplete
-data, the imputation model(s), and evaluation of the imputed data.
+data, the imputation model(s), and evaluation of the imputed data:
 
-\hypertarget{case-study-i-popularity}{%
-\section{Case Study I: Popularity}\label{case-study-i-popularity}}
+A. Choose an analysis model (so the imputation model will be compatible
+with the analyses); B. Evaluate the incomplete data; C. Develop the
+imputation model(s); D. Impute the missingness; E. Evaluate the
+imputations.
+
+\hypertarget{how-not-to-impute-case-study-i-popularity}{%
+\section{How (not) to impute (Case Study I:
+Popularity)}\label{how-not-to-impute-case-study-i-popularity}}
 
@@
 
-\begin{CodeChunk}
-\begin{figure}
+\hypertarget{complete-case-analysis-not-recommended}{%
+\subsubsection{Complete case analysis (not
+recommended)}\label{complete-case-analysis-not-recommended}}
 
-{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-1}
+Complete case analysis ignores the observations with missingness
+altogether, which may even introduce bias in MCAR situations.
 
-}
-
-\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist}
-\end{figure}
-\end{CodeChunk}
-
-\hypertarget{imputation-model}{%
-\subsubsection{Imputation model}\label{imputation-model}}
+\hypertarget{imputation-ignoring-the-cluster-variable-not-recommended}{%
+\subsubsection{Imputation ignoring the cluster variable (not
+recommended)}\label{imputation-ignoring-the-cluster-variable-not-recommended}}
 
@@
 \end{CodeChunk}
-
-\hypertarget{imputed-data}{%
-\subsubsection{Imputed data}\label{imputed-data}}
 
@@
 
-\hypertarget{imputation-model-1}{%
-\subsubsection{Imputation model}\label{imputation-model-1}}
+\hypertarget{imputation-with-the-cluster-variable-as-predictor-not-recommended}{%
+\subsubsection{Imputation with the cluster variable as predictor (not
+recommended)}\label{imputation-with-the-cluster-variable-as-predictor-not-recommended}}
 
@@
 variables. This is still not recommended practice, since it only works
-under certain circumstances and results may be biased. But at least, it
-includes some multilevel aspect. Colloquially, this is `multilevel
-imputation for dummies'.
+under certain circumstances and results may be biased
+\citep{drec15, ende16}. But at least, it includes some multilevel
+aspect. This method is also called `fixed cluster imputation', and uses
+N-1 indicator variables representing allocation of N clusters as a fixed
+factor in the model \citep{reit06, ende2016}. Colloquially, this is
+`multilevel imputation for dummies'.
 
@@
 \end{CodeChunk}
-
-\hypertarget{imputed-data-1}{%
-\subsubsection{Imputed data}\label{imputed-data-1}}
 
@@
 Now, we can clearly see that the imputed values of \texttt{texp} are
-higher than the observed values, which is in line with right-tailed
-MNAR.
+higher than the observed values, which is in line with right-tailed MAR.
 
@@
 
-\hypertarget{imputation-model-2}{%
-\subsubsection{Imputation model}\label{imputation-model-2}}
+\hypertarget{imputation-with-random-effects}{%
+\subsubsection{Imputation with random
+effects}\label{imputation-with-random-effects}}
 
-To include\ldots{}
+With \texttt{2l.norm} we impute the outcome with a multilevel model
+assuming random slopes for each variable in the imputation model and
+homogeneous within-cluster variance.
 
@@
 \end{CodeChunk}
+
+\hypertarget{imputation-with-random-effects-and-heterogeneity}{%
+\subsubsection{Imputation with random effects and
+heterogeneity}\label{imputation-with-random-effects-and-heterogeneity}}
+
+This method assumes random slopes for each variable in the imputation
+model. In contrast to \texttt{2l.norm} this method allows a
+cluster-specific residual error variance.
 
@@
 
-\hypertarget{case-study-ii-impact}{%
-\section{Case study II: IMPACT}\label{case-study-ii-impact}}
+\hypertarget{how-to-handle-non-random-selection-case-study-ii-hiv}{%
+\section{How to handle non-random selection (Case study II:
+HIV)}\label{how-to-handle-non-random-selection-case-study-ii-hiv}}
+
+Data are simulated and included in the \texttt{GJRM} package. We will
+use the following variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{region} Cluster variable,
+\item
+  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{marital} Marital status,
+\item
+  \texttt{condom} Condom use during last intercourse,
+\item
+  \texttt{smoke} Smoker (levels; inclusion restriction variable).
+\end{itemize}
+
+The imputation of these date is based on the toy example from
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}.
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/hiv-1} \end{center}
+
+\end{CodeChunk}
+
+From the missing data pattern we see that we can set \texttt{maxit} to
+1, since there is only one variable with missingness.
+
+The inclusion restriction variable should be a predictor of the the
+actual value of the variable of interest, but \emph{not} of missingness
+indicator for the variable of interest. In this case, the data were
+simulated to adhere to this requirement. Namely, \(\beta_{smoke}\) =
+-0.064, 95\% CI {[}-0.256, 0.126{]} for the analysis model
+(\texttt{formula\ =\ hiv\ \textasciitilde{}\ .}), and \(\beta_{smoke}\)
+= -0.265, 95\% CI {[}-0.422, -0.11{]} for the selection model
+(\texttt{formula\ =\ is.na(hiv)\ \textasciitilde{}\ .}). This means the
+assumptions for the Heckman-type selection model are met.
+
+\hypertarget{how-to-handle-multivariate-missingness-case-study-iii-impact}{%
+\section{How to handle multivariate missingness (Case study III:
+IMPACT)}\label{how-to-handle-multivariate-missingness-case-study-iii-impact}}
 
@@
 
-\hypertarget{case-study-iii-hiv}{%
-\section{Case study III: HIV}\label{case-study-iii-hiv}}
-
-Toy example from
-\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{Heckman
-Github repo}. We will use the following variables:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{region} Cluster variable,
-\item
-  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
-\item
-  \texttt{age} Age of the patient,
-\item
-  \texttt{marital} Marital status,
-\item
-  \texttt{condom} Condom use during last intercourse,
-\item
-  \texttt{smoke} Smoker (levels).
-\end{itemize}
-
 \begin{CodeChunk}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/hiv-1} \end{center}
+\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-2} \end{center}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-3} \end{center}
 
 \end{CodeChunk}
-
-Here we can set \texttt{maxit} to 1\ldots{}
 

### 2022-01-28 References/Huque2020/Simulation/Clustered data/100 clusters/Simulation scenario (iii)/Data analysis/Combined results from all methods/combinedDataHPC.R

@@ -1,0 +1,41 @@
+setwd("~/Simulation/Clustered data/100 clusters/Simulation scenario (iii)/Data analysis")
+
+### JM-MLMM
+resultMLMM1<-as.data.frame(read.csv("JM-MLMM/Results/resultpanRISM.csv",sep=",",header = FALSE))
+resultMLMM<-resultMLMM1[,3:ncol(resultMLMM1)]
+names(resultMLMM)<-c("cons","bmiz","seifa","varIntercept","varSlope","varResid","secons","sebmiz","seseifa","pbmiz","pSeifa","pCons")
+resultMLMM$Method<-rep(4,nrow(resultMLMM))
+### JM-FJ
+resultFJ1<-as.data.frame(read.csv("JM-FJ/Results/resultFJ.csv",sep=",",header = FALSE))
+resultFJ<-resultFJ1[,3:ncol(resultFJ1)]
+names(resultFJ)<-c("cons","bmiz","seifa","varIntercept","varSlope","varResid","secons","sebmiz","seseifa","pbmiz","pSeifa","pCons")
+resultFJ$Method<-rep(5,nrow(resultFJ))
+### JM-FJ-het
+resultFJhet1<-as.data.frame(read.csv("JM-FJ-het/Results/resultFJhet.csv",sep=",",header = FALSE))
+resultFJhet<-resultFJhet1[,3:ncol(resultFJhet1)]
+names(resultFJhet)<-c("cons","bmiz","seifa","varIntercept","varSlope","varResid","secons","sebmiz","seseifa","pbmiz","pSeifa","pCons")
+resultFJhet$Method<-rep(6,nrow(resultFJhet))
+
+### JM-SMC
+resultSMC1<-as.data.frame(read.csv("JM-SMC/Results/resultJomoSMC.csv",sep=",",header = FALSE))
+resultSMC<-resultSMC1[,3:ncol(resultSMC1)]
+names(resultSMC)<-c("cons","bmiz","seifa","varIntercept","varSlope","varResid","secons","sebmiz","seseifa","pbmiz","pSeifa","pCons")
+resultSMC$Method<-rep(7,nrow(resultSMC))
+### JM-SMC-het
+resultSMChet1<-as.data.frame(read.csv("JM-SMC-het/Results/resultJomoSMChet.csv",sep=",",header = FALSE))
+resultSMChet<-resultSMChet1[,3:ncol(resultSMChet1)]
[…]

### 2022-01-28 References/Huque2020/Simulation/Clustered data/100 clusters/Simulation scenario (iii)/Data analysis/JM-MLMM/codes/MLMMpan.R

@@ -1,0 +1,41 @@
+library(mitml)
+library(reshape2)
+library(mgcv)
+library(lme4)
+library(plyr)
+library(mice)
+library(foreign)
+library(jomo)
+
+strindex=Sys.getenv("PBS_ARRAY_INDEX")
+myindex=strtoi(strindex)
+lowerbound=(myindex-1) * 10  + 1
+upperbound=(myindex) * 10
+
+for (i in lowerbound:upperbound)
+{
+  dataname <- paste0("data", i,".dta")
+  datapan<-read.dta(dataname)
+  # Setup imputation model
+  fm1<-bmiz~1+QoL+seifa+sep+(1+QoL|ID)
+  imp<-panImpute(data=datapan,formula=fm1,m=30,seed=12345)
+  #summary(imp)
+  impList<-mitmlComplete(imp,print="all")
+  fit.lmm<-with(impList,lmer(QoL~bmiz+seifa+(1|ID) +(0+bmiz|ID)))
+  
+  lmm.fit<-testEstimates(fit.lmm,var.comp = TRUE)
+  
+  beta.lmm<-lmm.fit$estimates[,1]
+  se.lmm<-lmm.fit$estimates[,2]
+  varcom.lmm<-c(sqrt(lmm.fit$var.comp[1,1]),sqrt(lmm.fit$var.comp[2,1]),sqrt(lmm.fit$var.comp[3,1]))
+  
+  pm3<-0
+  pm3[(lmm.fit$estimates[1,1]+qt(0.025, lmm.fit$estimates[1,4])*lmm.fit$estimates[1,2]) < 0.50 &  (lmm.fit$estimates[1,1]+qt(0.975, lmm.fit$estimates[1,4])*lmm.fit$estimates[1,2]) > 0.50]<-1
+  pm1<-0
+  pm1[(lmm.fit$estimates[2,1]+qt(0.025, lmm.fit$estimates[2,4])*lmm.fit$estimates[2,2]) < -0.20  &  (lmm.fit$estimates[2,1]+qt(0.975, lmm.fit$estimates[2,4])*lmm.fit$estimates[2,2]) > -0.20]<-1
+  pm2<-0
[…]

### 2022-01-28 References/Huque2020/Simulation/Clustered data/100 clusters/Simulation scenario (iii)/Data analysis/JM-SMC-het/Codes/MLMMJomoSMChet.R

@@ -1,0 +1,45 @@
+library(mitml)
+library(reshape2)
+library(mgcv)
+library(lme4)
+library(plyr)
+#install.packages("mice")
+library(mice)
+library(foreign)
+library(jomo)
+
+strindex=Sys.getenv("PBS_ARRAY_INDEX")
+myindex=strtoi(strindex)
+lowerbound=(myindex-1) * 10  + 1
+upperbound=(myindex) * 10
+
+for (i in lowerbound:upperbound)
+{
+  dataname <- paste0("data", i,".dta")
+  datajomo<-read.dta(dataname)
+  #substantive model
+  formula<-QoL~bmiz+seifa+(1+bmiz|ID)
+  imp2<-jomo.lmer(formula,datajomo,level=c(2,2,1,1,1),nimp=30,output=2)
+  # Setup imputation model
+  impList <- jomo2mitml.list(imp2)
+  ## Analysis model 2
+  
+  imp2.fitMM <- lapply(impList, FUN=function(x){
+    lmer(QoL~bmiz+seifa+(1|clus)+(0+bmiz|clus),data=x)
+  })
+  
+  Jomo2l.fit<-testEstimates(imp2.fitMM,var.comp = TRUE)
+  beta.Jomo2l<-Jomo2l.fit$estimates[,1]
+  se.Jomo2l<-Jomo2l.fit$estimates[,2]
+  varcom.Jomo2l<-c(sqrt(Jomo2l.fit$var.comp[1,1]),sqrt(Jomo2l.fit$var.comp[2,1]),sqrt(Jomo2l.fit$var.comp[3,1]))
+  
+  pm3<-0
+  pm3[(Jomo2l.fit$estimates[1,1]+qt(0.025, Jomo2l.fit$estimates[1,4])*Jomo2l.fit$estimates[1,2]) < 0.50 &  (Jomo2l.fit$estimates[1,1]+qt(0.975, Jomo2l.fit$estimates[1,4])*Jomo2l.fit$estimates[1,2]) > 0.50]<-1
+  pm1<-0
[…]

### 2022-02-02 Manuscript/Manuscript.Rmd
Add some more refs to the intro

@@
 
-## Multilevel data
-
-In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.^[Add IPDMA back in: patient level data from multiple studies.] Ignoring such structures may be harmful to the inferences and can yield biased results.
-On the other hand, analyzing such multilevel data requires specialized techniques that take the clustered structure into account.
+In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.
+For example, students clustered within schools, or patients clustered within studies in individual patient data meta-analyses.
+Analyzing such multilevel data requires specialized techniques that take the clustered structure into account, since ignoring it may be harmful to the statistical inferences and can yield biased results.
 Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
-The cluster to which a unit belongs may then influence the unit-level observations--and vice versa--for each of the units that make up the cluster [@hox17].
+The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster [@hox17].
 These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
@@
 Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters.
-Additionally, there may be random predictor effects and/or random error terms (residual error variances), see e.g.
-@hox17 and @jong21.^[Add that heterogeneity refers to variability within clusters.] There are many names for models that take clustering into account.
+Additionally, models may include random effects of predictors across clusters, and random error terms (heterogeneous residual error variances; see e.g.
+@hox17 and @jong21).
+There are many names for models that take clustering into account.
 Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
 
-## Missing data
+## Incomplete multilevel data
 
@@
 
-The process of analyzing multilevel data is further complicated when not all observations are observed.
+The process of analyzing multilevel data is further complicated when not all data entries are observed.
 Just as with single level data, missingness may occur at the unit level.
 But with multiple levels of data comes the potential for clustered missingness.
-Missingness in multilevel data can therefore be categorized into two general patterns: systematic missingness and sporadic missingness, see @resc13.
+Missingness in multilevel data can therefore be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13].
 We have visualized the difference between these types of missingness in Figure 1.
-The figure shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n=5$ units and $p=3$ columns distributed over 3 clusters.
+The figure shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns.
 Column `X1` is completely observed.
-Column `X2` is systematically missing and column `X3` is sporadically missing.^[Explain why.
-GV: not necessary]
-
+Column `X2` is systematically missing and column `X3` is sporadically missing.
 Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters).
@@
 The optimal strategy for dealing with the missingness may therefore depend on the observed missing data pattern.
-^[add missing data mechanisms here?
-GV: No, patterns are different from mechanisms]
+
+Another characteristic of the missing data to take into account in analyses is the mechanism behind the missingness.
+Although the essence of the true non-response mechanism may not be known, it can be inferred or assumed to be one of the following:
+
+- Missing Completely At Random (MCAR), where the probability to be missing is equal across all data entries;
+- Missing At Random (MAR), where the probability to be missing depends on observed information;
+- Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94].
+Depending on the assumed missingness mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.
 
@@
 \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
-In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.^[Discuss missingness mechanisms before this point, add references @yuce08 and @hox15.]
+In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.
 
@@
 We provide practical guidelines and code snippets for different missing data situations.
-For reasons of brevity, we focus on imputation by chained equations^[add that JOMO is available in \pkg{mice} as well?].
-Other useful packages for incomplete multilevel data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed.
+For reasons of brevity, we focus on imputation by chained equations exclusively (although the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well).
+Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and the empirical work by @audi18 and @grun18.
+Note that this tutorial paper assumes a basic level of knowledge on multilevel models.
 We're providing an overview of implementations.
 It's up-to the reader to decide which multilevel strategy suits their data.
-So we won't go into detail for the different methods (and equations).
-Refer to @meng94, @audi18, and @grun18.
-This paper is just a software tutorial.
-We'll keep it practical.]
+We won't go into detail for the different methods (and equations).
+This paper is just a software tutorial, so we'll keep it practical.
 
-We structure this tutorial around three case studies:
+To illustrate how to impute incomplete multilevel data, we structure this tutorial around three case studies:
 
-- `mice::popmis` (simulated data on school kids, n = 199 units, N = 10 clusters, with MAR missingness in the outcome);
+- `mice::popmis` (simulated data on perceived popularity, $n = 199$ pupils across $N = 10$ classes, with univariate MAR missingness);
+- `GJRM::hiv` (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, with univariate MNAR missingness)
+- `metamisc::impact` (empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, without `NA`s).
 
-- `GJRM::hiv` (simulated patient data on HIV, with MNAR missingness in the outcome)
-
-- `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s);^[add cross-level interaction?]
-
-For each case study we focus on a different aspect to illustrate how to impute incomplete multilevel data.
-In the `mice::popmis` data, we show the advantages of including the multilevel structure of the data into the imputation model.
-In the `metamisc::impact` data we'll show how to induce missingness and solve it in real-world data.
-In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model.
-For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data:
-
-A. Choose an analysis model (so the imputation model will be compatible with the analyses);
-B. Evaluate the incomplete data;
-C. Develop the imputation model(s);
-D. Impute the missingness;
-E. Evaluate the imputations.
+In each case study we highlight one aspect of the imputation workflow.
+With the `mice::popmis` data, we show different ways of designing an imputation model and what happens if you misspecify this model.
+With the `GJRM::hiv` data we extend the imputation model to include Heckman-type selection-inclusion methods.
+And with the `metamisc::impact` data we provide a example of multivariate missingness in real-world data.
+For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data: A.
+Choose an analysis model (so the imputation model will be compatible with the analyses); B.
+Evaluate the incomplete data; C.
+Develop the imputation model(s); D.
+Impute the missingness; E.
+Evaluate the imputations.
 
@@
 
+Add: Assumes exchangability between units.
+
 ### Imputation with the cluster variable as predictor (not recommended)
@@
 Colloquially, this is 'multilevel imputation for dummies'.
+
+Add: doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.
+
+The more the random effects are of interest, the more you need ml models.
 

### 2022-02-02 Manuscript/Manuscript.tex
Add some more refs to the intro

@@
 
-\hypertarget{multilevel-data}{%
-\subsection{Multilevel data}\label{multilevel-data}}
-
 In many contemporary data analysis efforts, some form of hierarchical or
-clustered structure is recorded.\footnote{Add IPDMA back in: patient
-  level data from multiple studies.} Ignoring such structures may be
-harmful to the inferences and can yield biased results. On the other
-hand, analyzing such multilevel data requires specialized techniques
-that take the clustered structure into account. Imagine a case where
-cross-level interactions between unit-level variables and cluster-level
-variables are present. The cluster to which a unit belongs may then
-influence the unit-level observations--and vice versa--for each of the
-units that make up the cluster \citep{hox17}. These relations can and
-should be taken into account when developing analysis models for
-multilevel data for the simple reason that groups of observations share
-some common variance.
+clustered structure is recorded. For example, students clustered within
+schools, or patients clustered within studies in individual patient data
+meta-analyses. Analyzing such multilevel data requires specialized
+techniques that take the clustered structure into account, since
+ignoring it may be harmful to the statistical inferences and can yield
+biased results. Imagine a case where cross-level interactions between
+unit-level variables and cluster-level variables are present. The
+cluster to which a unit belongs may then influence the unit-level
+observations--and vice versa for each of the units that make up the
+cluster \citep{hox17}. These relations can and should be taken into
+account when developing analysis models for multilevel data for the
+simple reason that groups of observations share some common variance.
 
@@
 relieve the restriction imposed by single-level models: equal group
-means across clusters. Additionally, there may be random predictor
-effects and/or random error terms (residual error variances), see e.g.
-\citet{hox17} and \citet{jong21}.\footnote{Add that heterogeneity refers
-  to variability within clusters.} There are many names for models that
-take clustering into account. Some popular examples are `multilevel
-models', `hierarchical models', `mixed effect models' and `random effect
-models'.
+means across clusters. Additionally, models may include random effects
+of predictors across clusters, and random error terms (heterogeneous
+residual error variances; see e.g. \citet{hox17} and \citet{jong21}).
+There are many names for models that take clustering into account. Some
+popular examples are `multilevel models', `hierarchical models', `mixed
+effect models' and `random effect models'.
 
-\hypertarget{missing-data}{%
-\subsection{Missing data}\label{missing-data}}
+\hypertarget{incomplete-multilevel-data}{%
+\subsection{Incomplete multilevel
+data}\label{incomplete-multilevel-data}}
 
@@
 The process of analyzing multilevel data is further complicated when not
-all observations are observed. Just as with single level data,
+all data entries are observed. Just as with single level data,
 missingness may occur at the unit level. But with multiple levels of
@@
 multilevel data can therefore be categorized into two general patterns:
-systematic missingness and sporadic missingness, see \citet{resc13}. We
-have visualized the difference between these types of missingness in
-Figure 1. The figure shows an \(n \times p\) set
-\(\mathbf{X} = X_1, \dots, X_p\), with \(n=5\) units and \(p=3\) columns
-distributed over 3 clusters. Column \texttt{X1} is completely observed.
-Column \texttt{X2} is systematically missing and column \texttt{X3} is
-sporadically missing.\footnote{Explain why. GV: not necessary}
+systematic missingness and sporadic missingness \citep{resc13}. We have
+visualized the difference between these types of missingness in Figure
+1. The figure shows an \(n \times p\) set
+\(\mathbf{X} = X_1, \dots, X_p\), with \(n\) units distributed over
+\(N\) clusters, and \(p=3\) columns. Column \texttt{X1} is completely
+observed. Column \texttt{X2} is systematically missing and column
+\texttt{X3} is sporadically missing. Systematic missingness can be
+further subdivided into unobserved constants (i.e., the same value
+within clusters) and non-measured random variables (which may differ per
+unit within clusters). In Figure 1, the former would imply that the
+unobserved values for units 3 and 4 on column \texttt{X2} are identical.
+With the latter, the values would differ. The optimal strategy for
+dealing with the missingness may therefore depend on the observed
+missing data pattern.
 
-Systematic missingness can be further subdivided into unobserved
-constants (i.e., the same value within clusters) and non-measured random
-variables (which may differ per unit within clusters). In Figure 1, the
-former would imply that the unobserved values for units 3 and 4 on
-column \texttt{X2} are identical. With the latter, the values would
-differ. The optimal strategy for dealing with the missingness may
-therefore depend on the observed missing data pattern. \footnote{add
-  missing data mechanisms here? GV: No, patterns are different from
-  mechanisms}
+Another characteristic of the missing data to take into account in
+analyses is the mechanism behind the missingness. Although the essence
+of the true non-response mechanism may not be known, it can be inferred
+or assumed to be one of the following:
+
+\begin{itemize}
+\tightlist
+\item
+  Missing Completely At Random (MCAR), where the probability to be
+  missing is equal across all data entries;
+\item
+  Missing At Random (MAR), where the probability to be missing depends
+  on observed information;
+\item
+  Missing Not At Random (MNAR), where the probability to be missing
+  depends on unrecorded information, making the missingness
+  non-ignorable \citep{rubi76, meng94}. Depending on the assumed
+  missingness mechanism, missing data handling strategies may be more or
+  less suitable, see e.g., \citet{yuce08} and \citet{hox15}.
+\end{itemize}
 
@@
 \citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
-the context of multilevel data, under varying missing data
-mechanisms.\footnote{Discuss missingness mechanisms before this point,
-  add references \citet{yuce08} and \citet{hox15}.}
+the context of multilevel data, under varying missing data mechanisms.
 
@@
 different missing data situations. For reasons of brevity, we focus on
-imputation by chained equations\footnote{add that JOMO is available in
-  \pkg{mice} as well?}. Other useful packages for incomplete multilevel
-data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.\footnote{Rephrase:
-  Some level of knowledge on multilevel models is assumed. We're
-  providing an overview of implementations. It's up-to the reader to
-  decide which multilevel strategy suits their data. So we won't go into
-  detail for the different methods (and equations). Refer to
-  \citet{meng94}, \citet{audi18}, and \citet{grun18}. This paper is just
-  a software tutorial. We'll keep it practical.}
+imputation by chained equations exclusively (although the alternative,
+joint modeling imputation for multilevel data or \pkg{jomo}
+\citet{jomo}, has been implemented in \pkg{mice} as well). Other useful
+resources for the analysis of incomplete multilevel data include the
+\proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and
+the empirical work by \citet{audi18} and \citet{grun18}. Note that this
+tutorial paper assumes a basic level of knowledge on multilevel models.
+We're providing an overview of implementations. It's up-to the reader to
+decide which multilevel strategy suits their data. We won't go into
+detail for the different methods (and equations). This paper is just a
+software tutorial, so we'll keep it practical.
 
-We structure this tutorial around three case studies:
+To illustrate how to impute incomplete multilevel data, we structure
+this tutorial around three case studies:
 
 \begin{itemize}
+\tightlist
 \item
-  \texttt{mice::popmis} (simulated data on school kids, n = 199 units, N
-  = 10 clusters, with MAR missingness in the outcome);
+  \texttt{mice::popmis} (simulated data on perceived popularity,
+  \(n = 199\) pupils across \(N = 10\) classes, with univariate MAR
+  missingness);
 \item
-  \texttt{GJRM::hiv} (simulated patient data on HIV, with MNAR
-  missingness in the outcome)
+  \texttt{GJRM::hiv} (simulated data on HIV diagnoses, \(n = 6,416\)
+  patients across \(N = 9\) regions, with univariate MNAR missingness)
 \item
-  \texttt{metamisc::impact} (real IPD on traumatic brain injuries,
-  without \texttt{NA}s);\footnote{add cross-level interaction?}
+  \texttt{metamisc::impact} (empirical data on traumatic brain injuries,
+  \(n = 11,022\) patients across \(N = 15\) studies, without
+  \texttt{NA}s).
 \end{itemize}
 
-For each case study we focus on a different aspect to illustrate how to
-impute incomplete multilevel data. In the \texttt{mice::popmis} data, we
-show the advantages of including the multilevel structure of the data
-into the imputation model. In the \texttt{metamisc::impact} data we'll
-show how to induce missingness and solve it in real-world data. In the
-\texttt{GJRM::hiv} we provide novel methodology\footnote{not really, the
-  methods exist already, but how to show that this is something new and
-  exciting?} for imputing MNAR missingness according to the Heckman
-model. For all case studies we discuss the nature of the incomplete
-data, the imputation model(s), and evaluation of the imputed data:
-
-A. Choose an analysis model (so the imputation model will be compatible
-with the analyses); B. Evaluate the incomplete data; C. Develop the
-imputation model(s); D. Impute the missingness; E. Evaluate the
-imputations.
+In each case study we highlight one aspect of the imputation workflow.
+With the \texttt{mice::popmis} data, we show different ways of designing
+an imputation model and what happens if you misspecify this model. With
+the \texttt{GJRM::hiv} data we extend the imputation model to include
+Heckman-type selection-inclusion methods. And with the
+\texttt{metamisc::impact} data we provide a example of multivariate
+missingness in real-world data. For all case studies we discuss the
+nature of the incomplete data, the imputation model(s), and evaluation
+of the imputed data: A. Choose an analysis model (so the imputation
+model will be compatible with the analyses); B. Evaluate the incomplete
+data; C. Develop the imputation model(s); D. Impute the missingness; E.
+Evaluate the imputations.
 
@@
 
+Add: Assumes exchangability between units.
+
 \hypertarget{imputation-with-the-cluster-variable-as-predictor-not-recommended}{%
@@
 `multilevel imputation for dummies'.
+
+Add: doesn't work with syst missing (only sporadically). There's some
+pro's and con's. May not differ much if the number of clusters is low.
+
+The more the random effects are of interest, the more you need ml
+models.
 

### 2022-02-02 R/plot_imputed.R
New plotting function for imputed values to split by cluster

@@ -115,3 +115,18 @@ plot_box <- function(imp, x, cluster = NULL, strip = FALSE) {
     # output
   return(p)
 }
+
+# new plotting function for imputed data
+vrb = "popular"
+# y = "popular"
+obs <- do.call(rbind, replicate(imp$m, !imp$where, simplify=FALSE)) %>%
+  rbind(matrix(FALSE, nrow(imp$data), ncol(imp$data)), .) %>% 
+  cbind(.imp = FALSE, .id = FALSE, .)
+comp <- imp %>% 
+  mice::complete("long", include = TRUE) 
+comp[obs] <- NA
+
+comp %>% ggplot(aes(x = .imp, y = get(vrb))) +
+  geom_point() +
+  labs(x = "Imputation",
+       y = vrb)

### 2022-02-02 R/plot_imputed.R
Make plotting function for imputations work for densityplot, bwplot and stripplot

@@ -117,16 +117,51 @@ plot_box <- function(imp, x, cluster = NULL, strip = FALSE) {
 }
 
 # new plotting function for imputed data
-vrb = "popular"
-# y = "popular"
-obs <- do.call(rbind, replicate(imp$m, !imp$where, simplify=FALSE)) %>%
-  rbind(matrix(FALSE, nrow(imp$data), ncol(imp$data)), .) %>% 
-  cbind(.imp = FALSE, .id = FALSE, .)
-comp <- imp %>% 
-  mice::complete("long", include = TRUE) 
-comp[obs] <- NA
+plot_mids <- function(imp, vrb, type = c("bwplot", "stripplot", "densityplot")) {
+  
+  # process mids object for plotting
+  obs <- do.call(rbind, replicate(imp$m, !imp$where, simplify = FALSE)) %>%
+    rbind(matrix(FALSE, nrow(imp$data), ncol(imp$data)), .) %>% 
+    cbind(.imp = FALSE, .id = FALSE, .) 
+  comp <- mice::complete(imp, "long", include = TRUE) %>% 
+    mutate(.imp = factor(.imp, ordered = TRUE))
+  comp[obs] <- NA
+  
+  # bwplot or stripplot
+  if (type == "stripplot" | type == "bwplot"){
+    p <- ggplot(comp, aes(x = .imp, y = get(vrb), color = ifelse(.imp < 1, "Observed", "Imputed"))) +
+    labs(x = "Imputation",
+         y = vrb,
+         color = "",
+         fill = "") 
+    if(type == "stripplot"){
+      p <- p + geom_point(position = position_jitter(width = 0.25, height = 0))
+    }
+    p <- p + geom_boxplot(size = 1,
+                 width = 0.5,
+                 alpha = 0.5,
+                 outlier.shape = NA)
+  }
+  
+  # densityplot
+  if (type == "densityplot"){
+  p <- ggplot(comp, aes(x = get(vrb), group = .imp, color = ifelse(.imp < 1, "Observed", "Imputed"))) +
+    geom_density() +
+    labs(x = vrb,
+         color = "",
+         fill = "")
+  }
+  
+  # # xyplot
+  # ggplot(comp, aes(x = extrav, y = popular, color = ifelse(.imp < 1, "Observed", "Imputed"))) +
+  #   geom_point() +
+  #   geom_point(aes())
+  #   labs(color = "")
+  
+  # output
[…]

### 2022-02-02 R/plot_imputed.R
Add ggplot version of mice::xyplot()

@@ -117,19 +117,24 @@ plot_box <- function(imp, x, cluster = NULL, strip = FALSE) {
 }
 
 # new plotting function for imputed data
-plot_mids <- function(imp, vrb, type = c("bwplot", "stripplot", "densityplot")) {
+plot_mids <- function(imp, x, y = NULL, type = c("bwplot", "stripplot", "densityplot")) {
   
   # process mids object for plotting
-  obs <- do.call(rbind, replicate(imp$m, !imp$where, simplify = FALSE)) %>%
-    rbind(matrix(FALSE, nrow(imp$data), ncol(imp$data)), .) %>% 
-    cbind(.imp = FALSE, .id = FALSE, .) 
-  comp <- mice::complete(imp, "long", include = TRUE) %>% 
+  if(is.null(y)){y <- x}
+  comp <- mice::complete(imp, "long")[is.na(imp$data[, x]) | is.na(imp$data[, y]), ] %>% 
+    rbind(cbind(.imp = 0, .id = rownames(imp$data), imp$data), .) %>% 
     mutate(.imp = factor(.imp, ordered = TRUE))
-  comp[obs] <- NA
+  
+  # obs <- do.call(rbind, replicate(imp$m, !imp$where, simplify = FALSE)) %>%
+  #   rbind(matrix(FALSE, nrow(imp$data), ncol(imp$data)), .) %>% 
+  #   cbind(.imp = FALSE, .id = FALSE, .) 
+  # comp <- mice::complete(imp, "long", include = TRUE) %>% 
+  #   mutate(.imp = factor(.imp, ordered = TRUE))
+  # comp[obs] <- NA
   
   # bwplot or stripplot
   if (type == "stripplot" | type == "bwplot"){
-    p <- ggplot(comp, aes(x = .imp, y = get(vrb), color = ifelse(.imp < 1, "Observed", "Imputed"))) +
+    p <- ggplot(comp, aes(x = .imp, y = get(vrb), group = .imp, color = ifelse(.imp < 1, "Observed", "Imputed"))) +
     labs(x = "Imputation",
          y = vrb,
          color = "",
@@ -153,15 +158,23 @@ plot_mids <- function(imp, vrb, type = c("bwplot", "stripplot", "densityplot")) 
   }
   
   # # xyplot
-  # ggplot(comp, aes(x = extrav, y = popular, color = ifelse(.imp < 1, "Observed", "Imputed"))) +
-  #   geom_point() +
[…]

### 2022-02-04 R/plot_imputed.R
Add convergence plot function, but should actually just be included in {mice}

@@ -1,137 +1,129 @@
-# plotting functions for imputed data
-
-# basic pre-processing of imputations
-plot_imps <- function(imp, x, y = NULL) {
-  # parse inputs
-  if (is.null(y)) {
-    y <- x
-  }
-  # combine observed and imputed data
-  xy_obs <- imp$data %>%
-    cbind(datapoint = "observed",
-          .imp = 0,
-          .id = 1:nrow(.),
-          .) %>%
-    .[!is.na(imp$data[[x]]) &
-        !is.na(imp$data[[y]]),]
-  xy_imps <- imp %>%
-    mice::complete("long") %>%
-    cbind(datapoint = "imputed", .) %>%
-    .[.$.id %nin% xy_obs$.id, ]
-  xy_dat <- rbind(xy_obs, xy_imps) %>%
-    dplyr::mutate(datapoint = factor(datapoint, levels = c("observed", "imputed")))
-  # initialize plot
-  p <- xy_dat %>%
-    ggplot2::ggplot() +
-    ggplot2::theme_classic() +
-    ggplot2::theme(legend.position = "top") +
-    ggplot2::scale_color_manual(
-      values = c(
-      "observed" = mice:::mdc(1),
-      # "missing" = mice:::mdc(2),
-      "imputed" = mice:::mdc(2)
-      )
-  ) +
-    ggplot2::scale_fill_manual(
-    values = c(
-      "observed" = mice:::mdc(1),
-      # "missing" = mice:::mdc(2),
-      "imputed" = mice:::mdc(2)
-    ))
-  # output
-  return(p)
-}
-
-# stripplot
-plot_strip <- function(imp, x) {
-  # plot individual values (stripplot)
-  p <- imp %>% plot_imps(x) +
-    ggplot2::geom_jitter(
-      ggplot2::aes(
-        y = factor(.imp, ordered = TRUE),
-        x = .data[[x]],
-        color = datapoint
-      ),
-      height = 0.25,
-      width = 0
-    ) +
-    scale_y_discrete(limits = rev) +
-    ggplot2::ylab("Imputation (0 = observed data)")
-  # output
-  return(p)
-}
-
-# boxplot
-plot_box <- function(imp, x, cluster = NULL, strip = FALSE) {
-  # plot
-  if (strip){
-    p <- imp %>% plot_imps(x) +
-      ggplot2::geom_jitter(
-        ggplot2::aes(
-          y = factor(.imp, ordered = TRUE),
-          x = .data[[x]],
[…]

### 2022-02-04 Manuscript/Manuscript.Rmd
Add popularity data v1 to conform to popmis data from mice

@@
 # randomly select a few classes
-clusters <- sample(1:100, 10)
+# clusters <- sample(1:100, 10)
 
 # load data 
-pop <- readRDS("../Data/popNCR2.RDS") %>% 
-  filter(class %in% clusters) %>% 
-  mutate(class = as.factor(class))
-
-# ICCs
-icc_pop <- icc(popular ~ class, data = pop)
-icc_teach <- icc(popteach ~ class, data = pop)
+popmis <- mice::popmis %>% mutate(
+  class = factor(school), 
+  .keep = "unused")#readRDS("../Data/popNCR2.RDS") %>% 
+  #filter(class %in% clusters) %>% 
+  #mutate(class = as.factor(class))
+popcomp <- foreign::read.spss("../Data/popular2.sav", to.data.frame = TRUE)
 ```
 
-`popNCR2` is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:
+In this section we'll go over the different steps involved with imputing incomplete multilevel data.
+The data we're using is the `popmis` dataset from the `mice` package.
+This is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:
 
@@
   - `popular`	Pupil popularity,
-  - `popteach`	Teacher popularity.
+  - `teachpop`	Teacher popularity.
+
+The analysis model corresponding to this dataset is multilevel regression with `popular` as the outcome variable, `sex` and `extrav` as unit-level predictors, and `texp` and `teachpop` as cluster-level predictors.
+Since the data is simulated and the missingness is induced, we can compare our imputations to the true data generating mechanism.
+
+The data is created in such a way that the clustering variable `class` explains quite some variance in the outcome variable `popular`.
+We express this using the intra-class correlation,  ICC $=$ `r round(icc(popular ~ as.factor(class), data = popcomp), 2)`.
+This tells us that it would be wise to use multilevel modeling.
 
@@
 # missingness
-plot_md_pat(pop)
+plot_md_pat(popmis)
 # TODO: add pattern by cluster, text mining example square size
@@
 
-The popularity data is created such that there are strong relations between the incomplete variables and the clustering variable `class`.
-We can express this using the intra-class correlation (ICC).
-For `popular` the ICC is `r round(icc_pop, 2)`.
-For `popteach` it is `r round(icc_teach, 2)`.
-It would thus be wise to use multilevel modeling.
+The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}.
+The ICC in the incomplete data is `r round(icc(popular ~ as.factor(class), data = na.omit(popmis)), 2)`.
 
-The missingness in this dataset is induced conform MAR and MNAR mechanisms.
-The missing data pattern, Figure \ref{fig:pop_pat}, shows that just one variable is incomplete **[the next part is not yet updated to reflect this]**.
+To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables.
+We'll highlight only one other variable to illustrate, but ideally one would inspect all relations.
+The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`teachpop`)?'.
+This can be evaluated statistically, but visual inspection usually suffices.
+We'll make a histogram of `teachpop` separately for the pupils with known popularity and missing popularity.
 
-To develop the best imputation model, we need to know whether the missingness in one variable depends on the observed values of other variables.
-Visual inspection usually suffices.
-We'll highlight only two variables to illustrate, but ideally one would inspect all relations.
-The questions we'll ask are: 'Does the missing data of `popular` depend on `popteach`?' and 'Does the missingness in teacher popularity depend on pupil popularity?' We'll evaluate this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity, and the other way around.
-
-In Figure \ref{fig:pop_dist} we see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
+In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
 This would indicate a right-tailed MAR missingness.
-In fact, this is exactly what happens, because the missingness in these data was created manually.
-Now, we've made it observable by examining the relations between the missingness in popular and the observed data in `popteach`.
-There is also a dependency between the missingness in teacher popularity and pupil popularity.
-The relation seems to be right-tailed as well.
+(In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `teachpop`.
+<!-- There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well. -->
 
 ```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-# plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
+# popmis %>% 
+#   filter(as.numeric(class) < 5) %>% 
+#   ggplot(aes(x = teachpop, color = class)) +
+#     geom_density() +
+#     facet_wrap(~factor(is.na(popular), labels = c("popular missing", "popular observed")), ncol = 1) +
+#     # scale_color_manual(values = plot_col, name = "popular") +
+#     theme_classic() # +
+#     # theme(legend.position = "bottom")
+
+plot_conditional(popmis, x = "teachpop", z = "popular", cluster = "class")
 # TODO: check if up-side-down plot works
@@
 # make densities proportional
-# 
+
 ```
@@
 # dry run to get imputation parameters
-ini <- mice(pop, maxit = 0)
+ini <- mice(popmis, maxit = 0)
 
@@
 # impute the data, ignoring the cluster structure
-imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
+imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
 ```
@@
 # # compare descriptives before and after imputation
-# psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+# psych::describe(popmis)[, c("n", "mean", "median", "min", "max", "sd")]
 # psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
@@
 # further inspection of the imputations
-plot_box(imp_ignored, x = "popular", strip = TRUE)
+plot_imps(imp_ignored, type = "stripplot", x = "popular")
 # TODO: think about multimodality
@@
 ICCs <- data.frame(
-  vars = c("popular", "popteach", "texp"), 
-  CCA = c(icc(popular ~ class, pop), 
-               icc(popteach ~ class, pop),
-               icc(texp ~ class, pop)), 
+  vars = c("popular", "teachpop", "texp"), 
+  CCA = c(icc(popular ~ class, popmis), 
+               icc(teachpop ~ class, popmis),
+               icc(texp ~ class, popmis)), 
   ignored = c(icc(popular ~ class, complete(imp_ignored)), 
-              icc(popteach ~ class, complete(imp_ignored)), 
+              icc(teachpop ~ class, complete(imp_ignored)), 
               icc(texp ~ class, complete(imp_ignored)))
@@
 # impute the data, cluster as predictor
-imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)
+imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
 
@@
 # head(imp_predictor$loggedEvents)
-## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
+## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and teachpop.
 
@@
 # # compare descriptives before and after imputation
-# psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
+# psych::describe(popmis)[, c("n", "mean", "median", "min", "max", "sd")]
 # psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
@@
 # further inspection of the imputations
-plot_box(imp_predictor, x = "popular", strip = TRUE)
+plot_imps(imp_predictor, x = "popular", type = "stripplot")
 
@@
            predictor = c(icc(popular ~ class, complete(imp_predictor)), 
-                        icc(popteach ~ class, complete(imp_predictor)), 
+                        icc(teachpop ~ class, complete(imp_predictor)), 
                         icc(texp ~ class, complete(imp_predictor)))
@@
   mice(
-    pop %>% mutate(class = as.integer(class)),
+    popmis %>% mutate(class = as.integer(class)),
     pred = pred,
@@
 # plot(imp_norm)
-plot_box(imp_norm_2l, x = "popular", strip = TRUE)
+plot_imps(imp_norm_2l, x = "popular", type = "stripplot")
 
@@
            norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
-                    icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
+                    icc(teachpop ~ as.factor(class), complete(imp_norm_2l)), 
                     icc(texp ~ as.factor(class), complete(imp_norm_2l)))
@@
 # plot(imp_pan)
-plot_box(imp_pan_2l, x = "popular", strip = TRUE)
+plot_imps(imp_pan_2l, x = "popular", type = "stripplot")
 ICCs <- ICCs %>% mutate(
            pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
-                    icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
+                    icc(teachpop ~ as.factor(class), complete(imp_pan_2l)), 
                     icc(texp ~ as.factor(class), complete(imp_pan_2l)))

### 2022-02-04 Manuscript/Manuscript.tex
Add popularity data v1 to conform to popmis data from mice

@@
 
-\texttt{popNCR2} is a simulated dataset with pupils clustered in
-classes, where the number of units \(n = 2000\), and the number of
-clusters \(N = 100\), on 7 variables:
+In this section we'll go over the different steps involved with imputing
+incomplete multilevel data. The data we're using is the \texttt{popmis}
+dataset from the \texttt{mice} package. This is a simulated dataset with
+pupils clustered in classes, where the number of units \(n = 2000\), and
+the number of clusters \(N = 100\), on 7 variables:
 
@@
 \item
-  \texttt{popteach} Teacher popularity.
+  \texttt{teachpop} Teacher popularity.
 \end{itemize}
+
+The analysis model corresponding to this dataset is multilevel
+regression with \texttt{popular} as the outcome variable, \texttt{sex}
+and \texttt{extrav} as unit-level predictors, and \texttt{texp} and
+\texttt{teachpop} as cluster-level predictors. Since the data is
+simulated and the missingness is induced, we can compare our imputations
+to the true data generating mechanism.
+
+The data is created in such a way that the clustering variable
+\texttt{class} explains quite some variance in the outcome variable
+\texttt{popular}. We express this using the intra-class correlation, ICC
+\(=\) 0.36. This tells us that it would be wise to use multilevel
+modeling.
 
@@
 
-The popularity data is created such that there are strong relations
-between the incomplete variables and the clustering variable
-\texttt{class}. We can express this using the intra-class correlation
-(ICC). For \texttt{popular} the ICC is 0.4. For \texttt{popteach} it is
-0.36. It would thus be wise to use multilevel modeling.
+The missingness is univariate and sporadic, which is illustrated in the
+missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the
+incomplete data is 0.56.
 
-The missingness in this dataset is induced conform MAR and MNAR
-mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows
-that just one variable is incomplete \textbf{{[}the next part is not yet
-updated to reflect this{]}}.
+To develop the best imputation model for the incomplete variable
+\texttt{popular}, we need to know whether the missingness depends on the
+observed values of other variables. We'll highlight only one other
+variable to illustrate, but ideally one would inspect all relations. The
+questions we'll ask are: `Does the missing data of pupil popularity
+(\texttt{popular}) depend on observed teacher popularity
+(\texttt{teachpop})?'. This can be evaluated statistically, but visual
+inspection usually suffices. We'll make a histogram of \texttt{teachpop}
+separately for the pupils with known popularity and missing popularity.
 
-To develop the best imputation model, we need to know whether the
-missingness in one variable depends on the observed values of other
-variables. Visual inspection usually suffices. We'll highlight only two
-variables to illustrate, but ideally one would inspect all relations.
-The questions we'll ask are: `Does the missing data of \texttt{popular}
-depend on \texttt{popteach}?' and `Does the missingness in teacher
-popularity depend on pupil popularity?' We'll evaluate this by making a
-histogram of \texttt{popteach} separately for the pupils with known
-popularity and missing popularity, and the other way around.
+In Figure \ref{fig:pop_dist} we see that \textbf{{[}update this part{]}}
+the distribution for the missing \texttt{popular} is further to the
+right than the distribution for observed \texttt{popular}. This would
+indicate a right-tailed MAR missingness. (In fact, this is exactly what
+happens, because the missingness in these data was created manually.)
+We've made it observable by examining the relations between the
+missingness in popular and the observed data in \texttt{teachpop}.
 
-In Figure \ref{fig:pop_dist} we see that the distribution for the
-missing \texttt{popular} is further to the right than the distribution
-for observed \texttt{popular}. This would indicate a right-tailed MAR
-missingness. In fact, this is exactly what happens, because the
-missingness in these data was created manually. Now, we've made it
-observable by examining the relations between the missingness in popular
-and the observed data in \texttt{popteach}. There is also a dependency
-between the missingness in teacher popularity and pupil popularity. The
-relation seems to be right-tailed as well.
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-1}
+
+}
+
+\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist}
+\end{figure}
+\end{CodeChunk}
 
@@
 R> # dry run to get imputation parameters
-R> ini <- mice(pop, maxit = 0)
+R> ini <- mice(popmis, maxit = 0)
 R> 
@@
 R> # impute the data, ignoring the cluster structure
-R> imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
+R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
 \end{CodeInput}
@@
       vars       CCA   ignored
-1  popular 0.3989463 0.4235643
-2 popteach 0.3605902 0.3605902
+1  popular 0.5583543 0.4739527
+2 teachpop 0.6823380 0.6823380
 3     texp 1.0000000 1.0000000
@@
 R> # impute the data, cluster as predictor
-R> imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)
+R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
 \end{CodeInput}
@@
       vars       CCA   ignored predictor
-1  popular 0.3989463 0.4235643 0.3904372
-2 popteach 0.3605902 0.3605902 0.3605902
+1  popular 0.5583543 0.4739527 0.5660108
+2 teachpop 0.6823380 0.6823380 0.6823380
 3     texp 1.0000000 1.0000000 1.0000000
@@
 +   mice(
-+     pop %>% mutate(class = as.integer(class)),
++     popmis %>% mutate(class = as.integer(class)),
 +     pred = pred,
@@
 \end{CodeInput}
+\begin{CodeOutput}
+Warning: Number of logged events: 1
+\end{CodeOutput}
 \end{CodeChunk}
@@
 \begin{CodeChunk}
+\begin{CodeOutput}
+Warning: Removed 5088 rows containing non-finite values (stat_boxplot).
+\end{CodeOutput}
+\begin{CodeOutput}
+Warning: Removed 5088 rows containing missing values (geom_point).
+\end{CodeOutput}
 
@@
       vars       CCA   ignored predictor      norm
-1  popular 0.3989463 0.4235643 0.3904372 0.4165006
-2 popteach 0.3605902 0.3605902 0.3605902 0.3605902
+1  popular 0.5583543 0.4739527 0.5660108 0.5583543
+2 teachpop 0.6823380 0.6823380 0.6823380 0.6823380
 3     texp 1.0000000 1.0000000 1.0000000 1.0000000
@@
       vars       CCA   ignored predictor      norm       pan
-1  popular 0.3989463 0.4235643 0.3904372 0.4165006 0.3827595
-2 popteach 0.3605902 0.3605902 0.3605902 0.3605902 0.3605902
+1  popular 0.5583543 0.4739527 0.5660108 0.5583543 0.5583543
+2 teachpop 0.6823380 0.6823380 0.6823380 0.6823380 0.6823380
 3     texp 1.0000000 1.0000000 1.0000000 1.0000000 1.0000000

### 2022-02-04 R/load_popularity.R

@@ -1,15 +1,15 @@
 # generate incomplete data based on the popularity dataset
 
-# load complete data
-library(lme4)
-pop_complete <-
-  foreign::read.spss("../Data/popular2.sav", to.data.frame = TRUE)[, 2:7]
-
-mod <- lmer(popular ~ sex + extrav + texp + extrav * texp + (1 | class) + (1 | texp) +  (1 | extrav), pop_complete) %>% broom.mixed::tidy()
-
-# parameters
-N = 5
-n = 25
+# # load complete data
+# library(lme4)
+# pop_complete <-
+#   foreign::read.spss("../Data/popular2.sav", to.data.frame = TRUE)[, 2:7]
+# 
+# mod <- lmer(popular ~ sex + extrav + texp + extrav * texp + (1 | class) + (1 | texp) +  (1 | extrav), pop_complete) %>% broom.mixed::tidy()
+# 
+# # parameters
+# N = 5
+# n = 25
 
 # # synthesize some more observations
 # icc(popular ~as.factor(class), pop_complete)

### 2022-02-04 R/plot_imputed.R

@@ -117,53 +117,60 @@
 # }
 
 # new plotting function for imputed data
-plot_imps <- function(imp, x, y = NULL, type = c("bwplot", "stripplot", "densityplot")) {
-  # process mids object for plotting
+plot_imps <- function(imp, type = c("bwplot", "stripplot", "densityplot"), x, y = NULL) {
+  # pre-process mids object for plotting
   if(is.null(y)){y <- x}
   comp <- mice::complete(imp, "long")[is.na(imp$data[, x]) | is.na(imp$data[, y]), ] %>% 
     rbind(cbind(.imp = 0, .id = rownames(imp$data), imp$data), .) %>% 
     mutate(.imp = factor(.imp, ordered = TRUE))
-
+  
+  # basic plot object
+  p <- ggplot2::ggplot(comp) +
+    ggplot2::theme_classic() +
+    ggplot2::scale_color_manual(
+      values = c(mice:::mdc(2), mice:::mdc(1)),
+      labels = c("Imputed", "Observed"))
+  
   # bwplot or stripplot
+  if (type == "stripplot") {
+    p <- p + geom_point(
+      aes(x = .imp, y = get(x), group = .imp, color = ifelse(.imp < 1, "Observed", "Imputed")),
+      position = position_jitter(width = 0.25, height = 0),
+      data = p$data) 
+  }  
   if (type == "stripplot" | type == "bwplot"){
-    p <- ggplot(comp, aes(x = .imp, y = get(vrb), group = .imp, color = ifelse(.imp < 1, "Observed", "Imputed"))) +
-    labs(x = "Imputation",
-         y = vrb,
-         color = "",
-         fill = "") 
-    if(type == "stripplot"){
-      p <- p + geom_point(position = position_jitter(width = 0.25, height = 0))
-    }
-    p <- p + geom_boxplot(size = 1,
-                 width = 0.5,
-                 alpha = 0.5,
-                 outlier.shape = NA)
+    p <- p + geom_boxplot(
+      aes(x = .imp, y = get(x), group = .imp, color = ifelse(.imp < 1, "Observed", "Imputed")),
+      size = 1,
+      width = 0.5,
+      alpha = 0.5,
+      outlier.shape = NA,
+      data = p$data) + 
+      labs(x = "Imputation",
+           y = x,
+           color = "",
+           fill = "") 
   }
   
   # densityplot
   if (type == "densityplot"){
[…]

### 2022-02-04 Manuscript/Manuscript.Rmd
Add analysis model popularity data

@@
 
-The variability due to clustering is often measured by means of the intra-class coefficient (ICC).
+The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
 The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.
@@
 This paper is just a software tutorial, so we'll keep it practical.
+Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the `magrittr` package and avidly used among `tidy` workflows, and the `lme4` notation for multilevel models.
 
@@
 
-- `mice::popmis` (simulated data on perceived popularity, $n = 199$ pupils across $N = 10$ classes, with univariate MAR missingness);
+- `mice::popmis` (simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, with univariate MAR missingness);
 - `GJRM::hiv` (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, with univariate MNAR missingness)
@@
 # load data 
-popmis <- mice::popmis %>% mutate(
-  class = factor(school), 
-  .keep = "unused")#readRDS("../Data/popNCR2.RDS") %>% 
+# popmis <- mice::popmis 
+#readRDS("../Data/popNCR2.RDS") %>% 
   #filter(class %in% clusters) %>% 
   #mutate(class = as.factor(class))
-popcomp <- foreign::read.spss("../Data/popular2.sav", to.data.frame = TRUE)
+popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE) %>% 
+  mutate(school = SCHOOL,
+         popular = as.numeric(POPULAR),
+         sex = SEX,
+         texp = TEXP,
+         .keep = "none")
+comp_est <- lme4::lmer(popular ~ 1 + sex + texp + (1 + sex | school), popcomp) %>% broom.mixed::tidy()
 ```
@@
 The data we're using is the `popmis` dataset from the `mice` package.
-This is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:
+This is a simulated dataset with pupils ($n = 2000$) clustered within schools ($N = 100$).
+In this tutorial we'll use the following variables:
 
-  - `pupil`	Pupil number within class,
-  - `class`	Class number,
-  - `extrav`	Pupil extraversion,
-  - `sex`	Pupil gender,
-  - `texp`	Teacher experience (years),
-  - `popular`	Pupil popularity,
-  - `teachpop`	Teacher popularity.
+  - `school`,     school identification number (clustering variable);
+  - `popular`,    pupil popularity (self-rating between 0 and 10; unit-level);
+  - `sex`,	      pupil sex (0=boy, 1=girl; unit-level);
+  - `texp`,	      teacher experience (in years; cluster-level).
 
-The analysis model corresponding to this dataset is multilevel regression with `popular` as the outcome variable, `sex` and `extrav` as unit-level predictors, and `texp` and `teachpop` as cluster-level predictors.
-Since the data is simulated and the missingness is induced, we can compare our imputations to the true data generating mechanism.
+The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction.
+The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`.
+The regression equation and `lme4` notation for this model are
 
-The data is created in such a way that the clustering variable `class` explains quite some variance in the outcome variable `popular`.
-We express this using the intra-class correlation,  ICC $=$ `r round(icc(popular ~ as.factor(class), data = popcomp), 2)`.
-This tells us that it would be wise to use multilevel modeling.
+$$
+\text{popular}_{ij} =
+\gamma_{00} + 
+\gamma_{10} \text{ sex}_{ij} + 
+\gamma_{01} \text{ texp}_{j} + 
+\gamma_{11} \text{ texp}_{j} \times \text{sex}_{ij} + 
+u_{0j} + 
+u_{1j} \text{ sex}_{ij} + 
+e_{ij} \\
+$$
+
+$$
+\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
+$$
+
+<!-- Intercept-only model: lmer(formula = popular ~ 1 + (1|class)) -->
+<!-- First level predictors: lmer(formula = popular ~ 1 + sex + (1|class)) -->
+<!-- First and second level predictors: lmer(popular ~ 1 + sex + texp + (1 | class)) -->
+<!-- First and second level predictors with random slopes: lmer(formula = popular ~ 1 + sex + texp + (1 + sex | class) -->
+<!-- First and second level predictors with random slopes and cross-level interaction: lmer(formula = popular ~ 1 + sex + texp + sex:texp + (1 + sex | class)) -->
+
+Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data.
+The data is created in such a way that the clustering variable `school` explains quite some variance in the outcome variable `popular`.
+We express this using the intraclass correlation,  ICC $=$ `r round(icc(popular ~ as.factor(school), data = popcomp), 2)`.
+We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effect of `sex` on `popular` $\gamma_{10} = 0.843$.
 
 ### Incomplete data
+
+Load the data into the environment and select the relevant variables:
+
+```{r}
+popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
+```
 
@@
 The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}.
-The ICC in the incomplete data is `r round(icc(popular ~ as.factor(class), data = na.omit(popmis)), 2)`.
+The ICC in the incomplete data is `r round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.
 
@@
 We'll highlight only one other variable to illustrate, but ideally one would inspect all relations.
-The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`teachpop`)?'.
+The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'.
 This can be evaluated statistically, but visual inspection usually suffices.
-We'll make a histogram of `teachpop` separately for the pupils with known popularity and missing popularity.
+We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.
 
@@
 This would indicate a right-tailed MAR missingness.
-(In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `teachpop`.
+(In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `texp`.
 <!-- There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well. -->
@@
 # popmis %>% 
-#   filter(as.numeric(class) < 5) %>% 
-#   ggplot(aes(x = teachpop, color = class)) +
+#   filter(as.numeric(school) < 5) %>% 
+#   ggplot(aes(x = teachpop, color = school)) +
 #     geom_density() +
@@
 
-plot_conditional(popmis, x = "teachpop", z = "popular", cluster = "class")
+plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
 # TODO: check if up-side-down plot works
@@
 We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`.
-We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.
+We do not use the observation identifier `pupil` or cluster identifier `school` as predictors to impute other variables.
 
@@
 pred <- ini$pred
-pred[, c("class", "pupil")] <- 0
+pred[, c("school")] <- 0
 
@@
 
-# compare ICCs before and after imputation
-ICCs <- data.frame(
-  vars = c("popular", "teachpop", "texp"), 
-  CCA = c(icc(popular ~ class, popmis), 
-               icc(teachpop ~ class, popmis),
-               icc(texp ~ class, popmis)), 
-  ignored = c(icc(popular ~ class, complete(imp_ignored)), 
-              icc(teachpop ~ class, complete(imp_ignored)), 
-              icc(texp ~ class, complete(imp_ignored)))
-  )
-ICCs
+# # compare ICCs before and after imputation
+# ICCs <- data.frame(
+#   vars = c("popular", "teachpop", "texp"), 
+#   CCA = c(icc(popular ~ school, popmis), 
+#                icc(teachpop ~ school, popmis),
+#                icc(texp ~ school, popmis)), 
+#   ignored = c(icc(popular ~ school, complete(imp_ignored)), 
+#               icc(teachpop ~ school, complete(imp_ignored)), 
+#               icc(texp ~ school, complete(imp_ignored)))
+#   )
+# ICCs
 
@@
 
-As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `class`.
+As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `school`.
 This tells us that the multilevel structure of the data should be taken into account.
@@
 Higher values for `texp` have a larger probability to be missing.
-This may not a problem, however, if at least one pupil in each class has teacher experience recorded, we can deductively impute the correct (i.e.
-true) value for every pupil in the class.
+This may not a problem, however, if at least one pupil in each school has teacher experience recorded, we can deductively impute the correct (i.e.
+true) value for every pupil in the school.
 
@@
 
-We'll now use `class` as a predictor to impute all other variables.
+We'll now use `school` as a predictor to impute all other variables.
 This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
@@
 pred <- ini$pred 
-pred[, "pupil"] <- 0
+# pred[, "pupil"] <- 0
 
@@
 
-# compare ICCs before and after imputation
-ICCs <- ICCs %>% mutate(
-           predictor = c(icc(popular ~ class, complete(imp_predictor)), 
-                        icc(teachpop ~ class, complete(imp_predictor)), 
-                        icc(texp ~ class, complete(imp_predictor)))
-           )
-ICCs
+# # compare ICCs before and after imputation
+# ICCs <- ICCs %>% mutate(
+#            predictor = c(icc(popular ~ school, complete(imp_predictor)), 
+#                         icc(teachpop ~ school, complete(imp_predictor)), 
+#                         icc(texp ~ school, complete(imp_predictor)))
+#            )
+# ICCs
 ```
@@
 pred <- ini$pred
-pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
+pred["popular", ] <- c(-2, 2, 2, 2) 
 #-2 for the cluster variable, 2 for random effects
 meth <- ini$meth
-meth <- c("", "", "", "", "", "2l.norm", "")
+meth <- c("", "2l.norm", "",  "")
 imp_norm_2l <-
   mice(
-    popmis %>% mutate(class = as.integer(class)),
+    popmis %>% mutate(school = as.integer(school)),
     pred = pred,
@@
 
-ICCs <- ICCs %>% mutate(
-           norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
-                    icc(teachpop ~ as.factor(class), complete(imp_norm_2l)), 
-                    icc(texp ~ as.factor(class), complete(imp_norm_2l)))
-           )
-ICCs
+# ICCs <- ICCs %>% mutate(
+#            norm = c(icc(popular ~ as.factor(school), complete(imp_norm_2l)), 
+#                     icc(teachpop ~ as.factor(school), complete(imp_norm_2l)), 
+#                     icc(texp ~ as.factor(school), complete(imp_norm_2l)))
+#            )
+# ICCs
 ```
@@
 ```{r pop_pan}
-pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
-meth <- c("", "", "", "", "", "2l.pan", "")
+pred["popular", ] <- c(-2, 2, 1, 2)
+meth <- c("", "2l.pan", "", "")
 imp_pan_2l <-
   mice(
-    popmis %>% mutate(class = as.integer(class)),
+    popmis %>% mutate(school = as.integer(school)),
     pred = pred,
@@
 plot_imps(imp_pan_2l, x = "popular", type = "stripplot")
-ICCs <- ICCs %>% mutate(
-           pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
-                    icc(teachpop ~ as.factor(class), complete(imp_pan_2l)), 
-                    icc(texp ~ as.factor(class), complete(imp_pan_2l)))
-           )
-ICCs
+# ICCs <- ICCs %>% mutate(
+#            pan = c(icc(popular ~ as.factor(school), complete(imp_pan_2l)), 
+#                     icc(teachpop ~ as.factor(school), complete(imp_pan_2l)), 
+#                     icc(texp ~ as.factor(school), complete(imp_pan_2l)))
+#            )
+# ICCs
 ```

### 2022-02-04 Manuscript/Manuscript.tex
Add analysis model popularity data

@@
 The variability due to clustering is often measured by means of the
-intra-class coefficient (ICC). The ICC can be seen as the percentage of
+intraclass coefficient (ICC). The ICC can be seen as the percentage of
 variance that can be attributed to the cluster-level, where a high ICC
@@
 detail for the different methods (and equations). This paper is just a
-software tutorial, so we'll keep it practical.
+software tutorial, so we'll keep it practical. Assumed knowledge also
+includes the use of the `piping operator', \texttt{\%\textgreater{}\%},
+adopted from the \texttt{magrittr} package and avidly used among
+\texttt{tidy} workflows, and the \texttt{lme4} notation for multilevel
+models.
 
@@
   \texttt{mice::popmis} (simulated data on perceived popularity,
-  \(n = 199\) pupils across \(N = 10\) classes, with univariate MAR
+  \(n = 2,000\) pupils across \(N = 100\) schools, with univariate MAR
   missingness);
@@
 dataset from the \texttt{mice} package. This is a simulated dataset with
-pupils clustered in classes, where the number of units \(n = 2000\), and
-the number of clusters \(N = 100\), on 7 variables:
+pupils (\(n = 2000\)) clustered within schools (\(N = 100\)). In this
+tutorial we'll use the following variables:
 
@@
 \item
-  \texttt{pupil} Pupil number within class,
+  \texttt{school}, school identification number (clustering variable);
 \item
-  \texttt{class} Class number,
+  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
+  unit-level);
 \item
-  \texttt{extrav} Pupil extraversion,
+  \texttt{sex}, pupil sex (0=boy, 1=girl; unit-level);
 \item
-  \texttt{sex} Pupil gender,
-\item
-  \texttt{texp} Teacher experience (years),
-\item
-  \texttt{popular} Pupil popularity,
-\item
-  \texttt{teachpop} Teacher popularity.
+  \texttt{texp}, teacher experience (in years; cluster-level).
 \end{itemize}
@@
 The analysis model corresponding to this dataset is multilevel
-regression with \texttt{popular} as the outcome variable, \texttt{sex}
-and \texttt{extrav} as unit-level predictors, and \texttt{texp} and
-\texttt{teachpop} as cluster-level predictors. Since the data is
-simulated and the missingness is induced, we can compare our imputations
-to the true data generating mechanism.
+regression with random intercepts, random slopes and a cross-level
+interaction. The outcome variable is \texttt{popular}, which is
+predicted from the unit-level variable \texttt{sex} and the
+cluster-level variable \texttt{texp}. The regression equation and
+\texttt{lme4} notation for this model are
 
-The data is created in such a way that the clustering variable
-\texttt{class} explains quite some variance in the outcome variable
-\texttt{popular}. We express this using the intra-class correlation, ICC
-\(=\) 0.36. This tells us that it would be wise to use multilevel
-modeling.
+\[
+\text{popular}_{ij} =
+\gamma_{00} + 
+\gamma_{10} \text{ sex}_{ij} + 
+\gamma_{01} \text{ texp}_{j} + 
+\gamma_{11} \text{ texp}_{j} \times \text{sex}_{ij} + 
+u_{0j} + 
+u_{1j} \text{ sex}_{ij} + 
+e_{ij} \\
+\]
+
+\[
+\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
+\]
+
+Since the data is simulated and the missingness is induced, we can
+compare our inferences after imputation to the true complete data. The
+data is created in such a way that the clustering variable
+\texttt{school} explains quite some variance in the outcome variable
+\texttt{popular}. We express this using the intraclass correlation, ICC
+\(=\) 0.58. We'll evaluate the ICC after each missing data strategy, and
+compare the estimated fixed effect of \texttt{sex} on \texttt{popular}
+\(\gamma_{10} = 0.843\).
 
@@
 \subsubsection{Incomplete data}\label{incomplete-data}}
+
+Load the data into the environment and select the relevant variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 (\texttt{popular}) depend on observed teacher popularity
-(\texttt{teachpop})?'. This can be evaluated statistically, but visual
-inspection usually suffices. We'll make a histogram of \texttt{teachpop}
+(\texttt{texp})?'. This can be evaluated statistically, but visual
+inspection usually suffices. We'll make a histogram of \texttt{texp}
 separately for the pupils with known popularity and missing popularity.
@@
 We've made it observable by examining the relations between the
-missingness in popular and the observed data in \texttt{teachpop}.
+missingness in popular and the observed data in \texttt{texp}.
 
@@
 do not use the observation identifier \texttt{pupil} or cluster
-identifier \texttt{class} as predictors to impute other variables.
+identifier \texttt{school} as predictors to impute other variables.
 
@@
 R> pred <- ini$pred
-R> pred[, c("class", "pupil")] <- 0
+R> pred[, c("school")] <- 0
 R> 
@@
 
-\begin{CodeOutput}
-      vars       CCA   ignored
-1  popular 0.5583543 0.4739527
-2 teachpop 0.6823380 0.6823380
-3     texp 1.0000000 1.0000000
-\end{CodeOutput}
 \end{CodeChunk}
@@
 As the original ICCs show, 100\% of the variance in \texttt{texp} can be
-attributed to the clustering variable \texttt{class}. This tells us that
-the multilevel structure of the data should be taken into account. If we
-don't, we'll end up with incorrect imputations, biasing the effect of
-the clusters towards zero.
+attributed to the clustering variable \texttt{school}. This tells us
+that the multilevel structure of the data should be taken into account.
+If we don't, we'll end up with incorrect imputations, biasing the effect
+of the clusters towards zero.
 
@@
 values for \texttt{texp} have a larger probability to be missing. This
-may not a problem, however, if at least one pupil in each class has
+may not a problem, however, if at least one pupil in each school has
 teacher experience recorded, we can deductively impute the correct
-(i.e.~true) value for every pupil in the class.
+(i.e.~true) value for every pupil in the school.
 
@@
 
-We'll now use \texttt{class} as a predictor to impute all other
+We'll now use \texttt{school} as a predictor to impute all other
 variables. This is still not recommended practice, since it only works
@@
 R> pred <- ini$pred 
-R> pred[, "pupil"] <- 0
+R> # pred[, "pupil"] <- 0
 R> 
@@
 
-\begin{CodeOutput}
-      vars       CCA   ignored predictor
-1  popular 0.5583543 0.4739527 0.5660108
-2 teachpop 0.6823380 0.6823380 0.6823380
-3     texp 1.0000000 1.0000000 1.0000000
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> pred <- ini$pred
-R> pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
+R> pred["popular", ] <- c(-2, 2, 2, 2) 
 R> #-2 for the cluster variable, 2 for random effects
 R> meth <- ini$meth
-R> meth <- c("", "", "", "", "", "2l.norm", "")
+R> meth <- c("", "2l.norm", "",  "")
 R> imp_norm_2l <-
 +   mice(
-+     popmis %>% mutate(class = as.integer(class)),
++     popmis %>% mutate(school = as.integer(school)),
 +     pred = pred,
@@
 \end{CodeInput}
-\begin{CodeOutput}
-Warning: Number of logged events: 1
-\end{CodeOutput}
 \end{CodeChunk}
@@
 \begin{CodeOutput}
-Warning: Removed 5088 rows containing non-finite values (stat_boxplot).
+Warning: Removed 848 rows containing non-finite values (stat_boxplot).
 \end{CodeOutput}
 \begin{CodeOutput}
-Warning: Removed 5088 rows containing missing values (geom_point).
+Warning: Removed 848 rows containing missing values (geom_point).
 \end{CodeOutput}
@@
 
-\begin{CodeOutput}
-      vars       CCA   ignored predictor      norm
-1  popular 0.5583543 0.4739527 0.5660108 0.5583543
-2 teachpop 0.6823380 0.6823380 0.6823380 0.6823380
-3     texp 1.0000000 1.0000000 1.0000000 1.0000000
-\end{CodeOutput}
 \end{CodeChunk}
@@
 \begin{CodeInput}
-R> pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
-R> meth <- c("", "", "", "", "", "2l.pan", "")
+R> pred["popular", ] <- c(-2, 2, 1, 2)
+R> meth <- c("", "2l.pan", "", "")
 R> imp_pan_2l <-
 +   mice(
-+     popmis %>% mutate(class = as.integer(class)),
++     popmis %>% mutate(school = as.integer(school)),
 +     pred = pred,
@@
 
-\begin{CodeOutput}
-      vars       CCA   ignored predictor      norm       pan
-1  popular 0.5583543 0.4739527 0.5660108 0.5583543 0.5583543
-2 teachpop 0.6823380 0.6823380 0.6823380 0.6823380 0.6823380
-3     texp 1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
-\end{CodeOutput}
 \end{CodeChunk}

### 2022-02-07 Manuscript/Manuscript.Rmd
Feedback on analysis model notation and missing data pattern by cluster

@@
     url: https://hanneoberman.github.io/
-  - name: Johanna Munoz Avila
+  - name: Johanna Mu\~noz Avila
     affiliation: University Medical Center Utrecht \AND
@@
 
+TODO: add all estimimates, seed, and version nr., this also shows the workflow for pooling etc.
+
 Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data.
@@
 # TODO: try to make it a shepley plot
+
+# popcomp <- popcomp %>% 
+#   cbind(R = is.na(popmis$popular))
+# t.test(x = popcomp$popular[popcomp$R], y = popcomp$popular[!popcomp$R]) 
+# ggplot(popcomp, aes(y=popular, x = R)) +
+#   geom_boxplot()
+
 ```
@@
 
+Schafer book fortran workflows.
+
 <!-- "By default, 2l.pan includes the intercept as both a fixed and a random effect.
@@
 
+TODO: rotate labels, make missingness % gradient for clusters?? Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of ct on unfortunate outcomes.
+
+gradient would be the number of clusters for which this variable is missing
+
 # Discussion
@@
 
+- env dump in repo
+
 # References

### 2022-02-07 Manuscript/Manuscript.tex
Feedback on analysis model notation and missing data pattern by cluster

@@
 \author{
-Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
-Medical Center Utrecht \AND Valentijn de Jong\\University Medical
-Center Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
-Debray\\University Medical Center Utrecht
+Hanne Oberman\\Utrecht University \And Johanna Mu\textasciitilde noz
+Avila\\University Medical Center Utrecht \AND Valentijn de
+Jong\\University Medical Center Utrecht \And Gerko Vink\\Utrecht
+University \AND Thomas Debray\\University Medical Center Utrecht
 }
@@
 
-\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
-Jong, Gerko Vink, Thomas Debray}
+\Plainauthor{Hanne Oberman, Johanna Mu\textasciitilde noz
+Avila, Valentijn de Jong, Gerko Vink, Thomas Debray}
 \Plaintitle{Imputation of Incomplete Multilevel Data with mice}
@@
 
+TODO: add all estimimates, seed, and version nr., this also shows the
+workflow for pooling etc.
+
 Since the data is simulated and the missingness is induced, we can
@@
 
+Schafer book fortran workflows.
+
 \begin{CodeChunk}
@@
 
+TODO: rotate labels, make missingness \% gradient for clusters?? Look at
+analysis model, maybe copy from GREAT data example e.g., adjusted
+prognostic effect of ct on unfortunate outcomes.
+
+gradient would be the number of clusters for which this variable is
+missing
+
 \hypertarget{discussion}{%
@@
   doesn't hurt.
+\item
+  env dump in repo
 \end{itemize}

### 2022-02-07 R/plot_imputed.R

@@ -116,6 +116,9 @@
 #   return(p)
 # }
 
+# option 1: bwplot etc with argument
+# option 2: ggmice which overwrites the bw functions ect.
+
 # new plotting function for imputed data
 plot_imps <- function(imp, type = c("bwplot", "stripplot", "densityplot"), x, y = NULL) {
   # pre-process mids object for plotting

### 2022-02-08 Manuscript/Manuscript.Rmd
Add complete data estimates for popularity case study

@@
 `%nin%` <- Negate(`%in%`)
+round3 <- function(x){format(round(x, 3), nsmall = 3)}
 icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
@@
 
+<!-- Table 7.1: Questions to gauge the complexity of a multilevel imputation task. -->
+<!-- 1.	Will the complete-data model include random slopes? -->
+<!-- 2.	Will the data contain systematically missing values? -->
+<!-- 3.	Will the distribution of the residuals be non-normal? -->
+<!-- 4.	Will the error variance differ over clusters? -->
+<!-- 5.	Will there be small clusters? -->
+<!-- 6.	Will there be a small number of clusters? -->
+<!-- 7.	Will the complete-data model have cross-level interactions? -->
+<!-- 8.	Will the dataset be very large? (fimd, section 7.3)-->
+
 ## Aim of this paper
@@
   #mutate(class = as.factor(class))
-popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE) %>% 
+popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
   mutate(school = SCHOOL,
-         popular = as.numeric(POPULAR),
+         popular = POPULAR,
          sex = SEX,
@@
          .keep = "none")
-comp_est <- lme4::lmer(popular ~ 1 + sex + texp + (1 + sex | school), popcomp) %>% broom.mixed::tidy()
+est_comp <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popcomp) 
+# %>% 
+#   broom.mixed::tidy(conf.int = TRUE) 
+# %>% 
+  # mutate(
+  #   term = factor(term, levels = c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation"), labels = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor.
+cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error"), ordered = TRUE))
 ```
@@
 The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`.
-The regression equation and `lme4` notation for this model are
+The regression equation^[add the 'level notation' (Bryk and Raudenbush, 1992) and/or matrix notation ('linear mixed effects model'; Laird and Ware, 1982) too?] and `lme4` notation for this model are
 
@@
 We express this using the intraclass correlation,  ICC $=$ `r round(icc(popular ~ as.factor(school), data = popcomp), 2)`.
-We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effect of `sex` on `popular` $\gamma_{10} = 0.843$.
+We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effects:
+
+```{r echo=FALSE}
+# lvl <- c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation")
+# lbl = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor.
+cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error")
+est_comp %>% 
+  broom.mixed::tidy("fixed", conf.int = TRUE) %>% 
+  mutate(term = term,
+         estimate  = paste0(
+           round3(estimate), " [", 
+           round3(conf.low), ", ", 
+           round3(conf.high), "]"), 
+         .keep = "none")
+
+est_comp %>% 
+  broom.mixed::tidy("ran_pars") %>% 
+  mutate(term = term, #factor(term, levels = unique(.$term), labels = c()),
+         estimate  = round3(estimate), 
+         .keep = "none")
+```
 
@@
 But at least, it includes some multilevel aspect.
-This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende2016].
+This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16].
 Colloquially, this is 'multilevel imputation for dummies'.
@@
 
+"Van Buuren (2011) considered the homoscedastic linear mixed model as invalid for imputing incomplete predictors, and investigated only the 2l.norm method, which allows for heterogeneous error variances" [@buur18].
+
 ```{r pop_norm}
@@
 meth <- c("", "2l.norm", "",  "")
+# meth <- c("", "2l.pmm", "",  "")
 imp_norm_2l <-

### 2022-02-08 Manuscript/Manuscript.tex
Add complete data estimates for popularity case study

@@
 predicted from the unit-level variable \texttt{sex} and the
-cluster-level variable \texttt{texp}. The regression equation and
-\texttt{lme4} notation for this model are
+cluster-level variable \texttt{texp}. The regression equation\footnote{add
+  the `level notation' (Bryk and Raudenbush, 1992) and/or matrix
+  notation (`linear mixed effects model'; Laird and Ware, 1982) too?}
+and \texttt{lme4} notation for this model are
 
@@
 \(=\) 0.58. We'll evaluate the ICC after each missing data strategy, and
-compare the estimated fixed effect of \texttt{sex} on \texttt{popular}
-\(\gamma_{10} = 0.843\).
+compare the estimated fixed effects:
+
+\begin{CodeChunk}
+\begin{CodeOutput}
+# A tibble: 4 x 2
+  term        estimate                 
+  <chr>       <chr>                    
+1 (Intercept) " 3.314 [ 2.998,  3.629]"
+2 sex         " 1.330 [ 1.069,  1.590]"
+3 texp        " 0.110 [ 0.090,  0.130]"
+4 sex:texp    "-0.034 [-0.051, -0.017]"
+\end{CodeOutput}
+\begin{CodeOutput}
+# A tibble: 4 x 2
+  term                 estimate
+  <chr>                <chr>   
+1 sd__(Intercept)      0.642   
+2 cor__(Intercept).sex 0.077   
+3 sd__sex              0.476   
+4 sd__Observation      0.626   
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 N-1 indicator variables representing allocation of N clusters as a fixed
-factor in the model \citep{reit06, ende2016}. Colloquially, this is
+factor in the model \citep{reit06, ende16}. Colloquially, this is
 `multilevel imputation for dummies'.
@@
 
+``Van Buuren (2011) considered the homoscedastic linear mixed model as
+invalid for imputing incomplete predictors, and investigated only the
+2l.norm method, which allows for heterogeneous error variances''
+\citep{buur18}.
+
 \begin{CodeChunk}
@@
 R> meth <- c("", "2l.norm", "",  "")
+R> # meth <- c("", "2l.pmm", "",  "")
 R> imp_norm_2l <-

### 2022-02-11 Manuscript/Manuscript.Rmd
Clean up introduction section

@@
 abstract: >
-  Multilevel data is not spared the ubiquitous problem of missing information.
   This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
-Including methods for ignorable and non-ignorable missingness.
 Footnotes in the current version show work in progress/under construction.
@@
 plot_col <- mice:::mdc(1:2) %>% setNames(c("observed", "missing"))
+
+# data
+popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
+  mutate(school = SCHOOL,
+         popular = POPULAR,
+         sex = SEX,
+         texp = TEXP,
+         .keep = "none")
+popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
 ```
@@
 In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.
-For example, students clustered within schools, or patients clustered within studies in individual patient data meta-analyses.
-Analyzing such multilevel data requires specialized techniques that take the clustered structure into account, since ignoring it may be harmful to the statistical inferences and can yield biased results.
+For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses.
+Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and bias results.
 Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
@@
 These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
-
-The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
+The variability due to clustering is often measured by means of the intraclass coefficient [ICC, @fish25].
 The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.
-Multilevel models typically accommodate for this variability by including separate intercepts for each cluster.
-Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters.
-Additionally, models may include random effects of predictors across clusters, and random error terms (heterogeneous residual error variances; see e.g.
-@hox17 and @jong21).
+Multilevel models typically accommodate for this variability by including a separate group mean for each cluster.
+In addition to random intercepts, multilevel models can also include random effects and heterogeneous residual error variances across clusters [see e.g.
+@gelm06, @hox17 and @jong21].
 There are many names for models that take clustering into account.
@@
 
-## Incomplete multilevel data
+## Missingness in multilevel data
 
@@
 But with multiple levels of data comes the potential for clustered missingness.
-Missingness in multilevel data can therefore be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13].
-We have visualized the difference between these types of missingness in Figure 1.
-The figure shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns.
-Column `X1` is completely observed.
-Column `X2` is systematically missing and column `X3` is sporadically missing.
-Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters).
-In Figure 1, the former would imply that the unobserved values for units 3 and 4 on column `X2` are identical.
+Incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13].
+Systematic missingness implies that one or more variables are never observed in a certain cluster.
+With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18].
+We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns.
+Column $X1$ is completely observed, column $X2$ is systematically missing, and column $X3$ is sporadically missing.
+To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account.
+For example, the sporadic missingness in $X3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters).
+We could then just extrapolate the true (but missing) value of $X3$ for unit 1 from unit 2, and the value for unit 4 from unit 3.
+If $X3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but need some kind of missing data method or discard the incomplete units altogether (i.e., list-wise deletion/complete case analysis).
+And, with the systematic missingness in $X2$, it is impossible to fit a multilevel model since we cannot estimate the intercept of cluster 2.
+We would have to exclude this cluster from our analyses entirely to obtain any results.
+Obviously, excluding observations is not a desirable workflow.
+
+<!-- While the sporadic missingness in $X1$ would at least provide some information for each cluster.
+can be further subdivided depending on the underlying true value of the missing entries: this can either be  constant within clusters, or a non-measured random variable (which may differ per unit within clusters).
+In Figure 1, the former would imply that the unobserved values for units 3 and 4 on column $X2$ are identical.
 With the latter, the values would differ.
-The optimal strategy for dealing with the missingness may therefore depend on the observed missing data pattern.
-
-Another characteristic of the missing data to take into account in analyses is the mechanism behind the missingness.
-Although the essence of the true non-response mechanism may not be known, it can be inferred or assumed to be one of the following:
-
-- Missing Completely At Random (MCAR), where the probability to be missing is equal across all data entries;
-- Missing At Random (MAR), where the probability to be missing depends on observed information;
-- Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94].
-Depending on the assumed missingness mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.
+The pattern of the missing data across clusters determines the optimal strategy for dealing with the missingness.
+-->
 
 Ignoring the missingness in analyses can be extremely harmful to inferences.
-Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power.
+Complete case analysis can introduce bias in statistical inferences and lowers statistical power.
 Instead, the missingness should be accommodated \emph{before} or _within_ the analysis of scientific interest.
-Especially the former is very generic and popular and is widely known as imputation.
+Especially the former is very generic and popular.
 Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed.
@@
 \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
-In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.
+In this paper, we will discuss how to use \pkg{mice} in the context of multilevel data.
 
@@
 
+<!-- Another characteristic of the missing data to take into account in analyses is the mechanism behind the missingness.
+Although the essence of the true non-response mechanism may not be known, it can be inferred or assumed to be one of the following: -->
+
+<!-- - Missing Completely At Random (MCAR), where the probability to be missing is equal across all data entries; -->
+<!-- - Missing At Random (MAR), where the probability to be missing depends on observed information; -->
+<!-- - Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]. -->
+<!-- Depending on the assumed missingness mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15. -->
+
 ## Aim of this paper
@@
 This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}.
-We provide practical guidelines and code snippets for different missing data situations.
-For reasons of brevity, we focus on imputation by chained equations exclusively (although the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well).
-Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and the empirical work by @audi18 and @grun18.
-Note that this tutorial paper assumes a basic level of knowledge on multilevel models.
-We're providing an overview of implementations.
+We provide practical guidelines and code snippets for different missing data situations, including missing not at random (MNAR) mechanisms [where the probability to be missing depends on unrecorded information, making the missingness non-ignorable, @rubi76; @meng94].
+For reasons of brevity, we focus on imputation by chained equations wit \pkg{mice} exclusively^[Note that the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well but is outside the scope of this tutorial.].
+Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work by @audi18 and @grun18.
+Please note that this tutorial paper assumes a basic level of knowledge on multilevel models.^[Note to self: We're providing an overview of implementations.
 It's up-to the reader to decide which multilevel strategy suits their data.
 We won't go into detail for the different methods (and equations).
-This paper is just a software tutorial, so we'll keep it practical.
-Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the `magrittr` package and avidly used among `tidy` workflows, and the `lme4` notation for multilevel models.
+This paper is just a software tutorial, so we'll keep it practical.] Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the \pkg{magrittr} package, and the \pkg{lme4} notation for multilevel models.^[Add environment info, seed and version number(s) somewhere!]
 
-To illustrate how to impute incomplete multilevel data, we structure this tutorial around three case studies:
+We illustrate how to impute incomplete multilevel data by means of three case studies:
 
-- `mice::popmis` (simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, with univariate MAR missingness);
-- `GJRM::hiv` (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, with univariate MNAR missingness)
-- `metamisc::impact` (empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, without `NA`s).
+- `popmis` from the \pkg{mice} package (simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools);
+- `hiv` from the \pkg{GJRM} package (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions);
+- `impact` from the \pkg{metamisc} package (empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies).
 
-In each case study we highlight one aspect of the imputation workflow.
-With the `mice::popmis` data, we show different ways of designing an imputation model and what happens if you misspecify this model.
-With the `GJRM::hiv` data we extend the imputation model to include Heckman-type selection-inclusion methods.
-And with the `metamisc::impact` data we provide a example of multivariate missingness in real-world data.
-For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data: A.
-Choose an analysis model (so the imputation model will be compatible with the analyses); B.
-Evaluate the incomplete data; C.
-Develop the imputation model(s); D.
-Impute the missingness; E.
-Evaluate the imputations.
+For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific asect of the imputation workflow.
+With the `popmis` data, we show how (not) to develop an imputation model.
+With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods.
+And with the `impact` data we provide a example of multivariate missingness in real-world data.
+Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.
+
+---
+
+\begin{center}
+REVIEW UNTIL THIS POINT SVP
+\end{center}
+
+---
 
@@
   #mutate(class = as.factor(class))
-popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
-  mutate(school = SCHOOL,
-         popular = POPULAR,
-         sex = SEX,
-         texp = TEXP,
-         .keep = "none")
+# popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
+#   mutate(school = SCHOOL,
+#          popular = POPULAR,
+#          sex = SEX,
+#          texp = TEXP,
+#          .keep = "none")
 est_comp <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popcomp) 
@@
 
-TODO: add all estimimates, seed, and version nr., this also shows the workflow for pooling etc.
+<!-- TODO: add all estimimates, seed, and version nr., this also shows the workflow for pooling etc. -->
 
@@
 
-est_comp %>% 
-  broom.mixed::tidy("ran_pars") %>% 
-  mutate(term = term, #factor(term, levels = unique(.$term), labels = c()),
-         estimate  = round3(estimate), 
-         .keep = "none")
+# est_comp %>% 
+#   broom.mixed::tidy("ran_pars") %>% 
+#   mutate(term = term, #factor(term, levels = unique(.$term), labels = c()),
+#          estimate  = round3(estimate), 
+#          .keep = "none")
 ```
+
+<!--  Let's apply the following five methods: -->
+
+<!-- sample: Find imputations by random sampling from the observed values in lpo. This method ignores sch; -->
+
+<!-- pmm: Single-level predictive mean matching with the school indicator coded as a dummy variable; -->
+
+<!-- 2l.pan: Multilevel method using the linear mixed model to draw univariate imputations; -->
+
+<!-- 2l.norm: Multilevel method using the linear mixed model with heterogeneous error variances; -->
+
+<!-- 2l.pmm: Predictive mean matching based on predictions from the linear mixed model, with random draws from the regression coefficients and the random effects, using five donors. -->
 
@@
 
-TODO: rotate labels, make missingness % gradient for clusters?? Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of ct on unfortunate outcomes.
+TODO: Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of ct on unfortunate outcomes.
+Add something about systematically missing data here.
+In the md pattern plot: rotate labels, make missingness % gradient for clusters (ie the number of clusters for which this variable is missing).
 
-gradient would be the number of clusters for which this variable is missing
+<!-- Resche-Rigon and White (2018) proposed a two-stage estimator.
+At step 1, a linear regression model is fitted to each observed cluster.
+Any sporadically missing data are imputed, and the model per cluster ignores any systematically missing variables.
+At step 2, estimates obtained from each cluster are combined using meta-analysis.
+Systematically missing variables are modeled through a linear random effect model across clusters.
+-->
 

### 2022-02-11 Manuscript/Manuscript.tex
Clean up introduction section

@@
 \Abstract{
-Multilevel data is not spared the ubiquitous problem of missing
-information. This is a tutorial paper on imputing incomplete multilevel
-data with \pkg{mice}. Including methods for ignorable and non-ignorable
-missingness. Footnotes in the current version show work in
-progress/under construction. The last section is not part of the
-manuscript, but purely for reminders.
+This is a tutorial paper on imputing incomplete multilevel data with
+\pkg{mice}. Footnotes in the current version show work in progress/under
+construction. The last section is not part of the manuscript, but purely
+for reminders.
 }
@@
 In many contemporary data analysis efforts, some form of hierarchical or
-clustered structure is recorded. For example, students clustered within
-schools, or patients clustered within studies in individual patient data
-meta-analyses. Analyzing such multilevel data requires specialized
-techniques that take the clustered structure into account, since
-ignoring it may be harmful to the statistical inferences and can yield
-biased results. Imagine a case where cross-level interactions between
-unit-level variables and cluster-level variables are present. The
-cluster to which a unit belongs may then influence the unit-level
-observations--and vice versa for each of the units that make up the
-cluster \citep{hox17}. These relations can and should be taken into
-account when developing analysis models for multilevel data for the
+clustered structure is recorded. For example, students may be clustered
+in classes in psychometrics research, or patients may be clustered in
+studies in individual patient data meta-analyses. Ignoring the clustered
+structure of such multilevel data can be harmful to the statistical
+inferences and bias results. Imagine a case where cross-level
+interactions between unit-level variables and cluster-level variables
+are present. The cluster to which a unit belongs may then influence the
+unit-level observations--and vice versa for each of the units that make
+up the cluster \citep{hox17}. These relations can and should be taken
+into account when developing analysis models for multilevel data for the
 simple reason that groups of observations share some common variance.
+The variability due to clustering is often measured by means of the
+intraclass coefficient \citep[ICC,][]{fish25}. The ICC can be seen as
+the percentage of variance that can be attributed to the cluster-level,
+where a high ICC would indicate that a lot of variability is due to the
+cluster structure. Multilevel models typically accommodate for this
+variability by including a separate group mean for each cluster. In
+addition to random intercepts, multilevel models can also include random
+effects and heterogeneous residual error variances across clusters
+\citep[see e.g.][\citet{hox17} and \citet{jong21}]{gelm06}. There are
+many names for models that take clustering into account. Some popular
+examples are `multilevel models', `hierarchical models', `mixed effect
+models' and `random effect models'.
 
-The variability due to clustering is often measured by means of the
-intraclass coefficient (ICC). The ICC can be seen as the percentage of
-variance that can be attributed to the cluster-level, where a high ICC
-would indicate that a lot of variability is due to the cluster
-structure. Multilevel models typically accommodate for this variability
-by including separate intercepts for each cluster. Such fixed effects
-relieve the restriction imposed by single-level models: equal group
-means across clusters. Additionally, models may include random effects
-of predictors across clusters, and random error terms (heterogeneous
-residual error variances; see e.g. \citet{hox17} and \citet{jong21}).
-There are many names for models that take clustering into account. Some
-popular examples are `multilevel models', `hierarchical models', `mixed
-effect models' and `random effect models'.
-
-\hypertarget{incomplete-multilevel-data}{%
-\subsection{Incomplete multilevel
-data}\label{incomplete-multilevel-data}}
+\hypertarget{missingness-in-multilevel-data}{%
+\subsection{Missingness in multilevel
+data}\label{missingness-in-multilevel-data}}
 
@@
 missingness may occur at the unit level. But with multiple levels of
-data comes the potential for clustered missingness. Missingness in
-multilevel data can therefore be categorized into two general patterns:
-systematic missingness and sporadic missingness \citep{resc13}. We have
-visualized the difference between these types of missingness in Figure
-1. The figure shows an \(n \times p\) set
-\(\mathbf{X} = X_1, \dots, X_p\), with \(n\) units distributed over
-\(N\) clusters, and \(p=3\) columns. Column \texttt{X1} is completely
-observed. Column \texttt{X2} is systematically missing and column
-\texttt{X3} is sporadically missing. Systematic missingness can be
-further subdivided into unobserved constants (i.e., the same value
-within clusters) and non-measured random variables (which may differ per
-unit within clusters). In Figure 1, the former would imply that the
-unobserved values for units 3 and 4 on column \texttt{X2} are identical.
-With the latter, the values would differ. The optimal strategy for
-dealing with the missingness may therefore depend on the observed
-missing data pattern.
-
-Another characteristic of the missing data to take into account in
-analyses is the mechanism behind the missingness. Although the essence
-of the true non-response mechanism may not be known, it can be inferred
-or assumed to be one of the following:
-
-\begin{itemize}
-\tightlist
-\item
-  Missing Completely At Random (MCAR), where the probability to be
-  missing is equal across all data entries;
-\item
-  Missing At Random (MAR), where the probability to be missing depends
-  on observed information;
-\item
-  Missing Not At Random (MNAR), where the probability to be missing
-  depends on unrecorded information, making the missingness
-  non-ignorable \citep{rubi76, meng94}. Depending on the assumed
-  missingness mechanism, missing data handling strategies may be more or
-  less suitable, see e.g., \citet{yuce08} and \citet{hox15}.
-\end{itemize}
+data comes the potential for clustered missingness. Incomplete
+multilevel data can be categorized into two general patterns: systematic
+missingness and sporadic missingness \citep{resc13}. Systematic
+missingness implies that one or more variables are never observed in a
+certain cluster. With sporadic missingness there may be observed data
+for some but not all units in a cluster \citep{buur18, jola18}. We have
+visualized this difference in Figure 1, which shows an \(n \times p\)
+set \(\mathbf{X} = X_1, \dots, X_p\), with \(n\) units distributed over
+\(N\) clusters, and \(p=3\) columns. Column \(X1\) is completely
+observed, column \(X2\) is systematically missing, and column \(X3\) is
+sporadically missing. To analyze these incomplete data, we have to take
+the nature of the missingness and the cluster structure into account.
+For example, the sporadic missingness in \(X3\) could be easily amended
+if this would be a cluster-level variable (and thus constant within
+clusters). We could then just extrapolate the true (but missing) value
+of \(X3\) for unit 1 from unit 2, and the value for unit 4 from unit 3.
+If \(X3\) would instead be a unit-level variable (which may vary within
+clusters), we could not just recover the unobserved `truth', but need
+some kind of missing data method or discard the incomplete units
+altogether (i.e., list-wise deletion/complete case analysis). And, with
+the systematic missingness in \(X2\), it is impossible to fit a
+multilevel model since we cannot estimate the intercept of cluster 2. We
+would have to exclude this cluster from our analyses entirely to obtain
+any results. Obviously, excluding observations is not a desirable
+workflow.
 
 Ignoring the missingness in analyses can be extremely harmful to
-inferences. Complete case analysis (i.e., excluding all units with one
-or more missing entries) can introduce bias in statistical inference and
-lowers statistical power. Instead, the missingness should be
-accommodated \emph{before} or \emph{within} the analysis of scientific
-interest. Especially the former is very generic and popular and is
-widely known as imputation. Imputing (i.e., filling in) the missing
-values separates the missing data problem from the scientific problem:
-missing data are replaced by plausible values whereafter the completed
-data is analysed as if it were completely observed. The \proglang{R}
-package \pkg{mice} has become the de-facto standard for imputation by
-chained equations, which iteratively solves the missingness on a
-variable-by-variable basis. \pkg{mice} is known to yield valid
-inferences under many different missing data circumstances
-\citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
-the context of multilevel data, under varying missing data mechanisms.
+inferences. Complete case analysis can introduce bias in statistical
+inferences and lowers statistical power. Instead, the missingness should
+be accommodated \emph{before} or \emph{within} the analysis of
+scientific interest. Especially the former is very generic and popular.
+Imputing (i.e., filling in) the missing values separates the missing
+data problem from the scientific problem: missing data are replaced by
+plausible values whereafter the completed data is analysed as if it were
+completely observed. The \proglang{R} package \pkg{mice} has become the
+de-facto standard for imputation by chained equations, which iteratively
+solves the missingness on a variable-by-variable basis. \pkg{mice} is
+known to yield valid inferences under many different missing data
+circumstances \citep{buur18}. In this paper, we will discuss how to use
+\pkg{mice} in the context of multilevel data.
 
@@
 with \pkg{mice}. We provide practical guidelines and code snippets for
-different missing data situations. For reasons of brevity, we focus on
-imputation by chained equations exclusively (although the alternative,
-joint modeling imputation for multilevel data or \pkg{jomo}
-\citet{jomo}, has been implemented in \pkg{mice} as well). Other useful
-resources for the analysis of incomplete multilevel data include the
-\proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and
-the empirical work by \citet{audi18} and \citet{grun18}. Note that this
-tutorial paper assumes a basic level of knowledge on multilevel models.
-We're providing an overview of implementations. It's up-to the reader to
-decide which multilevel strategy suits their data. We won't go into
-detail for the different methods (and equations). This paper is just a
-software tutorial, so we'll keep it practical. Assumed knowledge also
-includes the use of the `piping operator', \texttt{\%\textgreater{}\%},
-adopted from the \texttt{magrittr} package and avidly used among
-\texttt{tidy} workflows, and the \texttt{lme4} notation for multilevel
-models.
+different missing data situations, including missing not at random
+(MNAR) mechanisms \citep[where the probability to be missing depends on
+unrecorded information, making the missingness
+non-ignorable,][]{rubi76, meng94}. For reasons of brevity, we focus on
+imputation by chained equations wit \pkg{mice} exclusively\footnote{Note
+  that the alternative, joint modeling imputation for multilevel data or
+  \pkg{jomo} \citet{jomo}, has been implemented in \pkg{mice} as well
+  but is outside the scope of this tutorial.}. Other useful resources
+for the analysis of incomplete multilevel data include the \proglang{R}
+packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work
+by \citet{audi18} and \citet{grun18}. Please note that this tutorial
+paper assumes a basic level of knowledge on multilevel
+models.\footnote{Note to self: We're providing an overview of
+  implementations. It's up-to the reader to decide which multilevel
+  strategy suits their data. We won't go into detail for the different
+  methods (and equations). This paper is just a software tutorial, so
+  we'll keep it practical.} Assumed knowledge also includes the use of
+the `piping operator', \texttt{\%\textgreater{}\%}, adopted from the
+\pkg{magrittr} package, and the \pkg{lme4} notation for multilevel
+models.\footnote{Add environment info, seed and version number(s)
+  somewhere!}
 
-To illustrate how to impute incomplete multilevel data, we structure
-this tutorial around three case studies:
+We illustrate how to impute incomplete multilevel data by means of three
+case studies:
 
@@
 \item
-  \texttt{mice::popmis} (simulated data on perceived popularity,
-  \(n = 2,000\) pupils across \(N = 100\) schools, with univariate MAR
-  missingness);
+  \texttt{popmis} from the \pkg{mice} package (simulated data on
+  perceived popularity, \(n = 2,000\) pupils across \(N = 100\)
+  schools);
 \item
-  \texttt{GJRM::hiv} (simulated data on HIV diagnoses, \(n = 6,416\)
-  patients across \(N = 9\) regions, with univariate MNAR missingness)
+  \texttt{hiv} from the \pkg{GJRM} package (simulated data on HIV
+  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions);
 \item
-  \texttt{metamisc::impact} (empirical data on traumatic brain injuries,
-  \(n = 11,022\) patients across \(N = 15\) studies, without
-  \texttt{NA}s).
+  \texttt{impact} from the \pkg{metamisc} package (empirical data on
+  traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
+  studies).
 \end{itemize}
 
-In each case study we highlight one aspect of the imputation workflow.
-With the \texttt{mice::popmis} data, we show different ways of designing
-an imputation model and what happens if you misspecify this model. With
-the \texttt{GJRM::hiv} data we extend the imputation model to include
-Heckman-type selection-inclusion methods. And with the
-\texttt{metamisc::impact} data we provide a example of multivariate
-missingness in real-world data. For all case studies we discuss the
-nature of the incomplete data, the imputation model(s), and evaluation
-of the imputed data: A. Choose an analysis model (so the imputation
-model will be compatible with the analyses); B. Evaluate the incomplete
-data; C. Develop the imputation model(s); D. Impute the missingness; E.
-Evaluate the imputations.
+For each of these datasets, we will discuss the nature of the
+missingness, choose one or more imputation models and evaluate the
+imputed data, but we will also highlight one specific asect of the
+imputation workflow. With the \texttt{popmis} data, we show how (not) to
+develop an imputation model. With the \texttt{hiv} data we focus on
+extending the imputation model to include Heckman-type
+selection-inclusion methods. And with the \texttt{impact} data we
+provide a example of multivariate missingness in real-world data.
+Together, this should give enough scaffolding for applied researchers
+who are faced with incomplete multilevel data.
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\begin{center}
+REVIEW UNTIL THIS POINT SVP
+\end{center}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
 
@@
 
-TODO: add all estimimates, seed, and version nr., this also shows the
-workflow for pooling etc.
-
 Since the data is simulated and the missingness is induced, we can
@@
 4 sex:texp    "-0.034 [-0.051, -0.017]"
-\end{CodeOutput}
-\begin{CodeOutput}
-# A tibble: 4 x 2
-  term                 estimate
-  <chr>                <chr>   
-1 sd__(Intercept)      0.642   
-2 cor__(Intercept).sex 0.077   
-3 sd__sex              0.476   
-4 sd__Observation      0.626   
 \end{CodeOutput}
@@
 
-TODO: rotate labels, make missingness \% gradient for clusters?? Look at
-analysis model, maybe copy from GREAT data example e.g., adjusted
-prognostic effect of ct on unfortunate outcomes.
-
-gradient would be the number of clusters for which this variable is
-missing
+TODO: Look at analysis model, maybe copy from GREAT data example e.g.,
+adjusted prognostic effect of ct on unfortunate outcomes. Add something
+about systematically missing data here. In the md pattern plot: rotate
+labels, make missingness \% gradient for clusters (ie the number of
+clusters for which this variable is missing).
 

### 2022-02-11 R/plot_incomplete.R

@@ -31,7 +31,7 @@ miss_ind %>%
   geom_tile(fill = "white", size = 1.2) +
   geom_text(aes(label = text), color = "black") +
   scale_x_discrete(position = "top", labels = c("cluster", "X1", "X2", "X3")) +
-  scale_y_discrete() +
+  scale_y_discrete(labels = c("n", "", "4", "3", "2", "1")) +
   scale_color_manual(values = c(plot_col, blank = "white"), name = "Legend:") +
   theme_minimal() +
   labs(x = "", y = "") +
@@ -42,7 +42,7 @@ miss_ind %>%
   # annotate(geom = "text", x = "X0", y = "2", label = "{", vjust = 0, hjust = 0, size = 16) +
   # annotate(geom = "text", x = "X0", y = "0", label = "{...", vjust = 0, hjust = 0, size = 16) +
   theme(
-    axis.text.y = element_blank(),
+    # axis.text.y = element_blank(),
     # text=element_text(family="Arial Unicode MS"),
     legend.position = "none", 
     panel.grid.major = element_blank())

### 2022-02-11 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Clean-up and rename manuscript file

--- 2022-02-11 Manuscript/Manuscript.Rmd
@@
  - \usepackage{ulem}
-output: rticles::jss_article
+output: 
+    rticles::jss_article: default
+    word_document: default
 bibliography: ../References/multilevelmice.bib
@@
 plot_md_set()
-# TODO: make cluster part f the data (blue)
-# TODO: dots between 2 and N, for each variable
 ```
@@
 
-<!-- While the sporadic missingness in $X1$ would at least provide some information for each cluster.
-can be further subdivided depending on the underlying true value of the missing entries: this can either be  constant within clusters, or a non-measured random variable (which may differ per unit within clusters).
-In Figure 1, the former would imply that the unobserved values for units 3 and 4 on column $X2$ are identical.
-With the latter, the values would differ.
-The pattern of the missing data across clusters determines the optimal strategy for dealing with the missingness.
--->
-
 Ignoring the missingness in analyses can be extremely harmful to inferences.
@@
 Especially the former is very generic and popular.
-Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed.
+Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analyzed as if it were completely observed.
 The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
@@
 
-<!-- "We find it useful to distinguish the missingness pattern, which describes which values are missing and observed in the data matrix and the missingness mechanism (or mechanisms), which concerns the relationship between missingness and the values of variables in the data matrix" (Little and Rubin, p.
-8).
--->
-
-<!-- Table 7.1: Questions to gauge the complexity of a multilevel imputation task. -->
-<!-- 1.	Will the complete-data model include random slopes? -->
-<!-- 2.	Will the data contain systematically missing values? -->
-<!-- 3.	Will the distribution of the residuals be non-normal? -->
-<!-- 4.	Will the error variance differ over clusters? -->
-<!-- 5.	Will there be small clusters? -->
-<!-- 6.	Will there be a small number of clusters? -->
-<!-- 7.	Will the complete-data model have cross-level interactions? -->
-<!-- 8.	Will the dataset be very large? (fimd, section 7.3)-->
-
 <!-- Another characteristic of the missing data to take into account in analyses is the mechanism behind the missingness.
 Although the essence of the true non-response mechanism may not be known, it can be inferred or assumed to be one of the following: -->
-
 <!-- - Missing Completely At Random (MCAR), where the probability to be missing is equal across all data entries; -->
@@
 <!-- Depending on the assumed missingness mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15. -->
+<!-- "We find it useful to distinguish the missingness pattern, which describes which values are missing and observed in the data matrix and the missingness mechanism (or mechanisms), which concerns the relationship between missingness and the values of variables in the data matrix" (Little and Rubin, p.
+8).
+-->
 
@@
 
-For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific asect of the imputation workflow.
+For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
 With the `popmis` data, we show how (not) to develop an imputation model.
 With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods.
-And with the `impact` data we provide a example of multivariate missingness in real-world data.
+And with the `impact` data we provide an example of multivariate missingness in real-world data.
 Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.
+
+<!-- Table 7.1: Questions to gauge the complexity of a multilevel imputation task. -->
+<!-- 1.	Will the complete-data model include random slopes? -->
+<!-- 2.	Will the data contain systematically missing values? -->
+<!-- 3.	Will the distribution of the residuals be non-normal? -->
+<!-- 4.	Will the error variance differ over clusters? -->
+<!-- 5.	Will there be small clusters? -->
+<!-- 6.	Will there be a small number of clusters? -->
+<!-- 7.	Will the complete-data model have cross-level interactions? -->
+<!-- 8.	Will the dataset be very large? (fimd, section 7.3)-->
 
@@
 ```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
-# randomly select a few classes
-# clusters <- sample(1:100, 10)
+# estimates complete popularity data
+est_comp <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popcomp) 
+# fixed effects only
+est_comp_fixed <- est_comp %>% 
+  broom.mixed::tidy("fixed", conf.int = TRUE) %>% 
+  mutate(term = term,
+         estimate  = paste0(
+           round3(estimate), " [", 
+           round3(conf.low), ", ", 
+           round3(conf.high), "]"), 
+         .keep = "none") %>% 
+  as.data.frame()
+# # random effects
+# est_comp %>% 
+#   broom.mixed::tidy("ran_pars") %>% 
+#   mutate(term = term, #factor(term, levels = unique(.$term), labels = c()),
+#          estimate  = round3(estimate), 
+#          .keep = "none")
+# # rename broom.mixed estimates
+# lvl = c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation")
+# lbl = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor.
+cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error")
 
-# load data 
-# popmis <- mice::popmis 
-#readRDS("../Data/popNCR2.RDS") %>% 
-  #filter(class %in% clusters) %>% 
-  #mutate(class = as.factor(class))
-# popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
-#   mutate(school = SCHOOL,
-#          popular = POPULAR,
-#          sex = SEX,
-#          texp = TEXP,
-#          .keep = "none")
-est_comp <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popcomp) 
-# %>% 
-#   broom.mixed::tidy(conf.int = TRUE) 
-# %>% 
-  # mutate(
-  #   term = factor(term, levels = c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation"), labels = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor.
-cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error"), ordered = TRUE))
+# compare complete data to missingness
+# popcomp <- popcomp %>% 
+#   cbind(R = is.na(popmis$popular))
+# t.test(x = popcomp$popular[popcomp$R], y = popcomp$popular[!popcomp$R]) 
+# ggplot(popcomp, aes(y=popular, x = R)) +
+#   geom_boxplot()
 ```
@@
 
-<!-- TODO: add all estimimates, seed, and version nr., this also shows the workflow for pooling etc. -->
-
 Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data.
@@
 
-```{r echo=FALSE}
-# lvl <- c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation")
-# lbl = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor.
-cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error")
-est_comp %>% 
-  broom.mixed::tidy("fixed", conf.int = TRUE) %>% 
-  mutate(term = term,
-         estimate  = paste0(
-           round3(estimate), " [", 
-           round3(conf.low), ", ", 
-           round3(conf.high), "]"), 
-         .keep = "none")
-
-# est_comp %>% 
-#   broom.mixed::tidy("ran_pars") %>% 
-#   mutate(term = term, #factor(term, levels = unique(.$term), labels = c()),
-#          estimate  = round3(estimate), 
-#          .keep = "none")
+```{r comp_est, echo=FALSE}
+est_comp_fixed %>% setNames(c("", "Estimate with 95% CI"))
 ```
-
-<!--  Let's apply the following five methods: -->
-
-<!-- sample: Find imputations by random sampling from the observed values in lpo. This method ignores sch; -->
-
-<!-- pmm: Single-level predictive mean matching with the school indicator coded as a dummy variable; -->
-
-<!-- 2l.pan: Multilevel method using the linear mixed model to draw univariate imputations; -->
-
-<!-- 2l.norm: Multilevel method using the linear mixed model with heterogeneous error variances; -->
-
-<!-- 2l.pmm: Predictive mean matching based on predictions from the linear mixed model, with random draws from the regression coefficients and the random effects, using five donors. -->
 
@@
 plot_md_pat(popmis)
-# TODO: add pattern by cluster, text mining example square size
-# TODO: try to make it a shepley plot
-
-# popcomp <- popcomp %>% 
-#   cbind(R = is.na(popmis$popular))
-# t.test(x = popcomp$popular[popcomp$R], y = popcomp$popular[!popcomp$R]) 
-# ggplot(popcomp, aes(y=popular, x = R)) +
-#   geom_boxplot()
-
 ```
@@
 (In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `texp`.
-<!-- There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well. -->
 
 ```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-# popmis %>% 
-#   filter(as.numeric(school) < 5) %>% 
-#   ggplot(aes(x = teachpop, color = school)) +
-#     geom_density() +
-#     facet_wrap(~factor(is.na(popular), labels = c("popular missing", "popular observed")), ncol = 1) +
-#     # scale_color_manual(values = plot_col, name = "popular") +
-#     theme_classic() # +
-#     # theme(legend.position = "bottom")
-
 plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
@@
 # TODO: add functions to mice
-# TODO: add title and informative legend
-# TODO: add facets for some clusters
-# or add propensity score distribution `is.na(popular) ~ .`
-# TODO: make it average of cluster dens, not marginal
-# alternatively add a quartile line around the density with geom_ribbon
+# TODO: add title and informative legend (pop obs and pop miss)
+# TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .`
+# TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon
+# TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?)
 
-## comments 24/1
-# pop obs and pop miss in legend
-# add to caption
-# add fill
-# overlap coeff??
-# make densities proportional
-
+# # maybe use facets instead
+# popmis %>%
+#   filter(as.numeric(school) < 5) %>%
+#   ggplot(aes(x = texp, color = school)) +
+#     geom_density() +
+#     facet_wrap(~factor(is.na(popular), labels = c("popular missing", "popular observed")), ncol = 1) +
+#     scale_color_manual(values = plot_col, name = "popular") +
+#     theme_classic() # +
+#     # theme(legend.position = "bottom")
 ```
@@
 In this model, we ignore the multilevel structure of the data, despite the high ICCs.
+And assumes exchangeability between units.
 This is purely to illustrate the effects of ignoring the clustering in our imputation effort.
@@
 #TODO: think about interactive clusters for the strippplots
-#TODO: switch off outliers in boxplot?
 ```
@@
 
-Add: Assumes exchangability between units.
+<!-- pmm: Single-level predictive mean matching with the school indicator coded as a dummy variable; -->
+<!-- 2l.pan: Multilevel method using the linear mixed model to draw univariate imputations; -->
+<!-- 2l.norm: Multilevel method using the linear mixed model with heterogeneous error variances; -->
+<!-- 2l.pmm: Predictive mean matching based on predictions from the linear mixed model, with random draws from the regression coefficients and the random effects, using five donors. -->
 
@@
 
-Schafer book fortran workflows.
+<!-- Look up Schafer book fortran workflows. -->
 
@@
 
-The data is already imputed (Steyerberg et al, 2008), so we'll induce missingness ourselves.
-For example, MAR missingness varying by cluster.^[Observed data pattern should differ per cluster.
-So, in cluster 1, the missingness would depend on age, but not in cluster two.
-Split the dataframe and run `ampute()` on each cluster.]
+The analysis model for this dataset is a prediction model with `mort` as the outcome.^[Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of ct on unfortunate outcomes.
+Add something about systematically missing data here.] The data is already imputed (Steyerberg et al, 2008), so we've induced missingness again based on the missingness in the original data.
 
-```{r impact, echo=FALSE, fig.width=5}
+```{r impact, echo=FALSE, fig.width=6, fig.height=8}
 # load data
@@
 plot_md_pat(impact_NA)
+# TODO: rotate variable labels
+# TODO: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters like shepley plot
+# TODO: make missingness % gradient for clusters (ie the number of clusters for which this variable is missing).
 
+```
+
+```{r echo=FALSE}
 # md pattern in first cluster
@@
 plot_md_pat(impact_NA[impact_NA$name == "APOE", ])
-# TODO: rotate variable labels
-
-# TODO: check md pattern in clusters, should differ, otherwise create new missingness
-# MAYBE: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters instead
-# TODO: e-mail Joop for DGM or look at adrew and jennifer book for data
-# sim.multi from psych
 ```
-
-TODO: Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of ct on unfortunate outcomes.
-Add something about systematically missing data here.
-In the md pattern plot: rotate labels, make missingness % gradient for clusters (ie the number of clusters for which this variable is missing).
 

### 2022-02-11 Manuscript/Manuscript.tex
Clean-up and rename manuscript file

@@
 data problem from the scientific problem: missing data are replaced by
-plausible values whereafter the completed data is analysed as if it were
+plausible values whereafter the completed data is analyzed as if it were
 completely observed. The \proglang{R} package \pkg{mice} has become the
@@
 missingness, choose one or more imputation models and evaluate the
-imputed data, but we will also highlight one specific asect of the
+imputed data, but we will also highlight one specific aspect of the
 imputation workflow. With the \texttt{popmis} data, we show how (not) to
@@
 selection-inclusion methods. And with the \texttt{impact} data we
-provide a example of multivariate missingness in real-world data.
+provide an example of multivariate missingness in real-world data.
 Together, this should give enough scaffolding for applied researchers
@@
 \begin{CodeOutput}
-# A tibble: 4 x 2
-  term        estimate                 
-  <chr>       <chr>                    
-1 (Intercept) " 3.314 [ 2.998,  3.629]"
-2 sex         " 1.330 [ 1.069,  1.590]"
-3 texp        " 0.110 [ 0.090,  0.130]"
-4 sex:texp    "-0.034 [-0.051, -0.017]"
+                 Estimate with 95% CI
+1 (Intercept)  3.314 [ 2.998,  3.629]
+2         sex  1.330 [ 1.069,  1.590]
+3        texp  0.110 [ 0.090,  0.130]
+4    sex:texp -0.034 [-0.051, -0.017]
 \end{CodeOutput}
@@
 this model, we ignore the multilevel structure of the data, despite the
-high ICCs. This is purely to illustrate the effects of ignoring the
-clustering in our imputation effort.
+high ICCs. And assumes exchangeability between units. This is purely to
+illustrate the effects of ignoring the clustering in our imputation
+effort.
 
@@
 (i.e.~true) value for every pupil in the school.
-
-Add: Assumes exchangability between units.
 
@@
 cluster-specific residual error variance.
-
-Schafer book fortran workflows.
 
@@
 
-The data is already imputed (Steyerberg et al, 2008), so we'll induce
-missingness ourselves. For example, MAR missingness varying by
-cluster.\footnote{Observed data pattern should differ per cluster. So,
-  in cluster 1, the missingness would depend on age, but not in cluster
-  two. Split the dataframe and run \texttt{ampute()} on each cluster.}
+The analysis model for this dataset is a prediction model with
+\texttt{mort} as the outcome.\footnote{Look at analysis model, maybe
+  copy from GREAT data example e.g., adjusted prognostic effect of ct on
+  unfortunate outcomes. Add something about systematically missing data
+  here.} The data is already imputed (Steyerberg et al, 2008), so we've
+induced missingness again based on the missingness in the original data.
 
@@
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-2} \end{center}
+\end{CodeChunk}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-3} \end{center}
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\begin{center}\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-3-2} \end{center}
 
 \end{CodeChunk}
-
-TODO: Look at analysis model, maybe copy from GREAT data example e.g.,
-adjusted prognostic effect of ct on unfortunate outcomes. Add something
-about systematically missing data here. In the md pattern plot: rotate
-labels, make missingness \% gradient for clusters (ie the number of
-clusters for which this variable is missing).
 

### 2022-02-22 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Add analysis model impact data

@@
 The last section is not part of the manuscript, but purely for reminders.
+We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
 keywords:
@@
 library(miceadds)
+library(lme4)
+library(broom.mixed)
 # library(metamisc)
@@
   - `ct` Marshall Computerized Tomography classification,
-  - `hypox` Hypoxia (0=no, 1=yes),
-  - `hypots` Hypotension (0=no, 1=yes),
-  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
-  - `edh` Epidural hematoma (0=no, 1=yes),
+  <!-- - `hypox` Hypoxia (0=no, 1=yes), -->
+  <!-- - `hypots` Hypotension (0=no, 1=yes), -->
+  <!-- - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes), -->
+  <!-- - `edh` Epidural hematoma (0=no, 1=yes), -->
   - `mort` 6-month mortality (0=alive, 1=dead).
 
-The analysis model for this dataset is a prediction model with `mort` as the outcome.^[Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of ct on unfortunate outcomes.
+The analysis model for this dataset is a prediction model with `mort` as the outcome.^[Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of `ct` on unfortunate outcomes, we just want to know the adjusted odds ratio for `ct`.
 Add something about systematically missing data here.] The data is already imputed (Steyerberg et al, 2008), so we've induced missingness again based on the missingness in the original data.
+
+$$
+\text{mort}_{ij} =
+\gamma_{00} + 
+\gamma_{01} \text{ type}_{j} + 
+\gamma_{10} \text{ age}_{ij} + 
+\gamma_{10} \text{ moter\_score}_{ij} + 
+\gamma_{10} \text{ pupil}_{ij} + 
+\gamma_{10} \text{ ct}_{ij} + \\
+u_{0j} + 
+u_{1j} \text{ age}_{ij} + 
+u_{1j} \text{ moter\_score}_{ij} +  
+u_{1j} \text{ pupil}_{ij} +  
+u_{1j} \text{ ct}_{ij} +  
+e_{ij} \\
+$$
+
+`glmer(mort ~ 1 + type + age + motor_score + pupil + ct + hypox + hypots + tsah + edh + (1 | name))`
+
+<!-- \gamma_{10} \text{ hypox}_{ij} + -->
+<!-- \gamma_{10} \text{ hypots}_{ij} + -->
+<!-- \gamma_{10} \text{ tsah}_{ij} + -->
+<!-- \gamma_{10} \text{ edh}_{ij} + -->
+
+<!-- u_{1j} \text{ hypox}_{ij} + -->
+<!-- u_{1j} \text{ hypots}_{ij} + -->
+<!-- u_{1j} \text{ tsah}_{ij} + -->
+<!-- u_{1j} \text{ edh}_{ij} + -->
 
@@
 # TODO: make missingness % gradient for clusters (ie the number of clusters for which this variable is missing).
-
 ```
 
-```{r echo=FALSE}
+```{r impact_md, echo=FALSE}
 # md pattern in first cluster
@@
 -->
+
+```{r impact_cca}
+mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
+mod_cca %>% broom.mixed::tidy()
+
+```
+
+```{r impact_pred}
+quickpred(impact_NA)
+```
 

### 2022-03-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Feedback on introduction section

--- 2022-02-11 Manuscript/Manuscript.tex
@@
 construction. The last section is not part of the manuscript, but purely
-for reminders.
+for reminders. We aim to submit at JSS, so there is no word count limit
+(``There is no page limit, nor a limit on the number of figures or
+tables'').
 }
@@
 
-In many contemporary data analysis efforts, some form of hierarchical or
-clustered structure is recorded. For example, students may be clustered
-in classes in psychometrics research, or patients may be clustered in
-studies in individual patient data meta-analyses. Ignoring the clustered
-structure of such multilevel data can be harmful to the statistical
-inferences and bias results. Imagine a case where cross-level
-interactions between unit-level variables and cluster-level variables
-are present. The cluster to which a unit belongs may then influence the
-unit-level observations--and vice versa for each of the units that make
-up the cluster \citep{hox17}. These relations can and should be taken
-into account when developing analysis models for multilevel data for the
-simple reason that groups of observations share some common variance.
-The variability due to clustering is often measured by means of the
-intraclass coefficient \citep[ICC,][]{fish25}. The ICC can be seen as
-the percentage of variance that can be attributed to the cluster-level,
-where a high ICC would indicate that a lot of variability is due to the
-cluster structure. Multilevel models typically accommodate for this
-variability by including a separate group mean for each cluster. In
-addition to random intercepts, multilevel models can also include random
-effects and heterogeneous residual error variances across clusters
-\citep[see e.g.][\citet{hox17} and \citet{jong21}]{gelm06}. There are
-many names for models that take clustering into account. Some popular
-examples are `multilevel models', `hierarchical models', `mixed effect
-models' and `random effect models'.
+\sout{In many contemporary data analysis efforts, some form of
+hierarchical or clustered structure is recorded.} Contemporary decision
+making is driven increasingly often and sometimes even entirely dictated
+by data. Although data relevant for decision makers were initially
+collected using small and well-designed studies, there has been a
+growing need to address more complex questions at a wider scale. This
+not only requires to collect larger amounts of data, but also from more
+diverse settings and populations. For example, {[}TD: introduce IPDMA
+here; eg. HIV studies Johanna is referring to{]}. Another example is the
+use of large databases with information from thousands or even millions
+of individuals from multiple cities, regions or even countries. Finally,
+a third situation arises when data are collected from multicenter
+studies {[}TD: maybe refer to the schools/classes example here{]}. For
+example, students may be clustered in classes in psychometrics research,
+or patients may be clustered in studies in individual patient data
+meta-analyses in medical research. A common characteristic in
+aforementioned examples is the presence of clustering. {[}TD: briefly
+explain what is clustering?{]}
+
+Ignoring the clustered structure of such multilevel data can be harmful
+to the statistical inferences and introduce bias in estimators
+\citep{hox17}. \sout{Imagine a case where cross-level interactions
+between unit-level variables and cluster-level variables are present.
+The cluster to which a unit belongs may then influence the unit-level
+observations--and vice versa for each of the units that make up the
+cluster.} These relations can and should be taken into account when
+developing analysis models for multilevel data for the simple reason
+that groups of observations share some common variance. \sout{The
+variability due to clustering is often measured by means of the
+intraclass coefficient (ICC). The ICC can be seen as the percentage of
+variance that can be attributed to the cluster-level, where a high ICC
+would indicate that a lot of variability is due to the cluster
+structure.} Multilevel models typically accommodate for this variability
+by including a separate group mean for each cluster. In addition to
+random intercepts, multilevel models can also include random effects and
+heterogeneous residual error variances across clusters \citep[see
+e.g.][\citet{hox17} and \citet{jong21}]{gelm06}. There are many names
+for models that take clustering into account. Some popular examples are
+`multilevel models', `hierarchical models', `mixed effect models' and
+`random effect models'. {[}TD: emphasize when/why we need to account for
+clustering in the analysis of clustered data. Why is the presence of
+clustering relevant when considering multiple imputation of missing
+data? e.g.~distinction between systematically and sporadically missing
+data. But also: mechanisms of missing data (e.g.~MCAR, MAR, MNAR) may
+differ between clusters. But also: relation between observed data may
+differ between clusters? When/ why should we avoid using traditional
+imputation methods? e.g.~congeniality issues.{]}
 
@@
 
-{\centering \includegraphics{Manuscript_files/figure-latex/patterns-1}
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/patterns-1}
 
@@
 missingness may occur at the unit level. But with multiple levels of
-data comes the potential for clustered missingness. Incomplete
-multilevel data can be categorized into two general patterns: systematic
-missingness and sporadic missingness \citep{resc13}. Systematic
-missingness implies that one or more variables are never observed in a
-certain cluster. With sporadic missingness there may be observed data
-for some but not all units in a cluster \citep{buur18, jola18}. We have
-visualized this difference in Figure 1, which shows an \(n \times p\)
-set \(\mathbf{X} = X_1, \dots, X_p\), with \(n\) units distributed over
-\(N\) clusters, and \(p=3\) columns. Column \(X1\) is completely
-observed, column \(X2\) is systematically missing, and column \(X3\) is
-sporadically missing. To analyze these incomplete data, we have to take
-the nature of the missingness and the cluster structure into account.
-For example, the sporadic missingness in \(X3\) could be easily amended
-if this would be a cluster-level variable (and thus constant within
-clusters). We could then just extrapolate the true (but missing) value
-of \(X3\) for unit 1 from unit 2, and the value for unit 4 from unit 3.
-If \(X3\) would instead be a unit-level variable (which may vary within
-clusters), we could not just recover the unobserved `truth', but need
+data comes the potential for clustered missingness. Therefore,
+incomplete multilevel data can be categorized into two general patterns:
+systematic missingness and sporadic missingness \citep{resc13}.
+Systematic missingness implies that one or more variables are never
+observed in a certain cluster. With sporadic missingness there may be
+observed data for some but not all units in a cluster
+\citep{buur18, jola18}. We have visualized this difference in Figure 1,
+which shows an \(n \times p\) set \(\mathbf{X} = X_1, \dots, X_p\), with
+\(n\) units distributed over \(N\) clusters, and \(p=3\) columns. Column
+\(X1\) is completely observed, column \(X2\) is systematically missing
+in cluster 2, and column \(X3\) is sporadically missing. To analyze
+these incomplete data, we have to take the nature of the missingness and
+the cluster structure into account. For example, the sporadic
+missingness in \(X3\) could be easily amended if this would be a
+cluster-level variable (and thus constant within clusters). We could
+then just extrapolate the true (but missing) value of \(X3\) for unit 1
+from unit 2, and the value for unit 4 from unit 3. If \(X3\) would
+instead be a unit-level variable (which may vary within clusters), we
+could not just recover the unobserved `truth', but would need to use
 some kind of missing data method or discard the incomplete units
-altogether (i.e., list-wise deletion/complete case analysis). And, with
-the systematic missingness in \(X2\), it is impossible to fit a
+altogether (i.e., list-wise deletion/complete case analysis). Further,
+with the systematic missingness in \(X2\), it is impossible to fit a
 multilevel model since we cannot estimate the intercept of cluster 2. We
@@
 scientific interest. Especially the former is very generic and popular.
-Imputing (i.e., filling in) the missing values separates the missing
-data problem from the scientific problem: missing data are replaced by
-plausible values whereafter the completed data is analyzed as if it were
-completely observed. The \proglang{R} package \pkg{mice} has become the
-de-facto standard for imputation by chained equations, which iteratively
-solves the missingness on a variable-by-variable basis. \pkg{mice} is
-known to yield valid inferences under many different missing data
-circumstances \citep{buur18}. In this paper, we will discuss how to use
-\pkg{mice} in the context of multilevel data.
+{[}VJ: add why multiple values are imputed.{]} Imputing (i.e., filling
+in) the missing values separates the missing data problem from the
+scientific problem: missing data are replaced by plausible values
+whereafter the completed data is analyzed as if it were completely
+observed. The \proglang{R} package \pkg{mice} has become the de-facto
+standard for imputation by chained equations, which iteratively solves
+the missingness on a variable-by-variable basis. \pkg{mice} is known to
+yield valid inferences under many different missing data circumstances
+\citep{buur18}. In this paper, we will discuss how to use \pkg{mice} in
+the context of multilevel data.
 
@@
 \item
-  \texttt{popmis} from the \pkg{mice} package (simulated data on
+  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
   perceived popularity, \(n = 2,000\) pupils across \(N = 100\)
-  schools);
+  schools,][]{mice};
 \item
-  \texttt{hiv} from the \pkg{GJRM} package (simulated data on HIV
-  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions);
+  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
+  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions,][]{GJRM};
 \item
-  \texttt{impact} from the \pkg{metamisc} package (empirical data on
-  traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
-  studies).
+  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
+  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
+  studies,][]{metamisc}.
 \end{itemize}
@@
 imputed data, but we will also highlight one specific aspect of the
-imputation workflow. With the \texttt{popmis} data, we show how (not) to
-develop an imputation model. With the \texttt{hiv} data we focus on
-extending the imputation model to include Heckman-type
-selection-inclusion methods. And with the \texttt{impact} data we
-provide an example of multivariate missingness in real-world data.
-Together, this should give enough scaffolding for applied researchers
-who are faced with incomplete multilevel data.
+imputation workflow. With the \texttt{popmis} data, we show how (and how
+not) to develop an imputation model. With the \texttt{hiv} data we focus
+on extending the imputation model to include Heckman-type
+selection-inclusion methods. With the \texttt{impact} data we provide an
+example of multivariate missingness in real-world data. Together, this
+should give enough scaffolding for applied researchers who are faced
+with incomplete multilevel data.
 
@@
 
+TODO: make a `translation table' between model equation and lme4. Use
+betas instead of gamma's and mu's.
+
 Since the data is simulated and the missingness is induced, we can
@@
 
-{\centering \includegraphics{Manuscript_files/figure-latex/pop_pat-1}
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pat-1}
 
@@
 
-{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-1}
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_dist-1}
 
@@
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_ignored_eval-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_ignored_eval-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_predictor_eval-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor_eval-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_norm_eval-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_norm_eval-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_pan_eval-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pan_eval-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/hiv-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/hiv-1} \end{center}
 
@@
 \item
-  \texttt{hypox} Hypoxia (0=no, 1=yes),
-\item
-  \texttt{hypots} Hypotension (0=no, 1=yes),
-\item
-  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
-\item
-  \texttt{edh} Epidural hematoma (0=no, 1=yes),
-\item
   \texttt{mort} 6-month mortality (0=alive, 1=dead).
@@
 \texttt{mort} as the outcome.\footnote{Look at analysis model, maybe
-  copy from GREAT data example e.g., adjusted prognostic effect of ct on
-  unfortunate outcomes. Add something about systematically missing data
-  here.} The data is already imputed (Steyerberg et al, 2008), so we've
-induced missingness again based on the missingness in the original data.
+  copy from GREAT data example e.g., adjusted prognostic effect of
+  \texttt{ct} on unfortunate outcomes, we just want to know the adjusted
+  odds ratio for \texttt{ct}. Add something about systematically missing
+  data here.} The data is already imputed (Steyerberg et al, 2008), so
+we've induced missingness again based on the missingness in the original
+data.
+
+\[
+\text{mort}_{ij} =
+\gamma_{00} + 
+\gamma_{01} \text{ type}_{j} + 
+\gamma_{10} \text{ age}_{ij} + 
+\gamma_{10} \text{ moter\_score}_{ij} + 
+\gamma_{10} \text{ pupil}_{ij} + 
+\gamma_{10} \text{ ct}_{ij} + \\
+u_{0j} + 
+u_{1j} \text{ age}_{ij} + 
+u_{1j} \text{ moter\_score}_{ij} +  
+u_{1j} \text{ pupil}_{ij} +  
+u_{1j} \text{ ct}_{ij} +  
+e_{ij} \\
+\]
+
+\texttt{glmer(mort\ \textasciitilde{}\ 1\ +\ type\ +\ age\ +\ motor\_score\ +\ pupil\ +\ ct\ +\ hypox\ +\ hypots\ +\ tsah\ +\ edh\ +\ (1\ \textbar{}\ name))}
 
@@
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_md-1} \end{center}
 
-\begin{center}\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-3-2} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_md-2} \end{center}
 
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
+\end{CodeInput}
+\begin{CodeOutput}
+Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, : Model is nearly unidentifiable: very large eigenvalue
+ - Rescale variables?
+\end{CodeOutput}
+\begin{CodeInput}
+R> mod_cca %>% broom.mixed::tidy()
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 9 x 7
+  effect   group term            estimate std.error statistic   p.value
+  <chr>    <chr> <chr>              <dbl>     <dbl>     <dbl>     <dbl>
+1 fixed    <NA>  (Intercept)      -2.66     0.194      -13.7   1.11e-42
+2 fixed    <NA>  typeRCT          -0.412    0.172       -2.40  1.64e- 2
+3 fixed    <NA>  age               0.0298   0.00190     15.7   2.56e-55
+4 fixed    <NA>  motor_score3     -0.631    0.0823      -7.66  1.82e-14
+5 fixed    <NA>  motor_score4     -0.994    0.0845     -11.8   5.61e-32
+6 fixed    <NA>  motor_score5/6   -1.54     0.0868     -17.7   2.88e-70
+7 fixed    <NA>  pupil             0.438    0.0397      11.0   3.01e-28
+8 fixed    <NA>  ct                0.430    0.0343      12.5   5.34e-36
+9 ran_pars name  sd__(Intercept)   0.257   NA           NA    NA       
+\end{CodeOutput}
+\begin{CodeInput}
+R> # use glm with name as dummy
+R> # only random effect on ct, not the other variables
+R> # make forest plot of ct variable
+R> # is there heterogen in association of ct with mort?
+R> # is there any random effect on that variable at all?
+R> # 
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred <- quickpred(impact_NA)
+R> pred[pred == 1] <- 2
+R> pred[, "name"] <- -2
+R> pred["mort", ] <- 2
+R> pred[, "mort"] <- 2
+R> diag(pred) <- 0
+R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
+R> pred
+\end{CodeInput}
+\begin{CodeOutput}
+            name type age motor_score pupil ct hypox hypots tsah edh mort
+name           0    0   0           0     0  0     0      0    0   0    0
+type           0    0   0           0     0  0     0      0    0   0    0
+age            0    0   0           0     0  0     0      0    0   0    0
+motor_score    0    0   0           0     0  0     0      0    0   0    0
+pupil         -2    0   0           2     0  2     0      2    0   0    2
+ct            -2    2   2           2     2  0     0      0    2   2    2
+hypox         -2    2   0           2     0  0     0      2    0   0    2
+hypots        -2    0   0           2     2  0     2      0    0   0    2
+tsah          -2    0   2           0     0  2     0      0    0   0    2
+edh           -2    0   0           0     0  2     0      0    0   0    2
+mort           0    0   0           0     0  0     0      0    0   0    0
+\end{CodeOutput}
+\begin{CodeInput}
+R> meth <- make.method(impact_NA)
+R> meth[meth != ""] <- "2l.pmm"
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # imp_impact <- impact_NA %>%
+R> #   mutate(name = as.integer(name), motor_score = as.numeric(motor_score)) %>%
+R> #   mice::mice(., m = 2, maxit = 1, method = meth, predictorMatrix = pred)
+R> # look at jomo for categorical variables?
+R> # semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach
+R> # pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters
+R> # make assumptions of these methods explicit!
+\end{CodeInput}
 \end{CodeChunk}

### 2022-03-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

@@
 
-In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.
-For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses.
-Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and bias results.
-Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
-The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster [@hox17].
+~~In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.~~ Contemporary decision making is driven increasingly often and sometimes even entirely dictated by data.
+Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale.
+This not only requires to collect larger amounts of data, but also from more diverse settings and populations.
+For example, [TD: introduce IPDMA here; eg.
+HIV studies Johanna is referring to].
+Another example is the use of large databases with information from thousands or even millions of individuals from multiple cities, regions or even countries.
+Finally, a third situation arises when data are collected from multicenter studies [TD: maybe refer to the schools/classes example here].
+For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses in medical research.
+A common characteristic in aforementioned examples is the presence of clustering.
+[TD: briefly explain what is clustering?]
+
+Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and introduce bias in estimators [@hox17]. 
+~~Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
+The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster.~~
 These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
-The variability due to clustering is often measured by means of the intraclass coefficient [ICC, @fish25].
-The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.
+~~The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
+The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.~~
 Multilevel models typically accommodate for this variability by including a separate group mean for each cluster.
@@
 Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
+[TD: emphasize when/why we need to account for clustering in the analysis of clustered data.
+Why is the presence of clustering relevant when considering multiple imputation of missing data?
+e.g.
+distinction between systematically and sporadically missing data.
+But also: mechanisms of missing data (e.g.
+MCAR, MAR, MNAR) may differ between clusters.
+But also: relation between observed data may differ between clusters?
+When/ why should we avoid using traditional imputation methods?
+e.g.
+congeniality issues.]
 
@@
 But with multiple levels of data comes the potential for clustered missingness.
-Incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13].
+Therefore, incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13].
 Systematic missingness implies that one or more variables are never observed in a certain cluster.
@@
 We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns.
-Column $X1$ is completely observed, column $X2$ is systematically missing, and column $X3$ is sporadically missing.
+Column $X1$ is completely observed, column $X2$ is systematically missing in cluster 2, and column $X3$ is sporadically missing.
 To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account.
@@
 We could then just extrapolate the true (but missing) value of $X3$ for unit 1 from unit 2, and the value for unit 4 from unit 3.
-If $X3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but need some kind of missing data method or discard the incomplete units altogether (i.e., list-wise deletion/complete case analysis).
-And, with the systematic missingness in $X2$, it is impossible to fit a multilevel model since we cannot estimate the intercept of cluster 2.
+If $X3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method or discard the incomplete units altogether (i.e., list-wise deletion/complete case analysis).
+Further, with the systematic missingness in $X2$, it is impossible to fit a multilevel model since we cannot estimate the intercept of cluster 2.
 We would have to exclude this cluster from our analyses entirely to obtain any results.
@@
 Especially the former is very generic and popular.
-Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analyzed as if it were completely observed.
+[VJ: add why multiple values are imputed.] Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analyzed as if it were completely observed.
 The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
@@
 
-- `popmis` from the \pkg{mice} package (simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools);
-- `hiv` from the \pkg{GJRM} package (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions);
-- `impact` from the \pkg{metamisc} package (empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies).
+- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, @mice];
+- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, @GJRM];
+- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc].
 
 For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
-With the `popmis` data, we show how (not) to develop an imputation model.
+With the `popmis` data, we show how (and how not) to develop an imputation model.
 With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods.
-And with the `impact` data we provide an example of multivariate missingness in real-world data.
+With the `impact` data we provide an example of multivariate missingness in real-world data.
 Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.
@@
 <!-- First and second level predictors with random slopes and cross-level interaction: lmer(formula = popular ~ 1 + sex + texp + sex:texp + (1 + sex | class)) -->
+
+TODO: make a 'translation table' between model equation and lme4. Use betas instead of gamma's and mu's.
 
@@
 
+# use glm with name as dummy
+# only random effect on ct, not the other variables
+# make forest plot of ct variable
+# is there heterogen in association of ct with mort?
+# is there any random effect on that variable at all?
+# 
 ```
@@
 ```{r impact_pred}
-quickpred(impact_NA)
+pred <- quickpred(impact_NA)
+pred[pred == 1] <- 2
+pred[, "name"] <- -2
+pred["mort", ] <- 2
+pred[, "mort"] <- 2
+diag(pred) <- 0
+pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
+pred
+
+meth <- make.method(impact_NA)
+meth[meth != ""] <- "2l.pmm"
+```
+
+```{r imp_impact}
+# imp_impact <- impact_NA %>%
+#   mutate(name = as.integer(name), motor_score = as.numeric(motor_score)) %>%
+#   mice::mice(., m = 2, maxit = 1, method = meth, predictorMatrix = pred)
+# look at jomo for categorical variables?
+# semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach
+# pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters
+# make assumptions of these methods explicit!
 ```

### 2022-03-07 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Feedback md pattern plot

@@
 ```
+
+<!-- Missing data pattern with clustering -->
+<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
+<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
+<!-- TODO: check spss missing data pie chart? -->
+<!-- TODO: similarity of patterns over clusters -->
+<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
+<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
+<!-- TODO: look at vim -->
 
@@
 We won't go into detail for the different methods (and equations).
-This paper is just a software tutorial, so we'll keep it practical.] Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the \pkg{magrittr} package, and the \pkg{lme4} notation for multilevel models.^[Add environment info, seed and version number(s) somewhere!]
+This paper is just a software tutorial, so we'll keep it practical.] Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the \pkg{magrittr} package, and the \pkg{lme4} notation for multilevel models.^[TODO: Add environment info, seed and version number(s) somewhere!]
 
@@
 With the `impact` data we provide an example of multivariate missingness in real-world data.
-Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.
+Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph, linking multilevel equations to \pkg{lme4} formulas.]
 
@@
 est_comp_fixed %>% setNames(c("", "Estimate with 95% CI"))
+# TODO: make this a figure with the different methods as facets
 ```
@@
 meth <- ini$meth
-meth <- c("", "2l.norm", "",  "")
-# meth <- c("", "2l.pmm", "",  "")
-imp_norm_2l <-
+meth["popular"] <- "2l.pmm"
+imp_pmm_2l <-
   mice(
@@
 # plot(imp_norm)
-plot_imps(imp_norm_2l, x = "popular", type = "stripplot")
+plot_imps(imp_pmm_2l, x = "popular", type = "stripplot")
 
@@
   - `ct` Marshall Computerized Tomography classification,
-  <!-- - `hypox` Hypoxia (0=no, 1=yes), -->
-  <!-- - `hypots` Hypotension (0=no, 1=yes), -->
-  <!-- - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes), -->
-  <!-- - `edh` Epidural hematoma (0=no, 1=yes), -->
+  - `hypox` Hypoxia (0=no, 1=yes),
+  - `hypots` Hypotension (0=no, 1=yes),
+  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+  - `edh` Epidural hematoma (0=no, 1=yes),
   - `mort` 6-month mortality (0=alive, 1=dead).
@@
 
-`glmer(mort ~ 1 + type + age + motor_score + pupil + ct + hypox + hypots + tsah + edh + (1 | name))`
+`glmer(mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name))`
+
+<!-- + hypox + hypots + tsah + edh  -->
 
@@
 data("impact", package = "metamisc")
-impact_NA <- readRDS("../Data/impact_NA.RDS") 
+impact_NA <- readRDS("../Data/impact_NA.RDS") #%>% select(!c(hypox, hypots, tsah, edh))
+
 # # descriptive statistics
@@
 
-```{r impact_cca}
-mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
-mod_cca %>% broom.mixed::tidy()
+```{r impact_cca, echo=FALSE}
+# mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
+mod_cca_dummy <- glm(mort ~ type + age + motor_score + pupil + ct + as.factor(name), family = "binomial", data = impact_NA) 
+mod_cca_dummy %>% broom.mixed::tidy()
 
@@
 ```
+
+If we run the multilevel model on the incomplete date, we get a warning about unindentifyability.
+This means that with CCA as missing data method, we cannot trust the estimates.
+To still obtain some estimates to compare later, we fit a simplified analysis model: using the cluster variable as indicator in the model.
 

### 2022-03-07 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Feedback md pattern plot

@@
 \pkg{magrittr} package, and the \pkg{lme4} notation for multilevel
-models.\footnote{Add environment info, seed and version number(s)
+models.\footnote{TODO: Add environment info, seed and version number(s)
   somewhere!}
@@
 should give enough scaffolding for applied researchers who are faced
-with incomplete multilevel data.
+with incomplete multilevel data.\footnote{TODO: Add notation paragraph,
+  linking multilevel equations to \pkg{lme4} formulas.}
 
@@
 R> meth <- ini$meth
-R> meth <- c("", "2l.norm", "",  "")
-R> # meth <- c("", "2l.pmm", "",  "")
-R> imp_norm_2l <-
+R> meth["popular"] <- "2l.pmm"
+R> imp_pmm_2l <-
 +   mice(
@@
 \begin{CodeChunk}
-\begin{CodeOutput}
-Warning: Removed 848 rows containing non-finite values (stat_boxplot).
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning: Removed 848 rows containing missing values (geom_point).
-\end{CodeOutput}
 
@@
 \item
+  \texttt{hypox} Hypoxia (0=no, 1=yes),
+\item
+  \texttt{hypots} Hypotension (0=no, 1=yes),
+\item
+  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+\item
+  \texttt{edh} Epidural hematoma (0=no, 1=yes),
+\item
   \texttt{mort} 6-month mortality (0=alive, 1=dead).
@@
 
-\texttt{glmer(mort\ \textasciitilde{}\ 1\ +\ type\ +\ age\ +\ motor\_score\ +\ pupil\ +\ ct\ +\ hypox\ +\ hypots\ +\ tsah\ +\ edh\ +\ (1\ \textbar{}\ name))}
+\texttt{glmer(mort\ \textasciitilde{}\ 1\ +\ type\ +\ age\ +\ motor\_score\ +\ pupil\ +\ ct\ +\ (1\ \textbar{}\ name))}
 
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
-\end{CodeInput}
 \begin{CodeOutput}
-Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, : Model is nearly unidentifiable: very large eigenvalue
- - Rescale variables?
+# A tibble: 22 x 5
+   term                 estimate std.error statistic  p.value
+   <chr>                   <dbl>     <dbl>     <dbl>    <dbl>
+ 1 (Intercept)           -3.37     0.205     -16.4   1.97e-60
+ 2 typeRCT                0.395    0.177       2.23  2.59e- 2
+ 3 age                    0.0301   0.00191    15.8   5.01e-56
+ 4 motor_score3          -0.618    0.0829     -7.45  9.33e-14
+ 5 motor_score4          -1.00     0.0850    -11.8   6.07e-32
+ 6 motor_score5/6        -1.53     0.0874    -17.5   7.92e-69
+ 7 pupil                  0.434    0.0398     10.9   9.36e-28
+ 8 ct                     0.428    0.0344     12.4   1.50e-35
+ 9 as.factor(name)CSTAT  -0.177    0.199      -0.892 3.72e- 1
+10 as.factor(name)EBIC    0.560    0.174       3.22  1.30e- 3
+# ... with 12 more rows
 \end{CodeOutput}
-\begin{CodeInput}
-R> mod_cca %>% broom.mixed::tidy()
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 9 x 7
-  effect   group term            estimate std.error statistic   p.value
-  <chr>    <chr> <chr>              <dbl>     <dbl>     <dbl>     <dbl>
-1 fixed    <NA>  (Intercept)      -2.66     0.194      -13.7   1.11e-42
-2 fixed    <NA>  typeRCT          -0.412    0.172       -2.40  1.64e- 2
-3 fixed    <NA>  age               0.0298   0.00190     15.7   2.56e-55
-4 fixed    <NA>  motor_score3     -0.631    0.0823      -7.66  1.82e-14
-5 fixed    <NA>  motor_score4     -0.994    0.0845     -11.8   5.61e-32
-6 fixed    <NA>  motor_score5/6   -1.54     0.0868     -17.7   2.88e-70
-7 fixed    <NA>  pupil             0.438    0.0397      11.0   3.01e-28
-8 fixed    <NA>  ct                0.430    0.0343      12.5   5.34e-36
-9 ran_pars name  sd__(Intercept)   0.257   NA           NA    NA       
-\end{CodeOutput}
-\begin{CodeInput}
-R> # use glm with name as dummy
-R> # only random effect on ct, not the other variables
-R> # make forest plot of ct variable
-R> # is there heterogen in association of ct with mort?
-R> # is there any random effect on that variable at all?
-R> # 
-\end{CodeInput}
 \end{CodeChunk}
+
+If we run the multilevel model on the incomplete date, we get a warning
+about unindentifyability. This means that with CCA as missing data
+method, we cannot trust the estimates. To still obtain some estimates to
+compare later, we fit a simplified analysis model: using the cluster
+variable as indicator in the model.
 

### 2022-03-07 R/plot_incomplete.R

@@ -104,7 +104,7 @@ plot_md_pat <- function(dat) {
       sec.axis = ggplot2::dup_axis(labels = as.character(NA_per_pat),
                                    name = "Number of missing entries per pattern")
     ) +
-    ggplot2::coord_cartesian(clip = "off") +
+    #ggplot2::coord_cartesian() +
     # add labels
     ggplot2::labs(
       x = "Number of missing entries per variable",
@@ -138,7 +138,9 @@ plot_md_pat <- function(dat) {
       axis.title.y.right = ggplot2::element_text(margin = ggplot2::margin(l = 10)),
       panel.grid = ggplot2::element_blank()
     ) +
-    ggplot2::scale_fill_manual(values = c("1" = mice:::mdc(1), "0" = mice:::mdc(2)))
+    ggplot2::scale_fill_manual(values = c("1" = mice:::mdc(1), "0" = mice:::mdc(2))) +
+    coord_cartesian(clip = "off") 
+    
   # output
   return(p)
 }

### 2022-03-17 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update popularity case study with ggmice functions

@@
     url: https://hanneoberman.github.io/
-  - name: Johanna Mu\~noz Avila
+  - name: "Johanna Mu\\~{n}oz Avila"
     affiliation: University Medical Center Utrecht \AND
@@
 output: 
-    rticles::jss_article: default
-    word_document: default
+    rticles::jss_article:
+      keep_tex: yes
+      number_sections: yes
+    # word_document: default
 bibliography: ../References/multilevelmice.bib
@@
 ```{r, setup, include=FALSE}
+knitr::opts_chunk$set(message=FALSE, warning=FALSE)
 options(prompt = 'R> ', continue = '+ ')
@@
 With the `impact` data we provide an example of multivariate missingness in real-world data.
-Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph, linking multilevel equations to \pkg{lme4} formulas.]
+Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph or 'translation table' linking multilevel equations to \pkg{lme4} formulas.
+Use betas instead of gamma's and mu's.
+Add interpretation of values in predictormatrix (-2 for the cluster variable, 2 for random effects).]
 
@@
 
+Set-up the R environment and load the necessary packages:
+
+```{r}
+set.seed(2022)
+library(mice)
+library(ggmice)
+library(ggplot2)
+library(dplyr)
+library(lme4)
+```
+
 ---
@@
 
-# How (not) to impute (Case Study I: Popularity)
+# Case Study I: How (not) to impute
 
@@
 
-TODO: make a 'translation table' between model equation and lme4. Use betas instead of gamma's and mu's.
-
 Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data.
@@
 ```{r}
-popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
+popmis <- popmis[, c("school", "popular", "sex", "texp")] 
 ```
+Plot the missing data pattern:
 
-```{r pop_pat, echo=FALSE, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
-# missingness
-plot_md_pat(popmis)
+```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
+plot_pattern(popmis)
 ```
@@
 The ICC in the incomplete data is `r round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.
+This tells us that the multilevel structure of the data should probably be taken into account.
+If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.
+
+Plot the correlations in the incomplete data:
+
+```{r pop-corr}
+plot_corr(popmis)
+```
 
 To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables.
-We'll highlight only one other variable to illustrate, but ideally one would inspect all relations.
+We'll highlight one other variable to illustrate, but ideally one would inspect all relations.
 The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'.
@@
 
-In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
+Plot the histogram for teacher experience conditional on the missingness indicator of `popular`:
+
+```{r pop-hist}
+ggmice(popmis, aes(texp)) +
+  geom_histogram(fill = "white") +
+  facet_grid(. ~ is.na(popular), labeller = label_both)
+```
+
+```{r pop-hist-clust, include=FALSE}
+# popmis %>% 
+#   filter(school < 5) %>% 
+# ggmice(., aes(texp)) +
+#   geom_histogram(fill = "white") +
+#   facet_grid(school ~ is.na(popular), labeller = label_both)
+# # does not make sense because texp is a cluster level variable!
+```
+
+This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` ($t = $ `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
+
+<!-- In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
 This would indicate a right-tailed MAR missingness.
 (In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `texp`.
+-->
 
 ```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
+# plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
 # TODO: check if up-side-down plot works
@@
 
-Complete case analysis ignores the observations with missingness altogether, which may even introduce bias in MCAR situations.
+Complete case analysis ignores the observations with missingness altogether, which lowers statistical power and may even introduce bias in MCAR situations.
 
@@
 The first imputation model that we'll use is likely to be invalid.
-In this model, we ignore the multilevel structure of the data, despite the high ICCs.
-And assumes exchangeability between units.
-This is purely to illustrate the effects of ignoring the clustering in our imputation effort.
+We do *not* use the cluster identifier `school` as imputation model predictor.
+With this model, we ignore the multilevel structure of the data, despite the high ICC.
+This assumes exchangeability between units.
+We include it purely to illustrate the effects of ignoring the clustering in our imputation effort.
+We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).
 
-We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`.
-We do not use the observation identifier `pupil` or cluster identifier `school` as predictors to impute other variables.
+Create a predictor matrix and remove `school` as predictor for `popular`:
 
 ```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
-# dry run to get imputation parameters
-ini <- mice(popmis, maxit = 0)
-
-# extract predictor matrix and adjust
-pred <- ini$pred
-pred[, c("school")] <- 0
+# predictor matrix
+pred <- quickpred(popmis)
+plot_pred(pred)
 
@@
 
-As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `school`.
-This tells us that the multilevel structure of the data should be taken into account.
-If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.
-
-We can also observe that the teacher experience increases slightly after imputation.
+<!-- We can also observe that the teacher experience increases slightly after imputation.
 This is due to the MNAR missingness in `texp`.
@@
 true) value for every pupil in the school.
+-->
 
@@
 # adjust the predictor matrix
-pred <- ini$pred 
-# pred[, "pupil"] <- 0
+pred["popular", "school"] <- 1 
+plot_pred(pred)
 
@@
 imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
-
 ```
@@
 ```{r pop_norm}
-pred <- ini$pred
-pred["popular", ] <- c(-2, 2, 2, 2) 
-#-2 for the cluster variable, 2 for random effects
-meth <- ini$meth
+# adjust the predictor matrix
+pred["popular", ] <- c(school = -2, popular = 0, sex = 2, texp = 2) 
+plot_pred(pred) 
+meth <- make.method(popmis)
 meth["popular"] <- "2l.pmm"
@@
 
+- Show use case for deductive imputation for cluster level variables?
+
 - env dump in repo

### 2022-03-17 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update popularity case study with ggmice functions

@@
 ]{jss}
+
+%% recommended packages
+\usepackage{orcidlink,thumbpdf,lmodern}
 
@@
 \author{
-Hanne Oberman\\Utrecht University \And Johanna Mu\textasciitilde noz
+Hanne Oberman\\Utrecht University \And Johanna Mu\textasciitilde\{n\}oz
 Avila\\University Medical Center Utrecht \AND Valentijn de
@@
 
-\Plainauthor{Hanne Oberman, Johanna Mu\textasciitilde noz
+\Plainauthor{Hanne Oberman, Johanna Mu\textasciitilde\{n\}oz
 Avila, Valentijn de Jong, Gerko Vink, Thomas Debray}
@@
 should give enough scaffolding for applied researchers who are faced
-with incomplete multilevel data.\footnote{TODO: Add notation paragraph,
-  linking multilevel equations to \pkg{lme4} formulas.}
+with incomplete multilevel data.\footnote{TODO: Add notation paragraph
+  or `translation table' linking multilevel equations to \pkg{lme4}
+  formulas. Use betas instead of gamma's and mu's. Add interpretation of
+  values in predictormatrix (-2 for the cluster variable, 2 for random
+  effects).}
+
+Set-up the R environment and load the necessary packages:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> set.seed(2022)
+R> library(mice)
+R> library(ggmice)
+R> library(ggplot2)
+R> library(dplyr)
+R> library(lme4)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-\hypertarget{how-not-to-impute-case-study-i-popularity}{%
-\section{How (not) to impute (Case Study I:
-Popularity)}\label{how-not-to-impute-case-study-i-popularity}}
+\hypertarget{case-study-i-how-not-to-impute}{%
+\section{Case Study I: How (not) to
+impute}\label{case-study-i-how-not-to-impute}}
 
@@
 
-TODO: make a `translation table' between model equation and lme4. Use
-betas instead of gamma's and mu's.
-
 Since the data is simulated and the missingness is induced, we can
@@
 \begin{CodeInput}
-R> popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
+R> popmis <- popmis[, c("school", "popular", "sex", "texp")] 
 \end{CodeInput}
@@
 
+Plot the missing data pattern:
+
 \begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(popmis)
+\end{CodeInput}
 \begin{figure}
@@
 missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the
-incomplete data is 0.56.
+incomplete data is 0.56. This tells us that the multilevel structure of
+the data should probably be taken into account. If we don't, we'll may
+end up with incorrect imputations, biasing the effect of the clusters
+towards zero.
+
+Plot the correlations in the incomplete data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_corr(popmis)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-corr-1} \end{center}
+
+\end{CodeChunk}
 
@@
 \texttt{popular}, we need to know whether the missingness depends on the
-observed values of other variables. We'll highlight only one other
-variable to illustrate, but ideally one would inspect all relations. The
+observed values of other variables. We'll highlight one other variable
+to illustrate, but ideally one would inspect all relations. The
 questions we'll ask are: `Does the missing data of pupil popularity
@@
 
-In Figure \ref{fig:pop_dist} we see that \textbf{{[}update this part{]}}
-the distribution for the missing \texttt{popular} is further to the
-right than the distribution for observed \texttt{popular}. This would
-indicate a right-tailed MAR missingness. (In fact, this is exactly what
-happens, because the missingness in these data was created manually.)
-We've made it observable by examining the relations between the
-missingness in popular and the observed data in \texttt{texp}.
+Plot the histogram for teacher experience conditional on the missingness
+indicator of \texttt{popular}:
 
 \begin{CodeChunk}
-\begin{figure}
+\begin{CodeInput}
+R> ggmice(popmis, aes(texp)) +
++   geom_histogram(fill = "white") +
++   facet_grid(. ~ is.na(popular), labeller = label_both)
+\end{CodeInput}
 
-{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_dist-1}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-hist-1} \end{center}
 
-}
+\end{CodeChunk}
 
-\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist}
-\end{figure}
-\end{CodeChunk}
+This shows us that there are no apparent differences in the distribution
+of \texttt{texp} depending on the missingness indicator of
+\texttt{popular} (\$t = \$ -0.873, p = 0.383).
 
@@
 Complete case analysis ignores the observations with missingness
-altogether, which may even introduce bias in MCAR situations.
+altogether, which lowers statistical power and may even introduce bias
+in MCAR situations.
 
@@
 
-The first imputation model that we'll use is likely to be invalid. In
-this model, we ignore the multilevel structure of the data, despite the
-high ICCs. And assumes exchangeability between units. This is purely to
-illustrate the effects of ignoring the clustering in our imputation
-effort.
+The first imputation model that we'll use is likely to be invalid. We do
+\emph{not} use the cluster identifier \texttt{school} as imputation
+model predictor. With this model, we ignore the multilevel structure of
+the data, despite the high ICC. This assumes exchangeability between
+units. We include it purely to illustrate the effects of ignoring the
+clustering in our imputation effort. We'll use the default imputation
+methods in \texttt{mice()} (predictive mean matching to impute the
+continuous variables and logistic regression to impute binary
+variables).
 
-We'll use predictive mean matching to impute the continuous variables
-and logistic regression to impute the binary variable \texttt{sex}. We
-do not use the observation identifier \texttt{pupil} or cluster
-identifier \texttt{school} as predictors to impute other variables.
+Create a predictor matrix and remove \texttt{school} as predictor for
+\texttt{popular}:
 
@@
 \begin{CodeInput}
-R> # dry run to get imputation parameters
-R> ini <- mice(popmis, maxit = 0)
-R> 
-R> # extract predictor matrix and adjust
-R> pred <- ini$pred
-R> pred[, c("school")] <- 0
-R> 
+R> # predictor matrix
+R> pred <- quickpred(popmis)
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_ignored-1} \end{center}
+
+\begin{CodeInput}
 R> # impute the data, ignoring the cluster structure
@@
 \end{CodeChunk}
-
-As the original ICCs show, 100\% of the variance in \texttt{texp} can be
-attributed to the clustering variable \texttt{school}. This tells us
-that the multilevel structure of the data should be taken into account.
-If we don't, we'll end up with incorrect imputations, biasing the effect
-of the clusters towards zero.
-
-We can also observe that the teacher experience increases slightly after
-imputation. This is due to the MNAR missingness in \texttt{texp}. Higher
-values for \texttt{texp} have a larger probability to be missing. This
-may not a problem, however, if at least one pupil in each school has
-teacher experience recorded, we can deductively impute the correct
-(i.e.~true) value for every pupil in the school.
 
@@
 R> # adjust the predictor matrix
-R> pred <- ini$pred 
-R> # pred[, "pupil"] <- 0
-R> 
+R> pred["popular", "school"] <- 1 
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor-1} \end{center}
+
+\begin{CodeInput}
 R> # impute the data, cluster as predictor
@@
 \begin{CodeInput}
-R> pred <- ini$pred
-R> pred["popular", ] <- c(-2, 2, 2, 2) 
-R> #-2 for the cluster variable, 2 for random effects
-R> meth <- ini$meth
+R> # adjust the predictor matrix
+R> pred["popular", ] <- c(school = -2, popular = 0, sex = 2, texp = 2) 
+R> plot_pred(pred) 
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_norm-1} \end{center}
+
+\begin{CodeInput}
+R> meth <- make.method(popmis)
 R> meth["popular"] <- "2l.pmm"
@@
 \item
+  Show use case for deductive imputation for cluster level variables?
+\item
   env dump in repo

### 2022-04-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Estimate multilevel model after imputation

@@
 
+<!-- Comments Thomas: -->
+<!-- .	We need to remind the reader about general knowledge about missing data: it usually leads to loss of power and bias; we distinguish between MCAR, MAR and MNAR in theory but in practice this distinction is less clear; multiple imputation is generally recommended to avoid bias (if data are MCAR or MAR, but also works reasonable in some situations where data are partially MNAR) -->
+<!-- .	We need to clarify what is clustered data, and provide common types (IPDMA, registries with data from multiple hospitals/regions, and multicenter studies).
+You can directly refer to the examples you will use in the paper here, and build on the text you already had.
+-->
+<!-- .	We need to clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implemention ("though shall not simply run mice() on any incomplete dataset") -->
+<!-- .	Maybe we should also emphasise that in more modern types of big data sources (E.g. EHR data ), measurement of data is often selective and MAR assumptions are less tenable.  -->
+
 # Introduction
 
-~~In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.~~ Contemporary decision making is driven increasingly often and sometimes even entirely dictated by data.
+~~In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.~~ Contemporary decision-making is  increasingly often data-driven and sometimes even entirely dictated by data.
 Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale.
@@
 Use betas instead of gamma's and mu's.
-Add interpretation of values in predictormatrix (-2 for the cluster variable, 2 for random effects).]
+Add interpretation of values in predictormatrix (-2 for the cluster variable, 2 for random effects).
+Add ICC and congeneality here as well.
+And make missingness mechanism table as well.]
 
@@
 library(lme4)
+library(mitml)
 ```
@@
 
+TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?
+
 ```{r pop-hist-clust, include=FALSE}
@@
 
-This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` ($t = $ `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
+This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
 
@@
 
+```{r pop-cca, echo=FALSE}
+est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis)
+
+```
+
 ### Imputation ignoring the cluster variable (not recommended)
@@
 
-Create a predictor matrix and remove `school` as predictor for `popular`:
+Create a methods vector and predictor matrix for `popular`, and make sure `school` is not included as predictor:
 
-```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
-# predictor matrix
-pred <- quickpred(popmis)
+```{r pop-ignored-pred, echo=TRUE, message=FALSE, warning=FALSE}
+meth <- make.method(popmis) # methods vector
+pred <- quickpred(popmis)   # predictor matrix
 plot_pred(pred)
+```
 
-# impute the data, ignoring the cluster structure
+Impute the data, ignoring the cluster structure:
+```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
 imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+```
+
+Analyze the imputations:
+```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
+fit_ignored <- imp_ignored %>% 
+  with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+fit_ignored %>% 
+  pool() %>% 
+  broom.mixed::tidy()
+testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
 ```

### 2022-04-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Estimate multilevel model after imputation

@@
 \sout{In many contemporary data analysis efforts, some form of
-hierarchical or clustered structure is recorded.} Contemporary decision
-making is driven increasingly often and sometimes even entirely dictated
-by data. Although data relevant for decision makers were initially
-collected using small and well-designed studies, there has been a
-growing need to address more complex questions at a wider scale. This
-not only requires to collect larger amounts of data, but also from more
-diverse settings and populations. For example, {[}TD: introduce IPDMA
-here; eg. HIV studies Johanna is referring to{]}. Another example is the
-use of large databases with information from thousands or even millions
-of individuals from multiple cities, regions or even countries. Finally,
-a third situation arises when data are collected from multicenter
-studies {[}TD: maybe refer to the schools/classes example here{]}. For
-example, students may be clustered in classes in psychometrics research,
-or patients may be clustered in studies in individual patient data
-meta-analyses in medical research. A common characteristic in
-aforementioned examples is the presence of clustering. {[}TD: briefly
-explain what is clustering?{]}
+hierarchical or clustered structure is recorded.} Contemporary
+decision-making is increasingly often data-driven and sometimes even
+entirely dictated by data. Although data relevant for decision makers
+were initially collected using small and well-designed studies, there
+has been a growing need to address more complex questions at a wider
+scale. This not only requires to collect larger amounts of data, but
+also from more diverse settings and populations. For example, {[}TD:
+introduce IPDMA here; eg. HIV studies Johanna is referring to{]}.
+Another example is the use of large databases with information from
+thousands or even millions of individuals from multiple cities, regions
+or even countries. Finally, a third situation arises when data are
+collected from multicenter studies {[}TD: maybe refer to the
+schools/classes example here{]}. For example, students may be clustered
+in classes in psychometrics research, or patients may be clustered in
+studies in individual patient data meta-analyses in medical research. A
+common characteristic in aforementioned examples is the presence of
+clustering. {[}TD: briefly explain what is clustering?{]}
 
@@
   values in predictormatrix (-2 for the cluster variable, 2 for random
-  effects).}
+  effects). Add ICC and congeneality here as well. And make missingness
+  mechanism table as well.}
 
@@
 
+TODO: think about what is a meaningful rule of thumb to signal that the
+user should be worried?
+
 This shows us that there are no apparent differences in the distribution
 of \texttt{texp} depending on the missingness indicator of
-\texttt{popular} (\$t = \$ -0.873, p = 0.383).
+\texttt{popular} (t = -0.873, p = 0.383).
 
@@
 in MCAR situations.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis) 
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_ignored-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-1} \end{center}
 

### 2022-04-01 MS_treat_effect.R

@@ -1,0 +1,359 @@
+## code to prepare `countExample` dataset
+
+library(tidyverse, warn.conflicts = FALSE)
+library(MASS)
+library(truncnorm)
+library(magrittr)
+library(reshape2)
+library(fastDummies)
+library(dplyr)
+library(optmatch) #match
+library(MatchIt)
+library(WeightIt) #IPW
+library(cobalt) # imbalance covariates
+library(ggdag)
+library(dagitty)
+library(mice)
+library(data.table)
+library(PSweight)
+library(sandwich) 
+library(lmtest)
+
+# Function to simulate MS data
+simcountdata <- function(n, seed = 999,
+                         beta = c(-0.5, -0.25, 0, 0.25, 0.5),
+                         beta.x = c(-1.54, -0.01, 0.06, 0.25, 0.5, 0.13, 0.0000003),
+                         percentiles = seq(0, 1, by = 0.2)){
+  
+  #' Generate simulated count data with settings based on real-world data
+  #' Assume randomized treatment and independent covariates
+  #'
+  #' @param n sample size; integer
+  #' @param seed randomization seed; integer
+  #' @param beta coefficients characterizing treatment effect heterogeneity; vector of length 5
+  #'             beta[1]*trt*I(high responder to DMF) +
+  #'             beta[2]*trt*I(moderate responder to DMF) +
+  #'             beta[3]*trt*I(neutral) +
+  #'             beta[4]*trt*I(moderate responder to TERI) +
+  #'             beta[5]*trt*I(high responder to TERI)
+  #'             In the absence of treatment effect heterogeneity, set all beta[1:5] with the same values
+  #' @param beta.x coefficients for main effects of other covariates in the rate; vector of length 7
+  #'               beta.x[1] (intercept)
+  #'               beta.x[2]*ageatindex_centered
+  #'               beta.x[3]*female
+  #'               beta.x[4]*prerelapse_num
+  #'               beta.x[5]*prevDMTefficacy== mediumhigh (reference low efficacy)
+  #'               beta.x[6]*prevDMTefficacy== none (reference low efficacy)
+  #'               beta.x[7]*premedicalcost
[…]

### 2022-04-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Impute impact data

@@
 
+TODO: remove the broom.mixed output, use mitml only
+
 Analyze the imputations:
@@
   with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
-fit_ignored %>% 
-  pool() %>% 
-  broom.mixed::tidy()
 testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
+```
+```{r eval=FALSE, include=FALSE}
+# fit_ignored %>% 
+#   pool() %>% 
+#   broom.mixed::tidy()
 ```
@@
 data("impact", package = "metamisc")
-impact_NA <- readRDS("../Data/impact_NA.RDS") #%>% select(!c(hypox, hypots, tsah, edh))
+impact_NA <- readRDS("../Data/impact_NA.RDS") %>% 
+  mutate(name = as.integer(name),
+         motor_score = as.numeric(motor_score))#%>% select(!c(hypox, hypots, tsah, edh))
 
@@
 # # missingness
-plot_md_pat(impact_NA) 
-# TODO: rotate variable labels
+plot_pattern(impact_NA, rotate = TRUE) 
 # TODO: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters like shepley plot
@@
 ```{r impact_md, echo=FALSE}
-# md pattern in first cluster
-plot_md_pat(impact_NA[impact_NA$name == "TINT", ])
-plot_md_pat(impact_NA[impact_NA$name == "APOE", ])
+# # md pattern in first cluster
+# plot_pattern(impact_NA[impact_NA$name == "TINT", ])
+# plot_pattern(impact_NA[impact_NA$name == "APOE", ])
 ```
@@
 ```{r impact_cca, echo=FALSE}
-# mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
-mod_cca_dummy <- glm(mort ~ type + age + motor_score + pupil + ct + as.factor(name), family = "binomial", data = impact_NA) 
-mod_cca_dummy %>% broom.mixed::tidy()
+mod_cca <- glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
+mod_cca %>% broom.mixed::tidy()
+# mod_cca_dummy <- glm(mort ~ type + age + motor_score + pupil + ct + name, family = "binomial", data = impact_NA) 
+# mod_cca_dummy %>% broom.mixed::tidy()
 
@@
 
-If we run the multilevel model on the incomplete date, we get a warning about unindentifyability.
+<!-- If we run the multilevel model on the incomplete date, we get a warning about unindentifyability.
 This means that with CCA as missing data method, we cannot trust the estimates.
 To still obtain some estimates to compare later, we fit a simplified analysis model: using the cluster variable as indicator in the model.
+-->
 
-```{r impact_pred}
+We should impute the variables `ct` and `pupil` and any auxiliary variables we might want to use to impute these incomplete analysis model variables.
+We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations:
+
+```{r impact_corr}
+plot_corr(impact_NA, label = T)
+```
+
+This shows us that `hypox` and `hypot` would not be useful auxiliary variables for imputing `ct`.
+Depending on the minimum required correlation, `tsah` could be useful, while `edh` has the strongest correlation with `ct` out of all the variables in the data and should definitely be included in the imputation model.
+For the imputation of `pupil`, none of the potential auxiliary variables has a very strong relation, but `hypots` could be used.
+We conclude that we can exclude `hypox` from the data since this is neither an analysis model variable, nor an auxiliary variable for imputation.
+
+```{r impact-pred}
+impact_NA <- select(impact_NA, -c(hypox, hypots))
 pred <- quickpred(impact_NA)
+plot_pred(pred)
+
 pred[pred == 1] <- 2
-pred[, "name"] <- -2
 pred["mort", ] <- 2
 pred[, "mort"] <- 2
+pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
+pred[, "name"] <- -2
 diag(pred) <- 0
-pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
-pred
+plot_pred(pred)
 
@@
 ```{r imp_impact}
+imp <- mice(impact_NA, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
 # imp_impact <- impact_NA %>%
@@
 # make assumptions of these methods explicit!
+```
+```{r impact-fit}
+fit <- imp_impact_pmm %>% 
+  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
+tidy(pool(fit))
+as.mitml.result(fit)
+# testEstimates(as.mitml.result(fit))
+```
+
+Compare estimates to complete data
+
+```{r}
+fit <- glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial", data = impact) 
+tidy(fit)
 ```

### 2022-04-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Impute impact data

@@
 R> library(lme4)
+R> library(mitml)
 \end{CodeInput}
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis) 
-\end{CodeInput}
-\end{CodeChunk}
-
 \hypertarget{imputation-ignoring-the-cluster-variable-not-recommended}{%
@@
 
-Create a predictor matrix and remove \texttt{school} as predictor for
-\texttt{popular}:
+Create a methods vector and predictor matrix for \texttt{popular}, and
+make sure \texttt{school} is not included as predictor:
 
@@
 \begin{CodeInput}
-R> # predictor matrix
-R> pred <- quickpred(popmis)
+R> meth <- make.method(popmis) # methods vector
+R> pred <- quickpred(popmis)   # predictor matrix
 R> plot_pred(pred)
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-pred-1} \end{center}
 
+\end{CodeChunk}
+
+Impute the data, ignoring the cluster structure:
+
+\begin{CodeChunk}
 \begin{CodeInput}
-R> # impute the data, ignoring the cluster structure
 R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
 \end{CodeInput}
+\end{CodeChunk}
+
+TODO: remove the broom.mixed output, use mitml only
+
+Analyze the imputations:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit_ignored <- imp_ignored %>% 
++   with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+R> testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+
+testEstimates(model = as.mitml.result(fit_ignored), var.comp = TRUE)
+
+Final parameter estimates and inferences obtained from 5 imputed data sets.
+
+             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
+(Intercept)     3.354     0.245    13.704     6.831     0.000     3.260     0.813 
+sex             1.247     0.332     3.759     6.154     0.009     4.160     0.849 
+texp            0.110     0.017     6.320     6.084     0.001     4.286     0.852 
+sex:texp       -0.027     0.022    -1.225     5.914     0.267     4.631     0.862
+
+                            Estimate 
+Intercept~~Intercept|school    0.177 
+sex~~sex|school                0.232 
+Intercept~~sex|school         -0.043 
+Residual~~Residual             0.670 
+ICC|school                     0.209
+
+Unadjusted hypothesis test as appropriate in larger samples.
+\end{CodeOutput}
 \end{CodeChunk}

### 2022-04-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
New version of figure with incomplete dataset

@@
 
-```{r patterns, fig.height=3, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
-plot_md_set()
+```{r patterns, fig.height=2.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
+# blankcol <- c(rep("", 4), "...", "")
+# dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
+#   cbind(text = c("1", "1", "2", "2", "...", "N", rep(blankcol, 3), rep("...", 6), blankcol),
+#         miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))
+dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
+  cbind(text = c("1", "1", "2", "2", "", "N", rep("", 30)),
+        miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))
+
+ggplot(dat, aes(x = cols, y = rows)) +
+  geom_tile(fill = "white", color = "black") +
+  geom_text(aes(label = text), color = "black") +
+  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
+  scale_x_continuous(breaks = 1:6, labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])), name = NULL, position = "top") +
+  scale_y_continuous(breaks = 1:6, labels = c(1:4, "...", "n"), name = NULL, trans = "reverse") +
+  theme_minimal() +
+  theme(panel.grid = element_blank())
+
 ```
@@
 
-```{r imp_impact}
+```{r imp_impact, eval = FALSE}
 imp <- mice(impact_NA, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
@@
 ```
-```{r impact-fit}
+```{r impact-fit, eval = FALSE}
 fit <- imp_impact_pmm %>% 

### 2022-04-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
New version of figure with incomplete dataset

@@
 \begin{CodeChunk}
+\begin{CodeOutput}
+# A tibble: 9 x 7
+  effect   group term                    estimate std.error statistic   p.value
+  <chr>    <chr> <chr>                      <dbl>     <dbl>     <dbl>     <dbl>
+1 fixed    <NA>  (Intercept)              -2.66     0.194      -13.7   1.11e-42
+2 fixed    <NA>  typeRCT                  -0.412    0.172       -2.40  1.64e- 2
+3 fixed    <NA>  age                       0.0298   0.00190     15.7   2.56e-55
+4 fixed    <NA>  as.factor(motor_score)2  -0.631    0.0823      -7.66  1.82e-14
+5 fixed    <NA>  as.factor(motor_score)3  -0.994    0.0845     -11.8   5.61e-32
+6 fixed    <NA>  as.factor(motor_score)4  -1.54     0.0868     -17.7   2.88e-70
+7 fixed    <NA>  pupil                     0.438    0.0397      11.0   3.01e-28
+8 fixed    <NA>  ct                        0.430    0.0343      12.5   5.34e-36
+9 ran_pars name  sd__(Intercept)           0.257   NA           NA    NA       
+\end{CodeOutput}
+\end{CodeChunk}
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_md-1} \end{center}
+We should impute the variables \texttt{ct} and \texttt{pupil} and any
+auxiliary variables we might want to use to impute these incomplete
+analysis model variables. We can evaluate which variables may be useful
+auxiliaries by plotting the pairwise complete correlations:
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_md-2} \end{center}
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_corr(impact_NA, label = T)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_corr-1} \end{center}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeOutput}
-# A tibble: 22 x 5
-   term                 estimate std.error statistic  p.value
-   <chr>                   <dbl>     <dbl>     <dbl>    <dbl>
- 1 (Intercept)           -3.37     0.205     -16.4   1.97e-60
- 2 typeRCT                0.395    0.177       2.23  2.59e- 2
- 3 age                    0.0301   0.00191    15.8   5.01e-56
- 4 motor_score3          -0.618    0.0829     -7.45  9.33e-14
- 5 motor_score4          -1.00     0.0850    -11.8   6.07e-32
- 6 motor_score5/6        -1.53     0.0874    -17.5   7.92e-69
- 7 pupil                  0.434    0.0398     10.9   9.36e-28
- 8 ct                     0.428    0.0344     12.4   1.50e-35
- 9 as.factor(name)CSTAT  -0.177    0.199      -0.892 3.72e- 1
-10 as.factor(name)EBIC    0.560    0.174       3.22  1.30e- 3
-# ... with 12 more rows
-\end{CodeOutput}
-\end{CodeChunk}
-
-If we run the multilevel model on the incomplete date, we get a warning
-about unindentifyability. This means that with CCA as missing data
-method, we cannot trust the estimates. To still obtain some estimates to
-compare later, we fit a simplified analysis model: using the cluster
-variable as indicator in the model.
+This shows us that \texttt{hypox} and \texttt{hypot} would not be useful
+auxiliary variables for imputing \texttt{ct}. Depending on the minimum
+required correlation, \texttt{tsah} could be useful, while \texttt{edh}
+has the strongest correlation with \texttt{ct} out of all the variables
+in the data and should definitely be included in the imputation model.
+For the imputation of \texttt{pupil}, none of the potential auxiliary
+variables has a very strong relation, but \texttt{hypots} could be used.
+We conclude that we can exclude \texttt{hypox} from the data since this
+is neither an analysis model variable, nor an auxiliary variable for
+imputation.
 
@@
 \begin{CodeInput}
+R> impact_NA <- select(impact_NA, -c(hypox, hypots))
 R> pred <- quickpred(impact_NA)
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-pred-1} \end{center}
+
+\begin{CodeInput}
 R> pred[pred == 1] <- 2
-R> pred[, "name"] <- -2
 R> pred["mort", ] <- 2
 R> pred[, "mort"] <- 2
+R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
+R> pred[, "name"] <- -2
 R> diag(pred) <- 0
-R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
-R> pred
+R> plot_pred(pred)
 \end{CodeInput}
-\begin{CodeOutput}
-            name type age motor_score pupil ct hypox hypots tsah edh mort
-name           0    0   0           0     0  0     0      0    0   0    0
-type           0    0   0           0     0  0     0      0    0   0    0
-age            0    0   0           0     0  0     0      0    0   0    0
-motor_score    0    0   0           0     0  0     0      0    0   0    0
-pupil         -2    0   0           2     0  2     0      2    0   0    2
-ct            -2    2   2           2     2  0     0      0    2   2    2
-hypox         -2    2   0           2     0  0     0      2    0   0    2
-hypots        -2    0   0           2     2  0     2      0    0   0    2
-tsah          -2    0   2           0     0  2     0      0    0   0    2
-edh           -2    0   0           0     0  2     0      0    0   0    2
-mort           0    0   0           0     0  0     0      0    0   0    0
-\end{CodeOutput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-pred-2} \end{center}
+
 \begin{CodeInput}
@@
 \begin{CodeInput}
+R> imp <- mice(impact_NA, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
 R> # imp_impact <- impact_NA %>%
@@
 \end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- imp_impact_pmm %>% 
++   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
+R> tidy(pool(fit))
+R> as.mitml.result(fit)
+R> # testEstimates(as.mitml.result(fit))
+\end{CodeInput}
+\end{CodeChunk}
+
+Compare estimates to complete data
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial", data = impact) 
+R> tidy(fit)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 11 x 7
+   effect   group term                   estimate std.error statistic    p.value
+   <chr>    <chr> <chr>                     <dbl>     <dbl>     <dbl>      <dbl>
+ 1 fixed    <NA>  (Intercept)             -2.29     0.171      -13.4   4.94e- 41
+ 2 fixed    <NA>  typeRCT                 -0.338    0.172       -1.96  5.01e-  2
+ 3 fixed    <NA>  age                      0.0322   0.00160     20.2   2.13e- 90
+ 4 fixed    <NA>  as.factor(motor_score~  -0.593    0.0698      -8.50  1.95e- 17
+ 5 fixed    <NA>  as.factor(motor_score~  -0.904    0.0713     -12.7   8.43e- 37
+ 6 fixed    <NA>  as.factor(motor_score~  -1.29     0.0732     -17.6   1.67e- 69
+ 7 fixed    <NA>  pupilNone                1.32     0.0616      21.4   2.09e-101
+ 8 fixed    <NA>  pupilOne                 0.626    0.0711       8.80  1.36e- 18
+ 9 fixed    <NA>  ctIII                    0.810    0.0700      11.6   5.12e- 31
+10 fixed    <NA>  ctIV/V                   0.832    0.0593      14.0   9.47e- 45
+11 ran_pars name  sd__(Intercept)          0.277   NA           NA    NA        
+\end{CodeOutput}
 \end{CodeChunk}

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Integrate all comments first part introduction

@@
     url: https://hanneoberman.github.io/
-  - name: "Johanna Mu\\~{n}oz Avila"
+  - name: "Johanna Munoz Avila"
     affiliation: University Medical Center Utrecht \AND
@@
 
-~~In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.~~ Contemporary decision-making is  increasingly often data-driven and sometimes even entirely dictated by data.
+In many contemporary data analysis efforts, some form of hierarchical or clustered data structures are recorded.
+In the simplest case, such a structure entails the nesting of units within clusters (e.g., students within school classes).
+More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., data from diverse settings and populations within large databases).
+The clustered structure of multilevel data should be taken into account when developing analysis models: 1) for the simple reason that groups of observations share some common variance, and 2) because ignoring multilevel structures can be harmful to the statistical inferences and introduce bias in estimators [@hox17].
+There are many names for models that take clustering into account.
+Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
+Table \ref{tab:def} provides an overview of some key concepts in multilevel modeling.
+
+<!-- Contemporary decision-making is  increasingly often data-driven and sometimes even entirely dictated by data.
 Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale.
@@
 A common characteristic in aforementioned examples is the presence of clustering.
-[TD: briefly explain what is clustering?]
+[TD: briefly explain what is clustering?] -->
 
-Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and introduce bias in estimators [@hox17]. 
-~~Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
-The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster.~~
-These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
-~~The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
-The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.~~
-Multilevel models typically accommodate for this variability by including a separate group mean for each cluster.
+<!-- Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and introduce bias in estimators [@hox17].  -->
+<!-- ~~Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
+The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster.~~  -->
+<!-- These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance. -->
+<!-- ~~The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
+The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.~~  -->
+<!-- Multilevel models typically accommodate for variability by including a separate group mean for each cluster.
 In addition to random intercepts, multilevel models can also include random effects and heterogeneous residual error variances across clusters [see e.g.
 @gelm06, @hox17 and @jong21].
-There are many names for models that take clustering into account.
+-->
+
+<!-- There are many names for models that take clustering into account.
 Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
@@
 e.g.
-congeniality issues.]
+congeniality issues.] -->
+
+\begin{table}[tb]
+\caption{Concepts in multilevel modeling}
+\label{tab:def}
+\centering
+\begin{tabular}{ll}
+\hline
+\textbf{Concept} & \textbf{Details}   \\
+\hline
+ICC                 & The variability due to clustering is often measured by means of the \\
+                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
+                    & of variance that can be attributed to the cluster-level, where a high \\
+                    & ICC would indicate that a lot of variability is due to the cluster \\
+                    & structure. \\
+Random intercept    & Multilevel models typically accommodate for variability by including \\
+                    & a separate group mean for each cluster. In addition to random \\
+                    & intercepts, multilevel models can also include random effects and \\
+                    & heterogeneous residual error variances across clusters [see e.g. \\
+                    & @gelm06, @hox17 and @jong21]. \\
+\hline
+\end{tabular}
+\end{table}
 
 ## Missingness in multilevel data
+
+The process of analyzing multilevel data is further complicated when not all data entries are observed.
+Just as with single level data, missingness may occur at the unit level.
+But with multiple levels of data comes the potential for clustered missingness.
+Therefore, incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13].
+Systematic missingness implies that one or more variables are never observed in a certain cluster.
+With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18].
+We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns.
+Column $X1$ is completely observed, column $X2$ is systematically missing in cluster 2, and column $X3$ is sporadically missing.
+To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account.
+For example, the sporadic missingness in $X3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters).
+We could then just extrapolate the true (but missing) value of $X3$ for unit 1 from unit 2, and the value for unit 4 from unit 3.
+If $X3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method or discard the incomplete units altogether (i.e., list-wise deletion/complete case analysis).
+Further, with the systematic missingness in $X2$, it is impossible to fit a multilevel model since we cannot estimate the intercept of cluster 2.
+We would have to exclude this cluster from our analyses entirely to obtain any results.
+Obviously, excluding observations is not a desirable workflow.
 
@@
   theme(panel.grid = element_blank())
-
 ```
-
-The process of analyzing multilevel data is further complicated when not all data entries are observed.
-Just as with single level data, missingness may occur at the unit level.
-But with multiple levels of data comes the potential for clustered missingness.
-Therefore, incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13].
-Systematic missingness implies that one or more variables are never observed in a certain cluster.
-With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18].
-We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns.
-Column $X1$ is completely observed, column $X2$ is systematically missing in cluster 2, and column $X3$ is sporadically missing.
-To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account.
-For example, the sporadic missingness in $X3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters).
-We could then just extrapolate the true (but missing) value of $X3$ for unit 1 from unit 2, and the value for unit 4 from unit 3.
-If $X3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method or discard the incomplete units altogether (i.e., list-wise deletion/complete case analysis).
-Further, with the systematic missingness in $X2$, it is impossible to fit a multilevel model since we cannot estimate the intercept of cluster 2.
-We would have to exclude this cluster from our analyses entirely to obtain any results.
-Obviously, excluding observations is not a desirable workflow.
 

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Integrate all comments first part introduction

@@
 \author{
-Hanne Oberman\\Utrecht University \And Johanna Mu\textasciitilde\{n\}oz
-Avila\\University Medical Center Utrecht \AND Valentijn de
-Jong\\University Medical Center Utrecht \And Gerko Vink\\Utrecht
-University \AND Thomas Debray\\University Medical Center Utrecht
+Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
+Medical Center Utrecht \AND Valentijn de Jong\\University Medical
+Center Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
+Debray\\University Medical Center Utrecht
 }
@@
 
-\Plainauthor{Hanne Oberman, Johanna Mu\textasciitilde\{n\}oz
-Avila, Valentijn de Jong, Gerko Vink, Thomas Debray}
+\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
+Jong, Gerko Vink, Thomas Debray}
 \Plaintitle{Imputation of Incomplete Multilevel Data with mice}
@@
 
-\sout{In many contemporary data analysis efforts, some form of
-hierarchical or clustered structure is recorded.} Contemporary
-decision-making is increasingly often data-driven and sometimes even
-entirely dictated by data. Although data relevant for decision makers
-were initially collected using small and well-designed studies, there
-has been a growing need to address more complex questions at a wider
-scale. This not only requires to collect larger amounts of data, but
-also from more diverse settings and populations. For example, {[}TD:
-introduce IPDMA here; eg. HIV studies Johanna is referring to{]}.
-Another example is the use of large databases with information from
-thousands or even millions of individuals from multiple cities, regions
-or even countries. Finally, a third situation arises when data are
-collected from multicenter studies {[}TD: maybe refer to the
-schools/classes example here{]}. For example, students may be clustered
-in classes in psychometrics research, or patients may be clustered in
-studies in individual patient data meta-analyses in medical research. A
-common characteristic in aforementioned examples is the presence of
-clustering. {[}TD: briefly explain what is clustering?{]}
+In many contemporary data analysis efforts, some form of hierarchical or
+clustered data structures are recorded. In the simplest case, such a
+structure entails the nesting of units within clusters (e.g., students
+within school classes). More complex clustered structures may occur when
+there are multiple hierarchical levels (e.g., patients within hospitals
+within regions or countries), or when the clustering is non-nested
+(e.g., data from diverse settings and populations within large
+databases). The clustered structure of multilevel data should be taken
+into account when developing analysis models: 1) for the simple reason
+that groups of observations share some common variance, and 2) because
+ignoring multilevel structures can be harmful to the statistical
+inferences and introduce bias in estimators \citep{hox17}. There are
+many names for models that take clustering into account. Some popular
+examples are `multilevel models', `hierarchical models', `mixed effect
+models' and `random effect models'. Table \ref{tab:def} provides an
+overview of some key concepts in multilevel modeling.
 
-Ignoring the clustered structure of such multilevel data can be harmful
-to the statistical inferences and introduce bias in estimators
-\citep{hox17}. \sout{Imagine a case where cross-level interactions
-between unit-level variables and cluster-level variables are present.
-The cluster to which a unit belongs may then influence the unit-level
-observations--and vice versa for each of the units that make up the
-cluster.} These relations can and should be taken into account when
-developing analysis models for multilevel data for the simple reason
-that groups of observations share some common variance. \sout{The
-variability due to clustering is often measured by means of the
-intraclass coefficient (ICC). The ICC can be seen as the percentage of
-variance that can be attributed to the cluster-level, where a high ICC
-would indicate that a lot of variability is due to the cluster
-structure.} Multilevel models typically accommodate for this variability
-by including a separate group mean for each cluster. In addition to
-random intercepts, multilevel models can also include random effects and
-heterogeneous residual error variances across clusters \citep[see
-e.g.][\citet{hox17} and \citet{jong21}]{gelm06}. There are many names
-for models that take clustering into account. Some popular examples are
-`multilevel models', `hierarchical models', `mixed effect models' and
-`random effect models'. {[}TD: emphasize when/why we need to account for
-clustering in the analysis of clustered data. Why is the presence of
-clustering relevant when considering multiple imputation of missing
-data? e.g.~distinction between systematically and sporadically missing
-data. But also: mechanisms of missing data (e.g.~MCAR, MAR, MNAR) may
-differ between clusters. But also: relation between observed data may
-differ between clusters? When/ why should we avoid using traditional
-imputation methods? e.g.~congeniality issues.{]}
+\begin{table}[tb]
+\caption{Key concepts in multilevel modeling}
+\label{tab:def}
+\centering
+\begin{tabular}{ll}
+\hline
+\textbf{Concept} & \textbf{Details}   \\
+\hline
+ICC                 & The variability due to clustering is often measured by means of the \\
+                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
+                    & of variance that can be attributed to the cluster-level, where a high \\
+                    & ICC would indicate that a lot of variability is due to the cluster \\
+                    & structure. \\
+Random intercept    & Multilevel models typically accommodate for variability by including \\
+                    & a separate group mean for each cluster. In addition to random \\
+                    & intercepts, multilevel models can also include random effects and \\
+                    & heterogeneous residual error variances across clusters [see e.g. \\
+                    & @gelm06, @hox17 and @jong21]. \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 data}\label{missingness-in-multilevel-data}}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/patterns-1}
-
-}
-
-\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
-\end{figure}
-\end{CodeChunk}
 
@@
 workflow.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/patterns-1}
+
+}
+
+\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
+\end{figure}
+\end{CodeChunk}
 

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Before moving case study 1 and 2 to a separate document

@@
 With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18].
-We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns.
-Column $X1$ is completely observed, column $X2$ is systematically missing in cluster 2, and column $X3$ is sporadically missing.
+We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters and $p$ variables.
+Column $X_1$ is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing.
 To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account.
-For example, the sporadic missingness in $X3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters).
-We could then just extrapolate the true (but missing) value of $X3$ for unit 1 from unit 2, and the value for unit 4 from unit 3.
-If $X3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method or discard the incomplete units altogether (i.e., list-wise deletion/complete case analysis).
-Further, with the systematic missingness in $X2$, it is impossible to fit a multilevel model since we cannot estimate the intercept of cluster 2.
-We would have to exclude this cluster from our analyses entirely to obtain any results.
-Obviously, excluding observations is not a desirable workflow.
+For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters).
+We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3.
+If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis).
+Complete case analysis can however introduce bias in statistical inferences and lowers statistical power.
+Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way.
+Complete case analysis in that case would mean excluding the entire cluster from the analyses.
+The wrong choice of missing data handling method can thus be extremely harmful to the inferences.
+
+<!-- Obviously, excluding observations is not a desirable workflow.  -->
 
@@
 
-Ignoring the missingness in analyses can be extremely harmful to inferences.
-Complete case analysis can introduce bias in statistical inferences and lowers statistical power.
-Instead, the missingness should be accommodated \emph{before} or _within_ the analysis of scientific interest.
-Especially the former is very generic and popular.
-[VJ: add why multiple values are imputed.] Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analyzed as if it were completely observed.
+Since excluding observations is not a desirable workflow, the missingness in multilevel data should be accommodated \emph{before} or _within_ the analysis of scientific interest.
+In this paper, we focus on the former approach: imputing (i.e., filling in) the missing data with plausible values, whereafter the completed data may be analyzed as if it were completely observed.
+Imputation separates the missing data problem from the scientific problem, which makes the missing data strategy very generic and popular.
+If each missing value is replaced multiple times, the resulting inferences may validly convey the uncertainty due to missingness [c.f.
+@rubi76].
 The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
@@
 
----
+<!-- --- -->
 
-\begin{center}
-REVIEW UNTIL THIS POINT SVP
-\end{center}
+<!-- \begin{center} -->
+<!-- REVIEW UNTIL THIS POINT SVP -->
+<!-- \end{center} -->
 
----
+<!-- --- -->
 

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Before moving case study 1 and 2 to a separate document

@@
 \begin{table}[tb]
-\caption{Key concepts in multilevel modeling}
+\caption{Concepts in multilevel modeling}
 \label{tab:def}
@@
 which shows an \(n \times p\) set \(\mathbf{X} = X_1, \dots, X_p\), with
-\(n\) units distributed over \(N\) clusters, and \(p=3\) columns. Column
-\(X1\) is completely observed, column \(X2\) is systematically missing
-in cluster 2, and column \(X3\) is sporadically missing. To analyze
+\(n\) units distributed over \(N\) clusters and \(p\) variables. Column
+\(X_1\) is completely observed, column \(X_2\) is systematically missing
+in cluster 2, and column \(X_3\) is sporadically missing. To analyze
 these incomplete data, we have to take the nature of the missingness and
 the cluster structure into account. For example, the sporadic
-missingness in \(X3\) could be easily amended if this would be a
+missingness in \(X_3\) could be easily amended if this would be a
 cluster-level variable (and thus constant within clusters). We could
-then just extrapolate the true (but missing) value of \(X3\) for unit 1
-from unit 2, and the value for unit 4 from unit 3. If \(X3\) would
+then just extrapolate the true (but missing) value of \(X_3\) for unit 1
+from unit 2, and the value for unit 4 from unit 3. If \(X_3\) would
 instead be a unit-level variable (which may vary within clusters), we
 could not just recover the unobserved `truth', but would need to use
-some kind of missing data method or discard the incomplete units
-altogether (i.e., list-wise deletion/complete case analysis). Further,
-with the systematic missingness in \(X2\), it is impossible to fit a
-multilevel model since we cannot estimate the intercept of cluster 2. We
-would have to exclude this cluster from our analyses entirely to obtain
-any results. Obviously, excluding observations is not a desirable
-workflow.
+some kind of missing data method, or discard the incomplete units
+altogether (i.e., complete case analysis). Complete case analysis can
+however introduce bias in statistical inferences and lowers statistical
+power. Further, with the systematic missingness in \(X_2\), it would be
+impossible to fit a multilevel model without accommodating the
+missingness in some way. Complete case analysis in that case would mean
+excluding the entire cluster from the analyses. The wrong choice of
+missing data handling method can thus be extremely harmful to the
+inferences.
 
@@
 
-Ignoring the missingness in analyses can be extremely harmful to
-inferences. Complete case analysis can introduce bias in statistical
-inferences and lowers statistical power. Instead, the missingness should
-be accommodated \emph{before} or \emph{within} the analysis of
-scientific interest. Especially the former is very generic and popular.
-{[}VJ: add why multiple values are imputed.{]} Imputing (i.e., filling
-in) the missing values separates the missing data problem from the
-scientific problem: missing data are replaced by plausible values
-whereafter the completed data is analyzed as if it were completely
-observed. The \proglang{R} package \pkg{mice} has become the de-facto
-standard for imputation by chained equations, which iteratively solves
-the missingness on a variable-by-variable basis. \pkg{mice} is known to
-yield valid inferences under many different missing data circumstances
+Since excluding observations is not a desirable workflow, the
+missingness in multilevel data should be accommodated \emph{before} or
+\emph{within} the analysis of scientific interest. In this paper, we
+focus on the former approach: imputing (i.e., filling in) the missing
+data with plausible values, whereafter the completed data may be
+analyzed as if it were completely observed. Imputation separates the
+missing data problem from the scientific problem, which makes the
+missing data strategy very generic and popular. If each missing value is
+replaced multiple times, the resulting inferences may validly convey the
+uncertainty due to missingness \citep[c.f.][]{rubi87}. The \proglang{R}
+package \pkg{mice} has become the de-facto standard for imputation by
+chained equations, which iteratively solves the missingness on a
+variable-by-variable basis. \pkg{mice} is known to yield valid
+inferences under many different missing data circumstances
 \citep{buur18}. In this paper, we will discuss how to use \pkg{mice} in

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Clean up manuscript file

@@
 We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
+[Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
 keywords:
@@
 
-# seed
+# environment
 set.seed(123)
-
-# packages
-library(tidyverse)
-library(mice)
-library(miceadds)
-library(lme4)
-library(broom.mixed)
-# library(metamisc)
-# library(GJRM)
-# library(pan)
-
-# functions
-`%nin%` <- Negate(`%in%`)
-round3 <- function(x){format(round(x, 3), nsmall = 3)}
-icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
-# icc for imputed data as function in mice
-miceadds::source.all(path = "../R")
-
-# plot parameters
-plot_col <- mice:::mdc(1:2) %>% setNames(c("observed", "missing"))
-
-# data
-popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
-  mutate(school = SCHOOL,
-         popular = POPULAR,
-         sex = SEX,
-         texp = TEXP,
-         .keep = "none")
-popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
+library(dplyr)
+library(ggplot2)
 ```
@@
 <!-- TODO: look at vim -->
+<!-- TODO: check if shepley plot is an option -->
 
-<!-- Comments Thomas: -->
-<!-- .	We need to remind the reader about general knowledge about missing data: it usually leads to loss of power and bias; we distinguish between MCAR, MAR and MNAR in theory but in practice this distinction is less clear; multiple imputation is generally recommended to avoid bias (if data are MCAR or MAR, but also works reasonable in some situations where data are partially MNAR) -->
-<!-- .	We need to clarify what is clustered data, and provide common types (IPDMA, registries with data from multiple hospitals/regions, and multicenter studies).
-You can directly refer to the examples you will use in the paper here, and build on the text you already had.
--->
-<!-- .	We need to clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implemention ("though shall not simply run mice() on any incomplete dataset") -->
-<!-- .	Maybe we should also emphasise that in more modern types of big data sources (E.g. EHR data ), measurement of data is often selective and MAR assumptions are less tenable.  -->
+<!-- Conditional plots missingness indicator -->
+<!-- TODO: check if up-side-down plot works -->
+<!-- TODO: check smoothing in geom_density function and make it the inverse of the sample size -->
+<!-- TODO: add functions to mice -->
+<!-- TODO: add title and informative legend (pop obs and pop miss) -->
+<!-- TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .` -->
+<!-- TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon -->
+<!-- TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?) -->
 
@@
 In the simplest case, such a structure entails the nesting of units within clusters (e.g., students within school classes).
-More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., data from diverse settings and populations within large databases).
+More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases).
 The clustered structure of multilevel data should be taken into account when developing analysis models: 1) for the simple reason that groups of observations share some common variance, and 2) because ignoring multilevel structures can be harmful to the statistical inferences and introduce bias in estimators [@hox17].
@@
 Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
-Table \ref{tab:def} provides an overview of some key concepts in multilevel modeling.
-
-<!-- Contemporary decision-making is  increasingly often data-driven and sometimes even entirely dictated by data.
-Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale.
-This not only requires to collect larger amounts of data, but also from more diverse settings and populations.
-For example, [TD: introduce IPDMA here; eg.
-HIV studies Johanna is referring to].
-Another example is the use of large databases with information from thousands or even millions of individuals from multiple cities, regions or even countries.
-Finally, a third situation arises when data are collected from multicenter studies [TD: maybe refer to the schools/classes example here].
-For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses in medical research.
-A common characteristic in aforementioned examples is the presence of clustering.
-[TD: briefly explain what is clustering?] -->
-
-<!-- Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and introduce bias in estimators [@hox17].  -->
-<!-- ~~Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
-The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster.~~  -->
-<!-- These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance. -->
-<!-- ~~The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
-The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.~~  -->
-<!-- Multilevel models typically accommodate for variability by including a separate group mean for each cluster.
-In addition to random intercepts, multilevel models can also include random effects and heterogeneous residual error variances across clusters [see e.g.
-@gelm06, @hox17 and @jong21].
--->
-
-<!-- There are many names for models that take clustering into account.
-Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
-[TD: emphasize when/why we need to account for clustering in the analysis of clustered data.
-Why is the presence of clustering relevant when considering multiple imputation of missing data?
-e.g.
-distinction between systematically and sporadically missing data.
-But also: mechanisms of missing data (e.g.
-MCAR, MAR, MNAR) may differ between clusters.
-But also: relation between observed data may differ between clusters?
-When/ why should we avoid using traditional imputation methods?
-e.g.
-congeniality issues.] -->
+Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.
 
 \begin{table}[tb]
-\caption{Concepts in multilevel modeling}
-\label{tab:def}
+\caption{Concepts in multilevel methods}
+\label{tab:clus}
 \centering
@@
 We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters and $p$ variables.
-Column $X_1$ is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing.
+
+```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
+dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
+  cbind(text = c("1", "1", "2", "2", "", "N", rep("", 30)),
+        miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))
+ggplot(dat, aes(x = cols, y = rows)) +
+  geom_tile(fill = "white", color = "black", size = 0.5) +
+  geom_text(aes(label = text), color = "black", size = 3) +
+  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
+  scale_x_continuous(
+    breaks = 1:6,
+    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
+    name = NULL,
+    position = "top"
+  ) +
+  scale_y_continuous(breaks = 1:6, labels = c(1:4, "...", "n"), name = NULL, trans = "reverse") +
+  coord_cartesian(expand = c(0,0,0,0)) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        plot.margin = unit(c(0,0,0,0), "pt"))
+```
+Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing.
 To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account.
@@
 
-<!-- Obviously, excluding observations is not a desirable workflow.  -->
+A key characteristic of the missing data to take into account in analyses is the mechanism behind the missingness.
+We distinguish between MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice this distinction is less clear.
+Since the essence of the true non-response mechanism may not be known, it is generally inferred or assumed to be ignorable (i.e., MCAR or MAR).
+[TODO: add that this assumption may not always be valid, especially with modern types of big dasta sources.] Depending on the actual missingness-generating mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.
 
-```{r patterns, fig.height=2.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
-# blankcol <- c(rep("", 4), "...", "")
-# dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
-#   cbind(text = c("1", "1", "2", "2", "...", "N", rep(blankcol, 3), rep("...", 6), blankcol),
-#         miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))
-dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
-  cbind(text = c("1", "1", "2", "2", "", "N", rep("", 30)),
-        miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))
-
-ggplot(dat, aes(x = cols, y = rows)) +
-  geom_tile(fill = "white", color = "black") +
-  geom_text(aes(label = text), color = "black") +
-  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
-  scale_x_continuous(breaks = 1:6, labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])), name = NULL, position = "top") +
-  scale_y_continuous(breaks = 1:6, labels = c(1:4, "...", "n"), name = NULL, trans = "reverse") +
-  theme_minimal() +
-  theme(panel.grid = element_blank())
-```
+\begin{table}[tb]
+\caption{Concepts in missing data methods}
+\label{tab:miss}
+\centering
+\begin{tabular}{ll}
+\hline
+\textbf{Concept} & \textbf{Details}   \\
+\hline
+MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
+& across all data entries \\
+MAR     & Missing At Random, where the probability to be missing depends on observed \\
+& information \\
+MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
+& depends on unrecorded information, making the missingness non-ignorable \\
+& [@rubi76; @meng94]. \\
+& [TODO: add congeniality.] \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 
-<!-- Another characteristic of the missing data to take into account in analyses is the mechanism behind the missingness.
-Although the essence of the true non-response mechanism may not be known, it can be inferred or assumed to be one of the following: -->
-<!-- - Missing Completely At Random (MCAR), where the probability to be missing is equal across all data entries; -->
-<!-- - Missing At Random (MAR), where the probability to be missing depends on observed information; -->
-<!-- - Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]. -->
-<!-- Depending on the assumed missingness mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15. -->
-<!-- "We find it useful to distinguish the missingness pattern, which describes which values are missing and observed in the data matrix and the missingness mechanism (or mechanisms), which concerns the relationship between missingness and the values of variables in the data matrix" (Little and Rubin, p.
-8).
--->
+[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").
+And add overview of possible predictor matrix values.]
 
@@
 
-This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}.
-We provide practical guidelines and code snippets for different missing data situations, including missing not at random (MNAR) mechanisms [where the probability to be missing depends on unrecorded information, making the missingness non-ignorable, @rubi76; @meng94].
-For reasons of brevity, we focus on imputation by chained equations wit \pkg{mice} exclusively^[Note that the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well but is outside the scope of this tutorial.].
-Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work by @audi18 and @grun18.
-Please note that this tutorial paper assumes a basic level of knowledge on multilevel models.^[Note to self: We're providing an overview of implementations.
-It's up-to the reader to decide which multilevel strategy suits their data.
-We won't go into detail for the different methods (and equations).
-This paper is just a software tutorial, so we'll keep it practical.] Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the \pkg{magrittr} package, and the \pkg{lme4} notation for multilevel models.^[TODO: Add environment info, seed and version number(s) somewhere!]
+This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice} in \proglang{R}.
+We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms.
+For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial.
+Assumed knowledge includes basic familiarity with multilevel imputation [see e.g.
+@audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).
 
-We illustrate how to impute incomplete multilevel data by means of three case studies:
+\begin{table}[tb]
+\caption{Concepts in missing data methods}
+\label{tab:mod}
+\centering
+\begin{tabular}{ll}
+\hline
+\textbf{Concept} & \textbf{Details}   \\
+\hline
+& [TODO: explain \pkg{lme4} notation here] \\
+\hline
+\end{tabular}
+\end{table}
 
-- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, @mice];
-- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, @GJRM];
-- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc].
+# Case study
 
-For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
-With the `popmis` data, we show how (and how not) to develop an imputation model.
-With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods.
-With the `impact` data we provide an example of multivariate missingness in real-world data.
-Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph or 'translation table' linking multilevel equations to \pkg{lme4} formulas.
-Use betas instead of gamma's and mu's.
-Add interpretation of values in predictormatrix (-2 for the cluster variable, 2 for random effects).
-Add ICC and congeneality here as well.
-And make missingness mechanism table as well.]
-
-<!-- Table 7.1: Questions to gauge the complexity of a multilevel imputation task. -->
-<!-- 1.	Will the complete-data model include random slopes? -->
-<!-- 2.	Will the data contain systematically missing values? -->
-<!-- 3.	Will the distribution of the residuals be non-normal? -->
-<!-- 4.	Will the error variance differ over clusters? -->
-<!-- 5.	Will there be small clusters? -->
-<!-- 6.	Will there be a small number of clusters? -->
-<!-- 7.	Will the complete-data model have cross-level interactions? -->
-<!-- 8.	Will the dataset be very large? (fimd, section 7.3)-->
-
-Set-up the R environment and load the necessary packages:
-
-```{r}
-set.seed(2022)
-library(mice)
-library(ggmice)
-library(ggplot2)
-library(dplyr)
-library(lme4)
-library(mitml)
-```
-
-<!-- --- -->
-
-<!-- \begin{center} -->
-<!-- REVIEW UNTIL THIS POINT SVP -->
-<!-- \end{center} -->
-
-<!-- --- -->
-
-# Case Study I: How (not) to impute
-
-```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
-# estimates complete popularity data
-est_comp <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popcomp) 
-# fixed effects only
-est_comp_fixed <- est_comp %>% 
-  broom.mixed::tidy("fixed", conf.int = TRUE) %>% 
-  mutate(term = term,
-         estimate  = paste0(
-           round3(estimate), " [", 
-           round3(conf.low), ", ", 
-           round3(conf.high), "]"), 
-         .keep = "none") %>% 
-  as.data.frame()
-# # random effects
-# est_comp %>% 
-#   broom.mixed::tidy("ran_pars") %>% 
-#   mutate(term = term, #factor(term, levels = unique(.$term), labels = c()),
-#          estimate  = round3(estimate), 
-#          .keep = "none")
-# # rename broom.mixed estimates
-# lvl = c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation")
-# lbl = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor.
-cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error")
-
-# compare complete data to missingness
-# popcomp <- popcomp %>% 
-#   cbind(R = is.na(popmis$popular))
-# t.test(x = popcomp$popular[popcomp$R], y = popcomp$popular[!popcomp$R]) 
-# ggplot(popcomp, aes(y=popular, x = R)) +
-#   geom_boxplot()
-```
-
-In this section we'll go over the different steps involved with imputing incomplete multilevel data.
-The data we're using is the `popmis` dataset from the `mice` package.
-This is a simulated dataset with pupils ($n = 2000$) clustered within schools ($N = 100$).
-In this tutorial we'll use the following variables:
-
-  - `school`,     school identification number (clustering variable);
-  - `popular`,    pupil popularity (self-rating between 0 and 10; unit-level);
-  - `sex`,	      pupil sex (0=boy, 1=girl; unit-level);
-  - `texp`,	      teacher experience (in years; cluster-level).
-
-The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction.
-The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`.
-The regression equation^[add the 'level notation' (Bryk and Raudenbush, 1992) and/or matrix notation ('linear mixed effects model'; Laird and Ware, 1982) too?] and `lme4` notation for this model are
-
-$$
-\text{popular}_{ij} =
-\gamma_{00} + 
-\gamma_{10} \text{ sex}_{ij} + 
-\gamma_{01} \text{ texp}_{j} + 
-\gamma_{11} \text{ texp}_{j} \times \text{sex}_{ij} + 
-u_{0j} + 
-u_{1j} \text{ sex}_{ij} + 
-e_{ij} \\
-$$
-
-$$
-\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
-$$
-
-<!-- Intercept-only model: lmer(formula = popular ~ 1 + (1|class)) -->
-<!-- First level predictors: lmer(formula = popular ~ 1 + sex + (1|class)) -->
-<!-- First and second level predictors: lmer(popular ~ 1 + sex + texp + (1 | class)) -->
-<!-- First and second level predictors with random slopes: lmer(formula = popular ~ 1 + sex + texp + (1 + sex | class) -->
-<!-- First and second level predictors with random slopes and cross-level interaction: lmer(formula = popular ~ 1 + sex + texp + sex:texp + (1 + sex | class)) -->
-
-Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data.
-The data is created in such a way that the clustering variable `school` explains quite some variance in the outcome variable `popular`.
-We express this using the intraclass correlation,  ICC $=$ `r round(icc(popular ~ as.factor(school), data = popcomp), 2)`.
-We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effects:
-
-```{r comp_est, echo=FALSE}
-est_comp_fixed %>% setNames(c("", "Estimate with 95% CI"))
-# TODO: make this a figure with the different methods as facets
-```
-
-### Incomplete data
-
-Load the data into the environment and select the relevant variables:
-
-```{r}
-popmis <- popmis[, c("school", "popular", "sex", "texp")] 
-```
-Plot the missing data pattern:
-
-```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
-plot_pattern(popmis)
-```
-
-The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}.
-The ICC in the incomplete data is `r round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.
-This tells us that the multilevel structure of the data should probably be taken into account.
-If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.
-
-Plot the correlations in the incomplete data:
-
-```{r pop-corr}
-plot_corr(popmis)
-```
-
-To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables.
-We'll highlight one other variable to illustrate, but ideally one would inspect all relations.
-The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'.
-This can be evaluated statistically, but visual inspection usually suffices.
-We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.
-
-Plot the histogram for teacher experience conditional on the missingness indicator of `popular`:
-
-```{r pop-hist}
-ggmice(popmis, aes(texp)) +
-  geom_histogram(fill = "white") +
-  facet_grid(. ~ is.na(popular), labeller = label_both)
-```
-
-TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?
-
-```{r pop-hist-clust, include=FALSE}
-# popmis %>% 
-#   filter(school < 5) %>% 
-# ggmice(., aes(texp)) +
-#   geom_histogram(fill = "white") +
-#   facet_grid(school ~ is.na(popular), labeller = label_both)
-# # does not make sense because texp is a cluster level variable!
-```
-
-This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
-
-<!-- In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
-This would indicate a right-tailed MAR missingness.
-(In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `texp`.
--->
-
-```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-# plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
-# TODO: check if up-side-down plot works
-# TODO: check smoothing in geom_density function and make it the inverse of the sample size
-# TODO: add functions to mice
-# TODO: add title and informative legend (pop obs and pop miss)
-# TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .`
-# TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon
-# TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?)
-
-# # maybe use facets instead
-# popmis %>%
-#   filter(as.numeric(school) < 5) %>%
-#   ggplot(aes(x = texp, color = school)) +
-#     geom_density() +
-#     facet_wrap(~factor(is.na(popular), labels = c("popular missing", "popular observed")), ncol = 1) +
-#     scale_color_manual(values = plot_col, name = "popular") +
-#     theme_classic() # +
-#     # theme(legend.position = "bottom")
-```
-
-### Complete case analysis (not recommended)
-
-Complete case analysis ignores the observations with missingness altogether, which lowers statistical power and may even introduce bias in MCAR situations.
-
-```{r pop-cca, echo=FALSE}
-est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis)
-
-```
-
-### Imputation ignoring the cluster variable (not recommended)
-
-The first imputation model that we'll use is likely to be invalid.
-We do *not* use the cluster identifier `school` as imputation model predictor.
-With this model, we ignore the multilevel structure of the data, despite the high ICC.
-This assumes exchangeability between units.
-We include it purely to illustrate the effects of ignoring the clustering in our imputation effort.
-We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).
-
-Create a methods vector and predictor matrix for `popular`, and make sure `school` is not included as predictor:
-
-```{r pop-ignored-pred, echo=TRUE, message=FALSE, warning=FALSE}
-meth <- make.method(popmis) # methods vector
-pred <- quickpred(popmis)   # predictor matrix
-plot_pred(pred)
-```
-
-Impute the data, ignoring the cluster structure:
-```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
-imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
-```
-
-TODO: remove the broom.mixed output, use mitml only
-
-Analyze the imputations:
-```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
-fit_ignored <- imp_ignored %>% 
-  with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
-testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
-```
-```{r eval=FALSE, include=FALSE}
-# fit_ignored %>% 
-#   pool() %>% 
-#   broom.mixed::tidy()
-```
-
-```{r pop_ignored_eval, echo=FALSE, message=FALSE, warning=FALSE}
-# # check convergence of the imputation model
-# plot(imp_ignored)
-
-# # compare descriptives before and after imputation
-# psych::describe(popmis)[, c("n", "mean", "median", "min", "max", "sd")]
-# psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
-# TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
-# TODO: pool mean median and sd
-# feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
-# TODO: add FMI for each of the estimates? at least for the mean
-
-# further inspection of the imputations
-plot_imps(imp_ignored, type = "stripplot", x = "popular")
-# TODO: think about multimodality
-
-# # compare ICCs before and after imputation
-# ICCs <- data.frame(
-#   vars = c("popular", "teachpop", "texp"), 
-#   CCA = c(icc(popular ~ school, popmis), 
-#                icc(teachpop ~ school, popmis),
-#                icc(texp ~ school, popmis)), 
-#   ignored = c(icc(popular ~ school, complete(imp_ignored)), 
-#               icc(teachpop ~ school, complete(imp_ignored)), 
-#               icc(texp ~ school, complete(imp_ignored)))
-#   )
-# ICCs
-
-#TODO: think about interactive clusters for the strippplots
-```
-
-<!-- We can also observe that the teacher experience increases slightly after imputation.
-This is due to the MNAR missingness in `texp`.
-Higher values for `texp` have a larger probability to be missing.
-This may not a problem, however, if at least one pupil in each school has teacher experience recorded, we can deductively impute the correct (i.e.
-true) value for every pupil in the school.
--->
-
-<!-- pmm: Single-level predictive mean matching with the school indicator coded as a dummy variable; -->
-<!-- 2l.pan: Multilevel method using the linear mixed model to draw univariate imputations; -->
-<!-- 2l.norm: Multilevel method using the linear mixed model with heterogeneous error variances; -->
-<!-- 2l.pmm: Predictive mean matching based on predictions from the linear mixed model, with random draws from the regression coefficients and the random effects, using five donors. -->
-
-### Imputation with the cluster variable as predictor (not recommended)
-
-We'll now use `school` as a predictor to impute all other variables.
-This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
-But at least, it includes some multilevel aspect.
-This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16].
-Colloquially, this is 'multilevel imputation for dummies'.
-
-Add: doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.
-
-The more the random effects are of interest, the more you need ml models.
-
-<!-- "Beyond the potential bias, there are other reasons why a multilevel imputation model should be preferred if possible.
-First, the model is guaranteed to produce unbiased results even in extreme cases with very high missingness rates and/or small intraclass correlations.
-Second, with the multilevel model, it is possible to include additional variables that are constant on the cluster level, for example, information on the teachers if the cluster levels are chosen to be students within classes or background information on the schools if the clustering is modeled as students within schools.
-This approach is not possible with fixed effects imputation.
-While this limitation should not be problematic in terms of bias (the combined school effects are still modeled through the cluster specific intercepts), using this additional information will make the random effects imputation model more efficient." -->
-
-```{r pop_predictor, message=FALSE, warning=FALSE}
-# adjust the predictor matrix
-pred["popular", "school"] <- 1 
-plot_pred(pred)
-
-# impute the data, cluster as predictor
-imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
-```
-
-```{r pop_predictor_eval, echo=FALSE, message=FALSE, warning=FALSE}
-# # check logged events
-# head(imp_predictor$loggedEvents)
-## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and teachpop.
-
-# # check convergence of the imputation model
-# plot(imp_predictor)
-
-# # compare descriptives before and after imputation
-# psych::describe(popmis)[, c("n", "mean", "median", "min", "max", "sd")]
-# psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
-
-# further inspection of the imputations
-plot_imps(imp_predictor, x = "popular", type = "stripplot")
-
-# # compare ICCs before and after imputation
-# ICCs <- ICCs %>% mutate(
-#            predictor = c(icc(popular ~ school, complete(imp_predictor)), 
-#                         icc(teachpop ~ school, complete(imp_predictor)), 
-#                         icc(texp ~ school, complete(imp_predictor)))
-#            )
-# ICCs
-```
-
-Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MAR.
-
-The ICCs are way more in line with the ICCs in the incomplete data. But this is a quick and dirty way of imputing multilevel data. We *should* be using a multilevel model.
-
-### Imputation with random effects
-
-With `2l.norm` we impute the outcome with a multilevel model  assuming random slopes for each variable in the imputation model and homogeneous within-cluster variance.
-
-"Van Buuren (2011) considered the homoscedastic linear mixed model as invalid for imputing incomplete predictors, and investigated only the 2l.norm method, which allows for heterogeneous error variances" [@buur18].
-
-```{r pop_norm}
-# adjust the predictor matrix
-pred["popular", ] <- c(school = -2, popular = 0, sex = 2, texp = 2) 
-plot_pred(pred) 
-meth <- make.method(popmis)
-meth["popular"] <- "2l.pmm"
-imp_pmm_2l <-
-  mice(
-    popmis %>% mutate(school = as.integer(school)),
-    pred = pred,
-    meth = meth,
-    maxit = 1,
-    print = FALSE
-  )
-```
-
-```{r pop_norm_eval, echo=FALSE}
-# plot(imp_norm)
-plot_imps(imp_pmm_2l, x = "popular", type = "stripplot")
-
-# ICCs <- ICCs %>% mutate(
-#            norm = c(icc(popular ~ as.factor(school), complete(imp_norm_2l)), 
-#                     icc(teachpop ~ as.factor(school), complete(imp_norm_2l)), 
-#                     icc(texp ~ as.factor(school), complete(imp_norm_2l)))
-#            )
-# ICCs
-```
-
-### Imputation with random effects and heterogeneity
-
-This method assumes random slopes for each variable in the imputation model. In contrast to `2l.norm` this method allows a cluster-specific residual error variance.
-
-<!-- Look up Schafer book fortran workflows. -->
-
-<!-- "By default, 2l.pan includes the intercept as both a fixed and a random effect.
-...
-There are 5 different codes you can use in the predictor matrix for variables imputed with 2l.pan.
-The person identifier is coded as -2 (this is different from 2l.norm).
-To include predictor variables with fixed or random effects, these variables are coded with 1 or 2, respectively.
-If coded as 2, the corresponding fixed effect is automatically included.
-In addition, 2l.pan offers the codes 3 and 4, which have similar meanings as 1 and 2 but will include an additional fixed effect for the person mean of that variable.
-This is useful if you're trying to model within- and between-person effects of time-varying predictor variables." -->
-
-```{r pop_pan}
-pred["popular", ] <- c(-2, 2, 1, 2)
-meth <- c("", "2l.pan", "", "")
-imp_pan_2l <-
-  mice(
-    popmis %>% mutate(school = as.integer(school)),
-    pred = pred,
-    meth = meth,
-    maxit = 1,
-    print = FALSE
-  )
-
-```
-
-```{r pop_pan_eval, echo=FALSE}
-# plot(imp_pan)
-plot_imps(imp_pan_2l, x = "popular", type = "stripplot")
-# ICCs <- ICCs %>% mutate(
-#            pan = c(icc(popular ~ as.factor(school), complete(imp_pan_2l)), 
-#                     icc(teachpop ~ as.factor(school), complete(imp_pan_2l)), 
-#                     icc(texp ~ as.factor(school), complete(imp_pan_2l)))
-#            )
-# ICCs
-```
-
-# How to handle non-random selection (Case study II: HIV)
-
-Data are simulated and included in the `GJRM` package. We will use the following variables:
-
-  - `region` Cluster variable,
-  - `hiv` HIV diagnosis (0=no, 1=yes),
-  - `age` Age of the patient,
-  - `marital` Marital status,
-  - `condom` Condom use during last intercourse,
-  - `smoke` Smoker (levels; inclusion restriction variable).
-
-The imputation of these date is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
-
-<!-- - main model = analysis model = predicting the prevalence of hiv -->
-<!-- - people who have hiv are more reluctant to taking the test, not to reveal their status -->
-<!-- - solution: take variables from analysis model like condom use, marital status, age, etc. that relate to hiv prevalence -->
-<!-- - exclusion restriction variable = instrumental variable (from economics) = e.g.
-how kind the nurse was = smoke in this dataset = should be correlated to the missingness indicator, but *not* the variable itself -->
-<!-- - include these variables in the selection equation -->
-<!-- - take the missingness indicator of the variable of interest for the selection equation as DV -->
-<!-- - heckman models estimate BOTH selection and mean models simultaneously, because linked through the error term -->
-<!-- - if the errors are not correlated, the models are not linked at all -> not MNAR, but just MAR -->
-<!-- - if the rho is significantly different from zero, the errors are correlated and the models are related so we have MNAR -->
-<!-- - now, we add clustering -->
-<!-- - copulas = estimate 2 equations separately and use another equation to link them -->
-<!-- - for each of the clusters, est the copula separately, so we have all coeffs for each cluster, incl the rho and sigma (the errors) -->
-<!-- - put together all of the estimates through a meta-analysis model to get population distribution of the estimates -->
-<!-- - use the population parameter estimates to draw marginal parameters from the distribution to get a cluster/study parameter -->
-<!-- - with systematical miss we can est the values of the imputed values based on the marginal distribution -->
-<!-- - optional to add shrinkage: the marginal in account as well -->
-<!-- - the hiv data was simulated by the copula function/model developer -->
-<!-- - we could add a systematically missing variable in the data, because these are all sporadically -->
-
-```{r hiv, echo=FALSE}
-# data
-data("hiv", package = "GJRM")
-hiv <- hiv %>% 
-  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
-  mutate(region = as.integer(region))
-# md pattern
-plot_md_pat(hiv)
-```
-
-From the missing data pattern we see that we can set `maxit` to 1, since there is only one variable with missingness.
-
-```{r include=FALSE}
-# inclusion restriction variable 
-mod_main <- glm(hiv ~ ., family = "binomial", data = hiv) %>% 
-  broom::tidy(conf.int = TRUE) %>% 
-  .[7, c(2,6,7)] %>% 
-  round(., 3)
-mod_select <- glm(is.na(hiv) ~ ., family = "binomial", data = hiv) %>% 
-  broom::tidy(conf.int = TRUE) %>% 
-  .[7, c(2,6,7)] %>% 
-  round(., 3)
-```
-
-The inclusion restriction variable should be a predictor of the the actual value of the variable of interest, but *not* of missingness indicator for the variable of interest.
-In this case, the data were simulated to adhere to this requirement.
-Namely, $\beta_{smoke}$ = `r mod_main$estimate`, 95% CI [`r paste0(mod_main$conf.low, ", ", mod_main$conf.high)`] for the analysis model (`formula = hiv ~ .`), and $\beta_{smoke}$ = `r mod_select$estimate`, 95% CI [`r paste0(mod_select$conf.low, ", ", mod_select$conf.high)`] for the selection model (`formula = is.na(hiv) ~ .`).
-This means the assumptions for the Heckman-type selection model are met.
-
-# How to handle multivariate missingness (Case study III: IMPACT)
-
-`impact` is traumatic brain injury data with patients, $n = 11022$, clustered in studies, $N = 15$. With the following 11 variables:
+We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc].
+`impact` is traumatic brain injury data with patients, $n = 11022$, clustered in studies, $N = 15$.
+With the following 11 variables:
 
@@
 
-The analysis model for this dataset is a prediction model with `mort` as the outcome.^[Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of `ct` on unfortunate outcomes, we just want to know the adjusted odds ratio for `ct`.
-Add something about systematically missing data here.] The data is already imputed (Steyerberg et al, 2008), so we've induced missingness again based on the missingness in the original data.
-
-$$
-\text{mort}_{ij} =
-\gamma_{00} + 
-\gamma_{01} \text{ type}_{j} + 
-\gamma_{10} \text{ age}_{ij} + 
-\gamma_{10} \text{ moter\_score}_{ij} + 
-\gamma_{10} \text{ pupil}_{ij} + 
-\gamma_{10} \text{ ct}_{ij} + \\
-u_{0j} + 
-u_{1j} \text{ age}_{ij} + 
-u_{1j} \text{ moter\_score}_{ij} +  
-u_{1j} \text{ pupil}_{ij} +  
-u_{1j} \text{ ct}_{ij} +  
-e_{ij} \\
-$$
-
-`glmer(mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name))`
-
-<!-- + hypox + hypots + tsah + edh  -->
-
-<!-- \gamma_{10} \text{ hypox}_{ij} + -->
-<!-- \gamma_{10} \text{ hypots}_{ij} + -->
-<!-- \gamma_{10} \text{ tsah}_{ij} + -->
-<!-- \gamma_{10} \text{ edh}_{ij} + -->
-
-<!-- u_{1j} \text{ hypox}_{ij} + -->
-<!-- u_{1j} \text{ hypots}_{ij} + -->
-<!-- u_{1j} \text{ tsah}_{ij} + -->
-<!-- u_{1j} \text{ edh}_{ij} + -->
-
-```{r impact, echo=FALSE, fig.width=6, fig.height=8}
-# load data
-data("impact", package = "metamisc")
-impact_NA <- readRDS("../Data/impact_NA.RDS") %>% 
-  mutate(name = as.integer(name),
-         motor_score = as.numeric(motor_score))#%>% select(!c(hypox, hypots, tsah, edh))
-
-# # descriptive statistics
-# by(impact, impact$name, summary) 
-# psych::describe(impact)[,c(2:5,8:9)]
-# # missingness
-plot_pattern(impact_NA, rotate = TRUE) 
-# TODO: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters like shepley plot
-# TODO: make missingness % gradient for clusters (ie the number of clusters for which this variable is missing).
+[TODO: add more info about the complete data.] The analysis model for this dataset is a prediction model with `mort` as the outcome.
+In this tutorial we'll estimate the adjusted prognostic effect of `ct` on unfortunate outcomes.
+The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model:
+```{r mod}
+mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name)
 ```
 
-```{r impact_md, echo=FALSE}
-# # md pattern in first cluster
-# plot_pattern(impact_NA[impact_NA$name == "TINT", ])
-# plot_pattern(impact_NA[impact_NA$name == "APOE", ])
+Note that variables `hypots`, `hypox`, `tsah` and `edh` are not part of the analysis model, and may thus serve as auxiliary variables for imputation.
+
+## Setup
+
+[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
+```{r env, message=FALSE, warning=FALSE}
+set.seed(2022)
+library(mice)         # for imputation
+library(ggmice)       # for visualization
+library(ggplot2)      # for visualization
+library(dplyr)        # for data wrangling
+library(lme4)         # for multilevel modeling
+library(broom.mixed)  # for multilevel estimates
+library(mitml)        # for multilevel pooling
 ```
 
-<!-- Resche-Rigon and White (2018) proposed a two-stage estimator.
-At step 1, a linear regression model is fitted to each observed cluster.
-Any sporadically missing data are imputed, and the model per cluster ignores any systematically missing variables.
-At step 2, estimates obtained from each cluster are combined using meta-analysis.
-Systematically missing variables are modeled through a linear random effect model across clusters.
+The `impact` data included in the \pkg{metamisc} package is a complete data set.
+The original data has already been imputed once (Steyerberg et al, 2008).
+For the purpose of this tutorial we have induced missingness (mimicking the missing data in the original data set before imputation).
+The resulting incomplete data can be accessed from [zenodo link to be created](https://zenodo.com).
+
+Load the complete and incomplete data into the R workspace:
+```{r data, eval = FALSE}
+data("impact", package = "metamisc")
+dat <- read.table("link/to/the/data.txt")
+```
+
+```{r data-actually, include=FALSE}
+data("impact", package = "metamisc")
+dat <- readRDS("../Data/impact_incomplete.RDS")
+```
+
+We will use the following estimates as comparative truth in this tutorial [TODO: make this a table or forest plot instead, to see if there is heterogeneity in the association of `ct` with `mort`]:
+```{r truth}
+fit <- glmer(mod, family = "binomial", data = impact) 
+tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
+```
+
+[TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account.
+If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.]
+
+[TODO: add descriptive statistics of the complete and incomplete data.]
+
+## Missingness
+
+To explore the missingness, it is wise to look at the missing data pattern:
+```{r pattern, fig.height=7.1}
+plot_pattern(dat, rotate = TRUE)
+```
+
+This shows... [TODO: fill in that we need to impute `ct` and `pupil`.]
+
+To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables.
+To see whether the missingness depends on the observed values of other variables, we...
+[TODO: fill in that we can test this statistically or use visual inspection (e.g.
+a histogram faceted by the missingness indicator).]
+
+We should impute the variables `ct` and `pupil` and any auxiliary variables we might want to use to impute these incomplete analysis model variables.
+We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations:
+```{r impact_corr}
+plot_corr(dat, rotate = TRUE)
+```
+
+This shows us that `hypox` and `hypot` would not be useful auxiliary variables for imputing `ct`.
+Depending on the minimum required correlation, `tsah` could be useful, while `edh` has the strongest correlation with `ct` out of all the variables in the data and should definitely be included in the imputation model.
+For the imputation of `pupil`, none of the potential auxiliary variables has a very strong relation, but `hypots` could be used.
+We conclude that we can exclude `hypox` from the data, since this is neither an analysis model variable nor an auxiliary variable for imputation:
+```{r}
+dat <- select(dat, !hypox)
+```
+
+<!-- We'll highlight one other variable to illustrate, but ideally one would inspect all relations.
+The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'.
+This can be evaluated statistically, but visual inspection usually suffices.
+We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.
 -->
 
-```{r impact_cca, echo=FALSE}
-mod_cca <- glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
-mod_cca %>% broom.mixed::tidy()
-# mod_cca_dummy <- glm(mort ~ type + age + motor_score + pupil + ct + name, family = "binomial", data = impact_NA) 
-# mod_cca_dummy %>% broom.mixed::tidy()
+## Complete case analysis
 
-# use glm with name as dummy
-# only random effect on ct, not the other variables
-# make forest plot of ct variable
-# is there heterogen in association of ct with mort?
-# is there any random effect on that variable at all?
-# 
+As previously stated, complete case analysis lowers statistical power and may bias results. The complete case analysis estimates are:
+```{r cca}
+fit <- glmer(mod, family = "binomial", data = na.omit(dat)) 
+tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
 ```
+
+As we can see... [TODO: fill in.]
 
@@
 
-We should impute the variables `ct` and `pupil` and any auxiliary variables we might want to use to impute these incomplete analysis model variables.
-We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations:
+## Imputation model
 
-```{r impact_corr}
-plot_corr(impact_NA, label = T)
+<!-- Table 7.1: Questions to gauge the complexity of a multilevel imputation task. -->
+<!-- 1.	Will the complete-data model include random slopes? -->
+<!-- 2.	Will the data contain systematically missing values? -->
+<!-- 3.	Will the distribution of the residuals be non-normal? -->
+<!-- 4.	Will the error variance differ over clusters? -->
+<!-- 5.	Will there be small clusters? -->
+<!-- 6.	Will there be a small number of clusters? -->
+<!-- 7.	Will the complete-data model have cross-level interactions? -->
+<!-- 8.	Will the dataset be very large? (fimd, section 7.3)-->
+
+The first imputation model that we'll use is likely to be invalid.
+We do *not* use the cluster identifier `name` as imputation model predictor.
+With this model, we ignore the multilevel structure of the data, despite the high ICC.
+This assumes exchangeability between units.
+We include it purely to illustrate the effects of ignoring the clustering in our imputation effort.
+We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).
+
+---
+
+\begin{center}
+
+Updated until here!
+
+\end{center}
+
+---
+
+Create a methods vector and predictor matrix, and make sure `name` is not included as predictor:
+```{r ignore}
+meth <- make.method(dat) # methods vector
+pred <- quickpred(dat)   # predictor matrix
+plot_pred(pred)
 ```
 
-This shows us that `hypox` and `hypot` would not be useful auxiliary variables for imputing `ct`.
-Depending on the minimum required correlation, `tsah` could be useful, while `edh` has the strongest correlation with `ct` out of all the variables in the data and should definitely be included in the imputation model.
-For the imputation of `pupil`, none of the potential auxiliary variables has a very strong relation, but `hypots` could be used.
-We conclude that we can exclude `hypox` from the data since this is neither an analysis model variable, nor an auxiliary variable for imputation.
+<!-- We'll now use `school` as a predictor to impute all other variables.
+This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
+But at least, it includes some multilevel aspect.
+This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16].
+Colloquially, this is 'multilevel imputation for dummies'.
+-->
 
-```{r impact-pred}
-impact_NA <- select(impact_NA, -c(hypox, hypots))
-pred <- quickpred(impact_NA)
+<!-- Add: doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->
+
+<!-- The more the random effects are of interest, the more you need ml models. -->
+
+```{r impact-pred, include=FALSE, eval=FALSE}
+pred <- quickpred(dat)
 plot_pred(pred)
@@
 
-meth <- make.method(impact_NA)
+meth <- make.method(dat)
 meth[meth != ""] <- "2l.pmm"
@@
 
-```{r imp_impact, eval = FALSE}
-imp <- mice(impact_NA, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
-# imp_impact <- impact_NA %>%
-#   mutate(name = as.integer(name), motor_score = as.numeric(motor_score)) %>%
-#   mice::mice(., m = 2, maxit = 1, method = meth, predictorMatrix = pred)
-# look at jomo for categorical variables?
-# semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach
-# pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters
-# make assumptions of these methods explicit!
+[TODO: mutate data to get the right data types for imputation (e.g. integer for clustering variable).]
+
+```{r imp_impact, eval = FALSE, include=FALSE}
+imp <- mice(dat, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
 ```
-```{r impact-fit, eval = FALSE}
+
+```{r impact-fit, eval = FALSE, include=FALSE}
 fit <- imp_impact_pmm %>% 
@@
 
-Compare estimates to complete data
-
-```{r}
-fit <- glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial", data = impact) 
-tidy(fit)
-```
-
 # Discussion
@@
 - JOMO in \pkg{mice} -> on the side for now
+<!-- # look at jomo for categorical variables? -->
+<!-- # semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach -->
+<!-- # pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters -->
+<!-- # make assumptions of these methods explicit! -->
 

### 2022-04-06 hiv.Rmd
Clean up manuscript file

--- 2022-02-07 Manuscript/Manuscript.Rmd
@@
 ---
-documentclass: jss
-author:
-  - name: Hanne Oberman
-    affiliation: Utrecht University
-    address: |
-      | Padualaan 14
-      | 3584 CH Utrecht
-    email: \email{h.i.oberman@uu.nl}
-    url: https://hanneoberman.github.io/
-  - name: Johanna Mu\~noz Avila
-    affiliation: University Medical Center Utrecht \AND
-  - name: Valentijn de Jong
-    affiliation: University Medical Center Utrecht 
-  - name: Gerko Vink
-    affiliation: Utrecht University \AND
-  - name: Thomas Debray
-    affiliation: University Medical Center Utrecht 
-title:
-  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
-  plain:     "Imputation of Incomplete Multilevel Data with mice"
-  short:     "Multilevel \\pkg{mice}"
-abstract: >
-  Multilevel data is not spared the ubiquitous problem of missing information.
-This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
-Including methods for ignorable and non-ignorable missingness.
-Footnotes in the current version show work in progress/under construction.
-The last section is not part of the manuscript, but purely for reminders.
-keywords:
-  # at least one keyword must be supplied
-  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
-  plain:     [missing data, multilevel, clustering, mice, R]
-preamble: >
-  \usepackage{amsmath}
-header-includes: 
- - \usepackage{graphicx} 
- - \usepackage{mathtools}
- - \usepackage{ulem}
-output: rticles::jss_article
-bibliography: ../References/multilevelmice.bib
-editor_options: 
-  chunk_output_type: inline
+  title: "Removed case studies"
+author: "Hanne Oberman"
+date: "6-4-2022"
+output: html_document
 ---
 
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
+
 ```{r, setup, include=FALSE}
+knitr::opts_chunk$set(message=FALSE, warning=FALSE)
 options(prompt = 'R> ', continue = '+ ')
@@
 plot_col <- mice:::mdc(1:2) %>% setNames(c("observed", "missing"))
-```
 
-# Introduction
-
-In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.
-For example, students clustered within schools, or patients clustered within studies in individual patient data meta-analyses.
-Analyzing such multilevel data requires specialized techniques that take the clustered structure into account, since ignoring it may be harmful to the statistical inferences and can yield biased results.
-Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present.
-The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster [@hox17].
-These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
-
-The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
-The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.
-Multilevel models typically accommodate for this variability by including separate intercepts for each cluster.
-Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters.
-Additionally, models may include random effects of predictors across clusters, and random error terms (heterogeneous residual error variances; see e.g.
-@hox17 and @jong21).
-There are many names for models that take clustering into account.
-Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
-
-## Incomplete multilevel data
-
-```{r patterns, fig.height=3, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
-plot_md_set()
-# TODO: make cluster part f the data (blue)
-# TODO: dots between 2 and N, for each variable
-```
-
-The process of analyzing multilevel data is further complicated when not all data entries are observed.
-Just as with single level data, missingness may occur at the unit level.
-But with multiple levels of data comes the potential for clustered missingness.
-Missingness in multilevel data can therefore be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13].
-We have visualized the difference between these types of missingness in Figure 1.
-The figure shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns.
-Column `X1` is completely observed.
-Column `X2` is systematically missing and column `X3` is sporadically missing.
-Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters).
-In Figure 1, the former would imply that the unobserved values for units 3 and 4 on column `X2` are identical.
-With the latter, the values would differ.
-The optimal strategy for dealing with the missingness may therefore depend on the observed missing data pattern.
-
-Another characteristic of the missing data to take into account in analyses is the mechanism behind the missingness.
-Although the essence of the true non-response mechanism may not be known, it can be inferred or assumed to be one of the following:
-
-- Missing Completely At Random (MCAR), where the probability to be missing is equal across all data entries;
-- Missing At Random (MAR), where the probability to be missing depends on observed information;
-- Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94].
-Depending on the assumed missingness mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.
-
-Ignoring the missingness in analyses can be extremely harmful to inferences.
-Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power.
-Instead, the missingness should be accommodated \emph{before} or _within_ the analysis of scientific interest.
-Especially the former is very generic and popular and is widely known as imputation.
-Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed.
-The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
-\pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
-In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.
-
-<!-- "We find it useful to distinguish the missingness pattern, which describes which values are missing and observed in the data matrix and the missingness mechanism (or mechanisms), which concerns the relationship between missingness and the values of variables in the data matrix" (Little and Rubin, p.
-8).
--->
-
-## Aim of this paper
-
-This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}.
-We provide practical guidelines and code snippets for different missing data situations.
-For reasons of brevity, we focus on imputation by chained equations exclusively (although the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well).
-Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and the empirical work by @audi18 and @grun18.
-Note that this tutorial paper assumes a basic level of knowledge on multilevel models.
-We're providing an overview of implementations.
-It's up-to the reader to decide which multilevel strategy suits their data.
-We won't go into detail for the different methods (and equations).
-This paper is just a software tutorial, so we'll keep it practical.
-Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the `magrittr` package and avidly used among `tidy` workflows, and the `lme4` notation for multilevel models.
-
-To illustrate how to impute incomplete multilevel data, we structure this tutorial around three case studies:
-
-- `mice::popmis` (simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, with univariate MAR missingness);
-- `GJRM::hiv` (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, with univariate MNAR missingness)
-- `metamisc::impact` (empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, without `NA`s).
-
-In each case study we highlight one aspect of the imputation workflow.
-With the `mice::popmis` data, we show different ways of designing an imputation model and what happens if you misspecify this model.
-With the `GJRM::hiv` data we extend the imputation model to include Heckman-type selection-inclusion methods.
-And with the `metamisc::impact` data we provide a example of multivariate missingness in real-world data.
-For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data: A.
-Choose an analysis model (so the imputation model will be compatible with the analyses); B.
-Evaluate the incomplete data; C.
-Develop the imputation model(s); D.
-Impute the missingness; E.
-Evaluate the imputations.
-
-# How (not) to impute (Case Study I: Popularity)
-
-```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
-# randomly select a few classes
-# clusters <- sample(1:100, 10)
-
-# load data 
-# popmis <- mice::popmis 
-#readRDS("../Data/popNCR2.RDS") %>% 
-  #filter(class %in% clusters) %>% 
-  #mutate(class = as.factor(class))
-popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE) %>% 
+# data
+popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
   mutate(school = SCHOOL,
-         popular = as.numeric(POPULAR),
+         popular = POPULAR,
          sex = SEX,
@@
          .keep = "none")
-comp_est <- lme4::lmer(popular ~ 1 + sex + texp + (1 + sex | school), popcomp) %>% broom.mixed::tidy()
-```
-
-In this section we'll go over the different steps involved with imputing incomplete multilevel data.
-The data we're using is the `popmis` dataset from the `mice` package.
-This is a simulated dataset with pupils ($n = 2000$) clustered within schools ($N = 100$).
-In this tutorial we'll use the following variables:
-
-  - `school`,     school identification number (clustering variable);
-  - `popular`,    pupil popularity (self-rating between 0 and 10; unit-level);
-  - `sex`,	      pupil sex (0=boy, 1=girl; unit-level);
-  - `texp`,	      teacher experience (in years; cluster-level).
-
-The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction.
-The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`.
-The regression equation and `lme4` notation for this model are
-
-$$
-\text{popular}_{ij} =
-\gamma_{00} + 
-\gamma_{10} \text{ sex}_{ij} + 
-\gamma_{01} \text{ texp}_{j} + 
-\gamma_{11} \text{ texp}_{j} \times \text{sex}_{ij} + 
-u_{0j} + 
-u_{1j} \text{ sex}_{ij} + 
-e_{ij} \\
-$$
-
-$$
-\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
-$$
-
-<!-- Intercept-only model: lmer(formula = popular ~ 1 + (1|class)) -->
-<!-- First level predictors: lmer(formula = popular ~ 1 + sex + (1|class)) -->
-<!-- First and second level predictors: lmer(popular ~ 1 + sex + texp + (1 | class)) -->
-<!-- First and second level predictors with random slopes: lmer(formula = popular ~ 1 + sex + texp + (1 + sex | class) -->
-<!-- First and second level predictors with random slopes and cross-level interaction: lmer(formula = popular ~ 1 + sex + texp + sex:texp + (1 + sex | class)) -->
-
-TODO: add all estimimates, seed, and version nr., this also shows the workflow for pooling etc.
-
-Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data.
-The data is created in such a way that the clustering variable `school` explains quite some variance in the outcome variable `popular`.
-We express this using the intraclass correlation,  ICC $=$ `r round(icc(popular ~ as.factor(school), data = popcomp), 2)`.
-We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effect of `sex` on `popular` $\gamma_{10} = 0.843$.
-
-### Incomplete data
-
-Load the data into the environment and select the relevant variables:
-
-```{r}
 popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
-```
-
-```{r pop_pat, echo=FALSE, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
-# missingness
-plot_md_pat(popmis)
-# TODO: add pattern by cluster, text mining example square size
-# TODO: try to make it a shepley plot
-
-# popcomp <- popcomp %>% 
-#   cbind(R = is.na(popmis$popular))
-# t.test(x = popcomp$popular[popcomp$R], y = popcomp$popular[!popcomp$R]) 
-# ggplot(popcomp, aes(y=popular, x = R)) +
-#   geom_boxplot()
-
-```
-
-The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}.
-The ICC in the incomplete data is `r round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.
-
-To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables.
-We'll highlight only one other variable to illustrate, but ideally one would inspect all relations.
-The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'.
-This can be evaluated statistically, but visual inspection usually suffices.
-We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.
-
-In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
-This would indicate a right-tailed MAR missingness.
-(In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `texp`.
-<!-- There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well. -->
-
-```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-# popmis %>% 
-#   filter(as.numeric(school) < 5) %>% 
-#   ggplot(aes(x = teachpop, color = school)) +
-#     geom_density() +
-#     facet_wrap(~factor(is.na(popular), labels = c("popular missing", "popular observed")), ncol = 1) +
-#     # scale_color_manual(values = plot_col, name = "popular") +
-#     theme_classic() # +
-#     # theme(legend.position = "bottom")
-
-plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
-# TODO: check if up-side-down plot works
-# TODO: check smoothing in geom_density function and make it the inverse of the sample size
-# TODO: add functions to mice
-# TODO: add title and informative legend
-# TODO: add facets for some clusters
-# or add propensity score distribution `is.na(popular) ~ .`
-# TODO: make it average of cluster dens, not marginal
-# alternatively add a quartile line around the density with geom_ribbon
-
-## comments 24/1
-# pop obs and pop miss in legend
-# add to caption
-# add fill
-# overlap coeff??
-# make densities proportional
-
-```
-
-### Complete case analysis (not recommended)
-
-Complete case analysis ignores the observations with missingness altogether, which may even introduce bias in MCAR situations.
-
-### Imputation ignoring the cluster variable (not recommended)
-
-The first imputation model that we'll use is likely to be invalid.
-In this model, we ignore the multilevel structure of the data, despite the high ICCs.
-This is purely to illustrate the effects of ignoring the clustering in our imputation effort.
-
-We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`.
-We do not use the observation identifier `pupil` or cluster identifier `school` as predictors to impute other variables.
-
-```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
-# dry run to get imputation parameters
-ini <- mice(popmis, maxit = 0)
-
-# extract predictor matrix and adjust
-pred <- ini$pred
-pred[, c("school")] <- 0
-
-# impute the data, ignoring the cluster structure
-imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
-```
-
-```{r pop_ignored_eval, echo=FALSE, message=FALSE, warning=FALSE}
-# # check convergence of the imputation model
-# plot(imp_ignored)
-
-# # compare descriptives before and after imputation
-# psych::describe(popmis)[, c("n", "mean", "median", "min", "max", "sd")]
-# psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
-# TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
-# TODO: pool mean median and sd
-# feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
-# TODO: add FMI for each of the estimates? at least for the mean
-
-# further inspection of the imputations
-plot_imps(imp_ignored, type = "stripplot", x = "popular")
-# TODO: think about multimodality
-
-# # compare ICCs before and after imputation
-# ICCs <- data.frame(
-#   vars = c("popular", "teachpop", "texp"), 
-#   CCA = c(icc(popular ~ school, popmis), 
-#                icc(teachpop ~ school, popmis),
-#                icc(texp ~ school, popmis)), 
-#   ignored = c(icc(popular ~ school, complete(imp_ignored)), 
-#               icc(teachpop ~ school, complete(imp_ignored)), 
-#               icc(texp ~ school, complete(imp_ignored)))
-#   )
-# ICCs
-
-#TODO: think about interactive clusters for the strippplots
-#TODO: switch off outliers in boxplot?
-```
-
-As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `school`.
-This tells us that the multilevel structure of the data should be taken into account.
-If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.
-
-We can also observe that the teacher experience increases slightly after imputation.
-This is due to the MNAR missingness in `texp`.
-Higher values for `texp` have a larger probability to be missing.
-This may not a problem, however, if at least one pupil in each school has teacher experience recorded, we can deductively impute the correct (i.e.
-true) value for every pupil in the school.
-
-Add: Assumes exchangability between units.
-
-### Imputation with the cluster variable as predictor (not recommended)
-
-We'll now use `school` as a predictor to impute all other variables.
-This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
-But at least, it includes some multilevel aspect.
-This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende2016].
-Colloquially, this is 'multilevel imputation for dummies'.
-
-Add: doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.
-
-The more the random effects are of interest, the more you need ml models.
-
-<!-- "Beyond the potential bias, there are other reasons why a multilevel imputation model should be preferred if possible.
-First, the model is guaranteed to produce unbiased results even in extreme cases with very high missingness rates and/or small intraclass correlations.
-Second, with the multilevel model, it is possible to include additional variables that are constant on the cluster level, for example, information on the teachers if the cluster levels are chosen to be students within classes or background information on the schools if the clustering is modeled as students within schools.
-This approach is not possible with fixed effects imputation.
-While this limitation should not be problematic in terms of bias (the combined school effects are still modeled through the cluster specific intercepts), using this additional information will make the random effects imputation model more efficient." -->
-
-```{r pop_predictor, message=FALSE, warning=FALSE}
-# adjust the predictor matrix
-pred <- ini$pred 
-# pred[, "pupil"] <- 0
-
-# impute the data, cluster as predictor
-imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
-
-```
-
-```{r pop_predictor_eval, echo=FALSE, message=FALSE, warning=FALSE}
-# # check logged events
-# head(imp_predictor$loggedEvents)
-## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and teachpop.
-
-# # check convergence of the imputation model
-# plot(imp_predictor)
-
-# # compare descriptives before and after imputation
-# psych::describe(popmis)[, c("n", "mean", "median", "min", "max", "sd")]
-# psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
-
-# further inspection of the imputations
-plot_imps(imp_predictor, x = "popular", type = "stripplot")
-
-# # compare ICCs before and after imputation
-# ICCs <- ICCs %>% mutate(
-#            predictor = c(icc(popular ~ school, complete(imp_predictor)), 
-#                         icc(teachpop ~ school, complete(imp_predictor)), 
-#                         icc(texp ~ school, complete(imp_predictor)))
-#            )
-# ICCs
-```
-
-Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MAR.
-
-The ICCs are way more in line with the ICCs in the incomplete data. But this is a quick and dirty way of imputing multilevel data. We *should* be using a multilevel model.
-
-### Imputation with random effects
-
-With `2l.norm` we impute the outcome with a multilevel model  assuming random slopes for each variable in the imputation model and homogeneous within-cluster variance.
-
-```{r pop_norm}
-pred <- ini$pred
-pred["popular", ] <- c(-2, 2, 2, 2) 
-#-2 for the cluster variable, 2 for random effects
-meth <- ini$meth
-meth <- c("", "2l.norm", "",  "")
-imp_norm_2l <-
-  mice(
-    popmis %>% mutate(school = as.integer(school)),
-    pred = pred,
-    meth = meth,
-    maxit = 1,
-    print = FALSE
-  )
-```
-
-```{r pop_norm_eval, echo=FALSE}
-# plot(imp_norm)
-plot_imps(imp_norm_2l, x = "popular", type = "stripplot")
-
-# ICCs <- ICCs %>% mutate(
-#            norm = c(icc(popular ~ as.factor(school), complete(imp_norm_2l)), 
-#                     icc(teachpop ~ as.factor(school), complete(imp_norm_2l)), 
-#                     icc(texp ~ as.factor(school), complete(imp_norm_2l)))
-#            )
-# ICCs
-```
-
-### Imputation with random effects and heterogeneity
-
-This method assumes random slopes for each variable in the imputation model. In contrast to `2l.norm` this method allows a cluster-specific residual error variance.
-
-Schafer book fortran workflows.
-
-<!-- "By default, 2l.pan includes the intercept as both a fixed and a random effect.
-...
-There are 5 different codes you can use in the predictor matrix for variables imputed with 2l.pan.
-The person identifier is coded as -2 (this is different from 2l.norm).
-To include predictor variables with fixed or random effects, these variables are coded with 1 or 2, respectively.
-If coded as 2, the corresponding fixed effect is automatically included.
-In addition, 2l.pan offers the codes 3 and 4, which have similar meanings as 1 and 2 but will include an additional fixed effect for the person mean of that variable.
-This is useful if you're trying to model within- and between-person effects of time-varying predictor variables." -->
-
-```{r pop_pan}
-pred["popular", ] <- c(-2, 2, 1, 2)
-meth <- c("", "2l.pan", "", "")
-imp_pan_2l <-
-  mice(
-    popmis %>% mutate(school = as.integer(school)),
-    pred = pred,
-    meth = meth,
-    maxit = 1,
-    print = FALSE
-  )
-
-```
-
-```{r pop_pan_eval, echo=FALSE}
-# plot(imp_pan)
-plot_imps(imp_pan_2l, x = "popular", type = "stripplot")
-# ICCs <- ICCs %>% mutate(
-#            pan = c(icc(popular ~ as.factor(school), complete(imp_pan_2l)), 
-#                     icc(teachpop ~ as.factor(school), complete(imp_pan_2l)), 
-#                     icc(texp ~ as.factor(school), complete(imp_pan_2l)))
-#            )
-# ICCs
 ```
@@
 This means the assumptions for the Heckman-type selection model are met.
-
-# How to handle multivariate missingness (Case study III: IMPACT)
-
-`impact` is traumatic brain injury data with patients, $n = 11022$, clustered in studies, $N = 15$. With the following 11 variables:
-
-  - `name` Name of the study,
-  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
-  - `age` Age of the patient,
-  - `motor_score` Glasgow Coma Scale motor score,
-  - `pupil` Pupillary reactivity,
-  - `ct` Marshall Computerized Tomography classification,
-  - `hypox` Hypoxia (0=no, 1=yes),
-  - `hypots` Hypotension (0=no, 1=yes),
-  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
-  - `edh` Epidural hematoma (0=no, 1=yes),
-  - `mort` 6-month mortality (0=alive, 1=dead).
-
-The data is already imputed (Steyerberg et al, 2008), so we'll induce missingness ourselves.
-For example, MAR missingness varying by cluster.^[Observed data pattern should differ per cluster.
-So, in cluster 1, the missingness would depend on age, but not in cluster two.
-Split the dataframe and run `ampute()` on each cluster.]
-
-```{r impact, echo=FALSE, fig.width=5}
-# load data
-data("impact", package = "metamisc")
-impact_NA <- readRDS("../Data/impact_NA.RDS")
-# # descriptive statistics
-# by(impact, impact$name, summary) 
-# psych::describe(impact)[,c(2:5,8:9)]
-# # missingness
-plot_md_pat(impact_NA)
-
-# md pattern in first cluster
-plot_md_pat(impact_NA[impact_NA$name == "TINT", ])
-plot_md_pat(impact_NA[impact_NA$name == "APOE", ])
-# TODO: rotate variable labels
-
-# TODO: check md pattern in clusters, should differ, otherwise create new missingness
-# MAYBE: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters instead
-# TODO: e-mail Joop for DGM or look at adrew and jennifer book for data
-# sim.multi from psych
-```
-
-TODO: rotate labels, make missingness % gradient for clusters?? Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of ct on unfortunate outcomes.
-
-gradient would be the number of clusters for which this variable is missing
-
-# Discussion
-
-- JOMO in \pkg{mice} -> on the side for now
-
-- Additional levels of clustering
-
-- More complex data types: timeseries and polynomial relationship in the clustering.
-
-# Think about
-
-- Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model.
-
-- Adding a `multilevel_ampute()` wrapper function in mice.
-
-- Exporting `mids` objects to other packages like `lme4` or `coxme`?
-
-- Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt.
-
-- env dump in repo
-
-# References

### 2022-04-06 popmis.Rmd
Clean up manuscript file

--- 2022-02-07 Manuscript/Manuscript.Rmd
@@
 ---
-documentclass: jss
-author:
-  - name: Hanne Oberman
-    affiliation: Utrecht University
-    address: |
-      | Padualaan 14
-      | 3584 CH Utrecht
-    email: \email{h.i.oberman@uu.nl}
-    url: https://hanneoberman.github.io/
-  - name: Johanna Mu\~noz Avila
-    affiliation: University Medical Center Utrecht \AND
-  - name: Valentijn de Jong
-    affiliation: University Medical Center Utrecht 
-  - name: Gerko Vink
-    affiliation: Utrecht University \AND
-  - name: Thomas Debray
-    affiliation: University Medical Center Utrecht 
-title:
-  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
-  plain:     "Imputation of Incomplete Multilevel Data with mice"
-  short:     "Multilevel \\pkg{mice}"
-abstract: >
-  Multilevel data is not spared the ubiquitous problem of missing information.
-This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
-Including methods for ignorable and non-ignorable missingness.
-Footnotes in the current version show work in progress/under construction.
-The last section is not part of the manuscript, but purely for reminders.
-keywords:
-  # at least one keyword must be supplied
-  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
-  plain:     [missing data, multilevel, clustering, mice, R]
-preamble: >
-  \usepackage{amsmath}
-header-includes: 
- - \usepackage{graphicx} 
- - \usepackage{mathtools}
- - \usepackage{ulem}
-output: rticles::jss_article
-bibliography: ../References/multilevelmice.bib
-editor_options: 
-  chunk_output_type: inline
+title: "Removed case studies"
+author: "Hanne Oberman"
+date: "6-4-2022"
+output: html_document
 ---
 
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
+
 ```{r, setup, include=FALSE}
+knitr::opts_chunk$set(message=FALSE, warning=FALSE)
 options(prompt = 'R> ', continue = '+ ')
@@
          .keep = "none")
-comp_est <- lme4::lmer(popular ~ 1 + sex + texp + (1 + sex | school), popcomp) %>% broom.mixed::tidy()
+popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
+```
+
+<!-- ^[Note that the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well but is outside the scope of this tutorial.].
+Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work by @audi18 and @grun18.
+Please note that this tutorial paper assumes a basic level of knowledge on multilevel models.^[Note to self: We're providing an overview of implementations.
+It's up-to the reader to decide which multilevel strategy suits their data.
+We won't go into detail for the different methods (and equations).
+This paper is just a software tutorial, so we'll keep it practical.] Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the \pkg{magrittr} package, and the \pkg{lme4} notation for multilevel models.^[TODO: Add environment info, seed and version number(s) somewhere!] -->
+
+We illustrate how to impute incomplete multilevel data by means of three case studies:
+
+- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, @mice];
+- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, @GJRM];
+- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc].
+
+For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
+With the `popmis` data, we show how (and how not) to develop an imputation model.
+With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods.
+With the `impact` data we provide an example of multivariate missingness in real-world data.
+Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph or 'translation table' linking multilevel equations to \pkg{lme4} formulas.
+Use betas instead of gamma's and mu's.
+Add interpretation of values in predictormatrix (-2 for the cluster variable, 2 for random effects).
+Add ICC and congeneality here as well.
+And make missingness mechanism table as well.]
+
+# Case Study I: How (not) to impute
+
+```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
+# estimates complete popularity data
+est_comp <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popcomp) 
+# fixed effects only
+est_comp_fixed <- est_comp %>% 
+  broom.mixed::tidy("fixed", conf.int = TRUE) %>% 
+  mutate(term = term,
+         estimate  = paste0(
+           round3(estimate), " [", 
+           round3(conf.low), ", ", 
+           round3(conf.high), "]"), 
+         .keep = "none") %>% 
+  as.data.frame()
+# # random effects
+# est_comp %>% 
+#   broom.mixed::tidy("ran_pars") %>% 
+#   mutate(term = term, #factor(term, levels = unique(.$term), labels = c()),
+#          estimate  = round3(estimate), 
+#          .keep = "none")
+# # rename broom.mixed estimates
+# lvl = c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation")
+# lbl = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor.
+cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error")
+
+# compare complete data to missingness
+# popcomp <- popcomp %>% 
+#   cbind(R = is.na(popmis$popular))
+# t.test(x = popcomp$popular[popcomp$R], y = popcomp$popular[!popcomp$R]) 
+# ggplot(popcomp, aes(y=popular, x = R)) +
+#   geom_boxplot()
 ```
@@
 
-TODO: add all estimimates, seed, and version nr., this also shows the workflow for pooling etc.
-
 Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data.
@@
 We express this using the intraclass correlation,  ICC $=$ `r round(icc(popular ~ as.factor(school), data = popcomp), 2)`.
-We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effect of `sex` on `popular` $\gamma_{10} = 0.843$.
+We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effects:
+
+```{r comp_est, echo=FALSE}
+est_comp_fixed %>% setNames(c("", "Estimate with 95% CI"))
+# TODO: make this a figure with the different methods as facets
+```
 
@@
 ```{r}
-popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
+popmis <- popmis[, c("school", "popular", "sex", "texp")] 
 ```
+Plot the missing data pattern:
 
-```{r pop_pat, echo=FALSE, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
-# missingness
-plot_md_pat(popmis)
-# TODO: add pattern by cluster, text mining example square size
-# TODO: try to make it a shepley plot
-
-# popcomp <- popcomp %>% 
-#   cbind(R = is.na(popmis$popular))
-# t.test(x = popcomp$popular[popcomp$R], y = popcomp$popular[!popcomp$R]) 
-# ggplot(popcomp, aes(y=popular, x = R)) +
-#   geom_boxplot()
-
+```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
+plot_pattern(popmis)
 ```
@@
 
-In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
+Plot the histogram for teacher experience conditional on the missingness indicator of `popular`:
+
+```{r pop-hist}
+ggmice(popmis, aes(texp)) +
+  geom_histogram(fill = "white") +
+  facet_grid(. ~ is.na(popular), labeller = label_both)
+```
+
+TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?
+
+```{r pop-hist-clust, include=FALSE}
+# popmis %>% 
+#   filter(school < 5) %>% 
+# ggmice(., aes(texp)) +
+#   geom_histogram(fill = "white") +
+#   facet_grid(school ~ is.na(popular), labeller = label_both)
+# # does not make sense because texp is a cluster level variable!
+```
+
+This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
+
+<!-- In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`.
 This would indicate a right-tailed MAR missingness.
 (In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `texp`.
-<!-- There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well. -->
+-->
 
 ```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
-# popmis %>% 
-#   filter(as.numeric(school) < 5) %>% 
-#   ggplot(aes(x = teachpop, color = school)) +
-#     geom_density() +
-#     facet_wrap(~factor(is.na(popular), labels = c("popular missing", "popular observed")), ncol = 1) +
-#     # scale_color_manual(values = plot_col, name = "popular") +
-#     theme_classic() # +
-#     # theme(legend.position = "bottom")
-
-plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
+# plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
 # TODO: check if up-side-down plot works
@@
 
-Complete case analysis ignores the observations with missingness altogether, which may even introduce bias in MCAR situations.
+Complete case analysis ignores the observations with missingness altogether, which lowers statistical power and may even introduce bias in MCAR situations.
+
+```{r pop-cca, echo=FALSE}
+est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis)
+
+```
 
@@
 The first imputation model that we'll use is likely to be invalid.
-In this model, we ignore the multilevel structure of the data, despite the high ICCs.
-This is purely to illustrate the effects of ignoring the clustering in our imputation effort.
+We do *not* use the cluster identifier `school` as imputation model predictor.
+With this model, we ignore the multilevel structure of the data, despite the high ICC.
+This assumes exchangeability between units.
+We include it purely to illustrate the effects of ignoring the clustering in our imputation effort.
+We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).
 
-We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`.
-We do not use the observation identifier `pupil` or cluster identifier `school` as predictors to impute other variables.
+Create a methods vector and predictor matrix for `popular`, and make sure `school` is not included as predictor:
 
-```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
-# dry run to get imputation parameters
-ini <- mice(popmis, maxit = 0)
+```{r pop-ignored-pred, echo=TRUE, message=FALSE, warning=FALSE}
+meth <- make.method(popmis) # methods vector
+pred <- quickpred(popmis)   # predictor matrix
+plot_pred(pred)
+```
 
-# extract predictor matrix and adjust
-pred <- ini$pred
-pred[, c("school")] <- 0
+Impute the data, ignoring the cluster structure:
+```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
+imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+```
 
-# impute the data, ignoring the cluster structure
-imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+TODO: remove the broom.mixed output, use mitml only
+
+Analyze the imputations:
+```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
+fit_ignored <- imp_ignored %>% 
+  with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
+```
+
+```{r eval=FALSE, include=FALSE}
+# fit_ignored %>% 
+#   pool() %>% 
+#   broom.mixed::tidy()
 ```
@@
 #TODO: think about interactive clusters for the strippplots
-#TODO: switch off outliers in boxplot?
 ```
 
-As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `school`.
-This tells us that the multilevel structure of the data should be taken into account.
-If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.
-
-We can also observe that the teacher experience increases slightly after imputation.
+<!-- We can also observe that the teacher experience increases slightly after imputation.
 This is due to the MNAR missingness in `texp`.
@@
 true) value for every pupil in the school.
+-->
 
-Add: Assumes exchangability between units.
+<!-- pmm: Single-level predictive mean matching with the school indicator coded as a dummy variable; -->
+<!-- 2l.pan: Multilevel method using the linear mixed model to draw univariate imputations; -->
+<!-- 2l.norm: Multilevel method using the linear mixed model with heterogeneous error variances; -->
+<!-- 2l.pmm: Predictive mean matching based on predictions from the linear mixed model, with random draws from the regression coefficients and the random effects, using five donors. -->
 
@@
 
+"Van Buuren (2011) considered the homoscedastic linear mixed model as invalid for imputing incomplete predictors, and investigated only the 2l.norm method, which allows for heterogeneous error variances" [@buur18].
+
 ```{r pop_norm}
-pred <- ini$pred
-pred["popular", ] <- c(-2, 2, 2, 2) 
-#-2 for the cluster variable, 2 for random effects
-meth <- ini$meth
-meth <- c("", "2l.norm", "",  "")
-imp_norm_2l <-
+# adjust the predictor matrix
+pred["popular", ] <- c(school = -2, popular = 0, sex = 2, texp = 2) 
+plot_pred(pred) 
+meth <- make.method(popmis)
+meth["popular"] <- "2l.pmm"
+imp_pmm_2l <-
   mice(
@@
 ```
-
-# How to handle non-random selection (Case study II: HIV)
-
-Data are simulated and included in the `GJRM` package. We will use the following variables:
-
-  - `region` Cluster variable,
-  - `hiv` HIV diagnosis (0=no, 1=yes),
-  - `age` Age of the patient,
-  - `marital` Marital status,
-  - `condom` Condom use during last intercourse,
-  - `smoke` Smoker (levels; inclusion restriction variable).
-
-The imputation of these date is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
-
-<!-- - main model = analysis model = predicting the prevalence of hiv -->
-<!-- - people who have hiv are more reluctant to taking the test, not to reveal their status -->
-<!-- - solution: take variables from analysis model like condom use, marital status, age, etc. that relate to hiv prevalence -->
-<!-- - exclusion restriction variable = instrumental variable (from economics) = e.g.
-how kind the nurse was = smoke in this dataset = should be correlated to the missingness indicator, but *not* the variable itself -->
-<!-- - include these variables in the selection equation -->
-<!-- - take the missingness indicator of the variable of interest for the selection equation as DV -->
-<!-- - heckman models estimate BOTH selection and mean models simultaneously, because linked through the error term -->
-<!-- - if the errors are not correlated, the models are not linked at all -> not MNAR, but just MAR -->
-<!-- - if the rho is significantly different from zero, the errors are correlated and the models are related so we have MNAR -->
-<!-- - now, we add clustering -->
-<!-- - copulas = estimate 2 equations separately and use another equation to link them -->
-<!-- - for each of the clusters, est the copula separately, so we have all coeffs for each cluster, incl the rho and sigma (the errors) -->
-<!-- - put together all of the estimates through a meta-analysis model to get population distribution of the estimates -->
-<!-- - use the population parameter estimates to draw marginal parameters from the distribution to get a cluster/study parameter -->
-<!-- - with systematical miss we can est the values of the imputed values based on the marginal distribution -->
-<!-- - optional to add shrinkage: the marginal in account as well -->
-<!-- - the hiv data was simulated by the copula function/model developer -->
-<!-- - we could add a systematically missing variable in the data, because these are all sporadically -->
-
-```{r hiv, echo=FALSE}
-# data
-data("hiv", package = "GJRM")
-hiv <- hiv %>% 
-  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
-  mutate(region = as.integer(region))
-# md pattern
-plot_md_pat(hiv)
-```
-
-From the missing data pattern we see that we can set `maxit` to 1, since there is only one variable with missingness.
-
-```{r include=FALSE}
-# inclusion restriction variable 
-mod_main <- glm(hiv ~ ., family = "binomial", data = hiv) %>% 
-  broom::tidy(conf.int = TRUE) %>% 
-  .[7, c(2,6,7)] %>% 
-  round(., 3)
-mod_select <- glm(is.na(hiv) ~ ., family = "binomial", data = hiv) %>% 
-  broom::tidy(conf.int = TRUE) %>% 
-  .[7, c(2,6,7)] %>% 
-  round(., 3)
-```
-
-The inclusion restriction variable should be a predictor of the the actual value of the variable of interest, but *not* of missingness indicator for the variable of interest.
-In this case, the data were simulated to adhere to this requirement.
-Namely, $\beta_{smoke}$ = `r mod_main$estimate`, 95% CI [`r paste0(mod_main$conf.low, ", ", mod_main$conf.high)`] for the analysis model (`formula = hiv ~ .`), and $\beta_{smoke}$ = `r mod_select$estimate`, 95% CI [`r paste0(mod_select$conf.low, ", ", mod_select$conf.high)`] for the selection model (`formula = is.na(hiv) ~ .`).
-This means the assumptions for the Heckman-type selection model are met.
-
-# How to handle multivariate missingness (Case study III: IMPACT)
-
-`impact` is traumatic brain injury data with patients, $n = 11022$, clustered in studies, $N = 15$. With the following 11 variables:
-
-  - `name` Name of the study,
-  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
-  - `age` Age of the patient,
-  - `motor_score` Glasgow Coma Scale motor score,
-  - `pupil` Pupillary reactivity,
-  - `ct` Marshall Computerized Tomography classification,
-  - `hypox` Hypoxia (0=no, 1=yes),
-  - `hypots` Hypotension (0=no, 1=yes),
-  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
-  - `edh` Epidural hematoma (0=no, 1=yes),
-  - `mort` 6-month mortality (0=alive, 1=dead).
-
-The data is already imputed (Steyerberg et al, 2008), so we'll induce missingness ourselves.
-For example, MAR missingness varying by cluster.^[Observed data pattern should differ per cluster.
-So, in cluster 1, the missingness would depend on age, but not in cluster two.
-Split the dataframe and run `ampute()` on each cluster.]
-
-```{r impact, echo=FALSE, fig.width=5}
-# load data
-data("impact", package = "metamisc")
-impact_NA <- readRDS("../Data/impact_NA.RDS")
-# # descriptive statistics
-# by(impact, impact$name, summary) 
-# psych::describe(impact)[,c(2:5,8:9)]
-# # missingness
-plot_md_pat(impact_NA)
-
-# md pattern in first cluster
-plot_md_pat(impact_NA[impact_NA$name == "TINT", ])
-plot_md_pat(impact_NA[impact_NA$name == "APOE", ])
-# TODO: rotate variable labels
-
-# TODO: check md pattern in clusters, should differ, otherwise create new missingness
-# MAYBE: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters instead
-# TODO: e-mail Joop for DGM or look at adrew and jennifer book for data
-# sim.multi from psych
-```
-
-TODO: rotate labels, make missingness % gradient for clusters?? Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of ct on unfortunate outcomes.
-
-gradient would be the number of clusters for which this variable is missing
-
-# Discussion
-
-- JOMO in \pkg{mice} -> on the side for now
-
-- Additional levels of clustering
-
-- More complex data types: timeseries and polynomial relationship in the clustering.
-
-# Think about
-
-- Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model.
-
-- Adding a `multilevel_ampute()` wrapper function in mice.
-
-- Exporting `mids` objects to other packages like `lme4` or `coxme`?
-
-- Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt.
-
-- env dump in repo
-
-# References

### 2022-04-06 R/load_impact.R

@@ -124,7 +124,9 @@ ampute_impact <- function(impact, p_obs){
     cbind(id = rownames(.), .) %>% 
     left_join(pr, by = "id") %>% 
     mutate(pupil = ifelse(R_pupil == 1, pupil, NA),
+           pupil = factor(pupil, levels = 1:3, labels = c("Both", "None", "One")),
            ct = ifelse(R_ct == 1, ct, NA),
+           ct = factor(ct, levels = 1:3, labels = c("I/II", "III", "IV/V")),
            hypox = ifelse(R_hypoxia == 1, hypox, NA),
            hypots = ifelse(R_hypotens == 1, hypots, NA),
            tsah = ifelse(R_tsah == 1, tsah, NA),

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

@@
 (``There is no page limit, nor a limit on the number of figures or
-tables'').
+tables''). {[}Just adding some text to get a better guess of what the
+actura abstract will look like: Lorem ipsum dolor sit amet, consectetur
+adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore
+magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco
+laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla
+pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa
+qui officia deserunt mollit anim id est laborum.{]}
 }
@@
 within regions or countries), or when the clustering is non-nested
-(e.g., data from diverse settings and populations within large
-databases). The clustered structure of multilevel data should be taken
-into account when developing analysis models: 1) for the simple reason
-that groups of observations share some common variance, and 2) because
-ignoring multilevel structures can be harmful to the statistical
-inferences and introduce bias in estimators \citep{hox17}. There are
-many names for models that take clustering into account. Some popular
-examples are `multilevel models', `hierarchical models', `mixed effect
-models' and `random effect models'. Table \ref{tab:def} provides an
-overview of some key concepts in multilevel modeling.
+(e.g., electronic health record data from diverse settings and
+populations within large databases). The clustered structure of
+multilevel data should be taken into account when developing analysis
+models: 1) for the simple reason that groups of observations share some
+common variance, and 2) because ignoring multilevel structures can be
+harmful to the statistical inferences and introduce bias in estimators
+\citep{hox17}. There are many names for models that take clustering into
+account. Some popular examples are `multilevel models', `hierarchical
+models', `mixed effect models' and `random effect models'. Table
+\ref{tab:clus} provides an overview of some key concepts in multilevel
+modeling.
 
 \begin{table}[tb]
-\caption{Concepts in multilevel modeling}
-\label{tab:def}
+\caption{Concepts in multilevel methods}
+\label{tab:clus}
 \centering
@@
 which shows an \(n \times p\) set \(\mathbf{X} = X_1, \dots, X_p\), with
-\(n\) units distributed over \(N\) clusters and \(p\) variables. Column
-\(X_1\) is completely observed, column \(X_2\) is systematically missing
-in cluster 2, and column \(X_3\) is sporadically missing. To analyze
-these incomplete data, we have to take the nature of the missingness and
-the cluster structure into account. For example, the sporadic
-missingness in \(X_3\) could be easily amended if this would be a
-cluster-level variable (and thus constant within clusters). We could
+\(n\) units distributed over \(N\) clusters and \(p\) variables.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/patterns-1}
+
+}
+
+\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
+\end{figure}
+\end{CodeChunk}
+
+Column \(X_1\) in Figure 1 is completely observed, column \(X_2\) is
+systematically missing in cluster 2, and column \(X_3\) is sporadically
+missing. To analyze these incomplete data, we have to take the nature of
+the missingness and the cluster structure into account. For example, the
+sporadic missingness in \(X_3\) could be easily amended if this would be
+a cluster-level variable (and thus constant within clusters). We could
 then just extrapolate the true (but missing) value of \(X_3\) for unit 1
@@
 
-\begin{CodeChunk}
-\begin{figure}
+A key characteristic of the missing data to take into account in
+analyses is the mechanism behind the missingness. We distinguish between
+MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice
+this distinction is less clear. Since the essence of the true
+non-response mechanism may not be known, it is generally inferred or
+assumed to be ignorable (i.e., MCAR or MAR). {[}TODO: add that this
+assumption may not always be valid, especially with modern types of big
+dasta sources.{]} Depending on the actual missingness-generating
+mechanism, missing data handling strategies may be more or less
+suitable, see e.g., \citet{yuce08} and \citet{hox15}.
 
-{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/patterns-1}
-
-}
-
-\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
-\end{figure}
-\end{CodeChunk}
+\begin{table}[tb]
+\caption{Concepts in missing data methods}
+\label{tab:miss}
+\centering
+\begin{tabular}{ll}
+\hline
+\textbf{Concept} & \textbf{Details}   \\
+\hline
+MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
+& across all data entries \\
+MAR     & Missing At Random, where the probability to be missing depends on observed \\
+& information \\
+MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
+& depends on unrecorded information, making the missingness non-ignorable \\
+& [@rubi76; @meng94]. \\
+& [TODO: add congeniality.] \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 replaced multiple times, the resulting inferences may validly convey the
-uncertainty due to missingness \citep[c.f.][]{rubi87}. The \proglang{R}
+uncertainty due to missingness \citep[c.f.][]{rubi76}. The \proglang{R}
 package \pkg{mice} has become the de-facto standard for imputation by
@@
 
+{[}TODO: clarify why clustering is relevant during imputation, and why
+this exposes the need for specialized imputation methods and more
+attention during their implementation (``thou shall not simply run
+\texttt{mice()} on any incomplete dataset''). And add overview of
+possible predictor matrix values.{]}
+
 \hypertarget{aim-of-this-paper}{%
@@
 This papers serves as a tutorial for imputing incomplete multilevel data
-with \pkg{mice}. We provide practical guidelines and code snippets for
-different missing data situations, including missing not at random
-(MNAR) mechanisms \citep[where the probability to be missing depends on
-unrecorded information, making the missingness
-non-ignorable,][]{rubi76, meng94}. For reasons of brevity, we focus on
-imputation by chained equations wit \pkg{mice} exclusively\footnote{Note
-  that the alternative, joint modeling imputation for multilevel data or
-  \pkg{jomo} \citet{jomo}, has been implemented in \pkg{mice} as well
-  but is outside the scope of this tutorial.}. Other useful resources
-for the analysis of incomplete multilevel data include the \proglang{R}
-packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work
-by \citet{audi18} and \citet{grun18}. Please note that this tutorial
-paper assumes a basic level of knowledge on multilevel
-models.\footnote{Note to self: We're providing an overview of
-  implementations. It's up-to the reader to decide which multilevel
-  strategy suits their data. We won't go into detail for the different
-  methods (and equations). This paper is just a software tutorial, so
-  we'll keep it practical.} Assumed knowledge also includes the use of
-the `piping operator', \texttt{\%\textgreater{}\%}, adopted from the
-\pkg{magrittr} package, and the \pkg{lme4} notation for multilevel
-models.\footnote{TODO: Add environment info, seed and version number(s)
-  somewhere!}
+with \pkg{mice} in \proglang{R}. We provide practical guidelines and
+code snippets for different missing data situations, including
+non-ignorable mechanisms. For reasons of brevity, we focus on multilevel
+imputation by chained equations with \pkg{mice} exclusively; other
+imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are
+outside the scope of this tutorial. Assumed knowledge includes basic
+familiarity with multilevel imputation \citep[see e.g.][ and
+\citet{grun18}]{audi18} and the \pkg{lme4} notation for multilevel
+models (see Table \ref{tab:mod}).
 
-We illustrate how to impute incomplete multilevel data by means of three
-case studies:
+\begin{table}[tb]
+\caption{Concepts in missing data methods}
+\label{tab:mod}
+\centering
+\begin{tabular}{ll}
+\hline
+\textbf{Concept} & \textbf{Details}   \\
+\hline
+& [TODO: explain \pkg{lme4} notation here] \\
+\hline
+\end{tabular}
+\end{table}
 
-\begin{itemize}
-\tightlist
-\item
-  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
-  perceived popularity, \(n = 2,000\) pupils across \(N = 100\)
-  schools,][]{mice};
-\item
-  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
-  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions,][]{GJRM};
-\item
-  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
-  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
-  studies,][]{metamisc}.
-\end{itemize}
+\hypertarget{case-study}{%
+\section{Case study}\label{case-study}}
 
-For each of these datasets, we will discuss the nature of the
-missingness, choose one or more imputation models and evaluate the
-imputed data, but we will also highlight one specific aspect of the
-imputation workflow. With the \texttt{popmis} data, we show how (and how
-not) to develop an imputation model. With the \texttt{hiv} data we focus
-on extending the imputation model to include Heckman-type
-selection-inclusion methods. With the \texttt{impact} data we provide an
-example of multivariate missingness in real-world data. Together, this
-should give enough scaffolding for applied researchers who are faced
-with incomplete multilevel data.\footnote{TODO: Add notation paragraph
-  or `translation table' linking multilevel equations to \pkg{lme4}
-  formulas. Use betas instead of gamma's and mu's. Add interpretation of
-  values in predictormatrix (-2 for the cluster variable, 2 for random
-  effects). Add ICC and congeneality here as well. And make missingness
-  mechanism table as well.}
-
-Set-up the R environment and load the necessary packages:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> set.seed(2022)
-R> library(mice)
-R> library(ggmice)
-R> library(ggplot2)
-R> library(dplyr)
-R> library(lme4)
-R> library(mitml)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\begin{center}
-REVIEW UNTIL THIS POINT SVP
-\end{center}
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\hypertarget{case-study-i-how-not-to-impute}{%
-\section{Case Study I: How (not) to
-impute}\label{case-study-i-how-not-to-impute}}
-
-In this section we'll go over the different steps involved with imputing
-incomplete multilevel data. The data we're using is the \texttt{popmis}
-dataset from the \texttt{mice} package. This is a simulated dataset with
-pupils (\(n = 2000\)) clustered within schools (\(N = 100\)). In this
-tutorial we'll use the following variables:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{school}, school identification number (clustering variable);
-\item
-  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
-  unit-level);
-\item
-  \texttt{sex}, pupil sex (0=boy, 1=girl; unit-level);
-\item
-  \texttt{texp}, teacher experience (in years; cluster-level).
-\end{itemize}
-
-The analysis model corresponding to this dataset is multilevel
-regression with random intercepts, random slopes and a cross-level
-interaction. The outcome variable is \texttt{popular}, which is
-predicted from the unit-level variable \texttt{sex} and the
-cluster-level variable \texttt{texp}. The regression equation\footnote{add
-  the `level notation' (Bryk and Raudenbush, 1992) and/or matrix
-  notation (`linear mixed effects model'; Laird and Ware, 1982) too?}
-and \texttt{lme4} notation for this model are
-
-\[
-\text{popular}_{ij} =
-\gamma_{00} + 
-\gamma_{10} \text{ sex}_{ij} + 
-\gamma_{01} \text{ texp}_{j} + 
-\gamma_{11} \text{ texp}_{j} \times \text{sex}_{ij} + 
-u_{0j} + 
-u_{1j} \text{ sex}_{ij} + 
-e_{ij} \\
-\]
-
-\[
-\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
-\]
-
-Since the data is simulated and the missingness is induced, we can
-compare our inferences after imputation to the true complete data. The
-data is created in such a way that the clustering variable
-\texttt{school} explains quite some variance in the outcome variable
-\texttt{popular}. We express this using the intraclass correlation, ICC
-\(=\) 0.58. We'll evaluate the ICC after each missing data strategy, and
-compare the estimated fixed effects:
-
-\begin{CodeChunk}
-\begin{CodeOutput}
-                 Estimate with 95% CI
-1 (Intercept)  3.314 [ 2.998,  3.629]
-2         sex  1.330 [ 1.069,  1.590]
-3        texp  0.110 [ 0.090,  0.130]
-4    sex:texp -0.034 [-0.051, -0.017]
-\end{CodeOutput}
-\end{CodeChunk}
-
-\hypertarget{incomplete-data}{%
-\subsubsection{Incomplete data}\label{incomplete-data}}
-
-Load the data into the environment and select the relevant variables:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> popmis <- popmis[, c("school", "popular", "sex", "texp")] 
-\end{CodeInput}
-\end{CodeChunk}
-
-Plot the missing data pattern:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_pattern(popmis)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pat-1}
-
-}
-
-\caption[Missing data pattern in the popularity data]{Missing data pattern in the popularity data}\label{fig:pop_pat}
-\end{figure}
-\end{CodeChunk}
-
-The missingness is univariate and sporadic, which is illustrated in the
-missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the
-incomplete data is 0.56. This tells us that the multilevel structure of
-the data should probably be taken into account. If we don't, we'll may
-end up with incorrect imputations, biasing the effect of the clusters
-towards zero.
-
-Plot the correlations in the incomplete data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_corr(popmis)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-corr-1} \end{center}
-
-\end{CodeChunk}
-
-To develop the best imputation model for the incomplete variable
-\texttt{popular}, we need to know whether the missingness depends on the
-observed values of other variables. We'll highlight one other variable
-to illustrate, but ideally one would inspect all relations. The
-questions we'll ask are: `Does the missing data of pupil popularity
-(\texttt{popular}) depend on observed teacher popularity
-(\texttt{texp})?'. This can be evaluated statistically, but visual
-inspection usually suffices. We'll make a histogram of \texttt{texp}
-separately for the pupils with known popularity and missing popularity.
-
-Plot the histogram for teacher experience conditional on the missingness
-indicator of \texttt{popular}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> ggmice(popmis, aes(texp)) +
-+   geom_histogram(fill = "white") +
-+   facet_grid(. ~ is.na(popular), labeller = label_both)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-hist-1} \end{center}
-
-\end{CodeChunk}
-
-TODO: think about what is a meaningful rule of thumb to signal that the
-user should be worried?
-
-This shows us that there are no apparent differences in the distribution
-of \texttt{texp} depending on the missingness indicator of
-\texttt{popular} (t = -0.873, p = 0.383).
-
-\hypertarget{complete-case-analysis-not-recommended}{%
-\subsubsection{Complete case analysis (not
-recommended)}\label{complete-case-analysis-not-recommended}}
-
-Complete case analysis ignores the observations with missingness
-altogether, which lowers statistical power and may even introduce bias
-in MCAR situations.
-
-\hypertarget{imputation-ignoring-the-cluster-variable-not-recommended}{%
-\subsubsection{Imputation ignoring the cluster variable (not
-recommended)}\label{imputation-ignoring-the-cluster-variable-not-recommended}}
-
-The first imputation model that we'll use is likely to be invalid. We do
-\emph{not} use the cluster identifier \texttt{school} as imputation
-model predictor. With this model, we ignore the multilevel structure of
-the data, despite the high ICC. This assumes exchangeability between
-units. We include it purely to illustrate the effects of ignoring the
-clustering in our imputation effort. We'll use the default imputation
-methods in \texttt{mice()} (predictive mean matching to impute the
-continuous variables and logistic regression to impute binary
-variables).
-
-Create a methods vector and predictor matrix for \texttt{popular}, and
-make sure \texttt{school} is not included as predictor:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> meth <- make.method(popmis) # methods vector
-R> pred <- quickpred(popmis)   # predictor matrix
-R> plot_pred(pred)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-pred-1} \end{center}
-
-\end{CodeChunk}
-
-Impute the data, ignoring the cluster structure:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
-
-TODO: remove the broom.mixed output, use mitml only
-
-Analyze the imputations:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> fit_ignored <- imp_ignored %>% 
-+   with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
-R> testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-
-testEstimates(model = as.mitml.result(fit_ignored), var.comp = TRUE)
-
-Final parameter estimates and inferences obtained from 5 imputed data sets.
-
-             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.354     0.245    13.704     6.831     0.000     3.260     0.813 
-sex             1.247     0.332     3.759     6.154     0.009     4.160     0.849 
-texp            0.110     0.017     6.320     6.084     0.001     4.286     0.852 
-sex:texp       -0.027     0.022    -1.225     5.914     0.267     4.631     0.862
-
-                            Estimate 
-Intercept~~Intercept|school    0.177 
-sex~~sex|school                0.232 
-Intercept~~sex|school         -0.043 
-Residual~~Residual             0.670 
-ICC|school                     0.209
-
-Unadjusted hypothesis test as appropriate in larger samples.
-\end{CodeOutput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_ignored_eval-1} \end{center}
-
-\end{CodeChunk}
-
-\hypertarget{imputation-with-the-cluster-variable-as-predictor-not-recommended}{%
-\subsubsection{Imputation with the cluster variable as predictor (not
-recommended)}\label{imputation-with-the-cluster-variable-as-predictor-not-recommended}}
-
-We'll now use \texttt{school} as a predictor to impute all other
-variables. This is still not recommended practice, since it only works
-under certain circumstances and results may be biased
-\citep{drec15, ende16}. But at least, it includes some multilevel
-aspect. This method is also called `fixed cluster imputation', and uses
-N-1 indicator variables representing allocation of N clusters as a fixed
-factor in the model \citep{reit06, ende16}. Colloquially, this is
-`multilevel imputation for dummies'.
-
-Add: doesn't work with syst missing (only sporadically). There's some
-pro's and con's. May not differ much if the number of clusters is low.
-
-The more the random effects are of interest, the more you need ml
-models.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # adjust the predictor matrix
-R> pred["popular", "school"] <- 1 
-R> plot_pred(pred)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor-1} \end{center}
-
-\begin{CodeInput}
-R> # impute the data, cluster as predictor
-R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor_eval-1} \end{center}
-
-\end{CodeChunk}
-
-Now, we can clearly see that the imputed values of \texttt{texp} are
-higher than the observed values, which is in line with right-tailed MAR.
-
-The ICCs are way more in line with the ICCs in the incomplete data. But
-this is a quick and dirty way of imputing multilevel data. We
-\emph{should} be using a multilevel model.
-
-\hypertarget{imputation-with-random-effects}{%
-\subsubsection{Imputation with random
-effects}\label{imputation-with-random-effects}}
-
-With \texttt{2l.norm} we impute the outcome with a multilevel model
-assuming random slopes for each variable in the imputation model and
-homogeneous within-cluster variance.
-
-``Van Buuren (2011) considered the homoscedastic linear mixed model as
-invalid for imputing incomplete predictors, and investigated only the
-2l.norm method, which allows for heterogeneous error variances''
-\citep{buur18}.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # adjust the predictor matrix
-R> pred["popular", ] <- c(school = -2, popular = 0, sex = 2, texp = 2) 
-R> plot_pred(pred) 
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_norm-1} \end{center}
-
-\begin{CodeInput}
-R> meth <- make.method(popmis)
-R> meth["popular"] <- "2l.pmm"
-R> imp_pmm_2l <-
-+   mice(
-+     popmis %>% mutate(school = as.integer(school)),
-+     pred = pred,
-+     meth = meth,
-+     maxit = 1,
-+     print = FALSE
-+   )
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_norm_eval-1} \end{center}
-
-\end{CodeChunk}
-
-\hypertarget{imputation-with-random-effects-and-heterogeneity}{%
-\subsubsection{Imputation with random effects and
-heterogeneity}\label{imputation-with-random-effects-and-heterogeneity}}
-
-This method assumes random slopes for each variable in the imputation
-model. In contrast to \texttt{2l.norm} this method allows a
-cluster-specific residual error variance.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> pred["popular", ] <- c(-2, 2, 1, 2)
-R> meth <- c("", "2l.pan", "", "")
-R> imp_pan_2l <-
-+   mice(
-+     popmis %>% mutate(school = as.integer(school)),
-+     pred = pred,
-+     meth = meth,
-+     maxit = 1,
-+     print = FALSE
-+   )
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pan_eval-1} \end{center}
-
-\end{CodeChunk}
-
-\hypertarget{how-to-handle-non-random-selection-case-study-ii-hiv}{%
-\section{How to handle non-random selection (Case study II:
-HIV)}\label{how-to-handle-non-random-selection-case-study-ii-hiv}}
-
-Data are simulated and included in the \texttt{GJRM} package. We will
-use the following variables:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{region} Cluster variable,
-\item
-  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
-\item
-  \texttt{age} Age of the patient,
-\item
-  \texttt{marital} Marital status,
-\item
-  \texttt{condom} Condom use during last intercourse,
-\item
-  \texttt{smoke} Smoker (levels; inclusion restriction variable).
-\end{itemize}
-
-The imputation of these date is based on the toy example from
-\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
-Heckman Github repo}.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/hiv-1} \end{center}
-
-\end{CodeChunk}
-
-From the missing data pattern we see that we can set \texttt{maxit} to
-1, since there is only one variable with missingness.
-
-The inclusion restriction variable should be a predictor of the the
-actual value of the variable of interest, but \emph{not} of missingness
-indicator for the variable of interest. In this case, the data were
-simulated to adhere to this requirement. Namely, \(\beta_{smoke}\) =
--0.064, 95\% CI {[}-0.256, 0.126{]} for the analysis model
-(\texttt{formula\ =\ hiv\ \textasciitilde{}\ .}), and \(\beta_{smoke}\)
-= -0.265, 95\% CI {[}-0.422, -0.11{]} for the selection model
-(\texttt{formula\ =\ is.na(hiv)\ \textasciitilde{}\ .}). This means the
-assumptions for the Heckman-type selection model are met.
-
-\hypertarget{how-to-handle-multivariate-missingness-case-study-iii-impact}{%
-\section{How to handle multivariate missingness (Case study III:
-IMPACT)}\label{how-to-handle-multivariate-missingness-case-study-iii-impact}}
-
-\texttt{impact} is traumatic brain injury data with patients,
-\(n = 11022\), clustered in studies, \(N = 15\). With the following 11
-variables:
+We illustrate how to impute incomplete multilevel data by means of a
+case study: \texttt{impact} from the \pkg{metamisc} package
+\citep[empirical data on traumatic brain injuries, \(n = 11,022\)
+patients across \(N = 15\) studies,][]{metamisc}. \texttt{impact} is
+traumatic brain injury data with patients, \(n = 11022\), clustered in
+studies, \(N = 15\). With the following 11 variables:
 
@@
 
-The analysis model for this dataset is a prediction model with
-\texttt{mort} as the outcome.\footnote{Look at analysis model, maybe
-  copy from GREAT data example e.g., adjusted prognostic effect of
-  \texttt{ct} on unfortunate outcomes, we just want to know the adjusted
-  odds ratio for \texttt{ct}. Add something about systematically missing
-  data here.} The data is already imputed (Steyerberg et al, 2008), so
-we've induced missingness again based on the missingness in the original
-data.
-
-\[
-\text{mort}_{ij} =
-\gamma_{00} + 
-\gamma_{01} \text{ type}_{j} + 
-\gamma_{10} \text{ age}_{ij} + 
-\gamma_{10} \text{ moter\_score}_{ij} + 
-\gamma_{10} \text{ pupil}_{ij} + 
-\gamma_{10} \text{ ct}_{ij} + \\
-u_{0j} + 
-u_{1j} \text{ age}_{ij} + 
-u_{1j} \text{ moter\_score}_{ij} +  
-u_{1j} \text{ pupil}_{ij} +  
-u_{1j} \text{ ct}_{ij} +  
-e_{ij} \\
-\]
-
-\texttt{glmer(mort\ \textasciitilde{}\ 1\ +\ type\ +\ age\ +\ motor\_score\ +\ pupil\ +\ ct\ +\ (1\ \textbar{}\ name))}
+{[}TODO: add more info about the complete data.{]} The analysis model
+for this dataset is a prediction model with \texttt{mort} as the
+outcome. In this tutorial we'll estimate the adjusted prognostic effect
+of \texttt{ct} on unfortunate outcomes. The estimand is the adjusted
+odds ratio for \texttt{ct}, after including \texttt{type}, \texttt{age}
+\texttt{motor\_score} and \texttt{pupil} into the analysis model:
 
 \begin{CodeChunk}
+\begin{CodeInput}
+R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name)
+\end{CodeInput}
+\end{CodeChunk}
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-1} \end{center}
+Note that variables \texttt{hypots}, \texttt{hypox}, \texttt{tsah} and
+\texttt{edh} are not part of the analysis model, and may thus serve as
+auxiliary variables for imputation.
+
+\hypertarget{setup}{%
+\subsection{Setup}\label{setup}}
+
+{[}TODO: Add environment info, seed and version number(s) somewhere.{]}
+Set up the R environment and load the necessary packages:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> set.seed(2022)
+R> library(mice)         # for imputation
+R> library(ggmice)       # for visualization
+R> library(ggplot2)      # for visualization
+R> library(dplyr)        # for data wrangling
+R> library(lme4)         # for multilevel modeling
+R> library(broom.mixed)  # for multilevel estimates
+R> library(mitml)        # for multilevel pooling
+\end{CodeInput}
+\end{CodeChunk}
+
+The \texttt{impact} data included in the \pkg{metamisc} package is a
+complete data set. The original data has already been imputed once
+(Steyerberg et al, 2008). For the purpose of this tutorial we have
+induced missingness (mimicking the missing data in the original data set
+before imputation). The resulting incomplete data can be accessed from
+\href{https://zenodo.com}{zenodo link to be created}.
+
+Load the complete and incomplete data into the R workspace:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> data("impact", package = "metamisc")
+R> dat <- read.table("link/to/the/data.txt")
+\end{CodeInput}
+\end{CodeChunk}
+
+We will use the following estimates as comparative truth in this
+tutorial {[}TODO: make this a table or forest plot instead, to see if
+there is heterogeneity in the association of \texttt{ct} with
+\texttt{mort}{]}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- glmer(mod, family = "binomial", data = impact) 
+R> tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 11 x 9
+   effect group term  estimate std.error statistic    p.value conf.low conf.high
+   <chr>  <chr> <chr>    <dbl>     <dbl>     <dbl>      <dbl>    <dbl>     <dbl>
+ 1 fixed  <NA>  (Int~    0.101   0.0173     -13.4   4.94e- 41   0.0723     0.141
+ 2 fixed  <NA>  type~    0.713   0.123       -1.96  5.01e-  2   0.509      1.00 
+ 3 fixed  <NA>  age      1.03    0.00165     20.2   2.13e- 90   1.03       1.04 
+ 4 fixed  <NA>  moto~    0.553   0.0386      -8.50  1.95e- 17   0.482      0.634
+ 5 fixed  <NA>  moto~    0.405   0.0289     -12.7   8.43e- 37   0.352      0.466
+ 6 fixed  <NA>  moto~    0.275   0.0202     -17.6   1.67e- 69   0.239      0.318
+ 7 fixed  <NA>  pupi~    3.73    0.230       21.4   2.09e-101   3.31       4.21 
+ 8 fixed  <NA>  pupi~    1.87    0.133        8.80  1.36e- 18   1.63       2.15 
+ 9 fixed  <NA>  ctIII    2.25    0.157       11.6   5.12e- 31   1.96       2.58 
+10 fixed  <NA>  ctIV~    2.30    0.136       14.0   9.47e- 45   2.05       2.58 
+11 ran_p~ name  sd__~    0.277  NA           NA    NA          NA         NA    
+\end{CodeOutput}
+\end{CodeChunk}
+
+{[}TODO: add ICC before/after imputation and interpret: This tells us
+that the multilevel structure of the data should probably be taken into
+account. If we don't, we'll may end up with incorrect imputations,
+biasing the effect of the clusters towards zero.{]}
+
+{[}TODO: add descriptive statistics of the complete and incomplete
+data.{]}
+
+\hypertarget{missingness}{%
+\subsection{Missingness}\label{missingness}}
+
+To explore the missingness, it is wise to look at the missing data
+pattern:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(dat, rotate = TRUE)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pattern-1} \end{center}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeOutput}
-# A tibble: 9 x 7
-  effect   group term                    estimate std.error statistic   p.value
-  <chr>    <chr> <chr>                      <dbl>     <dbl>     <dbl>     <dbl>
-1 fixed    <NA>  (Intercept)              -2.66     0.194      -13.7   1.11e-42
-2 fixed    <NA>  typeRCT                  -0.412    0.172       -2.40  1.64e- 2
-3 fixed    <NA>  age                       0.0298   0.00190     15.7   2.56e-55
-4 fixed    <NA>  as.factor(motor_score)2  -0.631    0.0823      -7.66  1.82e-14
-5 fixed    <NA>  as.factor(motor_score)3  -0.994    0.0845     -11.8   5.61e-32
-6 fixed    <NA>  as.factor(motor_score)4  -1.54     0.0868     -17.7   2.88e-70
-7 fixed    <NA>  pupil                     0.438    0.0397      11.0   3.01e-28
-8 fixed    <NA>  ct                        0.430    0.0343      12.5   5.34e-36
-9 ran_pars name  sd__(Intercept)           0.257   NA           NA    NA       
-\end{CodeOutput}
-\end{CodeChunk}
+This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
+and \texttt{pupil}.{]}
+
+To develop the best imputation model, we need to investigate the
+relations between the observed values of the incomplete variables and
+the observed values of other variables, and the relation between the
+missingness indicators of the incomplete variables and the observed
+values of the other variables. To see whether the missingness depends on
+the observed values of other variables, we\ldots{} {[}TODO: fill in that
+we can test this statistically or use visual inspection (e.g.~a
+histogram faceted by the missingness indicator).{]}
 
@@
 \begin{CodeInput}
-R> plot_corr(impact_NA, label = T)
+R> plot_corr(dat, rotate = TRUE)
 \end{CodeInput}
@@
 variables has a very strong relation, but \texttt{hypots} could be used.
-We conclude that we can exclude \texttt{hypox} from the data since this
-is neither an analysis model variable, nor an auxiliary variable for
-imputation.
+We conclude that we can exclude \texttt{hypox} from the data, since this
+is neither an analysis model variable nor an auxiliary variable for
+imputation:
 
@@
 \begin{CodeInput}
-R> impact_NA <- select(impact_NA, -c(hypox, hypots))
-R> pred <- quickpred(impact_NA)
+R> dat <- select(dat, !hypox)
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{complete-case-analysis}{%
+\subsection{Complete case analysis}\label{complete-case-analysis}}
+
+As previously stated, complete case analysis lowers statistical power
+and may bias results. The complete case analysis estimates are:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) 
+R> tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 11 x 9
+   effect  group term  estimate std.error statistic   p.value conf.low conf.high
+   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
+ 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
+ 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
+ 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
+ 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
+ 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
+ 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
+ 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
+ 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
+ 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
+10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
+11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
+\end{CodeOutput}
+\end{CodeChunk}
+
+As we can see\ldots{} {[}TODO: fill in.{]}
+
+\hypertarget{imputation-model}{%
+\subsection{Imputation model}\label{imputation-model}}
+
+The first imputation model that we'll use is likely to be invalid. We do
+\emph{not} use the cluster identifier \texttt{name} as imputation model
+predictor. With this model, we ignore the multilevel structure of the
+data, despite the high ICC. This assumes exchangeability between units.
+We include it purely to illustrate the effects of ignoring the
+clustering in our imputation effort. We'll use the default imputation
+methods in \texttt{mice()} (predictive mean matching to impute the
+continuous variables and logistic regression to impute binary
+variables).
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\begin{center}
+
+Updated until here!
+
+\end{center}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+Create a methods vector and predictor matrix, and make sure
+\texttt{name} is not included as predictor:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth <- make.method(dat) # methods vector
+R> pred <- quickpred(dat)   # predictor matrix
 R> plot_pred(pred)
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-pred-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/ignore-1} \end{center}
 
-\begin{CodeInput}
-R> pred[pred == 1] <- 2
-R> pred["mort", ] <- 2
-R> pred[, "mort"] <- 2
-R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
-R> pred[, "name"] <- -2
-R> diag(pred) <- 0
-R> plot_pred(pred)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-pred-2} \end{center}
-
-\begin{CodeInput}
-R> meth <- make.method(impact_NA)
-R> meth[meth != ""] <- "2l.pmm"
-\end{CodeInput}
 \end{CodeChunk}
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> imp <- mice(impact_NA, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
-R> # imp_impact <- impact_NA %>%
-R> #   mutate(name = as.integer(name), motor_score = as.numeric(motor_score)) %>%
-R> #   mice::mice(., m = 2, maxit = 1, method = meth, predictorMatrix = pred)
-R> # look at jomo for categorical variables?
-R> # semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach
-R> # pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters
-R> # make assumptions of these methods explicit!
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> fit <- imp_impact_pmm %>% 
-+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
-R> tidy(pool(fit))
-R> as.mitml.result(fit)
-R> # testEstimates(as.mitml.result(fit))
-\end{CodeInput}
-\end{CodeChunk}
-
-Compare estimates to complete data
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> fit <- glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial", data = impact) 
-R> tidy(fit)
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 11 x 7
-   effect   group term                   estimate std.error statistic    p.value
-   <chr>    <chr> <chr>                     <dbl>     <dbl>     <dbl>      <dbl>
- 1 fixed    <NA>  (Intercept)             -2.29     0.171      -13.4   4.94e- 41
- 2 fixed    <NA>  typeRCT                 -0.338    0.172       -1.96  5.01e-  2
- 3 fixed    <NA>  age                      0.0322   0.00160     20.2   2.13e- 90
- 4 fixed    <NA>  as.factor(motor_score~  -0.593    0.0698      -8.50  1.95e- 17
- 5 fixed    <NA>  as.factor(motor_score~  -0.904    0.0713     -12.7   8.43e- 37
- 6 fixed    <NA>  as.factor(motor_score~  -1.29     0.0732     -17.6   1.67e- 69
- 7 fixed    <NA>  pupilNone                1.32     0.0616      21.4   2.09e-101
- 8 fixed    <NA>  pupilOne                 0.626    0.0711       8.80  1.36e- 18
- 9 fixed    <NA>  ctIII                    0.810    0.0700      11.6   5.12e- 31
-10 fixed    <NA>  ctIV/V                   0.832    0.0593      14.0   9.47e- 45
-11 ran_pars name  sd__(Intercept)          0.277   NA           NA    NA        
-\end{CodeOutput}
-\end{CodeChunk}
+{[}TODO: mutate data to get the right data types for imputation
+(e.g.~integer for clustering variable).{]}
 

### 2022-04-07 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Add forest plot effect ct

@@
 
-We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc].
-`impact` is traumatic brain injury data with patients, $n = 11022$, clustered in studies, $N = 15$.
-With the following 11 variables:
+We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc].
+[TODO: add more info about the complete data.] The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:
 
@@
 
-[TODO: add more info about the complete data.] The analysis model for this dataset is a prediction model with `mort` as the outcome.
+The analysis model for this dataset is a prediction model with `mort` as the outcome.
 In this tutorial we'll estimate the adjusted prognostic effect of `ct` on unfortunate outcomes.
@@
 ```{r mod}
-mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name)
+mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) # TODO: add random effects ct?
 ```
@@
 ```{r data, eval = FALSE}
-data("impact", package = "metamisc")
-dat <- read.table("link/to/the/data.txt")
+data("impact", package = "metamisc")      # complete data
+dat <- read.table("link/to/the/data.txt") # incomplete data
 ```
@@
 ```{r truth}
-fit <- glmer(mod, family = "binomial", data = impact) 
-tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
+fit <- glmer(mod, family = "binomial", data = impact) # fit the model
+tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
+```
+
+```{r forest, echo=FALSE}
+fits <-
+  purrr::map_dfr(
+    unique(impact$name),
+    ~ {
+      glm(
+        mort ~ 1 + age + motor_score + pupil + ct,
+        family = "binomial",
+        data = filter(impact, name == .x)
+      )
+    } %>%
+      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
+      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
+      cbind(name = as.character(.x), .)
+  )
+filter(fits, term %in% c("ctIII")) %>% #, "ctIV/V"
+  ggplot(aes(y = name, x = estimate)) +
+  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
+  geom_point() +
+  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high)) +
+  labs(y = "Study", x = "Odds ratio", color = NULL) +
+  theme_classic()
 ```
@@
 ```{r pattern, fig.height=7.1}
-plot_pattern(dat, rotate = TRUE)
+plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
 ```
@@
 ```{r impact_corr}
-plot_corr(dat, rotate = TRUE)
+plot_corr(dat, rotate = TRUE) # plot correlations 
 ```
@@
 ```{r}
-dat <- select(dat, !hypox)
+dat <- select(dat, !hypox)  # remove variable
 ```
@@
 ```{r cca}
-fit <- glmer(mod, family = "binomial", data = na.omit(dat)) 
-tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
+fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
+tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
 ```

### 2022-04-07 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Add forest plot effect ct

@@
 case study: \texttt{impact} from the \pkg{metamisc} package
-\citep[empirical data on traumatic brain injuries, \(n = 11,022\)
-patients across \(N = 15\) studies,][]{metamisc}. \texttt{impact} is
-traumatic brain injury data with patients, \(n = 11022\), clustered in
-studies, \(N = 15\). With the following 11 variables:
+\citep[empirical data on traumatic brain injuries, \(n = 11,022\) units
+across \(N = 15\) clusters,][]{metamisc}. {[}TODO: add more info about
+the complete data.{]} The \texttt{impact} data set contains traumatic
+brain injury data on \(n = 11022\) patients clustered in \(N = 15\)
+studies with the following 11 variables:
 
@@
 
-{[}TODO: add more info about the complete data.{]} The analysis model
-for this dataset is a prediction model with \texttt{mort} as the
-outcome. In this tutorial we'll estimate the adjusted prognostic effect
-of \texttt{ct} on unfortunate outcomes. The estimand is the adjusted
-odds ratio for \texttt{ct}, after including \texttt{type}, \texttt{age}
-\texttt{motor\_score} and \texttt{pupil} into the analysis model:
+The analysis model for this dataset is a prediction model with
+\texttt{mort} as the outcome. In this tutorial we'll estimate the
+adjusted prognostic effect of \texttt{ct} on unfortunate outcomes. The
+estimand is the adjusted odds ratio for \texttt{ct}, after including
+\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
+into the analysis model:
 
@@
 \begin{CodeInput}
-R> data("impact", package = "metamisc")
-R> dat <- read.table("link/to/the/data.txt")
+R> data("impact", package = "metamisc")      # complete data
+R> dat <- read.table("link/to/the/data.txt") # incomplete data
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> fit <- glmer(mod, family = "binomial", data = impact) 
-R> tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
+R> fit <- glmer(mod, family = "binomial", data = impact) # fit the model
+R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
 \end{CodeInput}
@@
 
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/forest-1} \end{center}
+
+\end{CodeChunk}
+
 {[}TODO: add ICC before/after imputation and interpret: This tells us
@@
 \begin{CodeInput}
-R> plot_pattern(dat, rotate = TRUE)
+R> plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> plot_corr(dat, rotate = TRUE)
+R> plot_corr(dat, rotate = TRUE) # plot correlations 
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> dat <- select(dat, !hypox)
+R> dat <- select(dat, !hypox)  # remove variable
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) 
-R> tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
+R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
+R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
 \end{CodeInput}

### 2022-04-11 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Feedback on cleaned up manuscript (add popularity case study back in!)

@@
                     & structure. \\
-Random intercept    & Multilevel models typically accommodate for variability by including \\
+Random effect       & Multilevel models typically accommodate for variability by including \\
                     & a separate group mean for each cluster. In addition to random \\
@@
                     & heterogeneous residual error variances across clusters [see e.g. \\
-                    & @gelm06, @hox17 and @jong21]. \\
+                    & @gelm06, @hox17 and @jong21]. TODO: add stratification. \\
 \hline
@@
 & [@rubi76; @meng94]. \\
-& [TODO: add congeniality.] \\
+& [TODO: add congeniality, but maybe in-text?] \\
 \hline
@@
 \begin{table}[tb]
-\caption{Concepts in missing data methods}
+\caption{Notation}
 \label{tab:mod}
@@
 \end{table}
+
+TODO: add paragraph about novice and more experienced readers.
 
@@
 The resulting incomplete data can be accessed from [zenodo link to be created](https://zenodo.com).
+TODO: email script to thomas.
 
@@
   geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high)) +
-  labs(y = "Study", x = "Odds ratio", color = NULL) +
+  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
   theme_classic()
+# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
+# TODO: missingness prop inside clusters
+# TODO: make row height proportional to pattern frequency
 ```
@@
 
-This shows... [TODO: fill in that we need to impute `ct` and `pupil`.]
+This shows... [TODO: fill in that we need to impute `ct` and `pupil`.] TODO: remove axis labels
 

### 2022-04-11 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Feedback on cleaned up manuscript (add popularity case study back in!)

@@
 \begin{CodeInput}
-R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name)
+R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) # TODO: add random effects ct?
 \end{CodeInput}
@@
 before imputation). The resulting incomplete data can be accessed from
-\href{https://zenodo.com}{zenodo link to be created}.
+\href{https://zenodo.com}{zenodo link to be created}. TODO: email script
+to thomas.
 
@@
 This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
-and \texttt{pupil}.{]}
+and \texttt{pupil}.{]} TODO: remove axis labels
 

### 2022-04-29 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Add case studies back in

@@
         plot.margin = unit(c(0,0,0,0), "pt"))
+# TODO: add fifth case with cluster 3
+# TODO: use math font
 ```
@@
 
-TODO: add paragraph about novice and more experienced readers.
+We illustrate how to impute incomplete multilevel data by means of three case studies:
 
-# Case study
+- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, @mice];
+- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, @GJRM];
+- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc].
+
+For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
+With the `popmis` data, we show how (and how not) to develop an imputation model.
+With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods.
+With the `impact` data we provide an example of multivariate missingness in real-world data.
+Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.
+
+**TODO: add paragraph about novice and more experienced readers.**
+
+## Setup
+
+[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
+```{r env, message=FALSE, warning=FALSE}
+set.seed(2022)
+library(mice)         # for imputation
+library(ggmice)       # for visualization
+library(ggplot2)      # for visualization
+library(dplyr)        # for data wrangling
+library(lme4)         # for multilevel modeling
+library(broom.mixed)  # for multilevel estimates
+library(mitml)        # for multilevel pooling
+```
+
+# Case study I: popularity data
+
+In this section we'll go over the different steps involved with imputing incomplete multilevel data.
+The data we're using is the `popmis` dataset from the `mice` package.
+This is a simulated dataset with pupils ($n = 2000$) clustered within schools ($N = 100$).
+In this tutorial we'll use the following variables:
+
+  - `school`,     school identification number (clustering variable);
+  - `popular`,    pupil popularity (self-rating between 0 and 10; unit-level);
+  - `sex`,	      pupil sex (0=boy, 1=girl; unit-level);
+  - `texp`,	      teacher experience (in years; cluster-level).
+
+The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction.
+The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`:
+
+$$
+\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
+$$
+
+```{r include=FALSE}
+# data
+popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
+  mutate(school = SCHOOL,
+         popular = POPULAR,
+         sex = SEX,
+         texp = TEXP,
+         .keep = "none")
+popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
+```
+
+Load the data into the environment and select the relevant variables:
+
+```{r}
+popmis <- popmis[, c("school", "popular", "sex", "texp")] 
+```
+Plot the missing data pattern:
+
+```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
+plot_pattern(popmis)
+```
+
+The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}.
+This tells us that the multilevel structure of the data should probably be taken into account.
+If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.
+
+ <!-- The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. -->
+
+Plot the correlations in the incomplete data:
+
+```{r pop-corr}
+plot_corr(popmis)
+```
+
+To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables.
+We'll highlight one other variable to illustrate, but ideally one would inspect all relations.
+The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'.
+This can be evaluated statistically, but visual inspection usually suffices.
+We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.
+
+Plot the histogram for teacher experience conditional on the missingness indicator of `popular`:
+
+```{r pop-hist}
+ggmice(popmis, aes(texp)) +
+  geom_histogram(fill = "white") +
+  facet_grid(. ~ is.na(popular), labeller = label_both)
+```
+
+This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
+ **TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?**
+
+### Complete case analysis (not recommended)
+
+Complete case analysis ignores the observations with missingness altogether, which lowers statistical power and may even introduce bias in MCAR situations.
+
+```{r pop-cca, echo=FALSE}
+est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis)
+
+```
+
+### Imputation ignoring the cluster variable (not recommended)
+
+The first imputation model that we'll use is likely to be invalid.
+We do *not* use the cluster identifier `school` as imputation model predictor.
+With this model, we ignore the multilevel structure of the data, despite the high ICC.
+This assumes exchangeability between units.
+We include it purely to illustrate the effects of ignoring the clustering in our imputation effort.
+We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).
+
+Create a methods vector and predictor matrix for `popular`, and make sure `school` is not included as predictor:
+
+```{r pop-ignored-pred, echo=TRUE, message=FALSE, warning=FALSE}
+meth <- make.method(popmis) # methods vector
+pred <- quickpred(popmis)   # predictor matrix
+plot_pred(pred)
+```
+
+Impute the data, ignoring the cluster structure:
+```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
+imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+```
+
+TODO: remove the broom.mixed output, use mitml only
+
+Analyze the imputations:
+```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
+fit_ignored <- imp_ignored %>% 
+  with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
+```
+
+### Imputation with the cluster variable as predictor (not recommended)
+
+We'll now use `school` as a predictor to impute all other variables.
+This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
+But at least, it includes some multilevel aspect.
+This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16].
+Colloquially, this is 'multilevel imputation for dummies'.
+
+TODO: Add that it doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.
+
+TODO: Add that the more the random effects are of interest, the more you need ml models.
+
+```{r pop_predictor, message=FALSE, warning=FALSE}
+# adjust the predictor matrix
+pred["popular", "school"] <- 1 
+plot_pred(pred)
+
+# impute the data, cluster as predictor
+imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+```
+
+# Case study II: HIV data
+
+Data are simulated and included in the `GJRM` package. We will use the following variables:
+
+  - `region` Cluster variable,
+- `hiv` HIV diagnosis (0=no, 1=yes),
+- `age` Age of the patient,
+- `marital` Marital status,
+- `condom` Condom use during last intercourse,
+- `smoke` Smoker (levels; inclusion restriction variable).
+
+The imputation of these date is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
+
+```{r hiv, echo=FALSE}
+# data
+data("hiv", package = "GJRM")
+hiv <- hiv %>% 
+  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
+  mutate(region = as.integer(region))
+# md pattern
+plot_pattern(hiv)
+```
+
+# Case study III: IMPACT data
 
@@
 Note that variables `hypots`, `hypox`, `tsah` and `edh` are not part of the analysis model, and may thus serve as auxiliary variables for imputation.
-
-## Setup
-
-[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
-```{r env, message=FALSE, warning=FALSE}
-set.seed(2022)
-library(mice)         # for imputation
-library(ggmice)       # for visualization
-library(ggplot2)      # for visualization
-library(dplyr)        # for data wrangling
-library(lme4)         # for multilevel modeling
-library(broom.mixed)  # for multilevel estimates
-library(mitml)        # for multilevel pooling
-```
 

### 2022-04-29 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Add case studies back in

@@
 
-TODO: add paragraph about novice and more experienced readers.
+We illustrate how to impute incomplete multilevel data by means of three
+case studies:
 
-\hypertarget{case-study}{%
-\section{Case study}\label{case-study}}
+\begin{itemize}
+\tightlist
+\item
+  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
+  perceived popularity, \(n = 2,000\) pupils across \(N = 100\)
+  schools,][]{mice};
+\item
+  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
+  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions,][]{GJRM};
+\item
+  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
+  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
+  studies,][]{metamisc}.
+\end{itemize}
+
+For each of these datasets, we will discuss the nature of the
+missingness, choose one or more imputation models and evaluate the
+imputed data, but we will also highlight one specific aspect of the
+imputation workflow. With the \texttt{popmis} data, we show how (and how
+not) to develop an imputation model. With the \texttt{hiv} data we focus
+on extending the imputation model to include Heckman-type
+selection-inclusion methods. With the \texttt{impact} data we provide an
+example of multivariate missingness in real-world data. Together, this
+should give enough scaffolding for applied researchers who are faced
+with incomplete multilevel data.
+
+\textbf{TODO: add paragraph about novice and more experienced readers.}
+
+\hypertarget{setup}{%
+\subsection{Setup}\label{setup}}
+
+{[}TODO: Add environment info, seed and version number(s) somewhere.{]}
+Set up the R environment and load the necessary packages:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> set.seed(2022)
+R> library(mice)         # for imputation
+R> library(ggmice)       # for visualization
+R> library(ggplot2)      # for visualization
+R> library(dplyr)        # for data wrangling
+R> library(lme4)         # for multilevel modeling
+R> library(broom.mixed)  # for multilevel estimates
+R> library(mitml)        # for multilevel pooling
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{case-study-i-popularity-data}{%
+\section{Case study I: popularity
+data}\label{case-study-i-popularity-data}}
+
+In this section we'll go over the different steps involved with imputing
+incomplete multilevel data. The data we're using is the \texttt{popmis}
+dataset from the \texttt{mice} package. This is a simulated dataset with
+pupils (\(n = 2000\)) clustered within schools (\(N = 100\)). In this
+tutorial we'll use the following variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{school}, school identification number (clustering variable);
+\item
+  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
+  unit-level);
+\item
+  \texttt{sex}, pupil sex (0=boy, 1=girl; unit-level);
+\item
+  \texttt{texp}, teacher experience (in years; cluster-level).
+\end{itemize}
+
+The analysis model corresponding to this dataset is multilevel
+regression with random intercepts, random slopes and a cross-level
+interaction. The outcome variable is \texttt{popular}, which is
+predicted from the unit-level variable \texttt{sex} and the
+cluster-level variable \texttt{texp}:
+
+\[
+\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
+\]
+
+Load the data into the environment and select the relevant variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> popmis <- popmis[, c("school", "popular", "sex", "texp")] 
+\end{CodeInput}
+\end{CodeChunk}
+
+Plot the missing data pattern:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(popmis)
+\end{CodeInput}
+\begin{figure}
+
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pat-1}
+
+}
+
+\caption[Missing data pattern in the popularity data]{Missing data pattern in the popularity data}\label{fig:pop_pat}
+\end{figure}
+\end{CodeChunk}
+
+The missingness is univariate and sporadic, which is illustrated in the
+missing data pattern in Figure \ref{fig:pop_pat}. This tells us that the
+multilevel structure of the data should probably be taken into account.
+If we don't, we'll may end up with incorrect imputations, biasing the
+effect of the clusters towards zero.
+
+Plot the correlations in the incomplete data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_corr(popmis)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-corr-1} \end{center}
+
+\end{CodeChunk}
+
+To develop the best imputation model for the incomplete variable
+\texttt{popular}, we need to know whether the missingness depends on the
+observed values of other variables. We'll highlight one other variable
+to illustrate, but ideally one would inspect all relations. The
+questions we'll ask are: `Does the missing data of pupil popularity
+(\texttt{popular}) depend on observed teacher popularity
+(\texttt{texp})?'. This can be evaluated statistically, but visual
+inspection usually suffices. We'll make a histogram of \texttt{texp}
+separately for the pupils with known popularity and missing popularity.
+
+Plot the histogram for teacher experience conditional on the missingness
+indicator of \texttt{popular}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> ggmice(popmis, aes(texp)) +
++   geom_histogram(fill = "white") +
++   facet_grid(. ~ is.na(popular), labeller = label_both)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-hist-1} \end{center}
+
+\end{CodeChunk}
+
+This shows us that there are no apparent differences in the distribution
+of \texttt{texp} depending on the missingness indicator of
+\texttt{popular} (t = -0.873, p = 0.383). \textbf{TODO: think about what
+is a meaningful rule of thumb to signal that the user should be
+worried?}
+
+\hypertarget{complete-case-analysis-not-recommended}{%
+\subsubsection{Complete case analysis (not
+recommended)}\label{complete-case-analysis-not-recommended}}
+
+Complete case analysis ignores the observations with missingness
+altogether, which lowers statistical power and may even introduce bias
+in MCAR situations.
+
+\hypertarget{imputation-ignoring-the-cluster-variable-not-recommended}{%
+\subsubsection{Imputation ignoring the cluster variable (not
+recommended)}\label{imputation-ignoring-the-cluster-variable-not-recommended}}
+
+The first imputation model that we'll use is likely to be invalid. We do
+\emph{not} use the cluster identifier \texttt{school} as imputation
+model predictor. With this model, we ignore the multilevel structure of
+the data, despite the high ICC. This assumes exchangeability between
+units. We include it purely to illustrate the effects of ignoring the
+clustering in our imputation effort. We'll use the default imputation
+methods in \texttt{mice()} (predictive mean matching to impute the
+continuous variables and logistic regression to impute binary
+variables).
+
+Create a methods vector and predictor matrix for \texttt{popular}, and
+make sure \texttt{school} is not included as predictor:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth <- make.method(popmis) # methods vector
+R> pred <- quickpred(popmis)   # predictor matrix
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-pred-1} \end{center}
+
+\end{CodeChunk}
+
+Impute the data, ignoring the cluster structure:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+TODO: remove the broom.mixed output, use mitml only
+
+Analyze the imputations:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit_ignored <- imp_ignored %>% 
++   with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+R> testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+
+testEstimates(model = as.mitml.result(fit_ignored), var.comp = TRUE)
+
+Final parameter estimates and inferences obtained from 5 imputed data sets.
+
+             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
+(Intercept)     3.354     0.245    13.704     6.831     0.000     3.260     0.813 
+sex             1.247     0.332     3.759     6.154     0.009     4.160     0.849 
+texp            0.110     0.017     6.320     6.084     0.001     4.286     0.852 
+sex:texp       -0.027     0.022    -1.225     5.914     0.267     4.631     0.862
+
+                            Estimate 
+Intercept~~Intercept|school    0.177 
+sex~~sex|school                0.232 
+Intercept~~sex|school         -0.043 
+Residual~~Residual             0.670 
+ICC|school                     0.209
+
+Unadjusted hypothesis test as appropriate in larger samples.
+\end{CodeOutput}
+\end{CodeChunk}
+
+\hypertarget{imputation-with-the-cluster-variable-as-predictor-not-recommended}{%
+\subsubsection{Imputation with the cluster variable as predictor (not
+recommended)}\label{imputation-with-the-cluster-variable-as-predictor-not-recommended}}
+
+We'll now use \texttt{school} as a predictor to impute all other
+variables. This is still not recommended practice, since it only works
+under certain circumstances and results may be biased
+\citep{drec15, ende16}. But at least, it includes some multilevel
+aspect. This method is also called `fixed cluster imputation', and uses
+N-1 indicator variables representing allocation of N clusters as a fixed
+factor in the model \citep{reit06, ende16}. Colloquially, this is
+`multilevel imputation for dummies'.
+
+TODO: Add that it doesn't work with syst missing (only sporadically).
+There's some pro's and con's. May not differ much if the number of
+clusters is low.
+
+TODO: Add that the more the random effects are of interest, the more you
+need ml models.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # adjust the predictor matrix
+R> pred["popular", "school"] <- 1 
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor-1} \end{center}
+
+\begin{CodeInput}
+R> # impute the data, cluster as predictor
+R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{case-study-ii-hiv-data}{%
+\section{Case study II: HIV data}\label{case-study-ii-hiv-data}}
+
+Data are simulated and included in the \texttt{GJRM} package. We will
+use the following variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{region} Cluster variable,
+\item
+  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{marital} Marital status,
+\item
+  \texttt{condom} Condom use during last intercourse,
+\item
+  \texttt{smoke} Smoker (levels; inclusion restriction variable).
+\end{itemize}
+
+The imputation of these date is based on the toy example from
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}.
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/hiv-1} \end{center}
+
+\end{CodeChunk}
+
+\hypertarget{case-study-iii-impact-data}{%
+\section{Case study III: IMPACT data}\label{case-study-iii-impact-data}}
 
@@
 auxiliary variables for imputation.
-
-\hypertarget{setup}{%
-\subsection{Setup}\label{setup}}
-
-{[}TODO: Add environment info, seed and version number(s) somewhere.{]}
-Set up the R environment and load the necessary packages:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> set.seed(2022)
-R> library(mice)         # for imputation
-R> library(ggmice)       # for visualization
-R> library(ggplot2)      # for visualization
-R> library(dplyr)        # for data wrangling
-R> library(lme4)         # for multilevel modeling
-R> library(broom.mixed)  # for multilevel estimates
-R> library(mitml)        # for multilevel pooling
-\end{CodeInput}
-\end{CodeChunk}
 

### 2022-05-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Clean up repo

@@
                     & a separate group mean for each cluster. In addition to random \\
-                    & intercepts, multilevel models can also include random effects and \\
-                    & heterogeneous residual error variances across clusters [see e.g. \\
+                    & intercepts, multilevel models can also include random coefficients \\
+                    & and heterogeneous residual error variances across clusters [see e.g. \\
                     & @gelm06, @hox17 and @jong21]. TODO: add stratification. \\
@@
 [TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").
+Add that the more the random effects are of interest, the more you need ml models.
 And add overview of possible predictor matrix values.]
@@
 
-For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
+**TODO: add where novice and more experienced readers should start reading.** For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
 With the `popmis` data, we show how (and how not) to develop an imputation model.
@@
 Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.
-
-**TODO: add paragraph about novice and more experienced readers.**
 
@@
 
-$$
-\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
-$$
+<!-- $$ -->
+<!-- \texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)} -->
+<!-- $$ -->
+```{r}
+mod <- popular ~ 1 + sex + texp + sex:texp + (1 + sex | school)
+```
 
@@
          .keep = "none")
+
+est_true <- lme4::lmer(mod, popcomp) %>% broom.mixed::tidy()
+
 popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
@@
 ```
+
 Plot the missing data pattern:
@@
 The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}.
+
+ <!-- The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.
 This tells us that the multilevel structure of the data should probably be taken into account.
-If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.
+If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.-->
 
- <!-- The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. -->
-
-Plot the correlations in the incomplete data:
+To develop the best imputation model for the incomplete variable `popular`, we need to know whether the observed values of `popular` are related to observed values of other variables.
+Plot the pair-wise complete correlations in the incomplete data:
 
@@
 
-To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables.
+This shows us that both `sex` and `texp` may be useful imputation model predictors.
+Moreover, the missingness in `popular` may depend on the observed values of other variables.
 We'll highlight one other variable to illustrate, but ideally one would inspect all relations.
@@
   geom_histogram(fill = "white") +
-  facet_grid(. ~ is.na(popular), labeller = label_both)
+  facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
 ```
@@
 
-### Complete case analysis (not recommended)
+<!-- ### Complete case analysis (not recommended) -->
 
-Complete case analysis ignores the observations with missingness altogether, which lowers statistical power and may even introduce bias in MCAR situations.
+<!-- Complete case analysis ignores the observations with missingness altogether, which lowers statistical power and may even introduce bias in MCAR situations.  -->
 
-```{r pop-cca, echo=FALSE}
-est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis)
-
-```
+<!-- ```{r pop-cca, echo=FALSE} -->
+<!-- est_cca <- lme4::lmer(mod, popmis) %>% broom.mixed::tidy() -->
+<!-- # as.matrix(est_cca$estimate) - as.matrix(est_true) -->
+<!-- est_cca$estimate - est_true$estimate -->
+<!-- ``` -->
 
@@
 We include it purely to illustrate the effects of ignoring the clustering in our imputation effort.
-We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).
+<!-- We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).  -->
 
@@
 
-TODO: remove the broom.mixed output, use mitml only
+**TODO: remove the broom.mixed output, use mitml only**
 
@@
 
-TODO: Add that it doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.
-
-TODO: Add that the more the random effects are of interest, the more you need ml models.
+**TODO: Add that it doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.**
 
@@
 ```{r mod}
-mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) # TODO: add random effects ct?
+mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
 ```

### 2022-05-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Clean up repo

@@
 attention during their implementation (``thou shall not simply run
-\texttt{mice()} on any incomplete dataset''). And add overview of
-possible predictor matrix values.{]}
+\texttt{mice()} on any incomplete dataset''). Add that the more the
+random effects are of interest, the more you need ml models. And add
+overview of possible predictor matrix values.{]}
 
@@
 
-For each of these datasets, we will discuss the nature of the
+\textbf{TODO: add where novice and more experienced readers should start
+reading.} For each of these datasets, we will discuss the nature of the
 missingness, choose one or more imputation models and evaluate the
@@
 with incomplete multilevel data.
-
-\textbf{TODO: add paragraph about novice and more experienced readers.}
 
@@
 
-\[
-\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
-\]
+\begin{CodeChunk}
+\begin{CodeInput}
+R> mod <- popular ~ 1 + sex + texp + sex:texp + (1 + sex | school)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 The missingness is univariate and sporadic, which is illustrated in the
-missing data pattern in Figure \ref{fig:pop_pat}. This tells us that the
-multilevel structure of the data should probably be taken into account.
-If we don't, we'll may end up with incorrect imputations, biasing the
-effect of the clusters towards zero.
+missing data pattern in Figure \ref{fig:pop_pat}.
 
-Plot the correlations in the incomplete data:
+To develop the best imputation model for the incomplete variable
+\texttt{popular}, we need to know whether the observed values of
+\texttt{popular} are related to observed values of other variables. Plot
+the pair-wise complete correlations in the incomplete data:
 
@@
 
-To develop the best imputation model for the incomplete variable
-\texttt{popular}, we need to know whether the missingness depends on the
-observed values of other variables. We'll highlight one other variable
-to illustrate, but ideally one would inspect all relations. The
-questions we'll ask are: `Does the missing data of pupil popularity
-(\texttt{popular}) depend on observed teacher popularity
-(\texttt{texp})?'. This can be evaluated statistically, but visual
-inspection usually suffices. We'll make a histogram of \texttt{texp}
-separately for the pupils with known popularity and missing popularity.
+This shows us that both \texttt{sex} and \texttt{texp} may be useful
+imputation model predictors. Moreover, the missingness in
+\texttt{popular} may depend on the observed values of other variables.
+We'll highlight one other variable to illustrate, but ideally one would
+inspect all relations. The questions we'll ask are: `Does the missing
+data of pupil popularity (\texttt{popular}) depend on observed teacher
+popularity (\texttt{texp})?'. This can be evaluated statistically, but
+visual inspection usually suffices. We'll make a histogram of
+\texttt{texp} separately for the pupils with known popularity and
+missing popularity.
 
@@
 +   geom_histogram(fill = "white") +
-+   facet_grid(. ~ is.na(popular), labeller = label_both)
++   facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
 \end{CodeInput}
@@
 
-\hypertarget{complete-case-analysis-not-recommended}{%
-\subsubsection{Complete case analysis (not
-recommended)}\label{complete-case-analysis-not-recommended}}
-
-Complete case analysis ignores the observations with missingness
-altogether, which lowers statistical power and may even introduce bias
-in MCAR situations.
-
 \hypertarget{imputation-ignoring-the-cluster-variable-not-recommended}{%
@@
 units. We include it purely to illustrate the effects of ignoring the
-clustering in our imputation effort. We'll use the default imputation
-methods in \texttt{mice()} (predictive mean matching to impute the
-continuous variables and logistic regression to impute binary
-variables).
+clustering in our imputation effort.
 
@@
 
-TODO: remove the broom.mixed output, use mitml only
+\textbf{TODO: remove the broom.mixed output, use mitml only}
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.354     0.245    13.704     6.831     0.000     3.260     0.813 
-sex             1.247     0.332     3.759     6.154     0.009     4.160     0.849 
-texp            0.110     0.017     6.320     6.084     0.001     4.286     0.852 
-sex:texp       -0.027     0.022    -1.225     5.914     0.267     4.631     0.862
+(Intercept)     3.489     0.179    19.542     9.227     0.000     1.927     0.714 
+sex             1.138     0.197     5.776    11.588     0.000     1.424     0.644 
+texp            0.100     0.014     7.339     6.914     0.000     3.177     0.809 
+sex:texp       -0.019     0.015    -1.217     7.708     0.260     2.576     0.773
 
                             Estimate 
-Intercept~~Intercept|school    0.177 
-sex~~sex|school                0.232 
-Intercept~~sex|school         -0.043 
-Residual~~Residual             0.670 
-ICC|school                     0.209
+Intercept~~Intercept|school    0.123 
+sex~~sex|school                0.143 
+Intercept~~sex|school          0.014 
+Residual~~Residual             0.656 
+ICC|school                     0.158
 
@@
 
-TODO: Add that it doesn't work with syst missing (only sporadically).
-There's some pro's and con's. May not differ much if the number of
-clusters is low.
-
-TODO: Add that the more the random effects are of interest, the more you
-need ml models.
+\textbf{TODO: Add that it doesn't work with syst missing (only
+sporadically). There's some pro's and con's. May not differ much if the
+number of clusters is low.}
 
@@
 \begin{CodeInput}
-R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) # TODO: add random effects ct?
+R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
 \end{CodeInput}

### 2022-05-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Clean up manuscript before feedback

@@
                     & and heterogeneous residual error variances across clusters [see e.g. \\
-                    & @gelm06, @hox17 and @jong21]. TODO: add stratification. \\
+                    & @gelm06, @hox17 and @jong21]. [TODO: add stratification.] \\
 \hline
@@
 ```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
-dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
-  cbind(text = c("1", "1", "2", "2", "", "N", rep("", 30)),
-        miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))
+dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
+  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
+        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
 ggplot(dat, aes(x = cols, y = rows)) +
@@
   ) +
-  scale_y_continuous(breaks = 1:6, labels = c(1:4, "...", "n"), name = NULL, trans = "reverse") +
+  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
   coord_cartesian(expand = c(0,0,0,0)) +
@@
         plot.margin = unit(c(0,0,0,0), "pt"))
-# TODO: add fifth case with cluster 3
-# TODO: use math font
+# TODO: use math font for n and N
 ```
@@
 Since the essence of the true non-response mechanism may not be known, it is generally inferred or assumed to be ignorable (i.e., MCAR or MAR).
-[TODO: add that this assumption may not always be valid, especially with modern types of big dasta sources.] Depending on the actual missingness-generating mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.
+[TODO: add that this assumption may not always be valid, especially with modern types of big data sources.] Depending on the actual missingness-generating mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.
 
@@
 
-[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").
-Add that the more the random effects are of interest, the more you need ml models.
-And add overview of possible predictor matrix values.]
+[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.]
 
@@
 
-**TODO: add where novice and more experienced readers should start reading.** For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
+[TODO: add where novice vs more experienced readers should start reading.] For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
 With the `popmis` data, we show how (and how not) to develop an imputation model.
@@
 library(lme4)         # for multilevel modeling
-library(broom.mixed)  # for multilevel estimates
 library(mitml)        # for multilevel pooling
 ```
+<!-- library(broom.mixed)  # for multilevel estimates -->
 
@@
 est_true <- lme4::lmer(mod, popcomp) %>% broom.mixed::tidy()
+# testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
 
@@
 This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
- **TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?**
+ [TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?]
 
@@
 ```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
-imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+imp <- mice(popmis, pred = pred, print = FALSE)
 ```
 
-**TODO: remove the broom.mixed output, use mitml only**
+<!-- [TODO: remove the broom.mixed output, use mitml only] -->
 
@@
 ```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
-fit_ignored <- imp_ignored %>% 
-  with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
-testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
+fit <- with(imp, 
+            lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+```
+
+Print the estimates:
+```{r}
+testEstimates(as.mitml.result(fit), extra.pars = TRUE)
 ```
@@
 
-**TODO: Add that it doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.**
+[TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.]
 
@@
 # impute the data, cluster as predictor
-imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+imp <- mice(popmis, pred = pred, print = FALSE)
+```
+
+Analyze the imputations:
+```{r, echo=TRUE, message=FALSE, warning=FALSE}
+fit <- with(imp, 
+            lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+```
+
+Print the estimates:
+```{r}
+testEstimates(as.mitml.result(fit), extra.pars = TRUE)
 ```
@@
 The resulting incomplete data can be accessed from [zenodo link to be created](https://zenodo.com).
-TODO: email script to thomas.
 
@@
 
-We will use the following estimates as comparative truth in this tutorial [TODO: make this a table or forest plot instead, to see if there is heterogeneity in the association of `ct` with `mort`]:
-```{r truth}
-fit <- glmer(mod, family = "binomial", data = impact) # fit the model
-tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
-```
+The estimated effects in the complete data are visualized in Figure \ref{}.
 
@@
 ```
+<!-- We will use the following estimates as comparative truth in this tutorial: -->
+```{r truth}
+# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
+# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
+```
 
@@
 
-This shows... [TODO: fill in that we need to impute `ct` and `pupil`.] TODO: remove axis labels
+This shows... [TODO: fill in that we need to impute `ct` and `pupil`.]
 
@@
 
-The first imputation model that we'll use is likely to be invalid.
+<!-- The first imputation model that we'll use is likely to be invalid.
 We do *not* use the cluster identifier `name` as imputation model predictor.
@@
 We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).
+-->
 
----
+<!-- --- -->
 
-\begin{center}
+<!-- \begin{center} -->
 
-Updated until here!
+<!-- Updated until here! -->
 
-\end{center}
+<!-- \end{center} -->
 
----
+<!-- --- -->
 

### 2022-05-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Clean up manuscript before feedback

@@
 assumption may not always be valid, especially with modern types of big
-dasta sources.{]} Depending on the actual missingness-generating
+data sources.{]} Depending on the actual missingness-generating
 mechanism, missing data handling strategies may be more or less
@@
 attention during their implementation (``thou shall not simply run
-\texttt{mice()} on any incomplete dataset''). Add that the more the
-random effects are of interest, the more you need ml models. And add
-overview of possible predictor matrix values.{]}
+\texttt{mice()} on any incomplete dataset'').{]} {[}TODO: Add that the
+more the random effects are of interest, the more you need multilevel
+imputation models.{]} {[}TODO: Add an overview of all possible predictor
+matrix values in manuscript or \texttt{ggmice} legend.{]}
 
@@
 
-\textbf{TODO: add where novice and more experienced readers should start
-reading.} For each of these datasets, we will discuss the nature of the
-missingness, choose one or more imputation models and evaluate the
+{[}TODO: add where novice vs more experienced readers should start
+reading.{]} For each of these datasets, we will discuss the nature of
+the missingness, choose one or more imputation models and evaluate the
 imputed data, but we will also highlight one specific aspect of the
@@
 R> library(lme4)         # for multilevel modeling
-R> library(broom.mixed)  # for multilevel estimates
 R> library(mitml)        # for multilevel pooling
@@
 of \texttt{texp} depending on the missingness indicator of
-\texttt{popular} (t = -0.873, p = 0.383). \textbf{TODO: think about what
-is a meaningful rule of thumb to signal that the user should be
-worried?}
+\texttt{popular} (t = -0.873, p = 0.383). {[}TODO: think about what is a
+meaningful rule of thumb to signal that the user should be worried?{]}
 
@@
 \begin{CodeInput}
-R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+R> imp <- mice(popmis, pred = pred, print = FALSE)
 \end{CodeInput}
 \end{CodeChunk}
-
-\textbf{TODO: remove the broom.mixed output, use mitml only}
 
@@
 \begin{CodeInput}
-R> fit_ignored <- imp_ignored %>% 
-+   with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
-R> testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
+R> fit <- with(imp, 
++             lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+\end{CodeInput}
+\end{CodeChunk}
+
+Print the estimates:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
 \end{CodeInput}
@@
 
-testEstimates(model = as.mitml.result(fit_ignored), var.comp = TRUE)
+testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.489     0.179    19.542     9.227     0.000     1.927     0.714 
-sex             1.138     0.197     5.776    11.588     0.000     1.424     0.644 
-texp            0.100     0.014     7.339     6.914     0.000     3.177     0.809 
-sex:texp       -0.019     0.015    -1.217     7.708     0.260     2.576     0.773
+(Intercept)     3.445     0.153    22.473    20.745     0.000     0.783     0.486 
+sex             1.175     0.205     5.728    13.364     0.000     1.208     0.602 
+texp            0.103     0.010     9.991    15.982     0.000     1.001     0.553 
+sex:texp       -0.020     0.014    -1.460    11.610     0.171     1.421     0.644
 
                             Estimate 
-Intercept~~Intercept|school    0.123 
-sex~~sex|school                0.143 
-Intercept~~sex|school          0.014 
-Residual~~Residual             0.656 
-ICC|school                     0.158
+Intercept~~Intercept|school    0.158 
+sex~~sex|school                0.186 
+Intercept~~sex|school         -0.025 
+Residual~~Residual             0.704 
+ICC|school                     0.183
 
@@
 
-\textbf{TODO: Add that it doesn't work with syst missing (only
-sporadically). There's some pro's and con's. May not differ much if the
-number of clusters is low.}
+{[}TODO: Add that it doesn't work with systematic missingness (only with
+sporadic). There's some pros and cons, and it may not even differ much
+if the number of clusters is low.{]}
 
@@
 R> # impute the data, cluster as predictor
-R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
+R> imp <- mice(popmis, pred = pred, print = FALSE)
 \end{CodeInput}
+\end{CodeChunk}
+
+Analyze the imputations:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- with(imp, 
++             lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+\end{CodeInput}
+\end{CodeChunk}
+
+Print the estimates:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+
+testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
+
+Final parameter estimates and inferences obtained from 5 imputed data sets.
+
+             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
+(Intercept)     3.341     0.179    18.648    14.134     0.000     1.137     0.587 
+sex             1.267     0.216     5.866    12.916     0.000     1.255     0.612 
+texp            0.108     0.012     9.061    12.109     0.000     1.351     0.631 
+sex:texp       -0.030     0.014    -2.058    11.313     0.063     1.467     0.651
+
+                            Estimate 
+Intercept~~Intercept|school    0.198 
+sex~~sex|school                0.231 
+Intercept~~sex|school         -0.042 
+Residual~~Residual             0.622 
+ICC|school                     0.242
+
+Unadjusted hypothesis test as appropriate in larger samples.
+\end{CodeOutput}
 \end{CodeChunk}
@@
 before imputation). The resulting incomplete data can be accessed from
-\href{https://zenodo.com}{zenodo link to be created}. TODO: email script
-to thomas.
+\href{https://zenodo.com}{zenodo link to be created}.
 
@@
 
-We will use the following estimates as comparative truth in this
-tutorial {[}TODO: make this a table or forest plot instead, to see if
-there is heterogeneity in the association of \texttt{ct} with
-\texttt{mort}{]}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> fit <- glmer(mod, family = "binomial", data = impact) # fit the model
-R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 11 x 9
-   effect group term  estimate std.error statistic    p.value conf.low conf.high
-   <chr>  <chr> <chr>    <dbl>     <dbl>     <dbl>      <dbl>    <dbl>     <dbl>
- 1 fixed  <NA>  (Int~    0.101   0.0173     -13.4   4.94e- 41   0.0723     0.141
- 2 fixed  <NA>  type~    0.713   0.123       -1.96  5.01e-  2   0.509      1.00 
- 3 fixed  <NA>  age      1.03    0.00165     20.2   2.13e- 90   1.03       1.04 
- 4 fixed  <NA>  moto~    0.553   0.0386      -8.50  1.95e- 17   0.482      0.634
- 5 fixed  <NA>  moto~    0.405   0.0289     -12.7   8.43e- 37   0.352      0.466
- 6 fixed  <NA>  moto~    0.275   0.0202     -17.6   1.67e- 69   0.239      0.318
- 7 fixed  <NA>  pupi~    3.73    0.230       21.4   2.09e-101   3.31       4.21 
- 8 fixed  <NA>  pupi~    1.87    0.133        8.80  1.36e- 18   1.63       2.15 
- 9 fixed  <NA>  ctIII    2.25    0.157       11.6   5.12e- 31   1.96       2.58 
-10 fixed  <NA>  ctIV~    2.30    0.136       14.0   9.47e- 45   2.05       2.58 
-11 ran_p~ name  sd__~    0.277  NA           NA    NA          NA         NA    
-\end{CodeOutput}
-\end{CodeChunk}
+The estimated effects in the complete data are visualized in Figure
+\ref{}.
 
@@
 
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # fit <- glmer(mod, family = "binomial", data = impact) # fit the model
+R> # tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
+\end{CodeInput}
 \end{CodeChunk}
@@
 This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
-and \texttt{pupil}.{]} TODO: remove axis labels
+and \texttt{pupil}.{]}
 
@@
 
-The first imputation model that we'll use is likely to be invalid. We do
-\emph{not} use the cluster identifier \texttt{name} as imputation model
-predictor. With this model, we ignore the multilevel structure of the
-data, despite the high ICC. This assumes exchangeability between units.
-We include it purely to illustrate the effects of ignoring the
-clustering in our imputation effort. We'll use the default imputation
-methods in \texttt{mice()} (predictive mean matching to impute the
-continuous variables and logistic regression to impute binary
-variables).
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\begin{center}
-
-Updated until here!
-
-\end{center}
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
 Create a methods vector and predictor matrix, and make sure

### 2022-06-20 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

@@
 abstract: >
-  This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
+  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}.
 Footnotes in the current version show work in progress/under construction.
@@
 
-In many contemporary data analysis efforts, some form of hierarchical or clustered data structures are recorded.
-In the simplest case, such a structure entails the nesting of units within clusters (e.g., students within school classes).
+Many datasets include individuals from multiple settings, geographic regions, or even different studies.
+In the simplest case, individuals (e.g., students) are nested within so-called clusters (e.g., school classes).
 More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases).
-The clustered structure of multilevel data should be taken into account when developing analysis models: 1) for the simple reason that groups of observations share some common variance, and 2) because ignoring multilevel structures can be harmful to the statistical inferences and introduce bias in estimators [@hox17].
-There are many names for models that take clustering into account.
-Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
+In general, individuals from the same cluster tend to be more similar than individuals from other clusters.
+In statistical terms, this implies that observations from the same cluster are correlated.
+If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01].
+[TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation.] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
+These models are also know as 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
 Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.
@@
 
-The process of analyzing multilevel data is further complicated when not all data entries are observed.
-Just as with single level data, missingness may occur at the unit level.
-But with multiple levels of data comes the potential for clustered missingness.
-Therefore, incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13].
-Systematic missingness implies that one or more variables are never observed in a certain cluster.
+Like any other dataset, clustered datasets are prone to missing data.
+Several strategies can be used to handle missing data, including complete case analysis and imputation.
+We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values.
+Afterwards, the completed data can be analyzed as if they were completely observed.
+In contrast to single imputation (where missing data are only replaced once), multiple imputation allows to preserve uncertainty due to missingness and is therefore recommended (c.f.
+Rubin 1976).
+
+When clustered datasets are affected by missing values, we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13].
+Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18].
+For example, it is possible that test results are missing for several students in one or more classes.
+[TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are systematically missing.
+[TODO: Refer to Figure 1 and put interpretation in the figure caption.]
+
+<!-- Systematic missingness implies that one or more variables are never observed in a certain cluster.
 With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18].
 We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters and $p$ variables.
+-->
 
@@
 ```
-Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing.
+<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing.
 To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account.
@@
 The wrong choice of missing data handling method can thus be extremely harmful to the inferences.
+-->
 
-A key characteristic of the missing data to take into account in analyses is the mechanism behind the missingness.
-We distinguish between MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice this distinction is less clear.
-Since the essence of the true non-response mechanism may not be known, it is generally inferred or assumed to be ignorable (i.e., MCAR or MAR).
-[TODO: add that this assumption may not always be valid, especially with modern types of big data sources.] Depending on the actual missingness-generating mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.
+Imputation of missing data requires to consider the mechanism behind the missingness.
+Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}).
+For each of these three missingness generating mechanisms, different imputation strategies are warranted @yuce08 and @hox15.
+We here consider the general case that data are MAR, and expand on special MNAR situations.
 
@@
 
-Since excluding observations is not a desirable workflow, the missingness in multilevel data should be accommodated \emph{before} or _within_ the analysis of scientific interest.
+<!-- Since excluding observations is not a desirable workflow, the missingness in multilevel data should be accommodated \emph{before} or _within_ the analysis of scientific interest.
 In this paper, we focus on the former approach: imputing (i.e., filling in) the missing data with plausible values, whereafter the completed data may be analyzed as if it were completely observed.
@@
 @rubi76].
+-->
+
 The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
 \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
-In this paper, we will discuss how to use \pkg{mice} in the context of multilevel data.
+However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent.
+For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.
 
@@
 
-We illustrate how to impute incomplete multilevel data by means of three case studies:
+We illustrate imputation of incomplete multilevel data using three case studies:
 
-- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, @mice];
-- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, @GJRM];
-- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc].
+- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools with data that are MAR, @mice];
+- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies with data that are MAR, @metamisc];
+- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions with data that are MNAR, @GJRM].
 
-[TODO: add where novice vs more experienced readers should start reading.] For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
-With the `popmis` data, we show how (and how not) to develop an imputation model.
+For each of these datasets, we discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
+
+This tutorial is dedicated to readers who are unfamiliar with multiple imptuation.
+More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study IMPACT) or  under MNAR conditions (case study HIV).
+
+<!-- With the `popmis` data, we show how (and how not) to develop an imputation model.
 With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods.
@@
 Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.
+-->
 
@@
 
-In this section we'll go over the different steps involved with imputing incomplete multilevel data.
-The data we're using is the `popmis` dataset from the `mice` package.
-This is a simulated dataset with pupils ($n = 2000$) clustered within schools ($N = 100$).
-In this tutorial we'll use the following variables:
+In this section we'll go over the different steps involved with imputing incomplete multilevel data with the R package mice.
+We consider the simulated `popmis` dataset, which included pupils ($n = 2000$) clustered within schools ($N = 100$).
+The following variables are of primary interest:
 
@@
 
+The research objective of the popmis dataset is to predict the pupils' popularity based on their gender and the experience of the teacher.
 The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction.
@@
 ```
+
+[TODO: think about introducing systematically missing hiv.]
 

### 2022-06-20 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

@@
 \Abstract{
-This is a tutorial paper on imputing incomplete multilevel data with
-\pkg{mice}. Footnotes in the current version show work in progress/under
-construction. The last section is not part of the manuscript, but purely
-for reminders. We aim to submit at JSS, so there is no word count limit
-(``There is no page limit, nor a limit on the number of figures or
-tables''). {[}Just adding some text to get a better guess of what the
-actura abstract will look like: Lorem ipsum dolor sit amet, consectetur
-adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore
-magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco
-laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla
-pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa
-qui officia deserunt mollit anim id est laborum.{]}
+This tutorial illustrates the imputation of incomplete multilevel data
+with the \proglang{R} packackage \pkg{mice}. Footnotes in the current
+version show work in progress/under construction. The last section is
+not part of the manuscript, but purely for reminders. We aim to submit
+at JSS, so there is no word count limit (``There is no page limit, nor a
+limit on the number of figures or tables''). {[}Just adding some text to
+get a better guess of what the actura abstract will look like: Lorem
+ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor
+incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam,
+quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo
+consequat. Duis aute irure dolor in reprehenderit in voluptate velit
+esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat
+cupidatat non proident, sunt in culpa qui officia deserunt mollit anim
+id est laborum.{]}
 }
@@
 
-In many contemporary data analysis efforts, some form of hierarchical or
-clustered data structures are recorded. In the simplest case, such a
-structure entails the nesting of units within clusters (e.g., students
-within school classes). More complex clustered structures may occur when
-there are multiple hierarchical levels (e.g., patients within hospitals
-within regions or countries), or when the clustering is non-nested
-(e.g., electronic health record data from diverse settings and
-populations within large databases). The clustered structure of
-multilevel data should be taken into account when developing analysis
-models: 1) for the simple reason that groups of observations share some
-common variance, and 2) because ignoring multilevel structures can be
-harmful to the statistical inferences and introduce bias in estimators
-\citep{hox17}. There are many names for models that take clustering into
-account. Some popular examples are `multilevel models', `hierarchical
-models', `mixed effect models' and `random effect models'. Table
-\ref{tab:clus} provides an overview of some key concepts in multilevel
-modeling.
+Many datasets include individuals from multiple settings, geographic
+regions, or even different studies. In the simplest case, individuals
+(e.g., students) are nested within so-called clusters (e.g., school
+classes). More complex clustered structures may occur when there are
+multiple hierarchical levels (e.g., patients within hospitals within
+regions or countries), or when the clustering is non-nested (e.g.,
+electronic health record data from diverse settings and populations
+within large databases). In general, individuals from the same cluster
+tend to be more similar than individuals from other clusters. In
+statistical terms, this implies that observations from the same cluster
+are correlated. If this correlation is left unaddressed, estimates of
+\emph{p} values, confidence intervals even model parameters are prone to
+bias \citep{loca01}. {[}TODO: make a link to imputation methods, which
+require adequate handling and propagation of variance; we are not
+recommending the adoption of multilevel models for data analysis here,
+but rather for imputation.{]} Statistical methods for clustered data
+typically adopt hierarchical models that explicitly describe the
+grouping of observations. These models are also know as `multilevel
+models', `hierarchical models', `mixed effect models' and `random effect
+models'. Table \ref{tab:clus} provides an overview of some key concepts
+in multilevel modeling.
 
@@
 
-The process of analyzing multilevel data is further complicated when not
-all data entries are observed. Just as with single level data,
-missingness may occur at the unit level. But with multiple levels of
-data comes the potential for clustered missingness. Therefore,
-incomplete multilevel data can be categorized into two general patterns:
-systematic missingness and sporadic missingness \citep{resc13}.
-Systematic missingness implies that one or more variables are never
-observed in a certain cluster. With sporadic missingness there may be
-observed data for some but not all units in a cluster
-\citep{buur18, jola18}. We have visualized this difference in Figure 1,
-which shows an \(n \times p\) set \(\mathbf{X} = X_1, \dots, X_p\), with
-\(n\) units distributed over \(N\) clusters and \(p\) variables.
+Like any other dataset, clustered datasets are prone to missing data.
+Several strategies can be used to handle missing data, including
+complete case analysis and imputation. We focus on the latter approach
+and discuss statistical methods for replacing the missing data with one
+or more plausible values. Afterwards, the completed data can be analyzed
+as if they were completely observed. In contrast to single imputation
+(where missing data are only replaced once), multiple imputation allows
+to preserve uncertainty due to missingness and is therefore recommended
+(c.f. Rubin 1976).
+
+When clustered datasets are affected by missing values, we can
+distinguish between two types of missing data: sporadic missingness and
+systematic missingness \citep{resc13}. Sporadic missingness arises when
+variables are missing for some but not all of the units in a cluster
+\citep{buur18, jola18}. For example, it is possible that test results
+are missing for several students in one or more classes. {[}TODO:
+Provide an example for one of the case studies below.{]} When all
+observations are missing within one or more clusters, data are
+systematically missing. {[}TODO: Refer to Figure 1 and put
+interpretation in the figure caption.{]}
 
@@
 
-Column \(X_1\) in Figure 1 is completely observed, column \(X_2\) is
-systematically missing in cluster 2, and column \(X_3\) is sporadically
-missing. To analyze these incomplete data, we have to take the nature of
-the missingness and the cluster structure into account. For example, the
-sporadic missingness in \(X_3\) could be easily amended if this would be
-a cluster-level variable (and thus constant within clusters). We could
-then just extrapolate the true (but missing) value of \(X_3\) for unit 1
-from unit 2, and the value for unit 4 from unit 3. If \(X_3\) would
-instead be a unit-level variable (which may vary within clusters), we
-could not just recover the unobserved `truth', but would need to use
-some kind of missing data method, or discard the incomplete units
-altogether (i.e., complete case analysis). Complete case analysis can
-however introduce bias in statistical inferences and lowers statistical
-power. Further, with the systematic missingness in \(X_2\), it would be
-impossible to fit a multilevel model without accommodating the
-missingness in some way. Complete case analysis in that case would mean
-excluding the entire cluster from the analyses. The wrong choice of
-missing data handling method can thus be extremely harmful to the
-inferences.
-
-A key characteristic of the missing data to take into account in
-analyses is the mechanism behind the missingness. We distinguish between
-MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice
-this distinction is less clear. Since the essence of the true
-non-response mechanism may not be known, it is generally inferred or
-assumed to be ignorable (i.e., MCAR or MAR). {[}TODO: add that this
-assumption may not always be valid, especially with modern types of big
-data sources.{]} Depending on the actual missingness-generating
-mechanism, missing data handling strategies may be more or less
-suitable, see e.g., \citet{yuce08} and \citet{hox15}.
+Imputation of missing data requires to consider the mechanism behind the
+missingness. Rubin proposed to distinguish between data that are missing
+completely at random (MCAR), data that are missing at random (MAR) and
+data that are missing not at random (MNAR; see Table \ref{tab:miss}).
+For each of these three missingness generating mechanisms, different
+imputation strategies are warranted \citet{yuce08} and \citet{hox15}. We
+here consider the general case that data are MAR, and expand on special
+MNAR situations.
 
@@
 
-Since excluding observations is not a desirable workflow, the
-missingness in multilevel data should be accommodated \emph{before} or
-\emph{within} the analysis of scientific interest. In this paper, we
-focus on the former approach: imputing (i.e., filling in) the missing
-data with plausible values, whereafter the completed data may be
-analyzed as if it were completely observed. Imputation separates the
-missing data problem from the scientific problem, which makes the
-missing data strategy very generic and popular. If each missing value is
-replaced multiple times, the resulting inferences may validly convey the
-uncertainty due to missingness \citep[c.f.][]{rubi76}. The \proglang{R}
-package \pkg{mice} has become the de-facto standard for imputation by
-chained equations, which iteratively solves the missingness on a
-variable-by-variable basis. \pkg{mice} is known to yield valid
-inferences under many different missing data circumstances
-\citep{buur18}. In this paper, we will discuss how to use \pkg{mice} in
-the context of multilevel data.
+The \proglang{R} package \pkg{mice} has become the de-facto standard for
+imputation by chained equations, which iteratively solves the
+missingness on a variable-by-variable basis. \pkg{mice} is known to
+yield valid inferences under many different missing data circumstances
+\citep{buur18}. However, commonly used imputation methods were not
+designed for use in clustered data and usually generate observations
+that are independent. For this reason, we discuss how the \proglang{R}
+package \pkg{mice} can be used to impute multilevel data.
 
@@
 
-We illustrate how to impute incomplete multilevel data by means of three
-case studies:
+We illustrate imputation of incomplete multilevel data using three case
+studies:
 
@@
   \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
-  perceived popularity, \(n = 2,000\) pupils across \(N = 100\)
-  schools,][]{mice};
-\item
-  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
-  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions,][]{GJRM};
+  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
+  with data that are MAR,][]{mice};
 \item
@@
   on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
-  studies,][]{metamisc}.
+  studies with data that are MAR,][]{metamisc};
+\item
+  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
+  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions with data
+  that are MNAR,][]{GJRM}.
 \end{itemize}
 
-{[}TODO: add where novice vs more experienced readers should start
-reading.{]} For each of these datasets, we will discuss the nature of
-the missingness, choose one or more imputation models and evaluate the
-imputed data, but we will also highlight one specific aspect of the
-imputation workflow. With the \texttt{popmis} data, we show how (and how
-not) to develop an imputation model. With the \texttt{hiv} data we focus
-on extending the imputation model to include Heckman-type
-selection-inclusion methods. With the \texttt{impact} data we provide an
-example of multivariate missingness in real-world data. Together, this
-should give enough scaffolding for applied researchers who are faced
-with incomplete multilevel data.
+For each of these datasets, we discuss the nature of the missingness,
+choose one or more imputation models and evaluate the imputed data, but
+we will also highlight one specific aspect of the imputation workflow.
+
+This tutorial is dedicated to readers who are unfamiliar with multiple
+imptuation. More experienced readers can skip the introduction (case
+study 1) and directly head to practical applications of multilevel
+imputation under MAR conditions (case study IMPACT) or under MNAR
+conditions (case study HIV).
 
@@
 In this section we'll go over the different steps involved with imputing
-incomplete multilevel data. The data we're using is the \texttt{popmis}
-dataset from the \texttt{mice} package. This is a simulated dataset with
-pupils (\(n = 2000\)) clustered within schools (\(N = 100\)). In this
-tutorial we'll use the following variables:
+incomplete multilevel data with the R package mice. We consider the
+simulated \texttt{popmis} dataset, which included pupils (\(n = 2000\))
+clustered within schools (\(N = 100\)). The following variables are of
+primary interest:
 
@@
 
-The analysis model corresponding to this dataset is multilevel
-regression with random intercepts, random slopes and a cross-level
-interaction. The outcome variable is \texttt{popular}, which is
-predicted from the unit-level variable \texttt{sex} and the
-cluster-level variable \texttt{texp}:
+The research objective of the popmis dataset is to predict the pupils'
+popularity based on their gender and the experience of the teacher. The
+analysis model corresponding to this dataset is multilevel regression
+with random intercepts, random slopes and a cross-level interaction. The
+outcome variable is \texttt{popular}, which is predicted from the
+unit-level variable \texttt{sex} and the cluster-level variable
+\texttt{texp}:
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.445     0.153    22.473    20.745     0.000     0.783     0.486 
-sex             1.175     0.205     5.728    13.364     0.000     1.208     0.602 
-texp            0.103     0.010     9.991    15.982     0.000     1.001     0.553 
-sex:texp       -0.020     0.014    -1.460    11.610     0.171     1.421     0.644
+(Intercept)     3.469     0.139    24.958    32.413     0.000     0.542     0.388 
+sex             1.128     0.167     6.744    22.962     0.000     0.716     0.462 
+texp            0.102     0.011     9.237    11.848     0.000     1.387     0.637 
+sex:texp       -0.018     0.011    -1.616    16.878     0.125     0.949     0.538
 
                             Estimate 
-Intercept~~Intercept|school    0.158 
-sex~~sex|school                0.186 
-Intercept~~sex|school         -0.025 
-Residual~~Residual             0.704 
-ICC|school                     0.183
+Intercept~~Intercept|school    0.150 
+sex~~sex|school                0.143 
+Intercept~~sex|school          0.002 
+Residual~~Residual             0.677 
+ICC|school                     0.182
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.341     0.179    18.648    14.134     0.000     1.137     0.587 
-sex             1.267     0.216     5.866    12.916     0.000     1.255     0.612 
-texp            0.108     0.012     9.061    12.109     0.000     1.351     0.631 
-sex:texp       -0.030     0.014    -2.058    11.313     0.063     1.467     0.651
+(Intercept)     3.369     0.163    20.666    17.473     0.000     0.917     0.529 
+sex             1.228     0.178     6.910    44.805     0.000     0.426     0.328 
+texp            0.105     0.012     8.713    10.517     0.000     1.609     0.673 
+sex:texp       -0.027     0.011    -2.416    57.000     0.019     0.360     0.289
 
                             Estimate 
-Intercept~~Intercept|school    0.198 
-sex~~sex|school                0.231 
-Intercept~~sex|school         -0.042 
-Residual~~Residual             0.622 
-ICC|school                     0.242
+Intercept~~Intercept|school    0.178 
+sex~~sex|school                0.255 
+Intercept~~sex|school         -0.036 
+Residual~~Residual             0.628 
+ICC|school                     0.220
 
@@
 \end{CodeChunk}
+
+{[}TODO: think about introducing systematically missing hiv.{]}
 

### 2022-06-21 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Make forest plot dependent on sample size

@@
 
+[TODO: explain case study]
+
 In this section we'll go over the different steps involved with imputing incomplete multilevel data with the R package mice.
@@
 
+### Imputation with multilevel model
+
+```{r pop_multilevel, message=FALSE, warning=FALSE}
+# adjust the predictor matrix
+pred["popular", "school"] <- -2 
+plot_pred(pred)
+
+# impute the data, cluster as predictor
+imp <- mice(popmis, pred = pred, print = FALSE)
+```
+
+Analyze the imputations:
+```{r, echo=TRUE, message=FALSE, warning=FALSE}
+fit <- with(imp, 
+            lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+```
+
+Print the estimates:
+```{r}
+testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+```
+
 # Case study II: HIV data
+
+[TODO: change the order of case study 2 and 3]
 
@@
 
-[TODO: think about introducing systematically missing hiv.]
+# Case study III: IMPACT data
 
-# Case study III: IMPACT data
+[TODO: check if there is systematic missingness in this dataset, if not make Marshall Computerized Tomography classification (ct) systematically missing.]
 
@@
       cbind(name = as.character(.x), .)
-  )
-filter(fits, term %in% c("ctIII")) %>% #, "ctIV/V"
-  ggplot(aes(y = name, x = estimate)) +
+  ) %>% filter(term %in% c("ctIII", "ctIV/V"))
+
+n <- impact %>% group_by(name) %>% 
+  tally()
+
+fits %>% left_join(n, by = "name") %>% 
+  ggplot(aes(y = name, x = estimate, color = term, size = n)) +
   geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
-  geom_point() +
-  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high)) +
+  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, size = 0.5), position = position_dodge(width = 0.5)) +
+  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
   labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
-  theme_classic()
+  theme_classic() +
+  scale_size(guide = 'none')
 # TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
@@
 ```
+
+[TODO: show how much variance there is after different methods]
 

### 2022-06-21 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Make forest plot dependent on sample size

@@
 
+{[}TODO: explain case study{]}
+
 In this section we'll go over the different steps involved with imputing
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.469     0.139    24.958    32.413     0.000     0.542     0.388 
-sex             1.128     0.167     6.744    22.962     0.000     0.716     0.462 
-texp            0.102     0.011     9.237    11.848     0.000     1.387     0.637 
-sex:texp       -0.018     0.011    -1.616    16.878     0.125     0.949     0.538
+(Intercept)     3.436     0.179    19.207    11.230     0.000     1.480     0.653 
+sex             1.210     0.208     5.812    13.898     0.000     1.157     0.591 
+texp            0.101     0.014     7.261     7.456     0.000     2.738     0.784 
+sex:texp       -0.020     0.017    -1.153     7.534     0.284     2.686     0.780
 
                             Estimate 
-Intercept~~Intercept|school    0.150 
-sex~~sex|school                0.143 
-Intercept~~sex|school          0.002 
-Residual~~Residual             0.677 
-ICC|school                     0.182
+Intercept~~Intercept|school    0.158 
+sex~~sex|school                0.212 
+Intercept~~sex|school         -0.026 
+Residual~~Residual             0.663 
+ICC|school                     0.188
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.369     0.163    20.666    17.473     0.000     0.917     0.529 
-sex             1.228     0.178     6.910    44.805     0.000     0.426     0.328 
-texp            0.105     0.012     8.713    10.517     0.000     1.609     0.673 
-sex:texp       -0.027     0.011    -2.416    57.000     0.019     0.360     0.289
+(Intercept)     3.361     0.177    18.967    16.581     0.000     0.965     0.543 
+sex             1.250     0.243     5.145     9.696     0.000     1.796     0.699 
+texp            0.105     0.014     7.425     8.751     0.000     2.087     0.731 
+sex:texp       -0.026     0.019    -1.388     6.889     0.208     3.202     0.810
 
                             Estimate 
-Intercept~~Intercept|school    0.178 
-sex~~sex|school                0.255 
-Intercept~~sex|school         -0.036 
-Residual~~Residual             0.628 
-ICC|school                     0.220
+Intercept~~Intercept|school    0.216 
+sex~~sex|school                0.240 
+Intercept~~sex|school         -0.045 
+Residual~~Residual             0.615 
+ICC|school                     0.259
+
+Unadjusted hypothesis test as appropriate in larger samples.
+\end{CodeOutput}
+\end{CodeChunk}
+
+\hypertarget{imputation-with-multilevel-model}{%
+\subsubsection{Imputation with multilevel
+model}\label{imputation-with-multilevel-model}}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # adjust the predictor matrix
+R> pred["popular", "school"] <- -2 
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_multilevel-1} \end{center}
+
+\begin{CodeInput}
+R> # impute the data, cluster as predictor
+R> imp <- mice(popmis, pred = pred, print = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+Analyze the imputations:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- with(imp, 
++             lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+\end{CodeInput}
+\end{CodeChunk}
+
+Print the estimates:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+
+testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
+
+Final parameter estimates and inferences obtained from 5 imputed data sets.
+
+             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
+(Intercept)     3.333     0.177    18.832    16.459     0.000     0.972     0.545 
+sex             1.356     0.193     7.014    37.019     0.000     0.490     0.362 
+texp            0.106     0.012     8.870    13.221     0.000     1.222     0.606 
+sex:texp       -0.034     0.012    -2.901    51.319     0.005     0.387     0.306
+
+                            Estimate 
+Intercept~~Intercept|school    0.213 
+sex~~sex|school                0.307 
+Intercept~~sex|school         -0.084 
+Residual~~Residual             0.622 
+ICC|school                     0.254
 
@@
 \section{Case study II: HIV data}\label{case-study-ii-hiv-data}}
+
+{[}TODO: change the order of case study 2 and 3{]}
 
@@
 
-{[}TODO: think about introducing systematically missing hiv.{]}
-
 \hypertarget{case-study-iii-impact-data}{%
 \section{Case study III: IMPACT data}\label{case-study-iii-impact-data}}
+
+{[}TODO: check if there is systematic missingness in this dataset, if
+not make Marshall Computerized Tomography classification (ct)
+systematically missing.{]}
 
@@
 \end{CodeChunk}
+
+{[}TODO: show how much variance there is after different methods{]}
 

### 2022-06-23 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Add forest plot for popularity data

@@
 ```{r}
-mod <- popular ~ 1 + sex + texp + sex:texp + (1 + sex | school)
+mod <- popular ~ 1 + sex + (1 | school)
 ```
@@
          sex = SEX,
-         texp = TEXP,
          .keep = "none")
@@
 # testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
+```
 
-popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
+The true effect is:
+
+```{r echo=FALSE}
+fits <-
+  purrr::map_dfr(
+    unique(popcomp$school),
+    ~ {
+      lm(
+        mod,
+        data = filter(popcomp, school == .x)
+      )
+    } %>%
+      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
+      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
+      cbind(school = as.character(.x), .)
+  ) %>% filter(term %in% "sex")
+
+n <- popcomp %>% group_by(school) %>% 
+  tally() %>% 
+  mutate(school = as.character(school))
+
+fits %>% left_join(n, by = "school") %>% 
+  ggplot(aes(y = school, x = estimate, size = n)) +
+  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
+  geom_linerange(aes(y = school, xmin = conf.low, xmax = conf.high, size = 0.5), position = position_dodge(width = 0.5)) +
+  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
+  labs(y = "Study", x = "", color = NULL) +
+  theme_classic() +
+  scale_size(guide = 'none')
 ```
@@
 ```{r}
-popmis <- popmis[, c("school", "popular", "sex", "texp")] 
+popmis <- popmis[, c("school", "popular", "sex")] 
 ```
@@
 
-This shows us that both `sex` and `texp` may be useful imputation model predictors.
-Moreover, the missingness in `popular` may depend on the observed values of other variables.
-We'll highlight one other variable to illustrate, but ideally one would inspect all relations.
+This shows us that  `sex`  may be a useful imputation model predictor. Moreover, the missingness in `popular` may depend on the observed values of other variables.
+
+<!-- We'll highlight one other variable to illustrate, but ideally one would inspect all relations.
 The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'.
@@
 We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.
+-->
 
-Plot the histogram for teacher experience conditional on the missingness indicator of `popular`:
+<!-- Plot the histogram for teacher experience conditional on the missingness indicator of `popular`: -->
 
 ```{r pop-hist}
-ggmice(popmis, aes(texp)) +
-  geom_histogram(fill = "white") +
-  facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
+# ggmice(popmis, aes(sex)) +
+#   geom_histogram(fill = "white") +
+#   facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
+
+ggplot(popmis, aes(y = popular, group = sex)) +
+  geom_boxplot() + 
+  theme_classic()
 ```
 
-This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
- [TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?]
+<!-- This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")).
+-->
+<!--  [TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?] -->
 
@@
 fit <- with(imp, 
-            lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+            lmer(popular ~ 1 + sex  + (1 | school))) 
 ```
@@
 fit <- with(imp, 
-            lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
+            lmer(popular ~ 1 + sex + (1 | school))) 
 ```

### 2022-06-23 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Add forest plot for popularity data

@@
 \begin{CodeInput}
-R> mod <- popular ~ 1 + sex + texp + sex:texp + (1 + sex | school)
+R> mod <- popular ~ 1 + sex + (1 | school)
 \end{CodeInput}
+\end{CodeChunk}
+
+The true effect is:
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
 \end{CodeChunk}
@@
 \begin{CodeInput}
-R> popmis <- popmis[, c("school", "popular", "sex", "texp")] 
+R> popmis <- popmis[, c("school", "popular", "sex")] 
 \end{CodeInput}
@@
 
-This shows us that both \texttt{sex} and \texttt{texp} may be useful
-imputation model predictors. Moreover, the missingness in
-\texttt{popular} may depend on the observed values of other variables.
-We'll highlight one other variable to illustrate, but ideally one would
-inspect all relations. The questions we'll ask are: `Does the missing
-data of pupil popularity (\texttt{popular}) depend on observed teacher
-popularity (\texttt{texp})?'. This can be evaluated statistically, but
-visual inspection usually suffices. We'll make a histogram of
-\texttt{texp} separately for the pupils with known popularity and
-missing popularity.
-
-Plot the histogram for teacher experience conditional on the missingness
-indicator of \texttt{popular}:
+This shows us that \texttt{sex} may be a useful imputation model
+predictor. Moreover, the missingness in \texttt{popular} may depend on
+the observed values of other variables.
 
@@
 \begin{CodeInput}
-R> ggmice(popmis, aes(texp)) +
-+   geom_histogram(fill = "white") +
-+   facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
+R> # ggmice(popmis, aes(sex)) +
+R> #   geom_histogram(fill = "white") +
+R> #   facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
+R> 
+R> ggplot(popmis, aes(y = popular, group = sex)) +
++   geom_boxplot() + 
++   theme_classic()
 \end{CodeInput}
@@
 \end{CodeChunk}
-
-This shows us that there are no apparent differences in the distribution
-of \texttt{texp} depending on the missingness indicator of
-\texttt{popular} (t = -0.873, p = 0.383). {[}TODO: think about what is a
-meaningful rule of thumb to signal that the user should be worried?{]}
 
@@
 R> fit <- with(imp, 
-+             lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
++             lmer(popular ~ 1 + sex  + (1 | school))) 
 \end{CodeInput}
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.436     0.179    19.207    11.230     0.000     1.480     0.653 
-sex             1.210     0.208     5.812    13.898     0.000     1.157     0.591 
-texp            0.101     0.014     7.261     7.456     0.000     2.738     0.784 
-sex:texp       -0.020     0.017    -1.153     7.534     0.284     2.686     0.780
+(Intercept)     4.820     0.353    13.648     4.245     0.000    33.200     0.979 
+sex             0.971     0.376     2.581     4.128     0.059    63.141     0.989
 
                             Estimate 
-Intercept~~Intercept|school    0.158 
-sex~~sex|school                0.212 
-Intercept~~sex|school         -0.026 
-Residual~~Residual             0.663 
-ICC|school                     0.188
+Intercept~~Intercept|school    0.259 
+Residual~~Residual             1.065 
+ICC|school                     0.197
 
@@
 R> fit <- with(imp, 
-+             lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
++             lmer(popular ~ 1 + sex + (1 | school))) 
 \end{CodeInput}
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.361     0.177    18.967    16.581     0.000     0.965     0.543 
-sex             1.250     0.243     5.145     9.696     0.000     1.796     0.699 
-texp            0.105     0.014     7.425     8.751     0.000     2.087     0.731 
-sex:texp       -0.026     0.019    -1.388     6.889     0.208     3.202     0.810
+(Intercept)     5.121     0.390    13.125     4.242     0.000    33.503     0.979 
+sex             0.673     0.418     1.611     4.121     0.180    66.343     0.989
 
                             Estimate 
-Intercept~~Intercept|school    0.216 
-sex~~sex|school                0.240 
-Intercept~~sex|school         -0.045 
-Residual~~Residual             0.615 
-ICC|school                     0.259
+Intercept~~Intercept|school    0.317 
+Residual~~Residual             1.253 
+ICC|school                     0.213
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     3.333     0.177    18.832    16.459     0.000     0.972     0.545 
-sex             1.356     0.193     7.014    37.019     0.000     0.490     0.362 
-texp            0.106     0.012     8.870    13.221     0.000     1.222     0.606 
-sex:texp       -0.034     0.012    -2.901    51.319     0.005     0.387     0.306
+(Intercept)     4.941     0.422    11.708     4.203     0.000    39.844     0.982 
+sex             1.029     0.537     1.916     4.076     0.127   105.671     0.993
 
                             Estimate 
-Intercept~~Intercept|school    0.213 
-sex~~sex|school                0.307 
-Intercept~~sex|school         -0.084 
-Residual~~Residual             0.622 
-ICC|school                     0.254
+Intercept~~Intercept|school    0.306 
+Residual~~Residual             1.308 
+ICC|school                     0.195
 

### 2022-10-07 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update to work on HIV example together

@@
 \section{Introduction}\label{introduction}}
+
+TODO: send complete manuscript by september 30th!
 
@@
 
+TODO: explicit statement about not going into workings of the methods.
+Galimer 2l methods.
+
 \hypertarget{setup}{%
@@
 R> library(mice)         # for imputation
+R> library(miceadds)     # for imputation
 R> library(ggmice)       # for visualization
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.820     0.353    13.648     4.245     0.000    33.200     0.979 
-sex             0.971     0.376     2.581     4.128     0.059    63.141     0.989
+(Intercept)     5.010     0.316    15.841     4.324     0.000    25.159     0.972 
+sex             0.748     0.413     1.812     4.112     0.142    72.195     0.990
 
                             Estimate 
-Intercept~~Intercept|school    0.259 
-Residual~~Residual             1.065 
-ICC|school                     0.197
+Intercept~~Intercept|school    0.271 
+Residual~~Residual             1.125 
+ICC|school                     0.213
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.121     0.390    13.125     4.242     0.000    33.503     0.979 
-sex             0.673     0.418     1.611     4.121     0.180    66.343     0.989
+(Intercept)     5.043     0.293    17.187     4.469     0.000    17.558     0.961 
+sex             0.799     0.329     2.428     4.206     0.069    39.298     0.982
 
                             Estimate 
-Intercept~~Intercept|school    0.317 
-Residual~~Residual             1.253 
-ICC|school                     0.213
+Intercept~~Intercept|school    0.335 
+Residual~~Residual             1.296 
+ICC|school                     0.208
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.941     0.422    11.708     4.203     0.000    39.844     0.982 
-sex             1.029     0.537     1.916     4.076     0.127   105.671     0.993
+(Intercept)     5.051     0.259    19.541     4.645     0.000    12.876     0.947 
+sex             0.820     0.235     3.484     4.426     0.021    19.256     0.964
 
                             Estimate 
-Intercept~~Intercept|school    0.306 
-Residual~~Residual             1.308 
-ICC|school                     0.195
+Intercept~~Intercept|school    0.350 
+Residual~~Residual             1.318 
+ICC|school                     0.217
 
@@
 
-\hypertarget{case-study-ii-hiv-data}{%
-\section{Case study II: HIV data}\label{case-study-ii-hiv-data}}
-
-{[}TODO: change the order of case study 2 and 3{]}
-
-Data are simulated and included in the \texttt{GJRM} package. We will
-use the following variables:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{region} Cluster variable,
-\item
-  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
-\item
-  \texttt{age} Age of the patient,
-\item
-  \texttt{marital} Marital status,
-\item
-  \texttt{condom} Condom use during last intercourse,
-\item
-  \texttt{smoke} Smoker (levels; inclusion restriction variable).
-\end{itemize}
-
-The imputation of these date is based on the toy example from
-\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
-Heckman Github repo}.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/hiv-1} \end{center}
-
-\end{CodeChunk}
-
-\hypertarget{case-study-iii-impact-data}{%
-\section{Case study III: IMPACT data}\label{case-study-iii-impact-data}}
+\hypertarget{case-study-ii-impact-data-syst-missingness-pred-matrix}{%
+\section{Case study II: IMPACT data (syst missingness, pred
+matrix)}\label{case-study-ii-impact-data-syst-missingness-pred-matrix}}
 
@@
 \item
-  \texttt{ct} Marshall Computerized Tomography classification,
+  \texttt{ct} Marshall Computerized Tomography classification, {[}TODO:
+  make this one var? also shows that you don't always need random
+  effects everywhere?{]}
 \item
@@
 
-\hypertarget{complete-case-analysis}{%
-\subsection{Complete case analysis}\label{complete-case-analysis}}
+\hypertarget{complete-case-analysis-todo-remove-this}{%
+\subsection{Complete case analysis {[}TODO: remove
+this?{]}}\label{complete-case-analysis-todo-remove-this}}
 
@@
 # A tibble: 11 x 9
-   effect  group term  estimate std.error statistic   p.value conf.low conf.high
+   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
    <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
- 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
- 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
+ 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
+ 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
  3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
- 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
- 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
- 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
- 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
- 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
+ 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
+ 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
+ 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
+ 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
+ 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
  9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
-10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
-11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
+10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.55e-19  1.92     2.76 
+11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
+# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
+#   4: conf.low, 5: conf.high
 \end{CodeOutput}
@@
 Create a methods vector and predictor matrix, and make sure
-\texttt{name} is not included as predictor:
+\texttt{name} is not included as predictor, but as clustering variable:
 
@@
 (e.g.~integer for clustering variable).{]}
+
+\hypertarget{case-study-iii-hiv-data}{%
+\section{Case study III: HIV data}\label{case-study-iii-hiv-data}}
+
+Data are simulated and included in the \texttt{GJRM} package. We will
+use the following variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{region} Cluster variable,
+\item
+  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{marital} Marital status,
+\item
+  \texttt{condom} Condom use during last intercourse,
+\item
+  \texttt{smoke} Smoker (levels; inclusion restriction variable).
+\end{itemize}
+
+The imputation of these date is based on the toy example from
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}.
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/hiv-1} \end{center}
+
+\end{CodeChunk}
+
+R=region+language+(1\textbar InterviewID) model with with interviewer as
+random effects, because the observations are not independent. Interviews
+are not allocated randomly. In theory we expect the
+inclusion-restriction variable to be randomly assigned, that's why we're
+adding region and language to compensate for non-random allocation.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> data("hiv",  package = "GJRM")
+R> # We select  5 predictor variables over 9 regions
+R> colnames(hiv)
+\end{CodeInput}
+\begin{CodeOutput}
+ [1] "hivconsent"      "hiv"             "age"             "education"      
+ [5] "wealth"          "region"          "marital"         "std"            
+ [9] "age1sex_cat"     "highhiv"         "partner"         "condom"         
+[13] "aidscare"        "knowsdiedofaids" "evertestedHIV"   "smoke"          
+[17] "religion"        "ethnicity"       "language"        "interviewerID"  
+[21] "sw"             
+\end{CodeOutput}
+\begin{CodeInput}
+R> hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
+R> # Study/group variable has to be recoded as integer
+R> hivdata$region<-as.integer(hivdata$region)
+R> 
+R> # Categorical variables have to be recoded as factor
+R> # to use the 2l.binary imputation method, it is required that the level names should not have any space between words
+R> 
+R> hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
+R> #the missing variable is stored as a factor in the dataset, otherwise the method will 
+R> #apply the imputation correction for a  missing continuous variable instead for 
+R> #a missing binary variable,which is in this case the binary response of the hiv test.
+R> 
+R> hivdata$marital <- as.factor(hivdata$marital)
+R> levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
+R> hivdata$condom <- as.factor(hivdata$condom)
+R> levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
+R> hivdata$highhiv <- as.factor(hivdata$highhiv)
+R> levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
+R> hivdata$interviewerID <- as.factor(hivdata$interviewerID)
+R> 
+R> hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
+R> interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
+R> 
+R> summary(hivdata)
+\end{CodeInput}
+\begin{CodeOutput}
+   hiv         hivconsent          age                     marital    
+ 0   :4457   Min.   :0.0000   Min.   :15.00   never_married    :2513  
+ 1   : 641   1st Qu.:1.0000   1st Qu.:20.00   currently_married:3582  
+ NA's:1318   Median :1.0000   Median :28.00   formerly_married : 321  
+             Mean   :0.7946   Mean   :30.12                           
+             3rd Qu.:1.0000   3rd Qu.:37.00                           
+             Max.   :1.0000   Max.   :59.00
+
+                        condom                     highhiv     interviewerID 
+ No_Condom_Last_Intercourse:5319   Not_High_Risk_of_HIV:4655   404    : 306  
+ Condom_Last_Intercourse   :1097   High_Risk_of_HIV    :1761   303    : 248  
+                                                               807    : 247  
+                                                               403    : 246  
+                                                               903    : 241  
+                                                               305    : 232  
+                                                               (Other):4896  
+     region                   language   
+ Min.   :1.000   Language: bemba  :2347  
+ 1st Qu.:3.000   Language: nyanja :1927  
+ Median :5.000   Language: other  : 606  
+ Mean   :4.946   Language: lozi   : 604  
+ 3rd Qu.:7.000   Language: tonga  : 476  
+ Max.   :9.000   Language: english: 456  
+                 (Other)          :   0  
+\end{CodeOutput}
+\begin{CodeInput}
+R> #  We obtain here the random effects for each interviewer, this is an approximation 
+R> #of the interviewer's skill which will be used as an exclusion constraint. 
+R> #Here, since the location of the interviewer was not randomly assigned to the 
+R> #subjects, the assignment was corrected for region and language. 
+R> 
+R> hivdata$hivconsent <- as.factor(hivdata$hivconsent)
+R> ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
+R> reffect  <- ranef(ID_mixed)$interviewerID
+R> reffect$interviewerID <- levels(hivdata$interviewerID)
+R> colnames(reffect)     <- c("IDreffect","interviewerID")
+R> hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)
+R> 
+R> hivdata$interviewerID<-NULL
+R> hivdata$hivconsent<-NULL
+R> hivdata$language<-NULL
+R> summary(hivdata)
+\end{CodeInput}
+\begin{CodeOutput}
+   hiv            age                     marital    
+ 0   :4457   Min.   :15.00   never_married    :2513  
+ 1   : 641   1st Qu.:20.00   currently_married:3582  
+ NA's:1318   Median :28.00   formerly_married : 321  
+             Mean   :30.12                           
+             3rd Qu.:37.00                           
+             Max.   :59.00                           
+                        condom                     highhiv         region     
+ No_Condom_Last_Intercourse:5319   Not_High_Risk_of_HIV:4655   Min.   :1.000  
+ Condom_Last_Intercourse   :1097   High_Risk_of_HIV    :1761   1st Qu.:3.000  
+                                                               Median :5.000  
+                                                               Mean   :4.946  
+                                                               3rd Qu.:7.000  
+                                                               Max.   :9.000  
+   IDreffect        
+ Min.   :-0.682850  
+ 1st Qu.:-0.195361  
+ Median :-0.080531  
+ Mean   :-0.007813  
+ 3rd Qu.: 0.257165  
+ Max.   : 0.667147  
+\end{CodeOutput}
+\begin{CodeInput}
+R> #Set prediction matrix and methods
+R> ini <- mice(hivdata, maxit = 0)
+R> meth<-ini$method
+R> meth["hiv"]<-"2l.binary"
+R> pred <- ini$pred
+R> pred[,"region"] <- 0
+R> pred["region",] <- 0
+R> pred["hiv","region"]<- -2
+R> pred["hiv","IDreffect"]<- 0
+R> 
+R> # MAR imputation model, we used the 2l.binary from the miceadds package
+R> data_imp_mar <- mice( data= hivdata, # dataset with missing values
++                       m = 10,   # number of imputations
++                       meth = meth, #imputation method vector
++                       pred = pred, #imputation predictors matrix
++                       maxit=1)
+\end{CodeInput}
+\begin{CodeOutput}
+
+ iter imp variable
+  1   1  hiv
+  1   2  hiv
+  1   3  hiv
+  1   4  hiv
+  1   5  hiv
+  1   6  hiv
+  1   7  hiv
+  1   8  hiv
+  1   9  hiv
+  1   10  hiv
+\end{CodeOutput}
+\begin{CodeInput}
+R> # source("mice.impute.2l.heckman.R")
+R> # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["1"]]))
+R> # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["1"]]))
+R> # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["2"]]))
+R> # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["2"]]))
+R> 
+R> # Heckman model
+R> pred["hiv","IDreffect"] <- -3
+R> meth<-ini$method
+R> meth["hiv"]<-"2l.heckman"
+R> 
+R> data_imp_heckp <- mice( hivdata, # dataset with missing values
++                         m = 10,   # number of imputations
++                         seed = 1234, #seed attached to the dataID
++                         meth = meth, #imputation method vector
++                         pred = pred, #imputation predictors matrix
++                         print = T,
++                         meta_method="reml",
++                         pmm=FALSE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+ iter imp variable
+  1   1  hiv
+  1   2  hiv
+  1   3  hiv
+  1   4  hiv
+  1   5  hiv
+  1   6  hiv
+  1   7  hiv
+  1   8  hiv
+  1   9  hiv
+  1   10  hiv
+  2   1  hiv
+  2   2  hiv
+  2   3  hiv
+  2   4  hiv
+  2   5  hiv
+  2   6  hiv
+  2   7  hiv
+  2   8  hiv
+  2   9  hiv
+  2   10  hiv
+  3   1  hiv
+  3   2  hiv
+  3   3  hiv
+  3   4  hiv
+  3   5  hiv
+  3   6  hiv
+  3   7  hiv
+  3   8  hiv
+  3   9  hiv
+  3   10  hiv
+  4   1  hiv
+  4   2  hiv
+  4   3  hiv
+  4   4  hiv
+  4   5  hiv
+  4   6  hiv
+  4   7  hiv
+  4   8  hiv
+  4   9  hiv
+  4   10  hiv
+  5   1  hiv
+  5   2  hiv
+  5   3  hiv
+  5   4  hiv
+  5   5  hiv
+  5   6  hiv
+  5   7  hiv
+  5   8  hiv
+  5   9  hiv
+  5   10  hiv
+\end{CodeOutput}
+\end{CodeChunk}
 

### 2022-10-11 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

@@
 # Introduction
+
+TODO: send complete manuscript by september 30th!
 
@@
 
+TODO: explicit statement about not going into workings of the methods. Galimer 2l methods.
+
 ## Setup
@@
 library(mice)         # for imputation
+library(miceadds)     # for imputation
 library(ggmice)       # for visualization
@@
 
-# Case study II: HIV data
-
-[TODO: change the order of case study 2 and 3]
-
-Data are simulated and included in the `GJRM` package. We will use the following variables:
-
-  - `region` Cluster variable,
-- `hiv` HIV diagnosis (0=no, 1=yes),
-- `age` Age of the patient,
-- `marital` Marital status,
-- `condom` Condom use during last intercourse,
-- `smoke` Smoker (levels; inclusion restriction variable).
-
-The imputation of these date is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
-
-```{r hiv, echo=FALSE}
-# data
-data("hiv", package = "GJRM")
-hiv <- hiv %>% 
-  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
-  mutate(region = as.integer(region))
-# md pattern
-plot_pattern(hiv)
-```
-
-# Case study III: IMPACT data
+# Case study II: IMPACT data (syst missingness, pred matrix)
 
@@
   - `pupil` Pupillary reactivity,
-  - `ct` Marshall Computerized Tomography classification,
+  - `ct` Marshall Computerized Tomography classification, [TODO: make this one var? also shows that you don't always need random effects everywhere?]
   - `hypox` Hypoxia (0=no, 1=yes),
@@
 
-## Complete case analysis
+## Complete case analysis [TODO: remove this?]
 
@@
 
-Create a methods vector and predictor matrix, and make sure `name` is not included as predictor:
+Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
 ```{r ignore}
@@
 ```{r imp_impact, eval = FALSE, include=FALSE}
+dat <- dat %>% mutate(across(everything(), as.integer))
 imp <- mice(dat, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
@@
 ```{r impact-fit, eval = FALSE, include=FALSE}
-fit <- imp_impact_pmm %>% 
+fit <- imp %>% 
   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
@@
 # testEstimates(as.mitml.result(fit))
+```
+
+# Case study III: HIV data
+
+Data are simulated and included in the `GJRM` package. We will use the following variables:
+
+- `region` Cluster variable,
+- `hiv` HIV diagnosis (0=no, 1=yes),
+- `age` Age of the patient,
+- `marital` Marital status,
+- `condom` Condom use during last intercourse,
+- `smoke` Smoker (levels; inclusion restriction variable).
+
+The imputation of these date is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
+
+```{r hiv, echo=FALSE}
+# data
+data("hiv", package = "GJRM")
+hiv <- hiv %>% 
+  select(c("region","hiv", "age", "marital", "condom", "smoke", "interviewerID")) %>% 
+  mutate(region = as.integer(region))
+# md pattern
+plot_pattern(hiv)
+```
+
+R=region+language+(1|InterviewID) model with with interviewer as random effects, because the observations are not independent.
+Interviews are not allocated randomly.
+In theory we expect the inclusion-restriction variable to be randomly assigned, that's why we're adding region and language to compensate for non-random allocation.
+
+```{r eval=FALSE, include=FALSE}
+meth <- make.method(hiv)
+meth["hiv"] <- "2l.heckman"
+imp <- mice(hiv, method = meth)
+pred <- make.predictorMatrix(hiv)
+pred[, "region"] <- -2
+pred[, "smoke"] <- -3
+imp2 <- mice(hiv, method = meth, predictorMatrix = pred)
+```
+
+```{r}
+data("hiv",  package = "GJRM")
+# We select  5 predictor variables over 9 regions
+colnames(hiv)
+hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
+# Study/group variable has to be recoded as integer
+hivdata$region<-as.integer(hivdata$region)
+
+# Categorical variables have to be recoded as factor
+# to use the 2l.binary imputation method, it is required that the level names should not have any space between words
+
+hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
+#the missing variable is stored as a factor in the dataset, otherwise the method will 
+#apply the imputation correction for a  missing continuous variable instead for 
+#a missing binary variable,which is in this case the binary response of the hiv test.
+
+hivdata$marital <- as.factor(hivdata$marital)
+levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
+hivdata$condom <- as.factor(hivdata$condom)
+levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
+hivdata$highhiv <- as.factor(hivdata$highhiv)
+levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
+hivdata$interviewerID <- as.factor(hivdata$interviewerID)
+
+hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
+interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
+
+summary(hivdata)
+#  We obtain here the random effects for each interviewer, this is an approximation 
+#of the interviewer's skill which will be used as an exclusion constraint. 
+#Here, since the location of the interviewer was not randomly assigned to the 
+#subjects, the assignment was corrected for region and language.
+
+hivdata$hivconsent <- as.factor(hivdata$hivconsent)
+ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
+reffect  <- ranef(ID_mixed)$interviewerID
+reffect$interviewerID <- levels(hivdata$interviewerID)
+colnames(reffect)     <- c("IDreffect","interviewerID")
+hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)
+
+hivdata$interviewerID<-NULL
+hivdata$hivconsent<-NULL
+hivdata$language<-NULL
+summary(hivdata)
+
+# #Set prediction matrix and methods
+# ini <- mice(hivdata, maxit = 0)
+# meth<-ini$method
+# meth["hiv"]<-"2l.binary"
+# pred <- ini$pred
+# pred[,"region"] <- 0
+# pred["region",] <- 0
+# pred["hiv","region"]<- -2
+# pred["hiv","IDreffect"]<- 0
+# 
+# # MAR imputation model, we used the 2l.binary from the miceadds package
+# data_imp_mar <- mice( data= hivdata, # dataset with missing values
+#                       m = 10,   # number of imputations
+#                       meth = meth, #imputation method vector
+#                       pred = pred, #imputation predictors matrix
+#                       maxit=1)
+# 
+# # source("mice.impute.2l.heckman.R")
+# # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["1"]]))
+# # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["1"]]))
+# # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["2"]]))
+# # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["2"]]))
+# 
+# # Heckman model
+# pred["hiv","IDreffect"] <- -3
+# meth<-ini$method
+# meth["hiv"]<-"2l.heckman"
+# 
+# data_imp_heckp <- mice( hivdata, # dataset with missing values
+#                         m = 10,   # number of imputations
+#                         seed = 1234, #seed attached to the dataID
+#                         meth = meth, #imputation method vector
+#                         pred = pred, #imputation predictors matrix
+#                         print = T,
+#                         meta_method="reml",
+#                         pmm=FALSE)
+
+#Set prediction matrix and methods
+ini <- mice(hivdata, maxit = 0)
+meth<-ini$method
+meth["hiv"]<-"2l.binary"
+pred <- ini$pred
+pred[,"region"] <- 0
+pred["region",] <- 0
+pred["hiv","region"]<- -2
+pred["hiv","IDreffect"]<- 0
+
+# MAR imputation model, we used the 2l.binary from the miceadds package
+data_imp_mar <- mice( data= hivdata, # dataset with missing values
+                      m = 20,   # number of imputations
+                      meth = meth, #imputation method vector
+                      pred = pred, #imputation predictors matrix
+                      maxit=1)
+
+source("mice.impute.2l.heckman.R")
+
+# Heckman model
+pred["hiv","IDreffect"] <- -3
+meth<-ini$method
+meth["hiv"]<-"2l.heckman"
+
+data_imp_heckp <- mice( hivdata, # dataset with missing values
+                        m = 20,   # number of imputations
+                        seed = 1234, #seed attached to the dataID
+                        meth = meth, #imputation method vector
+                        pred = pred, #imputation predictors matrix
+                        print = T,
+                        maxit=1,
+                        meta_method="reml",
+                        pmm=FALSE)
 ```

### 2022-10-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

@@
 The last section is not part of the manuscript, but purely for reminders.
+See also all of the TODOs that need to be worked out.
 We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
@@
 # Introduction
-
-TODO: send complete manuscript by september 30th!
 
@@
 
-The research objective of the popmis dataset is to predict the pupils' popularity based on their gender and the experience of the teacher.
+The research objective of the `popmis` dataset is to predict the pupils' popularity based on their gender and the experience of the teacher.
 The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction.
@@
 <!-- --- -->
+Mutate data to get the right data types for imputation (e.g. integer for clustering variable).
+
+```{r}
+dat <- dat %>% mutate(across(everything(), as.integer))
+```
 
 Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
-```{r ignore}
+```{r impact}
 meth <- make.method(dat) # methods vector
@@
 plot_pred(pred)
+
+pred[pred == 1] <- 2
+pred["mort", ] <- 2
+pred[, "mort"] <- 2
+pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
+pred[, "name"] <- -2
+diag(pred) <- 0
+plot_pred(pred)
+
+meth <- make.method(dat)
+meth
 ```
@@
 
-```{r impact-pred, include=FALSE, eval=FALSE}
-pred <- quickpred(dat)
-plot_pred(pred)
-
-pred[pred == 1] <- 2
-pred["mort", ] <- 2
-pred[, "mort"] <- 2
-pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
-pred[, "name"] <- -2
-diag(pred) <- 0
-plot_pred(pred)
-
-meth <- make.method(dat)
-meth[meth != ""] <- "2l.pmm"
+Impute the incomplete data
+```{r imp_impact}
+imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
 ```
 
-[TODO: mutate data to get the right data types for imputation (e.g. integer for clustering variable).]
-
-```{r imp_impact, eval = FALSE, include=FALSE}
-dat <- dat %>% mutate(across(everything(), as.integer))
-imp <- mice(dat, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
-```
-
-```{r impact-fit, eval = FALSE, include=FALSE}
+```{r impact-fit}
 fit <- imp %>% 
@@
 
-```{r eval=FALSE, include=FALSE}
-meth <- make.method(hiv)
-meth["hiv"] <- "2l.heckman"
-imp <- mice(hiv, method = meth)
-pred <- make.predictorMatrix(hiv)
-pred[, "region"] <- -2
-pred[, "smoke"] <- -3
-imp2 <- mice(hiv, method = meth, predictorMatrix = pred)
-```
-
+Load the data:
 ```{r}
@@
 hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
+```
+
+Recode variables:
+```{r}
 # Study/group variable has to be recoded as integer
@@
 interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
+```
 
-summary(hivdata)
-#  We obtain here the random effects for each interviewer, this is an approximation 
-#of the interviewer's skill which will be used as an exclusion constraint. 
-#Here, since the location of the interviewer was not randomly assigned to the 
-#subjects, the assignment was corrected for region and language.
-
+We obtain here the random effects for each interviewer, this is an approximation of the interviewer's skill which will be used as an exclusion constraint.
+Here, since the location of the interviewer was not randomly assigned to the subjects, the assignment was corrected for region and language.
+```{r}
 hivdata$hivconsent <- as.factor(hivdata$hivconsent)
@@
 hivdata$language<-NULL
-summary(hivdata)
+```
 
-# #Set prediction matrix and methods
-# ini <- mice(hivdata, maxit = 0)
-# meth<-ini$method
-# meth["hiv"]<-"2l.binary"
-# pred <- ini$pred
-# pred[,"region"] <- 0
-# pred["region",] <- 0
-# pred["hiv","region"]<- -2
-# pred["hiv","IDreffect"]<- 0
-# 
-# # MAR imputation model, we used the 2l.binary from the miceadds package
-# data_imp_mar <- mice( data= hivdata, # dataset with missing values
-#                       m = 10,   # number of imputations
-#                       meth = meth, #imputation method vector
-#                       pred = pred, #imputation predictors matrix
-#                       maxit=1)
-# 
-# # source("mice.impute.2l.heckman.R")
-# # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["1"]]))
-# # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["1"]]))
-# # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["2"]]))
-# # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["2"]]))
-# 
-# # Heckman model
-# pred["hiv","IDreffect"] <- -3
-# meth<-ini$method
-# meth["hiv"]<-"2l.heckman"
-# 
-# data_imp_heckp <- mice( hivdata, # dataset with missing values
-#                         m = 10,   # number of imputations
-#                         seed = 1234, #seed attached to the dataID
-#                         meth = meth, #imputation method vector
-#                         pred = pred, #imputation predictors matrix
-#                         print = T,
-#                         meta_method="reml",
-#                         pmm=FALSE)
-
+Set the Heckman model as imputation method
+```{r}
 #Set prediction matrix and methods
@@
 
-# MAR imputation model, we used the 2l.binary from the miceadds package
-data_imp_mar <- mice( data= hivdata, # dataset with missing values
-                      m = 20,   # number of imputations
-                      meth = meth, #imputation method vector
-                      pred = pred, #imputation predictors matrix
-                      maxit=1)
-
-source("mice.impute.2l.heckman.R")
-
 # Heckman model
@@
 meth["hiv"]<-"2l.heckman"
+```
 
-data_imp_heckp <- mice( hivdata, # dataset with missing values
-                        m = 20,   # number of imputations
+Impute the data:
+```{r}
+imp <- mice( hivdata, # dataset with missing values
+                        m = 10,   # number of imputations
+                        maxit = 1,
                         seed = 1234, #seed attached to the dataID
@@
                         print = T,
-                        maxit=1,
                         meta_method="reml",
@@
 
+- Adding evaluations of the imputations such as convergence checks
+
 - Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model.

### 2022-10-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

--- 2022-06-23 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 version show work in progress/under construction. The last section is
-not part of the manuscript, but purely for reminders. We aim to submit
-at JSS, so there is no word count limit (``There is no page limit, nor a
-limit on the number of figures or tables''). {[}Just adding some text to
-get a better guess of what the actura abstract will look like: Lorem
-ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor
-incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam,
-quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo
-consequat. Duis aute irure dolor in reprehenderit in voluptate velit
-esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat
-cupidatat non proident, sunt in culpa qui officia deserunt mollit anim
-id est laborum.{]}
+not part of the manuscript, but purely for reminders. See also all of
+the TODOs that need to be worked out. We aim to submit at JSS, so there
+is no word count limit (``There is no page limit, nor a limit on the
+number of figures or tables''). {[}Just adding some text to get a better
+guess of what the actura abstract will look like: Lorem ipsum dolor sit
+amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut
+labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud
+exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum
+dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non
+proident, sunt in culpa qui officia deserunt mollit anim id est
+laborum.{]}
 }
@@
 
-The research objective of the popmis dataset is to predict the pupils'
-popularity based on their gender and the experience of the teacher. The
-analysis model corresponding to this dataset is multilevel regression
-with random intercepts, random slopes and a cross-level interaction. The
-outcome variable is \texttt{popular}, which is predicted from the
-unit-level variable \texttt{sex} and the cluster-level variable
-\texttt{texp}:
+The research objective of the \texttt{popmis} dataset is to predict the
+pupils' popularity based on their gender and the experience of the
+teacher. The analysis model corresponding to this dataset is multilevel
+regression with random intercepts, random slopes and a cross-level
+interaction. The outcome variable is \texttt{popular}, which is
+predicted from the unit-level variable \texttt{sex} and the
+cluster-level variable \texttt{texp}:
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.820     0.353    13.648     4.245     0.000    33.200     0.979 
-sex             0.971     0.376     2.581     4.128     0.059    63.141     0.989
+(Intercept)     4.932     0.274    18.012     4.420     0.000    19.514     0.964 
+sex             0.859     0.284     3.026     4.216     0.036    37.601     0.981
 
                             Estimate 
-Intercept~~Intercept|school    0.259 
-Residual~~Residual             1.065 
-ICC|school                     0.197
+Intercept~~Intercept|school    0.265 
+Residual~~Residual             1.007 
+ICC|school                     0.213
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.121     0.390    13.125     4.242     0.000    33.503     0.979 
-sex             0.673     0.418     1.611     4.121     0.180    66.343     0.989
+(Intercept)     5.054     0.266    19.029     4.554     0.000    14.936     0.954 
+sex             0.732     0.306     2.393     4.233     0.071    34.886     0.980
 
                             Estimate 
-Intercept~~Intercept|school    0.317 
-Residual~~Residual             1.253 
-ICC|school                     0.213
+Intercept~~Intercept|school    0.318 
+Residual~~Residual             1.257 
+ICC|school                     0.208
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.941     0.422    11.708     4.203     0.000    39.844     0.982 
-sex             1.029     0.537     1.916     4.076     0.127   105.671     0.993
+(Intercept)     4.714     0.193    24.432     5.032     0.000     8.224     0.919 
+sex             1.286     0.288     4.460     4.218     0.010    37.215     0.981
 
                             Estimate 
-Intercept~~Intercept|school    0.306 
-Residual~~Residual             1.308 
-ICC|school                     0.195
+Intercept~~Intercept|school    0.299 
+Residual~~Residual             1.050 
+ICC|school                     0.228
 
@@
 
+Mutate data to get the right data types for imputation (e.g.~integer for
+clustering variable).
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> dat <- dat %>% mutate(across(everything(), as.integer))
+\end{CodeInput}
+\end{CodeChunk}
+
 Create a methods vector and predictor matrix, and make sure
-\texttt{name} is not included as predictor:
+\texttt{name} is not included as predictor, but as clustering variable:
 
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/ignore-1} \end{center}
-
-\end{CodeChunk}
-
-{[}TODO: mutate data to get the right data types for imputation
-(e.g.~integer for clustering variable).{]}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-1} \end{center}
+
+\begin{CodeInput}
+R> pred[pred == 1] <- 2
+R> pred["mort", ] <- 2
+R> pred[, "mort"] <- 2
+R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
+R> pred[, "name"] <- -2
+R> diag(pred) <- 0
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-2} \end{center}
+
+\begin{CodeInput}
+R> meth <- make.method(dat)
+R> meth
+\end{CodeInput}
+\begin{CodeOutput}
+       name        type         age motor_score       pupil          ct 
+         ""          ""          ""          ""       "pmm"       "pmm" 
+     hypots        tsah         edh        mort 
+      "pmm"       "pmm"       "pmm"          "" 
+\end{CodeOutput}
+\end{CodeChunk}
+
+Impute the incomplete data
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- imp %>% 
++   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
+R> tidy(pool(fit))
+\end{CodeInput}
+\begin{CodeOutput}
+                     term    estimate   std.error  statistic      p.value
+1             (Intercept) -2.33813790 0.340459131  -6.867602 6.978862e-12
+2                    type -0.41509929 0.179936773  -2.306918 2.107792e-02
+3                     age  0.03034813 0.001561834  19.431086 0.000000e+00
+4 as.factor(motor_score)2 -0.66052409 0.068846376  -9.594174 0.000000e+00
+5 as.factor(motor_score)3 -1.05030912 0.070059114 -14.991756 0.000000e+00
+6 as.factor(motor_score)4 -1.49637069 0.071492961 -20.930322 0.000000e+00
+7                   pupil  0.48922641 0.038536483  12.695149 0.000000e+00
+8                      ct  0.42634415 0.028388807  15.018037 0.000000e+00
+             b          df dfcom          fmi       lambda m          riv
+1 9.395965e-04  8668.10341 11013 0.0099557149 0.0097273074 5 0.0098228573
+2 1.729929e-05 10991.51028 11013 0.0008229571 0.0006411648 5 0.0006415761
+3 8.332782e-09 10482.95010 11013 0.0042891736 0.0040992241 5 0.0041160969
+4 5.649577e-05  6979.24803 11013 0.0145856016 0.0143032580 5 0.0145108098
+5 3.003051e-05  9526.87220 11013 0.0075503299 0.0073420043 5 0.0073963080
+6 3.329522e-05  9362.42789 11013 0.0080288174 0.0078169354 5 0.0078785213
+7 3.337070e-04    54.63795 11013 0.2949938851 0.2696512493 5 0.3692088868
+8 1.180517e-05  5893.23780 11013 0.0179108151 0.0175775781 5 0.0178920775
+          ubar
+1 1.147849e-01
+2 3.235648e-02
+3 2.429325e-06
+4 4.672029e-03
+5 4.872243e-03
+6 5.071289e-03
+7 1.084612e-03
+8 7.917581e-04
+\end{CodeOutput}
+\begin{CodeInput}
+R> as.mitml.result(fit)
+\end{CodeInput}
+\begin{CodeOutput}
+[[1]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+      AIC       BIC    logLik  deviance  df.resid 
+10522.944 10588.713 -5252.472 10504.944     11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2913  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.32713                 -0.41953                  0.03031  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.65651                 -1.04719                 -1.49327  
+                  pupil                       ct  
+                0.48990                  0.42335  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+[[2]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+      AIC       BIC    logLik  deviance  df.resid 
+10525.723 10591.492 -5253.861 10507.723     11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2937  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.30150                 -0.41940                  0.03024  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.65497                 -1.04451                 -1.49454  
+                  pupil                       ct  
+                0.47281                  0.42888  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+[[3]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+     AIC      BIC   logLik deviance df.resid 
+10512.28 10578.05 -5247.14 10494.28    11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2852  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.35320                 -0.41028                  0.03034  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.66933                 -1.05474                 -1.50379  
+                  pupil                       ct  
+                0.49491                  0.42442  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+[[4]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+     AIC      BIC   logLik deviance df.resid 
+10530.02 10595.79 -5256.01 10512.02    11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2923  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.32670                 -0.41286                  0.03037  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.66801                 -1.04770                 -1.50070  
+                  pupil                       ct  
+                0.47210                  0.43109  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+[[5]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+      AIC       BIC    logLik  deviance  df.resid 
+10488.615 10554.383 -5235.307 10470.615     11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2939  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.38215                 -0.41343                  0.03049  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.65379                 -1.05740                 -1.48956  
+                  pupil                       ct  
+                0.51641                  0.42397  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+attr(,"class")
+[1] "mitml.result" "list"        
+\end{CodeOutput}
+\begin{CodeInput}
+R> # testEstimates(as.mitml.result(fit))
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{case-study-iii-hiv-data}{%
+\section{Case study III: HIV data}\label{case-study-iii-hiv-data}}
+
+Data are simulated and included in the \texttt{GJRM} package. We will
+use the following variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{region} Cluster variable,
+\item
+  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{marital} Marital status,
+\item
+  \texttt{condom} Condom use during last intercourse,
+\item
+  \texttt{smoke} Smoker (levels; inclusion restriction variable).
+\end{itemize}
+
+The imputation of these date is based on the toy example from
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}.
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/hiv-1} \end{center}
+
+\end{CodeChunk}
+
+R=region+language+(1\textbar InterviewID) model with with interviewer as
+random effects, because the observations are not independent. Interviews
+are not allocated randomly. In theory we expect the
+inclusion-restriction variable to be randomly assigned, that's why we're
+adding region and language to compensate for non-random allocation.
+
+Load the data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> data("hiv",  package = "GJRM")
+R> # We select  5 predictor variables over 9 regions
+R> colnames(hiv)
+\end{CodeInput}
+\begin{CodeOutput}
+ [1] "hivconsent"      "hiv"             "age"             "education"      
+ [5] "wealth"          "region"          "marital"         "std"            
+ [9] "age1sex_cat"     "highhiv"         "partner"         "condom"         
+[13] "aidscare"        "knowsdiedofaids" "evertestedHIV"   "smoke"          
+[17] "religion"        "ethnicity"       "language"        "interviewerID"  
+[21] "sw"             
+\end{CodeOutput}
+\begin{CodeInput}
+R> hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
+\end{CodeInput}
+\end{CodeChunk}
+
+Recode variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # Study/group variable has to be recoded as integer
+R> hivdata$region<-as.integer(hivdata$region)
+R> 
+R> # Categorical variables have to be recoded as factor
+R> # to use the 2l.binary imputation method, it is required that the level names should not have any space between words
+R> 
+R> hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
+R> #the missing variable is stored as a factor in the dataset, otherwise the method will 
+R> #apply the imputation correction for a  missing continuous variable instead for 
+R> #a missing binary variable,which is in this case the binary response of the hiv test.
+R> 
+R> hivdata$marital <- as.factor(hivdata$marital)
+R> levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
+R> hivdata$condom <- as.factor(hivdata$condom)
+R> levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
+R> hivdata$highhiv <- as.factor(hivdata$highhiv)
+R> levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
+R> hivdata$interviewerID <- as.factor(hivdata$interviewerID)
+R> 
+R> hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
+R> interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
+\end{CodeInput}
+\end{CodeChunk}
+
+We obtain here the random effects for each interviewer, this is an
+approximation of the interviewer's skill which will be used as an
+exclusion constraint. Here, since the location of the interviewer was
+not randomly assigned to the subjects, the assignment was corrected for
+region and language.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> hivdata$hivconsent <- as.factor(hivdata$hivconsent)
+R> ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
+R> reffect  <- ranef(ID_mixed)$interviewerID
+R> reffect$interviewerID <- levels(hivdata$interviewerID)
+R> colnames(reffect)     <- c("IDreffect","interviewerID")
+R> hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)
+R> 
+R> hivdata$interviewerID<-NULL
+R> hivdata$hivconsent<-NULL
+R> hivdata$language<-NULL
+\end{CodeInput}
+\end{CodeChunk}
+
+Set the Heckman model as imputation method
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> #Set prediction matrix and methods
+R> ini <- mice(hivdata, maxit = 0)
+R> meth<-ini$method
+R> meth["hiv"]<-"2l.binary"
+R> pred <- ini$pred
+R> pred[,"region"] <- 0
+R> pred["region",] <- 0
+R> pred["hiv","region"]<- -2
+R> pred["hiv","IDreffect"]<- 0
+R> 
+R> # Heckman model
+R> pred["hiv","IDreffect"] <- -3
+R> meth<-ini$method
+R> meth["hiv"]<-"2l.heckman"
+\end{CodeInput}
+\end{CodeChunk}
+
+Impute the data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp <- mice( hivdata, # dataset with missing values
++                         m = 10,   # number of imputations
++                         maxit = 1,
++                         seed = 1234, #seed attached to the dataID
++                         meth = meth, #imputation method vector
++                         pred = pred, #imputation predictors matrix
++                         print = T,
++                         meta_method="reml",
++                         pmm=FALSE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+ iter imp variable
+  1   1  hiv
+  1   2  hiv
+  1   3  hiv
+  1   4  hiv
+  1   5  hiv
+  1   6  hiv
+  1   7  hiv
+  1   8  hiv
+  1   9  hiv
+  1   10  hiv
+\end{CodeOutput}
+\end{CodeChunk}
 

### 2022-11-17 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Little comments

@@
 
+<!-- Maybe besides CC and imputation name also likelihood based methods -->
 Like any other dataset, clustered datasets are prone to missing data.
@@
 Plot the pair-wise complete correlations in the incomplete data:
+<!-- Excelent, but i dont know if you can refer to the outflux criteria here or elsewhere, cuz in practice it is used to select the predictors in the imputation model from dataset with a lot of features-->
 
@@
 [TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.]
+
+<!-- A fixed cluster model is better if we have data with all the clusters in the population and we are interested in knowing the effect of a treatment on the specific cluster eg.
+Cluster No.123.
+If the population is too large (10000 schools) and we have a sample (200 schools) then an random cluster is better and saves us degrees of freedom because some of the parameters are random variables.
+Here we have to say that the imputation model should be more in line with the analysis model (that defines in part how the cluster is included) and consider how the estimation of the imputation model could be affected when N is too large (as we add N dummies in the fixed cluster model) or if ni is too small in some clusters, so the model can not be estimated in them..
+-->
 

### 2022-11-18 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update Imputation_of_Incomplete_Multilevel_Data.Rmd

@@
 
+-  I don't know if in your article you cover something about model complexity, for example sometimes I have to switch from 2l.
+methods to 1l.
+methods just because the model didn't converge..
+this is due to the considered imputation model is very complex regarding the amount of information counted...
+I know that a solution for an imputation model with many predictors is to check correlation plots as you did or use quickpred()..
+(maybe you can add this somewhere after the correlation plots).....But as for cluster specification, I don't know if besides plots of distribution per cluster there is something else can be done to see if i have to use 1l.
+or 2l.
+for a given variable, also I have no idea..
+how to test which is better between 2l.norm or 2l.2stage.norm.
+
+Translated with www.DeepL.com/Translator (free version)
+
 # References

### 2022-12-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Comments Valentijn

--- 2022-10-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 author:
-  - name: Hanne Oberman
+  - name: Hanne I. Oberman
     affiliation: Utrecht University
@@
   - name: "Johanna Munoz Avila"
-    affiliation: University Medical Center Utrecht \AND
-  - name: Valentijn de Jong
-    affiliation: University Medical Center Utrecht 
+    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
+  - name: Thomas P. A. Debray
+    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
   - name: Gerko Vink
-    affiliation: Utrecht University \AND
-  - name: Thomas Debray
-    affiliation: University Medical Center Utrecht 
+    affiliation: Utrecht University 
+  - name: Valentijn M. T. de Jong
+    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands \AND Data Analytics and Methods Task Force, European Medicines Agency, Amsterdam, The Netherlands
+
 title:
@@
 If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01].
-[TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation.] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
-These models are also know as 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
+[TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation.
+VMTdJ: "I would only make that link after introducing missing data, and missing data is not mentioned yet, so I would do that in the next section"] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
+These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'.
 Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.
@@
 
-Imputation of missing data requires to consider the mechanism behind the missingness.
+Imputation of missing data requires consideration of the mechanism behind the missingness.
 Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}).
 For each of these three missingness generating mechanisms, different imputation strategies are warranted @yuce08 and @hox15.
-We here consider the general case that data are MAR, and expand on special MNAR situations.
+We here consider the general case that data are MAR, and expand on certain MNAR situations.
 
@@
 \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
-However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent.
+However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent whereas multilevel data are dependent.
 For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.
@@
 For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial.
-Assumed knowledge includes basic familiarity with multilevel imputation [see e.g.
-@audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).
+[TODO: change this sentence because we do not assume familiarity: "Assumed knowledge includes basic familiarity with multilevel imputation [see e.g.
+@audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod})"].
 
@@
 
-This tutorial is dedicated to readers who are unfamiliar with multiple imptuation.
+This tutorial is dedicated to readers who are unfamiliar with multiple imputation.
 More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study IMPACT) or  under MNAR conditions (case study HIV).

### 2022-12-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
update pdf
some code changes to be able to run

@@
         miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
+
 ggplot(dat, aes(x = cols, y = rows)) +
@@
   scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
-  coord_cartesian(expand = c(0,0,0,0)) +
+  coord_cartesian(expand = c(0)) + #Changed to run GV
+  # coord_cartesian(expand = c(0,0,0,0)) +
   theme_minimal() +
@@
 Impute the data:
-```{r}
+```{r eval = FALSE}
 imp <- mice( hivdata, # dataset with missing values

### 2022-12-07 Data/obesity.R
Add obesity example data for Heckman method

@@ -1,0 +1,93 @@
+library(data.table)
+library(truncnorm)
+library(nlme)
+library(devtools)
+library(dplyr)
+library(ggplot2)
+library(mice)
+
+#0. Load imputation method and original dataset ----
+# Source imputation function
+source_url("https://github.com/johamunoz/Statsmed_Heckman/blob/main/4.Codes/mice.impute.2l.heckman.R?raw=TRUE")
+# Dataset obtained from "https://www.kaggle.com/datasets/fabinmndez/obesitydata?select=ObesityDataSet_raw_and_data_sinthetic.csv"
+# and it is described here https://www.sciencedirect.com/science/article/pii/S2352340919306985?via%3Dihub
+
+# data<-as.data.table(read.csv2('./Data/obesity.csv', stringsAsFactors=TRUE, fileEncoding="latin1"))
+data <- as.data.table(read.csv('./Data/obesity_data.csv', stringsAsFactors=FALSE))
+#1. Modify dataset in order to get a variable weight with MNAR missing mechanism ----
+# Generate countries where survey was taken
+data<-data[,c("Weight","Gender","Age","Height","FAVC")] # Select some of the variables of the original dataset
+data <- data %>% mutate(across(c("Weight", "Age", "Height"), ~as.numeric(.x)))
+data$Cluster<-c(rep(1,500),rep(2,500),rep(3,500),rep(4,500),rep(5,111)) # in the imputation method is necessary cluster is included as numeric
+summary(data)
+#  Generation of the exclusion restriction (ERV): The time variable describes the time it takes for a person to answer the weight question.
+data$Time<-rtruncnorm(n=nrow(data), a = 1, b = 10, mean = 5, sd = 2) 
+data[,Gender:=ifelse(Gender=="Female",1,0)]
+data[,FAVC:=ifelse(FAVC=="yes",1,0)] 
+MAR_model<- lme(Weight~Gender+Age+Height+FAVC,random = ~1|Cluster, data = data) 
+# Generate selection and outcome variable (weight)
+data[,XOBO:=predict(MAR_model)]
+data[,XSBS:=15-0.7*Gender-0.06*Age-2*Time] # Simulated selection model
+
+# Simulate bivariate normal correlated errors
+rho=-0.6 # We assume that non-responders are more likely to have a high weight.
+d <- diag(2)
+d[1,1]<-MAR_model$sigma^2
+d[2,1] <- d[1,2] <- rho*MAR_model$sigma
[…]

### 2022-12-09 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

@@
 \author{
-Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
-Medical Center Utrecht \AND Valentijn de Jong\\University Medical
-Center Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
-Debray\\University Medical Center Utrecht
+Hanne I. Oberman\\Utrecht University \And Johanna Munoz Avila\\Julius
+Center for Health Sciences and Primary Care, University Medical Center
+Utrecht, Utrecht University, Utrecht, The Netherlands \And Thomas P. A.
+Debray\\Julius Center for Health Sciences and Primary Care, University
+Medical Center Utrecht, Utrecht University, Utrecht, The
+Netherlands \And Gerko Vink\\Utrecht University \And Valentijn M. T. de
+Jong\\Julius Center for Health Sciences and Primary Care, University
+Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands
+\AND Data Analytics and Methods Task Force, European Medicines Agency,
+Amsterdam, The Netherlands
 }
@@
 
-\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
-Jong, Gerko Vink, Thomas Debray}
+\Plainauthor{Hanne I. Oberman, Johanna Munoz Avila, Thomas P. A.
+Debray, Gerko Vink, Valentijn M. T. de Jong}
 \Plaintitle{Imputation of Incomplete Multilevel Data with mice}
@@
 \Address{
-    Hanne Oberman\\
+    Hanne I. Oberman\\
     Utrecht University\\
@@
 recommending the adoption of multilevel models for data analysis here,
-but rather for imputation.{]} Statistical methods for clustered data
-typically adopt hierarchical models that explicitly describe the
-grouping of observations. These models are also know as `multilevel
-models', `hierarchical models', `mixed effect models' and `random effect
-models'. Table \ref{tab:clus} provides an overview of some key concepts
-in multilevel modeling.
+but rather for imputation. VMTdJ: ``I would only make that link after
+introducing missing data, and missing data is not mentioned yet, so I
+would do that in the next section''{]} Statistical methods for clustered
+data typically adopt hierarchical models that explicitly describe the
+grouping of observations. These models are also known as `multilevel
+models', `hierarchical models', `mixed effect models', `random effect
+models', and in the context of time-to-event data as `frailty models'.
+Table \ref{tab:clus} provides an overview of some key concepts in
+multilevel modeling.
 
@@
 
-Imputation of missing data requires to consider the mechanism behind the
-missingness. Rubin proposed to distinguish between data that are missing
-completely at random (MCAR), data that are missing at random (MAR) and
-data that are missing not at random (MNAR; see Table \ref{tab:miss}).
-For each of these three missingness generating mechanisms, different
-imputation strategies are warranted \citet{yuce08} and \citet{hox15}. We
-here consider the general case that data are MAR, and expand on special
-MNAR situations.
+Imputation of missing data requires consideration of the mechanism
+behind the missingness. Rubin proposed to distinguish between data that
+are missing completely at random (MCAR), data that are missing at random
+(MAR) and data that are missing not at random (MNAR; see Table
+\ref{tab:miss}). For each of these three missingness generating
+mechanisms, different imputation strategies are warranted \citet{yuce08}
+and \citet{hox15}. We here consider the general case that data are MAR,
+and expand on certain MNAR situations.
 
@@
 designed for use in clustered data and usually generate observations
-that are independent. For this reason, we discuss how the \proglang{R}
-package \pkg{mice} can be used to impute multilevel data.
+that are independent whereas multilevel data are dependent. For this
+reason, we discuss how the \proglang{R} package \pkg{mice} can be used
+to impute multilevel data.
 
@@
 imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are
-outside the scope of this tutorial. Assumed knowledge includes basic
+outside the scope of this tutorial. {[}TODO: change this sentence
+because we do not assume familiarity: ``Assumed knowledge includes basic
 familiarity with multilevel imputation \citep[see e.g.][ and
 \citet{grun18}]{audi18} and the \pkg{lme4} notation for multilevel
-models (see Table \ref{tab:mod}).
+models (see Table \ref{tab:mod})''{]}.
 
@@
 This tutorial is dedicated to readers who are unfamiliar with multiple
-imptuation. More experienced readers can skip the introduction (case
+imputation. More experienced readers can skip the introduction (case
 study 1) and directly head to practical applications of multilevel
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.932     0.274    18.012     4.420     0.000    19.514     0.964 
-sex             0.859     0.284     3.026     4.216     0.036    37.601     0.981
+(Intercept)     4.986     0.194    25.704     4.917     0.000     9.201     0.927 
+sex             0.818     0.357     2.290     4.130     0.082    62.169     0.989
 
                             Estimate 
-Intercept~~Intercept|school    0.265 
-Residual~~Residual             1.007 
-ICC|school                     0.213
+Intercept~~Intercept|school    0.272 
+Residual~~Residual             0.975 
+ICC|school                     0.227
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.054     0.266    19.029     4.554     0.000    14.936     0.954 
-sex             0.732     0.306     2.393     4.233     0.071    34.886     0.980
+(Intercept)     5.186     0.254    20.457     4.706     0.000    11.816     0.942 
+sex             0.760     0.219     3.479     4.609     0.020    13.627     0.950
 
                             Estimate 
-Intercept~~Intercept|school    0.318 
-Residual~~Residual             1.257 
-ICC|school                     0.208
+Intercept~~Intercept|school    0.345 
+Residual~~Residual             1.578 
+ICC|school                     0.182
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.714     0.193    24.432     5.032     0.000     8.224     0.919 
-sex             1.286     0.288     4.460     4.218     0.010    37.215     0.981
+(Intercept)     5.000     0.356    14.027     4.283     0.000    28.723     0.976 
+sex             1.056     0.465     2.270     4.105     0.084    76.708     0.991
 
                             Estimate 
-Intercept~~Intercept|school    0.299 
-Residual~~Residual             1.050 
-ICC|school                     0.228
+Intercept~~Intercept|school    0.294 
+Residual~~Residual             1.347 
+ICC|school                     0.187
 
@@
    <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
- 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
+ 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   3.00e-31  0.0571   0.130
  2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
@@
  9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
-10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.55e-19  1.92     2.76 
+10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.56e-19  1.92     2.76 
 11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
@@
                      term    estimate   std.error  statistic      p.value
-1             (Intercept) -2.33813790 0.340459131  -6.867602 6.978862e-12
-2                    type -0.41509929 0.179936773  -2.306918 2.107792e-02
-3                     age  0.03034813 0.001561834  19.431086 0.000000e+00
-4 as.factor(motor_score)2 -0.66052409 0.068846376  -9.594174 0.000000e+00
-5 as.factor(motor_score)3 -1.05030912 0.070059114 -14.991756 0.000000e+00
-6 as.factor(motor_score)4 -1.49637069 0.071492961 -20.930322 0.000000e+00
-7                   pupil  0.48922641 0.038536483  12.695149 0.000000e+00
-8                      ct  0.42634415 0.028388807  15.018037 0.000000e+00
+1             (Intercept) -2.33931814 0.337744863  -6.926288 4.565459e-12
+2                    type -0.41950933 0.179122441  -2.342026 1.919717e-02
+3                     age  0.03042682 0.001565356  19.437637 0.000000e+00
+4 as.factor(motor_score)2 -0.66613691 0.069302767  -9.611981 0.000000e+00
+5 as.factor(motor_score)3 -1.04915293 0.070119780 -14.962297 0.000000e+00
+6 as.factor(motor_score)4 -1.50206863 0.071907169 -20.888997 0.000000e+00
+7                   pupil  0.49317740 0.035150534  14.030438 0.000000e+00
+8                      ct  0.42802910 0.029616105  14.452579 0.000000e+00
              b         df dfcom          fmi       lambda m          riv
-1 9.395965e-04  8668.10341 11013 0.0099557149 0.0097273074 5 0.0098228573
-2 1.729929e-05 10991.51028 11013 0.0008229571 0.0006411648 5 0.0006415761
-3 8.332782e-09 10482.95010 11013 0.0042891736 0.0040992241 5 0.0041160969
-4 5.649577e-05  6979.24803 11013 0.0145856016 0.0143032580 5 0.0145108098
-5 3.003051e-05  9526.87220 11013 0.0075503299 0.0073420043 5 0.0073963080
-6 3.329522e-05  9362.42789 11013 0.0080288174 0.0078169354 5 0.0078785213
-7 3.337070e-04    54.63795 11013 0.2949938851 0.2696512493 5 0.3692088868
-8 1.180517e-05  5893.23780 11013 0.0179108151 0.0175775781 5 0.0178920775
+1 2.240527e-04 10819.9749 11013 0.0025413258 0.0023569693 5 0.0023625377
+2 2.066282e-05 10984.4464 11013 0.0009546917 0.0007728065 5 0.0007734042
+3 1.927783e-08  8774.5121 11013 0.0096665946 0.0094408908 5 0.0095308707
+4 1.096200e-04  3559.8620 11013 0.0279345878 0.0273886158 5 0.0281598759
+5 3.652274e-05  8968.6823 11013 0.0091347491 0.0089138126 5 0.0089939832
+6 9.004098e-05  4952.3514 11013 0.0212918407 0.0208966706 5 0.0213426612
+7 1.183186e-04   293.7825 11013 0.1208778877 0.1149133407 5 0.1298328695
+8 6.936809e-05   425.1630 11013 0.0991319347 0.0949041246 5 0.1048553276
           ubar
-1 1.147849e-01
-2 3.235648e-02
-3 2.429325e-06
-4 4.672029e-03
-5 4.872243e-03
-6 5.071289e-03
-7 1.084612e-03
-8 7.917581e-04
+1 1.138027e-01
+2 3.206005e-02
+3 2.427206e-06
+4 4.671329e-03
+5 4.872956e-03
+6 5.062592e-03
+7 1.093578e-03
+8 7.938720e-04
 \end{CodeOutput}
@@
       AIC       BIC    logLik  deviance  df.resid 
-10522.944 10588.713 -5252.472 10504.944     11013 
+10509.778 10575.547 -5245.889 10491.778     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2913  
+ name   (Intercept) 0.2939  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.32713                 -0.41953                  0.03031  
+                -2.3395                  -0.4185                   0.0304  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65651                 -1.04719                 -1.49327  
+                -0.6597                  -1.0504                  -1.5006  
                   pupil                       ct  
-                0.48990                  0.42335  
+                 0.4893                   0.4302  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10525.723 10591.492 -5253.861 10507.723     11013 
+10501.337 10567.106 -5241.668 10483.337     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2937  
+ name   (Intercept) 0.2896  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.30150                 -0.41940                  0.03024  
+               -2.35407                 -0.42416                  0.03023  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65497                 -1.04451                 -1.49454  
+               -0.66302                 -1.03875                 -1.50219  
                   pupil                       ct  
-                0.47281                  0.42888  
+                0.50367                  0.43246  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10512.28 10578.05 -5247.14 10494.28    11013 
+10500.594 10566.363 -5241.297 10482.594     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2852  
+ name   (Intercept) 0.2902  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.35320                 -0.41028                  0.03034  
+               -2.35434                 -0.41988                  0.03042  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66933                 -1.05474                 -1.50379  
+               -0.65515                 -1.04987                 -1.49095  
                   pupil                       ct  
-                0.49491                  0.42442  
+                0.49221                  0.43460  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10530.02 10595.79 -5256.01 10512.02    11013 
+10520.673 10586.442 -5251.336 10502.673     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2923  
+ name   (Intercept) 0.2891  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.32670                 -0.41286                  0.03037  
+               -2.32285                 -0.42261                  0.03062  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66801                 -1.04770                 -1.50070  
+               -0.67111                 -1.05287                 -1.49941  
                   pupil                       ct  
-                0.47210                  0.43109  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+                0.50321                  0.41360  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
 
@@
       AIC       BIC    logLik  deviance  df.resid 
-10488.615 10554.383 -5235.307 10470.615     11013 
+10532.391 10598.160 -5257.196 10514.391     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2939  
+ name   (Intercept) 0.2866  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.38215                 -0.41343                  0.03049  
+               -2.32583                 -0.41243                  0.03045  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65379                 -1.05740                 -1.48956  
+               -0.68169                 -1.05384                 -1.51715  
                   pupil                       ct  
-                0.51641                  0.42397  
+                0.47746                  0.42928  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
 \end{CodeInput}
-\begin{CodeOutput}
-
- iter imp variable
-  1   1  hiv
-  1   2  hiv
-  1   3  hiv
-  1   4  hiv
-  1   5  hiv
-  1   6  hiv
-  1   7  hiv
-  1   8  hiv
-  1   9  hiv
-  1   10  hiv
-\end{CodeOutput}
 \end{CodeChunk}
@@
 \item
+  Adding evaluations of the imputations such as convergence checks
+\item
   Adding some kind of help function to mice that suggests a suitable

### 2022-12-09 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Fix overlapping affiliations

--- 2022-12-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
   - name: "Johanna Munoz Avila"
-    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
+    affiliation: |
+      | Julius Center for Health Sciences and Primary Care, 
+      | University Medical Center Utrecht, Utrecht University, 
+      | Utrecht, The Netherlands \AND
   - name: Thomas P. A. Debray
-    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
+    affiliation: Julius Center for Health Sciences and Primary Care
+    address: |
+      | University Medical Center Utrecht, Utrecht University, 
+      | Utrecht, The Netherlands   
   - name: Gerko Vink
-    affiliation: Utrecht University 
+    affiliation: Utrecht University \AND
   - name: Valentijn M. T. de Jong
-    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands \AND Data Analytics and Methods Task Force, European Medicines Agency, Amsterdam, The Netherlands
+    affiliation: |
+      | Julius Center for Health Sciences and Primary Care, 
+      | University Medical Center Utrecht, Utrecht University, 
+      | Utrecht, The Netherlands 
+    affiliation2: |
+      | Data Analytics and Methods Task Force, 
+      | European Medicines Agency, Amsterdam, The Netherlands
 
@@
 
-```{r, setup, include=FALSE}
+```{r, setup, include=FALSE, eval=TRUE}
 knitr::opts_chunk$set(message=FALSE, warning=FALSE)
@@
 [TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are systematically missing.
-[TODO: Refer to Figure 1 and put interpretation in the figure caption.]
+[TODO: Refer to Figure 1 and put interpretation in the figure caption.
+VMTdJ: "Mention in text that this is sporadic missingness?"]
 
@@
   scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
-  coord_cartesian(expand = c(0,0,0,0)) +
+  # coord_cartesian(expand = c(0,0)) +
   theme_minimal() +
@@
 The analysis model for this dataset is a prediction model with `mort` as the outcome.
-In this tutorial we'll estimate the adjusted prognostic effect of `ct` on unfortunate outcomes.
+In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes.
 The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model:
@@
 
-As we can see... [TODO: fill in.]
+As we can see...
+[TODO: fill in: This means that a higher ct (Marshall Computerized Tomography classification) is associated with a lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI ...
+to ..., when controlling for ...]
 
@@
 
-R=region+language+(1|InterviewID) model with with interviewer as random effects, because the observations are not independent.
+R=region+language+(1|InterviewID) model with interviewer as random effects, because the observations are not independent.
 Interviews are not allocated randomly.
 In theory we expect the inclusion-restriction variable to be randomly assigned, that's why we're adding region and language to compensate for non-random allocation.
+[TODO: Explain exclusion-restriction, refer to paper Johanna].
 
@@
 # Categorical variables have to be recoded as factor
-# to use the 2l.binary imputation method, it is required that the level names should not have any space between words
+# to use the 2l.binary imputation method, it is required 
+# that the level names should not have any space between words
 
@@
 Impute the data:
-```{r}
+```{r eval=FALSE}
 imp <- mice( hivdata, # dataset with missing values
@@
 
+# Funding
+
+This project has received funding from the European Union's Horizon 2020 research and innovation programme under ReCoDID grant agreement No 825746.
+
+The views expressed in this paper are the personal views of the authors and may not be understood or quoted as being made on behalf of or reflecting the position of the regulatory agency/agencies or organizations with which the authors are employed/affiliated.
+
 # References

### 2022-12-09 Data/obesity.R
Replace Heckman example

@@ -89,5 +89,7 @@ imp_pmm <- mice(data = data, meth = meth, pred = pred, maxit = 1, m = 5, seed = 
 
 densityplot(imp_pmm)
 
-obesity <- data
+obesity <- data %>% janitor::clean_names() %>% mutate(weight = round(weight, 2), rt = round(time, 2), .keep = "unused")
+obesity$bmi = round(b$weight / (b$height)^2, 2)
+obesity = obesity[, c("cluster", "gender", "age", "height", "weight", "bmi", "rt")]
 save(obesity, file = "./Data/obesity.Rdata")

### 2022-12-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
tables

--- 2022-12-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 author:
-  - name: Hanne I. Oberman
+  - name: Hanne Oberman
     affiliation: Utrecht University
@@
     url: https://hanneoberman.github.io/
-  - name: "Johanna Munoz Avila"
-    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
-  - name: Thomas P. A. Debray
-    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
+  - name: "Johanna Munoz"
+    affiliation: University Medical Center Utrecht \AND
+  - name: Valentijn de Jong
+    affiliation: University Medical Center Utrecht 
   - name: Gerko Vink
-    affiliation: Utrecht University 
-  - name: Valentijn M. T. de Jong
-    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands \AND Data Analytics and Methods Task Force, European Medicines Agency, Amsterdam, The Netherlands
-
+    affiliation: Utrecht University \AND
+  - name: Thomas Debray
+    affiliation: University Medical Center Utrecht 
 title:
@@
 If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01].
-[TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation.
-VMTdJ: "I would only make that link after introducing missing data, and missing data is not mentioned yet, so I would do that in the next section"] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
-These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'.
+[TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation.] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
+These models are also know as 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.
 Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.
@@
 \hline
+Sample unit         & Units of the population from which measurements are taken in a sample.\\                           & These should be distinct and not overlapping, e.g., patients, students.\\ 
+Hierarchical levels & Data are grouped into clusters at different levels.
+A three-level \\                               & hierarchical model could occur by collecting observations of math scores\\                         & (lower level) which can be measured over the whole school year for each\\                          & student, and then grouped by student (middle level) and these in turn can\\                        & be aggregated into classrooms (high level).\\
+Fixed effect        & Here we assume that the values of an independent variable are fixed, i.e., \\ 
+  & the values observed in the study are representative of all values in the \\                        & population we are interested in.
+In the case of a fixed factor variable, for \\                     & example treatment type, we are only concerned with assessing the difference \\
+                & in the dependent variable y e.g., blood pressure between treatments A and B.\\                     & There may be additional treatments, but based on the objective of the study,\\                     & which in this case is the equivalence effect of A and B, the other treatment\\                     & types are irrelevant and therefore the treatment type is included as a fixed \\                     & effect with only the A and B levels in the hierarchical model.
+\\
+Random effect       & The values of an independent variable are assumed to be randomly drawn from \\                     & a population with more values.
+In a random factor, the levels observed in the\\                     & study represent a sample of all possible levels, e.g., hospitals, classrooms.\\                     & For example, in a study to predict the blood pressure of patients on \\
+& admission we might select only certain hospitals that are representative of \\                     & the entire population of hospitals. Here we are not interested in measuring \\ 
+& the difference of y between individual hospitals, but rather the variation of\\                     & y between hospitals, to create a model that can be generalized to the entire\\                     & population of hospitals.
+\\
+
 ICC                 & The variability due to clustering is often measured by means of the \\
@@
 \end{table}
+```
 
@@
 
-Imputation of missing data requires consideration of the mechanism behind the missingness.
+Imputation of missing data requires to consider the mechanism behind the missingness.
 Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}).
 For each of these three missingness generating mechanisms, different imputation strategies are warranted @yuce08 and @hox15.
-We here consider the general case that data are MAR, and expand on certain MNAR situations.
+We here consider the general case that data are MAR, and expand on special MNAR situations.
 
@@
 \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
-However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent whereas multilevel data are dependent.
+However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent.
 For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.
@@
 For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial.
-[TODO: change this sentence because we do not assume familiarity: "Assumed knowledge includes basic familiarity with multilevel imputation [see e.g.
-@audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod})"].
+Assumed knowledge includes basic familiarity with multilevel imputation [see e.g.
+@audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).
 
+```{=tex}
 \begin{table}[tb]
@@
 \centering
-\begin{tabular}{ll}
+\begin{tabular}{lll}
 \hline
-\textbf{Concept} & \textbf{Details}   \\
+\textbf{Formula lme4} & \textbf{Details}   \\
 \hline
-& [TODO: explain \pkg{lme4} notation here] \\
+y$\sim$ x1 + (1 | g1)                                       & Fixed x1 predictor with random intercept  \\
+                                                         & varying among g\\
+y$\sim$x1*x2+ (1| g1)                                      & Interactions of x1 and x2 only in fixed effect\\
+y$\sim$x1*x2+ (x2| g1)                                     & Interactions of x1 and x2 only in fixed effect\\
+                                                         & with slope of x2 randomly varying among g1\\
+y$\sim$x1*x2+ (x1*x2| g1)                                  & variance-covariance matrix estimated only with \\
+                                                         & the variance terms of intercept, slope of x1, \\
+                                                         & slope of x2 and interaction x1*x2\\
+y$\sim$x1*x2+ (x1| g1)+ (x2| g1)                           & variance-covariance matrix estimated separately,\\
+                                                         & i.e, one for intercept and x1 and another for \\
+                                                         & intercept and x2\\
+y$\sim$x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)            & Fixed x1 with correlated random intercept and \\
+                                                         & random slope of x\\
+y$\sim$x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1) & Fixed x1 with uncorrelated random intercept\\
+                                                         & and random slope of x1\\  
+y$\sim$(1 | g1) + (1 | g2)                                 & Random intercept varying among g1 and among g2\\
+y$\sim$(1 | g1/g2) or ~(1|g1)+(1|g1:g2)                    & Random intercept varying among g1 and g2 
+                                                         & withing g1\\
+
+& [TODO: explain \pkg{lme4} notation here
+https://datalorax.github.io/equatiomatic/articles/lmer4-lmer.html
+https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf
+] \\
 \hline
@@
 
-This tutorial is dedicated to readers who are unfamiliar with multiple imputation.
+This tutorial is dedicated to readers who are unfamiliar with multiple imptuation.
 More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study IMPACT) or  under MNAR conditions (case study HIV).
@@
 
-Impute the data:
-```{r}
-imp <- mice( hivdata, # dataset with missing values
-                        m = 10,   # number of imputations
-                        maxit = 1,
-                        seed = 1234, #seed attached to the dataID
-                        meth = meth, #imputation method vector
-                        pred = pred, #imputation predictors matrix
-                        print = T,
-                        meta_method="reml",
-                        pmm=FALSE)
-```
+<!-- Impute the data: -->
+<!-- ```{r} -->
+<!-- imp <- mice( hivdata, # dataset with missing values -->
+<!--                         m = 10,   # number of imputations -->
+<!--                         maxit = 1, -->
+<!--                         seed = 1234, #seed attached to the dataID -->
+<!--                         meth = meth, #imputation method vector -->
+<!--                         pred = pred, #imputation predictors matrix -->
+<!--                         print = T, -->
+<!--                         meta_method="reml", -->
+<!--                         pmm=FALSE) -->
+<!-- ``` -->
 

### 2022-12-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
tables

--- 2022-10-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 
-%% recommended packages
-\usepackage{orcidlink,thumbpdf,lmodern}
-
 \usepackage[utf8]{inputenc}
 
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 \author{
-Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
-Medical Center Utrecht \AND Valentijn de Jong\\University Medical
-Center Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
+Hanne Oberman\\Utrecht University \And Johanna Munoz\\University Medical
+Center Utrecht \AND Valentijn de Jong\\University Medical Center
+Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
 Debray\\University Medical Center Utrecht
@@
 
-\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
-Jong, Gerko Vink, Thomas Debray}
+\Plainauthor{Hanne Oberman, Johanna Munoz, Valentijn de Jong, Gerko
+Vink, Thomas Debray}
 \Plaintitle{Imputation of Incomplete Multilevel Data with mice}
@@
 
-% tightlist command for lists without linebreak
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+% Pandoc syntax highlighting
+
+% Pandoc citation processing
 
@@
 
-\hypertarget{missingness-in-multilevel-data}{%
-\subsection{Missingness in multilevel
-data}\label{missingness-in-multilevel-data}}
-
+\begin{verbatim}
+
+## Missingness in multilevel data
+
+<!-- Maybe besides CC and imputation name also likelihood based methods -->
 Like any other dataset, clustered datasets are prone to missing data.
-Several strategies can be used to handle missing data, including
-complete case analysis and imputation. We focus on the latter approach
-and discuss statistical methods for replacing the missing data with one
-or more plausible values. Afterwards, the completed data can be analyzed
-as if they were completely observed. In contrast to single imputation
-(where missing data are only replaced once), multiple imputation allows
-to preserve uncertainty due to missingness and is therefore recommended
-(c.f. Rubin 1976).
-
-When clustered datasets are affected by missing values, we can
-distinguish between two types of missing data: sporadic missingness and
-systematic missingness \citep{resc13}. Sporadic missingness arises when
-variables are missing for some but not all of the units in a cluster
-\citep{buur18, jola18}. For example, it is possible that test results
-are missing for several students in one or more classes. {[}TODO:
-Provide an example for one of the case studies below.{]} When all
-observations are missing within one or more clusters, data are
-systematically missing. {[}TODO: Refer to Figure 1 and put
-interpretation in the figure caption.{]}
-
+Several strategies can be used to handle missing data, including complete case analysis and imputation.
+We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values.
+Afterwards, the completed data can be analyzed as if they were completely observed.
+In contrast to single imputation (where missing data are only replaced once), multiple imputation allows to preserve uncertainty due to missingness and is therefore recommended (c.f.
+Rubin 1976).
+
+When clustered datasets are affected by missing values, we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13].
+Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18].
+For example, it is possible that test results are missing for several students in one or more classes.
+[TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are systematically missing.
+[TODO: Refer to Figure 1 and put interpretation in the figure caption.]
+
+<!-- Systematic missingness implies that one or more variables are never observed in a certain cluster.
+With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18].
+We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters and $p$ variables.
+-->
+
+```{=latex}
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+\end{verbatim}
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.932     0.274    18.012     4.420     0.000    19.514     0.964 
-sex             0.859     0.284     3.026     4.216     0.036    37.601     0.981
+(Intercept)     5.012     0.295    16.994     4.362     0.000    22.587     0.969 
+sex             0.695     0.251     2.768     4.287     0.047    28.390     0.975
 
                             Estimate 
-Intercept~~Intercept|school    0.265 
-Residual~~Residual             1.007 
-ICC|school                     0.213
+Intercept~~Intercept|school    0.266 
+Residual~~Residual             1.035 
+ICC|school                     0.208
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.054     0.266    19.029     4.554     0.000    14.936     0.954 
-sex             0.732     0.306     2.393     4.233     0.071    34.886     0.980
+(Intercept)     4.915     0.217    22.642     4.926     0.000     9.110     0.926 
+sex             0.975     0.283     3.444     4.250     0.024    32.504     0.978
 
                             Estimate 
-Intercept~~Intercept|school    0.318 
-Residual~~Residual             1.257 
-ICC|school                     0.208
+Intercept~~Intercept|school    0.351 
+Residual~~Residual             1.153 
+ICC|school                     0.233
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.714     0.193    24.432     5.032     0.000     8.224     0.919 
-sex             1.286     0.288     4.460     4.218     0.010    37.215     0.981
+(Intercept)     5.011     0.410    12.222     4.226     0.000    35.955     0.980 
+sex             0.928     0.381     2.434     4.168     0.069    48.221     0.985
 
                             Estimate 
-Intercept~~Intercept|school    0.299 
-Residual~~Residual             1.050 
-ICC|school                     0.228
+Intercept~~Intercept|school    0.313 
+Residual~~Residual             1.428 
+ICC|school                     0.188
 
@@
    <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
- 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
+ 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.98e-31  0.0571   0.130
  2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
- 3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
+ 3 fixed    <NA>  age          1.03    0.00265   12.9   7.39e-38  1.03     1.04 
  4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
@@
  6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
- 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
+ 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.19e-32  2.63     3.88 
  8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
@@
                      term    estimate   std.error  statistic      p.value
-1             (Intercept) -2.33813790 0.340459131  -6.867602 6.978862e-12
-2                    type -0.41509929 0.179936773  -2.306918 2.107792e-02
-3                     age  0.03034813 0.001561834  19.431086 0.000000e+00
-4 as.factor(motor_score)2 -0.66052409 0.068846376  -9.594174 0.000000e+00
-5 as.factor(motor_score)3 -1.05030912 0.070059114 -14.991756 0.000000e+00
-6 as.factor(motor_score)4 -1.49637069 0.071492961 -20.930322 0.000000e+00
-7                   pupil  0.48922641 0.038536483  12.695149 0.000000e+00
-8                      ct  0.42634415 0.028388807  15.018037 0.000000e+00
+1             (Intercept) -2.35201485 0.340106796  -6.915519 4.943219e-12
+2                    type -0.41266362 0.180240358  -2.289518 2.206828e-02
+3                     age  0.03049017 0.001570142  19.418732 1.231236e-81
+4 as.factor(motor_score)2 -0.66765094 0.068736642  -9.713174 3.473396e-22
+5 as.factor(motor_score)3 -1.05520421 0.070218236 -15.027495 2.532464e-50
+6 as.factor(motor_score)4 -1.51238926 0.072304429 -20.916966 1.894925e-90
+7                   pupil  0.48421322 0.038983420  12.421004 6.807219e-17
+8                      ct  0.43474506 0.029967898  14.507025 2.318762e-36
              b          df dfcom         fmi      lambda m         riv
-1 9.395965e-04  8668.10341 11013 0.0099557149 0.0097273074 5 0.0098228573
-2 1.729929e-05 10991.51028 11013 0.0008229571 0.0006411648 5 0.0006415761
-3 8.332782e-09 10482.95010 11013 0.0042891736 0.0040992241 5 0.0041160969
-4 5.649577e-05  6979.24803 11013 0.0145856016 0.0143032580 5 0.0145108098
-5 3.003051e-05  9526.87220 11013 0.0075503299 0.0073420043 5 0.0073963080
-6 3.329522e-05  9362.42789 11013 0.0080288174 0.0078169354 5 0.0078785213
-7 3.337070e-04    54.63795 11013 0.2949938851 0.2696512493 5 0.3692088868
-8 1.180517e-05  5893.23780 11013 0.0179108151 0.0175775781 5 0.0178920775
+1 5.283158e-04 10118.53147 11013 0.005677319 0.005480804 5 0.005511008
+2 4.875905e-05 10894.06456 11013 0.001984282 0.001801077 5 0.001804327
+3 3.333802e-08  6323.24413 11013 0.016538196 0.016227183 5 0.016494848
+4 4.185462e-05  8330.19444 11013 0.010867814 0.010630361 5 0.010744580
+5 5.135737e-05  7632.09368 11013 0.012757930 0.012499256 5 0.012657465
+6 1.404727e-04  2826.80320 11013 0.032927570 0.032243595 5 0.033317884
+7 3.572093e-04    49.95954 11013 0.309174397 0.282061687 5 0.392877330
+8 8.584367e-05   294.82759 11013 0.120648604 0.114703577 5 0.129565165
           ubar
-1 1.147849e-01
-2 3.235648e-02
-3 2.429325e-06
-4 4.672029e-03
-5 4.872243e-03
-6 5.071289e-03
-7 1.084612e-03
-8 7.917581e-04
+1 1.150387e-01
+2 3.242808e-02
+3 2.425341e-06
+4 4.674500e-03
+5 4.868972e-03
+6 5.059363e-03
+7 1.091056e-03
+8 7.950625e-04
 \end{CodeOutput}
@@
       AIC       BIC    logLik  deviance  df.resid 
-10522.944 10588.713 -5252.472 10504.944     11013 
+10495.423 10561.192 -5238.712 10477.423     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2913  
+ name   (Intercept) 0.2843  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.32713                 -0.41953                  0.03031  
+               -2.37193                 -0.41015                  0.03052  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65651                 -1.04719                 -1.49327  
+               -0.65803                 -1.04612                 -1.51245  
                   pupil                       ct  
-                0.48990                  0.42335  
+                0.50405                  0.42496  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
      AIC      BIC   logLik deviance df.resid 
-10525.723 10591.492 -5253.861 10507.723     11013 
+10500.88 10566.65 -5241.44 10482.88    11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2937  
+ name   (Intercept) 0.2917  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.30150                 -0.41940                  0.03024  
+               -2.37717                 -0.41511                  0.03067  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65497                 -1.04451                 -1.49454  
+               -0.66935                 -1.05210                 -1.49428  
                   pupil                       ct  
-                0.47281                  0.42888  
+                0.49013                  0.43835  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10512.28 10578.05 -5247.14 10494.28    11013 
+10505.026 10570.795 -5243.513 10487.026     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2852  
+ name   (Intercept) 0.2908  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.35320                 -0.41028                  0.03034  
+               -2.32339                 -0.42359                  0.03023  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66933                 -1.05474                 -1.50379  
+               -0.67142                 -1.05776                 -1.51038  
                   pupil                       ct  
-                0.49491                  0.42442  
+                0.49756                  0.42474  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10530.02 10595.79 -5256.01 10512.02    11013 
+10519.511 10585.280 -5250.755 10501.511     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2923  
+ name   (Intercept) 0.2961  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.32670                 -0.41286                  0.03037  
+               -2.33576                 -0.40872                  0.03039  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66801                 -1.04770                 -1.50070  
+               -0.66477                 -1.05453                 -1.51860  
                   pupil                       ct  
-                0.47210                  0.43109  
+                0.45928                  0.44419  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10488.615 10554.383 -5235.307 10470.615     11013 
+10522.038 10587.807 -5252.019 10504.038     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2939  
+ name   (Intercept) 0.2955  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.38215                 -0.41343                  0.03049  
+               -2.35182                 -0.40575                  0.03064  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65379                 -1.05740                 -1.48956  
+               -0.67468                 -1.06551                 -1.52623  
                   pupil                       ct  
-                0.51641                  0.42397  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+                0.47006                  0.44148  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
 
@@
 
-Impute the data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> imp <- mice( hivdata, # dataset with missing values
-+                         m = 10,   # number of imputations
-+                         maxit = 1,
-+                         seed = 1234, #seed attached to the dataID
-+                         meth = meth, #imputation method vector
-+                         pred = pred, #imputation predictors matrix
-+                         print = T,
-+                         meta_method="reml",
-+                         pmm=FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-
- iter imp variable
-  1   1  hiv
-  1   2  hiv
-  1   3  hiv
-  1   4  hiv
-  1   5  hiv
-  1   6  hiv
-  1   7  hiv
-  1   8  hiv
-  1   9  hiv
-  1   10  hiv
-\end{CodeOutput}
-\end{CodeChunk}
-
 \hypertarget{discussion}{%
@@
   env dump in repo
+\item
+  I don't know if in your article you cover something about model
+  complexity, for example sometimes I have to switch from 2l. methods to
+  1l. methods just because the model didn't converge.. this is due to
+  the considered imputation model is very complex regarding the amount
+  of information counted\ldots{} I know that a solution for an
+  imputation model with many predictors is to check correlation plots as
+  you did or use quickpred().. (maybe you can add this somewhere after
+  the correlation plots)\ldots..But as for cluster specification, I
+  don't know if besides plots of distribution per cluster there is
+  something else can be done to see if i have to use 1l. or 2l. for a
+  given variable, also I have no idea.. how to test which is better
+  between 2l.norm or 2l.2stage.norm.
 \end{itemize}
+
+Translated with www.DeepL.com/Translator (free version)
 

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

--- 2022-12-09 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 library(mitml)        # for multilevel pooling
+library(miceheckman)  # for imputation
 ```
@@
 ```{r pattern, fig.height=7.1}
-plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
+plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
 ```
@@
 ```{r}
-data("hiv",  package = "GJRM")
-# We select  5 predictor variables over 9 regions
-colnames(hiv)
-hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
-```
-
-Recode variables:
-```{r}
-# Study/group variable has to be recoded as integer
-hivdata$region<-as.integer(hivdata$region)
-
-# Categorical variables have to be recoded as factor
-# to use the 2l.binary imputation method, it is required 
-# that the level names should not have any space between words
-
-hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
-#the missing variable is stored as a factor in the dataset, otherwise the method will 
-#apply the imputation correction for a  missing continuous variable instead for 
-#a missing binary variable,which is in this case the binary response of the hiv test.
-
-hivdata$marital <- as.factor(hivdata$marital)
-levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
-hivdata$condom <- as.factor(hivdata$condom)
-levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
-hivdata$highhiv <- as.factor(hivdata$highhiv)
-levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
-hivdata$interviewerID <- as.factor(hivdata$interviewerID)
-
-hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
-interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
+data("obesity",  package = "miceheckman")
 ```
@@
 Here, since the location of the interviewer was not randomly assigned to the subjects, the assignment was corrected for region and language.
-```{r}
-hivdata$hivconsent <- as.factor(hivdata$hivconsent)
-ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
-reffect  <- ranef(ID_mixed)$interviewerID
-reffect$interviewerID <- levels(hivdata$interviewerID)
-colnames(reffect)     <- c("IDreffect","interviewerID")
-hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)
-
-hivdata$interviewerID<-NULL
-hivdata$hivconsent<-NULL
-hivdata$language<-NULL
-```
 
@@
 #Set prediction matrix and methods
-ini <- mice(hivdata, maxit = 0)
+ini <- mice(obesity, maxit = 0)
 meth<-ini$method
-meth["hiv"]<-"2l.binary"
+meth["weight"]<-"2l.heckman"
 pred <- ini$pred
-pred[,"region"] <- 0
-pred["region",] <- 0
-pred["hiv","region"]<- -2
-pred["hiv","IDreffect"]<- 0
-
-# Heckman model
-pred["hiv","IDreffect"] <- -3
-meth<-ini$method
-meth["hiv"]<-"2l.heckman"
+pred["weight","cluster"]<- -2
+pred["weight","rt"] <- -3
 ```
@@
 ```{r eval=FALSE}
-imp <- mice( hivdata, # dataset with missing values
+imp <- mice(obesity, # dataset with missing values
                         m = 10,   # number of imputations

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

--- 2022-12-09 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 Hanne I. Oberman\\Utrecht University \And Johanna Munoz Avila\\Julius
-Center for Health Sciences and Primary Care, University Medical Center
-Utrecht, Utrecht University, Utrecht, The Netherlands \And Thomas P. A.
-Debray\\Julius Center for Health Sciences and Primary Care, University
-Medical Center Utrecht, Utrecht University, Utrecht, The
-Netherlands \And Gerko Vink\\Utrecht University \And Valentijn M. T. de
-Jong\\Julius Center for Health Sciences and Primary Care, University
-Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands
-\AND Data Analytics and Methods Task Force, European Medicines Agency,
-Amsterdam, The Netherlands
+Center for Health Sciences and Primary Care,\\
+University Medical Center Utrecht, Utrecht University,\\
+Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
+for Health Sciences and Primary Care \And Gerko Vink\\Utrecht University
+\AND Valentijn M. T. de Jong\\Julius Center for Health Sciences and
+Primary Care,\\
+University Medical Center Utrecht, Utrecht University,\\
+Utrecht, The Netherlands
 }
@@
   URL: \url{https://hanneoberman.github.io/}\\~\\
+        Thomas P. A. Debray\\
+    Julius Center for Health Sciences and Primary Care\\
+    University Medical Center Utrecht, Utrecht University,\\
+Utrecht, The Netherlands\\
+
       }
@@
 systematically missing. {[}TODO: Refer to Figure 1 and put
-interpretation in the figure caption.{]}
+interpretation in the figure caption. VMTdJ: ``Mention in text that this
+is sporadic missingness?''{]}
 
@@
 R> library(mitml)        # for multilevel pooling
+R> library(miceheckman)  # for imputation
 \end{CodeInput}
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.986     0.194    25.704     4.917     0.000     9.201     0.927 
-sex             0.818     0.357     2.290     4.130     0.082    62.169     0.989
+(Intercept)     5.012     0.295    16.994     4.362     0.000    22.587     0.969 
+sex             0.695     0.251     2.768     4.287     0.047    28.390     0.975
 
                             Estimate 
-Intercept~~Intercept|school    0.272 
-Residual~~Residual             0.975 
-ICC|school                     0.227
+Intercept~~Intercept|school    0.266 
+Residual~~Residual             1.035 
+ICC|school                     0.208
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.186     0.254    20.457     4.706     0.000    11.816     0.942 
-sex             0.760     0.219     3.479     4.609     0.020    13.627     0.950
+(Intercept)     4.808     0.197    24.354     5.072     0.000     7.930     0.916 
+sex             1.086     0.135     8.056     5.271     0.000     6.761     0.902
 
                             Estimate 
-Intercept~~Intercept|school    0.345 
-Residual~~Residual             1.578 
-ICC|school                     0.182
+Intercept~~Intercept|school    0.324 
+Residual~~Residual             1.130 
+ICC|school                     0.226
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.000     0.356    14.027     4.283     0.000    28.723     0.976 
-sex             1.056     0.465     2.270     4.105     0.084    76.708     0.991
+(Intercept)     4.808     0.197    24.354     5.072     0.000     7.930     0.916 
+sex             1.086     0.135     8.056     5.271     0.000     6.761     0.902
 
                             Estimate 
-Intercept~~Intercept|school    0.294 
-Residual~~Residual             1.347 
-ICC|school                     0.187
+Intercept~~Intercept|school    0.324 
+Residual~~Residual             1.130 
+ICC|school                     0.226
 
@@
 \texttt{mort} as the outcome. In this tutorial we'll estimate the
-adjusted prognostic effect of \texttt{ct} on unfortunate outcomes. The
+adjusted prognostic effect of \texttt{ct} on mortality outcomes. The
 estimand is the adjusted odds ratio for \texttt{ct}, after including
@@
 \begin{CodeInput}
-R> plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
+R> plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
 \end{CodeInput}
@@
    <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
- 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   3.00e-31  0.0571   0.130
+ 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
  2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
@@
  9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
-10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.56e-19  1.92     2.76 
+10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.55e-19  1.92     2.76 
 11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
@@
 
-As we can see\ldots{} {[}TODO: fill in.{]}
+As we can see\ldots{} {[}TODO: fill in: This means that a higher ct
+(Marshall Computerized Tomography classification) is associated with a
+lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI
+\ldots{} to \ldots, when controlling for \ldots{]}
 
@@
                      term   estimate   std.error  statistic      p.value
-1             (Intercept) -2.33931814 0.337744863  -6.926288 4.565459e-12
-2                    type -0.41950933 0.179122441  -2.342026 1.919717e-02
-3                     age  0.03042682 0.001565356  19.437637 0.000000e+00
-4 as.factor(motor_score)2 -0.66613691 0.069302767  -9.611981 0.000000e+00
-5 as.factor(motor_score)3 -1.04915293 0.070119780 -14.962297 0.000000e+00
-6 as.factor(motor_score)4 -1.50206863 0.071907169 -20.888997 0.000000e+00
-7                   pupil  0.49317740 0.035150534  14.030438 0.000000e+00
-8                      ct  0.42802910 0.029616105  14.452579 0.000000e+00
+1             (Intercept) -2.3054756 0.340854445  -6.763813 1.425704e-11
+2                    type -0.4200458 0.180837517  -2.322780 2.021049e-02
+3                     age  0.0302518 0.001561625  19.372002 0.000000e+00
+4 as.factor(motor_score)2 -0.6619736 0.069527977  -9.520967 0.000000e+00
+5 as.factor(motor_score)3 -1.0520025 0.070748986 -14.869507 0.000000e+00
+6 as.factor(motor_score)4 -1.5087772 0.072240455 -20.885489 0.000000e+00
+7                   pupil  0.4731045 0.038187004  12.389150 0.000000e+00
+8                      ct  0.4291135 0.032126389  13.357041 0.000000e+00
              b          df dfcom         fmi      lambda m         riv
-1 2.240527e-04 10819.9749 11013 0.0025413258 0.0023569693 5 0.0023625377
-2 2.066282e-05 10984.4464 11013 0.0009546917 0.0007728065 5 0.0007734042
-3 1.927783e-08  8774.5121 11013 0.0096665946 0.0094408908 5 0.0095308707
-4 1.096200e-04  3559.8620 11013 0.0279345878 0.0273886158 5 0.0281598759
-5 3.652274e-05  8968.6823 11013 0.0091347491 0.0089138126 5 0.0089939832
-6 9.004098e-05  4952.3514 11013 0.0212918407 0.0208966706 5 0.0213426612
-7 1.183186e-04   293.7825 11013 0.1208778877 0.1149133407 5 0.1298328695
-8 6.936809e-05   425.1630 11013 0.0991319347 0.0949041246 5 0.1048553276
+1 7.937082e-04  9227.60708 11013 0.008412824 0.008197930 5 0.008265691
+2 1.386594e-04 10229.67142 11013 0.005282534 0.005088076 5 0.005114097
+3 1.138007e-08 10083.78177 11013 0.005796971 0.005599802 5 0.005631337
+4 1.427540e-04  2450.43742 11013 0.036222753 0.035436457 5 0.036738334
+5 1.147119e-04  3540.28145 11013 0.028049975 0.027501049 5 0.028278744
+6 1.377409e-04  2902.11217 11013 0.032339200 0.031672563 5 0.032708526
+7 3.160434e-04    58.71181 11013 0.284053948 0.260073900 5 0.351486317
+8 1.985834e-04    74.37520 11013 0.250767587 0.230887509 5 0.300199921
           ubar
-1 1.138027e-01
-2 3.206005e-02
-3 2.427206e-06
-4 4.671329e-03
-5 4.872956e-03
-6 5.062592e-03
-7 1.093578e-03
-8 7.938720e-04
+1 1.152293e-01
+2 3.253582e-02
+3 2.425016e-06
+4 4.662835e-03
+5 4.867765e-03
+6 5.053394e-03
+7 1.078995e-03
+8 7.938047e-04
 \end{CodeOutput}
@@
       AIC       BIC    logLik  deviance  df.resid 
-10509.778 10575.547 -5245.889 10491.778     11013 
+10494.825 10560.593 -5238.412 10476.825     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2939  
+ name   (Intercept) 0.2917  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-                -2.3395                  -0.4185                   0.0304  
+               -2.33693                 -0.43143                  0.03013  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-                -0.6597                  -1.0504                  -1.5006  
+               -0.67764                 -1.05120                 -1.49989  
                   pupil                       ct  
-                 0.4893                   0.4302  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+                0.49336                  0.44123  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
 
@@
       AIC       BIC    logLik  deviance  df.resid 
-10501.337 10567.106 -5241.668 10483.337     11013 
+10543.607 10609.376 -5262.803 10525.607     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2896  
+ name   (Intercept) 0.2937  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.35407                 -0.42416                  0.03023  
+               -2.28159                 -0.42088                  0.03026  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66302                 -1.03875                 -1.50219  
+               -0.65610                 -1.04928                 -1.49432  
                   pupil                       ct  
-                0.50367                  0.43246  
+                0.48055                  0.41014  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10500.594 10566.363 -5241.297 10482.594     11013 
+10543.170 10608.939 -5262.585 10525.170     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2902  
+ name   (Intercept) 0.2901  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.35434                 -0.41988                  0.03042  
+               -2.29948                 -0.40181                  0.03038  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65515                 -1.04987                 -1.49095  
+               -0.65959                 -1.06391                 -1.52075  
                   pupil                       ct  
-                0.49221                  0.43460  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+                0.45837                  0.42186  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
 
@@
       AIC       BIC    logLik  deviance  df.resid 
-10520.673 10586.442 -5251.336 10502.673     11013 
+10527.222 10592.991 -5254.611 10509.222     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2891  
+ name   (Intercept) 0.2962  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.32285                 -0.42261                  0.03062  
+               -2.27660                 -0.42912                  0.03016  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67111                 -1.05287                 -1.49941  
+               -0.66963                 -1.05952                 -1.51971  
                   pupil                       ct  
-                0.50321                  0.41360  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
+                0.45085                  0.44439  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
 
@@
       AIC       BIC    logLik  deviance  df.resid 
-10532.391 10598.160 -5257.196 10514.391     11013 
+10519.178 10584.947 -5250.589 10501.178     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2866  
+ name   (Intercept) 0.2897  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.32583                 -0.41243                  0.03045  
+               -2.33278                 -0.41699                  0.03032  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.68169                 -1.05384                 -1.51715  
+               -0.64691                 -1.03611                 -1.50921  
                   pupil                       ct  
-                0.47746                  0.42928  
+                0.48240                  0.42794  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
 
-R=region+language+(1\textbar InterviewID) model with with interviewer as
+R=region+language+(1\textbar InterviewID) model with interviewer as
 random effects, because the observations are not independent. Interviews
@@
 adding region and language to compensate for non-random allocation.
+{[}TODO: Explain exclusion-restriction, refer to paper Johanna{]}.
 
@@
 \begin{CodeInput}
-R> data("hiv",  package = "GJRM")
-R> # We select  5 predictor variables over 9 regions
-R> colnames(hiv)
-\end{CodeInput}
-\begin{CodeOutput}
- [1] "hivconsent"      "hiv"             "age"             "education"      
- [5] "wealth"          "region"          "marital"         "std"            
- [9] "age1sex_cat"     "highhiv"         "partner"         "condom"         
-[13] "aidscare"        "knowsdiedofaids" "evertestedHIV"   "smoke"          
-[17] "religion"        "ethnicity"       "language"        "interviewerID"  
-[21] "sw"             
-\end{CodeOutput}
-\begin{CodeInput}
-R> hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
-\end{CodeInput}
-\end{CodeChunk}
-
-Recode variables:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # Study/group variable has to be recoded as integer
-R> hivdata$region<-as.integer(hivdata$region)
-R> 
-R> # Categorical variables have to be recoded as factor
-R> # to use the 2l.binary imputation method, it is required that the level names should not have any space between words
-R> 
-R> hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
-R> #the missing variable is stored as a factor in the dataset, otherwise the method will 
-R> #apply the imputation correction for a  missing continuous variable instead for 
-R> #a missing binary variable,which is in this case the binary response of the hiv test.
-R> 
-R> hivdata$marital <- as.factor(hivdata$marital)
-R> levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
-R> hivdata$condom <- as.factor(hivdata$condom)
-R> levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
-R> hivdata$highhiv <- as.factor(hivdata$highhiv)
-R> levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
-R> hivdata$interviewerID <- as.factor(hivdata$interviewerID)
-R> 
-R> hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
-R> interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
+R> data("obesity",  package = "miceheckman")
 \end{CodeInput}
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> hivdata$hivconsent <- as.factor(hivdata$hivconsent)
-R> ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
-R> reffect  <- ranef(ID_mixed)$interviewerID
-R> reffect$interviewerID <- levels(hivdata$interviewerID)
-R> colnames(reffect)     <- c("IDreffect","interviewerID")
-R> hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)
-R> 
-R> hivdata$interviewerID<-NULL
-R> hivdata$hivconsent<-NULL
-R> hivdata$language<-NULL
-\end{CodeInput}
-\end{CodeChunk}
-
 Set the Heckman model as imputation method
@@
 R> #Set prediction matrix and methods
-R> ini <- mice(hivdata, maxit = 0)
+R> ini <- mice(obesity, maxit = 0)
 R> meth<-ini$method
-R> meth["hiv"]<-"2l.binary"
+R> meth["weight"]<-"2l.heckman"
 R> pred <- ini$pred
-R> pred[,"region"] <- 0
-R> pred["region",] <- 0
-R> pred["hiv","region"]<- -2
-R> pred["hiv","IDreffect"]<- 0
-R> 
-R> # Heckman model
-R> pred["hiv","IDreffect"] <- -3
-R> meth<-ini$method
-R> meth["hiv"]<-"2l.heckman"
+R> pred["weight","cluster"]<- -2
+R> pred["weight","rt"] <- -3
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> imp <- mice( hivdata, # dataset with missing values
+R> imp <- mice(obesity, # dataset with missing values
 +                         m = 10,   # number of imputations
@@
 
+\hypertarget{funding}{%
+\section{Funding}\label{funding}}
+
+This project has received funding from the European Union's Horizon 2020
+research and innovation programme under ReCoDID grant agreement No
+825746.
+
+The views expressed in this paper are the personal views of the authors
+and may not be understood or quoted as being made on behalf of or
+reflecting the position of the regulatory agency/agencies or
+organizations with which the authors are employed/affiliated.
+
 \renewcommand\refname{References}

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

@@
     url: https://hanneoberman.github.io/
-  - name: "Johanna Munoz Avila"
+  - name: "Johanna Munoz"
     affiliation: |
@@
 
+```{=tex}
 \begin{table}[tb]
@@
 
-Impute the data:
 ```{r eval=FALSE}
@@
 
+-  I don't know if in your article you cover something about model complexity, for example sometimes I have to switch from 2l.
+methods to 1l.
+methods just because the model didn't converge..
+this is due to the considered imputation model is very complex regarding the amount of information counted...
+I know that a solution for an imputation model with many predictors is to check correlation plots as you did or use quickpred()..
+(maybe you can add this somewhere after the correlation plots).....But as for cluster specification, I don't know if besides plots of distribution per cluster there is something else can be done to see if i have to use 1l.
+or 2l.
+for a given variable, also I have no idea..
+how to test which is better between 2l.norm or 2l.2stage.norm.
+
 # Funding

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update pdf

@@
 \author{
-Hanne I. Oberman\\Utrecht University \And Johanna Munoz Avila\\Julius
-Center for Health Sciences and Primary Care,\\
+Hanne I. Oberman\\Utrecht University \And Johanna Munoz\\Julius Center
+for Health Sciences and Primary Care,\\
 University Medical Center Utrecht, Utrecht University,\\
@@
 
-\Plainauthor{Hanne I. Oberman, Johanna Munoz Avila, Thomas P. A.
-Debray, Gerko Vink, Valentijn M. T. de Jong}
+\Plainauthor{Hanne I. Oberman, Johanna Munoz, Thomas P. A. Debray, Gerko
+Vink, Valentijn M. T. de Jong}
 \Plaintitle{Imputation of Incomplete Multilevel Data with mice}
@@
 
-Impute the data:
-
 \begin{CodeChunk}
@@
   env dump in repo
+\item
+  I don't know if in your article you cover something about model
+  complexity, for example sometimes I have to switch from 2l. methods to
+  1l. methods just because the model didn't converge.. this is due to
+  the considered imputation model is very complex regarding the amount
+  of information counted\ldots{} I know that a solution for an
+  imputation model with many predictors is to check correlation plots as
+  you did or use quickpred().. (maybe you can add this somewhere after
+  the correlation plots)\ldots..But as for cluster specification, I
+  don't know if besides plots of distribution per cluster there is
+  something else can be done to see if i have to use 1l. or 2l. for a
+  given variable, also I have no idea.. how to test which is better
+  between 2l.norm or 2l.2stage.norm.
 \end{itemize}

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
add ñ

@@
     url: https://hanneoberman.github.io/
-  - name: "Johanna Munoz"
+  - name: "Johanna Muñoz"
     affiliation: |

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
add ñ

@@
 \author{
-Hanne I. Oberman\\Utrecht University \And Johanna Munoz\\Julius Center
+Hanne I. Oberman\\Utrecht University \And Johanna Muñoz\\Julius Center
 for Health Sciences and Primary Care,\\
@@
 
-\Plainauthor{Hanne I. Oberman, Johanna Munoz, Thomas P. A. Debray, Gerko
+\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
 Vink, Valentijn M. T. de Jong}
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.808     0.197    24.354     5.072     0.000     7.930     0.916 
-sex             1.086     0.135     8.056     5.271     0.000     6.761     0.902
+(Intercept)     4.915     0.217    22.642     4.926     0.000     9.110     0.926 
+sex             0.975     0.283     3.444     4.250     0.024    32.504     0.978
 
                             Estimate 
-Intercept~~Intercept|school    0.324 
-Residual~~Residual             1.130 
-ICC|school                     0.226
+Intercept~~Intercept|school    0.351 
+Residual~~Residual             1.153 
+ICC|school                     0.233
 
@@
              Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.808     0.197    24.354     5.072     0.000     7.930     0.916 
-sex             1.086     0.135     8.056     5.271     0.000     6.761     0.902
+(Intercept)     5.011     0.410    12.222     4.226     0.000    35.955     0.980 
+sex             0.928     0.381     2.434     4.168     0.069    48.221     0.985
 
                             Estimate 
-Intercept~~Intercept|school    0.324 
-Residual~~Residual             1.130 
-ICC|school                     0.226
+Intercept~~Intercept|school    0.313 
+Residual~~Residual             1.428 
+ICC|school                     0.188
 
@@
                      term    estimate   std.error  statistic      p.value
-1             (Intercept) -2.3054756 0.340854445  -6.763813 1.425704e-11
-2                    type -0.4200458 0.180837517  -2.322780 2.021049e-02
-3                     age  0.0302518 0.001561625  19.372002 0.000000e+00
-4 as.factor(motor_score)2 -0.6619736 0.069527977  -9.520967 0.000000e+00
-5 as.factor(motor_score)3 -1.0520025 0.070748986 -14.869507 0.000000e+00
-6 as.factor(motor_score)4 -1.5087772 0.072240455 -20.885489 0.000000e+00
-7                   pupil  0.4731045 0.038187004  12.389150 0.000000e+00
-8                      ct  0.4291135 0.032126389  13.357041 0.000000e+00
+1             (Intercept) -2.35203726 0.340181747  -6.914061 4.994037e-12
+2                    type -0.41265892 0.180274846  -2.289054 2.209524e-02
+3                     age  0.03049023 0.001570162  19.418521 1.238416e-81
+4 as.factor(motor_score)2 -0.66764920 0.068737865  -9.712975 3.480413e-22
+5 as.factor(motor_score)3 -1.05520001 0.070218940 -15.027285 2.540225e-50
+6 as.factor(motor_score)4 -1.51238349 0.072304262 -20.916934 1.850073e-90
+7                   pupil  0.48421447 0.038982800  12.421234 6.772069e-17
+8                      ct  0.43474621 0.029968474  14.506785 2.342253e-36
              b          df dfcom         fmi      lambda m         riv
-1 7.937082e-04  9227.60708 11013 0.008412824 0.008197930 5 0.008265691
-2 1.386594e-04 10229.67142 11013 0.005282534 0.005088076 5 0.005114097
-3 1.138007e-08 10083.78177 11013 0.005796971 0.005599802 5 0.005631337
-4 1.427540e-04  2450.43742 11013 0.036222753 0.035436457 5 0.036738334
-5 1.147119e-04  3540.28145 11013 0.028049975 0.027501049 5 0.028278744
-6 1.377409e-04  2902.11217 11013 0.032339200 0.031672563 5 0.032708526
-7 3.160434e-04    58.71181 11013 0.284053948 0.260073900 5 0.351486317
-8 1.985834e-04    74.37520 11013 0.250767587 0.230887509 5 0.300199921
+1 5.281599e-04 10119.70041 11013 0.005673265 0.005476771 5 0.005506932
+2 4.881699e-05 10893.89378 11013 0.001985736 0.001802528 5 0.001805783
+3 3.335358e-08  6320.89582 11013 0.016545467 0.016234340 5 0.016502243
+4 4.188703e-05  8327.24771 11013 0.010875748 0.010638213 5 0.010752602
+5 5.135721e-05  7632.20045 11013 0.012757637 0.012498967 5 0.012657168
+6 1.403058e-04  2831.75462 11013 0.032888241 0.032205434 5 0.033277139
+7 3.571497e-04    49.97295 11013 0.309130930 0.282023647 5 0.392803531
+8 8.586645e-05   294.69765 11013 0.120677044 0.114729598 5 0.129598366
           ubar
-1 1.152293e-01
-2 3.253582e-02
-3 2.425016e-06
-4 4.662835e-03
-5 4.867765e-03
-6 5.053394e-03
-7 1.078995e-03
-8 7.938047e-04
+1 1.150898e-01
+2 3.244044e-02
+3 2.425385e-06
+4 4.674630e-03
+5 4.869071e-03
+6 5.059539e-03
+7 1.091079e-03
+8 7.950697e-04
 \end{CodeOutput}
@@
       AIC       BIC    logLik  deviance  df.resid 
-10494.825 10560.593 -5238.412 10476.825     11013 
+10495.423 10561.192 -5238.712 10477.423     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2917  
+ name   (Intercept) 0.2843  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.33693                 -0.43143                  0.03013  
+               -2.37195                 -0.41014                  0.03052  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67764                 -1.05120                 -1.49989  
+               -0.65802                 -1.04611                 -1.51245  
                   pupil                       ct  
-                0.49336                  0.44123  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
+                0.50405                  0.42496  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
 
@@
      AIC      BIC   logLik deviance df.resid 
-10543.607 10609.376 -5262.803 10525.607     11013 
+10500.88 10566.65 -5241.44 10482.88    11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2937  
+ name   (Intercept) 0.2917  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.28159                 -0.42088                  0.03026  
+               -2.37718                 -0.41511                  0.03067  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65610                 -1.04928                 -1.49432  
+               -0.66935                 -1.05211                 -1.49429  
                   pupil                       ct  
-                0.48055                  0.41014  
+                0.49013                  0.43835  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10543.170 10608.939 -5262.585 10525.170     11013 
+10505.026 10570.795 -5243.513 10487.026     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2901  
+ name   (Intercept) 0.2908  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.29948                 -0.40181                  0.03038  
+               -2.32339                 -0.42359                  0.03023  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65959                 -1.06391                 -1.52075  
+               -0.67142                 -1.05776                 -1.51038  
                   pupil                       ct  
-                0.45837                  0.42186  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
+                0.49756                  0.42474  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
 
@@
       AIC       BIC    logLik  deviance  df.resid 
-10527.222 10592.991 -5254.611 10509.222     11013 
+10519.511 10585.280 -5250.755 10501.511     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2962  
+ name   (Intercept) 0.2961  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.27660                 -0.42912                  0.03016  
+               -2.33581                 -0.40871                  0.03039  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66963                 -1.05952                 -1.51971  
+               -0.66477                 -1.05451                 -1.51858  
                   pupil                       ct  
-                0.45085                  0.44439  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+                0.45928                  0.44419  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
 
@@
       AIC       BIC    logLik  deviance  df.resid 
-10519.178 10584.947 -5250.589 10501.178     11013 
+10522.038 10587.807 -5252.019 10504.038     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2897  
+ name   (Intercept) 0.2955  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.33278                 -0.41699                  0.03032  
+               -2.35187                 -0.40574                  0.03064  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.64691                 -1.03611                 -1.50921  
+               -0.67468                 -1.06551                 -1.52622  
                   pupil                       ct  
-                0.48240                  0.42794  
+                0.47006                  0.44148  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
update

--- 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
   - name: Hanne I. Oberman
-    affiliation: Utrecht University
+    affiliation: |
+      | Methodology and Statistics
+      | Utrecht University
     address: |
@@
     url: https://hanneoberman.github.io/
-  - name: "Johanna Munoz Avila"
+  - name: "Johanna Muñoz"
     affiliation: |
@@
   - name: Gerko Vink
-    affiliation: Utrecht University \AND
+    affiliation: |
+      | Methodology and Statistics
+      | Utrecht University \AND
   - name: Valentijn M. T. de Jong
@@
 
-\begin{table}[tb]
-\caption{Concepts in multilevel methods}
-\label{tab:clus}
-\centering
-\begin{tabular}{ll}
-\hline
-\textbf{Concept} & \textbf{Details}   \\
-\hline
-ICC                 & The variability due to clustering is often measured by means of the \\
-                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
-                    & of variance that can be attributed to the cluster-level, where a high \\
-                    & ICC would indicate that a lot of variability is due to the cluster \\
-                    & structure. \\
-Random effect       & Multilevel models typically accommodate for variability by including \\
-                    & a separate group mean for each cluster. In addition to random \\
-                    & intercepts, multilevel models can also include random coefficients \\
-                    & and heterogeneous residual error variances across clusters [see e.g. \\
-                    & @gelm06, @hox17 and @jong21]. [TODO: add stratification.] \\
-\hline
-\end{tabular}
-\end{table}
+Table 1: Concepts in multilevel methods
+| **Concept**         | **Details**                                                                   |
+|---------------------|-------------------------------------------------------------------------------|
+| Sample unit         | Units of the population from which measurements are taken in a sample.        |
+| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
+| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
+|                     | the values observed in the study are representative of all values in the      |
+|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
+| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
+|                     | admission we might select only certain hospitals that are representative of   |
+|                     | the difference of y between individual hospitals, but rather the variation of |
+| ICC                 | The variability due to clustering is often measured by means of the           |
+|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
+|                     | of variance that can be attributed to the cluster-level, where a high         |
+|                     | ICC would indicate that a lot of variability is due to the cluster            |
+|                     | structure.                                                                    |
+| Random effect       | Multilevel models typically accommodate for variability by including          |
+|                     | a separate group mean for each cluster. In addition to random                 |
+|                     | intercepts, multilevel models can also include random coefficients            |
+|                     | and heterogeneous residual error variances across clusters [see e.g.          |
+|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratification.]                     |
 
@@
 
-\begin{table}[tb]
-\caption{Concepts in missing data methods}
-\label{tab:miss}
-\centering
-\begin{tabular}{ll}
-\hline
-\textbf{Concept} & \textbf{Details}   \\
-\hline
-MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
-& across all data entries \\
-MAR     & Missing At Random, where the probability to be missing depends on observed \\
-& information \\
-MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
-& depends on unrecorded information, making the missingness non-ignorable \\
-& [@rubi76; @meng94]. \\
-& [TODO: add congeniality, but maybe in-text?] \\
-\hline
-\end{tabular}
-\end{table}
+Table 2: Concepts in missing data methods
+| **Concept**    | **Details**                                                                |
+|----------------|----------------------------------------------------------------------------|
+| MCAR           | Missing Completely At Random, where the probability to be missing is equal |
+|                | across all data entries                                                    |
+| MAR            | Missing At Random, where the probability to be missing depends on observed |
+|                | information                                                                |
+| MNAR           | Missing Not At Random (MNAR), where the probability to be missing          |
+|                | depends on unrecorded information, making the missingness non-ignorable    |
+|                | [@rubi76; @meng94].                                                        |
+|                | [TODO: add congeniality, but maybe in-text?]                               |
 
@@
 
-\begin{table}[tb]
-\caption{Notation}
-\label{tab:mod}
-\centering
-\begin{tabular}{ll}
-\hline
-\textbf{Concept} & \textbf{Details}   \\
-\hline
-& [TODO: explain \pkg{lme4} notation here] \\
-\hline
-\end{tabular}
-\end{table}
+Table 3: Notation
+
+| **Formula \pkg{lme4}**         | **Details**                                 |
+|-------------------------------------------------------------|--------------------------------------------------|
+| `y ~ x1 + (1 | g1)`                                      | Fixed `x1` predictor with random intercept         |
+|                                                             | varying among `g1`                                  |
+| `y~x1*x2+ (1| g1)`                                       | Interactions of `x1` and `x2` only in fixed effect   |
+| `y ~ x1*x2+ (x2| g1)`                                      | Interactions of `x1` and `x2` only in fixed effect   |
+|                                                             | with slope of `x2` randomly varying among `g1`       |
+| `y ~ x1*x2+ (x1*x2| g1)`                                   | variance-covariance matrix estimated only with   |
+|                                                             | the variance terms of intercept, slope of `x1`,    |
+|                                                             | slope of `x2` and interaction `x1*x2`                |
+| `y ~ x1*x2+ (x1 | g1)+ (x2| g1)`                            | variance-covariance matrix estimated separately, |
+|                                                             | i.e, one for intercept and `x1` and another for    |
+|                                                             | intercept and `x2`                                 |
+| `y ~ x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)`             | Fixed `x1` with correlated random intercept and    |
+|                                                             | random slope of `x`                                |
+| `y ~ x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1)` | Fixed `x1` with uncorrelated random intercept      |
+|                                                             | and random slope of `x1`                           |
+| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2   |
+| `y ~ (1 | g1/g2) or ~ (1 | g1) + (1 | g1:g2)`                     | Random intercept varying among `g1` and `g2`         |
+|                                                             | withing `g1`                                       |
+|                                                             | [TODO: explain \pkg{lme4} notation here          |
 
@@
 fits %>% left_join(n, by = "school") %>% 
-  ggplot(aes(y = school, x = estimate, size = n)) +
-  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
-  geom_linerange(aes(y = school, xmin = conf.low, xmax = conf.high, size = 0.5), position = position_dodge(width = 0.5)) +
+  ggplot(aes(y = school, x = estimate, linewidth = n)) +
+  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
+  geom_linerange(aes(y = school, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
   geom_point(shape = 20, position = position_dodge(width = 0.5)) +
@@
 fits %>% left_join(n, by = "name") %>% 
-  ggplot(aes(y = name, x = estimate, color = term, size = n)) +
-  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
-  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, size = 0.5), position = position_dodge(width = 0.5)) +
+  ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
+  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
+  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
   geom_point(shape = 20, position = position_dodge(width = 0.5)) +

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
update

@@
 \author{
-Hanne I. Oberman\\Utrecht University \And Johanna Muñoz\\Julius Center
-for Health Sciences and Primary Care,\\
+Hanne I. Oberman\\Methodology and Statistics\\
+Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
+and Primary Care,\\
 University Medical Center Utrecht, Utrecht University,\\
 Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
-for Health Sciences and Primary Care \And Gerko Vink\\Utrecht University
-\AND Valentijn M. T. de Jong\\Julius Center for Health Sciences and
-Primary Care,\\
+for Health Sciences and Primary Care \And Gerko Vink\\Methodology and
+Statistics\\
+Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
+Health Sciences and Primary Care,\\
 University Medical Center Utrecht, Utrecht University,\\
@@
     Hanne I. Oberman\\
+    Methodology and Statistics\\
 Utrecht University\\
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 
@@
 
-\begin{table}[tb]
-\caption{Concepts in multilevel methods}
-\label{tab:clus}
-\centering
-\begin{tabular}{ll}
-\hline
-\textbf{Concept} & \textbf{Details}   \\
-\hline
-Sample unit         & Units of the population from which measurements are taken in a sample.\\                           & These should be distinct and not overlapping, e.g., patients, students.\\ 
-Hierarchical levels & Data are grouped into clusters at different levels.
-A three-level \\                               & hierarchical model could occur by collecting observations of math scores\\                         & (lower level) which can be measured over the whole school year for each\\                          & student, and then grouped by student (middle level) and these in turn can\\                        & be aggregated into classrooms (high level).\\
-Fixed effect        & Here we assume that the values of an independent variable are fixed, i.e., \\ 
-  & the values observed in the study are representative of all values in the \\                        & population we are interested in.
-In the case of a fixed factor variable, for \\                     & example treatment type, we are only concerned with assessing the difference \\
-                & in the dependent variable y e.g., blood pressure between treatments A and B.\\                     & There may be additional treatments, but based on the objective of the study,\\                     & which in this case is the equivalence effect of A and B, the other treatment\\                     & types are irrelevant and therefore the treatment type is included as a fixed \\                     & effect with only the A and B levels in the hierarchical model.
-\\
-Random effect       & The values of an independent variable are assumed to be randomly drawn from \\                     & a population with more values.
-In a random factor, the levels observed in the\\                     & study represent a sample of all possible levels, e.g., hospitals, classrooms.\\                     & For example, in a study to predict the blood pressure of patients on \\
-& admission we might select only certain hospitals that are representative of \\                     & the entire population of hospitals. Here we are not interested in measuring \\ 
-& the difference of y between individual hospitals, but rather the variation of\\                     & y between hospitals, to create a model that can be generalized to the entire\\                     & population of hospitals.
-\\
-
-ICC                 & The variability due to clustering is often measured by means of the \\
-                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
-                    & of variance that can be attributed to the cluster-level, where a high \\
-                    & ICC would indicate that a lot of variability is due to the cluster \\
-                    & structure. \\
-Random effect       & Multilevel models typically accommodate for variability by including \\
-                    & a separate group mean for each cluster. In addition to random \\
-                    & intercepts, multilevel models can also include random coefficients \\
-                    & and heterogeneous residual error variances across clusters [see e.g. \\
-                    & @gelm06, @hox17 and @jong21]. [TODO: add stratification.] \\
-\hline
-\end{tabular}
-\end{table}
+Table 1: Concepts in multilevel methods \textbar{} \textbf{Concept}
+\textbar{} \textbf{Details} \textbar{}
+\textbar---------------------\textbar-------------------------------------------------------------------------------\textbar{}
+\textbar{} Sample unit \textbar{} Units of the population from which
+measurements are taken in a sample. \textbar{} \textbar{} Hierarchical
+levels \textbar{} Data are grouped into clusters at different levels. A
+three-level \textbar{} \textbar{} Fixed effect \textbar{} Here we assume
+that the values of an independent variable are fixed, i.e., \textbar{}
+\textbar{} \textbar{} the values observed in the study are
+representative of all values in the \textbar{} \textbar{} \textbar{} in
+the dependent variable y e.g., blood pressure between treatments A and
+B. \textbar{} \textbar{} Random effect \textbar{} The values of an
+independent variable are assumed to be randomly drawn from \textbar{}
+\textbar{} \textbar{} admission we might select only certain hospitals
+that are representative of \textbar{} \textbar{} \textbar{} the
+difference of y between individual hospitals, but rather the variation
+of \textbar{} \textbar{} ICC \textbar{} The variability due to
+clustering is often measured by means of the \textbar{} \textbar{}
+\textbar{} intraclass coefficient (ICC). The ICC can be seen as the
+percentage \textbar{} \textbar{} \textbar{} of variance that can be
+attributed to the cluster-level, where a high \textbar{} \textbar{}
+\textbar{} ICC would indicate that a lot of variability is due to the
+cluster \textbar{} \textbar{} \textbar{} structure. \textbar{}
+\textbar{} Random effect \textbar{} Multilevel models typically
+accommodate for variability by including \textbar{} \textbar{}
+\textbar{} a separate group mean for each cluster. In addition to random
+\textbar{} \textbar{} \textbar{} intercepts, multilevel models can also
+include random coefficients \textbar{} \textbar{} \textbar{} and
+heterogeneous residual error variances across clusters \citep[see
+e.g.~\textbar{} \textbar{} \textbar{}][\citet{hox17} and
+\citet{jong21}]{gelm06}. {[}TODO: add stratification.{]} \textbar{}
 
@@
 
-\begin{table}[tb]
-\caption{Concepts in missing data methods}
-\label{tab:miss}
-\centering
-\begin{tabular}{ll}
-\hline
-\textbf{Concept} & \textbf{Details}   \\
-\hline
-MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
-& across all data entries \\
-MAR     & Missing At Random, where the probability to be missing depends on observed \\
-& information \\
-MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
-& depends on unrecorded information, making the missingness non-ignorable \\
-& [@rubi76; @meng94]. \\
-& [TODO: add congeniality, but maybe in-text?] \\
-\hline
-\end{tabular}
-\end{table}
+Table 2: Concepts in missing data methods \textbar{} \textbf{Concept}
+\textbar{} \textbf{Details} \textbar{}
+\textbar----------------\textbar----------------------------------------------------------------------------\textbar{}
+\textbar{} MCAR \textbar{} Missing Completely At Random, where the
+probability to be missing is equal \textbar{} \textbar{} \textbar{}
+across all data entries \textbar{} \textbar{} MAR \textbar{} Missing At
+Random, where the probability to be missing depends on observed
+\textbar{} \textbar{} \textbar{} information \textbar{} \textbar{} MNAR
+\textbar{} Missing Not At Random (MNAR), where the probability to be
+missing \textbar{} \textbar{} \textbar{} depends on unrecorded
+information, making the missingness non-ignorable \textbar{} \textbar{}
+\textbar{} \citep{rubi76, meng94}. \textbar{} \textbar{} \textbar{}
+{[}TODO: add congeniality, but maybe in-text?{]} \textbar{}
 
@@
 
-\begin{table}[tb]
-\caption{Notation}
-\label{tab:mod}
-\centering
-\begin{tabular}{lll}
-\hline
-\textbf{Formula lme4} & \textbf{Details}   \\
-\hline
-y$\sim$ x1 + (1 | g1)                                       & Fixed x1 predictor with random intercept  \\
-                                                         & varying among g\\
-y$\sim$x1*x2+ (1| g1)                                      & Interactions of x1 and x2 only in fixed effect\\
-y$\sim$x1*x2+ (x2| g1)                                     & Interactions of x1 and x2 only in fixed effect\\
-                                                         & with slope of x2 randomly varying among g1\\
-y$\sim$x1*x2+ (x1*x2| g1)                                  & variance-covariance matrix estimated only with \\
-                                                         & the variance terms of intercept, slope of x1, \\
-                                                         & slope of x2 and interaction x1*x2\\
-y$\sim$x1*x2+ (x1| g1)+ (x2| g1)                           & variance-covariance matrix estimated separately,\\
-                                                         & i.e, one for intercept and x1 and another for \\
-                                                         & intercept and x2\\
-y$\sim$x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)            & Fixed x1 with correlated random intercept and \\
-                                                         & random slope of x\\
-y$\sim$x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1) & Fixed x1 with uncorrelated random intercept\\
-                                                         & and random slope of x1\\  
-y$\sim$(1 | g1) + (1 | g2)                                 & Random intercept varying among g1 and among g2\\
-y$\sim$(1 | g1/g2) or ~(1|g1)+(1|g1:g2)                    & Random intercept varying among g1 and g2 
-                                                         & withing g1\\
-
-& [TODO: explain \pkg{lme4} notation here
-https://datalorax.github.io/equatiomatic/articles/lmer4-lmer.html
-https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf
-] \\
-\hline
-\end{tabular}
-\end{table}
+Table 3: Notation
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Formula \pkg{lme4}}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Details}
+\end{minipage} \\
+\midrule()
+\endhead
+\texttt{y\ \textasciitilde{}\ x1\ +\ (1\ \textbar{}\ g1)} & Fixed
+\texttt{x1} predictor with random intercept \\
+& varying among \texttt{g1} \\
+\texttt{y\textasciitilde{}x1*x2+\ (1\textbar{}\ g1)} & Interactions of
+\texttt{x1} and \texttt{x2} only in fixed effect \\
+\texttt{y\ \textasciitilde{}\ x1*x2+\ (x2\textbar{}\ g1)} & Interactions
+of \texttt{x1} and \texttt{x2} only in fixed effect \\
+& with slope of \texttt{x2} randomly varying among \texttt{g1} \\
+\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1*x2\textbar{}\ g1)} &
+variance-covariance matrix estimated only with \\
+& the variance terms of intercept, slope of \texttt{x1}, \\
+& slope of \texttt{x2} and interaction \texttt{x1*x2} \\
+\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1\ \textbar{}\ g1)+\ (x2\textbar{}\ g1)}
+& variance-covariance matrix estimated separately, \\
+& i.e, one for intercept and \texttt{x1} and another for \\
+& intercept and \texttt{x2} \\
+\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ +\ x1\ \textbar{}\ g1)}
+& Fixed \texttt{x1} with correlated random intercept and \\
+& random slope of \texttt{x} \\
+\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ \textbar{}\ g1)\ +\ (0\ +\ x1\ \textbar{}\ g1)}
+& Fixed \texttt{x1} with uncorrelated random intercept \\
+& and random slope of \texttt{x1} \\
+\texttt{y\ \textasciitilde{}\ (1\ \textbar{}\ g1)\ +\ (1\ \textbar{}\ g2)}
+& Random intercept varying among \texttt{g1} and among
+\texttt{g2\ \ \ \textbar{}\ \textbar{}}y \textasciitilde{} (1 \\
+& {[}TODO: explain \pkg{lme4} notation here \\
+\bottomrule()
+\end{longtable}
 

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
correct table caption spacing

@@
 Table 1: Concepts in multilevel methods
+
 | **Concept**         | **Details**                                                                   |
@@
 Table 2: Concepts in missing data methods
+
 | **Concept**    | **Details**                                                                |

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

@@
 
-Table 1: Concepts in multilevel methods \textbar{} \textbf{Concept}
-\textbar{} \textbf{Details} \textbar{}
-\textbar---------------------\textbar-------------------------------------------------------------------------------\textbar{}
-\textbar{} Sample unit \textbar{} Units of the population from which
-measurements are taken in a sample. \textbar{} \textbar{} Hierarchical
-levels \textbar{} Data are grouped into clusters at different levels. A
-three-level \textbar{} \textbar{} Fixed effect \textbar{} Here we assume
-that the values of an independent variable are fixed, i.e., \textbar{}
-\textbar{} \textbar{} the values observed in the study are
-representative of all values in the \textbar{} \textbar{} \textbar{} in
-the dependent variable y e.g., blood pressure between treatments A and
-B. \textbar{} \textbar{} Random effect \textbar{} The values of an
-independent variable are assumed to be randomly drawn from \textbar{}
-\textbar{} \textbar{} admission we might select only certain hospitals
-that are representative of \textbar{} \textbar{} \textbar{} the
-difference of y between individual hospitals, but rather the variation
-of \textbar{} \textbar{} ICC \textbar{} The variability due to
-clustering is often measured by means of the \textbar{} \textbar{}
-\textbar{} intraclass coefficient (ICC). The ICC can be seen as the
-percentage \textbar{} \textbar{} \textbar{} of variance that can be
-attributed to the cluster-level, where a high \textbar{} \textbar{}
-\textbar{} ICC would indicate that a lot of variability is due to the
-cluster \textbar{} \textbar{} \textbar{} structure. \textbar{}
-\textbar{} Random effect \textbar{} Multilevel models typically
-accommodate for variability by including \textbar{} \textbar{}
-\textbar{} a separate group mean for each cluster. In addition to random
-\textbar{} \textbar{} \textbar{} intercepts, multilevel models can also
-include random coefficients \textbar{} \textbar{} \textbar{} and
-heterogeneous residual error variances across clusters \citep[see
-e.g.~\textbar{} \textbar{} \textbar{}][\citet{hox17} and
-\citet{jong21}]{gelm06}. {[}TODO: add stratification.{]} \textbar{}
+Table 1: Concepts in multilevel methods
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Concept}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Details}
+\end{minipage} \\
+\midrule()
+\endhead
+Sample unit & Units of the population from which measurements are taken
+in a sample. \\
+Hierarchical levels & Data are grouped into clusters at different
+levels. A three-level \\
+Fixed effect & Here we assume that the values of an independent variable
+are fixed, i.e., \\
+& the values observed in the study are representative of all values in
+the \\
+& in the dependent variable y e.g., blood pressure between treatments A
+and B. \\
+Random effect & The values of an independent variable are assumed to be
+randomly drawn from \\
+& admission we might select only certain hospitals that are
+representative of \\
+& the difference of y between individual hospitals, but rather the
+variation of \\
+ICC & The variability due to clustering is often measured by means of
+the \\
+& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
+& of variance that can be attributed to the cluster-level, where a
+high \\
+& ICC would indicate that a lot of variability is due to the cluster \\
+& structure. \\
+Random effect & Multilevel models typically accommodate for variability
+by including \\
+& a separate group mean for each cluster. In addition to random \\
+& intercepts, multilevel models can also include random coefficients \\
+& and heterogeneous residual error variances across clusters {[}see
+e.g. \\
+& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
+stratification.{]} \\
+\bottomrule()
+\end{longtable}
 
@@
 
-Table 2: Concepts in missing data methods \textbar{} \textbf{Concept}
-\textbar{} \textbf{Details} \textbar{}
-\textbar----------------\textbar----------------------------------------------------------------------------\textbar{}
-\textbar{} MCAR \textbar{} Missing Completely At Random, where the
-probability to be missing is equal \textbar{} \textbar{} \textbar{}
-across all data entries \textbar{} \textbar{} MAR \textbar{} Missing At
-Random, where the probability to be missing depends on observed
-\textbar{} \textbar{} \textbar{} information \textbar{} \textbar{} MNAR
-\textbar{} Missing Not At Random (MNAR), where the probability to be
-missing \textbar{} \textbar{} \textbar{} depends on unrecorded
-information, making the missingness non-ignorable \textbar{} \textbar{}
-\textbar{} \citep{rubi76, meng94}. \textbar{} \textbar{} \textbar{}
-{[}TODO: add congeniality, but maybe in-text?{]} \textbar{}
+Table 2: Concepts in missing data methods
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Concept}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Details}
+\end{minipage} \\
+\midrule()
+\endhead
+MCAR & Missing Completely At Random, where the probability to be missing
+is equal \\
+& across all data entries \\
+MAR & Missing At Random, where the probability to be missing depends on
+observed \\
+& information \\
+MNAR & Missing Not At Random (MNAR), where the probability to be
+missing \\
+& depends on unrecorded information, making the missingness
+non-ignorable \\
+& \citep{rubi76, meng94}. \\
+& {[}TODO: add congeniality, but maybe in-text?{]} \\
+\bottomrule()
+\end{longtable}
 

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
changes GV

@@
 
-Many datasets include individuals from multiple settings, geographic regions, or even different studies.
-In the simplest case, individuals (e.g., students) are nested within so-called clusters (e.g., school classes).
-More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases).
-In general, individuals from the same cluster tend to be more similar than individuals from other clusters.
-In statistical terms, this implies that observations from the same cluster are correlated.
+Many datasets include individuals that are clustered together, for example in geographic regions, or even different studies.
+In the simplest case, individuals (e.g., students) are nested within a single cluster (e.g., school classes).
+More complex clustered structures may occur when there are multiple hierarchical levels (e.g., students in different schools or patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases).
+With clustered data we generally assume that individuals from the same cluster tend to be more similar than individuals from other clusters.
+In statistical terms, this implies that observations from the same cluster are not independent and may in fact be correlated.
 If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01].
@@
 <!-- Maybe besides CC and imputation name also likelihood based methods -->
-Like any other dataset, clustered datasets are prone to missing data.
+As with any other dataset, clustered datasets may be impacted by missingness in much the same way.
 Several strategies can be used to handle missing data, including complete case analysis and imputation.
 We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values.
-Afterwards, the completed data can be analyzed as if they were completely observed.
-In contrast to single imputation (where missing data are only replaced once), multiple imputation allows to preserve uncertainty due to missingness and is therefore recommended (c.f.
+Imputation seperates the missing data problem from the analysis and the completed data can be analyzed as if it were completely observed.
+It is generally recommended to impute the missing values more than once to preserve uncertainty due to missingness and to allow for valid inferences (c.f.
 Rubin 1976).
 
-When clustered datasets are affected by missing values, we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13].
+With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13].
 Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18].
 For example, it is possible that test results are missing for several students in one or more classes.
-[TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are systematically missing.
+[TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are said to be systematically missing.
 [TODO: Refer to Figure 1 and put interpretation in the figure caption.
@@
 Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}).
-For each of these three missingness generating mechanisms, different imputation strategies are warranted @yuce08 and @hox15.
+For each of these three missingness generating mechanisms, different imputation strategies are warranted (@yuce08 and @hox15).
 We here consider the general case that data are MAR, and expand on certain MNAR situations.
@@
 
-The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
-\pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
-However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent whereas multilevel data are dependent.
+[GV: DELETE THIS SECTION?]However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent whereas multilevel data are dependent.
 For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.
@@
 This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice} in \proglang{R}.
+\pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
+\pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
+
 We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms.
@@
 ## Setup
+* GV: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB?
 
@@
 ```{r env, message=FALSE, warning=FALSE}
-set.seed(2022)
+set.seed(2022)        # for reproducibility
 library(mice)         # for imputation
-library(miceadds)     # for imputation
-library(ggmice)       # for visualization
+library(miceadds)     # for additional imputation routines
+library(ggmice)       # for incomplete/imputed data visualization
 library(ggplot2)      # for visualization
@@
 library(lme4)         # for multilevel modeling
-library(mitml)        # for multilevel pooling
-library(miceheckman)  # for imputation
+library(mitml)        # for multilevel parameter pooling
+library(miceheckman)  # for imputation cf heckman models
 ```

### 2022-12-16 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Add `hiceheckman` package

--- 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 - `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies with data that are MAR, @metamisc];
-- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions with data that are MNAR, @GJRM].
+- `obesity` from the \pkg{miceheckman} package [simulated data on obesity, $n = 2,111$ patients across $N = 5$ regions with data that are MNAR].
 
@@
 ## Setup
+
+Install non-CRAN packages:
+```{r packages}
+devtools::install_github("amices/ggmice")
+devtools::install_github("hanneoberman/miceheckman")
+```
 
@@
 
-# Case study III: HIV data
+# Case study III: obesity data
 
-Data are simulated and included in the `GJRM` package. We will use the following variables:
+Data are simulated and included in the `miceheckman` package. We will use the following variables:
 
 - `region` Cluster variable,
-- `hiv` HIV diagnosis (0=no, 1=yes),
+- `gender` Gender (0=male, 1=female),
 - `age` Age of the patient,
-- `marital` Marital status,
-- `condom` Condom use during last intercourse,
-- `smoke` Smoker (levels; inclusion restriction variable).
+- `height` Height in meters,
+- `weight` Weight in kilograms,
+- `rt`` Response time in minutes (inclusion-restriction variable).
 
-The imputation of these date is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
-
-```{r hiv, echo=FALSE}
-# data
-data("hiv", package = "GJRM")
-hiv <- hiv %>% 
-  select(c("region","hiv", "age", "marital", "condom", "smoke", "interviewerID")) %>% 
-  mutate(region = as.integer(region))
-# md pattern
-plot_pattern(hiv)
-```
-
-R=region+language+(1|InterviewID) model with interviewer as random effects, because the observations are not independent.
-Interviews are not allocated randomly.
-In theory we expect the inclusion-restriction variable to be randomly assigned, that's why we're adding region and language to compensate for non-random allocation.
-[TODO: Explain exclusion-restriction, refer to paper Johanna].
+The imputation of these data is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R) [TODO: add reference Arxiv paper Johanna].
 
 Load the data:
-```{r}
+```{r obesity}
 data("obesity", package = "miceheckman")
@@
 
-We obtain here the random effects for each interviewer, this is an approximation of the interviewer's skill which will be used as an exclusion constraint.
-Here, since the location of the interviewer was not randomly assigned to the subjects, the assignment was corrected for region and language.
-
-Set the Heckman model as imputation method
-```{r}
-#Set prediction matrix and methods
-ini <- mice(obesity, maxit = 0)
-meth<-ini$method
-meth["weight"]<-"2l.heckman"
-pred <- ini$pred
-pred["weight","cluster"]<- -2
-pred["weight","rt"] <- -3
+Select relevant variables:
+```{r bmi}
+dat <- select(obesity, -"bmi")
 ```
 
+Visualize missing data pattern:
+```{r obesity-md}
+plot_pattern(dat)
+```
+
+Create predictor matrix:
+```{r obesity-pred}
+pred <- quickpred(dat)   # predictor matrix
+pred["weight","cluster"]<- -2 # clustering variable
+pred["weight","rt"] <- -3 #inclusion-restriction variable
+plot_pred(pred)
+```
+
+Set the Heckman model as imputation method:
+```{r meth}
+meth <- make.method(dat) # methods vector
+meth["weight"]<-"2l.heckman"
+```
+
+Impute the missingness:
 ```{r eval=FALSE}
-imp <- mice(obesity, # dataset with missing values
-                        m = 10,   # number of imputations
-                        maxit = 1,
-                        seed = 1234, #seed attached to the dataID
+imp <- mice(
+  dat, # dataset with missing values
+  m = 5, # number of imputations
+  maxit = 10,
   meth = meth, #imputation method vector
   pred = pred, #imputation predictors matrix
-                        print = T,
-                        meta_method="reml",
-                        pmm=FALSE)
+  meta_method = "reml",
+  printFlag = FALSE
+)
 ```

### 2022-12-16 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

--- 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 * GV: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB?
+
+Install non-CRAN packages:
+```{r packages}
+devtools::install_github("amices/ggmice")
+devtools::install_github("hanneoberman/miceheckman")
+```
 

### 2022-12-16 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

@@
 
-Many datasets include individuals from multiple settings, geographic
-regions, or even different studies. In the simplest case, individuals
-(e.g., students) are nested within so-called clusters (e.g., school
-classes). More complex clustered structures may occur when there are
-multiple hierarchical levels (e.g., patients within hospitals within
-regions or countries), or when the clustering is non-nested (e.g.,
-electronic health record data from diverse settings and populations
-within large databases). In general, individuals from the same cluster
-tend to be more similar than individuals from other clusters. In
-statistical terms, this implies that observations from the same cluster
-are correlated. If this correlation is left unaddressed, estimates of
-\emph{p} values, confidence intervals even model parameters are prone to
-bias \citep{loca01}. {[}TODO: make a link to imputation methods, which
+Many datasets include individuals that are clustered together, for
+example in geographic regions, or even different studies. In the
+simplest case, individuals (e.g., students) are nested within a single
+cluster (e.g., school classes). More complex clustered structures may
+occur when there are multiple hierarchical levels (e.g., students in
+different schools or patients within hospitals within regions across
+countries), or when the clustering is non-nested (e.g., electronic
+health record data from diverse settings and populations within large
+databases). With clustered data we generally assume that individuals
+from the same cluster tend to be more similar than individuals from
+other clusters. In statistical terms, this implies that observations
+from the same cluster are not independent and may in fact be correlated.
+If this correlation is left unaddressed, estimates of \emph{p} values,
+confidence intervals even model parameters are prone to bias
+\citep{loca01}. {[}TODO: make a link to imputation methods, which
 require adequate handling and propagation of variance; we are not
@@
 
-Like any other dataset, clustered datasets are prone to missing data.
-Several strategies can be used to handle missing data, including
-complete case analysis and imputation. We focus on the latter approach
-and discuss statistical methods for replacing the missing data with one
-or more plausible values. Afterwards, the completed data can be analyzed
-as if they were completely observed. In contrast to single imputation
-(where missing data are only replaced once), multiple imputation allows
-to preserve uncertainty due to missingness and is therefore recommended
-(c.f. Rubin 1976).
-
-When clustered datasets are affected by missing values, we can
-distinguish between two types of missing data: sporadic missingness and
-systematic missingness \citep{resc13}. Sporadic missingness arises when
-variables are missing for some but not all of the units in a cluster
-\citep{buur18, jola18}. For example, it is possible that test results
-are missing for several students in one or more classes. {[}TODO:
-Provide an example for one of the case studies below.{]} When all
-observations are missing within one or more clusters, data are
-systematically missing. {[}TODO: Refer to Figure 1 and put
-interpretation in the figure caption. VMTdJ: ``Mention in text that this
-is sporadic missingness?''{]}
+As with any other dataset, clustered datasets may be impacted by
+missingness in much the same way. Several strategies can be used to
+handle missing data, including complete case analysis and imputation. We
+focus on the latter approach and discuss statistical methods for
+replacing the missing data with one or more plausible values. Imputation
+seperates the missing data problem from the analysis and the completed
+data can be analyzed as if it were completely observed. It is generally
+recommended to impute the missing values more than once to preserve
+uncertainty due to missingness and to allow for valid inferences (c.f.
+Rubin 1976).
+
+With incomplete clustered datasets we can distinguish between two types
+of missing data: sporadic missingness and systematic missingness
+\citep{resc13}. Sporadic missingness arises when variables are missing
+for some but not all of the units in a cluster \citep{buur18, jola18}.
+For example, it is possible that test results are missing for several
+students in one or more classes. {[}TODO: Provide an example for one of
+the case studies below.{]} When all observations are missing within one
+or more clusters, data are said to be systematically missing. {[}TODO:
+Refer to Figure 1 and put interpretation in the figure caption. VMTdJ:
+``Mention in text that this is sporadic missingness?''{]}
 
@@
 \ref{tab:miss}). For each of these three missingness generating
-mechanisms, different imputation strategies are warranted \citet{yuce08}
-and \citet{hox15}. We here consider the general case that data are MAR,
-and expand on certain MNAR situations.
+mechanisms, different imputation strategies are warranted
+(\citet{yuce08} and \citet{hox15}). We here consider the general case
+that data are MAR, and expand on certain MNAR situations.
 
@@
 
-The \proglang{R} package \pkg{mice} has become the de-facto standard for
-imputation by chained equations, which iteratively solves the
-missingness on a variable-by-variable basis. \pkg{mice} is known to
-yield valid inferences under many different missing data circumstances
-\citep{buur18}. However, commonly used imputation methods were not
-designed for use in clustered data and usually generate observations
-that are independent whereas multilevel data are dependent. For this
-reason, we discuss how the \proglang{R} package \pkg{mice} can be used
-to impute multilevel data.
+{[}GV: DELETE THIS SECTION?{]}However, commonly used imputation methods
+were not designed for use in clustered data and usually generate
+observations that are independent whereas multilevel data are dependent.
+For this reason, we discuss how the \proglang{R} package \pkg{mice} can
+be used to impute multilevel data.
 
@@
 This papers serves as a tutorial for imputing incomplete multilevel data
-with \pkg{mice} in \proglang{R}. We provide practical guidelines and
-code snippets for different missing data situations, including
-non-ignorable mechanisms. For reasons of brevity, we focus on multilevel
-imputation by chained equations with \pkg{mice} exclusively; other
-imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are
-outside the scope of this tutorial. {[}TODO: change this sentence
-because we do not assume familiarity: ``Assumed knowledge includes basic
-familiarity with multilevel imputation \citep[see e.g.][ and
-\citet{grun18}]{audi18} and the \pkg{lme4} notation for multilevel
-models (see Table \ref{tab:mod})''{]}.
+with \pkg{mice} in \proglang{R}. \pkg{mice} has become the de-facto
+standard for imputation by chained equations, which iteratively solves
+the missingness on a variable-by-variable basis. \pkg{mice} is known to
+yield valid inferences under many different missing data circumstances
+\citep{buur18}.
+
+We provide practical guidelines and code snippets for different missing
+data situations, including non-ignorable mechanisms. For reasons of
+brevity, we focus on multilevel imputation by chained equations with
+\pkg{mice} exclusively; other imputation methods and packages (e.g.,
+\pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial.
+{[}TODO: change this sentence because we do not assume familiarity:
+``Assumed knowledge includes basic familiarity with multilevel
+imputation \citep[see e.g.][ and \citet{grun18}]{audi18} and the
+\pkg{lme4} notation for multilevel models (see Table
+\ref{tab:mod})''{]}.
 
@@
 
+\begin{itemize}
+\tightlist
+\item
+  GV: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB?
+\end{itemize}
+
 {[}TODO: Add environment info, seed and version number(s) somewhere.{]}
@@
 \begin{CodeInput}
-R> set.seed(2022)
+R> set.seed(2022)        # for reproducibility
 R> library(mice)         # for imputation
-R> library(miceadds)     # for imputation
-R> library(ggmice)       # for visualization
+R> library(miceadds)     # for additional imputation routines
+R> library(ggmice)       # for incomplete/imputed data visualization
 R> library(ggplot2)      # for visualization
@@
 R> library(lme4)         # for multilevel modeling
-R> library(mitml)        # for multilevel pooling
-R> library(miceheckman)  # for imputation
+R> library(mitml)        # for multilevel parameter pooling
+R> library(miceheckman)  # for imputation cf heckman models
 \end{CodeInput}

### 2022-12-16 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Resolve TODOs in manuscript

@@
   - name: Thomas P. A. Debray
-    affiliation: Julius Center for Health Sciences and Primary Care
-    address: |
+    affiliation: |
+      | Julius Center for Health Sciences and Primary Care, 
       | University Medical Center Utrecht, Utrecht University, 
@@
   This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}.
-Footnotes in the current version show work in progress/under construction.
-The last section is not part of the manuscript, but purely for reminders.
-See also all of the TODOs that need to be worked out.
 We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
-[Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
 keywords:
@@
 
-<!-- Missing data pattern with clustering -->
-<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
-<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
-<!-- TODO: check spss missing data pie chart? -->
-<!-- TODO: similarity of patterns over clusters -->
-<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
-<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
-<!-- TODO: look at vim -->
-<!-- TODO: check if shepley plot is an option -->
-
-<!-- Conditional plots missingness indicator -->
-<!-- TODO: check if up-side-down plot works -->
-<!-- TODO: check smoothing in geom_density function and make it the inverse of the sample size -->
-<!-- TODO: add functions to mice -->
-<!-- TODO: add title and informative legend (pop obs and pop miss) -->
-<!-- TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .` -->
-<!-- TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon -->
-<!-- TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?) -->
-
 # Introduction
@@
 If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01].
-[TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation.
-VMTdJ: "I would only make that link after introducing missing data, and missing data is not mentioned yet, so I would do that in the next section"] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
+Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
 These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'.
@@
 |                     | and heterogeneous residual error variances across clusters [see e.g.          |
-|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratification.]                     |
+|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |
 
@@
 
-<!-- Maybe besides CC and imputation name also likelihood based methods -->
 As with any other dataset, clustered datasets may be impacted by missingness in much the same way.
@@
 We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values.
-Imputation seperates the missing data problem from the analysis and the completed data can be analyzed as if it were completely observed.
+Imputation separates the missing data problem from the analysis and the completed data can be analyzed as if it were completely observed.
 It is generally recommended to impute the missing values more than once to preserve uncertainty due to missingness and to allow for valid inferences (c.f.
@@
 For example, it is possible that test results are missing for several students in one or more classes.
-[TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are said to be systematically missing.
-[TODO: Refer to Figure 1 and put interpretation in the figure caption.
-VMTdJ: "Mention in text that this is sporadic missingness?"]
+When all observations are missing within one or more clusters, data are said to be systematically missing.
+<!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"] -->
+<!-- TODO: Provide an example for one of the case studies below.] -->
+<!-- TODO: add that indicator method doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->
 
@@
 
-```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
+```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE}
 dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
@@
 |                | [@rubi76; @meng94].                                                        |
-|                | [TODO: add congeniality, but maybe in-text?]                               |
+|                |                              |
+
+<!-- [TODO: add paragraph about congeniality and smc-fcs!]  -->
 
@@
 
-[GV: DELETE THIS SECTION?]However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent whereas multilevel data are dependent.
-For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.
-
-[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.]
+<!-- [TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] -->
+<!-- [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.] -->
 
@@
 We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms.
-For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial.
-[TODO: change this sentence because we do not assume familiarity: "Assumed knowledge includes basic familiarity with multilevel imputation [see e.g.
-@audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod})"].
+For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages [see e.g.
+@audi18 and @grun18] are outside the scope of this tutorial.
+Assumed knowledge includes basic familiarity with the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).
+
+<!-- TODO: mention other packages? \pkg{jomo} and \pkg{mdmb} -->
 
@@
 |                                                             | withing `g1`                                       |
-|                                                             | [TODO: explain \pkg{lme4} notation here          |
+|                                                             |          |
 
@@
 This tutorial is dedicated to readers who are unfamiliar with multiple imputation.
-More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study IMPACT) or under MNAR conditions (case study HIV).
+More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2) or under MNAR conditions (case study 3).
 
@@
 
-TODO: explicit statement about not going into workings of the methods. Galimer 2l methods.
+<!-- TODO: explicit statement about not going into workings of the methods. Galimer 2l methods. -->
 
 ## Setup
-* GV: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB?
+<!-- * TODO: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB? -->
 
-Install non-CRAN packages:
-```{r packages}
+Install non-CRAN packages if necessary:
+```{r packages, eval=FALSE, include=TRUE}
 devtools::install_github("amices/ggmice")
@@
 
-[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
+<!-- [TODO: Add environment info, seed and version number(s) somewhere.]  -->
+
+Set up the R environment and load the necessary packages:
 ```{r env, message=FALSE, warning=FALSE}
@@
 
-[TODO: explain case study]
+<!-- [TODO: explain case study] -->
 
@@
 ```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
-plot_pattern(popmis)
+plot_pattern(popmis, square = TRUE)
 ```
@@
 
-[TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.]
+<!-- [TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.] -->
 
@@
 
-[TODO: check if there is systematic missingness in this dataset, if not make Marshall Computerized Tomography classification (ct) systematically missing.]
+<!-- [TODO: check if there is systematic missingness in this dataset, if not make Marshall Computerized Tomography classification (ct) systematically missing.] -->
 
 We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc].
-[TODO: add more info about the complete data.] The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:
+<!-- [TODO: add more info about the complete data.]  -->
+The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:
 
@@
   - `pupil` Pupillary reactivity,
-  - `ct` Marshall Computerized Tomography classification, [TODO: make this one var? also shows that you don't always need random effects everywhere?]
+  - `ct` Marshall Computerized Tomography classification, 
   - `hypox` Hypoxia (0=no, 1=yes),
@@
   - `mort` 6-month mortality (0=alive, 1=dead).
+
+<!-- [TODO: make ct levels one var? also shows that you don't always need random effects everywhere?] -->
 
@@
 
-[TODO: show how much variance there is after different methods]
+<!-- [TODO: show how much variance there is after different methods] -->
 
-[TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account.
-If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.]
+<!-- [TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account.
+If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.] -->
 
-[TODO: add descriptive statistics of the complete and incomplete data.]
+<!-- [TODO: add descriptive statistics of the complete and incomplete data.] -->
 
@@
 
-To explore the missingness, it is wise to look at the missing data pattern:
+To explore the missingness, it is wise to look at the missing data pattern. The ten most frequent missingness patterns are shown:
 ```{r pattern, fig.height=7.1}
@@
 
-This shows... [TODO: fill in that we need to impute `ct` and `pupil`.]
+This shows that we need to impute `ct` and `pupil`.
 
 To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables.
-To see whether the missingness depends on the observed values of other variables, we...
-[TODO: fill in that we can test this statistically or use visual inspection (e.g.
-a histogram faceted by the missingness indicator).]
+To see whether the missingness depends on the observed values of other variables, we can test this statistically or use visual inspection (e.g.
+a histogram faceted by the missingness indicator).
 
@@
 
-## Complete case analysis [TODO: remove this?]
+## Complete case analysis
 
@@
 
-As we can see...
-[TODO: fill in: This means that a higher ct (Marshall Computerized Tomography classification) is associated with a lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI ...
-to ..., when controlling for ...]
+As we can see, a higher `ct` (Marshall Computerized Tomography classification) is associated with a lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI ...
+to ..., when controlling for...
+
+<!-- TODO: fill in results above -->
 
@@
 pred <- quickpred(dat)   # predictor matrix
-plot_pred(pred)
+plot_pred(pred, rotate = TRUE)
 
@@
 diag(pred) <- 0
-plot_pred(pred)
+plot_pred(pred, rotate = TRUE)
 
@@
 ```
-
-<!-- We'll now use `school` as a predictor to impute all other variables.
-This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
-But at least, it includes some multilevel aspect.
-This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16].
-Colloquially, this is 'multilevel imputation for dummies'.
--->
-
-<!-- Add: doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->
-
-<!-- The more the random effects are of interest, the more you need ml models. -->
 
@@
 
-The imputation of these data is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R) [TODO: add reference Arxiv paper Johanna].
+The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
+
+<!-- [TODO: add reference Arxiv paper Johanna].  -->
 
@@
 
-- JOMO in \pkg{mice} -> on the side for now
+<!-- - JOMO in \pkg{mice} -> on the side for now -->
 <!-- # look at jomo for categorical variables? -->
@@
 
-# Think about
+<!-- # Think about -->
 
-- Adding evaluations of the imputations such as convergence checks
+<!-- - Adding evaluations of the imputations such as convergence checks -->
 
-- Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model.
+<!-- TODO in the future: Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model. -->
 
-- Adding a `multilevel_ampute()` wrapper function in mice.
+<!-- TODO in the future: Adding a `multilevel_ampute()` wrapper function in mice. -->
 
-- Exporting `mids` objects to other packages like `lme4` or `coxme`?
+<!-- - Exporting `mids` objects to other packages like `lme4` or `coxme`?  -->
 
-- Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt.
+<!-- - Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt. -->
 
-- Show use case for deductive imputation for cluster level variables?
+<!-- - Show use case for deductive imputation for cluster level variables? -->
 
-- env dump in repo
+<!-- - env dump in repo -->
 
--  I don't know if in your article you cover something about model complexity, for example sometimes I have to switch from 2l.
+<!-- -  I don't know if in your article you cover something about model complexity, for example sometimes I have to switch from 2l.
 methods to 1l.
@@
 how to test which is better between 2l.norm or 2l.2stage.norm.
+-->
 
@@
 # References
+
+<!-- Comment dump -->
+<!-- Missing data pattern with clustering -->
+<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
+<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
+<!-- TODO: check spss missing data pie chart? -->
+<!-- TODO: similarity of patterns over clusters -->
+<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
+<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
+<!-- TODO: look at vim -->
+<!-- TODO: check if shepley plot is an option -->
+
+<!-- Conditional plots missingness indicator -->
+<!-- TODO: check if up-side-down plot works -->
+<!-- TODO: check smoothing in geom_density function and make it the inverse of the sample size -->
+<!-- TODO: add functions to mice -->
+<!-- TODO: add title and informative legend (pop obs and pop miss) -->
+<!-- TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .` -->
+<!-- TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon -->
+<!-- TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?) -->

### 2023-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update for re-start project

@@
 Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
-for Health Sciences and Primary Care \And Gerko Vink\\Methodology and
-Statistics\\
+for Health Sciences and Primary Care,\\
+University Medical Center Utrecht, Utrecht University,\\
+Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
 Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
@@
 This tutorial illustrates the imputation of incomplete multilevel data
-with the \proglang{R} packackage \pkg{mice}. Footnotes in the current
-version show work in progress/under construction. The last section is
-not part of the manuscript, but purely for reminders. See also all of
-the TODOs that need to be worked out. We aim to submit at JSS, so there
-is no word count limit (``There is no page limit, nor a limit on the
-number of figures or tables''). {[}Just adding some text to get a better
-guess of what the actura abstract will look like: Lorem ipsum dolor sit
-amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut
-labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud
-exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum
-dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non
-proident, sunt in culpa qui officia deserunt mollit anim id est
-laborum.{]}
+with the \proglang{R} packackage \pkg{mice}.
 }
@@
   URL: \url{https://hanneoberman.github.io/}\\~\\
-        Thomas P. A. Debray\\
-    Julius Center for Health Sciences and Primary Care\\
-    University Medical Center Utrecht, Utrecht University,\\
-Utrecht, The Netherlands\\
-
           }
@@
 confidence intervals even model parameters are prone to bias
-\citep{loca01}. {[}TODO: make a link to imputation methods, which
-require adequate handling and propagation of variance; we are not
-recommending the adoption of multilevel models for data analysis here,
-but rather for imputation. VMTdJ: ``I would only make that link after
-introducing missing data, and missing data is not mentioned yet, so I
-would do that in the next section''{]} Statistical methods for clustered
-data typically adopt hierarchical models that explicitly describe the
-grouping of observations. These models are also known as `multilevel
-models', `hierarchical models', `mixed effect models', `random effect
-models', and in the context of time-to-event data as `frailty models'.
-Table \ref{tab:clus} provides an overview of some key concepts in
-multilevel modeling.
+\citep{loca01}. Statistical methods for clustered data typically adopt
+hierarchical models that explicitly describe the grouping of
+observations. These models are also known as `multilevel models',
+`hierarchical models', `mixed effect models', `random effect models',
+and in the context of time-to-event data as `frailty models'. Table
+\ref{tab:clus} provides an overview of some key concepts in multilevel
+modeling.
 
@@
 & \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
-stratification.{]} \\
+stratified intercept as concept..{]} \\
 \bottomrule()
@@
 replacing the missing data with one or more plausible values. Imputation
-seperates the missing data problem from the analysis and the completed
+separates the missing data problem from the analysis and the completed
 data can be analyzed as if it were completely observed. It is generally
@@
 For example, it is possible that test results are missing for several
-students in one or more classes. {[}TODO: Provide an example for one of
-the case studies below.{]} When all observations are missing within one
-or more clusters, data are said to be systematically missing. {[}TODO:
-Refer to Figure 1 and put interpretation in the figure caption. VMTdJ:
-``Mention in text that this is sporadic missingness?''{]}
+students in one or more classes. When all observations are missing
+within one or more clusters, data are said to be systematically missing.
 
@@
 
-\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
+\caption[Sporadic missingness in multilevel data]{Sporadic missingness in multilevel data}\label{fig:patterns}
 \end{figure}
@@
 & \citep{rubi76, meng94}. \\
-& {[}TODO: add congeniality, but maybe in-text?{]} \\
+& \\
 \bottomrule()
 \end{longtable}
-
-{[}GV: DELETE THIS SECTION?{]}However, commonly used imputation methods
-were not designed for use in clustered data and usually generate
-observations that are independent whereas multilevel data are dependent.
-For this reason, we discuss how the \proglang{R} package \pkg{mice} can
-be used to impute multilevel data.
-
-{[}TODO: clarify why clustering is relevant during imputation, and why
-this exposes the need for specialized imputation methods and more
-attention during their implementation (``thou shall not simply run
-\texttt{mice()} on any incomplete dataset'').{]} {[}TODO: Add that the
-more the random effects are of interest, the more you need multilevel
-imputation models.{]} {[}TODO: Add an overview of all possible predictor
-matrix values in manuscript or \texttt{ggmice} legend.{]}
 
@@
 brevity, we focus on multilevel imputation by chained equations with
-\pkg{mice} exclusively; other imputation methods and packages (e.g.,
-\pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial.
-{[}TODO: change this sentence because we do not assume familiarity:
-``Assumed knowledge includes basic familiarity with multilevel
-imputation \citep[see e.g.][ and \citet{grun18}]{audi18} and the
-\pkg{lme4} notation for multilevel models (see Table
-\ref{tab:mod})''{]}.
+\pkg{mice} exclusively; other imputation methods and packages \citep[see
+e.g.][ and \citet{grun18}]{audi18} are outside the scope of this
+tutorial. Assumed knowledge includes basic familiarity with the
+\pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).
 
@@
 \texttt{g2\ \ \ \textbar{}\ \textbar{}}y \textasciitilde{} (1 \\
-& {[}TODO: explain \pkg{lme4} notation here \\
+& \\
 \bottomrule()
@@
 \item
-  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
-  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions with data
-  that are MNAR,][]{GJRM}.
+  \texttt{obesity} from the \pkg{miceheckman} package {[}simulated data
+  on obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
+  that are MNAR{]}.
 \end{itemize}
@@
 study 1) and directly head to practical applications of multilevel
-imputation under MAR conditions (case study IMPACT) or under MNAR
-conditions (case study HIV).
-
-TODO: explicit statement about not going into workings of the methods.
-Galimer 2l methods.
+imputation under MAR conditions (case study 2) or under MNAR conditions
+(case study 3).
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  GV: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB?
-\end{itemize}
-
-{[}TODO: Add environment info, seed and version number(s) somewhere.{]}
+Install non-CRAN packages if necessary:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> devtools::install_github("amices/ggmice")
+R> devtools::install_github("hanneoberman/miceheckman")
+\end{CodeInput}
+\end{CodeChunk}
+
 Set up the R environment and load the necessary packages:
@@
 
-{[}TODO: explain case study{]}
-
 In this section we'll go over the different steps involved with imputing
@@
 \begin{CodeInput}
-R> plot_pattern(popmis)
+R> plot_pattern(popmis, square = TRUE)
 \end{CodeInput}
@@
 
-{[}TODO: Add that it doesn't work with systematic missingness (only with
-sporadic). There's some pros and cons, and it may not even differ much
-if the number of clusters is low.{]}
-
 \begin{CodeChunk}
@@
 
-{[}TODO: check if there is systematic missingness in this dataset, if
-not make Marshall Computerized Tomography classification (ct)
-systematically missing.{]}
-
 We illustrate how to impute incomplete multilevel data by means of a
@@
 \citep[empirical data on traumatic brain injuries, \(n = 11,022\) units
-across \(N = 15\) clusters,][]{metamisc}. {[}TODO: add more info about
-the complete data.{]} The \texttt{impact} data set contains traumatic
-brain injury data on \(n = 11022\) patients clustered in \(N = 15\)
-studies with the following 11 variables:
+across \(N = 15\) clusters,][]{metamisc}. The \texttt{impact} data set
+contains traumatic brain injury data on \(n = 11022\) patients clustered
+in \(N = 15\) studies with the following 11 variables:
 
@@
 \item
-  \texttt{ct} Marshall Computerized Tomography classification, {[}TODO:
-  make this one var? also shows that you don't always need random
-  effects everywhere?{]}
+  \texttt{ct} Marshall Computerized Tomography classification,
 \item
@@
 
-{[}TODO: show how much variance there is after different methods{]}
-
-{[}TODO: add ICC before/after imputation and interpret: This tells us
-that the multilevel structure of the data should probably be taken into
-account. If we don't, we'll may end up with incorrect imputations,
-biasing the effect of the clusters towards zero.{]}
-
-{[}TODO: add descriptive statistics of the complete and incomplete
-data.{]}
-
 \hypertarget{missingness}{%
@@
 To explore the missingness, it is wise to look at the missing data
-pattern:
+pattern. The ten most frequent missingness patterns are shown:
 
@@
 
-This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
-and \texttt{pupil}.{]}
+This shows that we need to impute \texttt{ct} and \texttt{pupil}.
 
@@
 values of the other variables. To see whether the missingness depends on
-the observed values of other variables, we\ldots{} {[}TODO: fill in that
-we can test this statistically or use visual inspection (e.g.~a
-histogram faceted by the missingness indicator).{]}
+the observed values of other variables, we can test this statistically
+or use visual inspection (e.g.~a histogram faceted by the missingness
+indicator).
 
@@
 
-\hypertarget{complete-case-analysis-todo-remove-this}{%
-\subsection{Complete case analysis {[}TODO: remove
-this?{]}}\label{complete-case-analysis-todo-remove-this}}
+\hypertarget{complete-case-analysis}{%
+\subsection{Complete case analysis}\label{complete-case-analysis}}
 
@@
 
-As we can see\ldots{} {[}TODO: fill in: This means that a higher ct
-(Marshall Computerized Tomography classification) is associated with a
-lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI
-\ldots{} to \ldots, when controlling for \ldots{]}
+As we can see, a higher \texttt{ct} (Marshall Computerized Tomography
+classification) is associated with a lower odds of 6-month mortality,
+given by the odds ratio exp(0.42), CI \ldots{} to \ldots, when
+controlling for\ldots{}
 
@@
 R> pred <- quickpred(dat)   # predictor matrix
-R> plot_pred(pred)
+R> plot_pred(pred, rotate = TRUE)
 \end{CodeInput}
@@
 R> diag(pred) <- 0
-R> plot_pred(pred)
+R> plot_pred(pred, rotate = TRUE)
 \end{CodeInput}
@@
                      term    estimate   std.error  statistic      p.value
-1             (Intercept) -2.35203726 0.340181747  -6.914061 4.994037e-12
-2                    type -0.41265892 0.180274846  -2.289054 2.209524e-02
-3                     age  0.03049023 0.001570162  19.418521 1.238416e-81
-4 as.factor(motor_score)2 -0.66764920 0.068737865  -9.712975 3.480413e-22
-5 as.factor(motor_score)3 -1.05520001 0.070218940 -15.027285 2.540225e-50
-6 as.factor(motor_score)4 -1.51238349 0.072304262 -20.916934 1.850073e-90
-7                   pupil  0.48421447 0.038982800  12.421234 6.772069e-17
-8                      ct  0.43474621 0.029968474  14.506785 2.342253e-36
+1             (Intercept) -2.35199614 0.340174191  -6.914093 4.993068e-12
+2                    type -0.41267936 0.180271952  -2.289204 2.208652e-02
+3                     age  0.03049024 0.001570154  19.418637 1.233010e-81
+4 as.factor(motor_score)2 -0.66765325 0.068737838  -9.713038 3.478858e-22
+5 as.factor(motor_score)3 -1.05520488 0.070218678 -15.027410 2.536132e-50
+6 as.factor(motor_score)4 -1.51238928 0.072304221 -20.917026 1.857373e-90
+7                   pupil  0.48421506 0.038982924  12.421209 6.781796e-17
+8                      ct  0.43474556 0.029968441  14.506779 2.341875e-36
              b          df dfcom         fmi      lambda m         riv
-1 5.281599e-04 10119.70041 11013 0.005673265 0.005476771 5 0.005506932
-2 4.881699e-05 10893.89378 11013 0.001985736 0.001802528 5 0.001805783
-3 3.335358e-08  6320.89582 11013 0.016545467 0.016234340 5 0.016502243
-4 4.188703e-05  8327.24771 11013 0.010875748 0.010638213 5 0.010752602
-5 5.135721e-05  7632.20045 11013 0.012757637 0.012498967 5 0.012657168
-6 1.403058e-04  2831.75462 11013 0.032888241 0.032205434 5 0.033277139
-7 3.571497e-04    49.97295 11013 0.309130930 0.282023647 5 0.392803531
-8 8.586645e-05   294.69765 11013 0.120677044 0.114729598 5 0.129598366
+1 5.301620e-04 10113.60453 11013 0.005694385 0.005497777 5 0.005528170
+2 4.905995e-05 10892.82806 11013 0.001994781 0.001811557 5 0.001814845
+3 3.333628e-08  6323.59900 11013 0.016537098 0.016226102 5 0.016493731
+4 4.194026e-05  8322.17049 11013 0.010889419 0.010651742 5 0.010766424
+5 5.136864e-05  7631.13821 11013 0.012760548 0.012501842 5 0.012660117
+6 1.403437e-04  2830.62134 11013 0.032897234 0.032214161 5 0.033286456
+7 3.571667e-04    49.96887 11013 0.309144148 0.282035215 5 0.392825972
+8 8.586575e-05   294.70102 11013 0.120676307 0.114728924 5 0.129597506
           ubar
-1 1.150898e-01
-2 3.244044e-02
-3 2.425385e-06
-4 4.674630e-03
-5 4.869071e-03
-6 5.059539e-03
-7 1.091079e-03
-8 7.950697e-04
+1 1.150823e-01
+2 3.243910e-02
+3 2.425379e-06
+4 4.674562e-03
+5 4.869020e-03
+6 5.059488e-03
+7 1.091068e-03
+8 7.950686e-04
 \end{CodeOutput}
@@
             (Intercept)                     type                      age  
-               -2.37195                 -0.41014                  0.03052  
+               -2.37194                 -0.41015                  0.03052  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
@@
             (Intercept)                     type                      age  
-               -2.32339                 -0.42359                  0.03023  
+               -2.32327                 -0.42365                  0.03023  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67142                 -1.05776                 -1.51038  
+               -0.67143                 -1.05776                 -1.51039  
                   pupil                       ct  
@@
             (Intercept)                     type                      age  
-               -2.33581                 -0.40871                  0.03039  
+               -2.33576                 -0.40872                  0.03039  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66477                 -1.05451                 -1.51858  
+               -0.66477                 -1.05453                 -1.51860  
                   pupil                       ct  
                 0.45928                  0.44419  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
 
@@
             (Intercept)                     type                      age  
-               -2.35187                 -0.40574                  0.03064  
+               -2.35183                 -0.40576                  0.03064  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67468                 -1.06551                 -1.52622  
+               -0.67469                 -1.06551                 -1.52622  
                   pupil                       ct  
@@
 
-\hypertarget{case-study-iii-hiv-data}{%
-\section{Case study III: HIV data}\label{case-study-iii-hiv-data}}
-
-Data are simulated and included in the \texttt{GJRM} package. We will
-use the following variables:
+\hypertarget{case-study-iii-obesity-data}{%
+\section{Case study III: obesity
+data}\label{case-study-iii-obesity-data}}
+
+Data are simulated and included in the \texttt{miceheckman} package. We
+will use the following variables:
 
@@
 \item
-  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
+  \texttt{gender} Gender (0=male, 1=female),
 \item
@@
 \item
-  \texttt{marital} Marital status,
-\item
-  \texttt{condom} Condom use during last intercourse,
-\item
-  \texttt{smoke} Smoker (levels; inclusion restriction variable).
+  \texttt{height} Height in meters,
+\item
+  \texttt{weight} Weight in kilograms,
+\item
+  `rt`` Response time in minutes (inclusion-restriction variable).
 \end{itemize}
 
-The imputation of these date is based on the toy example from
+The imputation of these data is based on the
 \href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
-Heckman Github repo}.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/hiv-1} \end{center}
-
-\end{CodeChunk}
-
-R=region+language+(1\textbar InterviewID) model with interviewer as
-random effects, because the observations are not independent. Interviews
-are not allocated randomly. In theory we expect the
-inclusion-restriction variable to be randomly assigned, that's why we're
-adding region and language to compensate for non-random allocation.
-{[}TODO: Explain exclusion-restriction, refer to paper Johanna{]}.
+Heckman Github repo}
 
@@
 
-We obtain here the random effects for each interviewer, this is an
-approximation of the interviewer's skill which will be used as an
-exclusion constraint. Here, since the location of the interviewer was
-not randomly assigned to the subjects, the assignment was corrected for
-region and language.
-
-Set the Heckman model as imputation method
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> #Set prediction matrix and methods
-R> ini <- mice(obesity, maxit = 0)
-R> meth<-ini$method
+Select relevant variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> dat <- select(obesity, -"bmi")
+\end{CodeInput}
+\end{CodeChunk}
+
+Visualize missing data pattern:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(dat)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1} \end{center}
+
+\end{CodeChunk}
+
+Create predictor matrix:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred <- quickpred(dat)   # predictor matrix
+R> pred["weight","cluster"]<- -2 # clustering variable
+R> pred["weight","rt"] <- -3 #inclusion-restriction variable
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-pred-1} \end{center}
+
+\end{CodeChunk}
+
+Set the Heckman model as imputation method:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth <- make.method(dat) # methods vector
 R> meth["weight"]<-"2l.heckman"
-R> pred <- ini$pred
-R> pred["weight","cluster"]<- -2
-R> pred["weight","rt"] <- -3
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> imp <- mice(obesity, # dataset with missing values
-+                         m = 10,   # number of imputations
-+                         maxit = 1,
-+                         seed = 1234, #seed attached to the dataID
+\end{CodeInput}
+\end{CodeChunk}
+
+Impute the missingness:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp <- mice(
++   dat, # dataset with missing values
++   m = 5, # number of imputations
++   maxit = 10,
 +   meth = meth, #imputation method vector
 +   pred = pred, #imputation predictors matrix
-+                         print = T,
-+                         meta_method="reml",
-+                         pmm=FALSE)
++   meta_method = "reml",
++   printFlag = FALSE
++ )
 \end{CodeInput}
@@
 \item
-  JOMO in \pkg{mice} -\textgreater{} on the side for now
-\item
   Additional levels of clustering
@@
   clustering.
-\end{itemize}
-
-\hypertarget{think-about}{%
-\section{Think about}\label{think-about}}
-
-\begin{itemize}
-\item
-  Adding evaluations of the imputations such as convergence checks
-\item
-  Adding some kind of help function to mice that suggests a suitable
-  predictor matrix to the user, given a certain analysis model.
-\item
-  Adding a \texttt{multilevel\_ampute()} wrapper function in mice.
-\item
-  Exporting \texttt{mids} objects to other packages like \texttt{lme4}
-  or \texttt{coxme}?
-\item
-  Adding a ICC=0 dataset to show that even if there is no clustering it
-  doesn't hurt.
-\item
-  Show use case for deductive imputation for cluster level variables?
-\item
-  env dump in repo
-\item
-  I don't know if in your article you cover something about model
-  complexity, for example sometimes I have to switch from 2l. methods to
-  1l. methods just because the model didn't converge.. this is due to
-  the considered imputation model is very complex regarding the amount
-  of information counted\ldots{} I know that a solution for an
-  imputation model with many predictors is to check correlation plots as
-  you did or use quickpred().. (maybe you can add this somewhere after
-  the correlation plots)\ldots..But as for cluster specification, I
-  don't know if besides plots of distribution per cluster there is
-  something else can be done to see if i have to use 1l. or 2l. for a
-  given variable, also I have no idea.. how to test which is better
-  between 2l.norm or 2l.2stage.norm.
 \end{itemize}
@@
 
-\renewcommand\refname{References}
+\hypertarget{references}{%
+\section{References}\label{references}}
+
 \bibliography{../References/multilevelmice.bib}

### 2023-05-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Discussion section Johanna

@@
   This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. 
-We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
 keywords:
@@
 
+In hierarchical datasets, clustering is a concern because the homoscedasticity in the error terms cannot be assumed across clusters and the relationship among variables may vary at different hierarchical levels.
+When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Audigier, 2021), e.g.
+if the main model accounts for the hierarchical structure also imputation model should do it.
+Not including clustering into the imputation process may lead to effect estimates with smaller standard errors and inflated type I error.
+
+There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/)
+
+The selection of each strategy depends mainly on the assumptions in the main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
+The last restriction is also required when imputations are performed on each cluster separately.
+When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modelled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
+However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.
+the number of clusters is adequate.
+(Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008).
+Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020).
+For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
+(Zhang,2023).
+
 # Funding

### 2023-05-10 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Comments Valentijn

@@
 
-In this section we'll go over the different steps involved with imputing incomplete multilevel data with the R package mice.
+In this section we will go over the different steps involved with imputing incomplete multilevel data with the R package mice.
 We consider the simulated `popmis` dataset, which included pupils ($n = 2000$) clustered within schools ($N = 100$).
@@
 The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`:
+
+<!-- TODO: choose sex or gender -->
 
@@
 
-```{r echo=FALSE}
+<!-- TODO: True effect?
+We have not discussed any effects yet.
+Maybe say that "we consider the associations in the full data set to be the true associations''.
+And mention association instead of effect (effect has a causal interpretation, association can be either causal or predictive) -->
+
+```{r echo=FALSE, fig.cap = "True effect in the popularity data"}
 fits <-
@@
 
-Plot the missing data pattern:
+First we plot the pattern of missing data within categories of the relevant variables. Plot the missing data pattern:
 
@@
 Print the estimates:
+
+<!-- TODO: explain what this function does -->
+
 ```{r}
@@
 
+- FIML vs MI
+
+An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML).
+This method does not require the imputation of any missing values.
+Whereas MI consists of  imputation, analyses and pooling steps, FIML analyses the data in a single step.
+When the assumptions are met the two approaches should produce equivalent results.
+[REF] As FIML requires specialised software, not all analyses can be performed with standard software.
+[REF]
+
+- Survival / TTE, this could be put in the paragraph on congeniality
+
+When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]
+
 <!-- # Think about -->
@@
 In hierarchical datasets, clustering is a concern because the homoscedasticity in the error terms cannot be assumed across clusters and the relationship among variables may vary at different hierarchical levels.
-When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Audigier, 2021), e.g.
-if the main model accounts for the hierarchical structure also imputation model should do it.
+When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g.
+if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021).
 Not including clustering into the imputation process may lead to effect estimates with smaller standard errors and inflated type I error.

### 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Cache chunks manuscript

@@
 library(miceheckman)  # for imputation cf heckman models
+library(metamisc)     # for ???
 ```
@@
 Impute the data, ignoring the cluster structure:
-```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
+```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
 imp <- mice(popmis, pred = pred, print = FALSE)
@@
 Analyze the imputations:
-```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
+```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
 fit <- with(imp, 
@@
 
-```{r pop_predictor, message=FALSE, warning=FALSE}
+```{r pop_predictor, message=FALSE, warning=FALSE, cache=TRUE}
 # adjust the predictor matrix
@@
 Analyze the imputations:
-```{r, echo=TRUE, message=FALSE, warning=FALSE}
+```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
 fit <- with(imp, 
@@
 
-```{r pop_multilevel, message=FALSE, warning=FALSE}
+```{r pop_multilevel, message=FALSE, warning=FALSE, cache=TRUE}
 # adjust the predictor matrix
@@
 Impute the incomplete data
-```{r imp_impact}
+```{r imp_impact, cache=TRUE}
 imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
@@
 
-```{r impact-fit}
+```{r impact-fit, cache=TRUE}
 fit <- imp %>% 
@@
 ```{r bmi}
-dat <- select(obesity, -"bmi")
+dat <- obesity
 ```
@@
 pred <- quickpred(dat)   # predictor matrix
-pred["weight","cluster"]<- -2 # clustering variable
-pred["weight","rt"] <- -3 #inclusion-restriction variable
+pred["Weight","Cluster"]<- -2 # clustering variable
+pred["Weight", ] <- -3 #inclusion-restriction variable
 plot_pred(pred)
@@
 Impute the missingness:
-```{r eval=FALSE}
+```{r eval=FALSE, cache=TRUE}
 imp <- mice(

### 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Cache chunks manuscript

@@
 R> library(miceheckman)  # for imputation cf heckman models
+R> library(metamisc)     # for ???
 \end{CodeInput}
@@
 
-In this section we'll go over the different steps involved with imputing
-incomplete multilevel data with the R package mice. We consider the
-simulated \texttt{popmis} dataset, which included pupils (\(n = 2000\))
-clustered within schools (\(N = 100\)). The following variables are of
-primary interest:
+In this section we will go over the different steps involved with
+imputing incomplete multilevel data with the R package mice. We consider
+the simulated \texttt{popmis} dataset, which included pupils
+(\(n = 2000\)) clustered within schools (\(N = 100\)). The following
+variables are of primary interest:
 
@@
 \begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-3-1} \end{center}
-
+\begin{figure}
+
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-3-1}
+
+}
+
+\caption[True effect in the popularity data]{True effect in the popularity data}\label{fig:unnamed-chunk-3}
+\end{figure}
 \end{CodeChunk}
@@
 
-Plot the missing data pattern:
+First we plot the pattern of missing data within categories of the
+relevant variables. Plot the missing data pattern:
 
@@
 # A tibble: 11 x 9
-   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
+   effect  group term  estimate std.error statistic   p.value conf.low conf.high
    <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
- 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
- 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
+ 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
+ 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
  3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
- 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
- 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
- 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
- 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
- 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
+ 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
+ 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
+ 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
+ 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
+ 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
  9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
-10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.55e-19  1.92     2.76 
-11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
-# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
-#   4: conf.low, 5: conf.high
+10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
+11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
 \end{CodeOutput}
@@
 \begin{CodeInput}
-R> dat <- select(obesity, -"bmi")
+R> dat <- obesity
 \end{CodeInput}
@@
 R> pred <- quickpred(dat)   # predictor matrix
-R> pred["weight","cluster"]<- -2 # clustering variable
-R> pred["weight","rt"] <- -3 #inclusion-restriction variable
+R> pred["Weight","Cluster"]<- -2 # clustering variable
+R> pred["Weight", ] <- -3 #inclusion-restriction variable
 R> plot_pred(pred)
@@
   clustering.
+\item
+  FIML vs MI
 \end{itemize}
+
+An alternative approach to missing data is to use Full Information
+Maximum Likelihood (FIML). This method does not require the imputation
+of any missing values. Whereas MI consists of imputation, analyses and
+pooling steps, FIML analyses the data in a single step. When the
+assumptions are met the two approaches should produce equivalent
+results. {[}REF{]} As FIML requires specialised software, not all
+analyses can be performed with standard software. {[}REF{]}
+
+\begin{itemize}
+\tightlist
+\item
+  Survival / TTE, this could be put in the paragraph on congeniality
+\end{itemize}
+
+When the outcome is time-to-event, the Nelson-Aalen estimate of the time
+to event should be included as a covariate in the imputation model
+{[}REF{]}
+
+In hierarchical datasets, clustering is a concern because the
+homoscedasticity in the error terms cannot be assumed across clusters
+and the relationship among variables may vary at different hierarchical
+levels. When multiple imputation is used to deal with missing data, as
+the imputation and analysis process is performed separately, it is
+necessary that imputation model being congenial with the main analysis
+model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
+structure also imputation model should do it (Audigier, 2021). Not
+including clustering into the imputation process may lead to effect
+estimates with smaller standard errors and inflated type I error.
+
+There are different strategies that can be adopted in the imputation
+process that account for clustering: inclusion of cluster indicator
+variable, performing a separate imputation process for each cluster, or
+performing a simultaneous imputation process by using an imputation
+method that accounts for clustering.(Stata:
+\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
+
+The selection of each strategy depends mainly on the assumptions in the
+main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of
+cluster indicator variables is restricted in datasets where there are
+not many clusters and many observations per cluster (Graham, 2009). The
+last restriction is also required when imputations are performed on each
+cluster separately. When this restriction cannot be achieved, one can
+use an imputation model that simultaneously imputes all clusters using a
+hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters
+are correlated and this correlation is modelled by a random effect so
+the hierarchical model can be estimated even when there are few
+observations per cluster. However, this strategy is best suited for
+balanced data (Grund, 2017) and when random effects model is
+appropriated, i.e.~the number of clusters is adequate. (Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main
+model, for instance by using the cluster indicator strategy may lead to
+bias estimates when the model is based on a hierarchical model
+(Taaljard,2008). Even when an imputation strategy congenial with the
+main model is preferred, it is important to consider whether it is
+appropriate for the data as a less complex imputation strategies may
+also lead to unbiased estimates in certain scenarios(Bailey 2020). For
+instance, in causal effect analysis, separately imputation may lead to
+smaller bias when the size of the smaller exposure cluster is large,
+compared with an imputation model that includes exposure-confounder
+interactions. (Zhang,2023).
 

### 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update figures

@@
       | Utrecht, The Netherlands 
-    affiliation2: |
       | Data Analytics and Methods Task Force, 
-      | European Medicines Agency, Amsterdam, The Netherlands
+      | European Medicines Agency, 
+      | Amsterdam, The Netherlands
 
@@
 
-```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE}
+```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE, fig.margin=TRUE}
 dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
@@
 
-Table 3: Notation
-
-| **Formula \pkg{lme4}**         | **Details**                                 |
-|-------------------------------------------------------------|--------------------------------------------------|
-| `y ~ x1 + (1 | g1)`                                      | Fixed `x1` predictor with random intercept         |
-|                                                             | varying among `g1`                                  |
-| `y~x1*x2+ (1| g1)`                                       | Interactions of `x1` and `x2` only in fixed effect   |
-| `y ~ x1*x2+ (x2| g1)`                                      | Interactions of `x1` and `x2` only in fixed effect   |
-|                                                             | with slope of `x2` randomly varying among `g1`       |
-| `y ~ x1*x2+ (x1*x2| g1)`                                   | variance-covariance matrix estimated only with   |
-|                                                             | the variance terms of intercept, slope of `x1`,    |
-|                                                             | slope of `x2` and interaction `x1*x2`                |
-| `y ~ x1*x2+ (x1 | g1)+ (x2| g1)`                            | variance-covariance matrix estimated separately, |
-|                                                             | i.e, one for intercept and `x1` and another for    |
-|                                                             | intercept and `x2`                                 |
-| `y ~ x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)`             | Fixed `x1` with correlated random intercept and    |
-|                                                             | random slope of `x`                                |
-| `y ~ x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1)` | Fixed `x1` with uncorrelated random intercept      |
-|                                                             | and random slope of `x1`                           |
-| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2   |
-| `y ~ (1 | g1/g2) or ~ (1 | g1) + (1 | g1:g2)`                     | Random intercept varying among `g1` and `g2`         |
-|                                                             | withing `g1`                                       |
-|                                                             |          |
-
 We illustrate imputation of incomplete multilevel data using three case studies:
@@
 library(mitml)        # for multilevel parameter pooling
-library(miceheckman)  # for imputation cf heckman models
-library(metamisc)     # for ???
+library(miceheckman)  # for imputation cf. heckman models
+library(metamisc)     # for case study data
 ```
@@
 
-The true effect is:
-
 <!-- TODO: True effect?
@@
 And mention association instead of effect (effect has a causal interpretation, association can be either causal or predictive) -->
-
-```{r echo=FALSE, fig.cap = "True effect in the popularity data"}
-fits <-
-  purrr::map_dfr(
-    unique(popcomp$school),
-    ~ {
-      lm(
-        mod,
-        data = filter(popcomp, school == .x)
-      )
-    } %>%
-      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
-      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
-      cbind(school = as.character(.x), .)
-  ) %>% filter(term %in% "sex")
-
-n <- popcomp %>% group_by(school) %>% 
-  tally() %>% 
-  mutate(school = as.character(school))
-
-fits %>% left_join(n, by = "school") %>% 
-  ggplot(aes(y = school, x = estimate, linewidth = n)) +
-  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
-  geom_linerange(aes(y = school, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
-  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
-  labs(y = "Study", x = "", color = NULL) +
-  theme_classic() +
-  scale_size(guide = 'none')
-```
 
@@
 
-```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
-plot_pattern(popmis, square = TRUE)
+```{r pop_pat, fig.cap = "Missing data pattern in the popularity data", fig.margin=TRUE}
+plot_pattern(popmis)
 ```
@@
 <!-- TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?) -->
+
+# Appendix
+
+Table 3: Notation
+
+| **Formula \pkg{lme4}**         | **Details**                                 |
+|-------------------------------------------------------------|--------------------------------------------------|
+| `y ~ x1 + (1 | g1)`                                      | Fixed `x1` predictor with random intercept         |
+|                                                             | varying among `g1`                                  |
+| `y~x1*x2+ (1| g1)`                                       | Interactions of `x1` and `x2` only in fixed effect   |
+| `y ~ x1*x2+ (x2| g1)`                                      | Interactions of `x1` and `x2` only in fixed effect   |
+|                                                             | with slope of `x2` randomly varying among `g1`       |
+| `y ~ x1*x2+ (x1*x2| g1)`                                   | variance-covariance matrix estimated only with   |
+|                                                             | the variance terms of intercept, slope of `x1`,    |
+|                                                             | slope of `x2` and interaction `x1*x2`                |
+| `y ~ x1*x2+ (x1 | g1)+ (x2| g1)`                            | variance-covariance matrix estimated separately, |
+|                                                             | i.e, one for intercept and `x1` and another for    |
+|                                                             | intercept and `x2`                                 |
+| `y ~ x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)`             | Fixed `x1` with correlated random intercept and    |
+|                                                             | random slope of `x`                                |
+| `y ~ x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1)` | Fixed `x1` with uncorrelated random intercept      |
+|                                                             | and random slope of `x1`                           |
+| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2   |
+| `y ~ (1 | g1/g2) or ~ (1 | g1) + (1 | g1:g2)`                     | Random intercept varying among `g1` and `g2`         |
+|                                                             | withing `g1`                                       |
+|                                                             |          |

### 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update figures

@@
 University Medical Center Utrecht, Utrecht University,\\
-Utrecht, The Netherlands
+Utrecht, The Netherlands\\
+Data Analytics and Methods Task Force,\\
+European Medicines Agency,\\
+Amsterdam, The Netherlands
 }
@@
 
+We illustrate imputation of incomplete multilevel data using three case
+studies:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
+  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
+  with data that are MAR,][]{mice};
+\item
+  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
+  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
+  studies with data that are MAR,][]{metamisc};
+\item
+  \texttt{obesity} from the \pkg{miceheckman} package {[}simulated data
+  on obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
+  that are MNAR{]}.
+\end{itemize}
+
+For each of these datasets, we discuss the nature of the missingness,
+choose one or more imputation models and evaluate the imputed data, but
+we will also highlight one specific aspect of the imputation workflow.
+
+This tutorial is dedicated to readers who are unfamiliar with multiple
+imputation. More experienced readers can skip the introduction (case
+study 1) and directly head to practical applications of multilevel
+imputation under MAR conditions (case study 2) or under MNAR conditions
+(case study 3).
+
+\hypertarget{setup}{%
+\subsection{Setup}\label{setup}}
+
+Install non-CRAN packages if necessary:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> devtools::install_github("amices/ggmice")
+R> devtools::install_github("hanneoberman/miceheckman")
+\end{CodeInput}
+\end{CodeChunk}
+
+Set up the R environment and load the necessary packages:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> set.seed(2022)        # for reproducibility
+R> library(mice)         # for imputation
+R> library(miceadds)     # for additional imputation routines
+R> library(ggmice)       # for incomplete/imputed data visualization
+R> library(ggplot2)      # for visualization
+R> library(dplyr)        # for data wrangling
+R> library(lme4)         # for multilevel modeling
+R> library(mitml)        # for multilevel parameter pooling
+R> library(miceheckman)  # for imputation cf. heckman models
+R> library(metamisc)     # for case study data
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{case-study-i-popularity-data}{%
+\section{Case study I: popularity
+data}\label{case-study-i-popularity-data}}
+
+In this section we will go over the different steps involved with
+imputing incomplete multilevel data with the R package mice. We consider
+the simulated \texttt{popmis} dataset, which included pupils
+(\(n = 2000\)) clustered within schools (\(N = 100\)). The following
+variables are of primary interest:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{school}, school identification number (clustering variable);
+\item
+  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
+  unit-level);
+\item
+  \texttt{sex}, pupil sex (0=boy, 1=girl; unit-level);
+\item
+  \texttt{texp}, teacher experience (in years; cluster-level).
+\end{itemize}
+
+The research objective of the \texttt{popmis} dataset is to predict the
+pupils' popularity based on their gender and the experience of the
+teacher. The analysis model corresponding to this dataset is multilevel
+regression with random intercepts, random slopes and a cross-level
+interaction. The outcome variable is \texttt{popular}, which is
+predicted from the unit-level variable \texttt{sex} and the
+cluster-level variable \texttt{texp}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> mod <- popular ~ 1 + sex + (1 | school)
+\end{CodeInput}
+\end{CodeChunk}
+
+Load the data into the environment and select the relevant variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> popmis <- popmis[, c("school", "popular", "sex")] 
+\end{CodeInput}
+\end{CodeChunk}
+
+First we plot the pattern of missing data within categories of the
+relevant variables. Plot the missing data pattern:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(popmis, square = TRUE)
+\end{CodeInput}
+\begin{figure}
+
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pat-1}
+
+}
+
+\caption[Missing data pattern in the popularity data]{Missing data pattern in the popularity data}\label{fig:pop_pat}
+\end{figure}
+\end{CodeChunk}
+
+The missingness is univariate and sporadic, which is illustrated in the
+missing data pattern in Figure \ref{fig:pop_pat}.
+
+To develop the best imputation model for the incomplete variable
+\texttt{popular}, we need to know whether the observed values of
+\texttt{popular} are related to observed values of other variables. Plot
+the pair-wise complete correlations in the incomplete data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_corr(popmis)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-corr-1} \end{center}
+
+\end{CodeChunk}
+
+This shows us that \texttt{sex} may be a useful imputation model
+predictor. Moreover, the missingness in \texttt{popular} may depend on
+the observed values of other variables.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # ggmice(popmis, aes(sex)) +
+R> #   geom_histogram(fill = "white") +
+R> #   facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
+R> 
+R> ggplot(popmis, aes(y = popular, group = sex)) +
++   geom_boxplot() + 
++   theme_classic()
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-hist-1} \end{center}
+
+\end{CodeChunk}
+
+\hypertarget{imputation-ignoring-the-cluster-variable-not-recommended}{%
+\subsubsection{Imputation ignoring the cluster variable (not
+recommended)}\label{imputation-ignoring-the-cluster-variable-not-recommended}}
+
+The first imputation model that we'll use is likely to be invalid. We do
+\emph{not} use the cluster identifier \texttt{school} as imputation
+model predictor. With this model, we ignore the multilevel structure of
+the data, despite the high ICC. This assumes exchangeability between
+units. We include it purely to illustrate the effects of ignoring the
+clustering in our imputation effort.
+
+Create a methods vector and predictor matrix for \texttt{popular}, and
+make sure \texttt{school} is not included as predictor:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth <- make.method(popmis) # methods vector
+R> pred <- quickpred(popmis)   # predictor matrix
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-pred-1} \end{center}
+
+\end{CodeChunk}
+
+Impute the data, ignoring the cluster structure:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp <- mice(popmis, pred = pred, print = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+Analyze the imputations:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- with(imp, 
++             lmer(popular ~ 1 + sex  + (1 | school))) 
+\end{CodeInput}
+\end{CodeChunk}
+
+Print the estimates:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+
+testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
+
+Final parameter estimates and inferences obtained from 5 imputed data sets.
+
+             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
+(Intercept)     5.012     0.295    16.994     4.362     0.000    22.587     0.969 
+sex             0.695     0.251     2.768     4.287     0.047    28.390     0.975
+
+                            Estimate 
+Intercept~~Intercept|school    0.266 
+Residual~~Residual             1.035 
+ICC|school                     0.208
+
+Unadjusted hypothesis test as appropriate in larger samples.
+\end{CodeOutput}
+\end{CodeChunk}
+
+\hypertarget{imputation-with-the-cluster-variable-as-predictor-not-recommended}{%
+\subsubsection{Imputation with the cluster variable as predictor (not
+recommended)}\label{imputation-with-the-cluster-variable-as-predictor-not-recommended}}
+
+We'll now use \texttt{school} as a predictor to impute all other
+variables. This is still not recommended practice, since it only works
+under certain circumstances and results may be biased
+\citep{drec15, ende16}. But at least, it includes some multilevel
+aspect. This method is also called `fixed cluster imputation', and uses
+N-1 indicator variables representing allocation of N clusters as a fixed
+factor in the model \citep{reit06, ende16}. Colloquially, this is
+`multilevel imputation for dummies'.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # adjust the predictor matrix
+R> pred["popular", "school"] <- 1 
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor-1} \end{center}
+
+\begin{CodeInput}
+R> # impute the data, cluster as predictor
+R> imp <- mice(popmis, pred = pred, print = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+Analyze the imputations:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- with(imp, 
++             lmer(popular ~ 1 + sex + (1 | school))) 
+\end{CodeInput}
+\end{CodeChunk}
+
+Print the estimates:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+
+testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
+
+Final parameter estimates and inferences obtained from 5 imputed data sets.
+
+             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
+(Intercept)     4.915     0.217    22.642     4.926     0.000     9.110     0.926 
+sex             0.975     0.283     3.444     4.250     0.024    32.504     0.978
+
+                            Estimate 
+Intercept~~Intercept|school    0.351 
+Residual~~Residual             1.153 
+ICC|school                     0.233
+
+Unadjusted hypothesis test as appropriate in larger samples.
+\end{CodeOutput}
+\end{CodeChunk}
+
+\hypertarget{imputation-with-multilevel-model}{%
+\subsubsection{Imputation with multilevel
+model}\label{imputation-with-multilevel-model}}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # adjust the predictor matrix
+R> pred["popular", "school"] <- -2 
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_multilevel-1} \end{center}
+
+\begin{CodeInput}
+R> # impute the data, cluster as predictor
+R> imp <- mice(popmis, pred = pred, print = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+Analyze the imputations:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- with(imp, 
++             lmer(popular ~ 1 + sex + (1 | school))) 
+\end{CodeInput}
+\end{CodeChunk}
+
+Print the estimates:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+
+testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
+
+Final parameter estimates and inferences obtained from 5 imputed data sets.
+
+             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
+(Intercept)     5.011     0.410    12.222     4.226     0.000    35.955     0.980 
+sex             0.928     0.381     2.434     4.168     0.069    48.221     0.985
+
+                            Estimate 
+Intercept~~Intercept|school    0.313 
+Residual~~Residual             1.428 
+ICC|school                     0.188
+
+Unadjusted hypothesis test as appropriate in larger samples.
+\end{CodeOutput}
+\end{CodeChunk}
+
+\hypertarget{case-study-ii-impact-data-syst-missingness-pred-matrix}{%
+\section{Case study II: IMPACT data (syst missingness, pred
+matrix)}\label{case-study-ii-impact-data-syst-missingness-pred-matrix}}
+
+We illustrate how to impute incomplete multilevel data by means of a
+case study: \texttt{impact} from the \pkg{metamisc} package
+\citep[empirical data on traumatic brain injuries, \(n = 11,022\) units
+across \(N = 15\) clusters,][]{metamisc}. The \texttt{impact} data set
+contains traumatic brain injury data on \(n = 11022\) patients clustered
+in \(N = 15\) studies with the following 11 variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{name} Name of the study,
+\item
+  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
+  observational cohort),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{motor\_score} Glasgow Coma Scale motor score,
+\item
+  \texttt{pupil} Pupillary reactivity,
+\item
+  \texttt{ct} Marshall Computerized Tomography classification,
+\item
+  \texttt{hypox} Hypoxia (0=no, 1=yes),
+\item
+  \texttt{hypots} Hypotension (0=no, 1=yes),
+\item
+  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+\item
+  \texttt{edh} Epidural hematoma (0=no, 1=yes),
+\item
+  \texttt{mort} 6-month mortality (0=alive, 1=dead).
+\end{itemize}
+
+The analysis model for this dataset is a prediction model with
+\texttt{mort} as the outcome. In this tutorial we'll estimate the
+adjusted prognostic effect of \texttt{ct} on mortality outcomes. The
+estimand is the adjusted odds ratio for \texttt{ct}, after including
+\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
+into the analysis model:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
+\end{CodeInput}
+\end{CodeChunk}
+
+Note that variables \texttt{hypots}, \texttt{hypox}, \texttt{tsah} and
+\texttt{edh} are not part of the analysis model, and may thus serve as
+auxiliary variables for imputation.
+
+The \texttt{impact} data included in the \pkg{metamisc} package is a
+complete data set. The original data has already been imputed once
+(Steyerberg et al, 2008). For the purpose of this tutorial we have
+induced missingness (mimicking the missing data in the original data set
+before imputation). The resulting incomplete data can be accessed from
+\href{https://zenodo.com}{zenodo link to be created}.
+
+Load the complete and incomplete data into the R workspace:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> data("impact", package = "metamisc")      # complete data
+R> dat <- read.table("link/to/the/data.txt") # incomplete data
+\end{CodeInput}
+\end{CodeChunk}
+
+The estimated effects in the complete data are visualized in Figure
+\ref{}.
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/forest-1} \end{center}
+
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # fit <- glmer(mod, family = "binomial", data = impact) # fit the model
+R> # tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{missingness}{%
+\subsection{Missingness}\label{missingness}}
+
+To explore the missingness, it is wise to look at the missing data
+pattern. The ten most frequent missingness patterns are shown:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pattern-1} \end{center}
+
+\end{CodeChunk}
+
+This shows that we need to impute \texttt{ct} and \texttt{pupil}.
+
+To develop the best imputation model, we need to investigate the
+relations between the observed values of the incomplete variables and
+the observed values of other variables, and the relation between the
+missingness indicators of the incomplete variables and the observed
+values of the other variables. To see whether the missingness depends on
+the observed values of other variables, we can test this statistically
+or use visual inspection (e.g.~a histogram faceted by the missingness
+indicator).
+
+We should impute the variables \texttt{ct} and \texttt{pupil} and any
+auxiliary variables we might want to use to impute these incomplete
+analysis model variables. We can evaluate which variables may be useful
+auxiliaries by plotting the pairwise complete correlations:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_corr(dat, rotate = TRUE) # plot correlations 
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_corr-1} \end{center}
+
+\end{CodeChunk}
+
+This shows us that \texttt{hypox} and \texttt{hypot} would not be useful
+auxiliary variables for imputing \texttt{ct}. Depending on the minimum
+required correlation, \texttt{tsah} could be useful, while \texttt{edh}
+has the strongest correlation with \texttt{ct} out of all the variables
+in the data and should definitely be included in the imputation model.
+For the imputation of \texttt{pupil}, none of the potential auxiliary
+variables has a very strong relation, but \texttt{hypots} could be used.
+We conclude that we can exclude \texttt{hypox} from the data, since this
+is neither an analysis model variable nor an auxiliary variable for
+imputation:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> dat <- select(dat, !hypox)  # remove variable
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{complete-case-analysis}{%
+\subsection{Complete case analysis}\label{complete-case-analysis}}
+
+As previously stated, complete case analysis lowers statistical power
+and may bias results. The complete case analysis estimates are:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
+R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 11 x 9
+   effect  group term  estimate std.error statistic   p.value conf.low conf.high
+   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
+ 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
+ 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
+ 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
+ 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
+ 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
+ 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
+ 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
+ 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
+ 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
+10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
+11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
+\end{CodeOutput}
+\end{CodeChunk}
+
+As we can see, a higher \texttt{ct} (Marshall Computerized Tomography
+classification) is associated with a lower odds of 6-month mortality,
+given by the odds ratio exp(0.42), CI \ldots{} to \ldots, when
+controlling for\ldots{}
+
+\hypertarget{imputation-model}{%
+\subsection{Imputation model}\label{imputation-model}}
+
+Mutate data to get the right data types for imputation (e.g.~integer for
+clustering variable).
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> dat <- dat %>% mutate(across(everything(), as.integer))
+\end{CodeInput}
+\end{CodeChunk}
+
+Create a methods vector and predictor matrix, and make sure
+\texttt{name} is not included as predictor, but as clustering variable:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth <- make.method(dat) # methods vector
+R> pred <- quickpred(dat)   # predictor matrix
+R> plot_pred(pred, rotate = TRUE)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-1} \end{center}
+
+\begin{CodeInput}
+R> pred[pred == 1] <- 2
+R> pred["mort", ] <- 2
+R> pred[, "mort"] <- 2
+R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
+R> pred[, "name"] <- -2
+R> diag(pred) <- 0
+R> plot_pred(pred, rotate = TRUE)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-2} \end{center}
+
+\begin{CodeInput}
+R> meth <- make.method(dat)
+R> meth
+\end{CodeInput}
+\begin{CodeOutput}
+       name        type         age motor_score       pupil          ct 
+         ""          ""          ""          ""       "pmm"       "pmm" 
+     hypots        tsah         edh        mort 
+      "pmm"       "pmm"       "pmm"          "" 
+\end{CodeOutput}
+\end{CodeChunk}
+
+Impute the incomplete data
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fit <- imp %>% 
++   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
+R> tidy(pool(fit))
+\end{CodeInput}
+\begin{CodeOutput}
+                     term    estimate   std.error  statistic      p.value
+1             (Intercept) -2.35199614 0.340174191  -6.914093 4.993068e-12
+2                    type -0.41267936 0.180271952  -2.289204 2.208652e-02
+3                     age  0.03049024 0.001570154  19.418637 1.233010e-81
+4 as.factor(motor_score)2 -0.66765325 0.068737838  -9.713038 3.478858e-22
+5 as.factor(motor_score)3 -1.05520488 0.070218678 -15.027410 2.536132e-50
+6 as.factor(motor_score)4 -1.51238928 0.072304221 -20.917026 1.857373e-90
+7                   pupil  0.48421506 0.038982924  12.421209 6.781796e-17
+8                      ct  0.43474556 0.029968441  14.506779 2.341875e-36
+             b          df dfcom         fmi      lambda m         riv
+1 5.301620e-04 10113.60453 11013 0.005694385 0.005497777 5 0.005528170
+2 4.905995e-05 10892.82806 11013 0.001994781 0.001811557 5 0.001814845
+3 3.333628e-08  6323.59900 11013 0.016537098 0.016226102 5 0.016493731
+4 4.194026e-05  8322.17049 11013 0.010889419 0.010651742 5 0.010766424
+5 5.136864e-05  7631.13821 11013 0.012760548 0.012501842 5 0.012660117
+6 1.403437e-04  2830.62134 11013 0.032897234 0.032214161 5 0.033286456
+7 3.571667e-04    49.96887 11013 0.309144148 0.282035215 5 0.392825972
+8 8.586575e-05   294.70102 11013 0.120676307 0.114728924 5 0.129597506
+          ubar
+1 1.150823e-01
+2 3.243910e-02
+3 2.425379e-06
+4 4.674562e-03
+5 4.869020e-03
+6 5.059488e-03
+7 1.091068e-03
+8 7.950686e-04
+\end{CodeOutput}
+\begin{CodeInput}
+R> as.mitml.result(fit)
+\end{CodeInput}
+\begin{CodeOutput}
+[[1]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+      AIC       BIC    logLik  deviance  df.resid 
+10495.423 10561.192 -5238.712 10477.423     11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2843  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.37194                 -0.41015                  0.03052  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.65802                 -1.04611                 -1.51245  
+                  pupil                       ct  
+                0.50405                  0.42496  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+[[2]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+     AIC      BIC   logLik deviance df.resid 
+10500.88 10566.65 -5241.44 10482.88    11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2917  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.37718                 -0.41511                  0.03067  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.66935                 -1.05211                 -1.49429  
+                  pupil                       ct  
+                0.49013                  0.43835  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+[[3]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+      AIC       BIC    logLik  deviance  df.resid 
+10505.026 10570.795 -5243.513 10487.026     11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2908  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.32327                 -0.42365                  0.03023  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.67143                 -1.05776                 -1.51039  
+                  pupil                       ct  
+                0.49756                  0.42474  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+[[4]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+      AIC       BIC    logLik  deviance  df.resid 
+10519.511 10585.280 -5250.755 10501.511     11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2961  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.33576                 -0.40872                  0.03039  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.66477                 -1.05453                 -1.51860  
+                  pupil                       ct  
+                0.45928                  0.44419  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+[[5]]
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial  ( logit )
+Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
+    name)
+      AIC       BIC    logLik  deviance  df.resid 
+10522.038 10587.807 -5252.019 10504.038     11013 
+Random effects:
+ Groups Name        Std.Dev.
+ name   (Intercept) 0.2955  
+Number of obs: 11022, groups:  name, 15
+Fixed Effects:
+            (Intercept)                     type                      age  
+               -2.35183                 -0.40576                  0.03064  
+as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
+               -0.67469                 -1.06551                 -1.52622  
+                  pupil                       ct  
+                0.47006                  0.44148  
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+
+attr(,"class")
+[1] "mitml.result" "list"        
+\end{CodeOutput}
+\begin{CodeInput}
+R> # testEstimates(as.mitml.result(fit))
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{case-study-iii-obesity-data}{%
+\section{Case study III: obesity
+data}\label{case-study-iii-obesity-data}}
+
+Data are simulated and included in the \texttt{miceheckman} package. We
+will use the following variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{region} Cluster variable,
+\item
+  \texttt{gender} Gender (0=male, 1=female),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{height} Height in meters,
+\item
+  \texttt{weight} Weight in kilograms,
+\item
+  `rt`` Response time in minutes (inclusion-restriction variable).
+\end{itemize}
+
+The imputation of these data is based on the
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}
+
+Load the data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> data("obesity", package = "miceheckman")
+\end{CodeInput}
+\end{CodeChunk}
+
+Select relevant variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> dat <- obesity
+\end{CodeInput}
+\end{CodeChunk}
+
+Visualize missing data pattern:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(dat)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1} \end{center}
+
+\end{CodeChunk}
+
+Create predictor matrix:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred <- quickpred(dat)   # predictor matrix
+R> pred["Weight","Cluster"]<- -2 # clustering variable
+R> pred["Weight", ] <- -3 #inclusion-restriction variable
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-pred-1} \end{center}
+
+\end{CodeChunk}
+
+Set the Heckman model as imputation method:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth <- make.method(dat) # methods vector
+R> meth["weight"]<-"2l.heckman"
+\end{CodeInput}
+\end{CodeChunk}
+
+Impute the missingness:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp <- mice(
++   dat, # dataset with missing values
++   m = 5, # number of imputations
++   maxit = 10,
++   meth = meth, #imputation method vector
++   pred = pred, #imputation predictors matrix
++   meta_method = "reml",
++   printFlag = FALSE
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+\begin{itemize}
+\item
+  Additional levels of clustering
+\item
+  More complex data types: timeseries and polynomial relationship in the
+  clustering.
+\item
+  FIML vs MI
+\end{itemize}
+
+An alternative approach to missing data is to use Full Information
+Maximum Likelihood (FIML). This method does not require the imputation
+of any missing values. Whereas MI consists of imputation, analyses and
+pooling steps, FIML analyses the data in a single step. When the
+assumptions are met the two approaches should produce equivalent
+results. {[}REF{]} As FIML requires specialised software, not all
+analyses can be performed with standard software. {[}REF{]}
+
+\begin{itemize}
+\tightlist
+\item
+  Survival / TTE, this could be put in the paragraph on congeniality
+\end{itemize}
+
+When the outcome is time-to-event, the Nelson-Aalen estimate of the time
+to event should be included as a covariate in the imputation model
+{[}REF{]}
+
+In hierarchical datasets, clustering is a concern because the
+homoscedasticity in the error terms cannot be assumed across clusters
+and the relationship among variables may vary at different hierarchical
+levels. When multiple imputation is used to deal with missing data, as
+the imputation and analysis process is performed separately, it is
+necessary that imputation model being congenial with the main analysis
+model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
+structure also imputation model should do it (Audigier, 2021). Not
+including clustering into the imputation process may lead to effect
+estimates with smaller standard errors and inflated type I error.
+
+There are different strategies that can be adopted in the imputation
+process that account for clustering: inclusion of cluster indicator
+variable, performing a separate imputation process for each cluster, or
+performing a simultaneous imputation process by using an imputation
+method that accounts for clustering.(Stata:
+\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
+
+The selection of each strategy depends mainly on the assumptions in the
+main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of
+cluster indicator variables is restricted in datasets where there are
+not many clusters and many observations per cluster (Graham, 2009). The
+last restriction is also required when imputations are performed on each
+cluster separately. When this restriction cannot be achieved, one can
+use an imputation model that simultaneously imputes all clusters using a
+hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters
+are correlated and this correlation is modelled by a random effect so
+the hierarchical model can be estimated even when there are few
+observations per cluster. However, this strategy is best suited for
+balanced data (Grund, 2017) and when random effects model is
+appropriated, i.e.~the number of clusters is adequate. (Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main
+model, for instance by using the cluster indicator strategy may lead to
+bias estimates when the model is based on a hierarchical model
+(Taaljard,2008). Even when an imputation strategy congenial with the
+main model is preferred, it is important to consider whether it is
+appropriate for the data as a less complex imputation strategies may
+also lead to unbiased estimates in certain scenarios(Bailey 2020). For
+instance, in causal effect analysis, separately imputation may lead to
+smaller bias when the size of the smaller exposure cluster is large,
+compared with an imputation model that includes exposure-confounder
+interactions. (Zhang,2023).
+
+\hypertarget{funding}{%
+\section{Funding}\label{funding}}
+
+This project has received funding from the European Union's Horizon 2020
+research and innovation programme under ReCoDID grant agreement No
+825746.
+
+The views expressed in this paper are the personal views of the authors
+and may not be understood or quoted as being made on behalf of or
+reflecting the position of the regulatory agency/agencies or
+organizations with which the authors are employed/affiliated.
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
 Table 3: Notation
@@
 
-We illustrate imputation of incomplete multilevel data using three case
-studies:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
-  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
-  with data that are MAR,][]{mice};
-\item
-  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
-  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
-  studies with data that are MAR,][]{metamisc};
-\item
-  \texttt{obesity} from the \pkg{miceheckman} package {[}simulated data
-  on obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
-  that are MNAR{]}.
-\end{itemize}
-
-For each of these datasets, we discuss the nature of the missingness,
-choose one or more imputation models and evaluate the imputed data, but
-we will also highlight one specific aspect of the imputation workflow.
-
-This tutorial is dedicated to readers who are unfamiliar with multiple
-imputation. More experienced readers can skip the introduction (case
-study 1) and directly head to practical applications of multilevel
-imputation under MAR conditions (case study 2) or under MNAR conditions
-(case study 3).
-
-\hypertarget{setup}{%
-\subsection{Setup}\label{setup}}
-
-Install non-CRAN packages if necessary:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> devtools::install_github("amices/ggmice")
-R> devtools::install_github("hanneoberman/miceheckman")
-\end{CodeInput}
-\end{CodeChunk}
-
-Set up the R environment and load the necessary packages:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> set.seed(2022)        # for reproducibility
-R> library(mice)         # for imputation
-R> library(miceadds)     # for additional imputation routines
-R> library(ggmice)       # for incomplete/imputed data visualization
-R> library(ggplot2)      # for visualization
-R> library(dplyr)        # for data wrangling
-R> library(lme4)         # for multilevel modeling
-R> library(mitml)        # for multilevel parameter pooling
-R> library(miceheckman)  # for imputation cf heckman models
-R> library(metamisc)     # for ???
-\end{CodeInput}
-\end{CodeChunk}
-
-\hypertarget{case-study-i-popularity-data}{%
-\section{Case study I: popularity
-data}\label{case-study-i-popularity-data}}
-
-In this section we will go over the different steps involved with
-imputing incomplete multilevel data with the R package mice. We consider
-the simulated \texttt{popmis} dataset, which included pupils
-(\(n = 2000\)) clustered within schools (\(N = 100\)). The following
-variables are of primary interest:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{school}, school identification number (clustering variable);
-\item
-  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
-  unit-level);
-\item
-  \texttt{sex}, pupil sex (0=boy, 1=girl; unit-level);
-\item
-  \texttt{texp}, teacher experience (in years; cluster-level).
-\end{itemize}
-
-The research objective of the \texttt{popmis} dataset is to predict the
-pupils' popularity based on their gender and the experience of the
-teacher. The analysis model corresponding to this dataset is multilevel
-regression with random intercepts, random slopes and a cross-level
-interaction. The outcome variable is \texttt{popular}, which is
-predicted from the unit-level variable \texttt{sex} and the
-cluster-level variable \texttt{texp}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> mod <- popular ~ 1 + sex + (1 | school)
-\end{CodeInput}
-\end{CodeChunk}
-
-The true effect is:
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-3-1}
-
-}
-
-\caption[True effect in the popularity data]{True effect in the popularity data}\label{fig:unnamed-chunk-3}
-\end{figure}
-\end{CodeChunk}
-
-Load the data into the environment and select the relevant variables:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> popmis <- popmis[, c("school", "popular", "sex")] 
-\end{CodeInput}
-\end{CodeChunk}
-
-First we plot the pattern of missing data within categories of the
-relevant variables. Plot the missing data pattern:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_pattern(popmis, square = TRUE)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pat-1}
-
-}
-
-\caption[Missing data pattern in the popularity data]{Missing data pattern in the popularity data}\label{fig:pop_pat}
-\end{figure}
-\end{CodeChunk}
-
-The missingness is univariate and sporadic, which is illustrated in the
-missing data pattern in Figure \ref{fig:pop_pat}.
-
-To develop the best imputation model for the incomplete variable
-\texttt{popular}, we need to know whether the observed values of
-\texttt{popular} are related to observed values of other variables. Plot
-the pair-wise complete correlations in the incomplete data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_corr(popmis)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-corr-1} \end{center}
-
-\end{CodeChunk}
-
-This shows us that \texttt{sex} may be a useful imputation model
-predictor. Moreover, the missingness in \texttt{popular} may depend on
-the observed values of other variables.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # ggmice(popmis, aes(sex)) +
-R> #   geom_histogram(fill = "white") +
-R> #   facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
-R> 
-R> ggplot(popmis, aes(y = popular, group = sex)) +
-+   geom_boxplot() + 
-+   theme_classic()
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-hist-1} \end{center}
-
-\end{CodeChunk}
-
-\hypertarget{imputation-ignoring-the-cluster-variable-not-recommended}{%
-\subsubsection{Imputation ignoring the cluster variable (not
-recommended)}\label{imputation-ignoring-the-cluster-variable-not-recommended}}
-
-The first imputation model that we'll use is likely to be invalid. We do
-\emph{not} use the cluster identifier \texttt{school} as imputation
-model predictor. With this model, we ignore the multilevel structure of
-the data, despite the high ICC. This assumes exchangeability between
-units. We include it purely to illustrate the effects of ignoring the
-clustering in our imputation effort.
-
-Create a methods vector and predictor matrix for \texttt{popular}, and
-make sure \texttt{school} is not included as predictor:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> meth <- make.method(popmis) # methods vector
-R> pred <- quickpred(popmis)   # predictor matrix
-R> plot_pred(pred)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-pred-1} \end{center}
-
-\end{CodeChunk}
-
-Impute the data, ignoring the cluster structure:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> imp <- mice(popmis, pred = pred, print = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
-
-Analyze the imputations:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> fit <- with(imp, 
-+             lmer(popular ~ 1 + sex  + (1 | school))) 
-\end{CodeInput}
-\end{CodeChunk}
-
-Print the estimates:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-
-testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
-
-Final parameter estimates and inferences obtained from 5 imputed data sets.
-
-             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.012     0.295    16.994     4.362     0.000    22.587     0.969 
-sex             0.695     0.251     2.768     4.287     0.047    28.390     0.975
-
-                            Estimate 
-Intercept~~Intercept|school    0.266 
-Residual~~Residual             1.035 
-ICC|school                     0.208
-
-Unadjusted hypothesis test as appropriate in larger samples.
-\end{CodeOutput}
-\end{CodeChunk}
-
-\hypertarget{imputation-with-the-cluster-variable-as-predictor-not-recommended}{%
-\subsubsection{Imputation with the cluster variable as predictor (not
-recommended)}\label{imputation-with-the-cluster-variable-as-predictor-not-recommended}}
-
-We'll now use \texttt{school} as a predictor to impute all other
-variables. This is still not recommended practice, since it only works
-under certain circumstances and results may be biased
-\citep{drec15, ende16}. But at least, it includes some multilevel
-aspect. This method is also called `fixed cluster imputation', and uses
-N-1 indicator variables representing allocation of N clusters as a fixed
-factor in the model \citep{reit06, ende16}. Colloquially, this is
-`multilevel imputation for dummies'.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # adjust the predictor matrix
-R> pred["popular", "school"] <- 1 
-R> plot_pred(pred)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor-1} \end{center}
-
-\begin{CodeInput}
-R> # impute the data, cluster as predictor
-R> imp <- mice(popmis, pred = pred, print = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
-
-Analyze the imputations:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> fit <- with(imp, 
-+             lmer(popular ~ 1 + sex + (1 | school))) 
-\end{CodeInput}
-\end{CodeChunk}
-
-Print the estimates:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-
-testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
-
-Final parameter estimates and inferences obtained from 5 imputed data sets.
-
-             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.915     0.217    22.642     4.926     0.000     9.110     0.926 
-sex             0.975     0.283     3.444     4.250     0.024    32.504     0.978
-
-                            Estimate 
-Intercept~~Intercept|school    0.351 
-Residual~~Residual             1.153 
-ICC|school                     0.233
-
-Unadjusted hypothesis test as appropriate in larger samples.
-\end{CodeOutput}
-\end{CodeChunk}
-
-\hypertarget{imputation-with-multilevel-model}{%
-\subsubsection{Imputation with multilevel
-model}\label{imputation-with-multilevel-model}}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # adjust the predictor matrix
-R> pred["popular", "school"] <- -2 
-R> plot_pred(pred)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_multilevel-1} \end{center}
-
-\begin{CodeInput}
-R> # impute the data, cluster as predictor
-R> imp <- mice(popmis, pred = pred, print = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
-
-Analyze the imputations:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> fit <- with(imp, 
-+             lmer(popular ~ 1 + sex + (1 | school))) 
-\end{CodeInput}
-\end{CodeChunk}
-
-Print the estimates:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-
-testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
-
-Final parameter estimates and inferences obtained from 5 imputed data sets.
-
-             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.011     0.410    12.222     4.226     0.000    35.955     0.980 
-sex             0.928     0.381     2.434     4.168     0.069    48.221     0.985
-
-                            Estimate 
-Intercept~~Intercept|school    0.313 
-Residual~~Residual             1.428 
-ICC|school                     0.188
-
-Unadjusted hypothesis test as appropriate in larger samples.
-\end{CodeOutput}
-\end{CodeChunk}
-
-\hypertarget{case-study-ii-impact-data-syst-missingness-pred-matrix}{%
-\section{Case study II: IMPACT data (syst missingness, pred
-matrix)}\label{case-study-ii-impact-data-syst-missingness-pred-matrix}}
-
-We illustrate how to impute incomplete multilevel data by means of a
-case study: \texttt{impact} from the \pkg{metamisc} package
-\citep[empirical data on traumatic brain injuries, \(n = 11,022\) units
-across \(N = 15\) clusters,][]{metamisc}. The \texttt{impact} data set
-contains traumatic brain injury data on \(n = 11022\) patients clustered
-in \(N = 15\) studies with the following 11 variables:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{name} Name of the study,
-\item
-  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
-  observational cohort),
-\item
-  \texttt{age} Age of the patient,
-\item
-  \texttt{motor\_score} Glasgow Coma Scale motor score,
-\item
-  \texttt{pupil} Pupillary reactivity,
-\item
-  \texttt{ct} Marshall Computerized Tomography classification,
-\item
-  \texttt{hypox} Hypoxia (0=no, 1=yes),
-\item
-  \texttt{hypots} Hypotension (0=no, 1=yes),
-\item
-  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
-\item
-  \texttt{edh} Epidural hematoma (0=no, 1=yes),
-\item
-  \texttt{mort} 6-month mortality (0=alive, 1=dead).
-\end{itemize}
-
-The analysis model for this dataset is a prediction model with
-\texttt{mort} as the outcome. In this tutorial we'll estimate the
-adjusted prognostic effect of \texttt{ct} on mortality outcomes. The
-estimand is the adjusted odds ratio for \texttt{ct}, after including
-\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
-into the analysis model:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that variables \texttt{hypots}, \texttt{hypox}, \texttt{tsah} and
-\texttt{edh} are not part of the analysis model, and may thus serve as
-auxiliary variables for imputation.
-
-The \texttt{impact} data included in the \pkg{metamisc} package is a
-complete data set. The original data has already been imputed once
-(Steyerberg et al, 2008). For the purpose of this tutorial we have
-induced missingness (mimicking the missing data in the original data set
-before imputation). The resulting incomplete data can be accessed from
-\href{https://zenodo.com}{zenodo link to be created}.
-
-Load the complete and incomplete data into the R workspace:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> data("impact", package = "metamisc")      # complete data
-R> dat <- read.table("link/to/the/data.txt") # incomplete data
-\end{CodeInput}
-\end{CodeChunk}
-
-The estimated effects in the complete data are visualized in Figure
-\ref{}.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/forest-1} \end{center}
-
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # fit <- glmer(mod, family = "binomial", data = impact) # fit the model
-R> # tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
-\end{CodeInput}
-\end{CodeChunk}
-
-\hypertarget{missingness}{%
-\subsection{Missingness}\label{missingness}}
-
-To explore the missingness, it is wise to look at the missing data
-pattern. The ten most frequent missingness patterns are shown:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pattern-1} \end{center}
-
-\end{CodeChunk}
-
-This shows that we need to impute \texttt{ct} and \texttt{pupil}.
-
-To develop the best imputation model, we need to investigate the
-relations between the observed values of the incomplete variables and
-the observed values of other variables, and the relation between the
-missingness indicators of the incomplete variables and the observed
-values of the other variables. To see whether the missingness depends on
-the observed values of other variables, we can test this statistically
-or use visual inspection (e.g.~a histogram faceted by the missingness
-indicator).
-
-We should impute the variables \texttt{ct} and \texttt{pupil} and any
-auxiliary variables we might want to use to impute these incomplete
-analysis model variables. We can evaluate which variables may be useful
-auxiliaries by plotting the pairwise complete correlations:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_corr(dat, rotate = TRUE) # plot correlations 
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_corr-1} \end{center}
-
-\end{CodeChunk}
-
-This shows us that \texttt{hypox} and \texttt{hypot} would not be useful
-auxiliary variables for imputing \texttt{ct}. Depending on the minimum
-required correlation, \texttt{tsah} could be useful, while \texttt{edh}
-has the strongest correlation with \texttt{ct} out of all the variables
-in the data and should definitely be included in the imputation model.
-For the imputation of \texttt{pupil}, none of the potential auxiliary
-variables has a very strong relation, but \texttt{hypots} could be used.
-We conclude that we can exclude \texttt{hypox} from the data, since this
-is neither an analysis model variable nor an auxiliary variable for
-imputation:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> dat <- select(dat, !hypox)  # remove variable
-\end{CodeInput}
-\end{CodeChunk}
-
-\hypertarget{complete-case-analysis}{%
-\subsection{Complete case analysis}\label{complete-case-analysis}}
-
-As previously stated, complete case analysis lowers statistical power
-and may bias results. The complete case analysis estimates are:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
-R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 11 x 9
-   effect  group term  estimate std.error statistic   p.value conf.low conf.high
-   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
- 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
- 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
- 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
- 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
- 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
- 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
- 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
- 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
- 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
-10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
-11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
-\end{CodeOutput}
-\end{CodeChunk}
-
-As we can see, a higher \texttt{ct} (Marshall Computerized Tomography
-classification) is associated with a lower odds of 6-month mortality,
-given by the odds ratio exp(0.42), CI \ldots{} to \ldots, when
-controlling for\ldots{}
-
-\hypertarget{imputation-model}{%
-\subsection{Imputation model}\label{imputation-model}}
-
-Mutate data to get the right data types for imputation (e.g.~integer for
-clustering variable).
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> dat <- dat %>% mutate(across(everything(), as.integer))
-\end{CodeInput}
-\end{CodeChunk}
-
-Create a methods vector and predictor matrix, and make sure
-\texttt{name} is not included as predictor, but as clustering variable:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> meth <- make.method(dat) # methods vector
-R> pred <- quickpred(dat)   # predictor matrix
-R> plot_pred(pred, rotate = TRUE)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-1} \end{center}
-
-\begin{CodeInput}
-R> pred[pred == 1] <- 2
-R> pred["mort", ] <- 2
-R> pred[, "mort"] <- 2
-R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
-R> pred[, "name"] <- -2
-R> diag(pred) <- 0
-R> plot_pred(pred, rotate = TRUE)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-2} \end{center}
-
-\begin{CodeInput}
-R> meth <- make.method(dat)
-R> meth
-\end{CodeInput}
-\begin{CodeOutput}
-       name        type         age motor_score       pupil          ct 
-         ""          ""          ""          ""       "pmm"       "pmm" 
-     hypots        tsah         edh        mort 
-      "pmm"       "pmm"       "pmm"          "" 
-\end{CodeOutput}
-\end{CodeChunk}
-
-Impute the incomplete data
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> fit <- imp %>% 
-+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
-R> tidy(pool(fit))
-\end{CodeInput}
-\begin{CodeOutput}
-                     term    estimate   std.error  statistic      p.value
-1             (Intercept) -2.35199614 0.340174191  -6.914093 4.993068e-12
-2                    type -0.41267936 0.180271952  -2.289204 2.208652e-02
-3                     age  0.03049024 0.001570154  19.418637 1.233010e-81
-4 as.factor(motor_score)2 -0.66765325 0.068737838  -9.713038 3.478858e-22
-5 as.factor(motor_score)3 -1.05520488 0.070218678 -15.027410 2.536132e-50
-6 as.factor(motor_score)4 -1.51238928 0.072304221 -20.917026 1.857373e-90
-7                   pupil  0.48421506 0.038982924  12.421209 6.781796e-17
-8                      ct  0.43474556 0.029968441  14.506779 2.341875e-36
-             b          df dfcom         fmi      lambda m         riv
-1 5.301620e-04 10113.60453 11013 0.005694385 0.005497777 5 0.005528170
-2 4.905995e-05 10892.82806 11013 0.001994781 0.001811557 5 0.001814845
-3 3.333628e-08  6323.59900 11013 0.016537098 0.016226102 5 0.016493731
-4 4.194026e-05  8322.17049 11013 0.010889419 0.010651742 5 0.010766424
-5 5.136864e-05  7631.13821 11013 0.012760548 0.012501842 5 0.012660117
-6 1.403437e-04  2830.62134 11013 0.032897234 0.032214161 5 0.033286456
-7 3.571667e-04    49.96887 11013 0.309144148 0.282035215 5 0.392825972
-8 8.586575e-05   294.70102 11013 0.120676307 0.114728924 5 0.129597506
-          ubar
-1 1.150823e-01
-2 3.243910e-02
-3 2.425379e-06
-4 4.674562e-03
-5 4.869020e-03
-6 5.059488e-03
-7 1.091068e-03
-8 7.950686e-04
-\end{CodeOutput}
-\begin{CodeInput}
-R> as.mitml.result(fit)
-\end{CodeInput}
-\begin{CodeOutput}
-[[1]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10495.423 10561.192 -5238.712 10477.423     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2843  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.37194                 -0.41015                  0.03052  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65802                 -1.04611                 -1.51245  
-                  pupil                       ct  
-                0.50405                  0.42496  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[2]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-     AIC      BIC   logLik deviance df.resid 
-10500.88 10566.65 -5241.44 10482.88    11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2917  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.37718                 -0.41511                  0.03067  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66935                 -1.05211                 -1.49429  
-                  pupil                       ct  
-                0.49013                  0.43835  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[3]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10505.026 10570.795 -5243.513 10487.026     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2908  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.32327                 -0.42365                  0.03023  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67143                 -1.05776                 -1.51039  
-                  pupil                       ct  
-                0.49756                  0.42474  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[4]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10519.511 10585.280 -5250.755 10501.511     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2961  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.33576                 -0.40872                  0.03039  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66477                 -1.05453                 -1.51860  
-                  pupil                       ct  
-                0.45928                  0.44419  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[5]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10522.038 10587.807 -5252.019 10504.038     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2955  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.35183                 -0.40576                  0.03064  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67469                 -1.06551                 -1.52622  
-                  pupil                       ct  
-                0.47006                  0.44148  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-attr(,"class")
-[1] "mitml.result" "list"        
-\end{CodeOutput}
-\begin{CodeInput}
-R> # testEstimates(as.mitml.result(fit))
-\end{CodeInput}
-\end{CodeChunk}
-
-\hypertarget{case-study-iii-obesity-data}{%
-\section{Case study III: obesity
-data}\label{case-study-iii-obesity-data}}
-
-Data are simulated and included in the \texttt{miceheckman} package. We
-will use the following variables:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{region} Cluster variable,
-\item
-  \texttt{gender} Gender (0=male, 1=female),
-\item
-  \texttt{age} Age of the patient,
-\item
-  \texttt{height} Height in meters,
-\item
-  \texttt{weight} Weight in kilograms,
-\item
-  `rt`` Response time in minutes (inclusion-restriction variable).
-\end{itemize}
-
-The imputation of these data is based on the
-\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
-Heckman Github repo}
-
-Load the data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> data("obesity", package = "miceheckman")
-\end{CodeInput}
-\end{CodeChunk}
-
-Select relevant variables:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> dat <- obesity
-\end{CodeInput}
-\end{CodeChunk}
-
-Visualize missing data pattern:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_pattern(dat)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1} \end{center}
-
-\end{CodeChunk}
-
-Create predictor matrix:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> pred <- quickpred(dat)   # predictor matrix
-R> pred["Weight","Cluster"]<- -2 # clustering variable
-R> pred["Weight", ] <- -3 #inclusion-restriction variable
-R> plot_pred(pred)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-pred-1} \end{center}
-
-\end{CodeChunk}
-
-Set the Heckman model as imputation method:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> meth <- make.method(dat) # methods vector
-R> meth["weight"]<-"2l.heckman"
-\end{CodeInput}
-\end{CodeChunk}
-
-Impute the missingness:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> imp <- mice(
-+   dat, # dataset with missing values
-+   m = 5, # number of imputations
-+   maxit = 10,
-+   meth = meth, #imputation method vector
-+   pred = pred, #imputation predictors matrix
-+   meta_method = "reml",
-+   printFlag = FALSE
-+ )
-\end{CodeInput}
-\end{CodeChunk}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\begin{itemize}
-\item
-  Additional levels of clustering
-\item
-  More complex data types: timeseries and polynomial relationship in the
-  clustering.
-\item
-  FIML vs MI
-\end{itemize}
-
-An alternative approach to missing data is to use Full Information
-Maximum Likelihood (FIML). This method does not require the imputation
-of any missing values. Whereas MI consists of imputation, analyses and
-pooling steps, FIML analyses the data in a single step. When the
-assumptions are met the two approaches should produce equivalent
-results. {[}REF{]} As FIML requires specialised software, not all
-analyses can be performed with standard software. {[}REF{]}
-
-\begin{itemize}
-\tightlist
-\item
-  Survival / TTE, this could be put in the paragraph on congeniality
-\end{itemize}
-
-When the outcome is time-to-event, the Nelson-Aalen estimate of the time
-to event should be included as a covariate in the imputation model
-{[}REF{]}
-
-In hierarchical datasets, clustering is a concern because the
-homoscedasticity in the error terms cannot be assumed across clusters
-and the relationship among variables may vary at different hierarchical
-levels. When multiple imputation is used to deal with missing data, as
-the imputation and analysis process is performed separately, it is
-necessary that imputation model being congenial with the main analysis
-model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
-structure also imputation model should do it (Audigier, 2021). Not
-including clustering into the imputation process may lead to effect
-estimates with smaller standard errors and inflated type I error.
-
-There are different strategies that can be adopted in the imputation
-process that account for clustering: inclusion of cluster indicator
-variable, performing a separate imputation process for each cluster, or
-performing a simultaneous imputation process by using an imputation
-method that accounts for clustering.(Stata:
-\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
-
-The selection of each strategy depends mainly on the assumptions in the
-main analysis and also on the restriction of the analyzed data.
-
-Regarding the restrictions imposed by the data, for instance, the use of
-cluster indicator variables is restricted in datasets where there are
-not many clusters and many observations per cluster (Graham, 2009). The
-last restriction is also required when imputations are performed on each
-cluster separately. When this restriction cannot be achieved, one can
-use an imputation model that simultaneously imputes all clusters using a
-hierarchical model (Allison 2002).
-
-Under this hierarchical imputation model, observations within clusters
-are correlated and this correlation is modelled by a random effect so
-the hierarchical model can be estimated even when there are few
-observations per cluster. However, this strategy is best suited for
-balanced data (Grund, 2017) and when random effects model is
-appropriated, i.e.~the number of clusters is adequate. (Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main
-model, for instance by using the cluster indicator strategy may lead to
-bias estimates when the model is based on a hierarchical model
-(Taaljard,2008). Even when an imputation strategy congenial with the
-main model is preferred, it is important to consider whether it is
-appropriate for the data as a less complex imputation strategies may
-also lead to unbiased estimates in certain scenarios(Bailey 2020). For
-instance, in causal effect analysis, separately imputation may lead to
-smaller bias when the size of the smaller exposure cluster is large,
-compared with an imputation model that includes exposure-confounder
-interactions. (Zhang,2023).
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-This project has received funding from the European Union's Horizon 2020
-research and innovation programme under ReCoDID grant agreement No
-825746.
-
-The views expressed in this paper are the personal views of the authors
-and may not be understood or quoted as being made on behalf of or
-reflecting the position of the regulatory agency/agencies or
-organizations with which the authors are employed/affiliated.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
 \bibliography{../References/multilevelmice.bib}

### 2023-05-24 Manuscript/Archive.Rmd

---
title: "Archive"
author: "Hanne Oberman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, fig.cap = "True effect in the popularity data"}
fits <-
  purrr::map_dfr(
    unique(popcomp$school),
    ~ {
      lm(
        mod,
        data = filter(popcomp, school == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(school = as.character(.x), .)
  ) %>% filter(term %in% "sex")

n <- popcomp %>% group_by(school) %>% 
  tally() %>% 
  mutate(school = as.character(school))

fits %>% left_join(n, by = "school") %>% 
  ggplot(aes(y = school, x = estimate, linewidth = n)) +
  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
  geom_linerange(aes(y = school, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
  labs(y = "Study", x = "", color = NULL) +
  theme_classic() +
  scale_size(guide = 'none')
```

### 2023-05-26 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Feedback

@@
 
+TODO: explain exclusion restriction.
+
 Data are simulated and included in the `miceheckman` package. We will use the following variables:
@@
 - `weight` Weight in kilograms,
-- `rt`` Response time in minutes (inclusion-restriction variable).
+- `time`` Response time in minutes (inclusion-restriction variable).
 
@@
 
+The matrix only shows the predictors for the main model, not the selection model.
+
 Create predictor matrix:
@@
 pred <- quickpred(dat)   # predictor matrix
-pred["Weight","Cluster"]<- -2 # clustering variable
-pred["Weight", ] <- -3 #inclusion-restriction variable
+pred[,"Cluster"]<- -2 # clustering variable
+pred[, "Weight"] <- -3 #inclusion-restriction variable
 plot_pred(pred)
@@
 <!-- # make assumptions of these methods explicit! -->
+ORDER:
+
+- summary
+- congeniality, then in hierarchical models
+- look whether we can fit cong. back in the main body
+- alt. methods
+- conclusion: mice is really easy!
 
@@
 
-There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/)
+There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.
 
@@
 
-Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modelled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
+Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
 However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.

### 2023-05-31 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
update table 1 and 2 tables in appendix
Please just refer to changes in tables, other changes i did was to solve running problems

--- 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 ```{r pattern, fig.height=7.1}
-plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
+plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
 ```

### 2023-05-31 Manuscript/Imputation_of_Incomplete_Multilevel_Data2.Rmd
update table 1 and 2 tables in appendix
Please just refer to changes in tables, other changes i did was to solve running problems

--- 2023-05-31 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
   - name: Hanne I. Oberman
-    affiliation: |
-      | Methodology and Statistics
-      | Utrecht University
+    affiliation: Utrecht University
     address: |
@@
     url: https://hanneoberman.github.io/
-  - name: "Johanna Muñoz"
+  - name: "Johanna Munoz"
     affiliation: |
@@
   - name: Thomas P. A. Debray
-    affiliation: |
-      | Julius Center for Health Sciences and Primary Care, 
+    affiliation: Julius Center for Health Sciences and Primary Care
+    address: |
       | University Medical Center Utrecht, Utrecht University, 
@@
   - name: Gerko Vink
-    affiliation: |
-      | Methodology and Statistics
-      | Utrecht University \AND
+    affiliation: Utrecht University \AND
   - name: Valentijn M. T. de Jong
@@
       | Utrecht, The Netherlands 
+    affiliation2: |
       | Data Analytics and Methods Task Force, 
-      | European Medicines Agency, 
-      | Amsterdam, The Netherlands
+      | European Medicines Agency, Amsterdam, The Netherlands
 
@@
   This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}.
+Footnotes in the current version show work in progress/under construction.
+The last section is not part of the manuscript, but purely for reminders.
+See also all of the TODOs that need to be worked out.
+We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
+[Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
 keywords:
@@
   chunk_output_type: inline
+  markdown: 
+    wrap: 72
 ---
@@
 
+<!-- Missing data pattern with clustering -->
+<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
+<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
+<!-- TODO: check spss missing data pie chart? -->
+<!-- TODO: similarity of patterns over clusters -->
+<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
+<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
+<!-- TODO: look at vim -->
+<!-- TODO: check if shepley plot is an option -->
+
+<!-- Conditional plots missingness indicator -->
+<!-- TODO: check if up-side-down plot works -->
+<!-- TODO: check smoothing in geom_density function and make it the inverse of the sample size -->
+<!-- TODO: add functions to mice -->
+<!-- TODO: add title and informative legend (pop obs and pop miss) -->
+<!-- TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .` -->
+<!-- TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon -->
+<!-- TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?) -->
+
 # Introduction
 
-Many datasets include individuals that are clustered together, for example in geographic regions, or even different studies.
-In the simplest case, individuals (e.g., students) are nested within a single cluster (e.g., school classes).
-More complex clustered structures may occur when there are multiple hierarchical levels (e.g., students in different schools or patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases).
-With clustered data we generally assume that individuals from the same cluster tend to be more similar than individuals from other clusters.
-In statistical terms, this implies that observations from the same cluster are not independent and may in fact be correlated.
+Many datasets include individuals from multiple settings, geographic regions, or even different studies.
+In the simplest case, individuals (e.g., students) are nested within so-called clusters (e.g., school classes).
+More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases).
+In general, individuals from the same cluster tend to be more similar than individuals from other clusters.
+In statistical terms, this implies that observations from the same cluster are correlated.
 If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01].
-Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
+[TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation.
+VMTdJ: "I would only make that link after introducing missing data, and missing data is not mentioned yet, so I would do that in the next section"] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
 These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'.
@@
 
-Table 1: Concepts in multilevel methods
-
-| **Concept**         | **Details**                                                                   |
-|---------------------|-------------------------------------------------------------------------------|
-| Sample unit         | Units of the population from which measurements are taken in a sample.        |
-| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
-| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
-|                     | the values observed in the study are representative of all values in the      |
-|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
-| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
-|                     | admission we might select only certain hospitals that are representative of   |
-|                     | the difference of y between individual hospitals, but rather the variation of |
-| ICC                 | The variability due to clustering is often measured by means of the           |
-|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
-|                     | of variance that can be attributed to the cluster-level, where a high         |
-|                     | ICC would indicate that a lot of variability is due to the cluster            |
-|                     | structure.                                                                    |
-| Random effect       | Multilevel models typically accommodate for variability by including          |
-|                     | a separate group mean for each cluster. In addition to random                 |
-|                     | intercepts, multilevel models can also include random coefficients            |
-|                     | and heterogeneous residual error variances across clusters [see e.g.          |
-|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |
-
 ## Missingness in multilevel data
 
-As with any other dataset, clustered datasets may be impacted by missingness in much the same way.
+<!-- Maybe besides CC and imputation name also likelihood based methods -->
+Like any other dataset, clustered datasets are prone to missing data.
 Several strategies can be used to handle missing data, including complete case analysis and imputation.
 We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values.
-Imputation separates the missing data problem from the analysis and the completed data can be analyzed as if it were completely observed.
-It is generally recommended to impute the missing values more than once to preserve uncertainty due to missingness and to allow for valid inferences (c.f.
+Afterwards, the completed data can be analyzed as if they were completely observed.
+In contrast to single imputation (where missing data are only replaced once), multiple imputation allows to preserve uncertainty due to missingness and is therefore recommended (c.f.
 Rubin 1976).
 
-With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13].
+When clustered datasets are affected by missing values, we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13].
 Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18].
 For example, it is possible that test results are missing for several students in one or more classes.
-When all observations are missing within one or more clusters, data are said to be systematically missing.
+[TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are systematically missing.
+[TODO: Refer to Figure 1 and put interpretation in the figure caption.
+VMTdJ: "Mention in text that this is sporadic missingness?"]
+
+# Introduction
+
+Many datasets include individuals that are clustered together, for
+example in geographic regions, or even different studies. In the
+simplest case, individuals (e.g., students) are nested within a single
+cluster (e.g., school classes). More complex clustered structures may
+occur when there are multiple hierarchical levels (e.g., students in
+different schools or patients within hospitals within regions across
+countries), or when the clustering is non-nested (e.g., electronic
+health record data from diverse settings and populations within large
+databases). With clustered data we generally assume that individuals
+from the same cluster tend to be more similar than individuals from
+other clusters. In statistical terms, this implies that observations
+from the same cluster are not independent and may in fact be correlated.
+If this correlation is left unaddressed, estimates of *p* values,
+confidence intervals even model parameters are prone to bias [@loca01].
+Statistical methods for clustered data typically adopt hierarchical
+models that explicitly describe the grouping of observations. These
+models are also known as 'multilevel models', 'hierarchical models',
+'mixed effect models', 'random effect models', and in the context of
+time-to-event data as 'frailty models'. Table \ref{tab:clus} provides an
+overview of some key concepts in multilevel modeling.
+
+Table 1: Concepts in multilevel methods
+
+| **Concept**          | **Details**                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      |
+|-----------------|-------------------------------------------------------|
+| Sample unit          | Units of the population from which measurements are taken in a sample.
+|
+| Hierarchical levels  | Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.
+|
+| Fixed effect         | Effects that are constant across all sample units, e.g.
+something that researchers control for and can repeat, such as the administration of a drug.
+|
+| Random effect        | Effects that are a source of random variation in the data, and whose levels are not fully sampled.
+e.g.
+individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.
+|
+| Mixed effect         | Includes fixed and random effects, e.g.
+the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
+Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
+In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters [see e.g.
+@gelm06, @hox17 and @jong21].
+|
+| ICC                  | The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
+The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.
+|
+| Stratified intercept |                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                  |
+
+## Missingness in multilevel data
+
+As with any other dataset, clustered datasets may be impacted by
+missingness in much the same way. Several strategies can be used to
+handle missing data, including complete case analysis and imputation. We
+focus on the latter approach and discuss statistical methods for
+replacing the missing data with one or more plausible values. Imputation
+separates the missing data problem from the analysis and the completed
+data can be analyzed as if it were completely observed. It is generally
+recommended to impute the missing values more than once to preserve
+uncertainty due to missingness and to allow for valid inferences (c.f.
+Rubin 1976).
+
+With incomplete clustered datasets we can distinguish between two types
+of missing data: sporadic missingness and systematic missingness
+[@resc13]. Sporadic missingness arises when variables are missing for
+some but not all of the units in a cluster [@buur18; @jola18]. For
+example, it is possible that test results are missing for several
+students in one or more classes. When all observations are missing
+within one or more clusters, data are said to be systematically missing.
 <!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"] -->
@@
 
-```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE, fig.margin=TRUE}
+```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
 dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
@@
 ```
+
 <!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing.
@@
 Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}).
-For each of these three missingness generating mechanisms, different imputation strategies are warranted (@yuce08 and @hox15).
+For each of these three missingness generating mechanisms, different imputation strategies are warranted @yuce08 and @hox15.
 We here consider the general case that data are MAR, and expand on certain MNAR situations.
 
-Table 2: Concepts in missing data methods
-
-| **Concept**    | **Details**                                                                |
-|----------------|----------------------------------------------------------------------------|
-| MCAR           | Missing Completely At Random, where the probability to be missing is equal |
-|                | across all data entries                                                    |
-| MAR            | Missing At Random, where the probability to be missing depends on observed |
-|                | information                                                                |
-| MNAR           | Missing Not At Random (MNAR), where the probability to be missing          |
-|                | depends on unrecorded information, making the missingness non-ignorable    |
-|                | [@rubi76; @meng94].                                                        |
-|                |                              |
-
-<!-- [TODO: add paragraph about congeniality and smc-fcs!]  -->
+\begin{table}[tb]
+\caption{Concepts in missing data methods}
+\label{tab:miss}
+\centering
+\begin{tabular}{ll}
+\hline
+\textbf{Concept} & \textbf{Details}   \\
+\hline
+MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
+& across all data entries \\
+MAR     & Missing At Random, where the probability to be missing depends on observed \\
+& information \\
+MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
+& depends on unrecorded information, making the missingness non-ignorable \\
+& [@rubi76; @meng94]. \\
+& [TODO: add congeniality, but maybe in-text?] \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 
-<!-- [TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] -->
-<!-- [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.] -->
+The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
+\pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
+However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent whereas multilevel data are dependent.
+For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.
+
+[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.]
 
@@
 This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice} in \proglang{R}.
-\pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
-\pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18].
-
 We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms.
-For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages [see e.g.
-@audi18 and @grun18] are outside the scope of this tutorial.
-Assumed knowledge includes basic familiarity with the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).
-
-<!-- TODO: mention other packages? \pkg{jomo} and \pkg{mdmb} -->
+For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial.
+[TODO: change this sentence because we do not assume familiarity: "Assumed knowledge includes basic familiarity with multilevel imputation [see e.g.
+@audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod})"].
 
@@
 - `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies with data that are MAR, @metamisc];
-- `obesity` from the \pkg{miceheckman} package [simulated data on obesity, $n = 2,111$ patients across $N = 5$ regions with data that are MNAR].
+- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions with data that are MNAR, @GJRM].
 
@@
 This tutorial is dedicated to readers who are unfamiliar with multiple imputation.
-More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2) or under MNAR conditions (case study 3).
+More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study IMPACT) or  under MNAR conditions (case study HIV).
+
+Imputation of missing data requires consideration of the mechanism
+behind the missingness. Rubin proposed to distinguish between data that
+are missing completely at random (MCAR), data that are missing at random
+(MAR) and data that are missing not at random (MNAR; see Table
+\ref{tab:miss}). For each of these three missingness generating
+mechanisms, different imputation strategies are warranted (@yuce08 and
+@hox15). We here consider the general case that data are MAR, and expand
+on certain MNAR situations.
+
+Table 2: Concepts in missing data methods
+
+| **Concept** | **Details**                                                                                                                                                   |
+|----------------|--------------------------------------------------------|
+| MCAR        | Missing Completely At Random, where the probability to be missing is equal across all data entries                                                            |
+| MAR         | Missing At Random, where the probability to be missing depends on observed information                                                                        |
+| MNAR        | Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]. |
+
+<!-- [TODO: add paragraph about congeniality and smc-fcs!]  -->
+
+<!-- Since excluding observations is not a desirable workflow, the missingness in multilevel data should be accommodated \emph{before} or _within_ the analysis of scientific interest.
+In this paper, we focus on the former approach: imputing (i.e., filling in) the missing data with plausible values, whereafter the completed data may be analyzed as if it were completely observed.
+Imputation separates the missing data problem from the scientific problem, which makes the missing data strategy very generic and popular.
+If each missing value is replaced multiple times, the resulting inferences may validly convey the uncertainty due to missingness [c.f.
+@rubi76].
+-->
+
+<!-- [TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] -->
+
+<!-- [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.] -->
+
+## Aim of this paper
+
+This papers serves as a tutorial for imputing incomplete multilevel data
+with \pkg{mice} in \proglang{R}. \pkg{mice} has become the de-facto
+standard for imputation by chained equations, which iteratively solves
+the missingness on a variable-by-variable basis. \pkg{mice} is known to
+yield valid inferences under many different missing data circumstances
+[@buur18].
+
+We provide practical guidelines and code snippets for different missing
+data situations, including non-ignorable mechanisms. For reasons of
+brevity, we focus on multilevel imputation by chained equations with
+\pkg{mice} exclusively; other imputation methods and packages [see e.g.
+@audi18 and @grun18] are outside the scope of this tutorial. Assumed
+knowledge includes basic familiarity with the \pkg{lme4} notation for
+multilevel models (see Table \ref{tab:mod}).
+
+<!-- TODO: mention other packages? \pkg{jomo} and \pkg{mdmb} -->
+
+We illustrate imputation of incomplete multilevel data using three case
+studies:
+
+-   `popmis` from the \pkg{mice} package [simulated data on perceived
+    popularity, $n = 2,000$ pupils across $N = 100$ schools with data
+    that are MAR, @mice];
+-   `impact` from the \pkg{metamisc} package [empirical data on
+    traumatic brain injuries, $n = 11,022$ patients across $N = 15$
+    studies with data that are MAR, @metamisc];
+-   `obesity` from the \pkg{miceheckman} package [simulated data on
+    obesity, $n = 2,111$ patients across $N = 5$ regions with data that
+    are MNAR].
+
+For each of these datasets, we discuss the nature of the missingness,
+choose one or more imputation models and evaluate the imputed data, but
+we will also highlight one specific aspect of the imputation workflow.
+
+This tutorial is dedicated to readers who are unfamiliar with multiple
+imputation. More experienced readers can skip the introduction (case
+study 1) and directly head to practical applications of multilevel
+imputation under MAR conditions (case study 2) or under MNAR conditions
+(case study 3).
 
@@
 
-<!-- TODO: explicit statement about not going into workings of the methods. Galimer 2l methods. -->
+TODO: explicit statement about not going into workings of the methods. Galimer 2l methods.
 
 ## Setup
+
+[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
+
 <!-- * TODO: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB? -->
@@
 Install non-CRAN packages if necessary:
+
 ```{r packages, eval=FALSE, include=TRUE}
@@
 
-<!-- [TODO: Add environment info, seed and version number(s) somewhere.]  -->
-
 Set up the R environment and load the necessary packages:
+
 ```{r env, message=FALSE, warning=FALSE}
-set.seed(2022)        # for reproducibility
+set.seed(2022)
 library(mice)         # for imputation
-library(miceadds)     # for additional imputation routines
-library(ggmice)       # for incomplete/imputed data visualization
+library(miceadds)     # for imputation
+library(ggmice)       # for visualization
 library(ggplot2)      # for visualization
@@
 library(lme4)         # for multilevel modeling
-library(mitml)        # for multilevel parameter pooling
+library(mitml)        # for multilevel pooling
 library(miceheckman)       # for imputation cf. heckman models
 library(metamisc)     # for case study data
-```
+
+```
+
 <!-- library(broom.mixed)  # for multilevel estimates -->
@@
 
-<!-- [TODO: explain case study] -->
-
-In this section we will go over the different steps involved with imputing incomplete multilevel data with the R package mice.
+[TODO: explain case study]
+
+In this section we'll go over the different steps involved with imputing incomplete multilevel data with the R package mice.
 We consider the simulated `popmis` dataset, which included pupils ($n = 2000$) clustered within schools ($N = 100$).
@@
 
-<!-- TODO: choose sex or gender -->
-
 <!-- $$ -->
+
 <!-- \texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)} -->
+
 <!-- $$ -->
+
 ```{r}
@@
 
+The true effect is:
+
+```{r echo=FALSE}
+fits <-
+  purrr::map_dfr(
+    unique(popcomp$school),
+    ~ {
+      lm(
+        mod,
+        data = filter(popcomp, school == .x)
+      )
+    } %>%
+      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
+      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
+      cbind(school = as.character(.x), .)
+  ) %>% filter(term %in% "sex")
+
+n <- popcomp %>% group_by(school) %>% 
+  tally() %>% 
+  mutate(school = as.character(school))
+
+fits %>% left_join(n, by = "school") %>% 
+  ggplot(aes(y = school, x = estimate, size = n)) +
+  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
+  geom_linerange(aes(y = school, xmin = conf.low, xmax = conf.high, size = 0.5), position = position_dodge(width = 0.5)) +
+  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
+  labs(y = "Study", x = "", color = NULL) +
+  theme_classic() +
+  scale_size(guide = 'none')
+```
+
 <!-- TODO: True effect?
@@
 
-First we plot the pattern of missing data within categories of the relevant variables. Plot the missing data pattern:
-
-```{r pop_pat, fig.cap = "Missing data pattern in the popularity data", fig.margin=TRUE}
+Plot the missing data pattern:
+
+```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
 plot_pattern(popmis)
@@
 
-The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}.
+The missingness is univariate and sporadic, which is illustrated in the
+missing data pattern in Figure \ref{fig:pop_pat}.
 
@@
 
-To develop the best imputation model for the incomplete variable `popular`, we need to know whether the observed values of `popular` are related to observed values of other variables.
-Plot the pair-wise complete correlations in the incomplete data:
+To develop the best imputation model for the incomplete variable
+`popular`, we need to know whether the observed values of `popular` are
+related to observed values of other variables. Plot the pair-wise
+complete correlations in the incomplete data:
 <!-- Excelent, but i dont know if you can refer to the outflux criteria here or elsewhere, cuz in practice it is used to select the predictors in the imputation model from dataset with a lot of features-->
@@
 
-This shows us that  `sex`  may be a useful imputation model predictor. Moreover, the missingness in `popular` may depend on the observed values of other variables.
+This shows us that `sex` may be a useful imputation model predictor.
+Moreover, the missingness in `popular` may depend on the observed values
+of other variables.
 
@@
 -->
+
 <!--  [TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?] -->
@@
 <!-- ```{r pop-cca, echo=FALSE} -->
+
 <!-- est_cca <- lme4::lmer(mod, popmis) %>% broom.mixed::tidy() -->
+
 <!-- # as.matrix(est_cca$estimate) - as.matrix(est_true) -->
+
 <!-- est_cca$estimate - est_true$estimate -->
+
 <!-- ``` -->
@@
 
-The first imputation model that we'll use is likely to be invalid.
-We do *not* use the cluster identifier `school` as imputation model predictor.
-With this model, we ignore the multilevel structure of the data, despite the high ICC.
-This assumes exchangeability between units.
-We include it purely to illustrate the effects of ignoring the clustering in our imputation effort.
+The first imputation model that we'll use is likely to be invalid. We do
+*not* use the cluster identifier `school` as imputation model predictor.
+With this model, we ignore the multilevel structure of the data, despite
+the high ICC. This assumes exchangeability between units. We include it
+purely to illustrate the effects of ignoring the clustering in our
+imputation effort.
 <!-- We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).  -->
 
-Create a methods vector and predictor matrix for `popular`, and make sure `school` is not included as predictor:
+Create a methods vector and predictor matrix for `popular`, and make
+sure `school` is not included as predictor:
 
@@
 Impute the data, ignoring the cluster structure:
-```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
+
+```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
 imp <- mice(popmis, pred = pred, print = FALSE)
@@
 Analyze the imputations:
-```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
+```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
 fit <- with(imp, 
@@
 Print the estimates:
-
-<!-- TODO: explain what this function does -->
-
 ```{r}
@@
 We'll now use `school` as a predictor to impute all other variables.
-This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
-But at least, it includes some multilevel aspect.
-This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16].
-Colloquially, this is 'multilevel imputation for dummies'.
-
-<!-- [TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.] -->
+This is still not recommended practice, since it only works under
+certain circumstances and results may be biased [@drec15; @ende16]. But
+at least, it includes some multilevel aspect. This method is also called
+'fixed cluster imputation', and uses N-1 indicator variables
+representing allocation of N clusters as a fixed factor in the model
+[@reit06; @ende16]. Colloquially, this is 'multilevel imputation for
+dummies'.
+
+[TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.]
 
@@
 
-```{r pop_predictor, message=FALSE, warning=FALSE, cache=TRUE}
+```{r pop_predictor, message=FALSE, warning=FALSE}
 # adjust the predictor matrix
@@
 Analyze the imputations:
-```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
+
+```{r, echo=TRUE, message=FALSE, warning=FALSE}
 fit <- with(imp, 
@@
 Print the estimates:
+
 ```{r}
@@
 
-```{r pop_multilevel, message=FALSE, warning=FALSE, cache=TRUE}
+```{r pop_multilevel, message=FALSE, warning=FALSE}
 # adjust the predictor matrix
@@
 
-<!-- [TODO: check if there is systematic missingness in this dataset, if not make Marshall Computerized Tomography classification (ct) systematically missing.] -->
+[TODO: check if there is systematic missingness in this dataset, if not make Marshall Computerized Tomography classification (ct) systematically missing.]
 
 We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc].
-<!-- [TODO: add more info about the complete data.]  -->
-The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:
+[TODO: add more info about the complete data.] The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:
 
@@
   - `pupil` Pupillary reactivity,
-  - `ct` Marshall Computerized Tomography classification, 
+  - `ct` Marshall Computerized Tomography classification, [TODO: make this one var? also shows that you don't always need random effects everywhere?]
   - `hypox` Hypoxia (0=no, 1=yes),
@@
 
-<!-- [TODO: make ct levels one var? also shows that you don't always need random effects everywhere?] -->
-
 The analysis model for this dataset is a prediction model with `mort` as the outcome.
@@
 The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model:
+
 ```{r mod}
@@
 
-Note that variables `hypots`, `hypox`, `tsah` and `edh` are not part of the analysis model, and may thus serve as auxiliary variables for imputation.
-
-The `impact` data included in the \pkg{metamisc} package is a complete data set.
-The original data has already been imputed once (Steyerberg et al, 2008).
-For the purpose of this tutorial we have induced missingness (mimicking the missing data in the original data set before imputation).
-The resulting incomplete data can be accessed from [zenodo link to be created](https://zenodo.com).
+Note that variables `hypots`, `hypox`, `tsah` and `edh` are not part of
+the analysis model, and may thus serve as auxiliary variables for
+imputation.
+
+The `impact` data included in the \pkg{metamisc} package is a complete
+data set. The original data has already been imputed once (Steyerberg et
+al, 2008). For the purpose of this tutorial we have induced missingness
+(mimicking the missing data in the original data set before imputation).
+The resulting incomplete data can be accessed from [zenodo link to be
+created](https://zenodo.com).
 
 Load the complete and incomplete data into the R workspace:
+
 ```{r data, eval = FALSE}
@@
 
-The estimated effects in the complete data are visualized in Figure \ref{}.
+The estimated effects in the complete data are visualized in Figure
+\ref{}.
 
@@
 fits %>% left_join(n, by = "name") %>% 
-  ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
-  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
-  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
+  ggplot(aes(y = name, x = estimate, color = term, size = n)) +
+  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
+  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, size = 0.5), position = position_dodge(width = 0.5)) +
   geom_point(shape = 20, position = position_dodge(width = 0.5)) +
@@
 ```
+
 <!-- We will use the following estimates as comparative truth in this tutorial: -->
+
 ```{r truth}
@@
 
-<!-- [TODO: show how much variance there is after different methods] -->
-
-<!-- [TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account.
-If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.] -->
-
-<!-- [TODO: add descriptive statistics of the complete and incomplete data.] -->
+[TODO: show how much variance there is after different methods]
+
+[TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account.
+If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.]
+
+[TODO: add descriptive statistics of the complete and incomplete data.]
 
@@
 
-To explore the missingness, it is wise to look at the missing data pattern. The ten most frequent missingness patterns are shown:
+To explore the missingness, it is wise to look at the missing data
+pattern. The ten most frequent missingness patterns are shown:
+
+To explore the missingness, it is wise to look at the missing data pattern:
+
 ```{r pattern, fig.height=7.1}
@@
 
-This shows that we need to impute `ct` and `pupil`.
+This shows... [TODO: fill in that we need to impute `ct` and `pupil`.]
 
 To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables.
-To see whether the missingness depends on the observed values of other variables, we can test this statistically or use visual inspection (e.g.
-a histogram faceted by the missingness indicator).
-
-We should impute the variables `ct` and `pupil` and any auxiliary variables we might want to use to impute these incomplete analysis model variables.
-We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations:
+To see whether the missingness depends on the observed values of other variables, we...
+[TODO: fill in that we can test this statistically or use visual inspection (e.g.
+a histogram faceted by the missingness indicator).]
+
+We should impute the variables `ct` and `pupil` and any auxiliary
+variables we might want to use to impute these incomplete analysis model
+variables. We can evaluate which variables may be useful auxiliaries by
+plotting the pairwise complete correlations:
+
 ```{r impact_corr}
@@
 
-This shows us that `hypox` and `hypot` would not be useful auxiliary variables for imputing `ct`.
-Depending on the minimum required correlation, `tsah` could be useful, while `edh` has the strongest correlation with `ct` out of all the variables in the data and should definitely be included in the imputation model.
-For the imputation of `pupil`, none of the potential auxiliary variables has a very strong relation, but `hypots` could be used.
-We conclude that we can exclude `hypox` from the data, since this is neither an analysis model variable nor an auxiliary variable for imputation:
+This shows us that `hypox` and `hypot` would not be useful auxiliary
+variables for imputing `ct`. Depending on the minimum required
+correlation, `tsah` could be useful, while `edh` has the strongest
+correlation with `ct` out of all the variables in the data and should
+definitely be included in the imputation model. For the imputation of
+`pupil`, none of the potential auxiliary variables has a very strong
+relation, but `hypots` could be used. We conclude that we can exclude
+`hypox` from the data, since this is neither an analysis model variable
+nor an auxiliary variable for imputation:
+
 ```{r}
@@
 
-## Complete case analysis
-
-As previously stated, complete case analysis lowers statistical power and may bias results. The complete case analysis estimates are:
+## Complete case analysis [TODO: remove this?]
+
+As previously stated, complete case analysis lowers statistical power
+and may bias results. The complete case analysis estimates are:
+
 ```{r cca}
@@
 
-As we can see, a higher `ct` (Marshall Computerized Tomography classification) is associated with a lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI ...
-to ..., when controlling for...
+As we can see...
+[TODO: fill in: This means that a higher ct (Marshall Computerized Tomography classification) is associated with a lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI ...
+to ..., when controlling for ...]
 
@@
 <!-- Table 7.1: Questions to gauge the complexity of a multilevel imputation task. -->
+
 <!-- 1. Will the complete-data model include random slopes? -->
+
 <!-- 2. Will the data contain systematically missing values? -->
+
 <!-- 3. Will the distribution of the residuals be non-normal? -->
+
 <!-- 4. Will the error variance differ over clusters? -->
+
 <!-- 5. Will there be small clusters? -->
+
 <!-- 6. Will there be a small number of clusters? -->
+
 <!-- 7. Will the complete-data model have cross-level interactions? -->
+
 <!-- 8. Will the dataset be very large? (fimd, section 7.3)-->
@@
 <!-- --- -->
-Mutate data to get the right data types for imputation (e.g. integer for clustering variable).
+
+Mutate data to get the right data types for imputation (e.g. integer for
+clustering variable).
 
@@
 
-Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
+Create a methods vector and predictor matrix, and make sure `name` is
+not included as predictor, but as clustering variable:
+
 ```{r impact}
@@
 pred <- quickpred(dat)   # predictor matrix
-plot_pred(pred, rotate = TRUE)
+plot_pred(pred)
 
@@
 diag(pred) <- 0
-plot_pred(pred, rotate = TRUE)
+plot_pred(pred)
 
@@
 
+<!-- We'll now use `school` as a predictor to impute all other variables.
+This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
+But at least, it includes some multilevel aspect.
+This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16].
+Colloquially, this is 'multilevel imputation for dummies'.
+-->
+
+<!-- Add: doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->
+
+<!-- The more the random effects are of interest, the more you need ml models. -->
+
 Impute the incomplete data
-```{r imp_impact, cache=TRUE}
+
+```{r imp_impact}
 imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
@@
 
-```{r impact-fit, cache=TRUE}
+```{r impact-fit}
 fit <- imp %>% 
@@
 
-# Case study III: obesity data
-
-Data are simulated and included in the `miceheckman` package. We will use the following variables:
+# Case study III: HIV data
+
+Data are simulated and included in the `miceheckman` package. We will
+use the following variables:
+
+Data are simulated and included in the `GJRM` package. We will use the following variables:
+
+- `region` Cluster variable,
+- `hiv` HIV diagnosis (0=no, 1=yes),
+- `age` Age of the patient,
+- `marital` Marital status,
+- `condom` Condom use during last intercourse,
+- `smoke` Smoker (levels; inclusion restriction variable).
+
+The imputation of these date is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).
+
+```{r hiv, echo=FALSE}
+# data
+data("hiv", package = "GJRM")
+hiv <- hiv %>% 
+  select(c("region","hiv", "age", "marital", "condom", "smoke", "interviewerID")) %>% 
+  mutate(region = as.integer(region))
+# md pattern
+plot_pattern(hiv)
+```
 
@@
 -   `weight` Weight in kilograms,
-- `rt`` Response time in minutes (inclusion-restriction variable).
-
-The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
-
-<!-- [TODO: add reference Arxiv paper Johanna].  -->
-
-Load the data:
-```{r obesity}
+-   \`time\`\` Response time in minutes (exclusion restriction
+    variable).
+
+The imputation of these data is based on the [IPDMA Heckman Github
+repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
+
+load the data
+```{r}
 data("obesity",  package = "miceheckman")
+
+```
+
+We obtain here the random effects for each interviewer, this is an approximation of the interviewer's skill which will be used as an exclusion constraint.
+Here, since the location of the interviewer was not randomly assigned to the subjects, the assignment was corrected for region and language.
+
+Set the Heckman model as imputation method
+```{r}
+# Set prediction matrix and methods
+#ini <- mice(obesity, maxit = 0)
+#meth<-ini$method
+#meth["weight"]<-"2l.heckman"
+#pred <- ini$pred
+#pred["weight","cluster"]<- -2
+#pred["weight","rt"] <- -3
 ```
@@
 Select relevant variables:
+
 ```{r bmi}
-dat <- obesity
+#dat <- select(obesity, -"bmi")
 ```
@@
 Visualize missing data pattern:
+
 ```{r obesity-md}
-plot_pattern(dat)
+#plot_pattern(dat)
 ```
@@
 Create predictor matrix:
+
 ```{r obesity-pred}
-pred <- quickpred(dat)   # predictor matrix
-pred["Weight","Cluster"]<- -2 # clustering variable
-pred["Weight", ] <- -3 #inclusion-restriction variable
-plot_pred(pred)
+#pred <- quickpred(dat)   # predictor matrix
+#pred["weight","cluster"]<- -2 # clustering variable
+#pred["weight","rt"] <- -3 #inclusion-restriction variable
+#plot_pred(pred)
 ```
@@
 Set the Heckman model as imputation method:
+
 ```{r meth}
-meth <- make.method(dat) # methods vector
-meth["weight"]<-"2l.heckman"
+#meth <- make.method(dat) # methods vector
+#meth["weight"]<-"2l.heckman"
 ```
@@
 Impute the missingness:
-```{r eval=FALSE, cache=TRUE}
-imp <- mice(
-  dat, # dataset with missing values
-  m = 5, # number of imputations
-  maxit = 10,
-  meth = meth, #imputation method vector
-  pred = pred, #imputation predictors matrix
-  meta_method = "reml",
-  printFlag = FALSE
-)
+```{r eval=FALSE}
+#imp <- mice(obesity, # dataset with missing values
+#                        m = 10,   # number of imputations
+#                        maxit = 1,
+#                        seed = 1234, #seed attached to the dataID
+#                        meth = meth, #imputation method vector
+#                        pred = pred, #imputation predictors matrix
+#                        print = T,
+#                        meta_method="reml",
+#                        pmm=FALSE)
+
 ```
@@
 
-<!-- - JOMO in \pkg{mice} -> on the side for now -->
+- JOMO in \pkg{mice} -> on the side for now
+
 <!-- # look at jomo for categorical variables? -->
+
 <!-- # semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach -->
+
 <!-- # pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters -->
+
 <!-- # make assumptions of these methods explicit! -->
@@
 
+-   More complex data types: timeseries and polynomial relationship in
+    the clustering.
+
+-   FIML vs MI
+
+An alternative approach to missing data is to use Full Information
+Maximum Likelihood (FIML). This method does not require the imputation
+of any missing values. Whereas MI consists of imputation, analyses and
+pooling steps, FIML analyses the data in a single step. When the
+assumptions are met the two approaches should produce equivalent
+results. [REF] As FIML requires specialised software, not all analyses
+can be performed with standard software. [REF]
+
+-   Survival / TTE, this could be put in the paragraph on congeniality
+
+When the outcome is time-to-event, the Nelson-Aalen estimate of the time
+to event should be included as a covariate in the imputation model [REF]
+
+- Additional levels of clustering
+
 - More complex data types: timeseries and polynomial relationship in the clustering.
 
-- FIML vs MI
-
-An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML).
-This method does not require the imputation of any missing values.
-Whereas MI consists of  imputation, analyses and pooling steps, FIML analyses the data in a single step.
-When the assumptions are met the two approaches should produce equivalent results.
-[REF] As FIML requires specialised software, not all analyses can be performed with standard software.
-[REF]
-
-- Survival / TTE, this could be put in the paragraph on congeniality
-
-When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]
-
-<!-- # Think about -->
-
-<!-- - Adding evaluations of the imputations such as convergence checks -->
-
-<!-- TODO in the future: Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model. -->
-
-<!-- TODO in the future: Adding a `multilevel_ampute()` wrapper function in mice. -->
-
-<!-- - Exporting `mids` objects to other packages like `lme4` or `coxme`?  -->
-
-<!-- - Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt. -->
-
-<!-- - Show use case for deductive imputation for cluster level variables? -->
-
-<!-- - env dump in repo -->
-
-<!-- -  I don't know if in your article you cover something about model complexity, for example sometimes I have to switch from 2l.
+# Think about
+
+- Adding evaluations of the imputations such as convergence checks
+
+- Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model.
+
+- Adding a `multilevel_ampute()` wrapper function in mice.
+
+- Exporting `mids` objects to other packages like `lme4` or `coxme`?
+
+- Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt.
+
+- Show use case for deductive imputation for cluster level variables?
+
+- env dump in repo
+
+-  I don't know if in your article you cover something about model complexity, for example sometimes I have to switch from 2l.
 methods to 1l.
@@
 how to test which is better between 2l.norm or 2l.2stage.norm.
--->
-
-In hierarchical datasets, clustering is a concern because the homoscedasticity in the error terms cannot be assumed across clusters and the relationship among variables may vary at different hierarchical levels.
-When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g.
-if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021).
-Not including clustering into the imputation process may lead to effect estimates with smaller standard errors and inflated type I error.
-
-There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/)
-
-The selection of each strategy depends mainly on the assumptions in the main analysis and also on the restriction of the analyzed data.
-
-Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
-The last restriction is also required when imputations are performed on each cluster separately.
-When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).
-
-Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modelled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
-However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.
-the number of clusters is adequate.
-(Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008).
-Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020).
-For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
+
+In hierarchical datasets, clustering is a concern because the
+homoscedasticity in the error terms cannot be assumed across clusters
+and the relationship among variables may vary at different hierarchical
+levels. When multiple imputation is used to deal with missing data, as
+the imputation and analysis process is performed separately, it is
+necessary that imputation model being congenial with the main analysis
+model (Meng, 1994), e.g. if the main model accounts for the hierarchical
+structure also imputation model should do it (Audigier, 2021). Not
+including clustering into the imputation process may lead to effect
+estimates with smaller standard errors and inflated type I error.
+
+There are different strategies that can be adopted in the imputation
+process that account for clustering: inclusion of cluster indicator
+variable, performing a separate imputation process for each cluster, or
+performing a simultaneous imputation process by using an imputation
+method that accounts for clustering.(Stata:
+<https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/>)
+
+The selection of each strategy depends mainly on the assumptions in the
+main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of
+cluster indicator variables is restricted in datasets where there are
+not many clusters and many observations per cluster (Graham, 2009). The
+last restriction is also required when imputations are performed on each
+cluster separately. When this restriction cannot be achieved, one can
+use an imputation model that simultaneously imputes all clusters using a
+hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters
+are correlated and this correlation is modelled by a random effect so
+the hierarchical model can be estimated even when there are few
+observations per cluster. However, this strategy is best suited for
+balanced data (Grund, 2017) and when random effects model is
+appropriated, i.e. the number of clusters is adequate. (Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main
+model, for instance by using the cluster indicator strategy may lead to
+bias estimates when the model is based on a hierarchical model
+(<https://onlinelibrary.wiley.com/doi/full/10.1002/bimj.201900051>).
+Even when an imputation strategy congenial with the main model is
+preferred, it is important to consider whether it is appropriate for the
+data as a less complex imputation strategies may also lead to unbiased
+estimates in certain scenarios(Bailey 2020). For instance, in causal
+effect analysis, separately imputation may lead to smaller bias when the
+size of the smaller exposure cluster is large, compared with an
+imputation model that includes exposure-confounder interactions.
 (Zhang,2023).
@@
 
-This project has received funding from the European Union's Horizon 2020 research and innovation programme under ReCoDID grant agreement No 825746.
-
-The views expressed in this paper are the personal views of the authors and may not be understood or quoted as being made on behalf of or reflecting the position of the regulatory agency/agencies or organizations with which the authors are employed/affiliated.
+This project has received funding from the European Union's Horizon 2020
+research and innovation programme under ReCoDID grant agreement No
+825746.
+
+The views expressed in this paper are the personal views of the authors
+and may not be understood or quoted as being made on behalf of or
+reflecting the position of the regulatory agency/agencies or
+organizations with which the authors are employed/affiliated.
 
@@
 <!-- Comment dump -->
+
 <!-- Missing data pattern with clustering -->
+
 <!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
+
 <!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
+
 <!-- TODO: check spss missing data pie chart? -->
+
 <!-- TODO: similarity of patterns over clusters -->
+
 <!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
+
 <!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
+
 <!-- TODO: look at vim -->
+
 <!-- TODO: check if shepley plot is an option -->
@@
 <!-- Conditional plots missingness indicator -->
+
 <!-- TODO: check if up-side-down plot works -->
+
 <!-- TODO: check smoothing in geom_density function and make it the inverse of the sample size -->
+
 <!-- TODO: add functions to mice -->
+
 <!-- TODO: add title and informative legend (pop obs and pop miss) -->
+
 <!-- TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .` -->
+
 <!-- TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon -->
+
 <!-- TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?) -->
@@
 
-Table 3: Notation
-
-| **Formula \pkg{lme4}**         | **Details**                                 |
-|-------------------------------------------------------------|--------------------------------------------------|
-| `y ~ x1 + (1 | g1)`                                      | Fixed `x1` predictor with random intercept         |
-|                                                             | varying among `g1`                                  |
-| `y~x1*x2+ (1| g1)`                                       | Interactions of `x1` and `x2` only in fixed effect   |
-| `y ~ x1*x2+ (x2| g1)`                                      | Interactions of `x1` and `x2` only in fixed effect   |
-|                                                             | with slope of `x2` randomly varying among `g1`       |
-| `y ~ x1*x2+ (x1*x2| g1)`                                   | variance-covariance matrix estimated only with   |
-|                                                             | the variance terms of intercept, slope of `x1`,    |
-|                                                             | slope of `x2` and interaction `x1*x2`                |
-| `y ~ x1*x2+ (x1 | g1)+ (x2| g1)`                            | variance-covariance matrix estimated separately, |
-|                                                             | i.e, one for intercept and `x1` and another for    |
-|                                                             | intercept and `x2`                                 |
-| `y ~ x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)`             | Fixed `x1` with correlated random intercept and    |
-|                                                             | random slope of `x`                                |
-| `y ~ x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1)` | Fixed `x1` with uncorrelated random intercept      |
-|                                                             | and random slope of `x1`                           |
-| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2   |
-| `y ~ (1 | g1/g2) or ~ (1 | g1) + (1 | g1:g2)`                     | Random intercept varying among `g1` and `g2`         |
-|                                                             | withing `g1`                                       |
-|                                                             |          |
+Table 3: Notation based on https://datalorax.github.io/equatiomatic/articles/lmer4-lmer.html
+https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf
+
+| **Formula** \pkg{lme4}                                     | **Details**                                                                                                                                 |
+|---------------------------------------|---------------------------------|
+| `y ~ x1 + (1|g1)`                                          | Fixed `x1` predictor with random intercept varying among `g1`                                                                               |
+| `y ~ x1 + (x1 | g1) or 1 + x1 + ( 1+x1 | g1)`              | Fixed `x1` with correlated random intercept and random slope of `x1`                                                                        |
+| `y ~ x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1)` | Fixed `x1` with uncorrelated random intercept and random slope of `x1`                                                                      |
+| `y ~ x1 + x2 + (x1|g1)+ (x2|g1)`                           | variance-covariance matrix estimated separately for each fixed predictor, i.e, one for intercept and `x1` and another forintercept and `x2` |
+| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2`                                                                                          |
+
+Table 4: Imputation methods for hierarchical data, based on Audigier 2018 study
+comparison.
+
+| Type approach | Imputation method                                      | Type of variables | Systematically missing data | Missing mechanism | Hetero-scedasticity | Details                                                                         |
+|---------------|--------------------------------------------------------|-------------------|-----------------------------|-------------------|---------------------|---------------------------------------------------------------------------------|
+| JM            | Pan                                                    | C-B               | Yes                         | MAR               |                     |                                                                                 |
+|               | Jomo (Quartagno and Carpenter,2016)                    | C-B               | Yes                         | MAR               | Y                   | Time consuming,recommended for large clusters when proportion of binary is high |
+| FCS           | Pan                                                    | C                 | Yes                         | MAR               |                     |                                                                                 |
+|               | 2lnorm                                                 | C                 | Yes                         | MAR               |                     |                                                                                 |
+|               | GLM (Jolani,2017)                                      | C-B               | No                          | MAR               | N                   | Recommended for small clusters                                                  |
+|               | 2l.stage (Resche-Rigon and White,2016, Audigier, 2018) | C-B               | Yes                         | MAR/MNAR          | Y                   | Combine estimates with random intercept model, recommended for large studies    |

### 2023-05-31 Manuscript/Imputation_of_Incomplete_Multilevel_Data2.tex
update table 1 and 2 tables in appendix
Please just refer to changes in tables, other changes i did was to solve running problems

--- 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 \author{
-Hanne I. Oberman\\Methodology and Statistics\\
-Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
-and Primary Care,\\
+Hanne I. Oberman\\Utrecht University \And Johanna Munoz\\Julius Center
+for Health Sciences and Primary Care,\\
 University Medical Center Utrecht, Utrecht University,\\
 Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
-for Health Sciences and Primary Care,\\
-University Medical Center Utrecht, Utrecht University,\\
-Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
-Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
-Health Sciences and Primary Care,\\
+for Health Sciences and Primary Care \And Gerko Vink\\Utrecht University
+\AND Valentijn M. T. de Jong\\Julius Center for Health Sciences and
+Primary Care,\\
 University Medical Center Utrecht, Utrecht University,\\
@@
 
-\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
+\Plainauthor{Hanne I. Oberman, Johanna Munoz, Thomas P. A. Debray, Gerko
 Vink, Valentijn M. T. de Jong}
@@
 This tutorial illustrates the imputation of incomplete multilevel data
-with the \proglang{R} packackage \pkg{mice}.
+with the \proglang{R} packackage \pkg{mice}. Footnotes in the current
+version show work in progress/under construction. The last section is
+not part of the manuscript, but purely for reminders. See also all of
+the TODOs that need to be worked out. We aim to submit at JSS, so there
+is no word count limit (``There is no page limit, nor a limit on the
+number of figures or tables''). {[}Just adding some text to get a better
+guess of what the actura abstract will look like: Lorem ipsum dolor sit
+amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut
+labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud
+exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum
+dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non
+proident, sunt in culpa qui officia deserunt mollit anim id est
+laborum.{]}
 }
@@
     Hanne I. Oberman\\
-    Methodology and Statistics\\
     Utrecht University\\
@@
 \section{Introduction}\label{introduction}}
+
+Many datasets include individuals from multiple settings, geographic
+regions, or even different studies. In the simplest case, individuals
+(e.g., students) are nested within so-called clusters (e.g., school
+classes). More complex clustered structures may occur when there are
+multiple hierarchical levels (e.g., patients within hospitals within
+regions or countries), or when the clustering is non-nested (e.g.,
+electronic health record data from diverse settings and populations
+within large databases). In general, individuals from the same cluster
+tend to be more similar than individuals from other clusters. In
+statistical terms, this implies that observations from the same cluster
+are correlated. If this correlation is left unaddressed, estimates of
+\emph{p} values, confidence intervals even model parameters are prone to
+bias \citep{loca01}. {[}TODO: make a link to imputation methods, which
+require adequate handling and propagation of variance; we are not
+recommending the adoption of multilevel models for data analysis here,
+but rather for imputation. VMTdJ: ``I would only make that link after
+introducing missing data, and missing data is not mentioned yet, so I
+would do that in the next section''{]} Statistical methods for clustered
+data typically adopt hierarchical models that explicitly describe the
+grouping of observations. These models are also known as `multilevel
+models', `hierarchical models', `mixed effect models', `random effect
+models', and in the context of time-to-event data as `frailty models'.
+Table \ref{tab:clus} provides an overview of some key concepts in
+multilevel modeling.
+
+\hypertarget{missingness-in-multilevel-data}{%
+\subsection{Missingness in multilevel
+data}\label{missingness-in-multilevel-data}}
+
+Like any other dataset, clustered datasets are prone to missing data.
+Several strategies can be used to handle missing data, including
+complete case analysis and imputation. We focus on the latter approach
+and discuss statistical methods for replacing the missing data with one
+or more plausible values. Afterwards, the completed data can be analyzed
+as if they were completely observed. In contrast to single imputation
+(where missing data are only replaced once), multiple imputation allows
+to preserve uncertainty due to missingness and is therefore recommended
+(c.f. Rubin 1976).
+
+When clustered datasets are affected by missing values, we can
+distinguish between two types of missing data: sporadic missingness and
+systematic missingness \citep{resc13}. Sporadic missingness arises when
+variables are missing for some but not all of the units in a cluster
+\citep{buur18, jola18}. For example, it is possible that test results
+are missing for several students in one or more classes. {[}TODO:
+Provide an example for one of the case studies below.{]} When all
+observations are missing within one or more clusters, data are
+systematically missing. {[}TODO: Refer to Figure 1 and put
+interpretation in the figure caption. VMTdJ: ``Mention in text that this
+is sporadic missingness?''{]}
+
+\hypertarget{introduction-1}{%
+\section{Introduction}\label{introduction-1}}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2361}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7639}}@{}}
 \toprule()
@@
 Hierarchical levels & Data are grouped into clusters at different
-levels. A three-level \\
-Fixed effect & Here we assume that the values of an independent variable
-are fixed, i.e., \\
-& the values observed in the study are representative of all values in
-the \\
-& in the dependent variable y e.g., blood pressure between treatments A
-and B. \\
-Random effect & The values of an independent variable are assumed to be
-randomly drawn from \\
-& admission we might select only certain hospitals that are
-representative of \\
-& the difference of y between individual hospitals, but rather the
-variation of \\
+levels, observations belonging to the same cluster are expected to share
+certain characteristics. \\
+Fixed effect & Effects that are constant across all sample units,
+e.g.~something that researchers control for and can repeat, such as the
+administration of a drug. \\
+Random effect & Effects that are a source of random variation in the
+data, and whose levels are not fully sampled. e.g.~individuals are drawn
+from a population of hospitals, here it is not possible to sample all
+hospitals but drug effects could vary between hospitals. \\
+Mixed effect & Includes fixed and random effects, e.g.~the fixed effect
+would be the treatment effect of a drug and the random effect would be
+the ID of the hospital where the patient is treated. Multilevel models
+typically accommodate for variability by including a separate group mean
+for each cluster e.g random intercept on hospitals. In addition to
+random intercepts, multilevel models can also include random
+coefficients and heterogeneous residual error variances across clusters
+\citep[see e.g.][\citet{hox17} and \citet{jong21}]{gelm06}. \\
 ICC & The variability due to clustering is often measured by means of
-the \\
-& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
-& of variance that can be attributed to the cluster-level, where a
-high \\
-& ICC would indicate that a lot of variability is due to the cluster \\
-& structure. \\
-Random effect & Multilevel models typically accommodate for variability
-by including \\
-& a separate group mean for each cluster. In addition to random \\
-& intercepts, multilevel models can also include random coefficients \\
-& and heterogeneous residual error variances across clusters {[}see
-e.g. \\
-& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
-stratified intercept as concept..{]} \\
+the intraclass coefficient (ICC). The ICC can be seen as the percentage
+of variance that can be attributed to the cluster-level, where a high
+ICC would indicate that a lot of variability is due to the cluster
+structure. \\
+Stratified intercept & \\
 \bottomrule()
@@
 
-\hypertarget{missingness-in-multilevel-data}{%
+\hypertarget{missingness-in-multilevel-data-1}{%
 \subsection{Missingness in multilevel
-data}\label{missingness-in-multilevel-data}}
+data}\label{missingness-in-multilevel-data-1}}
 
@@
 
-{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/patterns-1}
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/patterns-1}
 
@@
 
-\caption[Sporadic missingness in multilevel data]{Sporadic missingness in multilevel data}\label{fig:patterns}
+\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
 \end{figure}
 \end{CodeChunk}
+
+Imputation of missing data requires consideration of the mechanism
+behind the missingness. Rubin proposed to distinguish between data that
+are missing completely at random (MCAR), data that are missing at random
+(MAR) and data that are missing not at random (MNAR; see Table
+\ref{tab:miss}). For each of these three missingness generating
+mechanisms, different imputation strategies are warranted \citet{yuce08}
+and \citet{hox15}. We here consider the general case that data are MAR,
+and expand on certain MNAR situations.
+
+\begin{table}[tb]
+\caption{Concepts in missing data methods}
+\label{tab:miss}
+\centering
+\begin{tabular}{ll}
+\hline
+\textbf{Concept} & \textbf{Details}   \\
+\hline
+MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
+& across all data entries \\
+MAR     & Missing At Random, where the probability to be missing depends on observed \\
+& information \\
+MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
+& depends on unrecorded information, making the missingness non-ignorable \\
+& [@rubi76; @meng94]. \\
+& [TODO: add congeniality, but maybe in-text?] \\
+\hline
+\end{tabular}
+\end{table}
+
+The \proglang{R} package \pkg{mice} has become the de-facto standard for
+imputation by chained equations, which iteratively solves the
+missingness on a variable-by-variable basis. \pkg{mice} is known to
+yield valid inferences under many different missing data circumstances
+\citep{buur18}. However, commonly used imputation methods were not
+designed for use in clustered data and usually generate observations
+that are independent whereas multilevel data are dependent. For this
+reason, we discuss how the \proglang{R} package \pkg{mice} can be used
+to impute multilevel data.
+
+{[}TODO: clarify why clustering is relevant during imputation, and why
+this exposes the need for specialized imputation methods and more
+attention during their implementation (``thou shall not simply run
+\texttt{mice()} on any incomplete dataset'').{]} {[}TODO: Add that the
+more the random effects are of interest, the more you need multilevel
+imputation models.{]} {[}TODO: Add an overview of all possible predictor
+matrix values in manuscript or \texttt{ggmice} legend.{]}
+
+\hypertarget{aim-of-this-paper}{%
+\subsection{Aim of this paper}\label{aim-of-this-paper}}
+
+This papers serves as a tutorial for imputing incomplete multilevel data
+with \pkg{mice} in \proglang{R}. We provide practical guidelines and
+code snippets for different missing data situations, including
+non-ignorable mechanisms. For reasons of brevity, we focus on multilevel
+imputation by chained equations with \pkg{mice} exclusively; other
+imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are
+outside the scope of this tutorial. {[}TODO: change this sentence
+because we do not assume familiarity: ``Assumed knowledge includes basic
+familiarity with multilevel imputation \citep[see e.g.][ and
+\citet{grun18}]{audi18} and the \pkg{lme4} notation for multilevel
+models (see Table \ref{tab:mod})''{]}.
+
+We illustrate imputation of incomplete multilevel data using three case
+studies:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
+  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
+  with data that are MAR,][]{mice};
+\item
+  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
+  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
+  studies with data that are MAR,][]{metamisc};
+\item
+  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
+  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions with data
+  that are MNAR,][]{GJRM}.
+\end{itemize}
+
+For each of these datasets, we discuss the nature of the missingness,
+choose one or more imputation models and evaluate the imputed data, but
+we will also highlight one specific aspect of the imputation workflow.
+
+This tutorial is dedicated to readers who are unfamiliar with multiple
+imputation. More experienced readers can skip the introduction (case
+study 1) and directly head to practical applications of multilevel
+imputation under MAR conditions (case study IMPACT) or under MNAR
+conditions (case study HIV).
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2222}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7778}}@{}}
 \toprule()
@@
 MCAR & Missing Completely At Random, where the probability to be missing
-is equal \\
-& across all data entries \\
+is equal across all data entries \\
 MAR & Missing At Random, where the probability to be missing depends on
-observed \\
-& information \\
-MNAR & Missing Not At Random (MNAR), where the probability to be
-missing \\
-& depends on unrecorded information, making the missingness
-non-ignorable \\
-& \citep{rubi76, meng94}. \\
-& \\
+observed information \\
+MNAR & Missing Not At Random (MNAR), where the probability to be missing
+depends on unrecorded information, making the missingness non-ignorable
+\citep{rubi76, meng94}. \\
 \bottomrule()
@@
 
-\hypertarget{aim-of-this-paper}{%
-\subsection{Aim of this paper}\label{aim-of-this-paper}}
+\hypertarget{aim-of-this-paper-1}{%
+\subsection{Aim of this paper}\label{aim-of-this-paper-1}}
 
@@
 
-Table 3: Notation
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Formula \pkg{lme4}}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Details}
-\end{minipage} \\
-\midrule()
-\endhead
-\texttt{y\ \textasciitilde{}\ x1\ +\ (1\ \textbar{}\ g1)} & Fixed
-\texttt{x1} predictor with random intercept \\
-& varying among \texttt{g1} \\
-\texttt{y\textasciitilde{}x1*x2+\ (1\textbar{}\ g1)} & Interactions of
-\texttt{x1} and \texttt{x2} only in fixed effect \\
-\texttt{y\ \textasciitilde{}\ x1*x2+\ (x2\textbar{}\ g1)} & Interactions
-of \texttt{x1} and \texttt{x2} only in fixed effect \\
-& with slope of \texttt{x2} randomly varying among \texttt{g1} \\
-\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1*x2\textbar{}\ g1)} &
-variance-covariance matrix estimated only with \\
-& the variance terms of intercept, slope of \texttt{x1}, \\
-& slope of \texttt{x2} and interaction \texttt{x1*x2} \\
-\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1\ \textbar{}\ g1)+\ (x2\textbar{}\ g1)}
-& variance-covariance matrix estimated separately, \\
-& i.e, one for intercept and \texttt{x1} and another for \\
-& intercept and \texttt{x2} \\
-\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ +\ x1\ \textbar{}\ g1)}
-& Fixed \texttt{x1} with correlated random intercept and \\
-& random slope of \texttt{x} \\
-\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ \textbar{}\ g1)\ +\ (0\ +\ x1\ \textbar{}\ g1)}
-& Fixed \texttt{x1} with uncorrelated random intercept \\
-& and random slope of \texttt{x1} \\
-\texttt{y\ \textasciitilde{}\ (1\ \textbar{}\ g1)\ +\ (1\ \textbar{}\ g2)}
-& Random intercept varying among \texttt{g1} and among
-\texttt{g2\ \ \ \textbar{}\ \textbar{}}y \textasciitilde{} (1 \\
-& \\
-\bottomrule()
-\end{longtable}
-
 We illustrate imputation of incomplete multilevel data using three case
@@
 
+{[}TODO: Add environment info, seed and version number(s) somewhere.{]}
+Set up the R environment and load the necessary packages:
+
 Install non-CRAN packages if necessary:
@@
 \begin{CodeInput}
-R> set.seed(2022)        # for reproducibility
+R> set.seed(2022)
 R> library(mice)         # for imputation
-R> library(miceadds)     # for additional imputation routines
-R> library(ggmice)       # for incomplete/imputed data visualization
+R> library(miceadds)     # for imputation
+R> library(ggmice)       # for visualization
 R> library(ggplot2)      # for visualization
@@
 R> library(lme4)         # for multilevel modeling
-R> library(mitml)        # for multilevel parameter pooling
-R> library(miceheckman)  # for imputation cf heckman models
-R> library(metamisc)     # for ???
+R> library(mitml)        # for multilevel pooling
+R> library(miceheckman)       # for imputation cf. heckman models
+R> library(metamisc)     # for case study data
 \end{CodeInput}
@@
 
-In this section we will go over the different steps involved with
-imputing incomplete multilevel data with the R package mice. We consider
-the simulated \texttt{popmis} dataset, which included pupils
-(\(n = 2000\)) clustered within schools (\(N = 100\)). The following
-variables are of primary interest:
+{[}TODO: explain case study{]}
+
+In this section we'll go over the different steps involved with imputing
+incomplete multilevel data with the R package mice. We consider the
+simulated \texttt{popmis} dataset, which included pupils (\(n = 2000\))
+clustered within schools (\(N = 100\)). The following variables are of
+primary interest:
 
@@
 \begin{CodeChunk}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\end{CodeChunk}
+
+Load the data into the environment and select the relevant variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> popmis <- popmis[, c("school", "popular", "sex")] 
+\end{CodeInput}
+\end{CodeChunk}
+
+Plot the missing data pattern:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(popmis)
+\end{CodeInput}
 \begin{figure}
 
-{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-3-1}
-
-}
-
-\caption[True effect in the popularity data]{True effect in the popularity data}\label{fig:unnamed-chunk-3}
-\end{figure}
-\end{CodeChunk}
-
-Load the data into the environment and select the relevant variables:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> popmis <- popmis[, c("school", "popular", "sex")] 
-\end{CodeInput}
-\end{CodeChunk}
-
-First we plot the pattern of missing data within categories of the
-relevant variables. Plot the missing data pattern:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_pattern(popmis, square = TRUE)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pat-1}
+{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop_pat-1}
 
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-corr-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop-corr-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-hist-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop-hist-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-pred-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop-ignored-pred-1} \end{center}
 
@@
 
+{[}TODO: Add that it doesn't work with systematic missingness (only with
+sporadic). There's some pros and cons, and it may not even differ much
+if the number of clusters is low.{]}
+
 \begin{CodeChunk}
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop_predictor-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_multilevel-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop_multilevel-1} \end{center}
 
@@
 
+{[}TODO: check if there is systematic missingness in this dataset, if
+not make Marshall Computerized Tomography classification (ct)
+systematically missing.{]}
+
 We illustrate how to impute incomplete multilevel data by means of a
@@
 \citep[empirical data on traumatic brain injuries, \(n = 11,022\) units
-across \(N = 15\) clusters,][]{metamisc}. The \texttt{impact} data set
-contains traumatic brain injury data on \(n = 11022\) patients clustered
-in \(N = 15\) studies with the following 11 variables:
+across \(N = 15\) clusters,][]{metamisc}. {[}TODO: add more info about
+the complete data.{]} The \texttt{impact} data set contains traumatic
+brain injury data on \(n = 11022\) patients clustered in \(N = 15\)
+studies with the following 11 variables:
 
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/forest-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/forest-1} \end{center}
 
@@
 
+{[}TODO: show how much variance there is after different methods{]}
+
+{[}TODO: add ICC before/after imputation and interpret: This tells us
+that the multilevel structure of the data should probably be taken into
+account. If we don't, we'll may end up with incorrect imputations,
+biasing the effect of the clusters towards zero.{]}
+
+{[}TODO: add descriptive statistics of the complete and incomplete
+data.{]}
+
 \hypertarget{missingness}{%
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pattern-1} \end{center}
-
-\end{CodeChunk}
-
-This shows that we need to impute \texttt{ct} and \texttt{pupil}.
+To explore the missingness, it is wise to look at the missing data
+pattern:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pattern-1} \end{center}
+
+\end{CodeChunk}
+
+This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
+and \texttt{pupil}.{]}
 
@@
 values of the other variables. To see whether the missingness depends on
-the observed values of other variables, we can test this statistically
-or use visual inspection (e.g.~a histogram faceted by the missingness
-indicator).
+the observed values of other variables, we\ldots{} {[}TODO: fill in that
+we can test this statistically or use visual inspection (e.g.~a
+histogram faceted by the missingness indicator).{]}
 
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_corr-1} \end{center}
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/impact_corr-1} \end{center}
 
@@
    <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
- 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
+ 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   3.00e-31   0.0571     0.130
  2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
@@
  9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
-10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
+10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.56e-19   1.92       2.76 
 11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
@@
 
-As we can see, a higher \texttt{ct} (Marshall Computerized Tomography
-classification) is associated with a lower odds of 6-month mortality,
-given by the odds ratio exp(0.42), CI \ldots{} to \ldots, when
-controlling for\ldots{}
+As we can see\ldots{} {[}TODO: fill in: This means that a higher ct
+(Marshall Computerized Tomography classification) is associated with a
+lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI
+\ldots{} to \ldots, when controlling for \ldots{]}
 
@@
 R> pred <- quickpred(dat)   # predictor matrix
-R> plot_pred(pred, rotate = TRUE)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-1} \end{center}
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/impact-1} \end{center}
 
@@
 R> diag(pred) <- 0
-R> plot_pred(pred, rotate = TRUE)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-2} \end{center}
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/impact-2} \end{center}
 
@@
                      term    estimate   std.error  statistic      p.value
-1             (Intercept) -2.35199614 0.340174191  -6.914093 4.993068e-12
-2                    type -0.41267936 0.180271952  -2.289204 2.208652e-02
-3                     age  0.03049024 0.001570154  19.418637 1.233010e-81
-4 as.factor(motor_score)2 -0.66765325 0.068737838  -9.713038 3.478858e-22
-5 as.factor(motor_score)3 -1.05520488 0.070218678 -15.027410 2.536132e-50
-6 as.factor(motor_score)4 -1.51238928 0.072304221 -20.917026 1.857373e-90
-7                   pupil  0.48421506 0.038982924  12.421209 6.781796e-17
-8                      ct  0.43474556 0.029968441  14.506779 2.341875e-36
+1             (Intercept) -2.35203726 0.340181747  -6.914061 4.994037e-12
+2                    type -0.41265892 0.180274846  -2.289054 2.209524e-02
+3                     age  0.03049023 0.001570162  19.418521 1.238416e-81
+4 as.factor(motor_score)2 -0.66764920 0.068737865  -9.712975 3.480413e-22
+5 as.factor(motor_score)3 -1.05520001 0.070218940 -15.027285 2.540225e-50
+6 as.factor(motor_score)4 -1.51238349 0.072304262 -20.916934 1.850073e-90
+7                   pupil  0.48421447 0.038982800  12.421234 6.772069e-17
+8                      ct  0.43474621 0.029968474  14.506785 2.342253e-36
              b          df dfcom         fmi      lambda m         riv
-1 5.301620e-04 10113.60453 11013 0.005694385 0.005497777 5 0.005528170
-2 4.905995e-05 10892.82806 11013 0.001994781 0.001811557 5 0.001814845
-3 3.333628e-08  6323.59900 11013 0.016537098 0.016226102 5 0.016493731
-4 4.194026e-05  8322.17049 11013 0.010889419 0.010651742 5 0.010766424
-5 5.136864e-05  7631.13821 11013 0.012760548 0.012501842 5 0.012660117
-6 1.403437e-04  2830.62134 11013 0.032897234 0.032214161 5 0.033286456
-7 3.571667e-04    49.96887 11013 0.309144148 0.282035215 5 0.392825972
-8 8.586575e-05   294.70102 11013 0.120676307 0.114728924 5 0.129597506
+1 5.281599e-04 10119.70041 11013 0.005673265 0.005476771 5 0.005506932
+2 4.881699e-05 10893.89378 11013 0.001985736 0.001802528 5 0.001805783
+3 3.335358e-08  6320.89582 11013 0.016545467 0.016234340 5 0.016502243
+4 4.188703e-05  8327.24771 11013 0.010875748 0.010638213 5 0.010752602
+5 5.135721e-05  7632.20045 11013 0.012757637 0.012498967 5 0.012657168
+6 1.403058e-04  2831.75462 11013 0.032888241 0.032205434 5 0.033277139
+7 3.571497e-04    49.97295 11013 0.309130930 0.282023647 5 0.392803531
+8 8.586645e-05   294.69765 11013 0.120677044 0.114729598 5 0.129598366
           ubar
-1 1.150823e-01
-2 3.243910e-02
-3 2.425379e-06
-4 4.674562e-03
-5 4.869020e-03
-6 5.059488e-03
-7 1.091068e-03
-8 7.950686e-04
+1 1.150898e-01
+2 3.244044e-02
+3 2.425385e-06
+4 4.674630e-03
+5 4.869071e-03
+6 5.059539e-03
+7 1.091079e-03
+8 7.950697e-04
 \end{CodeOutput}
@@
             (Intercept)                     type                      age  
-               -2.37194                 -0.41015                  0.03052  
+               -2.37195                 -0.41014                  0.03052  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
@@
             (Intercept)                     type                      age  
-               -2.32327                 -0.42365                  0.03023  
+               -2.32339                 -0.42359                  0.03023  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67143                 -1.05776                 -1.51039  
+               -0.67142                 -1.05776                 -1.51038  
                   pupil                       ct  
@@
             (Intercept)                     type                      age  
-               -2.33576                 -0.40872                  0.03039  
+               -2.33581                 -0.40871                  0.03039  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66477                 -1.05453                 -1.51860  
+               -0.66477                 -1.05451                 -1.51858  
                   pupil                       ct  
                 0.45928                  0.44419  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
+optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
 
@@
             (Intercept)                     type                      age  
-               -2.35183                 -0.40576                  0.03064  
+               -2.35187                 -0.40574                  0.03064  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67469                 -1.06551                 -1.52622  
+               -0.67468                 -1.06551                 -1.52622  
                   pupil                       ct  
@@
 
-\hypertarget{case-study-iii-obesity-data}{%
-\section{Case study III: obesity
-data}\label{case-study-iii-obesity-data}}
+\hypertarget{case-study-iii-hiv-data}{%
+\section{Case study III: HIV data}\label{case-study-iii-hiv-data}}
 
@@
 
+Data are simulated and included in the \texttt{GJRM} package. We will
+use the following variables:
+
 \begin{itemize}
@@
 \item
+  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{marital} Marital status,
+\item
+  \texttt{condom} Condom use during last intercourse,
+\item
+  \texttt{smoke} Smoker (levels; inclusion restriction variable).
+\end{itemize}
+
+The imputation of these date is based on the toy example from
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}.
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/hiv-1} \end{center}
+
+\end{CodeChunk}
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{region} Cluster variable,
+\item
   \texttt{gender} Gender (0=male, 1=female),
@@
 \item
-  `rt`` Response time in minutes (inclusion-restriction variable).
+  `time`` Response time in minutes (exclusion restriction variable).
 \end{itemize}
@@
 
-Load the data:
+load the data
 
@@
 
+We obtain here the random effects for each interviewer, this is an
+approximation of the interviewer's skill which will be used as an
+exclusion constraint. Here, since the location of the interviewer was
+not randomly assigned to the subjects, the assignment was corrected for
+region and language.
+
+Set the Heckman model as imputation method
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # Set prediction matrix and methods
+R> #ini <- mice(obesity, maxit = 0)
+R> #meth<-ini$method
+R> #meth["weight"]<-"2l.heckman"
+R> #pred <- ini$pred
+R> #pred["weight","cluster"]<- -2
+R> #pred["weight","rt"] <- -3
+\end{CodeInput}
+\end{CodeChunk}
+
 Select relevant variables:
@@
 \begin{CodeInput}
-R> dat <- obesity
+R> #dat <- select(obesity, -"bmi")
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> plot_pattern(dat)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1} \end{center}
-
+R> #plot_pattern(dat)
+\end{CodeInput}
 \end{CodeChunk}
@@
 \begin{CodeInput}
-R> pred <- quickpred(dat)   # predictor matrix
-R> pred["Weight","Cluster"]<- -2 # clustering variable
-R> pred["Weight", ] <- -3 #inclusion-restriction variable
-R> plot_pred(pred)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-pred-1} \end{center}
-
+R> #pred <- quickpred(dat)   # predictor matrix
+R> #pred["weight","cluster"]<- -2 # clustering variable
+R> #pred["weight","rt"] <- -3 #inclusion-restriction variable
+R> #plot_pred(pred)
+\end{CodeInput}
 \end{CodeChunk}
@@
 \begin{CodeInput}
-R> meth <- make.method(dat) # methods vector
-R> meth["weight"]<-"2l.heckman"
+R> #meth <- make.method(dat) # methods vector
+R> #meth["weight"]<-"2l.heckman"
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> imp <- mice(
-+   dat, # dataset with missing values
-+   m = 5, # number of imputations
-+   maxit = 10,
-+   meth = meth, #imputation method vector
-+   pred = pred, #imputation predictors matrix
-+   meta_method = "reml",
-+   printFlag = FALSE
-+ )
+R> #imp <- mice(obesity, # dataset with missing values
+R> #                        m = 10,   # number of imputations
+R> #                        maxit = 1,
+R> #                        seed = 1234, #seed attached to the dataID
+R> #                        meth = meth, #imputation method vector
+R> #                        pred = pred, #imputation predictors matrix
+R> #                        print = T,
+R> #                        meta_method="reml",
+R> #                        pmm=FALSE)
 \end{CodeInput}
@@
 \section{Discussion}\label{discussion}}
+
+\begin{itemize}
+\tightlist
+\item
+  JOMO in \pkg{mice} -\textgreater{} on the side for now
+\end{itemize}
 
@@
 {[}REF{]}
+
+\begin{itemize}
+\item
+  Additional levels of clustering
+\item
+  More complex data types: timeseries and polynomial relationship in the
+  clustering.
+\end{itemize}
+
+\hypertarget{think-about}{%
+\section{Think about}\label{think-about}}
+
+\begin{itemize}
+\item
+  Adding evaluations of the imputations such as convergence checks
+\item
+  Adding some kind of help function to mice that suggests a suitable
+  predictor matrix to the user, given a certain analysis model.
+\item
+  Adding a \texttt{multilevel\_ampute()} wrapper function in mice.
+\item
+  Exporting \texttt{mids} objects to other packages like \texttt{lme4}
+  or \texttt{coxme}?
+\item
+  Adding a ICC=0 dataset to show that even if there is no clustering it
+  doesn't hurt.
+\item
+  Show use case for deductive imputation for cluster level variables?
+\item
+  env dump in repo
+\item
+  I don't know if in your article you cover something about model
+  complexity, for example sometimes I have to switch from 2l. methods to
+  1l. methods just because the model didn't converge.. this is due to
+  the considered imputation model is very complex regarding the amount
+  of information counted\ldots{} I know that a solution for an
+  imputation model with many predictors is to check correlation plots as
+  you did or use quickpred().. (maybe you can add this somewhere after
+  the correlation plots)\ldots..But as for cluster specification, I
+  don't know if besides plots of distribution per cluster there is
+  something else can be done to see if i have to use 1l. or 2l. for a
+  given variable, also I have no idea.. how to test which is better
+  between 2l.norm or 2l.2stage.norm.
+\end{itemize}
 
@@
 bias estimates when the model is based on a hierarchical model
-(Taaljard,2008). Even when an imputation strategy congenial with the
-main model is preferred, it is important to consider whether it is
-appropriate for the data as a less complex imputation strategies may
-also lead to unbiased estimates in certain scenarios(Bailey 2020). For
-instance, in causal effect analysis, separately imputation may lead to
-smaller bias when the size of the smaller exposure cluster is large,
-compared with an imputation model that includes exposure-confounder
-interactions. (Zhang,2023).
+(\url{https://onlinelibrary.wiley.com/doi/full/10.1002/bimj.201900051}).
+Even when an imputation strategy congenial with the main model is
+preferred, it is important to consider whether it is appropriate for the
+data as a less complex imputation strategies may also lead to unbiased
+estimates in certain scenarios(Bailey 2020). For instance, in causal
+effect analysis, separately imputation may lead to smaller bias when the
+size of the smaller exposure cluster is large, compared with an
+imputation model that includes exposure-confounder interactions.
+(Zhang,2023).
 
@@
 
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
+Table 3: Notation based on
+\url{https://datalorax.github.io/equatiomatic/articles/lmer4-lmer.html}
+\url{https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf}
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5417}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4583}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Formula} \pkg{lme4}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Details}
+\end{minipage} \\
+\midrule()
+\endhead
+\texttt{y\ \textasciitilde{}\ x1\ +\ (1\textbar{}g1)} & Fixed
+\texttt{x1} predictor with random intercept varying among \texttt{g1} \\
+\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (\ 1+x1\ \textbar{}\ g1)}
+& Fixed \texttt{x1} with correlated random intercept and random slope of
+\texttt{x1} \\
+\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ \textbar{}\ g1)\ +\ (0\ +\ x1\ \textbar{}\ g1)}
+& Fixed \texttt{x1} with uncorrelated random intercept and random slope
+of \texttt{x1} \\
+\texttt{y\ \textasciitilde{}\ x1\ +\ x2\ +\ (x1\textbar{}g1)+\ (x2\textbar{}g1)}
+& variance-covariance matrix estimated separately for each fixed
+predictor, i.e, one for intercept and \texttt{x1} and another
+forintercept and \texttt{x2} \\
+\texttt{y\ \textasciitilde{}\ (1\ \textbar{}\ g1)\ +\ (1\ \textbar{}\ g2)}
+& Random intercept varying among \texttt{g1} and among \texttt{g2} \\
+\bottomrule()
+\end{longtable}
+
+Table 4: Imputation methods for hierarchical data, based on Audigier
+2018 study comparison.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0625}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.2333}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0792}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1208}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0792}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0875}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.3375}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Type approach
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Imputation method
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Type of variables
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Systematically missing data
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Missing mechanism
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Hetero-scedasticity
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Details
+\end{minipage} \\
+\midrule()
+\endhead
+JM & Pan & C-B & Yes & MAR & & \\
+& Jomo (Quartagno and Carpenter,2016) & C-B & Yes & MAR & Y & Time
+consuming,recommended for large clusters when proportion of binary is
+high \\
+FCS & Pan & C & Yes & MAR & & \\
+& 2lnorm & C & Yes & MAR & & \\
+& GLM (Jolani,2017) & C-B & No & MAR & N & Recommended for small
+clusters \\
+& 2l.stage (Resche-Rigon and White,2016, Audigier, 2018) & C-B & Yes &
+MAR/MNAR & Y & Combine estimates with random intercept model,
+recommended for large studies \\
+\bottomrule()
+\end{longtable}
+
 \bibliography{../References/multilevelmice.bib}

### 2023-10-02 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Re-knit pdf

--- 2023-05-26 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
   This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}.
+Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data.
+More complex models can be accomodated, but are outside the scope of this paper.
 keywords:
@@
 devtools::install_github("amices/ggmice")
-devtools::install_github("hanneoberman/miceheckman")
+# devtools::install_github("hanneoberman/miceheckman")
 ```
@@
 library(mitml)        # for multilevel parameter pooling
-library(miceheckman)  # for imputation cf. heckman models
+library(micemd)  # for imputation cf. heckman models
 library(metamisc)     # for case study data
@@
 ```{r obesity}
-data("obesity", package = "miceheckman")
+data("data_heckman", package = "micemd")
 ```
@@
 ```{r bmi}
-dat <- obesity
+dat <- data_heckman
 ```

### 2023-10-02 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Re-knit pdf

@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 Sample unit & Units of the population from which measurements are taken
@@
 stratified intercept as concept..{]} \\
-\bottomrule()
 \end{longtable}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 MCAR & Missing Completely At Random, where the probability to be missing
@@
 & \\
-\bottomrule()
 \end{longtable}
@@
 R> devtools::install_github("amices/ggmice")
-R> devtools::install_github("hanneoberman/miceheckman")
+R> # devtools::install_github("hanneoberman/miceheckman")
 \end{CodeInput}
@@
 R> library(mitml)        # for multilevel parameter pooling
-R> library(miceheckman)  # for imputation cf. heckman models
+R> library(micemd)  # for imputation cf. heckman models
 R> library(metamisc)     # for case study data
@@
 \begin{CodeInput}
-R> plot_pattern(popmis, square = TRUE)
+R> plot_pattern(popmis)
 \end{CodeInput}
@@
 
+TODO: explain exclusion restriction.
+
 Data are simulated and included in the \texttt{miceheckman} package. We
@@
 \item
-  `rt`` Response time in minutes (inclusion-restriction variable).
+  `time`` Response time in minutes (inclusion-restriction variable).
 \end{itemize}
@@
 \begin{CodeInput}
-R> data("obesity", package = "miceheckman")
+R> data("data_heckman", package = "micemd")
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> dat <- obesity
+R> dat <- data_heckman
 \end{CodeInput}
@@
 
+The matrix only shows the predictors for the main model, not the
+selection model.
+
 Create predictor matrix:
@@
 R> pred <- quickpred(dat)   # predictor matrix
-R> pred["Weight","Cluster"]<- -2 # clustering variable
-R> pred["Weight", ] <- -3 #inclusion-restriction variable
+R> pred[,"Cluster"]<- -2 # clustering variable
+R> pred[, "Weight"] <- -3 #inclusion-restriction variable
 R> plot_pred(pred)
@@
 
+ORDER:
+
 \begin{itemize}
+\item
+  summary
+\item
+  congeniality, then in hierarchical models
+\item
+  look whether we can fit cong. back in the main body
+\item
+  alt. methods
+\item
+  conclusion: mice is really easy!
 \item
@@
 \url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
+TODO: replace ref.
 
@@
 Under this hierarchical imputation model, observations within clusters
-are correlated and this correlation is modelled by a random effect so
-the hierarchical model can be estimated even when there are few
-observations per cluster. However, this strategy is best suited for
-balanced data (Grund, 2017) and when random effects model is
-appropriated, i.e.~the number of clusters is adequate. (Austin,2018).
+are correlated and this correlation is modeled by a random effect so the
+hierarchical model can be estimated even when there are few observations
+per cluster. However, this strategy is best suited for balanced data
+(Grund, 2017) and when random effects model is appropriated, i.e.~the
+number of clusters is adequate. (Austin,2018).
 
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 \texttt{y\ \textasciitilde{}\ x1\ +\ (1\ \textbar{}\ g1)} & Fixed

### 2023-10-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update case3

@@
 
-Table 1: Concepts in multilevel methods
-
-| **Concept**         | **Details**                                                                   |
-|---------------------|-------------------------------------------------------------------------------|
-| Sample unit         | Units of the population from which measurements are taken in a sample.        |
-| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
-| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
-|                     | the values observed in the study are representative of all values in the      |
-|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
-| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
-|                     | admission we might select only certain hospitals that are representative of   |
-|                     | the difference of y between individual hospitals, but rather the variation of |
-| ICC                 | The variability due to clustering is often measured by means of the           |
-|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
-|                     | of variance that can be attributed to the cluster-level, where a high         |
-|                     | ICC would indicate that a lot of variability is due to the cluster            |
-|                     | structure.                                                                    |
-| Random effect       | Multilevel models typically accommodate for variability by including          |
-|                     | a separate group mean for each cluster. In addition to random                 |
-|                     | intercepts, multilevel models can also include random coefficients            |
-|                     | and heterogeneous residual error variances across clusters [see e.g.          |
-|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |
+```{r clus,  echo = FALSE, warning = FALSE,cache = TRUE}
+concepts <- data.frame(Concept=c("Sample unit","Hierarchical levels","Fixed effect","Random effect","Mixed effect","ICC","Stratified intercept"),
+                       Details=c("Units of the population from which measurements are taken in a sample.",
+                                 "Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.",
+                                 "Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.",
+                                 "Effects that are a source of random variation in the data, and whose levels are not fully sampled.
+e.g.
+individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.",
+                                 "Includes fixed and random effects, e.g.
+the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
+Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
+In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
+@gelm06, @hox17 and @jong21).",
+                                 "The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
+of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
+                                 ""))
+library(kableExtra)
+concepts%>%
+  kable( booktabs = T, escape = F,
+         caption = "Concepts in multilevel methods",
+         col.names = c("Concept","Details"), align = "l")%>%
+  column_spec(c(1), width = "3cm")%>%
+  column_spec(c(2), width = "12cm")
+```
 
@@
 - `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies with data that are MAR, @metamisc];
-- `obesity` from the \pkg{miceheckman} package [simulated data on obesity, $n = 2,111$ patients across $N = 5$ regions with data that are MNAR].
+- `obesity` from the \pkg{micemd} package [simulated data on obesity, $n = 2,111$ patients across $N = 5$ regions with data that are MNAR].
 
@@
 devtools::install_github("amices/ggmice")
-# devtools::install_github("hanneoberman/miceheckman")
+
 ```
@@
 
-TODO: explain exclusion restriction.
-
-Data are simulated and included in the `miceheckman` package. We will use the following variables:
+In this example, we demonstrate a multilevel imputation of an intercept and slope random model with a continuous response.
+We utilize the obesity dataset included in the `micemd` package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries.
+To easy the explanations we simulate data for 5 clusters and we reduce the dataset to the following variables:
 
@@
 - `time`` Response time in minutes (inclusion-restriction variable).
+- `FamOb`` Family obesity history (yes or not)
 
-The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
+Age and FamOb are missing according to a Missing At Random (MAR) mechanism, whereas the weight variable is subject to selection bias, following an indirect Missing Not At Random (MNAR) mechanism, that usually happens when it is unobserved or omitted a variable affecting both the value of incomplete variable (here weight) and its probability of missing.
 
-<!-- [TODO: add reference Arxiv paper Johanna].  -->
+The analysis model considers BMI as the dependent variable, predicted by Age, Height, Gender, and FAV. Random slopes are considered for the Age variable, which contains missing values.
 
-Load the data:
+The model is represented as:
+$$BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FAV_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}$$
+We load the data:
 ```{r obesity}
 data("data_heckman", package = "micemd")
-```
-
-Select relevant variables:
-```{r bmi}
 dat <- data_heckman
@@
 
-Visualize missing data pattern:
+Now, let's begin by examining the missing patterns in the data by cluster:
 ```{r obesity-md}
@@
 
-The matrix only shows the predictors for the main model, not the selection model.
+We observe that the proportion of missingness for the Age and FAVC variables is quite similar across the clusters.
+However, regarding the weight variable, we notice a heterogeneous proportion of missingness, with systematic missingness in cluster 5.
+In order to evaluate if it is required a imputation method for 1-level or 2-level we assess the Intraclass Correlation Coefficient (ICC) for the outcome variable, we use the "performance" package:
+
+```{r obesity icc}
+dat$bmi=dat$Weight/(dat$Height)^2
+Nulmodel <- lme4::lmer(bmi ~ 1 + (1|Cluster), data = dat)
+#Performance::ICC(Nulmodel)
+```
+
+Since the ICC is above 0.1, we decide to employ two-level (2l) imputation methods.
+In this imputation process, we include all predictor variables in the main model and since BMI is a composite variable.
+We also incorporate weight and height variables.
+
+To determine the best imputation method for these variables, we can use the `find.defaultMethod`function provided in the `micemd` package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters.
+It suggests using "2l.2stage.bin" for the FAV variable and "2l.2stage.norm" for the age and weight variables.
+However, after inspecting the age density plot, we consider modifying its method to "2l.2stage.ppm." For the BMI variable, we employ deterministic imputation.
+```{r obesity meth}
+meth_mar <- find.defaultMethod(dat, ind.clust=7, I.small = 7, ni.small = 100, prop.small = 0.4)
+```
+For these imputation models, it is necessary to specify the prediction matrix, with the cluster variable labeled as -2 and the predictor variable labeled as 2, encompassing all variables.
+
+After confirming convergence, we proceed to fit the main model and save the results for future use.
+Users are considering the possibility that the weight variable may not have been selected randomly.
+It's likely that an omitted variable, like self-esteem, could influence this selection.
+For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.
+
+To address this situation, two approaches have been proposed for dealing with Missing Not at Random (MNAR) data: pattern-mixed models and selection models.
+Within pattern-mixed models, methods like the delta method and more advanced ones like NARFS have been suggested.
+In contrast, the selection model approach includes methods such as the Heckman model, which can be particularly useful in this case.
+Several methods, including Galimar2017, Hammon2021, and the recent Munoz method, designed for two-level data, allowing for variations in intercepts and exposure effects (random intercept and slope).
 
@@
 ```
+
+To apply the Munoz method, the weight variable should be specified as "2l.2stage.heckman" found in the micemd package.
+Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based on certain variables.
+
+TODO: explain exclusion restriction.
+In this specific case, it's assumed that height does not influence selection (-4), and an exclusion restriction variable is introduced as an instrumental variable affecting selection but not the outcome.
+Time of response is used for this purpose, as it's believed that subjects taking more time to complete the survey are more likely to not respond due to poor connectivity or lack of familiarity with electronic surveys (specified as -3).
+
+After implementing these imputation procedures, it's crucial to check for convergence.
+If convergence issues arise and imputed values are far from plausible, especially affecting BMI values, a post-imputation procedure is employed to constrain BMI values within a plausible range.
+
+Then after this modification we proceed to compare the effects on the model
+###Plot estimates
+
+The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
+
+<!-- [TODO: add reference Arxiv paper Johanna].  -->
+
+Visualize missing data pattern:
+
+The matrix only shows the predictors for the main model, not the selection model.
 

### 2023-10-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update case3

@@
 This tutorial illustrates the imputation of incomplete multilevel data
-with the \proglang{R} packackage \pkg{mice}.
+with the \proglang{R} packackage \pkg{mice}. Our scope is only simple
+multilevel models, to show how imputation can yield less biased
+estimates from incomplete clustered data. More complex models can be
+accomodated, but are outside the scope of this paper.
 }
@@
 
-Table 1: Concepts in multilevel methods
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Concept}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Details}
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Sample unit & Units of the population from which measurements are taken
-in a sample. \\
-Hierarchical levels & Data are grouped into clusters at different
-levels. A three-level \\
-Fixed effect & Here we assume that the values of an independent variable
-are fixed, i.e., \\
-& the values observed in the study are representative of all values in
-the \\
-& in the dependent variable y e.g., blood pressure between treatments A
-and B. \\
-Random effect & The values of an independent variable are assumed to be
-randomly drawn from \\
-& admission we might select only certain hospitals that are
-representative of \\
-& the difference of y between individual hospitals, but rather the
-variation of \\
-ICC & The variability due to clustering is often measured by means of
-the \\
-& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
-& of variance that can be attributed to the cluster-level, where a
-high \\
-& ICC would indicate that a lot of variability is due to the cluster \\
-& structure. \\
-Random effect & Multilevel models typically accommodate for variability
-by including \\
-& a separate group mean for each cluster. In addition to random \\
-& intercepts, multilevel models can also include random coefficients \\
-& and heterogeneous residual error variances across clusters {[}see
-e.g. \\
-& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
-stratified intercept as concept..{]} \\
-\end{longtable}
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:clus}Concepts in multilevel methods}
+\centering
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
+\toprule
+Concept & Details\\
+\midrule
+Sample unit & Units of the population from which measurements are taken in a sample.\\
+Hierarchical levels & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
+Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.\\
+Random effect & Effects that are a source of random variation in the data, and whose levels are not fully sampled.
+e.g.
+individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.\\
+Mixed effect & Includes fixed and random effects, e.g.
+the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
+Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
+In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
+@gelm06, @hox17 and @jong21).\\
+\addlinespace
+ICC & The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
+of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.\\
+Stratified intercept & \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
 
@@
 \item
-  \texttt{obesity} from the \pkg{miceheckman} package {[}simulated data
-  on obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
+  \texttt{obesity} from the \pkg{micemd} package {[}simulated data on
+  obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
   that are MNAR{]}.
@@
 R> devtools::install_github("amices/ggmice")
-R> # devtools::install_github("hanneoberman/miceheckman")
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
+R> plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
 \end{CodeInput}
@@
 
-TODO: explain exclusion restriction.
-
-Data are simulated and included in the \texttt{miceheckman} package. We
-will use the following variables:
+In this example, we demonstrate a multilevel imputation of an intercept
+and slope random model with a continuous response. We utilize the
+obesity dataset included in the \texttt{micemd} package, a simulated
+dataset that emulates an electronic survey in which individuals are
+asked to provide information about their weight and consumption habits
+in different countries. To easy the explanations we simulate data for 5
+clusters and we reduce the dataset to the following variables:
 
@@
   `time`` Response time in minutes (inclusion-restriction variable).
+\item
+  `FamOb`` Family obesity history (yes or not)
 \end{itemize}
 
-The imputation of these data is based on the
-\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
-Heckman Github repo}
-
-Load the data:
+Age and FamOb are missing according to a Missing At Random (MAR)
+mechanism, whereas the weight variable is subject to selection bias,
+following an indirect Missing Not At Random (MNAR) mechanism, that
+usually happens when it is unobserved or omitted a variable affecting
+both the value of incomplete variable (here weight) and its probability
+of missing.
+
+The analysis model considers BMI as the dependent variable, predicted by
+Age, Height, Gender, and FAV. Random slopes are considered for the Age
+variable, which contains missing values.
+
+The model is represented as:
+\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FAV_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
+We load the data:
 
@@
 R> data("data_heckman", package = "micemd")
-\end{CodeInput}
-\end{CodeChunk}
-
-Select relevant variables:
-
-\begin{CodeChunk}
-\begin{CodeInput}
 R> dat <- data_heckman
@@
 
-Visualize missing data pattern:
+Now, let's begin by examining the missing patterns in the data by
+cluster:
 
@@
 
-The matrix only shows the predictors for the main model, not the
-selection model.
+We observe that the proportion of missingness for the Age and FAVC
+variables is quite similar across the clusters. However, regarding the
+weight variable, we notice a heterogeneous proportion of missingness,
+with systematic missingness in cluster 5. In order to evaluate if it is
+required a imputation method for 1-level or 2-level we assess the
+Intraclass Correlation Coefficient (ICC) for the outcome variable, we
+use the ``performance'' package:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> dat$bmi=dat$Weight/(dat$Height)^2
+R> Nulmodel <- lme4::lmer(bmi ~ 1 + (1|Cluster), data = dat)
+R> #Performance::ICC(Nulmodel)
+\end{CodeInput}
+\end{CodeChunk}
+
+Since the ICC is above 0.1, we decide to employ two-level (2l)
+imputation methods. In this imputation process, we include all predictor
+variables in the main model and since BMI is a composite variable. We
+also incorporate weight and height variables.
+
+To determine the best imputation method for these variables, we can use
+the \texttt{find.defaultMethod}function provided in the \texttt{micemd}
+package, which suggests an appropriate method for MAR variables based on
+the type of variable, number of observations in the cluster, and number
+of clusters. It suggests using ``2l.2stage.bin'' for the FAV variable
+and ``2l.2stage.norm'' for the age and weight variables. However, after
+inspecting the age density plot, we consider modifying its method to
+``2l.2stage.ppm.'' For the BMI variable, we employ deterministic
+imputation.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth_mar <- find.defaultMethod(dat, ind.clust=7, I.small = 7, ni.small = 100, prop.small = 0.4)
+\end{CodeInput}
+\end{CodeChunk}
+
+For these imputation models, it is necessary to specify the prediction
+matrix, with the cluster variable labeled as -2 and the predictor
+variable labeled as 2, encompassing all variables.
+
+After confirming convergence, we proceed to fit the main model and save
+the results for future use. Users are considering the possibility that
+the weight variable may not have been selected randomly. It's likely
+that an omitted variable, like self-esteem, could influence this
+selection. For instance, individuals with lower self-esteem might have
+higher weight values, impacting their willingness to provide honest
+information due to embarrassment.
+
+To address this situation, two approaches have been proposed for dealing
+with Missing Not at Random (MNAR) data: pattern-mixed models and
+selection models. Within pattern-mixed models, methods like the delta
+method and more advanced ones like NARFS have been suggested. In
+contrast, the selection model approach includes methods such as the
+Heckman model, which can be particularly useful in this case. Several
+methods, including Galimar2017, Hammon2021, and the recent Munoz method,
+designed for two-level data, allowing for variations in intercepts and
+exposure effects (random intercept and slope).
 
@@
 \end{CodeChunk}
+
+To apply the Munoz method, the weight variable should be specified as
+``2l.2stage.heckman'' found in the micemd package. Additionally, the
+prediction matrix needs modification because this method involves
+specifying two equations: one for the outcome, describing the incomplete
+variable in terms of partially observed predictors (in this case, all
+variables from the main model), and the other for the selection model,
+explaining the probability of being observed based on certain variables.
+
+TODO: explain exclusion restriction. In this specific case, it's assumed
+that height does not influence selection (-4), and an exclusion
+restriction variable is introduced as an instrumental variable affecting
+selection but not the outcome. Time of response is used for this
+purpose, as it's believed that subjects taking more time to complete the
+survey are more likely to not respond due to poor connectivity or lack
+of familiarity with electronic surveys (specified as -3).
+
+After implementing these imputation procedures, it's crucial to check
+for convergence. If convergence issues arise and imputed values are far
+from plausible, especially affecting BMI values, a post-imputation
+procedure is employed to constrain BMI values within a plausible range.
+
+Then after this modification we proceed to compare the effects on the
+model \#\#\#Plot estimates
+
+The imputation of these data is based on the
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}
+
+Visualize missing data pattern:
+
+The matrix only shows the predictors for the main model, not the
+selection model.
 

### 2023-10-27 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Add convergence plots

--- 2023-10-02 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 title:
-  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
-  plain:     "Imputation of Incomplete Multilevel Data with mice"
+  formatted: "Imputation of Incomplete Multilevel Data with \\proglang{R}"
+  plain:     "Imputation of Incomplete Multilevel Data with R"
   short:     "Multilevel \\pkg{mice}"
@@
 # Introduction
+
+## Multilevel data
 
@@
 When all observations are missing within one or more clusters, data are said to be systematically missing.
+Sporadic missingness is visualized in Figure XYZ.
 <!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"] -->
@@
 
-Install non-CRAN packages if necessary:
-```{r packages, eval=FALSE, include=TRUE}
-devtools::install_github("amices/ggmice")
-# devtools::install_github("hanneoberman/miceheckman")
-```
-
 <!-- [TODO: Add environment info, seed and version number(s) somewhere.]  -->
@@
 ```{r env, message=FALSE, warning=FALSE}
-set.seed(2022)        # for reproducibility
+set.seed(123)         # for reproducibility
 library(mice)         # for imputation
@@
 library(mitml)        # for multilevel parameter pooling
-library(micemd)  # for imputation cf. heckman models
+library(micemd)       # for case study data and imputation cf. heckman models
 library(metamisc)     # for case study data
+library(broom.mixed)  # for multilevel estimates
 ```
-<!-- library(broom.mixed)  # for multilevel estimates -->
+
+TODO: add table with predictor matrix values
+
+- -2 = cluster variable 
+- 1 = overall effect
+- 3 = overall + group-level effect 
+- 4 = individual-level (random) and group-level (fixed) effect
 
@@
 ```
+
+The estimated effects in the complete data are presented in Table XYZ. We consider the associations in the full data set to be the true associations.
 
@@
 
- <!-- The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.
+The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.
 This tells us that the multilevel structure of the data should probably be taken into account.
-If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.-->
+If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.
 
@@
 
+Evaluate the convergence of the algorithm:
+```{r pop-ignored-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
+plot_trace(imp)
+```
 <!-- [TODO: remove the broom.mixed output, use mitml only] -->
@@
 
-```{r}
+```{r pop-print, eval=FALSE}
 testEstimates(as.mitml.result(fit), extra.pars = TRUE)
@@
 
+Evaluate the convergence of the algorithm:
+```{r pop-predictor-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
+plot_trace(imp)
+```
+
 Analyze the imputations:
@@
 Print the estimates:
-```{r}
+```{r, eval=FALSE}
 testEstimates(as.mitml.result(fit), extra.pars = TRUE)
@@
 
+Evaluate the convergence of the algorithm:
+```{r pop-multilevel-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
+plot_trace(imp)
+```
+
 Analyze the imputations:
@@
 
-The estimated effects in the complete data are visualized in Figure \ref{}.
+The estimated effects in the complete data are presented in Table XYZ.
+
+<!-- visualized in Figure \ref{}. -->
 
 ```{r forest, echo=FALSE}
-fits <-
-  purrr::map_dfr(
-    unique(impact$name),
-    ~ {
-      glm(
-        mort ~ 1 + age + motor_score + pupil + ct,
-        family = "binomial",
-        data = filter(impact, name == .x)
-      )
-    } %>%
-      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
-      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
-      cbind(name = as.character(.x), .)
-  ) %>% filter(term %in% c("ctIII", "ctIV/V"))
-
-n <- impact %>% group_by(name) %>% 
-  tally()
-
-fits %>% left_join(n, by = "name") %>% 
-  ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
-  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
-  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
-  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
-  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
-  theme_classic() +
-  scale_size(guide = 'none')
-# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
-# TODO: missingness prop inside clusters
-# TODO: make row height proportional to pattern frequency
+# fits <-
+#   purrr::map_dfr(
+#     unique(impact$name),
+#     ~ {
+#       glm(
+#         mort ~ 1 + age + motor_score + pupil + ct,
+#         family = "binomial",
+#         data = filter(impact, name == .x)
+#       )
+#     } %>%
+#       broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
+#       .[, c("term", "estimate", "conf.low", "conf.high")] %>%
+#       cbind(name = as.character(.x), .)
+#   ) %>% filter(term %in% c("ctIII", "ctIV/V"))
+# 
+# n <- impact %>% group_by(name) %>% 
+#   tally()
+# 
+# fits %>% left_join(n, by = "name") %>% 
+#   ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
+#   geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
+#   geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
+#   geom_point(shape = 20, position = position_dodge(width = 0.5)) +
+#   labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
+#   theme_classic() +
+#   scale_size(guide = 'none')
+# # TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
+# # TODO: missingness prop inside clusters
+# # TODO: make row height proportional to pattern frequency
 ```
@@
 
+Evaluate the convergence of the algorithm:
+```{r impact-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
+plot_trace(imp)
+```
+
+Analyze the imputed data:
 ```{r impact-fit, cache=TRUE}
@@
 
-Data are simulated and included in the `miceheckman` package. We will use the following variables:
+Data are simulated and included in the `micemd` package. We will use the following variables:
 

### 2023-10-27 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Add convergence plots

--- 2023-10-02 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 }
-\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}
+\title{Imputation of Incomplete Multilevel Data with \proglang{R}}
 
@@
 Vink, Valentijn M. T. de Jong}
-\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
+\Plaintitle{Imputation of Incomplete Multilevel Data with R}
 \Shorttitle{Multilevel \pkg{mice}}
@@
 \section{Introduction}\label{introduction}}
+
+\hypertarget{multilevel-data}{%
+\subsection{Multilevel data}\label{multilevel-data}}
 
@@
 within one or more clusters, data are said to be systematically missing.
+Sporadic missingness is visualized in Figure XYZ.
 
@@
 
-Install non-CRAN packages if necessary:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> devtools::install_github("amices/ggmice")
-R> # devtools::install_github("hanneoberman/miceheckman")
-\end{CodeInput}
-\end{CodeChunk}
-
 Set up the R environment and load the necessary packages:
@@
 \begin{CodeInput}
-R> set.seed(2022)        # for reproducibility
+R> set.seed(123)         # for reproducibility
 R> library(mice)         # for imputation
@@
 R> library(mitml)        # for multilevel parameter pooling
-R> library(micemd)  # for imputation cf. heckman models
+R> library(micemd)       # for case study data and imputation cf. heckman models
 R> library(metamisc)     # for case study data
-\end{CodeInput}
-\end{CodeChunk}
+R> library(broom.mixed)  # for multilevel estimates
+\end{CodeInput}
+\end{CodeChunk}
+
+TODO: add table with predictor matrix values
+
+\begin{itemize}
+\tightlist
+\item
+  -2 = cluster variable
+\item
+  1 = overall effect
+\item
+  3 = overall + group-level effect
+\item
+  4 = individual-level (random) and group-level (fixed) effect
+\end{itemize}
 
@@
 
+The estimated effects in the complete data are presented in Table XYZ.
+We consider the associations in the full data set to be the true
+associations.
+
 Load the data into the environment and select the relevant variables:
@@
 missing data pattern in Figure \ref{fig:pop_pat}.
+
+The ICC in the incomplete data is
+\texttt{round(icc(popular\ \textasciitilde{}\ as.factor(school),\ data\ =\ na.omit(popmis)),\ 2)}.
+This tells us that the multilevel structure of the data should probably
+be taken into account. If we don't, we'll may end up with incorrect
+imputations, biasing the effect of the clusters towards zero.
 
@@
 
+Evaluate the convergence of the algorithm:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_trace(imp)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-conv-1} \end{center}
+
+\end{CodeChunk}
+
 Analyze the imputations:
@@
 \end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-
-testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
-
-Final parameter estimates and inferences obtained from 5 imputed data sets.
-
-             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.012     0.295    16.994     4.362     0.000    22.587     0.969 
-sex             0.695     0.251     2.768     4.287     0.047    28.390     0.975
-
-                            Estimate 
-Intercept~~Intercept|school    0.266 
-Residual~~Residual             1.035 
-ICC|school                     0.208
-
-Unadjusted hypothesis test as appropriate in larger samples.
-\end{CodeOutput}
 \end{CodeChunk}
@@
 
+Evaluate the convergence of the algorithm:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_trace(imp)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-predictor-conv-1} \end{center}
+
+\end{CodeChunk}
+
 Analyze the imputations:
@@
 \end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-
-testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
-
-Final parameter estimates and inferences obtained from 5 imputed data sets.
-
-             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     4.915     0.217    22.642     4.926     0.000     9.110     0.926 
-sex             0.975     0.283     3.444     4.250     0.024    32.504     0.978
-
-                            Estimate 
-Intercept~~Intercept|school    0.351 
-Residual~~Residual             1.153 
-ICC|school                     0.233
-
-Unadjusted hypothesis test as appropriate in larger samples.
-\end{CodeOutput}
 \end{CodeChunk}
@@
 
+Evaluate the convergence of the algorithm:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_trace(imp)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-multilevel-conv-1} \end{center}
+
+\end{CodeChunk}
+
 Analyze the imputations:
@@
 \end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-
-testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
-
-Final parameter estimates and inferences obtained from 5 imputed data sets.
-
-             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     5.011     0.410    12.222     4.226     0.000    35.955     0.980 
-sex             0.928     0.381     2.434     4.168     0.069    48.221     0.985
-
-                            Estimate 
-Intercept~~Intercept|school    0.313 
-Residual~~Residual             1.428 
-ICC|school                     0.188
-
-Unadjusted hypothesis test as appropriate in larger samples.
-\end{CodeOutput}
 \end{CodeChunk}
@@
 
-The estimated effects in the complete data are visualized in Figure
-\ref{}.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/forest-1} \end{center}
-
-\end{CodeChunk}
+The estimated effects in the complete data are presented in Table XYZ.
 
@@
 
+Evaluate the convergence of the algorithm:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_trace(imp)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-conv-1} \end{center}
+
+\end{CodeChunk}
+
+Analyze the imputed data:
+
 \begin{CodeChunk}
@@
                      term    estimate   std.error  statistic      p.value
-1             (Intercept) -2.35199614 0.340174191  -6.914093 4.993068e-12
-2                    type -0.41267936 0.180271952  -2.289204 2.208652e-02
-3                     age  0.03049024 0.001570154  19.418637 1.233010e-81
-4 as.factor(motor_score)2 -0.66765325 0.068737838  -9.713038 3.478858e-22
-5 as.factor(motor_score)3 -1.05520488 0.070218678 -15.027410 2.536132e-50
-6 as.factor(motor_score)4 -1.51238928 0.072304221 -20.917026 1.857373e-90
-7                   pupil  0.48421506 0.038982924  12.421209 6.781796e-17
-8                      ct  0.43474556 0.029968441  14.506779 2.341875e-36
+1             (Intercept) -2.29521346 0.349767178  -6.562118 7.886872e-11
+2                    type -0.41750657 0.180984169  -2.306868 2.108127e-02
+3                     age  0.03038268 0.001579024  19.241433 2.785453e-78
+4 as.factor(motor_score)2 -0.65834010 0.068555350  -9.603045 9.808985e-22
+5 as.factor(motor_score)3 -1.05935395 0.070432441 -15.040711 4.685873e-50
+6 as.factor(motor_score)4 -1.51332652 0.072497350 -20.874232 1.222074e-88
+7                   pupil  0.45660452 0.043112821  10.590922 2.255987e-10
+8                      ct  0.43389476 0.034110953  12.720101 1.856548e-15
              b          df dfcom         fmi      lambda m         riv
-1 5.301620e-04 10113.60453 11013 0.005694385 0.005497777 5 0.005528170
-2 4.905995e-05 10892.82806 11013 0.001994781 0.001811557 5 0.001814845
-3 3.333628e-08  6323.59900 11013 0.016537098 0.016226102 5 0.016493731
-4 4.194026e-05  8322.17049 11013 0.010889419 0.010651742 5 0.010766424
-5 5.136864e-05  7631.13821 11013 0.012760548 0.012501842 5 0.012660117
-6 1.403437e-04  2830.62134 11013 0.032897234 0.032214161 5 0.033286456
-7 3.571667e-04    49.96887 11013 0.309144148 0.282035215 5 0.392825972
-8 8.586575e-05   294.70102 11013 0.120676307 0.114728924 5 0.129597506
+1 5.551031e-03  1194.39341 11013 0.056029215 0.054449867 5 0.057585384
+2 9.060570e-05 10652.43106 11013 0.003506444 0.003319369 5 0.003330424
+3 6.016742e-08  3298.53424 11013 0.029546025 0.028957788 5 0.029821348
+4 3.222589e-05  9216.80652 11013 0.008443314 0.008228174 5 0.008296439
+5 8.436583e-05  5080.45332 11013 0.020793493 0.020408089 5 0.020833256
+6 1.633007e-04  2263.14149 11013 0.038133867 0.037284215 5 0.038728164
+7 6.411714e-04    23.25984 11013 0.458579946 0.413944862 5 0.706324089
+8 3.096973e-04    39.00604 11013 0.351802055 0.319397049 5 0.469285431
           ubar
-1 1.150823e-01
-2 3.243910e-02
-3 2.425379e-06
-4 4.674562e-03
-5 4.869020e-03
-6 5.059488e-03
-7 1.091068e-03
-8 7.950686e-04
+1 1.156758e-01
+2 3.264654e-02
+3 2.421115e-06
+4 4.661165e-03
+5 4.859490e-03
+6 5.059905e-03
+7 1.089310e-03
+8 7.919204e-04
 \end{CodeOutput}
@@
       AIC       BIC    logLik  deviance  df.resid 
-10495.423 10561.192 -5238.712 10477.423     11013 
+10498.989 10564.757 -5240.494 10480.989     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2843  
+ name   (Intercept) 0.2895  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.37194                 -0.41015                  0.03052  
+               -2.39617                 -0.40607                  0.03021  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65802                 -1.04611                 -1.51245  
+               -0.65393                 -1.05692                 -1.49555  
                   pupil                       ct  
-                0.50405                  0.42496  
+                0.48407                  0.45095  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10500.88 10566.65 -5241.44 10482.88    11013 
+10502.536 10568.305 -5242.268 10484.536     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2917  
+ name   (Intercept) 0.2926  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.37718                 -0.41511                  0.03067  
+                -2.3350                  -0.4268                   0.0301  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66935                 -1.05211                 -1.49429  
+                -0.6647                  -1.0565                  -1.5065  
                   pupil                       ct  
-                0.49013                  0.43835  
+                 0.4827                   0.4481  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10505.026 10570.795 -5243.513 10487.026     11013 
+10574.673 10640.442 -5278.337 10556.673     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2908  
+ name   (Intercept) 0.2921  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.32327                 -0.42365                  0.03023  
+               -2.23276                 -0.41737                  0.03068  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67143                 -1.05776                 -1.51039  
+               -0.66414                 -1.07209                 -1.52198  
                   pupil                       ct  
-                0.49756                  0.42474  
+                0.44804                  0.40791  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10519.511 10585.280 -5250.755 10501.511     11013 
+10548.019 10613.788 -5265.009 10530.019     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2961  
+ name   (Intercept) 0.2953  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.33576                 -0.40872                  0.03039  
+               -2.29768                 -0.41020                  0.03058  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66477                 -1.05453                 -1.51860  
+               -0.65615                 -1.04746                 -1.52793  
                   pupil                       ct  
-                0.45928                  0.44419  
+                0.43858                  0.43668  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
       AIC       BIC    logLik  deviance  df.resid 
-10522.038 10587.807 -5252.019 10504.038     11013 
+10562.267 10628.035 -5272.133 10544.267     11013 
 Random effects:
  Groups Name        Std.Dev.
- name   (Intercept) 0.2955  
+ name   (Intercept) 0.2949  
 Number of obs: 11022, groups:  name, 15
@@
             (Intercept)                     type                      age  
-               -2.35183                 -0.40576                  0.03064  
+               -2.21445                 -0.42713                  0.03034  
 as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67469                 -1.06551                 -1.52622  
+               -0.65280                 -1.06379                 -1.51469  
                   pupil                       ct  
-                0.47006                  0.44148  
+                0.42960                  0.42587  
 optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
@@
 
-Data are simulated and included in the \texttt{miceheckman} package. We
-will use the following variables:
+Data are simulated and included in the \texttt{micemd} package. We will
+use the following variables:
 

### 2023-10-27 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
fix analysis after imputation case study 2

@@
 ```{r mod}
-mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
+mod <- mort ~ type + age + motor_score + pupil + ct + (1 | name) 
 ```
@@
 ```
-
-The estimated effects in the complete data are presented in Table XYZ.
 
@@
 ```
-<!-- We will use the following estimates as comparative truth in this tutorial: -->
-```{r truth}
-# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
-# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
+
+We will use the complete data estimates as comparative truth in this tutorial. The estimated effects in the complete data are presented in Table XYZ. 
+```{r impact_truth, include=FALSE, echo=TRUE}
+fit <- glmer(mod, family = "binomial", data = impact) # fit the model
+tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
 ```
@@
 dat <- select(dat, !hypox)  # remove variable
+dat <- mutate(dat, motor_score = as.factor(motor_score))
 ```
@@
 fit <- imp %>% 
-  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
-tidy(pool(fit))
-as.mitml.result(fit)
+  with(glmer(
+    mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
+    family = "binomial"
+    )) 
+# tidy(pool(fit))
+# as.mitml.result(fit)
 # testEstimates(as.mitml.result(fit))
 ```
+The estimated effects after imputation are presented in Table XYZ.
 

### 2023-10-30 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update Obesity dataset

--- 2023-10-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 In this example, we demonstrate a multilevel imputation of an intercept and slope random model with a continuous response.
-We utilize the obesity dataset included in the `micemd` package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries.
+We use the obesity dataset included in the `micemd`@ package, a synthetic dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries.
 To easy the explanations we simulate data for 5 clusters and we reduce the dataset to the following variables:
@@
 
-Age and FamOb are missing according to a Missing At Random (MAR) mechanism, whereas the weight variable is subject to selection bias, following an indirect Missing Not At Random (MNAR) mechanism, that usually happens when it is unobserved or omitted a variable affecting both the value of incomplete variable (here weight) and its probability of missing.
+In this dataset, Age and FamOb are MAR, while the weight variable is affected by selection bias, attributed to an indirect MNAR mechanism.
+This MNAR mechanism typically arises when an unobserved or omitted variable influences both the value of the incomplete variable (in this case, Weight) and its likelihood of being missing (denoted as R).
 
-The analysis model considers BMI as the dependent variable, predicted by Age, Height, Gender, and FAV. Random slopes are considered for the Age variable, which contains missing values.
+In the primary analysis model, BMI serves as the dependent variable, with Age, Gender, and FamOb as predictors.
+Researchers, based on observable data, are considering the inclusion of random intercept effects, as well as introducing a random slope for the Age variable.
 
-The model is represented as:
-$$BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FAV_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}$$
-We load the data:
+The main model is given by:
+$$BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}$$
+We start by loading the data:
+```{r, include=FALSE}
+library(here)
+library(performance)
+source(here("heckman","ipdfunction.R"),local =knitr::knit_global())
+load(here("heckman","Obesity.rda"))
+```
+
 ```{r obesity}
-data("data_heckman", package = "micemd")
-dat <- data_heckman
+#data("data_heckman", package = "micemd")
+#dat <- data_heckman
 ```
@@
 ```{r obesity-md}
-plot_pattern(dat)
+require(patchwork)
+  myplots <- lapply(1:5, function(i) {
+         ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
+           ggtitle(paste0("Cluster", i))+ 
+      theme(legend.position = ifelse(i==1,"bottom","none"))
+      })
+  myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]
+
 ```
 
-We observe that the proportion of missingness for the Age and FAVC variables is quite similar across the clusters.
-However, regarding the weight variable, we notice a heterogeneous proportion of missingness, with systematic missingness in cluster 5.
+We observe that the missing pattern is quite similar across the clusters. However, regarding the weight variable, we notice that is systematically missing in cluster 3.
 In order to evaluate if it is required a imputation method for 1-level or 2-level we assess the Intraclass Correlation Coefficient (ICC) for the outcome variable, we use the "performance" package:
@@
 ```{r obesity icc}
-dat$bmi=dat$Weight/(dat$Height)^2
-Nulmodel <- lme4::lmer(bmi ~ 1 + (1|Cluster), data = dat)
-#Performance::ICC(Nulmodel)
+Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+performance::icc(Nulmodel)
 ```
 
-Since the ICC is above 0.1, we decide to employ two-level (2l) imputation methods.
+Since the ICC is above 0.1 and as the main analyis will be consider a mixed model, we decide to use two-level (2l) imputation methods.
 In this imputation process, we include all predictor variables in the main model and since BMI is a composite variable.
@@
 To determine the best imputation method for these variables, we can use the `find.defaultMethod`function provided in the `micemd` package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters.
-It suggests using "2l.2stage.bin" for the FAV variable and "2l.2stage.norm" for the age and weight variables.
-However, after inspecting the age density plot, we consider modifying its method to "2l.2stage.ppm." For the BMI variable, we employ deterministic imputation.
-```{r obesity meth}
-meth_mar <- find.defaultMethod(dat, ind.clust=7, I.small = 7, ni.small = 100, prop.small = 0.4)
+It suggests using "2l.2stage.bin" for the FAV variable and "2l.2stage.norm" for the age variable.
+However, after inspecting the age density plot, we consider modifying its method to "2l.2stage.pmm." For the BMI variable, we employ deterministic imputation.
+```{r obsmar_met}
+meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
+meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
+meth_mar["Age"]<-"2l.2stage.pmm" 
 ```
 For these imputation models, it is necessary to specify the prediction matrix, with the cluster variable labeled as -2 and the predictor variable labeled as 2, encompassing all variables.
+We need to suprime the variable Time as this variable is not especified in the main model and modify also in the prediction matrix the relationship between BMI, weight and height to avoid circular predictions.
+Then we proced to run the imputation model.
 
-After confirming convergence, we proceed to fit the main model and save the results for future use.
+```{r obsmar_pred,echo = T}
+pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
+pred_mar[,"Cluster"]<- -2 # clustering variable
+pred_mar[,"Time"]<-0
+pred_mar[pred_mar==1]<-2
+pred_mar[c("Height", "Weight"), "BMI"] <- 0
+ggmice::plot_pred(pred_mar)
+```
+```{r obsmar_mice,echo = T,results="hide",include=FALSE}
+imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar, m=10, seed = 123)
+```
+
+```{r obsmar_plot}
+summary(complete(imp_mar,"long")$Weight)
+plot(imp_mar)
+```
+We are also contemplating the utilization of the "pmm" option, given that the imputed values' weights are lower than those of the observable values.
+
+```{r obsmarp_mice,results="hide",include=FALSE}
+meth_mar["Weight"]<-"2l.2stage.pmm" 
+imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar, m=10, seed = 123)
+```
+```{r obsmarp_plot}
+summary(complete(imp_mar_pmm,"long")$Weight)
+plot(imp_mar_pmm)
+```
+After confirming convergence, we proceed to save the results for future use.
 Users are considering the possibility that the weight variable may not have been selected randomly.
@@
 
-Create predictor matrix:
-```{r obesity-pred}
-pred <- quickpred(dat)   # predictor matrix
-pred[,"Cluster"]<- -2 # clustering variable
-pred[, "Weight"] <- -3 #inclusion-restriction variable
-plot_pred(pred)
+To apply the heckman method, the weight variable should be specified as "2l.2stage.heckman" found in the micemd package.
+Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.
+
+Here for the outcome equation we consider is the same imputation model that we used for the MAR case (main model). 
+$$Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}$$
+Regarding the selection equation, we include the same predictors as those in the primary model, as well as a time variable.
+Here the time variable serves as a restriction exclusion variable specifically explaining the probability of being observed but not affecting the incomplete value (Weight).
+In this context, we assume that the time a user spends completing the survey serves as a proxy for the barriers they may encounter in survey completion, such as familiarity with the survey content or internet speed.
+These factors may lead the user to skip specific questions or even the entire survey.
+Also we assume the time is not have any influence on the weight assigned to the subject.
+$$R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}$$
+These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution.
+For a more comprehensive understanding of the model and the exclusion restriction, you can refer to (Munoz).
+
+To integrate information from both equations, we must adjust the prediction matrix.
+The cluster variable remains specified as before (-2).
+In this imputation method, all the variables present in both the selection and outcome equations are included with a random effect.
+However, it's essential to distinguish which of these variables appear in each equation.
+
+In this framework, when a variable is shared between both equations, it is denoted as (2).
+Predictors exclusive to the outcome equation are indicated as (-4), while those exclusive to the selection equation are labeled as (-3).
+Consequently, the only alteration needed in the predictor matrix pertains to the variable "Time."
+
+```{r obsmnar_pred}
+pred_mnar <- pred_mar
+pred_mnar["Weight","Time"]<- -3
+ggmice::plot_pred(pred_mnar)
 ```
 
-Set the Heckman model as imputation method:
-```{r meth}
-meth <- make.method(dat) # methods vector
-meth["weight"]<-"2l.heckman"
+We also need to modify the method of the weight variable
+```{r obsmnar_meth}
+meth_mnar <- meth_mar
+meth_mnar["Weight"]<- "3l.2stage.heckman"
 ```
 
+Then we proceed to run the imputation model as before 
 Impute the missingness:
-```{r eval=FALSE, cache=TRUE}
-imp <- mice(
-  dat, # dataset with missing values
-  m = 5, # number of imputations
+```{r obsmnar_mice,results="hide",include=FALSE}
+imp_mnar <- mice(
+  Obesity, # dataset with missing values
+  m = 10, # number of imputations
   maxit = 10,
-  meth = meth, #imputation method vector
-  pred = pred, #imputation predictors matrix
+  meth = meth_mnar, #imputation method vector
+  pred = pred_mnar, #imputation predictors matrix
   meta_method = "reml",
+  pred_std=FALSE,
   printFlag = FALSE
@@
 
-To apply the Munoz method, the weight variable should be specified as "2l.2stage.heckman" found in the micemd package.
-Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based on certain variables.
+After executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
+Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive).
+Consequently, we are contemplating a different approach, specifically, the use of "pmm," which involves imputing values from donor observations.
+This approach ensures that the imputed values remain within the range of observable values.
 
-TODO: explain exclusion restriction.
-In this specific case, it's assumed that height does not influence selection (-4), and an exclusion restriction variable is introduced as an instrumental variable affecting selection but not the outcome.
-Time of response is used for this purpose, as it's believed that subjects taking more time to complete the survey are more likely to not respond due to poor connectivity or lack of familiarity with electronic surveys (specified as -3).
+```{r obsmnar_plot}
+summary(complete(imp_mnar,"long")$Weight)
+plot(imp_mnar)
+```
 
-After implementing these imputation procedures, it's crucial to check for convergence.
-If convergence issues arise and imputed values are far from plausible, especially affecting BMI values, a post-imputation procedure is employed to constrain BMI values within a plausible range.
+We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by seeting the pmm parameter as true.
+```{r obsmnarp_mice,results="hide",include=FALSE}
+imp_mnar_pmm <- mice(
+  Obesity, # dataset with missing values
+  m = 10, # number of imputations
+  maxit = 10,
+  meth = meth_mnar, #imputation method vector
+  pred = pred_mnar, #imputation predictors matrix
+  meta_method = "reml",
+  pred_std=FALSE,
+  pmm=TRUE,
+  printFlag = FALSE
+)
+```
 
-Then after this modification we proceed to compare the effects on the model
-###Plot estimates
+```{r obsmnarp_plot}
+summary(complete(imp_mnar_pmm,"long"))
+plot(imp_mnar_pmm)
+```
 
-The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
+Then after this modification we proceed to compare the effects on the model, therefore we run the analysis model on each of the completed datasets along with the dataset where the incomplete values are removed (CC).
+
+```{r models}
+library(ggplot2)
+cc_rs<- with(Obesity[complete.cases(Obesity),],lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
+plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
+```
+
+We note that there is minimal disparity in the age effect, FamObs, across the various imputation models under consideration.
+Nonetheless, with respect to Gender, we observe a notable distinction: in the M(C)AR assumption, the effect is significant, whereas under the MNAR assumption, it becomes insignificant.
+An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
 

### 2023-10-30 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update Obesity dataset

--- 2023-10-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 In this example, we demonstrate a multilevel imputation of an intercept
-and slope random model with a continuous response. We utilize the
-obesity dataset included in the \texttt{micemd} package, a simulated
-dataset that emulates an electronic survey in which individuals are
-asked to provide information about their weight and consumption habits
-in different countries. To easy the explanations we simulate data for 5
+and slope random model with a continuous response. We use the obesity
+dataset included in the \texttt{micemd}@ package, a synthetic dataset
+that emulates an electronic survey in which individuals are asked to
+provide information about their weight and consumption habits in
+different countries. To easy the explanations we simulate data for 5
 clusters and we reduce the dataset to the following variables:
@@
 
-Age and FamOb are missing according to a Missing At Random (MAR)
-mechanism, whereas the weight variable is subject to selection bias,
-following an indirect Missing Not At Random (MNAR) mechanism, that
-usually happens when it is unobserved or omitted a variable affecting
-both the value of incomplete variable (here weight) and its probability
-of missing.
-
-The analysis model considers BMI as the dependent variable, predicted by
-Age, Height, Gender, and FAV. Random slopes are considered for the Age
-variable, which contains missing values.
-
-The model is represented as:
-\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FAV_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
-We load the data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> data("data_heckman", package = "micemd")
-R> dat <- data_heckman
+In this dataset, Age and FamOb are MAR, while the weight variable is
+affected by selection bias, attributed to an indirect MNAR mechanism.
+This MNAR mechanism typically arises when an unobserved or omitted
+variable influences both the value of the incomplete variable (in this
+case, Weight) and its likelihood of being missing (denoted as R).
+
+In the primary analysis model, BMI serves as the dependent variable,
+with Age, Gender, and FamOb as predictors. Researchers, based on
+observable data, are considering the inclusion of random intercept
+effects, as well as introducing a random slope for the Age variable.
+
+The main model is given by:
+\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
+We start by loading the data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> #data("data_heckman", package = "micemd")
+R> #dat <- data_heckman
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> plot_pattern(dat)
+R> require(patchwork)
+R>   myplots <- lapply(1:5, function(i) {
++          ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
++            ggtitle(paste0("Cluster", i))+ 
++       theme(legend.position = ifelse(i==1,"bottom","none"))
++       })
+R>   myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]
 \end{CodeInput}
@@
 
-We observe that the proportion of missingness for the Age and FAVC
-variables is quite similar across the clusters. However, regarding the
-weight variable, we notice a heterogeneous proportion of missingness,
-with systematic missingness in cluster 5. In order to evaluate if it is
+We observe that the missing pattern is quite similar across the
+clusters. However, regarding the weight variable, we notice that is
+systematically missing in cluster 3. In order to evaluate if it is
 required a imputation method for 1-level or 2-level we assess the
@@
 \begin{CodeInput}
-R> dat$bmi=dat$Weight/(dat$Height)^2
-R> Nulmodel <- lme4::lmer(bmi ~ 1 + (1|Cluster), data = dat)
-R> #Performance::ICC(Nulmodel)
-\end{CodeInput}
-\end{CodeChunk}
-
-Since the ICC is above 0.1, we decide to employ two-level (2l)
-imputation methods. In this imputation process, we include all predictor
-variables in the main model and since BMI is a composite variable. We
-also incorporate weight and height variables.
+R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+R> performance::icc(Nulmodel)
+\end{CodeInput}
+\begin{CodeOutput}
+# Intraclass Correlation Coefficient
+
+    Adjusted ICC: 0.342
+  Unadjusted ICC: 0.342
+\end{CodeOutput}
+\end{CodeChunk}
+
+Since the ICC is above 0.1 and as the main analyis will be consider a
+mixed model, we decide to use two-level (2l) imputation methods. In this
+imputation process, we include all predictor variables in the main model
+and since BMI is a composite variable. We also incorporate weight and
+height variables.
 
@@
 of clusters. It suggests using ``2l.2stage.bin'' for the FAV variable
-and ``2l.2stage.norm'' for the age and weight variables. However, after
-inspecting the age density plot, we consider modifying its method to
-``2l.2stage.ppm.'' For the BMI variable, we employ deterministic
+and ``2l.2stage.norm'' for the age variable. However, after inspecting
+the age density plot, we consider modifying its method to
+``2l.2stage.pmm.'' For the BMI variable, we employ deterministic
 imputation.
@@
 \begin{CodeInput}
-R> meth_mar <- find.defaultMethod(dat, ind.clust=7, I.small = 7, ni.small = 100, prop.small = 0.4)
+R> meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
+R> meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
+R> meth_mar["Age"]<-"2l.2stage.pmm" 
 \end{CodeInput}
@@
 matrix, with the cluster variable labeled as -2 and the predictor
-variable labeled as 2, encompassing all variables.
-
-After confirming convergence, we proceed to fit the main model and save
-the results for future use. Users are considering the possibility that
-the weight variable may not have been selected randomly. It's likely
-that an omitted variable, like self-esteem, could influence this
-selection. For instance, individuals with lower self-esteem might have
-higher weight values, impacting their willingness to provide honest
-information due to embarrassment.
+variable labeled as 2, encompassing all variables. We need to suprime
+the variable Time as this variable is not especified in the main model
+and modify also in the prediction matrix the relationship between BMI,
+weight and height to avoid circular predictions. Then we proced to run
+the imputation model.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
+R> pred_mar[,"Cluster"]<- -2 # clustering variable
+R> pred_mar[,"Time"]<-0
+R> pred_mar[pred_mar==1]<-2
+R> pred_mar[c("Height", "Weight"), "BMI"] <- 0
+R> ggmice::plot_pred(pred_mar)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_pred-1} \end{center}
+
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(complete(imp_mar,"long")$Weight)
+\end{CodeInput}
+\begin{CodeOutput}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+  2.409  69.703  82.925  82.476  95.310 158.420 
+\end{CodeOutput}
+\begin{CodeInput}
+R> plot(imp_mar)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-1} \end{center}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-2} \end{center}
+
+\end{CodeChunk}
+
+We are also contemplating the utilization of the ``pmm'' option, given
+that the imputed values' weights are lower than those of the observable
+values.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(complete(imp_mar_pmm,"long")$Weight)
+\end{CodeInput}
+\begin{CodeOutput}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+  27.50   71.02   83.82   83.76   95.55  135.98 
+\end{CodeOutput}
+\begin{CodeInput}
+R> plot(imp_mar_pmm)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-1} \end{center}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-2} \end{center}
+
+\end{CodeChunk}
+
+After confirming convergence, we proceed to save the results for future
+use. Users are considering the possibility that the weight variable may
+not have been selected randomly. It's likely that an omitted variable,
+like self-esteem, could influence this selection. For instance,
+individuals with lower self-esteem might have higher weight values,
+impacting their willingness to provide honest information due to
+embarrassment.
 
@@
 
-Create predictor matrix:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> pred <- quickpred(dat)   # predictor matrix
-R> pred[,"Cluster"]<- -2 # clustering variable
-R> pred[, "Weight"] <- -3 #inclusion-restriction variable
-R> plot_pred(pred)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-pred-1} \end{center}
-
-\end{CodeChunk}
-
-Set the Heckman model as imputation method:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> meth <- make.method(dat) # methods vector
-R> meth["weight"]<-"2l.heckman"
-\end{CodeInput}
-\end{CodeChunk}
-
-Impute the missingness:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> imp <- mice(
-+   dat, # dataset with missing values
-+   m = 5, # number of imputations
-+   maxit = 10,
-+   meth = meth, #imputation method vector
-+   pred = pred, #imputation predictors matrix
-+   meta_method = "reml",
-+   printFlag = FALSE
-+ )
-\end{CodeInput}
-\end{CodeChunk}
-
-To apply the Munoz method, the weight variable should be specified as
+To apply the heckman method, the weight variable should be specified as
 ``2l.2stage.heckman'' found in the micemd package. Additionally, the
@@
 variables from the main model), and the other for the selection model,
-explaining the probability of being observed based on certain variables.
-
-TODO: explain exclusion restriction. In this specific case, it's assumed
-that height does not influence selection (-4), and an exclusion
-restriction variable is introduced as an instrumental variable affecting
-selection but not the outcome. Time of response is used for this
-purpose, as it's believed that subjects taking more time to complete the
-survey are more likely to not respond due to poor connectivity or lack
-of familiarity with electronic surveys (specified as -3).
-
-After implementing these imputation procedures, it's crucial to check
-for convergence. If convergence issues arise and imputed values are far
-from plausible, especially affecting BMI values, a post-imputation
-procedure is employed to constrain BMI values within a plausible range.
+explaining the probability of being observed based (R) on certain
+variables.
+
+Here for the outcome equation we consider is the same imputation model
+that we used for the MAR case (main model).
+\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
+Regarding the selection equation, we include the same predictors as
+those in the primary model, as well as a time variable. Here the time
+variable serves as a restriction exclusion variable specifically
+explaining the probability of being observed but not affecting the
+incomplete value (Weight). In this context, we assume that the time a
+user spends completing the survey serves as a proxy for the barriers
+they may encounter in survey completion, such as familiarity with the
+survey content or internet speed. These factors may lead the user to
+skip specific questions or even the entire survey. Also we assume the
+time is not have any influence on the weight assigned to the subject.
+\[R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
+These two equations are jointly estimated under the assumption that the
+error terms are interconnected with a bivariate normal distribution. For
+a more comprehensive understanding of the model and the exclusion
+restriction, you can refer to (Munoz).
+
+To integrate information from both equations, we must adjust the
+prediction matrix. The cluster variable remains specified as before
+(-2). In this imputation method, all the variables present in both the
+selection and outcome equations are included with a random effect.
+However, it's essential to distinguish which of these variables appear
+in each equation.
+
+In this framework, when a variable is shared between both equations, it
+is denoted as (2). Predictors exclusive to the outcome equation are
+indicated as (-4), while those exclusive to the selection equation are
+labeled as (-3). Consequently, the only alteration needed in the
+predictor matrix pertains to the variable ``Time.''
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred_mnar <- pred_mar
+R> pred_mnar["Weight","Time"]<- -3
+R> ggmice::plot_pred(pred_mnar)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_pred-1} \end{center}
+
+\end{CodeChunk}
+
+We also need to modify the method of the weight variable
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth_mnar <- meth_mar
+R> meth_mnar["Weight"]<- "3l.2stage.heckman"
+\end{CodeInput}
+\end{CodeChunk}
+
+Then we proceed to run the imputation model as before Impute the
+missingness:
+
+After executing these imputation procedures, it is essential to assess
+convergence and the coherence of the imputed values. Upon examining the
+weight variable, we noticed that the imputed range falls outside the
+realm of plausible values (as weight should be positive). Consequently,
+we are contemplating a different approach, specifically, the use of
+``pmm,'' which involves imputing values from donor observations. This
+approach ensures that the imputed values remain within the range of
+observable values.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(complete(imp_mnar,"long")$Weight)
+\end{CodeInput}
+\begin{CodeOutput}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+  10.26   73.41   86.94   88.27  101.73  186.47 
+\end{CodeOutput}
+\begin{CodeInput}
+R> plot(imp_mnar)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-1} \end{center}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-2} \end{center}
+
+\end{CodeChunk}
+
+We then run the imputation model but this time using the option of pmm,
+to assure that weight values are in the range of the observable data,
+this can be implemented by seeting the pmm parameter as true.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(complete(imp_mnar_pmm,"long"))
+\end{CodeInput}
+\begin{CodeOutput}
+      .imp           .id          Cluster           Time          Gender     
+ Min.   : 1.0   Min.   :   1   Min.   :1.000   Min.   :1.016   Female:10720  
+ 1st Qu.: 3.0   1st Qu.: 528   1st Qu.:2.000   1st Qu.:3.724   Male  :10390  
+ Median : 5.5   Median :1056   Median :3.000   Median :5.151                 
+ Mean   : 5.5   Mean   :1056   Mean   :2.868   Mean   :5.211                 
+ 3rd Qu.: 8.0   3rd Qu.:1584   3rd Qu.:4.000   3rd Qu.:6.698                 
+ Max.   :10.0   Max.   :2111   Max.   :5.000   Max.   :9.957                 
+     Height      FamOb            Age            Weight            BMI       
+ Min.   :1.450   no : 3794   Min.   :14.00   Min.   : 27.50   Min.   : 9.59  
+ 1st Qu.:1.642   yes:17316   1st Qu.:20.00   1st Qu.: 75.19   1st Qu.:25.84  
+ Median :1.701               Median :25.00   Median : 88.88   Median :30.57  
+ Mean   :1.702               Mean   :24.82   Mean   : 89.61   Mean   :31.02  
+ 3rd Qu.:1.760               3rd Qu.:29.00   3rd Qu.:103.78   3rd Qu.:35.89  
+ Max.   :1.961               Max.   :47.00   Max.   :135.98   Max.   :63.21  
+\end{CodeOutput}
+\begin{CodeInput}
+R> plot(imp_mnar_pmm)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-1} \end{center}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-2} \end{center}
+
+\end{CodeChunk}
 
 Then after this modification we proceed to compare the effects on the
-model \#\#\#Plot estimates
-
-The imputation of these data is based on the
-\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
-Heckman Github repo}
+model, therefore we run the analysis model on each of the completed
+datasets along with the dataset where the incomplete values are removed
+(CC).
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library(ggplot2)
+R> cc_rs<- with(Obesity[complete.cases(Obesity),],lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+R> mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+R> mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+R> mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+R> mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+R> list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
+R> plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/models-1} \end{center}
+
+\end{CodeChunk}
+
+We note that there is minimal disparity in the age effect, FamObs,
+across the various imputation models under consideration. Nonetheless,
+with respect to Gender, we observe a notable distinction: in the M(C)AR
+assumption, the effect is significant, whereas under the MNAR
+assumption, it becomes insignificant. An analysis of the intercept
+reveals that, under the MNAR assumption, a higher average BMI is
+anticipated compared to the MAR assumption.
 

### 2023-10-30 heckman/ipdfunction.R

@@ -1,0 +1,472 @@
+library(lme4)
+library(data.table)
+library(mvtnorm)
+library(fMultivar)
+library(data.table)
+library(mgcv)
+library(GJRM)
+library(Matrix)
+library(mixmeta)
+library(mvtnorm)
+library(pbivnorm)
+library(mice)
+library(micemd)
+library(glmmTMB)
+
+
+copulaIPD <- function(data, sel, out, family, send) {
+  
+  fit_ind <- 0 # None model estimable for the cluster
+  # A. Estimate Heckman model
+  fit <- try(GJRM::gjrm( formula = list(sel, out),
+                         data = data,
+                         margins = c("probit", ifelse(family=="binomial","probit","N")),
+                         Model = "BSS",
+                         gamlssfit = TRUE,
+                         extra.regI = "sED",
+                         parscale = TRUE),
+             silent = TRUE)
+  
+  if (!any(inherits(fit, "try-error"))) {
+    # model is estimable
+    ev <- eigen(fit$fit$hessian, symmetric = TRUE, only.values = TRUE)$values
+    convh <- min(ev) > 0 # convergence based on hessian positive definiteness
+    convg <- max(abs(fit$fit$gradient)) < 10 # convergence based on abs max gradient
+    
+    #MAR indication 
+    CIcon<-summary(fit)$ CItheta 
+    MNAR_ind<-!(CIcon[[1]]>-0.001&CIcon[[2]]<0.001) # exclusion of cases that blow up variance
+    
+    if (MNAR_ind){
+      fit_ind <- 2
+      if (convh & convg) {
+        # MNAR estimable
+        fit_ind <- 1 # Heckman model estimable for the cluster
+      }
+    }
+  }
+  
+  if( fit_ind ==2){
+    fit <- NULL
+    gam1 <- try(mgcv::gam(formula=sel,data = data,family = "binomial", method ="REML"))
+    gam2 <- try(mgcv::gam(formula=out,data = data,family = family, method ="REML"))
+    fit_ind <- 0
+    if(!any(inherits(gam1, "try-error"))&!any(inherits(gam2, "try-error"))){
+      coefficients <- c(gam1$coefficients,gam2$coefficients)
+      vp <- c(diag(gam1$Vp),diag(gam2$Vp))
+      if(all(c(!is.na(coefficients),vp!=0))){
[…]

### 2023-10-31 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update figures

@@
          ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
-           ggtitle(paste0("Cluster", i))+ 
-      theme(legend.position = ifelse(i==1,"bottom","none"))
+         ggplot2::ggtitle(paste0("Cluster", i))+ 
+         ggplot2::theme(legend.position = ifelse(i==1,"bottom","none"))
       })
@@
 list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
-plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
+plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
 ```

### 2023-10-31 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update figures

@@
 +          ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
-+            ggtitle(paste0("Cluster", i))+ 
-+       theme(legend.position = ifelse(i==1,"bottom","none"))
++          ggplot2::ggtitle(paste0("Cluster", i))+ 
++          ggplot2::theme(legend.position = ifelse(i==1,"bottom","none"))
 +       })
@@
 
-    Adjusted ICC: 0.342
-  Unadjusted ICC: 0.342
+    Adjusted ICC: 0.362
+  Unadjusted ICC: 0.362
 \end{CodeOutput}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  2.409  69.703  82.925  82.476  95.310 158.420 
+ -19.48   68.45   81.65   81.28   94.12  160.76 
 \end{CodeOutput}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  27.50   71.02   83.82   83.76   95.55  135.98 
+  28.35   69.55   82.54   82.47   94.83  134.61 
 \end{CodeOutput}
@@
     Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
-  10.26   73.41   86.94   88.27  101.73  186.47 
+ -0.3457  71.4561  84.9177  85.1453  98.5923 182.3955 
 \end{CodeOutput}
@@
      Height      FamOb            Age            Weight            BMI        
- Min.   :1.450   no : 3794   Min.   :14.00   Min.   : 27.50   Min.   : 9.59  
- 1st Qu.:1.642   yes:17316   1st Qu.:20.00   1st Qu.: 75.19   1st Qu.:25.84  
- Median :1.701               Median :25.00   Median : 88.88   Median :30.57  
- Mean   :1.702               Mean   :24.82   Mean   : 89.61   Mean   :31.02  
- 3rd Qu.:1.760               3rd Qu.:29.00   3rd Qu.:103.78   3rd Qu.:35.89  
- Max.   :1.961               Max.   :47.00   Max.   :135.98   Max.   :63.21  
+ Min.   :1.450   no : 3954   Min.   :14.00   Min.   : 28.35   Min.   : 8.191  
+ 1st Qu.:1.642   yes:17156   1st Qu.:21.00   1st Qu.: 71.74   1st Qu.:24.699  
+ Median :1.701               Median :25.00   Median : 85.36   Median :29.396  
+ Mean   :1.702               Mean   :24.88   Mean   : 85.99   Mean   :29.717  
+ 3rd Qu.:1.761               3rd Qu.:29.00   3rd Qu.: 99.26   3rd Qu.:34.216  
+ Max.   :2.012               Max.   :47.00   Max.   :134.61   Max.   :61.911  
 \end{CodeOutput}
@@
 R> list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
-R> plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
+R> plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
 \end{CodeInput}

### 2023-11-08 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update Heckman and imputation flow

--- 2023-10-27 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 title:
-  formatted: "Imputation of Incomplete Multilevel Data with \\proglang{R}"
-  plain:     "Imputation of Incomplete Multilevel Data with R"
+  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
+  plain:     "Imputation of Incomplete Multilevel Data with mice"
   short:     "Multilevel \\pkg{mice}"
 abstract: >
-  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}.
-Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data.
-More complex models can be accomodated, but are outside the scope of this paper.
+  Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy.
+In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed.This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R.
+We aim to show how imputation can yield less biased estimates from incomplete clustered data.
+We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms.
+For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
 keywords:
@@
 set.seed(123)
+library(formatR)
 library(dplyr)
@@
 
+\newpage
+
 # Introduction
-
-## Multilevel data
 
@@
 
-Table 1: Concepts in multilevel methods
-
-| **Concept**         | **Details**                                                                   |
-|---------------------|-------------------------------------------------------------------------------|
-| Sample unit         | Units of the population from which measurements are taken in a sample.        |
-| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
-| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
-|                     | the values observed in the study are representative of all values in the      |
-|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
-| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
-|                     | admission we might select only certain hospitals that are representative of   |
-|                     | the difference of y between individual hospitals, but rather the variation of |
-| ICC                 | The variability due to clustering is often measured by means of the           |
-|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
-|                     | of variance that can be attributed to the cluster-level, where a high         |
-|                     | ICC would indicate that a lot of variability is due to the cluster            |
-|                     | structure.                                                                    |
-| Random effect       | Multilevel models typically accommodate for variability by including          |
-|                     | a separate group mean for each cluster. In addition to random                 |
-|                     | intercepts, multilevel models can also include random coefficients            |
-|                     | and heterogeneous residual error variances across clusters [see e.g.          |
-|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |
+```{r clus0,  echo = FALSE, warning = FALSE,cache = TRUE}
+concepts <- data.frame(Concept=c("Sample units","Cluster","Hierarchical data","Level-1","Level-2","Hierarchical model","Fixed effect","Random effect","Mixed effect","ICC","Stratified intercept"),
+                       Details=c("Units of the population from which measurements are taken in a sample, e.g., students.", "Variable that specify the cluster or agruppation, e.g., Classroom","Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.","Variable that varies within a cluster, eg.
+Test score","Variable that does not vary within a cluster but between, e.g.
+teacher experience.","Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)" ,"Effects that are constant across all sample units, e.g.
+something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)","Effects that are a source of random variation in the data, and whose levels are not fully sampled.
+e.g.
+test score tendency during academic year between students due to no controlled factors such as  genetic,family history","Includes fixed and random effects, e.g.
+the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
+Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
+In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
+@gelm06, @hox17 and @jong21).","The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
+The ICC can be seen as the percentage
+of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
+                                 ""))
+library(kableExtra)
+concepts%>%
+  kable( booktabs = T, escape = F,
+         caption = "Concepts in multilevel methods",
+         col.names = c("Concept","Details"), align = "l")%>%
+  column_spec(c(1), width = "3cm")%>%
+  column_spec(c(2), width = "12cm")
+```
 
@@
 When all observations are missing within one or more clusters, data are said to be systematically missing.
-Sporadic missingness is visualized in Figure XYZ.
-<!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"] -->
-<!-- TODO: Provide an example for one of the case studies below.] -->
-<!-- TODO: add that indicator method doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->
+<!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption.
+VMTdJ: "Mention in text that this is sporadic missingness?"] --> <!-- TODO: Provide an example for one of the case studies below.] --> <!-- TODO: add that indicator method doesn't work with syst missing (only sporadically).
+There's some pro's and con's.
+May not differ much if the number of clusters is low.
+-->
 
@@
 
-Table 2: Concepts in missing data methods
-
-| **Concept**    | **Details**                                                                |
-|----------------|----------------------------------------------------------------------------|
-| MCAR           | Missing Completely At Random, where the probability to be missing is equal |
-|                | across all data entries                                                    |
-| MAR            | Missing At Random, where the probability to be missing depends on observed |
-|                | information                                                                |
-| MNAR           | Missing Not At Random (MNAR), where the probability to be missing          |
-|                | depends on unrecorded information, making the missingness non-ignorable    |
-|                | [@rubi76; @meng94].                                                        |
-|                |                              |
+```{r mechanism,  echo = FALSE, warning = FALSE,cache = TRUE}
+concepts <- data.frame(Concept=c("MCAR","MAR","MNAR"),
+                       Details=c("Missing Completely At Random, where the probability to be missing is equal across all data entries.","Missing At Random, where the probability to be missing depends on observed information","Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]"))
+library(kableExtra)
+concepts%>%
+  kable( booktabs = T, escape = F,
+         caption = "Concepts in missing data methods",
+         col.names = c("Concept","Details"), align = "l")%>%
+  column_spec(c(1), width = "3cm")%>%
+  column_spec(c(2), width = "12cm")
+```
 
@@
 <!-- [TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] -->
+
 <!-- [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.] -->
@@
 
+## Imputation workflow
+Below we provide a imputation workflow that can be used in general to impute cluster data.
+
+### Main analysis
+When imputing clustered data, our first consideration should be the research question and the type of analysis that researchers intend to perform, assuming that non-incomplete values were present.
+This initial assessment provides valuable insights into the data's structure (see the level table), relationships among variables, and the causes of missingness.
+
+When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g.
+if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021).
+
+### Exploration of available data
+Then it is required to explore available information across clusters.
+This can involve evaluating the Intra-class Correlation Coefficient (ICC) to determine if a one-level imputation method is sufficient, particularly if there are no substantial differences between clusters.
+Additionally, it can be done by visualizing the data using available package as **ggmice**, for instance examining density plots can help us understand the variable type (normal, Poisson, categorical) and its range of plausible values.
+Exploring relationships between covariates through bivariate plots can be valuable in identifying potential interactions or non-linearities.
+
+In the context of missingness, we should assess patterns of systematic and sporadic missing values and explore the relationships between the missingness of an incomplete variable and observable data.
+This evaluation allows us to consider the plausibility of Missing at Random (MAR) mechanisms over Missing Completely at Random (MCAR) methods and helps in selecting prediction variables to be included in the imputation model (considering overflux).
+
+### Selecting plausible imputation methods
+Before to implement multiple imputation it is requiered to evaluate if simpler imputation methods are suffice: 
+- For example, certain incomplete variables may not necessitate stochastic imputation methods like MICE.
+Instead, they can be effectively addressed through deductive imputation, where incomplete values can be inferred from logic and deterministic relationships between variables.
+This approach is particularly useful for variables that are functions of others, e.g.
+a person's BMI from their weight and height, determining values for one-level variables from two-level ones, e.g.
+in the context of Individual Participant Data (IPD), incomplete information can be extracted from metadata, such as deducing incomplete data about abortion in a country where abortion is illegal, or engaging in cross-temporal or protocol's deduction, like imputing missing test values for deceased patients.
+
+- Under monotonic pattern, the use of regression models
+- CC when the proportion of missing is minimal <5%.
+- Analysis that are robust to missing data, for instance FIML models such as mixed model do not require imputation models when only the outcome variable is missing.An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML).
+This method does not require the imputation of any missing values.
+Whereas MI consists of imputation, analyses and pooling steps, FIML analyses the data in a single step.
+When the assumptions are met the two approaches should produce equivalent results.
+[REF] As FIML requires specialised software, not all analyses can be performed with standard software.
+[REF]
+
+- Add table describing 2l- imputation models. like this https://bookdown.org/mwheymans/bookmi/multiple-imputation-models-for-multilevel-data.html
+
+### Imputation process
+
+#### Clustering specification
+There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: <https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/>) TODO: replace ref.
+
+The selection of each strategy depends mainly on the assumptions in the main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
+The last restriction is also required when imputations are performed on each cluster separately.
+When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
+However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.
+the number of clusters is adequate.
+(Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008).
+Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020).
+For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
+(Zhang,2023).
+
+#### Selection of variables in the imputation model
+- Inclusion of outcome variable [Moons] When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]
+- Imputation for interaction term, JAV, Machine lerarning methods (Random forest) and imputation separated by treatment group
+- Prediction matrix specfication (table)
+
+#### Convergency problems during imputation
+- Check log 
+- Problems of convergence on some 2l methods can be solved by use of overflux or quickpred. 
+- Consider the use of scaled variables Vanburen. 
+- Consider to swap imputation mehtod in problematic variables: The use of non parametrical imputation methods such as pmm. or downgrade to 1l methods.
+- Also by the specification of post-processing in some variables can alivianate this. 
+- Separated imputation process for each endpoint analysis
+
+### Convergency 
+Plots and also the prediction plots.
+
+### Sensibility analysis
+About the possibility of assumptions about Missing mechanism.
+
 <!-- With the `popmis` data, we show how (and how not) to develop an imputation model.
@@
 ## Setup
+
 <!-- * TODO: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB? -->
 
+Install non-CRAN packages if necessary:
+
+```{r packages, eval=FALSE, include=TRUE}
+devtools::install_github("amices/ggmice")
+
+```
+
 <!-- [TODO: Add environment info, seed and version number(s) somewhere.]  -->
@@
 Set up the R environment and load the necessary packages:
+
 ```{r env, message=FALSE, warning=FALSE}
-set.seed(123)         # for reproducibility
+set.seed(2022)        # for reproducibility
 library(mice)         # for imputation
@@
 library(mitml)        # for multilevel parameter pooling
-library(micemd)       # for case study data and imputation cf. heckman models
+library(micemd)  # for imputation cf. heckman models
 library(metamisc)     # for case study data
-library(broom.mixed)  # for multilevel estimates
-```
-
-TODO: add table with predictor matrix values
-
-- -2 = cluster variable 
-- 1 = overall effect
-- 3 = overall + group-level effect 
-- 4 = individual-level (random) and group-level (fixed) effect
+```
+
+<!-- library(broom.mixed)  # for multilevel estimates -->
 
@@
 <!-- $$ -->
+
 <!-- \texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)} -->
+
 <!-- $$ -->
+
 ```{r}
@@
 ```
-
-The estimated effects in the complete data are presented in Table XYZ. We consider the associations in the full data set to be the true associations.
 
@@
 
-The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.
+<!-- The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.
 This tells us that the multilevel structure of the data should probably be taken into account.
-If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.
+If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.-->
 
 To develop the best imputation model for the incomplete variable `popular`, we need to know whether the observed values of `popular` are related to observed values of other variables.
-Plot the pair-wise complete correlations in the incomplete data:
-<!-- Excelent, but i dont know if you can refer to the outflux criteria here or elsewhere, cuz in practice it is used to select the predictors in the imputation model from dataset with a lot of features-->
+Plot the pair-wise complete correlations in the incomplete data: <!-- Excelent, but i dont know if you can refer to the outflux criteria here or elsewhere, cuz in practice it is used to select the predictors in the imputation model from dataset with a lot of features-->
 
@@
 We include it purely to illustrate the effects of ignoring the clustering in our imputation effort.
-<!-- We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).  -->
+<!-- We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).
+-->
 
@@
 Impute the data, ignoring the cluster structure:
+
 ```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
@@
 
-Evaluate the convergence of the algorithm:
-```{r pop-ignored-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
-plot_trace(imp)
-```
 <!-- [TODO: remove the broom.mixed output, use mitml only] -->
@@
 Analyze the imputations:
+
 ```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
@@
 
-```{r pop-print, eval=FALSE}
+```{r}
 testEstimates(as.mitml.result(fit), extra.pars = TRUE)
@@
 
-Evaluate the convergence of the algorithm:
-```{r pop-predictor-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
-plot_trace(imp)
-```
-
 Analyze the imputations:
+
 ```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
@@
 Print the estimates:
-```{r, eval=FALSE}
+
+```{r}
 testEstimates(as.mitml.result(fit), extra.pars = TRUE)
@@
 
-Evaluate the convergence of the algorithm:
-```{r pop-multilevel-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
-plot_trace(imp)
-```
-
 Analyze the imputations:
+
 ```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
@@
 We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc].
-<!-- [TODO: add more info about the complete data.]  -->
-The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:
+<!-- [TODO: add more info about the complete data.]  --> The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:
 
@@
 The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model:
+
 ```{r mod}
-mod <- mort ~ type + age + motor_score + pupil + ct + (1 | name) 
+mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
 ```
@@
 Load the complete and incomplete data into the R workspace:
+
 ```{r data, eval = FALSE}
@@
 
-<!-- visualized in Figure \ref{}. -->
+The estimated effects in the complete data are visualized in Figure \ref{}.
 
 ```{r forest, echo=FALSE}
-# fits <-
-#   purrr::map_dfr(
-#     unique(impact$name),
-#     ~ {
-#       glm(
-#         mort ~ 1 + age + motor_score + pupil + ct,
-#         family = "binomial",
-#         data = filter(impact, name == .x)
-#       )
-#     } %>%
-#       broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
-#       .[, c("term", "estimate", "conf.low", "conf.high")] %>%
-#       cbind(name = as.character(.x), .)
-#   ) %>% filter(term %in% c("ctIII", "ctIV/V"))
-# 
-# n <- impact %>% group_by(name) %>% 
-#   tally()
-# 
-# fits %>% left_join(n, by = "name") %>% 
-#   ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
-#   geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
-#   geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
-#   geom_point(shape = 20, position = position_dodge(width = 0.5)) +
-#   labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
-#   theme_classic() +
-#   scale_size(guide = 'none')
-# # TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
-# # TODO: missingness prop inside clusters
-# # TODO: make row height proportional to pattern frequency
-```
-
-We will use the complete data estimates as comparative truth in this tutorial. The estimated effects in the complete data are presented in Table XYZ. 
-```{r impact_truth, include=FALSE, echo=TRUE}
-fit <- glmer(mod, family = "binomial", data = impact) # fit the model
-tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
+fits <-
+  purrr::map_dfr(
+    unique(impact$name),
+    ~ {
+      glm(
+        mort ~ 1 + age + motor_score + pupil + ct,
+        family = "binomial",
+        data = filter(impact, name == .x)
+      )
+    } %>%
+      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
+      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
+      cbind(name = as.character(.x), .)
+  ) %>% filter(term %in% c("ctIII", "ctIV/V"))
+
+n <- impact %>% group_by(name) %>% 
+  tally()
+
+fits %>% left_join(n, by = "name") %>% 
+  ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
+  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
+  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
+  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
+  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
+  theme_classic() +
+  scale_size(guide = 'none')
+# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
+# TODO: missingness prop inside clusters
+# TODO: make row height proportional to pattern frequency
+```
+
+<!-- We will use the following estimates as comparative truth in this tutorial: -->
+
+```{r truth}
+# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
+# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
 ```
@@
 To explore the missingness, it is wise to look at the missing data pattern. The ten most frequent missingness patterns are shown:
+
 ```{r pattern, fig.height=7.1}
-plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
+plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
 ```
@@
 We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations:
+
 ```{r impact_corr}
@@
 We conclude that we can exclude `hypox` from the data, since this is neither an analysis model variable nor an auxiliary variable for imputation:
+
 ```{r}
 dat <- select(dat, !hypox)  # remove variable
-dat <- mutate(dat, motor_score = as.factor(motor_score))
 ```
@@
 As previously stated, complete case analysis lowers statistical power and may bias results. The complete case analysis estimates are:
+
 ```{r cca}
@@
 <!-- --- -->
+
 Mutate data to get the right data types for imputation (e.g. integer for clustering variable).
@@
 Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
+
 ```{r impact}
@@
 Impute the incomplete data
+
 ```{r imp_impact, cache=TRUE}
@@
 
-Evaluate the convergence of the algorithm:
-```{r impact-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
-plot_trace(imp)
-```
-
-Analyze the imputed data:
 ```{r impact-fit, cache=TRUE}
 fit <- imp %>% 
-  with(glmer(
-    mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
-    family = "binomial"
-    )) 
-# tidy(pool(fit))
-# as.mitml.result(fit)
+  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
+tidy(pool(fit))
+as.mitml.result(fit)
 # testEstimates(as.mitml.result(fit))
 ```
-The estimated effects after imputation are presented in Table XYZ.
 
@@
 
-TODO: explain exclusion restriction.
-
-Data are simulated and included in the `micemd` package. We will use the following variables:
-
-- `region` Cluster variable,
-- `gender` Gender (0=male, 1=female),
-- `age` Age of the patient,
-- `height` Height in meters,
-- `weight` Weight in kilograms,
-- `time`` Response time in minutes (inclusion-restriction variable).
+In this example, we demonstrate a multilevel imputation of random intercept and random slope model with a continuous response.
+We utilize the obesity dataset included in the `micemd`\@ package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries.
+We simulate data for 5 clusters so that the true values are known.
+We use the following variables from the dataset:
+
+-   **Cluster:** Region of the patients' healthcare provider (Cluster variable),
+-   **Gender:** Subjects' Gender (0=male, 1=female),
+-   **Age:** Subjects' age,
+-   **Height:** Subjects' height in metres,
+-   **Weight:** Subjects' weight in kilograms,
+-   **BMI:** Subjects' body mass index,
+-   **FamOb:** Family obesity history (yes or no),
+-   **Time:** Response time in minutes (exclusion-restriction variable).
+
+In this dataset, Age and FamOb are MAR, while the weight variable is affected by selection bias, attributed to an indirect MNAR mechanism.
+This MNAR mechanism typically arises when an unobserved or omitted variable influences both the value of the incomplete variable (in this case, Weight) and its likelihood of being missing (denoted as R).
+
+In the primary analysis model, BMI serves as the dependent variable, with Age, Gender, and FamOb as predictors.
+Because of the clustered nature of the data, which is quantified with the Intraclass Correlation Coefficient (ICC) below, we include random intercepts, as well as a random slope for the Age variable.
+The model is represented as:
+\begin{equation}
+\label{eqn:main}
+BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}
+\end{equation}
+
+```{r, include=FALSE}
+library(here)
+source(here("heckman","ipdfunction.R"),local =knitr::knit_global())
+load(here("heckman","Obesity.rda"))
+```
+
+We start by loading the data:
+
+```{r obesity}
+#data("data_heckman", package = "micemd")
+#dat <- data_heckman
+```
+
+Now, let's begin by examining the missing patterns in the data by cluster:
+
+```{r obesity-md, dpi = 300,out.width="70%", fig.cap="Missing pattern"}
+library(ggpubr)
+myplots <- lapply(1:5, function(i) {
+  ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
+  ggplot2::ggtitle(paste0("Cluster", i))
+ })
+ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
+```
+
+We observe that the missing pattern is non-monotonic and quite similar across the clusters.
+However, regarding the weight variable, we notice that is systematically missing in cluster 3.
+In order to evaluate
+if we require a imputation method that accounts for clustering  we assess the Intraclass Correlation
+
+```{r obesity icc}
+Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+#Performance::ICC(Nulmodel)
+```
+
+Since the ICC is above 0.1 and as the main analysis will be use a mixed model, we decide
+to use two-level (2l) imputation methods. In this imputation process, we include all predictor
+variables from equation \ref{eqn:main} in the main model. However, since BMI is a composite of
+weight and height, we use deterministic imputation for these, which is described below.
+
+We use the **find.defaultMethodfunction** provided in the **micemd** package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters.
+
+It suggests using '2l.2stage.bin' for the FAV variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
+'2l.2stage.pmm'. For the BMI variable, we employ deterministic imputation.
+
+```{r obesity meth}
+meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+                               ni.small = 100, prop.small = 0.4)
+meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
+meth_mar["Age"]<-"2l.2stage.pmm" 
+```
+
+For these imputation models, it is necessary to specify the prediction matrix, with the cluster
+variable labelled as -2 and the predictor variable measured within clusters labelled as 2, encompassing all variables.
+We need to suprime the variable Time as this variable is not specified in the main model.
+We also  modify the relationship between BMI, weight and height in the prediction matrix to avoid circular predictions.
+Then we proceed to run the imputation model.
+```{r obesity-predmar,out.width="70%"}
+pred_mar <- quickpred(Obesity)   # predictor matrix
+pred_mar[,"Cluster"] <- -2 # clustering variable
+pred_mar[,"Time"] <- 0
+pred_mar[pred_mar==1] <- 2
+pred_mar[c("Height", "Weight"), "BMI"] <- 0
+ggmice::plot_pred(pred_mar)
+imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+                m=10, seed = 123, printFlag = FALSE)
+```
+
+```{r obesity-predmar1,eval=TRUE}
+summary(complete(imp_mar,"long")$Weight)
+```
+We are also contemplating the utilisation of the ???pmm??? option, as the values imputed using a fully parametric method may be implausibly low for some patients.
+
+```{r obesity-predmar_pmm}
+meth_mar["Weight"]<-"2l.2stage.pmm" 
+imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+                    m=10, seed = 123, printFlag = FALSE)
+```
+
+```{r obesity-predmar_pmm1,eval=TRUE}
+summary(complete(imp_mar_pmm,"long")$Weight)
+#ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+
+```
+After confirming convergence, we proceed to save the results for future use.
+We consider the possibility that patients may not have been selected randomly, which would then have led to a distribution for weight that does not reflect the weight in the population.
+It???s likely that an omitted variable, like self-esteem, could influence this selection.
+For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.
+
+To address this situation, two approaches have been proposed for dealing with Missing Not at Random (MNAR) data: pattern-mixed models and selection models.
+Within pattern-mixed models, methods like the delta method and more advanced ones like NARFS have been suggested.
+The selection model approach includes methods such as the Heckman model, which can be particularly useful in this case.
+Several methods, including those by \cite{Galimar_2017,Hammon_2021}, and the recently a Heckman method designed for two-level data, allow for variations in intercepts and exposure effects (random intercept and slope).
+
+To apply the **2l.2stage.heckman** method, the weight variable should be specified as '2l.2stage.heckman' found in the micemd package.
+Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.
+For the outcome equation we consider the same imputation model that we used for
+the MAR case (main model).
+ $$Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}$$ 
+Regarding the selection equation, we include the same predictors as those in the main model, as well as a time variable.
+Here the time variable serves as a restriction exclusion variable specifically explaining the probability of being observed but not affecting the incomplete value (Weight).
+In this context, we assume that the time a user spends completing the survey serves as a proxy for the barriers they may encounter in survey completion, such as familiarity with the survey content or internet speed.
+These factors may lead the user to skip specific questions or even the entire survey.
+Also, we assume the time does not have any influence on the subject???s weight.
+ $$R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}$$
+
+These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution.
+For a more comprehensive understanding of the model and the exclusion restriction, see \cite{Munoz_2022}.
+
+To use information from both equations, we must adjust the prediction matrix.
+The cluster variable remains specified as before (-2).
+In this imputation method, all the variables present in both the selection and outcome equations are included with a random effect.
+
+However, it is essential to distinguish which of these variables appear in each equation.
+In this framework, when a variable is shared between both equations, it is denoted as (2).
+Predictors exclusive to the outcome equation are indicated as (-4), while those exclusive to the selection equation are labelled as (-3).
+Consequently, the only alteration needed in the predictor matrix pertains to the variable 'Time'.
+
+```{r obesity-predmnar}
+pred_mnar <- pred_mar
+pred_mnar["Weight","Time"]<- -3
+#ggmice::plot_pred(pred_mnar)
+```
+
+We also need to modify the method of the weight variable.
+
+```{r obesity-predmnar1}
+meth_mnar <- meth_mar
+meth_mnar["Weight"]<- "3l.2stage.heckman"
+```
+Then we proceed to run the imputation model as before, after executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
+```{r obesity-predmnar2,eval=TRUE}
+imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+                m=10, seed = 123, printFlag = FALSE)
+summary(complete(imp_mnar,"long")$Weight)
+```
+
+Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive).
+
+```{r obesity-predmnar3,eval=TRUE}
+summary(complete(imp_mnar,"long")$Weight)
+```
+Consequently, as before we  use the 'pmm', option but this time for the Heckman imputation, this approach ensures that the imputed values remain within the range of observable values.
+We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by setting the pmm parameter to true.
+
+```{r obesity-predmnarp,eval=TRUE}
+imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+                     m=10, seed = 123, pmm = T,  printFlag = FALSE)
+
+```
+
+We check the convergency of the results
+```{r obesity-predmnarp1,eval=TRUE}
+summary(complete(imp_mnar_pmm,"long")$Weight)
+#ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+```
+
+After this modification we proceed to compare the effects on the model. We
+run the analysis model on each of the completed datasets as well asthe dataset where the
+incomplete values are removed (Complete Case analysis, CC).
+
+```{r models, eval=TRUE}
+library(ggplot2)
+cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
+             lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
+mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+
+list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
+plot_models(list_models,
+            mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
+```
+
+We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration.
+An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
+Nonetheless, with respect to precision of estimates, we notice that in general MNAR imputation leads to wider confidence intervals, in this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on significant test and therefore lead to contradictory conclusions.
+
+# Additional
 
@@
 
-Load the data:
-```{r obesity}
-data("data_heckman", package = "micemd")
-```
-
-Select relevant variables:
-```{r bmi}
-dat <- data_heckman
-```
-
 Visualize missing data pattern:
-```{r obesity-md}
-plot_pattern(dat)
-```
 
@@
 
-Create predictor matrix:
-```{r obesity-pred}
-pred <- quickpred(dat)   # predictor matrix
-pred[,"Cluster"]<- -2 # clustering variable
-pred[, "Weight"] <- -3 #inclusion-restriction variable
-plot_pred(pred)
-```
-
-Set the Heckman model as imputation method:
-```{r meth}
-meth <- make.method(dat) # methods vector
-meth["weight"]<-"2l.heckman"
-```
-
-Impute the missingness:
-```{r eval=FALSE, cache=TRUE}
-imp <- mice(
-  dat, # dataset with missing values
-  m = 5, # number of imputations
-  maxit = 10,
-  meth = meth, #imputation method vector
-  pred = pred, #imputation predictors matrix
-  meta_method = "reml",
-  printFlag = FALSE
-)
-```
-
 # Discussion
@@
 <!-- - JOMO in \pkg{mice} -> on the side for now -->
+
 <!-- # look at jomo for categorical variables? -->
+
 <!-- # semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach -->
+
 <!-- # pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters -->
+
 <!-- # make assumptions of these methods explicit! -->
+
 ORDER:
@@
 -   summary
+
 -   congeniality, then in hierarchical models
+
 -   look whether we can fit cong. back in the main body
+
 -   alt. methods
+
 -   conclusion: mice is really easy!
@@
 
-In hierarchical datasets, clustering is a concern because the homoscedasticity in the error terms cannot be assumed across clusters and the relationship among variables may vary at different hierarchical levels.
-When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g.
-if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021).
-Not including clustering into the imputation process may lead to effect estimates with smaller standard errors and inflated type I error.
-
-There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.
-
-The selection of each strategy depends mainly on the assumptions in the main analysis and also on the restriction of the analyzed data.
-
-Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
-The last restriction is also required when imputations are performed on each cluster separately.
-When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).
-
-Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
-However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.
-the number of clusters is adequate.
-(Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008).
-Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020).
-For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
-(Zhang,2023).
-
 # Funding
@@
 
-| **Formula \pkg{lme4}**         | **Details**                                 |
-|-------------------------------------------------------------|--------------------------------------------------|
+| **Formula** \pkg{lme4}                                     | **Details**                                                     |
+|---------------------------------------|---------------------------------|
 | `y ~ x1 + (1 | g1)`                                        | Fixed `x1` predictor with random intercept                      |
@@
 |                                                            | and random slope of `x1`                                        |
-| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2   |
-| `y ~ (1 | g1/g2) or ~ (1 | g1) + (1 | g1:g2)`                     | Random intercept varying among `g1` and `g2`         |
-|                                                             | withing `g1`                                       |
+| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2   | |`y \~ (1 |
 |                                                            |                                                                 |

### 2023-11-08 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update Heckman and imputation flow

--- 2023-10-30 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 Hanne I. Oberman\\Methodology and Statistics\\
-Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
+Utrecht University \And Johanna Munoz\\Julius Center for Health Sciences
 and Primary Care,\\
@@
 \Abstract{
-This tutorial illustrates the imputation of incomplete multilevel data
-with the \proglang{R} packackage \pkg{mice}. Our scope is only simple
-multilevel models, to show how imputation can yield less biased
-estimates from incomplete clustered data. More complex models can be
-accomodated, but are outside the scope of this paper.
+Incomplete multilevel data requires careful consideration of the missing
+data problem and analysis strategy. In this tutorial, we focus on a
+popular strategy for accommodating missingness in multilevel data:
+replacing the missing data with one or more plausible values, i.e.,
+imputation.Imputation separates the missing data problem from the main
+analysis and the completed data can be analyzed as if it has been fully
+observed.This tutorial illustrates the imputation of incomplete
+multilevel data with the statistical programming language R. We aim to
+show how imputation can yield less biased estimates from incomplete
+clustered data. We provide practical guidelines and code snippets for
+different missing data situations, including non-ignorable missingness
+mechanisms. For brevity, we focus on multilevel imputation using chained
+equations with the R mice package and its adjacent packages.
 }
@@
 \begin{document}
+
+\newpage
 
@@
 
-\caption{\label{tab:clus}Concepts in multilevel methods}
+\caption{\label{tab:clus0}Concepts in multilevel methods}
 \centering
@@
 \midrule
-Sample unit & Units of the population from which measurements are taken in a sample.\\
-Hierarchical levels & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
-Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.\\
+Sample units & Units of the population from which measurements are taken in a sample, e.g., students.\\
+Cluster & Variable that specify the cluster or agruppation, e.g., Classroom\\
+Hierarchical data & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
+Level-1 & Variable that varies within a cluster, eg. Test score\\
+Level-2 & Variable that does not vary within a cluster but between, e.g. teacher experience.\\
+\addlinespace
+Hierarchical model & Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)\\
+Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)\\
 Random effect & Effects that are a source of random variation in the data, and whose levels are not fully sampled.
 e.g.
-individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.\\
+test score tendency during academic year between students due to no controlled factors such as  genetic,family history\\
 Mixed effect & Includes fixed and random effects, e.g.
@@
 @gelm06, @hox17 and @jong21).\\
-\addlinespace
 ICC & The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
 of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.\\
+\addlinespace
 Stratified intercept & \\
@@
 
-Table 2: Concepts in missing data methods
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Concept}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Details}
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-MCAR & Missing Completely At Random, where the probability to be missing
-is equal \\
-& across all data entries \\
-MAR & Missing At Random, where the probability to be missing depends on
-observed \\
-& information \\
-MNAR & Missing Not At Random (MNAR), where the probability to be
-missing \\
-& depends on unrecorded information, making the missingness
-non-ignorable \\
-& \citep{rubi76, meng94}. \\
-& \\
-\end{longtable}
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:mechanism}Concepts in missing data methods}
+\centering
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
+\toprule
+Concept & Details\\
+\midrule
+MCAR & Missing Completely At Random, where the probability to be missing is equal across all data entries.\\
+MAR & Missing At Random, where the probability to be missing depends on observed information\\
+MNAR & Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
 
@@
 (case study 3).
+
+\hypertarget{imputation-workflow}{%
+\subsection{Imputation workflow}\label{imputation-workflow}}
+
+Below we provide a imputation workflow that can be used in general to
+impute cluster data.
+
+\hypertarget{main-analysis}{%
+\subsubsection{Main analysis}\label{main-analysis}}
+
+When imputing clustered data, our first consideration should be the
+research question and the type of analysis that researchers intend to
+perform, assuming that non-incomplete values were present. This initial
+assessment provides valuable insights into the data's structure (see the
+level table), relationships among variables, and the causes of
+missingness.
+
+When multiple imputation is used to deal with missing data, as the
+imputation and analysis process is performed separately, it is necessary
+that imputation model being congenial with the main analysis model
+(Meng, 1994), e.g.~if the main model accounts for the hierarchical
+structure also imputation model should do it (Audigier, 2021).
+
+\hypertarget{exploration-of-available-data}{%
+\subsubsection{Exploration of available
+data}\label{exploration-of-available-data}}
+
+Then it is required to explore available information across clusters.
+This can involve evaluating the Intra-class Correlation Coefficient
+(ICC) to determine if a one-level imputation method is sufficient,
+particularly if there are no substantial differences between clusters.
+Additionally, it can be done by visualizing the data using available
+package as \textbf{ggmice}, for instance examining density plots can
+help us understand the variable type (normal, Poisson, categorical) and
+its range of plausible values. Exploring relationships between
+covariates through bivariate plots can be valuable in identifying
+potential interactions or non-linearities.
+
+In the context of missingness, we should assess patterns of systematic
+and sporadic missing values and explore the relationships between the
+missingness of an incomplete variable and observable data. This
+evaluation allows us to consider the plausibility of Missing at Random
+(MAR) mechanisms over Missing Completely at Random (MCAR) methods and
+helps in selecting prediction variables to be included in the imputation
+model (considering overflux).
+
+\hypertarget{selecting-plausible-imputation-methods}{%
+\subsubsection{Selecting plausible imputation
+methods}\label{selecting-plausible-imputation-methods}}
+
+Before to implement multiple imputation it is requiered to evaluate if
+simpler imputation methods are suffice: - For example, certain
+incomplete variables may not necessitate stochastic imputation methods
+like MICE. Instead, they can be effectively addressed through deductive
+imputation, where incomplete values can be inferred from logic and
+deterministic relationships between variables. This approach is
+particularly useful for variables that are functions of others, e.g.~a
+person's BMI from their weight and height, determining values for
+one-level variables from two-level ones, e.g.~in the context of
+Individual Participant Data (IPD), incomplete information can be
+extracted from metadata, such as deducing incomplete data about abortion
+in a country where abortion is illegal, or engaging in cross-temporal or
+protocol's deduction, like imputing missing test values for deceased
+patients.
+
+\begin{itemize}
+\item
+  Under monotonic pattern, the use of regression models
+\item
+  CC when the proportion of missing is minimal \textless5\%.
+\item
+  Analysis that are robust to missing data, for instance FIML models
+  such as mixed model do not require imputation models when only the
+  outcome variable is missing.An alternative approach to missing data is
+  to use Full Information Maximum Likelihood (FIML). This method does
+  not require the imputation of any missing values. Whereas MI consists
+  of imputation, analyses and pooling steps, FIML analyses the data in a
+  single step. When the assumptions are met the two approaches should
+  produce equivalent results. {[}REF{]} As FIML requires specialised
+  software, not all analyses can be performed with standard software.
+  {[}REF{]}
+\item
+  Add table describing 2l- imputation models. like this
+  \url{https://bookdown.org/mwheymans/bookmi/multiple-imputation-models-for-multilevel-data.html}
+\end{itemize}
+
+\hypertarget{imputation-process}{%
+\subsubsection{Imputation process}\label{imputation-process}}
+
+\hypertarget{clustering-specification}{%
+\paragraph{Clustering specification}\label{clustering-specification}}
+
+There are different strategies that can be adopted in the imputation
+process that account for clustering: inclusion of cluster indicator
+variable, performing a separate imputation process for each cluster, or
+performing a simultaneous imputation process by using an imputation
+method that accounts for clustering.(Stata:
+\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
+TODO: replace ref.
+
+The selection of each strategy depends mainly on the assumptions in the
+main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of
+cluster indicator variables is restricted in datasets where there are
+not many clusters and many observations per cluster (Graham, 2009). The
+last restriction is also required when imputations are performed on each
+cluster separately. When this restriction cannot be achieved, one can
+use an imputation model that simultaneously imputes all clusters using a
+hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters
+are correlated and this correlation is modeled by a random effect so the
+hierarchical model can be estimated even when there are few observations
+per cluster. However, this strategy is best suited for balanced data
+(Grund, 2017) and when random effects model is appropriated, i.e.~the
+number of clusters is adequate. (Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main
+model, for instance by using the cluster indicator strategy may lead to
+bias estimates when the model is based on a hierarchical model
+(Taaljard,2008). Even when an imputation strategy congenial with the
+main model is preferred, it is important to consider whether it is
+appropriate for the data as a less complex imputation strategies may
+also lead to unbiased estimates in certain scenarios(Bailey 2020). For
+instance, in causal effect analysis, separately imputation may lead to
+smaller bias when the size of the smaller exposure cluster is large,
+compared with an imputation model that includes exposure-confounder
+interactions. (Zhang,2023).
+
+\hypertarget{selection-of-variables-in-the-imputation-model}{%
+\paragraph{Selection of variables in the imputation
+model}\label{selection-of-variables-in-the-imputation-model}}
+
+\begin{itemize}
+\tightlist
+\item
+  Inclusion of outcome variable {[}Moons{]} When the outcome is
+  time-to-event, the Nelson-Aalen estimate of the time to event should
+  be included as a covariate in the imputation model {[}REF{]}
+\item
+  Imputation for interaction term, JAV, Machine lerarning methods
+  (Random forest) and imputation separated by treatment group
+\item
+  Prediction matrix specfication (table)
+\end{itemize}
+
+\hypertarget{convergency-problems-during-imputation}{%
+\paragraph{Convergency problems during
+imputation}\label{convergency-problems-during-imputation}}
+
+\begin{itemize}
+\tightlist
+\item
+  Check log
+\item
+  Problems of convergence on some 2l methods can be solved by use of
+  overflux or quickpred.
+\item
+  Consider the use of scaled variables Vanburen.
+\item
+  Consider to swap imputation mehtod in problematic variables: The use
+  of non parametrical imputation methods such as pmm. or downgrade to 1l
+  methods.
+\item
+  Also by the specification of post-processing in some variables can
+  alivianate this.
+\item
+  Separated imputation process for each endpoint analysis
+\end{itemize}
+
+\hypertarget{convergency}{%
+\subsubsection{Convergency}\label{convergency}}
+
+Plots and also the prediction plots.
+
+\hypertarget{sensibility-analysis}{%
+\subsubsection{Sensibility analysis}\label{sensibility-analysis}}
+
+About the possibility of assumptions about Missing mechanism.
 
@@
 
-In this example, we demonstrate a multilevel imputation of an intercept
-and slope random model with a continuous response. We use the obesity
-dataset included in the \texttt{micemd}@ package, a synthetic dataset
-that emulates an electronic survey in which individuals are asked to
-provide information about their weight and consumption habits in
-different countries. To easy the explanations we simulate data for 5
-clusters and we reduce the dataset to the following variables:
+In this example, we demonstrate a multilevel imputation of random
+intercept and random slope model with a continuous response. We utilize
+the obesity dataset included in the \texttt{micemd}@ package, a
+simulated dataset that emulates an electronic survey in which
+individuals are asked to provide information about their weight and
+consumption habits in different countries. We simulate data for 5
+clusters so that the true values are known. We use the following
+variables from the dataset:
 
@@
 \item
-  \texttt{region} Cluster variable,
-\item
-  \texttt{gender} Gender (0=male, 1=female),
-\item
-  \texttt{age} Age of the patient,
-\item
-  \texttt{height} Height in meters,
-\item
-  \texttt{weight} Weight in kilograms,
-\item
-  `time`` Response time in minutes (inclusion-restriction variable).
-\item
-  `FamOb`` Family obesity history (yes or not)
+  \textbf{Cluster:} Region of the patients' healthcare provider (Cluster
+  variable),
+\item
+  \textbf{Gender:} Subjects' Gender (0=male, 1=female),
+\item
+  \textbf{Age:} Subjects' age,
+\item
+  \textbf{Height:} Subjects' height in metres,
+\item
+  \textbf{Weight:} Subjects' weight in kilograms,
+\item
+  \textbf{BMI:} Subjects' body mass index,
+\item
+  \textbf{FamOb:} Family obesity history (yes or no),
+\item
+  \textbf{Time:} Response time in minutes (exclusion-restriction
+  variable).
 \end{itemize}
@@
 In the primary analysis model, BMI serves as the dependent variable,
-with Age, Gender, and FamOb as predictors. Researchers, based on
-observable data, are considering the inclusion of random intercept
-effects, as well as introducing a random slope for the Age variable.
-
-The main model is given by:
-\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
+with Age, Gender, and FamOb as predictors. Because of the clustered
+nature of the data, which is quantified with the Intraclass Correlation
+Coefficient (ICC) below, we include random intercepts, as well as a
+random slope for the Age variable. The model is represented as:
+\begin{equation}
+\label{eqn:main}
+BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}
+\end{equation}
+
 We start by loading the data:
@@
 \begin{CodeInput}
-R> require(patchwork)
+R> library(ggpubr)
 R> myplots <- lapply(1:5, function(i) {
 +   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
-+            ggtitle(paste0("Cluster", i))+ 
-+       theme(legend.position = ifelse(i==1,"bottom","none"))
++   ggplot2::ggtitle(paste0("Cluster", i))
 +  })
-R>   myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1} \end{center}
-
-\end{CodeChunk}
-
-We observe that the missing pattern is quite similar across the
-clusters. However, regarding the weight variable, we notice that is
-systematically missing in cluster 3. In order to evaluate if it is
-required a imputation method for 1-level or 2-level we assess the
-Intraclass Correlation Coefficient (ICC) for the outcome variable, we
-use the ``performance'' package:
+R> ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
+\end{CodeInput}
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1}
+
+}
+
+\caption[Missing pattern]{Missing pattern}\label{fig:obesity-md}
+\end{figure}
+\end{CodeChunk}
+
+We observe that the missing pattern is non-monotonic and quite similar
+across the clusters. However, regarding the weight variable, we notice
+that is systematically missing in cluster 3. In order to evaluate if we
+require a imputation method that accounts for clustering we assess the
+Intraclass Correlation
 
@@
 R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-R> performance::icc(Nulmodel)
-\end{CodeInput}
-\begin{CodeOutput}
-# Intraclass Correlation Coefficient
-
-    Adjusted ICC: 0.342
-  Unadjusted ICC: 0.342
-\end{CodeOutput}
-\end{CodeChunk}
-
-Since the ICC is above 0.1 and as the main analyis will be consider a
-mixed model, we decide to use two-level (2l) imputation methods. In this
-imputation process, we include all predictor variables in the main model
-and since BMI is a composite variable. We also incorporate weight and
-height variables.
-
-To determine the best imputation method for these variables, we can use
-the \texttt{find.defaultMethod}function provided in the \texttt{micemd}
-package, which suggests an appropriate method for MAR variables based on
-the type of variable, number of observations in the cluster, and number
-of clusters. It suggests using ``2l.2stage.bin'' for the FAV variable
-and ``2l.2stage.norm'' for the age variable. However, after inspecting
-the age density plot, we consider modifying its method to
-``2l.2stage.pmm.'' For the BMI variable, we employ deterministic
-imputation.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
+R> #Performance::ICC(Nulmodel)
+\end{CodeInput}
+\end{CodeChunk}
+
+Since the ICC is above 0.1 and as the main analysis will be use a mixed
+model, we decide to use two-level (2l) imputation methods. In this
+imputation process, we include all predictor variables from equation
+\ref{eqn:main} in the main model. However, since BMI is a composite of
+weight and height, we use deterministic imputation for these, which is
+described below.
+
+We use the \textbf{find.defaultMethodfunction} provided in the
+\textbf{micemd} package, which suggests an appropriate method for MAR
+variables based on the type of variable, number of observations in the
+cluster, and number of clusters.
+
+It suggests using `2l.2stage.bin' for the FAV variable and
+`2l.2stage.norm' for the age variable. However, after inspecting the age
+density plot, we consider modifying its method to `2l.2stage.pmm'. For
+the BMI variable, we employ deterministic imputation.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
++                                ni.small = 100, prop.small = 0.4)
 R> meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
@@
 For these imputation models, it is necessary to specify the prediction
-matrix, with the cluster variable labeled as -2 and the predictor
-variable labeled as 2, encompassing all variables. We need to suprime
-the variable Time as this variable is not especified in the main model
-and modify also in the prediction matrix the relationship between BMI,
-weight and height to avoid circular predictions. Then we proced to run
-the imputation model.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
-R> pred_mar[,"Cluster"]<- -2 # clustering variable
-R> pred_mar[,"Time"]<-0
-R> pred_mar[pred_mar==1]<-2
+matrix, with the cluster variable labelled as -2 and the predictor
+variable measured within clusters labelled as 2, encompassing all
+variables. We need to suprime the variable Time as this variable is not
+specified in the main model. We also modify the relationship between
+BMI, weight and height in the prediction matrix to avoid circular
+predictions. Then we proceed to run the imputation model.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred_mar <- quickpred(Obesity)   # predictor matrix
+R> pred_mar[,"Cluster"] <- -2 # clustering variable
+R> pred_mar[,"Time"] <- 0
+R> pred_mar[pred_mar==1] <- 2
 R> pred_mar[c("Height", "Weight"), "BMI"] <- 0
@@
 
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_pred-1} \end{center}
-
+\begin{center}\includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmar-1} \end{center}
+
+\begin{CodeInput}
+R> imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
++                 m=10, seed = 123, printFlag = FALSE)
+\end{CodeInput}
 \end{CodeChunk}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  2.409  69.703  82.925  82.476  95.310 158.420 
+ -9.826  69.500  82.789  82.337  94.919 148.712 
 \end{CodeOutput}
-\begin{CodeInput}
-R> plot(imp_mar)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-1} \end{center}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-2} \end{center}
-
-\end{CodeChunk}
-
-We are also contemplating the utilization of the ``pmm'' option, given
-that the imputed values' weights are lower than those of the observable
-values.
+\end{CodeChunk}
+
+We are also contemplating the utilisation of the ???pmm??? option, as
+the values imputed using a fully parametric method may be implausibly
+low for some patients.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth_mar["Weight"]<-"2l.2stage.pmm" 
+R> imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
++                     m=10, seed = 123, printFlag = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  27.50   71.02   83.82   83.76   95.55  135.98 
+  28.35   67.80   80.53   80.90   92.94  134.61 
 \end{CodeOutput}
 \begin{CodeInput}
-R> plot(imp_mar_pmm)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-1} \end{center}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-2} \end{center}
-
+R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+\end{CodeInput}
 \end{CodeChunk}
@@
 After confirming convergence, we proceed to save the results for future
-use. Users are considering the possibility that the weight variable may
-not have been selected randomly. It's likely that an omitted variable,
-like self-esteem, could influence this selection. For instance,
-individuals with lower self-esteem might have higher weight values,
-impacting their willingness to provide honest information due to
-embarrassment.
+use. We consider the possibility that patients may not have been
+selected randomly, which would then have led to a distribution for
+weight that does not reflect the weight in the population. It???s likely
+that an omitted variable, like self-esteem, could influence this
+selection. For instance, individuals with lower self-esteem might have
+higher weight values, impacting their willingness to provide honest
+information due to embarrassment.
 
@@
 selection models. Within pattern-mixed models, methods like the delta
-method and more advanced ones like NARFS have been suggested. In
-contrast, the selection model approach includes methods such as the
-Heckman model, which can be particularly useful in this case. Several
-methods, including Galimar2017, Hammon2021, and the recent Munoz method,
-designed for two-level data, allowing for variations in intercepts and
-exposure effects (random intercept and slope).
-
-To apply the heckman method, the weight variable should be specified as
-``2l.2stage.heckman'' found in the micemd package. Additionally, the
-prediction matrix needs modification because this method involves
-specifying two equations: one for the outcome, describing the incomplete
-variable in terms of partially observed predictors (in this case, all
-variables from the main model), and the other for the selection model,
-explaining the probability of being observed based (R) on certain
-variables.
-
-Here for the outcome equation we consider is the same imputation model
-that we used for the MAR case (main model).
-\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
+method and more advanced ones like NARFS have been suggested. The
+selection model approach includes methods such as the Heckman model,
+which can be particularly useful in this case. Several methods,
+including those by \cite{Galimar_2017,Hammon_2021}, and the recently a
+Heckman method designed for two-level data, allow for variations in
+intercepts and exposure effects (random intercept and slope).
+
+To apply the \textbf{2l.2stage.heckman} method, the weight variable
+should be specified as `2l.2stage.heckman' found in the micemd package.
+Additionally, the prediction matrix needs modification because this
+method involves specifying two equations: one for the outcome,
+describing the incomplete variable in terms of partially observed
+predictors (in this case, all variables from the main model), and the
+other for the selection model, explaining the probability of being
+observed based (R) on certain variables. For the outcome equation we
+consider the same imputation model that we used for the MAR case (main
+model).
+\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
 Regarding the selection equation, we include the same predictors as
-those in the primary model, as well as a time variable. Here the time
+those in the main model, as well as a time variable. Here the time
 variable serves as a restriction exclusion variable specifically
@@
 survey content or internet speed. These factors may lead the user to
-skip specific questions or even the entire survey. Also we assume the
-time is not have any influence on the weight assigned to the subject.
-\[R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
+skip specific questions or even the entire survey. Also, we assume the
+time does not have any influence on the subject???s weight.
+\[R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
+
 These two equations are jointly estimated under the assumption that the
@@
 a more comprehensive understanding of the model and the exclusion
-restriction, you can refer to (Munoz).
-
-To integrate information from both equations, we must adjust the
-prediction matrix. The cluster variable remains specified as before
-(-2). In this imputation method, all the variables present in both the
-selection and outcome equations are included with a random effect.
-However, it's essential to distinguish which of these variables appear
-in each equation.
-
-In this framework, when a variable is shared between both equations, it
-is denoted as (2). Predictors exclusive to the outcome equation are
-indicated as (-4), while those exclusive to the selection equation are
-labeled as (-3). Consequently, the only alteration needed in the
-predictor matrix pertains to the variable ``Time.''
+restriction, see \cite{Munoz_2022}.
+
+To use information from both equations, we must adjust the prediction
+matrix. The cluster variable remains specified as before (-2). In this
+imputation method, all the variables present in both the selection and
+outcome equations are included with a random effect.
+
+However, it is essential to distinguish which of these variables appear
+in each equation. In this framework, when a variable is shared between
+both equations, it is denoted as (2). Predictors exclusive to the
+outcome equation are indicated as (-4), while those exclusive to the
+selection equation are labelled as (-3). Consequently, the only
+alteration needed in the predictor matrix pertains to the variable
+`Time'.
 
@@
 R> pred_mnar["Weight","Time"]<- -3
-R> ggmice::plot_pred(pred_mnar)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_pred-1} \end{center}
-
-\end{CodeChunk}
-
-We also need to modify the method of the weight variable
+R> #ggmice::plot_pred(pred_mnar)
+\end{CodeInput}
+\end{CodeChunk}
+
+We also need to modify the method of the weight variable.
 
@@
 
-Then we proceed to run the imputation model as before Impute the
-missingness:
-
-After executing these imputation procedures, it is essential to assess
-convergence and the coherence of the imputed values. Upon examining the
-weight variable, we noticed that the imputed range falls outside the
-realm of plausible values (as weight should be positive). Consequently,
-we are contemplating a different approach, specifically, the use of
-``pmm,'' which involves imputing values from donor observations. This
-approach ensures that the imputed values remain within the range of
-observable values.
-
-\begin{CodeChunk}
-\begin{CodeInput}
+Then we proceed to run the imputation model as before, after executing
+these imputation procedures, it is essential to assess convergence and
+the coherence of the imputed values.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
++                 m=10, seed = 123, printFlag = FALSE)
 R> summary(complete(imp_mnar,"long")$Weight)
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  10.26   73.41   86.94   88.27  101.73  186.47 
+ -7.334  69.553  83.722  83.086  96.738 183.338 
 \end{CodeOutput}
-\begin{CodeInput}
-R> plot(imp_mnar)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-1} \end{center}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-2} \end{center}
-
-\end{CodeChunk}
-
-We then run the imputation model but this time using the option of pmm,
-to assure that weight values are in the range of the observable data,
-this can be implemented by seeting the pmm parameter as true.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(complete(imp_mnar_pmm,"long"))
+\end{CodeChunk}
+
+Upon examining the weight variable, we noticed that the imputed range
+falls outside the realm of plausible values (as weight should be
+positive).
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(complete(imp_mnar,"long")$Weight)
 \end{CodeInput}
 \begin{CodeOutput}
-      .imp           .id          Cluster           Time          Gender     
- Min.   : 1.0   Min.   :   1   Min.   :1.000   Min.   :1.016   Female:10720  
- 1st Qu.: 3.0   1st Qu.: 528   1st Qu.:2.000   1st Qu.:3.724   Male  :10390  
- Median : 5.5   Median :1056   Median :3.000   Median :5.151                 
- Mean   : 5.5   Mean   :1056   Mean   :2.868   Mean   :5.211                 
- 3rd Qu.: 8.0   3rd Qu.:1584   3rd Qu.:4.000   3rd Qu.:6.698                 
- Max.   :10.0   Max.   :2111   Max.   :5.000   Max.   :9.957                 
-     Height      FamOb            Age            Weight            BMI       
- Min.   :1.450   no : 3794   Min.   :14.00   Min.   : 27.50   Min.   : 9.59  
- 1st Qu.:1.642   yes:17316   1st Qu.:20.00   1st Qu.: 75.19   1st Qu.:25.84  
- Median :1.701               Median :25.00   Median : 88.88   Median :30.57  
- Mean   :1.702               Mean   :24.82   Mean   : 89.61   Mean   :31.02  
- 3rd Qu.:1.760               3rd Qu.:29.00   3rd Qu.:103.78   3rd Qu.:35.89  
- Max.   :1.961               Max.   :47.00   Max.   :135.98   Max.   :63.21  
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+ -7.334  69.553  83.722  83.086  96.738 183.338 
 \end{CodeOutput}
-\begin{CodeInput}
-R> plot(imp_mnar_pmm)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-1} \end{center}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-2} \end{center}
-
-\end{CodeChunk}
-
-Then after this modification we proceed to compare the effects on the
-model, therefore we run the analysis model on each of the completed
-datasets along with the dataset where the incomplete values are removed
-(CC).
+\end{CodeChunk}
+
+Consequently, as before we use the `pmm', option but this time for the
+Heckman imputation, this approach ensures that the imputed values remain
+within the range of observable values. We then run the imputation model
+but this time using the option of pmm, to assure that weight values are
+in the range of the observable data, this can be implemented by setting
+the pmm parameter to true.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
++                      m=10, seed = 123, pmm = T,  printFlag = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+We check the convergency of the results
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(complete(imp_mnar_pmm,"long")$Weight)
+\end{CodeInput}
+\begin{CodeOutput}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+  28.35   71.68   85.07   85.07   98.21  134.61 
+\end{CodeOutput}
+\begin{CodeInput}
+R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+\end{CodeInput}
+\end{CodeChunk}
+
+After this modification we proceed to compare the effects on the model.
+We run the analysis model on each of the completed datasets as well
+asthe dataset where the incomplete values are removed (Complete Case
+analysis, CC).
 
@@
 R> library(ggplot2)
-R> cc_rs<- with(Obesity[complete.cases(Obesity),],lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+R> cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
++              lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
 R> mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
@@
 R> mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+R> 
 R> list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
-R> plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
+R> plot_models(list_models,
++             mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
 \end{CodeInput}
@@
 
-We note that there is minimal disparity in the age effect, FamObs,
-across the various imputation models under consideration. Nonetheless,
-with respect to Gender, we observe a notable distinction: in the M(C)AR
-assumption, the effect is significant, whereas under the MNAR
-assumption, it becomes insignificant. An analysis of the intercept
-reveals that, under the MNAR assumption, a higher average BMI is
-anticipated compared to the MAR assumption.
+We note that there is minimal disparity in the age effect, FamObs, or
+Gender across the various imputation models under consideration. An
+analysis of the intercept reveals that, under the MNAR assumption, a
+higher average BMI is anticipated compared to the MAR assumption.
+Nonetheless, with respect to precision of estimates, we notice that in
+general MNAR imputation leads to wider confidence intervals, in this
+case it does not have any influence on the final result but there could
+be cases where variation in the assumed missing mechanism could lead
+also to differences on significant test and therefore lead to
+contradictory conclusions.
+
+\hypertarget{additional}{%
+\section{Additional}\label{additional}}
+
+The imputation of these data is based on the
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}
 
@@
 
-In hierarchical datasets, clustering is a concern because the
-homoscedasticity in the error terms cannot be assumed across clusters
-and the relationship among variables may vary at different hierarchical
-levels. When multiple imputation is used to deal with missing data, as
-the imputation and analysis process is performed separately, it is
-necessary that imputation model being congenial with the main analysis
-model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
-structure also imputation model should do it (Audigier, 2021). Not
-including clustering into the imputation process may lead to effect
-estimates with smaller standard errors and inflated type I error.
-
-There are different strategies that can be adopted in the imputation
-process that account for clustering: inclusion of cluster indicator
-variable, performing a separate imputation process for each cluster, or
-performing a simultaneous imputation process by using an imputation
-method that accounts for clustering.(Stata:
-\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
-TODO: replace ref.
-
-The selection of each strategy depends mainly on the assumptions in the
-main analysis and also on the restriction of the analyzed data.
-
-Regarding the restrictions imposed by the data, for instance, the use of
-cluster indicator variables is restricted in datasets where there are
-not many clusters and many observations per cluster (Graham, 2009). The
-last restriction is also required when imputations are performed on each
-cluster separately. When this restriction cannot be achieved, one can
-use an imputation model that simultaneously imputes all clusters using a
-hierarchical model (Allison 2002).
-
-Under this hierarchical imputation model, observations within clusters
-are correlated and this correlation is modeled by a random effect so the
-hierarchical model can be estimated even when there are few observations
-per cluster. However, this strategy is best suited for balanced data
-(Grund, 2017) and when random effects model is appropriated, i.e.~the
-number of clusters is adequate. (Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main
-model, for instance by using the cluster indicator strategy may lead to
-bias estimates when the model is based on a hierarchical model
-(Taaljard,2008). Even when an imputation strategy congenial with the
-main model is preferred, it is important to consider whether it is
-appropriate for the data as a less complex imputation strategies may
-also lead to unbiased estimates in certain scenarios(Bailey 2020). For
-instance, in causal effect analysis, separately imputation may lead to
-smaller bias when the size of the smaller exposure cluster is large,
-compared with an imputation model that includes exposure-confounder
-interactions. (Zhang,2023).
-
 \hypertarget{funding}{%
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5417}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4583}}@{}}
 \toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Formula \pkg{lme4}}
+\textbf{Formula} \pkg{lme4}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright

### 2023-11-10 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Johanna commit 2e8b414c631dc873416177dee20c7d44108de096

--- 2023-10-30 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 
+## Multilevel data
+
 Many datasets include individuals that are clustered together, for example in geographic regions, or even different studies.
@@
 
-```{r clus,  echo = FALSE, warning = FALSE,cache = TRUE}
-concepts <- data.frame(Concept=c("Sample unit","Hierarchical levels","Fixed effect","Random effect","Mixed effect","ICC","Stratified intercept"),
-                       Details=c("Units of the population from which measurements are taken in a sample.",
-                                 "Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.",
-                                 "Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.",
-                                 "Effects that are a source of random variation in the data, and whose levels are not fully sampled.
-e.g.
-individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.",
-                                 "Includes fixed and random effects, e.g.
-the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
-Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
-In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
-@gelm06, @hox17 and @jong21).",
-                                 "The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
-of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
-                                 ""))
-library(kableExtra)
-concepts%>%
-  kable( booktabs = T, escape = F,
-         caption = "Concepts in multilevel methods",
-         col.names = c("Concept","Details"), align = "l")%>%
-  column_spec(c(1), width = "3cm")%>%
-  column_spec(c(2), width = "12cm")
-```
+Table 1: Concepts in multilevel methods
+
+| **Concept**         | **Details**                                                                   |
+|---------------------|-------------------------------------------------------------------------------|
+| Sample unit         | Units of the population from which measurements are taken in a sample.        |
+| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
+| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
+|                     | the values observed in the study are representative of all values in the      |
+|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
+| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
+|                     | admission we might select only certain hospitals that are representative of   |
+|                     | the difference of y between individual hospitals, but rather the variation of |
+| ICC                 | The variability due to clustering is often measured by means of the           |
+|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
+|                     | of variance that can be attributed to the cluster-level, where a high         |
+|                     | ICC would indicate that a lot of variability is due to the cluster            |
+|                     | structure.                                                                    |
+| Random effect       | Multilevel models typically accommodate for variability by including          |
+|                     | a separate group mean for each cluster. In addition to random                 |
+|                     | intercepts, multilevel models can also include random coefficients            |
+|                     | and heterogeneous residual error variances across clusters [see e.g.          |
+|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |
 
@@
 
-Install non-CRAN packages if necessary:
-```{r packages, eval=FALSE, include=TRUE}
-devtools::install_github("amices/ggmice")
-
-```
-
 <!-- [TODO: Add environment info, seed and version number(s) somewhere.]  -->
@@
 
-The estimated effects in the complete data are visualized in Figure \ref{}.
+<!-- visualized in Figure \ref{}. -->
 
 ```{r forest, echo=FALSE}
-fits <-
-  purrr::map_dfr(
-    unique(impact$name),
-    ~ {
-      glm(
-        mort ~ 1 + age + motor_score + pupil + ct,
-        family = "binomial",
-        data = filter(impact, name == .x)
-      )
-    } %>%
-      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
-      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
-      cbind(name = as.character(.x), .)
-  ) %>% filter(term %in% c("ctIII", "ctIV/V"))
+# fits <-
+#   purrr::map_dfr(
+#     unique(impact$name),
+#     ~ {
+#       glm(
+#         mort ~ 1 + age + motor_score + pupil + ct,
+#         family = "binomial",
+#         data = filter(impact, name == .x)
+#       )
+#     } %>%
+#       broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
+#       .[, c("term", "estimate", "conf.low", "conf.high")] %>%
+#       cbind(name = as.character(.x), .)
+#   ) %>% filter(term %in% c("ctIII", "ctIV/V"))
+# 
+# n <- impact %>% group_by(name) %>% 
+#   tally()
+# 
+# fits %>% left_join(n, by = "name") %>% 
+#   ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
+#   geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
+#   geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
+#   geom_point(shape = 20, position = position_dodge(width = 0.5)) +
+#   labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
+#   theme_classic() +
+#   scale_size(guide = 'none')
+# # TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
+# # TODO: missingness prop inside clusters
+# # TODO: make row height proportional to pattern frequency
+```
 
-n <- impact %>% group_by(name) %>% 
-  tally()
-
-fits %>% left_join(n, by = "name") %>% 
-  ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
-  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
-  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
-  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
-  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
-  theme_classic() +
-  scale_size(guide = 'none')
-# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
-# TODO: missingness prop inside clusters
-# TODO: make row height proportional to pattern frequency
-```
-<!-- We will use the following estimates as comparative truth in this tutorial: -->
-```{r truth}
-# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
-# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
+We will use the complete data estimates as comparative truth in this tutorial. The estimated effects in the complete data are presented in Table XYZ. 
+```{r impact_truth, include=FALSE, echo=TRUE}
+fit <- glmer(mod, family = "binomial", data = impact) # fit the model
+tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
 ```
@@
 
+Evaluate the convergence of the algorithm:
+```{r impact-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
+plot_trace(imp)
+```
+
+Analyze the imputed data:
 ```{r impact-fit, cache=TRUE}
 fit <- imp %>% 
-  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
-tidy(pool(fit))
-as.mitml.result(fit)
+  with(glmer(
+    mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
+    family = "binomial"
+    )) 
+# tidy(pool(fit))
+# as.mitml.result(fit)
 # testEstimates(as.mitml.result(fit))
 ```
+The estimated effects after imputation are presented in Table XYZ.
 
@@
 
-In this example, we demonstrate a multilevel imputation of an intercept and slope random model with a continuous response.
-We use the obesity dataset included in the `micemd`@ package, a synthetic dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries.
-To easy the explanations we simulate data for 5 clusters and we reduce the dataset to the following variables:
+TODO: explain exclusion restriction.
+
+Data are simulated and included in the `micemd` package. We will use the following variables:
 
@@
 - `time`` Response time in minutes (inclusion-restriction variable).
-- `FamOb`` Family obesity history (yes or not)
 
-In this dataset, Age and FamOb are MAR, while the weight variable is affected by selection bias, attributed to an indirect MNAR mechanism.
-This MNAR mechanism typically arises when an unobserved or omitted variable influences both the value of the incomplete variable (in this case, Weight) and its likelihood of being missing (denoted as R).
+The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
 
-In the primary analysis model, BMI serves as the dependent variable, with Age, Gender, and FamOb as predictors.
-Researchers, based on observable data, are considering the inclusion of random intercept effects, as well as introducing a random slope for the Age variable.
+<!-- [TODO: add reference Arxiv paper Johanna].  -->
 
-The main model is given by:
-$$BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}$$
-We start by loading the data:
-```{r, include=FALSE}
-library(here)
-library(performance)
-source(here("heckman","ipdfunction.R"),local =knitr::knit_global())
-load(here("heckman","Obesity.rda"))
+Load the data:
+```{r obesity}
+data("data_heckman", package = "micemd")
 ```
 
-```{r obesity}
-#data("data_heckman", package = "micemd")
-#dat <- data_heckman
+Load the data:
+```{r}
+#data("obesity", package = "miceheckman")
 ```
 
-Now, let's begin by examining the missing patterns in the data by cluster:
-```{r obesity-md}
-require(patchwork)
-  myplots <- lapply(1:5, function(i) {
-         ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
-           ggtitle(paste0("Cluster", i))+ 
-      theme(legend.position = ifelse(i==1,"bottom","none"))
-      })
-  myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]
-
+Select relevant variables:
+```{r bmi}
+#dat <- select(obesity, -"bmi")
 ```
 
-We observe that the missing pattern is quite similar across the clusters. However, regarding the weight variable, we notice that is systematically missing in cluster 3.
-In order to evaluate if it is required a imputation method for 1-level or 2-level we assess the Intraclass Correlation Coefficient (ICC) for the outcome variable, we use the "performance" package:
-
-```{r obesity icc}
-Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-performance::icc(Nulmodel)
+Create predictor matrix:
+```{r}
+#pred <- quickpred(dat)   # predictor matrix
+#pred["weight","cluster"]<- -2 # clustering variable
+#pred["weight","rt"] <- -3 #inclusion-restriction variable
+#plot_pred(pred)
 ```
 
-Since the ICC is above 0.1 and as the main analyis will be consider a mixed model, we decide to use two-level (2l) imputation methods.
-In this imputation process, we include all predictor variables in the main model and since BMI is a composite variable.
-We also incorporate weight and height variables.
-
-To determine the best imputation method for these variables, we can use the `find.defaultMethod`function provided in the `micemd` package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters.
-It suggests using "2l.2stage.bin" for the FAV variable and "2l.2stage.norm" for the age variable.
-However, after inspecting the age density plot, we consider modifying its method to "2l.2stage.pmm." For the BMI variable, we employ deterministic imputation.
-```{r obsmar_met}
-meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
-meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
-meth_mar["Age"]<-"2l.2stage.pmm" 
-```
-For these imputation models, it is necessary to specify the prediction matrix, with the cluster variable labeled as -2 and the predictor variable labeled as 2, encompassing all variables.
-We need to suprime the variable Time as this variable is not especified in the main model and modify also in the prediction matrix the relationship between BMI, weight and height to avoid circular predictions.
-Then we proced to run the imputation model.
-
-```{r obsmar_pred,echo = T}
-pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
-pred_mar[,"Cluster"]<- -2 # clustering variable
-pred_mar[,"Time"]<-0
-pred_mar[pred_mar==1]<-2
-pred_mar[c("Height", "Weight"), "BMI"] <- 0
-ggmice::plot_pred(pred_mar)
-```
-```{r obsmar_mice,echo = T,results="hide",include=FALSE}
-imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar, m=10, seed = 123)
+Set the Heckman model as imputation method:
+```{r}
+#meth <- make.method(dat) # methods vector
+#meth["weight"]<-"2l.heckman"
 ```
 
-```{r obsmar_plot}
-summary(complete(imp_mar,"long")$Weight)
-plot(imp_mar)
-```
-We are also contemplating the utilization of the "pmm" option, given that the imputed values' weights are lower than those of the observable values.
-
-```{r obsmarp_mice,results="hide",include=FALSE}
-meth_mar["Weight"]<-"2l.2stage.pmm" 
-imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar, m=10, seed = 123)
-```
-```{r obsmarp_plot}
-summary(complete(imp_mar_pmm,"long")$Weight)
-plot(imp_mar_pmm)
-```
-After confirming convergence, we proceed to save the results for future use.
-Users are considering the possibility that the weight variable may not have been selected randomly.
-It's likely that an omitted variable, like self-esteem, could influence this selection.
-For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.
-
-To address this situation, two approaches have been proposed for dealing with Missing Not at Random (MNAR) data: pattern-mixed models and selection models.
-Within pattern-mixed models, methods like the delta method and more advanced ones like NARFS have been suggested.
-In contrast, the selection model approach includes methods such as the Heckman model, which can be particularly useful in this case.
-Several methods, including Galimar2017, Hammon2021, and the recent Munoz method, designed for two-level data, allowing for variations in intercepts and exposure effects (random intercept and slope).
-
-To apply the heckman method, the weight variable should be specified as "2l.2stage.heckman" found in the micemd package.
-Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.
-
-Here for the outcome equation we consider is the same imputation model that we used for the MAR case (main model). 
-$$Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}$$
-Regarding the selection equation, we include the same predictors as those in the primary model, as well as a time variable.
-Here the time variable serves as a restriction exclusion variable specifically explaining the probability of being observed but not affecting the incomplete value (Weight).
-In this context, we assume that the time a user spends completing the survey serves as a proxy for the barriers they may encounter in survey completion, such as familiarity with the survey content or internet speed.
-These factors may lead the user to skip specific questions or even the entire survey.
-Also we assume the time is not have any influence on the weight assigned to the subject.
-$$R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}$$
-These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution.
-For a more comprehensive understanding of the model and the exclusion restriction, you can refer to (Munoz).
-
-To integrate information from both equations, we must adjust the prediction matrix.
-The cluster variable remains specified as before (-2).
-In this imputation method, all the variables present in both the selection and outcome equations are included with a random effect.
-However, it's essential to distinguish which of these variables appear in each equation.
-
-In this framework, when a variable is shared between both equations, it is denoted as (2).
-Predictors exclusive to the outcome equation are indicated as (-4), while those exclusive to the selection equation are labeled as (-3).
-Consequently, the only alteration needed in the predictor matrix pertains to the variable "Time."
-
-```{r obsmnar_pred}
-pred_mnar <- pred_mar
-pred_mnar["Weight","Time"]<- -3
-ggmice::plot_pred(pred_mnar)
+Visualize missing data pattern:
+```{r obesity-md}
+plot_pattern(dat)
 ```
 
-We also need to modify the method of the weight variable
-```{r obsmnar_meth}
-meth_mnar <- meth_mar
-meth_mnar["Weight"]<- "3l.2stage.heckman"
+The matrix only shows the predictors for the main model, not the selection model.
+
+Create predictor matrix:
+```{r obesity-pred, eval = FALSE}
+pred <- quickpred(dat)   # predictor matrix
+pred[,"Cluster"] <- -2 # clustering variable
+pred[, "Weight"] <- -3 #inclusion-restriction variable
+plot_pred(pred)
 ```
 
-Then we proceed to run the imputation model as before 
+Set the Heckman model as imputation method:
+```{r, eval = FALSE}
+meth <- make.method(dat) # methods vector
+meth["weight"]<-"2l.heckman"
+```
+
 Impute the missingness:
-```{r obsmnar_mice,results="hide",include=FALSE}
-imp_mnar <- mice(
-  Obesity, # dataset with missing values
-  m = 10, # number of imputations
+```{r eval=FALSE, cache=TRUE}
+imp <- mice(
+  dat, # dataset with missing values
+  m = 5, # number of imputations
   maxit = 10,
-  meth = meth_mnar, #imputation method vector
-  pred = pred_mnar, #imputation predictors matrix
+  meth = meth, #imputation method vector
+  pred = pred, #imputation predictors matrix
   meta_method = "reml",
-  pred_std=FALSE,
   printFlag = FALSE
@@
 ```
-
-After executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
-Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive).
-Consequently, we are contemplating a different approach, specifically, the use of "pmm," which involves imputing values from donor observations.
-This approach ensures that the imputed values remain within the range of observable values.
-
-```{r obsmnar_plot}
-summary(complete(imp_mnar,"long")$Weight)
-plot(imp_mnar)
-```
-
-We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by seeting the pmm parameter as true.
-```{r obsmnarp_mice,results="hide",include=FALSE}
-imp_mnar_pmm <- mice(
-  Obesity, # dataset with missing values
-  m = 10, # number of imputations
-  maxit = 10,
-  meth = meth_mnar, #imputation method vector
-  pred = pred_mnar, #imputation predictors matrix
-  meta_method = "reml",
-  pred_std=FALSE,
-  pmm=TRUE,
-  printFlag = FALSE
-)
-```
-
-```{r obsmnarp_plot}
-summary(complete(imp_mnar_pmm,"long"))
-plot(imp_mnar_pmm)
-```
-
-Then after this modification we proceed to compare the effects on the model, therefore we run the analysis model on each of the completed datasets along with the dataset where the incomplete values are removed (CC).
-
-```{r models}
-library(ggplot2)
-cc_rs<- with(Obesity[complete.cases(Obesity),],lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
-plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
-```
-
-We note that there is minimal disparity in the age effect, FamObs, across the various imputation models under consideration.
-Nonetheless, with respect to Gender, we observe a notable distinction: in the M(C)AR assumption, the effect is significant, whereas under the MNAR assumption, it becomes insignificant.
-An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
-
-<!-- [TODO: add reference Arxiv paper Johanna].  -->
-
-Visualize missing data pattern:
-
-The matrix only shows the predictors for the main model, not the selection model.
 

### 2023-11-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

@@
 More complex models can be accomodated, but are outside the scope of this paper.
+Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy.
+In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed.This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R.
+We aim to show how imputation can yield less biased estimates from incomplete clustered data.
+We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms.
+For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
 keywords:
@@
 
-| **Concept**         | **Details**                                                                   |
-|---------------------|-------------------------------------------------------------------------------|
-| Sample unit         | Units of the population from which measurements are taken in a sample.        |
-| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
-| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
-|                     | the values observed in the study are representative of all values in the      |
-|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
-| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
-|                     | admission we might select only certain hospitals that are representative of   |
-|                     | the difference of y between individual hospitals, but rather the variation of |
-| ICC                 | The variability due to clustering is often measured by means of the           |
-|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
-|                     | of variance that can be attributed to the cluster-level, where a high         |
-|                     | ICC would indicate that a lot of variability is due to the cluster            |
-|                     | structure.                                                                    |
-| Random effect       | Multilevel models typically accommodate for variability by including          |
-|                     | a separate group mean for each cluster. In addition to random                 |
-|                     | intercepts, multilevel models can also include random coefficients            |
-|                     | and heterogeneous residual error variances across clusters [see e.g.          |
-|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |
+```{r clus0,  echo = FALSE, warning = FALSE,cache = TRUE}
+concepts <- data.frame(Concept=c("Sample units","Cluster","Hierarchical data","Level-1","Level-2","Hierarchical model","Fixed effect","Random effect","Mixed effect","ICC","Stratified intercept"),
+                       Details=c("Units of the population from which measurements are taken in a sample, e.g., students.", "Variable that specify the cluster or agruppation, e.g., Classroom","Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.","Variable that varies within a cluster, eg.
+Test score","Variable that does not vary within a cluster but between, e.g.
+teacher experience.","Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)" ,"Effects that are constant across all sample units, e.g.
+something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)","Effects that are a source of random variation in the data, and whose levels are not fully sampled.
+e.g.
+test score tendency during academic year between students due to no controlled factors such as  genetic,family history","Includes fixed and random effects, e.g.
+the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
+Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
+In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
+@gelm06, @hox17 and @jong21).","The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
+The ICC can be seen as the percentage
+of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
+                                 ""))
+library(kableExtra)
+concepts%>%
+  kable( booktabs = T, escape = F,
+         caption = "Concepts in multilevel methods",
+         col.names = c("Concept","Details"), align = "l")%>%
+  column_spec(c(1), width = "3cm")%>%
+  column_spec(c(2), width = "12cm")
+```
 
@@
 
+## Imputation workflow
+Below we provide a imputation workflow that can be used in general to impute cluster data.
+
+### Main analysis
+When imputing clustered data, our first consideration should be the research question and the type of analysis that researchers intend to perform, assuming that non-incomplete values were present.
+This initial assessment provides valuable insights into the data's structure (see the level table), relationships among variables, and the causes of missingness.
+
+When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g.
+if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021).
+
+### Exploration of available data
+Then it is required to explore available information across clusters.
+This can involve evaluating the Intra-class Correlation Coefficient (ICC) to determine if a one-level imputation method is sufficient, particularly if there are no substantial differences between clusters.
+Additionally, it can be done by visualizing the data using available package as **ggmice**, for instance examining density plots can help us understand the variable type (normal, Poisson, categorical) and its range of plausible values.
+Exploring relationships between covariates through bivariate plots can be valuable in identifying potential interactions or non-linearities.
+
+In the context of missingness, we should assess patterns of systematic and sporadic missing values and explore the relationships between the missingness of an incomplete variable and observable data.
+This evaluation allows us to consider the plausibility of Missing at Random (MAR) mechanisms over Missing Completely at Random (MCAR) methods and helps in selecting prediction variables to be included in the imputation model (considering overflux).
+
+### Selecting plausible imputation methods
+Before to implement multiple imputation it is requiered to evaluate if simpler imputation methods are suffice: 
+- For example, certain incomplete variables may not necessitate stochastic imputation methods like MICE.
+Instead, they can be effectively addressed through deductive imputation, where incomplete values can be inferred from logic and deterministic relationships between variables.
+This approach is particularly useful for variables that are functions of others, e.g.
+a person's BMI from their weight and height, determining values for one-level variables from two-level ones, e.g.
+in the context of Individual Participant Data (IPD), incomplete information can be extracted from metadata, such as deducing incomplete data about abortion in a country where abortion is illegal, or engaging in cross-temporal or protocol's deduction, like imputing missing test values for deceased patients.
+
+- Under monotonic pattern, the use of regression models
+- CC when the proportion of missing is minimal <5%.
+- Analysis that are robust to missing data, for instance FIML models such as mixed model do not require imputation models when only the outcome variable is missing.An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML).
+This method does not require the imputation of any missing values.
+Whereas MI consists of imputation, analyses and pooling steps, FIML analyses the data in a single step.
+When the assumptions are met the two approaches should produce equivalent results.
+[REF] As FIML requires specialised software, not all analyses can be performed with standard software.
+[REF]
+
+- Add table describing 2l- imputation models. like this https://bookdown.org/mwheymans/bookmi/multiple-imputation-models-for-multilevel-data.html
+
+### Imputation process
+
+#### Clustering specification
+There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: <https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/>) TODO: replace ref.
+
+The selection of each strategy depends mainly on the assumptions in the main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
+The last restriction is also required when imputations are performed on each cluster separately.
+When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
+However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.
+the number of clusters is adequate.
+(Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008).
+Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020).
+For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
+(Zhang,2023).
+
+#### Selection of variables in the imputation model
+- Inclusion of outcome variable [Moons] When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]
+- Imputation for interaction term, JAV, Machine lerarning methods (Random forest) and imputation separated by treatment group
+- Prediction matrix specfication (table)
+
+#### Convergency problems during imputation
+- Check log 
+- Problems of convergence on some 2l methods can be solved by use of overflux or quickpred. 
+- Consider the use of scaled variables Vanburen. 
+- Consider to swap imputation mehtod in problematic variables: The use of non parametrical imputation methods such as pmm. or downgrade to 1l methods.
+- Also by the specification of post-processing in some variables can alivianate this. 
+- Separated imputation process for each endpoint analysis
+
+### Convergency 
+Plots and also the prediction plots.
+
+### Sensibility analysis
+About the possibility of assumptions about Missing mechanism.
+
+<!-- With the `popmis` data, we show how (and how not) to develop an imputation model.
+With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods.
+With the `impact` data we provide an example of multivariate missingness in real-world data.
+Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.
+-->
+
+<!-- TODO: explicit statement about not going into workings of the methods. Galimer 2l methods. -->
+
 ## Setup
@@
 
-We'll now use `school` as a predictor to impute all other variables.
+We will now use `school` as a predictor to impute all other variables.
 This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16].
@@
 Impute the incomplete data
-```{r imp_impact, cache=TRUE}
+```{r imp_impact}
 imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
@@
 Analyze the imputed data:
-```{r impact-fit, cache=TRUE}
+```{r impact-fit}
 fit <- imp %>% 
@@
 # Case study III: obesity data
+
+In this example, we demonstrate a multilevel imputation of random intercept and random slope model with a continuous response.
+We utilize the obesity dataset included in the `micemd`\@ package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries.
+We simulate data for 5 clusters so that the true values are known.
+We use the following variables from the dataset:
+
+-   **Cluster:** Region of the patients' healthcare provider (Cluster variable),
+-   **Gender:** Subjects' Gender (0=male, 1=female),
+-   **Age:** Subjects' age,
+-   **Height:** Subjects' height in metres,
+-   **Weight:** Subjects' weight in kilograms,
+-   **BMI:** Subjects' body mass index,
+-   **FamOb:** Family obesity history (yes or no),
+-   **Time:** Response time in minutes (exclusion-restriction variable).
+
+In this dataset, Age and FamOb are MAR, while the weight variable is affected by selection bias, attributed to an indirect MNAR mechanism.
+This MNAR mechanism typically arises when an unobserved or omitted variable influences both the value of the incomplete variable (in this case, Weight) and its likelihood of being missing (denoted as R).
+
+In the primary analysis model, BMI serves as the dependent variable, with Age, Gender, and FamOb as predictors.
+Because of the clustered nature of the data, which is quantified with the Intraclass Correlation Coefficient (ICC) below, we include random intercepts, as well as a random slope for the Age variable.
+The model is represented as:
+\begin{equation}
+\label{eqn:main}
+BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}
+\end{equation}
+
+```{r, include=FALSE, eval=FALSE}
+library(here)
+source(here("heckman","ipdfunction.R"),local =knitr::knit_global())
+load(here("heckman","Obesity.rda"))
+```
+
+We start by loading the data:
+
+```{r}
+#data("data_heckman", package = "micemd")
+#dat <- data_heckman
+```
+
+Now, let's begin by examining the missing patterns in the data by cluster:
+
+```{r, dpi = 300,out.width="70%", fig.cap="Missing pattern", eval=FALSE}
+library(ggpubr)
+myplots <- lapply(1:5, function(i) {
+  ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
+  ggplot2::ggtitle(paste0("Cluster", i))
+ })
+ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
+```
+
+We observe that the missing pattern is non-monotonic and quite similar across the clusters.
+However, regarding the weight variable, we notice that is systematically missing in cluster 3.
+In order to evaluate
+if we require a imputation method that accounts for clustering  we assess the Intraclass Correlation
+
+```{r obesity-icc, eval=FALSE}
+Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+#Performance::ICC(Nulmodel)
+```
+
+Since the ICC is above 0.1 and as the main analysis will be use a mixed model, we decide
+to use two-level (2l) imputation methods. In this imputation process, we include all predictor
+variables from equation \ref{eqn:main} in the main model. However, since BMI is a composite of
+weight and height, we use deterministic imputation for these, which is described below.
+
+We use the **find.defaultMethodfunction** provided in the **micemd** package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters.
+
+It suggests using '2l.2stage.bin' for the FAV variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
+'2l.2stage.pmm'. For the BMI variable, we employ deterministic imputation.
+
+```{r obesitymeth, eval=FALSE}
+meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+                               ni.small = 100, prop.small = 0.4)
+meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
+meth_mar["Age"]<-"2l.2stage.pmm" 
+```
+
+For these imputation models, it is necessary to specify the prediction matrix, with the cluster
+variable labelled as -2 and the predictor variable measured within clusters labelled as 2, encompassing all variables.
+We need to suprime the variable Time as this variable is not specified in the main model.
+We also  modify the relationship between BMI, weight and height in the prediction matrix to avoid circular predictions.
+Then we proceed to run the imputation model.
+```{r obesity-predmar,out.width="70%", eval=FALSE}
+pred_mar <- quickpred(Obesity)   # predictor matrix
+pred_mar[,"Cluster"] <- -2 # clustering variable
+pred_mar[,"Time"] <- 0
+pred_mar[pred_mar==1] <- 2
+pred_mar[c("Height", "Weight"), "BMI"] <- 0
+ggmice::plot_pred(pred_mar)
+imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+                m=10, seed = 123, printFlag = FALSE)
+```
+
+```{r obesity-predmar1,eval=FALSE}
+summary(complete(imp_mar,"long")$Weight)
+```
+We are also contemplating the utilisation of the ???pmm??? option, as the values imputed using a fully parametric method may be implausibly low for some patients.
+
+```{r obesity-predmar_pmm, eval=FALSE}
+meth_mar["Weight"]<-"2l.2stage.pmm" 
+imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+                    m=10, seed = 123, printFlag = FALSE)
+```
+
+```{r obesity-predmar_pmm1,eval=FALSE}
+summary(complete(imp_mar_pmm,"long")$Weight)
+#ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+```
+
+After confirming convergence, we proceed to save the results for future use.
+We consider the possibility that patients may not have been selected randomly, which would then have led to a distribution for weight that does not reflect the weight in the population.
+It???s likely that an omitted variable, like self-esteem, could influence this selection.
+For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.
+
+To address this situation, two approaches have been proposed for dealing with Missing Not at Random (MNAR) data: pattern-mixed models and selection models.
+Within pattern-mixed models, methods like the delta method and more advanced ones like NARFS have been suggested.
+The selection model approach includes methods such as the Heckman model, which can be particularly useful in this case.
+Several methods, including those by \cite{Galimar_2017,Hammon_2021}, and the recently a Heckman method designed for two-level data, allow for variations in intercepts and exposure effects (random intercept and slope).
+
+To apply the **2l.2stage.heckman** method, the weight variable should be specified as '2l.2stage.heckman' found in the micemd package.
+Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.
+For the outcome equation we consider the same imputation model that we used for
+the MAR case (main model).
+ $$Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}$$ 
+Regarding the selection equation, we include the same predictors as those in the main model, as well as a time variable.
+Here the time variable serves as a restriction exclusion variable specifically explaining the probability of being observed but not affecting the incomplete value (Weight).
+In this context, we assume that the time a user spends completing the survey serves as a proxy for the barriers they may encounter in survey completion, such as familiarity with the survey content or internet speed.
+These factors may lead the user to skip specific questions or even the entire survey.
+Also, we assume the time does not have any influence on the subject???s weight.
+ $$R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}$$
+
+These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution.
+For a more comprehensive understanding of the model and the exclusion restriction, see \cite{Munoz_2022}.
+
+To use information from both equations, we must adjust the prediction matrix.
+The cluster variable remains specified as before (-2).
+In this imputation method, all the variables present in both the selection and outcome equations are included with a random effect.
+
+However, it is essential to distinguish which of these variables appear in each equation.
+In this framework, when a variable is shared between both equations, it is denoted as (2).
+Predictors exclusive to the outcome equation are indicated as (-4), while those exclusive to the selection equation are labelled as (-3).
+Consequently, the only alteration needed in the predictor matrix pertains to the variable 'Time'.
+
+```{r obesity-predmnar, eval=FALSE}
+pred_mnar <- pred_mar
+pred_mnar["Weight","Time"]<- -3
+#ggmice::plot_pred(pred_mnar)
+```
+
+We also need to modify the method of the weight variable.
+
+```{r obesity-predmnar1, eval=FALSE}
+meth_mnar <- meth_mar
+meth_mnar["Weight"]<- "3l.2stage.heckman"
+```
+Then we proceed to run the imputation model as before, after executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
+```{r obesity-predmnar2,eval=FALSE}
+imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+                m=10, seed = 123, printFlag = FALSE)
+summary(complete(imp_mnar,"long")$Weight)
+```
+
+Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive).
+
+```{r obesity-predmnar3,eval=FALSE}
+summary(complete(imp_mnar,"long")$Weight)
+```
+Consequently, as before we  use the 'pmm', option but this time for the Heckman imputation, this approach ensures that the imputed values remain within the range of observable values.
+We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by setting the pmm parameter to true.
+
+```{r obesity-predmnarp, eval=FALSE}
+imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+                     m=10, seed = 123, pmm = T,  printFlag = FALSE)
+```
+
+We check the convergency of the results
+```{r obesity-predmnarp1, eval=FALSE}
+summary(complete(imp_mnar_pmm,"long")$Weight)
+#ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+```
+
+After this modification we proceed to compare the effects on the model. We
+run the analysis model on each of the completed datasets as well asthe dataset where the
+incomplete values are removed (Complete Case analysis, CC).
+
+```{r models, eval=FALSE}
+library(ggplot2)
+cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
+             lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
+mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
+plot_models(list_models,
+            mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
+```
+
+We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration.
+An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
+Nonetheless, with respect to precision of estimates, we notice that in general MNAR imputation leads to wider confidence intervals, in this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on significant test and therefore lead to contradictory conclusions.
+
+# Additional
+
+The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
+
+<!-- [TODO: add reference Arxiv paper Johanna].  -->
+
+Visualize missing data pattern:
+
+The matrix only shows the predictors for the main model, not the selection model.
 
@@
 # Discussion
+
+ORDER:
+
+-   summary
+
+-   congeniality, then in hierarchical models
+
+-   look whether we can fit cong. back in the main body
+
+-   alt. methods
+
+-   conclusion: mice is really easy!
+
+-   Additional levels of clustering
+
+-   More complex data types: timeseries and polynomial relationship in the clustering.
+
+-   FIML vs MI
+
+An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML).
+This method does not require the imputation of any missing values.
+Whereas MI consists of imputation, analyses and pooling steps, FIML analyses the data in a single step.
+When the assumptions are met the two approaches should produce equivalent results.
+[REF] As FIML requires specialised software, not all analyses can be performed with standard software.
+[REF]
+
+-   Survival / TTE, this could be put in the paragraph on congeniality
+
+When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]
 

### 2023-11-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

--- 2023-10-30 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:clus}Concepts in multilevel methods}
-\centering
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
-\toprule
-Concept & Details\\
-\midrule
-Sample unit & Units of the population from which measurements are taken in a sample.\\
-Hierarchical levels & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
-Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.\\
-Random effect & Effects that are a source of random variation in the data, and whose levels are not fully sampled.
-e.g.
-individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.\\
-Mixed effect & Includes fixed and random effects, e.g.
-the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
-Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
-In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
-@gelm06, @hox17 and @jong21).\\
-\addlinespace
-ICC & The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
-of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.\\
-Stratified intercept & \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
+Table 1: Concepts in multilevel methods
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Concept}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Details}
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Sample unit & Units of the population from which measurements are taken
+in a sample. \\
+Hierarchical levels & Data are grouped into clusters at different
+levels. A three-level \\
+Fixed effect & Here we assume that the values of an independent variable
+are fixed, i.e., \\
+& the values observed in the study are representative of all values in
+the \\
+& in the dependent variable y e.g., blood pressure between treatments A
+and B. \\
+Random effect & The values of an independent variable are assumed to be
+randomly drawn from \\
+& admission we might select only certain hospitals that are
+representative of \\
+& the difference of y between individual hospitals, but rather the
+variation of \\
+ICC & The variability due to clustering is often measured by means of
+the \\
+& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
+& of variance that can be attributed to the cluster-level, where a
+high \\
+& ICC would indicate that a lot of variability is due to the cluster \\
+& structure. \\
+Random effect & Multilevel models typically accommodate for variability
+by including \\
+& a separate group mean for each cluster. In addition to random \\
+& intercepts, multilevel models can also include random coefficients \\
+& and heterogeneous residual error variances across clusters {[}see
+e.g. \\
+& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
+stratified intercept as concept..{]} \\
+\end{longtable}
 
@@
 
-Install non-CRAN packages if necessary:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> devtools::install_github("amices/ggmice")
-\end{CodeInput}
-\end{CodeChunk}
-
 Set up the R environment and load the necessary packages:
@@
 \begin{CodeInput}
-R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
+R> mod <- mort ~ type + age + motor_score + pupil + ct + (1 | name) 
 \end{CodeInput}
@@
 
-The estimated effects in the complete data are visualized in Figure
-\ref{}.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/forest-1} \end{center}
-
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # fit <- glmer(mod, family = "binomial", data = impact) # fit the model
-R> # tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
-\end{CodeInput}
-\end{CodeChunk}
+We will use the complete data estimates as comparative truth in this
+tutorial. The estimated effects in the complete data are presented in
+Table XYZ.
 
@@
 R> dat <- select(dat, !hypox)  # remove variable
+R> dat <- mutate(dat, motor_score = as.factor(motor_score))
 \end{CodeInput}
@@
    <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
- 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   3.00e-31   0.0571     0.130
+ 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
  2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
@@
  9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
-10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.56e-19   1.92       2.76 
+10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
 11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
@@
 R> fit <- imp %>% 
-+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
-R> tidy(pool(fit))
-\end{CodeInput}
-\begin{CodeOutput}
-                     term    estimate   std.error  statistic      p.value
-1             (Intercept) -2.35203726 0.340181747  -6.914061 4.994037e-12
-2                    type -0.41265892 0.180274846  -2.289054 2.209524e-02
-3                     age  0.03049023 0.001570162  19.418521 1.238416e-81
-4 as.factor(motor_score)2 -0.66764920 0.068737865  -9.712975 3.480413e-22
-5 as.factor(motor_score)3 -1.05520001 0.070218940 -15.027285 2.540225e-50
-6 as.factor(motor_score)4 -1.51238349 0.072304262 -20.916934 1.850073e-90
-7                   pupil  0.48421447 0.038982800  12.421234 6.772069e-17
-8                      ct  0.43474621 0.029968474  14.506785 2.342253e-36
-             b          df dfcom         fmi      lambda m         riv
-1 5.281599e-04 10119.70041 11013 0.005673265 0.005476771 5 0.005506932
-2 4.881699e-05 10893.89378 11013 0.001985736 0.001802528 5 0.001805783
-3 3.335358e-08  6320.89582 11013 0.016545467 0.016234340 5 0.016502243
-4 4.188703e-05  8327.24771 11013 0.010875748 0.010638213 5 0.010752602
-5 5.135721e-05  7632.20045 11013 0.012757637 0.012498967 5 0.012657168
-6 1.403058e-04  2831.75462 11013 0.032888241 0.032205434 5 0.033277139
-7 3.571497e-04    49.97295 11013 0.309130930 0.282023647 5 0.392803531
-8 8.586645e-05   294.69765 11013 0.120677044 0.114729598 5 0.129598366
-          ubar
-1 1.150898e-01
-2 3.244044e-02
-3 2.425385e-06
-4 4.674630e-03
-5 4.869071e-03
-6 5.059539e-03
-7 1.091079e-03
-8 7.950697e-04
-\end{CodeOutput}
-\begin{CodeInput}
-R> as.mitml.result(fit)
-\end{CodeInput}
-\begin{CodeOutput}
-[[1]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10495.423 10561.192 -5238.712 10477.423     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2843  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.37195                 -0.41014                  0.03052  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65802                 -1.04611                 -1.51245  
-                  pupil                       ct  
-                0.50405                  0.42496  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[2]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-     AIC      BIC   logLik deviance df.resid 
-10500.88 10566.65 -5241.44 10482.88    11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2917  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.37718                 -0.41511                  0.03067  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66935                 -1.05211                 -1.49429  
-                  pupil                       ct  
-                0.49013                  0.43835  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[3]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10505.026 10570.795 -5243.513 10487.026     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2908  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.32339                 -0.42359                  0.03023  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67142                 -1.05776                 -1.51038  
-                  pupil                       ct  
-                0.49756                  0.42474  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[4]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10519.511 10585.280 -5250.755 10501.511     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2961  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.33581                 -0.40871                  0.03039  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66477                 -1.05451                 -1.51858  
-                  pupil                       ct  
-                0.45928                  0.44419  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
-
-[[5]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10522.038 10587.807 -5252.019 10504.038     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2955  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.35187                 -0.40574                  0.03064  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67468                 -1.06551                 -1.52622  
-                  pupil                       ct  
-                0.47006                  0.44148  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-attr(,"class")
-[1] "mitml.result" "list"        
-\end{CodeOutput}
-\begin{CodeInput}
++   with(glmer(
++     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
++     family = "binomial")) 
+R> # tidy(pool(fit))
+R> # as.mitml.result(fit)
 R> # testEstimates(as.mitml.result(fit))
@@
 \end{CodeChunk}
+
+The estimated effects after imputation are presented in Table XYZ.
 
@@
 
-In this example, we demonstrate a multilevel imputation of an intercept
-and slope random model with a continuous response. We use the obesity
-dataset included in the \texttt{micemd}@ package, a synthetic dataset
-that emulates an electronic survey in which individuals are asked to
-provide information about their weight and consumption habits in
-different countries. To easy the explanations we simulate data for 5
-clusters and we reduce the dataset to the following variables:
+TODO: explain exclusion restriction.
+
+Data are simulated and included in the \texttt{micemd} package. We will
+use the following variables:
 
@@
   `time`` Response time in minutes (inclusion-restriction variable).
-\item
-  `FamOb`` Family obesity history (yes or not)
 \end{itemize}
 
-In this dataset, Age and FamOb are MAR, while the weight variable is
-affected by selection bias, attributed to an indirect MNAR mechanism.
-This MNAR mechanism typically arises when an unobserved or omitted
-variable influences both the value of the incomplete variable (in this
-case, Weight) and its likelihood of being missing (denoted as R).
-
-In the primary analysis model, BMI serves as the dependent variable,
-with Age, Gender, and FamOb as predictors. Researchers, based on
-observable data, are considering the inclusion of random intercept
-effects, as well as introducing a random slope for the Age variable.
-
-The main model is given by:
-\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
-We start by loading the data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> #data("data_heckman", package = "micemd")
-R> #dat <- data_heckman
-\end{CodeInput}
-\end{CodeChunk}
-
-Now, let's begin by examining the missing patterns in the data by
-cluster:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> require(patchwork)
-R>   myplots <- lapply(1:5, function(i) {
-+          ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
-+            ggtitle(paste0("Cluster", i))+ 
-+       theme(legend.position = ifelse(i==1,"bottom","none"))
-+       })
-R>   myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]
+The imputation of these data is based on the
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}
+
+Load the data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> data("data_heckman", package = "micemd")
+\end{CodeInput}
+\end{CodeChunk}
+
+Select relevant variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> dat <- data_heckman
+\end{CodeInput}
+\end{CodeChunk}
+
+Visualize missing data pattern:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(dat)
 \end{CodeInput}
@@
 \end{CodeChunk}
-
-We observe that the missing pattern is quite similar across the
-clusters. However, regarding the weight variable, we notice that is
-systematically missing in cluster 3. In order to evaluate if it is
-required a imputation method for 1-level or 2-level we assess the
-Intraclass Correlation Coefficient (ICC) for the outcome variable, we
-use the ``performance'' package:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-R> performance::icc(Nulmodel)
-\end{CodeInput}
-\begin{CodeOutput}
-# Intraclass Correlation Coefficient
-
-    Adjusted ICC: 0.342
-  Unadjusted ICC: 0.342
-\end{CodeOutput}
-\end{CodeChunk}
-
-Since the ICC is above 0.1 and as the main analyis will be consider a
-mixed model, we decide to use two-level (2l) imputation methods. In this
-imputation process, we include all predictor variables in the main model
-and since BMI is a composite variable. We also incorporate weight and
-height variables.
-
-To determine the best imputation method for these variables, we can use
-the \texttt{find.defaultMethod}function provided in the \texttt{micemd}
-package, which suggests an appropriate method for MAR variables based on
-the type of variable, number of observations in the cluster, and number
-of clusters. It suggests using ``2l.2stage.bin'' for the FAV variable
-and ``2l.2stage.norm'' for the age variable. However, after inspecting
-the age density plot, we consider modifying its method to
-``2l.2stage.pmm.'' For the BMI variable, we employ deterministic
-imputation.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
-R> meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
-R> meth_mar["Age"]<-"2l.2stage.pmm" 
-\end{CodeInput}
-\end{CodeChunk}
-
-For these imputation models, it is necessary to specify the prediction
-matrix, with the cluster variable labeled as -2 and the predictor
-variable labeled as 2, encompassing all variables. We need to suprime
-the variable Time as this variable is not especified in the main model
-and modify also in the prediction matrix the relationship between BMI,
-weight and height to avoid circular predictions. Then we proced to run
-the imputation model.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
-R> pred_mar[,"Cluster"]<- -2 # clustering variable
-R> pred_mar[,"Time"]<-0
-R> pred_mar[pred_mar==1]<-2
-R> pred_mar[c("Height", "Weight"), "BMI"] <- 0
-R> ggmice::plot_pred(pred_mar)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_pred-1} \end{center}
-
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(complete(imp_mar,"long")$Weight)
-\end{CodeInput}
-\begin{CodeOutput}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  2.409  69.703  82.925  82.476  95.310 158.420 
-\end{CodeOutput}
-\begin{CodeInput}
-R> plot(imp_mar)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-1} \end{center}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-2} \end{center}
-
-\end{CodeChunk}
-
-We are also contemplating the utilization of the ``pmm'' option, given
-that the imputed values' weights are lower than those of the observable
-values.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(complete(imp_mar_pmm,"long")$Weight)
-\end{CodeInput}
-\begin{CodeOutput}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  27.50   71.02   83.82   83.76   95.55  135.98 
-\end{CodeOutput}
-\begin{CodeInput}
-R> plot(imp_mar_pmm)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-1} \end{center}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-2} \end{center}
-
-\end{CodeChunk}
-
-After confirming convergence, we proceed to save the results for future
-use. Users are considering the possibility that the weight variable may
-not have been selected randomly. It's likely that an omitted variable,
-like self-esteem, could influence this selection. For instance,
-individuals with lower self-esteem might have higher weight values,
-impacting their willingness to provide honest information due to
-embarrassment.
-
-To address this situation, two approaches have been proposed for dealing
-with Missing Not at Random (MNAR) data: pattern-mixed models and
-selection models. Within pattern-mixed models, methods like the delta
-method and more advanced ones like NARFS have been suggested. In
-contrast, the selection model approach includes methods such as the
-Heckman model, which can be particularly useful in this case. Several
-methods, including Galimar2017, Hammon2021, and the recent Munoz method,
-designed for two-level data, allowing for variations in intercepts and
-exposure effects (random intercept and slope).
-
-To apply the heckman method, the weight variable should be specified as
-``2l.2stage.heckman'' found in the micemd package. Additionally, the
-prediction matrix needs modification because this method involves
-specifying two equations: one for the outcome, describing the incomplete
-variable in terms of partially observed predictors (in this case, all
-variables from the main model), and the other for the selection model,
-explaining the probability of being observed based (R) on certain
-variables.
-
-Here for the outcome equation we consider is the same imputation model
-that we used for the MAR case (main model).
-\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
-Regarding the selection equation, we include the same predictors as
-those in the primary model, as well as a time variable. Here the time
-variable serves as a restriction exclusion variable specifically
-explaining the probability of being observed but not affecting the
-incomplete value (Weight). In this context, we assume that the time a
-user spends completing the survey serves as a proxy for the barriers
-they may encounter in survey completion, such as familiarity with the
-survey content or internet speed. These factors may lead the user to
-skip specific questions or even the entire survey. Also we assume the
-time is not have any influence on the weight assigned to the subject.
-\[R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
-These two equations are jointly estimated under the assumption that the
-error terms are interconnected with a bivariate normal distribution. For
-a more comprehensive understanding of the model and the exclusion
-restriction, you can refer to (Munoz).
-
-To integrate information from both equations, we must adjust the
-prediction matrix. The cluster variable remains specified as before
-(-2). In this imputation method, all the variables present in both the
-selection and outcome equations are included with a random effect.
-However, it's essential to distinguish which of these variables appear
-in each equation.
-
-In this framework, when a variable is shared between both equations, it
-is denoted as (2). Predictors exclusive to the outcome equation are
-indicated as (-4), while those exclusive to the selection equation are
-labeled as (-3). Consequently, the only alteration needed in the
-predictor matrix pertains to the variable ``Time.''
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> pred_mnar <- pred_mar
-R> pred_mnar["Weight","Time"]<- -3
-R> ggmice::plot_pred(pred_mnar)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_pred-1} \end{center}
-
-\end{CodeChunk}
-
-We also need to modify the method of the weight variable
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> meth_mnar <- meth_mar
-R> meth_mnar["Weight"]<- "3l.2stage.heckman"
-\end{CodeInput}
-\end{CodeChunk}
-
-Then we proceed to run the imputation model as before Impute the
-missingness:
-
-After executing these imputation procedures, it is essential to assess
-convergence and the coherence of the imputed values. Upon examining the
-weight variable, we noticed that the imputed range falls outside the
-realm of plausible values (as weight should be positive). Consequently,
-we are contemplating a different approach, specifically, the use of
-``pmm,'' which involves imputing values from donor observations. This
-approach ensures that the imputed values remain within the range of
-observable values.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(complete(imp_mnar,"long")$Weight)
-\end{CodeInput}
-\begin{CodeOutput}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  10.26   73.41   86.94   88.27  101.73  186.47 
-\end{CodeOutput}
-\begin{CodeInput}
-R> plot(imp_mnar)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-1} \end{center}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-2} \end{center}
-
-\end{CodeChunk}
-
-We then run the imputation model but this time using the option of pmm,
-to assure that weight values are in the range of the observable data,
-this can be implemented by seeting the pmm parameter as true.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(complete(imp_mnar_pmm,"long"))
-\end{CodeInput}
-\begin{CodeOutput}
-      .imp           .id          Cluster           Time          Gender     
- Min.   : 1.0   Min.   :   1   Min.   :1.000   Min.   :1.016   Female:10720  
- 1st Qu.: 3.0   1st Qu.: 528   1st Qu.:2.000   1st Qu.:3.724   Male  :10390  
- Median : 5.5   Median :1056   Median :3.000   Median :5.151                 
- Mean   : 5.5   Mean   :1056   Mean   :2.868   Mean   :5.211                 
- 3rd Qu.: 8.0   3rd Qu.:1584   3rd Qu.:4.000   3rd Qu.:6.698                 
- Max.   :10.0   Max.   :2111   Max.   :5.000   Max.   :9.957                 
-     Height      FamOb            Age            Weight            BMI       
- Min.   :1.450   no : 3794   Min.   :14.00   Min.   : 27.50   Min.   : 9.59  
- 1st Qu.:1.642   yes:17316   1st Qu.:20.00   1st Qu.: 75.19   1st Qu.:25.84  
- Median :1.701               Median :25.00   Median : 88.88   Median :30.57  
- Mean   :1.702               Mean   :24.82   Mean   : 89.61   Mean   :31.02  
- 3rd Qu.:1.760               3rd Qu.:29.00   3rd Qu.:103.78   3rd Qu.:35.89  
- Max.   :1.961               Max.   :47.00   Max.   :135.98   Max.   :63.21  
-\end{CodeOutput}
-\begin{CodeInput}
-R> plot(imp_mnar_pmm)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-1} \end{center}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-2} \end{center}
-
-\end{CodeChunk}
-
-Then after this modification we proceed to compare the effects on the
-model, therefore we run the analysis model on each of the completed
-datasets along with the dataset where the incomplete values are removed
-(CC).
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library(ggplot2)
-R> cc_rs<- with(Obesity[complete.cases(Obesity),],lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-R> mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-R> mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-R> mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-R> mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-R> list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
-R> plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/models-1} \end{center}
-
-\end{CodeChunk}
-
-We note that there is minimal disparity in the age effect, FamObs,
-across the various imputation models under consideration. Nonetheless,
-with respect to Gender, we observe a notable distinction: in the M(C)AR
-assumption, the effect is significant, whereas under the MNAR
-assumption, it becomes insignificant. An analysis of the intercept
-reveals that, under the MNAR assumption, a higher average BMI is
-anticipated compared to the MAR assumption.
-
-Visualize missing data pattern:
 
@@
 selection model.
+
+Create predictor matrix:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred <- quickpred(dat)   # predictor matrix
+R> pred[,"Cluster"]<- -2 # clustering variable
+R> pred[, "Weight"] <- -3 #inclusion-restriction variable
+R> plot_pred(pred)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-pred-1} \end{center}
+
+\end{CodeChunk}
+
+Set the Heckman model as imputation method:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth <- make.method(dat) # methods vector
+R> meth["weight"]<-"2l.heckman"
+\end{CodeInput}
+\end{CodeChunk}
+
+Impute the missingness:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp <- mice(
++   dat, # dataset with missing values
++   m = 5, # number of imputations
++   maxit = 10,
++   meth = meth, #imputation method vector
++   pred = pred, #imputation predictors matrix
++   meta_method = "reml",
++   printFlag = FALSE
++ )
+\end{CodeInput}
+\end{CodeChunk}
 

### 2023-11-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
update case III

@@
 
-```{r, include=FALSE, eval=FALSE}
+```{r, include=FALSE, eval=TRUE}
 library(here)
@@
 
-```{r, dpi = 300,out.width="70%", fig.cap="Missing pattern", eval=FALSE}
+```{r, dpi = 300,out.width="70%", fig.cap="Missing pattern", eval=TRUE}
+
+#ggmice::plot_pattern(data=Obesity,cluster="Cluster") does not work!! bug
+#ggmice::plot_pattern(data=Obesity,cluster=Obesity$Cluster) does not work!! bug
 library(ggpubr)
@@
 
-```{r obesity-icc, eval=FALSE}
+```{r obesity-icc, eval=TRUE}
 Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-#Performance::ICC(Nulmodel)
+performance::icc(Nulmodel)
 ```
@@
 
-```{r obesitymeth, eval=FALSE}
-meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+```{r obesitymeth, eval=TRUE}
+library(micemd)
+meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
                                ni.small = 100, prop.small = 0.4)
@@
 Then we proceed to run the imputation model.
-```{r obesity-predmar,out.width="70%", eval=FALSE}
-pred_mar <- quickpred(Obesity)   # predictor matrix
+```{r obesity-predmar,out.width="70%", eval=TRUE}
+pred_mar <- mice(Obesity, maxit = 0)$pred 
 pred_mar[,"Cluster"] <- -2 # clustering variable
@@
 ggmice::plot_pred(pred_mar)
-imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+imp_mar <- mice::mice(data = Obesity, meth = meth_mar, pred = pred_mar,
                 m=10, seed = 123, printFlag = FALSE)
@@
 
-```{r obesity-predmar1,eval=FALSE}
+```{r obesity-predmar1,eval=TRUE}
 summary(complete(imp_mar,"long")$Weight)
@@
 
-```{r obesity-predmar_pmm, eval=FALSE}
+```{r obesity-predmar_pmm, eval=TRUE}
 meth_mar["Weight"]<-"2l.2stage.pmm" 
@@
 
-```{r obesity-predmar_pmm1,eval=FALSE}
+```{r obesity-predmar_pmm1,eval=TRUE}
 summary(complete(imp_mar_pmm,"long")$Weight)
-#ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+ggmice::plot_trace(imp_mar_pmm, "Weight")
 ```
@@
 
-```{r obesity-predmnar, eval=FALSE}
+```{r obesity-predmnar, eval=TRUE}
 pred_mnar <- pred_mar
 pred_mnar["Weight","Time"]<- -3
-#ggmice::plot_pred(pred_mnar)
+ggmice::plot_pred(pred_mnar)
 ```
@@
 
-```{r obesity-predmnar1, eval=FALSE}
+```{r obesity-predmnar1, eval=TRUE}
 meth_mnar <- meth_mar
@@
 Then we proceed to run the imputation model as before, after executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
-```{r obesity-predmnar2,eval=FALSE}
+```{r obesity-predmnar2,eval=TRUE}
 imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
@@
 
-```{r obesity-predmnar3,eval=FALSE}
+```{r obesity-predmnar3,eval=TRUE}
 summary(complete(imp_mnar,"long")$Weight)
@@
 
-```{r obesity-predmnarp, eval=FALSE}
+```{r obesity-predmnarp, eval=TRUE}
 imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
@@
 We check the convergency of the results
-```{r obesity-predmnarp1, eval=FALSE}
+```{r obesity-predmnarp1, eval=TRUE}
 summary(complete(imp_mnar_pmm,"long")$Weight)
-#ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+ggmice::plot_trace(imp_mar_pmm, "Weight")
 ```
@@
 
-```{r models, eval=FALSE}
+```{r models, eval=TRUE}
 library(ggplot2)
@@
 
-Data are simulated and included in the `micemd` package. We will use the following variables:
-
-- `region` Cluster variable,
-- `gender` Gender (0=male, 1=female),
-- `age` Age of the patient,
-- `height` Height in meters,
-- `weight` Weight in kilograms,
-- `time`` Response time in minutes (inclusion-restriction variable).
-
-The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
-
-<!-- [TODO: add reference Arxiv paper Johanna].  -->
-
-Load the data:
-```{r obesity}
-data("data_heckman", package = "micemd")
-```
-
-Load the data:
-```{r}
-#data("obesity", package = "miceheckman")
-```
-
-Select relevant variables:
-```{r bmi}
-#dat <- select(obesity, -"bmi")
-```
-
-Create predictor matrix:
-```{r}
-#pred <- quickpred(dat)   # predictor matrix
-#pred["weight","cluster"]<- -2 # clustering variable
-#pred["weight","rt"] <- -3 #inclusion-restriction variable
-#plot_pred(pred)
-```
-
-Set the Heckman model as imputation method:
-```{r}
-#meth <- make.method(dat) # methods vector
-#meth["weight"]<-"2l.heckman"
-```
-
-Visualize missing data pattern:
-```{r obesity-md}
-plot_pattern(dat)
-```
-
-The matrix only shows the predictors for the main model, not the selection model.
-
-Create predictor matrix:
-```{r obesity-pred, eval = FALSE}
-pred <- quickpred(dat)   # predictor matrix
-pred[,"Cluster"] <- -2 # clustering variable
-pred[, "Weight"] <- -3 #inclusion-restriction variable
-plot_pred(pred)
-```
-
-Set the Heckman model as imputation method:
-```{r, eval = FALSE}
-meth <- make.method(dat) # methods vector
-meth["weight"]<-"2l.heckman"
-```
-
-Impute the missingness:
-```{r eval=FALSE, cache=TRUE}
-imp <- mice(
-  dat, # dataset with missing values
-  m = 5, # number of imputations
-  maxit = 10,
-  meth = meth, #imputation method vector
-  pred = pred, #imputation predictors matrix
-  meta_method = "reml",
-  printFlag = FALSE
-)
-```
-
-# Discussion
-
 ORDER:
@@
 (Zhang,2023).
+
+# Conclusion
 

### 2023-11-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
update case III

--- 2023-11-08 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 Hanne I. Oberman\\Methodology and Statistics\\
-Utrecht University \And Johanna Munoz\\Julius Center for Health Sciences
+Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
 and Primary Care,\\
@@
 }
-\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}
-
-\Plainauthor{Hanne I. Oberman, Johanna Munoz, Thomas P. A. Debray, Gerko
+\title{Imputation of Incomplete Multilevel Data with \proglang{R}}
+
+\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
 Vink, Valentijn M. T. de Jong}
-\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
+\Plaintitle{Imputation of Incomplete Multilevel Data with R}
 \Shorttitle{Multilevel \pkg{mice}}
@@
 \Abstract{
-Incomplete multilevel data requires careful consideration of the missing
-data problem and analysis strategy. In this tutorial, we focus on a
-popular strategy for accommodating missingness in multilevel data:
-replacing the missing data with one or more plausible values, i.e.,
+This tutorial illustrates the imputation of incomplete multilevel data
+with the \proglang{R} packackage \pkg{mice}. Our scope is only simple
+multilevel models, to show how imputation can yield less biased
+estimates from incomplete clustered data. More complex models can be
+accomodated, but are outside the scope of this paper. Incomplete
+multilevel data requires careful consideration of the missing data
+problem and analysis strategy. In this tutorial, we focus on a popular
+strategy for accommodating missingness in multilevel data: replacing the
+missing data with one or more plausible values, i.e.,
 imputation.Imputation separates the missing data problem from the main
@@
 
-\newpage
-
 \hypertarget{introduction}{%
 \section{Introduction}\label{introduction}}
+
+\hypertarget{multilevel-data}{%
+\subsection{Multilevel data}\label{multilevel-data}}
 
@@
 modeling.
+
+Table 1: Concepts in multilevel methods
 
@@
 
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:mechanism}Concepts in missing data methods}
-\centering
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
-\toprule
-Concept & Details\\
-\midrule
-MCAR & Missing Completely At Random, where the probability to be missing is equal across all data entries.\\
-MAR & Missing At Random, where the probability to be missing depends on observed information\\
-MNAR & Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
+Table 2: Concepts in missing data methods
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Concept}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Details}
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+MCAR & Missing Completely At Random, where the probability to be missing
+is equal \\
+& across all data entries \\
+MAR & Missing At Random, where the probability to be missing depends on
+observed \\
+& information \\
+MNAR & Missing Not At Random (MNAR), where the probability to be
+missing \\
+& depends on unrecorded information, making the missingness
+non-ignorable \\
+& \citep{rubi76, meng94}. \\
+& \\
+\end{longtable}
 
@@
 
-We'll now use \texttt{school} as a predictor to impute all other
+We will now use \texttt{school} as a predictor to impute all other
 variables. This is still not recommended practice, since it only works
@@
 R> fit <- imp %>% 
-+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
-R> tidy(pool(fit))
-\end{CodeInput}
-\begin{CodeOutput}
-                     term    estimate   std.error  statistic      p.value
-1             (Intercept) -2.35203726 0.340181747  -6.914061 4.994037e-12
-2                    type -0.41265892 0.180274846  -2.289054 2.209524e-02
-3                     age  0.03049023 0.001570162  19.418521 1.238416e-81
-4 as.factor(motor_score)2 -0.66764920 0.068737865  -9.712975 3.480413e-22
-5 as.factor(motor_score)3 -1.05520001 0.070218940 -15.027285 2.540225e-50
-6 as.factor(motor_score)4 -1.51238349 0.072304262 -20.916934 1.850073e-90
-7                   pupil  0.48421447 0.038982800  12.421234 6.772069e-17
-8                      ct  0.43474621 0.029968474  14.506785 2.342253e-36
-             b          df dfcom         fmi      lambda m         riv
-1 5.281599e-04 10119.70041 11013 0.005673265 0.005476771 5 0.005506932
-2 4.881699e-05 10893.89378 11013 0.001985736 0.001802528 5 0.001805783
-3 3.335358e-08  6320.89582 11013 0.016545467 0.016234340 5 0.016502243
-4 4.188703e-05  8327.24771 11013 0.010875748 0.010638213 5 0.010752602
-5 5.135721e-05  7632.20045 11013 0.012757637 0.012498967 5 0.012657168
-6 1.403058e-04  2831.75462 11013 0.032888241 0.032205434 5 0.033277139
-7 3.571497e-04    49.97295 11013 0.309130930 0.282023647 5 0.392803531
-8 8.586645e-05   294.69765 11013 0.120677044 0.114729598 5 0.129598366
-          ubar
-1 1.150898e-01
-2 3.244044e-02
-3 2.425385e-06
-4 4.674630e-03
-5 4.869071e-03
-6 5.059539e-03
-7 1.091079e-03
-8 7.950697e-04
-\end{CodeOutput}
-\begin{CodeInput}
-R> as.mitml.result(fit)
-\end{CodeInput}
-\begin{CodeOutput}
-[[1]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10495.423 10561.192 -5238.712 10477.423     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2843  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.37195                 -0.41014                  0.03052  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.65802                 -1.04611                 -1.51245  
-                  pupil                       ct  
-                0.50405                  0.42496  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[2]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-     AIC      BIC   logLik deviance df.resid 
-10500.88 10566.65 -5241.44 10482.88    11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2917  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.37718                 -0.41511                  0.03067  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66935                 -1.05211                 -1.49429  
-                  pupil                       ct  
-                0.49013                  0.43835  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[3]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10505.026 10570.795 -5243.513 10487.026     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2908  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.32339                 -0.42359                  0.03023  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67142                 -1.05776                 -1.51038  
-                  pupil                       ct  
-                0.49756                  0.42474  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-[[4]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10519.511 10585.280 -5250.755 10501.511     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2961  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.33581                 -0.40871                  0.03039  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.66477                 -1.05451                 -1.51858  
-                  pupil                       ct  
-                0.45928                  0.44419  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings
-
-[[5]]
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
-    name)
-      AIC       BIC    logLik  deviance  df.resid 
-10522.038 10587.807 -5252.019 10504.038     11013 
-Random effects:
- Groups Name        Std.Dev.
- name   (Intercept) 0.2955  
-Number of obs: 11022, groups:  name, 15
-Fixed Effects:
-            (Intercept)                     type                      age  
-               -2.35187                 -0.40574                  0.03064  
-as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
-               -0.67468                 -1.06551                 -1.52622  
-                  pupil                       ct  
-                0.47006                  0.44148  
-optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
-
-attr(,"class")
-[1] "mitml.result" "list"        
-\end{CodeOutput}
-\begin{CodeInput}
++   with(glmer(
++     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
++     family = "binomial"
++     )) 
+R> # tidy(pool(fit))
+R> # as.mitml.result(fit)
 R> # testEstimates(as.mitml.result(fit))
@@
 \begin{CodeInput}
+R> #ggmice::plot_pattern(data=Obesity,cluster="Cluster") does not work!! bug
+R> #ggmice::plot_pattern(data=Obesity,cluster=Obesity$Cluster) does not work!! bug
 R> library(ggpubr)
@@
 
-{\centering \includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1}
+{\centering \includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-12-1}
 
@@
 
-\caption[Missing pattern]{Missing pattern}\label{fig:obesity-md}
+\caption[Missing pattern]{Missing pattern}\label{fig:unnamed-chunk-12}
 \end{figure}
@@
 R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-R> #Performance::ICC(Nulmodel)
-\end{CodeInput}
+R> performance::icc(Nulmodel)
+\end{CodeInput}
+\begin{CodeOutput}
+# Intraclass Correlation Coefficient
+
+    Adjusted ICC: 0.362
+  Unadjusted ICC: 0.362
+\end{CodeOutput}
 \end{CodeChunk}
@@
 \begin{CodeInput}
-R> meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+R> library(micemd)
+R> meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
 +                                ni.small = 100, prop.small = 0.4)
@@
 \begin{CodeInput}
-R> pred_mar <- quickpred(Obesity)   # predictor matrix
+R> pred_mar <- mice(Obesity, maxit = 0)$pred 
 R> pred_mar[,"Cluster"] <- -2 # clustering variable
@@
 \begin{CodeInput}
-R> imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+R> imp_mar <- mice::mice(data = Obesity, meth = meth_mar, pred = pred_mar,
 +                 m=10, seed = 123, printFlag = FALSE)
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
- -9.826  69.500  82.789  82.337  94.919 148.712 
+ -19.48   68.45   81.65   81.28   94.12  160.76 
 \end{CodeOutput}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  28.35   67.80   80.53   80.90   92.94  134.61 
+  28.35   69.25   82.43   82.41   94.76  134.61 
 \end{CodeOutput}
 \begin{CodeInput}
-R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
-\end{CodeInput}
+R> ggmice::plot_trace(imp_mar_pmm, "Weight")
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmar_pmm1-1} \end{center}
+
 \end{CodeChunk}
@@
 R> pred_mnar["Weight","Time"]<- -3
-R> #ggmice::plot_pred(pred_mnar)
-\end{CodeInput}
+R> ggmice::plot_pred(pred_mnar)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmnar-1} \end{center}
+
 \end{CodeChunk}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
- -7.334  69.553  83.722  83.086  96.738 183.338 
+  14.09   71.68   85.33   85.79   99.31  186.67 
 \end{CodeOutput}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  28.35   71.68   85.07   85.07   98.21  134.61 
+  28.35   71.51   85.09   85.64   99.00  134.61 
 \end{CodeOutput}
 \begin{CodeInput}
-R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
-\end{CodeInput}
+R> ggmice::plot_trace(imp_mar_pmm, "Weight")
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmnarp1-1} \end{center}
+
 \end{CodeChunk}
@@
 R> mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-R> 
 R> list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+TODO: explain exclusion restriction.
 
@@
 
+ORDER:
+
+\begin{itemize}
+\item
+  summary
+\item
+  congeniality, then in hierarchical models
+\item
+  look whether we can fit cong. back in the main body
+\item
+  alt. methods
+\item
+  conclusion: mice is really easy!
+\item
+  Additional levels of clustering
+\item
+  More complex data types: timeseries and polynomial relationship in the
+  clustering.
+\item
+  FIML vs MI
+\end{itemize}
+
+An alternative approach to missing data is to use Full Information
+Maximum Likelihood (FIML). This method does not require the imputation
+of any missing values. Whereas MI consists of imputation, analyses and
+pooling steps, FIML analyses the data in a single step. When the
+assumptions are met the two approaches should produce equivalent
+results. {[}REF{]} As FIML requires specialised software, not all
+analyses can be performed with standard software. {[}REF{]}
+
+\begin{itemize}
+\tightlist
+\item
+  Survival / TTE, this could be put in the paragraph on congeniality
+\end{itemize}
+
+When the outcome is time-to-event, the Nelson-Aalen estimate of the time
+to event should be included as a covariate in the imputation model
+{[}REF{]}
+
+In hierarchical datasets, clustering is a concern because the
+homoscedasticity in the error terms cannot be assumed across clusters
+and the relationship among variables may vary at different hierarchical
+levels. When multiple imputation is used to deal with missing data, as
+the imputation and analysis process is performed separately, it is
+necessary that imputation model being congenial with the main analysis
+model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
+structure also imputation model should do it (Audigier, 2021). Not
+including clustering into the imputation process may lead to effect
+estimates with smaller standard errors and inflated type I error.
+
+There are different strategies that can be adopted in the imputation
+process that account for clustering: inclusion of cluster indicator
+variable, performing a separate imputation process for each cluster, or
+performing a simultaneous imputation process by using an imputation
+method that accounts for clustering.(Stata:
+\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
+TODO: replace ref.
+
+The selection of each strategy depends mainly on the assumptions in the
+main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of
+cluster indicator variables is restricted in datasets where there are
+not many clusters and many observations per cluster (Graham, 2009). The
+last restriction is also required when imputations are performed on each
+cluster separately. When this restriction cannot be achieved, one can
+use an imputation model that simultaneously imputes all clusters using a
+hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters
+are correlated and this correlation is modeled by a random effect so the
+hierarchical model can be estimated even when there are few observations
+per cluster. However, this strategy is best suited for balanced data
+(Grund, 2017) and when random effects model is appropriated, i.e.~the
+number of clusters is adequate. (Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main
+model, for instance by using the cluster indicator strategy may lead to
+bias estimates when the model is based on a hierarchical model
+(Taaljard,2008). Even when an imputation strategy congenial with the
+main model is preferred, it is important to consider whether it is
+appropriate for the data as a less complex imputation strategies may
+also lead to unbiased estimates in certain scenarios(Bailey 2020). For
+instance, in causal effect analysis, separately imputation may lead to
+smaller bias when the size of the smaller exposure cluster is large,
+compared with an imputation model that includes exposure-confounder
+interactions. (Zhang,2023).
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
 \hypertarget{funding}{%
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5417}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4583}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
 \toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Formula} \pkg{lme4}
+\textbf{Formula \pkg{lme4}}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright

### 2023-11-14 heckman/ipdfunction.R

@@ -21,7 +21,7 @@ copulaIPD <- function(data, sel, out, family, send) {
   fit <- try(GJRM::gjrm( formula = list(sel, out),
                          data = data,
                          margins = c("probit", ifelse(family=="binomial","probit","N")),
-                         Model = "BSS",
+                         model = "BSS",
                          gamlssfit = TRUE,
                          extra.regI = "sED",
                          parscale = TRUE),

### 2023-11-21 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
move to quarto

--- 2023-11-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 
-Many datasets include individuals that are clustered together, for example in geographic regions, or even different studies.
-In the simplest case, individuals (e.g., students) are nested within a single cluster (e.g., school classes).
-More complex clustered structures may occur when there are multiple hierarchical levels (e.g., students in different schools or patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases).
+Understanding and effectively handling missing data within multilevel structures are crucial for preserving the integrity of analyses and deriving accurate inferences.
+
+Many datasets include individuals that are clustered together, for example by geographic region.
+In the simplest case, each individual is a unit (e.g., a student) which is nested within a single cluster (e.g., a school class).
+More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases).
+The hierarchical structures of such clustered data may be captured using multilevel analysis techniques.
+
 With clustered data we generally assume that individuals from the same cluster tend to be more similar than individuals from other clusters.
@@
 Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.
+
+In the context of multilevel data, missingness can occur at different levels, complicating the interpretation and reliability of results.
 

### 2023-11-21 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
move to quarto

@@
 \usepackage{ulem}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
@@
 \begin{CodeInput}
-R> #ggmice::plot_pattern(data=Obesity,cluster="Cluster") does not work!! bug
-R> #ggmice::plot_pattern(data=Obesity,cluster=Obesity$Cluster) does not work!! bug
 R> library(ggpubr)
@@
 \end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-12-1}
-
-}
-
-\caption[Missing pattern]{Missing pattern}\label{fig:unnamed-chunk-12}
-\end{figure}
 \end{CodeChunk}
@@
 R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-R> performance::icc(Nulmodel)
-\end{CodeInput}
-\begin{CodeOutput}
-# Intraclass Correlation Coefficient
-
-    Adjusted ICC: 0.362
-  Unadjusted ICC: 0.362
-\end{CodeOutput}
+R> #Performance::ICC(Nulmodel)
+\end{CodeInput}
 \end{CodeChunk}
@@
 \begin{CodeInput}
-R> library(micemd)
-R> meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+R> meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
 +                                ni.small = 100, prop.small = 0.4)
@@
 \begin{CodeInput}
-R> pred_mar <- mice(Obesity, maxit = 0)$pred 
+R> pred_mar <- quickpred(Obesity)   # predictor matrix
 R> pred_mar[,"Cluster"] <- -2 # clustering variable
@@
 R> ggmice::plot_pred(pred_mar)
-\end{CodeInput}
-
-\begin{center}\includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmar-1} \end{center}
-
-\begin{CodeInput}
-R> imp_mar <- mice::mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+R> imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
 +                 m=10, seed = 123, printFlag = FALSE)
@@
 \end{CodeInput}
-\begin{CodeOutput}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
- -19.48   68.45   81.65   81.28   94.12  160.76 
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> summary(complete(imp_mar_pmm,"long")$Weight)
-\end{CodeInput}
-\begin{CodeOutput}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  28.35   69.25   82.43   82.41   94.76  134.61 
-\end{CodeOutput}
-\begin{CodeInput}
-R> ggmice::plot_trace(imp_mar_pmm, "Weight")
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmar_pmm1-1} \end{center}
-
+R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> pred_mnar["Weight","Time"]<- -3
-R> ggmice::plot_pred(pred_mnar)
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmnar-1} \end{center}
-
+R> #ggmice::plot_pred(pred_mnar)
+\end{CodeInput}
 \end{CodeChunk}
@@
 \end{CodeInput}
-\begin{CodeOutput}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  14.09   71.68   85.33   85.79   99.31  186.67 
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> summary(complete(imp_mnar_pmm,"long")$Weight)
-\end{CodeInput}
-\begin{CodeOutput}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  28.35   71.51   85.09   85.64   99.00  134.61 
-\end{CodeOutput}
-\begin{CodeInput}
-R> ggmice::plot_trace(imp_mar_pmm, "Weight")
-\end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmnarp1-1} \end{center}
-
+R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
+\end{CodeInput}
 \end{CodeChunk}
@@
 \end{CodeInput}
-
-\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/models-1} \end{center}
-
 \end{CodeChunk}
@@
 TODO: explain exclusion restriction.
+
+Data are simulated and included in the \texttt{micemd} package. We will
+use the following variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{region} Cluster variable,
+\item
+  \texttt{gender} Gender (0=male, 1=female),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{height} Height in meters,
+\item
+  \texttt{weight} Weight in kilograms,
+\item
+  `time`` Response time in minutes (inclusion-restriction variable).
+\end{itemize}
+
+The imputation of these data is based on the
+\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
+Heckman Github repo}
+
+Load the data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> data("data_heckman", package = "micemd")
+\end{CodeInput}
+\end{CodeChunk}
+
+Load the data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> #data("obesity", package = "miceheckman")
+\end{CodeInput}
+\end{CodeChunk}
+
+Select relevant variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> #dat <- select(obesity, -"bmi")
+\end{CodeInput}
+\end{CodeChunk}
+
+Create predictor matrix:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> #pred <- quickpred(dat)   # predictor matrix
+R> #pred["weight","cluster"]<- -2 # clustering variable
+R> #pred["weight","rt"] <- -3 #inclusion-restriction variable
+R> #plot_pred(pred)
+\end{CodeInput}
+\end{CodeChunk}
+
+Set the Heckman model as imputation method:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> #meth <- make.method(dat) # methods vector
+R> #meth["weight"]<-"2l.heckman"
+\end{CodeInput}
+\end{CodeChunk}
+
+Visualize missing data pattern:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot_pattern(dat)
+\end{CodeInput}
+
+\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1} \end{center}
+
+\end{CodeChunk}
+
+The matrix only shows the predictors for the main model, not the
+selection model.
+
+Create predictor matrix:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pred <- quickpred(dat)   # predictor matrix
+R> pred[,"Cluster"] <- -2 # clustering variable
+R> pred[, "Weight"] <- -3 #inclusion-restriction variable
+R> plot_pred(pred)
+\end{CodeInput}
+\end{CodeChunk}
+
+Set the Heckman model as imputation method:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> meth <- make.method(dat) # methods vector
+R> meth["weight"]<-"2l.heckman"
+\end{CodeInput}
+\end{CodeChunk}
+
+Impute the missingness:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> imp <- mice(
++   dat, # dataset with missing values
++   m = 5, # number of imputations
++   maxit = 10,
++   meth = meth, #imputation method vector
++   pred = pred, #imputation predictors matrix
++   meta_method = "reml",
++   printFlag = FALSE
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 interactions. (Zhang,2023).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
 

### 2023-11-21 docs/manuscript.qmd
move to quarto

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: Achim.Zeileis@R-project.org
    url: https://www.zeileis.org/
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed. This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.

keywords: [missing data, multilevel, clustering, mice, R]
keywords-formatted: [missing data, multilevel, clustering, "[mice]{.pkg}", "[R]{.proglang}"]

bibliography: bibliography.bib  
---

## Introduction: Clustering and incomplete data {#sec-intro}

1. missing data occur often in data with human subjects
2. missing data may be resolved, but need to be handled in accordance with the analysis of scientific interest
3. in human-subjects research, there is often clustering, which may be captured with multilevel modeling techniques
4. if the analysis of scientific interest is a multilevel model, the missing data handling method should accommodate the multilevel structure of the data
4. both missingness and multilevel structures require advanced statistical techniques
5. this tutorial sets out to facilitate empirical researchers in accommodating both multilevel structures as well as missing data.
6. we illustrate the use of the software by means of three case studies from the social and biomedical sciences.

### overview of software

The popular [mice]{.pkg} package in [R]{.proglang} @R...

### scope

## Background {#sec-models}

### concepts in multilevel data

### concepts in missing data

::: callout
Box 1. The intraclass correlation coefficient.
:::

The [R]{.proglang} package [mice]{.pkg} provides a framework for imputing incomplete data on a variable-by-variable basis. The [mice]{.fct} function allows users to flexibly specify how many times and under what model the missing data should be imputed. This is reflected in the first four function arguments

```r
mice(data, m, method, predictorMatrix, ...)
```

where `data` refers to the incomplete dataset, `m` determines the number of imputations, `method` denotes the functional form of the imputation model and `predictorMatrix` specifies the interrelational dependencies between variables and imputation models (i.e., the set of predictors to be used for imputing each incomplete variable).

## Illustrations {#sec-illustrations}

In this section, we demonstrate the workflow using three case studies.

## Summary and discussion {#sec-summary}

What is missing from this manuscript...

## Computational details {.unnumbered}

The results in this paper were obtained using [R]{.proglang}~4.3.0. [R]{.proglang} itself and all packages used are available from the Comprehensive [R]{.proglang} Archive Network (CRAN) at [https://CRAN.R-project.org/].

## Acknowledgments {.unnumbered}

This project has received funding from the European Union's Horizon 2020 research and innovation programme under ReCoDID grant agreement No 825746.

## References {.unnumbered}

:::{#refs}

:::

{{< pagebreak >}}

## More technical details {#sec-techdetails .unnumbered}

:::{.callout}

Appendices can be included after the bibliography (with a page break). Each
section within the appendix should have a proper section title (rather than
just _Appendix_).

For more technical style details, please check out JSS's style FAQ at
[https://www.jstatsoft.org/pages/view/style#frequently-asked-questions]
which includes the following topics:

- Title vs. sentence case.
- Graphics formatting.
- Naming conventions.
- Turning JSS manuscripts into [R]{.proglang} package vignettes.
- Trouble shooting.
- Many other potentially helpful details...

:::

## Using BibTeX {#sec-bibtex .unnumbered}

:::{.callout}

References need to be provided in a {{< bibtex >}} file (`.bib`). All
references should be made with `@cite` syntax. This commands yield different
formats of author-year citations and allow to include additional details (e.g.,pages, chapters, \dots) in brackets. In case you are not familiar with these
commands see the JSS style FAQ for details.

Cleaning up {{< bibtex >}} files is a somewhat tedious task -- especially
when acquiring the entries automatically from mixed online sources. However,
it is important that informations are complete and presented in a consistent
style to avoid confusions. JSS requires the following format.

- item JSS-specific markup (`\proglang`, `\pkg`, `\code`) should be used in the references.
- item Titles should be in title case.
- item Journal titles should not be abbreviated and in title case.
- item DOIs should be included where available.
- item Software should be properly cited as well. For [R]{.proglang} packages `citation("pkgname")` typically provides a good starting point.

:::

### 2023-11-21 docs/manuscript.tex
move to quarto

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  article]{jss}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{orcidlink,thumbpdf,lmodern}

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Imputation of Incomplete Multilevel Data with R},
  pdfauthor={Hanne I. Oberman; Johanna Muñoz; Valentijn M.T. de Jong; Gerko Vink; Thomas P.A. Debray},
  pdfkeywords={missing data, multilevel, clustering, mice, R},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

%% -- Article metainformation (author, title, ...) -----------------------------

%% Author information
\author{Hanne I. Oberman~\orcidlink{0000-0003-3276-2141}\\Utrecht
University \And Johanna
Muñoz~\orcidlink{0000-0002-2384-5415}\\University Medical Center
Utrecht \AND Valentijn M.T. de
Jong~\orcidlink{0000-0001-9921-3468}\\University Medical Center
Utrecht \And Gerko Vink~\orcidlink{0000-0001-9767-1924}\\University
Medical Center Utrecht \AND Thomas P.A.
Debray~\orcidlink{0000-0002-1790-2719}\\University Medical Center
Utrecht}
\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Valentijn M.T. de
Jong, Gerko Vink, Thomas P.A. Debray} %% comma-separated

\title{Imputation of Incomplete Multilevel Data with R}
\Plaintitle{Imputation of Incomplete Multilevel Data with
R} %% without formatting

%% an abstract and keywords
\Abstract{This tutorial illustrates the imputation of incomplete
multilevel data with the \proglang{R} packackage \pkg{mice}. Our scope
is only simple multilevel models, to show how imputation can yield less
biased estimates from incomplete clustered data. More complex models can
be accomodated, but are outside the scope of this paper. Incomplete
multilevel data requires careful consideration of the missing data
problem and analysis strategy. In this tutorial, we focus on a popular
strategy for accommodating missingness in multilevel data: replacing the
missing data with one or more plausible values, i.e.,
imputation.Imputation separates the missing data problem from the main
analysis and the completed data can be analyzed as if it has been fully
observed. This tutorial illustrates the imputation of incomplete
multilevel data with the statistical programming language R. We aim to
show how imputation can yield less biased estimates from incomplete
clustered data. We provide practical guidelines and code snippets for
different missing data situations, including non-ignorable missingness
mechanisms. For brevity, we focus on multilevel imputation using chained
equations with the R mice package and its adjacent packages.}

%% at least one keyword must be supplied
\Keywords{missing
data, multilevel, clustering, \pkg{mice}, \proglang{R}}
\Plainkeywords{missing data, multilevel, clustering, mice, R}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}
%% \setcounter{page}{1}
%% \Pages{1--xx}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Hanne I. Oberman\\
Methodology and Statistics\\
Padualaan 14\\
Utrecht The Netherlands\\
E-mail: \email{Achim.Zeileis@R-project.org}\\
URL: \url{https://www.zeileis.org/}\\
\\~
Johanna Muñoz\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Valentijn M.T. de Jong\\
Julius Centre for Health Sciences and Primary Care\\
Utrecht The Netherlands\\
\\~
Gerko Vink\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Thomas P.A. Debray\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~

}

\begin{document}
\maketitle
\hypertarget{sec-intro}{%
\section{Introduction: Clustering and incomplete data}\label{sec-intro}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  missing data occur often in data with human subjects
\item
  missing data may be resolved, but need to be handled in accordance
  with the analysis of scientific interest
\item
  in human-subjects research, there is often clustering, which may be
  captured with multilevel modeling techniques
\item
  if the analysis of scientific interest is a multilevel model, the
  missing data handling method should accommodate the multilevel
  structure of the data
\item
  both missingness and multilevel structures require advanced
  statistical techniques
\item
  this tutorial sets out to facilitate empirical researchers in
  accommodating both multilevel structures as well as missing data.
\item
  we illustrate the use of the software by means of three case studies
  from the social and biomedical sciences.
\end{enumerate}

\hypertarget{overview-of-software}{%
\subsection{overview of software}\label{overview-of-software}}

The popular \pkg{mice} package in \proglang{R} \citet{R}\ldots{}

\hypertarget{scope}{%
\subsection{scope}\label{scope}}

\hypertarget{sec-models}{%
\section{Background}\label{sec-models}}

\hypertarget{concepts-in-multilevel-data}{%
\subsection{concepts in multilevel
data}\label{concepts-in-multilevel-data}}

\hypertarget{concepts-in-missing-data}{%
\subsection{concepts in missing data}\label{concepts-in-missing-data}}

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, left=2mm, breakable, arc=.35mm, opacityback=0, leftrule=.75mm, toprule=.15mm, bottomrule=.15mm]

Box 1. The intraclass correlation coefficient.

\end{tcolorbox}

The \proglang{R} package \pkg{mice} provides a framework for imputing
incomplete data on a variable-by-variable basis. The \fct{mice} function
allows users to flexibly specify how many times and under what model the
missing data should be imputed. This is reflected in the first four
function arguments

\begin{verbatim}
mice(data, m, method, predictorMatrix, ...)
\end{verbatim}

where \texttt{data} refers to the incomplete dataset, \texttt{m}
determines the number of imputations, \texttt{method} denotes the
functional form of the imputation model and \texttt{predictorMatrix}
specifies the interrelational dependencies between variables and
imputation models (i.e., the set of predictors to be used for imputing
each incomplete variable).

\hypertarget{sec-illustrations}{%
\section{Illustrations}\label{sec-illustrations}}

In this section, we demonstrate the workflow using three case studies.

\hypertarget{sec-summary}{%
\section{Summary and discussion}\label{sec-summary}}

What is missing from this manuscript\ldots{}

\hypertarget{computational-details}{%
\section*{Computational details}\label{computational-details}}
\addcontentsline{toc}{section}{Computational details}

The results in this paper were obtained using
\proglang{R}\textasciitilde4.3.0. \proglang{R} itself and all packages
used are available from the Comprehensive \proglang{R} Archive Network
(CRAN) at {[}https://CRAN.R-project.org/{]}.

\hypertarget{acknowledgments}{%
\section*{Acknowledgments}\label{acknowledgments}}
\addcontentsline{toc}{section}{Acknowledgments}

This project has received funding from the European Union's Horizon 2020
research and innovation programme under ReCoDID grant agreement No
825746.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\renewcommand{\bibsection}{}
\bibliography{bibliography.bib}

\newpage{}

\hypertarget{sec-techdetails}{%
\section*{More technical details}\label{sec-techdetails}}
\addcontentsline{toc}{section}{More technical details}

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, left=2mm, breakable, arc=.35mm, opacityback=0, leftrule=.75mm, toprule=.15mm, bottomrule=.15mm]

Appendices can be included after the bibliography (with a page break).
Each section within the appendix should have a proper section title
(rather than just \emph{Appendix}).

For more technical style details, please check out JSS's style FAQ at
{[}https://www.jstatsoft.org/pages/view/style\#frequently-asked-questions{]}
which includes the following topics:

\begin{itemize}
\tightlist
\item
  Title vs.~sentence case.
\item
  Graphics formatting.
\item
  Naming conventions.
\item
  Turning JSS manuscripts into \proglang{R} package vignettes.
\item
  Trouble shooting.
\item
  Many other potentially helpful details\ldots{}
\end{itemize}

\end{tcolorbox}

\hypertarget{sec-bibtex}{%
\section*{Using BibTeX}\label{sec-bibtex}}
\addcontentsline{toc}{section}{Using BibTeX}

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, left=2mm, breakable, arc=.35mm, opacityback=0, leftrule=.75mm, toprule=.15mm, bottomrule=.15mm]

References need to be provided in a \textsc{Bib}{\TeX} file
(\texttt{.bib}). All references should be made with \texttt{@cite}
syntax. This commands yield different formats of author-year citations
and allow to include additional details (e.g.,pages, chapters, \dots) in
brackets. In case you are not familiar with these commands see the JSS
style FAQ for details.

Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task --
especially when acquiring the entries automatically from mixed online
sources. However, it is important that informations are complete and
presented in a consistent style to avoid confusions. JSS requires the
following format.

\begin{itemize}
\tightlist
\item
  item JSS-specific markup (\texttt{\textbackslash{}proglang},
  \texttt{\textbackslash{}pkg}, \texttt{\textbackslash{}code}) should be
  used in the references.
\item
  item Titles should be in title case.
\item
  item Journal titles should not be abbreviated and in title case.
\item
  item DOIs should be included where available.
\item
  item Software should be properly cited as well. For \proglang{R}
  packages \texttt{citation("pkgname")} typically provides a good
  starting point.
\end{itemize}

\end{tcolorbox}

\end{document}

### 2023-11-21 docs/template.qmd
move to quarto

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: Achim.Zeileis@R-project.org
    url: https://www.zeileis.org/
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This short article illustrates how to write a manuscript for the
  _Journal of Statistical Software_ (JSS) using its {{< latex >}} style files.
  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
  it is recommended to keep the {{< latex >}} code as simple as possible,
  i.e., avoid inclusion of packages/commands that are not necessary.
  For outlining the typical structure of a JSS article some brief text snippets are employed that have been inspired by @ZeileisKleiberJackman2008, discussing count data regression in [R]{.proglang}. Editorial comments and instructions are marked by vertical bars.

keywords: [JSS, style guide, comma-separated, not capitalized, R]
keywords-formatted: [JSS, style guide, comma-separated, not capitalized, "[R]{.proglang}"]

bibliography: bibliography.bib  
---

## Introduction: Count data regression in R {#sec-intro}

::: callout
The introduction is in principle "as usual". However, it should usually embed both the implemented *methods* and the *software* into the respective relevant literature. For the latter both competing and complementary software should be discussed (within the same software environment and beyond), bringing out relative (dis)advantages. All software mentioned should be properly `@cited`'d. (See also [Using BibTeX] for more details on {{< bibtex >}}.)

For writing about software JSS requires authors to use the markup `[]{.proglang}` (programming languages and large programmable systems), `[]{.pkg}` (software packages), back ticks like \`code\` for code (functions, commands, arguments, etc.).

If there is such markup in (sub)section titles (as above), a plain text version has to be provided in the {{< latex >}} command as well. Below we also illustrate how abbrevations should be introduced and citation commands can be employed. See the {{< latex >}} code for more details.
:::

Modeling count variables is a common task in economics and the social sciences. The classical Poisson regression model for count data is often of limited use in these disciplines because empirical count data sets typically exhibit overdispersion and/or an excess number of zeros. The former issue can be addressed by extending the plain Poisson regression model in various directions: e.g., using sandwich covariances or estimating an additional dispersion parameter (in a so-called quasi-Poisson model). Another more formal way is to use a negative binomial (NB) regression. All of these models belong to the family of generalized linear models (GLMs). However, although these models typically can capture overdispersion rather well, they are in many applications not sufficient for modeling excess zeros. Since @Mullahy1986 there is increased interest in zero-augmented models that address this issue by a second model component capturing zero counts. An overview of count data models in econometrics, including hurdle and zero-inflated models, is provided in @CameronTrivedi2013.

In [R]{.proglang} @R, GLMs are provided by the model fitting functions [glm]{.fct} in the [stats]{.pkg} package and [glm.nb]{.fct} in the [MASS]{.pkg} package [@VenablesRipley2002] along with associated methods for diagnostics and inference. The manuscript that this document is based on [@ZeileisKleiberJackman2008] then introduced hurdle and zero-inflated count models in the functions [hurdle]{.fct} and [zeroinfl]{.fct} in the [pcsl]{.pkg} package @Jackman2015. Of course, much more software could be discussed here, including (but not limited to) generalized additive models for count data as available in the [R]{.proglang} packages [mgcv]{.pkg} @Wood2006, [gamlss]{.pkg} @StasinopoulosRigby2007, or [VGAM]{.pkg} @Yee2009.

## Models and software {#sec-models}

The basic Poisson regression model for count data is a special case of the GLM framework @McCullaghNelder1989. It describes the dependence of a count response variable $y_i$ ($i = 1, \dots, n$) by assuming a Poisson distribution $y_i \sim \mathrm{Pois}(\mu_i)$. The dependence of the conditional mean $E[y_i \, | \, x_i] = \mu_i$ on the regressors $x_i$ is then specified via a log link and a linear predictor

$$
\log(\mu_i) \quad = \quad x_i^\top \beta,
$$ {#eq-mean}

where the regression coefficients $\beta$ are estimated by maximum likelihood (ML) using the iterative weighted least squares (IWLS) algorithm.

::: callout
TODO: Note that around the equation above there should be no spaces (avoided in the {{< latex >}} code by `%` lines) so that "normal" spacing is used and not a new paragraph started.
:::

[R]{.proglang} provides a very flexible implementation of the general GLM framework in the function [glm]{.fct} @ChambersHastie1992 in the [stats]{.pkg} package. Its most important arguments are

```r
glm(formula, data, subset, na.action, weights, offset,
  family = gaussian, start = NULL, control = glm.control(...),
  model = TRUE, y = TRUE, x = FALSE, ...)
```

where `formula` plus `data` is the now standard way of specifying regression relationships in [R]{.proglang}/[S]{.proglang} introduced in @ChambersHastie1992. The remaining arguments in the first line (`subset`, `na.action`, `weights`, and `offset`) are also standard for setting up formula-based regression models in [R]{.proglang}/[S]{.proglang}. The arguments in the second line control aspects specific to GLMs while the arguments in the last line specify which components are returned in the fitted model object (of class [glm]{.class} which inherits from [lm]{.class}). For further arguments to [glm]{.fct} (including alternative specifications of starting values) see `?glm`. For estimating a Poisson model `family = poisson` has to be specified.

::: callout
As the synopsis above is a code listing that is not meant to be executed, one can use either the dedicated `{Code}` environment or a simple `{verbatim}` environment for this. Again, spaces before and after should be avoided.

Finally, there might be a reference to a `{table}` such as @tbl-overview. Usually, these are placed at the top of the page (`[t!]`), centered (`\centering`), with a caption below the table, column headers and captions in sentence style, and if possible avoiding vertical lines.
:::

| Type           | Distribution | Method   | Description                                                                                                                                                                                  |
|---------|---------|---------|-----------------------------------------------|
| GLM            | Poisson      | ML       | Poisson regression: classical GLM, estimated by maximum likelihood (ML)                                                                                                                      |
|                |              | Quasi    | "Quasi-Poisson regression'': same mean function, estimated by quasi-ML (QML) or equivalently generalized estimating equations (GEE), inference adjustment via estimated dispersion parameter |
|                |              | Adjusted | "Adjusted Poisson regression'': same mean function, estimated by QML/GEE, inference adjustment via sandwich covariances                                                                      |
|                | NB           | ML       | NB regression: extended GLM, estimated by ML including additional shape parameter                                                                                                            |
| Zero-augmented | Poisson      | ML       | Zero-inflated Poisson (ZIP), hurdle Poisson                                                                                                                                                  |
|                | NB           | ML       | Zero-inflated NB (ZINB), hurdle NB                                                                                                                                                           |
: Overview of various count regression models. The table is usually placed at the top of the page (`[t!]`), centered (`centering`), has a caption below the table, column headers and captions are in sentence style, and if possible vertical lines should be avoided. {#tbl-overview}

## Illustrations {#sec-illustrations}

For a simple illustration of basic Poisson and NB count regression the
`quine` data from the [MASS]{.pkg} package is used. This provides the number
of `Days` that children were absent from school in Australia in a
particular year, along with several covariates that can be employed as regressors. The data can be loaded by

```{R}
#| prompt: true
data("quine", package = "MASS")
```
and a basic frequency distribution of the response variable is displayed in
@fig-quine.

:::{.callout}
For code input and output, the style files provide dedicated environments.
Either the "agnostic" `{CodeInput}` and `{CodeOutput}` can be used
or, equivalently, the environments `{Sinput}` and `{Soutput}` as
produced by [Sweave]{.fct} or [knitr]{.pkg} when using the `render_sweave()`
hook. Please make sure that all code is properly spaced, e.g., using
`y = a + b * x` and _not_ `y=a+b*x`. Moreover, code input should use "the usual" command prompt in the respective software system. For
[R]{.proglang} code, the prompt `R> ` should be used with `+  ` as
the continuation prompt. Generally, comments within the code chunks should be
avoided -- and made in the regular {{< latex >}} text instead. Finally, empty lines before and after code input/output should be avoided (see above).
:::

::: {#fig-quine}

![](article-visualization.pdf)

Frequency distribution for number of days absent from school.
:::

As a first model for the `quine` data, we fit the basic Poisson regression
model. (Note that JSS prefers when the second line of code is indented by two
spaces.)

```{R}
#| prompt: true
m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine, family = poisson)
```

To account for potential overdispersion we also consider a negative binomial
GLM.

```{R}
#| prompt: true
library("MASS")
m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
```

In a comparison with the BIC the latter model is clearly preferred.

```{R}
#| prompt: true
library("MASS")
BIC(m_pois, m_nbin)
```

Hence, the full summary of that model is shown below.

```{R}
#| prompt: true
summary(m_nbin)
```

## Summary and discussion {#sec-summary}

:::{.callout}

As usual...

:::

## Computational details {.unnumbered}

:::{.callout}

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be included
in an unnumbered section. Also, auxiliary packages (say, for visualizations,
maps, tables, ...) that are not cited in the main text can be credited here.

:::

The results in this paper were obtained using [R]{.proglang}~3.4.1 with the
[MASS]{.pkg}~7.3.47 package. [R]{.proglang} itself and all packages used are available from the Comprehensive [R]{.proglang} Archive Network (CRAN) at
[https://CRAN.R-project.org/].

## Acknowledgments {.unnumbered}

:::{.callout}

All acknowledgments (note the AE spelling) should be collected in this
unnumbered section before the references. It may contain the usual information
about funding and feedback from colleagues/reviewers/etc. Furthermore,
information such as relative contributions of the authors may be added here
(if any).

:::

## References {.unnumbered}

:::{#refs}

:::

{{< pagebreak >}}

## More technical details {#sec-techdetails .unnumbered}

:::{.callout}

Appendices can be included after the bibliography (with a page break). Each
section within the appendix should have a proper section title (rather than
just _Appendix_).

For more technical style details, please check out JSS's style FAQ at
[https://www.jstatsoft.org/pages/view/style#frequently-asked-questions]
which includes the following topics:

- Title vs. sentence case.
- Graphics formatting.
- Naming conventions.
- Turning JSS manuscripts into [R]{.proglang} package vignettes.
- Trouble shooting.
- Many other potentially helpful details...

:::

## Using BibTeX {#sec-bibtex .unnumbered}

:::{.callout}

References need to be provided in a {{< bibtex >}} file (`.bib`). All
references should be made with `@cite` syntax. This commands yield different
formats of author-year citations and allow to include additional details (e.g.,pages, chapters, \dots) in brackets. In case you are not familiar with these
commands see the JSS style FAQ for details.

Cleaning up {{< bibtex >}} files is a somewhat tedious task -- especially
when acquiring the entries automatically from mixed online sources. However,
it is important that informations are complete and presented in a consistent
style to avoid confusions. JSS requires the following format.

- item JSS-specific markup (`\proglang`, `\pkg`, `\code`) should be used in the references.
- item Titles should be in title case.
- item Journal titles should not be abbreviated and in title case.
- item DOIs should be included where available.
- item Software should be properly cited as well. For [R]{.proglang} packages `citation("pkgname")` typically provides a good starting point.

:::

### 2023-11-21 docs/manuscript.qmd
test r chunks

@@
 
-### concepts in missing data
-
 ::: callout
@@
 :::
+
+In [R]{.proglang}, multilevel models may be fitted using the package [lme4]{.pkg}. For linear mixed-effects models, the function
+
+```r
+lmer(formula, data, ...)
+```
+
+### concepts in missing data
 
@@
 
+::: callout
+Box 2. The `methods`.
+:::
+
+::: callout
+Box 2. The predictor matrix.
+:::
+
 ## Illustrations {#sec-illustrations}
@@
 In this section, we demonstrate the workflow using three case studies.
+
+### Setup
+
+```{R}
+#| prompt: true
+#| warning: false
+#| message: false
+library(mice)
+library(ggmice)
+```
+
+### Popularity data
+
+```{R}
+#| prompt: true
+data("popmis", package = "mice")
+```
+
+```{R}
+#| prompt: true
+dat <- popmis[, c("school", "popular", "sex")] 
+```
+
+```{R}
+#| prompt: true
+#| label: fig-pattern
+#| fig-cap: "Missing data pattern."
+plot_pattern(dat)
+```
+
+```{R}
+#| prompt: true
+#| fig-cap: "Pair-wise correlations."
+plot_corr(dat)
+```
+
+```{R}
+#| prompt: true
+meth <- make.method(dat)
+```
+
+```{R}
+#| prompt: true
+pred <- quickpred(dat)
+plot_pred(pred, method = meth)
+```
+
+```{R}
+#| prompt: true
+imp <- mice(
+  data = dat,
+  method = meth,
+  predictorMatrix = pred,
+  printFlag = FALSE
+)
+```
+
+```{R}
+#| prompt: true
+plot_trace(imp)
+```
 

### 2023-11-21 docs/manuscript.tex
test r chunks

@@
 
-\hypertarget{concepts-in-missing-data}{%
-\subsection{concepts in missing data}\label{concepts-in-missing-data}}
-
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, left=2mm, breakable, arc=.35mm, opacityback=0, leftrule=.75mm, toprule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, arc=.35mm, colback=white, left=2mm, toprule=.15mm, rightrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 \end{tcolorbox}
+
+In \proglang{R}, multilevel models may be fitted using the package
+\pkg{lme4}. For linear mixed-effects models, the function
+
+\begin{verbatim}
+lmer(formula, data, ...)
+\end{verbatim}
+
+\hypertarget{concepts-in-missing-data}{%
+\subsection{concepts in missing data}\label{concepts-in-missing-data}}
 
@@
 
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, arc=.35mm, colback=white, left=2mm, toprule=.15mm, rightrule=.15mm, leftrule=.75mm, opacityback=0]
+
+Box 2. The \texttt{methods}.
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, arc=.35mm, colback=white, left=2mm, toprule=.15mm, rightrule=.15mm, leftrule=.75mm, opacityback=0]
+
+Box 2. The predictor matrix.
+
+\end{tcolorbox}
+
 \hypertarget{sec-illustrations}{%
@@
 In this section, we demonstrate the workflow using three case studies.
+
+\hypertarget{setup}{%
+\subsection{Setup}\label{setup}}
+
+\begin{verbatim}
+R> library(mice)
+R> library(ggmice)
+\end{verbatim}
+
+\hypertarget{popularity-data}{%
+\subsection{Popularity data}\label{popularity-data}}
+
+\begin{verbatim}
+R> data("popmis", package = "mice")
+\end{verbatim}
+
+\begin{verbatim}
+R> dat <- popmis[, c("school", "popular", "sex")] 
+\end{verbatim}
+
+\begin{verbatim}
+R> plot_pattern(dat)
+\end{verbatim}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/fig-pattern-1.pdf}
+
+}
+
+\caption{\label{fig-pattern}Missing data pattern.}
+
+\end{figure}
+
+\begin{verbatim}
+R> plot_corr(dat)
+\end{verbatim}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-5-1.pdf}
+
+}
+
+\caption{Pair-wise correlations.}
+
+\end{figure}
+
+\begin{verbatim}
+R> meth <- make.method(dat)
+\end{verbatim}
+
+\begin{verbatim}
+R> pred <- quickpred(dat)
+R> plot_pred(pred, method = meth)
+\end{verbatim}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}
+
+}
+
+\end{figure}
+
+\begin{verbatim}
+R> imp <- mice(
++  data = dat,
++  method = meth,
++  predictorMatrix = pred,
++  printFlag = FALSE
++)
+\end{verbatim}
+
+\begin{verbatim}
+R> plot_trace(imp)
+\end{verbatim}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-9-1.pdf}
+
+}
+
+\end{figure}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, left=2mm, breakable, arc=.35mm, opacityback=0, leftrule=.75mm, toprule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, arc=.35mm, colback=white, left=2mm, toprule=.15mm, rightrule=.15mm, leftrule=.75mm, opacityback=0]
 

### 2023-11-22 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
render

@@
 \usepackage{ulem}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 
@@
 
+Understanding and effectively handling missing data within multilevel
+structures are crucial for preserving the integrity of analyses and
+deriving accurate inferences.
+
 Many datasets include individuals that are clustered together, for
-example in geographic regions, or even different studies. In the
-simplest case, individuals (e.g., students) are nested within a single
-cluster (e.g., school classes). More complex clustered structures may
-occur when there are multiple hierarchical levels (e.g., students in
-different schools or patients within hospitals within regions across
-countries), or when the clustering is non-nested (e.g., electronic
-health record data from diverse settings and populations within large
-databases). With clustered data we generally assume that individuals
-from the same cluster tend to be more similar than individuals from
-other clusters. In statistical terms, this implies that observations
-from the same cluster are not independent and may in fact be correlated.
-If this correlation is left unaddressed, estimates of \emph{p} values,
-confidence intervals even model parameters are prone to bias
-\citep{loca01}. Statistical methods for clustered data typically adopt
-hierarchical models that explicitly describe the grouping of
-observations. These models are also known as `multilevel models',
-`hierarchical models', `mixed effect models', `random effect models',
-and in the context of time-to-event data as `frailty models'. Table
-\ref{tab:clus} provides an overview of some key concepts in multilevel
-modeling.
+example by geographic region. In the simplest case, each individual is a
+unit (e.g., a student) which is nested within a single cluster (e.g., a
+school class). More complex clustered structures may occur when there
+are multiple hierarchical levels (e.g., patients within hospitals within
+regions across countries), or when the clustering is non-nested (e.g.,
+electronic health record data from diverse settings and populations
+within large databases). The hierarchical structures of such clustered
+data may be captured using multilevel analysis techniques.
+
+With clustered data we generally assume that individuals from the same
+cluster tend to be more similar than individuals from other clusters. In
+statistical terms, this implies that observations from the same cluster
+are not independent and may in fact be correlated. If this correlation
+is left unaddressed, estimates of \emph{p} values, confidence intervals
+even model parameters are prone to bias \citep{loca01}. Statistical
+methods for clustered data typically adopt hierarchical models that
+explicitly describe the grouping of observations. These models are also
+known as `multilevel models', `hierarchical models', `mixed effect
+models', `random effect models', and in the context of time-to-event
+data as `frailty models'. Table \ref{tab:clus} provides an overview of
+some key concepts in multilevel modeling.
+
+In the context of multilevel data, missingness can occur at different
+levels, complicating the interpretation and reliability of results.
 

### 2023-11-24 docs/manuscript.qmd
figure positioning

@@
         keep-tex: true
+        fig-pos: 't'
     jss-html: default
@@
     orcid: 0000-0003-3276-2141
-    email: Achim.Zeileis@R-project.org
-    url: https://www.zeileis.org/
+    email: h.i.oberman@uu.nl
+    url: https://www.hanneoberman.github.io
   - name: Johanna Muñoz
@@
 #| fig-cap: "Missing data pattern."
+#| fig-pos: 't'
 plot_pattern(dat)
+```
+
+With the `ggmice` unction `plot_pattern` we can visualize this.
+
+```{R}
+#| layout-ncol: 2
+#| echo: false
+plot_pattern(dat)
+plot_corr(dat)
 ```

### 2023-11-24 docs/manuscript.tex
figure positioning

@@
 Utrecht The Netherlands\\
-E-mail: \email{Achim.Zeileis@R-project.org}\\
-URL: \url{https://www.zeileis.org/}\\
+E-mail: \email{h.i.oberman@uu.nl}\\
+URL: \url{https://www.hanneoberman.github.io}\\
 \\~
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, arc=.35mm, colback=white, left=2mm, toprule=.15mm, rightrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, breakable, leftrule=.75mm, colback=white, bottomrule=.15mm, toprule=.15mm, left=2mm, rightrule=.15mm, opacityback=0]
 
@@
 
-\begin{figure}[H]
+\begin{figure}[t]
 
@@
 
+With the \texttt{ggmice} unction \texttt{plot\_pattern} we can visualize
+this.
+
+\begin{figure}
+
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{manuscript_files/figure-pdf/unnamed-chunk-5-1.pdf}
+
+}
+
+}
+
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{manuscript_files/figure-pdf/unnamed-chunk-5-2.pdf}
+
+}
+
+}
+
+\end{minipage}%
+
+\end{figure}
+
 \begin{verbatim}
@@
 
-\begin{figure}[H]
+\begin{figure}[t]
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-5-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-6-1.pdf}
 
@@
 
-\begin{figure}[H]
+\begin{figure}[t]
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-8-1.pdf}
 
@@
 
-\begin{figure}[H]
+\begin{figure}[t]
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-9-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}
 

### 2023-11-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update imputation workflow

--- 2023-11-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
@@
 
-### Main analysis
-When imputing clustered data, our first consideration should be the research question and the type of analysis that researchers intend to perform, assuming that non-incomplete values were present.
-This initial assessment provides valuable insights into the data's structure (see the level table), relationships among variables, and the causes of missingness.
-
-When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g.
-if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021).
-
-### Exploration of available data
-Then it is required to explore available information across clusters.
-This can involve evaluating the Intra-class Correlation Coefficient (ICC) to determine if a one-level imputation method is sufficient, particularly if there are no substantial differences between clusters.
-Additionally, it can be done by visualizing the data using available package as **ggmice**, for instance examining density plots can help us understand the variable type (normal, Poisson, categorical) and its range of plausible values.
-Exploring relationships between covariates through bivariate plots can be valuable in identifying potential interactions or non-linearities.
-
-In the context of missingness, we should assess patterns of systematic and sporadic missing values and explore the relationships between the missingness of an incomplete variable and observable data.
-This evaluation allows us to consider the plausibility of Missing at Random (MAR) mechanisms over Missing Completely at Random (MCAR) methods and helps in selecting prediction variables to be included in the imputation model (considering overflux).
-
-### Selecting plausible imputation methods
-Before to implement multiple imputation it is requiered to evaluate if simpler imputation methods are suffice: 
-- For example, certain incomplete variables may not necessitate stochastic imputation methods like MICE.
-Instead, they can be effectively addressed through deductive imputation, where incomplete values can be inferred from logic and deterministic relationships between variables.
-This approach is particularly useful for variables that are functions of others, e.g.
-a person's BMI from their weight and height, determining values for one-level variables from two-level ones, e.g.
-in the context of Individual Participant Data (IPD), incomplete information can be extracted from metadata, such as deducing incomplete data about abortion in a country where abortion is illegal, or engaging in cross-temporal or protocol's deduction, like imputing missing test values for deceased patients.
-
-- Under monotonic pattern, the use of regression models
-- CC when the proportion of missing is minimal <5%.
-- Analysis that are robust to missing data, for instance FIML models such as mixed model do not require imputation models when only the outcome variable is missing.An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML).
-This method does not require the imputation of any missing values.
-Whereas MI consists of imputation, analyses and pooling steps, FIML analyses the data in a single step.
-When the assumptions are met the two approaches should produce equivalent results.
-[REF] As FIML requires specialised software, not all analyses can be performed with standard software.
-[REF]
-
-- Add table describing 2l- imputation models. like this https://bookdown.org/mwheymans/bookmi/multiple-imputation-models-for-multilevel-data.html
-
-### Imputation process
-
-#### Clustering specification
-There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: <https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/>) TODO: replace ref.
-
-The selection of each strategy depends mainly on the assumptions in the main analysis and also on the restriction of the analyzed data.
-
-Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
-The last restriction is also required when imputations are performed on each cluster separately.
-When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).
-
-Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
-However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.
-the number of clusters is adequate.
-(Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008).
-Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020).
-For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
-(Zhang,2023).
-
-#### Selection of variables in the imputation model
-- Inclusion of outcome variable [Moons] When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]
-- Imputation for interaction term, JAV, Machine lerarning methods (Random forest) and imputation separated by treatment group
-- Prediction matrix specfication (table)
-
-#### Convergency problems during imputation
-- Check log 
-- Problems of convergence on some 2l methods can be solved by use of overflux or quickpred. 
-- Consider the use of scaled variables Vanburen. 
-- Consider to swap imputation mehtod in problematic variables: The use of non parametrical imputation methods such as pmm. or downgrade to 1l methods.
-- Also by the specification of post-processing in some variables can alivianate this. 
-- Separated imputation process for each endpoint analysis
-
-### Convergency 
-Plots and also the prediction plots.
-
-### Sensibility analysis
-About the possibility of assumptions about Missing mechanism.
+### Focus on the Main Analysis
+When imputing clustered data, we initially focus on the research question and the intended analysis, assuming the presence of non-incomplete values.
+This assessment sheds light on research hypotheses and the contemplated main(s) statistical model(s) and it provides insights into the data's structure (refer to the level table), variable types (e.g., confounders, auxiliary variables), and other considerations about variables relationships such as interactions or polynomial terms.
+
+### Exploration of Available Data
+Next, it is necessary to explore the information available in the data.
+Exploring individual variables using tools such as histograms or QQ plots can help to delineate variable distributions and plausible ranges of values and also identify input errors or outliers.
+Evaluating interactions between predictors using scatter plots or conditional plots, assessing collinearity using VIF or correlations, can help to glimpse nonlinear relationships that may affect the original formulation of the main model.
+Further testing of assumptions related to the response variable, including variance (homogeneity using conditional box plots), independence (e.g., ACF, variograms) and response vs.
+predictor relationships (conditional plots - x vs.
+y), may serve to choose an imputation model more suitable to the data.
+In addition, the intraclass correlation coefficient (ICC) can be examined to assess cluster differences, aiding in the choice between the 2l and 1l methods for imputation.
+
+One can also explore missingness, specifically missing patterns at the cluster level to identify types of patterns (e.g., non-monotonic, systematic) to guide the selection of an appropriate imputation approach in terms of computational efficiency (e.g., simpler regression imputation versus FCS in univariate patterns).
+Similarly, these missing patterns can also be used to identify potential predictors in individual imputation models by means of inflow criteria (quantifying the connections between missing data in one variable and other observed variables) and outflow criteria (determining the connections between observed values in one variable and missing data in other variables).
+Although several packages exist for visualizing missing data, this tutorial focuses on the recently implemented ggmice package.
+
+### Assess Estimation Procedure Robustness to Missing Data
+Prior to implementing multiple imputation, it is critical to assess the robustness of the estimation model when confronted with missing data.
+\cite{Alisson_2002} highlighted scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods.
+Notably, when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}.
+Furthermore, depending on the proportion of missingness (usually is <5%), opting for a simpler complete case analysis might be sufficient.
+
+### Pre-imputation
+Before proceeding with the imputation model, it is necessary to filter the dataset to include only the minimal variables required for the main model.
+If additional procedures are conducted during the analysis, it is essential to include variables associated with these procedures, such as confounders in the case of balancing techniques.
+Other relevant variables, like instrumental variables or auxiliary variables, may enhance parameter estimates even if not included in the main model, as they could be linked to the probability of missingness for some incomplete variables.
+
+Furthermore, at the variable level, it is crucial to assess whether proxy variables should be considered or if deterministic imputation alone is adequate for imputing missing values.
+For instance, certain incomplete variables may not necessitate stochastic imputation methods like MICE.
+Instead, they can be effectively addressed through deductive imputation, where incomplete values are inferred from logical and deterministic relationships between variables.
+This approach is especially beneficial for variables that are functions of others, such as deriving a person's BMI from their weight and height.
+It also proves valuable for determining values for one-level variables from two-level ones, as seen in the context of Individual Participant Data (IPD).
+In such cases, incomplete information can be deduced from metadata, like inferring incomplete data about abortion in a country where abortion is illegal, or through cross-temporal or protocol-based deduction, such as imputing missing test values for deceased patients.
+
+### Setting Imputation Model
+
+##### Clustering Inclusion
+Various strategies can be employed in the imputation process to account for clustering, including the inclusion of a cluster indicator variable, conducting a separate imputation process for each cluster, or employing a simultaneous imputation method that considers clustering (Stata: Clustering and MI Impute) TODO: Replace ref.
+
+The choice of strategy depends primarily on the assumptions made in the main analysis and the constraints imposed by the analyzed data.Concerning data restrictions, if there are few clusters with many observations per cluster, the cluster indicator or imputation separated by groups may be appropriate \cite{Graham_2009}.
+Conversely, an imputation model that simultaneously imputes all clusters using a hierarchical model can be employed when there are more clusters or fewer observations per cluster \cite{Allison_2002}.
+
+In a hierarchical imputation model, correlations between observations within clusters are modeled by a random effect, allowing estimation even with a limited number of observations per cluster.
+Multiple imputation models based on hierarchical models have been proposed, each with different assumptions.
+
+<!--TODO: Add table 2l.imputation models -->
+
+Before choosing a particular strategy, it is crucial to verify if the assumptions of the imputation model align with those of the main analysis.
+For example, using the cluster indicator strategy may introduce bias estimates when the model is based on a hierarchical structure \cite{Taaljard_2008}.
+Even if an imputation strategy congruent with the main model is preferred, it is essential to assess its appropriateness for the data, considering that less complex imputation strategies may lead to unbiased estimates in certain scenarios \cite(Bailey_2020).
+
+#### Choice of Individual Imputation Methods
+The initial step involves selecting the individual imputation model for each incomplete variable in the dataset.
+The mice package automatically proposes imputation methods based on the variable type for 1l variables.
+For 2l level variables, the micemd package's () function can be utilized.
+This function selects among the 2l imputation methods based on the size of the clusters and the proportion of missingness in each cluster.
+
+In addition to the package-defined imputation methods, users can specify custom methods using the "I formula".
+This flexibility allows for the calculation of derived variables internally or adjustments to imputation methods based on specific conditions, such as conditioning the imputation model on the level of an incomplete covariate (e.g., pregnancy test on females).
+
+#### Model Specification
+The imputation model must be congenial with the main model \cite{Meng_1994}.
+Congeniality issues arise when the imputation model and the main model make different assumptions, often due to the omission of a polynomial or interaction term or the use of transformed variables.
+
+The imputation model can include additional terms than the main model terms that do not lead to uncongeniality.
+For instance, it is advisable to include the outcome variable in the imputation model for prediction variables \cite{Moons}.
+In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be included as a covariate in the imputation model [REF].
+Furthermore, the inclusion of auxiliary variables, even if not part of the main model, can be associated with the probability of missingness, enhancing the likelihood of satisfying the Missing at Random (MAR) assumption and improving estimation efficiency \cite{Hardt_2012}.
+
+The specification of imputation models is done on a variable basis, either using a prediction matrix where the type of prediction variable should be specified accordingly (see table) or through a list of formulas in the formula parameter.
+The latter option proves useful in formulating imputation models with polynomial terms or interactions compared to the prediction matrix specification, which requires the inclusion of additional terms as Just Another Variable (JAV).
+
+<!--TODO: Add a table describing 2l-imputation models codes -->
+
+However, even when considering an interaction term, depending on the adapted individual imputation model, it has been suggested, for instance, for treatment interaction effects, to conduct separated imputation by treatment group \cite{Zhang_2023}.
+Additionally, there are imputation models based on random forest or deep learning that can handle interaction and non-linear terms that do not require the explicit inclusion of the non-linear terms.
+
+### Post-Imputation
+
+During the imputation process, certain issues may arise that halt the process.
+In hierarchical model imputations, many issues are linked to methods based on parametric hierarchical models, which may struggle to estimate a model midway through the process.
+In such cases, it is advisable to examine the imputation log file to identify variables with imputation issues.
+Subsequently, reducing the number of predictors in the imputation model could be explored, either by using functions like quickpred or by considering variable transformations, such as scaling.
+Adjusting the level of the hierarchical model (e.g., using a homogenous variance model or a 1l model) can also be beneficial.
+Additionally, checking the range of imputed variables is crucial, as excessively large imputed values for one predictor could lead to convergence issues in other variables.
+This can be mitigated by including post-processing specifications on problematic variables or by using imputation models like Predicted Mean Matching (PMM), ensuring imputed values align with the observable values.
+In certain cases, a separate imputation strategy may be considered.
+For instance, in analyses involving multiple endpoints, conducting separate imputation processes for each endpoint might be preferable to a unified imputation process.
+
+### Convergence and Sensitivity Analysis
+
+Before commencing the analysis, it is imperative to verify the convergence of the imputation process. This is typically achieved through trace plots or by evaluating prediction plots.
+
+While the majority of Multiple Imputation by Chained Equations (MICE) methods are based on Missing at Random (MAR) assumptions, field expert input may suggest that the Missing Not at Random (MNAR) mechanism could be plausible for certain variables.
+An MNAR variable is one in which the probability of missingness depends on an unobservable variable.
+This can occur when missingness is associated with the incomplete value itself (self-marking) or when there is an unobserved variable linked to both the value and the probability of missingness of the incomplete value (indirectly non-informative).
+Specifically, for the indirectly non-informative case in hierarchical datasets, imputation methods based on the Heckman method, such as those proposed by Hammon (2021) and Munoz (2023), can be considered.
 

### 2023-11-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update imputation workflow

--- 2023-11-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
@@
 
-\hypertarget{main-analysis}{%
-\subsubsection{Main analysis}\label{main-analysis}}
-
-When imputing clustered data, our first consideration should be the
-research question and the type of analysis that researchers intend to
-perform, assuming that non-incomplete values were present. This initial
-assessment provides valuable insights into the data's structure (see the
-level table), relationships among variables, and the causes of
-missingness.
-
-When multiple imputation is used to deal with missing data, as the
-imputation and analysis process is performed separately, it is necessary
-that imputation model being congenial with the main analysis model
-(Meng, 1994), e.g.~if the main model accounts for the hierarchical
-structure also imputation model should do it (Audigier, 2021).
+\hypertarget{focus-on-the-main-analysis}{%
+\subsubsection{Focus on the Main
+Analysis}\label{focus-on-the-main-analysis}}
+
+When imputing clustered data, we initially focus on the research
+question and the intended analysis, assuming the presence of
+non-incomplete values. This assessment sheds light on research
+hypotheses and the contemplated main(s) statistical model(s) and it
+provides insights into the data's structure (refer to the level table),
+variable types (e.g., confounders, auxiliary variables), and other
+considerations about variables relationships such as interactions or
+polynomial terms.
 
 \hypertarget{exploration-of-available-data}{%
-\subsubsection{Exploration of available
-data}\label{exploration-of-available-data}}
-
-Then it is required to explore available information across clusters.
-This can involve evaluating the Intra-class Correlation Coefficient
-(ICC) to determine if a one-level imputation method is sufficient,
-particularly if there are no substantial differences between clusters.
-Additionally, it can be done by visualizing the data using available
-package as \textbf{ggmice}, for instance examining density plots can
-help us understand the variable type (normal, Poisson, categorical) and
-its range of plausible values. Exploring relationships between
-covariates through bivariate plots can be valuable in identifying
-potential interactions or non-linearities.
-
-In the context of missingness, we should assess patterns of systematic
-and sporadic missing values and explore the relationships between the
-missingness of an incomplete variable and observable data. This
-evaluation allows us to consider the plausibility of Missing at Random
-(MAR) mechanisms over Missing Completely at Random (MCAR) methods and
-helps in selecting prediction variables to be included in the imputation
-model (considering overflux).
-
-\hypertarget{selecting-plausible-imputation-methods}{%
-\subsubsection{Selecting plausible imputation
-methods}\label{selecting-plausible-imputation-methods}}
-
-Before to implement multiple imputation it is requiered to evaluate if
-simpler imputation methods are suffice: - For example, certain
+\subsubsection{Exploration of Available
+Data}\label{exploration-of-available-data}}
+
+Next, it is necessary to explore the information available in the data.
+Exploring individual variables using tools such as histograms or QQ
+plots can help to delineate variable distributions and plausible ranges
+of values and also identify input errors or outliers. Evaluating
+interactions between predictors using scatter plots or conditional
+plots, assessing collinearity using VIF or correlations, can help to
+glimpse nonlinear relationships that may affect the original formulation
+of the main model. Further testing of assumptions related to the
+response variable, including variance (homogeneity using conditional box
+plots), independence (e.g., ACF, variograms) and response vs.~predictor
+relationships (conditional plots - x vs.~y), may serve to choose an
+imputation model more suitable to the data. In addition, the intraclass
+correlation coefficient (ICC) can be examined to assess cluster
+differences, aiding in the choice between the 2l and 1l methods for
+imputation.
+
+One can also explore missingness, specifically missing patterns at the
+cluster level to identify types of patterns (e.g., non-monotonic,
+systematic) to guide the selection of an appropriate imputation approach
+in terms of computational efficiency (e.g., simpler regression
+imputation versus FCS in univariate patterns). Similarly, these missing
+patterns can also be used to identify potential predictors in individual
+imputation models by means of inflow criteria (quantifying the
+connections between missing data in one variable and other observed
+variables) and outflow criteria (determining the connections between
+observed values in one variable and missing data in other variables).
+Although several packages exist for visualizing missing data, this
+tutorial focuses on the recently implemented ggmice package.
+
+\hypertarget{assess-estimation-procedure-robustness-to-missing-data}{%
+\subsubsection{Assess Estimation Procedure Robustness to Missing
+Data}\label{assess-estimation-procedure-robustness-to-missing-data}}
+
+Prior to implementing multiple imputation, it is critical to assess the
+robustness of the estimation model when confronted with missing data.
+\cite{Alisson_2002} highlighted scenarios in which specific Maximum
+Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
+methods. Notably, when the response variable is the sole incomplete
+variable, mixed models demonstrate robustness to missing data under the
+Missing at Random (MAR) assumption and with a correct
+variance-covariance specification.\cite{Molenberghs_2007}. Furthermore,
+depending on the proportion of missingness (usually is \textless5\%),
+opting for a simpler complete case analysis might be sufficient.
+
+\hypertarget{pre-imputation}{%
+\subsubsection{Pre-imputation}\label{pre-imputation}}
+
+Before proceeding with the imputation model, it is necessary to filter
+the dataset to include only the minimal variables required for the main
+model. If additional procedures are conducted during the analysis, it is
+essential to include variables associated with these procedures, such as
+confounders in the case of balancing techniques. Other relevant
+variables, like instrumental variables or auxiliary variables, may
+enhance parameter estimates even if not included in the main model, as
+they could be linked to the probability of missingness for some
+incomplete variables.
+
+Furthermore, at the variable level, it is crucial to assess whether
+proxy variables should be considered or if deterministic imputation
+alone is adequate for imputing missing values. For instance, certain
 incomplete variables may not necessitate stochastic imputation methods
 like MICE. Instead, they can be effectively addressed through deductive
-imputation, where incomplete values can be inferred from logic and
+imputation, where incomplete values are inferred from logical and
 deterministic relationships between variables. This approach is
-particularly useful for variables that are functions of others, e.g.~a
-person's BMI from their weight and height, determining values for
-one-level variables from two-level ones, e.g.~in the context of
-Individual Participant Data (IPD), incomplete information can be
-extracted from metadata, such as deducing incomplete data about abortion
-in a country where abortion is illegal, or engaging in cross-temporal or
-protocol's deduction, like imputing missing test values for deceased
-patients.
-
-\begin{itemize}
-\item
-  Under monotonic pattern, the use of regression models
-\item
-  CC when the proportion of missing is minimal \textless5\%.
-\item
-  Analysis that are robust to missing data, for instance FIML models
-  such as mixed model do not require imputation models when only the
-  outcome variable is missing.An alternative approach to missing data is
-  to use Full Information Maximum Likelihood (FIML). This method does
-  not require the imputation of any missing values. Whereas MI consists
-  of imputation, analyses and pooling steps, FIML analyses the data in a
-  single step. When the assumptions are met the two approaches should
-  produce equivalent results. {[}REF{]} As FIML requires specialised
-  software, not all analyses can be performed with standard software.
-  {[}REF{]}
-\item
-  Add table describing 2l- imputation models. like this
-  \url{https://bookdown.org/mwheymans/bookmi/multiple-imputation-models-for-multilevel-data.html}
-\end{itemize}
-
-\hypertarget{imputation-process}{%
-\subsubsection{Imputation process}\label{imputation-process}}
-
-\hypertarget{clustering-specification}{%
-\paragraph{Clustering specification}\label{clustering-specification}}
-
-There are different strategies that can be adopted in the imputation
-process that account for clustering: inclusion of cluster indicator
-variable, performing a separate imputation process for each cluster, or
-performing a simultaneous imputation process by using an imputation
-method that accounts for clustering.(Stata:
-\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
-TODO: replace ref.
-
-The selection of each strategy depends mainly on the assumptions in the
-main analysis and also on the restriction of the analyzed data.
-
-Regarding the restrictions imposed by the data, for instance, the use of
-cluster indicator variables is restricted in datasets where there are
-not many clusters and many observations per cluster (Graham, 2009). The
-last restriction is also required when imputations are performed on each
-cluster separately. When this restriction cannot be achieved, one can
-use an imputation model that simultaneously imputes all clusters using a
-hierarchical model (Allison 2002).
-
-Under this hierarchical imputation model, observations within clusters
-are correlated and this correlation is modeled by a random effect so the
-hierarchical model can be estimated even when there are few observations
-per cluster. However, this strategy is best suited for balanced data
-(Grund, 2017) and when random effects model is appropriated, i.e.~the
-number of clusters is adequate. (Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main
-model, for instance by using the cluster indicator strategy may lead to
-bias estimates when the model is based on a hierarchical model
-(Taaljard,2008). Even when an imputation strategy congenial with the
-main model is preferred, it is important to consider whether it is
-appropriate for the data as a less complex imputation strategies may
-also lead to unbiased estimates in certain scenarios(Bailey 2020). For
-instance, in causal effect analysis, separately imputation may lead to
-smaller bias when the size of the smaller exposure cluster is large,
-compared with an imputation model that includes exposure-confounder
-interactions. (Zhang,2023).
-
-\hypertarget{selection-of-variables-in-the-imputation-model}{%
-\paragraph{Selection of variables in the imputation
-model}\label{selection-of-variables-in-the-imputation-model}}
-
-\begin{itemize}
-\tightlist
-\item
-  Inclusion of outcome variable {[}Moons{]} When the outcome is
-  time-to-event, the Nelson-Aalen estimate of the time to event should
-  be included as a covariate in the imputation model {[}REF{]}
-\item
-  Imputation for interaction term, JAV, Machine lerarning methods
-  (Random forest) and imputation separated by treatment group
-\item
-  Prediction matrix specfication (table)
-\end{itemize}
-
-\hypertarget{convergency-problems-during-imputation}{%
-\paragraph{Convergency problems during
-imputation}\label{convergency-problems-during-imputation}}
-
-\begin{itemize}
-\tightlist
-\item
-  Check log
-\item
-  Problems of convergence on some 2l methods can be solved by use of
-  overflux or quickpred.
-\item
-  Consider the use of scaled variables Vanburen.
-\item
-  Consider to swap imputation mehtod in problematic variables: The use
-  of non parametrical imputation methods such as pmm. or downgrade to 1l
-  methods.
-\item
-  Also by the specification of post-processing in some variables can
-  alivianate this.
-\item
-  Separated imputation process for each endpoint analysis
-\end{itemize}
-
-\hypertarget{convergency}{%
-\subsubsection{Convergency}\label{convergency}}
-
-Plots and also the prediction plots.
-
-\hypertarget{sensibility-analysis}{%
-\subsubsection{Sensibility analysis}\label{sensibility-analysis}}
-
-About the possibility of assumptions about Missing mechanism.
+especially beneficial for variables that are functions of others, such
+as deriving a person's BMI from their weight and height. It also proves
+valuable for determining values for one-level variables from two-level
+ones, as seen in the context of Individual Participant Data (IPD). In
+such cases, incomplete information can be deduced from metadata, like
+inferring incomplete data about abortion in a country where abortion is
+illegal, or through cross-temporal or protocol-based deduction, such as
+imputing missing test values for deceased patients.
+
+\hypertarget{setting-imputation-model}{%
+\subsubsection{Setting Imputation
+Model}\label{setting-imputation-model}}
+
+\hypertarget{clustering-inclusion}{%
+\subparagraph{Clustering Inclusion}\label{clustering-inclusion}}
+
+Various strategies can be employed in the imputation process to account
+for clustering, including the inclusion of a cluster indicator variable,
+conducting a separate imputation process for each cluster, or employing
+a simultaneous imputation method that considers clustering (Stata:
+Clustering and MI Impute) TODO: Replace ref.
+
+The choice of strategy depends primarily on the assumptions made in the
+main analysis and the constraints imposed by the analyzed
+data.Concerning data restrictions, if there are few clusters with many
+observations per cluster, the cluster indicator or imputation separated
+by groups may be appropriate \cite{Graham_2009}. Conversely, an
+imputation model that simultaneously imputes all clusters using a
+hierarchical model can be employed when there are more clusters or fewer
+observations per cluster \cite{Allison_2002}.
+
+In a hierarchical imputation model, correlations between observations
+within clusters are modeled by a random effect, allowing estimation even
+with a limited number of observations per cluster. Multiple imputation
+models based on hierarchical models have been proposed, each with
+different assumptions.
+
+Before choosing a particular strategy, it is crucial to verify if the
+assumptions of the imputation model align with those of the main
+analysis. For example, using the cluster indicator strategy may
+introduce bias estimates when the model is based on a hierarchical
+structure \cite{Taaljard_2008}. Even if an imputation strategy congruent
+with the main model is preferred, it is essential to assess its
+appropriateness for the data, considering that less complex imputation
+strategies may lead to unbiased estimates in certain scenarios
+\textbackslash cite(Bailey\_2020).
+
+\hypertarget{choice-of-individual-imputation-methods}{%
+\paragraph{Choice of Individual Imputation
+Methods}\label{choice-of-individual-imputation-methods}}
+
+The initial step involves selecting the individual imputation model for
+each incomplete variable in the dataset. The mice package automatically
+proposes imputation methods based on the variable type for 1l variables.
+For 2l level variables, the micemd package's () function can be
+utilized. This function selects among the 2l imputation methods based on
+the size of the clusters and the proportion of missingness in each
+cluster.
+
+In addition to the package-defined imputation methods, users can specify
+custom methods using the ``I formula''. This flexibility allows for the
+calculation of derived variables internally or adjustments to imputation
+methods based on specific conditions, such as conditioning the
+imputation model on the level of an incomplete covariate (e.g.,
+pregnancy test on females).
+
+\hypertarget{model-specification}{%
+\paragraph{Model Specification}\label{model-specification}}
+
+The imputation model must be congenial with the main model
+\cite{Meng_1994}. Congeniality issues arise when the imputation model
+and the main model make different assumptions, often due to the omission
+of a polynomial or interaction term or the use of transformed variables.
+
+The imputation model can include additional terms than the main model
+terms that do not lead to uncongeniality. For instance, it is advisable
+to include the outcome variable in the imputation model for prediction
+variables \cite{Moons}. In cases where the outcome is time-to-event, the
+Nelson-Aalen estimate of the time to the event should be included as a
+covariate in the imputation model {[}REF{]}. Furthermore, the inclusion
+of auxiliary variables, even if not part of the main model, can be
+associated with the probability of missingness, enhancing the likelihood
+of satisfying the Missing at Random (MAR) assumption and improving
+estimation efficiency \cite{Hardt_2012}.
+
+The specification of imputation models is done on a variable basis,
+either using a prediction matrix where the type of prediction variable
+should be specified accordingly (see table) or through a list of
+formulas in the formula parameter. The latter option proves useful in
+formulating imputation models with polynomial terms or interactions
+compared to the prediction matrix specification, which requires the
+inclusion of additional terms as Just Another Variable (JAV).
+
+However, even when considering an interaction term, depending on the
+adapted individual imputation model, it has been suggested, for
+instance, for treatment interaction effects, to conduct separated
+imputation by treatment group \cite{Zhang_2023}. Additionally, there are
+imputation models based on random forest or deep learning that can
+handle interaction and non-linear terms that do not require the explicit
+inclusion of the non-linear terms.
+
+\hypertarget{post-imputation}{%
+\subsubsection{Post-Imputation}\label{post-imputation}}
+
+During the imputation process, certain issues may arise that halt the
+process. In hierarchical model imputations, many issues are linked to
+methods based on parametric hierarchical models, which may struggle to
+estimate a model midway through the process. In such cases, it is
+advisable to examine the imputation log file to identify variables with
+imputation issues. Subsequently, reducing the number of predictors in
+the imputation model could be explored, either by using functions like
+quickpred or by considering variable transformations, such as scaling.
+Adjusting the level of the hierarchical model (e.g., using a homogenous
+variance model or a 1l model) can also be beneficial. Additionally,
+checking the range of imputed variables is crucial, as excessively large
+imputed values for one predictor could lead to convergence issues in
+other variables. This can be mitigated by including post-processing
+specifications on problematic variables or by using imputation models
+like Predicted Mean Matching (PMM), ensuring imputed values align with
+the observable values. In certain cases, a separate imputation strategy
+may be considered. For instance, in analyses involving multiple
+endpoints, conducting separate imputation processes for each endpoint
+might be preferable to a unified imputation process.
+
+\hypertarget{convergence-and-sensitivity-analysis}{%
+\subsubsection{Convergence and Sensitivity
+Analysis}\label{convergence-and-sensitivity-analysis}}
+
+Before commencing the analysis, it is imperative to verify the
+convergence of the imputation process. This is typically achieved
+through trace plots or by evaluating prediction plots.
+
+While the majority of Multiple Imputation by Chained Equations (MICE)
+methods are based on Missing at Random (MAR) assumptions, field expert
+input may suggest that the Missing Not at Random (MNAR) mechanism could
+be plausible for certain variables. An MNAR variable is one in which the
+probability of missingness depends on an unobservable variable. This can
+occur when missingness is associated with the incomplete value itself
+(self-marking) or when there is an unobserved variable linked to both
+the value and the probability of missingness of the incomplete value
+(indirectly non-informative). Specifically, for the indirectly
+non-informative case in hierarchical datasets, imputation methods based
+on the Heckman method, such as those proposed by Hammon (2021) and Munoz
+(2023), can be considered.
 

### 2023-12-07 docs/manuscript.qmd
Add scatterplots

@@
         fig-pos: 't'
+        fig-width: 7
+        fig-height: 3.5
     jss-html: default
@@
 #| message: false
+set.seed(123)
 library(mice)
 library(ggmice)
+library(ggplot2)
+library(miceadds)
+library(lme4)
+library(mitml)
+library(broom.mixed)
 ```
@@
 #| prompt: true
-dat <- popmis[, c("school", "popular", "sex")] 
+dat <- popmis[, c("school", "teachpop", "popular", "texp", "sex")] 
 ```
+
+```{R}
+#| fig-cap: "Polar axis plot"
+#| fig-alt: "A line plot on a polar axis"
+ggmice(dat, aes(popular, teachpop)) + 
+  geom_jitter()
+```
+
+With the `ggmice` unction `plot_pattern` we can visualize this.
 
@@
 
-With the `ggmice` unction `plot_pattern` we can visualize this.
-
-```{R}
+```{R, eval=FALSE}
 #| layout-ncol: 2
 #| echo: false
-plot_pattern(dat)
-plot_corr(dat)
+# plot_pattern(dat)
+# plot_corr(dat)
 ```
@@
 meth <- make.method(dat)
+meth
 ```
@@
 pred <- quickpred(dat)
+pred
+```
+
+Adjust the methods vector.
+```{R}
+#| prompt: true
+meth["popular"] <- "2l.pmm"
+```
+
+Adjust the predictor matrix.
+```{R}
+#| prompt: true
+pred["popular", "school"] <- -2
+pred["popular", "sex"] <- 2
+```
+
+Visualize the imputation methods and predictors.
+```{R}
 plot_pred(pred, method = meth)
@@
 
+Impute the data.
 ```{R}
@@
 
+Evaluate the convergence.
 ```{R}
@@
 plot_trace(imp)
+```
+
+Evaluate the distribution of imputed values.
+```{R}
+ggmice(imp, aes(popular, group = .imp)) + 
+  geom_density() 
+```
+
+Evaluate the distribution of imputed values.
+```{R}
+ggmice(imp, aes(.imp, popular)) + 
+  geom_jitter(alpha = 0.05) +
+    geom_boxplot()
+```
+
+```{R}
+ggmice(imp, aes(popular, teachpop)) + 
+  geom_jitter() +
+  facet_wrap(~ .imp)
+```
+
+Analyze the imputed data.
+```{R}
+fit <- with(
+  imp,
+  lmer(teachpop ~ popular + texp + (1 | school))
+)
+```
+
+Pool the estimates.
+```{R}
+pool(fit)
+```
+
+Display results in table.
+```{R}
+testEstimates(as.mitml.result(fit), extra.pars = TRUE)
 ```

### 2023-12-07 docs/manuscript.tex
Add scatterplots

@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, breakable, leftrule=.75mm, colback=white, bottomrule=.15mm, toprule=.15mm, left=2mm, rightrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, breakable, colback=white, opacityback=0, arc=.35mm, rightrule=.15mm]
 
@@
 \begin{verbatim}
+R> set.seed(123)
 R> library(mice)
 R> library(ggmice)
+R> library(ggplot2)
+R> library(miceadds)
+R> library(lme4)
+R> library(mitml)
+R> library(broom.mixed)
 \end{verbatim}
@@
 \begin{verbatim}
-R> dat <- popmis[, c("school", "popular", "sex")] 
+R> dat <- popmis[, c("school", "teachpop", "popular", "texp", "sex")] 
 \end{verbatim}
+
+\begin{verbatim}
+ggmice(dat, aes(popular, teachpop)) + 
+  geom_jitter()
+\end{verbatim}
+
+\begin{figure}[t]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-4-1.pdf}
+
+}
+
+\caption{Polar axis plot}
+
+\end{figure}
+
+With the \texttt{ggmice} unction \texttt{plot\_pattern} we can visualize
+this.
 
@@
 
-With the \texttt{ggmice} unction \texttt{plot\_pattern} we can visualize
-this.
-
 \begin{figure}
-
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{manuscript_files/figure-pdf/unnamed-chunk-5-1.pdf}
-
-}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{manuscript_files/figure-pdf/unnamed-chunk-5-2.pdf}
-
-}
-
-}
-
-\end{minipage}%
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-6-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}
 
@@
 R> meth <- make.method(dat)
+R> meth
+\end{verbatim}
+
+\begin{verbatim}
+  school teachpop  popular     texp      sex 
+      ""       ""    "pmm"       ""       "" 
 \end{verbatim}
@@
 R> pred <- quickpred(dat)
-R> plot_pred(pred, method = meth)
+R> pred
+\end{verbatim}
+
+\begin{verbatim}
+         school teachpop popular texp sex
+school        0        0       0    0   0
+teachpop      0        0       0    0   0
+popular       0        1       0    1   1
+texp          0        0       0    0   0
+sex           0        0       0    0   0
+\end{verbatim}
+
+Adjust the methods vector.
+
+\begin{verbatim}
+R> meth["popular"] <- "2l.pmm"
+\end{verbatim}
+
+Adjust the predictor matrix.
+
+\begin{verbatim}
+R> pred["popular", "school"] <- -2
+R> pred["popular", "sex"] <- 2
+\end{verbatim}
+
+Visualize the imputation methods and predictors.
+
+\begin{verbatim}
+plot_pred(pred, method = meth)
 \end{verbatim}
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-8-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-12-1.pdf}
 
@@
 \end{figure}
+
+Impute the data.
 
@@
 
+Evaluate the convergence.
+
 \begin{verbatim}
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-14-1.pdf}
 
@@
 \end{figure}
+
+Evaluate the distribution of imputed values.
+
+\begin{verbatim}
+ggmice(imp, aes(popular, group = .imp)) + 
+  geom_density() 
+\end{verbatim}
+
+\begin{figure}[t]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-15-1.pdf}
+
+}
+
+\end{figure}
+
+Evaluate the distribution of imputed values.
+
+\begin{verbatim}
+ggmice(imp, aes(.imp, popular)) + 
+  geom_jitter(alpha = 0.05) +
+    geom_boxplot()
+\end{verbatim}
+
+\begin{figure}[t]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-16-1.pdf}
+
+}
+
+\end{figure}
+
+\begin{verbatim}
+ggmice(imp, aes(popular, teachpop)) + 
+  geom_jitter() +
+  facet_wrap(~ .imp)
+\end{verbatim}
+
+\begin{figure}[t]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-17-1.pdf}
+
+}
+
+\end{figure}
+
+Analyze the imputed data.
+
+\begin{verbatim}
+fit <- with(
+  imp,
+  lmer(teachpop ~ popular + texp + (1 | school))
+)
+\end{verbatim}
+
+Pool the estimates.
+
+\begin{verbatim}
+pool(fit)
+\end{verbatim}
+
+\begin{verbatim}
+Class: mipo    m = 5 
+         term m  estimate         ubar            b            t dfcom
+1 (Intercept) 5 2.4091354 2.241304e-02 1.712964e-03 2.446860e-02  1995
+2     popular 5 0.2597284 2.353344e-04 1.209648e-04 3.804922e-04  1995
+3        texp 5 0.0484727 7.728295e-05 3.236252e-06 8.116646e-05  1995
+         df        riv     lambda        fmi
+1 432.50579 0.09171257 0.08400798 0.08821454
+2  26.88403 0.61681474 0.38149995 0.42289329
+3 909.68447 0.05025044 0.04784615 0.04993264
+\end{verbatim}
+
+Display results in table.
+
+\begin{verbatim}
+testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+\end{verbatim}
+
+\begin{verbatim}
+
+Call:
+
+testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
+
+Final parameter estimates and inferences obtained from 5 imputed data sets.
+
+             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
+(Intercept)     2.409     0.156    15.401   566.786     0.000     0.092     0.087 
+popular         0.260     0.020    13.315    27.483     0.000     0.617     0.422 
+texp            0.048     0.009     5.380  1747.294     0.000     0.050     0.049
+
+                            Estimate 
+Intercept~~Intercept|school    0.310 
+Residual~~Residual             0.307 
+ICC|school                     0.502
+
+Unadjusted hypothesis test as appropriate in larger samples.
+\end{verbatim}
 

### 2023-12-08 docs/manuscript.qmd
Add introduction text

@@
 For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
-
 keywords: [missing data, multilevel, clustering, mice, R]
 keywords-formatted: [missing data, multilevel, clustering, "[mice]{.pkg}", "[R]{.proglang}"]
-
 bibliography: bibliography.bib  
@@
 
+```{r}
+#| label: setup
+#| echo: false
+library(ggplot2)
+source("../docs/utils.R")
+```
+
 ### overview of software
 
-The popular [mice]{.pkg} package in [R]{.proglang} @R...
+The popular [mice]{.pkg} package in [R]{.proglang} @R
 
 ### scope
+
+This papers serves as a tutorial for imputing incomplete multilevel data with [mice]{.pkg} in [R]{.proglang}.
+[mice]{.pkg} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis.
+[mice]{.pkg} is known to yield valid inferences under many different missing data circumstances [@buur18].
+
+We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms.
+For reasons of brevity, we focus on multilevel imputation by chained equations with [mice]{.pkg} exclusively; other imputation methods and packages [see e.g.
+@audi18 and @grun18] are outside the scope of this tutorial.
+Assumed knowledge includes basic familiarity with the [lme4]{.pkg} notation for multilevel models (see Table \ref{tab:mod}).
+
+We illustrate imputation of incomplete multilevel data using three case studies:
+
+- `popmis` from the [mice]{.pkg} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools with data that are MAR, @mice];
+- `impact` from the [metamisc]{.pkg} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies with data that are MAR, @metamisc];
+- `obesity` from the [micemd]{.pkg} package [simulated data on obesity, $n = 2,111$ patients across $N = 5$ regions with data that are MNAR].
+
+For each of these datasets, we discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.
+
+This tutorial is dedicated to readers who are unfamiliar with multiple imputation.
+More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2) or under MNAR conditions (case study 3).
 
@@
 
+Many datasets include individuals that are clustered together, for example in geographic regions, or even different studies.
+In the simplest case, individuals (e.g., students) are nested within a single cluster (e.g., school classes).
+More complex clustered structures may occur when there are multiple hierarchical levels (e.g., students in different schools or patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases).
+With clustered data we generally assume that individuals from the same cluster tend to be more similar than individuals from other clusters.
+In statistical terms, this implies that observations from the same cluster are not independent and may in fact be correlated.
+If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01].
+Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations.
+These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'.
+Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.
+
 ::: callout
@@
 :::
+
+```{R}
+#| echo: false
+concepts <-
+  data.frame(
+    Concept = c(
+      "Sample units",
+      "Cluster",
+      "Hierarchical data",
+      "Level-1",
+      "Level-2",
+      "Hierarchical model",
+      "Fixed effect",
+      "Random effect",
+      "Mixed effect",
+      "ICC",
+      "Stratified intercept"
+    ),
+    Details = c(
+      "Units of the population from which measurements are taken in a sample, e.g., students.",
+      "Variable that specify the cluster or agruppation, e.g., Classroom",
+      "Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.",
+      "Variable that varies within a cluster, eg. Test score",
+      "Variable that does not vary within a cluster but between, e.g. teacher experience.",
+      "Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)" ,
+      "Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)",
+      "Effects that are a source of random variation in the data, and whose levels are not fully sampled.
+e.g.
+test score tendency during academic year between students due to no controlled factors such as  genetic,family history",
+      "Includes fixed and random effects, e.g.
+the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
+Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
+In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
+@gelm06, @hox17 and @jong21).",
+      "The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
+of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
+""
+    )
+  )
+# concepts |>
+#   kable(
+#     booktabs = T,
+#     escape = F,
+#     caption = "Concepts in multilevel methods",
+#     col.names = c("Concept", "Details"),
+#     align = "l"
+#   ) |>
+#   column_spec(c(1), width = "3cm") |>
+#   column_spec(c(2), width = "12cm")
+```
 
@@
 ### concepts in missing data
+
+missing data mechanisms etc.
+
+As with any other dataset, clustered datasets may be impacted by missingness in much the same way.
+Several strategies can be used to handle missing data, including complete case analysis and imputation.
+We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values.
+Imputation separates the missing data problem from the analysis and the completed data can be analyzed as if it were completely observed.
+It is generally recommended to impute the missing values more than once to preserve uncertainty due to missingness and to allow for valid inferences (c.f.
+Rubin 1976).
+
+With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13].
+Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18].
+For example, it is possible that test results are missing for several students in one or more classes.
+When all observations are missing within one or more clusters, data are said to be systematically missing.
+Sporadic missingness is visualized in Figure XYZ.
+
+```{R}
+#| echo: false
+plot_na()
+```
+
+Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing.
+To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account.
+For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters).
+We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3.
+If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis).
+Complete case analysis can however introduce bias in statistical inferences and lowers statistical power.
+Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way.
+Complete case analysis in that case would mean excluding the entire cluster from the analyses.
+The wrong choice of missing data handling method can thus be extremely harmful to the inferences.
+
+Imputation of missing data requires consideration of the mechanism behind the missingness.
+Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}).
+For each of these three missingness generating mechanisms, different imputation strategies are warranted (@yuce08 and @hox15).
+We here consider the general case that data are MAR, and expand on certain MNAR situations.
+
+### imputation with mice
 
@@
 ::: callout
-Box 2. The predictor matrix.
+Box 3.
+The predictor matrix.
+The entries corresponding to the level-1 predictors are coded with a 3, indicating that both the original values as well as the cluster means of the predictor are included into the imputation model.
+The entry of 4 in the predictor matrix adds three variables to the imputation model for the imputation model predictor: the value of the predictor, the cluster means of the predictor and the random slopes of the predictor.
 :::

### 2023-12-08 docs/manuscript.tex
Add introduction text

@@
 
+\begin{verbatim}
+Warning: package 'ggplot2' was built under R version 4.3.2
+\end{verbatim}
+
 \hypertarget{overview-of-software}{%
@@
 
-The popular \pkg{mice} package in \proglang{R} \citet{R}\ldots{}
+The popular \pkg{mice} package in \proglang{R} \citet{R}
 
@@
 \subsection{scope}\label{scope}}
+
+This papers serves as a tutorial for imputing incomplete multilevel data
+with \pkg{mice} in \proglang{R}. \pkg{mice} has become the de-facto
+standard for imputation by chained equations, which iteratively solves
+the missingness on a variable-by-variable basis. \pkg{mice} is known to
+yield valid inferences under many different missing data circumstances
+\citep{buur18}.
+
+We provide practical guidelines and code snippets for different missing
+data situations, including non-ignorable mechanisms. For reasons of
+brevity, we focus on multilevel imputation by chained equations with
+\pkg{mice} exclusively; other imputation methods and packages \citep[see
+e.g.][ and \citet{grun18}]{audi18} are outside the scope of this
+tutorial. Assumed knowledge includes basic familiarity with the
+\pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).
+
+We illustrate imputation of incomplete multilevel data using three case
+studies:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
+  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
+  with data that are MAR,][]{mice};
+\item
+  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
+  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
+  studies with data that are MAR,][]{metamisc};
+\item
+  \texttt{obesity} from the \pkg{micemd} package {[}simulated data on
+  obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
+  that are MNAR{]}.
+\end{itemize}
+
+For each of these datasets, we discuss the nature of the missingness,
+choose one or more imputation models and evaluate the imputed data, but
+we will also highlight one specific aspect of the imputation workflow.
+
+This tutorial is dedicated to readers who are unfamiliar with multiple
+imputation. More experienced readers can skip the introduction (case
+study 1) and directly head to practical applications of multilevel
+imputation under MAR conditions (case study 2) or under MNAR conditions
+(case study 3).
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, breakable, colback=white, opacityback=0, arc=.35mm, rightrule=.15mm]
+Many datasets include individuals that are clustered together, for
+example in geographic regions, or even different studies. In the
+simplest case, individuals (e.g., students) are nested within a single
+cluster (e.g., school classes). More complex clustered structures may
+occur when there are multiple hierarchical levels (e.g., students in
+different schools or patients within hospitals within regions across
+countries), or when the clustering is non-nested (e.g., electronic
+health record data from diverse settings and populations within large
+databases). With clustered data we generally assume that individuals
+from the same cluster tend to be more similar than individuals from
+other clusters. In statistical terms, this implies that observations
+from the same cluster are not independent and may in fact be correlated.
+If this correlation is left unaddressed, estimates of \emph{p} values,
+confidence intervals even model parameters are prone to bias
+\citep{loca01}. Statistical methods for clustered data typically adopt
+hierarchical models that explicitly describe the grouping of
+observations. These models are also known as `multilevel models',
+`hierarchical models', `mixed effect models', `random effect models',
+and in the context of time-to-event data as `frailty models'. Table
+\ref{tab:clus} provides an overview of some key concepts in multilevel
+modeling.
+
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomrule=.15mm, toprule=.15mm, breakable, leftrule=.75mm, left=2mm, colback=white, opacityback=0, rightrule=.15mm]
 
@@
 \subsection{concepts in missing data}\label{concepts-in-missing-data}}
+
+missing data mechanisms etc.
+
+As with any other dataset, clustered datasets may be impacted by
+missingness in much the same way. Several strategies can be used to
+handle missing data, including complete case analysis and imputation. We
+focus on the latter approach and discuss statistical methods for
+replacing the missing data with one or more plausible values. Imputation
+separates the missing data problem from the analysis and the completed
+data can be analyzed as if it were completely observed. It is generally
+recommended to impute the missing values more than once to preserve
+uncertainty due to missingness and to allow for valid inferences (c.f.
+Rubin 1976).
+
+With incomplete clustered datasets we can distinguish between two types
+of missing data: sporadic missingness and systematic missingness
+\citep{resc13}. Sporadic missingness arises when variables are missing
+for some but not all of the units in a cluster \citep{buur18, jola18}.
+For example, it is possible that test results are missing for several
+students in one or more classes. When all observations are missing
+within one or more clusters, data are said to be systematically missing.
+Sporadic missingness is visualized in Figure XYZ.
+
+\begin{verbatim}
+plot_na()
+\end{verbatim}
+
+\begin{figure}[t]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-3-1.pdf}
+
+}
+
+\end{figure}
+
+Column \(X_1\) in Figure 1 is completely observed, column \(X_2\) is
+systematically missing in cluster 2, and column \(X_3\) is sporadically
+missing. To analyze these incomplete data, we have to take the nature of
+the missingness and the cluster structure into account. For example, the
+sporadic missingness in \(X_3\) could be easily amended if this would be
+a cluster-level variable (and thus constant within clusters). We could
+then just extrapolate the true (but missing) value of \(X_3\) for unit 1
+from unit 2, and the value for unit 4 from unit 3. If \(X_3\) would
+instead be a unit-level variable (which may vary within clusters), we
+could not just recover the unobserved `truth', but would need to use
+some kind of missing data method, or discard the incomplete units
+altogether (i.e., complete case analysis). Complete case analysis can
+however introduce bias in statistical inferences and lowers statistical
+power. Further, with the systematic missingness in \(X_2\), it would be
+impossible to fit a multilevel model without accommodating the
+missingness in some way. Complete case analysis in that case would mean
+excluding the entire cluster from the analyses. The wrong choice of
+missing data handling method can thus be extremely harmful to the
+inferences.
+
+Imputation of missing data requires consideration of the mechanism
+behind the missingness. Rubin proposed to distinguish between data that
+are missing completely at random (MCAR), data that are missing at random
+(MAR) and data that are missing not at random (MNAR; see Table
+\ref{tab:miss}). For each of these three missingness generating
+mechanisms, different imputation strategies are warranted
+(\citet{yuce08} and \citet{hox15}). We here consider the general case
+that data are MAR, and expand on certain MNAR situations.
+
+\hypertarget{imputation-with-mice}{%
+\subsection{imputation with mice}\label{imputation-with-mice}}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, breakable, colback=white, opacityback=0, arc=.35mm, rightrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomrule=.15mm, toprule=.15mm, breakable, leftrule=.75mm, left=2mm, colback=white, opacityback=0, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, breakable, colback=white, opacityback=0, arc=.35mm, rightrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomrule=.15mm, toprule=.15mm, breakable, leftrule=.75mm, left=2mm, colback=white, opacityback=0, rightrule=.15mm]
 
-Box 2. The predictor matrix.
+Box 3. The predictor matrix. The entries corresponding to the level-1
+predictors are coded with a 3, indicating that both the original values
+as well as the cluster means of the predictor are included into the
+imputation model. The entry of 4 in the predictor matrix adds three
+variables to the imputation model for the imputation model predictor:
+the value of the predictor, the cluster means of the predictor and the
+random slopes of the predictor.
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-4-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-12-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-15-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-14-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-17-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-15-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-18-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-16-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-19-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-17-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-20-1.pdf}
 

### 2023-12-11 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
update conclusion

@@
 Prior to implementing multiple imputation, it is critical to assess the robustness of the estimation model when confronted with missing data.
-\cite{Alisson_2002} highlighted scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods.
-Notably, when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}.
+There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods.
+For instance, when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}.
 Furthermore, depending on the proportion of missingness (usually is <5%), opting for a simpler complete case analysis might be sufficient.
@@
 ##### Clustering Inclusion
-Various strategies can be employed in the imputation process to account for clustering, including the inclusion of a cluster indicator variable, conducting a separate imputation process for each cluster, or employing a simultaneous imputation method that considers clustering (Stata: Clustering and MI Impute) TODO: Replace ref.
-
-The choice of strategy depends primarily on the assumptions made in the main analysis and the constraints imposed by the analyzed data.Concerning data restrictions, if there are few clusters with many observations per cluster, the cluster indicator or imputation separated by groups may be appropriate \cite{Graham_2009}.
+Various strategies can be employed in the imputation process to account for clustering, including the inclusion of a cluster indicator variable, conducting a separate imputation process for each cluster, or employing a simultaneous imputation method that considers clustering\cite{eddings}
+
+The choice of strategy depends primarily on the assumptions made in the main analysis and the constraints imposed by the analyzed data.
+Concerning data restrictions when the analysis do not contemplate an hierarchical model (eg.
+descriptive way), if there are few clusters with many observations per cluster, the cluster indicator or imputation separated by groups may be appropriate \cite{graham2009}.
 Conversely, an imputation model that simultaneously imputes all clusters using a hierarchical model can be employed when there are more clusters or fewer observations per cluster \cite{Allison_2002}.
@@
 Multiple imputation models based on hierarchical models have been proposed, each with different assumptions.
-
-<!--TODO: Add table 2l.imputation models -->
-
-Before choosing a particular strategy, it is crucial to verify if the assumptions of the imputation model align with those of the main analysis.
-For example, using the cluster indicator strategy may introduce bias estimates when the model is based on a hierarchical structure \cite{Taaljard_2008}.
-Even if an imputation strategy congruent with the main model is preferred, it is essential to assess its appropriateness for the data, considering that less complex imputation strategies may lead to unbiased estimates in certain scenarios \cite(Bailey_2020).
+\cite{audigier}
+
+In the particular case that the analysis contemplates the use of a hierarchical model (i.e.
+main model), before choosing a particular strategy, it is crucial to verify if the assumptions of the imputation model align with those of the main analysis.
+For example, using the cluster indicator strategy may introduce bias estimates when the model is based on a hierarchical structure \cite{taljaard2008,speidel2018}.
+Even if an imputation strategy congruent with the main model is preferred, it is essential to assess its appropriateness for the data, considering that less complex imputation strategies may lead to unbiased estimates in certain scenarios \cite{bailey2020}.
 
@@
 #### Model Specification
-The imputation model must be congenial with the main model \cite{Meng_1994}.
+The imputation model must be congenial with the main model \cite{meng1994}.
 Congeniality issues arise when the imputation model and the main model make different assumptions, often due to the omission of a polynomial or interaction term or the use of transformed variables.
@@
 The imputation model can include additional terms than the main model terms that do not lead to uncongeniality.
-For instance, it is advisable to include the outcome variable in the imputation model for prediction variables \cite{Moons}.
+For instance, it is advisable to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}.
 In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be included as a covariate in the imputation model [REF].
-Furthermore, the inclusion of auxiliary variables, even if not part of the main model, can be associated with the probability of missingness, enhancing the likelihood of satisfying the Missing at Random (MAR) assumption and improving estimation efficiency \cite{Hardt_2012}.
+Furthermore, the inclusion of auxiliary variables, even if not part of the main model, can be associated with the probability of missingness, enhancing the likelihood of satisfying the Missing at Random (MAR) assumption and improving estimation efficiency \cite{hardt2012a}.
 
 The specification of imputation models is done on a variable basis, either using a prediction matrix where the type of prediction variable should be specified accordingly (see table) or through a list of formulas in the formula parameter.
-The latter option proves useful in formulating imputation models with polynomial terms or interactions compared to the prediction matrix specification, which requires the inclusion of additional terms as Just Another Variable (JAV).
+The latter option proves useful in formulating imputation models with polynomial terms or interactions compared to the prediction matrix specification, which requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}.
 
@@
 
-However, even when considering an interaction term, depending on the adapted individual imputation model, it has been suggested, for instance, for treatment interaction effects, to conduct separated imputation by treatment group \cite{Zhang_2023}.
+However, even when considering an interaction term, depending on the adapted individual imputation model, it has been suggested, for instance, for treatment interaction effects, to conduct separated imputation by treatment group \cite{zhang2023}.
 Additionally, there are imputation models based on random forest or deep learning that can handle interaction and non-linear terms that do not require the explicit inclusion of the non-linear terms.
@@
 
-Before commencing the analysis, it is imperative to verify the convergence of the imputation process. This is typically achieved through trace plots or by evaluating prediction plots.
+Before starting into the analysis of each imputed dataset, it is crucial to validate the convergence of the imputation process.
+This is commonly accomplished through trace plots that depict the mean and variance of the incomplete variables across iterations.
+These plots serve to uncover potential circular issues or the need for additional iterations.
+Additionally,  it is also important to verify that imputed values fall within a plausible range and also to check the distribution of imputed variables, ensuring that the imputed variable distribution aligns with the distribution of observed values (under the MAR assumption).
+An alternative approach involves assessing the prediction accuracy of the imputation method \cite{cai2023}.
 
@@
 This can occur when missingness is associated with the incomplete value itself (self-marking) or when there is an unobserved variable linked to both the value and the probability of missingness of the incomplete value (indirectly non-informative).
-Specifically, for the indirectly non-informative case in hierarchical datasets, imputation methods based on the Heckman method, such as those proposed by Hammon (2021) and Munoz (2023), can be considered.
+Specifically, for the indirectly non-informative case in hierarchical datasets, imputation methods based on the Heckman method can be considered.\cite{hammon2020,hammon2022,munoz2023}
 
@@
 library(here)
+library(micemd)
 source(here("heckman","ipdfunction.R"),local =knitr::knit_global())
@@
 ```
-We are also contemplating the utilisation of the ???pmm??? option, as the values imputed using a fully parametric method may be implausibly low for some patients.
+We are also contemplating the utilisation of the predictive mean matching (pmm) option, as the values imputed using a fully parametric method may be implausibly low for some patients.
 
@@
 The selection model approach includes methods such as the Heckman model, which can be particularly useful in this case.
-Several methods, including those by \cite{Galimar_2017,Hammon_2021}, and the recently a Heckman method designed for two-level data, allow for variations in intercepts and exposure effects (random intercept and slope).
+Several methods, including those by \cite{galimard2018}, and the recently a Heckman method designed for two-level data, allow for variations in intercepts and exposure effects (random intercept and slope) \cite{munoz2023}.
 
@@
 These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution.
-For a more comprehensive understanding of the model and the exclusion restriction, see \cite{Munoz_2022}.
+For a more comprehensive understanding of the model and the exclusion restriction, see \cite{munoz2023a}.
 
@@
 
-# Additional
-
-The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)
-
-<!-- [TODO: add reference Arxiv paper Johanna].  -->
-
-Visualize missing data pattern:
-
-The matrix only shows the predictors for the main model, not the selection model.
-
-TODO: explain exclusion restriction.
-
-ORDER:
-
--   summary
-
--   congeniality, then in hierarchical models
-
--   look whether we can fit cong. back in the main body
-
--   alt. methods
-
--   conclusion: mice is really easy!
-
--   Additional levels of clustering
-
--   More complex data types: timeseries and polynomial relationship in the clustering.
-
--   FIML vs MI
-
-An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML).
-This method does not require the imputation of any missing values.
-Whereas MI consists of imputation, analyses and pooling steps, FIML analyses the data in a single step.
-When the assumptions are met the two approaches should produce equivalent results.
-[REF] As FIML requires specialised software, not all analyses can be performed with standard software.
-[REF]
-
--   Survival / TTE, this could be put in the paragraph on congeniality
-
-When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]
-
-<!-- - JOMO in \pkg{mice} -> on the side for now -->
-<!-- # look at jomo for categorical variables? -->
-<!-- # semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach -->
-<!-- # pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters -->
-<!-- # make assumptions of these methods explicit! -->
-ORDER:
-
-- summary
-- congeniality, then in hierarchical models
-- look whether we can fit cong. back in the main body
-- alt. methods
-- conclusion: mice is really easy!
-
-- Additional levels of clustering
-
-- More complex data types: timeseries and polynomial relationship in the clustering.
-
-- FIML vs MI
-
-An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML).
-This method does not require the imputation of any missing values.
-Whereas MI consists of  imputation, analyses and pooling steps, FIML analyses the data in a single step.
-When the assumptions are met the two approaches should produce equivalent results.
-[REF] As FIML requires specialised software, not all analyses can be performed with standard software.
-[REF]
-
-- Survival / TTE, this could be put in the paragraph on congeniality
-
-When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]
-
-<!-- # Think about -->
-
-<!-- - Adding evaluations of the imputations such as convergence checks -->
-
-<!-- TODO in the future: Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model. -->
-
-<!-- TODO in the future: Adding a `multilevel_ampute()` wrapper function in mice. -->
-
-<!-- - Exporting `mids` objects to other packages like `lme4` or `coxme`?  -->
-
-<!-- - Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt. -->
-
-<!-- - Show use case for deductive imputation for cluster level variables? -->
-
-<!-- - env dump in repo -->
-
-<!-- -  I don't know if in your article you cover something about model complexity, for example sometimes I have to switch from 2l.
-methods to 1l.
-methods just because the model didn't converge..
-this is due to the considered imputation model is very complex regarding the amount of information counted...
-I know that a solution for an imputation model with many predictors is to check correlation plots as you did or use quickpred()..
-(maybe you can add this somewhere after the correlation plots).....But as for cluster specification, I don't know if besides plots of distribution per cluster there is something else can be done to see if i have to use 1l.
-or 2l.
-for a given variable, also I have no idea..
-how to test which is better between 2l.norm or 2l.2stage.norm.
--->
-
-In hierarchical datasets, clustering is a concern because the homoscedasticity in the error terms cannot be assumed across clusters and the relationship among variables may vary at different hierarchical levels.
-When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g.
-if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021).
-Not including clustering into the imputation process may lead to effect estimates with smaller standard errors and inflated type I error.
-
-There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.
-
-The selection of each strategy depends mainly on the assumptions in the main analysis and also on the restriction of the analyzed data.
-
-Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
-The last restriction is also required when imputations are performed on each cluster separately.
-When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).
-
-Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
-However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.
-the number of clusters is adequate.
-(Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008).
-Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020).
-For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
-(Zhang,2023).
-
 # Conclusion
+This paper is dedicated to exploring the imputation process for incomplete datasets, with a primary focus on utilizing a hierarchical model for analysis.
+Initially, users are encouraged to consider the main analysis within the context of the incomplete dataset, along with insights provided by domain experts, to gain a better understanding of variable relationships.
+Employing clear data visualization is instrumental in comprehending the missing data patterns, establishing a missing mechanism, and aiding in the selection of suitable imputation methods.
+
+The "Mice" and "mice"-based R packages offer a range of imputation methods tailored for hierarchical data, easily adaptable to the dataset's structure.
+Before proceeding with the analysis of the imputed dataset, it is essential to assess the convergence of the imputation method.
+This evaluation can reveal issues such as circular problems, the need for additional iterations, or challenges associated with the chosen imputation method.
 
@@
 # References
-
-<!-- Comment dump -->
-<!-- Missing data pattern with clustering -->
-<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
-<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
-<!-- TODO: check spss missing data pie chart? -->
-<!-- TODO: similarity of patterns over clusters -->
-<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
-<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
-<!-- TODO: look at vim -->
-<!-- TODO: check if shepley plot is an option -->
-
-<!-- Conditional plots missingness indicator -->
-<!-- TODO: check if up-side-down plot works -->
-<!-- TODO: check smoothing in geom_density function and make it the inverse of the sample size -->
-<!-- TODO: add functions to mice -->
-<!-- TODO: add title and informative legend (pop obs and pop miss) -->
-<!-- TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .` -->
-<!-- TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon -->
-<!-- TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?) -->
-
-# Appendix
-
-Table 3: Notation
-
-| **Formula \pkg{lme4}**         | **Details**                                 |
-|-------------------------------------------------------------|--------------------------------------------------|
-| `y ~ x1 + (1 | g1)`                                      | Fixed `x1` predictor with random intercept         |
-|                                                             | varying among `g1`                                  |
-| `y~x1*x2+ (1| g1)`                                       | Interactions of `x1` and `x2` only in fixed effect   |
-| `y ~ x1*x2+ (x2| g1)`                                      | Interactions of `x1` and `x2` only in fixed effect   |
-|                                                             | with slope of `x2` randomly varying among `g1`       |
-| `y ~ x1*x2+ (x1*x2| g1)`                                   | variance-covariance matrix estimated only with   |
-|                                                             | the variance terms of intercept, slope of `x1`,    |
-|                                                             | slope of `x2` and interaction `x1*x2`                |
-| `y ~ x1*x2+ (x1 | g1)+ (x2| g1)`                            | variance-covariance matrix estimated separately, |
-|                                                             | i.e, one for intercept and `x1` and another for    |
-|                                                             | intercept and `x2`                                 |
-| `y ~ x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)`             | Fixed `x1` with correlated random intercept and    |
-|                                                             | random slope of `x`                                |
-| `y ~ x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1)` | Fixed `x1` with uncorrelated random intercept      |
-|                                                             | and random slope of `x1`                           |
-| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2   |
-| `y ~ (1 | g1/g2) or ~ (1 | g1) + (1 | g1:g2)`                     | Random intercept varying among `g1` and `g2`         |
-|                                                             | withing `g1`                                       |
-|                                                             |          |

### 2023-12-11 docs/manuscript.qmd
Add complete case study 1 and 2, and text of 3

@@
         keep-tex: true
-        fig-pos: 't'
+        fig-pos: 'h'
         fig-width: 7
-        fig-height: 3.5
+        fig-height: 2.5
     jss-html: default
@@
 
+In hierarchical datasets, clustering is a concern because the homoscedasticity in the error terms cannot be assumed across clusters and the relationship among variables may vary at different hierarchical levels.
+When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g.
+if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021).
+Not including clustering into the imputation process may lead to effect estimates with smaller standard errors and inflated type I error.
+
 ```{r}
@@
 #| echo: false
+#| warning: false
+#| message: false
+library(knitr)
 library(ggplot2)
 source("../docs/utils.R")
+source("../docs/tables.R")
 ```
@@
 #| echo: false
-concepts <-
-  data.frame(
-    Concept = c(
-      "Sample units",
-      "Cluster",
-      "Hierarchical data",
-      "Level-1",
-      "Level-2",
-      "Hierarchical model",
-      "Fixed effect",
-      "Random effect",
-      "Mixed effect",
-      "ICC",
-      "Stratified intercept"
-    ),
-    Details = c(
-      "Units of the population from which measurements are taken in a sample, e.g., students.",
-      "Variable that specify the cluster or agruppation, e.g., Classroom",
-      "Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.",
-      "Variable that varies within a cluster, eg. Test score",
-      "Variable that does not vary within a cluster but between, e.g. teacher experience.",
-      "Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)" ,
-      "Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)",
-      "Effects that are a source of random variation in the data, and whose levels are not fully sampled.
-e.g.
-test score tendency during academic year between students due to no controlled factors such as  genetic,family history",
-      "Includes fixed and random effects, e.g.
-the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
-Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
-In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
-@gelm06, @hox17 and @jong21).",
-      "The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
-of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
-""
-    )
-  )
 # concepts |>
@@
 #| echo: false
+#| fig-height: 1.5
 plot_na()
@@
 The entry of 4 in the predictor matrix adds three variables to the imputation model for the imputation model predictor: the value of the predictor, the cluster means of the predictor and the random slopes of the predictor.
+- -2 = cluster variable 
+- 1 = overall effect
+- 3 = overall + group-level effect 
+- 4 = individual-level (random) and group-level (fixed) effect 
 :::
+
+## Multilevel imputation workflow {#sec-workflow}
+
+There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.
+
+The selection of each strategy depends mainly on the assumptions in the main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
+The last restriction is also required when imputations are performed on each cluster separately.
+When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
+However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.
+the number of clusters is adequate.
+(Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008).
+Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020).
+For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
+(Zhang,2023).
+
+Below we provide a imputation workflow that can be used in general to impute cluster data.
+
+### Focus on the Main Analysis
+When imputing clustered data, we initially focus on the research question and the intended analysis, assuming the presence of non-incomplete values.
+This assessment sheds light on research hypotheses and the contemplated main(s) statistical model(s) and it provides insights into the data's structure (refer to the level table), variable types (e.g., confounders, auxiliary variables), and other considerations about variables relationships such as interactions or polynomial terms.
+
+### Exploration of Available Data
+Next, it is necessary to explore the information available in the data.
+Exploring individual variables using tools such as histograms or QQ plots can help to delineate variable distributions and plausible ranges of values and also identify input errors or outliers.
+Evaluating interactions between predictors using scatter plots or conditional plots, assessing collinearity using VIF or correlations, can help to glimpse nonlinear relationships that may affect the original formulation of the main model.
+Further testing of assumptions related to the response variable, including variance (homogeneity using conditional box plots), independence (e.g., ACF, variograms) and response vs.
+predictor relationships (conditional plots - x vs.
+y), may serve to choose an imputation model more suitable to the data.
+In addition, the intraclass correlation coefficient (ICC) can be examined to assess cluster differences, aiding in the choice between the 2l and 1l methods for imputation.
+
+One can also explore missingness, specifically missing patterns at the cluster level to identify types of patterns (e.g., non-monotonic, systematic) to guide the selection of an appropriate imputation approach in terms of computational efficiency (e.g., simpler regression imputation versus FCS in univariate patterns).
+Similarly, these missing patterns can also be used to identify potential predictors in individual imputation models by means of inflow criteria (quantifying the connections between missing data in one variable and other observed variables) and outflow criteria (determining the connections between observed values in one variable and missing data in other variables).
+Although several packages exist for visualizing missing data, this tutorial focuses on the recently implemented ggmice package.
+
+### Assess Estimation Procedure Robustness to Missing Data
+Prior to implementing multiple imputation, it is critical to assess the robustness of the estimation model when confronted with missing data.
+There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods.
+For instance, when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}.
+Furthermore, depending on the proportion of missingness (usually is <5%), opting for a simpler complete case analysis might be sufficient.
+
+### Pre-imputation
+Before proceeding with the imputation model, it is necessary to filter the dataset to include only the minimal variables required for the main model.
+If additional procedures are conducted during the analysis, it is essential to include variables associated with these procedures, such as confounders in the case of balancing techniques.
+Other relevant variables, like instrumental variables or auxiliary variables, may enhance parameter estimates even if not included in the main model, as they could be linked to the probability of missingness for some incomplete variables.
+
+Furthermore, at the variable level, it is crucial to assess whether proxy variables should be considered or if deterministic imputation alone is adequate for imputing missing values.
+For instance, certain incomplete variables may not necessitate stochastic imputation methods like MICE.
+Instead, they can be effectively addressed through deductive imputation, where incomplete values are inferred from logical and deterministic relationships between variables.
+This approach is especially beneficial for variables that are functions of others, such as deriving a person's BMI from their weight and height.
+It also proves valuable for determining values for one-level variables from two-level ones, as seen in the context of Individual Participant Data (IPD).
+In such cases, incomplete information can be deduced from metadata, like inferring incomplete data about abortion in a country where abortion is illegal, or through cross-temporal or protocol-based deduction, such as imputing missing test values for deceased patients.
+
+### Setting Imputation Model
+
+##### Clustering Inclusion
+Various strategies can be employed in the imputation process to account for clustering, including the inclusion of a cluster indicator variable, conducting a separate imputation process for each cluster, or employing a simultaneous imputation method that considers clustering\cite{eddings}
+
+The choice of strategy depends primarily on the assumptions made in the main analysis and the constraints imposed by the analyzed data.
+Concerning data restrictions when the analysis do not contemplate an hierarchical model (eg.
+descriptive way), if there are few clusters with many observations per cluster, the cluster indicator or imputation separated by groups may be appropriate \cite{graham2009}.
+Conversely, an imputation model that simultaneously imputes all clusters using a hierarchical model can be employed when there are more clusters or fewer observations per cluster \cite{Allison_2002}.
+
+In a hierarchical imputation model, correlations between observations within clusters are modeled by a random effect, allowing estimation even with a limited number of observations per cluster.
+Multiple imputation models based on hierarchical models have been proposed, each with different assumptions.
+\cite{audigier}
+
+In the particular case that the analysis contemplates the use of a hierarchical model (i.e.
+main model), before choosing a particular strategy, it is crucial to verify if the assumptions of the imputation model align with those of the main analysis.
+For example, using the cluster indicator strategy may introduce bias estimates when the model is based on a hierarchical structure \cite{taljaard2008,speidel2018}.
+Even if an imputation strategy congruent with the main model is preferred, it is essential to assess its appropriateness for the data, considering that less complex imputation strategies may lead to unbiased estimates in certain scenarios \cite{bailey2020}.
+
+#### Choice of Individual Imputation Methods
+The initial step involves selecting the individual imputation model for each incomplete variable in the dataset.
+The mice package automatically proposes imputation methods based on the variable type for 1l variables.
+For 2l level variables, the micemd package's () function can be utilized.
+This function selects among the 2l imputation methods based on the size of the clusters and the proportion of missingness in each cluster.
+
+In addition to the package-defined imputation methods, users can specify custom methods using the "I formula".
+This flexibility allows for the calculation of derived variables internally or adjustments to imputation methods based on specific conditions, such as conditioning the imputation model on the level of an incomplete covariate (e.g., pregnancy test on females).
+
+#### Model Specification
+The imputation model must be congenial with the main model \cite{meng1994}.
+Congeniality issues arise when the imputation model and the main model make different assumptions, often due to the omission of a polynomial or interaction term or the use of transformed variables.
+
+The imputation model can include additional terms than the main model terms that do not lead to uncongeniality.
+For instance, it is advisable to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}.
+In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be included as a covariate in the imputation model [REF].
+Furthermore, the inclusion of auxiliary variables, even if not part of the main model, can be associated with the probability of missingness, enhancing the likelihood of satisfying the Missing at Random (MAR) assumption and improving estimation efficiency \cite{hardt2012a}.
+
+The specification of imputation models is done on a variable basis, either using a prediction matrix where the type of prediction variable should be specified accordingly (see table) or through a list of formulas in the formula parameter.
+The latter option proves useful in formulating imputation models with polynomial terms or interactions compared to the prediction matrix specification, which requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}.
+
+However, even when considering an interaction term, depending on the adapted individual imputation model, it has been suggested, for instance, for treatment interaction effects, to conduct separated imputation by treatment group \cite{zhang2023}.
+Additionally, there are imputation models based on random forest or deep learning that can handle interaction and non-linear terms that do not require the explicit inclusion of the non-linear terms.
+
+### Post-Imputation
+
+During the imputation process, certain issues may arise that halt the process.
+In hierarchical model imputations, many issues are linked to methods based on parametric hierarchical models, which may struggle to estimate a model midway through the process.
+In such cases, it is advisable to examine the imputation log file to identify variables with imputation issues.
+Subsequently, reducing the number of predictors in the imputation model could be explored, either by using functions like quickpred or by considering variable transformations, such as scaling.
+Adjusting the level of the hierarchical model (e.g., using a homogenous variance model or a 1l model) can also be beneficial.
+Additionally, checking the range of imputed variables is crucial, as excessively large imputed values for one predictor could lead to convergence issues in other variables.
+This can be mitigated by including post-processing specifications on problematic variables or by using imputation models like Predicted Mean Matching (PMM), ensuring imputed values align with the observable values.
+In certain cases, a separate imputation strategy may be considered.
+For instance, in analyses involving multiple endpoints, conducting separate imputation processes for each endpoint might be preferable to a unified imputation process.
+
+### Convergence and Sensitivity Analysis
+
+Before starting into the analysis of each imputed dataset, it is crucial to validate the convergence of the imputation process.
+This is commonly accomplished through trace plots that depict the mean and variance of the incomplete variables across iterations.
+These plots serve to uncover potential circular issues or the need for additional iterations.
+Additionally,  it is also important to verify that imputed values fall within a plausible range and also to check the distribution of imputed variables, ensuring that the imputed variable distribution aligns with the distribution of observed values (under the MAR assumption).
+An alternative approach involves assessing the prediction accuracy of the imputation method \cite{cai2023}.
+
+While the majority of Multiple Imputation by Chained Equations (MICE) methods are based on Missing at Random (MAR) assumptions, field expert input may suggest that the Missing Not at Random (MNAR) mechanism could be plausible for certain variables.
+An MNAR variable is one in which the probability of missingness depends on an unobservable variable.
+This can occur when missingness is associated with the incomplete value itself (self-marking) or when there is an unobserved variable linked to both the value and the probability of missingness of the incomplete value (indirectly non-informative).
+Specifically, for the indirectly non-informative case in hierarchical datasets, imputation methods based on the Heckman method can be considered.\cite{hammon2020,hammon2022,munoz2023}
+
+---
 
@@
 #| message: false
-set.seed(123)
-library(mice)
-library(ggmice)
-library(ggplot2)
-library(miceadds)
-library(lme4)
-library(mitml)
-library(broom.mixed)
+set.seed(123)         # for reproducibility
+library(lme4)         # for multilevel modeling
+library(mice)         # for imputation
+library(miceadds)     # for multilevel imputation methods
+library(micemd)       # for selection-model imputation methods
+library(mitml)        # for multilevel parameter pooling
+library(broom)        # for clean model estimates
+library(broom.mixed)  # for multilevel model estimates
+library(dplyr)        # for data wrangling
+library(ggmice)       # for visualization
+library(ggplot2)      # for visualization
 ```
@@
 
+In this section we will go over the different steps involved with imputing incomplete multilevel data with the R package mice.
+We consider the simulated `popmis` dataset, which included pupils ($n = 2000$) clustered within schools ($N = 100$).
+The following variables are of primary interest:
+
+  - `school`,     school identification number (clustering variable);
+  - `popular`,    pupil popularity (self-rating between 0 and 10; unit-level);
+  - `sex`,	      pupil sex (0 = boy, 1 = girl; unit-level);
+  - `texp`,	      teacher experience (in years; cluster-level).
+
+The analysis model corresponding to this dataset is multilevel regression with random intercepts for the different schools.
+We will estimate the association between the pupils' sex and their popularity score.
+This model can be expressed in `lme4` code as:
+
+```r
+popular ~ 1 + sex + (1 | school)
+```
+
+We load the data into the environment with
 ```{R}
@@
 
+and select the relevant variables
 ```{R}
 #| prompt: true
-dat <- popmis[, c("school", "teachpop", "popular", "texp", "sex")] 
+dat <- popmis[, c("school", "popular", "texp", "sex")] 
 ```
 
+which results in the following data structure.
+```{r}
+#| prompt: true
+head(dat)
+```
+
+The association of interest can be visualized with `ggmice`,
 ```{R}
-#| fig-cap: "Polar axis plot"
-#| fig-alt: "A line plot on a polar axis"
-ggmice(dat, aes(popular, teachpop)) + 
+#| fig-cap: "Scatterplot of student popularity by sex"
+#| fig-alt: ""
+ggmice(dat, aes(sex, popular)) +
   geom_jitter()
 ```
+where missing datapoints in the `popular` variable are represented by red points on the X-axis of the figure.
 
-With the `ggmice` unction `plot_pattern` we can visualize this.
-
+With the `ggmice` function `plot_pattern` we can visualize the missing data pattern 
 ```{R}
@@
 #| fig-cap: "Missing data pattern."
-#| fig-pos: 't'
 plot_pattern(dat)
@@
 
-```{R, eval=FALSE}
-#| layout-ncol: 2
-#| echo: false
-# plot_pattern(dat)
-# plot_corr(dat)
-```
+which shows us that the missingness is univariate and sporadic.
 
+To develop the best imputation model for the incomplete variable `popular`, we need to know whether the observed values of `popular` are related to observed values of other variables.
+Plot the pair-wise complete correlations in the incomplete data
 ```{R}
@@
 ```
+
+This shows us that  not just the analysis-model variable `sex`, but also the cluster-level covariate teacher experience, `texp`, may be a useful as an imputation model predictor.
+Moreover, the missingness in `popular` may depend on the observed values of other variables.
+With `ggmice()` we can visualize the distribution of the teacher experience for cases where `popular` is observed and cases where `popular` is missing.
+```{r}
+#| prompt: true
+ggmice(dat, aes(texp)) + 
+  geom_density() +
+  facet_wrap(~is.na(popular))
+```
+
+It appears that students with a missing value for `popular` are in clusters with a slightly higher `texp` value.
+```{r}
+t.test(dat$texp ~ is.na(dat$popular)) |> 
+  tidy() |>
+  kable()
+```
+
+Although there are no significant differences in the distribution of `texp` depending on the missingness indicator of `popular`, this variable can serve as auxiliary variable in the imputation of `popular`.
 
@@
 ```
+
+The `pmm` method is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters.
 
@@
 
-Evaluate the distribution of imputed values.
+Troubleshoot non-convergence of imputation model. Evaluate the distribution of imputed values.
 ```{R}
+#| prompt: true
 ggmice(imp, aes(popular, group = .imp)) + 
@@
 ```{R}
+#| prompt: true
 ggmice(imp, aes(.imp, popular)) + 
@@
 ```{R}
-ggmice(imp, aes(popular, teachpop)) + 
+#| prompt: true
+ggmice(imp, aes(sex, popular)) +
   geom_jitter() +
@@
 ```{R}
+#| prompt: true
 fit <- with(
   imp,
-  lmer(teachpop ~ popular + texp + (1 | school))
+  lmer(popular ~ texp + sex + (1 | school))
 )
@@
 
-Pool the estimates.
+Pooling the estimates does not provide estimates of the variance components.
 ```{R}
+#| prompt: true
 pool(fit)
@@
 
+Therefore, `mitml` is used.
+```{R}
+#| prompt: true
+est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+```
+
 Display results in table.
-```{R}
-testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+```{r}
+#| prompt: true
+est$estimates |> 
+  round(3) |>
+  kable()
 ```
+
+```{r}
+#| prompt: true
+est$extra.pars |> 
+  round(3) |>
+  kable()
+```
+
+### IMPACT data
+
+The second case study is the `impact` data from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc].
+
+The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:
+
+  - `name` Name of the study,
+  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
+  - `age` Age of the patient,
+  - `motor_score` Glasgow Coma Scale motor score,
+  - `pupil` Pupillary reactivity,
+  - `ct` Marshall Computerized Tomography classification, 
+  - `hypox` Hypoxia (0=no, 1=yes),
+  - `hypots` Hypotension (0=no, 1=yes),
+  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+  - `edh` Epidural hematoma (0=no, 1=yes),
+  - `mort` 6-month mortality (0=alive, 1=dead).
+
+Check if there is systematic missingness in this dataset. For illustration purposes, we made Marshall Computerized Tomography classification (ct) systematically missing.
+
+The analysis model for this dataset is a prediction model with `mort` as the outcome.
+In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes.
+The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model:
+```r 
+mort ~ type + age + motor_score + pupil + ct + (1 | name) 
+```
+
+Note that variables `hypots`, `hypox`, `tsah` and `edh` are not part of the analysis model, and may thus serve as auxiliary variables for imputation.
+
+The `impact` data included in the \pkg{metamisc} package is a complete data set.
+The original data has already been imputed once (Steyerberg et al, 2008).
+For the purpose of this tutorial we have induced missingness (mimicking the missing data in the original data set before imputation).
+The resulting incomplete data can be accessed from [zenodo link to be created](https://zenodo.com).
+
+Load the incomplete data into the R workspace:
+```{r data, eval = FALSE}
+#| prompt: true
+dat <- read.table("link/to/the/data.txt") 
+```
+
+```{r data-actually, include=FALSE}
+#| prompt: true
+dat <- readRDS("../Data/impact_incomplete.RDS")
+```
+
+To explore the missingness, we should look at the missing data pattern. The ten most frequent missingness patterns are shown with
+```{r}
+#| prompt: true
+#| fig-height: 4
+plot_pattern(dat, rotate = TRUE, npat = 10L)  
+```
+
+This shows that we need to impute `ct` and `pupil`.
+To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables.
+To see whether the missingness depends on the observed values of other variables, we can test this statistically or use visual inspection (e.g.
+a histogram faceted by the missingness indicator).
+
+We should impute the variables `ct` and `pupil` and any auxiliary variables we might want to use to impute these incomplete analysis model variables.
+We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations
+```{r}
+#| prompt: true
+#| fig-height: 4
+plot_corr(dat, rotate = TRUE)
+```
+
+This shows us that `hypox` and `hypot` would not be useful auxiliary variables for imputing `ct`.
+Depending on the minimum required correlation, `tsah` could be useful, while `edh` has the strongest correlation with `ct` out of all the variables in the data and should definitely be included in the imputation model.
+For the imputation of `pupil`, none of the potential auxiliary variables has a very strong relation, but `hypots` could be used.
+We conclude that we can exclude `hypox` from the data, since this is neither an analysis model variable nor an auxiliary variable for imputation
+```{r}
+#| prompt: true
+dat <- select(dat, !hypox)
+```
+
+Mutate data to get the right data types for imputation (e.g. integer for clustering variable).
+```{r}
+#| prompt: true
+dat <- dat |>
+  mutate(name = as.integer(name))
+```
+
+Create an initial methods vector for the incomplete variables
+```{r}
+#| prompt: true
+meth <- make.method(dat)
+meth
+```
+
+which should be adjusted to the appropriate `2l` methods.
+```{r}
+#| prompt: true
+meth[c("pupil", "ct")] <- "2l.pmm"
+meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
+```
+
+Create an initial predictor matrix 
+```{r}
+#| prompt: true
+pred <- quickpred(dat)
+```
+
+This predictor matrix is too large to display inline. A visualization of the adapted predictor matrix is presented in Figure XYZ.
+
+We should make sure `name` is used as clustering variable
+```{r}
+#| prompt: true
+pred[, "name"] <- -2
+```
+
+and the analysis-model outcome should be used as a predictor in all imputation models
+```{r}
+#| prompt: true
+pred[, "mort"] <- 2
+```
+
+the resulting predictor matrix is visualized with
+```{r}
+#| prompt: true
+#| fig-height: 4
+plot_pred(pred, method = meth, rotate = TRUE)
+```
+
+Impute the incomplete data with
+```{r}
+#| prompt: true
+# imp <- mice(
+#   dat,
+#   method = meth,
+#   predictorMatrix = pred,
+#   printFlag = FALSE
+# )
+```
+
+Evaluate the convergence of the algorithm
+```{r}
+#| prompt: true
+# plot_trace(imp)
+```
+
+Analyze the imputed data:
+```{r}
+#| prompt: true
+# fit <- imp %>% 
+#   with(glmer(
+#     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
+#     family = "binomial"
+#     )) 
+```
+The estimated effects after imputation are presented in Table XYZ.
+
+### obesity data
+
+In this example, we demonstrate a multilevel imputation of random intercept and random slope model with a continuous response.
+We utilize the obesity dataset included in the `micemd`\@ package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries.
+We simulate data for 5 clusters so that the true values are known.
+We use the following variables from the dataset:
+
+-   **Cluster:** Region of the patients' healthcare provider (Cluster variable),
+-   **Gender:** Subjects' Gender (0=male, 1=female),
+-   **Age:** Subjects' age,
+-   **Height:** Subjects' height in metres,
+-   **Weight:** Subjects' weight in kilograms,
+-   **BMI:** Subjects' body mass index,
+-   **FamOb:** Family obesity history (yes or no),
+-   **Time:** Response time in minutes (exclusion-restriction variable).
+
+In this dataset, Age and FamOb are MAR, while the weight variable is affected by selection bias, attributed to an indirect MNAR mechanism.
+This MNAR mechanism typically arises when an unobserved or omitted variable influences both the value of the incomplete variable (in this case, Weight) and its likelihood of being missing (denoted as R).
+
+In the primary analysis model, BMI serves as the dependent variable, with Age, Gender, and FamOb as predictors.
+Because of the clustered nature of the data, which is quantified with the Intraclass Correlation Coefficient (ICC) below, we include random intercepts, as well as a random slope for the Age variable.
+The model is represented as:
+\begin{equation}
+\label{eqn:main}
+BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}
+\end{equation}
+
+We start by loading the data:
+```{r}
+#data("data_heckman", package = "micemd")
+#dat <- data_heckman
+```
+
+Now, let's begin by examining the missing patterns in the data by cluster:
+
+```{r, dpi = 300,out.width="70%", fig.cap="Missing pattern", eval=TRUE}
+# myplots <- lapply(1:5, function(i) {
+#   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
+#   ggplot2::ggtitle(paste0("Cluster", i))
+#  })
+# ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
+```
+
+We observe that the missing pattern is non-monotonic and quite similar across the clusters.
+However, regarding the weight variable, we notice that is systematically missing in cluster 3.
+In order to evaluate
+if we require a imputation method that accounts for clustering  we assess the Intraclass Correlation
+
+```{r obesity-icc, eval=TRUE}
+# Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+# performance::icc(Nulmodel)
+```
+
+Since the ICC is above 0.1 and as the main analysis will be use a mixed model, we decide
+to use two-level (2l) imputation methods. In this imputation process, we include all predictor
+variables from equation \ref{eqn:main} in the main model. However, since BMI is a composite of
+weight and height, we use deterministic imputation for these, which is described below.
+
+We use the **find.defaultMethodfunction** provided in the **micemd** package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters.
+
+It suggests using '2l.2stage.bin' for the FAV variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
+'2l.2stage.pmm'. For the BMI variable, we employ deterministic imputation.
+
+```{r obesitymeth, eval=TRUE}
+# meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+#                                ni.small = 100, prop.small = 0.4)
+# meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
+# meth_mar["Age"]<-"2l.2stage.pmm" 
+```
+
+For these imputation models, it is necessary to specify the prediction matrix, with the cluster
+variable labelled as -2 and the predictor variable measured within clusters labelled as 2, encompassing all variables.
+We need to suprime the variable Time as this variable is not specified in the main model.
+We also  modify the relationship between BMI, weight and height in the prediction matrix to avoid circular predictions.
+Then we proceed to run the imputation model.
+```{r obesity-predmar,out.width="70%", eval=TRUE}
+# pred_mar <- mice(Obesity, maxit = 0)$pred 
+# pred_mar[,"Cluster"] <- -2 # clustering variable
+# pred_mar[,"Time"] <- 0
+# pred_mar[pred_mar==1] <- 2
+# pred_mar[c("Height", "Weight"), "BMI"] <- 0
+# ggmice::plot_pred(pred_mar)
+# imp_mar <- mice::mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+#                 m=10, seed = 123, printFlag = FALSE)
+```
+
+```{r obesity-predmar1,eval=TRUE}
+# summary(complete(imp_mar,"long")$Weight)
+```
+We are also contemplating the utilisation of the predictive mean matching (pmm) option, as the values imputed using a fully parametric method may be implausibly low for some patients.
+
+```{r obesity-predmar_pmm, eval=TRUE}
+# meth_mar["Weight"]<-"2l.2stage.pmm" 
+# imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+#                     m=10, seed = 123, printFlag = FALSE)
+```
+
+```{r obesity-predmar_pmm1,eval=TRUE}
+# summary(complete(imp_mar_pmm,"long")$Weight)
+# ggmice::plot_trace(imp_mar_pmm, "Weight")
+```
+
+After confirming convergence, we proceed to save the results for future use.
+We consider the possibility that patients may not have been selected randomly, which would then have led to a distribution for weight that does not reflect the weight in the population.
+It???s likely that an omitted variable, like self-esteem, could influence this selection.
+For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.
+
+To address this situation, two approaches have been proposed for dealing with Missing Not at Random (MNAR) data: pattern-mixed models and selection models.
+Within pattern-mixed models, methods like the delta method and more advanced ones like NARFS have been suggested.
+The selection model approach includes methods such as the Heckman model, which can be particularly useful in this case.
+Several methods, including those by \cite{galimard2018}, and the recently a Heckman method designed for two-level data, allow for variations in intercepts and exposure effects (random intercept and slope) \cite{munoz2023}.
+
+To apply the **2l.2stage.heckman** method, the weight variable should be specified as '2l.2stage.heckman' found in the micemd package.
+Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.
+For the outcome equation we consider the same imputation model that we used for
+the MAR case (main model).
+ $$Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}$$ 
+Regarding the selection equation, we include the same predictors as those in the main model, as well as a time variable.
+Here the time variable serves as a restriction exclusion variable specifically explaining the probability of being observed but not affecting the incomplete value (Weight).
+In this context, we assume that the time a user spends completing the survey serves as a proxy for the barriers they may encounter in survey completion, such as familiarity with the survey content or internet speed.
+These factors may lead the user to skip specific questions or even the entire survey.
+Also, we assume the time does not have any influence on the subject???s weight.
+ $$R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}$$
+
+These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution.
+For a more comprehensive understanding of the model and the exclusion restriction, see \cite{munoz2023a}.
+
+To use information from both equations, we must adjust the prediction matrix.
+The cluster variable remains specified as before (-2).
+In this imputation method, all the variables present in both the selection and outcome equations are included with a random effect.
+
+However, it is essential to distinguish which of these variables appear in each equation.
+In this framework, when a variable is shared between both equations, it is denoted as (2).
+Predictors exclusive to the outcome equation are indicated as (-4), while those exclusive to the selection equation are labelled as (-3).
+Consequently, the only alteration needed in the predictor matrix pertains to the variable 'Time'.
+
+```{r obesity-predmnar, eval=TRUE}
+# pred_mnar <- pred_mar
+# pred_mnar["Weight","Time"]<- -3
+# ggmice::plot_pred(pred_mnar)
+```
+
+We also need to modify the method of the weight variable.
+
+```{r obesity-predmnar1, eval=TRUE}
+# meth_mnar <- meth_mar
+# meth_mnar["Weight"]<- "3l.2stage.heckman"
+```
+Then we proceed to run the imputation model as before, after executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
+```{r obesity-predmnar2,eval=TRUE}
+# imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+#                 m=10, seed = 123, printFlag = FALSE)
+# summary(complete(imp_mnar,"long")$Weight)
+```
+
+Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive).
+
+```{r obesity-predmnar3,eval=TRUE}
+# summary(complete(imp_mnar,"long")$Weight)
+```
+Consequently, as before we  use the 'pmm', option but this time for the Heckman imputation, this approach ensures that the imputed values remain within the range of observable values.
+We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by setting the pmm parameter to true.
+
+```{r obesity-predmnarp, eval=TRUE}
+# imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+#                      m=10, seed = 123, pmm = T,  printFlag = FALSE)
+```
+
+We check the convergency of the results
+```{r obesity-predmnarp1, eval=TRUE}
+# summary(complete(imp_mnar_pmm,"long")$Weight)
+# ggmice::plot_trace(imp_mar_pmm, "Weight")
+```
+
+After this modification we proceed to compare the effects on the model. We
+run the analysis model on each of the completed datasets as well asthe dataset where the
+incomplete values are removed (Complete Case analysis, CC).
+
+```{r models, eval=TRUE}
+# cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
+#              lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
+# mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+# mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+# mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+# mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+# list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
+# plot_models(list_models,
+#             mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
+```
+
+We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration.
+An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
+Nonetheless, with respect to precision of estimates, we notice that in general MNAR imputation leads to wider confidence intervals, in this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on significant test and therefore lead to contradictory conclusions.
+
+## Conclusion
+This paper is dedicated to exploring the imputation process for incomplete datasets, with a primary focus on utilizing a hierarchical model for analysis.
+Initially, users are encouraged to consider the main analysis within the context of the incomplete dataset, along with insights provided by domain experts, to gain a better understanding of variable relationships.
+Employing clear data visualization is instrumental in comprehending the missing data patterns, establishing a missing mechanism, and aiding in the selection of suitable imputation methods.
+
+The "Mice" and "mice"-based R packages offer a range of imputation methods tailored for hierarchical data, easily adaptable to the dataset's structure.
+Before proceeding with the analysis of the imputed dataset, it is essential to assess the convergence of the imputation method.
+This evaluation can reveal issues such as circular problems, the need for additional iterations, or challenges associated with the chosen imputation method.
+
+## Funding
+
+This project has received funding from the European Union's Horizon 2020 research and innovation programme under ReCoDID grant agreement No 825746.
+
+The views expressed in this paper are the personal views of the authors and may not be understood or quoted as being made on behalf of or reflecting the position of the regulatory agency/agencies or organizations with which the authors are employed/affiliated.
 

### 2023-12-11 docs/manuscript.tex
Add complete case study 1 and 2, and text of 3

@@
 
-\begin{verbatim}
-Warning: package 'ggplot2' was built under R version 4.3.2
-\end{verbatim}
+In hierarchical datasets, clustering is a concern because the
+homoscedasticity in the error terms cannot be assumed across clusters
+and the relationship among variables may vary at different hierarchical
+levels. When multiple imputation is used to deal with missing data, as
+the imputation and analysis process is performed separately, it is
+necessary that imputation model being congenial with the main analysis
+model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
+structure also imputation model should do it (Audigier, 2021). Not
+including clustering into the imputation process may lead to effect
+estimates with smaller standard errors and inflated type I error.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomrule=.15mm, toprule=.15mm, breakable, leftrule=.75mm, left=2mm, colback=white, opacityback=0, rightrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, arc=.35mm, colback=white, breakable, toprule=.15mm, bottomrule=.15mm, rightrule=.15mm, leftrule=.75mm, left=2mm]
 
@@
 
-\begin{verbatim}
-plot_na()
-\end{verbatim}
-
-\begin{figure}[t]
+\begin{figure}[h]
 
@@
 the value of the predictor, the cluster means of the predictor and the
-random slopes of the predictor.
+random slopes of the predictor. - -2 = cluster variable - 1 = overall
+effect - 3 = overall + group-level effect - 4 = individual-level
+(random) and group-level (fixed) effect
 
 \end{tcolorbox}
+
+\hypertarget{sec-workflow}{%
+\section{Multilevel imputation workflow}\label{sec-workflow}}
+
+There are different strategies that can be adopted in the imputation
+process that account for clustering: inclusion of cluster indicator
+variable, performing a separate imputation process for each cluster, or
+performing a simultaneous imputation process by using an imputation
+method that accounts for clustering.(Stata:
+https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/)
+TODO: replace ref.
+
+The selection of each strategy depends mainly on the assumptions in the
+main analysis and also on the restriction of the analyzed data.
+
+Regarding the restrictions imposed by the data, for instance, the use of
+cluster indicator variables is restricted in datasets where there are
+not many clusters and many observations per cluster (Graham, 2009). The
+last restriction is also required when imputations are performed on each
+cluster separately. When this restriction cannot be achieved, one can
+use an imputation model that simultaneously imputes all clusters using a
+hierarchical model (Allison 2002).
+
+Under this hierarchical imputation model, observations within clusters
+are correlated and this correlation is modeled by a random effect so the
+hierarchical model can be estimated even when there are few observations
+per cluster. However, this strategy is best suited for balanced data
+(Grund, 2017) and when random effects model is appropriated, i.e.~the
+number of clusters is adequate. (Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main
+model, for instance by using the cluster indicator strategy may lead to
+bias estimates when the model is based on a hierarchical model
+(Taaljard,2008). Even when an imputation strategy congenial with the
+main model is preferred, it is important to consider whether it is
+appropriate for the data as a less complex imputation strategies may
+also lead to unbiased estimates in certain scenarios(Bailey 2020). For
+instance, in causal effect analysis, separately imputation may lead to
+smaller bias when the size of the smaller exposure cluster is large,
+compared with an imputation model that includes exposure-confounder
+interactions. (Zhang,2023).
+
+Below we provide a imputation workflow that can be used in general to
+impute cluster data.
+
+\hypertarget{focus-on-the-main-analysis}{%
+\subsection{Focus on the Main
+Analysis}\label{focus-on-the-main-analysis}}
+
+When imputing clustered data, we initially focus on the research
+question and the intended analysis, assuming the presence of
+non-incomplete values. This assessment sheds light on research
+hypotheses and the contemplated main(s) statistical model(s) and it
+provides insights into the data's structure (refer to the level table),
+variable types (e.g., confounders, auxiliary variables), and other
+considerations about variables relationships such as interactions or
+polynomial terms.
+
+\hypertarget{exploration-of-available-data}{%
+\subsection{Exploration of Available
+Data}\label{exploration-of-available-data}}
+
+Next, it is necessary to explore the information available in the data.
+Exploring individual variables using tools such as histograms or QQ
+plots can help to delineate variable distributions and plausible ranges
+of values and also identify input errors or outliers. Evaluating
+interactions between predictors using scatter plots or conditional
+plots, assessing collinearity using VIF or correlations, can help to
+glimpse nonlinear relationships that may affect the original formulation
+of the main model. Further testing of assumptions related to the
+response variable, including variance (homogeneity using conditional box
+plots), independence (e.g., ACF, variograms) and response vs.~predictor
+relationships (conditional plots - x vs.~y), may serve to choose an
+imputation model more suitable to the data. In addition, the intraclass
+correlation coefficient (ICC) can be examined to assess cluster
+differences, aiding in the choice between the 2l and 1l methods for
+imputation.
+
+One can also explore missingness, specifically missing patterns at the
+cluster level to identify types of patterns (e.g., non-monotonic,
+systematic) to guide the selection of an appropriate imputation approach
+in terms of computational efficiency (e.g., simpler regression
+imputation versus FCS in univariate patterns). Similarly, these missing
+patterns can also be used to identify potential predictors in individual
+imputation models by means of inflow criteria (quantifying the
+connections between missing data in one variable and other observed
+variables) and outflow criteria (determining the connections between
+observed values in one variable and missing data in other variables).
+Although several packages exist for visualizing missing data, this
+tutorial focuses on the recently implemented ggmice package.
+
+\hypertarget{assess-estimation-procedure-robustness-to-missing-data}{%
+\subsection{Assess Estimation Procedure Robustness to Missing
+Data}\label{assess-estimation-procedure-robustness-to-missing-data}}
+
+Prior to implementing multiple imputation, it is critical to assess the
+robustness of the estimation model when confronted with missing data.
+There are scenarios in which specific Maximum Likelihood (ML) estimation
+methods outperform Multiple Imputation (MI) methods. For instance, when
+the response variable is the sole incomplete variable, mixed models
+demonstrate robustness to missing data under the Missing at Random (MAR)
+assumption and with a correct variance-covariance
+specification.\cite{Molenberghs_2007}. Furthermore, depending on the
+proportion of missingness (usually is \textless5\%), opting for a
+simpler complete case analysis might be sufficient.
+
+\hypertarget{pre-imputation}{%
+\subsection{Pre-imputation}\label{pre-imputation}}
+
+Before proceeding with the imputation model, it is necessary to filter
+the dataset to include only the minimal variables required for the main
+model. If additional procedures are conducted during the analysis, it is
+essential to include variables associated with these procedures, such as
+confounders in the case of balancing techniques. Other relevant
+variables, like instrumental variables or auxiliary variables, may
+enhance parameter estimates even if not included in the main model, as
+they could be linked to the probability of missingness for some
+incomplete variables.
+
+Furthermore, at the variable level, it is crucial to assess whether
+proxy variables should be considered or if deterministic imputation
+alone is adequate for imputing missing values. For instance, certain
+incomplete variables may not necessitate stochastic imputation methods
+like MICE. Instead, they can be effectively addressed through deductive
+imputation, where incomplete values are inferred from logical and
+deterministic relationships between variables. This approach is
+especially beneficial for variables that are functions of others, such
+as deriving a person's BMI from their weight and height. It also proves
+valuable for determining values for one-level variables from two-level
+ones, as seen in the context of Individual Participant Data (IPD). In
+such cases, incomplete information can be deduced from metadata, like
+inferring incomplete data about abortion in a country where abortion is
+illegal, or through cross-temporal or protocol-based deduction, such as
+imputing missing test values for deceased patients.
+
+\hypertarget{setting-imputation-model}{%
+\subsection{Setting Imputation Model}\label{setting-imputation-model}}
+
+\hypertarget{clustering-inclusion}{%
+\paragraph{Clustering Inclusion}\label{clustering-inclusion}}
+
+Various strategies can be employed in the imputation process to account
+for clustering, including the inclusion of a cluster indicator variable,
+conducting a separate imputation process for each cluster, or employing
+a simultaneous imputation method that considers clustering\cite{eddings}
+
+The choice of strategy depends primarily on the assumptions made in the
+main analysis and the constraints imposed by the analyzed data.
+Concerning data restrictions when the analysis do not contemplate an
+hierarchical model (eg. descriptive way), if there are few clusters with
+many observations per cluster, the cluster indicator or imputation
+separated by groups may be appropriate \cite{graham2009}. Conversely, an
+imputation model that simultaneously imputes all clusters using a
+hierarchical model can be employed when there are more clusters or fewer
+observations per cluster \cite{Allison_2002}.
+
+In a hierarchical imputation model, correlations between observations
+within clusters are modeled by a random effect, allowing estimation even
+with a limited number of observations per cluster. Multiple imputation
+models based on hierarchical models have been proposed, each with
+different assumptions. \cite{audigier}
+
+In the particular case that the analysis contemplates the use of a
+hierarchical model (i.e.~main model), before choosing a particular
+strategy, it is crucial to verify if the assumptions of the imputation
+model align with those of the main analysis. For example, using the
+cluster indicator strategy may introduce bias estimates when the model
+is based on a hierarchical structure \cite{taljaard2008,speidel2018}.
+Even if an imputation strategy congruent with the main model is
+preferred, it is essential to assess its appropriateness for the data,
+considering that less complex imputation strategies may lead to unbiased
+estimates in certain scenarios \cite{bailey2020}.
+
+\hypertarget{choice-of-individual-imputation-methods}{%
+\subsubsection{Choice of Individual Imputation
+Methods}\label{choice-of-individual-imputation-methods}}
+
+The initial step involves selecting the individual imputation model for
+each incomplete variable in the dataset. The mice package automatically
+proposes imputation methods based on the variable type for 1l variables.
+For 2l level variables, the micemd package's () function can be
+utilized. This function selects among the 2l imputation methods based on
+the size of the clusters and the proportion of missingness in each
+cluster.
+
+In addition to the package-defined imputation methods, users can specify
+custom methods using the ``I formula''. This flexibility allows for the
+calculation of derived variables internally or adjustments to imputation
+methods based on specific conditions, such as conditioning the
+imputation model on the level of an incomplete covariate (e.g.,
+pregnancy test on females).
+
+\hypertarget{model-specification}{%
+\subsubsection{Model Specification}\label{model-specification}}
+
+The imputation model must be congenial with the main model
+\cite{meng1994}. Congeniality issues arise when the imputation model and
+the main model make different assumptions, often due to the omission of
+a polynomial or interaction term or the use of transformed variables.
+
+The imputation model can include additional terms than the main model
+terms that do not lead to uncongeniality. For instance, it is advisable
+to include the outcome variable in the imputation model for prediction
+variables \cite{moons2006a}. In cases where the outcome is
+time-to-event, the Nelson-Aalen estimate of the time to the event should
+be included as a covariate in the imputation model {[}REF{]}.
+Furthermore, the inclusion of auxiliary variables, even if not part of
+the main model, can be associated with the probability of missingness,
+enhancing the likelihood of satisfying the Missing at Random (MAR)
+assumption and improving estimation efficiency \cite{hardt2012a}.
+
+The specification of imputation models is done on a variable basis,
+either using a prediction matrix where the type of prediction variable
+should be specified accordingly (see table) or through a list of
+formulas in the formula parameter. The latter option proves useful in
+formulating imputation models with polynomial terms or interactions
+compared to the prediction matrix specification, which requires the
+inclusion of additional terms as Just Another Variable (JAV)
+\cite{buuren2018a}.
+
+However, even when considering an interaction term, depending on the
+adapted individual imputation model, it has been suggested, for
+instance, for treatment interaction effects, to conduct separated
+imputation by treatment group \cite{zhang2023}. Additionally, there are
+imputation models based on random forest or deep learning that can
+handle interaction and non-linear terms that do not require the explicit
+inclusion of the non-linear terms.
+
+\hypertarget{post-imputation}{%
+\subsection{Post-Imputation}\label{post-imputation}}
+
+During the imputation process, certain issues may arise that halt the
+process. In hierarchical model imputations, many issues are linked to
+methods based on parametric hierarchical models, which may struggle to
+estimate a model midway through the process. In such cases, it is
+advisable to examine the imputation log file to identify variables with
+imputation issues. Subsequently, reducing the number of predictors in
+the imputation model could be explored, either by using functions like
+quickpred or by considering variable transformations, such as scaling.
+Adjusting the level of the hierarchical model (e.g., using a homogenous
+variance model or a 1l model) can also be beneficial. Additionally,
+checking the range of imputed variables is crucial, as excessively large
+imputed values for one predictor could lead to convergence issues in
+other variables. This can be mitigated by including post-processing
+specifications on problematic variables or by using imputation models
+like Predicted Mean Matching (PMM), ensuring imputed values align with
+the observable values. In certain cases, a separate imputation strategy
+may be considered. For instance, in analyses involving multiple
+endpoints, conducting separate imputation processes for each endpoint
+might be preferable to a unified imputation process.
+
+\hypertarget{convergence-and-sensitivity-analysis}{%
+\subsection{Convergence and Sensitivity
+Analysis}\label{convergence-and-sensitivity-analysis}}
+
+Before starting into the analysis of each imputed dataset, it is crucial
+to validate the convergence of the imputation process. This is commonly
+accomplished through trace plots that depict the mean and variance of
+the incomplete variables across iterations. These plots serve to uncover
+potential circular issues or the need for additional iterations.
+Additionally, it is also important to verify that imputed values fall
+within a plausible range and also to check the distribution of imputed
+variables, ensuring that the imputed variable distribution aligns with
+the distribution of observed values (under the MAR assumption). An
+alternative approach involves assessing the prediction accuracy of the
+imputation method \cite{cai2023}.
+
+While the majority of Multiple Imputation by Chained Equations (MICE)
+methods are based on Missing at Random (MAR) assumptions, field expert
+input may suggest that the Missing Not at Random (MNAR) mechanism could
+be plausible for certain variables. An MNAR variable is one in which the
+probability of missingness depends on an unobservable variable. This can
+occur when missingness is associated with the incomplete value itself
+(self-marking) or when there is an unobserved variable linked to both
+the value and the probability of missingness of the incomplete value
+(indirectly non-informative). Specifically, for the indirectly
+non-informative case in hierarchical datasets, imputation methods based
+on the Heckman method can be
+considered.\cite{hammon2020,hammon2022,munoz2023}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
 
@@
 \begin{verbatim}
-R> set.seed(123)
-R> library(mice)
-R> library(ggmice)
-R> library(ggplot2)
-R> library(miceadds)
-R> library(lme4)
-R> library(mitml)
-R> library(broom.mixed)
+R> set.seed(123)         # for reproducibility
+R> library(lme4)         # for multilevel modeling
+R> library(mice)         # for imputation
+R> library(miceadds)     # for multilevel imputation methods
+R> library(micemd)       # for selection-model imputation methods
+R> library(mitml)        # for multilevel parameter pooling
+R> library(broom)        # for clean model estimates
+R> library(broom.mixed)  # for multilevel model estimates
+R> library(dplyr)        # for data wrangling
+R> library(ggmice)       # for visualization
+R> library(ggplot2)      # for visualization
 \end{verbatim}
@@
 
+In this section we will go over the different steps involved with
+imputing incomplete multilevel data with the R package mice. We consider
+the simulated \texttt{popmis} dataset, which included pupils
+(\(n = 2000\)) clustered within schools (\(N = 100\)). The following
+variables are of primary interest:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{school}, school identification number (clustering variable);
+\item
+  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
+  unit-level);
+\item
+  \texttt{sex}, pupil sex (0 = boy, 1 = girl; unit-level);
+\item
+  \texttt{texp}, teacher experience (in years; cluster-level).
+\end{itemize}
+
+The analysis model corresponding to this dataset is multilevel
+regression with random intercepts for the different schools. We will
+estimate the association between the pupils' sex and their popularity
+score. This model can be expressed in \texttt{lme4} code as:
+
+\begin{verbatim}
+popular ~ 1 + sex + (1 | school)
+\end{verbatim}
+
+We load the data into the environment with
+
 \begin{verbatim}
@@
 
-\begin{verbatim}
-R> dat <- popmis[, c("school", "teachpop", "popular", "texp", "sex")] 
-\end{verbatim}
-
-\begin{verbatim}
-ggmice(dat, aes(popular, teachpop)) + 
+and select the relevant variables
+
+\begin{verbatim}
+R> dat <- popmis[, c("school", "popular", "texp", "sex")] 
+\end{verbatim}
+
+which results in the following data structure.
+
+\begin{verbatim}
+R> head(dat)
+\end{verbatim}
+
+\begin{verbatim}
+  school popular texp sex
+1      1      NA   24   1
+2      1      NA   24   0
+3      1       7   24   1
+4      1      NA   24   1
+5      1      NA   24   1
+6      1       7   24   0
+\end{verbatim}
+
+The association of interest can be visualized with \texttt{ggmice},
+
+\begin{verbatim}
+ggmice(dat, aes(sex, popular)) +
   geom_jitter()
@@
 
-\begin{figure}[t]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-8-1.pdf}
 
@@
 
-\caption{Polar axis plot}
+\caption{Scatterplot of student popularity by sex}
 
@@
 
-With the \texttt{ggmice} unction \texttt{plot\_pattern} we can visualize
-this.
+where missing datapoints in the \texttt{popular} variable are
+represented by red points on the X-axis of the figure.
+
+With the \texttt{ggmice} function \texttt{plot\_pattern} we can
+visualize the missing data pattern
 
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 
@@
 
-\begin{figure}
+which shows us that the missingness is univariate and sporadic.
+
+To develop the best imputation model for the incomplete variable
+\texttt{popular}, we need to know whether the observed values of
+\texttt{popular} are related to observed values of other variables. Plot
+the pair-wise complete correlations in the incomplete data
+
+\begin{verbatim}
+R> plot_corr(dat)
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}
+
+}
+
+\caption{Pair-wise correlations.}
 
@@
 
-\begin{verbatim}
-R> plot_corr(dat)
-\end{verbatim}
-
-\begin{figure}[t]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}
+This shows us that not just the analysis-model variable \texttt{sex},
+but also the cluster-level covariate teacher experience, \texttt{texp},
+may be a useful as an imputation model predictor. Moreover, the
+missingness in \texttt{popular} may depend on the observed values of
+other variables. With \texttt{ggmice()} we can visualize the
+distribution of the teacher experience for cases where \texttt{popular}
+is observed and cases where \texttt{popular} is missing.
+
+\begin{verbatim}
+R> ggmice(dat, aes(texp)) + 
++  geom_density() +
++  facet_wrap(~is.na(popular))
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-11-1.pdf}
 
@@
 
-\caption{Pair-wise correlations.}
-
 \end{figure}
+
+It appears that students with a missing value for \texttt{popular} are
+in clusters with a slightly higher \texttt{texp} value.
+
+\begin{verbatim}
+t.test(dat$texp ~ is.na(dat$popular)) |> 
+  tidy() |>
+  kable()
+\end{verbatim}
+
+\begin{longtable}[]{@{}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0924}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0924}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0924}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.2017}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.1008}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedleft
+estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+estimate1
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+estimate2
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+statistic
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+p.value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+parameter
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+conf.low
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+conf.high
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+method
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+alternative
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+-0.2599581 & 14.15278 & 14.41274 & -0.8731291 & 0.3827095 & 1796.115 &
+-0.8438947 & 0.3239786 & Welch Two Sample t-test & two.sided \\
+\end{longtable}
+
+Although there are no significant differences in the distribution of
+\texttt{texp} depending on the missingness indicator of
+\texttt{popular}, this variable can serve as auxiliary variable in the
+imputation of \texttt{popular}.
 
@@
 \begin{verbatim}
-  school teachpop  popular     texp      sex 
-      ""       ""    "pmm"       ""       "" 
+ school popular    texp     sex 
+     ""   "pmm"      ""      "" 
 \end{verbatim}
@@
 \begin{verbatim}
-         school teachpop popular texp sex
-school        0        0       0    0   0
-teachpop      0        0       0    0   0
-popular       0        1       0    1   1
-texp          0        0       0    0   0
-sex           0        0       0    0   0
+        school popular texp sex
+school       0       0    0   0
+popular      0       0    1   1
+texp         0       0    0   0
+sex          0       0    0   0
 \end{verbatim}
@@
 \end{verbatim}
+
+The \texttt{pmm} method is better (more efficient) because it will still
+look for donors (maybe outside of cluster) based on predictive distance,
+even for very small clusters.
 
@@
 
-\begin{figure}[t]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-15-1.pdf}
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-17-1.pdf}
 
@@
 
-\begin{figure}[t]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-17-1.pdf}
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-19-1.pdf}
 
@@
 
+Troubleshoot non-convergence of imputation model. Evaluate the
+distribution of imputed values.
+
+\begin{verbatim}
+R> ggmice(imp, aes(popular, group = .imp)) + 
++  geom_density() 
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-20-1.pdf}
+
+}
+
+\end{figure}
+
 Evaluate the distribution of imputed values.
@@
 \begin{verbatim}
-ggmice(imp, aes(popular, group = .imp)) + 
-  geom_density() 
-\end{verbatim}
-
-\begin{figure}[t]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-18-1.pdf}
+R> ggmice(imp, aes(.imp, popular)) + 
++  geom_jitter(alpha = 0.05) +
++    geom_boxplot()
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-21-1.pdf}
 
@@
 
-Evaluate the distribution of imputed values.
-
-\begin{verbatim}
-ggmice(imp, aes(.imp, popular)) + 
-  geom_jitter(alpha = 0.05) +
-    geom_boxplot()
-\end{verbatim}
-
-\begin{figure}[t]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-19-1.pdf}
+\begin{verbatim}
+R> ggmice(imp, aes(sex, popular)) +
++  geom_jitter() +
++  facet_wrap(~ .imp)
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-22-1.pdf}
 
@@
 
-\begin{verbatim}
-ggmice(imp, aes(popular, teachpop)) + 
-  geom_jitter() +
-  facet_wrap(~ .imp)
-\end{verbatim}
-
-\begin{figure}[t]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-20-1.pdf}
-
-}
-
-\end{figure}
-
 Analyze the imputed data.
@@
 \begin{verbatim}
-fit <- with(
-  imp,
-  lmer(teachpop ~ popular + texp + (1 | school))
-)
-\end{verbatim}
-
-Pool the estimates.
-
-\begin{verbatim}
-pool(fit)
+R> fit <- with(
++  imp,
++  lmer(popular ~ texp + sex + (1 | school))
++)
+\end{verbatim}
+
+Pooling the estimates does not provide estimates of the variance
+components.
+
+\begin{verbatim}
+R> pool(fit)
 \end{verbatim}
@@
          term m   estimate         ubar            b            t dfcom
-1 (Intercept) 5 2.4091354 2.241304e-02 1.712964e-03 2.446860e-02  1995
-2     popular 5 0.2597284 2.353344e-04 1.209648e-04 3.804922e-04  1995
-3        texp 5 0.0484727 7.728295e-05 3.236252e-06 8.116646e-05  1995
+1 (Intercept) 5 3.59902513 0.0284760238 3.461284e-03 0.0326295651  1995
+2        texp 5 0.09225928 0.0001140591 2.273382e-05 0.0001413397  1995
+3         sex 5 0.85187018 0.0009628869 4.946825e-04 0.0015565060  1995
         df       riv    lambda       fmi
-1 432.50579 0.09171257 0.08400798 0.08821454
-2  26.88403 0.61681474 0.38149995 0.42289329
-3 909.68447 0.05025044 0.04784615 0.04993264
+1 216.1756 0.1458610 0.1272938 0.1352573
+2 100.6502 0.2391794 0.1930144 0.2085857
+3  26.9008 0.6164992 0.3813792 0.4227574
+\end{verbatim}
+
+Therefore, \texttt{mitml} is used.
+
+\begin{verbatim}
+R> est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
 \end{verbatim}
@@
 \begin{verbatim}
-testEstimates(as.mitml.result(fit), extra.pars = TRUE)
-\end{verbatim}
-
-\begin{verbatim}
-
-Call:
-
-testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)
-
-Final parameter estimates and inferences obtained from 5 imputed data sets.
-
-             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
-(Intercept)     2.409     0.156    15.401   566.786     0.000     0.092     0.087 
-popular         0.260     0.020    13.315    27.483     0.000     0.617     0.422 
-texp            0.048     0.009     5.380  1747.294     0.000     0.050     0.049
-
+R> est$estimates |> 
++  round(3) |>
++  kable()
+\end{verbatim}
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1558}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1169}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1299}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2338}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
 Estimate
-Intercept~~Intercept|school    0.310 
-Residual~~Residual             0.307 
-ICC|school                     0.502
-
-Unadjusted hypothesis test as appropriate in larger samples.
-\end{verbatim}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Std.Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+t.value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+P(\textgreater\textbar t\textbar)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+RIV
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+FMI
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+(Intercept) & 3.599 & 0.181 & 19.924 & 246.857 & 0 & 0.146 & 0.134 \\
+texp & 0.092 & 0.012 & 7.760 & 107.369 & 0 & 0.239 & 0.208 \\
+sex & 0.852 & 0.039 & 21.592 & 27.501 & 0 & 0.616 & 0.422 \\
+\end{longtable}
+
+\begin{verbatim}
+R> est$extra.pars |> 
++  round(3) |>
++  kable()
+\end{verbatim}
+
+\begin{longtable}[]{@{}lr@{}}
+\toprule\noalign{}
+& Estimate \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Intercept\textasciitilde\textasciitilde Intercept\textbar school &
+0.470 \\
+Residual\textasciitilde\textasciitilde Residual & 0.462 \\
+ICC\textbar school & 0.504 \\
+\end{longtable}
+
+\hypertarget{impact-data}{%
+\subsection{IMPACT data}\label{impact-data}}
+
+The second case study is the \texttt{impact} data from the
+\pkg{metamisc} package \citep[empirical data on traumatic brain
+injuries, \(n = 11,022\) units across \(N = 15\) clusters,][]{metamisc}.
+
+The \texttt{impact} data set contains traumatic brain injury data on
+\(n = 11022\) patients clustered in \(N = 15\) studies with the
+following 11 variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{name} Name of the study,
+\item
+  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
+  observational cohort),
+\item
+  \texttt{age} Age of the patient,
+\item
+  \texttt{motor\_score} Glasgow Coma Scale motor score,
+\item
+  \texttt{pupil} Pupillary reactivity,
+\item
+  \texttt{ct} Marshall Computerized Tomography classification,
+\item
+  \texttt{hypox} Hypoxia (0=no, 1=yes),
+\item
+  \texttt{hypots} Hypotension (0=no, 1=yes),
+\item
+  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
+\item
+  \texttt{edh} Epidural hematoma (0=no, 1=yes),
+\item
+  \texttt{mort} 6-month mortality (0=alive, 1=dead).
+\end{itemize}
+
+Check if there is systematic missingness in this dataset. For
+illustration purposes, we made Marshall Computerized Tomography
+classification (ct) systematically missing.
+
+The analysis model for this dataset is a prediction model with
+\texttt{mort} as the outcome. In this tutorial we'll estimate the
+adjusted prognostic effect of \texttt{ct} on mortality outcomes. The
+estimand is the adjusted odds ratio for \texttt{ct}, after including
+\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
+into the analysis model:
+
+\begin{verbatim}
+mort ~ type + age + motor_score + pupil + ct + (1 | name) 
+\end{verbatim}
+
+Note that variables \texttt{hypots}, \texttt{hypox}, \texttt{tsah} and
+\texttt{edh} are not part of the analysis model, and may thus serve as
+auxiliary variables for imputation.
+
+The \texttt{impact} data included in the \pkg{metamisc} package is a
+complete data set. The original data has already been imputed once
+(Steyerberg et al, 2008). For the purpose of this tutorial we have
+induced missingness (mimicking the missing data in the original data set
+before imputation). The resulting incomplete data can be accessed from
+\href{https://zenodo.com}{zenodo link to be created}.
+
+Load the incomplete data into the R workspace:
+
+\begin{verbatim}
+R> dat <- read.table("link/to/the/data.txt") 
+\end{verbatim}
+
+To explore the missingness, we should look at the missing data pattern.
+The ten most frequent missingness patterns are shown with
+
+\begin{verbatim}
+R> plot_pattern(dat, rotate = TRUE, npat = 10L)  
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-28-1.pdf}
+
+}
+
+\end{figure}
+
+This shows that we need to impute \texttt{ct} and \texttt{pupil}. To
+develop the best imputation model, we need to investigate the relations
+between the observed values of the incomplete variables and the observed
+values of other variables, and the relation between the missingness
+indicators of the incomplete variables and the observed values of the
+other variables. To see whether the missingness depends on the observed
+values of other variables, we can test this statistically or use visual
+inspection (e.g.~a histogram faceted by the missingness indicator).
+
+We should impute the variables \texttt{ct} and \texttt{pupil} and any
+auxiliary variables we might want to use to impute these incomplete
+analysis model variables. We can evaluate which variables may be useful
+auxiliaries by plotting the pairwise complete correlations
+
+\begin{verbatim}
+R> plot_corr(dat, rotate = TRUE)
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-29-1.pdf}
+
+}
+
+\end{figure}
+
+This shows us that \texttt{hypox} and \texttt{hypot} would not be useful
+auxiliary variables for imputing \texttt{ct}. Depending on the minimum
+required correlation, \texttt{tsah} could be useful, while \texttt{edh}
+has the strongest correlation with \texttt{ct} out of all the variables
+in the data and should definitely be included in the imputation model.
+For the imputation of \texttt{pupil}, none of the potential auxiliary
+variables has a very strong relation, but \texttt{hypots} could be used.
+We conclude that we can exclude \texttt{hypox} from the data, since this
+is neither an analysis model variable nor an auxiliary variable for
+imputation
+
+\begin{verbatim}
+R> dat <- select(dat, !hypox)
+\end{verbatim}
+
+Mutate data to get the right data types for imputation (e.g.~integer for
+clustering variable).
+
+\begin{verbatim}
+R> dat <- dat |>
++  mutate(name = as.integer(name))
+\end{verbatim}
+
+Create an initial methods vector for the incomplete variables
+
+\begin{verbatim}
+R> meth <- make.method(dat)
+R> meth
+\end{verbatim}
+
+\begin{verbatim}
+       name        type         age motor_score       pupil          ct 
+         ""          ""          ""          ""   "polyreg"   "polyreg" 
+     hypots        tsah         edh        mort 
+      "pmm"       "pmm"       "pmm"          "" 
+\end{verbatim}
+
+which should be adjusted to the appropriate \texttt{2l} methods.
+
+\begin{verbatim}
+R> meth[c("pupil", "ct")] <- "2l.pmm"
+R> meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
+\end{verbatim}
+
+Create an initial predictor matrix
+
+\begin{verbatim}
+R> pred <- quickpred(dat)
+\end{verbatim}
+
+This predictor matrix is too large to display inline. A visualization of
+the adapted predictor matrix is presented in Figure XYZ.
+
+We should make sure \texttt{name} is used as clustering variable
+
+\begin{verbatim}
+R> pred[, "name"] <- -2
+\end{verbatim}
+
+and the analysis-model outcome should be used as a predictor in all
+imputation models
+
+\begin{verbatim}
+R> pred[, "mort"] <- 2
+\end{verbatim}
+
+the resulting predictor matrix is visualized with
+
+\begin{verbatim}
+R> plot_pred(pred, method = meth, rotate = TRUE)
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-37-1.pdf}
+
+}
+
+\end{figure}
+
+Impute the incomplete data with
+
+\begin{verbatim}
+R> # imp <- mice(
+R> #   dat,
+R> #   method = meth,
+R> #   predictorMatrix = pred,
+R> #   printFlag = FALSE
+R> # )
+\end{verbatim}
+
+Evaluate the convergence of the algorithm
+
+\begin{verbatim}
+R> # plot_trace(imp)
+\end{verbatim}
+
+Analyze the imputed data:
+
+\begin{verbatim}
+R> # fit <- imp %>% 
+R> #   with(glmer(
+R> #     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
+R> #     family = "binomial"
+R> #     )) 
+\end{verbatim}
+
+The estimated effects after imputation are presented in Table XYZ.
+
+\hypertarget{obesity-data}{%
+\subsection{obesity data}\label{obesity-data}}
+
+In this example, we demonstrate a multilevel imputation of random
+intercept and random slope model with a continuous response. We utilize
+the obesity dataset included in the \texttt{micemd}@ package, a
+simulated dataset that emulates an electronic survey in which
+individuals are asked to provide information about their weight and
+consumption habits in different countries. We simulate data for 5
+clusters so that the true values are known. We use the following
+variables from the dataset:
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Cluster:} Region of the patients' healthcare provider (Cluster
+  variable),
+\item
+  \textbf{Gender:} Subjects' Gender (0=male, 1=female),
+\item
+  \textbf{Age:} Subjects' age,
+\item
+  \textbf{Height:} Subjects' height in metres,
+\item
+  \textbf{Weight:} Subjects' weight in kilograms,
+\item
+  \textbf{BMI:} Subjects' body mass index,
+\item
+  \textbf{FamOb:} Family obesity history (yes or no),
+\item
+  \textbf{Time:} Response time in minutes (exclusion-restriction
+  variable).
+\end{itemize}
+
+In this dataset, Age and FamOb are MAR, while the weight variable is
+affected by selection bias, attributed to an indirect MNAR mechanism.
+This MNAR mechanism typically arises when an unobserved or omitted
+variable influences both the value of the incomplete variable (in this
+case, Weight) and its likelihood of being missing (denoted as R).
+
+In the primary analysis model, BMI serves as the dependent variable,
+with Age, Gender, and FamOb as predictors. Because of the clustered
+nature of the data, which is quantified with the Intraclass Correlation
+Coefficient (ICC) below, we include random intercepts, as well as a
+random slope for the Age variable. The model is represented as:
+\begin{equation}
+\label{eqn:main}
+BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}
+\end{equation}
+
+We start by loading the data:
+
+\begin{verbatim}
+#data("data_heckman", package = "micemd")
+#dat <- data_heckman
+\end{verbatim}
+
+Now, let's begin by examining the missing patterns in the data by
+cluster:
+
+\begin{verbatim}
+# myplots <- lapply(1:5, function(i) {
+#   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
+#   ggplot2::ggtitle(paste0("Cluster", i))
+#  })
+# ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
+\end{verbatim}
+
+We observe that the missing pattern is non-monotonic and quite similar
+across the clusters. However, regarding the weight variable, we notice
+that is systematically missing in cluster 3. In order to evaluate if we
+require a imputation method that accounts for clustering we assess the
+Intraclass Correlation
+
+\begin{verbatim}
+# Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+# performance::icc(Nulmodel)
+\end{verbatim}
+
+Since the ICC is above 0.1 and as the main analysis will be use a mixed
+model, we decide to use two-level (2l) imputation methods. In this
+imputation process, we include all predictor variables from equation
+\ref{eqn:main} in the main model. However, since BMI is a composite of
+weight and height, we use deterministic imputation for these, which is
+described below.
+
+We use the \textbf{find.defaultMethodfunction} provided in the
+\textbf{micemd} package, which suggests an appropriate method for MAR
+variables based on the type of variable, number of observations in the
+cluster, and number of clusters.
+
+It suggests using `2l.2stage.bin' for the FAV variable and
+`2l.2stage.norm' for the age variable. However, after inspecting the age
+density plot, we consider modifying its method to `2l.2stage.pmm'. For
+the BMI variable, we employ deterministic imputation.
+
+\begin{verbatim}
+# meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+#                                ni.small = 100, prop.small = 0.4)
+# meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
+# meth_mar["Age"]<-"2l.2stage.pmm" 
+\end{verbatim}
+
+For these imputation models, it is necessary to specify the prediction
+matrix, with the cluster variable labelled as -2 and the predictor
+variable measured within clusters labelled as 2, encompassing all
+variables. We need to suprime the variable Time as this variable is not
+specified in the main model. We also modify the relationship between
+BMI, weight and height in the prediction matrix to avoid circular
+predictions. Then we proceed to run the imputation model.
+
+\begin{verbatim}
+# pred_mar <- mice(Obesity, maxit = 0)$pred 
+# pred_mar[,"Cluster"] <- -2 # clustering variable
+# pred_mar[,"Time"] <- 0
+# pred_mar[pred_mar==1] <- 2
+# pred_mar[c("Height", "Weight"), "BMI"] <- 0
+# ggmice::plot_pred(pred_mar)
+# imp_mar <- mice::mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+#                 m=10, seed = 123, printFlag = FALSE)
+\end{verbatim}
+
+\begin{verbatim}
+# summary(complete(imp_mar,"long")$Weight)
+\end{verbatim}
+
+We are also contemplating the utilisation of the predictive mean
+matching (pmm) option, as the values imputed using a fully parametric
+method may be implausibly low for some patients.
+
+\begin{verbatim}
+# meth_mar["Weight"]<-"2l.2stage.pmm" 
+# imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+#                     m=10, seed = 123, printFlag = FALSE)
+\end{verbatim}
+
+\begin{verbatim}
+# summary(complete(imp_mar_pmm,"long")$Weight)
+# ggmice::plot_trace(imp_mar_pmm, "Weight")
+\end{verbatim}
+
+After confirming convergence, we proceed to save the results for future
+use. We consider the possibility that patients may not have been
+selected randomly, which would then have led to a distribution for
+weight that does not reflect the weight in the population. It???s likely
+that an omitted variable, like self-esteem, could influence this
+selection. For instance, individuals with lower self-esteem might have
+higher weight values, impacting their willingness to provide honest
+information due to embarrassment.
+
+To address this situation, two approaches have been proposed for dealing
+with Missing Not at Random (MNAR) data: pattern-mixed models and
+selection models. Within pattern-mixed models, methods like the delta
+method and more advanced ones like NARFS have been suggested. The
+selection model approach includes methods such as the Heckman model,
+which can be particularly useful in this case. Several methods,
+including those by \cite{galimard2018}, and the recently a Heckman
+method designed for two-level data, allow for variations in intercepts
+and exposure effects (random intercept and slope) \cite{munoz2023}.
+
+To apply the \textbf{2l.2stage.heckman} method, the weight variable
+should be specified as `2l.2stage.heckman' found in the micemd package.
+Additionally, the prediction matrix needs modification because this
+method involves specifying two equations: one for the outcome,
+describing the incomplete variable in terms of partially observed
+predictors (in this case, all variables from the main model), and the
+other for the selection model, explaining the probability of being
+observed based (R) on certain variables. For the outcome equation we
+consider the same imputation model that we used for the MAR case (main
+model).
+\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
+Regarding the selection equation, we include the same predictors as
+those in the main model, as well as a time variable. Here the time
+variable serves as a restriction exclusion variable specifically
+explaining the probability of being observed but not affecting the
+incomplete value (Weight). In this context, we assume that the time a
+user spends completing the survey serves as a proxy for the barriers
+they may encounter in survey completion, such as familiarity with the
+survey content or internet speed. These factors may lead the user to
+skip specific questions or even the entire survey. Also, we assume the
+time does not have any influence on the subject???s weight.
+\[R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
+
+These two equations are jointly estimated under the assumption that the
+error terms are interconnected with a bivariate normal distribution. For
+a more comprehensive understanding of the model and the exclusion
+restriction, see \cite{munoz2023a}.
+
+To use information from both equations, we must adjust the prediction
+matrix. The cluster variable remains specified as before (-2). In this
+imputation method, all the variables present in both the selection and
+outcome equations are included with a random effect.
+
+However, it is essential to distinguish which of these variables appear
+in each equation. In this framework, when a variable is shared between
+both equations, it is denoted as (2). Predictors exclusive to the
+outcome equation are indicated as (-4), while those exclusive to the
+selection equation are labelled as (-3). Consequently, the only
+alteration needed in the predictor matrix pertains to the variable
+`Time'.
+
+\begin{verbatim}
+# pred_mnar <- pred_mar
+# pred_mnar["Weight","Time"]<- -3
+# ggmice::plot_pred(pred_mnar)
+\end{verbatim}
+
+We also need to modify the method of the weight variable.
+
+\begin{verbatim}
+# meth_mnar <- meth_mar
+# meth_mnar["Weight"]<- "3l.2stage.heckman"
+\end{verbatim}
+
+Then we proceed to run the imputation model as before, after executing
+these imputation procedures, it is essential to assess convergence and
+the coherence of the imputed values.
+
+\begin{verbatim}
+# imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+#                 m=10, seed = 123, printFlag = FALSE)
+# summary(complete(imp_mnar,"long")$Weight)
+\end{verbatim}
+
+Upon examining the weight variable, we noticed that the imputed range
+falls outside the realm of plausible values (as weight should be
+positive).
+
+\begin{verbatim}
+# summary(complete(imp_mnar,"long")$Weight)
+\end{verbatim}
+
+Consequently, as before we use the `pmm', option but this time for the
+Heckman imputation, this approach ensures that the imputed values remain
+within the range of observable values. We then run the imputation model
+but this time using the option of pmm, to assure that weight values are
+in the range of the observable data, this can be implemented by setting
+the pmm parameter to true.
+
+\begin{verbatim}
+# imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+#                      m=10, seed = 123, pmm = T,  printFlag = FALSE)
+\end{verbatim}
+
+We check the convergency of the results
+
+\begin{verbatim}
+# summary(complete(imp_mnar_pmm,"long")$Weight)
+# ggmice::plot_trace(imp_mar_pmm, "Weight")
+\end{verbatim}
+
+After this modification we proceed to compare the effects on the model.
+We run the analysis model on each of the completed datasets as well
+asthe dataset where the incomplete values are removed (Complete Case
+analysis, CC).
+
+\begin{verbatim}
+# cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
+#              lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
+# mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+# mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+# mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+# mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
+# list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
+# plot_models(list_models,
+#             mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
+\end{verbatim}
+
+We note that there is minimal disparity in the age effect, FamObs, or
+Gender across the various imputation models under consideration. An
+analysis of the intercept reveals that, under the MNAR assumption, a
+higher average BMI is anticipated compared to the MAR assumption.
+Nonetheless, with respect to precision of estimates, we notice that in
+general MNAR imputation leads to wider confidence intervals, in this
+case it does not have any influence on the final result but there could
+be cases where variation in the assumed missing mechanism could lead
+also to differences on significant test and therefore lead to
+contradictory conclusions.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+This paper is dedicated to exploring the imputation process for
+incomplete datasets, with a primary focus on utilizing a hierarchical
+model for analysis. Initially, users are encouraged to consider the main
+analysis within the context of the incomplete dataset, along with
+insights provided by domain experts, to gain a better understanding of
+variable relationships. Employing clear data visualization is
+instrumental in comprehending the missing data patterns, establishing a
+missing mechanism, and aiding in the selection of suitable imputation
+methods.
+
+The ``Mice'' and ``mice''-based R packages offer a range of imputation
+methods tailored for hierarchical data, easily adaptable to the
+dataset's structure. Before proceeding with the analysis of the imputed
+dataset, it is essential to assess the convergence of the imputation
+method. This evaluation can reveal issues such as circular problems, the
+need for additional iterations, or challenges associated with the chosen
+imputation method.
+
+\hypertarget{funding}{%
+\section{Funding}\label{funding}}
+
+This project has received funding from the European Union's Horizon 2020
+research and innovation programme under ReCoDID grant agreement No
+825746.
+
+The views expressed in this paper are the personal views of the authors
+and may not be understood or quoted as being made on behalf of or
+reflecting the position of the regulatory agency/agencies or
+organizations with which the authors are employed/affiliated.
 

### 2023-12-12 docs/manuscript.qmd
Analysis case study 2, data case study 3

@@
 #| prompt: true
-dat <- dat |>
-  mutate(name = as.integer(name))
+# dat <- mutate(
+#   dat,
+#   name = as.integer(name))
+```
+
+This is necessary because otherwise PMM cannot be used for these factor variables.
+```{r}
+#| prompt: true
+dat <- mutate(
+  dat,
+  across(everything(), as.numeric))
 ```
@@
 #| prompt: true
-meth[c("pupil", "ct")] <- "2l.pmm"
-meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
+# meth[c("pupil", "ct")] <- "2l.pmm"
+# meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
+meth[meth == "pmm"] <- "2l.pmm"
 ```
@@
 pred[, "mort"] <- 2
+# pred[pred == 1] <- 2
 ```
@@
 #| prompt: true
-# imp <- mice(
-#   dat,
-#   method = meth,
-#   predictorMatrix = pred,
-#   printFlag = FALSE
-# )
+imp <- mice(
+  dat,
+  method = meth,
+  predictorMatrix = pred,
+  printFlag = FALSE
+)
 ```
@@
 #| prompt: true
-# plot_trace(imp)
+plot_trace(imp)
+```
+
+Evaluate the imputed values.
+```{r}
+#| prompt: true
+ggmice(imp, aes(.imp, pupil)) +
+  geom_jitter()
+```
+
+Convert the data back to factors.
+```{r}
+long <- complete(imp, "long", include = TRUE)
+long <- mutate(long, across(c("motor_score", "pupil", "ct"), as.factor))
+imp <- as.mids(long)
 ```
@@
 #| prompt: true
-# fit <- imp %>% 
-#   with(glmer(
-#     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
-#     family = "binomial"
-#     )) 
-```
+#| warning: false
+fit <- imp %>%
+  with(glmer(
+    mort ~ type + age + motor_score + pupil + ct + (1 | name),
+    family = "binomial"
+    ))
+```
+
 The estimated effects after imputation are presented in Table XYZ.
+```{R}
+#| prompt: true
+est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+```
+
+Display results in table.
+```{r}
+#| prompt: true
+est$estimates |> 
+  round(3) |>
+  kable()
+```
+
+```{r}
+#| prompt: true
+est$extra.pars |> 
+  round(3) |>
+  kable()
+```
 
@@
 ```{r}
-#data("data_heckman", package = "micemd")
-#dat <- data_heckman
+#| prompt: true
+data(Obesity, package = "micemd")
 ```
@@
 
-```{r, dpi = 300,out.width="70%", fig.cap="Missing pattern", eval=TRUE}
-# myplots <- lapply(1:5, function(i) {
-#   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
-#   ggplot2::ggtitle(paste0("Cluster", i))
-#  })
-# ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
+```{r}
+#| prompt: true
+plot_pattern(Obesity)
+```
+
+```{r}
+#| prompt: true
+Obesity |> 
+  split(~Cluster) |>
+  lapply(plot_pattern)
 ```

### 2023-12-12 docs/manuscript.tex
Analysis case study 2, data case study 3

@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, arc=.35mm, colback=white, breakable, toprule=.15mm, bottomrule=.15mm, rightrule=.15mm, leftrule=.75mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, toprule=.15mm, rightrule=.15mm, arc=.35mm, left=2mm, leftrule=.75mm, colback=white, opacityback=0]
 
@@
 \begin{verbatim}
-R> dat <- dat |>
-+  mutate(name = as.integer(name))
+R> # dat <- mutate(
+R> #   dat,
+R> #   name = as.integer(name))
+\end{verbatim}
+
+This is necessary because otherwise PMM cannot be used for these factor
+variables.
+
+\begin{verbatim}
+R> dat <- mutate(
++  dat,
++  across(everything(), as.numeric))
 \end{verbatim}
@@
        name        type         age motor_score       pupil          ct 
-         ""          ""          ""          ""   "polyreg"   "polyreg" 
+         ""          ""          ""          ""       "pmm"       "pmm" 
      hypots        tsah         edh        mort 
@@
 \begin{verbatim}
-R> meth[c("pupil", "ct")] <- "2l.pmm"
-R> meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
+R> # meth[c("pupil", "ct")] <- "2l.pmm"
+R> # meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
+R> meth[meth == "pmm"] <- "2l.pmm"
 \end{verbatim}
@@
 R> pred[, "mort"] <- 2
+R> # pred[pred == 1] <- 2
 \end{verbatim}
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-37-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-38-1.pdf}
 
@@
 \begin{verbatim}
-R> # imp <- mice(
-R> #   dat,
-R> #   method = meth,
-R> #   predictorMatrix = pred,
-R> #   printFlag = FALSE
-R> # )
+R> imp <- mice(
++  dat,
++  method = meth,
++  predictorMatrix = pred,
++  printFlag = FALSE
++)
 \end{verbatim}
@@
 \begin{verbatim}
-R> # plot_trace(imp)
+R> plot_trace(imp)
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-40-1.pdf}
+
+}
+
+\end{figure}
+
+Evaluate the imputed values.
+
+\begin{verbatim}
+R> ggmice(imp, aes(.imp, pupil)) +
++  geom_jitter()
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-41-1.pdf}
+
+}
+
+\end{figure}
+
+Convert the data back to factors.
+
+\begin{verbatim}
+long <- complete(imp, "long", include = TRUE)
+long <- mutate(long, across(c("motor_score", "pupil", "ct"), as.factor))
+imp <- as.mids(long)
 \end{verbatim}
@@
 \begin{verbatim}
-R> # fit <- imp %>% 
-R> #   with(glmer(
-R> #     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
-R> #     family = "binomial"
-R> #     )) 
+R> fit <- imp %>%
++  with(glmer(
++    mort ~ type + age + motor_score + pupil + ct + (1 | name),
++    family = "binomial"
++    ))
 \end{verbatim}
@@
 The estimated effects after imputation are presented in Table XYZ.
+
+\begin{verbatim}
+R> est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+\end{verbatim}
+
+Display results in table.
+
+\begin{verbatim}
+R> est$estimates |> 
++  round(3) |>
++  kable()
+\end{verbatim}
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1605}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1111}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1235}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0988}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1358}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2222}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Std.Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+t.value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+P(\textgreater\textbar t\textbar)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+RIV
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+FMI
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+(Intercept) & -1.892 & 0.333 & -5.682 & 11445.455 & 0.000 & 0.019 &
+0.019 \\
+type & -0.346 & 0.178 & -1.948 & 686356.768 & 0.051 & 0.002 & 0.002 \\
+age & 0.032 & 0.002 & 19.564 & 2418.910 & 0.000 & 0.042 & 0.041 \\
+motor\_score2 & -0.575 & 0.070 & -8.226 & 71295.018 & 0.000 & 0.008 &
+0.008 \\
+motor\_score3 & -0.891 & 0.072 & -12.440 & 66712.189 & 0.000 & 0.008 &
+0.008 \\
+motor\_score4 & -1.276 & 0.074 & -17.305 & 18426.087 & 0.000 & 0.015 &
+0.015 \\
+pupil2 & 1.315 & 0.068 & 19.339 & 110.826 & 0.000 & 0.235 & 0.204 \\
+pupil3 & 0.657 & 0.075 & 8.745 & 423.463 & 0.000 & 0.108 & 0.101 \\
+ct2 & 0.747 & 0.084 & 8.884 & 41.781 & 0.000 & 0.448 & 0.340 \\
+ct3 & 0.766 & 0.090 & 8.535 & 12.103 & 0.000 & 1.352 & 0.631 \\
+\end{longtable}
+
+\begin{verbatim}
+R> est$extra.pars |> 
++  round(3) |>
++  kable()
+\end{verbatim}
+
+\begin{longtable}[]{@{}lr@{}}
+\toprule\noalign{}
+& Estimate \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Intercept\textasciitilde\textasciitilde Intercept\textbar name &
+0.082 \\
+\end{longtable}
 
@@
 \begin{verbatim}
-#data("data_heckman", package = "micemd")
-#dat <- data_heckman
+R> data(Obesity, package = "micemd")
 \end{verbatim}
@@
 \begin{verbatim}
-# myplots <- lapply(1:5, function(i) {
-#   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
-#   ggplot2::ggtitle(paste0("Cluster", i))
-#  })
-# ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
-\end{verbatim}
+R> plot_pattern(Obesity)
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-48-1.pdf}
+
+}
+
+\end{figure}
+
+\begin{verbatim}
+R> Obesity |> 
++  split(~Cluster) |>
++  lapply(plot_pattern)
+\end{verbatim}
+
+\begin{verbatim}
+$`1`
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-1.pdf}
+
+}
+
+\end{figure}
+
+\begin{verbatim}
+
+$`2`
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-2.pdf}
+
+}
+
+\end{figure}
+
+\begin{verbatim}
+
+$`3`
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-3.pdf}
+
+}
+
+\end{figure}
+
+\begin{verbatim}
+
+$`4`
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-4.pdf}
+
+}
+
+\end{figure}
+
+\begin{verbatim}
+
+$`5`
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-5.pdf}
+
+}
+
+\end{figure}
 

### 2023-12-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update prediction matrix table,mixed models

@@
  - \usepackage{ulem}
+ - \usepackage{xcolor}
 output: 
@@
 |                | depends on unrecorded information, making the missingness non-ignorable    |
-|                | [@rubi76; @meng94].                                                        |
+|                | [@rubi76; @meng1994].                                                        |
 |                |                              |
@@
 ### Focus on the Main Analysis
-When imputing clustered data, we initially focus on the research question and the intended analysis, assuming the presence of non-incomplete values.
-This assessment sheds light on research hypotheses and the contemplated main(s) statistical model(s) and it provides insights into the data's structure (refer to the level table), variable types (e.g., confounders, auxiliary variables), and other considerations about variables relationships such as interactions or polynomial terms.
+
+When dealing with incomplete clustered data, start by looking at your research questions and planned analysis.
+Pretend there are no missing data at first.
+This will help you understand your research hypotheses, the main statistical model, and give you insights into your data's structure (refer to the level table), variable types (e.g., confounders, auxiliary variables), and other considerations such as variable relationships, interactions or polynomial terms.
 
 ### Exploration of Available Data
-Next, it is necessary to explore the information available in the data.
-Exploring individual variables using tools such as histograms or QQ plots can help to delineate variable distributions and plausible ranges of values and also identify input errors or outliers.
-Evaluating interactions between predictors using scatter plots or conditional plots, assessing collinearity using VIF or correlations, can help to glimpse nonlinear relationships that may affect the original formulation of the main model.
-Further testing of assumptions related to the response variable, including variance (homogeneity using conditional box plots), independence (e.g., ACF, variograms) and response vs.
-predictor relationships (conditional plots - x vs.
-y), may serve to choose an imputation model more suitable to the data.
+
+Now, explore what's in your data.
+Use simple tools like histograms and QQ plots to understand variable distributions, plausible range of values and identify errors or outliers.
+Check interactions between predictors with scatter plots, look for collinearity with VIF or correlations plots, this may also help you to identify non-considered relationships that may affect the main model.
+You might also consider to test assumptions related to your response variable such as variance homogeneity, independence between observations (eg.
+ACF, variagrams).
+This will help you choose an imputation model that suits your data.
 In addition, the intraclass correlation coefficient (ICC) can be examined to assess cluster differences, aiding in the choice between the 2l and 1l methods for imputation.
 
-One can also explore missingness, specifically missing patterns at the cluster level to identify types of patterns (e.g., non-monotonic, systematic) to guide the selection of an appropriate imputation approach in terms of computational efficiency (e.g., simpler regression imputation versus FCS in univariate patterns).
-Similarly, these missing patterns can also be used to identify potential predictors in individual imputation models by means of inflow criteria (quantifying the connections between missing data in one variable and other observed variables) and outflow criteria (determining the connections between observed values in one variable and missing data in other variables).
-Although several packages exist for visualizing missing data, this tutorial focuses on the recently implemented ggmice package.
+Next, explore the missingness.
+Look at the \textbf{proportion of missing values} in the dataset variables.
+This helps find potential predictors and cut down on unnecessary variables in the imputation model.
+Doing this can lower the risk of multicollinearity or computational issues, especially with certain parametric imputation methods.
+You can also identify predictors for the imputation model by using inflow criteria to see connections between missing data in one variable and observed variables, and outflow criteria to identify connections between observed values in one variable and missing data in others.
+
+Check the missing patterns at cluster level, this can help you to select the most appropiated imputation approach in terms of computational efficiency (e.g., simpler regression imputation versus FCS in univariate patterns).
 
 ### Assess Estimation Procedure Robustness to Missing Data
-Prior to implementing multiple imputation, it is critical to assess the robustness of the estimation model when confronted with missing data.
-There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods.
-For instance, when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}.
-Furthermore, depending on the proportion of missingness (usually is <5%), opting for a simpler complete case analysis might be sufficient.
+Before diving into imputation, make sure your estimation model can handle missing data.
+Sometimes, simpler methods like complete case analysis might be suffice, especially if your missingness is low (usually <5%).
+There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods, for instance when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}.
 
 ### Pre-imputation
-Before proceeding with the imputation model, it is necessary to filter the dataset to include only the minimal variables required for the main model.
-If additional procedures are conducted during the analysis, it is essential to include variables associated with these procedures, such as confounders in the case of balancing techniques.
-Other relevant variables, like instrumental variables or auxiliary variables, may enhance parameter estimates even if not included in the main model, as they could be linked to the probability of missingness for some incomplete variables.
-
-Furthermore, at the variable level, it is crucial to assess whether proxy variables should be considered or if deterministic imputation alone is adequate for imputing missing values.
-For instance, certain incomplete variables may not necessitate stochastic imputation methods like MICE.
-Instead, they can be effectively addressed through deductive imputation, where incomplete values are inferred from logical and deterministic relationships between variables.
-This approach is especially beneficial for variables that are functions of others, such as deriving a person's BMI from their weight and height.
-It also proves valuable for determining values for one-level variables from two-level ones, as seen in the context of Individual Participant Data (IPD).
-In such cases, incomplete information can be deduced from metadata, like inferring incomplete data about abortion in a country where abortion is illegal, or through cross-temporal or protocol-based deduction, such as imputing missing test values for deceased patients.
+Clean up your dataset before the imputation process.
+Keep initially the essential variables for your main model, but include extra variables if needed for specific procedures during analysis (e.g.
+confounders on balancing procedures).
+Think about adding other useful variables, even if they're not in the main model, such as instrumental or auxiliary variables which might boost your parameter estimates, especially if they're associated to the probability of missingness for some incomplete variables.
+
+Figure out if you can directly impute incomplete variables by just use deductive imputation.
+This involves inferring missing values based on logicallogical connections between variables.
+It's especially handy for variables that depend on each other, like calculating BMI from weight and height.
+
+Deductive imputation is also useful for getting values for level-1 variables from level-2 ones, like in Individual Participant Data (IPD).
+In these situations, you can guess missing information from metadata or by using or through deduction based on time or protocols.
+For example, you could deduce missing test values for patients who have passed away.
 
@@
 ##### Clustering Inclusion
-Various strategies can be employed in the imputation process to account for clustering, including the inclusion of a cluster indicator variable, conducting a separate imputation process for each cluster, or employing a simultaneous imputation method that considers clustering\cite{eddings}
-
-The choice of strategy depends primarily on the assumptions made in the main analysis and the constraints imposed by the analyzed data.
-Concerning data restrictions when the analysis do not contemplate an hierarchical model (eg.
-descriptive way), if there are few clusters with many observations per cluster, the cluster indicator or imputation separated by groups may be appropriate \cite{graham2009}.
-Conversely, an imputation model that simultaneously imputes all clusters using a hierarchical model can be employed when there are more clusters or fewer observations per cluster \cite{Allison_2002}.
-
-In a hierarchical imputation model, correlations between observations within clusters are modeled by a random effect, allowing estimation even with a limited number of observations per cluster.
-Multiple imputation models based on hierarchical models have been proposed, each with different assumptions.
-\cite{audigier}
-
-In the particular case that the analysis contemplates the use of a hierarchical model (i.e.
-main model), before choosing a particular strategy, it is crucial to verify if the assumptions of the imputation model align with those of the main analysis.
-For example, using the cluster indicator strategy may introduce bias estimates when the model is based on a hierarchical structure \cite{taljaard2008,speidel2018}.
-Even if an imputation strategy congruent with the main model is preferred, it is essential to assess its appropriateness for the data, considering that less complex imputation strategies may lead to unbiased estimates in certain scenarios \cite{bailey2020}.
+
+When it comes to handling clusters during the imputation process, you've got a few options.
+You can use a cluster indicator variable, run separate imputation for each cluster, or go for a simultaneous imputation method that takes clustering into account \cite{eddings}.
+
+Which strategy you choose depends on the assumptions in your main analysis and the limitations of your data.
+If your analysis do not use a hierarchical model (like a descriptive approach) and you have a small number of clusters with lots of observations in each, using a cluster indicator or separate imputation might be the way to go \cite{graham2009}.
+Conversely, if you have more clusters or fewer observations per cluster, you might want to try a simultaneous hierarchical imputation model \cite{Allison_2002}.
+
+In a hierarchical imputation model, random effects model the correlations between observations within clusters, making it possible to estimate even with a small number of observations per cluster.
+There are various proposed multiple imputation models based on hierarchical models, each with its own set of assumptions \cite{audigier}.
+
+If your analysis uses a hierarchical model, make sure the assumptions of your imputation model match up.
+For example, using a cluster indicator approach may lead to bias estimates if your model is based on a hierarchical structure \cite{taljaard2008,speidel2018}.
+Even if you prefer an imputation strategy that aligns with your main model, check if it suits your data; sometimes simpler strategies can give unbiased estimates in certain scenarios \cite{bailey2020}.
 
 #### Choice of Individual Imputation Methods
-The initial step involves selecting the individual imputation model for each incomplete variable in the dataset.
-The mice package automatically proposes imputation methods based on the variable type for 1l variables.
-For 2l level variables, the micemd package's () function can be utilized.
-This function selects among the 2l imputation methods based on the size of the clusters and the proportion of missingness in each cluster.
-
-In addition to the package-defined imputation methods, users can specify custom methods using the "I formula".
-This flexibility allows for the calculation of derived variables internally or adjustments to imputation methods based on specific conditions, such as conditioning the imputation model on the level of an incomplete covariate (e.g., pregnancy test on females).
+Start by choosing the imputation model for each incomplete variable in your dataset.
+The mice package suggests methods based on variable types for non-clustered variables, and for the cluster ones, you can use the micemd package's find.defaultMethod() function.
+This function selects from different 2l imputation methods based on cluster size and the proportion of missing data in each cluster.
+
+Besides the package-defined imputation methods, you can specify custom methods using the "I formula".
+This lets you calculate deterministic variables during the imputation or tweak imputation methods based on specific conditions, like conditioning the imputation model to the level of an incomplete covariate (e.g., a pregnancy test for females).
 
 #### Model Specification
+
 The imputation model must be congenial with the main model \cite{meng1994}.
@@
 
-The imputation model can include additional terms than the main model terms that do not lead to uncongeniality.
-For instance, it is advisable to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}.
-In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be included as a covariate in the imputation model [REF].
-Furthermore, the inclusion of auxiliary variables, even if not part of the main model, can be associated with the probability of missingness, enhancing the likelihood of satisfying the Missing at Random (MAR) assumption and improving estimation efficiency \cite{hardt2012a}.
-
-The specification of imputation models is done on a variable basis, either using a prediction matrix where the type of prediction variable should be specified accordingly (see table) or through a list of formulas in the formula parameter.
-The latter option proves useful in formulating imputation models with polynomial terms or interactions compared to the prediction matrix specification, which requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}.
-
-<!--TODO: Add a table describing 2l-imputation models codes -->
-
-However, even when considering an interaction term, depending on the adapted individual imputation model, it has been suggested, for instance, for treatment interaction effects, to conduct separated imputation by treatment group \cite{zhang2023}.
-Additionally, there are imputation models based on random forest or deep learning that can handle interaction and non-linear terms that do not require the explicit inclusion of the non-linear terms.
+The imputation model can incorporate additional terms compared to the main model without causing compatibility issues.
+For instance, it's recommended to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}.
+In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be added as a covariate in the imputation model [REF].
+Additionally, including auxiliary variables, even if not part of the main model, can be linked to the probability of missingness, improving the likelihood of meeting the Missing at Random (MAR) assumption and enhancing estimation efficiency \cite{hardt2012a}.
+
+Imputation models are specified on a variable basis, either using a prediction matrix in the pred parameter or through a list of formulas in the formula parameter.
+In the prediction matrix option, the type of each predictor variable is specified for each incomplete variable (see table).
+
+\begin{align}
+y_{ij} =& (\beta_0 + b_{0_i})+ \beta_1x_{1_{ij}}+\beta_2x_{2_{i.}} \nonumber \\
++& (\beta_3 +  b_{3_i})x_{3_{ij}} \nonumber \\
++& \beta_4x_{4_{ij}} + \beta_{m4}\overline{x_{4_{i.}}}\nonumber \\
++& (\beta_5 +  b_{5_i})x_{5_{ij}} +\beta_{m5}\overline{x_{5_{i.}}}\nonumber \\
+\end{align}
+
+\begin{tabular}{ll}
+     \toprule
+Type & Definition \\
+  \midrule
+-2  & Cluster variable, in this case the one defined by j index  \\
+ 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
+ 2  & Random variable, e.g., $x_{3_{ij}}$\\
+ 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
+ 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
+-3. & Random variable only included on selection model (Heckman model)\\
+-4. & Random variable only included on main model (Heckman model)\\
+      \bottomrule
+\end{tabular}
+
+* 3 and 4 type have been advised to used as the inclusion of the means of the cluster is beneficial on FCS \cite{mistler2017}
+
+Recipes have been proposed for imputing incomplete level-1 and level-2 variables for hierarchical models \cite{buuren2018a} ($\S$ 7.10), which can be convenient to follow when dealing with numerous variables (interaction terms at different levels) and models with many random effects that are prone to convergency problems due to overspecification in the imputation model.
+These rules were designed to ensure compatibility among the conditionally specified imputation models and congeniality between the imputation and the main model.
+However, they may not be applicable to all 2l imputation methods; for instance, the 2l.2stage imputation methods of the micemd package only allow the inclusion of random predictor variables (2).
+
+On the other hand, the formula option is useful in specifying complex imputation models with polynomial terms or interactions and compared with the prediction matrix method do not requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}($\S$ 6.4).
+
+For some interaction terms, it has been suggested, for instance, for treatment interaction effects, to conduct separate imputation by treatment group \cite{zhang2023}.
+Additionally, it can be used imputation models based on random forest or deep learning that can handle interaction and non-linear terms without requiring the explicit specification of an imputation model.
 
@@
 During the imputation process, certain issues may arise that halt the process.
-In hierarchical model imputations, many issues are linked to methods based on parametric hierarchical models, which may struggle to estimate a model midway through the process.
-In such cases, it is advisable to examine the imputation log file to identify variables with imputation issues.
-Subsequently, reducing the number of predictors in the imputation model could be explored, either by using functions like quickpred or by considering variable transformations, such as scaling.
-Adjusting the level of the hierarchical model (e.g., using a homogenous variance model or a 1l model) can also be beneficial.
-Additionally, checking the range of imputed variables is crucial, as excessively large imputed values for one predictor could lead to convergence issues in other variables.
-This can be mitigated by including post-processing specifications on problematic variables or by using imputation models like Predicted Mean Matching (PMM), ensuring imputed values align with the observable values.
-In certain cases, a separate imputation strategy may be considered.
-For instance, in analyses involving multiple endpoints, conducting separate imputation processes for each endpoint might be preferable to a unified imputation process.
+In hierarchical model imputations, many issues are related to over fitting the imputation model.
+To troubleshoot, it is recommended to inspect the imputation log file for variables causing problems.
+One approach to address this is to reduce the number of predictors, by using step by step the previous referred recepies or by using functions like quickpred.
+Another option is to consider variable transformations, such as scaling when model is invariant to linear transformations e.g., random intercept models.
+Also adjusting the level of the hierarchical model (e.g., using a homogeneous variance imputation method or a 1l model) can also be beneficial.
+
+It is crucial to check the range of imputed variables, as excessively large imputed values for one predictor may trigger convergence issues in other variables.
+To tackle this, including post-processing specifications on problematic variables or utilizing imputation models like Predicted Mean Matching (PMM) can ensure that imputed values align with observable values.
+
+In some situations, adopting a separate imputation strategy might be worth considering.
+For example, in analyses involving multiple endpoints, conducting distinct imputation processes for each endpoint could be more effective than a unified imputation approach.
 
@@
 The research objective of the `popmis` dataset is to predict the pupils' popularity based on their gender and the experience of the teacher.
-The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction.
-The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`:
+The analysis model corresponding to this dataset is multilevel regression with only random intercepts.
+The outcome variable is `popular`, which is predicted from the level-1 (student) variable `sex` and the level-2 (school) variable `texp`.
+Given the $j$-th student belonging to the $i$-th school, the main model can be formulated as:
+
+$$ popular_{ij} = (\beta_0 + b_i) + \beta_1sex_{ij}+ \beta_2texp_{ij}+\epsilon_{ij}$$
+where $\epsilon_{ij}$ corresponds to the error term.
 
@@
 <!-- [TODO: make ct levels one var? also shows that you don't always need random effects everywhere?] -->
-
+In this dataset all the variables are level-1 (patient), except by the level-2 (study) type variable.
 The analysis model for this dataset is a prediction model with `mort` as the outcome.
 In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes.
-The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model:
+The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil`.
+Therefore the main model for the $i$-th patient from the $j$-th study can be described as:
+
+$$ mort_{ij} = (\beta_0 + b_i) + \beta_1type_{.j}+ \beta_2age_{ij}+ \beta_3motorscore_{ij} + \beta_4pupil_{ij} + \beta_5ct_{ij}+\epsilon_{ij}$$
+where $\epsilon_{ij}$ corresponds to the error term.
+
 ```{r mod}
@@
 source(here("heckman","ipdfunction.R"),local =knitr::knit_global())
-load(here("heckman","Obesity.rda"))
 ```
@@
 ```{r}
-#data("data_heckman", package = "micemd")
-#dat <- data_heckman
+data("Obesity", package = "micemd")
+
 ```
@@
 meth_mnar <- meth_mar
-meth_mnar["Weight"]<- "3l.2stage.heckman"
+meth_mnar["Weight"]<- "2l.2stage.heckman"
 ```

### 2023-12-12 docs/manuscript.qmd
Case study 3

@@
 library(lme4)         # for multilevel modeling
+library(nlme)         # for multilevel modeling
 library(mice)         # for imputation
@@
 #| prompt: true
-Obesity |> 
-  split(~Cluster) |>
-  lapply(plot_pattern)
+# Obesity |> 
+#   split(~Cluster) |>
+#   lapply(plot_pattern)
 ```
@@
 
-```{r obesity-icc, eval=TRUE}
-# Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+```{r}
+#| prompt: true
+Nulmodel <- lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
 # performance::icc(Nulmodel)
@@
 ```{r obesitymeth, eval=TRUE}
-# meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
-#                                ni.small = 100, prop.small = 0.4)
-# meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
-# meth_mar["Age"]<-"2l.2stage.pmm" 
+meth_mar <-
+  find.defaultMethod(
+    Obesity,
+    ind.clust = 1,
+    I.small = 7,
+    ni.small = 100,
+    prop.small = 0.4
+  )
+meth_mar["BMI"] <- "~ I(Weight / (Height)^2)"
+meth_mar["Age"] <- "2l.2stage.pmm"
 ```
@@
 ```{r obesity-predmar,out.width="70%", eval=TRUE}
-# pred_mar <- mice(Obesity, maxit = 0)$pred 
-# pred_mar[,"Cluster"] <- -2 # clustering variable
-# pred_mar[,"Time"] <- 0
-# pred_mar[pred_mar==1] <- 2
-# pred_mar[c("Height", "Weight"), "BMI"] <- 0
-# ggmice::plot_pred(pred_mar)
-# imp_mar <- mice::mice(data = Obesity, meth = meth_mar, pred = pred_mar,
-#                 m=10, seed = 123, printFlag = FALSE)
+pred_mar <- quickpred(Obesity)
+pred_mar[, "Cluster"] <- -2 
+pred_mar[, "Time"] <- 0
+pred_mar[pred_mar == 1] <- 2
+pred_mar[c("Height", "Weight"), "BMI"] <- 0
+plot_pred(pred_mar)
+imp_mar <-
+  mice(
+    data = Obesity,
+    method = meth_mar,
+    predictorMatrix = pred_mar,
+    printFlag = FALSE
+  )
 ```
@@
 ```{r obesity-predmar1,eval=TRUE}
-# summary(complete(imp_mar,"long")$Weight)
+summary(complete(imp_mar,"long")$Weight)
 ```
@@
 ```{r obesity-predmar_pmm, eval=TRUE}
-# meth_mar["Weight"]<-"2l.2stage.pmm" 
-# imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
-#                     m=10, seed = 123, printFlag = FALSE)
+meth_mar["Weight"] <- "2l.2stage.pmm"
+imp_mar_pmm <- mice(
+  data = Obesity,
+  method = meth_mar,
+  predictorMatrix = pred_mar,
+  printFlag = FALSE
+)
 ```
@@
 ```{r obesity-predmar_pmm1,eval=TRUE}
-# summary(complete(imp_mar_pmm,"long")$Weight)
-# ggmice::plot_trace(imp_mar_pmm, "Weight")
+summary(complete(imp_mar_pmm, "long")$Weight)
+plot_trace(imp_mar_pmm, "Weight")
+ggmice(imp_mar_pmm, aes(.imp, Weight)) + 
+  geom_jitter()
+```
+
+Analyze the imputed data.
+```{R}
+#| prompt: true
+#| warning: false
+# fit <- with(
+#     imp_mar_pmm,
+#     lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+#     )
+fit <- with(
+    imp_mar_pmm,
+    lmer(BMI ~ Age + FamOb + Gender + (1 + Age | Cluster))
+    )
+est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+```
+
+Display results in table.
+```{r}
+#| prompt: true
+est$estimates |> 
+  round(3) |>
+  kable()
+```
+
+```{r}
+#| prompt: true
+est$extra.pars |> 
+  round(3) |>
+  kable()
 ```
@@
 ```{r obesity-predmnar, eval=TRUE}
-# pred_mnar <- pred_mar
-# pred_mnar["Weight","Time"]<- -3
-# ggmice::plot_pred(pred_mnar)
+pred_mnar <- pred_mar
+pred_mnar["Weight","Time"] <- -3
+plot_pred(pred_mnar)
 ```
@@
 ```{r obesity-predmnar1, eval=TRUE}
-# meth_mnar <- meth_mar
-# meth_mnar["Weight"]<- "3l.2stage.heckman"
+meth_mnar <- meth_mar
+meth_mnar["Weight"] <- "2l.2stage.heckman"
 ```
@@
 ```{r obesity-predmnar2,eval=TRUE}
-# imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
-#                 m=10, seed = 123, printFlag = FALSE)
-# summary(complete(imp_mnar,"long")$Weight)
+imp_mnar <- mice(
+  data = Obesity,
+  method = meth_mnar,
+  predictorMatrix = pred_mnar,
+  printFlag = FALSE
+)
+summary(complete(imp_mnar, "long")$Weight)
 ```
@@
 ```{r obesity-predmnar3,eval=TRUE}
-# summary(complete(imp_mnar,"long")$Weight)
+summary(complete(imp_mnar, "long")$Weight)
 ```
@@
 ```{r obesity-predmnarp, eval=TRUE}
-# imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
-#                      m=10, seed = 123, pmm = T,  printFlag = FALSE)
-```
-
-We check the convergency of the results
+imp_mnar_pmm <-
+  mice(
+    data = Obesity,
+    method = meth_mnar,
+    predictorMatrix = pred_mnar,
+    pmm = T,
+    printFlag = FALSE
+  )
+```
+
+We check the convergence of the results
 ```{r obesity-predmnarp1, eval=TRUE}
-# summary(complete(imp_mnar_pmm,"long")$Weight)
-# ggmice::plot_trace(imp_mar_pmm, "Weight")
+summary(complete(imp_mnar_pmm, "long")$Weight)
+plot_trace(imp_mar_pmm, "Weight")
 ```
@@
 After this modification we proceed to compare the effects on the model. We
-run the analysis model on each of the completed datasets as well asthe dataset where the
+run the analysis model on each of the completed datasets as well as the dataset where the
 incomplete values are removed (Complete Case analysis, CC).
@@
 ```{r models, eval=TRUE}
-# cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
-#              lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
-# mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
+# cc_rs <- with(setDT(Obesity)[complete.cases(Obesity), ],
+#               lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age |
+#                     Cluster))
+mar_rs <- with(
+  imp_mar, 
+  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+  )
+mar_pmm_rs <- with(
+    imp_mar_pmm,
+    lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+    )
+mnar_rs <- with(
+  imp_mnar, 
+  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+  )
+mnar_pmm_rs <- with(
+  imp_mnar_pmm,
+  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+  )
+# list_models <- list(cc_rs, mar_rs, mar_pmm_rs, mnar_rs, mnar_pmm_rs)
 # plot_models(list_models,
-#             mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
-```
-
+#             mod_name = c("Complete case", "MAR", "MAR_pmm", "MNAR", "MNAR_pmm"))
+```
+
+```{r}
+est <- testEstimates(as.mitml.result(mnar_pmm_rs), extra.pars = TRUE)
+```
+
+Display results in table.
+```{r}
+#| prompt: true
+est$estimates |> 
+  round(3) |>
+  kable()
+```
+
+```{r}
+#| prompt: true
+est$extra.pars |> 
+  round(3) |>
+  kable()
+```
 We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration.

### 2023-12-12 docs/manuscript.tex
Case study 3

@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, toprule=.15mm, rightrule=.15mm, arc=.35mm, left=2mm, leftrule=.75mm, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, breakable, colback=white, opacityback=0, left=2mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, rightrule=.15mm, leftrule=.75mm]
 
@@
 R> library(lme4)         # for multilevel modeling
+R> library(nlme)         # for multilevel modeling
 R> library(mice)         # for imputation
@@
 \begin{verbatim}
-R> Obesity |> 
-+  split(~Cluster) |>
-+  lapply(plot_pattern)
-\end{verbatim}
-
-\begin{verbatim}
-$`1`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-1.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-
-$`2`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-2.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-
-$`3`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-3.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-
-$`4`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-4.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-
-$`5`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-5.pdf}
-
-}
-
-\end{figure}
+R> # Obesity |> 
+R> #   split(~Cluster) |>
+R> #   lapply(plot_pattern)
+\end{verbatim}
 
@@
 \begin{verbatim}
-# Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-# performance::icc(Nulmodel)
+R> Nulmodel <- lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+R> # performance::icc(Nulmodel)
 \end{verbatim}
@@
 \begin{verbatim}
-# meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
-#                                ni.small = 100, prop.small = 0.4)
-# meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
-# meth_mar["Age"]<-"2l.2stage.pmm" 
+meth_mar <-
+  find.defaultMethod(
+    Obesity,
+    ind.clust = 1,
+    I.small = 7,
+    ni.small = 100,
+    prop.small = 0.4
+  )
+meth_mar["BMI"] <- "~ I(Weight / (Height)^2)"
+meth_mar["Age"] <- "2l.2stage.pmm"
 \end{verbatim}
@@
 \begin{verbatim}
-# pred_mar <- mice(Obesity, maxit = 0)$pred 
-# pred_mar[,"Cluster"] <- -2 # clustering variable
-# pred_mar[,"Time"] <- 0
-# pred_mar[pred_mar==1] <- 2
-# pred_mar[c("Height", "Weight"), "BMI"] <- 0
-# ggmice::plot_pred(pred_mar)
-# imp_mar <- mice::mice(data = Obesity, meth = meth_mar, pred = pred_mar,
-#                 m=10, seed = 123, printFlag = FALSE)
-\end{verbatim}
-
-\begin{verbatim}
-# summary(complete(imp_mar,"long")$Weight)
+pred_mar <- quickpred(Obesity)
+pred_mar[, "Cluster"] <- -2 
+pred_mar[, "Time"] <- 0
+pred_mar[pred_mar == 1] <- 2
+pred_mar[c("Height", "Weight"), "BMI"] <- 0
+plot_pred(pred_mar)
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics[width=0.7\textwidth,height=\textheight]{manuscript_files/figure-pdf/obesity-predmar-1.pdf}
+
+}
+
+\end{figure}
+
+\begin{verbatim}
+imp_mar <-
+  mice(
+    data = Obesity,
+    method = meth_mar,
+    predictorMatrix = pred_mar,
+    printFlag = FALSE
+  )
+\end{verbatim}
+
+\begin{verbatim}
+summary(complete(imp_mar,"long")$Weight)
+\end{verbatim}
+
+\begin{verbatim}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+  4.428  69.890  82.800  82.835  95.491 145.885 
 \end{verbatim}
@@
 \begin{verbatim}
-# meth_mar["Weight"]<-"2l.2stage.pmm" 
-# imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
-#                     m=10, seed = 123, printFlag = FALSE)
-\end{verbatim}
-
-\begin{verbatim}
-# summary(complete(imp_mar_pmm,"long")$Weight)
-# ggmice::plot_trace(imp_mar_pmm, "Weight")
-\end{verbatim}
+meth_mar["Weight"] <- "2l.2stage.pmm"
+imp_mar_pmm <- mice(
+  data = Obesity,
+  method = meth_mar,
+  predictorMatrix = pred_mar,
+  printFlag = FALSE
+)
+\end{verbatim}
+
+\begin{verbatim}
+summary(complete(imp_mar_pmm, "long")$Weight)
+\end{verbatim}
+
+\begin{verbatim}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+  28.35   71.04   83.65   83.87   95.66  134.61 
+\end{verbatim}
+
+\begin{verbatim}
+plot_trace(imp_mar_pmm, "Weight")
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/obesity-predmar_pmm1-1.pdf}
+
+}
+
+\end{figure}
+
+\begin{verbatim}
+ggmice(imp_mar_pmm, aes(.imp, Weight)) + 
+  geom_jitter()
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/obesity-predmar_pmm1-2.pdf}
+
+}
+
+\end{figure}
+
+Analyze the imputed data.
+
+\begin{verbatim}
+R> # fit <- with(
+R> #     imp_mar_pmm,
+R> #     lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+R> #     )
+R> fit <- with(
++    imp_mar_pmm,
++    lmer(BMI ~ Age + FamOb + Gender + (1 + Age | Cluster))
++    )
+R> est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
+\end{verbatim}
+
+Display results in table.
+
+\begin{verbatim}
+R> est$estimates |> 
++  round(3) |>
++  kable()
+\end{verbatim}
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1558}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1169}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1299}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2338}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Std.Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+t.value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+P(\textgreater\textbar t\textbar)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+RIV
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+FMI
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+(Intercept) & 28.445 & 2.262 & 12.578 & 158.191 & 0.000 & 0.189 &
+0.169 \\
+Age & 0.041 & 0.032 & 1.312 & 383.212 & 0.190 & 0.114 & 0.107 \\
+FamObyes & 0.000 & 0.316 & -0.001 & 207.642 & 0.999 & 0.161 & 0.147 \\
+GenderMale & -1.361 & 0.411 & -3.310 & 7.985 & 0.011 & 2.422 & 0.761 \\
+\end{longtable}
+
+\begin{verbatim}
+R> est$extra.pars |> 
++  round(3) |>
++  kable()
+\end{verbatim}
+
+\begin{longtable}[]{@{}lr@{}}
+\toprule\noalign{}
+& Estimate \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Intercept\textasciitilde\textasciitilde Intercept\textbar Cluster &
+19.862 \\
+Age\textasciitilde\textasciitilde Age\textbar Cluster & 0.002 \\
+Intercept\textasciitilde\textasciitilde Age\textbar Cluster & -0.067 \\
+Residual\textasciitilde\textasciitilde Residual & 26.010 \\
+ICC\textbar Cluster & 0.395 \\
+\end{longtable}
 
@@
 \begin{verbatim}
-# pred_mnar <- pred_mar
-# pred_mnar["Weight","Time"]<- -3
-# ggmice::plot_pred(pred_mnar)
-\end{verbatim}
+pred_mnar <- pred_mar
+pred_mnar["Weight","Time"] <- -3
+plot_pred(pred_mnar)
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/obesity-predmnar-1.pdf}
+
+}
+
+\end{figure}
 
@@
 \begin{verbatim}
-# meth_mnar <- meth_mar
-# meth_mnar["Weight"]<- "3l.2stage.heckman"
+meth_mnar <- meth_mar
+meth_mnar["Weight"] <- "2l.2stage.heckman"
 \end{verbatim}
@@
 \begin{verbatim}
-# imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
-#                 m=10, seed = 123, printFlag = FALSE)
-# summary(complete(imp_mnar,"long")$Weight)
+imp_mnar <- mice(
+  data = Obesity,
+  method = meth_mnar,
+  predictorMatrix = pred_mnar,
+  printFlag = FALSE
+)
+summary(complete(imp_mnar, "long")$Weight)
+\end{verbatim}
+
+\begin{verbatim}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+  16.26   70.95   85.21   86.14  100.14  190.95 
 \end{verbatim}
@@
 \begin{verbatim}
-# summary(complete(imp_mnar,"long")$Weight)
+summary(complete(imp_mnar, "long")$Weight)
+\end{verbatim}
+
+\begin{verbatim}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+  16.26   70.95   85.21   86.14  100.14  190.95 
 \end{verbatim}
@@
 \begin{verbatim}
-# imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
-#                      m=10, seed = 123, pmm = T,  printFlag = FALSE)
-\end{verbatim}
-
-We check the convergency of the results
-
-\begin{verbatim}
-# summary(complete(imp_mnar_pmm,"long")$Weight)
-# ggmice::plot_trace(imp_mar_pmm, "Weight")
-\end{verbatim}
+imp_mnar_pmm <-
+  mice(
+    data = Obesity,
+    method = meth_mnar,
+    predictorMatrix = pred_mnar,
+    pmm = T,
+    printFlag = FALSE
+  )
+\end{verbatim}
+
+We check the convergence of the results
+
+\begin{verbatim}
+summary(complete(imp_mnar_pmm, "long")$Weight)
+\end{verbatim}
+
+\begin{verbatim}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+  28.35   71.51   85.07   85.35   98.90  134.61 
+\end{verbatim}
+
+\begin{verbatim}
+plot_trace(imp_mar_pmm, "Weight")
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/obesity-predmnarp1-1.pdf}
+
+}
+
+\end{figure}
 
 After this modification we proceed to compare the effects on the model.
-We run the analysis model on each of the completed datasets as well
-asthe dataset where the incomplete values are removed (Complete Case
+We run the analysis model on each of the completed datasets as well as
+the dataset where the incomplete values are removed (Complete Case
 analysis, CC).
@@
 \begin{verbatim}
-# cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
-#              lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
-# mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
+# cc_rs <- with(setDT(Obesity)[complete.cases(Obesity), ],
+#               lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age |
+#                     Cluster))
+mar_rs <- with(
+  imp_mar, 
+  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+  )
+mar_pmm_rs <- with(
+    imp_mar_pmm,
+    lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+    )
+mnar_rs <- with(
+  imp_mnar, 
+  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+  )
+mnar_pmm_rs <- with(
+  imp_mnar_pmm,
+  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
+  )
+# list_models <- list(cc_rs, mar_rs, mar_pmm_rs, mnar_rs, mnar_pmm_rs)
 # plot_models(list_models,
-#             mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
-\end{verbatim}
+#             mod_name = c("Complete case", "MAR", "MAR_pmm", "MNAR", "MNAR_pmm"))
+\end{verbatim}
+
+\begin{verbatim}
+est <- testEstimates(as.mitml.result(mnar_pmm_rs), extra.pars = TRUE)
+\end{verbatim}
+
+Display results in table.
+
+\begin{verbatim}
+R> est$estimates |> 
++  round(3) |>
++  kable()
+\end{verbatim}
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1558}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1169}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1299}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2338}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Std.Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+t.value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+P(\textgreater\textbar t\textbar)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+RIV
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+FMI
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+(Intercept) & 28.880 & 2.391 & 12.078 & 56.876 & 0.000 & 0.361 &
+0.290 \\
+Age & 0.050 & 0.037 & 1.343 & 19.874 & 0.194 & 0.814 & 0.497 \\
+FamObyes & -0.027 & 0.490 & -0.056 & 11.685 & 0.957 & 1.410 & 0.642 \\
+GenderMale & -1.484 & 0.804 & -1.846 & 4.813 & 0.126 & 10.318 & 0.934 \\
+\end{longtable}
+
+\begin{verbatim}
+R> est$extra.pars |> 
++  round(3) |>
++  kable()
+\end{verbatim}
+
+\begin{longtable}[]{@{}lr@{}}
+\toprule\noalign{}
+& Estimate \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Intercept\textasciitilde\textasciitilde Intercept\textbar Cluster &
+19.111 \\
+Age\textasciitilde\textasciitilde Age\textbar Cluster & 0.001 \\
+Intercept\textasciitilde\textasciitilde Age\textbar Cluster & -0.046 \\
+Residual\textasciitilde\textasciitilde Residual & 30.071 \\
+ICC\textbar Cluster & 0.372 \\
+\end{longtable}
 

### 2023-12-12 docs/manuscript.tex
Integrate PR #42

@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, colback=white, opacityback=0, left=2mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, rightrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, arc=.35mm, rightrule=.15mm, breakable, opacityback=0, bottomrule=.15mm, colback=white, leftrule=.75mm, left=2mm]
 
@@
 
-When imputing clustered data, we initially focus on the research
-question and the intended analysis, assuming the presence of
-non-incomplete values. This assessment sheds light on research
-hypotheses and the contemplated main(s) statistical model(s) and it
-provides insights into the data's structure (refer to the level table),
-variable types (e.g., confounders, auxiliary variables), and other
-considerations about variables relationships such as interactions or
-polynomial terms.
+When dealing with incomplete clustered data, start by looking at your
+research questions and planned analysis. Pretend there are no missing
+data at first. This will help you understand your research hypotheses,
+the main statistical model, and give you insights into your data's
+structure (refer to the level table), variable types (e.g., confounders,
+auxiliary variables), and other considerations such as variable
+relationships, interactions or polynomial terms.
 
@@
 
-Next, it is necessary to explore the information available in the data.
-Exploring individual variables using tools such as histograms or QQ
-plots can help to delineate variable distributions and plausible ranges
-of values and also identify input errors or outliers. Evaluating
-interactions between predictors using scatter plots or conditional
-plots, assessing collinearity using VIF or correlations, can help to
-glimpse nonlinear relationships that may affect the original formulation
-of the main model. Further testing of assumptions related to the
-response variable, including variance (homogeneity using conditional box
-plots), independence (e.g., ACF, variograms) and response vs.~predictor
-relationships (conditional plots - x vs.~y), may serve to choose an
-imputation model more suitable to the data. In addition, the intraclass
-correlation coefficient (ICC) can be examined to assess cluster
-differences, aiding in the choice between the 2l and 1l methods for
-imputation.
-
-One can also explore missingness, specifically missing patterns at the
-cluster level to identify types of patterns (e.g., non-monotonic,
-systematic) to guide the selection of an appropriate imputation approach
-in terms of computational efficiency (e.g., simpler regression
-imputation versus FCS in univariate patterns). Similarly, these missing
-patterns can also be used to identify potential predictors in individual
-imputation models by means of inflow criteria (quantifying the
-connections between missing data in one variable and other observed
-variables) and outflow criteria (determining the connections between
-observed values in one variable and missing data in other variables).
-Although several packages exist for visualizing missing data, this
-tutorial focuses on the recently implemented ggmice package.
+Now, explore what's in your data. Use simple tools like histograms and
+QQ plots to understand variable distributions, plausible range of values
+and identify errors or outliers. Check interactions between predictors
+with scatter plots, look for collinearity with VIF or correlations
+plots, this may also help you to identify non-considered relationships
+that may affect the main model. You might also consider to test
+assumptions related to your response variable such as variance
+homogeneity, independence between observations (eg. ACF, variagrams).
+This will help you choose an imputation model that suits your data. In
+addition, the intraclass correlation coefficient (ICC) can be examined
+to assess cluster differences, aiding in the choice between the 2l and
+1l methods for imputation.
+
+Next, explore the missingness. Look at the
+\textbf{proportion of missing values} in the dataset variables. This
+helps find potential predictors and cut down on unnecessary variables in
+the imputation model. Doing this can lower the risk of multicollinearity
+or computational issues, especially with certain parametric imputation
+methods. You can also identify predictors for the imputation model by
+using inflow criteria to see connections between missing data in one
+variable and observed variables, and outflow criteria to identify
+connections between observed values in one variable and missing data in
+others.
+
+Check the missing patterns at cluster level, this can help you to select
+the most appropiated imputation approach in terms of computational
+efficiency (e.g., simpler regression imputation versus FCS in univariate
+patterns).
 
@@
 
-Prior to implementing multiple imputation, it is critical to assess the
-robustness of the estimation model when confronted with missing data.
-There are scenarios in which specific Maximum Likelihood (ML) estimation
-methods outperform Multiple Imputation (MI) methods. For instance, when
-the response variable is the sole incomplete variable, mixed models
-demonstrate robustness to missing data under the Missing at Random (MAR)
-assumption and with a correct variance-covariance
-specification.\cite{Molenberghs_2007}. Furthermore, depending on the
-proportion of missingness (usually is \textless5\%), opting for a
-simpler complete case analysis might be sufficient.
+Before diving into imputation, make sure your estimation model can
+handle missing data. Sometimes, simpler methods like complete case
+analysis might be suffice, especially if your missingness is low
+(usually \textless5\%). There are scenarios in which specific Maximum
+Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
+methods, for instance when the response variable is the sole incomplete
+variable, mixed models demonstrate robustness to missing data under the
+Missing at Random (MAR) assumption and with a correct
+variance-covariance specification.\cite{Molenberghs_2007}.
 
@@
 
-Before proceeding with the imputation model, it is necessary to filter
-the dataset to include only the minimal variables required for the main
-model. If additional procedures are conducted during the analysis, it is
-essential to include variables associated with these procedures, such as
-confounders in the case of balancing techniques. Other relevant
-variables, like instrumental variables or auxiliary variables, may
-enhance parameter estimates even if not included in the main model, as
-they could be linked to the probability of missingness for some
-incomplete variables.
-
-Furthermore, at the variable level, it is crucial to assess whether
-proxy variables should be considered or if deterministic imputation
-alone is adequate for imputing missing values. For instance, certain
-incomplete variables may not necessitate stochastic imputation methods
-like MICE. Instead, they can be effectively addressed through deductive
-imputation, where incomplete values are inferred from logical and
-deterministic relationships between variables. This approach is
-especially beneficial for variables that are functions of others, such
-as deriving a person's BMI from their weight and height. It also proves
-valuable for determining values for one-level variables from two-level
-ones, as seen in the context of Individual Participant Data (IPD). In
-such cases, incomplete information can be deduced from metadata, like
-inferring incomplete data about abortion in a country where abortion is
-illegal, or through cross-temporal or protocol-based deduction, such as
-imputing missing test values for deceased patients.
+Clean up your dataset before the imputation process. Keep initially the
+essential variables for your main model, but include extra variables if
+needed for specific procedures during analysis (e.g.~confounders on
+balancing procedures). Think about adding other useful variables, even
+if they're not in the main model, such as instrumental or auxiliary
+variables which might boost your parameter estimates, especially if
+they're associated to the probability of missingness for some incomplete
+variables.
+
+Figure out if you can directly impute incomplete variables by just use
+deductive imputation. This involves inferring missing values based on
+logicallogical connections between variables. It's especially handy for
+variables that depend on each other, like calculating BMI from weight
+and height.
+
+Deductive imputation is also useful for getting values for level-1
+variables from level-2 ones, like in Individual Participant Data (IPD).
+In these situations, you can guess missing information from metadata or
+by using or through deduction based on time or protocols. For example,
+you could deduce missing test values for patients who have passed away.
 
@@
 
-Various strategies can be employed in the imputation process to account
-for clustering, including the inclusion of a cluster indicator variable,
-conducting a separate imputation process for each cluster, or employing
-a simultaneous imputation method that considers clustering\cite{eddings}
-
-The choice of strategy depends primarily on the assumptions made in the
-main analysis and the constraints imposed by the analyzed data.
-Concerning data restrictions when the analysis do not contemplate an
-hierarchical model (eg. descriptive way), if there are few clusters with
-many observations per cluster, the cluster indicator or imputation
-separated by groups may be appropriate \cite{graham2009}. Conversely, an
-imputation model that simultaneously imputes all clusters using a
-hierarchical model can be employed when there are more clusters or fewer
-observations per cluster \cite{Allison_2002}.
-
-In a hierarchical imputation model, correlations between observations
-within clusters are modeled by a random effect, allowing estimation even
-with a limited number of observations per cluster. Multiple imputation
-models based on hierarchical models have been proposed, each with
-different assumptions. \cite{audigier}
-
-In the particular case that the analysis contemplates the use of a
-hierarchical model (i.e.~main model), before choosing a particular
-strategy, it is crucial to verify if the assumptions of the imputation
-model align with those of the main analysis. For example, using the
-cluster indicator strategy may introduce bias estimates when the model
-is based on a hierarchical structure \cite{taljaard2008,speidel2018}.
-Even if an imputation strategy congruent with the main model is
-preferred, it is essential to assess its appropriateness for the data,
-considering that less complex imputation strategies may lead to unbiased
+When it comes to handling clusters during the imputation process, you've
+got a few options. You can use a cluster indicator variable, run
+separate imputation for each cluster, or go for a simultaneous
+imputation method that takes clustering into account \cite{eddings}.
+
+Which strategy you choose depends on the assumptions in your main
+analysis and the limitations of your data. If your analysis do not use a
+hierarchical model (like a descriptive approach) and you have a small
+number of clusters with lots of observations in each, using a cluster
+indicator or separate imputation might be the way to go
+\cite{graham2009}. Conversely, if you have more clusters or fewer
+observations per cluster, you might want to try a simultaneous
+hierarchical imputation model \cite{Allison_2002}.
+
+In a hierarchical imputation model, random effects model the
+correlations between observations within clusters, making it possible to
+estimate even with a small number of observations per cluster. There are
+various proposed multiple imputation models based on hierarchical
+models, each with its own set of assumptions \cite{audigier}.
+
+If your analysis uses a hierarchical model, make sure the assumptions of
+your imputation model match up. For example, using a cluster indicator
+approach may lead to bias estimates if your model is based on a
+hierarchical structure \cite{taljaard2008,speidel2018}. Even if you
+prefer an imputation strategy that aligns with your main model, check if
+it suits your data; sometimes simpler strategies can give unbiased
 estimates in certain scenarios \cite{bailey2020}.
@@
 
-The initial step involves selecting the individual imputation model for
-each incomplete variable in the dataset. The mice package automatically
-proposes imputation methods based on the variable type for 1l variables.
-For 2l level variables, the micemd package's () function can be
-utilized. This function selects among the 2l imputation methods based on
-the size of the clusters and the proportion of missingness in each
-cluster.
-
-In addition to the package-defined imputation methods, users can specify
-custom methods using the ``I formula''. This flexibility allows for the
-calculation of derived variables internally or adjustments to imputation
-methods based on specific conditions, such as conditioning the
-imputation model on the level of an incomplete covariate (e.g.,
-pregnancy test on females).
+Start by choosing the imputation model for each incomplete variable in
+your dataset. The mice package suggests methods based on variable types
+for non-clustered variables, and for the cluster ones, you can use the
+micemd package's find.defaultMethod() function. This function selects
+from different 2l imputation methods based on cluster size and the
+proportion of missing data in each cluster.
+
+Besides the package-defined imputation methods, you can specify custom
+methods using the ``I formula''. This lets you calculate deterministic
+variables during the imputation or tweak imputation methods based on
+specific conditions, like conditioning the imputation model to the level
+of an incomplete covariate (e.g., a pregnancy test for females).
 
@@
 
-The imputation model can include additional terms than the main model
-terms that do not lead to uncongeniality. For instance, it is advisable
-to include the outcome variable in the imputation model for prediction
-variables \cite{moons2006a}. In cases where the outcome is
+The imputation model can incorporate additional terms compared to the
+main model without causing compatibility issues. For instance, it's
+recommended to include the outcome variable in the imputation model for
+prediction variables \cite{moons2006a}. In cases where the outcome is
 time-to-event, the Nelson-Aalen estimate of the time to the event should
-be included as a covariate in the imputation model {[}REF{]}.
-Furthermore, the inclusion of auxiliary variables, even if not part of
-the main model, can be associated with the probability of missingness,
-enhancing the likelihood of satisfying the Missing at Random (MAR)
-assumption and improving estimation efficiency \cite{hardt2012a}.
-
-The specification of imputation models is done on a variable basis,
-either using a prediction matrix where the type of prediction variable
-should be specified accordingly (see table) or through a list of
-formulas in the formula parameter. The latter option proves useful in
-formulating imputation models with polynomial terms or interactions
-compared to the prediction matrix specification, which requires the
-inclusion of additional terms as Just Another Variable (JAV)
-\cite{buuren2018a}.
-
-However, even when considering an interaction term, depending on the
-adapted individual imputation model, it has been suggested, for
-instance, for treatment interaction effects, to conduct separated
-imputation by treatment group \cite{zhang2023}. Additionally, there are
+be added as a covariate in the imputation model {[}REF{]}. Additionally,
+including auxiliary variables, even if not part of the main model, can
+be linked to the probability of missingness, improving the likelihood of
+meeting the Missing at Random (MAR) assumption and enhancing estimation
+efficiency \cite{hardt2012a}.
+
+Imputation models are specified on a variable basis, either using a
+prediction matrix in the pred parameter or through a list of formulas in
+the formula parameter. In the prediction matrix option, the type of each
+predictor variable is specified for each incomplete variable (see
+table).
+
+\begin{align}
+y_{ij} =& (\beta_0 + b_{0_i})+ \beta_1x_{1_{ij}}+\beta_2x_{2_{i.}} \nonumber \\
++& (\beta_3 +  b_{3_i})x_{3_{ij}} \nonumber \\
++& \beta_4x_{4_{ij}} + \beta_{m4}\overline{x_{4_{i.}}}\nonumber \\
++& (\beta_5 +  b_{5_i})x_{5_{ij}} +\beta_{m5}\overline{x_{5_{i.}}}\nonumber \\
+\end{align}
+
+\begin{tabular}{ll}
+     \toprule
+Type & Definition \\
+  \midrule
+-2  & Cluster variable, in this case the one defined by j index  \\
+ 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
+ 2  & Random variable, e.g., $x_{3_{ij}}$\\
+ 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
+ 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
+-3. & Random variable only included on selection model (Heckman model)\\
+-4. & Random variable only included on main model (Heckman model)\\
+      \bottomrule
+\end{tabular}
+
+\begin{itemize}
+\tightlist
+\item
+  3 and 4 type have been advised to used as the inclusion of the means
+  of the cluster is beneficial on FCS \cite{mistler2017}
+\end{itemize}
+
+Recipes have been proposed for imputing incomplete level-1 and level-2
+variables for hierarchical models \cite{buuren2018a} (\(\S\) 7.10),
+which can be convenient to follow when dealing with numerous variables
+(interaction terms at different levels) and models with many random
+effects that are prone to convergency problems due to overspecification
+in the imputation model. These rules were designed to ensure
+compatibility among the conditionally specified imputation models and
+congeniality between the imputation and the main model. However, they
+may not be applicable to all 2l imputation methods; for instance, the
+2l.2stage imputation methods of the micemd package only allow the
+inclusion of random predictor variables (2).
+
+On the other hand, the formula option is useful in specifying complex
+imputation models with polynomial terms or interactions and compared
+with the prediction matrix method do not requires the inclusion of
+additional terms as Just Another Variable (JAV)
+\cite{buuren2018a}(\(\S\) 6.4).
+
+For some interaction terms, it has been suggested, for instance, for
+treatment interaction effects, to conduct separate imputation by
+treatment group \cite{zhang2023}. Additionally, it can be used
 imputation models based on random forest or deep learning that can
-handle interaction and non-linear terms that do not require the explicit
-inclusion of the non-linear terms.
+handle interaction and non-linear terms without requiring the explicit
+specification of an imputation model.
 
@@
 During the imputation process, certain issues may arise that halt the
-process. In hierarchical model imputations, many issues are linked to
-methods based on parametric hierarchical models, which may struggle to
-estimate a model midway through the process. In such cases, it is
-advisable to examine the imputation log file to identify variables with
-imputation issues. Subsequently, reducing the number of predictors in
-the imputation model could be explored, either by using functions like
-quickpred or by considering variable transformations, such as scaling.
-Adjusting the level of the hierarchical model (e.g., using a homogenous
-variance model or a 1l model) can also be beneficial. Additionally,
-checking the range of imputed variables is crucial, as excessively large
-imputed values for one predictor could lead to convergence issues in
-other variables. This can be mitigated by including post-processing
-specifications on problematic variables or by using imputation models
-like Predicted Mean Matching (PMM), ensuring imputed values align with
-the observable values. In certain cases, a separate imputation strategy
-may be considered. For instance, in analyses involving multiple
-endpoints, conducting separate imputation processes for each endpoint
-might be preferable to a unified imputation process.
+process. In hierarchical model imputations, many issues are related to
+over fitting the imputation model. To troubleshoot, it is recommended to
+inspect the imputation log file for variables causing problems. One
+approach to address this is to reduce the number of predictors, by using
+step by step the previous referred recepies or by using functions like
+quickpred. Another option is to consider variable transformations, such
+as scaling when model is invariant to linear transformations e.g.,
+random intercept models. Also adjusting the level of the hierarchical
+model (e.g., using a homogeneous variance imputation method or a 1l
+model) can also be beneficial.
+
+It is crucial to check the range of imputed variables, as excessively
+large imputed values for one predictor may trigger convergence issues in
+other variables. To tackle this, including post-processing
+specifications on problematic variables or utilizing imputation models
+like Predicted Mean Matching (PMM) can ensure that imputed values align
+with observable values.
+
+In some situations, adopting a separate imputation strategy might be
+worth considering. For example, in analyses involving multiple
+endpoints, conducting distinct imputation processes for each endpoint
+could be more effective than a unified imputation approach.
 
@@
 \end{verbatim}
+
+Given the \(j\)-th student belonging to the \(i\)-th school, the main
+model can be formulated as:
+
+\[ popular_{ij} = (\beta_0 + b_i) + \beta_1sex_{ij}+ \beta_2texp_{ij}+\epsilon_{ij}\]
+where \(\epsilon_{ij}\) corresponds to the error term.
 
@@
 \end{verbatim}
+
+In this dataset all the variables are level-1 (patient), except by the
+level-2 (study) type variable. The analysis model for this dataset is a
+prediction model with \texttt{mort} as the outcome. In this tutorial
+we'll estimate the adjusted prognostic effect of \texttt{ct} on
+mortality outcomes. The estimand is the adjusted odds ratio for
+\texttt{ct}, after including \texttt{type}, \texttt{age}
+\texttt{motor\_score} and \texttt{pupil}. Therefore the main model for
+the \(i\)-th patient from the \(j\)-th study can be described as:
+
+\[ mort_{ij} = (\beta_0 + b_i) + \beta_1type_{.j}+ \beta_2age_{ij}+ \beta_3motorscore_{ij} + \beta_4pupil_{ij} + \beta_5ct_{ij}+\epsilon_{ij}\]
+where \(\epsilon_{ij}\) corresponds to the error term.
 

### 2023-12-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

@@
 \usepackage{ulem}
+\usepackage{xcolor}
 
@@
 non-ignorable \\
-& \citep{rubi76, meng94}. \\
+& \citep{rubi76, meng1994}. \\
 & \\
@@
 
-Prior to implementing multiple imputation, it is critical to assess the
-robustness of the estimation model when confronted with missing data.
-\cite{Alisson_2002} highlighted scenarios in which specific Maximum
+Before diving into imputation, make sure your estimation model can
+handle missing data. Sometimes, simpler methods like complete case
+analysis might be suffice, especially if your missingness is low
+(usually \textless5\%). There are scenarios in which specific Maximum
 Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
-methods. Notably, when the response variable is the sole incomplete
+methods, for instance when the response variable is the sole incomplete
 variable, mixed models demonstrate robustness to missing data under the
 Missing at Random (MAR) assumption and with a correct
-variance-covariance specification.\cite{Molenberghs_2007}. Furthermore,
-depending on the proportion of missingness (usually is \textless5\%),
-opting for a simpler complete case analysis might be sufficient.
+variance-covariance specification.\cite{Molenberghs_2007}.
 
@@
 
-Various strategies can be employed in the imputation process to account
-for clustering, including the inclusion of a cluster indicator variable,
-conducting a separate imputation process for each cluster, or employing
-a simultaneous imputation method that considers clustering (Stata:
-Clustering and MI Impute) TODO: Replace ref.
-
-The choice of strategy depends primarily on the assumptions made in the
-main analysis and the constraints imposed by the analyzed
-data.Concerning data restrictions, if there are few clusters with many
-observations per cluster, the cluster indicator or imputation separated
-by groups may be appropriate \cite{Graham_2009}. Conversely, an
-imputation model that simultaneously imputes all clusters using a
-hierarchical model can be employed when there are more clusters or fewer
-observations per cluster \cite{Allison_2002}.
-
-In a hierarchical imputation model, correlations between observations
-within clusters are modeled by a random effect, allowing estimation even
-with a limited number of observations per cluster. Multiple imputation
-models based on hierarchical models have been proposed, each with
-different assumptions.
-
-Before choosing a particular strategy, it is crucial to verify if the
-assumptions of the imputation model align with those of the main
-analysis. For example, using the cluster indicator strategy may
-introduce bias estimates when the model is based on a hierarchical
-structure \cite{Taaljard_2008}. Even if an imputation strategy congruent
-with the main model is preferred, it is essential to assess its
-appropriateness for the data, considering that less complex imputation
-strategies may lead to unbiased estimates in certain scenarios
-\textbackslash cite(Bailey\_2020).
+When it comes to handling clusters during the imputation process, you've
+got a few options. You can use a cluster indicator variable, run
+separate imputation for each cluster, or go for a simultaneous
+imputation method that takes clustering into account \cite{eddings}.
+
+Which strategy you choose depends on the assumptions in your main
+analysis and the limitations of your data. If your analysis do not use a
+hierarchical model (like a descriptive approach) and you have a small
+number of clusters with lots of observations in each, using a cluster
+indicator or separate imputation might be the way to go
+\cite{graham2009}. Conversely, if you have more clusters or fewer
+observations per cluster, you might want to try a simultaneous
+hierarchical imputation model \cite{Allison_2002}.
+
+In a hierarchical imputation model, random effects model the
+correlations between observations within clusters, making it possible to
+estimate even with a small number of observations per cluster. There are
+various proposed multiple imputation models based on hierarchical
+models, each with its own set of assumptions \cite{audigier}.
+
+If your analysis uses a hierarchical model, make sure the assumptions of
+your imputation model match up. For example, using a cluster indicator
+approach may lead to bias estimates if your model is based on a
+hierarchical structure \cite{taljaard2008,speidel2018}. Even if you
+prefer an imputation strategy that aligns with your main model, check if
+it suits your data; sometimes simpler strategies can give unbiased
+estimates in certain scenarios \cite{bailey2020}.
 
@@
 The imputation model must be congenial with the main model
-\cite{Meng_1994}. Congeniality issues arise when the imputation model
-and the main model make different assumptions, often due to the omission
-of a polynomial or interaction term or the use of transformed variables.
-
-The imputation model can include additional terms than the main model
-terms that do not lead to uncongeniality. For instance, it is advisable
-to include the outcome variable in the imputation model for prediction
-variables \cite{Moons}. In cases where the outcome is time-to-event, the
-Nelson-Aalen estimate of the time to the event should be included as a
-covariate in the imputation model {[}REF{]}. Furthermore, the inclusion
-of auxiliary variables, even if not part of the main model, can be
-associated with the probability of missingness, enhancing the likelihood
-of satisfying the Missing at Random (MAR) assumption and improving
-estimation efficiency \cite{Hardt_2012}.
-
-The specification of imputation models is done on a variable basis,
-either using a prediction matrix where the type of prediction variable
-should be specified accordingly (see table) or through a list of
-formulas in the formula parameter. The latter option proves useful in
-formulating imputation models with polynomial terms or interactions
-compared to the prediction matrix specification, which requires the
-inclusion of additional terms as Just Another Variable (JAV).
-
-However, even when considering an interaction term, depending on the
-adapted individual imputation model, it has been suggested, for
-instance, for treatment interaction effects, to conduct separated
-imputation by treatment group \cite{Zhang_2023}. Additionally, there are
+\cite{meng1994}. Congeniality issues arise when the imputation model and
+the main model make different assumptions, often due to the omission of
+a polynomial or interaction term or the use of transformed variables.
+
+The imputation model can incorporate additional terms compared to the
+main model without causing compatibility issues. For instance, it's
+recommended to include the outcome variable in the imputation model for
+prediction variables \cite{moons2006a}. In cases where the outcome is
+time-to-event, the Nelson-Aalen estimate of the time to the event should
+be added as a covariate in the imputation model {[}REF{]}. Additionally,
+including auxiliary variables, even if not part of the main model, can
+be linked to the probability of missingness, improving the likelihood of
+meeting the Missing at Random (MAR) assumption and enhancing estimation
+efficiency \cite{hardt2012a}.
+
+Imputation models are specified on a variable basis, either using a
+prediction matrix in the pred parameter or through a list of formulas in
+the formula parameter. In the prediction matrix option, the type of each
+predictor variable is specified for each incomplete variable (see
+table).
+
+\begin{align}
+y_{ij} =& (\beta_0 + b_{0_i})+ \beta_1x_{1_{ij}}+\beta_2x_{2_{i.}} \nonumber \\
++& (\beta_3 +  b_{3_i})x_{3_{ij}} \nonumber \\
++& \beta_4x_{4_{ij}} + \beta_{m4}\overline{x_{4_{i.}}}\nonumber \\
++& (\beta_5 +  b_{5_i})x_{5_{ij}} +\beta_{m5}\overline{x_{5_{i.}}}\nonumber \\
+\end{align}
+
+\begin{tabular}{ll}
+     \toprule
+Type & Definition \\
+  \midrule
+-2  & Cluster variable, in this case the one defined by j index  \\
+ 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
+ 2  & Random variable, e.g., $x_{3_{ij}}$\\
+ 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
+ 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
+-3. & Random variable only included on selection model (Heckman model)\\
+-4. & Random variable only included on main model (Heckman model)\\
+      \bottomrule
+\end{tabular}
+
+\begin{itemize}
+\tightlist
+\item
+  3 and 4 type have been advised to used as the inclusion of the means
+  of the cluster is beneficial on FCS \cite{mistler2017}
+\end{itemize}
+
+Recipes have been proposed for imputing incomplete level-1 and level-2
+variables for hierarchical models \cite{buuren2018a} (\(\S\) 7.10),
+which can be convenient to follow when dealing with numerous variables
+(interaction terms at different levels) and models with many random
+effects that are prone to convergency problems due to overspecification
+in the imputation model. These rules were designed to ensure
+compatibility among the conditionally specified imputation models and
+congeniality between the imputation and the main model. However, they
+may not be applicable to all 2l imputation methods; for instance, the
+2l.2stage imputation methods of the micemd package only allow the
+inclusion of random predictor variables (2).
+
+On the other hand, the formula option is useful in specifying complex
+imputation models with polynomial terms or interactions and compared
+with the prediction matrix method do not requires the inclusion of
+additional terms as Just Another Variable (JAV)
+\cite{buuren2018a}(\(\S\) 6.4).
+
+For some interaction terms, it has been suggested, for instance, for
+treatment interaction effects, to conduct separate imputation by
+treatment group \cite{zhang2023}. Additionally, it can be used
 imputation models based on random forest or deep learning that can
-handle interaction and non-linear terms that do not require the explicit
-inclusion of the non-linear terms.
+handle interaction and non-linear terms without requiring the explicit
+specification of an imputation model.
 
@@
 
-Before commencing the analysis, it is imperative to verify the
-convergence of the imputation process. This is typically achieved
-through trace plots or by evaluating prediction plots.
+Before starting into the analysis of each imputed dataset, it is crucial
+to validate the convergence of the imputation process. This is commonly
+accomplished through trace plots that depict the mean and variance of
+the incomplete variables across iterations. These plots serve to uncover
+potential circular issues or the need for additional iterations.
+Additionally, it is also important to verify that imputed values fall
+within a plausible range and also to check the distribution of imputed
+variables, ensuring that the imputed variable distribution aligns with
+the distribution of observed values (under the MAR assumption). An
+alternative approach involves assessing the prediction accuracy of the
+imputation method \cite{cai2023}.
 
@@
 non-informative case in hierarchical datasets, imputation methods based
-on the Heckman method, such as those proposed by Hammon (2021) and Munoz
-(2023), can be considered.
+on the Heckman method can be
+considered.\cite{hammon2020,hammon2022,munoz2023}
 
@@
 teacher. The analysis model corresponding to this dataset is multilevel
-regression with random intercepts, random slopes and a cross-level
-interaction. The outcome variable is \texttt{popular}, which is
-predicted from the unit-level variable \texttt{sex} and the
-cluster-level variable \texttt{texp}:
+regression with only random intercepts. The outcome variable is
+\texttt{popular}, which is predicted from the level-1 (student) variable
+\texttt{sex} and the level-2 (school) variable \texttt{texp}. Given the
+\(j\)-th student belonging to the \(i\)-th school, the main model can be
+formulated as:
+
+\[ popular_{ij} = (\beta_0 + b_i) + \beta_1sex_{ij}+ \beta_2texp_{ij}+\epsilon_{ij}\]
+where \(\epsilon_{ij}\) corresponds to the error term.
 
@@
 
-The analysis model for this dataset is a prediction model with
-\texttt{mort} as the outcome. In this tutorial we'll estimate the
-adjusted prognostic effect of \texttt{ct} on mortality outcomes. The
-estimand is the adjusted odds ratio for \texttt{ct}, after including
-\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
-into the analysis model:
+In this dataset all the variables are level-1 (patient), except by the
+level-2 (study) type variable. The analysis model for this dataset is a
+prediction model with \texttt{mort} as the outcome. In this tutorial
+we'll estimate the adjusted prognostic effect of \texttt{ct} on
+mortality outcomes. The estimand is the adjusted odds ratio for
+\texttt{ct}, after including \texttt{type}, \texttt{age}
+\texttt{motor\_score} and \texttt{pupil}. Therefore the main model for
+the \(i\)-th patient from the \(j\)-th study can be described as:
+
+\[ mort_{ij} = (\beta_0 + b_i) + \beta_1type_{.j}+ \beta_2age_{ij}+ \beta_3motorscore_{ij} + \beta_4pupil_{ij} + \beta_5ct_{ij}+\epsilon_{ij}\]
+where \(\epsilon_{ij}\) corresponds to the error term.
 
@@
 \begin{CodeInput}
-R> #data("data_heckman", package = "micemd")
-R> #dat <- data_heckman
+R> data("Obesity", package = "micemd")
 \end{CodeInput}
@@
 
-We are also contemplating the utilisation of the ???pmm??? option, as
-the values imputed using a fully parametric method may be implausibly
-low for some patients.
+We are also contemplating the utilisation of the predictive mean
+matching (pmm) option, as the values imputed using a fully parametric
+method may be implausibly low for some patients.
 
@@
 which can be particularly useful in this case. Several methods,
-including those by \cite{Galimar_2017,Hammon_2021}, and the recently a
-Heckman method designed for two-level data, allow for variations in
-intercepts and exposure effects (random intercept and slope).
+including those by \cite{galimard2018}, and the recently a Heckman
+method designed for two-level data, allow for variations in intercepts
+and exposure effects (random intercept and slope) \cite{munoz2023}.
 
@@
 a more comprehensive understanding of the model and the exclusion
-restriction, see \cite{Munoz_2022}.
+restriction, see \cite{munoz2023a}.
 
@@
 R> meth_mnar <- meth_mar
-R> meth_mnar["Weight"]<- "3l.2stage.heckman"
+R> meth_mnar["Weight"]<- "2l.2stage.heckman"
 \end{CodeInput}
@@
 
-\hypertarget{additional}{%
-\section{Additional}\label{additional}}
-
-The imputation of these data is based on the
-\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
-Heckman Github repo}
-
-Visualize missing data pattern:
-
-The matrix only shows the predictors for the main model, not the
-selection model.
-
-TODO: explain exclusion restriction.
-
-ORDER:
-
-\begin{itemize}
-\item
-  summary
-\item
-  congeniality, then in hierarchical models
-\item
-  look whether we can fit cong. back in the main body
-\item
-  alt. methods
-\item
-  conclusion: mice is really easy!
-\item
-  Additional levels of clustering
-\item
-  More complex data types: timeseries and polynomial relationship in the
-  clustering.
-\item
-  FIML vs MI
-\end{itemize}
-
-An alternative approach to missing data is to use Full Information
-Maximum Likelihood (FIML). This method does not require the imputation
-of any missing values. Whereas MI consists of imputation, analyses and
-pooling steps, FIML analyses the data in a single step. When the
-assumptions are met the two approaches should produce equivalent
-results. {[}REF{]} As FIML requires specialised software, not all
-analyses can be performed with standard software. {[}REF{]}
-
-\begin{itemize}
-\tightlist
-\item
-  Survival / TTE, this could be put in the paragraph on congeniality
-\end{itemize}
-
-When the outcome is time-to-event, the Nelson-Aalen estimate of the time
-to event should be included as a covariate in the imputation model
-{[}REF{]}
-
-ORDER:
-
-\begin{itemize}
-\item
-  summary
-\item
-  congeniality, then in hierarchical models
-\item
-  look whether we can fit cong. back in the main body
-\item
-  alt. methods
-\item
-  conclusion: mice is really easy!
-\item
-  Additional levels of clustering
-\item
-  More complex data types: timeseries and polynomial relationship in the
-  clustering.
-\item
-  FIML vs MI
-\end{itemize}
-
-An alternative approach to missing data is to use Full Information
-Maximum Likelihood (FIML). This method does not require the imputation
-of any missing values. Whereas MI consists of imputation, analyses and
-pooling steps, FIML analyses the data in a single step. When the
-assumptions are met the two approaches should produce equivalent
-results. {[}REF{]} As FIML requires specialised software, not all
-analyses can be performed with standard software. {[}REF{]}
-
-\begin{itemize}
-\tightlist
-\item
-  Survival / TTE, this could be put in the paragraph on congeniality
-\end{itemize}
-
-When the outcome is time-to-event, the Nelson-Aalen estimate of the time
-to event should be included as a covariate in the imputation model
-{[}REF{]}
-
-In hierarchical datasets, clustering is a concern because the
-homoscedasticity in the error terms cannot be assumed across clusters
-and the relationship among variables may vary at different hierarchical
-levels. When multiple imputation is used to deal with missing data, as
-the imputation and analysis process is performed separately, it is
-necessary that imputation model being congenial with the main analysis
-model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
-structure also imputation model should do it (Audigier, 2021). Not
-including clustering into the imputation process may lead to effect
-estimates with smaller standard errors and inflated type I error.
-
-There are different strategies that can be adopted in the imputation
-process that account for clustering: inclusion of cluster indicator
-variable, performing a separate imputation process for each cluster, or
-performing a simultaneous imputation process by using an imputation
-method that accounts for clustering.(Stata:
-\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
-TODO: replace ref.
-
-The selection of each strategy depends mainly on the assumptions in the
-main analysis and also on the restriction of the analyzed data.
-
-Regarding the restrictions imposed by the data, for instance, the use of
-cluster indicator variables is restricted in datasets where there are
-not many clusters and many observations per cluster (Graham, 2009). The
-last restriction is also required when imputations are performed on each
-cluster separately. When this restriction cannot be achieved, one can
-use an imputation model that simultaneously imputes all clusters using a
-hierarchical model (Allison 2002).
-
-Under this hierarchical imputation model, observations within clusters
-are correlated and this correlation is modeled by a random effect so the
-hierarchical model can be estimated even when there are few observations
-per cluster. However, this strategy is best suited for balanced data
-(Grund, 2017) and when random effects model is appropriated, i.e.~the
-number of clusters is adequate. (Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main
-model, for instance by using the cluster indicator strategy may lead to
-bias estimates when the model is based on a hierarchical model
-(Taaljard,2008). Even when an imputation strategy congenial with the
-main model is preferred, it is important to consider whether it is
-appropriate for the data as a less complex imputation strategies may
-also lead to unbiased estimates in certain scenarios(Bailey 2020). For
-instance, in causal effect analysis, separately imputation may lead to
-smaller bias when the size of the smaller exposure cluster is large,
-compared with an imputation model that includes exposure-confounder
-interactions. (Zhang,2023).
-
 \hypertarget{conclusion}{%
 \section{Conclusion}\label{conclusion}}
+
+This paper is dedicated to exploring the imputation process for
+incomplete datasets, with a primary focus on utilizing a hierarchical
+model for analysis. Initially, users are encouraged to consider the main
+analysis within the context of the incomplete dataset, along with
+insights provided by domain experts, to gain a better understanding of
+variable relationships. Employing clear data visualization is
+instrumental in comprehending the missing data patterns, establishing a
+missing mechanism, and aiding in the selection of suitable imputation
+methods.
+
+The ``Mice'' and ``mice''-based R packages offer a range of imputation
+methods tailored for hierarchical data, easily adaptable to the
+dataset's structure. Before proceeding with the analysis of the imputed
+dataset, it is essential to assess the convergence of the imputation
+method. This evaluation can reveal issues such as circular problems, the
+need for additional iterations, or challenges associated with the chosen
+imputation method.
 
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
-
-Table 3: Notation
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Formula \pkg{lme4}}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Details}
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-\texttt{y\ \textasciitilde{}\ x1\ +\ (1\ \textbar{}\ g1)} & Fixed
-\texttt{x1} predictor with random intercept \\
-& varying among \texttt{g1} \\
-\texttt{y\textasciitilde{}x1*x2+\ (1\textbar{}\ g1)} & Interactions of
-\texttt{x1} and \texttt{x2} only in fixed effect \\
-\texttt{y\ \textasciitilde{}\ x1*x2+\ (x2\textbar{}\ g1)} & Interactions
-of \texttt{x1} and \texttt{x2} only in fixed effect \\
-& with slope of \texttt{x2} randomly varying among \texttt{g1} \\
-\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1*x2\textbar{}\ g1)} &
-variance-covariance matrix estimated only with \\
-& the variance terms of intercept, slope of \texttt{x1}, \\
-& slope of \texttt{x2} and interaction \texttt{x1*x2} \\
-\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1\ \textbar{}\ g1)+\ (x2\textbar{}\ g1)}
-& variance-covariance matrix estimated separately, \\
-& i.e, one for intercept and \texttt{x1} and another for \\
-& intercept and \texttt{x2} \\
-\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ +\ x1\ \textbar{}\ g1)}
-& Fixed \texttt{x1} with correlated random intercept and \\
-& random slope of \texttt{x} \\
-\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ \textbar{}\ g1)\ +\ (0\ +\ x1\ \textbar{}\ g1)}
-& Fixed \texttt{x1} with uncorrelated random intercept \\
-& and random slope of \texttt{x1} \\
-\texttt{y\ \textasciitilde{}\ (1\ \textbar{}\ g1)\ +\ (1\ \textbar{}\ g2)}
-& Random intercept varying among \texttt{g1} and among
-\texttt{g2\ \ \ \textbar{}\ \textbar{}}y \textasciitilde{} (1 \\
-& \\
-\end{longtable}
-
+\renewcommand\refname{References}
 \bibliography{../References/multilevelmice.bib}

### 2023-12-12 R/heckman.R
Clean up repo

@@ -1,238 +1,0 @@
-mice.impute.HeckmanIpdma<- function(y,ry,x, type, shrinkage_o=TRUE,shrinkage_s=TRUE,...){
-  ##first 3 arguments are mandatory in mice
-  ##type = how a var is used in the predictor matrix, new type is -3 for exclusion restriction var
-  ##shrinkage in outcome/selection, may not be necessary 
-  
-  # 0. Define functions ----
-  
-  # 0.1. CopulaIPD: Apply Binomial or Gaussian model depending on y type
-  ## this is fitting the copula equation in each cluster
-  copulaIPD <- function(data=data,formula=formula,y=y){
-    data$y<-as.numeric(as.character(data$y))
-    
-    if(class(y) == "numeric"){ #y continuos
-      fit <- try(GJRM::copulaSampleSel(formula=formula, data=data), silent=TRUE)
-      if(class(fit) == "try-error"){ # model is not estimable
-        fit <- NA}
-    }else if (class(y)=="factor" & nlevels(y)==2){ #y binomial
-      fit <- try(GJRM::SemiParBIV(formula=formula, data=data, Model="BSS"), silent=TRUE)
-      if(class(fit) == "try-error"){
-        fit <- NA}
-    }
-    return(fit)
-  }
-  
-  
-  # 0.2. cov_mat_vector: Transform covariance matrix in a ordered vector
-  ## additional function to transform the outcome for the copula, used later to calculate the marginals for the random effect param
-  cov_mat_vector <- function(cov_mat,vnames){
-    
-    cov_mat[upper.tri(cov_mat)] <- "Up"
-    cov_vec <- as.vector(cov_mat[vnames,vnames])
-    cov_vec <- as.numeric(cov_vec[cov_vec!="Up"])
-    return(cov_vec)
-  }
-  
-  # 0.3. draw_theta_psi_mar: Estimate true effect size and draw a marginal theta and psi=var(theta) .
-  ## psi is the standard error
-  draw_theta_psi_mar<-function(coef_mat_s,Vb_list,vnames=NULL){
-    
-    if (is.null(vnames)){
-      vnames <- colnames(coef_mat_s)} #use all set of parameters
-    
-    # Get covariance matrix
-    coef_mat_s <- coef_mat_s[,vnames]
[…]

### 2023-12-12 R/plot_imputed.R

@@ -1,225 +1,0 @@
-# # plotting functions for imputed data
-# 
-# # basic pre-processing of imputations
-# plot_imps <- function(imp, x, y = NULL) {
-#   # parse inputs
-#   if (is.null(y)) {
-#     y <- x
-#   }
-#   # combine observed and imputed data
-#   xy_obs <- imp$data %>%
-#     cbind(datapoint = "observed",
-#           .imp = 0,
-#           .id = 1:nrow(.),
-#           .) %>%
-#     .[!is.na(imp$data[[x]]) &
-#         !is.na(imp$data[[y]]),]
-#   xy_imps <- imp %>%
-#     mice::complete("long") %>%
-#     cbind(datapoint = "imputed", .) %>%
-#     .[.$.id %nin% xy_obs$.id, ]
-#   xy_dat <- rbind(xy_obs, xy_imps) %>%
-#     dplyr::mutate(datapoint = factor(datapoint, levels = c("observed", "imputed")))
-#   # initialize plot
-#   p <- xy_dat %>%
-#     ggplot2::ggplot() +
-#     ggplot2::theme_classic() +
-#     ggplot2::theme(legend.position = "top") +
-#     ggplot2::scale_color_manual(
-#       values = c(
-#       "observed" = mice:::mdc(1),
-#       # "missing" = mice:::mdc(2),
-#       "imputed" = mice:::mdc(2)
-#       )
-#   ) +
-#     ggplot2::scale_fill_manual(
-#     values = c(
-#       "observed" = mice:::mdc(1),
-#       # "missing" = mice:::mdc(2),
-#       "imputed" = mice:::mdc(2)
-#     ))
-#   # output
-#   return(p)
-# }
-# 
-# # stripplot
-# plot_strip <- function(imp, x) {
-#   # plot individual values (stripplot)
-#   p <- imp %>% plot_imps(x) +
-#     ggplot2::geom_jitter(
-#       ggplot2::aes(
-#         y = factor(.imp, ordered = TRUE),
-#         x = .data[[x]],
-#         color = datapoint
-#       ),
-#       height = 0.25,
-#       width = 0
-#     ) +
-#     scale_y_discrete(limits = rev) +
-#     ggplot2::ylab("Imputation (0 = observed data)")
-#   # output
-#   return(p)
-# }
-# 
-# # boxplot
-# plot_box <- function(imp, x, cluster = NULL, strip = FALSE) {
-#   # plot
-#   if (strip){
-#     p <- imp %>% plot_imps(x) +
[…]

### 2023-12-12 docs/template/template.tex
Template manuscript

--- 2023-12-12 docs/manuscript.tex
@@
   pdfauthor={Hanne I. Oberman; Johanna Muñoz; Valentijn M.T. de Jong; Gerko Vink; Thomas P.A. Debray},
-  pdfkeywords={missing data, multilevel, clustering, mice, R},
+  pdfkeywords={JSS, style guide, comma-separated, not capitalized, R},
   colorlinks=true,
@@
 %% an abstract and keywords
-\Abstract{This tutorial illustrates the imputation of incomplete
-multilevel data with the \proglang{R} packackage \pkg{mice}. Our scope
-is only simple multilevel models, to show how imputation can yield less
-biased estimates from incomplete clustered data. More complex models can
-be accomodated, but are outside the scope of this paper. Incomplete
-multilevel data requires careful consideration of the missing data
-problem and analysis strategy. In this tutorial, we focus on a popular
-strategy for accommodating missingness in multilevel data: replacing the
-missing data with one or more plausible values, i.e.,
-imputation.Imputation separates the missing data problem from the main
-analysis and the completed data can be analyzed as if it has been fully
-observed. This tutorial illustrates the imputation of incomplete
-multilevel data with the statistical programming language R. We aim to
-show how imputation can yield less biased estimates from incomplete
-clustered data. We provide practical guidelines and code snippets for
-different missing data situations, including non-ignorable missingness
-mechanisms. For brevity, we focus on multilevel imputation using chained
-equations with the R mice package and its adjacent packages.}
+\Abstract{This short article illustrates how to write a manuscript for
+the \emph{Journal of Statistical Software} (JSS) using its {\LaTeX}
+style files. Generally, we ask to follow JSS's style guide and FAQs
+precisely. Also, it is recommended to keep the {\LaTeX} code as simple
+as possible, i.e., avoid inclusion of packages/commands that are not
+necessary. For outlining the typical structure of a JSS article some
+brief text snippets are employed that have been inspired by
+\citet{ZeileisKleiberJackman2008}, discussing count data regression in
+\proglang{R}. Editorial comments and instructions are marked by vertical
+bars.}
 
 %% at least one keyword must be supplied
-\Keywords{missing
-data, multilevel, clustering, \pkg{mice}, \proglang{R}}
-\Plainkeywords{missing data, multilevel, clustering, mice, R}
+\Keywords{JSS, style guide, comma-separated, not
+capitalized, \proglang{R}}
+\Plainkeywords{JSS, style guide, comma-separated, not capitalized, R}
 
@@
 Utrecht The Netherlands\\
-E-mail: \email{h.i.oberman@uu.nl}\\
-URL: \url{https://www.hanneoberman.github.io}\\
+E-mail: \email{Achim.Zeileis@R-project.org}\\
+URL: \url{https://www.zeileis.org/}\\
 \\~
@@
 \hypertarget{sec-intro}{%
-\section{Introduction: Clustering and incomplete data}\label{sec-intro}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  missing data occur often in data with human subjects
-\item
-  missing data may be resolved, but need to be handled in accordance
-  with the analysis of scientific interest
-\item
-  in human-subjects research, there is often clustering, which may be
-  captured with multilevel modeling techniques
-\item
-  if the analysis of scientific interest is a multilevel model, the
-  missing data handling method should accommodate the multilevel
-  structure of the data
-\item
-  both missingness and multilevel structures require advanced
-  statistical techniques
-\item
-  this tutorial sets out to facilitate empirical researchers in
-  accommodating both multilevel structures as well as missing data.
-\item
-  we illustrate the use of the software by means of three case studies
-  from the social and biomedical sciences.
-\end{enumerate}
-
-In hierarchical datasets, clustering is a concern because the
-homoscedasticity in the error terms cannot be assumed across clusters
-and the relationship among variables may vary at different hierarchical
-levels. When multiple imputation is used to deal with missing data, as
-the imputation and analysis process is performed separately, it is
-necessary that imputation model being congenial with the main analysis
-model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
-structure also imputation model should do it (Audigier, 2021). Not
-including clustering into the imputation process may lead to effect
-estimates with smaller standard errors and inflated type I error.
-
-\hypertarget{overview-of-software}{%
-\subsection{overview of software}\label{overview-of-software}}
-
-The popular \pkg{mice} package in \proglang{R} \citet{R}
-
-\hypertarget{scope}{%
-\subsection{scope}\label{scope}}
-
-This papers serves as a tutorial for imputing incomplete multilevel data
-with \pkg{mice} in \proglang{R}. \pkg{mice} has become the de-facto
-standard for imputation by chained equations, which iteratively solves
-the missingness on a variable-by-variable basis. \pkg{mice} is known to
-yield valid inferences under many different missing data circumstances
-\citep{buur18}.
-
-We provide practical guidelines and code snippets for different missing
-data situations, including non-ignorable mechanisms. For reasons of
-brevity, we focus on multilevel imputation by chained equations with
-\pkg{mice} exclusively; other imputation methods and packages \citep[see
-e.g.][ and \citet{grun18}]{audi18} are outside the scope of this
-tutorial. Assumed knowledge includes basic familiarity with the
-\pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).
-
-We illustrate imputation of incomplete multilevel data using three case
-studies:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
-  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
-  with data that are MAR,][]{mice};
-\item
-  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
-  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
-  studies with data that are MAR,][]{metamisc};
-\item
-  \texttt{obesity} from the \pkg{micemd} package {[}simulated data on
-  obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
-  that are MNAR{]}.
-\end{itemize}
-
-For each of these datasets, we discuss the nature of the missingness,
-choose one or more imputation models and evaluate the imputed data, but
-we will also highlight one specific aspect of the imputation workflow.
-
-This tutorial is dedicated to readers who are unfamiliar with multiple
-imputation. More experienced readers can skip the introduction (case
-study 1) and directly head to practical applications of multilevel
-imputation under MAR conditions (case study 2) or under MNAR conditions
-(case study 3).
+\section{Introduction: Count data regression in R}\label{sec-intro}}
+
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]
+
+The introduction is in principle ``as usual''. However, it should
+usually embed both the implemented \emph{methods} and the
+\emph{software} into the respective relevant literature. For the latter
+both competing and complementary software should be discussed (within
+the same software environment and beyond), bringing out relative
+(dis)advantages. All software mentioned should be properly
+\texttt{@cited}'d.~(See also \protect\hyperlink{sec-bibtex}{Using
+BibTeX} for more details on \textsc{Bib}{\TeX}.)
+
+For writing about software JSS requires authors to use the markup
+\texttt{{[}{]}\{.proglang\}} (programming languages and large
+programmable systems), \texttt{{[}{]}\{.pkg\}} (software packages), back
+ticks like `code` for code (functions, commands, arguments, etc.).
+
+If there is such markup in (sub)section titles (as above), a plain text
+version has to be provided in the {\LaTeX} command as well. Below we
+also illustrate how abbrevations should be introduced and citation
+commands can be employed. See the {\LaTeX} code for more details.
+
+\end{tcolorbox}
+
+Modeling count variables is a common task in economics and the social
+sciences. The classical Poisson regression model for count data is often
+of limited use in these disciplines because empirical count data sets
+typically exhibit overdispersion and/or an excess number of zeros. The
+former issue can be addressed by extending the plain Poisson regression
+model in various directions: e.g., using sandwich covariances or
+estimating an additional dispersion parameter (in a so-called
+quasi-Poisson model). Another more formal way is to use a negative
+binomial (NB) regression. All of these models belong to the family of
+generalized linear models (GLMs). However, although these models
+typically can capture overdispersion rather well, they are in many
+applications not sufficient for modeling excess zeros. Since
+\citet{Mullahy1986} there is increased interest in zero-augmented models
+that address this issue by a second model component capturing zero
+counts. An overview of count data models in econometrics, including
+hurdle and zero-inflated models, is provided in
+\citet{CameronTrivedi2013}.
+
+In \proglang{R} \citet{R}, GLMs are provided by the model fitting
+functions \fct{glm} in the \pkg{stats} package and \fct{glm.nb} in the
+\pkg{MASS} package \citep{VenablesRipley2002} along with associated
+methods for diagnostics and inference. The manuscript that this document
+is based on \citep{ZeileisKleiberJackman2008} then introduced hurdle and
+zero-inflated count models in the functions \fct{hurdle} and
+\fct{zeroinfl} in the \pkg{pcsl} package \citet{Jackman2015}. Of course,
+much more software could be discussed here, including (but not limited
+to) generalized additive models for count data as available in the
+\proglang{R} packages \pkg{mgcv} \citet{Wood2006}, \pkg{gamlss}
+\citet{StasinopoulosRigby2007}, or \pkg{VGAM} \citet{Yee2009}.
 
 \hypertarget{sec-models}{%
-\section{Background}\label{sec-models}}
-
-\hypertarget{concepts-in-multilevel-data}{%
-\subsection{concepts in multilevel
-data}\label{concepts-in-multilevel-data}}
-
-Many datasets include individuals that are clustered together, for
-example in geographic regions, or even different studies. In the
-simplest case, individuals (e.g., students) are nested within a single
-cluster (e.g., school classes). More complex clustered structures may
-occur when there are multiple hierarchical levels (e.g., students in
-different schools or patients within hospitals within regions across
-countries), or when the clustering is non-nested (e.g., electronic
-health record data from diverse settings and populations within large
-databases). With clustered data we generally assume that individuals
-from the same cluster tend to be more similar than individuals from
-other clusters. In statistical terms, this implies that observations
-from the same cluster are not independent and may in fact be correlated.
-If this correlation is left unaddressed, estimates of \emph{p} values,
-confidence intervals even model parameters are prone to bias
-\citep{loca01}. Statistical methods for clustered data typically adopt
-hierarchical models that explicitly describe the grouping of
-observations. These models are also known as `multilevel models',
-`hierarchical models', `mixed effect models', `random effect models',
-and in the context of time-to-event data as `frailty models'. Table
-\ref{tab:clus} provides an overview of some key concepts in multilevel
-modeling.
-
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, toprule=.15mm, rightrule=.15mm, arc=.35mm, left=2mm, leftrule=.75mm, colback=white, opacityback=0]
-
-Box 1. The intraclass correlation coefficient.
-
-\end{tcolorbox}
-
-In \proglang{R}, multilevel models may be fitted using the package
-\pkg{lme4}. For linear mixed-effects models, the function
-
-\begin{verbatim}
-lmer(formula, data, ...)
-\end{verbatim}
-
-\hypertarget{concepts-in-missing-data}{%
-\subsection{concepts in missing data}\label{concepts-in-missing-data}}
-
-missing data mechanisms etc.
-
-As with any other dataset, clustered datasets may be impacted by
-missingness in much the same way. Several strategies can be used to
-handle missing data, including complete case analysis and imputation. We
-focus on the latter approach and discuss statistical methods for
-replacing the missing data with one or more plausible values. Imputation
-separates the missing data problem from the analysis and the completed
-data can be analyzed as if it were completely observed. It is generally
-recommended to impute the missing values more than once to preserve
-uncertainty due to missingness and to allow for valid inferences (c.f.
-Rubin 1976).
-
-With incomplete clustered datasets we can distinguish between two types
-of missing data: sporadic missingness and systematic missingness
-\citep{resc13}. Sporadic missingness arises when variables are missing
-for some but not all of the units in a cluster \citep{buur18, jola18}.
-For example, it is possible that test results are missing for several
-students in one or more classes. When all observations are missing
-within one or more clusters, data are said to be systematically missing.
-Sporadic missingness is visualized in Figure XYZ.
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-3-1.pdf}
-
-}
-
-\end{figure}
-
-Column \(X_1\) in Figure 1 is completely observed, column \(X_2\) is
-systematically missing in cluster 2, and column \(X_3\) is sporadically
-missing. To analyze these incomplete data, we have to take the nature of
-the missingness and the cluster structure into account. For example, the
-sporadic missingness in \(X_3\) could be easily amended if this would be
-a cluster-level variable (and thus constant within clusters). We could
-then just extrapolate the true (but missing) value of \(X_3\) for unit 1
-from unit 2, and the value for unit 4 from unit 3. If \(X_3\) would
-instead be a unit-level variable (which may vary within clusters), we
-could not just recover the unobserved `truth', but would need to use
-some kind of missing data method, or discard the incomplete units
-altogether (i.e., complete case analysis). Complete case analysis can
-however introduce bias in statistical inferences and lowers statistical
-power. Further, with the systematic missingness in \(X_2\), it would be
-impossible to fit a multilevel model without accommodating the
-missingness in some way. Complete case analysis in that case would mean
-excluding the entire cluster from the analyses. The wrong choice of
-missing data handling method can thus be extremely harmful to the
-inferences.
-
-Imputation of missing data requires consideration of the mechanism
-behind the missingness. Rubin proposed to distinguish between data that
-are missing completely at random (MCAR), data that are missing at random
-(MAR) and data that are missing not at random (MNAR; see Table
-\ref{tab:miss}). For each of these three missingness generating
-mechanisms, different imputation strategies are warranted
-(\citet{yuce08} and \citet{hox15}). We here consider the general case
-that data are MAR, and expand on certain MNAR situations.
-
-\hypertarget{imputation-with-mice}{%
-\subsection{imputation with mice}\label{imputation-with-mice}}
-
-The \proglang{R} package \pkg{mice} provides a framework for imputing
-incomplete data on a variable-by-variable basis. The \fct{mice} function
-allows users to flexibly specify how many times and under what model the
-missing data should be imputed. This is reflected in the first four
-function arguments
-
-\begin{verbatim}
-mice(data, m, method, predictorMatrix, ...)
-\end{verbatim}
-
-where \texttt{data} refers to the incomplete dataset, \texttt{m}
-determines the number of imputations, \texttt{method} denotes the
-functional form of the imputation model and \texttt{predictorMatrix}
-specifies the interrelational dependencies between variables and
-imputation models (i.e., the set of predictors to be used for imputing
-each incomplete variable).
-
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, toprule=.15mm, rightrule=.15mm, arc=.35mm, left=2mm, leftrule=.75mm, colback=white, opacityback=0]
-
-Box 2. The \texttt{methods}.
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, toprule=.15mm, rightrule=.15mm, arc=.35mm, left=2mm, leftrule=.75mm, colback=white, opacityback=0]
-
-Box 3. The predictor matrix. The entries corresponding to the level-1
-predictors are coded with a 3, indicating that both the original values
-as well as the cluster means of the predictor are included into the
-imputation model. The entry of 4 in the predictor matrix adds three
-variables to the imputation model for the imputation model predictor:
-the value of the predictor, the cluster means of the predictor and the
-random slopes of the predictor. - -2 = cluster variable - 1 = overall
-effect - 3 = overall + group-level effect - 4 = individual-level
-(random) and group-level (fixed) effect
-
-\end{tcolorbox}
-
-\hypertarget{sec-workflow}{%
-\section{Multilevel imputation workflow}\label{sec-workflow}}
-
-There are different strategies that can be adopted in the imputation
-process that account for clustering: inclusion of cluster indicator
-variable, performing a separate imputation process for each cluster, or
-performing a simultaneous imputation process by using an imputation
-method that accounts for clustering.(Stata:
-https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/)
-TODO: replace ref.
-
-The selection of each strategy depends mainly on the assumptions in the
-main analysis and also on the restriction of the analyzed data.
-
-Regarding the restrictions imposed by the data, for instance, the use of
-cluster indicator variables is restricted in datasets where there are
-not many clusters and many observations per cluster (Graham, 2009). The
-last restriction is also required when imputations are performed on each
-cluster separately. When this restriction cannot be achieved, one can
-use an imputation model that simultaneously imputes all clusters using a
-hierarchical model (Allison 2002).
-
-Under this hierarchical imputation model, observations within clusters
-are correlated and this correlation is modeled by a random effect so the
-hierarchical model can be estimated even when there are few observations
-per cluster. However, this strategy is best suited for balanced data
-(Grund, 2017) and when random effects model is appropriated, i.e.~the
-number of clusters is adequate. (Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main
-model, for instance by using the cluster indicator strategy may lead to
-bias estimates when the model is based on a hierarchical model
-(Taaljard,2008). Even when an imputation strategy congenial with the
-main model is preferred, it is important to consider whether it is
-appropriate for the data as a less complex imputation strategies may
-also lead to unbiased estimates in certain scenarios(Bailey 2020). For
-instance, in causal effect analysis, separately imputation may lead to
-smaller bias when the size of the smaller exposure cluster is large,
-compared with an imputation model that includes exposure-confounder
-interactions. (Zhang,2023).
-
-Below we provide a imputation workflow that can be used in general to
-impute cluster data.
-
-\hypertarget{focus-on-the-main-analysis}{%
-\subsection{Focus on the Main
-Analysis}\label{focus-on-the-main-analysis}}
-
-When imputing clustered data, we initially focus on the research
-question and the intended analysis, assuming the presence of
-non-incomplete values. This assessment sheds light on research
-hypotheses and the contemplated main(s) statistical model(s) and it
-provides insights into the data's structure (refer to the level table),
-variable types (e.g., confounders, auxiliary variables), and other
-considerations about variables relationships such as interactions or
-polynomial terms.
-
-\hypertarget{exploration-of-available-data}{%
-\subsection{Exploration of Available
-Data}\label{exploration-of-available-data}}
-
-Next, it is necessary to explore the information available in the data.
-Exploring individual variables using tools such as histograms or QQ
-plots can help to delineate variable distributions and plausible ranges
-of values and also identify input errors or outliers. Evaluating
-interactions between predictors using scatter plots or conditional
-plots, assessing collinearity using VIF or correlations, can help to
-glimpse nonlinear relationships that may affect the original formulation
-of the main model. Further testing of assumptions related to the
-response variable, including variance (homogeneity using conditional box
-plots), independence (e.g., ACF, variograms) and response vs.~predictor
-relationships (conditional plots - x vs.~y), may serve to choose an
-imputation model more suitable to the data. In addition, the intraclass
-correlation coefficient (ICC) can be examined to assess cluster
-differences, aiding in the choice between the 2l and 1l methods for
-imputation.
-
-One can also explore missingness, specifically missing patterns at the
-cluster level to identify types of patterns (e.g., non-monotonic,
-systematic) to guide the selection of an appropriate imputation approach
-in terms of computational efficiency (e.g., simpler regression
-imputation versus FCS in univariate patterns). Similarly, these missing
-patterns can also be used to identify potential predictors in individual
-imputation models by means of inflow criteria (quantifying the
-connections between missing data in one variable and other observed
-variables) and outflow criteria (determining the connections between
-observed values in one variable and missing data in other variables).
-Although several packages exist for visualizing missing data, this
-tutorial focuses on the recently implemented ggmice package.
-
-\hypertarget{assess-estimation-procedure-robustness-to-missing-data}{%
-\subsection{Assess Estimation Procedure Robustness to Missing
-Data}\label{assess-estimation-procedure-robustness-to-missing-data}}
-
-Prior to implementing multiple imputation, it is critical to assess the
-robustness of the estimation model when confronted with missing data.
-There are scenarios in which specific Maximum Likelihood (ML) estimation
-methods outperform Multiple Imputation (MI) methods. For instance, when
-the response variable is the sole incomplete variable, mixed models
-demonstrate robustness to missing data under the Missing at Random (MAR)
-assumption and with a correct variance-covariance
-specification.\cite{Molenberghs_2007}. Furthermore, depending on the
-proportion of missingness (usually is \textless5\%), opting for a
-simpler complete case analysis might be sufficient.
-
-\hypertarget{pre-imputation}{%
-\subsection{Pre-imputation}\label{pre-imputation}}
-
-Before proceeding with the imputation model, it is necessary to filter
-the dataset to include only the minimal variables required for the main
-model. If additional procedures are conducted during the analysis, it is
-essential to include variables associated with these procedures, such as
-confounders in the case of balancing techniques. Other relevant
-variables, like instrumental variables or auxiliary variables, may
-enhance parameter estimates even if not included in the main model, as
-they could be linked to the probability of missingness for some
-incomplete variables.
-
-Furthermore, at the variable level, it is crucial to assess whether
-proxy variables should be considered or if deterministic imputation
-alone is adequate for imputing missing values. For instance, certain
-incomplete variables may not necessitate stochastic imputation methods
-like MICE. Instead, they can be effectively addressed through deductive
-imputation, where incomplete values are inferred from logical and
-deterministic relationships between variables. This approach is
-especially beneficial for variables that are functions of others, such
-as deriving a person's BMI from their weight and height. It also proves
-valuable for determining values for one-level variables from two-level
-ones, as seen in the context of Individual Participant Data (IPD). In
-such cases, incomplete information can be deduced from metadata, like
-inferring incomplete data about abortion in a country where abortion is
-illegal, or through cross-temporal or protocol-based deduction, such as
-imputing missing test values for deceased patients.
-
-\hypertarget{setting-imputation-model}{%
-\subsection{Setting Imputation Model}\label{setting-imputation-model}}
-
-\hypertarget{clustering-inclusion}{%
-\paragraph{Clustering Inclusion}\label{clustering-inclusion}}
-
-Various strategies can be employed in the imputation process to account
-for clustering, including the inclusion of a cluster indicator variable,
-conducting a separate imputation process for each cluster, or employing
-a simultaneous imputation method that considers clustering\cite{eddings}
-
-The choice of strategy depends primarily on the assumptions made in the
-main analysis and the constraints imposed by the analyzed data.
-Concerning data restrictions when the analysis do not contemplate an
-hierarchical model (eg. descriptive way), if there are few clusters with
-many observations per cluster, the cluster indicator or imputation
-separated by groups may be appropriate \cite{graham2009}. Conversely, an
-imputation model that simultaneously imputes all clusters using a
-hierarchical model can be employed when there are more clusters or fewer
-observations per cluster \cite{Allison_2002}.
-
-In a hierarchical imputation model, correlations between observations
-within clusters are modeled by a random effect, allowing estimation even
-with a limited number of observations per cluster. Multiple imputation
-models based on hierarchical models have been proposed, each with
-different assumptions. \cite{audigier}
-
-In the particular case that the analysis contemplates the use of a
-hierarchical model (i.e.~main model), before choosing a particular
-strategy, it is crucial to verify if the assumptions of the imputation
-model align with those of the main analysis. For example, using the
-cluster indicator strategy may introduce bias estimates when the model
-is based on a hierarchical structure \cite{taljaard2008,speidel2018}.
-Even if an imputation strategy congruent with the main model is
-preferred, it is essential to assess its appropriateness for the data,
-considering that less complex imputation strategies may lead to unbiased
-estimates in certain scenarios \cite{bailey2020}.
-
-\hypertarget{choice-of-individual-imputation-methods}{%
-\subsubsection{Choice of Individual Imputation
-Methods}\label{choice-of-individual-imputation-methods}}
-
-The initial step involves selecting the individual imputation model for
-each incomplete variable in the dataset. The mice package automatically
-proposes imputation methods based on the variable type for 1l variables.
-For 2l level variables, the micemd package's () function can be
-utilized. This function selects among the 2l imputation methods based on
-the size of the clusters and the proportion of missingness in each
-cluster.
-
-In addition to the package-defined imputation methods, users can specify
-custom methods using the ``I formula''. This flexibility allows for the
-calculation of derived variables internally or adjustments to imputation
-methods based on specific conditions, such as conditioning the
-imputation model on the level of an incomplete covariate (e.g.,
-pregnancy test on females).
-
-\hypertarget{model-specification}{%
-\subsubsection{Model Specification}\label{model-specification}}
-
-The imputation model must be congenial with the main model
-\cite{meng1994}. Congeniality issues arise when the imputation model and
-the main model make different assumptions, often due to the omission of
-a polynomial or interaction term or the use of transformed variables.
-
-The imputation model can include additional terms than the main model
-terms that do not lead to uncongeniality. For instance, it is advisable
-to include the outcome variable in the imputation model for prediction
-variables \cite{moons2006a}. In cases where the outcome is
-time-to-event, the Nelson-Aalen estimate of the time to the event should
-be included as a covariate in the imputation model {[}REF{]}.
-Furthermore, the inclusion of auxiliary variables, even if not part of
-the main model, can be associated with the probability of missingness,
-enhancing the likelihood of satisfying the Missing at Random (MAR)
-assumption and improving estimation efficiency \cite{hardt2012a}.
-
-The specification of imputation models is done on a variable basis,
-either using a prediction matrix where the type of prediction variable
-should be specified accordingly (see table) or through a list of
-formulas in the formula parameter. The latter option proves useful in
-formulating imputation models with polynomial terms or interactions
-compared to the prediction matrix specification, which requires the
-inclusion of additional terms as Just Another Variable (JAV)
-\cite{buuren2018a}.
-
-However, even when considering an interaction term, depending on the
-adapted individual imputation model, it has been suggested, for
-instance, for treatment interaction effects, to conduct separated
-imputation by treatment group \cite{zhang2023}. Additionally, there are
-imputation models based on random forest or deep learning that can
-handle interaction and non-linear terms that do not require the explicit
-inclusion of the non-linear terms.
-
-\hypertarget{post-imputation}{%
-\subsection{Post-Imputation}\label{post-imputation}}
-
-During the imputation process, certain issues may arise that halt the
-process. In hierarchical model imputations, many issues are linked to
-methods based on parametric hierarchical models, which may struggle to
-estimate a model midway through the process. In such cases, it is
-advisable to examine the imputation log file to identify variables with
-imputation issues. Subsequently, reducing the number of predictors in
-the imputation model could be explored, either by using functions like
-quickpred or by considering variable transformations, such as scaling.
-Adjusting the level of the hierarchical model (e.g., using a homogenous
-variance model or a 1l model) can also be beneficial. Additionally,
-checking the range of imputed variables is crucial, as excessively large
-imputed values for one predictor could lead to convergence issues in
-other variables. This can be mitigated by including post-processing
-specifications on problematic variables or by using imputation models
-like Predicted Mean Matching (PMM), ensuring imputed values align with
-the observable values. In certain cases, a separate imputation strategy
-may be considered. For instance, in analyses involving multiple
-endpoints, conducting separate imputation processes for each endpoint
-might be preferable to a unified imputation process.
-
-\hypertarget{convergence-and-sensitivity-analysis}{%
-\subsection{Convergence and Sensitivity
-Analysis}\label{convergence-and-sensitivity-analysis}}
-
-Before starting into the analysis of each imputed dataset, it is crucial
-to validate the convergence of the imputation process. This is commonly
-accomplished through trace plots that depict the mean and variance of
-the incomplete variables across iterations. These plots serve to uncover
-potential circular issues or the need for additional iterations.
-Additionally, it is also important to verify that imputed values fall
-within a plausible range and also to check the distribution of imputed
-variables, ensuring that the imputed variable distribution aligns with
-the distribution of observed values (under the MAR assumption). An
-alternative approach involves assessing the prediction accuracy of the
-imputation method \cite{cai2023}.
-
-While the majority of Multiple Imputation by Chained Equations (MICE)
-methods are based on Missing at Random (MAR) assumptions, field expert
-input may suggest that the Missing Not at Random (MNAR) mechanism could
-be plausible for certain variables. An MNAR variable is one in which the
-probability of missingness depends on an unobservable variable. This can
-occur when missingness is associated with the incomplete value itself
-(self-marking) or when there is an unobserved variable linked to both
-the value and the probability of missingness of the incomplete value
-(indirectly non-informative). Specifically, for the indirectly
-non-informative case in hierarchical datasets, imputation methods based
-on the Heckman method can be
-considered.\cite{hammon2020,hammon2022,munoz2023}
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\hypertarget{sec-illustrations}{%
-\section{Illustrations}\label{sec-illustrations}}
-
-In this section, we demonstrate the workflow using three case studies.
-
-\hypertarget{setup}{%
-\subsection{Setup}\label{setup}}
-
-\begin{verbatim}
-R> set.seed(123)         # for reproducibility
-R> library(lme4)         # for multilevel modeling
-R> library(mice)         # for imputation
-R> library(miceadds)     # for multilevel imputation methods
-R> library(micemd)       # for selection-model imputation methods
-R> library(mitml)        # for multilevel parameter pooling
-R> library(broom)        # for clean model estimates
-R> library(broom.mixed)  # for multilevel model estimates
-R> library(dplyr)        # for data wrangling
-R> library(ggmice)       # for visualization
-R> library(ggplot2)      # for visualization
-\end{verbatim}
-
-\hypertarget{popularity-data}{%
-\subsection{Popularity data}\label{popularity-data}}
-
-In this section we will go over the different steps involved with
-imputing incomplete multilevel data with the R package mice. We consider
-the simulated \texttt{popmis} dataset, which included pupils
-(\(n = 2000\)) clustered within schools (\(N = 100\)). The following
-variables are of primary interest:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{school}, school identification number (clustering variable);
-\item
-  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
-  unit-level);
-\item
-  \texttt{sex}, pupil sex (0 = boy, 1 = girl; unit-level);
-\item
-  \texttt{texp}, teacher experience (in years; cluster-level).
-\end{itemize}
-
-The analysis model corresponding to this dataset is multilevel
-regression with random intercepts for the different schools. We will
-estimate the association between the pupils' sex and their popularity
-score. This model can be expressed in \texttt{lme4} code as:
-
-\begin{verbatim}
-popular ~ 1 + sex + (1 | school)
-\end{verbatim}
-
-We load the data into the environment with
-
-\begin{verbatim}
-R> data("popmis", package = "mice")
-\end{verbatim}
-
-and select the relevant variables
-
-\begin{verbatim}
-R> dat <- popmis[, c("school", "popular", "texp", "sex")] 
-\end{verbatim}
-
-which results in the following data structure.
-
-\begin{verbatim}
-R> head(dat)
-\end{verbatim}
-
-\begin{verbatim}
-  school popular texp sex
-1      1      NA   24   1
-2      1      NA   24   0
-3      1       7   24   1
-4      1      NA   24   1
-5      1      NA   24   1
-6      1       7   24   0
-\end{verbatim}
-
-The association of interest can be visualized with \texttt{ggmice},
-
-\begin{verbatim}
-ggmice(dat, aes(sex, popular)) +
-  geom_jitter()
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-8-1.pdf}
-
-}
-
-\caption{Scatterplot of student popularity by sex}
-
-\end{figure}
-
-where missing datapoints in the \texttt{popular} variable are
-represented by red points on the X-axis of the figure.
-
-With the \texttt{ggmice} function \texttt{plot\_pattern} we can
-visualize the missing data pattern
-
-\begin{verbatim}
-R> plot_pattern(dat)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/fig-pattern-1.pdf}
-
-}
-
-\caption{\label{fig-pattern}Missing data pattern.}
-
-\end{figure}
-
-which shows us that the missingness is univariate and sporadic.
-
-To develop the best imputation model for the incomplete variable
-\texttt{popular}, we need to know whether the observed values of
-\texttt{popular} are related to observed values of other variables. Plot
-the pair-wise complete correlations in the incomplete data
-
-\begin{verbatim}
-R> plot_corr(dat)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}
-
-}
-
-\caption{Pair-wise correlations.}
-
-\end{figure}
-
-This shows us that not just the analysis-model variable \texttt{sex},
-but also the cluster-level covariate teacher experience, \texttt{texp},
-may be a useful as an imputation model predictor. Moreover, the
-missingness in \texttt{popular} may depend on the observed values of
-other variables. With \texttt{ggmice()} we can visualize the
-distribution of the teacher experience for cases where \texttt{popular}
-is observed and cases where \texttt{popular} is missing.
-
-\begin{verbatim}
-R> ggmice(dat, aes(texp)) + 
-+  geom_density() +
-+  facet_wrap(~is.na(popular))
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-11-1.pdf}
-
-}
-
-\end{figure}
-
-It appears that students with a missing value for \texttt{popular} are
-in clusters with a slightly higher \texttt{texp} value.
-
-\begin{verbatim}
-t.test(dat$texp ~ is.na(dat$popular)) |> 
-  tidy() |>
-  kable()
-\end{verbatim}
-
+\section{Models and software}\label{sec-models}}
+
+The basic Poisson regression model for count data is a special case of
+the GLM framework \citet{McCullaghNelder1989}. It describes the
+dependence of a count response variable \(y_i\) (\(i = 1, \dots, n\)) by
+assuming a Poisson distribution \(y_i \sim \mathrm{Pois}(\mu_i)\). The
+dependence of the conditional mean \(E[y_i \, | \, x_i] = \mu_i\) on the
+regressors \(x_i\) is then specified via a log link and a linear
+predictor
+
+\begin{equation}\protect\hypertarget{eq-mean}{}{
+\log(\mu_i) \quad = \quad x_i^\top \beta,
+}\label{eq-mean}\end{equation}
+
+where the regression coefficients \(\beta\) are estimated by maximum
+likelihood (ML) using the iterative weighted least squares (IWLS)
+algorithm.
+
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]
+
+TODO: Note that around the equation above there should be no spaces
+(avoided in the {\LaTeX} code by \texttt{\%} lines) so that ``normal''
+spacing is used and not a new paragraph started.
+
+\end{tcolorbox}
+
+\proglang{R} provides a very flexible implementation of the general GLM
+framework in the function \fct{glm} \citet{ChambersHastie1992} in the
+\pkg{stats} package. Its most important arguments are
+
+\begin{verbatim}
+glm(formula, data, subset, na.action, weights, offset,
+  family = gaussian, start = NULL, control = glm.control(...),
+  model = TRUE, y = TRUE, x = FALSE, ...)
+\end{verbatim}
+
+where \texttt{formula} plus \texttt{data} is the now standard way of
+specifying regression relationships in \proglang{R}/\proglang{S}
+introduced in \citet{ChambersHastie1992}. The remaining arguments in the
+first line (\texttt{subset}, \texttt{na.action}, \texttt{weights}, and
+\texttt{offset}) are also standard for setting up formula-based
+regression models in \proglang{R}/\proglang{S}. The arguments in the
+second line control aspects specific to GLMs while the arguments in the
+last line specify which components are returned in the fitted model
+object (of class \class{glm} which inherits from \class{lm}). For
+further arguments to \fct{glm} (including alternative specifications of
+starting values) see \texttt{?glm}. For estimating a Poisson model
+\texttt{family\ =\ poisson} has to be specified.
+
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]
+
+As the synopsis above is a code listing that is not meant to be
+executed, one can use either the dedicated \texttt{\{Code\}} environment
+or a simple \texttt{\{verbatim\}} environment for this. Again, spaces
+before and after should be avoided.
+
+Finally, there might be a reference to a \texttt{\{table\}} such as
+Table~\ref{tbl-overview}. Usually, these are placed at the top of the
+page (\texttt{{[}t!{]}}), centered (\texttt{\textbackslash{}centering}),
+with a caption below the table, column headers and captions in sentence
+style, and if possible avoiding vertical lines.
+
+\end{tcolorbox}
+
+\hypertarget{tbl-overview}{}
 \begin{longtable}[]{@{}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0924}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0924}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0924}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.2017}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.1008}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.6351}}@{}}
 \toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedleft
-estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-estimate1
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-estimate2
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-statistic
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-p.value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-parameter
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-conf.low
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-conf.high
+\begin{minipage}[b]{\linewidth}\raggedright
+Type
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-method
+Distribution
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-alternative
+Method
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Description
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Type
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Distribution
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Method
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Description
 \end{minipage} \\
@@
 \endlastfoot
--0.2599581 & 14.15278 & 14.41274 & -0.8731291 & 0.3827095 & 1796.115 &
--0.8438947 & 0.3239786 & Welch Two Sample t-test & two.sided \\
+GLM & Poisson & ML & Poisson regression: classical GLM, estimated by
+maximum likelihood (ML) \\
+& & Quasi & ``Quasi-Poisson regression'\,': same mean function,
+estimated by quasi-ML (QML) or equivalently generalized estimating
+equations (GEE), inference adjustment via estimated dispersion
+parameter \\
+& & Adjusted & ``Adjusted Poisson regression'\,': same mean function,
+estimated by QML/GEE, inference adjustment via sandwich covariances \\
+& NB & ML & NB regression: extended GLM, estimated by ML including
+additional shape parameter \\
+Zero-augmented & Poisson & ML & Zero-inflated Poisson (ZIP), hurdle
+Poisson \\
+& NB & ML & Zero-inflated NB (ZINB), hurdle NB \\
+\caption{\label{tbl-overview}Overview of various count regression
+models. The table is usually placed at the top of the page
+(\texttt{{[}t!{]}}), centered (\texttt{centering}), has a caption below
+the table, column headers and captions are in sentence style, and if
+possible vertical lines should be avoided.}\tabularnewline
 \end{longtable}
 
-Although there are no significant differences in the distribution of
-\texttt{texp} depending on the missingness indicator of
-\texttt{popular}, this variable can serve as auxiliary variable in the
-imputation of \texttt{popular}.
-
-\begin{verbatim}
-R> meth <- make.method(dat)
-R> meth
-\end{verbatim}
-
-\begin{verbatim}
- school popular    texp     sex 
-     ""   "pmm"      ""      "" 
-\end{verbatim}
-
-\begin{verbatim}
-R> pred <- quickpred(dat)
-R> pred
-\end{verbatim}
-
-\begin{verbatim}
-        school popular texp sex
-school       0       0    0   0
-popular      0       0    1   1
-texp         0       0    0   0
-sex          0       0    0   0
-\end{verbatim}
-
-Adjust the methods vector.
-
-\begin{verbatim}
-R> meth["popular"] <- "2l.pmm"
-\end{verbatim}
-
-The \texttt{pmm} method is better (more efficient) because it will still
-look for donors (maybe outside of cluster) based on predictive distance,
-even for very small clusters.
-
-Adjust the predictor matrix.
-
-\begin{verbatim}
-R> pred["popular", "school"] <- -2
-R> pred["popular", "sex"] <- 2
-\end{verbatim}
-
-Visualize the imputation methods and predictors.
-
-\begin{verbatim}
-plot_pred(pred, method = meth)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-17-1.pdf}
+\hypertarget{sec-illustrations}{%
+\section{Illustrations}\label{sec-illustrations}}
+
+For a simple illustration of basic Poisson and NB count regression the
+\texttt{quine} data from the \pkg{MASS} package is used. This provides
+the number of \texttt{Days} that children were absent from school in
+Australia in a particular year, along with several covariates that can
+be employed as regressors. The data can be loaded by
+
+\begin{verbatim}
+R> data("quine", package = "MASS")
+\end{verbatim}
+
+and a basic frequency distribution of the response variable is displayed
+in Figure~\ref{fig-quine}.
+
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]
+
+For code input and output, the style files provide dedicated
+environments. Either the ``agnostic'' \texttt{\{CodeInput\}} and
+\texttt{\{CodeOutput\}} can be used or, equivalently, the environments
+\texttt{\{Sinput\}} and \texttt{\{Soutput\}} as produced by \fct{Sweave}
+or \pkg{knitr} when using the \texttt{render\_sweave()} hook. Please
+make sure that all code is properly spaced, e.g., using
+\texttt{y\ =\ a\ +\ b\ *\ x} and \emph{not} \texttt{y=a+b*x}. Moreover,
+code input should use ``the usual'' command prompt in the respective
+software system. For \proglang{R} code, the prompt
+\texttt{R\textgreater{}} should be used with \texttt{+} as the
+continuation prompt. Generally, comments within the code chunks should
+be avoided -- and made in the regular {\LaTeX} text instead. Finally,
+empty lines before and after code input/output should be avoided (see
+above).
+
+\end{tcolorbox}
+
+\begin{figure}
+
+{\centering
+
+\includegraphics{article-visualization.pdf}
 
@@
 
+\caption{\label{fig-quine}Frequency distribution for number of days
+absent from school.}
+
 \end{figure}
 
-Impute the data.
-
-\begin{verbatim}
-R> imp <- mice(
-+  data = dat,
-+  method = meth,
-+  predictorMatrix = pred,
-+  printFlag = FALSE
-+)
-\end{verbatim}
-
-Evaluate the convergence.
-
-\begin{verbatim}
-R> plot_trace(imp)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-19-1.pdf}
-
-}
-
-\end{figure}
-
-Troubleshoot non-convergence of imputation model. Evaluate the
-distribution of imputed values.
-
-\begin{verbatim}
-R> ggmice(imp, aes(popular, group = .imp)) + 
-+  geom_density() 
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-20-1.pdf}
-
-}
-
-\end{figure}
-
-Evaluate the distribution of imputed values.
-
-\begin{verbatim}
-R> ggmice(imp, aes(.imp, popular)) + 
-+  geom_jitter(alpha = 0.05) +
-+    geom_boxplot()
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-21-1.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-R> ggmice(imp, aes(sex, popular)) +
-+  geom_jitter() +
-+  facet_wrap(~ .imp)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-22-1.pdf}
-
-}
-
-\end{figure}
-
-Analyze the imputed data.
-
-\begin{verbatim}
-R> fit <- with(
-+  imp,
-+  lmer(popular ~ texp + sex + (1 | school))
-+)
-\end{verbatim}
-
-Pooling the estimates does not provide estimates of the variance
-components.
-
-\begin{verbatim}
-R> pool(fit)
-\end{verbatim}
-
-\begin{verbatim}
-Class: mipo    m = 5 
-         term m   estimate         ubar            b            t dfcom
-1 (Intercept) 5 3.59902513 0.0284760238 3.461284e-03 0.0326295651  1995
-2        texp 5 0.09225928 0.0001140591 2.273382e-05 0.0001413397  1995
-3         sex 5 0.85187018 0.0009628869 4.946825e-04 0.0015565060  1995
-        df       riv    lambda       fmi
-1 216.1756 0.1458610 0.1272938 0.1352573
-2 100.6502 0.2391794 0.1930144 0.2085857
-3  26.9008 0.6164992 0.3813792 0.4227574
-\end{verbatim}
-
-Therefore, \texttt{mitml} is used.
-
-\begin{verbatim}
-R> est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
-\end{verbatim}
-
-Display results in table.
-
-\begin{verbatim}
-R> est$estimates |> 
-+  round(3) |>
-+  kable()
-\end{verbatim}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1558}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1169}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1299}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2338}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}@{}}
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Std.Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-t.value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-P(\textgreater\textbar t\textbar)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-RIV
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-FMI
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-(Intercept) & 3.599 & 0.181 & 19.924 & 246.857 & 0 & 0.146 & 0.134 \\
-texp & 0.092 & 0.012 & 7.760 & 107.369 & 0 & 0.239 & 0.208 \\
-sex & 0.852 & 0.039 & 21.592 & 27.501 & 0 & 0.616 & 0.422 \\
-\end{longtable}
-
-\begin{verbatim}
-R> est$extra.pars |> 
-+  round(3) |>
-+  kable()
-\end{verbatim}
-
-\begin{longtable}[]{@{}lr@{}}
-\toprule\noalign{}
-& Estimate \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Intercept\textasciitilde\textasciitilde Intercept\textbar school &
-0.470 \\
-Residual\textasciitilde\textasciitilde Residual & 0.462 \\
-ICC\textbar school & 0.504 \\
-\end{longtable}
-
-\hypertarget{impact-data}{%
-\subsection{IMPACT data}\label{impact-data}}
-
-The second case study is the \texttt{impact} data from the
-\pkg{metamisc} package \citep[empirical data on traumatic brain
-injuries, \(n = 11,022\) units across \(N = 15\) clusters,][]{metamisc}.
-
-The \texttt{impact} data set contains traumatic brain injury data on
-\(n = 11022\) patients clustered in \(N = 15\) studies with the
-following 11 variables:
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{name} Name of the study,
-\item
-  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
-  observational cohort),
-\item
-  \texttt{age} Age of the patient,
-\item
-  \texttt{motor\_score} Glasgow Coma Scale motor score,
-\item
-  \texttt{pupil} Pupillary reactivity,
-\item
-  \texttt{ct} Marshall Computerized Tomography classification,
-\item
-  \texttt{hypox} Hypoxia (0=no, 1=yes),
-\item
-  \texttt{hypots} Hypotension (0=no, 1=yes),
-\item
-  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
-\item
-  \texttt{edh} Epidural hematoma (0=no, 1=yes),
-\item
-  \texttt{mort} 6-month mortality (0=alive, 1=dead).
-\end{itemize}
-
-Check if there is systematic missingness in this dataset. For
-illustration purposes, we made Marshall Computerized Tomography
-classification (ct) systematically missing.
-
-The analysis model for this dataset is a prediction model with
-\texttt{mort} as the outcome. In this tutorial we'll estimate the
-adjusted prognostic effect of \texttt{ct} on mortality outcomes. The
-estimand is the adjusted odds ratio for \texttt{ct}, after including
-\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
-into the analysis model:
-
-\begin{verbatim}
-mort ~ type + age + motor_score + pupil + ct + (1 | name) 
-\end{verbatim}
-
-Note that variables \texttt{hypots}, \texttt{hypox}, \texttt{tsah} and
-\texttt{edh} are not part of the analysis model, and may thus serve as
-auxiliary variables for imputation.
-
-The \texttt{impact} data included in the \pkg{metamisc} package is a
-complete data set. The original data has already been imputed once
-(Steyerberg et al, 2008). For the purpose of this tutorial we have
-induced missingness (mimicking the missing data in the original data set
-before imputation). The resulting incomplete data can be accessed from
-\href{https://zenodo.com}{zenodo link to be created}.
-
-Load the incomplete data into the R workspace:
-
-\begin{verbatim}
-R> dat <- read.table("link/to/the/data.txt") 
-\end{verbatim}
-
-To explore the missingness, we should look at the missing data pattern.
-The ten most frequent missingness patterns are shown with
-
-\begin{verbatim}
-R> plot_pattern(dat, rotate = TRUE, npat = 10L)  
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-28-1.pdf}
-
-}
-
-\end{figure}
-
-This shows that we need to impute \texttt{ct} and \texttt{pupil}. To
-develop the best imputation model, we need to investigate the relations
-between the observed values of the incomplete variables and the observed
-values of other variables, and the relation between the missingness
-indicators of the incomplete variables and the observed values of the
-other variables. To see whether the missingness depends on the observed
-values of other variables, we can test this statistically or use visual
-inspection (e.g.~a histogram faceted by the missingness indicator).
-
-We should impute the variables \texttt{ct} and \texttt{pupil} and any
-auxiliary variables we might want to use to impute these incomplete
-analysis model variables. We can evaluate which variables may be useful
-auxiliaries by plotting the pairwise complete correlations
-
-\begin{verbatim}
-R> plot_corr(dat, rotate = TRUE)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-29-1.pdf}
-
-}
-
-\end{figure}
-
-This shows us that \texttt{hypox} and \texttt{hypot} would not be useful
-auxiliary variables for imputing \texttt{ct}. Depending on the minimum
-required correlation, \texttt{tsah} could be useful, while \texttt{edh}
-has the strongest correlation with \texttt{ct} out of all the variables
-in the data and should definitely be included in the imputation model.
-For the imputation of \texttt{pupil}, none of the potential auxiliary
-variables has a very strong relation, but \texttt{hypots} could be used.
-We conclude that we can exclude \texttt{hypox} from the data, since this
-is neither an analysis model variable nor an auxiliary variable for
-imputation
-
-\begin{verbatim}
-R> dat <- select(dat, !hypox)
-\end{verbatim}
-
-Mutate data to get the right data types for imputation (e.g.~integer for
-clustering variable).
-
-\begin{verbatim}
-R> # dat <- mutate(
-R> #   dat,
-R> #   name = as.integer(name))
-\end{verbatim}
-
-This is necessary because otherwise PMM cannot be used for these factor
-variables.
-
-\begin{verbatim}
-R> dat <- mutate(
-+  dat,
-+  across(everything(), as.numeric))
-\end{verbatim}
-
-Create an initial methods vector for the incomplete variables
-
-\begin{verbatim}
-R> meth <- make.method(dat)
-R> meth
-\end{verbatim}
-
-\begin{verbatim}
-       name        type         age motor_score       pupil          ct 
-         ""          ""          ""          ""       "pmm"       "pmm" 
-     hypots        tsah         edh        mort 
-      "pmm"       "pmm"       "pmm"          "" 
-\end{verbatim}
-
-which should be adjusted to the appropriate \texttt{2l} methods.
-
-\begin{verbatim}
-R> # meth[c("pupil", "ct")] <- "2l.pmm"
-R> # meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
-R> meth[meth == "pmm"] <- "2l.pmm"
-\end{verbatim}
-
-Create an initial predictor matrix
-
-\begin{verbatim}
-R> pred <- quickpred(dat)
-\end{verbatim}
-
-This predictor matrix is too large to display inline. A visualization of
-the adapted predictor matrix is presented in Figure XYZ.
-
-We should make sure \texttt{name} is used as clustering variable
-
-\begin{verbatim}
-R> pred[, "name"] <- -2
-\end{verbatim}
-
-and the analysis-model outcome should be used as a predictor in all
-imputation models
-
-\begin{verbatim}
-R> pred[, "mort"] <- 2
-R> # pred[pred == 1] <- 2
-\end{verbatim}
-
-the resulting predictor matrix is visualized with
-
-\begin{verbatim}
-R> plot_pred(pred, method = meth, rotate = TRUE)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-38-1.pdf}
-
-}
-
-\end{figure}
-
-Impute the incomplete data with
-
-\begin{verbatim}
-R> imp <- mice(
-+  dat,
-+  method = meth,
-+  predictorMatrix = pred,
-+  printFlag = FALSE
-+)
-\end{verbatim}
-
-Evaluate the convergence of the algorithm
-
-\begin{verbatim}
-R> plot_trace(imp)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-40-1.pdf}
-
-}
-
-\end{figure}
-
-Evaluate the imputed values.
-
-\begin{verbatim}
-R> ggmice(imp, aes(.imp, pupil)) +
-+  geom_jitter()
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-41-1.pdf}
-
-}
-
-\end{figure}
-
-Convert the data back to factors.
-
-\begin{verbatim}
-long <- complete(imp, "long", include = TRUE)
-long <- mutate(long, across(c("motor_score", "pupil", "ct"), as.factor))
-imp <- as.mids(long)
-\end{verbatim}
-
-Analyze the imputed data:
-
-\begin{verbatim}
-R> fit <- imp %>%
-+  with(glmer(
-+    mort ~ type + age + motor_score + pupil + ct + (1 | name),
-+    family = "binomial"
-+    ))
-\end{verbatim}
-
-The estimated effects after imputation are presented in Table XYZ.
-
-\begin{verbatim}
-R> est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
-\end{verbatim}
-
-Display results in table.
-
-\begin{verbatim}
-R> est$estimates |> 
-+  round(3) |>
-+  kable()
-\end{verbatim}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1605}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1111}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1235}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0988}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1358}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2222}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}@{}}
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Std.Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-t.value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-P(\textgreater\textbar t\textbar)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-RIV
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-FMI
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-(Intercept) & -1.892 & 0.333 & -5.682 & 11445.455 & 0.000 & 0.019 &
-0.019 \\
-type & -0.346 & 0.178 & -1.948 & 686356.768 & 0.051 & 0.002 & 0.002 \\
-age & 0.032 & 0.002 & 19.564 & 2418.910 & 0.000 & 0.042 & 0.041 \\
-motor\_score2 & -0.575 & 0.070 & -8.226 & 71295.018 & 0.000 & 0.008 &
-0.008 \\
-motor\_score3 & -0.891 & 0.072 & -12.440 & 66712.189 & 0.000 & 0.008 &
-0.008 \\
-motor\_score4 & -1.276 & 0.074 & -17.305 & 18426.087 & 0.000 & 0.015 &
-0.015 \\
-pupil2 & 1.315 & 0.068 & 19.339 & 110.826 & 0.000 & 0.235 & 0.204 \\
-pupil3 & 0.657 & 0.075 & 8.745 & 423.463 & 0.000 & 0.108 & 0.101 \\
-ct2 & 0.747 & 0.084 & 8.884 & 41.781 & 0.000 & 0.448 & 0.340 \\
-ct3 & 0.766 & 0.090 & 8.535 & 12.103 & 0.000 & 1.352 & 0.631 \\
-\end{longtable}
-
-\begin{verbatim}
-R> est$extra.pars |> 
-+  round(3) |>
-+  kable()
-\end{verbatim}
-
-\begin{longtable}[]{@{}lr@{}}
-\toprule\noalign{}
-& Estimate \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Intercept\textasciitilde\textasciitilde Intercept\textbar name &
-0.082 \\
-\end{longtable}
-
-\hypertarget{obesity-data}{%
-\subsection{obesity data}\label{obesity-data}}
-
-In this example, we demonstrate a multilevel imputation of random
-intercept and random slope model with a continuous response. We utilize
-the obesity dataset included in the \texttt{micemd}@ package, a
-simulated dataset that emulates an electronic survey in which
-individuals are asked to provide information about their weight and
-consumption habits in different countries. We simulate data for 5
-clusters so that the true values are known. We use the following
-variables from the dataset:
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Cluster:} Region of the patients' healthcare provider (Cluster
-  variable),
-\item
-  \textbf{Gender:} Subjects' Gender (0=male, 1=female),
-\item
-  \textbf{Age:} Subjects' age,
-\item
-  \textbf{Height:} Subjects' height in metres,
-\item
-  \textbf{Weight:} Subjects' weight in kilograms,
-\item
-  \textbf{BMI:} Subjects' body mass index,
-\item
-  \textbf{FamOb:} Family obesity history (yes or no),
-\item
-  \textbf{Time:} Response time in minutes (exclusion-restriction
-  variable).
-\end{itemize}
-
-In this dataset, Age and FamOb are MAR, while the weight variable is
-affected by selection bias, attributed to an indirect MNAR mechanism.
-This MNAR mechanism typically arises when an unobserved or omitted
-variable influences both the value of the incomplete variable (in this
-case, Weight) and its likelihood of being missing (denoted as R).
-
-In the primary analysis model, BMI serves as the dependent variable,
-with Age, Gender, and FamOb as predictors. Because of the clustered
-nature of the data, which is quantified with the Intraclass Correlation
-Coefficient (ICC) below, we include random intercepts, as well as a
-random slope for the Age variable. The model is represented as:
-\begin{equation}
-\label{eqn:main}
-BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}
-\end{equation}
-
-We start by loading the data:
-
-\begin{verbatim}
-R> data(Obesity, package = "micemd")
-\end{verbatim}
-
-Now, let's begin by examining the missing patterns in the data by
-cluster:
-
-\begin{verbatim}
-R> plot_pattern(Obesity)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-48-1.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-R> Obesity |> 
-+  split(~Cluster) |>
-+  lapply(plot_pattern)
-\end{verbatim}
-
-\begin{verbatim}
-$`1`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-1.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-
-$`2`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-2.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-
-$`3`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-3.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-
-$`4`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-4.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
-
-$`5`
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-5.pdf}
-
-}
-
-\end{figure}
-
-We observe that the missing pattern is non-monotonic and quite similar
-across the clusters. However, regarding the weight variable, we notice
-that is systematically missing in cluster 3. In order to evaluate if we
-require a imputation method that accounts for clustering we assess the
-Intraclass Correlation
-
-\begin{verbatim}
-# Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-# performance::icc(Nulmodel)
-\end{verbatim}
-
-Since the ICC is above 0.1 and as the main analysis will be use a mixed
-model, we decide to use two-level (2l) imputation methods. In this
-imputation process, we include all predictor variables from equation
-\ref{eqn:main} in the main model. However, since BMI is a composite of
-weight and height, we use deterministic imputation for these, which is
-described below.
-
-We use the \textbf{find.defaultMethodfunction} provided in the
-\textbf{micemd} package, which suggests an appropriate method for MAR
-variables based on the type of variable, number of observations in the
-cluster, and number of clusters.
-
-It suggests using `2l.2stage.bin' for the FAV variable and
-`2l.2stage.norm' for the age variable. However, after inspecting the age
-density plot, we consider modifying its method to `2l.2stage.pmm'. For
-the BMI variable, we employ deterministic imputation.
-
-\begin{verbatim}
-# meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
-#                                ni.small = 100, prop.small = 0.4)
-# meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
-# meth_mar["Age"]<-"2l.2stage.pmm" 
-\end{verbatim}
-
-For these imputation models, it is necessary to specify the prediction
-matrix, with the cluster variable labelled as -2 and the predictor
-variable measured within clusters labelled as 2, encompassing all
-variables. We need to suprime the variable Time as this variable is not
-specified in the main model. We also modify the relationship between
-BMI, weight and height in the prediction matrix to avoid circular
-predictions. Then we proceed to run the imputation model.
-
-\begin{verbatim}
-# pred_mar <- mice(Obesity, maxit = 0)$pred 
-# pred_mar[,"Cluster"] <- -2 # clustering variable
-# pred_mar[,"Time"] <- 0
-# pred_mar[pred_mar==1] <- 2
-# pred_mar[c("Height", "Weight"), "BMI"] <- 0
-# ggmice::plot_pred(pred_mar)
-# imp_mar <- mice::mice(data = Obesity, meth = meth_mar, pred = pred_mar,
-#                 m=10, seed = 123, printFlag = FALSE)
-\end{verbatim}
-
-\begin{verbatim}
-# summary(complete(imp_mar,"long")$Weight)
-\end{verbatim}
-
-We are also contemplating the utilisation of the predictive mean
-matching (pmm) option, as the values imputed using a fully parametric
-method may be implausibly low for some patients.
-
-\begin{verbatim}
-# meth_mar["Weight"]<-"2l.2stage.pmm" 
-# imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
-#                     m=10, seed = 123, printFlag = FALSE)
-\end{verbatim}
-
-\begin{verbatim}
-# summary(complete(imp_mar_pmm,"long")$Weight)
-# ggmice::plot_trace(imp_mar_pmm, "Weight")
-\end{verbatim}
-
-After confirming convergence, we proceed to save the results for future
-use. We consider the possibility that patients may not have been
-selected randomly, which would then have led to a distribution for
-weight that does not reflect the weight in the population. It???s likely
-that an omitted variable, like self-esteem, could influence this
-selection. For instance, individuals with lower self-esteem might have
-higher weight values, impacting their willingness to provide honest
-information due to embarrassment.
-
-To address this situation, two approaches have been proposed for dealing
-with Missing Not at Random (MNAR) data: pattern-mixed models and
-selection models. Within pattern-mixed models, methods like the delta
-method and more advanced ones like NARFS have been suggested. The
-selection model approach includes methods such as the Heckman model,
-which can be particularly useful in this case. Several methods,
-including those by \cite{galimard2018}, and the recently a Heckman
-method designed for two-level data, allow for variations in intercepts
-and exposure effects (random intercept and slope) \cite{munoz2023}.
-
-To apply the \textbf{2l.2stage.heckman} method, the weight variable
-should be specified as `2l.2stage.heckman' found in the micemd package.
-Additionally, the prediction matrix needs modification because this
-method involves specifying two equations: one for the outcome,
-describing the incomplete variable in terms of partially observed
-predictors (in this case, all variables from the main model), and the
-other for the selection model, explaining the probability of being
-observed based (R) on certain variables. For the outcome equation we
-consider the same imputation model that we used for the MAR case (main
-model).
-\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
-Regarding the selection equation, we include the same predictors as
-those in the main model, as well as a time variable. Here the time
-variable serves as a restriction exclusion variable specifically
-explaining the probability of being observed but not affecting the
-incomplete value (Weight). In this context, we assume that the time a
-user spends completing the survey serves as a proxy for the barriers
-they may encounter in survey completion, such as familiarity with the
-survey content or internet speed. These factors may lead the user to
-skip specific questions or even the entire survey. Also, we assume the
-time does not have any influence on the subject???s weight.
-\[R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
-
-These two equations are jointly estimated under the assumption that the
-error terms are interconnected with a bivariate normal distribution. For
-a more comprehensive understanding of the model and the exclusion
-restriction, see \cite{munoz2023a}.
-
-To use information from both equations, we must adjust the prediction
-matrix. The cluster variable remains specified as before (-2). In this
-imputation method, all the variables present in both the selection and
-outcome equations are included with a random effect.
-
-However, it is essential to distinguish which of these variables appear
-in each equation. In this framework, when a variable is shared between
-both equations, it is denoted as (2). Predictors exclusive to the
-outcome equation are indicated as (-4), while those exclusive to the
-selection equation are labelled as (-3). Consequently, the only
-alteration needed in the predictor matrix pertains to the variable
-`Time'.
-
-\begin{verbatim}
-# pred_mnar <- pred_mar
-# pred_mnar["Weight","Time"]<- -3
-# ggmice::plot_pred(pred_mnar)
-\end{verbatim}
-
-We also need to modify the method of the weight variable.
-
-\begin{verbatim}
-# meth_mnar <- meth_mar
-# meth_mnar["Weight"]<- "3l.2stage.heckman"
-\end{verbatim}
-
-Then we proceed to run the imputation model as before, after executing
-these imputation procedures, it is essential to assess convergence and
-the coherence of the imputed values.
-
-\begin{verbatim}
-# imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
-#                 m=10, seed = 123, printFlag = FALSE)
-# summary(complete(imp_mnar,"long")$Weight)
-\end{verbatim}
-
-Upon examining the weight variable, we noticed that the imputed range
-falls outside the realm of plausible values (as weight should be
-positive).
-
-\begin{verbatim}
-# summary(complete(imp_mnar,"long")$Weight)
-\end{verbatim}
-
-Consequently, as before we use the `pmm', option but this time for the
-Heckman imputation, this approach ensures that the imputed values remain
-within the range of observable values. We then run the imputation model
-but this time using the option of pmm, to assure that weight values are
-in the range of the observable data, this can be implemented by setting
-the pmm parameter to true.
-
-\begin{verbatim}
-# imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
-#                      m=10, seed = 123, pmm = T,  printFlag = FALSE)
-\end{verbatim}
-
-We check the convergency of the results
-
-\begin{verbatim}
-# summary(complete(imp_mnar_pmm,"long")$Weight)
-# ggmice::plot_trace(imp_mar_pmm, "Weight")
-\end{verbatim}
-
-After this modification we proceed to compare the effects on the model.
-We run the analysis model on each of the completed datasets as well
-asthe dataset where the incomplete values are removed (Complete Case
-analysis, CC).
-
-\begin{verbatim}
-# cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
-#              lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
-# mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
-# list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
-# plot_models(list_models,
-#             mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
-\end{verbatim}
-
-We note that there is minimal disparity in the age effect, FamObs, or
-Gender across the various imputation models under consideration. An
-analysis of the intercept reveals that, under the MNAR assumption, a
-higher average BMI is anticipated compared to the MAR assumption.
-Nonetheless, with respect to precision of estimates, we notice that in
-general MNAR imputation leads to wider confidence intervals, in this
-case it does not have any influence on the final result but there could
-be cases where variation in the assumed missing mechanism could lead
-also to differences on significant test and therefore lead to
-contradictory conclusions.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-This paper is dedicated to exploring the imputation process for
-incomplete datasets, with a primary focus on utilizing a hierarchical
-model for analysis. Initially, users are encouraged to consider the main
-analysis within the context of the incomplete dataset, along with
-insights provided by domain experts, to gain a better understanding of
-variable relationships. Employing clear data visualization is
-instrumental in comprehending the missing data patterns, establishing a
-missing mechanism, and aiding in the selection of suitable imputation
-methods.
-
-The ``Mice'' and ``mice''-based R packages offer a range of imputation
-methods tailored for hierarchical data, easily adaptable to the
-dataset's structure. Before proceeding with the analysis of the imputed
-dataset, it is essential to assess the convergence of the imputation
-method. This evaluation can reveal issues such as circular problems, the
-need for additional iterations, or challenges associated with the chosen
-imputation method.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-This project has received funding from the European Union's Horizon 2020
-research and innovation programme under ReCoDID grant agreement No
-825746.
-
-The views expressed in this paper are the personal views of the authors
-and may not be understood or quoted as being made on behalf of or
-reflecting the position of the regulatory agency/agencies or
-organizations with which the authors are employed/affiliated.
+As a first model for the \texttt{quine} data, we fit the basic Poisson
+regression model. (Note that JSS prefers when the second line of code is
+indented by two spaces.)
+
+\begin{verbatim}
+R> m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine, family = poisson)
+\end{verbatim}
+
+To account for potential overdispersion we also consider a negative
+binomial GLM.
+
+\begin{verbatim}
+R> library("MASS")
+R> m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
+\end{verbatim}
+
+In a comparison with the BIC the latter model is clearly preferred.
+
+\begin{verbatim}
+R> library("MASS")
+R> BIC(m_pois, m_nbin)
+\end{verbatim}
+
+\begin{verbatim}
+       df      BIC
+m_pois 18 2046.851
+m_nbin 19 1157.235
+\end{verbatim}
+
+Hence, the full summary of that model is shown below.
+
+\begin{verbatim}
+R> summary(m_nbin)
+\end{verbatim}
+
+\begin{verbatim}
+
+Call:
+glm.nb(formula = Days ~ (Eth + Sex + Age + Lrn)^2, data = quine, 
+    init.theta = 1.60364105, link = log)
+
+Coefficients: (1 not defined because of singularities)
+            Estimate Std. Error z value Pr(>|z|)    
+(Intercept)  3.00155    0.33709   8.904  < 2e-16 ***
+EthN        -0.24591    0.39135  -0.628  0.52977    
+SexM        -0.77181    0.38021  -2.030  0.04236 *  
+AgeF1       -0.02546    0.41615  -0.061  0.95121    
+AgeF2       -0.54884    0.54393  -1.009  0.31296    
+AgeF3       -0.25735    0.40558  -0.635  0.52574    
+LrnSL        0.38919    0.48421   0.804  0.42153    
+EthN:SexM    0.36240    0.29430   1.231  0.21818    
+EthN:AgeF1  -0.70000    0.43646  -1.604  0.10876    
+EthN:AgeF2  -1.23283    0.42962  -2.870  0.00411 ** 
+EthN:AgeF3   0.04721    0.44883   0.105  0.91622    
+EthN:LrnSL   0.06847    0.34040   0.201  0.84059    
+SexM:AgeF1   0.02257    0.47360   0.048  0.96198    
+SexM:AgeF2   1.55330    0.51325   3.026  0.00247 ** 
+SexM:AgeF3   1.25227    0.45539   2.750  0.00596 ** 
+SexM:LrnSL   0.07187    0.40805   0.176  0.86019    
+AgeF1:LrnSL -0.43101    0.47948  -0.899  0.36870    
+AgeF2:LrnSL  0.52074    0.48567   1.072  0.28363    
+AgeF3:LrnSL       NA         NA      NA       NA    
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for Negative Binomial(1.6036) family taken to be 1)
+
+    Null deviance: 235.23  on 145  degrees of freedom
+Residual deviance: 167.53  on 128  degrees of freedom
+AIC: 1100.5
+
+Number of Fisher Scoring iterations: 1
+
+              Theta:  1.604 
+          Std. Err.:  0.214
+
+ 2 x log-likelihood:  -1062.546 
+\end{verbatim}
 
@@
 
-What is missing from this manuscript\ldots{}
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]
+
+As usual\ldots{}
+
+\end{tcolorbox}
 
@@
 
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]
+
+If necessary or useful, information about certain computational details
+such as version numbers, operating systems, or compilers could be
+included in an unnumbered section. Also, auxiliary packages (say, for
+visualizations, maps, tables, \ldots) that are not cited in the main
+text can be credited here.
+
+\end{tcolorbox}
+
 The results in this paper were obtained using
-\proglang{R}\textasciitilde4.3.0. \proglang{R} itself and all packages
-used are available from the Comprehensive \proglang{R} Archive Network
-(CRAN) at {[}https://CRAN.R-project.org/{]}.
+\proglang{R}\textasciitilde3.4.1 with the
+\pkg{MASS}\textasciitilde7.3.47 package. \proglang{R} itself and all
+packages used are available from the Comprehensive \proglang{R} Archive
+Network (CRAN) at {[}https://CRAN.R-project.org/{]}.
 
@@
 
-This project has received funding from the European Union's Horizon 2020
-research and innovation programme under ReCoDID grant agreement No
-825746.
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]
+
+All acknowledgments (note the AE spelling) should be collected in this
+unnumbered section before the references. It may contain the usual
+information about funding and feedback from colleagues/reviewers/etc.
+Furthermore, information such as relative contributions of the authors
+may be added here (if any).
+
+\end{tcolorbox}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, toprule=.15mm, rightrule=.15mm, arc=.35mm, left=2mm, leftrule=.75mm, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]
 

### 2023-12-19 Manuscript/manuscript.qmd
Rename foldes

--- 2023-12-12 docs/manuscript.qmd
@@
 
-The imputation model can include additional terms than the main model terms that do not lead to uncongeniality.
-For instance, it is advisable to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}.
-In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be included as a covariate in the imputation model [REF].
-Furthermore, the inclusion of auxiliary variables, even if not part of the main model, can be associated with the probability of missingness, enhancing the likelihood of satisfying the Missing at Random (MAR) assumption and improving estimation efficiency \cite{hardt2012a}.
-
-The specification of imputation models is done on a variable basis, either using a prediction matrix where the type of prediction variable should be specified accordingly (see table) or through a list of formulas in the formula parameter.
-The latter option proves useful in formulating imputation models with polynomial terms or interactions compared to the prediction matrix specification, which requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}.
-
-However, even when considering an interaction term, depending on the adapted individual imputation model, it has been suggested, for instance, for treatment interaction effects, to conduct separated imputation by treatment group \cite{zhang2023}.
-Additionally, there are imputation models based on random forest or deep learning that can handle interaction and non-linear terms that do not require the explicit inclusion of the non-linear terms.
+The imputation model can incorporate additional terms compared to the main model without causing compatibility issues.
+For instance, it's recommended to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}.
+In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be added as a covariate in the imputation model [REF].
+Additionally, including auxiliary variables, even if not part of the main model, can be linked to the probability of missingness, improving the likelihood of meeting the Missing at Random (MAR) assumption and enhancing estimation efficiency \cite{hardt2012a}.
+
+Imputation models are specified on a variable basis, either using a prediction matrix in the pred parameter or through a list of formulas in the formula parameter.
+In the prediction matrix option, the type of each predictor variable is specified for each incomplete variable (see table).
+
+\begin{align}
+y_{ij} =& (\beta_0 + b_{0_i})+ \beta_1x_{1_{ij}}+\beta_2x_{2_{i.}} \nonumber \\
++& (\beta_3 +  b_{3_i})x_{3_{ij}} \nonumber \\
++& \beta_4x_{4_{ij}} + \beta_{m4}\overline{x_{4_{i.}}}\nonumber \\
++& (\beta_5 +  b_{5_i})x_{5_{ij}} +\beta_{m5}\overline{x_{5_{i.}}}\nonumber \\
+\end{align}
+
+\begin{tabular}{ll}
+     \toprule
+Type & Definition \\
+  \midrule
+-2  & Cluster variable, in this case the one defined by j index  \\
+ 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
+ 2  & Random variable, e.g., $x_{3_{ij}}$\\
+ 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
+ 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
+-3. & Random variable only included on selection model (Heckman model)\\
+-4. & Random variable only included on main model (Heckman model)\\
+      \bottomrule
+\end{tabular}
+
+* 3 and 4 type have been advised to used as the inclusion of the means of the cluster is beneficial on FCS \cite{mistler2017}
+
+Recipes have been proposed for imputing incomplete level-1 and level-2 variables for hierarchical models \cite{buuren2018a} ($\S$ 7.10), which can be convenient to follow when dealing with numerous variables (interaction terms at different levels) and models with many random effects that are prone to convergency problems due to overspecification in the imputation model.
+These rules were designed to ensure compatibility among the conditionally specified imputation models and congeniality between the imputation and the main model.
+However, they may not be applicable to all 2l imputation methods; for instance, the 2l.2stage imputation methods of the micemd package only allow the inclusion of random predictor variables (2).
+
+On the other hand, the formula option is useful in specifying complex imputation models with polynomial terms or interactions and compared with the prediction matrix method do not requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}($\S$ 6.4).
+
+For some interaction terms, it has been suggested, for instance, for treatment interaction effects, to conduct separate imputation by treatment group \cite{zhang2023}.
+Additionally, it can be used imputation models based on random forest or deep learning that can handle interaction and non-linear terms without requiring the explicit specification of an imputation model.
 
@@
 
+Given the $j$-th student belonging to the $i$-th school, the main model can be formulated as:
+
+$$ popular_{ij} = (\beta_0 + b_i) + \beta_1sex_{ij}+ \beta_2texp_{ij}+\epsilon_{ij}$$
+where $\epsilon_{ij}$ corresponds to the error term.
+
 We load the data into the environment with
@@
 ```
+
+In this dataset all the variables are level-1 (patient), except by the level-2 (study) type variable.
+The analysis model for this dataset is a prediction model with `mort` as the outcome.
+In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes.
+The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil`.
+Therefore the main model for the $i$-th patient from the $j$-th study can be described as:
+
+$$ mort_{ij} = (\beta_0 + b_i) + \beta_1type_{.j}+ \beta_2age_{ij}+ \beta_3motorscore_{ij} + \beta_4pupil_{ij} + \beta_5ct_{ij}+\epsilon_{ij}$$
+where $\epsilon_{ij}$ corresponds to the error term.
 

### 2023-12-22 Manuscript/manuscript.qmd
Re-run on different pc

@@
 library(ggplot2)
-source("../docs/utils.R")
-source("../docs/tables.R")
+source("../Manuscript/utils.R")
+source("../Manuscript/tables.R")
 ```

### 2023-12-22 Manuscript/manuscript.qmd
Render manuscript

--- 2023-12-19 Manuscript/manuscript.qmd
@@
 #| message: false
+set.seed(1)
 library(knitr)
 library(ggplot2)
-source("../docs/utils.R")
-source("../docs/tables.R")
+source("../Manuscript/utils.R")
+source("../Manuscript/tables.R")
 ```

### 2023-12-22 Manuscript/manuscript.tex
Render manuscript

--- 2023-12-12 docs/manuscript.tex
@@
 \usepackage{amsmath,amssymb}
+\usepackage{lmodern}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
 \fi
@@
 \makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
 \@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
@@
 \makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
 \makeatother
@@
 \maketitle
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
+
 \hypertarget{sec-intro}{%
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, arc=.35mm, rightrule=.15mm, breakable, opacityback=0, bottomrule=.15mm, colback=white, leftrule=.75mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-color-frame, opacityback=0, colback=white, rightrule=.15mm, left=2mm]
 
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.1008}}@{}}
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedleft
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 -0.2599581 & 14.15278 & 14.41274 & -0.8731291 & 0.3827095 & 1796.115 &
 -0.8438947 & 0.3239786 & Welch Two Sample t-test & two.sided \\
+\bottomrule()
 \end{longtable}
@@
   >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}@{}}
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 (Intercept) & 3.599 & 0.181 & 19.924 & 246.857 & 0 & 0.146 & 0.134 \\
@@
 sex & 0.852 & 0.039 & 21.592 & 27.501 & 0 & 0.616 & 0.422 \\
+\bottomrule()
 \end{longtable}
@@
 \begin{longtable}[]{@{}lr@{}}
-\toprule\noalign{}
+\toprule()
 & Estimate \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Intercept\textasciitilde\textasciitilde Intercept\textbar school &
@@
 ICC\textbar school & 0.504 \\
+\bottomrule()
 \end{longtable}
@@
   >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}@{}}
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 (Intercept) & -1.892 & 0.333 & -5.682 & 11445.455 & 0.000 & 0.019 &
@@
 ct3 & 0.766 & 0.090 & 8.535 & 12.103 & 0.000 & 1.352 & 0.631 \\
+\bottomrule()
 \end{longtable}
@@
 \begin{longtable}[]{@{}lr@{}}
-\toprule\noalign{}
+\toprule()
 & Estimate \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Intercept\textasciitilde\textasciitilde Intercept\textbar name &
 0.082 \\
+\bottomrule()
 \end{longtable}
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
-(Intercept) & 28.445 & 2.262 & 12.578 & 158.191 & 0.000 & 0.189 &
-0.169 \\
-Age & 0.041 & 0.032 & 1.312 & 383.212 & 0.190 & 0.114 & 0.107 \\
-FamObyes & 0.000 & 0.316 & -0.001 & 207.642 & 0.999 & 0.161 & 0.147 \\
-GenderMale & -1.361 & 0.411 & -3.310 & 7.985 & 0.011 & 2.422 & 0.761 \\
+(Intercept) & 28.445 & 2.259 & 12.592 & 157.553 & 0.000 & 0.190 &
+0.170 \\
+Age & 0.041 & 0.032 & 1.311 & 388.953 & 0.191 & 0.113 & 0.106 \\
+FamObyes & 0.000 & 0.316 & -0.001 & 207.648 & 0.999 & 0.161 & 0.147 \\
+GenderMale & -1.361 & 0.411 & -3.310 & 7.986 & 0.011 & 2.421 & 0.761 \\
+\bottomrule()
 \end{longtable}
@@
 \begin{longtable}[]{@{}lr@{}}
-\toprule\noalign{}
+\toprule()
 & Estimate \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Intercept\textasciitilde\textasciitilde Intercept\textbar Cluster &
-19.862 \\
-Age\textasciitilde\textasciitilde Age\textbar Cluster & 0.002 \\
-Intercept\textasciitilde\textasciitilde Age\textbar Cluster & -0.067 \\
+19.805 \\
+Age\textasciitilde\textasciitilde Age\textbar Cluster & 0.003 \\
+Intercept\textasciitilde\textasciitilde Age\textbar Cluster & -0.069 \\
 Residual\textasciitilde\textasciitilde Residual & 26.010 \\
-ICC\textbar Cluster & 0.395 \\
+ICC\textbar Cluster & 0.394 \\
+\bottomrule()
 \end{longtable}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  16.26   70.95   85.21   86.14  100.14  190.95 
+  16.31   72.14   86.97   87.68  102.43  190.44 
 \end{verbatim}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  28.35   71.51   85.07   85.35   98.90  134.61 
+  28.35   71.87   85.58   86.51  100.15  134.61 
 \end{verbatim}
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1558}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1169}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1299}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2338}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}@{}}
-\toprule\noalign{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1579}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1184}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1316}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1053}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0921}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2368}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0789}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0789}}@{}}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
-(Intercept) & 28.880 & 2.391 & 12.078 & 56.876 & 0.000 & 0.361 &
-0.290 \\
-Age & 0.050 & 0.037 & 1.343 & 19.874 & 0.194 & 0.814 & 0.497 \\
-FamObyes & -0.027 & 0.490 & -0.056 & 11.685 & 0.957 & 1.410 & 0.642 \\
-GenderMale & -1.484 & 0.804 & -1.846 & 4.813 & 0.126 & 10.318 & 0.934 \\
+(Intercept) & 29.199 & 2.766 & 10.555 & 55.483 & 0.000 & 0.367 &
+0.294 \\
+Age & 0.050 & 0.038 & 1.301 & 18.731 & 0.209 & 0.859 & 0.512 \\
+FamObyes & 0.016 & 0.445 & 0.036 & 15.910 & 0.972 & 1.006 & 0.554 \\
+GenderMale & -1.408 & 0.696 & -2.024 & 5.128 & 0.097 & 7.563 & 0.912 \\
+\bottomrule()
 \end{longtable}
@@
 \begin{longtable}[]{@{}lr@{}}
-\toprule\noalign{}
+\toprule()
 & Estimate \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Intercept\textasciitilde\textasciitilde Intercept\textbar Cluster &
-19.111 \\
-Age\textasciitilde\textasciitilde Age\textbar Cluster & 0.001 \\
-Intercept\textasciitilde\textasciitilde Age\textbar Cluster & -0.046 \\
-Residual\textasciitilde\textasciitilde Residual & 30.071 \\
-ICC\textbar Cluster & 0.372 \\
+26.114 \\
+Age\textasciitilde\textasciitilde Age\textbar Cluster & 0.002 \\
+Intercept\textasciitilde\textasciitilde Age\textbar Cluster & -0.088 \\
+Residual\textasciitilde\textasciitilde Residual & 29.741 \\
+ICC\textbar Cluster & 0.431 \\
+\bottomrule()
 \end{longtable}
@@
 
-\renewcommand{\bibsection}{}
-\bibliography{bibliography.bib}
+\hypertarget{refs}{}
+\begin{CSLReferences}{0}{0}
+\end{CSLReferences}
 
@@
 
+  \bibliography{bibliography.bib}
+
 \end{document}

### 2024-01-08 Manuscript/manuscript.qmd
small textual change in intro

@@
 This tutorial is dedicated to readers who are unfamiliar with multiple imputation.
-More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2) or under MNAR conditions (case study 3).
+More experienced readers can skip the remainder of the introduction section, background section and case study 1, and directly head to practical applications of multilevel imputation under MAR conditions (case study 2) or under MNAR conditions (case study 3).
 

### 2024-01-10 Manuscript/manuscript.qmd
Updated manuscript
I have removed the 'no pmm' part from the Heckman model. Additionally, I've included the plot function with echo=FALSE, making it suitable for copying and pasting into the appendix later on. However, the pattern function with cluster still does not work. Regarding the box 3 about the prediction matrix, previously i have added relevant information about this (with an equation) in the Imputation flow. Please review this information to ensure that the box is constructed appropriately, avoiding redundant information throughout the paper. Thanks.

--- 2023-12-22 Manuscript/manuscript.qmd
@@
 The popular [mice]{.pkg} package in [R]{.proglang} @R
-
-### scope
+TODO: @Hanne(Is this section relevant?)
+
+### Scope
 
@@
 
-### concepts in multilevel data
+### Concepts in multilevel data
 
@@
 These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'.
-Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.
-
-::: callout
-Box 1. The intraclass correlation coefficient. 
-:::
-
-```{R}
-#| echo: false
-# concepts |>
-#   kable(
-#     booktabs = T,
-#     escape = F,
-#     caption = "Concepts in multilevel methods",
-#     col.names = c("Concept", "Details"),
-#     align = "l"
-#   ) |>
-#   column_spec(c(1), width = "3cm") |>
-#   column_spec(c(2), width = "12cm")
-```
-
-In [R]{.proglang}, multilevel models may be fitted using the package [lme4]{.pkg}. For linear mixed-effects models, the function
+Table \ref{tab:clus0} provides an overview of some key concepts in multilevel modeling.
+
+```{r clus0,  echo = FALSE, warning = FALSE,cache = TRUE}
+concepts <- data.frame(Concept=c("Sample units","Cluster","Hierarchical data","Level-1","Level-2","Hierarchical model","Fixed effect","Random effect","Mixed effect","ICC","Stratified intercept"),
+                       Details=c("Units of the population from which measurements are taken in a sample, e.g., students.", "Variable that specify the cluster or agruppation, e.g., Classroom","Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.","Variable that varies within a cluster, eg.
+Test score","Variable that does not vary within a cluster but between, e.g.
+teacher experience.","Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)" ,"Effects that are constant across all sample units, e.g.
+something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)","Effects that are a source of random variation in the data, and whose levels are not fully sampled.
+e.g.
+test score tendency during academic year between students due to no controlled factors such as  genetic,family history","Includes fixed and random effects, e.g.
+the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
+Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
+In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
+@gelm06, @hox17 and @jong21).","The variability due to clustering is often measured by means of the intraclass coefficient (ICC).
+The ICC can be seen as the percentage
+of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
+                                 ""))
+library(kableExtra)
+concepts%>%
+  kable( booktabs = T, escape = F,
+         caption = "Concepts in multilevel methods",
+         col.names = c("Concept","Details"), align = "l")%>%
+  column_spec(c(1), width = "3cm")%>%
+  column_spec(c(2), width = "12cm")
+```
+
+In [R]{.proglang}, multilevel models may be fitted using package such as  [lme4]{.pkg} or [nlme]{.pkg}. For linear mixed-effects models, the function
 
@@
 
-### concepts in missing data
+### Concepts in missing data
 
@@
 
-### imputation with mice
+Table 2: Concepts in missing data methods
+
+| **Concept**    | **Details**                                                                |
+|----------------|----------------------------------------------------------------------------|
+| MCAR           | Missing Completely At Random, where the probability to be missing is equal |
+|                | across all data entries                                                    |
+| MAR            | Missing At Random, where the probability to be missing depends on observed |
+|                | information                                                                |
+| MNAR           | Missing Not At Random (MNAR), where the probability to be missing          |
+|                | depends on unrecorded information, making the missingness non-ignorable    |
+|                | [@rubi76; @meng94].                                                        |
+|                |                              |
+
+### Imputation with mice
 
@@
 ::: callout
+TODO: @ Hanne i added a wide explanation about hte prediction matrix in the Model Specification subsection, so you can check it and consider to modify your box..
+because i dont think is usefull to repeat information across the paper...
+
 Box 3.
@@
 
-There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.
+There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.\cite{zotero-516}
 
@@
 Sometimes, simpler methods like complete case analysis might be suffice, especially if your missingness is low (usually <5%).
-There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods, for instance when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}.
+There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods, for instance when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{molenberghs2007}.
 
@@
 The imputation model can incorporate additional terms compared to the main model without causing compatibility issues.
-For instance, it's recommended to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}.
+For instance, it's recommended to include the outcome variable in the imputation model for prediction variables \cite{moons2006}.
 In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be added as a covariate in the imputation model [REF].
-Additionally, including auxiliary variables, even if not part of the main model, can be linked to the probability of missingness, improving the likelihood of meeting the Missing at Random (MAR) assumption and enhancing estimation efficiency \cite{hardt2012a}.
+Additionally, including auxiliary variables, even if not part of the main model, can be linked to the probability of missingness, improving the likelihood of meeting the Missing at Random (MAR) assumption and enhancing estimation efficiency \cite{hardt2012}.
 
@@
 
-Recipes have been proposed for imputing incomplete level-1 and level-2 variables for hierarchical models \cite{buuren2018a} ($\S$ 7.10), which can be convenient to follow when dealing with numerous variables (interaction terms at different levels) and models with many random effects that are prone to convergency problems due to overspecification in the imputation model.
+Recipes have been proposed for imputing incomplete level-1 and level-2 variables for hierarchical models \cite{buur18} ($\S$ 7.10), which can be convenient to follow when dealing with numerous variables (interaction terms at different levels) and models with many random effects that are prone to convergency problems due to overspecification in the imputation model.
 These rules were designed to ensure compatibility among the conditionally specified imputation models and congeniality between the imputation and the main model.
@@
 
-On the other hand, the formula option is useful in specifying complex imputation models with polynomial terms or interactions and compared with the prediction matrix method do not requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}($\S$ 6.4).
+On the other hand, the formula option is useful in specifying complex imputation models with polynomial terms or interactions and compared with the prediction matrix method do not requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buur18}($\S$ 6.4).
 
@@
 This can occur when missingness is associated with the incomplete value itself (self-marking) or when there is an unobserved variable linked to both the value and the probability of missingness of the incomplete value (indirectly non-informative).
-Specifically, for the indirectly non-informative case in hierarchical datasets, imputation methods based on the Heckman method can be considered.\cite{hammon2020,hammon2022,munoz2023}
+Specifically, for the indirectly non-informative case in hierarchical datasets, imputation methods based on the Heckman method can be considered.\cite{hammon2020,hammon2022,munoz2023b}
 
@@
 
-### obesity data
+### Obesity data
 
@@
 #| prompt: true
-Nulmodel <- lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-# performance::icc(Nulmodel)
+Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+performance::icc(Nulmodel)
 ```
@@
 
-It suggests using '2l.2stage.bin' for the FAV variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
+It suggests using '2l.2stage.bin' for the FAV variable and '2l.2stage.norm' for the age variable. However, after inspecting the age and weight density plot, we consider modifying its method to
 '2l.2stage.pmm'. For the BMI variable, we employ deterministic imputation.
@@
 meth_mar["Age"] <- "2l.2stage.pmm"
+meth_mar["Weight"] <- "2l.2stage.pmm"
+
 ```
@@
 pred_mar <- quickpred(Obesity)
+pred_mar["Weight",] <- 2
+pred_mar["Weight","Weight"] <- 0
 pred_mar[, "Cluster"] <- -2 
@@
 pred_mar[pred_mar == 1] <- 2
+
 pred_mar[c("Height", "Weight"), "BMI"] <- 0
-plot_pred(pred_mar)
+#plot_pred(pred_mar)
 imp_mar <-
@@
 
-```{r obesity-predmar1,eval=TRUE}
-summary(complete(imp_mar,"long")$Weight)
-```
-We are also contemplating the utilisation of the predictive mean matching (pmm) option, as the values imputed using a fully parametric method may be implausibly low for some patients.
-
-```{r obesity-predmar_pmm, eval=TRUE}
-meth_mar["Weight"] <- "2l.2stage.pmm"
-imp_mar_pmm <- mice(
-  data = Obesity,
-  method = meth_mar,
-  predictorMatrix = pred_mar,
-  printFlag = FALSE
-)
-```
-
 ```{r obesity-predmar_pmm1,eval=TRUE}
-summary(complete(imp_mar_pmm, "long")$Weight)
-plot_trace(imp_mar_pmm, "Weight")
-ggmice(imp_mar_pmm, aes(.imp, Weight)) + 
+summary(complete(imp_mar, "long")$Weight)
+plot_trace(imp_mar, "Weight")
+ggmice(imp_mar, aes(.imp, Weight)) + 
   geom_jitter()
-```
-
-Analyze the imputed data.
-```{R}
-#| prompt: true
-#| warning: false
-# fit <- with(
-#     imp_mar_pmm,
-#     lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-#     )
-fit <- with(
-    imp_mar_pmm,
-    lmer(BMI ~ Age + FamOb + Gender + (1 + Age | Cluster))
-    )
-est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
-```
-
-Display results in table.
-```{r}
-#| prompt: true
-est$estimates |> 
-  round(3) |>
-  kable()
-```
-
-```{r}
-#| prompt: true
-est$extra.pars |> 
-  round(3) |>
-  kable()
 ```
@@
 These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution.
-For a more comprehensive understanding of the model and the exclusion restriction, see \cite{munoz2023a}.
+For a more comprehensive understanding of the model and the exclusion restriction, see \cite{munoz2023}.
 
@@
 ```
-Then we proceed to run the imputation model as before, after executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
-```{r obesity-predmnar2,eval=TRUE}
-imp_mnar <- mice(
-  data = Obesity,
-  method = meth_mnar,
-  predictorMatrix = pred_mnar,
-  printFlag = FALSE
-)
-summary(complete(imp_mnar, "long")$Weight)
-```
-
-Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive).
-
-```{r obesity-predmnar3,eval=TRUE}
-summary(complete(imp_mnar, "long")$Weight)
-```
-Consequently, as before we  use the 'pmm', option but this time for the Heckman imputation, this approach ensures that the imputed values remain within the range of observable values.
-We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by setting the pmm parameter to true.
+As before we  use the 'pmm', option but this time for the Heckman imputation, this approach ensures that the imputed values remain within the range of observable values.
+We then run the MNAR imputation model by setting the pmm parameter to true.
 
 ```{r obesity-predmnarp, eval=TRUE}
-imp_mnar_pmm <-
+imp_mnar <-
   mice(
@@
 ```{r obesity-predmnarp1, eval=TRUE}
-summary(complete(imp_mnar_pmm, "long")$Weight)
-plot_trace(imp_mar_pmm, "Weight")
-```
-
-After this modification we proceed to compare the effects on the model. We
+summary(complete(imp_mnar, "long")$Weight)
+plot_trace(imp_mar, "Weight")
+```
+
+We proceed to compare the effects on the model. We
 run the analysis model on each of the completed datasets as well as the dataset where the
 incomplete values are removed (Complete Case analysis, CC).
-
-```{r models, eval=TRUE}
-# cc_rs <- with(setDT(Obesity)[complete.cases(Obesity), ],
-#               lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age |
-#                     Cluster))
-mar_rs <- with(
-  imp_mar, 
-  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-  )
-mar_pmm_rs <- with(
-    imp_mar_pmm,
-    lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-    )
-mnar_rs <- with(
-  imp_mnar, 
-  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-  )
-mnar_pmm_rs <- with(
-  imp_mnar_pmm,
-  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-  )
-# list_models <- list(cc_rs, mar_rs, mar_pmm_rs, mnar_rs, mnar_pmm_rs)
-# plot_models(list_models,
-#             mod_name = c("Complete case", "MAR", "MAR_pmm", "MNAR", "MNAR_pmm"))
-```
-
-```{r}
-est <- testEstimates(as.mitml.result(mnar_pmm_rs), extra.pars = TRUE)
-```
-
-Display results in table.
-```{r}
-#| prompt: true
-est$estimates |> 
-  round(3) |>
-  kable()
-```
-
-```{r}
-#| prompt: true
-est$extra.pars |> 
-  round(3) |>
-  kable()
-```
+```{r plotfunction, eval=TRUE, echo=FALSE}
+library(broom.mixed)
+plot_models <- function(list_models, mod_name){
+  mod_table<-NULL
+  for(i in 1:length(list_models)){
+    mod <- list_models[[i]]  
+    if(class(mod)[[1]]=="mira"){
+      mod_tab <-summary(pool(mod), conf.int = TRUE)[,c("term","estimate","2.5 %","97.5 %","p.value")]
+    }else{
+      sum_cc <- (summary(mod,conf.int = TRUE))
+      mod_tab<-data.frame(nlme::intervals(mod, which = "fixed")$fixed)
+      mod_tab$term<-rownames(mod_tab)
+      mod_tab$pvalue<-sum_cc$tTable[,5]
+      mod_tab <-mod_tab[,c("term","est.","lower","upper","pvalue")]
+      rownames(mod_tab)<-NULL
+
+    }
+    colnames(mod_tab)<-c("Coefficient","Estimate","lower","upper","p.value")
+    mod_tab$model<-mod_name[i]
+    mod_table<-rbind(mod_table,mod_tab)
+  }
+
+  plot <-ggplot(mod_table, aes(x = Coefficient, y = Estimate, ymin = lower, ymax = upper, group=model, color=model)) + 
+    geom_hline(yintercept=0,linetype="dashed",color="black",size=0.1)+
+    geom_linerange( position = position_dodge(width = 0.8)) + 
+    geom_pointrange(size=0.1, position = position_dodge(width = 0.8))+coord_flip()+
+    theme_classic()+theme(legend.position = "bottom")+scale_color_viridis_d()
+  return(plot)
+}
+```
+We proceed to fit the mixed model by using the complete case database and the imputed datasets (MAR,MNAR), and we compare the estimates of each dataset by using a plot (the plot function can be found in the appendix).
+```{r models, eval=TRUE,warning=FALSE,message=FALSE}
+# Save all datasets in a list
+list_data <- list(Obesity |> filter(complete.cases(Obesity)), # complete case
+                  imp_mar, # imputed under mar
+                  imp_mnar) # imputed under mnar
+
+# Run mixed models on each dataset
+list_models <- lapply(list_data, 
+                   FUN = function(x) 
+                     with(x,nlme::lme(BMI ~ Age + FamOb + Gender, 
+                                      random =  ~ 1 + Age | Cluster)))
+# Plot coefficients
+plot_models(list_models,
+             mod_name = c("Complete case", "MAR", "MNAR"))
+```
+
 We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration.

### 2024-01-10 Manuscript/manuscript.tex
Updated manuscript
I have removed the 'no pmm' part from the Heckman model. Additionally, I've included the plot function with echo=FALSE, making it suitable for copying and pasting into the appendix later on. However, the pattern function with cluster still does not work. Regarding the box 3 about the prediction matrix, previously i have added relevant information about this (with an equation) in the Imputation flow. Please review this information to ensure that the box is constructed appropriately, avoiding redundant information throughout the paper. Thanks.

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, boxrule=0pt, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-The popular \pkg{mice} package in \proglang{R} \citet{R}
+The popular \pkg{mice} package in \proglang{R} \citet{R} TODO:
+\citet{Hanne}(Is this section relevant?)
 
 \hypertarget{scope}{%
-\subsection{scope}\label{scope}}
+\subsection{Scope}\label{scope}}
 
@@
 \hypertarget{concepts-in-multilevel-data}{%
-\subsection{concepts in multilevel
+\subsection{Concepts in multilevel
 data}\label{concepts-in-multilevel-data}}
@@
 and in the context of time-to-event data as `frailty models'. Table
-\ref{tab:clus} provides an overview of some key concepts in multilevel
+\ref{tab:clus0} provides an overview of some key concepts in multilevel
 modeling.
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-color-frame, opacityback=0, colback=white, rightrule=.15mm, left=2mm]
-
-Box 1. The intraclass correlation coefficient.
-
-\end{tcolorbox}
-
-In \proglang{R}, multilevel models may be fitted using the package
-\pkg{lme4}. For linear mixed-effects models, the function
+\begin{table}
+
+\caption{Concepts in multilevel methods}
+\centering
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
+\toprule
+Concept & Details\\
+\midrule
+Sample units & Units of the population from which measurements are taken in a sample, e.g., students.\\
+Cluster & Variable that specify the cluster or agruppation, e.g., Classroom\\
+Hierarchical data & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
+Level-1 & Variable that varies within a cluster, eg. Test score\\
+Level-2 & Variable that does not vary within a cluster but between, e.g. teacher experience.\\
+\addlinespace
+Hierarchical model & Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)\\
+Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)\\
+Random effect & Effects that are a source of random variation in the data, and whose levels are not fully sampled.
+e.g.
+test score tendency during academic year between students due to no controlled factors such as  genetic,family history\\
+Mixed effect & Includes fixed and random effects, e.g.
+the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated.
+Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals.
+In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g.
+@gelm06, @hox17 and @jong21).\\
+ICC & The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
+of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.\\
+\addlinespace
+Stratified intercept & \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+In \proglang{R}, multilevel models may be fitted using package such as
+\pkg{lme4} or \pkg{nlme}. For linear mixed-effects models, the function
 
@@
 \hypertarget{concepts-in-missing-data}{%
-\subsection{concepts in missing data}\label{concepts-in-missing-data}}
+\subsection{Concepts in missing data}\label{concepts-in-missing-data}}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-3-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-2-1.pdf}
 
@@
 
+Table 2: Concepts in missing data methods
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Concept}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Details}
+\end{minipage} \\
+\midrule()
+\endhead
+MCAR & Missing Completely At Random, where the probability to be missing
+is equal \\
+& across all data entries \\
+MAR & Missing At Random, where the probability to be missing depends on
+observed \\
+& information \\
+MNAR & Missing Not At Random (MNAR), where the probability to be
+missing \\
+& depends on unrecorded information, making the missingness
+non-ignorable \\
+& \citep{rubi76, meng94}. \\
+& \\
+\bottomrule()
+\end{longtable}
+
 \hypertarget{imputation-with-mice}{%
-\subsection{imputation with mice}\label{imputation-with-mice}}
+\subsection{Imputation with mice}\label{imputation-with-mice}}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-color-frame, opacityback=0, colback=white, rightrule=.15mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, leftrule=.75mm, toprule=.15mm, colframe=quarto-callout-color-frame, rightrule=.15mm, bottomrule=.15mm, colback=white, breakable, opacityback=0, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-color-frame, opacityback=0, colback=white, rightrule=.15mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, leftrule=.75mm, toprule=.15mm, colframe=quarto-callout-color-frame, rightrule=.15mm, bottomrule=.15mm, colback=white, breakable, opacityback=0, left=2mm]
+
+TODO: @ Hanne i added a wide explanation about hte prediction matrix in
+the Model Specification subsection, so you can check it and consider to
+modify your box.. because i dont think is usefull to repeat information
+across the paper\ldots{}
 
@@
 performing a simultaneous imputation process by using an imputation
-method that accounts for clustering.(Stata:
-https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/)
-TODO: replace ref.
+method that accounts for clustering.\cite{zotero-516}
 
@@
 Missing at Random (MAR) assumption and with a correct
-variance-covariance specification.\cite{Molenberghs_2007}.
+variance-covariance specification.\cite{molenberghs2007}.
 
@@
 recommended to include the outcome variable in the imputation model for
-prediction variables \cite{moons2006a}. In cases where the outcome is
+prediction variables \cite{moons2006}. In cases where the outcome is
 time-to-event, the Nelson-Aalen estimate of the time to the event should
@@
 meeting the Missing at Random (MAR) assumption and enhancing estimation
-efficiency \cite{hardt2012a}.
+efficiency \cite{hardt2012}.
 
@@
 Recipes have been proposed for imputing incomplete level-1 and level-2
-variables for hierarchical models \cite{buuren2018a} (\(\S\) 7.10),
-which can be convenient to follow when dealing with numerous variables
+variables for hierarchical models \cite{buur18} (\(\S\) 7.10), which can
+be convenient to follow when dealing with numerous variables
 (interaction terms at different levels) and models with many random
@@
 with the prediction matrix method do not requires the inclusion of
-additional terms as Just Another Variable (JAV)
-\cite{buuren2018a}(\(\S\) 6.4).
+additional terms as Just Another Variable (JAV) \cite{buur18}(\(\S\)
+6.4).
 
@@
 on the Heckman method can be
-considered.\cite{hammon2020,hammon2022,munoz2023}
+considered.\cite{hammon2020,hammon2022,munoz2023b}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-8-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-9-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-11-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-17-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-16-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-19-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-18-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-20-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-19-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-21-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-20-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-22-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-21-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-28-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-27-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-29-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-28-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-38-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-37-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-40-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-39-1.pdf}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-41-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-40-1.pdf}
 
@@
 \endhead
-(Intercept) & -1.892 & 0.333 & -5.682 & 11445.455 & 0.000 & 0.019 &
-0.019 \\
-type & -0.346 & 0.178 & -1.948 & 686356.768 & 0.051 & 0.002 & 0.002 \\
-age & 0.032 & 0.002 & 19.564 & 2418.910 & 0.000 & 0.042 & 0.041 \\
-motor\_score2 & -0.575 & 0.070 & -8.226 & 71295.018 & 0.000 & 0.008 &
-0.008 \\
-motor\_score3 & -0.891 & 0.072 & -12.440 & 66712.189 & 0.000 & 0.008 &
-0.008 \\
-motor\_score4 & -1.276 & 0.074 & -17.305 & 18426.087 & 0.000 & 0.015 &
-0.015 \\
-pupil2 & 1.315 & 0.068 & 19.339 & 110.826 & 0.000 & 0.235 & 0.204 \\
-pupil3 & 0.657 & 0.075 & 8.745 & 423.463 & 0.000 & 0.108 & 0.101 \\
-ct2 & 0.747 & 0.084 & 8.884 & 41.781 & 0.000 & 0.448 & 0.340 \\
-ct3 & 0.766 & 0.090 & 8.535 & 12.103 & 0.000 & 1.352 & 0.631 \\
+(Intercept) & -1.910 & 0.335 & -5.700 & 9617.768 & 0.000 & 0.021 &
+0.021 \\
+type & -0.347 & 0.179 & -1.941 & 256541.695 & 0.052 & 0.004 & 0.004 \\
+age & 0.032 & 0.002 & 19.550 & 1965.628 & 0.000 & 0.047 & 0.046 \\
+motor\_score2 & -0.576 & 0.070 & -8.195 & 14898.703 & 0.000 & 0.017 &
+0.017 \\
+motor\_score3 & -0.892 & 0.072 & -12.390 & 14760.294 & 0.000 & 0.017 &
+0.017 \\
+motor\_score4 & -1.272 & 0.074 & -17.203 & 10235.157 & 0.000 & 0.020 &
+0.020 \\
+pupil2 & 1.311 & 0.074 & 17.758 & 41.455 & 0.000 & 0.451 & 0.342 \\
+pupil3 & 0.661 & 0.077 & 8.533 & 165.867 & 0.000 & 0.184 & 0.165 \\
+ct2 & 0.783 & 0.075 & 10.461 & 264.558 & 0.000 & 0.140 & 0.130 \\
+ct3 & 0.786 & 0.095 & 8.280 & 10.443 & 0.000 & 1.624 & 0.676 \\
 \bottomrule()
@@
 Intercept\textasciitilde\textasciitilde Intercept\textbar name &
-0.082 \\
+0.083 \\
 \bottomrule()
@@
 \hypertarget{obesity-data}{%
-\subsection{obesity data}\label{obesity-data}}
+\subsection{Obesity data}\label{obesity-data}}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-48-1.pdf}
+{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-47-1.pdf}
 
@@
 \begin{verbatim}
-R> Nulmodel <- lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
-R> # performance::icc(Nulmodel)
+R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
+R> performance::icc(Nulmodel)
+\end{verbatim}
+
+\begin{verbatim}
+# Intraclass Correlation Coefficient
+
+    Adjusted ICC: 0.362
+  Unadjusted ICC: 0.362
 \end{verbatim}
@@
 `2l.2stage.norm' for the age variable. However, after inspecting the age
-density plot, we consider modifying its method to `2l.2stage.pmm'. For
-the BMI variable, we employ deterministic imputation.
+and weight density plot, we consider modifying its method to
+`2l.2stage.pmm'. For the BMI variable, we employ deterministic
+imputation.
 
@@
 meth_mar["Age"] <- "2l.2stage.pmm"
+meth_mar["Weight"] <- "2l.2stage.pmm"
 \end{verbatim}
@@
 pred_mar[pred_mar == 1] <- 2
+
 pred_mar[c("Height", "Weight"), "BMI"] <- 0
-plot_pred(pred_mar)
-\end{verbatim}
-
-\begin{figure}[h]
-
-{\centering \includegraphics[width=0.7\textwidth,height=\textheight]{manuscript_files/figure-pdf/obesity-predmar-1.pdf}
-
-}
-
-\end{figure}
-
-\begin{verbatim}
+#plot_pred(pred_mar)
 imp_mar <-
@@
 \begin{verbatim}
-summary(complete(imp_mar,"long")$Weight)
+summary(complete(imp_mar, "long")$Weight)
 \end{verbatim}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  4.428  69.890  82.800  82.835  95.491 145.885 
-\end{verbatim}
-
-We are also contemplating the utilisation of the predictive mean
-matching (pmm) option, as the values imputed using a fully parametric
-method may be implausibly low for some patients.
-
-\begin{verbatim}
-meth_mar["Weight"] <- "2l.2stage.pmm"
-imp_mar_pmm <- mice(
-  data = Obesity,
-  method = meth_mar,
-  predictorMatrix = pred_mar,
-  printFlag = FALSE
-)
-\end{verbatim}
-
-\begin{verbatim}
-summary(complete(imp_mar_pmm, "long")$Weight)
-\end{verbatim}
-
-\begin{verbatim}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  28.35   71.04   83.65   83.87   95.66  134.61 
-\end{verbatim}
-
-\begin{verbatim}
-plot_trace(imp_mar_pmm, "Weight")
+  28.35   69.38   81.96   82.36   94.46  134.61 
+\end{verbatim}
+
+\begin{verbatim}
+plot_trace(imp_mar, "Weight")
 \end{verbatim}
@@
 \begin{verbatim}
-ggmice(imp_mar_pmm, aes(.imp, Weight)) + 
+ggmice(imp_mar, aes(.imp, Weight)) + 
   geom_jitter()
@@
 \end{figure}
-
-Analyze the imputed data.
-
-\begin{verbatim}
-R> # fit <- with(
-R> #     imp_mar_pmm,
-R> #     lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-R> #     )
-R> fit <- with(
-+    imp_mar_pmm,
-+    lmer(BMI ~ Age + FamOb + Gender + (1 + Age | Cluster))
-+    )
-R> est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
-\end{verbatim}
-
-Display results in table.
-
-\begin{verbatim}
-R> est$estimates |> 
-+  round(3) |>
-+  kable()
-\end{verbatim}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1558}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1169}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1299}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2338}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}@{}}
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Std.Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-t.value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-P(\textgreater\textbar t\textbar)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-RIV
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-FMI
-\end{minipage} \\
-\midrule()
-\endhead
-(Intercept) & 28.445 & 2.259 & 12.592 & 157.553 & 0.000 & 0.190 &
-0.170 \\
-Age & 0.041 & 0.032 & 1.311 & 388.953 & 0.191 & 0.113 & 0.106 \\
-FamObyes & 0.000 & 0.316 & -0.001 & 207.648 & 0.999 & 0.161 & 0.147 \\
-GenderMale & -1.361 & 0.411 & -3.310 & 7.986 & 0.011 & 2.421 & 0.761 \\
-\bottomrule()
-\end{longtable}
-
-\begin{verbatim}
-R> est$extra.pars |> 
-+  round(3) |>
-+  kable()
-\end{verbatim}
-
-\begin{longtable}[]{@{}lr@{}}
-\toprule()
-& Estimate \\
-\midrule()
-\endhead
-Intercept\textasciitilde\textasciitilde Intercept\textbar Cluster &
-19.805 \\
-Age\textasciitilde\textasciitilde Age\textbar Cluster & 0.003 \\
-Intercept\textasciitilde\textasciitilde Age\textbar Cluster & -0.069 \\
-Residual\textasciitilde\textasciitilde Residual & 26.010 \\
-ICC\textbar Cluster & 0.394 \\
-\bottomrule()
-\end{longtable}
 
@@
 a more comprehensive understanding of the model and the exclusion
-restriction, see \cite{munoz2023a}.
+restriction, see \cite{munoz2023}.
 
@@
 
-Then we proceed to run the imputation model as before, after executing
-these imputation procedures, it is essential to assess convergence and
-the coherence of the imputed values.
-
-\begin{verbatim}
-imp_mnar <- mice(
-  data = Obesity,
-  method = meth_mnar,
-  predictorMatrix = pred_mnar,
-  printFlag = FALSE
-)
-summary(complete(imp_mnar, "long")$Weight)
-\end{verbatim}
-
-\begin{verbatim}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  16.31   72.14   86.97   87.68  102.43  190.44 
-\end{verbatim}
-
-Upon examining the weight variable, we noticed that the imputed range
-falls outside the realm of plausible values (as weight should be
-positive).
-
-\begin{verbatim}
-summary(complete(imp_mnar, "long")$Weight)
-\end{verbatim}
-
-\begin{verbatim}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  16.31   72.14   86.97   87.68  102.43  190.44 
-\end{verbatim}
-
-Consequently, as before we use the `pmm', option but this time for the
-Heckman imputation, this approach ensures that the imputed values remain
-within the range of observable values. We then run the imputation model
-but this time using the option of pmm, to assure that weight values are
-in the range of the observable data, this can be implemented by setting
-the pmm parameter to true.
-
-\begin{verbatim}
-imp_mnar_pmm <-
+As before we use the `pmm', option but this time for the Heckman
+imputation, this approach ensures that the imputed values remain within
+the range of observable values. We then run the MNAR imputation model by
+setting the pmm parameter to true.
+
+\begin{verbatim}
+imp_mnar <-
   mice(
@@
 \begin{verbatim}
-summary(complete(imp_mnar_pmm, "long")$Weight)
+summary(complete(imp_mnar, "long")$Weight)
 \end{verbatim}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-  28.35   71.87   85.58   86.51  100.15  134.61 
-\end{verbatim}
-
-\begin{verbatim}
-plot_trace(imp_mar_pmm, "Weight")
+  28.35   67.39   82.57   81.71   96.04  134.61 
+\end{verbatim}
+
+\begin{verbatim}
+plot_trace(imp_mar, "Weight")
 \end{verbatim}
@@
 
-After this modification we proceed to compare the effects on the model.
-We run the analysis model on each of the completed datasets as well as
-the dataset where the incomplete values are removed (Complete Case
-analysis, CC).
-
-\begin{verbatim}
-# cc_rs <- with(setDT(Obesity)[complete.cases(Obesity), ],
-#               lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age |
-#                     Cluster))
-mar_rs <- with(
-  imp_mar, 
-  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-  )
-mar_pmm_rs <- with(
-    imp_mar_pmm,
-    lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-    )
-mnar_rs <- with(
-  imp_mnar, 
-  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-  )
-mnar_pmm_rs <- with(
-  imp_mnar_pmm,
-  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-  )
-# list_models <- list(cc_rs, mar_rs, mar_pmm_rs, mnar_rs, mnar_pmm_rs)
-# plot_models(list_models,
-#             mod_name = c("Complete case", "MAR", "MAR_pmm", "MNAR", "MNAR_pmm"))
-\end{verbatim}
-
-\begin{verbatim}
-est <- testEstimates(as.mitml.result(mnar_pmm_rs), extra.pars = TRUE)
-\end{verbatim}
-
-Display results in table.
-
-\begin{verbatim}
-R> est$estimates |> 
-+  round(3) |>
-+  kable()
-\end{verbatim}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1579}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1184}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1316}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1053}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0921}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2368}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0789}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0789}}@{}}
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Std.Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-t.value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-P(\textgreater\textbar t\textbar)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-RIV
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-FMI
-\end{minipage} \\
-\midrule()
-\endhead
-(Intercept) & 29.199 & 2.766 & 10.555 & 55.483 & 0.000 & 0.367 &
-0.294 \\
-Age & 0.050 & 0.038 & 1.301 & 18.731 & 0.209 & 0.859 & 0.512 \\
-FamObyes & 0.016 & 0.445 & 0.036 & 15.910 & 0.972 & 1.006 & 0.554 \\
-GenderMale & -1.408 & 0.696 & -2.024 & 5.128 & 0.097 & 7.563 & 0.912 \\
-\bottomrule()
-\end{longtable}
-
-\begin{verbatim}
-R> est$extra.pars |> 
-+  round(3) |>
-+  kable()
-\end{verbatim}
-
-\begin{longtable}[]{@{}lr@{}}
-\toprule()
-& Estimate \\
-\midrule()
-\endhead
-Intercept\textasciitilde\textasciitilde Intercept\textbar Cluster &
-26.114 \\
-Age\textasciitilde\textasciitilde Age\textbar Cluster & 0.002 \\
-Intercept\textasciitilde\textasciitilde Age\textbar Cluster & -0.088 \\
-Residual\textasciitilde\textasciitilde Residual & 29.741 \\
-ICC\textbar Cluster & 0.431 \\
-\bottomrule()
-\end{longtable}
+We proceed to compare the effects on the model. We run the analysis
+model on each of the completed datasets as well as the dataset where the
+incomplete values are removed (Complete Case analysis, CC).
+
+We proceed to fit the mixed model by using the complete case database
+and the imputed datasets (MAR,MNAR), and we compare the estimates of
+each dataset by using a plot (the plot function can be found in the
+appendix).
+
+\begin{verbatim}
+# Save all datasets in a list
+list_data <- list(Obesity |> filter(complete.cases(Obesity)), # complete case
+                  imp_mar, # imputed under mar
+                  imp_mnar) # imputed under mnar
+
+# Run mixed models on each dataset
+list_models <- lapply(list_data, 
+                   FUN = function(x) 
+                     with(x,nlme::lme(BMI ~ Age + FamOb + Gender, 
+                                      random =  ~ 1 + Age | Cluster)))
+# Plot coefficients
+plot_models(list_models,
+             mod_name = c("Complete case", "MAR", "MNAR"))
+\end{verbatim}
+
+\begin{figure}[h]
+
+{\centering \includegraphics{manuscript_files/figure-pdf/models-1.pdf}
+
+}
+
+\end{figure}
 

### 2024-01-10 renv/library/R-4.3/aarch64-apple-darwin20/semEff/doc/predicting-effects.Rmd
Updated manuscript
I have removed the 'no pmm' part from the Heckman model. Additionally, I've included the plot function with echo=FALSE, making it suitable for copying and pasting into the appendix later on. However, the pattern function with cluster still does not work. Regarding the box 3 about the prediction matrix, previously i have added relevant information about this (with an equation) in the Imputation flow. Please review this information to ensure that the box is constructed appropriately, avoiding redundant information throughout the paper. Thanks.

---
title: "Predicting Effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`semEff` can be used to predict values for different types of effects in an SEM. For example, we may be interested in the predictions of an indirect effect of a variable vs. its direct effect, or we may simply wish to predict the total (combined) effect. Here we'll use [simulated data](https://murphymv.github.io/semEff/reference/shipley.html) from @shipley2009 on tree growth and survival for a single tree species to predict and plot a direct vs. indirect effect.

```{r message = FALSE}
# install.packages(c("semEff", "ggplot2"))
library(semEff)
library(ggplot2)
```

Let's preview the data (n = 1900):

```{r}
knitr::kable(head(shipley))
```

Tree measurements are taken for five individuals of the same species at 20 different sites every two years from 1970 to 2006 (repeated measures). The hypothesised SEM from @shipley2009 tests the indirect effect of site latitude on tree survival (0/1), via degree days to bud burst, date of first bud burst, and tree growth. This can be fit as a series of linear and generalised linear mixed models using the `lme4` package:

```{r eval = FALSE}
shipley.sem <- list(
  DD = lme4::lmer(DD ~ lat + (1 | site) + (1 | tree), data = shipley),
  Date = lme4::lmer(Date ~ DD + (1 | site) + (1 | tree), data = shipley),
  Growth = lme4::lmer(Growth ~ Date + (1 | site) + (1 | tree), data = shipley),
  Live = lme4::glmer(Live ~ Growth + (1 | site) + (1 | tree), data = shipley, 
                     family = binomial)
)
```

We can bootstrap estimates for these models:

```{r eval = FALSE}
shipley.sem.boot <- bootEff(shipley.sem, R = 1000, seed = 13, ran.eff = "site")
```

And use the bootstrap samples to calculate effects and confidence intervals:

```{r}
(shipley.sem.eff <- semEff(shipley.sem.boot))
```

For this example, let's compare the direct effects of *Date* (Julian date of first bud burst) vs. the indirect effects of *DD* (cumulative degree days to first bud burst, °C) on tree *Growth* (increase in stem diameter, mm):

```{r}
summary(shipley.sem.eff, "Growth")
```

We'll extract the (total) effects of each variable to use for prediction:

```{r}
tot <- getTotEff(shipley.sem.eff, "Growth")
tot.b <- getTotEff(shipley.sem.eff, "Growth", type = "boot")
```

Now we can predict values for *Growth* using [`predEff()`](https://murphymv.github.io/semEff/reference/predEff.html). We'll need the model object, and we'll generate 100 values of the predictors to predict from:

```{r}
dat <- na.omit(shipley)
mod <- shipley.sem$Growth
fit <- sapply(c("Date", "DD"), function(i) {
  x <- data.frame(seq(min(dat[i]), max(dat[i]), length = 100)); names(x) <- i
  f <- predEff(mod, newdata = x, effects = tot[i], eff.boot = tot.b)
  c(x, f)
}, simplify = FALSE)
```

OK let's plot the predictions. We'll use a custom plotting function which wraps `ggplot2::ggplot()`:

```{r}
plotFit <- function(x, y, fit, x.lab = NULL, y.lab = NULL) {
  d <- fit[[1]]
  f <- fit$fit
  ci.l <- fit$ci.lower
  ci.u <- fit$ci.upper
  ggplot () + 
    geom_point(aes(x, y)) +
    geom_ribbon(aes(d, ymin = ci.l, ymax = ci.u, alpha = "0.15"), fill = "blue") +
    geom_line(aes(d, f), color = "blue", size = 1) +
    xlab(x.lab) + 
    ylab(y.lab) +
    theme_bw() + 
    theme(legend.position = "none")
}
```

Effects of *Date* (direct):

```{r warning = FALSE}
plotFit(x = dat$Date, y = dat$Growth, fit = fit$Date, 
        x.lab = "Julian Date of Bud Burst (direct effect)", y.lab = "Stem Growth (mm)")
```

Effects of *DD* (indirect, via *Date*):

```{r warning = FALSE}
plotFit(x = dat$DD, y = dat$Growth, fit = fit$DD, 
        x.lab = "Degree Days to Bud Burst (indirect effect)", y.lab = "Stem Growth (mm)")
```

We can see that the effects are in opposite directions: positive for *Date* (direct) and negative for *DD* (indirect). This is because trees at higher latitudes require fewer degree days to break buds (cooler climate), break buds later, and grow more in the growing season than trees at lower latitudes. Thus *DD* negatively affects *Growth* by reducing the *Date* to first bud burst. This is just a toy example from the SEM to illustrate contrasting effects of similar magnitude. *DD* in this case is acting as a proxy for latitude (high *DD* = low latitude).

It's also evident from the plots that there is a huge amount of scatter around the fits. This is due to the random effects ('site' and 'tree') accounting for most variation in tree growth. We can get a rough breakdown of this using R-squared, here via [`R2()`](https://murphymv.github.io/semEff/reference/R2.html) (squared multiple correlation):

```{r}
c(r2_marg = R2(mod, re.form = NA)[[1]],
  r2_cond = R2(mod)[[1]])

```

Fixed ('marginal') effects account for only about 5% of variation, compared to almost 80% when adding all random effects ('conditional'). We can also condition on individual random effects, showing that 'tree' accounts for most random variation:

```{r}
c(r2_site = R2(mod, re.form = ~ 1 | site)[[1]],
  r2_tree = R2(mod, re.form = ~ 1 | tree)[[1]])
```

#### References

### 2024-01-10 Manuscript/manuscript.qmd
Update manuscript.qmd
Added link, changed affiliations and acknowledgements

--- 2023-12-22 Manuscript/manuscript.qmd
@@
       - name: University Medical Center Utrecht
-        department: Julius Centre for Health Sciences and Primary Care
+        department: Julius Center for Health Sciences and Primary Care
         address: Universiteitsweg 100
@@
         country: The Netherlands
-        postal-code: 3584 CG
+        postal-code: 3584 CX
       - Journal of Statistical Software
@@
       - name: University Medical Center Utrecht
-        department: Julius Centre for Health Sciences and Primary Care
+        department: Julius Center for Health Sciences and Primary Care
         city: Utrecht
@@
 This tutorial is dedicated to readers who are unfamiliar with multiple imputation.
-More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2) or under MNAR conditions (case study 3).
+More experienced readers can skip the background section and introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2, @sec-impact) or under MNAR conditions (case study 3).
 
@@
 
-### IMPACT data
+### IMPACT data {#sec-impact}
 
@@
 
-## Funding
-
-This project has received funding from the European Union's Horizon 2020 research and innovation programme under ReCoDID grant agreement No 825746.
-
-The views expressed in this paper are the personal views of the authors and may not be understood or quoted as being made on behalf of or reflecting the position of the regulatory agency/agencies or organizations with which the authors are employed/affiliated.
-
 ## Summary and discussion {#sec-summary}
@@
 [R]{.proglang} itself and all packages used are available from the Comprehensive [R]{.proglang} Archive Network (CRAN) at [https://CRAN.R-project.org/].
+
+## Disclaimer
+
+The views expressed in this paper are the personal views of the authors and may not be understood or quoted as being made on behalf of or reflecting the position of the regulatory agency/agencies or organizations with which the authors are employed/affiliated.
 

### 2024-01-16 Manuscript/manuscript.qmd
Suggestions for background section

@@
 
-There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.
+There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of a cluster indicator variable, performing a separate imputation process for each cluster, or performing a imputation for each cluster simultaneously by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.
 
@@
 
+<!-- I don't understand the following paragraph. What is meant with "is restricted in datasets..."? -->
 Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
@@
 Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
-However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e.
-the number of clusters is adequate.
-(Austin,2018).
-
-Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008).
-Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020).
-For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
+However, this strategy is best suited for balanced data (Grund, 2017) and when the number of clusters is adequate sufficient for the random effects model (Austin,2018).
+
+Here it is important to evaluate the assumptions imposed by the main model, for instance the cluster indicator strategy bias the estimator when the model is based on a hierarchical model (Taaljard,2008).
+Although an imputation strategy congenial with the main model is preferred, it is important to consider whether a complex imputation model is appropriate for the data, as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios (Bailey 2020).
+For instance, in causal effect analysis, separately imputating data sets may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
 (Zhang,2023).
+<!-- What is an exposure cluster? -->
 
@@
 
-### Focus on the Main Analysis
-
-When dealing with incomplete clustered data, start by looking at your research questions and planned analysis.
-Pretend there are no missing data at first.
-This will help you understand your research hypotheses, the main statistical model, and give you insights into your data's structure (refer to the level table), variable types (e.g., confounders, auxiliary variables), and other considerations such as variable relationships, interactions or polynomial terms.
-
 ### Exploration of Available Data
 
-Now, explore what's in your data.
-Use simple tools like histograms and QQ plots to understand variable distributions, plausible range of values and identify errors or outliers.
-Check interactions between predictors with scatter plots, look for collinearity with VIF or correlations plots, this may also help you to identify non-considered relationships that may affect the main model.
-You might also consider to test assumptions related to your response variable such as variance homogeneity, independence between observations (eg.
-ACF, variagrams).
+First, explore the data, preferably using visual tools.
 This will help you choose an imputation model that suits your data.
-In addition, the intraclass correlation coefficient (ICC) can be examined to assess cluster differences, aiding in the choice between the 2l and 1l methods for imputation.
+In addition, the intraclass correlation coefficient (ICC) can be examined to assess cluster differences, aiding in the choice between methods for dependent (such as clustering) or independent data.
 
@@
 Look at the \textbf{proportion of missing values} in the dataset variables.
-This helps find potential predictors and cut down on unnecessary variables in the imputation model.
-Doing this can lower the risk of multicollinearity or computational issues, especially with certain parametric imputation methods.
+Variables for which most data is missing may contribute little in the analysis, and may be removed from the analysis model.
+Removing them also from the imputation model can lower the risk of multicollinearity or computational issues, especially with certain parametric imputation methods.
 You can also identify predictors for the imputation model by using inflow criteria to see connections between missing data in one variable and observed variables, and outflow criteria to identify connections between observed values in one variable and missing data in others.
 
-Check the missing patterns at cluster level, this can help you to select the most appropiated imputation approach in terms of computational efficiency (e.g., simpler regression imputation versus FCS in univariate patterns).
+<!-- What are inflow and outflow criteria? -->
+
+Check the missing patterns at cluster level, this can help you to select the most appropriate imputation approach in terms of computational efficiency (e.g., simpler regression imputation versus FCS in univariate patterns).
 
@@
 
+<!-- What does it mean if an estimation model is not able to handle missing data? What do you mean with estimation model? -->
+<!-- AFAIK, when only the outcome is missing, MI and a ML based method lead to similar results. The text appears to suggest that in that case ML based methods are better. -->
+
 ### Pre-imputation
 Clean up your dataset before the imputation process.
-Keep initially the essential variables for your main model, but include extra variables if needed for specific procedures during analysis (e.g.
-confounders on balancing procedures).
-Think about adding other useful variables, even if they're not in the main model, such as instrumental or auxiliary variables which might boost your parameter estimates, especially if they're associated to the probability of missingness for some incomplete variables.
-
-Figure out if you can directly impute incomplete variables by just use deductive imputation.
-This involves inferring missing values based on logicallogical connections between variables.
-It's especially handy for variables that depend on each other, like calculating BMI from weight and height.
+For imputation, include the variables necessary for your analysis, depending on your analysis this may be the exposure or intervention, outcome, covariates, confounders, stratification factors, and the cluster variable.
+Think about adding other useful variables, even if they're not in the main model, such as instrumental or auxiliary variables which might improve the accuracy of the imputation model, especially if they're associated to the probability of missingness for some incomplete variables.
+
+Figure out if you can directly impute incomplete variables by using deductive imputation.
+This involves inferring missing values based on logical connections between variables.
+It's especially useful for variables that depend on each other, like calculating BMI from weight and height.
 
 Deductive imputation is also useful for getting values for level-1 variables from level-2 ones, like in Individual Participant Data (IPD).
-In these situations, you can guess missing information from metadata or by using or through deduction based on time or protocols.
+In some situations, you can derive missing information from metadata or protocols.
 For example, you could deduce missing test values for patients who have passed away.
+<!-- I don't think this is the best example.
+Usually, test scores after death are not of interest.
+The test score before death may be missing for a reason, but this should be derived based on clinical knowledge.
+The example is too vague (and I can't readily improve it.) -->
 

### 2024-01-16 Manuscript/manuscript.qmd
edits background

@@
 <!-- I don't understand the following paragraph. What is meant with "is restricted in datasets..."? -->
+<!-- Also, this info is also mentioned in ##### Clustering Inclusion -->
 Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
@@
 
+<!-- most of this subsection is also mentioned at the start of #sec-workflow -->
+
 Which strategy you choose depends on the assumptions in your main analysis and the limitations of your data.
@@
 
-In a hierarchical imputation model, random effects model the correlations between observations within clusters, making it possible to estimate even with a small number of observations per cluster.
+In a hierarchical imputation model, random intercepts and random effects model the correlations between observations within clusters, making it possible to estimate even with a small number of observations per cluster.
 There are various proposed multiple imputation models based on hierarchical models, each with its own set of assumptions \cite{audigier}.

### 2024-01-17 Manuscript/manuscript.qmd
more suggestions for the background

@@
 Start by choosing the imputation model for each incomplete variable in your dataset.
-The mice package suggests methods based on variable types for non-clustered variables, and for the cluster ones, you can use the micemd package's find.defaultMethod() function.
+The mice package suggests methods based on variable types for non-clustered variables, and for the clustered ones you can use the micemd package's find.defaultMethod() function.
 This function selects from different 2l imputation methods based on cluster size and the proportion of missing data in each cluster.
@@
 For instance, it's recommended to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}.
-In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be added as a covariate in the imputation model [REF].
+In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be added as a covariate in the imputation model, as well as the event indicator [REF: White IR, Royston P.
+Imputing missing covariate values for the Cox model.
+Stat Med.
+2009 Jul 10;28(15):1982-98.
+Available from: doi.org/10.1002/sim.3618].
 Additionally, including auxiliary variables, even if not part of the main model, can be linked to the probability of missingness, improving the likelihood of meeting the Missing at Random (MAR) assumption and enhancing estimation efficiency \cite{hardt2012a}.
@@
 
-* 3 and 4 type have been advised to used as the inclusion of the means of the cluster is beneficial on FCS \cite{mistler2017}
+<!-- Can we reorder this table? It starts from -2, then -1 and 0 are missing, 1 to 4 and then again 2 negative numbers. What about the following?  -->
+
+\begin{tabular}{ll}
+     \toprule
+Type & Definition \\
+  \midrule
+ 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
+ 2  & Random variable, e.g., $x_{3_{ij}}$\\
+ 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
+ 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
+  \midrule
+-2  & Cluster variable, in this case the one defined by j index  \\
+-3. & Random variable only included on selection model (Heckman model)\\
+-4. & Random variable only included on main model (Heckman model)\\
+      \bottomrule
+\end{tabular}
+
+* It has been advised to use variables of type 3 and 4, as the inclusion of the means of the cluster is beneficial on FCS \cite{mistler2017}.
 
@@
 These rules were designed to ensure compatibility among the conditionally specified imputation models and congeniality between the imputation and the main model.
-However, they may not be applicable to all 2l imputation methods; for instance, the 2l.2stage imputation methods of the micemd package only allow the inclusion of random predictor variables (2).
+However, they may not be applicable to all 2l imputation methods.
+For instance, the 2l.2stage imputation methods of the micemd package only allow the inclusion of random predictor variables (2).
 
@@
 
-For some interaction terms, it has been suggested, for instance, for treatment interaction effects, to conduct separate imputation by treatment group \cite{zhang2023}.
-Additionally, it can be used imputation models based on random forest or deep learning that can handle interaction and non-linear terms without requiring the explicit specification of an imputation model.
+For some interaction terms, including treatment interaction effects, it has been suggested to conduct separate imputation by treatment group \cite{zhang2023}.
+Additionally, one might apply random forest or deep learning methods, as these can handle interaction and non-linear terms without requiring the explicit specification of an imputation model.
 
@@
 During the imputation process, certain issues may arise that halt the process.
-In hierarchical model imputations, many issues are related to over fitting the imputation model.
+In hierarchical model imputations, many issues are related to overfitting the imputation model.
 To troubleshoot, it is recommended to inspect the imputation log file for variables causing problems.
-One approach to address this is to reduce the number of predictors, by using step by step the previous referred recepies or by using functions like quickpred.
-Another option is to consider variable transformations, such as scaling when model is invariant to linear transformations e.g., random intercept models.
-Also adjusting the level of the hierarchical model (e.g., using a homogeneous variance imputation method or a 1l model) can also be beneficial.
+One approach to address this is to reduce the number of predictors, by applying the previous referred recepies step by step  or by using functions like quickpred.
+Another option is to consider variable transformations, such as scaling when the model is invariant to linear transformations e.g., random intercept models.
+Also, adjusting the level of the hierarchical model (e.g., using a homogeneous variance imputation method or a 1l model) can also be beneficial.
 
@@
 
-Before starting into the analysis of each imputed dataset, it is crucial to validate the convergence of the imputation process.
+Before starting the analysis of each imputed dataset, it is crucial to validate the convergence of the imputation process.
 This is commonly accomplished through trace plots that depict the mean and variance of the incomplete variables across iterations.

### 2024-01-17 Manuscript/manuscript.qmd
Suggestions MNAR sections

@@
 
+<!-- What do you mean with "We simulate data for 5 clusters"? If this dataset is included in the package why do we need to simulate the data? -->
+
 -   **Cluster:** Region of the patients' healthcare provider (Cluster variable),
@@
 
+<!-- The sentence below ends after "In order to evaluate..."  -->
+
 We observe that the missing pattern is non-monotonic and quite similar across the clusters.
-However, regarding the weight variable, we notice that is systematically missing in cluster 3.
+However, regarding the weight variable, we notice that it is systematically missing in cluster 3.
 In order to evaluate
@@
 
-Since the ICC is above 0.1 and as the main analysis will be use a mixed model, we decide
+As the ICC is greater than 0.1 and as we will use a mixed model for the analysis, we decide
 to use two-level (2l) imputation methods. In this imputation process, we include all predictor
@@
 We use the **find.defaultMethodfunction** provided in the **micemd** package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters.
+
+<!-- What is the FAV variable? -->
 
@@
 ```
+
+<!-- What is suprime (below)? Do you mean suppress? -->
 
@@
 ```
-We are also contemplating the utilisation of the predictive mean matching (pmm) option, as the values imputed using a fully parametric method may be implausibly low for some patients.
+We also contemplate using the predictive mean matching (pmm) option, as the values imputed using a fully parametric method may be implausibly low for some patients.
 
@@
 We consider the possibility that patients may not have been selected randomly, which would then have led to a distribution for weight that does not reflect the weight in the population.
-It???s likely that an omitted variable, like self-esteem, could influence this selection.
+It's likely that an omitted variable, like self-esteem, could influence this selection.
 For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.
 
-To address this situation, two approaches have been proposed for dealing with Missing Not at Random (MNAR) data: pattern-mixed models and selection models.
-Within pattern-mixed models, methods like the delta method and more advanced ones like NARFS have been suggested.
-The selection model approach includes methods such as the Heckman model, which can be particularly useful in this case.
-Several methods, including those by \cite{galimard2018}, and the recently a Heckman method designed for two-level data, allow for variations in intercepts and exposure effects (random intercept and slope) \cite{munoz2023}.
+To address this situation where data are Missing Not At Random (MNAR), one approach is to apply the Heckman selection model.
+This method has recently been extended to allow for variations in intercepts and exposure effects (random intercept and slope) \cite{galimard2018, munoz2023}.
+This method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.
 
 To apply the **2l.2stage.heckman** method, the weight variable should be specified as '2l.2stage.heckman' found in the micemd package.
-Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.
+Additionally, the prediction matrix needs modification, as this model includes one submodel for the individuals' measured outcome, and one for the selection of individuals.
 For the outcome equation we consider the same imputation model that we used for
@@
 These factors may lead the user to skip specific questions or even the entire survey.
-Also, we assume the time does not have any influence on the subject???s weight.
+Also, we assume the time does not have any influence on the subject's weight.
  $$R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}$$
@@
 
-Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive).
+Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values, as weight should be positive.
 
@@
 ```
-Consequently, as before we  use the 'pmm', option but this time for the Heckman imputation, this approach ensures that the imputed values remain within the range of observable values.
-We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by setting the pmm parameter to true.
+Consequently, as before, we  use the 'pmm' option, but this time for the Heckman imputation.
+The pmm approach ensures that the imputed values remain within the range of observable values.
+This can be implemented by setting the pmm parameter to true.
 
@@
 An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
-Nonetheless, with respect to precision of estimates, we notice that in general MNAR imputation leads to wider confidence intervals, in this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on significant test and therefore lead to contradictory conclusions.
+Nonetheless, with respect to precision of the estimates, we notice that in general assuming MNAR leads to wider confidence intervals.
+In this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on the significance of a statistical test and therefore lead to contradictory conclusions.
+
+<!-- Contradictory? Do you mean if the results from an imputation assuming MAR and one assuming MNAR disagree? -->
 

### 2024-01-22 Manuscript/manuscript.qmd
comments conclusion
Some suggestions for the abstract, as there was some repetition. I removed the repetitions, and reordered a bit. Also added the three case studies.

@@
 abstract: |
-  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}.
-Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data.
-More complex models can be accomodated, but are outside the scope of this paper.
-Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy.
-In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed.
-This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R.
-We aim to show how imputation can yield less biased estimates from incomplete clustered data.
-We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms.
-For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
+  Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy, as it may introduce bias.
+We focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.
+Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed.
+Our scope is multilevel models for incomplete clustered data, to show how imputation can yield less biased estimates.
+This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} package [mice]{.pkg}.
+
+ We provide a background of missing data and multiple imputation, practical guidelines for different missing data situations, including non-ignorable missingness mechanisms.
+We apply the methods, and provide descriptions and code snippets for three case studies: data on 2000 pupils' popularity in 100 schools,  data on 11,002 patients across 15 studies on traumatic brain injuries with data Missing At Random, and 2,111 patients from 5 regions with data Missing Not At Random.
 keywords: [missing data, multilevel, clustering, mice, R]
@@
 
+<!-- What is the difference between Mice and mice? is there a Mice package? -->
+
 ## Summary and discussion {#sec-summary}

### 2024-01-23 Manuscript/manuscript.qmd

--- 2024-01-17 Manuscript/manuscript.qmd
@@
 
-There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of a cluster indicator variable, performing a separate imputation process for each cluster, or performing a imputation for each cluster simultaneously by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.
+There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.\cite{zotero-516}
 
@@
 
-<!-- I don't understand the following paragraph. What is meant with "is restricted in datasets..."? -->
+<!-- I don't understand the following paragraph.
+What is meant with "is restricted in datasets..."?
+I mean about the characteristics of the data such as sample size per cluster and balance of the data-->
 <!-- Also, this info is also mentioned in ##### Clustering Inclusion -->
-Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
+Regarding the restrictions imposed by the nature of the data, i.e., the sample size, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009).
 The last restriction is also required when imputations are performed on each cluster separately.
@@
 Although an imputation strategy congenial with the main model is preferred, it is important to consider whether a complex imputation model is appropriate for the data, as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios (Bailey 2020).
-For instance, in causal effect analysis, separately imputating data sets may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions.
-(Zhang,2023).
-<!-- What is an exposure cluster? -->
-
 Below we provide a imputation workflow that can be used in general to impute cluster data.
@@
 Sometimes, simpler methods like complete case analysis might be suffice, especially if your missingness is low (usually <5%).
-There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods, for instance when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}.
-
-<!-- What does it mean if an estimation model is not able to handle missing data? What do you mean with estimation model? -->
-<!-- AFAIK, when only the outcome is missing, MI and a ML based method lead to similar results. The text appears to suggest that in that case ML based methods are better. -->
+There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods, for instance when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{molenberghs2007}.
+
+<!-- What does it mean if an estimation model is not able to handle missing data?
+there are models that implicitaly are able to manage the missingness like the FML where in the likelihood it is used the information of the completed observations and the incomplete ones.
+What do you mean with estimation model?
+the main model you use to analyse the data-->
+<!-- AFAIK, when only the outcome is missing, MI and a ML based method lead to similar results.
+The text appears to suggest that in that case ML based methods are better.
+what i understood is ML are better in that case but please check it.-->
 
@@
 Deductive imputation is also useful for getting values for level-1 variables from level-2 ones, like in Individual Participant Data (IPD).
-In some situations, you can derive missing information from metadata or protocols.
-For example, you could deduce missing test values for patients who have passed away.
+For example, deductive imputation can be employed to infer missing test values for patients associated with a center unit that lacks a test procedure or machine, utilizing accessible metadata or protocol details.
 <!-- I don't think this is the best example.
@@
 The test score before death may be missing for a reason, but this should be derived based on clinical knowledge.
-The example is too vague (and I can't readily improve it.) -->
+The example is too vague (and I can't readily improve it.)?
+is the new example better for you?
+-->
 
@@
 
-<!-- What is the FAV variable? -->
-
-It suggests using '2l.2stage.bin' for the FAV variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
+It suggests using '2l.2stage.bin' for the FamOb variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
 '2l.2stage.pmm'. For the BMI variable, we employ deterministic imputation.
@@
 meth_mar["Age"] <- "2l.2stage.pmm"
-```
-
-<!-- What is suprime (below)? Do you mean suppress? -->
+meth_mar["Weight"] <- "2l.2stage.pmm"
+
+```
 
@@
 variable labelled as -2 and the predictor variable measured within clusters labelled as 2, encompassing all variables.
-We need to suprime the variable Time as this variable is not specified in the main model.
+We need to supress the variable Time as this variable is not specified in the main model.
 We also  modify the relationship between BMI, weight and height in the prediction matrix to avoid circular predictions.
 Then we proceed to run the imputation model.
+
 ```{r obesity-predmar,out.width="70%", eval=TRUE}
 pred_mar <- quickpred(Obesity)
+pred_mar["Weight",] <- 2
+pred_mar["Weight","Weight"] <- 0
 pred_mar[, "Cluster"] <- -2 
@@
 ```
-
-```{r obesity-predmar1,eval=TRUE}
-summary(complete(imp_mar,"long")$Weight)
-```
-We also contemplate using the predictive mean matching (pmm) option, as the values imputed using a fully parametric method may be implausibly low for some patients.
-
-```{r obesity-predmar_pmm, eval=TRUE}
-meth_mar["Weight"] <- "2l.2stage.pmm"
-imp_mar_pmm <- mice(
-  data = Obesity,
-  method = meth_mar,
-  predictorMatrix = pred_mar,
-  printFlag = FALSE
-)
-```
+After confirming convergence, we proceed to save the results for future use.
 
 ```{r obesity-predmar_pmm1,eval=TRUE}
-summary(complete(imp_mar_pmm, "long")$Weight)
-plot_trace(imp_mar_pmm, "Weight")
-ggmice(imp_mar_pmm, aes(.imp, Weight)) + 
+summary(complete(imp_mar, "long")$Weight)
+plot_trace(imp_mar, "Weight")
+ggmice(imp_mar, aes(.imp, Weight)) + 
   geom_jitter()
@@
 
-Analyze the imputed data.
-```{R}
-#| prompt: true
-#| warning: false
-# fit <- with(
-#     imp_mar_pmm,
-#     lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-#     )
-fit <- with(
-    imp_mar_pmm,
-    lmer(BMI ~ Age + FamOb + Gender + (1 + Age | Cluster))
-    )
-est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
-```
-
-Display results in table.
-```{r}
-#| prompt: true
-est$estimates |> 
-  round(3) |>
-  kable()
-```
-
-```{r}
-#| prompt: true
-est$extra.pars |> 
-  round(3) |>
-  kable()
-```
-
-After confirming convergence, we proceed to save the results for future use.
 We consider the possibility that patients may not have been selected randomly, which would then have led to a distribution for weight that does not reflect the weight in the population.
@@
 ```
-Then we proceed to run the imputation model as before, after executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
-```{r obesity-predmnar2,eval=TRUE}
-imp_mnar <- mice(
-  data = Obesity,
-  method = meth_mnar,
-  predictorMatrix = pred_mnar,
-  printFlag = FALSE
-)
-summary(complete(imp_mnar, "long")$Weight)
-```
-
-Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values, as weight should be positive.
-
-```{r obesity-predmnar3,eval=TRUE}
-summary(complete(imp_mnar, "long")$Weight)
-```
-Consequently, as before, we  use the 'pmm' option, but this time for the Heckman imputation.
-The pmm approach ensures that the imputed values remain within the range of observable values.
-This can be implemented by setting the pmm parameter to true.
+
+As before we  use the 'pmm', option but this time for the Heckman imputation, this approach ensures that the imputed values remain within the range of observable values.
+We then run the MNAR imputation model by setting the pmm parameter to true.
 
 ```{r obesity-predmnarp, eval=TRUE}
-imp_mnar_pmm <-
+imp_mnar <-
   mice(
@@
 ```{r obesity-predmnarp1, eval=TRUE}
-summary(complete(imp_mnar_pmm, "long")$Weight)
-plot_trace(imp_mar_pmm, "Weight")
-```
-
-After this modification we proceed to compare the effects on the model. We
+summary(complete(imp_mnar, "long")$Weight)
+plot_trace(imp_mar, "Weight")
+```
+
+We proceed to compare the effects on the model. We
 run the analysis model on each of the completed datasets as well as the dataset where the
@@
 
-```{r models, eval=TRUE}
-# cc_rs <- with(setDT(Obesity)[complete.cases(Obesity), ],
-#               lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age |
-#                     Cluster))
-mar_rs <- with(
-  imp_mar, 
-  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-  )
-mar_pmm_rs <- with(
-    imp_mar_pmm,
-    lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-    )
-mnar_rs <- with(
-  imp_mnar, 
-  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-  )
-mnar_pmm_rs <- with(
-  imp_mnar_pmm,
-  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
-  )
-# list_models <- list(cc_rs, mar_rs, mar_pmm_rs, mnar_rs, mnar_pmm_rs)
-# plot_models(list_models,
-#             mod_name = c("Complete case", "MAR", "MAR_pmm", "MNAR", "MNAR_pmm"))
-```
-
-```{r}
-est <- testEstimates(as.mitml.result(mnar_pmm_rs), extra.pars = TRUE)
-```
-
-Display results in table.
-```{r}
-#| prompt: true
-est$estimates |> 
-  round(3) |>
-  kable()
-```
-
-```{r}
-#| prompt: true
-est$extra.pars |> 
-  round(3) |>
-  kable()
-```
+```{r plotfunction, eval=TRUE, echo=FALSE}
+library(broom.mixed)
+plot_models <- function(list_models, mod_name){
+  mod_table<-NULL
+  for(i in 1:length(list_models)){
+    mod <- list_models[[i]]  
+    if(class(mod)[[1]]=="mira"){
+      mod_tab <-summary(pool(mod), conf.int = TRUE)[,c("term","estimate","2.5 %","97.5 %","p.value")]
+    }else{
+      sum_cc <- (summary(mod,conf.int = TRUE))
+      mod_tab<-data.frame(nlme::intervals(mod, which = "fixed")$fixed)
+      mod_tab$term<-rownames(mod_tab)
+      mod_tab$pvalue<-sum_cc$tTable[,5]
+      mod_tab <-mod_tab[,c("term","est.","lower","upper","pvalue")]
+      rownames(mod_tab)<-NULL
+
+    }
+    colnames(mod_tab)<-c("Coefficient","Estimate","lower","upper","p.value")
+    mod_tab$model<-mod_name[i]
+    mod_table<-rbind(mod_table,mod_tab)
+  }
+
+  plot <-ggplot(mod_table, aes(x = Coefficient, y = Estimate, ymin = lower, ymax = upper, group=model, color=model)) + 
+    geom_hline(yintercept=0,linetype="dashed",color="black",size=0.1)+
+    geom_linerange( position = position_dodge(width = 0.8)) + 
+    geom_pointrange(size=0.1, position = position_dodge(width = 0.8))+coord_flip()+
+    theme_classic()+theme(legend.position = "bottom")+scale_color_viridis_d()
+  return(plot)
+}
+```
+We proceed to fit the mixed model by using the complete case database and the imputed datasets (MAR,MNAR), and we compare the estimates of each dataset by using a plot (the plot function can be found in the appendix).
+```{r models, eval=TRUE,warning=FALSE,message=FALSE}
+# Save all datasets in a list
+list_data <- list(Obesity |> filter(complete.cases(Obesity)), # complete case
+                  imp_mar, # imputed under mar
+                  imp_mnar) # imputed under mnar
+
+# Run mixed models on each dataset
+list_models <- lapply(list_data, 
+                   FUN = function(x) 
+                     with(x,nlme::lme(BMI ~ Age + FamOb + Gender, 
+                                      random =  ~ 1 + Age | Cluster)))
+# Plot coefficients
+plot_models(list_models,
+             mod_name = c("Complete case", "MAR", "MNAR"))
+```
+
 We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration.
@@
 This evaluation can reveal issues such as circular problems, the need for additional iterations, or challenges associated with the chosen imputation method.
+
+<!-- What is the difference between Mice and mice? is there a Mice package? -->
 

### 2024-01-23 renv/library/R-4.3/aarch64-apple-darwin20/semEff/doc/semEff.Rmd

---
title: "semEff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{semEff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`semEff` automates calculation of effects for structural equation models (SEM) which use local estimation ('piecewise'; see @lefcheck2016). Here's a short example using an SEM testing fire impacts on plant species richness (@grace2006; data provided with the `piecewiseSEM` package).

```{r message = FALSE}
# install.packages(c("semEff", "piecewiseSEM"))
library(semEff)
data(keeley, package = "piecewiseSEM")
```

Let's have a look at the data. Observations represent a range of [vegetation parameters](https://rdrr.io/cran/piecewiseSEM/man/keeley.html) measured in 90 post-fire study plots after wildfires in southern California in 1993:

```{r}
knitr::kable(head(keeley))
```

An SEM fit to these data in @grace2006 tests the influence of landscape vs. local factors in the response of plants to fire, or more specifically, the direct vs. indirect effects of landscape location (distance from coast) on plant species richness via other biotic and abiotic mediators (see the paper for further details). We can reproduce this SEM (roughly) in a piecewise fashion by fitting a separate linear model for each observed response variable:

```{r}
keeley.sem <- list(
  lm(age ~ distance, data = keeley),
  lm(hetero ~ distance, data = keeley),
  lm(abiotic ~ distance, data = keeley),
  lm(firesev ~ age, data = keeley),
  lm(cover ~ firesev, data = keeley),
  lm(rich ~ distance + hetero + abiotic + cover, data = keeley)
)
```

And visualise it using `piecewiseSEM:::plot.psem()` (wrapper for `DiagrammeR::render_graph()`), which helps to clarify the directions of causal pathways in the system:

```{r message = FALSE, warning = FALSE}
piecewiseSEM:::plot.psem(
  piecewiseSEM::as.psem(keeley.sem),
  node_attrs = data.frame(shape = "rectangle", color = "black", fillcolor = "grey"),
  layout = "tree"
)
```

Now, to calculate effects, we first bootstrap and save the standardised direct effects:

```{r}
system.time(
  keeley.sem.boot <- bootEff(keeley.sem, R = 100, seed = 13, parallel = "no")
)
```

Note that here we use only 100 reps to save time, where typically we should use a lot more for greater accuracy of confidence intervals (e.g. 1,000--10,000+). Using the bootstrapped estimates we can now calculate direct, indirect, and total effects:

```{r}
(keeley.sem.eff <- semEff(keeley.sem.boot))
```

Let's examine effects for the final response variable, plant species richness, which has four direct and five indirect paths leading to it:

```{r}
summary(keeley.sem.eff, response = "rich")
```

We can see that the standardised total effect of distance from coast on plant species richness is moderately high at 0.453 (CIs: 0.373, 0.526), and is split almost equally between direct (0.233) and indirect (0.220) effects (involving five mediators). The mediator effects listed here are the sum of all the indirect paths from all predictors which involve each mediator, and can be useful as a rough indicator of the overall importance and net direction of effect of each. *hetero*, *abiotic*, and *cover* appear to be relatively important, with *cover* having the strongest influence (-0.155). However, this considers all predictors in the system collectively -- meaning that mediators are themselves counted as predictors. To focus only on the exogenous variable *distance*, let's recalculate:

```{r}
summary(
  semEff(keeley.sem.boot, predictor = "distance"),
  response = "rich"
)
```

Now we can see that *hetero* and *abiotic* are the important mediators for distance, while the relative influence of *cover* is much lower -- weakened due to being part of a single longer chain of effect (involving *age* and *firesev*). The total indirect effect of 0.220 can thus be broken down into three individual paths:

1.  *distance* -\> *age* -\> *firesev* -\> *cover* -\> *rich* (0.015)

2.  *distance* -\> *hetero* -\> *rich* (0.104)

3.  *distance* -\> *abiotic* -\> *rich* (0.101)

This is only one potential way of analysing paths in this SEM. Depending on the size of the system and the particular research questions, any number of predictors, mediators and/or response variables can be investigated separately or collectively for direct vs. indirect effects. If there are multiple hypotheses (competing or complementary), these can all be tested in a single SEM by comparing the relative importance of different variables and pathways of effect.

**NOTE**: All effects presented here are standardised unique effects (i.e. adjusted for multicollinearity, a.k.a. semipartial correlations), which is the only way to fully partition effects in the system. These will usually be a bit smaller than the unadjusted standardised coefficients, since most variables are correlated to some degree. If there are large differences between the two, consideration should be given to the impact (and relevance) of multicollinearity in the system. In this particular case, it's minimal, with standard errors of coefficients less than twice as large as they should be (checked using [`RVIF()`](https://murphymv.github.io/semEff/reference/RVIF.html) for the species richness model):

```{r}
RVIF(keeley.sem[[6]])
```

To use original standardised coefficients instead, specify `unique.eff = FALSE`, which can be passed to [`stdEff()`](https://murphymv.github.io/semEff/reference/stdEff.html), [`bootEff()`](https://murphymv.github.io/semEff/reference/bootEff.html), and/or [`semEff()`](https://murphymv.github.io/semEff/reference/semEff.html). These can be preferred if prediction is the primary goal rather than inference, for comparison with the unique effects or with other standardised coefficients, or for other reasons.

#### References

### 2024-01-23 Manuscript/manuscript.qmd
Update

@@
 
-Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
+Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modelled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster.
 However, this strategy is best suited for balanced data (Grund, 2017) and when the number of clusters is adequate sufficient for the random effects model (Austin,2018).
@@
 ### Assess Estimation Procedure Robustness to Missing Data
-Before diving into imputation, make sure your estimation model can handle missing data.
+Before diving into imputation, make sure your main model can handle missing data.
 Sometimes, simpler methods like complete case analysis might be suffice, especially if your missingness is low (usually <5%).
@@
 <!-- What does it mean if an estimation model is not able to handle missing data?
-there are models that implicitaly are able to manage the missingness like the FML where in the likelihood it is used the information of the completed observations and the incomplete ones.
+there are models that implicitely are able to manage the missingness like the FML where computation of the likelihood uses the information from the completed observations and also from the incomplete ones.
 What do you mean with estimation model?
-the main model you use to analyse the data-->
+the main model used to analyse the data-->
 <!-- AFAIK, when only the outcome is missing, MI and a ML based method lead to similar results.
 The text appears to suggest that in that case ML based methods are better.
-what i understood is ML are better in that case but please check it.-->
+what i understood is ML are better cause they lead to efficient results, user do not have to make a lot of choises and it provides a determinist result.
+check this: https://missingdata.org/why-multiple-imputation-is-better-than-maximum-likelihood/#:~:text=Maximum%20likelihood%20is%20faster%20and,every%20time%20you%20run%20it.
+-->
 
@@
 
-<!-- most of this subsection is also mentioned at the start of #sec-workflow -->
+<!-- most of this subsection is also mentioned at the start of #sec-workflow agree i dont know how Hanne wants to integrate this as she wrote the start section after we completed this section. -->
 
@@
 \end{align}
-
-\begin{tabular}{ll}
-     \toprule
-Type & Definition \\
-  \midrule
--2  & Cluster variable, in this case the one defined by j index  \\
- 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
- 2  & Random variable, e.g., $x_{3_{ij}}$\\
- 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
- 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
--3. & Random variable only included on selection model (Heckman model)\\
--4. & Random variable only included on main model (Heckman model)\\
-      \bottomrule
-\end{tabular}
-
-<!-- Can we reorder this table? It starts from -2, then -1 and 0 are missing, 1 to 4 and then again 2 negative numbers. What about the following?  -->
 
@@
 We utilize the obesity dataset included in the `micemd`\@ package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries.
-We simulate data for 5 clusters so that the true values are known.
 We use the following variables from the dataset:
-
-<!-- What do you mean with "We simulate data for 5 clusters"? If this dataset is included in the package why do we need to simulate the data? -->
 
@@
 Nonetheless, with respect to precision of the estimates, we notice that in general assuming MNAR leads to wider confidence intervals.
-In this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on the significance of a statistical test and therefore lead to contradictory conclusions.
-
-<!-- Contradictory? Do you mean if the results from an imputation assuming MAR and one assuming MNAR disagree? -->
+In this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on the significance of a statistical test and therefore lead to different conclusions.
+
+<!-- Contradictory? Do you mean if the results from an imputation assuming MAR and one assuming MNAR disagree? yes indeed -->
 
@@
 
-The "Mice" and "mice"-based R packages offer a range of imputation methods tailored for hierarchical data, easily adaptable to the dataset's structure.
+The "Mice" and "Mice"-based R packages offer a range of imputation methods tailored for hierarchical data, easily adaptable to the dataset's structure.
 Before proceeding with the analysis of the imputed dataset, it is essential to assess the convergence of the imputation method.

### 2024-02-05 Manuscript/manuscript.tex

@@
 Johanna Muñoz\\
-Julius Centre for Health Sciences and Primary Care\\
+Julius Center for Health Sciences and Primary Care\\
 Universiteitsweg 100\\
@@
 Valentijn M.T. de Jong\\
-Julius Centre for Health Sciences and Primary Care\\
+Julius Center for Health Sciences and Primary Care\\
 Utrecht The Netherlands\\
@@
 Gerko Vink\\
-Julius Centre for Health Sciences and Primary Care\\
+Julius Center for Health Sciences and Primary Care\\
 Universiteitsweg 100\\
@@
 Thomas P.A. Debray\\
-Julius Centre for Health Sciences and Primary Care\\
+Julius Center for Health Sciences and Primary Care\\
 Universiteitsweg 100\\
@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, boxrule=0pt, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 This tutorial is dedicated to readers who are unfamiliar with multiple
-imputation. More experienced readers can skip the introduction (case
-study 1) and directly head to practical applications of multilevel
-imputation under MAR conditions (case study 2) or under MNAR conditions
-(case study 3).
+imputation. More experienced readers can skip the background section and
+introduction (case study 1) and directly head to practical applications
+of multilevel imputation under MAR conditions (case study 2,
+Section~\ref{sec-impact}) or under MNAR conditions (case study 3).
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, leftrule=.75mm, toprule=.15mm, colframe=quarto-callout-color-frame, rightrule=.15mm, bottomrule=.15mm, colback=white, breakable, opacityback=0, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, breakable, bottomrule=.15mm, colback=white, left=2mm, colframe=quarto-callout-color-frame, rightrule=.15mm, leftrule=.75mm, arc=.35mm, toprule=.15mm]
 
@@
 
-Regarding the restrictions imposed by the data, for instance, the use of
-cluster indicator variables is restricted in datasets where there are
-not many clusters and many observations per cluster (Graham, 2009). The
-last restriction is also required when imputations are performed on each
-cluster separately. When this restriction cannot be achieved, one can
-use an imputation model that simultaneously imputes all clusters using a
-hierarchical model (Allison 2002).
+Regarding the restrictions imposed by the nature of the data, i.e., the
+sample size, the use of cluster indicator variables is restricted in
+datasets where there are not many clusters and many observations per
+cluster (Graham, 2009). The last restriction is also required when
+imputations are performed on each cluster separately. When this
+restriction cannot be achieved, one can use an imputation model that
+simultaneously imputes all clusters using a hierarchical model (Allison
+2002).
 
 Under this hierarchical imputation model, observations within clusters
-are correlated and this correlation is modeled by a random effect so the
-hierarchical model can be estimated even when there are few observations
-per cluster. However, this strategy is best suited for balanced data
-(Grund, 2017) and when random effects model is appropriated, i.e.~the
-number of clusters is adequate. (Austin,2018).
+are correlated and this correlation is modelled by a random effect so
+the hierarchical model can be estimated even when there are few
+observations per cluster. However, this strategy is best suited for
+balanced data (Grund, 2017) and when the number of clusters is adequate
+sufficient for the random effects model (Austin,2018).
 
 Here it is important to evaluate the assumptions imposed by the main
-model, for instance by using the cluster indicator strategy may lead to
-bias estimates when the model is based on a hierarchical model
-(Taaljard,2008). Even when an imputation strategy congenial with the
-main model is preferred, it is important to consider whether it is
-appropriate for the data as a less complex imputation strategies may
-also lead to unbiased estimates in certain scenarios(Bailey 2020). For
-instance, in causal effect analysis, separately imputation may lead to
-smaller bias when the size of the smaller exposure cluster is large,
-compared with an imputation model that includes exposure-confounder
-interactions. (Zhang,2023).
-
-Below we provide a imputation workflow that can be used in general to
-impute cluster data.
-
-\hypertarget{focus-on-the-main-analysis}{%
-\subsection{Focus on the Main
-Analysis}\label{focus-on-the-main-analysis}}
-
-When dealing with incomplete clustered data, start by looking at your
-research questions and planned analysis. Pretend there are no missing
-data at first. This will help you understand your research hypotheses,
-the main statistical model, and give you insights into your data's
-structure (refer to the level table), variable types (e.g., confounders,
-auxiliary variables), and other considerations such as variable
-relationships, interactions or polynomial terms.
+model, for instance the cluster indicator strategy bias the estimator
+when the model is based on a hierarchical model (Taaljard,2008).
+Although an imputation strategy congenial with the main model is
+preferred, it is important to consider whether a complex imputation
+model is appropriate for the data, as a less complex imputation
+strategies may also lead to unbiased estimates in certain scenarios
+(Bailey 2020). Below we provide a imputation workflow that can be used
+in general to impute cluster data.
 
@@
 
-Now, explore what's in your data. Use simple tools like histograms and
-QQ plots to understand variable distributions, plausible range of values
-and identify errors or outliers. Check interactions between predictors
-with scatter plots, look for collinearity with VIF or correlations
-plots, this may also help you to identify non-considered relationships
-that may affect the main model. You might also consider to test
-assumptions related to your response variable such as variance
-homogeneity, independence between observations (eg. ACF, variagrams).
-This will help you choose an imputation model that suits your data. In
-addition, the intraclass correlation coefficient (ICC) can be examined
-to assess cluster differences, aiding in the choice between the 2l and
-1l methods for imputation.
+First, explore the data, preferably using visual tools. This will help
+you choose an imputation model that suits your data. In addition, the
+intraclass correlation coefficient (ICC) can be examined to assess
+cluster differences, aiding in the choice between methods for dependent
+(such as clustering) or independent data.
 
 Next, explore the missingness. Look at the
-\textbf{proportion of missing values} in the dataset variables. This
-helps find potential predictors and cut down on unnecessary variables in
-the imputation model. Doing this can lower the risk of multicollinearity
-or computational issues, especially with certain parametric imputation
+\textbf{proportion of missing values} in the dataset variables.
+Variables for which most data is missing may contribute little in the
+analysis, and may be removed from the analysis model. Removing them also
+from the imputation model can lower the risk of multicollinearity or
+computational issues, especially with certain parametric imputation
 methods. You can also identify predictors for the imputation model by
@@
 Check the missing patterns at cluster level, this can help you to select
-the most appropiated imputation approach in terms of computational
+the most appropriate imputation approach in terms of computational
 efficiency (e.g., simpler regression imputation versus FCS in univariate
@@
 
-Before diving into imputation, make sure your estimation model can
-handle missing data. Sometimes, simpler methods like complete case
-analysis might be suffice, especially if your missingness is low
-(usually \textless5\%). There are scenarios in which specific Maximum
-Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
-methods, for instance when the response variable is the sole incomplete
-variable, mixed models demonstrate robustness to missing data under the
-Missing at Random (MAR) assumption and with a correct
-variance-covariance specification.\cite{molenberghs2007}.
+Before diving into imputation, make sure your main model can handle
+missing data. Sometimes, simpler methods like complete case analysis
+might be suffice, especially if your missingness is low (usually
+\textless5\%). There are scenarios in which specific Maximum Likelihood
+(ML) estimation methods outperform Multiple Imputation (MI) methods, for
+instance when the response variable is the sole incomplete variable,
+mixed models demonstrate robustness to missing data under the Missing at
+Random (MAR) assumption and with a correct variance-covariance
+specification.\cite{molenberghs2007}.
 
@@
 
-Clean up your dataset before the imputation process. Keep initially the
-essential variables for your main model, but include extra variables if
-needed for specific procedures during analysis (e.g.~confounders on
-balancing procedures). Think about adding other useful variables, even
-if they're not in the main model, such as instrumental or auxiliary
-variables which might boost your parameter estimates, especially if
-they're associated to the probability of missingness for some incomplete
-variables.
-
-Figure out if you can directly impute incomplete variables by just use
+Clean up your dataset before the imputation process. For imputation,
+include the variables necessary for your analysis, depending on your
+analysis this may be the exposure or intervention, outcome, covariates,
+confounders, stratification factors, and the cluster variable. Think
+about adding other useful variables, even if they're not in the main
+model, such as instrumental or auxiliary variables which might improve
+the accuracy of the imputation model, especially if they're associated
+to the probability of missingness for some incomplete variables.
+
+Figure out if you can directly impute incomplete variables by using
 deductive imputation. This involves inferring missing values based on
-logicallogical connections between variables. It's especially handy for
+logical connections between variables. It's especially useful for
 variables that depend on each other, like calculating BMI from weight
@@
 variables from level-2 ones, like in Individual Participant Data (IPD).
-In these situations, you can guess missing information from metadata or
-by using or through deduction based on time or protocols. For example,
-you could deduce missing test values for patients who have passed away.
+For example, deductive imputation can be employed to infer missing test
+values for patients associated with a center unit that lacks a test
+procedure or machine, utilizing accessible metadata or protocol details.
 
@@
 
-In a hierarchical imputation model, random effects model the
-correlations between observations within clusters, making it possible to
-estimate even with a small number of observations per cluster. There are
-various proposed multiple imputation models based on hierarchical
-models, each with its own set of assumptions \cite{audigier}.
+In a hierarchical imputation model, random intercepts and random effects
+model the correlations between observations within clusters, making it
+possible to estimate even with a small number of observations per
+cluster. There are various proposed multiple imputation models based on
+hierarchical models, each with its own set of assumptions
+\cite{audigier}.
 
@@
 your dataset. The mice package suggests methods based on variable types
-for non-clustered variables, and for the cluster ones, you can use the
+for non-clustered variables, and for the clustered ones you can use the
 micemd package's find.defaultMethod() function. This function selects
@@
 recommended to include the outcome variable in the imputation model for
-prediction variables \cite{moons2006}. In cases where the outcome is
+prediction variables \cite{moons2006a}. In cases where the outcome is
 time-to-event, the Nelson-Aalen estimate of the time to the event should
-be added as a covariate in the imputation model {[}REF{]}. Additionally,
-including auxiliary variables, even if not part of the main model, can
-be linked to the probability of missingness, improving the likelihood of
-meeting the Missing at Random (MAR) assumption and enhancing estimation
-efficiency \cite{hardt2012}.
+be added as a covariate in the imputation model, as well as the event
+indicator {[}REF: White IR, Royston P. Imputing missing covariate values
+for the Cox model. Stat Med. 2009 Jul 10;28(15):1982-98. Available from:
+doi.org/10.1002/sim.3618{]}. Additionally, including auxiliary
+variables, even if not part of the main model, can be linked to the
+probability of missingness, improving the likelihood of meeting the
+Missing at Random (MAR) assumption and enhancing estimation efficiency
+\cite{hardt2012a}.
 
@@
   \midrule
--2  & Cluster variable, in this case the one defined by j index  \\
  1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
@@
  4  & *Random variable with cluster mean $x_{5_{ij}}$\\
+  \midrule
+-2  & Cluster variable, in this case the one defined by j index  \\
 -3. & Random variable only included on selection model (Heckman model)\\
@@
 \item
-  3 and 4 type have been advised to used as the inclusion of the means
-  of the cluster is beneficial on FCS \cite{mistler2017}
+  It has been advised to use variables of type 3 and 4, as the inclusion
+  of the means of the cluster is beneficial on FCS \cite{mistler2017}.
 \end{itemize}
@@
 Recipes have been proposed for imputing incomplete level-1 and level-2
-variables for hierarchical models \cite{buur18} (\(\S\) 7.10), which can
-be convenient to follow when dealing with numerous variables
+variables for hierarchical models \cite{buuren2018a} (\(\S\) 7.10),
+which can be convenient to follow when dealing with numerous variables
 (interaction terms at different levels) and models with many random
@@
 congeniality between the imputation and the main model. However, they
-may not be applicable to all 2l imputation methods; for instance, the
+may not be applicable to all 2l imputation methods. For instance, the
 2l.2stage imputation methods of the micemd package only allow the
@@
 
-For some interaction terms, it has been suggested, for instance, for
-treatment interaction effects, to conduct separate imputation by
-treatment group \cite{zhang2023}. Additionally, it can be used
-imputation models based on random forest or deep learning that can
-handle interaction and non-linear terms without requiring the explicit
-specification of an imputation model.
+For some interaction terms, including treatment interaction effects, it
+has been suggested to conduct separate imputation by treatment group
+\cite{zhang2023}. Additionally, one might apply random forest or deep
+learning methods, as these can handle interaction and non-linear terms
+without requiring the explicit specification of an imputation model.
 
@@
 process. In hierarchical model imputations, many issues are related to
-over fitting the imputation model. To troubleshoot, it is recommended to
+overfitting the imputation model. To troubleshoot, it is recommended to
 inspect the imputation log file for variables causing problems. One
-approach to address this is to reduce the number of predictors, by using
-step by step the previous referred recepies or by using functions like
-quickpred. Another option is to consider variable transformations, such
-as scaling when model is invariant to linear transformations e.g.,
-random intercept models. Also adjusting the level of the hierarchical
-model (e.g., using a homogeneous variance imputation method or a 1l
-model) can also be beneficial.
+approach to address this is to reduce the number of predictors, by
+applying the previous referred recepies step by step or by using
+functions like quickpred. Another option is to consider variable
+transformations, such as scaling when the model is invariant to linear
+transformations e.g., random intercept models. Also, adjusting the level
+of the hierarchical model (e.g., using a homogeneous variance imputation
+method or a 1l model) can also be beneficial.
 
@@
 
-Before starting into the analysis of each imputed dataset, it is crucial
-to validate the convergence of the imputation process. This is commonly
+Before starting the analysis of each imputed dataset, it is crucial to
+validate the convergence of the imputation process. This is commonly
 accomplished through trace plots that depict the mean and variance of
@@
 
-\hypertarget{impact-data}{%
-\subsection{IMPACT data}\label{impact-data}}
+\hypertarget{sec-impact}{%
+\subsection{IMPACT data}\label{sec-impact}}
 
@@
 individuals are asked to provide information about their weight and
-consumption habits in different countries. We simulate data for 5
-clusters so that the true values are known. We use the following
+consumption habits in different countries. We use the following
 variables from the dataset:
@@
 across the clusters. However, regarding the weight variable, we notice
-that is systematically missing in cluster 3. In order to evaluate if we
-require a imputation method that accounts for clustering we assess the
-Intraclass Correlation
+that it is systematically missing in cluster 3. In order to evaluate if
+we require a imputation method that accounts for clustering we assess
+the Intraclass Correlation
 
@@
 
-Since the ICC is above 0.1 and as the main analysis will be use a mixed
-model, we decide to use two-level (2l) imputation methods. In this
+As the ICC is greater than 0.1 and as we will use a mixed model for the
+analysis, we decide to use two-level (2l) imputation methods. In this
 imputation process, we include all predictor variables from equation
@@
 
-It suggests using `2l.2stage.bin' for the FAV variable and
+It suggests using `2l.2stage.bin' for the FamOb variable and
 `2l.2stage.norm' for the age variable. However, after inspecting the age
-and weight density plot, we consider modifying its method to
-`2l.2stage.pmm'. For the BMI variable, we employ deterministic
-imputation.
+density plot, we consider modifying its method to `2l.2stage.pmm'. For
+the BMI variable, we employ deterministic imputation.
 
@@
 variable measured within clusters labelled as 2, encompassing all
-variables. We need to suprime the variable Time as this variable is not
+variables. We need to supress the variable Time as this variable is not
 specified in the main model. We also modify the relationship between
@@
 
+After confirming convergence, we proceed to save the results for future
+use.
+
 \begin{verbatim}
@@
 
-After confirming convergence, we proceed to save the results for future
-use. We consider the possibility that patients may not have been
-selected randomly, which would then have led to a distribution for
-weight that does not reflect the weight in the population. It???s likely
-that an omitted variable, like self-esteem, could influence this
-selection. For instance, individuals with lower self-esteem might have
-higher weight values, impacting their willingness to provide honest
-information due to embarrassment.
-
-To address this situation, two approaches have been proposed for dealing
-with Missing Not at Random (MNAR) data: pattern-mixed models and
-selection models. Within pattern-mixed models, methods like the delta
-method and more advanced ones like NARFS have been suggested. The
-selection model approach includes methods such as the Heckman model,
-which can be particularly useful in this case. Several methods,
-including those by \cite{galimard2018}, and the recently a Heckman
-method designed for two-level data, allow for variations in intercepts
-and exposure effects (random intercept and slope) \cite{munoz2023}.
+We consider the possibility that patients may not have been selected
+randomly, which would then have led to a distribution for weight that
+does not reflect the weight in the population. It's likely that an
+omitted variable, like self-esteem, could influence this selection. For
+instance, individuals with lower self-esteem might have higher weight
+values, impacting their willingness to provide honest information due to
+embarrassment.
+
+To address this situation where data are Missing Not At Random (MNAR),
+one approach is to apply the Heckman selection model. This method has
+recently been extended to allow for variations in intercepts and
+exposure effects (random intercept and slope)
+\cite{galimard2018, munoz2023}. This method involves specifying two
+equations: one for the outcome, describing the incomplete variable in
+terms of partially observed predictors (in this case, all variables from
+the main model), and the other for the selection model, explaining the
+probability of being observed based (R) on certain variables.
 
@@
 should be specified as `2l.2stage.heckman' found in the micemd package.
-Additionally, the prediction matrix needs modification because this
-method involves specifying two equations: one for the outcome,
-describing the incomplete variable in terms of partially observed
-predictors (in this case, all variables from the main model), and the
-other for the selection model, explaining the probability of being
-observed based (R) on certain variables. For the outcome equation we
-consider the same imputation model that we used for the MAR case (main
-model).
+Additionally, the prediction matrix needs modification, as this model
+includes one submodel for the individuals' measured outcome, and one for
+the selection of individuals. For the outcome equation we consider the
+same imputation model that we used for the MAR case (main model).
 \[Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
@@
 skip specific questions or even the entire survey. Also, we assume the
-time does not have any influence on the subject???s weight.
+time does not have any influence on the subject's weight.
 \[R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
@@
 higher average BMI is anticipated compared to the MAR assumption.
-Nonetheless, with respect to precision of estimates, we notice that in
-general MNAR imputation leads to wider confidence intervals, in this
+Nonetheless, with respect to precision of the estimates, we notice that
+in general assuming MNAR leads to wider confidence intervals. In this
 case it does not have any influence on the final result but there could
 be cases where variation in the assumed missing mechanism could lead
-also to differences on significant test and therefore lead to
-contradictory conclusions.
+also to differences on the significance of a statistical test and
+therefore lead to different conclusions.
 
@@
 
-The ``Mice'' and ``mice''-based R packages offer a range of imputation
+The ``Mice'' and ``Mice''-based R packages offer a range of imputation
 methods tailored for hierarchical data, easily adaptable to the
@@
 
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-This project has received funding from the European Union's Horizon 2020
-research and innovation programme under ReCoDID grant agreement No
-825746.
+\hypertarget{sec-summary}{%
+\section{Summary and discussion}\label{sec-summary}}
+
+What is missing from this manuscript\ldots{}
+
+\hypertarget{computational-details}{%
+\section*{Computational details}\label{computational-details}}
+\addcontentsline{toc}{section}{Computational details}
+
+The results in this paper were obtained using
+\proglang{R}\textasciitilde4.3.0. \proglang{R} itself and all packages
+used are available from the Comprehensive \proglang{R} Archive Network
+(CRAN) at {[}https://CRAN.R-project.org/{]}.
+
+\hypertarget{disclaimer}{%
+\section{Disclaimer}\label{disclaimer}}
 
@@
 
-\hypertarget{sec-summary}{%
-\section{Summary and discussion}\label{sec-summary}}
-
-What is missing from this manuscript\ldots{}
-
-\hypertarget{computational-details}{%
-\section*{Computational details}\label{computational-details}}
-\addcontentsline{toc}{section}{Computational details}
-
-The results in this paper were obtained using
-\proglang{R}\textasciitilde4.3.0. \proglang{R} itself and all packages
-used are available from the Comprehensive \proglang{R} Archive Network
-(CRAN) at {[}https://CRAN.R-project.org/{]}.
-
 \hypertarget{acknowledgments}{%
